### 2022-11-24 report/results.Rmd
First results compiled

---
title: "Modelling writing hesitations in text production as finite mixture model"
author: "Jens Roeser"
date: "Compiled `r format(Sys.Date(), '%b %d %Y')`"
output: 
  rmdformats::downcute:
    keep_md: true
    self_contained: true
    thumbnails: false # for images
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true # for cross references
bibliography: references.bib
csl: apa.csl
link-citations: yes
---

```{r include = F}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(gridExtra)
library(scales)
library(caret)
library(InformationValue)

# Calculate Bayes Factor
BF <- function(ps, prior_sd = 1){
  fit_posterior <- logspline(ps) 
  posterior <- dlogspline(0, fit_posterior) # Height of the posterior at 0 
  prior <- dnorm(0, 0, prior_sd) # Height of the prior at 0
  BF10 <- prior / posterior 
  return(BF10)
}

logit <- function(p) log(p / (1-p))
ilogit <- function(x) 1 / (1 + exp(-x))
inv_logit <- function(logit) exp(logit) / (1 + exp(logit))

se_bin <- function(x) sqrt((mean(x, na.rm = T)*(1 - mean(x, na.rm = T)))/length(x)) # se for binary data

lower <- function(x) quantile(x, prob = .025)
upper <- function(x) quantile(x, prob = .975)

# Remove leading zeros and round numbers
dezero <- function(x, dp){ # dp is decimal places
  fmt = paste0('%.',dp,'f')
  x = sprintf(fmt,x)
  x = str_replace(x, '(-|^)0\\.','\\1\\.')
  return(x)
}
dezero_plot <- function(x) dezero(x, 1)

theme_set(theme_bw(base_size = 14) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))

options(scipen = 999)

beta_label <- "Short key transition\nduration"
delta_label <- "Slowdown for hesitant key\ntransitions"
theta_label <- "Probability of hesitant\ntransitions"
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      comment = NA,
                      #                      dev = c("pdf", "png"), 
                      #                     fig.keep = "all", 
                      #                    fig.path = "figures/",
                      fig.align = 'center', 
                      #                   pdf.options(encoding = "ISOLatin9.enc"),
                      fig.width = 8, 
                      fig.height = 5,
                      width=90)
```

# SPL2

## Model comparisons

```{r}
files <- list.files("../stanout/spl2", pattern = "^modelcomp", full.names = T)
table <- purrr::map_dfr(files, read_csv) %>% 
  mutate(across(location, replace_na, "overall")) %>% 
  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
  mutate(across(where(is.numeric), round, 0),
         across(where(is.numeric), format, big.mark = ","),
         across(everything(), str_trim)) %>%
  transmute(across(location, ~str_c("Text location: ",.)),
            across(model, recode, lmm = "Unimodal log-normal",
                                  lmmgaus = "Unimodal Gaussian",
                                  mog = "Bimodal log-normal",
                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
            across(elpd_diff, str_c, " (", se_diff, ")"),
            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
  select(Model = model, 
         ends_with("overall"), 
         ends_with("sent"), 
         ends_with("beforeword"), 
         ends_with("withinword"))

```

```{r spl2, results='asis'}
names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)

kable(table, caption = "SPL2 data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
        align =c("l", rep("r", 8))) %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 11) %>%
  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
  add_header_above(c(" " = 1, 
                     "All transitions" = 2,
                     "Before sentence" = 2, 
                     "Before word" = 2,
                     "Within word" = 2)) %>%
  column_spec(1, width = "8em")

```

## Posterior parameter estimates of mixture model

```{r spl2post, fig.cap="SPL2. Posterior parameter distribution"}
file <- list.files(path = "../stanout/spl2", pattern = "mog.csv$",full.names = T)
ps <- read_csv(file) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2")) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(across(delta, ~.+beta),
         across(c(delta, beta), exp),
         across(delta, ~.-beta),
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(beta:prob) %>% 
  group_by(lang, location, name) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"))

beta <- filter(ps, name == "beta") %>% 
  mutate(name = beta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = lang,
             shape = lang)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = c(21, 23)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_continuous(labels = scales::comma) +
  labs(y = "Posterior estimate with 95% PIs",
       x = "",
       colour = "Language",
       shape = "Language")

delta <- filter(ps, name == "delta") %>% 
  mutate(name = delta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = lang,
             shape = lang)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = c(21, 23)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "",
       x = "Transition location",
       colour = "Language",
       shape = "Language") +
  theme(axis.title.y = element_blank())

theta <- filter(ps, name == "prob") %>% 
  mutate(name = theta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = lang,
             shape = lang)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = c(21, 23)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
  labs(y = "",
       x = "",
       colour = "Language",
       shape = "Language") +
  theme(axis.title.y = element_blank())

beta + delta + theta +
  plot_layout(guides = "collect")

```

# PlanTra

## Model comparisons

```{r}
files <- list.files("../stanout/plantra", pattern = "^modelcomp", full.names = T)
table <- purrr::map_dfr(files, read_csv) %>% 
  mutate(across(location, replace_na, "overall")) %>% 
  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
  mutate(across(where(is.numeric), round, 0),
         across(where(is.numeric), format, big.mark = ","),
         across(everything(), str_trim)) %>%
  transmute(across(location, ~str_c("Text location: ",.)),
            across(model, recode, lmm = "Unimodal log-normal",
                                  lmmgaus = "Unimodal Gaussian",
                                  mog = "Bimodal log-normal",
                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
            across(elpd_diff, str_c, " (", se_diff, ")"),
            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
  select(Model = model, 
         ends_with("overall"), 
         ends_with("sent"), 
         ends_with("beforeword"), 
         ends_with("withinword"))
```

```{r plantra, results='asis'}
names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)

kable(table, caption = "PlanTra data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
        align =c("l", rep("r", 8))) %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 11) %>%
  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
  add_header_above(c(" " = 1, 
                     "All transitions" = 2,
                     "Before sentence" = 2, 
                     "Before word" = 2,
                     "Within word" = 2)) %>%
  column_spec(1, width = "8em")

```

## Posterior parameter estimates of mixture model

```{r plantrapost, fig.cap="PlanTra. Posterior parameter distribution"}
file <- list.files(path = "../stanout/plantra", pattern = "mog.csv$",full.names = T)
ps <- read_csv(file) %>% 
  mutate(across(task, factor, levels = c("pre-test", "post-test"), ordered = T)) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(across(delta, ~.+beta),
         across(c(delta, beta), exp),
         across(delta, ~.-beta),
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(beta:prob) %>% 
  group_by(task, location, name) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"))

beta <- filter(ps, name == "beta") %>% 
  mutate(name = beta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = task,
             shape = task)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = c(21, 23)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "Posterior estimate with 95% PIs",
       x = "",
       colour = "Task",
       shape = "Task")

delta <- filter(ps, name == "delta") %>% 
  mutate(name = delta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = task,
             shape = task)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = c(21, 23)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "",
       x = "Transition location",
       colour = "Task",
       shape = "Task") +
  theme(axis.title.y = element_blank())

theta <- filter(ps, name == "prob") %>% 
  mutate(name = theta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = task,
             shape = task)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = c(21, 23)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
  labs(y = "",
       x = "",
       colour = "Task",
       shape = "Task") +
  theme(axis.title.y = element_blank())

beta + delta + theta +
  plot_layout(guides = "collect")

```

# LIFT

## Model comparisons

```{r}
files <- list.files("../stanout/lift", pattern = "^modelcomp", full.names = T)
table <- purrr::map_dfr(files, read_csv) %>% 
  mutate(across(location, replace_na, "overall")) %>% 
  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
  mutate(across(where(is.numeric), round, 0),
         across(where(is.numeric), format, big.mark = ","),
         across(everything(), str_trim)) %>%
  transmute(across(location, ~str_c("Text location: ",.)),
            across(model, recode, lmm = "Unimodal log-normal",
                                  lmmgaus = "Unimodal Gaussian",
                                  mog = "Bimodal log-normal",
                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
            across(elpd_diff, str_c, " (", se_diff, ")"),
            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
  select(Model = model, 
         ends_with("overall"), 
         ends_with("sent"), 
         ends_with("beforeword"), 
         ends_with("withinword"))
```

```{r lift, results='asis'}
names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)

kable(table, caption = "LIFT data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
        align =c("l", rep("r", 8))) %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 11) %>%
  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
  add_header_above(c(" " = 1, 
                     "All transitions" = 2,
                     "Before sentence" = 2, 
                     "Before word" = 2,
                     "Within word" = 2)) %>%
  column_spec(1, width = "8em")

```

## Posterior parameter estimates of mixture model

```{r}
file <- list.files(path = "../stanout/lift", pattern = "mog.csv$",full.names = T)
ps <- read_csv(file) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(across(delta, ~.+beta),
         across(c(delta, beta), exp),
         across(delta, ~.-beta),
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(beta:prob) %>% 
  group_by(genre, topic, location, name) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}")) %>% 
  ungroup() %>% 
  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre))
```

```{r liftpost, fig.cap="LIFT. Posterior parameter distribution"}
posd <- position_dodge(.65)
dotsize <- 2.5
beta <- filter(ps, name == "beta") %>% 
  mutate(name = beta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = group,
             shape = group)) +
  geom_point(size = dotsize, position = posd) +
  geom_errorbar(width = 0, position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1:8)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "Posterior estimate with 95% PIs",
       x = "",
       colour = "",
       shape = "")

delta <- filter(ps, name == "delta") %>% 
  mutate(name = delta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = group,
             shape = group)) +
  geom_point(size = dotsize, position = posd) +
  geom_errorbar(width = 0, position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1:8)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "",
       x = "Transition location",
       colour = "",
       shape = "") +
  theme(axis.title.y = element_blank())

theta <- filter(ps, name == "prob") %>% 
  mutate(name = theta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = group,
             shape = group)) +
  geom_point(size = dotsize, position = posd) +
  geom_errorbar(width = 0, position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1:8)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
  labs(y = "",
       x = "",
       colour = "",
       shape = "") +
  theme(axis.title.y = element_blank())

beta + delta + theta +
  plot_layout(guides = "collect")

```

# Cross-task / data set comparisons

```{r}
# Get the differences for tasks / across data set
# use effect size = beta / sd
# averaged across grouping variable

# Also look at by ppt estimates?

```

```{r}
files <- str_c("../stanout/", c("lift", "spl2", "plantra"), "/mog.csv")
ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
  mutate(data = .x)) %>% 
  mutate(across(data, str_remove_all, "\\.|\\/|stanout|\\mog.csv"),
         across(data, recode_factor, 
                "spl2" = "SPL2", 
                "plantra" = "PlanTra",
                "lift" = "LIFT", 
                .ordered = TRUE)) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(across(delta, ~.+beta),
         across(c(delta, beta), exp),
         across(delta, ~.-beta),
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(beta:prob) %>% 
  group_by(data, location, name) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper),
                   .names = "{.fn}")) %>% 
  ungroup()
```

```{r crossstudypost, fig.cap="Across studies. Posterior parameter distribution"}
posd <- position_dodge(.65)
dotsize <- 2.5
grouplabel <- "Data set"
beta <- filter(ps, name == "beta") %>% 
  mutate(name = beta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = data,
             shape = data)) +
  geom_point(size = dotsize, position = posd) +
  geom_errorbar(width = 0, position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1, 4, 7)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "Posterior estimate with 95% PIs",
       x = "",
       colour = grouplabel,
       shape = grouplabel)

delta <- filter(ps, name == "delta") %>% 
  mutate(name = delta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = data,
             shape = data)) +
  geom_point(size = dotsize, position = posd) +
  geom_errorbar(width = 0, position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1, 4, 7)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "",
       x = "Transition location",
       colour = grouplabel,
       shape = grouplabel) +
  theme(axis.title.y = element_blank())

theta <- filter(ps, name == "prob") %>% 
  mutate(name = theta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             colour = data,
             shape = data)) +
  geom_point(size = dotsize, position = posd) +
  geom_errorbar(width = 0, position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1, 4, 7)) +
  scale_colour_viridis_d(end = .7) +
  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
  labs(y = "",
       x = "",
       colour = grouplabel,
       shape = grouplabel) +
  theme(axis.title.y = element_blank())

beta + delta + theta +
  plot_layout(guides = "collect")

```

```{r eval = F}
# use posterior to calculate interactions, cell means and dataset differences
files <- str_c("../stanout/", c("lift", "spl2", "plantra"), "/mog.csv")

plantra <- read_csv(files[1], id = "data") %>% 
  pivot_wider(names_from = c(genre, topic), values_from = value) %>% 
  unnest(cols = c(ARG_A, ARG_B, ARG_C, ARG_D, INF_A, INF_B, INF_C, INF_D)) %>% 
  rowwise() %>% 
  mutate(value = mean(c_across(contains("_")))) %>% 
  ungroup() %>% 
  select(param, location, value, data)

spl2 <- read_csv(files[2], id = "data") %>% 
  pivot_wider(names_from = c(lang), values_from = value) %>% 
  unnest(cols = c(EN, ES)) %>% 
  rowwise() %>% 
  mutate(value = mean(c_across(c(EN, ES)))) %>% 
  ungroup() %>% 
  select(param, location, value, data)

lift <- read_csv(files[3], id = "data") %>% 
  pivot_wider(names_from = c(task), values_from = value) %>% 
  unnest(cols = c(`pre-test`, `post-test`)) %>% 
  rowwise() %>% 
  mutate(value = mean(c_across(contains("-test")))) %>% 
  ungroup() %>% 
  select(param, location, value, data)

ps <- bind_rows(plantra, spl2, lift) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(across(data, 
                str_remove_all, 
                "\\.|\\/|stanout|\\mog.csv"),
         theta = logit_scaled(prob)) %>% 
  select(-prob) %>%
  pivot_longer(beta:theta) %>%
  pivot_wider(names_from = c(data, location), 
              values_from = value, 
              names_sep = "_") %>% 
  unnest(contains("_"))

# Get main effects, interactions, nested effects
ps_effects <- ps %>%
  rowwise() %>%
  mutate(# Main effects
         DS_1 = mean(c_across(starts_with("lift_"))) - 
                mean(c_across(starts_with("spl2_"))),
         DS_2 = mean(c_across(starts_with("lift_"))) - 
                mean(c_across(starts_with("plantra_"))),
         DS_3 = mean(c_across(starts_with("spl2_"))) - 
                mean(c_across(starts_with("plantra_"))),
         TL_1 =  mean(c_across(contains(" sentence"))) - 
                mean(c_across(contains("_before word"))),
         TL_2 =  mean(c_across(contains("_before"))) - 
                mean(c_across(contains("_within word"))),
         # Two-way interactions
         `DS_1:TL_1` = (`lift_before sentence` - `lift_before word`) -
                       (`spl2_before sentence` - `spl2_before word`),

         `DS_2:TL_1` = (`lift_before sentence` - `lift_before word`) -
                       (`plantra_before sentence` - `plantra_before word`),

         `DS_3:TL_1` = (`spl2_before sentence` - `spl2_before word`) -
                       (`plantra_before sentence` - `plantra_before word`),

         `DS_1:TL_2` = (mean(c_across(starts_with("lift_") & contains("_before"))) - 
                        mean(c_across(starts_with("spl2_") & contains("_before")))) -
                       (`lift_within word` - `spl2_within word`),

         `DS_2:TL_2` = (mean(c_across(starts_with("lift_") & contains("_before"))) - 
                        mean(c_across(starts_with("plantra_") & contains("_before")))) -
                       (`lift_within word` - `plantra_within word`),

         `DS_3:TL_2` = (mean(c_across(starts_with("spl2_") & contains("_before"))) - 
                        mean(c_across(starts_with("plantra_") & contains("_before")))) -
                       (`spl2_within word` - `plantra_within word`)) %>%
  ungroup() %>%
  mutate(# data set effects
         before_word_lift_spl2 = `lift_before word` - `spl2_before word`,
         before_word_lift_plantra = `lift_before word` - `plantra_before word`,
         before_word_spl2_plantra = `spl2_before word` - `plantra_before word`,

         before_sentence_lift_spl2 = `lift_before sentence` - `spl2_before sentence`,
         before_sentence_lift_plantra = `lift_before sentence` - `plantra_before sentence`,
         before_sentence_spl2_plantra = `spl2_before sentence` - `plantra_before sentence`,

         within_word_lift_spl2 = `lift_within word` - `spl2_within word`,
         within_word_lift_plantra = `lift_within word` - `plantra_within word`,
         within_word_spl2_plantra = `spl2_within word` - `plantra_within word`) %>%
  rename(parameter = name)

write_csv(ps_effects, "tables/ps_effects_acrossdata.csv")

#conda <- c(2, 3, 4, 1)
#condb <- c(4, 5, 6, 1)
#condc <- c(1, 2, 3, 4)
#condd <- c(4, 3, 2, 1)

#me1 <- (conda + condb)/2 - (condc + condd)/2
#me2 <- (conda + condc)/2 - (condb + condd)/2
#int <- (conda - condc) - (condb - condd)

```

```{r}
ps_effects <- read_csv("tables/ps_effects_acrossdata.csv")

# Main effects and interaction summary
ps_effects_summary <- ps_effects %>% 
  select(parameter,
         starts_with("DS"), 
         starts_with("TL")) %>%
  pivot_longer(-parameter) %>%
  group_by(parameter, name) %>%
  summarise(
    across(value, list(est = mean,
                       lower = lower,
                       upper = upper,
                       BF = BF), .names = "{.fn}")) %>%
  ungroup() %>%
  mutate(across(where(is.numeric), round, 2),
         across(est, str_c, " [", lower, " -- ", upper, "]"),
         across(BF, ~ifelse( . > 100, "> 100", as.character(.))),
         across(name, recode_factor, 
                DS_1 = "Dataset 1 (LIFT, SPL2)",
                DS_2 = "Dataset 2 (LIFT, PlanTra)",
                DS_3 = "Dataset 1 (SPL2, PlanTra)",
                TL_1 = "Location 1 (before sentence, before word)",
                TL_2 = "Location 2 (before word / sentence, within word)",
               `DS_1:TL_1` = "Dataset 1 : Location 1",
               `DS_1:TL_2` = "Dataset 1 : Location 2",
               `DS_2:TL_1` = "Dataset 2 : Location 1",
               `DS_2:TL_2` = "Dataset 2 : Location 2",
               `DS_3:TL_1` = "Dataset 3 : Location 1",
               `DS_3:TL_2` = "Dataset 3 : Location 2", .ordered = T)) %>%
  select(-lower,-upper) %>%
  pivot_wider(names_from = parameter, values_from = c(est, BF)) %>%
  arrange(name) %>% 
  select(Predictor = name, ends_with("beta"), ends_with("delta"), ends_with("theta"))

```

```{r effects, results='asis'}
ps_effects_summary %>%
  knitr::kable(align =c("l", "r", "r", "r", "r", "r", "r"), 
               col.names = c("Predictor", rep(c("Estimate", "$BF_{10}$"), 3)),
               caption = "Mixture model results of transition duration with predictor estimates for the distribution of short and hesitant transition durations (on log scale) and the probability of hesitant transitions (on logit scale). Estimates are shown with 95% PI.") %>%
  kable_styling(position = "center", full_width = T, font_size = 10) %>%
  footnote("Colon indicates interactions. PI is the probability interval. $BF_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
           threeparttable = T) %>%
  add_header_above(c("", "Short transition duration" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2))  %>%
  group_rows("Main effects", 1, 5) %>%
  group_rows("Two-way interactions", 6, 11)

```

```{r eval = F}
cellmeans <- ps_effects %>% 
  select(parameter, contains(" ")) %>%
  pivot_longer(-parameter) %>%
  pivot_wider(names_from = parameter, values_from = value) %>%
  unnest() %>%
  mutate(delta = exp(beta + delta) - exp(beta),
         across(beta, exp),
         across(theta, ilogit)) %>%
  pivot_longer(cols = beta:theta, names_to = "param") %>% 
  group_by(name, param) %>%
  summarise(across(value, 
                   list(mean = mean, 
                        lower = lower, 
                        upper = upper),
                   .names = "{.fn}")) %>%
  ungroup() %>% 
  separate(name, into = c("data", "loc"), sep = "_") %>%
  mutate(across(data, recode, "lift" = "LIFT", "spl2" = "SPL2", "plantra" = "PlanTra")) %>% 
  pivot_wider(names_from = param, values_from = c(mean, lower, upper)) %>%
  mutate(across(c(ends_with("beta"), ends_with("delta")), round, 0),
         across(c(ends_with("beta"), ends_with("delta")), scales::comma, accuracy = 1),
         across(contains("theta"), dezero, 2),
         across(everything(), as.character)) %>%
  pivot_longer(contains("_"), 
               names_to = c(".value", "parameter"), 
               names_sep ="_") %>%
  mutate(across(mean, str_c, " [", lower, " -- ", upper, "]")) %>%
  select(-lower,-upper) %>% 
  pivot_wider(names_from = data, values_from = mean)

tmp <- ps_effects %>% 
  select(parameter, starts_with("before"), starts_with("within")) %>%
  pivot_longer(-parameter) %>%
  group_by(parameter, name) %>%
  summarise(across(value, 
                   list(mean = mean, 
                        lower = lower, 
                        upper = upper,
                        BF = BF),
                   .names = "{.fn}")) %>%
  ungroup() %>%
  mutate(across(where(is.numeric), round, 2),
         across(mean, str_c, " [", lower, " -- ", upper, "]"),
         across(BF, ~ifelse(.>100, "> 100", as.character(.)))) %>%
  rename(diff = mean) %>% 
  select(-lower,-upper) %>% 
  separate(name, into = c("loc", "loc2", "comp", "comp2")) %>% 
  unite("loc", loc:loc2, sep = " ") %>% 
  unite("comp", comp:comp2, sep = " - ") %>% 
  right_join(cellmeans) %>% 
  arrange(parameter, loc, comp) %>% 
  select(parameter, `Transition location` = loc, LIFT, PlanTra, SPL2, Contrasts = comp, `Difference` = diff, `$BF_{10}$`= BF)

write_csv(tmp, "tables/cellmeans_transdur.csv")
```

```{r cellmeans, results = 'asis'}
tmp <- read_csv("tables/cellmeans_transdur.csv") %>% 
  mutate(across(Contrasts, str_replace, "lift", "LIFT"),
         across(Contrasts, str_replace, "spl2", "SPL2"),
         across(Contrasts, str_replace, "plantra", "PlanTra"))

cellmeans <- tmp %>% select(1:5) %>% 
  pivot_longer(3:5) %>% 
  unique() %>% 
  arrange(parameter, `Transition location`) %>% 
  select(-parameter, -`Transition location`)

tmp %>% select(-LIFT:-SPL2) %>% 
  arrange(parameter, `Transition location`) %>% 
  pivot_wider(names_from = parameter, values_from = c(Difference, contains("BF"))) %>% 
  select(1, Contrasts, 
         ends_with("beta"), 
         ends_with("delta"), 
         ends_with("theta")) %>% 
  knitr::kable(
  align =c("l", "l", rep("r", 6)),
  col.names = c("Transition location", "Comparisons", rep(c("Est. with 95% PI", "BF$_{10}$"),3)),
  caption = "By-transition location differences for transition duration estimates inferred from mixture model. Differences are shown on the log scale (for durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.") %>%
  collapse_rows(columns = 1, valign = "middle") %>% 
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 10) %>%
  footnote("PIs are probability intervals. $BF_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
           threeparttable = T) %>% 
  add_header_above(c(" " = 2, "Short interval durations" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2)) #%>% 
#  group_rows("Short transition duration", 1, 9) %>%
#  group_rows("Slowdown for hesitant transitions", 10, 18) %>%
#  group_rows("Probability of hesitant transitions", 19, 27) #%>%
#  column_spec(1, "2.5cm") %>% 
#  column_spec(2:4, "1.75cm") %>% 
#  column_spec(5:6, "1.75cm") %>% 
#  column_spec(7, "1cm") 
```

# References

### 2022-11-24 scripts/lift/get_mog_posterior.R

@@ -1,0 +1,35 @@
+library(tidyverse)
+library(rstan)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  mutate(condition = factor(str_c(location, genre, topic, sep = "_")),
+         cond_num = as.character(as.integer(condition))) %>% 
+  select(condition, cond_num) %>% 
+  unique()
+
+count(d, condition, cond_num)
+
+# Load model
+m <- readRDS(file = "stanout/lift/mog.rda")
+
+# Select relevant parameters
+param <- c("beta", "delta", "prob") 
+
+# Traceplots
+traceplot(m, param, inc_warmup = F)
+
+# Extract posterior
+ps <- as.matrix(m, param) %>% as_tibble()
+
+# Combine posterior with conditions
+ps %>% 
+  pivot_longer(everything()) %>% 
+  #count(name) %>% as.data.frame()
+  separate(name, into = c("param", "cond_num")) %>% 
+  left_join(d) %>% 
+  separate(condition, into = c("location", "genre", "topic"), sep = "_") %>% 
+  select(-cond_num) %>% 
+  # Save posterior
+  write_csv("stanout/lift/mog.csv")
+

### 2022-11-24 scripts/lift/lmm.R

@@ -1,0 +1,112 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_samples <- 50 # number of random data points
+n_ppts <- 100 # number of random ppts
+
+ppt_keep <- read_csv("data/lift.csv") %>% 
+  count(ppt, topic, genre) %>% 
+  group_by(ppt) %>% 
+  summarise(n = n()) %>% 
+  ungroup() %>% 
+  filter(n == 4) %>% 
+  pull(ppt)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         ppt %in% ppt_keep) %>%
+  drop_na() %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(location, genre, topic, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_samples) %>% 
+  ungroup() %>% 
+  filter(keep,
+         ppt %in% sample(unique(ppt), n_ppts)) %>% select(-keep) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))))
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Check lmer for starting values
+m <- lme4::lmer(log(iki) ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5.4
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = .8
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
[…]

### 2022-11-24 scripts/lift/lmmgaus_beforesent.R

@@ -1,0 +1,111 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+ppt_keep <- read_csv("data/lift.csv") %>% 
+  count(ppt, topic, genre) %>% 
+  group_by(ppt) %>% 
+  summarise(n = n()) %>% 
+  ungroup() %>% 
+  filter(n == 4) %>% 
+  pull(ppt)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before sentence",
+         ppt %in% ppt_keep) %>%
+  drop_na(ppt) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(genre, topic, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Check lmer for starting values
+m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 530
+          , beta_sigma = 100
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1350
+          , sigma_sigma = 10
+          , sigma_raw = rep(100, dat$K)
+          , sigma_u = 290)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model ##
[…]

### 2022-11-24 scripts/lift/lmmgaus_beforeword.R

@@ -1,0 +1,107 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+ppt_keep <- read_csv("data/lift.csv") %>% 
+  count(ppt, topic, genre) %>% 
+  group_by(ppt) %>% 
+  summarise(n = n()) %>% 
+  ungroup() %>% 
+  filter(n == 4) %>% 
+  pull(ppt)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before word",
+         ppt %in% ppt_keep) %>%
+  drop_na(ppt) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(genre, topic, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Check lmer for starting values
+m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 450
+          , beta_sigma = 100
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 950
+          , sigma_sigma = 10
+          , sigma_raw = rep(100, dat$K)
+          , sigma_u = 120)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
[…]

### 2022-11-24 scripts/lift/lmmuneqvar_beforesent.R

@@ -1,0 +1,107 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+ppt_keep <- read_csv("data/lift.csv") %>% 
+  count(ppt, topic, genre) %>% 
+  group_by(ppt) %>% 
+  summarise(n = n()) %>% 
+  ungroup() %>% 
+  filter(n == 4) %>% 
+  pull(ppt)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before sentence",
+         ppt %in% ppt_keep) %>%
+  drop_na(ppt) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(genre, topic, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+m <- lme4::lmer(log(iki) ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5.59
+          , beta_sigma = 1.5
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = .87
+          , sigma_sigma = 1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.35)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
[…]

### 2022-11-24 scripts/lift/lmmuneqvar_beforeword.R

@@ -1,0 +1,110 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+ppt_keep <- read_csv("data/lift.csv") %>% 
+  count(ppt, topic, genre) %>% 
+  group_by(ppt) %>% 
+  summarise(n = n()) %>% 
+  ungroup() %>% 
+  filter(n == 4) %>% 
+  pull(ppt)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before word",
+         ppt %in% ppt_keep) %>%
+  drop_na(ppt) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(genre, topic, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+m <- lme4::lmer(log(iki) ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5.59
+          , beta_sigma = 1.5
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = .81
+          , sigma_sigma = 1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.27)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
[…]

### 2022-11-24 scripts/lift/mog_beforesent.R

@@ -1,0 +1,115 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+ppt_keep <- read_csv("data/lift.csv") %>% 
+  count(ppt, topic, genre) %>% 
+  group_by(ppt) %>% 
+  summarise(n = n()) %>% 
+  ungroup() %>% 
+  filter(n == 4) %>% 
+  pull(ppt)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before sentence",
+         ppt %in% ppt_keep) %>%
+  drop_na(ppt) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(genre, topic, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+m <- lme4::lmer(log(iki) ~ 1 + (1|ppt), data = d);m
+start <- function(chain_id = 1){
+  list(   beta_mu = 5.5
+          , beta_sigma = 1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = 1
+          , theta_sigma = .1
+          , tau = .1
+          , sigma = .87
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.35)}
+
[…]

### 2022-11-24 scripts/lift/mog_withinword.R

@@ -1,0 +1,118 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+ppt_keep <- read_csv("data/lift.csv") %>% 
+  count(ppt, topic, genre) %>% 
+  group_by(ppt) %>% 
+  summarise(n = n()) %>% 
+  ungroup() %>% 
+  filter(n == 4) %>% 
+  pull(ppt)
+
+# Load df
+d <- read_csv("data/lift.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "within word",
+         ppt %in% ppt_keep) %>%
+  drop_na(ppt) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(genre, topic, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+m <- lme4::lmer(log(iki) ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = 0
+          , theta_sigma = .5
+          , tau = .1
+          , sigma = .5
[…]

### 2022-11-24 scripts/lift/zzz/explore.R

@@ -1,0 +1,37 @@
+# Load library
+library(tidyverse)
+theme_set(theme_bw() +
+          theme(legend.position = "top",
+                legend.justification = "right"))
+
+# Load data
+lift <- read_csv("data/lift.csv")
+
+# Check out data
+glimpse(lift)
+
+select(lift, ppt, genre, topic) %>% 
+  unique() %>% 
+  count(ppt, genre) 
+
+ggplot(lift, aes(x = iki, colour = location, fill = location)) +
+  geom_density(alpha = .25) +
+  scale_x_log10() 
+
+ggplot(lift, aes(x = iki)) +
+  geom_density(alpha = .25) +
+  scale_x_log10(limits = c(5, 150)) 
+
+count(lift, location)
+
+lift %>% group_by(ppt, location) %>% 
+  summarise(across(iki, list(iki = mean, 
+                             n = length, 
+                             pause = ~mean(. > 2000) * 100), .names = "{.fn}")) %>% 
+  ungroup() %>% 
+  mutate(across(where(is.numeric), signif)) %>% 
+  select(-ppt) %>% 
+  group_by(location) %>% 
+  summarise(across(where(is.numeric), list(mean = mean, sd = sd)))
+
+pull(lift, ppt) %>% unique() %>% length()

### 2022-11-24 scripts/lift/zzz/preprocessing.R

@@ -1,0 +1,67 @@
+# Load packages
+library(tidyverse)
+library(janitor)
+source("scripts/functions.R")
+
+# Get log files
+files <- list.files(path = "data/WritingPhases/LIFT_Data/data_out", 
+                    pattern = "logs", 
+                    full.names = TRUE)
+
+ppts <- read_csv("data/WritingPhases/LIFT_Data/data_out/participants.csv") %>% 
+  select(logguid, matches("partic")) %>% 
+  rename(ppt = ID_Participant)
+
+# Load log files
+lift <- map_dfr(files, read_csv) %>% 
+  remove_constant() %>% 
+  remove_empty() %>% 
+  indicate_dels_and_inserts() %>% 
+  clean_names()
+
+# Get only writing events
+lift_small <- lift %>% 
+  filter(type == "keyboard", !arrowkeys,
+         !output %in% c("CAPS LOCK", "BACK", "RCTRL", "DELETE", "PAUSE"), 
+         !str_detect(output, "OEM|CTRL|SHIFT|ALT")) %>%
+  select(-arrowkeys, -type, -starts_with("deleted_")) %>% 
+  rename(iki = pause_time,
+         location = pause_location_full) %>%
+  mutate(output = ifelse(output %in% c("SPACE", "TAB", "RETURN"), "_", output),
+     #    ppt = as.numeric(factor(logguid)),
+         across(location, str_to_lower),
+         across(location, ~str_sub(., 1, nchar(.)-1))) %>% 
+  filter(str_detect(location, "before|within")) %>% 
+  select(task, logguid, output, iki, location) %>% 
+  separate(task, into = c("topic", "genre"), sep = "-") %>% 
+  left_join(ppts) %>% 
+  mutate(across(ppt, ~as.numeric(factor(.)))) %>% 
+  filter(location != "before paragraph") %>% 
+  group_by(logguid, location) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(logguid) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # min 10 sentences or more
+  ungroup()
+
+glimpse(lift_small)
+
+select(lift_small, ppt, topic, genre) %>% unique() %>% 
+  count(ppt, topic, genre)
+
+write_csv(lift_small, "data/lift.csv")
+
+# lift: Nina's phd; writing from sources 13-15 year old students 
+# (arg = argumentative; inf = informative) 
+# plantra: translation and rewrite text to plain language; professional translators
[…]

### 2022-11-24 scripts/plantra/lmm.R

@@ -1,0 +1,97 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences) %>%
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(location, task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2022-11-24 scripts/plantra/lmm_beforesent.R

@@ -1,0 +1,84 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before sentence") %>%
+  mutate(condition = factor(str_c(task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) %>% 
+  drop_na() %>% 
+  mutate(across(ppt, ~as.numeric(factor(.))))
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Stan model
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 14,
+                             adapt_delta = 0.96)
+)
+
+# Save model
+saveRDS(m, 
+        file = "stanout/plantra/lmm_beforesent.rda",
+        compress = "xz")
+
+
[…]

### 2022-11-24 scripts/plantra/lmmgaus.R

@@ -1,0 +1,96 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences) %>%
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(location, task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Check lmer for starting values
+m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 615
+          , beta_sigma = 100
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1850
+          , sigma_sigma = 10
+          , sigma_raw = rep(100, dat$K)
+          , sigma_u = 165)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+lmm <- stan_model(file = "stan/lmmgaus.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2022-11-24 scripts/plantra/lmmgaus_withinword.R

@@ -1,0 +1,101 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "within word") %>%
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Check lmer for starting values
+m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 220
+          , beta_sigma = 100
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 480
+          , sigma_sigma = 100
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 37)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmgaus.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
[…]

### 2022-11-24 scripts/plantra/lmmuneqvar_beforesent.R

@@ -1,0 +1,87 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before sentence") %>%
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5.5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1.25
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.4)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmuneqvar.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 14,
+                             adapt_delta = 0.96)
+)
+
+# Save model
+saveRDS(m, 
[…]

### 2022-11-24 scripts/plantra/mog.R

@@ -1,0 +1,119 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences) %>%
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(location, task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = 0
+          , theta_sigma = .1
+          , tau = .01
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Stan model
+# Load model
+mog <- stan_model(file = "stan/mog.stan")
+
+# Parameters to omit in output
+omit <- c("theta", "prob_tilde", "mu",  "_mu", "_raw", "_sigma", "z_u", "theta_s", "L_u")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
[…]

### 2022-11-24 scripts/plantra/mog_beforesent.R

@@ -1,0 +1,95 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "before sentence") %>%
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = -1
+          , theta_sigma = .1
+          , tau = .01
+          , sigma = 1
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Stan models 
+
+# Load model
+mog <- stan_model(file = "stan/mog.stan")
+
+# Parameters to omit in output
+omit <- c("theta", "prob_tilde", "mu",  "_mu", "_raw", "_sigma", "z_u", "theta_s", "L_u")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
[…]

### 2022-11-24 scripts/plantra/mog_withinword.R

@@ -1,0 +1,102 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/plantra.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         enough_sentences,
+         location == "within word") %>%
+  mutate(across(ppt, ~as.numeric(factor(.))),
+         condition = factor(str_c(task, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+m <- lme4::lmer(log(iki) ~ 1 + (1|ppt), data = d);m
+
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = 1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = 0
+          , theta_sigma = .5
+          , tau = .1
+          , sigma = .5
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.15)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Stan models 
+
+# Load model
+mog <- stan_model(file = "stan/mog.stan")
+
+# Parameters to omit in output
+omit <- c("theta", "prob_tilde", "mu",  "_mu", "_raw", "_sigma", "z_u", "theta_s", "L_u")
+
[…]

### 2022-11-24 scripts/plantra/zzz/explore.R

@@ -1,0 +1,38 @@
+# Load library
+library(tidyverse)
+theme_set(theme_bw() +
+          theme(legend.position = "top",
+                legend.justification = "right"))
+
+# Load data
+plantra <- read_csv("data/plantra.csv")
+
+# Check out data
+glimpse(plantra)
+
+select(plantra, ppt, task) %>% 
+  unique() %>% 
+  count(ppt) 
+
+ggplot(plantra, aes(x = iki, colour = location, fill = location)) +
+  geom_density(alpha = .25) +
+  scale_x_log10() +
+  facet_wrap(~task)
+
+ggplot(plantra, aes(x = iki)) +
+  geom_density(alpha = .25) +
+  scale_x_log10(limits = c(5, 150)) 
+
+count(plantra, location)
+
+plantra %>% group_by(ppt, location) %>% 
+  summarise(across(iki, list(iki = mean, 
+                             n = length, 
+                             pause = ~mean(. > 2000) * 100), .names = "{.fn}")) %>% 
+  ungroup() %>% 
+  mutate(across(where(is.numeric), signif)) %>% 
+  select(-ppt) %>% 
+  group_by(location) %>% 
+  summarise(across(where(is.numeric), list(mean = mean, sd = sd)))
+
+pull(plantra, ppt) %>% unique() %>% length()

### 2022-11-24 scripts/plantra/zzz/preprocessing.R

@@ -1,0 +1,41 @@
+library(tidyverse)
+library(janitor)
+#source("scripts/functions.R")
+
+# Get log files
+files <- list.files(path = "data/WritingPhases/PlanTra_Data/data_out", 
+                    full.names = TRUE);files
+
+# Process data: get only typing events
+plantra <- map_dfr(files, read_csv) %>% 
+  remove_constant() %>% 
+  remove_empty() %>% 
+  clean_names() %>% 
+  mutate(arrowkeys = output %in% c("RIGHT", "LEFT", "DOWN", "UP")) %>% 
+  filter(type == "keyboard", 
+         !is.na(position),
+         !arrowkeys,
+       !output %in% c("CAPS LOCK", "BACK", "RCTRL", "DELETE", "PAUSE"), 
+       !str_detect(output, "OEM|CTRL|SHIFT|ALT")) %>%
+  select(-arrowkeys, -type, -starts_with("deleted_")) %>% 
+  rename(iki = pause_time,
+         location = pause_location_full) %>% 
+  mutate(output = ifelse(output %in% c("SPACE", "TAB", "RETURN"), "_", output),
+         ppt = as.numeric(factor(participant)),
+         across(c(location, folder), str_to_lower),
+         across(location, ~str_sub(., 1, nchar(.)-1))) %>% 
+  filter(str_detect(location, "before|within")) %>% 
+  select(ppt, task = folder, logguid, output, iki, location) %>% 
+  filter(location != "before paragraph") %>% 
+  group_by(logguid, location) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(ppt) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # min 10 sentences or more
+  ungroup()
+
+glimpse(plantra)
+select(plantra, ppt, enough_sentences) %>% unique() %>% count(enough_sentences)
+count(plantra, ppt, location_count, enough_sentences) %>% 
+  as.data.frame()
+
+write_csv(plantra, "data/plantra.csv")

### 2022-11-24 scripts/spl2/lmm_beforesent.R

@@ -1,0 +1,94 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores = 3
+n_chain = 3
+iterations = 10000
+
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit",
+         transition_type == "sentence_before") %>%
+  group_by(SubNo, Lang) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences) %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(Lang),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 7
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2022-11-24 scripts/spl2/lmm_withinword.R

@@ -1,0 +1,104 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores = 3
+n_chain = 3
+iterations = 10000
+
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit") %>%
+  group_by(SubNo, Lang, transition_type) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences,
+         transition_type == "within_word") %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(Lang),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
+
+# Sample within each category 100 random data points per loc and ppt
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= 100) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
[…]

### 2022-11-24 scripts/spl2/lmmgaus_beforesent.R

@@ -1,0 +1,93 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores = 3
+n_chain = 3
+iterations = 10000
+
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit",
+         transition_type == "sentence_before") %>%
+  group_by(SubNo, Lang) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences) %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(Lang),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 250
+          , beta_sigma = 100
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 50
+          , sigma_sigma = 10
+          , sigma_raw = rep(100, dat$K)
+          , sigma_u = 10)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmgaus.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2022-11-24 scripts/spl2/lmmgaus_beforeword.R

@@ -1,0 +1,103 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores = 3
+n_chain = 3
+iterations = 10000
+
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit") %>%
+  group_by(SubNo, Lang, transition_type) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences,
+         transition_type == "word_before") %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(Lang),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
+
+# Sample within each category 100 random data points per loc and ppt
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= 100) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 250
+          , beta_sigma = 100
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 50
+          , sigma_sigma = 10
+          , sigma_raw = rep(100, dat$K)
+          , sigma_u = 10)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmgaus.stan")
+
+# Parameters to omit in output
[…]

### 2022-11-24 scripts/spl2/mog.R

@@ -1,0 +1,128 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores = 3
+n_chain = 3
+iterations = 10000
+
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit") %>%
+  group_by(SubNo, Lang, transition_type) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences) %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(str_c(Lang, transition_type, sep =  "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
+
+# Sample within each category 100 random data points per loc and ppt
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= 100) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = -1
+          , theta_sigma = .1
+          , tau = .01
+          , sigma = 1
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Stan model
+# Load model
[…]

### 2022-11-26 report/results.Rmd
Add predicted data

@@
 ---
-title: "Modelling writing hesitations in text production as finite mixture model"
+title: "Modelling writing hesitations in text writing as finite mixture process"
 author: "Jens Roeser"
@@
 
+# Analysis
+
+We reanalysed data sets including process information from participants writing text.
+For all data sets we fit a series of four models each with random effects for participants.
+Probability functions used were normal and log-normal in line with typically treatments used in the literature, a log-normal distribution with unequal variances for model predictions, and a bimodal mixed effects model.
+Stan code for mixture models was based on @roeser2021modelling.
+Text locations (levels: before sentence, before word, within word) was included as predictor in all models.
+
+Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
+The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
+Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 10,000 iterations on 3 chains with a warm-up of 5,000 iterations and no thinning.
+Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
+
+For model comparisons we used out-of-sample predictions estimated using Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
+Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and the difference $\Delta\widehat{elpd}$ between models.
+The advantage of using leave-one-out cross-validation is that models with more parameters are penalised to prevent overfit.
+
 # SPL2
 
+Data are going to be published in @torrance.
+
+## Methods
+
+Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
+CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
+We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
+
+Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
+
+## Data processing
+
 ## Model comparisons
+
+### Fit to data
+
+```{r predictionsspl2, fig.cap="SPL2 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+files <- list.files("../stanout/spl2", pattern = "sims", full.names = T)
+sims <- read_csv(files) %>% 
+  pivot_longer(cols = c(y_obs, y_tilde))  %>% 
+  mutate(across(name, recode_factor, 
+                y_tilde = "Simulated data",
+                y_obs = "Observed data",
+                .ordered = T)) %>% 
+  filter(name ==  "Simulated data" |
+         (name == "Observed data" & sim_idx == 1))
+
+max_iki <- 2000
+
+plot_mog <- sims %>%
+  filter(model == "mog", value < max_iki) %>% 
+  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmm <- sims %>%
+  filter(model == "lmm", value < max_iki) %>% 
+  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmgaus <- sims %>%
+  filter(model == "lmmgaus", value < max_iki) %>% 
+  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmuneqvar <- sims %>%
+  filter(model == "lmmuneqvar", value < max_iki) %>% 
+  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
+  plot_layout(guides = "collect") +
+  plot_annotation(tag_levels = 'A')
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Out-of-samples cross-validation
 
@@
             across(model, recode, lmm = "Unimodal log-normal",
-                                  lmmgaus = "Unimodal Gaussian",
+                                  lmmgaus = "Unimodal normal",
                                   mog = "Bimodal log-normal",
@@
 
+## Methods
+
+## Data processing
+
 ## Model comparisons
+
+### Fit to data
+
+```{r predictionsplantra, fig.cap="PlanTra data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+files <- list.files("../stanout/plantra", pattern = "sims", full.names = T)
+sims <- read_csv(files) %>% 
+  pivot_longer(cols = c(y_obs, y_tilde))  %>% 
+  mutate(across(name, recode_factor, 
+                y_tilde = "Simulated data",
+                y_obs = "Observed data",
+                .ordered = T)) %>% 
+  filter(name ==  "Simulated data" |
+         (name == "Observed data" & sim_idx == 1))
+
+max_iki <- 2000
+
+plot_mog <- sims %>%
+  filter(model == "mog", value < max_iki) %>% 
+  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmm <- sims %>%
+  filter(model == "lmm", value < max_iki) %>% 
+  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmgaus <- sims %>%
+  filter(model == "lmmgaus", value < max_iki) %>% 
+  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmuneqvar <- sims %>%
+  filter(model == "lmmuneqvar", value < max_iki) %>% 
+  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
+  plot_layout(guides = "collect") +
+  plot_annotation(tag_levels = 'A')
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Out-of-samples cross-validation
 
@@
 
+## Methods
+
+## Data processing
+
 ## Model comparisons
+
+### Fit to data
+
+```{r predictionslift, fig.cap="LIFT data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+files <- list.files("../stanout/lift", pattern = "sims", full.names = T)
+sims <- read_csv(files) %>% 
+  pivot_longer(cols = c(y_obs, y_tilde))  %>% 
+  mutate(across(name, recode_factor, 
+                y_tilde = "Simulated data",
+                y_obs = "Observed data",
+                .ordered = T)) %>% 
+  filter(name ==  "Simulated data" |
+         (name == "Observed data" & sim_idx == 1))
+
+max_iki <- 2000
+
+plot_mog <- sims %>%
+  filter(model == "mog", value < max_iki) %>% 
+  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmm <- sims %>%
+  filter(model == "lmm", value < max_iki) %>% 
+  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmgaus <- sims %>%
+  filter(model == "lmmgaus", value < max_iki) %>% 
+  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmuneqvar <- sims %>%
+  filter(model == "lmmuneqvar", value < max_iki) %>% 
+  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
+  plot_layout(guides = "collect") +
+  plot_annotation(tag_levels = 'A')
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Out-of-samples cross-validation
 
@@
   knitr::kable(align =c("l", "r", "r", "r", "r", "r", "r"), 
-               col.names = c("Predictor", rep(c("Estimate", "$BF_{10}$"), 3)),
+               col.names = c("Predictor", rep(c("Estimate", "BF$_{10}$"), 3)),
                caption = "Mixture model results of transition duration with predictor estimates for the distribution of short and hesitant transition durations (on log scale) and the probability of hesitant transitions (on logit scale).
@@
   kable_styling(position = "center", full_width = T, font_size = 10) %>%
-  footnote("Colon indicates interactions. PI is the probability interval. $BF_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
+  footnote("Colon indicates interactions. PI is the probability interval. BF$_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
            threeparttable = T) %>%
@@
   arrange(parameter, loc, comp) %>% 
-  select(parameter, `Transition location` = loc, LIFT, PlanTra, SPL2, Contrasts = comp, `Difference` = diff, `$BF_{10}$`= BF)
+  select(parameter, `Transition location` = loc, LIFT, PlanTra, SPL2, Contrasts = comp, `Difference` = diff, `BF$_{10}$`= BF)
 
@@
                 font_size = 10) %>%
-  footnote("PIs are probability intervals. $BF_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
+  footnote("PIs are probability intervals. BF$_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
            threeparttable = T) %>% 
-  add_header_above(c(" " = 2, "Short interval durations" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2)) #%>% 
+  add_header_above(c(" " = 2, "Short interval durations" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2)) %>% 
 #  group_rows("Short transition duration", 1, 9) %>%
@@
 #  group_rows("Probability of hesitant transitions", 19, 27) #%>%
-#  column_spec(1, "2.5cm") %>% 
+  column_spec(1, "1.25cm") 
 #  column_spec(2:4, "1.75cm") %>% 

### 2023-02-07 report/results.Rmd
Upload mixture model tutorial

@@
 library(scales)
-library(caret)
-library(InformationValue)
+#library(caret)
+#library(InformationValue)
 
@@
 
-theme_set(theme_bw(base_size = 14) +
+theme_set(theme_bw(base_size = 12) +
             theme(strip.background = element_blank(),
@@
 
+# Data processing
+
+TODO: Add a description with general data filtering and a table with how many data were removed from the analysis by study.
+
+# C2L1
+
+The C2L1 data set comprises data Norwegian 6th graders--*N*=133, mean age 11 years 10 months--published in @ronneberg2022process.
+The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
+
+```{r eval = F}
+d <- read_csv("../data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000) %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num)
+
+count(d, ppt) %>% 
+  filter(n < 100)
+pull(d, ppt) %>% unique() %>% length()
+```
+
+TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
+
+## Model comparisons
+
+### Fit to data
+
+```{r predictionsc2l1, fig.cap="C2L1 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+files <- list.files("../stanout/c2l1", pattern = "sims", full.names = T)
+sims <- read_csv(files) %>% 
+  pivot_longer(cols = c(y_obs, y_tilde)) %>% 
+  mutate(across(name, recode_factor, 
+                y_tilde = "Simulated data",
+                y_obs = "Observed data",
+                .ordered = T)) %>% 
+  filter(name ==  "Simulated data" |
+         (name == "Observed data" & sim_idx == 1))
+
+max_iki <- 2000
+
+plot_mog <- sims %>%
+  filter(model == "mog", value < max_iki) %>% 
+  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmm <- sims %>%
+  filter(model == "lmm", value < max_iki) %>% 
+  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmgaus <- sims %>%
+  filter(model == "lmmgaus", value < max_iki) %>% 
+  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmuneqvar <- sims %>%
+  filter(model == "lmmuneqvar", value < max_iki) %>% 
+  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
+  plot_layout(guides = "collect") +
+  plot_annotation(tag_levels = 'A')
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Out-of-samples cross-validation
+
+```{r}
+files <- list.files("../stanout/c2l1", pattern = "^modelcomp", full.names = T)
+table <- purrr::map_dfr(files, read_csv) %>% 
+  mutate(across(location, replace_na, "overall")) %>% 
+  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
+  mutate(across(where(is.numeric), round, 0),
+         across(where(is.numeric), format, big.mark = ","),
+         across(everything(), str_trim)) %>%
+  transmute(across(location, ~str_c("Text location: ",.)),
+            across(model, recode, lmm = "Unimodal log-normal",
+                                  lmmgaus = "Unimodal normal",
+                                  mog = "Bimodal log-normal",
+                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
+            across(elpd_diff, str_c, " (", se_diff, ")"),
+            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
+  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
+  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
+  select(Model = model, 
+         ends_with("overall"), 
+         ends_with("sent"), 
+         ends_with("beforeword"), 
+         ends_with("withinword"))
+
+```
+
+```{r c2l1, results='asis'}
+names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)
+
+kable(table, caption = "C2L1 data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+        align =c("l", rep("r", 8))) %>%
+  kable_styling(position = "center", 
+                full_width = T, 
+                font_size = 11) %>%
+  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
+  add_header_above(c(" " = 1, 
+                     "All transitions" = 2,
+                     "Before sentence" = 2, 
+                     "Before word" = 2,
+                     "Within word" = 2)) %>%
+  column_spec(1, width = "8em")
+
+```
+
+## Posterior parameter estimates of mixture model
+
+```{r c2l1post, fig.cap="C2l1. Posterior parameter distribution"}
+file <- list.files(path = "../stanout/c2l1", pattern = "mog.csv$",full.names = T)
+ps <- read_csv(file) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, delta, prob)) %>% 
+  mutate(across(delta, ~.+beta),
+         across(c(delta, beta), exp),
+         across(delta, ~.-beta),
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(beta:prob) %>% 
+  group_by(location, name) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"))
+
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  geom_errorbar(width = 0, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_viridis_d(end = .7) +
+  scale_y_continuous(labels = scales::comma) +
+  labs(y = "Posterior estimate with 95% PIs",
+       x = "",
+       colour = "",
+       shape = "") +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  geom_errorbar(width = 0, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_viridis_d(end = .7) +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  geom_errorbar(width = 0, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_viridis_d(end = .7) +
+  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+beta + delta + theta +
+  plot_layout(guides = "collect")
+
+```
+
+# CATO
+
+Data are published in @torrance2016adolescent. Norwegian upper
+secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
+
+## Model comparisons
+
+### Fit to data
+
+```{r predictionscato, fig.cap="CATO data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+files <- list.files("../stanout/cato", pattern = "sims", full.names = T)
+sims <- read_csv(files) %>% 
+  pivot_longer(cols = c(y_obs, y_tilde)) %>% 
+  mutate(across(name, recode_factor, 
+                y_tilde = "Simulated data",
+                y_obs = "Observed data",
+                .ordered = T)) %>% 
+  filter(name ==  "Simulated data" |
+         (name == "Observed data" & sim_idx == 1))
+
+max_iki <- 2000
+
+plot_mog <- sims %>%
+  filter(model == "mog", value < max_iki) %>% 
+  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmm <- sims %>%
+  filter(model == "lmm", value < max_iki) %>% 
+  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmgaus <- sims %>%
+  filter(model == "lmmgaus", value < max_iki) %>% 
+  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plot_lmmuneqvar <- sims %>%
+  filter(model == "lmmuneqvar", value < max_iki) %>% 
+  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
+  ggplot(aes(x = value, 
+             group = interaction(sim_idx,name), 
+             colour = name)) +
+  geom_density() +
+  facet_wrap(~model) +
+  scale_colour_manual(values = c("firebrick3", "black")) +
+  scale_x_continuous(labels = scales::comma) +
+  labs(colour = "") +
+  theme(axis.title = element_blank())
+
+plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
+  plot_layout(guides = "collect") +
+  plot_annotation(tag_levels = 'A')
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Out-of-samples cross-validation
+
+```{r}
+files <- list.files("../stanout/cato", pattern = "^modelcomp", full.names = T)
+table <- purrr::map_dfr(files, read_csv) %>% 
+  mutate(across(location, replace_na, "overall")) %>% 
+  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
+  mutate(across(where(is.numeric), round, 0),
+         across(where(is.numeric), format, big.mark = ","),
+         across(everything(), str_trim)) %>%
+  transmute(across(location, ~str_c("Text location: ",.)),
+            across(model, recode, lmm = "Unimodal log-normal",
+                                  lmmgaus = "Unimodal normal",
+                                  mog = "Bimodal log-normal",
+                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
+            across(elpd_diff, str_c, " (", se_diff, ")"),
+            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
+  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
+  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
+  select(Model = model, 
+         ends_with("overall"), 
+         ends_with("sent"), 
+         ends_with("beforeword"), 
+         ends_with("withinword"))
+
+```
+
+```{r cato, results='asis'}
+names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)
+
+kable(table, caption = "CATO data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+        align =c("l", rep("r", 8))) %>%
+  kable_styling(position = "center", 
+                full_width = T, 
+                font_size = 11) %>%
+  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
+  add_header_above(c(" " = 1, 
+                     "All transitions" = 2,
+                     "Before sentence" = 2, 
+                     "Before word" = 2,
+                     "Within word" = 2)) %>%
+  column_spec(1, width = "8em")
+
+```
+
+## Posterior parameter estimates of mixture model
+
+```{r catopost, fig.cap="CATO. Posterior parameter distribution"}
+file <- list.files(path = "../stanout/cato", pattern = "mog.csv$",full.names = T)
+ps <- read_csv(file) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, delta, prob)) %>% 
+  mutate(across(delta, ~.+beta),
+         across(c(delta, beta), exp),
+         across(delta, ~.-beta),
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(beta:prob) %>% 
+  group_by(task, group, location, name) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}")) %>% 
+  mutate(task = str_c("Task: ", task, "; Group: ", group))
+
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             colour = task,
+             shape = task)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  geom_errorbar(width = 0, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_viridis_d(end = .7) +
+  scale_y_continuous(labels = scales::comma) +
+  labs(y = "Posterior estimate with 95% PIs",
+       x = "",
+       colour = "",
+       shape = "") +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             colour = task,
+             shape = task)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  geom_errorbar(width = 0, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_viridis_d(end = .7) +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             colour = task,
+             shape = task)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  geom_errorbar(width = 0, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_viridis_d(end = .7) +
+  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+beta + delta + theta +
+  plot_layout(guides = "collect")
+
+```
+
 # SPL2
@@
 Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-## Data processing
 
@@
 
-# PlanTra
+# PLanTra
 
@@
 
-## Data processing
+The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
+In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
+Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
+During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
+making it easier to read for a lay audience--by applying what they had learned from their respective training.
+The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
+It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
 
@@
 
-```{r predictionsplantra, fig.cap="PlanTra data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+```{r predictionsplantra, fig.cap="PLanTra data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
 
@@
 
-kable(table, caption = "PlanTra data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+kable(table, caption = "PLanTra data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
         align =c("l", rep("r", 8))) %>%
@@
 
-```{r plantrapost, fig.cap="PlanTra. Posterior parameter distribution"}
+```{r plantrapost, fig.cap="PLanTra. Posterior parameter distribution"}
 file <- list.files(path = "../stanout/plantra", pattern = "mog.csv$",full.names = T)
@@
        colour = "Task",
-       shape = "Task")
+       shape = "Task") +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
 
@@
        shape = "Task") +
-  theme(axis.title.y = element_blank())
+  theme(axis.title.y = element_blank()) +
+    guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
 
@@
 
-## Data processing
+LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
 
@@
 ```{r}
-files <- str_c("../stanout/", c("lift", "spl2", "plantra"), "/mog.csv")
+files <- str_c("../stanout/", 
+               c("lift", "spl2", "plantra", "cato", "c2l1"), 
+               "/mog.csv")
 ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
   mutate(data = .x)) %>% 
-  mutate(across(data, str_remove_all, "\\.|\\/|stanout|\\mog.csv"),
+  mutate(across(data, str_remove_all, 
+                "\\.|\\/|stanout|\\mog.csv"),
          across(data, recode_factor, 
-                "spl2" = "SPL2", 
-                "plantra" = "PlanTra",
+                "cato" = "CATO (non-dyslexic\nunmasked)",
+                "spl2" = "SPL2 (L1)", 
+                "plantra" = "PLanTra",
                 "lift" = "LIFT",
+                "c2l1" = "C2L1",
                 .ordered = TRUE)) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
+  filter(!(str_detect(data, "SPL2") & lang == "ES"),
+         !(str_detect(data, "CATO") & group == "dyslexic"),
+         !(str_detect(data, "CATO") & task == "masked")) %>% 
+#  filter(str_detect(data, "CATO"))
+  pivot_wider(names_from = param, 
+              values_from = value) %>% 
   unnest(cols = c(beta, delta, prob)) %>% 
@@
   group_by(data, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
+  summarise(across(value, 
+                   list(mean = mean, 
                         lower = lower, 
@@
              colour = data,
-             shape = data)) +
+             shape = data,
+             group = interaction(data))) +
+  geom_line(size = .5, alpha = .35, position = posd) +
   geom_point(size = dotsize, position = posd) +
@@
   facet_grid(~name) +
-  scale_shape_manual(values = c(1, 4, 7)) +
+  scale_shape_manual(values = c(1, 4, 7, 8, 9)) +
   scale_colour_viridis_d(end = .7) +
@@
        colour = grouplabel,
-       shape = grouplabel)
+       shape = grouplabel) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
 
@@
        shape = grouplabel) +
-  theme(axis.title.y = element_blank())
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
 
@@
 # use posterior to calculate interactions, cell means and dataset differences
-files <- str_c("../stanout/", c("lift", "spl2", "plantra"), "/mog.csv")
+files <- str_c("../stanout/", c("lift", "spl2", "plantra", "cato", "c2l1"), "/mog.csv")
 
@@
 
-ps <- bind_rows(plantra, spl2, lift) %>% 
+cato <- read_csv(files[4], id = "data") %>% 
+  pivot_wider(names_from = c(task, group), values_from = value) %>% 
+  unnest() %>% 
+  rowwise() %>% 
+  mutate(value = mean(c_across(contains("_")))) %>% 
+  ungroup() %>% 
+  select(param, location, value, data)
+
+c2l1 <- read_csv(files[5], id = "data") %>% 
+  select(param, location, value, data)
+
+ps <- bind_rows(plantra, spl2, lift, cato, c2l1) %>% 
   pivot_wider(names_from = param, values_from = value) %>% 
@@
                 mean(c_across(starts_with("plantra_"))),
+         DS_4 = mean(c_across(starts_with("lift_"))) - 
+                mean(c_across(starts_with("cato_"))),
+         DS_5 = mean(c_across(starts_with("plantra_"))) - 
+                mean(c_across(starts_with("cato_"))),
+         DS_6 = mean(c_across(starts_with("spl2_"))) - 
+                mean(c_across(starts_with("cato_"))),
+         DS_7 = mean(c_across(starts_with("c2l1_"))) - 
+                mean(c_across(starts_with("plantra_"))),
+         DS_8 = mean(c_across(starts_with("c2l1_"))) - 
+                mean(c_across(starts_with("cato_"))),
+         DS_9 = mean(c_across(starts_with("c2l1_"))) - 
+                mean(c_across(starts_with("spl2_"))),
+         DS_10 = mean(c_across(starts_with("c2l1_"))) - 
+                mean(c_across(starts_with("lift_"))),
+
          TL_1 =  mean(c_across(contains(" sentence"))) - 
@@
 
+         `DS_4:TL_1` = (`lift_before sentence` - `lift_before word`) -
+                       (`cato_before sentence` - `cato_before word`),
+
+         `DS_5:TL_1` = (`plantra_before sentence` - `plantra_before word`) - 
+                       (`cato_before sentence` - `cato_before word`),
+
+         `DS_6:TL_1` = (`spl2_before sentence` - `spl2_before word`) -
+                       (`cato_before sentence` - `cato_before word`),
+
+         `DS_7:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
+                       (`plantra_before sentence` - `plantra_before word`),
+
+         `DS_8:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
+                       (`cato_before sentence` - `cato_before word`),
+
+         `DS_9:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
+                       (`spl2_before sentence` - `spl2_before word`),
+
+         `DS_10:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
+                        (`lift_before sentence` - `lift_before word`),
+
          `DS_1:TL_2` = (mean(c_across(starts_with("lift_") & contains("_before"))) - 
@@
                         mean(c_across(starts_with("plantra_") & contains("_before")))) -
-                       (`spl2_within word` - `plantra_within word`)) %>%
+                       (`spl2_within word` - `plantra_within word`),
+
+         `DS_4:TL_2` = (mean(c_across(starts_with("lift_") & contains("_before"))) - 
+                        mean(c_across(starts_with("cato_") & contains("_before")))) -
+                       (`lift_within word` - `cato_within word`),
+
+         `DS_5:TL_2` = (mean(c_across(starts_with("plantra_") & contains("_before"))) - 
+                        mean(c_across(starts_with("cato_") & contains("_before")))) -
+                       (`plantra_within word` - `cato_within word`),
+
+         `DS_6:TL_2` = (mean(c_across(starts_with("spl2_") & contains("_before"))) - 
+                        mean(c_across(starts_with("cato_") & contains("_before")))) -
+                       (`spl2_within word` - `cato_within word`),
+
+         `DS_7:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
+                        mean(c_across(starts_with("plantra_") & contains("_before")))) -
+                       (`c2l1_within word` - `plantra_within word`),
+         `DS_8:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
+                        mean(c_across(starts_with("cato_") & contains("_before")))) -
+                       (`c2l1_within word` - `cato_within word`),
+         `DS_9:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
+                        mean(c_across(starts_with("spl2_") & contains("_before")))) -
+                       (`c2l1_within word` - `spl2_within word`),
+         `DS_10:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
+                        mean(c_across(starts_with("lift_") & contains("_before")))) -
+                       (`c2l1_within word` - `lift_within word`)
+         ) %>%
   ungroup() %>%
-  mutate(# data set effects
-         before_word_lift_spl2 = `lift_before word` - `spl2_before word`,
-         before_word_lift_plantra = `lift_before word` - `plantra_before word`,
-         before_word_spl2_plantra = `spl2_before word` - `plantra_before word`,
-
-         before_sentence_lift_spl2 = `lift_before sentence` - `spl2_before sentence`,
-         before_sentence_lift_plantra = `lift_before sentence` - `plantra_before sentence`,
-         before_sentence_spl2_plantra = `spl2_before sentence` - `plantra_before sentence`,
-
-         within_word_lift_spl2 = `lift_within word` - `spl2_within word`,
-         within_word_lift_plantra = `lift_within word` - `plantra_within word`,
-         within_word_spl2_plantra = `spl2_within word` - `plantra_within word`) %>%
+  mutate(
+    # Locations effecs
+    lift_sent = `lift_before sentence` - `lift_before word`,
+    lift_word = `lift_before word` - `lift_within word`,
+    spl2_sent = `spl2_before sentence` - `spl2_before word`,
+    spl2_word = `spl2_before word` - `spl2_within word`,
+    cato_sent = `cato_before sentence` - `cato_before word`,
+    cato_word = `cato_before word` - `cato_within word`,
+    c2l1_sent = `c2l1_before sentence` - `c2l1_before word`,
+    c2l1_word = `c2l1_before word` - `c2l1_within word`,
+    plantra_sent = `plantra_before sentence` - `plantra_before word`,
+    plantra_word = `plantra_before word` - `plantra_within word`
+    # data set effects
+         # before_word_lift_spl2 = `lift_before word` - `spl2_before word`,
+         # before_word_lift_plantra = `lift_before word` - `plantra_before word`,
+         # before_word_spl2_plantra = `spl2_before word` - `plantra_before word`,
+         # before_word_cato_spl2 = `cato_before word` - `spl2_before word`,
+         # before_word_cato_plantra = `cato_before word` - `plantra_before word`,
+         # before_word_cato_lift = `cato_before word` - `lift_before word`,
+         # before_word_c2l1_plantra = `c2l1_before word` - `plantra_before word`,
+         # before_word_c2l1_cato = `c2l1_before word` - `cato_before word`,
+         # before_word_c2l1_spl2 = `c2l1_before word` - `spl2_before word`,
+         # before_word_c2l1_lift = `c2l1_before word` - `lift_before word`,
+         # 
+         # before_sentence_lift_spl2 = `lift_before sentence` - `spl2_before sentence`,
+         # before_sentence_lift_plantra = `lift_before sentence` - `plantra_before sentence`,
+         # before_sentence_spl2_plantra = `spl2_before sentence` - `plantra_before sentence`,
+         # before_sentence_cato_spl2 = `cato_before sentence` - `spl2_before sentence`,
+         # before_sentence_cato_plantra = `cato_before sentence` - `plantra_before sentence`,
+         # before_sentence_cato_lift = `cato_before sentence` - `lift_before sentence`,
+         # before_sentence_c2l1_plantra = `c2l1_before sentence` - `plantra_before sentence`,
+         # before_sentence_c2l1_cato = `c2l1_before sentence` - `cato_before sentence`,
+         # before_sentence_c2l1_spl2 = `c2l1_before sentence` - `spl2_before sentence`,
+         # before_sentence_c2l1_lift = `c2l1_before sentence` - `lift_before sentence`,
+         # 
+         # within_word_lift_spl2 = `lift_within word` - `spl2_within word`,
+         # within_word_lift_plantra = `lift_within word` - `plantra_within word`,
+         # within_word_spl2_plantra = `spl2_within word` - `plantra_within word`,
+         # within_word_cato_spl2 = `cato_within word` - `spl2_within word`,
+         # within_word_cato_plantra = `cato_within word` - `plantra_within word`,
+         # within_word_cato_lift = `cato_within word` - `lift_within word`,
+         # within_word_c2l1_plantra = `c2l1_within word` - `plantra_within word`,
+         # within_word_c2l1_cato = `c2l1_within word` - `cato_within word`,
+         # within_word_c2l1_spl2 = `c2l1_within word` - `spl2_within word`,
+         # within_word_c2l1_lift = `c2l1_within word` - `lift_within word`
+    ) %>%
   rename(parameter = name)
@@
 ps_effects_summary <- ps_effects %>% 
+#  mutate(across(-parameter, scale)) %>% 
   select(parameter,
@@
   mutate(across(where(is.numeric), round, 2),
-         across(est, str_c, " [", lower, " -- ", upper, "]"),
+#         across(est, str_c, " [", lower, " -- ", upper, "]"),
          across(BF, ~ifelse( . > 100, "> 100", as.character(.))),
@@
                 DS_1 = "Dataset 1 (LIFT, SPL2)",
-                DS_2 = "Dataset 2 (LIFT, PlanTra)",
-                DS_3 = "Dataset 1 (SPL2, PlanTra)",
+                DS_2 = "Dataset 2 (LIFT, PLanTra)",
+                DS_3 = "Dataset 3 (SPL2, PLanTra)",
+                DS_4 = "Dataset 4 (CATO, SPL2)",
+                DS_5 = "Dataset 5 (CATO, PLanTra)",
+                DS_6 = "Dataset 6 (CATO, LIFT)",
+                DS_7 = "Dataset 7 (C2L1, PLanTra)",
+                DS_8 = "Dataset 8 (C2L1, CATO)",
+                DS_9 = "Dataset 9 (C2L1, SPL2)",
+                DS_10 = "Dataset 10 (C2L1, LIFT)",
+
                 TL_1 = "Location 1 (before sentence, before word)",
@@
                `DS_3:TL_1` = "Dataset 3 : Location 1",
-               `DS_3:TL_2` = "Dataset 3 : Location 2", .ordered = T)) %>%
-  select(-lower,-upper) %>%
-  pivot_wider(names_from = parameter, values_from = c(est, BF)) %>%
+               `DS_3:TL_2` = "Dataset 3 : Location 2",
+               `DS_4:TL_1` = "Dataset 4 : Location 1",
+               `DS_4:TL_2` = "Dataset 4 : Location 2",
+               `DS_5:TL_1` = "Dataset 5 : Location 1",
+               `DS_5:TL_2` = "Dataset 5 : Location 2",
+               `DS_6:TL_1` = "Dataset 6 : Location 1",
+               `DS_6:TL_2` = "Dataset 6 : Location 2",
+               `DS_7:TL_1` = "Dataset 7 : Location 1",
+               `DS_7:TL_2` = "Dataset 7 : Location 2",
+               `DS_8:TL_1` = "Dataset 8 : Location 1",
+               `DS_8:TL_2` = "Dataset 8 : Location 2",
+               `DS_9:TL_1` = "Dataset 9 : Location 1",
+               `DS_9:TL_2` = "Dataset 9 : Location 2",
+               `DS_10:TL_1` = "Dataset 10 : Location 1",
+               `DS_10:TL_2` = "Dataset 10 : Location 2",
+               .ordered = T)) %>%
+#  select(-lower,-upper) %>%
+  pivot_wider(names_from = parameter, values_from = c(est,lower,upper,BF)) %>%
   arrange(name) %>% 
@@
 ps_effects_summary %>%
-  knitr::kable(align =c("l", "r", "r", "r", "r", "r", "r"), 
-               col.names = c("Predictor", rep(c("Estimate", "BF$_{10}$"), 3)),
-               caption = "Mixture model results of transition duration with predictor estimates for the distribution of short and hesitant transition durations (on log scale) and the probability of hesitant transitions (on logit scale).
-Estimates are shown with 95% PI.") %>%
-  kable_styling(position = "center", full_width = T, font_size = 10) %>%
-  footnote("Colon indicates interactions. PI is the probability interval. BF$_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
+  mutate(est_1 = "",
+         est_2 = "",
+         est_3 = "") %>%
+  select(Predictor, est_1, BF_beta, est_2, BF_delta, est_3, BF_theta) %>%
+  knitr::kable(align =c("l", rep("c", 6)), 
+               col.names = c("Predictor", rep(c("Est.", "BF"), 3)),
+               caption = "Mixture model results of transition duration with predictor estimates (main effects, interactions) for the distribution of short and hesitant transition durations (on log scale) and the probability of hesitant transitions (on logit scale).
+Estimates are shown with 95% PI.
+Dotted vertical line indicates 0.") %>%
+  kable_styling(position = "center", full_width = T, font_size = 12) %>%
+  column_spec(2, image = spec_pointrange(
+    x = ps_effects_summary$est_beta, 
+    xmin = ps_effects_summary$lower_beta, 
+    xmax = ps_effects_summary$upper_beta, 
+    vline = 0)
+    ) %>% 
+  column_spec(4, image = spec_pointrange(
+    x = ps_effects_summary$est_delta, 
+    xmin = ps_effects_summary$lower_delta, 
+    xmax = ps_effects_summary$upper_delta, 
+    vline = 0)
+  ) %>% 
+  column_spec(6, image = spec_pointrange(
+    x = ps_effects_summary$est_theta, 
+    xmin = ps_effects_summary$lower_theta, 
+    xmax = ps_effects_summary$upper_theta, 
+    vline = 0)
+  ) %>% 
+  footnote("Colon indicates interactions. PI is the probability interval. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
            threeparttable = T) %>%
-  add_header_above(c("", "Short transition duration" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2))  %>%
-  group_rows("Main effects", 1, 5) %>%
-  group_rows("Two-way interactions", 6, 11)
+  add_header_above(c("", 
+                     "Short transition duration" = 2, 
+                     "Slowdown for hesitant transitions" = 2, 
+                     "Probability of hesitant transitions" = 2))  %>%
+  group_rows("Main effects", 1, 12) %>%
+  group_rows("Two-way interactions", 13, 32)  %>% 
+  column_spec(1, width = "5.5cm") %>% 
+  column_spec(2:7, width = "3cm")
 
@@
 ```{r eval = F}
+ps_effects <- read_csv("tables/ps_effects_acrossdata.csv")
+
 cellmeans <- ps_effects %>% 
@@
   separate(name, into = c("data", "loc"), sep = "_") %>%
-  mutate(across(data, recode, "lift" = "LIFT", "spl2" = "SPL2", "plantra" = "PlanTra")) %>% 
+  mutate(across(data, recode, 
+                "lift" = "LIFT", 
+                "spl2" = "SPL2", 
+                "plantra" = "PLanTra",
+                "cato" = "CATO",
+                "c2l1" = "C2L1")) %>% 
   pivot_wider(names_from = param, values_from = c(mean, lower, upper)) %>%
@@
                names_sep ="_") %>%
-  mutate(across(mean, str_c, " [", lower, " -- ", upper, "]")) %>%
-  select(-lower,-upper) %>% 
-  pivot_wider(names_from = data, values_from = mean)
+#  mutate(across(mean, str_c, " [", lower, " -- ", upper, "]")) %>%
+#  select(-lower,-upper) %>% 
+  pivot_wider(names_from = loc, values_from = c(mean, lower, upper))
 
 tmp <- ps_effects %>% 
-  select(parameter, starts_with("before"), starts_with("within")) %>%
+  select(parameter, ends_with("_sent"), ends_with("_word")) %>%
   pivot_longer(-parameter) %>%
@@
   mutate(across(where(is.numeric), round, 2),
-         across(mean, str_c, " [", lower, " -- ", upper, "]"),
+#         across(mean, str_c, " [", lower, " -- ", upper, "]"),
          across(BF, ~ifelse(.>100, "> 100", as.character(.)))) %>%
   rename(diff = mean) %>% 
-  select(-lower,-upper) %>% 
-  separate(name, into = c("loc", "loc2", "comp", "comp2")) %>% 
-  unite("loc", loc:loc2, sep = " ") %>% 
-  unite("comp", comp:comp2, sep = " - ") %>% 
+#  select(-lower,-upper) %>% 
+  separate(name, into = c("data", "loc")) %>% 
+  mutate(across(data, recode, 
+                "lift" = "LIFT", 
+                "spl2" = "SPL2", 
+                "plantra" = "PLanTra",
+                "cato" = "CATO",
+                "c2l1" = "C2L1"),
+         across(loc, recode, 
+                "sent" = "before sentence - before word",
+                "word" = "before word - within word")) %>% 
   right_join(cellmeans) %>% 
-  arrange(parameter, loc, comp) %>% 
-  select(parameter, `Transition location` = loc, LIFT, PlanTra, SPL2, Contrasts = comp, `Difference` = diff, `BF$_{10}$`= BF)
+  arrange(parameter, data, loc) %>% 
+  select(parameter, data, #`Transition location` = loc, 
+         ends_with("before sentence"), 
+         ends_with("before word"), 
+         ends_with("within word"),
+         Contrasts = loc, diff, lower, upper, BF)
 
@@
 ```{r cellmeans, results = 'asis'}
-tmp <- read_csv("tables/cellmeans_transdur.csv") %>% 
-  mutate(across(Contrasts, str_replace, "lift", "LIFT"),
-         across(Contrasts, str_replace, "spl2", "SPL2"),
-         across(Contrasts, str_replace, "plantra", "PlanTra"))
-
-cellmeans <- tmp %>% select(1:5) %>% 
-  pivot_longer(3:5) %>% 
+tmp <- read_csv("tables/cellmeans_transdur.csv")
+
+cellmeans <- tmp %>% select(1:2, 
+                            starts_with("mean_"), 
+                            starts_with("lower_"),
+                            starts_with("upper_")) %>% 
+  pivot_longer(-1:-2) %>% 
   unique() %>% 
-  arrange(parameter, `Transition location`) %>% 
-  select(-parameter, -`Transition location`)
-
-tmp %>% select(-LIFT:-SPL2) %>% 
-  arrange(parameter, `Transition location`) %>% 
-  pivot_wider(names_from = parameter, values_from = c(Difference, contains("BF"))) %>% 
-  select(1, Contrasts, 
-         ends_with("beta"), 
-         ends_with("delta"), 
-         ends_with("theta")) %>% 
+  arrange(parameter, data) %>% 
+  select(-parameter, -data)
+
+table <- tmp %>% select(-contains("_")) %>% 
+  arrange(parameter, Contrasts, lower) %>% 
+  pivot_wider(names_from = parameter, 
+              values_from = c(diff, lower, upper, BF))
+
+table %>% 
+  mutate(est_1 = "",
+         est_2 = "",
+         est_3 = "") %>%  
+  select(data, #Contrasts, 
+         est_1, BF_beta,
+         est_2, BF_delta,
+         est_3, BF_theta) %>% 
   knitr::kable(
-  align =c("l", "l", rep("r", 6)),
-  col.names = c("Transition location", "Comparisons", rep(c("Est. with 95% PI", "BF$_{10}$"),3)),
-  caption = "By-transition location differences for transition duration estimates inferred from mixture model.
+  align =c("l", rep("c", 6)),
+  col.names = c("Data set", rep(c("Est.", "BF"),3)),
+  caption = "By-data set differences between transition locations for transition duration estimates inferred from mixture model.
 Differences are shown on the log scale (for durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.") %>%
-  collapse_rows(columns = 1, valign = "middle") %>% 
+95% PIs in brackets.
+Dotted line indicates 0.") %>%
+#  collapse_rows(columns = 1, valign = "middle") %>% 
   kable_styling(position = "center", 
                 full_width = T, 
-                font_size = 10) %>%
-  footnote("PIs are probability intervals. BF$_{10}$ is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
+                font_size = 9) %>%
+    column_spec(2, image = spec_pointrange(
+    x = table$diff_beta, 
+    xmin = table$lower_beta, 
+    xmax = table$upper_beta, 
+    vline = 0)
+    ) %>% 
+  column_spec(4, image = spec_pointrange(
+    x = table$diff_delta, 
+    xmin = table$lower_delta, 
+    xmax = table$upper_delta, 
+    vline = 0)
+  ) %>% 
+  column_spec(6, image = spec_pointrange(
+    x = table$diff_theta, 
+    xmin = table$lower_theta, 
+    xmax = table$upper_theta, 
+    vline = 0)
+  ) %>% 
+  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
            threeparttable = T) %>% 
-  add_header_above(c(" " = 2, "Short interval durations" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2)) %>% 
+  add_header_above(c(" " = 1, "Short interval durations" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2)) %>% 
 #  group_rows("Short transition duration", 1, 9) %>%
@@
 #  group_rows("Probability of hesitant transitions", 19, 27) #%>%
-  column_spec(1, "1.25cm") 
+#  column_spec(1, "1.75cm") %>%
+#  column_spec(2, "4cm") %>% 
+  group_rows("Contrast: before sentence -- before word", 1, 5) %>%
+  group_rows("Contrast: before word -- within word", 6, 10)
+
 #  column_spec(2:4, "1.75cm") %>% 

### 2023-02-07 tutorial/tutorial.Rmd
Upload mixture model tutorial

---
title: "Tutorial: fitting mixed mixture models on text writing data"
author: "Jens Roeser"
date: "Compiled `r Sys.Date()`"
bibliography: ref.bib
csl: apa.csl
link-citations: yes
output:
  rmdformats::readthedown:
    lightbox: true
    gallery: false
    highlight: "pygments"
    toc_depth: 3
    use_bookdown: false

---

```{r, include=FALSE}
library(rmdformats)
library(knitr)
library(tidyverse)
library(rstan)
library(brms)
```

```{r setup, include=FALSE}
options(scipen = 999) # Turn off scientific notation
opts_chunk$set(fig.width = 8, fig.height = 4.5)
opts_knit$set(width=90)
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, 
                      warning = FALSE,
                      message = FALSE)
theme_set(theme_bw() + theme(legend.justification = "top"))
```

This is a step-by-step walk-through that describes how to fit a finite mixture model of two log-Gaussian distributions using the statistical program R and the `rstan` package to interface with the probabilistic programming language Stan [@carpenter2016stan; @hoffman2014no; @rstan; @rstan2]. We will compare this mixture model to standard Gaussian and log-Gaussian models implemented in `brms` [@burkner2017brms; @R-brms_b].^[An alternative to writing models in Stan is the R package `brms` which provides a flexible framework to implement mixture models and many other types of probability models [@burkner2017brms; @R-brms_b]. In particular, `brms` has a `mixture` function to specify mixtures of various types of distributions. These could be Gaussian, skewed-Normal, shifted-Normal, ex-Gaussian etc. and combinations thereof. There is a large number of probability models for continuous data that are plausible candidates [for reaction time data see e.g. @matzke2009psychological].] For a rationale of using mixture models for writing process data see @roeser2021.

This guide shows how to fit a mixture model written in Stan to keystroke data to then calculate the differences between transition locations in the text. This tutorial is largely self-contained. To run the code below, the reader only needs to install the required packages. Data and Stan code can be loaded from the code presented below.

We require three packages: (1) `rstan` to use R to interface with Stan for fitting Bayesian models^[Instructions on how to install `rstan` can be found [here]( https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).]; (2) `tidyverse` for data processing and visualisation.^[Run `install.packages("tidyverse")` for installation.]; (3) `brms` for simple Bayesian models.^[Run `install.packages("brms")` for installation.]

```{r eval = F}
library(rstan)
library(tidyverse)
library(brms)
```

# Fitting the model

## Preparing the data

First we load a sample of keystroke data. Data are loaded from [OSF](https://osf.io/y3p4d/) using the URL as below and the `read_csv` function. For brevity we use a random subset. Data are from participants that completed an essay writing task in the context of the Prowrite project.

```{r}
# Load data
data <- read_csv("https://osf.io/e5gy6/download") 
```

We reduced the data to three variables: a participant identifier `participant`, the inter-keystroke intervals `iki`, and the transition location `location` (levels: within-word, before-word, before-sentence).

```{r}
data
```

The data must be transformed into a list to feed them into Stan. The information required from the data is determined in the Stan code and has to do with what we need to estimate the model parameters and how the model is implemented. The exert from the Stan code below shows which information is expected as input (e.g. int is integer), how they are named, and what the smallest (i.e. lower) and largest possible value (i.e. upper) are. The full Stan code can be found at the end of this document.

```{r eval=F, engine='Rcpp'}
// Do not run
// Data chunk of Stan code
data {
  int<lower=1> N;                    // number of observations
  int<lower=1> nS;                   // number of ppts
  int<lower=1, upper=nS> ppts[N];    // ppts identifier
  int<lower=1> nB;                   // number of bigrams
  int<lower=1, upper=nB> bigrams[N]; // ppts identifier
  int<lower=1> K;                    // number of conditions
  int condition[N];                  // condition identifiers
  vector[N] y;                       // outcome
}
```

Depending on the data we might have different numbers of participants, bigrams, and conditions. The Stan code is using indices to be able to fit varying numbers for each. In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices. We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants), bigram identifiers `bigrams` (numbers 1 through $J$ where $J$ is the maximum number of bigrams), and a numeric identifier `components` for each copy task component (levels: consonants = 1, LF = 2).  The returned values are indices for the parameters in the model: For example, for the components identifier, `beta[1]` is the population estimate for non-hesitant typing in the consonants task, `beta[2]` is same for LF-bigram; `theta[1]` and `theta[2]` are the disfluency probability for the consonants and LF-bigrams task, respectively.

```{r}
ppts <- as.numeric(factor(data$ppts))
bigrams <- as.numeric(factor(data$bigram))
condition <- as.numeric(factor(data$component))
```

In the below code, the names on the left side of the arrow must correspond to the names expected in the Stan code (above); names on the right side do not. We assign the participant, bigram, and component identifiers created above and their maximum values `nS`, `nB`, `K`, respectively. The keystroke data `IKI` are assigned to `y` and the total number of observations (number of participants $\times$ number of bigrams produced by participant) is assigned to `N` (number of rows in the data `nrow(data`).

```{r}
data_list <- within( list(), {
  ppts <- ppts  
  nS <- max(ppts) # max no of ppts
  bigrams <- bigrams 
  nB <- max(bigrams) # max no of bigrams
  condition <- condition
  K <- max(condition) # no of conditions
  y <- data$IKI
  N <- nrow(data)
} )
```

The information in the data list can be viewed using the `glimpse` function.

```{r}
glimpse(data_list)
```

## Load model

The Stan code described **in the manuscript** was extended to fit more than one copy-task component (i.e. a categorical predictor with $K$ levels). The code is stored in the file "MoG.stan" (i.e. mixture of Gaussians) and can be downloaded from the "walkthrough" folder on [OSF](https://osf.io/t34jn/). Alternatively, `readLines` and `writeLines` can be used to download the Stan code from OSF so it will be saved in R's current working directory.

```{r eval = F}
# Download Stan code
file <- readLines("https://osf.io/j6dzt/download")
# Save Stan code to current working directory
writeLines(file, "MoG.stan")
```

The Stan model can be loaded for data fitting using the `stan_model` function and assign it to the object `mog`. The "MoG.stan" file needs to be available in the current working directory of R or the file path needs to be indicated.

```{r eval = F}
mog <- stan_model(file = "MoG.stan")
```

This code fits the posterior of a categorical predictor with any number of levels. In other words, the posterior that model returns can be used to calculate simple effects, main effects and interactions of a factorial design. We demonstrate below how to calculate a simple effect. The same logic can be used to calculate main effects and interactions. Also, participant estimates are calculated for both mixture components and the mixing proportions by condition. For including more predictors, or removing, e.g., random error terms, or adjusting priors, the user will have to work directly in the Stan code.

This Stan code used is largely based on @sorensen2016bayesian and @vasishth2017. @sorensen2016bayesian presents a detailed tutorial on how to fit Stan models [see also @lambert2018student].

## Initiate sampling

For the model to converge we need to run a sufficient number of iterations. 30,000 iterations, as below, are a lot but does not guarantee convergence. We need to test convergence using the model's posterior (see blow). To test whether the model has converged, we need to run the model more than one time (i.e. different chains). These can be run at the same time (in parallel) using more than one core of your computer (three cores below).^[If you want to use R to check how many cores are available on your machine, run `parallel::detectCores()` (you might need to install the package `parallel`).] There is no need to use more cores and chains and using less cores would mean that at least one process has to wait until after other processes are completed. When the model has settled on a parameter values, we should observe that all three streams (chains) overlap. Running long chains (many iterations) is useful to get more accurate parameter estimates.

```{r}
iterations <- 30000       # No of iterations
warmup <- 15000           # Warmup samples to be discarded
n_chains <- n_cores <- 3  # No of chains / cores used (one chain per core)
```

The model is fitting a number of parameters that are not relevant for our inference. To reduce the size of the R object that containts the posterior we can select the parameters of interest and disregard all paremters that are not relevant for our inference.

```{r}
# Parameters to keep in output
pars <- c("beta",                        # fluent typing
          "delta",                       # disfluency slowdown
          "theta",                       # mixing proportion
          "beta_s",                      # by-participant typing speed
          "theta_s",                     # by-participant disfluency probability
          "sigma",                       # variance component
          "sigma_diff", "sigmap_e", "sigma_e",  # variance by mixture component
          "sigma_u", "sigma_w",          # variance for random ppt and bigram intercepts
          "log_lik",                     # log likelihood (for model comparison) 
          "y_tilde")                     # predicted data 
```

The `sampling` function applies the model to the data using the information specified above (iterations, warmup, chains, cores, parameters `pars` to be saved). `save_warmup` is set to `FALSE` to discard the warmup samples (which are not used for inference anyway) to reduce the size of the posterior. To allow reproducibility (and because Bayesian models involves random number generation) we set the seed. The seed can be any number but using the same number ensures the use of the same random number. Lastly, the control argument was specified with higher values for `adapt_delta` and `max_treedepth`: using higher values here mean the model runs slower but supports more careful parameter estimation. The model is assigned to the variable `m`.

```{r eval = F}
# Fit model
m <- sampling(mog, 
              data = data_list,
              iter = iterations,
              warmup = warmup,
              chains = n_chains, 
              cores = n_cores,
              pars = pars, # Parameters to keep.
              save_warmup = FALSE, # Don't save the warmup samples.
              seed = 365, 
              control = list(adapt_delta = .96, # default: .9
                             max_treedepth = 12)) # default: 10
```

```{r echo =F}
#sampler_params <- get_sampler_params(m, inc_warmup = FALSE)
#sampler_params_chain1 <- sampler_params[[3]]#colnames(sampler_params_chain1)
#sapply(sampler_params, function(x) mean(x[, "accept_stat__"]))
#sapply(sampler_params, function(x) max(x[, "treedepth__"]))
#inits <- get_inits(m)
#inits_chain1 <- inits[[1]]
#print(inits_chain1)
```

Running this model will take a while to complete sampling depending on your hardware specifications. The time it took my machine to complete this job can me viewed using `get_elapsed_time`. Therefore it is worth to not use all cores for this to run or to use a dedicated high performance machine.

```{r echo = F}
m <- readRDS("MoG.rda")
```

```{r}
get_elapsed_time(m)/60 # in mins
```

## Reusing model output

The output of the model (i.e. the posterior) can be saved, so we don't need to run the model again (which admittedly can take a while). The function requires the name of the fitted model `m`. We prefer to keep the name of the output similar to the name of the Stan code used. The model is stored as compressed ".rda" file.

```{r eval = F}
saveRDS(m, "MoG.rda", compress = "xz")
```

The model can be load into the environment of R using `readRDS`.

```{r eval = F}
m <- readRDS("MoG.rda")
```

# Model diagnostics

## Convergence

Model convergence can be established in two practically simple techniques. First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain. We will look at the population-level parameters (names are determined by the model) corresponding to the Greek symbols used **in the manuscript** and assign those to `pars`.

```{r}
pars <- c("beta", "delta", "theta", "sigma", "sigmap_e", "sigma_e")
```

To create trace plots, we can apply the `stan_trace` function to the model `m` and extract the MCMC chains for the parameters in `pars`. The `alpha` argument makes the colours slightly more transparent. If the chains overlap and look like "fat hairy caterpillars", chains have converged on the same target distribution.

```{r}
# Check convergence
stan_trace(m, pars = pars, alpha = .5)
```

Second, we can calculate the $\hat{R}$ statistic. Successful convergence is reflected in $\hat{R}$ values smaller than 1.05 [@gelman1992]. $\hat{R}$ is similar to the F statistic: it tells us how much bigger the variability between chains is compared to the variability within chains. A value of $\approx 1$ indicates that the variability is essentially identical signifying convergence. For this we can use the `rhat` function applied to the model and the population parameters.

```{r}
summary(m, pars=pars)$summary %>% as.data.frame() %>% 
  rownames_to_column("parameter") %>% 
  select(parameter, Rhat)
```

Convergence problems can have many reasons and therefore many solutions: running longer chains (in particular longer warmups), increasing the maximum treedepth and the average acceptance probability (`adapt_delta`), specifying starting values, adjusting priors, constraining parameters, or changing the parametrisation of the model. Severe convergence problems are indicative of a misspecification of the model.

## Posterior predictive check

The model used the posterior parameter values to simulate hypothetical data sets. This happened for every iteration for every chains.

```{r echo = F, include=F}
(total_samples <- (iterations - warmup) * n_chains)
```

We can draw predicted data from the model output and compare these to the observed data. A model that makes reasonable predictions should fit the observed data.

Using the `as.matrix` function we extract a matrix of `y_tilde`, the simulated data. This returns a matrix with the size `total_samples` $\times$ `nrow(data)` (so `r scales::comma(total_samples)` $\times$ `r scales::comma(nrow(data))`). We use the `sample` function to draw `N` randomly sampled hypothetical data sets. The `ppc_dens_oberlay` function from the `bayesplot` package is then mapping the observed data ($y$; thick blue line) to the simulated data ($y_{rep}$; thin lightblue lines).

```{r}
y_tilde <- as.matrix(m, pars = "y_tilde")          # extract simulated data sets
N <- 50                                            # number of simulations to use
total_samples <- (iterations - warmup) * n_chains  # total number of samples
rnd_sims <- sample(total_samples, N)               # created random indices
y_tilde_sample <- y_tilde[rnd_sims,]               # draw N random simulations 
```

```{r}
library(bayesplot)
ppc_dens_overlay(data_list$y, y_tilde_sample) +
  scale_x_continuous(limits = c(0, 3000))
```

# Posterior probability distribution

At the core of Bayesian inference is the posterior probability distribution. For each model parameter we have `r scales::comma(total_samples)` posterior samples that form a posterior probability distribution. These samples are the results of the number of iterations after warmup for all chains.

```{r }
(total_samples <- (iterations - warmup) * n_chains)
```

The distribution of parameter values represents the uncertainty about parameter values given the data. There is a large range of things one can do with a posterior. Below we will focus on summarising parameter estimates, comparing conditions, and extracting by-participant estimates.

The `names` function can be used to remind us of the parameter names used in the model (we reduced the output to the first 15 parameter names). The meaning of the parameters is described **in the manuscript** and can be obtained from the Stan code. Indices refer to the copy-task component (1 = consonants; 2 = LF bigrams) and to the participant identifier for by-participant parameters (indicated by `_s`).

```{r}
names(m)[1:15] 
```

## Parameter-value estimates

The parameter estimates can be visualised using the `stan_hist` function with the model variable `m` and the to-be visualised parameters `pars` as arguments. The function returns a histogram for each other parameter values. Each histogram is the probability distribution of the parameter values indicating which values are more or less probable to be the true parameter value.

```{r}
stan_hist(m, pars = pars)
```

The posterior samples of the parameter values can be summarised using the `print` function. The `probs` argument requires the lower and upper bound of the probability interval that we are interested in. A lower bound of .025 and an upper bound of .975 gives the 95% probability interval (PI; also called "credible intervals"), i.e. the range that contains the true parameter value with a 95% probability given the data.

The output summaries the most probable parameter value as mean with its standard error (`se_mean`) and standard deviation (`sd`), the effective sample size (`n_eff`) indicating sampling efficiency and the convergence metric $\hat{R}$ (`Rhat`) we introduced above.

```{r}
print(m, pars = pars, probs = c(.025,.975))
```

For the following steps, we will focus on the three parameters that have conceptually interesting interpretations: (1) the average fluent typing speed $\beta$, (2) the disfluency slowdown $\delta$, and (3) the disfluency probability $\theta$.

```{r}
pars <- c("beta", "delta", "theta")
```

The `plot` function shows the posterior probability distribution of the three parameters for the consonants task (indicated as 1) and the LF-bigrams task (indicated as 2) summarised as median and 95% PI.

```{r}
plot(m, pars = pars, ci_level = .95) # ci = credible interval
```

The values for $\beta$ and $\delta$ are shown on a log-scale. To transform their values back to msecs we can extract the posterior samples using the `as.data.frame` function. We prefer the use of tibble objects.

```{r}
posterior <- as.data.frame(m, pars) %>% as_tibble() 
posterior
```

The `pivot_longer` function is transforming the data above to a long format with an additional column for component and the model parameters kept as columns. The `names_pattern` argument is using a regular expression to extract the number in the squared brackets.

```{r}
posterior_long <- pivot_longer(posterior, everything(), 
             names_to = c(".value", "component"), 
             names_pattern = "(.*)\\[(.)\\]")

posterior_long
```

This code is then transforming `beta` and `delta` to msecs using the exponential function `exp` to un-log the values. In order to transform the slowdown `delta` into msecs we need to add `beta` before using the exponential function; we can then subtract beta again. The `recode` function changes the component indices from 1 to "Consonants" and 2 to "LF bigrams".

```{r}
posterior_in_msecs <- mutate(posterior_long,
                             delta = exp(beta + delta) - exp(beta),
                             beta = exp(beta),
                             component = recode(component, `1` = "Consonants",
                                                           `2` = "LF bigrams"))
posterior_in_msecs
```

The posterior distribution of the parameter estimates can then be visualised in, for example, histograms.

```{r}
pivot_longer(posterior_in_msecs, beta:theta, names_to = "parameter") %>%
  ggplot(aes(x = value, colour = component, fill = component)) +
  geom_histogram(position = "identity", alpha = .25) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  scale_fill_brewer("Copy-task\ncomponents", palette = "Dark2") +
  scale_color_brewer("Copy-task\ncomponents", palette = "Dark2") +
  labs(x = "Posterior parameter estimate")
```

## Difference between copy-task components

We can calculate the differences between the copy-task components for each parameter value. This is giving an indication of whether fluent typing ($\beta$) is influenced by the copy task, maybe the size of the disfluency slowdown ($\delta$) is impacted by the task, or the disfluency probability ($\theta$).

To determine these differences we change the data format above and create a variable that indicates the parameter and one column for the corresponding values for each copy-task component. The last line calculates the difference between the consonants task and the LF-bigrams task for each of the three parameter values.

```{r}
posterior_by_component <- posterior_in_msecs %>%
  pivot_longer(beta:theta, names_to = "parameter") %>%
  group_by(component) %>% mutate(id = row_number()) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  select(-id) %>% # drop id column
  mutate(diff = Consonants - `LF bigrams`,  # calculate the difference between copy-task components
         parameter = paste0(parameter, "[diff]"))

posterior_by_component
```

The difference between the conditions can be summarised using the mean, the 95% PIs and the probability that the difference between the components is small 0 (indicated as e.g. $P(\hat{\beta} <0)$ for the parameter $\beta$; the hat $\hat{.}$ symbol indicates that the value is an estimate of the population parameter value). This summary tells us whether the difference between the consonants task and the LF-bigrams task is different from zero and the most probable value for the difference between tasks.

```{r}
posterior_by_component %>%
  group_by(parameter) %>%                      # group by parameters
  summarise(mean = mean(diff),                 # mean difference
            lower = quantile(diff, .025),      # 2.5% lower bound of difference
            upper = quantile(diff, .975),      # 97.5% upper bound of difference
            p = mean(diff < 0)) %>%            # prob that difference is negative
  mutate(across(where(is.numeric), round, 2))  # round after 2nd decimal place
```

```{r echo=F}
ps <- posterior_by_component %>%
  group_by(parameter) %>%
  summarise(mean = mean(diff),
            lower = quantile(diff, .025),
            upper = quantile(diff, .975),
            p = mean(diff < 0)) %>% 
  mutate(across(where(is.numeric), round, 2),
         parameter = gsub("\\[diff\\]", "", parameter)) 
```

These differences can be viewed in histograms. The vertical line indicates a difference between tasks corresponding to a value of zero. The area of the histogram to the left of this line corresponds to the probability `p` to observe a value smaller than zero as in the output above. We observe difference between copy-task components for all three parameters. Fluent typing is `r round(ps[ps$parameter == "beta",]$mean)` msecs (95% PI: `r round(ps[ps$parameter == "beta",]$lower)`, `r round(ps[ps$parameter == "beta",]$upper)`) slower for the consonants task, and hesitations are `r round(ps[ps$parameter == "delta",]$mean)` msecs (95% PI: `r round(ps[ps$parameter == "delta",]$lower)`, `r round(ps[ps$parameter == "delta",]$upper)`) longer and `r ps[ps$parameter == "theta",]$mean` ($\approx$ `r round(ps[ps$parameter == "theta",]$mean*100)`%; 95% PI: `r ps[ps$parameter == "theta",]$lower`, `r ps[ps$parameter == "theta",]$upper`) times more likely. Interestingly fluent typing (i.e. planning and executing motor plans) is slower in the consonant task even after modeling hesitations as a separate process.

```{r}
ggplot(posterior_by_component, aes(x = diff)) +
  geom_histogram(alpha = .5) +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed", size = .5) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  labs(x = "Difference between copy-task components")
```

In practice we rarely believe that only an effect of exactly null is indicating that there is no difference between our groups. Instead there are values that are non-different from null; for example a slowdown of 1 msecs is meaningless in most, if not all,  contexts. To take this into account we can determine the *region of practical equivalence* [ROPE, @kruschke2010believe; @kruschke2011bayesian; @kruschke2014doing]. We can define the ROPE as the range of values that is _a priori_ considered non-different from a null effect. The ROPE value indicates to what extent the posterior can not rule out a negligible effect. A meaningful difference between groups (or effects more generally) should have a small proportion of posterior samples within the ROPE.

To calculate the ROPE for out differences between the two copy-task components we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference. We can calculate a vector with posterior difference for each of the three parameters, indicated as $\beta_\text{diff}$, $\delta_\text{diff}$, and $\theta_\text{diff}$.

```{r}
# Fluent typing
beta_diff <- filter(posterior_by_component, parameter == "beta[diff]") %>% pull(diff)

# Slowdown for hesitant typing
delta_diff <- filter(posterior_by_component, parameter == "delta[diff]") %>% pull(diff)

# Proportion of hesitations
theta_diff <- filter(posterior_by_component, parameter == "theta[diff]") %>% pull(diff)
```

Also, we need to define the `range` argument which specifies the range of values are not considerably different from null. Of course this range will depend on what the parameter represents and researcher intuition. For example, for fluent typing we would consider small differences, of say 10 msecs, more meaningful than for disfluencies. If we consider differences of 10 msecs or smaller equivalent to a null difference, we can define the `range` argument with a lower bound of -10 and the upper bound of 10.

```{r}
library(bayestestR)
rope(beta_diff, range = c(-10, 10))
```

The ROPE for fluent typing speed contains `r round(pull(rope(beta_diff, range = c(-10, 10)))*100,2)`% of the posterior samples. In other words, there is a `r round(pull(rope(beta_diff, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.

For disfluencies we might want to consider a larger range of values as negligible differences between two tasks (also because the variability in disfluencies is larger than the variability for fluent key transitions). Even if we define a ROPE of -250 through 250 msecs, there are no posterior samples inside the ROPE indicating that the difference in typing hesitations are related to difficulty that arise prior to processing on the motor level.

```{r}
rope(delta_diff, range = c(-250, 250))
```

As for the disfluency probably, we need to bear in mind that estimates for $\theta$ are bound between 0 and 1. Hence, the difference $\theta_\text{diff}$ cannot ever exceed -1 and 1. If we consider a difference of 5% or smaller as negligible, we can define the ROPE as ranging from -.05 through .05. We observe that `r round(pull(rope(theta_diff, range = c(-.05, .05)))*100,2)`% of the posterior difference $\theta_\text{diff}$ is within the ROPE. In other words, a large proportion of posterior samples is supporting a meaningful difference between the two copy-task components. We observe that hesitations are largely more frequent in the consonants task.

```{r}
rope(theta_diff, range = c(-.05, .05))
```

The Stan code presented in this walk through can be used to estimate the posterior for two and more conditions: We can calculate the difference between conditions from the posterior as well as main effects and interactions for more complex factorial designs as any factorial design can be reduced to a single variable.^[For example, say we have a 2 $\times$ 2 factorial design with factor 1 having two levels AB and CD and factor 2 having the corresponding levels AC and BD. These two factors render four conditions A, B, C, D. From posterior samples for each level A through D we can calculate main effect 1 as $\text{ME1}=(A+B) - (C+D)$, main effect 2 as $\text{ME2}=(A+C) - (B+D)$, and their interaction as $\text{Interaction}=(A-B) - (C-D)$ (or $\text{Interaction}=(A-C) - (B-D)$) and summary statistics as shown above.]

## By-participant estimates

In some contexts we would like to obtain by-participant estimates of the average fluent typing-speed of a participant after accounting for disfluencies (i.e. $\beta_s$); by-participant disfluency information can be estimated the proportion of disfluencies (i.e. $\theta_s$). These estimates inform us about the fluent typing speed but also about the prevalence of typing hesitations.

As before we can extract the posterior from the model `m`. This time we use the parameters that stored by-participant estimates `beta_s` and `theta_s` corresponding to the population estimates `beta` and `theta`.

```{r}
pars <- c("beta_s", "theta_s")
posterior_ppts <- as.data.frame(m, pars = pars) %>% as_tibble()
names(posterior_ppts)[1:10]
```

This posterior tibble has the format "$<\text{parameter name}>\text{_s}[<\text{component id}>, <\text{participant id}>]$" where indices indicate the copy-task component and the participant identifier of the parameters. The following code is converting the `beta`s to msecs, creates a long format with columns `parameter` with the names above as levels and a `value` column with their respective values. The `parameter` column is then separated into `parameter` (levels: `beta`, `theta`), `component` (levels: `1`, `2`), and participant (an index for each participant) using "," to separate the three variables. To use "," as separator we replace "_s" with "," in the line before. Copy-task component names were changed as before. A preview is below.

```{r}
posterior_ppts_long <- posterior_ppts %>%
  # use exp() on everything starting with "beta"
  mutate(across(starts_with("beta"), exp)) %>%             
  pivot_longer(everything(), names_to = "parameter") %>%
  # replace "_s" with "," 
  mutate(parameter = gsub("_s", ",", parameter)) %>%
  # separate into parameter, component, ppt using "," as separator
  separate(parameter, into = c("parameter", "component", "participant")) %>% 
  mutate(component = recode(component, `1` = "Consonants", `2` = "LF bigrams")) %>%
  group_by(parameter) %>% mutate(id = row_number()) %>%
  pivot_wider(names_from = parameter, values_from = value) %>% 
  select(-id) # drop id

posterior_ppts_long
```

We can then summarise by-participant estimates for each component and parameter with the most probable parameter value and the 95% PI as before.

```{r}
posterior_ppts_long_summary <- posterior_ppts_long %>%
  pivot_longer(beta:theta, names_to = "parameter", values_to = "value") %>%
  group_by(parameter, component, participant) %>%
  summarise(mean = mean(value),
            lower = quantile(value, .025),
            upper = quantile(value, .975))

posterior_ppts_long_summary
```

To get an idea how the by-participant estimates look like, we extracted participants 1 through 5 for visualisation. From this subset we see that, although we observed overall larger population estimates for the consonants task, some participants do not show a difference in fluent typing speed and a similar frequency of hesitations (e.g. participant 2) or even slightly more hesitations in the LF-bigrams task (e.g. participant 1).

```{r }
posterior_ppts_long_summary %>%
  filter(participant %in% 1:5) %>% # only participants 1 through 5
  ggplot(aes(x = participant, 
             y = mean, ymin = lower, ymax = upper,
            colour = component)) +
  geom_pointrange(position = position_dodge(.75)) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  coord_flip() +
  labs(x = "Participant id", y = "Parameter estimates") +
  scale_colour_brewer("Copy-task\ncomponent", palette = "Dark2")
```

Finally, we can calculate and visualise the by-participant differences.

```{r}
posterior_ppts_long_diff_summary <- posterior_ppts_long %>% group_by(component) %>% 
  mutate(id = row_number()) %>%
  pivot_wider(names_from = component, 
              values_from = c(beta, theta)) %>%
  pivot_longer(-c(participant, id),
             names_to = c("parameter", ".value"), 
             names_pattern = "(.*)_(.*)") %>%
  mutate(diff = Consonants - `LF bigrams`,
         parameter = paste0(parameter, "[diff]")) %>% 
  select(participant, parameter, diff) %>%
  group_by(parameter, participant) %>%
  summarise(mean = mean(diff),
            lower = quantile(diff, .025),
            upper = quantile(diff, .975))

posterior_ppts_long_diff_summary
```

For simplicity we show the differences for the first 10 participants only. This calculation and visualisation allows researchers to identify participants that do or do not differ in terms of fluent typing speed or hesitation frequency, in our example for tasks that do and do not involve lexical information.

```{r }
posterior_ppts_long_diff_summary %>%
  filter(participant %in% 1:10) %>% # only participants 1 through 10
  ggplot(aes(x = factor(as.numeric(participant)), 
             y = mean, ymin = lower, ymax = upper)) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  geom_pointrange(position = position_dodge(.75)) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  coord_flip() +
  labs(x = "Participant id", y = "Difference between copy-task components") +
  scale_colour_brewer("Copy-task\ncomponent", palette = "Dark2")
```

# References

<div id="refs"></div>

# Stan code

```{r echo = F}
cat(readLines("MoG.stan"), sep = "\n")
```

# Session info

```{r}
sessionInfo()
```

### 2023-02-07 scripts/c2l1/get_ytilde.R

@@ -1,0 +1,85 @@
+# Extract simulated data and compare to read data
+# for all four models
+
+# Load packages
+library(tidyverse)
+library(rstan)
+
+# Extract model simulations
+get_ytilde <- function(files, idx) {
+  # Load model
+  m <- readRDS(files)
+  
+  # Get model name
+  model_name <- str_remove_all(files, "stanout\\/|c2l1\\/|\\.rda")
+  
+  # Extract simulations
+  y_tilde <- as.matrix(m, pars = "y_tilde")
+    
+  # Extract 100 simulations
+  y_tilde_sims <- y_tilde[idx,]
+
+  # Make data frame
+  sims <- y_tilde_sims %>% 
+    as_tibble() %>% 
+    rownames_to_column("sim_idx") %>% 
+    pivot_longer(-sim_idx) %>% 
+    mutate(data_idx = parse_number(name)) %>% 
+    arrange(data_idx) %>% 
+    mutate(model = model_name) %>% 
+    select(-name)
+  
+  return(sims)
+}
+
+# Load data
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000) %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) %>% 
+  mutate(data_idx = 1:n())
+
+count(d, ppt, condition)
+
+# Number of simulations per model
+nsims <- 100
+
+# Max number of iterations
+maxiter <- 15000
+
+# Sample random idx for simulations
+idx <- sample(maxiter, nsims)
+
+# File names
+(files <- str_c("stanout/c2l1/", 
+                c("mog", "lmm", "lmmgaus", "lmmuneqvar"), 
+                ".rda"))
+
+# Apply get_ytilde function
+sims <- map_dfr(files, get_ytilde, idx = idx)
+
[…]

### 2023-02-07 scripts/c2l1/lmm_beforesent.R

@@ -1,0 +1,88 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+d <- read_csv("data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "before sentence") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 14,
+                             adapt_delta = 0.96)
+)
+
[…]

### 2023-02-07 scripts/c2l1/lmmgaus_beforeword.R

@@ -1,0 +1,98 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "before word") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Check lmer for starting values
+m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 1000
+          , beta_sigma = 100
+          , sigma_mu = 2000
+          , sigma_sigma = 100
+          , sigma_u = 470)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmgaus.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2023-02-07 scripts/c2l1/lmmuneqvar_beforesent.R

@@ -1,0 +1,93 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+d <- read_csv("data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "before sentence") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5.5
+          , beta_sigma = .1
+          , sigma_mu = 1.25
+          , sigma_sigma = .1
+          , sigma_u = 0.4)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmuneqvar.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
[…]

### 2023-02-07 scripts/c2l1/lmmuneqvar_withinword.R

@@ -1,0 +1,96 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "within word") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , sigma_mu = 1.25
+          , sigma_sigma = .1
+          , sigma_u = 0.4)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmuneqvar.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
[…]

### 2023-02-07 scripts/c2l1/modelcomparison_beforesent.R

@@ -1,0 +1,35 @@
+library(loo)
+library(tidyverse)
+
+path <- "stanout/c2l1"
+(files <- list.files(path, pattern = "_beforesent.rda$", full.names = T))
+
+for(file in files){
+    varname <- str_remove_all(file, ".rda|stanout/c2l1/")
+    m <- readRDS(file)
+    log_lik <- extract_log_lik(m, merge_chains = F) 
+    r_eff <- relative_eff(exp(log_lik)) 
+    assign(paste0("loo_",varname), loo(log_lik, r_eff = r_eff, cores = 2))  
+    print(varname); rm(list = "m"); gc()
+}
+
+(loos <- ls(pattern = "loo_.*_beforesent"))
+mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
+
+mcs <- mcs %>% as.data.frame() %>% 
+  rownames_to_column("model") %>%
+  as_tibble() %>%
+  mutate(model = recode(model, model1 = loos[1],
+                               model2 = loos[2],
+                               model3 = loos[3],
+                               model4 = loos[4]),
+         model = str_remove(model, "^loo_")) %>% 
+  separate(model, into = c("model", "location"), sep = "_");mcs
+
+file_out <- paste0("stanout/c2l1/modelcomparison_beforesent.csv")
+write_csv(mcs, file_out)
+
+
+
+
+

### 2023-02-07 scripts/c2l1/modelcomparison_beforeword.R

@@ -1,0 +1,35 @@
+library(loo)
+library(tidyverse)
+
+path <- "stanout/c2l1"
+(files <- list.files(path, pattern = "_beforeword.rda$", full.names = T))
+
+for(file in files){
+    varname <- str_remove_all(file, ".rda|stanout/c2l1/")
+    m <- readRDS(file)
+    log_lik <- extract_log_lik(m, merge_chains = F) 
+    r_eff <- relative_eff(exp(log_lik)) 
+    assign(paste0("loo_",varname), loo(log_lik, r_eff = r_eff, cores = 2))  
+    print(varname); rm(list = "m"); gc()
+}
+
+(loos <- ls(pattern = "loo_.*beforeword$"))
+mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
+
+mcs <- mcs %>% as.data.frame() %>% 
+  rownames_to_column("model") %>%
+  as_tibble() %>%
+  mutate(model = recode(model, model1 = loos[1],
+                               model2 = loos[2],
+                               model3 = loos[3],
+                               model4 = loos[4]),
+         model = str_remove(model, "^loo_")) %>% 
+  separate(model, into = c("model", "location"), sep = "_");mcs
+
+file_out <- paste0("stanout/c2l1/modelcomparison_beforeword.csv")
+write_csv(mcs, file_out)
+
+
+
+
+

### 2023-02-07 scripts/c2l1/modelcomparison_overall.R

@@ -1,0 +1,29 @@
+library(loo)
+library(tidyverse)
+
+path <- "stanout/c2l1"
+(files <- list.files(path, pattern = "^.[^\\_]*.rda$", full.names = T))
+
+for(file in files){
+    varname <- str_remove_all(file, ".rda|stanout/c2l1/")
+    m <- readRDS(file)
+    log_lik <- extract_log_lik(m, merge_chains = F) 
+    r_eff <- relative_eff(exp(log_lik)) 
+    assign(paste0("loo_",varname), loo(log_lik, r_eff = r_eff, cores = 2))  
+    print(varname); rm(list = "m"); gc()
+}
+
+(loos <- ls(pattern = "loo_.*"))
+mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
+
+mcs <- mcs %>% as.data.frame() %>% 
+  rownames_to_column("model") %>%
+  as_tibble() %>%
+  mutate(model = recode(model, model1 = loos[1],
+                               model2 = loos[2],
+                               model3 = loos[3],
+                               model4 = loos[4]),
+         model = str_remove(model, "^loo_"));mcs
+
+file_out <- paste0("stanout/c2l1/modelcomparison_overall.csv")
+write_csv(mcs, file_out)

### 2023-02-07 scripts/c2l1/modelcomparison_withinword.R

@@ -1,0 +1,35 @@
+library(loo)
+library(tidyverse)
+
+path <- "stanout/c2l1"
+(files <- list.files(path, pattern = "_withinword.rda$", full.names = T))
+
+for(file in files){
+    varname <- str_remove_all(file, ".rda|stanout/c2l1/")
+    m <- readRDS(file)
+    log_lik <- extract_log_lik(m, merge_chains = F) 
+    r_eff <- relative_eff(exp(log_lik)) 
+    assign(paste0("loo_",varname), loo(log_lik, r_eff = r_eff, cores = 2))  
+    print(varname); rm(list = "m"); gc()
+}
+
+(loos <- ls(pattern = "loo_.*_withinword"))
+mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
+
+mcs <- mcs %>% as.data.frame() %>% 
+  rownames_to_column("model") %>%
+  as_tibble() %>%
+  mutate(model = recode(model, model1 = loos[1],
+                               model2 = loos[2],
+                               model3 = loos[3],
+                               model4 = loos[4]),
+         model = str_remove(model, "^loo_")) %>% 
+  separate(model, into = c("model", "location"), sep = "_");mcs
+
+file_out <- paste0("stanout/c2l1/modelcomparison_withinword.csv")
+write_csv(mcs, file_out)
+
+
+
+
+

### 2023-02-07 scripts/c2l1/mog.R

@@ -1,0 +1,193 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000) %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = 0
+          , theta_sigma = .1
+          , tau = .01
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Stan model
+# Load model
+mog <- stan_model(file = "stan/mog.stan")
+
+# Parameters to omit in output
+omit <- c("theta", "prob_tilde", "mu",  "_mu", "_raw", "_sigma", "z_u", "theta_s", "L_u")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
[…]

### 2023-02-07 scripts/c2l1/mog_withinword.R

@@ -1,0 +1,98 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "within word") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = 1
+#          , beta_raw = rep(0, dat$K)
+#          , delta = rep(.1, 1)
+#          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+#          , theta_raw = rep(0, dat$K)
+          , theta_mu = 0
+          , theta_sigma = .5
+          , tau = .1
+          , sigma = .5
+ #         , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.15)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+mog <- stan_model(file = "stan/mog.stan")
+
+# Parameters to omit in output
+omit <- c("theta", "prob_tilde", "mu",  "_mu", "_raw", "_sigma", "z_u", "theta_s", "L_u")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
[…]

### 2023-02-07 scripts/cato/get_ytilde.R

@@ -1,0 +1,85 @@
+# Extract simulated data and compare to read data
+# for all four models
+
+# Load packages
+library(tidyverse)
+library(rstan)
+
+# Extract model simulations
+get_ytilde <- function(files, idx) {
+  # Load model
+  m <- readRDS(files)
+  
+  # Get model name
+  model_name <- str_remove_all(files, "stanout\\/|cato\\/|\\.rda")
+  
+  # Extract simulations
+  y_tilde <- as.matrix(m, pars = "y_tilde")
+    
+  # Extract 100 simulations
+  y_tilde_sims <- y_tilde[idx,]
+
+  # Make data frame
+  sims <- y_tilde_sims %>% 
+    as_tibble() %>% 
+    rownames_to_column("sim_idx") %>% 
+    pivot_longer(-sim_idx) %>% 
+    mutate(data_idx = parse_number(name)) %>% 
+    arrange(data_idx) %>% 
+    mutate(model = model_name) %>% 
+    select(-name)
+  
+  return(sims)
+}
+
+# Load data
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/cato.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000) %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, xn, dystyp, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep) %>% 
+  mutate(data_idx = 1:n())
+
+count(d, ppt, condition)
+
+# Number of simulations per model
+nsims <- 100
+
+# Max number of iterations
+maxiter <- 15000
+
+# Sample random idx for simulations
+idx <- sample(maxiter, nsims)
+
+# File names
+(files <- str_c("stanout/cato/", 
+                c("mog", "lmm", "lmmgaus", "lmmuneqvar"), 
+                ".rda"))
+
+# Apply get_ytilde function
+sims <- map_dfr(files, get_ytilde, idx = idx)
+
[…]

### 2023-02-07 scripts/cato/lmm_beforesent.R

@@ -1,0 +1,90 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+d <- read_csv("data/cato.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "before sentence") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(xn, dystyp, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
[…]

### 2023-02-07 scripts/cato/lmm_withinword.R

@@ -1,0 +1,98 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/cato.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "within word") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(xn, dystyp, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2023-02-07 scripts/cato/lmmuneqvar_beforeword.R

@@ -1,0 +1,98 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/cato.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "before word") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(xn, dystyp, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1.25
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.4)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmuneqvar.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2023-02-07 scripts/cato/lmmuneqvar_withinword.R

@@ -1,0 +1,98 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/cato.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "within word") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(xn, dystyp, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1.25
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.4)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmuneqvar.stan")
+
+# Parameters to omit in output
+omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+
+# Fit model
+m <- sampling(lmm, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
[…]

### 2023-02-07 scripts/cato/modelcomparison_beforeword.R

@@ -1,0 +1,35 @@
+library(loo)
+library(tidyverse)
+
+path <- "stanout/cato"
+(files <- list.files(path, pattern = "_beforeword.rda$", full.names = T))
+
+for(file in files){
+    varname <- str_remove_all(file, ".rda|stanout/cato/")
+    m <- readRDS(file)
+    log_lik <- extract_log_lik(m, merge_chains = F) 
+    r_eff <- relative_eff(exp(log_lik)) 
+    assign(paste0("loo_",varname), loo(log_lik, r_eff = r_eff, cores = 2))  
+    print(varname); rm(list = "m"); gc()
+}
+
+(loos <- ls(pattern = "loo_.*beforeword$"))
+mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
+
+mcs <- mcs %>% as.data.frame() %>% 
+  rownames_to_column("model") %>%
+  as_tibble() %>%
+  mutate(model = recode(model, model1 = loos[1],
+                               model2 = loos[2],
+                               model3 = loos[3],
+                               model4 = loos[4]),
+         model = str_remove(model, "^loo_")) %>% 
+  separate(model, into = c("model", "location"), sep = "_");mcs
+
+file_out <- paste0("stanout/cato/modelcomparison_beforeword.csv")
+write_csv(mcs, file_out)
+
+
+
+
+

### 2023-02-07 scripts/cato/modelcomparison_withinword.R

@@ -1,0 +1,35 @@
+library(loo)
+library(tidyverse)
+
+path <- "stanout/cato"
+(files <- list.files(path, pattern = "_withinword.rda$", full.names = T))
+
+for(file in files){
+    varname <- str_remove_all(file, ".rda|stanout/cato/")
+    m <- readRDS(file)
+    log_lik <- extract_log_lik(m, merge_chains = F) 
+    r_eff <- relative_eff(exp(log_lik)) 
+    assign(paste0("loo_",varname), loo(log_lik, r_eff = r_eff, cores = 2))  
+    print(varname); rm(list = "m"); gc()
+}
+
+(loos <- ls(pattern = "loo_.*_withinword"))
+mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
+
+mcs <- mcs %>% as.data.frame() %>% 
+  rownames_to_column("model") %>%
+  as_tibble() %>%
+  mutate(model = recode(model, model1 = loos[1],
+                               model2 = loos[2],
+                               model3 = loos[3],
+                               model4 = loos[4]),
+         model = str_remove(model, "^loo_")) %>% 
+  separate(model, into = c("model", "location"), sep = "_");mcs
+
+file_out <- paste0("stanout/cato/modelcomparison_withinword.csv")
+write_csv(mcs, file_out)
+
+
+
+
+

### 2023-02-07 scripts/cato/mog.R

@@ -1,0 +1,198 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+# Load df
+d <- read_csv("data/cato.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000) %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, xn, dystyp, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = 0
+          , theta_sigma = .1
+          , tau = .01
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Stan model
+# Load model
+mog <- stan_model(file = "stan/mog.stan")
+
+# Parameters to omit in output
+omit <- c("theta", "prob_tilde", "mu",  "_mu", "_raw", "_sigma", "z_u", "theta_s", "L_u")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
[…]

### 2023-02-07 scripts/cato/mog_beforesent.R

@@ -1,0 +1,94 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 10000
+n_sample <- 50 # number of random data points
+
+d <- read_csv("data/cato.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000,
+         location == "before sentence") %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(xn, dystyp, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num) 
+
+# Sample within each category random data points
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_sample) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta_raw = rep(0, dat$K)
+          , theta_mu = -1
+          , theta_sigma = .1
+          , tau = .01
+          , sigma = 1
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+mog <- stan_model(file = "stan/mog.stan")
+
+# Parameters to omit in output
+omit <- c("theta", "prob_tilde", "mu",  "_mu", "_raw", "_sigma", "z_u", "theta_s", "L_u")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
[…]

### 2023-02-07 scripts/cato/processing.R

@@ -1,0 +1,45 @@
+library(tidyverse)
+library(janitor)
+
+files <- list.files("data/MarksData/", pattern = "CATO", full.names = T)
+
+read_delim(files, delim = "\t") %>% 
+  clean_names() %>% 
+  filter(!is.na(ke_location), 
+         ke_location %in% c("a_^a", "a^a", "a._^a"),
+         pwd_curs_up == 0,
+         pwd_curs_down == 0,
+         pwd_curs_left == 0,
+         pwd_curs_right == 0,
+         pwd_deletes == 0,
+         ismodified == 0
+         ) %>% 
+  select(-id, -dtxn, -starts_with("log"), 
+         -condit, -starts_with("pwd"), 
+         -event_time, 
+         -iki_mod, 
+         -w_spell,
+         -pword_id,
+         -ismodified,
+         -ke_code,
+         -starts_with("word"), 
+         -inword,
+         -starts_with("is")) %>% 
+  mutate(across(dystyp, recode, 
+                D = "dyslexic", 
+                K = "non dyslexic"),
+         across(xn, recode, 
+                X = "masked", 
+                N = "unmasked"),
+         across(ke_location, recode,
+                `a_^a` = "before word",
+                `a^a` = "within word",
+                `a._^a` = "before sentence"),
+         across(subno, ~as.numeric(factor(.)))) %>% 
+  rename(location = ke_location) %>% 
+  remove_constant() %>% 
+  remove_empty() %>% 
+  #count(subno, xn, dystyp, location)
+  write_csv("data/cato.csv")
+
+  

### 2023-02-07 tutorial/tutorial.Rmd
Rename stan file

@@
 options(scipen = 999) # Turn off scientific notation
-opts_chunk$set(fig.width = 8, fig.height = 4.5)
-opts_knit$set(width=90)
-knitr::opts_chunk$set(echo = TRUE,
+knitr::opts_knit$set(width=90)
+knitr::opts_chunk$set(fig.width = 8, 
+                      fig.height = 4.5,
+                      echo = TRUE,
                       comment = NA, 
@@
 
-# Fitting the model
-
-## Preparing the data
+# Preparing the data
 
 First we load a sample of keystroke data.
-Data are loaded from [OSF](https://osf.io/y3p4d/) using the URL as below and the `read_csv` function.
+Data are loaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/) using the URL as below and the `read_csv` function.
 For brevity we use a random subset.
@@
 # Load data
-data <- read_csv("https://osf.io/e5gy6/download") 
+data <- read_csv("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/data/sampledata.csv") 
 ```
@@
 ```
+
+We reduced the number of keystroke transitions to `n` observations per participant and per location.
+
+```{r}
+count(data, participant, location)
+```
+
+# Uni-modal mixed-effects models
+
+## Gaussian mixed-effects model
+
+### Model
+
+$$
+\text{iki}_i \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
+\mu_i = \beta_\text{location[i]} + u_i\\
+u \sim \text{N}(0, \sigma_u) 
+$$
+
+```{r}
+formula <- bf(iki ~ 0 + location + (1|participant), family = gaussian())
+```
+
+### Priors
+
+```{r}
+get_prior(formula = formula, data = data)
+```
+
+```{r}
+prior <- set_prior('student_t(3, 350, 100)', class = 'b')
+```
+
+```{r}
+mean <- 350
+sd <- 100
+tibble(x = 0:1000) %>% 
+  mutate(df_3 = dstudent_t(x, 3, mean, sd),
+         df_10 = dstudent_t(x, 10, mean, sd),
+         df_100 = dstudent_t(x, 100, mean, sd)) %>% 
+  pivot_longer(-x) %>% 
+  mutate(across(name, str_remove, "df_"),
+         across(name, factor, levels = c(3, 10, 100), ordered = TRUE)) %>% 
+  ggplot(aes(x = x, y = value, colour = name)) +
+  geom_line() +
+  labs(colour = "Degrees of\nfreedom")
+```
+
+### Initiate sampling
+
+For the model to converge we need to run a sufficient number of iterations.
+30,000 iterations, as below, are a lot but does not guarantee convergence.
+We need to test convergence using the model's posterior (see blow).
+To test whether the model has converged, we need to run the model more than one time (i.e.
+different chains).
+These can be run at the same time (in parallel) using more than one core of your computer (three cores below).^[If you want to use R to check how many cores are available on your machine, run `parallel::detectCores()` (you might need to install the package `parallel`).] There is no need to use more cores and chains and using less cores would mean that at least one process has to wait until after other processes are completed.
+When the model has settled on a parameter values, we should observe that all three streams (chains) overlap.
+Running long chains (many iterations) is useful to get more accurate parameter estimates.
+
+```{r}
+iterations <- 30000       # No of iterations
+warmup <- iterations/2    # Warmup samples to be discarded
+nchains <- ncores <- 3  # No of chains / cores used (one chain per core)
+```
+
+### Fitting the model
+
+The following code combines the model, the data, and the prior and returns the posterior. This will take a couple of minutes.
+
+```{r eval = F}
+fit_gaus <- brm(formula = formula, 
+                 data = data,
+                 prior = prior,
+                 chains = nchains,
+                 cores = ncores,
+                 iter = iterations,
+                 warmup = warmup,
+                 sample_prior = TRUE,
+                 inits = 0,
+                 seed = 365)
+```
+
+Because Bayesian models tend to take more time to run, it makes sense to save the results using `saveRDS` for later re-use.
+
+```{r eval = F}
+saveRDS(object = fit_gaus,
+        file = "stan/gaussian.rda", 
+        compress = "xz")
+```
+
+To read the fitted model back into your working environment used `readRDS`.
+
+```{r}
+fit_gaus <- readRDS("stan/gaussian.rda")
+```
+
+The function `fixef` provide a summary of the model inference showing the posterior estimates as average and 95% probability interval by transition location in msecs.
+
+```{r}
+fixef(fit_gaus) %>% round()
+```
+
+## Log-Gaussian mixed-effects model
+
+### Model
+
+$$
+\text{iki}_i \sim \text{logN}(\mu_i, \sigma_\text{e}^2)\\
+\mu_i = \beta_\text{location[i]} + u_i\\
+u \sim \text{N}(0, \sigma_u) 
+$$
+
+```{r}
+formula <- bf(iki ~ 0 + location + (1|participant), family = lognormal())
+```
+
+### Priors
+
+```{r}
+get_prior(formula = formula, data = data)
+```
+
+```{r}
+prior <- set_prior('student_t(3, 5, .5)', class = 'b')
+```
+
+```{r}
+mean <- 5
+sd <- .5
+tibble(x = seq(3, 7, .01)) %>% 
+  mutate(df_3 = dstudent_t(x, 3, mean, sd),
+         df_10 = dstudent_t(x, 10, mean, sd),
+         df_100 = dstudent_t(x, 100, mean, sd)) %>% 
+  pivot_longer(-x) %>% 
+  mutate(across(name, str_remove, "df_"),
+         across(name, factor, levels = c(3, 10, 100), ordered = TRUE)) %>% 
+  ggplot(aes(x = x, y = value, colour = name)) +
+  geom_line() +
+  labs(colour = "Degrees of\nfreedom") 
+```
+
+### Fitting the model
+
+```{r eval = F}
+fit_lgaus <- brm(formula = formula, 
+                 data = data,
+                 prior = prior,
+                 chains = nchains,
+                 cores = ncores,
+                 iter = iterations,
+                 warmup = warmup,
+                 sample_prior = TRUE,
+                 inits = 0,
+                 seed = 365)
+```
+
+Save the results using `saveRDS`.
+
+```{r eval = F}
+saveRDS(object = fit_lgaus,
+        file = "stan/loggaussian.rda", 
+        compress = "xz")
+```
+
+The function `fixef` provide a summary of the inferred posterior estimates as average and 95% probability interval by transition location in log msecs.
+
+```{r echo = F}
+fit_lgaus <- readRDS("stan/loggaussian.rda")
+```
+
+```{r}
+fixef(fit_lgaus) %>% round(1)
+```
+
+## Model comparison
+
+To evaluate the predictive performance of the models we can use leave-one-out cross-validation.
+The out-of-sample predictive performance was determined via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
+Cross-validation, in contrast to more conventional model-comparison metrics such as $R^2$, prevent overfitting by penalising models with more parameters [see @mcelreath2016statistical; @lambert2018student].
+Predictive quality was summarised as the sum of the expected log predictive density (`elpd_loo`; $\widehat{elpd}$).
+The difference in `elpd_loo` is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
+
+```{r}
+loo_fit <- loo(fit_gaus, fit_lgaus)
+loo_fit$diffs %>% 
+  as.data.frame() %>% 
+  round(1)
+```
+The model comparison shows that the model `fit_lgaus` (the model with the log-Gaussian likelihood) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` times better than `fit_gaus` (try `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])`).
+
+# Bi-modal mixed-effects log-Gaussian mixture model
+
+The model returns population estimates for the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the frequency of hesitant transitions ($\theta$) for each level of a categorical predictor after taking into account three sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor; (3) keystroke intervals for some bigrams are slower than for others.
+
+$$
+
+$$
+
+## Mixture model app
 
@@
 
-Depending on the data we might have different numbers of participants, bigrams, and conditions.
+Depending on the data we might have different numbers of participants and conditions.
 The Stan code is using indices to be able to fit varying numbers for each.
 In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices.
-We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants), bigram identifiers `bigrams` (numbers 1 through $J$ where $J$ is the maximum number of bigrams), and a numeric identifier `components` for each copy task component (levels: consonants = 1, LF = 2).
-The returned values are indices for the parameters in the model: For example, for the components identifier, `beta[1]` is the population estimate for non-hesitant typing in the consonants task, `beta[2]` is same for LF-bigram; `theta[1]` and `theta[2]` are the disfluency probability for the consonants and LF-bigrams task, respectively.
+We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `location` for each transition locations (levels: before-sentence = 1, before-word = 2, within-word = 3).
+The returned values are indices for the parameters in the model: For example, for the location identifier, `beta[1]` is the population estimate for non-hesitant typing intervals at before-sentence transition locations, `beta[2]` is same for before-word transition locations; `theta[1]` and `theta[2]` are the disfluency probability on the logit scale for before-sentence and before-word transition locations, respectively.
 
 ```{r}
-ppts <- as.numeric(factor(data$ppts))
-bigrams <- as.numeric(factor(data$bigram))
-condition <- as.numeric(factor(data$component))
+ppts <- pull(data, participant) %>% factor() %>% as.numeric()
+condition <- pull(data, location) %>% factor() %>% as.numeric()
 ```
@@
 In the below code, the names on the left side of the arrow must correspond to the names expected in the Stan code (above); names on the right side do not.
-We assign the participant, bigram, and component identifiers created above and their maximum values `nS`, `nB`, `K`, respectively.
-The keystroke data `IKI` are assigned to `y` and the total number of observations (number of participants $\times$ number of bigrams produced by participant) is assigned to `N` (number of rows in the data `nrow(data`).
+We assign the participant and location identifiers created above and their maximum values `nS` and `K`, respectively.
+The keystroke data `iki` are assigned to `y` and the total number of observations is assigned to `N` (number of rows in the data `nrow(data`).
 
@@
   nS <- max(ppts) # max no of ppts
-  bigrams <- bigrams 
-  nB <- max(bigrams) # max no of bigrams
   condition <- condition
   K <- max(condition) # no of conditions
-  y <- data$IKI
+  y <- data$iki
   N <- nrow(data)
@@
 
-The Stan code described **in the manuscript** was extended to fit more than one copy-task component (i.e.
-a categorical predictor with $K$ levels).
-The code is stored in the file "MoG.stan" (i.e.
-mixture of Gaussians) and can be downloaded from the "walkthrough" folder on [OSF](https://osf.io/t34jn/).
-Alternatively, `readLines` and `writeLines` can be used to download the Stan code from OSF so it will be saved in R's current working directory.
+The Stan code can be downloaded from [GitHub](https://osf.io/t34jn/) using `readLines` and `writeLines`.
+The code will be saved in R's current working directory as "molg.stan" (mixture of log-Gaussians).
 
@@
 # Save Stan code to current working directory
-writeLines(file, "MoG.stan")
+writeLines(file, "molg.stan")
 ```
@@
 
-## Initiate sampling
-
-For the model to converge we need to run a sufficient number of iterations.
-30,000 iterations, as below, are a lot but does not guarantee convergence.
-We need to test convergence using the model's posterior (see blow).
-To test whether the model has converged, we need to run the model more than one time (i.e.
-different chains).
-These can be run at the same time (in parallel) using more than one core of your computer (three cores below).^[If you want to use R to check how many cores are available on your machine, run `parallel::detectCores()` (you might need to install the package `parallel`).] There is no need to use more cores and chains and using less cores would mean that at least one process has to wait until after other processes are completed.
-When the model has settled on a parameter values, we should observe that all three streams (chains) overlap.
-Running long chains (many iterations) is useful to get more accurate parameter estimates.
-
-```{r}
-iterations <- 30000       # No of iterations
-warmup <- 15000           # Warmup samples to be discarded
-n_chains <- n_cores <- 3  # No of chains / cores used (one chain per core)
-```
-
 The model is fitting a number of parameters that are not relevant for our inference.
-To reduce the size of the R object that containts the posterior we can select the parameters of interest and disregard all paremters that are not relevant for our inference.
+To reduce the size of the R object that contains the posterior we can select the parameters of interest and disregard all parameters that are not relevant for our inference.
 
@@
           "sigma_diff", "sigmap_e", "sigma_e",  # variance by mixture component
-          "sigma_u", "sigma_w",          # variance for random ppt and bigram intercepts
+          "sigma_u",                     # variance for random ppt 
           "log_lik",                     # log likelihood (for model comparison) 

### 2023-02-07 tutorial/tutorial.Rmd
Update prior

@@
 
+The output of the model (i.e.
+the posterior) can be saved, so we don't need to run the model again (which can take a while).
+The function requires the name of the fitted model `fit_gaus`.
+We prefer to keep the name of the output similar to the name of the Stan code used and ideally the R script that is running the model so it is easy to find the script that is running which model.
+The model is stored as compressed ".rda" file.
+
 ```{r eval = F}
@@
 data_list <- within( list(), {
-  ppts <- ppts  
+  ppt <- ppts  
   nS <- max(ppts) # max no of ppts
@@
 
-The Stan code can be downloaded from [GitHub](https://osf.io/t34jn/) using `readLines` and `writeLines`.
-The code will be saved in R's current working directory as "molg.stan" (mixture of log-Gaussians).
+The Stan code can be downloaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/blob/main/tutorial/stan/molg.stan) using `readLines` and `writeLines`.
+The code will be saved as "molg.stan" (mixture of log-Gaussians) in a directory called "stan" which needs to be available in your current working directory.
 
 ```{r eval = F}
+# Create an empty directory called "stan"
+dir.create("stan")
 # Download Stan code
-file <- readLines("https://osf.io/j6dzt/download")
+file <- readLines("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/stan/molg.stan")
 # Save Stan code to current working directory
-writeLines(file, "molg.stan")
+writeLines(file, "stan/molg.stan")
 ```
 
-The Stan model can be loaded for data fitting using the `stan_model` function and assign it to the object `mog`.
-The "MoG.stan" file needs to be available in the current working directory of R or the file path needs to be indicated.
+The Stan model can be loaded into R using the `stan_model` function and assign it to the object `molg`.
+The "molg.stan" file needs to be available in the directory "stan" located inside the current working directory.
 
 ```{r eval = F}
-mog <- stan_model(file = "MoG.stan")
+molg <- stan_model(file = "stan/molg.stan")
 ```
@@
           "delta",                       # disfluency slowdown
-          "theta",                       # mixing proportion
-          "beta_s",                      # by-participant typing speed
-          "theta_s",                     # by-participant disfluency probability
+          "theta",                       # mixing proportion on logit scale
+          "prob",                        # mixing proportion as probability
+          "theta_s",                     # by-participant disfluency probability (logit scale)
+          "prob_s",                      # by-participant disfluency probability
           "sigma",                       # variance component
@@
 # Fit model
-m <- sampling(mog, 
+fit_molg <- sampling(molg, 
                      data = data_list,
@@
                      warmup = warmup,
-              chains = n_chains, 
-              cores = n_cores,
+                     chains = nchains, 
+                     cores = ncores,
                      pars = pars, # Parameters to keep.
@@
                      seed = 365, 
-              control = list(adapt_delta = .96, # default: .9
-                             max_treedepth = 12)) # default: 10
+                     control = list(adapt_delta = 0.96,  # default: 0.9
+                                    max_treedepth = 14)) # default: 10
 ```
@@
 
+Save the results using `saveRDS`.
+
+```{r eval = F}
+saveRDS(object = fit_molg,
+        file = "stan/molg.rda", 
+        compress = "xz")
+```
+
+```{r echo = F}
+fit_molg <- readRDS("stan/molg.rda")
+```
+
+The function `fixef` provide a summary of the inferred posterior estimates as average and 95% probability interval by transition location in log msecs.
+
+```{r}
+print(fit_molg, pars = c("beta", "delta", "theta"), probs = c(.025, .975))
+```
+
 Running this model will take a while to complete sampling depending on your hardware specifications.
@@
 
-```{r echo = F}
-m <- readRDS("MoG.rda")
+```{r}
+get_elapsed_time(fit_molg)/60 # in mins
 ```
 
-```{r}
-get_elapsed_time(m)/60 # in mins
-```
+## Model diagnostics
 
-## Reusing model output
-
-The output of the model (i.e.
-the posterior) can be saved, so we don't need to run the model again (which admittedly can take a while).
-The function requires the name of the fitted model `m`.
-We prefer to keep the name of the output similar to the name of the Stan code used.
-The model is stored as compressed ".rda" file.
-
-```{r eval = F}
-saveRDS(m, "MoG.rda", compress = "xz")
-```
-
-The model can be load into the environment of R using `readRDS`.
-
-```{r eval = F}
-m <- readRDS("MoG.rda")
-```
-
-# Model diagnostics
-
-## Convergence
+### Convergence
 
@@
 
-## Posterior predictive check
+### Posterior predictive check
 
@@
 
-# Posterior probability distribution
+# Posterior probability distributions
 

### 2023-02-07 tutorial/tutorial.Rmd
Swap prob in stan model

@@
 
-The model returns population estimates for the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the frequency of hesitant transitions ($\theta$) for each level of a categorical predictor after taking into account three sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor; (3) keystroke intervals for some bigrams are slower than for others.
+The model returns population estimates for the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the frequency of hesitant transitions ($\theta$) for each level of a categorical predictor after taking into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
 
 $$
-
+\text{iki}_{i} \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_i, \sigma_{e'_\text{location[i]}}^2) +\\
+  (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_i, \sigma_{e_\text{location[i]}}^2)\\
+		\text{where}\\
+		\delta \sim \text{N}(0,1)\\
+		\text{constraint: } \delta > 0
 $$
@@
 
-The Stan model can be loaded into R using the `stan_model` function and assign it to the object `molg`.
+The Stan model can be loaded and translated into an S4 class object using the `stan_model` function and assign it to the object `molg`.
 The "molg.stan" file needs to be available in the directory "stan" located inside the current working directory.
@@
           "prob",                        # mixing proportion as probability
-          "theta_s",                     # by-participant disfluency probability (logit scale)
           "prob_s",                      # by-participant disfluency probability
@@
                      seed = 365, 
+                     # Improve sampling efficiency
                      control = list(adapt_delta = 0.96,  # default: 0.9
@@
 
-The function `fixef` provide a summary of the inferred posterior estimates as average and 95% probability interval by transition location in log msecs.
-
-```{r}
-print(fit_molg, pars = c("beta", "delta", "theta"), probs = c(.025, .975))
-```
-
 Running this model will take a while to complete sampling depending on your hardware specifications.
@@
 
+Inferred posterior estimates shown as average and 95% probability interval can be extracted using the `print` function for each transition location and the three parameter values of interest.
+
+```{r}
+print(fit_molg, pars = c("beta", "delta", "prob"), probs = c(.025, .975))
+```
+
+## Model comparsion
+
+Similar to above we will compare the three models -- the mixture of log-Gaussians, and the uni-model Gaussian and log-Gaussian models -- using leave-one-out cross-validation.
+Because the mixture model was fitted directly in Stan and the uni-model models were fitted in `brms`, we need to extract the fit statistics before comparing the models using functions provided by the package `loo`.
+
+```{r}
+library(loo)
+# Mixture of log-Gaussians
+log_lik <- extract_log_lik(fit_molg, merge_chains = F) 
+r_eff <- relative_eff(exp(log_lik)) # relative effective sample size
+loo_molg <- loo(log_lik, r_eff = r_eff)
+
+# Uni-modal log-Gaussian
+loo_lgaus <- loo(fit_lgaus)
+
+# Uni-modal Gaussian
+loo_gaus <- loo(fit_gaus)
+```
+
+As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to each the uni-modal Gaussian and log-Gaussian model.
+Their difference is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
+Models are ordered from the model with the highest predictive performance to the lowest (i.e.
+`elpd_loo`).
+
+```{r}
+loo_comp <- loo_compare(loo_gaus, loo_lgaus, loo_molg)
+```
+
+To extract the model comparisons and fit statistics:
+
+```{r}
+loo_comp %>% 
+  as.data.frame() %>% 
+  rownames_to_column("models") %>% 
+  mutate(across(models, recode, model3 = "fit_molg"),
+         across(where(is.numeric), round, 1))  
+```
+
+The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian ) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` times better than `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`).
+
 ## Model diagnostics
@@
 First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain.
-We will look at the population-level parameters (names are determined by the model) corresponding to the Greek symbols used **in the manuscript** and assign those to `pars`.
+We will look at the population-level parameters (names are determined in the Stan code) and assign those to `pars`.
 
 ```{r}
-pars <- c("beta", "delta", "theta", "sigma", "sigmap_e", "sigma_e")
+pars <- c("beta", "delta", "theta")
 ```
@@
 # Check convergence
-stan_trace(m, pars = pars, alpha = .5)
+stan_trace(fit_molg, pars = pars, alpha = .5)
 ```
@@
 ```{r}
-summary(m, pars=pars)$summary %>% as.data.frame() %>% 
+summary(fit_molg, pars=pars)$summary %>% 
+  as.data.frame() %>% 
   rownames_to_column("parameter") %>% 
@@
 
-The model used the posterior parameter values to simulate hypothetical data sets. This happened for every iteration for every chains.
+```{r echo = F, include = F}
+(total_samples <- (iterations - warmup) * nchains)
+```
 
-```{r echo = F, include=F}
-(total_samples <- (iterations - warmup) * n_chains)
+The model used the posterior parameter values to simulate hypothetical data sets.
+This happened for every iteration for every chains.
+In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) available to us because
+
+```{r}
+(total_samples <- (iterations - warmup) * nchains)
 ```
@@
 ```{r}
-y_tilde <- as.matrix(m, pars = "y_tilde")          # extract simulated data sets
+y_tilde <- as.matrix(fit_molg, pars = "y_tilde")   # extract simulated data sets
 N <- 50                                            # number of simulations to use
-total_samples <- (iterations - warmup) * n_chains  # total number of samples
 rnd_sims <- sample(total_samples, N)               # created random indices
@@
 
-```{r }
-(total_samples <- (iterations - warmup) * n_chains)
-```
-
 The distribution of parameter values represents the uncertainty about parameter values given the data.
@@
 The `names` function can be used to remind us of the parameter names used in the model (we reduced the output to the first 15 parameter names).
-The meaning of the parameters is described **in the manuscript** and can be obtained from the Stan code.
-Indices refer to the copy-task component (1 = consonants; 2 = LF bigrams) and to the participant identifier for by-participant parameters (indicated by `_s`).
+The meaning of the parameters is described above and is determined in the Stan code.
+Indices refer to the transition locations (1 = before-sentence; 2 = before-word; 3 = within-word) and to the participant identifier for by-participant parameters (indicated by `_s`).
 
 ```{r}
-names(m)[1:15] 
+names(fit_molg)[!str_detect(names(fit_molg), "log_lik|y_tilde")]
 ```
@@
 ```{r}
-stan_hist(m, pars = pars)
+stan_hist(fit_molg, pars = pars)
 ```
@@
 ```{r}
-print(m, pars = pars, probs = c(.025,.975))
+print(fit_molg, pars = pars, probs = c(.025,.975))
 ```
@@
 
-The `plot` function shows the posterior probability distribution of the three parameters for the consonants task (indicated as 1) and the LF-bigrams task (indicated as 2) summarised as median and 95% PI.
+The `plot` function shows the posterior probability distribution of the three parameters for before-sentence transitions (indicated as 1), before-word transitions (indicated as 2), and within-word transitions summarised as median and 95% PI.
 
 ```{r}
-plot(m, pars = pars, ci_level = .95) # ci = credible interval
+plot(fit_molg, pars = pars, ci_level = .95) # ci = credible interval
 ```
@@
 ```{r}
-posterior <- as.data.frame(m, pars) %>% as_tibble() 
+posterior <- as.data.frame(fit_molg, pars) %>% as_tibble() 
 posterior
@@
 posterior_long <- pivot_longer(posterior, everything(), 
-             names_to = c(".value", "component"), 
+             names_to = c(".value", "location"),
              names_pattern = "(.*)\\[(.)\\]") 
-
 posterior_long
@@
 In order to transform the slowdown `delta` into msecs we need to add `beta` before using the exponential function; we can then subtract beta again.
-The `recode` function changes the component indices from 1 to "Consonants" and 2 to "LF bigrams".
+`theta` samples are currently on the logit scale and can be transformed to probabilities using `brms::inv_logit_scaled`.
+The `recode` function changes the component indices from 1 to "before-sentence", 2 to "before-word", 3 to "within-word".
 
@@
                              delta = exp(beta + delta) - exp(beta),
-                             beta = exp(beta),
-                             component = recode(component, `1` = "Consonants",
-                                                           `2` = "LF bigrams"))
+                             across(beta, exp),
+                             across(theta, inv_logit_scaled),
+                             across(location, recode, `1` = "before-sentence",
+                                                      `2` = "before-word",
+                                                      `3` = "within-word"))
 posterior_in_msecs
@@
 pivot_longer(posterior_in_msecs, beta:theta, names_to = "parameter") %>%
-  ggplot(aes(x = value, colour = component, fill = component)) +
+  ggplot(aes(x = value, colour = location, fill = location)) +
   geom_histogram(position = "identity", alpha = .25) +
   facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
-  scale_fill_brewer("Copy-task\ncomponents", palette = "Dark2") +
-  scale_color_brewer("Copy-task\ncomponents", palette = "Dark2") +
+  scale_fill_brewer("Transition\nlocation", palette = "Dark2") +
+  scale_color_brewer("Transition\nlocation", palette = "Dark2") +
   labs(x = "Posterior parameter estimate")

### 2023-02-08 tutorial/tutorial.Rmd
Update tutorial.rmd which can not be compiled

@@
 ---
-title: "Tutorial: fitting mixed mixture models on text writing data"
+title: "Tutorial: fitting mixed-effects and log-Gaussian mixture models on text-writing data"
 author: "Jens Roeser"
@@
 ```{r setup, include=FALSE}
-options(scipen = 999) # Turn off scientific notation
+# Turn off scientific notation
+options(scipen = 999)
+
+# Chunk defaults
 knitr::opts_knit$set(width=90)
@@
                       fig.height = 4.5,
+                      size = "small",
                       echo = TRUE,
@@
                       message = FALSE)
-theme_set(theme_bw() + theme(legend.justification = "top"))
+
+# Plotting defaults
+theme_set(theme_bw() + 
+            theme(legend.justification = "top",
+                  panel.grid = element_blank()))
 ```
@@
 ```{r}
-iterations <- 30000       # No of iterations
-warmup <- iterations/2    # Warmup samples to be discarded
+iterations <- 10000      # No of iterations
+warmup <- iterations / 2 # Warmup samples to be discarded
 nchains <- ncores <- 3   # No of chains / cores used (one chain per core)
@@
 
-## Model comparison
+## TODO Model comparison
+
+ADD here info from slides on how to calculate elpd
 
@@
 
-## Mixture model app
+## TODO Mixture model simulation
+
+## Model fit
 
@@
 upper) are.
-The full Stan code can be found at the end of this document.
+The full Stan code can be found at the end of this tutorial.
 
@@
                      # Improve sampling efficiency
-                     control = list(adapt_delta = 0.96,  # default: 0.9
-                                    max_treedepth = 14)) # default: 10
+                     control = list(adapt_delta = 0.94,  # default: 0.9
+                                    max_treedepth = 12)) # default: 10
 ```
 
-```{r echo =F}
+```{r echo = F}
 #sampler_params <- get_sampler_params(m, inc_warmup = FALSE)
@@
 ```{r}
-print(fit_molg, pars = c("beta", "delta", "prob"), probs = c(.025, .975))
+print(fit_molg, 
+      pars = c("beta", "delta", "prob"), 
+      probs = c(.025, .975))
 ```
@@
 ```{r}
-loo_comp %>% 
-  as.data.frame() %>% 
+loo_comp %>% as.data.frame() %>% 
   rownames_to_column("models") %>% 
@@
   rownames_to_column("parameter") %>% 
+  mutate(across(Rhat, round, 3)) %>% 
   select(parameter, Rhat)
@@
 
-## Difference between copy-task components
+## Difference between transition locations
 
-We can calculate the differences between the copy-task components for each parameter value.
-This is giving an indication of whether fluent typing ($\beta$) is influenced by the copy task, maybe the size of the disfluency slowdown ($\delta$) is impacted by the task, or the disfluency probability ($\theta$).
+We can calculate the differences between the transition locations for each parameter.
+This is giving an indication of whether fluent typing ($\beta$) is influenced by linguistic edges, maybe the size of the disfluency slowdown ($\delta$), or the disfluency probability ($\theta$).
 
-To determine these differences we change the data format above and create a variable that indicates the parameter and one column for the corresponding values for each copy-task component.
-The last line calculates the difference between the consonants task and the LF-bigrams task for each of the three parameter values.
+To determine these differences we change the data format above and create a variable that indicates the parameter and one column for the corresponding values for each transition location.
+The last line calculates the difference between (1) transitions before sentences and words, and (2) transitions before and within words for each of the three parameters.
 
 ```{r}
-posterior_by_component <- posterior_in_msecs %>%
+posterior_diffs <- posterior_in_msecs %>%
   pivot_longer(beta:theta, names_to = "parameter") %>%
-  group_by(component) %>% mutate(id = row_number()) %>%
-  pivot_wider(names_from = component, values_from = value) %>%
+  group_by(location) %>% 
+  mutate(id = row_number()) %>%
+  pivot_wider(names_from = location, values_from = value) %>%
   select(-id) %>% # drop id column
-  mutate(diff = Consonants - `LF bigrams`,  # calculate the difference between copy-task components
-         parameter = paste0(parameter, "[diff]"))
+  mutate(diff_sent = `before-sentence` - `before-word`,
+         diff_word = `before-word` - `within-word`,
+         parameter = paste0(parameter, "[diff]")) %>% 
+  select(parameter, starts_with("diff"))
 
-posterior_by_component
+posterior_diffs
 ```
@@
 ```{r}
-posterior_by_component %>%
-  group_by(parameter) %>%                      # group by parameters
-  summarise(mean = mean(diff),                 # mean difference
-            lower = quantile(diff, .025),      # 2.5% lower bound of difference
-            upper = quantile(diff, .975),      # 97.5% upper bound of difference
-            p = mean(diff < 0)) %>%            # prob that difference is negative
-  mutate(across(where(is.numeric), round, 2))  # round after 2nd decimal place
+# Create some summary functions
+lower <- function(x) quantile(x, .025)
+upper <- function(x) quantile(x, .975)
+p_diff <- function(x) mean(x < 0)
+
+# Summarise differences
+posterior_diffs %>%
+  pivot_longer(starts_with("diff")) %>% 
+  group_by(parameter, name) %>%      # group by parameters, comparison
+  summarise(across(value, 
+            list(mean = base::mean,  # mean difference
+                 lower = lower,      # 2.5% lower bound of difference
+                 upper = upper,      # 97.5% upper bound of difference
+                 p_diff = p_diff),   # prob that difference is negative
+            .names = "{.fn}")) %>% 
+  ungroup() %>% 
+  mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
+  arrange(name)
 ```
@@
 ```{r echo=F}
-ps <- posterior_by_component %>%
-  group_by(parameter) %>%
-  summarise(mean = mean(diff),
-            lower = quantile(diff, .025),
-            upper = quantile(diff, .975),
-            p = mean(diff < 0)) %>% 
+ps <- posterior_diffs %>%
+  pivot_longer(-parameter) %>% 
+  group_by(parameter, name) %>%
+  summarise(across(value, 
+            list(mean = base::mean,  
+                 lower = lower,      
+                 upper = upper,      
+                 p_diff = p_diff),   
+            .names = "{.fn}")) %>% 
+  ungroup() %>% 
   mutate(across(where(is.numeric), round, 2),
@@
 The area of the histogram to the left of this line corresponds to the probability `p` to observe a value smaller than zero as in the output above.
-We observe difference between copy-task components for all three parameters.
-Fluent typing is `r round(ps[ps$parameter == "beta",]$mean)` msecs (95% PI: `r round(ps[ps$parameter == "beta",]$lower)`, `r round(ps[ps$parameter == "beta",]$upper)`) slower for the consonants task, and hesitations are `r round(ps[ps$parameter == "delta",]$mean)` msecs (95% PI: `r round(ps[ps$parameter == "delta",]$lower)`, `r round(ps[ps$parameter == "delta",]$upper)`) longer and `r ps[ps$parameter == "theta",]$mean` ($\approx$ `r round(ps[ps$parameter == "theta",]$mean*100)`%; 95% PI: `r ps[ps$parameter == "theta",]$lower`, `r ps[ps$parameter == "theta",]$upper`) times more likely.
-Interestingly fluent typing (i.e.
-planning and executing motor plans) is slower in the consonant task even after modeling hesitations as a separate process.
 
 ```{r}
-ggplot(posterior_by_component, aes(x = diff)) +
-  geom_histogram(alpha = .5) +
+posterior_diffs %>% 
+  pivot_longer(starts_with("diff")) %>% 
+  mutate(across(name, recode, diff_sent = "before-sentence vs before-word",
+                              diff_word = "before-word vs within word")) %>% 
+  ggplot(aes(x = value, colour = name, fill = name)) +
+  geom_histogram(position = "identity", alpha = .25) +
+  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
+  scale_fill_brewer("Difference", palette = "Dark2") +
+  scale_color_brewer("Difference", palette = "Dark2") +
   geom_vline(xintercept = 0, colour = "red", linetype = "dashed", size = .5) +
-  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
-  labs(x = "Difference between copy-task components")
+  labs(x = "Difference between transition locations") +
+  theme(legend.position = "top")
 ```
 
-In practice we rarely believe that only an effect of exactly null is indicating that there is no difference between our groups.
+We observe the following differences that can be written up in a report like this: We observe evidence for a higher probability to pause before sentences which is `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(upper)`) times more likely than before words, which was is `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(upper)`) more likely than within words.
+
+Similarly, differences were found for the hesitation difference.
+Hesitations before sentences are `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(upper)`) longer than before words, which were `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(upper)`) longer than within words.
+
+Interestingly though fluent transitions (i.e.
+planning and executing motor plans) were only faster within words by `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper)`) compared to before-word transitions but there was no difference for before-sentence transitions compared to before-word transitions ($\hat{\beta}$ = `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean)` msecs, 95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper)`).
+
+## ROPE
+
+In practice, we rarely believe that only an effect of exactly null is indicating that there is no difference between our groups (which is what a null hypothesis assumes).
 Instead there are values that are non-different from null; for example a slowdown of 1 msecs is meaningless in most, if not all, contexts.
@@
 
-To calculate the ROPE for out differences between the two copy-task components we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference.
-We can calculate a vector with posterior difference for each of the three parameters, indicated as $\beta_\text{diff}$, $\delta_\text{diff}$, and $\theta_\text{diff}$.
+To calculate the ROPE for out differences between two transition locations, we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference.
+We can create such a vector from our posterior differences.
+Let's focus on $\beta_\text{diff}$, the average typing speed for fluent transitions.
 
 ```{r}
-# Fluent typing
-beta_diff <- filter(posterior_by_component, parameter == "beta[diff]") %>% pull(diff)
+# Extract vector of differences
+# For sentence vs word
+beta_diff_sent <- filter(posterior_diffs, parameter == "beta[diff]") %>% pull(diff_sent)
 
-# Slowdown for hesitant typing
-delta_diff <- filter(posterior_by_component, parameter == "delta[diff]") %>% pull(diff)
-
-# Proportion of hesitations
-theta_diff <- filter(posterior_by_component, parameter == "theta[diff]") %>% pull(diff)
+# For before word vs within word
+beta_diff_word <- filter(posterior_diffs, parameter == "beta[diff]") %>% pull(diff_word)
 ```
@@
 library(bayestestR)
-rope(beta_diff, range = c(-10, 10))
+rope(beta_diff_sent, range = c(-10, 10))
 ```
 
-The ROPE for fluent typing speed contains `r round(pull(rope(beta_diff, range = c(-10, 10)))*100,2)`% of the posterior samples.
-In other words, there is a `r round(pull(rope(beta_diff, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.
-
-For disfluencies we might want to consider a larger range of values as negligible differences between two tasks (also because the variability in disfluencies is larger than the variability for fluent key transitions).
-Even if we define a ROPE of -250 through 250 msecs, there are no posterior samples inside the ROPE indicating that the difference in typing hesitations are related to difficulty that arise prior to processing on the motor level.
+The ROPE for fluent typing speed for the difference between before-sentence and before-word transitions contains `r round(pull(rope(beta_diff_sent, range = c(-10, 10)))*100,2)`% of the posterior samples.
+In other words, there is a `r round(pull(rope(beta_diff_sent, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.
 
 ```{r}
-rope(delta_diff, range = c(-250, 250))
+rope(beta_diff_word, range = c(-10, 10))
 ```
 
-As for the disfluency probably, we need to bear in mind that estimates for $\theta$ are bound between 0 and 1.
-Hence, the difference $\theta_\text{diff}$ cannot ever exceed -1 and 1.
-If we consider a difference of 5% or smaller as negligible, we can define the ROPE as ranging from -.05 through .05.
-We observe that `r round(pull(rope(theta_diff, range = c(-.05, .05)))*100,2)`% of the posterior difference $\theta_\text{diff}$ is within the ROPE.
-In other words, a large proportion of posterior samples is supporting a meaningful difference between the two copy-task components.
-We observe that hesitations are largely more frequent in the consonants task.
+However, the ROPE for fluent typing speed for the difference between before-word and within-word transitions contains `r round(pull(rope(beta_diff_word, range = c(-10, 10)))*100,2)`% of the posterior samples.
+In other words, there is a `r round(pull(rope(beta_diff_word, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.
+
+Let's see what happens if we determine a wider ROPE for negligible differences.
+What happens if we define a ROPE of -50 through 50 msecs for the difference between before-word and within-word transitions.
 
 ```{r}
-rope(theta_diff, range = c(-.05, .05))
+rope(beta_diff_word, range = c(-50, 50))
 ```
+
+The ROPE contains `r round(pull(rope(beta_diff_word, range = c(-50, 50)))*100,2)`% of the posterior samples.
+In other words, there is a `r round(pull(rope(beta_diff_word, range = c(-50, 50)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -50 and 50 msecs which we considered negligible.
+
+## TODO Bayes Factor
+
+Add info from Savage Dickey Bayes Factor slide
+
+## Note on using more than a single IV
 
@@
 
-In some contexts we would like to obtain by-participant estimates of the average fluent typing-speed of a participant after accounting for disfluencies (i.e.
-$\beta_s$); by-participant disfluency information can be estimated the proportion of disfluencies (i.e.
+In some contexts we would like to obtain by-participant estimates of the probability of a participant to exhibit a disfluency (i.e.
 $\theta_s$).
-These estimates inform us about the fluent typing speed but also about the prevalence of typing hesitations.
+These estimates inform us about the prevalence of typing hesitations.
 
-As before we can extract the posterior from the model `m`.
-This time we use the parameters that stored by-participant estimates `beta_s` and `theta_s` corresponding to the population estimates `beta` and `theta`.
+As before we can extract the posterior from the model `fit_molg`.
+This time we use the parameter that stored by-participant estimates `prob_s` (which is the inverse-logit of the `theta_s` which we didn't store in the model fit) corresponding to the population estimate `theta`.
 
 ```{r}
-pars <- c("beta_s", "theta_s")
-posterior_ppts <- as.data.frame(m, pars = pars) %>% as_tibble()
+posterior_ppts <- as.data.frame(fit_molg, pars = 'prob_s') %>% as_tibble()
 names(posterior_ppts)[1:10]
@@
 
-This posterior tibble has the format "$<\text{parameter name}>\text{_s}[<\text{component id}>, <\text{participant id}>]$" where indices indicate the copy-task component and the participant identifier of the parameters.
-The following code is converting the `beta`s to msecs, creates a long format with columns `parameter` with the names above as levels and a `value` column with their respective values.
-The `parameter` column is then separated into `parameter` (levels: `beta`, `theta`), `component` (levels: `1`, `2`), and participant (an index for each participant) using "," to separate the three variables.
-To use "," as separator we replace "_s" with "," in the line before.
-Copy-task component names were changed as before.
-A preview is below.
+This posterior tibble has the format
+
+| "$<\text{parameter name}>\text{_s}[<\text{transition location id}>, <\text{participant id}>]$"
+
+where indices indicate the transition-location id and the participant identifier of the parameters.
+The following code creates a long format with columns `name` with the names above as levels and a `value` column with their respective values (both `name` and `value` are `pivot_longer` defaults).
+The `name` column is then separated into `name` (levels: `prob`), `location` (levels: `1`, `2`, `3`), and participant (an index for each participant) using "," to separate the three variables.
 
@@
 posterior_ppts_long <- posterior_ppts %>%
-  # use exp() on everything starting with "beta"
-  mutate(across(starts_with("beta"), exp)) %>%             
-  pivot_longer(everything(), names_to = "parameter") %>%
-  # replace "_s" with "," 
-  mutate(parameter = gsub("_s", ",", parameter)) %>%
-  # separate into parameter, component, ppt using "," as separator
-  separate(parameter, into = c("parameter", "component", "participant")) %>% 
-  mutate(component = recode(component, `1` = "Consonants", `2` = "LF bigrams")) %>%
-  group_by(parameter) %>% mutate(id = row_number()) %>%
-  pivot_wider(names_from = parameter, values_from = value) %>% 
+  pivot_longer(everything()) %>%
+  # separate into location and ppt using "," as separator
+  separate(name, into = c("location", "participant"), sep = ",") %>% 
+  # extract number from new variables and recode location
+  mutate(across(c(location, participant), parse_number),
+         across(location, recode, 
+                `1` = "before-sentence",
+                `2` = "before-word",
+                `3` = "within-word")) %>%
+  # create different columns for location
+  group_by(location) %>% 
+  mutate(id = row_number()) %>%
+  pivot_wider(names_from = location, values_from = value) %>% 
   select(-id) # drop id
@@
 
-We can then summarise by-participant estimates for each component and parameter with the most probable parameter value and the 95% PI as before.
+We can then summarise by-participant estimates for each transition location with the most probable parameter value and the 95% PI as before.
 
@@
 posterior_ppts_long_summary <- posterior_ppts_long %>%
-  pivot_longer(beta:theta, names_to = "parameter", values_to = "value") %>%
-  group_by(parameter, component, participant) %>%
-  summarise(mean = mean(value),
-            lower = quantile(value, .025),
-            upper = quantile(value, .975))
+  pivot_longer(-participant, names_to = "location") %>%
+  group_by(location, participant) %>%
+  summarise(across(value, list(mean = base::mean, 
+                               lower = lower, 
+                               upper = upper), .names = '{.fn}')) %>% 
+  ungroup()
 
@@
 
-To get an idea how the by-participant estimates look like, we extracted participants 1 through 5 for visualisation.
-From this subset we see that, although we observed overall larger population estimates for the consonants task, some participants do not show a difference in fluent typing speed and a similar frequency of hesitations (e.g.
-participant 2) or even slightly more hesitations in the LF-bigrams task (e.g.
-participant 1).
+For all 10 participants we see the estimate probability of typing hesitations by transition location.
+We observe in general that hesitations are more frequent before sentences and words than within words.
+However, participants vary to the extend they hesitate before words and sentences suggesting that some participant plan sentences more globally and others plan sentences word-by-word (incrementally).
 
-```{r }
+```{r}
 posterior_ppts_long_summary %>%
-  filter(participant %in% 1:5) %>% # only participants 1 through 5
   ggplot(aes(x = participant, 
              y = mean, ymin = lower, ymax = upper,
-            colour = component)) +
+            colour = location,
+            shape = location)) +
   geom_pointrange(position = position_dodge(.75)) +
-  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
-  coord_flip() +
-  labs(x = "Participant id", y = "Parameter estimates") +
-  scale_colour_brewer("Copy-task\ncomponent", palette = "Dark2")
+  labs(x = "Participant id", 
+       y = "Parameter estimates with 95% PIs") +
+  scale_colour_brewer("Transition\nlocation", palette = "Dark2") +
+  scale_shape_manual("Transition\nlocation", values = c(21, 3, 8)) +
+  scale_x_continuous(breaks = 1:10) +
+  theme(legend.position = 'top')
 ```
@@
 ```{r}
-posterior_ppts_long_diff_summary <- posterior_ppts_long %>% group_by(component) %>% 
-  mutate(id = row_number()) %>%
-  pivot_wider(names_from = component, 
-              values_from = c(beta, theta)) %>%
-  pivot_longer(-c(participant, id),
-             names_to = c("parameter", ".value"), 
-             names_pattern = "(.*)_(.*)") %>%
-  mutate(diff = Consonants - `LF bigrams`,
-         parameter = paste0(parameter, "[diff]")) %>% 
-  select(participant, parameter, diff) %>%
-  group_by(parameter, participant) %>%
-  summarise(mean = mean(diff),
-            lower = quantile(diff, .025),
-            upper = quantile(diff, .975))
+posterior_ppts_long_diff_summary <- posterior_ppts_long %>% 
+  mutate(diff_sent = `before-sentence` - `before-word`,
+         diff_word = `before-word` - `within-word`) %>% 
+  select(participant, starts_with('diff')) %>%
+  pivot_longer(-participant, 
+               names_to = 'difference', 
+               values_to = 'diff') %>% 
+  group_by(participant, difference) %>%
+  summarise(across(diff, list(mean = base::mean, 
+                               lower = lower, 
+                               upper = upper), .names = '{.fn}')) %>% 
+  ungroup()
 
@@
 
-For simplicity we show the differences for the first 10 participants only.
-This calculation and visualisation allows researchers to identify participants that do or do not differ in terms of fluent typing speed or hesitation frequency, in our example for tasks that do and do not involve lexical information.
+This calculation and visualisation allows researchers to identify participants that do or do not differ in terms of hesitation frequency, in our example depending on transition locations.
 
@@
 posterior_ppts_long_diff_summary %>%
-  filter(participant %in% 1:10) %>% # only participants 1 through 10
+  mutate(across(difference, recode, 
+                diff_sent = "Comparison: before-sentence vs before-word",
+                diff_word = "Comparison: before-word vs within-word")) %>% 
   ggplot(aes(x = factor(as.numeric(participant)), 
@@
   geom_pointrange(position = position_dodge(.75)) +
-  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
+  facet_wrap(~difference) +
   coord_flip() +
-  labs(x = "Participant id", y = "Difference between copy-task components") +
-  scale_colour_brewer("Copy-task\ncomponent", palette = "Dark2")
+  labs(x = "Participant id", 
+       y = "Difference between transition locations") 
 ```
@@
 ```{r echo = F}
-cat(readLines("MoG.stan"), sep = "\n")
+cat(readLines("stan/molg.stan"), sep = "\n")
 ```

### 2023-05-05 report/results.Rmd
Add tutorial and draft results

@@
 library(scales)
-#library(caret)
-#library(InformationValue)
-
-# Calculate Bayes Factor
-BF <- function(ps, prior_sd = 1){
-  fit_posterior <- logspline(ps) 
-  posterior <- dlogspline(0, fit_posterior) # Height of the posterior at 0 
-  prior <- dnorm(0, 0, prior_sd) # Height of the prior at 0
-  BF10 <- prior / posterior 
-  return(BF10)
-}
-
-logit <- function(p) log(p / (1-p))
-ilogit <- function(x) 1 / (1 + exp(-x))
-inv_logit <- function(logit) exp(logit) / (1 + exp(logit))
-
-se_bin <- function(x) sqrt((mean(x, na.rm = T)*(1 - mean(x, na.rm = T)))/length(x)) # se for binary data
-
-lower <- function(x) quantile(x, prob = .025)
-upper <- function(x) quantile(x, prob = .975)
-
-# Remove leading zeros and round numbers
-dezero <- function(x, dp){ # dp is decimal places
-  fmt = paste0('%.',dp,'f')
-  x = sprintf(fmt,x)
-  x = str_replace(x, '(-|^)0\\.','\\1\\.')
-  return(x)
-}
-dezero_plot <- function(x) dezero(x, 1)
+library(ggthemes)
+source("../scripts/functions.R")
+source("scripts/get_pdens_plot.R")
 
@@
 
+# Introduction
+
+```{r eval = F}
+Take intro from Mark's draft of SPL2 paper
+
+Different views on pausing in text production, largely representing typing as a unimodal or bimodal process
+
+- pauses are represented in the right tail of the distribution (unimodal) explaining the long positive skew and therefore possibly making a log transformation redundant (there are people arguing against log transformation of keystrokes but I don't understand their argument)
+- as pauses are more likely at linguistic edges, the variance of the distribution is going to be larger
+- pauses are represented as an independent distribution / process
+  - short distribution: fluent key transitions only limited by the ability to move fingers (same across text locations)
+  - short distribution: processes involved in cascade are different when lexical and syntactic planning is involved (differences across text locations)
+
+```
+
+# Model definitions
+
+The models presented in the following can be divided into two general groups.
+The first three models are largely akin to models typically used in the literature.
+By this we mean models that assume a uni-modal process that generates keystroke data as is incorporated in statistical models such as analysis of variance and linear mixed-effects models.
+Second, the last two models model keystroke intervals as a combination of two weighted processes of which one presents a smooth information flow from mind into finger; the other component is more important as it represents moments at which the information flow was interrupted leading to longer latencies.
+The latter two models map directly on the idea of a cascading model of writing.
+
+## Uni-modal Gaussian
+
+We start with a Gaussian mixed-effects models similar to a standard analysis of variance.
+We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
+The mean can be decomposed into $\beta$ and $\text{participant}_i$.
+$\beta$ will be allowed to take on a different value for each transition location.
+Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.
+
+$$
+\begin{align}
+\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
+\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
+& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
+\end{align}
+$$
+
+## Uni-modal log-Gaussian
+
+This model is largely identical to the previous model, except we use a log-normal distribution instead of a normal distribution.
+The advantage of using a log-normal distribution is two-fold: (1) the log-normal distribution has a lower bound of zero.
+For our data we generally consider negative ikis as mistakes which occasionally occur when the next key was pressed before the current key.
+Other than that keystroke intervals are constrained by a persons ability to move their fingers and keyboard polling.
+(2) the log-scale is known to be a better match for data from human behaviour, in particular motor responses.
+In particular, a normal distribution assumes that units are linearly scaled.
+For example, a 50 msecs difference is the same between 100 msecs and 150 msecs than it is between 5 secs and 5 secs 50 msecs (i.e.
+5,050 msecs).
+This is not necessarily plausible for keystroke data.
+We would assume that differences that are due to motor activity (typing a high frequency bigram vs a low frequency bigram) are smaller than difference that are due to high levels of cognitive activity (retrieving a word in your L1 or L2).
+Log-normal distributions are a natural way of scaling units so that a 50 msecs difference on the lower end of the iki scale (motor activity) is more meaningful than a 50 msecs difference on the upper end of the iki scale (retrieving words, planning sentences).
+
+The model can be described like this:
+
+$$
+\begin{align}
+\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
+\text{where: } &
+\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
+& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
+\end{align}
+$$
+
+## Uni-modal unequal-variance log-Gaussian
+
+$$
+\begin{align}
+\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
+\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
+ & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
+\end{align}
+$$
+
+## Bi-modal log-Gaussian (constrained)
+
+This models extends the intuition from the previous model that higher levels of activation lead to longer pauses.
+Instead of assuming that there is one process that underlies the generation of ikis, we assume there are two.
+(1) activation can flow into keystrokes without interrupts.
+These fluent keystroke transitions are merely limited by a person's ability to move their finger and will be captured but the parameter $\beta$.
+In principle, there are no differences for fluent key-transitions between transition location.
+The next model with loosen this assumption (hence, "unconstrained model").
+(2) difficulty in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when spelling, words, or contents couldn't be retrieved in time.
+The size of these pauses will depend on the reason for delays which is typically associated with transition locations (contents are typically planned before sentences, words are retrieved before they are typed and spelling difficulty typically occurs when typing a word).
+Pauses will be captured by two model parameters: (1) the slowdown for these hesitant transitions will be captured by $\delta$ which is the deviation compared to normal typing intervals (constrained to be positive).
+(2) the frequency of hesitant transitions will be captured by $\theta$ for each level of a categorical predictor.
+
+$$
+\begin{align}
+\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e}^2)\\
+\text{where: }
+		&\delta \sim \text{N}(0,1)\\
+\text{constraint: } & \delta > 0\\
+		& \sigma_{e'}^2 > \sigma_{e}^2
+\end{align}
+$$
+This models takes into account two sources of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
+
+## Bi-modal log-Gaussian (unconstrained)
+
+This model is identical to the previous model with one exception.
+The distribution of fluent keystroke transitions captured by $\beta$ was fixed to be the same across transition locations is the previous model.
+In other words, the mean $\beta$ and it's standard deviation $\sigma_{e}^2$ was the same for before-sentence, before-word, and within-word transitions.
+This means, because of the naturally larger number of within-word transitions, the posterior is dominated by within-word transitions.
+
+In this model we will loosen this constraint and allow $\beta$ and $\sigma_{e}^2$ to vary by transition-location.
+
+$$
+\begin{align}
+\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
+	\text{where: } & \delta \sim \text{N}(0,1)\\
+\text{constraint: } & \delta > 0\\
+		& \sigma_{e'}^2 > \sigma_{e}^2
+\end{align}
+$$
+
 # Analysis
@@
 The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 10,000 iterations on 3 chains with a warm-up of 5,000 iterations and no thinning.
+Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
 Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
+
+# Datasets in overview
+
+Five dataset with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).
+
+```{r datasets, results='asis'}
+tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT"),
+                Source = c("@ronneberg2022process",
+                              "@torrance2016adolescent",
+                              "@torrance",
+                              "@rossetti2022s",
+                              "@nina_vandermeulen_2020_3893538"),
+                Keylogger = c("EyeWrite", 
+                           "EyeWrite", 
+                           "CyWrite", 
+                           "InputLog", 
+                           "InputLog"),
+                `Writing task` = c(
+                                 "Argumentative essays", 
+                                 "Expository texts",
+                                 "Argumentative essays",
+                                 "Text simplification",
+                                 "Synthesis"),
+                n_ppt = c(126, 26*2, 39, 47, 658),
+#                n_texts = c(1, 2, 2, 2, NA),
+#                n_sentences = c(),
+#                n_words = c(),
+                conditions = c(" ", 
+                               "weak decoders / control; masked / unmasked",
+                               "write in L1 / L2",
+                               "pre / post test trained in plain language principles and control", 
+                               "Various topics and genres"),
+                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95),
+                 Language = c( "Norwegian", 
+                               "Norwegian", 
+                               "English (L1) / Spanish (L2)", 
+                               "English (L2)", 
+                               "Dutch")) %>% 
+  kable(caption = "Datasets in brief.") %>%
+  kable_styling(position = "center", 
+                full_width = T, 
+                font_size = 12)
+
+```
+
+```{r eval = F}
+Data reduction using random samples of subsets of participants for larger data sets
+
+Also for larger data sets, only used participants that provided a min number of sentences
+
+Locations with example from Marks
+
+Removed transitions that were followed by an editing operation
+
+Extremely short and long intervals
+
+Add a description with general data filtering and a table with how many data were removed from the analysis by study.
+
+```
+
+# Out-of-samples cross-validation
 
@@
 
-# Data processing
-
-TODO: Add a description with general data filtering and a table with how many data were removed from the analysis by study.
-
-# C2L1
-
-The C2L1 data set comprises data Norwegian 6th graders--*N*=133, mean age 11 years 10 months--published in @ronneberg2022process.
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-
-```{r eval = F}
-d <- read_csv("../data/c2l1.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000) %>%
-  mutate(ppt = as.numeric(factor(subno)),
-         condition = factor(str_c(location, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num)
-
-count(d, ppt) %>% 
-  filter(n < 100)
-pull(d, ppt) %>% unique() %>% length()
-```
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
-
-## Model comparisons
-
-### Fit to data
-
-```{r predictionsc2l1, fig.cap="C2L1 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-files <- list.files("../stanout/c2l1", pattern = "sims", full.names = T)
-sims <- read_csv(files) %>% 
-  pivot_longer(cols = c(y_obs, y_tilde)) %>% 
-  mutate(across(name, recode_factor, 
-                y_tilde = "Simulated data",
-                y_obs = "Observed data",
-                .ordered = T)) %>% 
-  filter(name ==  "Simulated data" |
-         (name == "Observed data" & sim_idx == 1))
-
-max_iki <- 2000
-
-plot_mog <- sims %>%
-  filter(model == "mog", value < max_iki) %>% 
-  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmm <- sims %>%
-  filter(model == "lmm", value < max_iki) %>% 
-  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmgaus <- sims %>%
-  filter(model == "lmmgaus", value < max_iki) %>% 
-  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmuneqvar <- sims %>%
-  filter(model == "lmmuneqvar", value < max_iki) %>% 
-  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
-  plot_layout(guides = "collect") +
-  plot_annotation(tag_levels = 'A')
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Out-of-samples cross-validation
+Results for all data sets are shown in Table \@ref(tab:loos).
+For all data sets we found the same pattern.
+The mixture of log-normal distributions provided a substantially better fit than uni-modal distribution models.
+The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across conditions.
 
 ```{r}
-files <- list.files("../stanout/c2l1", pattern = "^modelcomp", full.names = T)
-table <- purrr::map_dfr(files, read_csv) %>% 
-  mutate(across(location, replace_na, "overall")) %>% 
-  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
-  mutate(across(where(is.numeric), round, 0),
+files <- list.files(
+  str_c("../stanout/", 
+  c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
+  pattern = "model.+.csv", 
+  full.names = T)
+
+table <- purrr::map_dfr(files, ~read_csv(.) %>% 
+                          mutate(dataset = .x)) %>% 
+  select(dataset, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
+  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
+         across(where(is.numeric), round, 0),
          across(where(is.numeric), format, big.mark = ","),
          across(everything(), str_trim)) %>%
-  transmute(across(location, ~str_c("Text location: ",.)),
+  transmute(
+     across(dataset, recode, 
+                "cato" = "CATO",
+                "spl2" = "SPL2", 
+                "spl2_shift" = "SPL2 (shift + C)", 
+                "plantra" = "PLanTra",
+                "lift" = "LIFT",
+                "c2l1" = "C2L1"),
     across(model, recode, lmm = "Unimodal log-normal",
                           lmmgaus = "Unimodal normal",
-                                  mog = "Bimodal log-normal",
+                          mogbetaconstr = "Bimodal log-normal (constrained)",
+                          mogbetacontr = "Bimodal log-normal (constrained)",
+                          mogbetaunconstr = "Bimodal log-normal (unconstrained)",
                           lmmuneqvar = "Unimodal log-normal (unequal variance)"),
@@
   mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
-  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
-  select(Model = model, 
-         ends_with("overall"), 
-         ends_with("sent"), 
-         ends_with("beforeword"), 
-         ends_with("withinword"))
-
-```
-
-```{r c2l1, results='asis'}
-names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)
-
-kable(table, caption = "C2L1 data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 8))) %>%
+  pivot_wider(names_from = dataset, values_from = contains("elpd")) %>% 
+  select(Model = model, ends_with("CATO"), ends_with("C2L1"), ends_with("LIFT"), ends_with("PLanTra"), ends_with("SPL2"), ends_with("+ C)"))
+
+```
+
+```{r loos, results='asis'}
+names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 6)
+
+kable(table, 
+      caption = "Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+        align =c("l", rep("r", 12))) %>%
   kable_styling(position = "center", 
@@
   add_header_above(c(" " = 1, 
-                     "All transitions" = 2,
-                     "Before sentence" = 2, 
-                     "Before word" = 2,
-                     "Within word" = 2)) %>%
+                     "CATO" = 2,
+                     "CL21" = 2, 
+                     "LIFT" = 2,
+                     "PLanTra" = 2,
+                     "SPL2" = 2,
+                     "SPL2 (shift + C)" = 2)) %>%
   column_spec(1, width = "8em")
@@
 
-## Posterior parameter estimates of mixture model
-
-```{r c2l1post, fig.cap="C2l1. Posterior parameter distribution"}
-file <- list.files(path = "../stanout/c2l1", pattern = "mog.csv$",full.names = T)
-ps <- read_csv(file) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(across(delta, ~.+beta),
-         across(c(delta, beta), exp),
-         across(delta, ~.-beta),
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:prob) %>% 
-  group_by(location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"))
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "",
-       shape = "") +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-# CATO
-
-Data are published in @torrance2016adolescent. Norwegian upper
-secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-
-## Model comparisons
-
-### Fit to data
-
-```{r predictionscato, fig.cap="CATO data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-files <- list.files("../stanout/cato", pattern = "sims", full.names = T)
-sims <- read_csv(files) %>% 
-  pivot_longer(cols = c(y_obs, y_tilde)) %>% 
-  mutate(across(name, recode_factor, 
-                y_tilde = "Simulated data",
-                y_obs = "Observed data",
-                .ordered = T)) %>% 
-  filter(name ==  "Simulated data" |
-         (name == "Observed data" & sim_idx == 1))
-
-max_iki <- 2000
-
-plot_mog <- sims %>%
-  filter(model == "mog", value < max_iki) %>% 
-  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmm <- sims %>%
-  filter(model == "lmm", value < max_iki) %>% 
-  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmgaus <- sims %>%
-  filter(model == "lmmgaus", value < max_iki) %>% 
-  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmuneqvar <- sims %>%
-  filter(model == "lmmuneqvar", value < max_iki) %>% 
-  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
-  plot_layout(guides = "collect") +
-  plot_annotation(tag_levels = 'A')
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Out-of-samples cross-validation
-
-```{r}
-files <- list.files("../stanout/cato", pattern = "^modelcomp", full.names = T)
-table <- purrr::map_dfr(files, read_csv) %>% 
-  mutate(across(location, replace_na, "overall")) %>% 
-  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
-  mutate(across(where(is.numeric), round, 0),
-         across(where(is.numeric), format, big.mark = ","),
-         across(everything(), str_trim)) %>%
-  transmute(across(location, ~str_c("Text location: ",.)),
-            across(model, recode, lmm = "Unimodal log-normal",
-                                  lmmgaus = "Unimodal normal",
-                                  mog = "Bimodal log-normal",
-                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
-            across(elpd_diff, str_c, " (", se_diff, ")"),
-            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
-  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
-  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
-  select(Model = model, 
-         ends_with("overall"), 
-         ends_with("sent"), 
-         ends_with("beforeword"), 
-         ends_with("withinword"))
-
-```
-
-```{r cato, results='asis'}
-names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)
-
-kable(table, caption = "CATO data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 8))) %>%
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 11) %>%
-  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
-  add_header_above(c(" " = 1, 
-                     "All transitions" = 2,
-                     "Before sentence" = 2, 
-                     "Before word" = 2,
-                     "Within word" = 2)) %>%
-  column_spec(1, width = "8em")
-
-```
-
-## Posterior parameter estimates of mixture model
-
-```{r catopost, fig.cap="CATO. Posterior parameter distribution"}
-file <- list.files(path = "../stanout/cato", pattern = "mog.csv$",full.names = T)
-ps <- read_csv(file) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(across(delta, ~.+beta),
-         across(c(delta, beta), exp),
-         across(delta, ~.-beta),
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:prob) %>% 
-  group_by(task, group, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}")) %>% 
-  mutate(task = str_c("Task: ", task, "; Group: ", group))
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = task,
-             shape = task)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "",
-       shape = "") +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = task,
-             shape = task)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = task,
-             shape = task)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-# SPL2
-
-Data are going to be published in @torrance.
-
-## Methods
-
-Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
-CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
-We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
-
-Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-## Model comparisons
-
-### Fit to data
-
-```{r predictionsspl2, fig.cap="SPL2 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-files <- list.files("../stanout/spl2", pattern = "sims", full.names = T)
-sims <- read_csv(files) %>% 
-  pivot_longer(cols = c(y_obs, y_tilde))  %>% 
-  mutate(across(name, recode_factor, 
-                y_tilde = "Simulated data",
-                y_obs = "Observed data",
-                .ordered = T)) %>% 
-  filter(name ==  "Simulated data" |
-         (name == "Observed data" & sim_idx == 1))
-
-max_iki <- 2000
-
-plot_mog <- sims %>%
-  filter(model == "mog", value < max_iki) %>% 
-  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmm <- sims %>%
-  filter(model == "lmm", value < max_iki) %>% 
-  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmgaus <- sims %>%
-  filter(model == "lmmgaus", value < max_iki) %>% 
-  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmuneqvar <- sims %>%
-  filter(model == "lmmuneqvar", value < max_iki) %>% 
-  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
-  plot_layout(guides = "collect") +
-  plot_annotation(tag_levels = 'A')
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Out-of-samples cross-validation
-
-```{r}
-files <- list.files("../stanout/spl2", pattern = "^modelcomp", full.names = T)
-table <- purrr::map_dfr(files, read_csv) %>% 
-  mutate(across(location, replace_na, "overall")) %>% 
-  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
-  mutate(across(where(is.numeric), round, 0),
-         across(where(is.numeric), format, big.mark = ","),
-         across(everything(), str_trim)) %>%
-  transmute(across(location, ~str_c("Text location: ",.)),
-            across(model, recode, lmm = "Unimodal log-normal",
-                                  lmmgaus = "Unimodal normal",
-                                  mog = "Bimodal log-normal",
-                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
-            across(elpd_diff, str_c, " (", se_diff, ")"),
-            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
-  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
-  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
-  select(Model = model, 
-         ends_with("overall"), 
-         ends_with("sent"), 
-         ends_with("beforeword"), 
-         ends_with("withinword"))
-
-```
-
-```{r spl2, results='asis'}
-names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)
-
-kable(table, caption = "SPL2 data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 8))) %>%
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 11) %>%
-  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
-  add_header_above(c(" " = 1, 
-                     "All transitions" = 2,
-                     "Before sentence" = 2, 
-                     "Before word" = 2,
-                     "Within word" = 2)) %>%
-  column_spec(1, width = "8em")
-
-```
-
-## Posterior parameter estimates of mixture model
-
-```{r spl2post, fig.cap="SPL2. Posterior parameter distribution"}
-file <- list.files(path = "../stanout/spl2", pattern = "mog.csv$",full.names = T)
-ps <- read_csv(file) %>% 
-  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(across(delta, ~.+beta),
-         across(c(delta, beta), exp),
-         across(delta, ~.-beta),
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:prob) %>% 
-  group_by(lang, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"))
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = lang,
-             shape = lang)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 23)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "Language",
-       shape = "Language")
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = lang,
-             shape = lang)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 23)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "Language",
-       shape = "Language") +
-  theme(axis.title.y = element_blank())
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = lang,
-             shape = lang)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 23)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "Language",
-       shape = "Language") +
-  theme(axis.title.y = element_blank())
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-# PLanTra
-
-## Methods
-
-The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
-In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
-Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
-During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
-making it easier to read for a lay audience--by applying what they had learned from their respective training.
-The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
-It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
-
-## Model comparisons
-
-### Fit to data
-
-```{r predictionsplantra, fig.cap="PLanTra data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-files <- list.files("../stanout/plantra", pattern = "sims", full.names = T)
-sims <- read_csv(files) %>% 
-  pivot_longer(cols = c(y_obs, y_tilde))  %>% 
-  mutate(across(name, recode_factor, 
-                y_tilde = "Simulated data",
-                y_obs = "Observed data",
-                .ordered = T)) %>% 
-  filter(name ==  "Simulated data" |
-         (name == "Observed data" & sim_idx == 1))
-
-max_iki <- 2000
-
-plot_mog <- sims %>%
-  filter(model == "mog", value < max_iki) %>% 
-  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmm <- sims %>%
-  filter(model == "lmm", value < max_iki) %>% 
-  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmgaus <- sims %>%
-  filter(model == "lmmgaus", value < max_iki) %>% 
-  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmuneqvar <- sims %>%
-  filter(model == "lmmuneqvar", value < max_iki) %>% 
-  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
-  plot_layout(guides = "collect") +
-  plot_annotation(tag_levels = 'A')
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Out-of-samples cross-validation
-
-```{r}
-files <- list.files("../stanout/plantra", pattern = "^modelcomp", full.names = T)
-table <- purrr::map_dfr(files, read_csv) %>% 
-  mutate(across(location, replace_na, "overall")) %>% 
-  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
-  mutate(across(where(is.numeric), round, 0),
-         across(where(is.numeric), format, big.mark = ","),
-         across(everything(), str_trim)) %>%
-  transmute(across(location, ~str_c("Text location: ",.)),
-            across(model, recode, lmm = "Unimodal log-normal",
-                                  lmmgaus = "Unimodal normal",
-                                  mog = "Bimodal log-normal",
-                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
-            across(elpd_diff, str_c, " (", se_diff, ")"),
-            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
-  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
-  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
-  select(Model = model, 
-         ends_with("overall"), 
-         ends_with("sent"), 
-         ends_with("beforeword"), 
-         ends_with("withinword"))
-```
-
-```{r plantra, results='asis'}
-names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)
-
-kable(table, caption = "PLanTra data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 8))) %>%
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 11) %>%
-  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
-  add_header_above(c(" " = 1, 
-                     "All transitions" = 2,
-                     "Before sentence" = 2, 
-                     "Before word" = 2,
-                     "Within word" = 2)) %>%
-  column_spec(1, width = "8em")
-
-```
-
-## Posterior parameter estimates of mixture model
-
-```{r plantrapost, fig.cap="PLanTra. Posterior parameter distribution"}
-file <- list.files(path = "../stanout/plantra", pattern = "mog.csv$",full.names = T)
-ps <- read_csv(file) %>% 
-  mutate(across(task, factor, levels = c("pre-test", "post-test"), ordered = T)) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(across(delta, ~.+beta),
-         across(c(delta, beta), exp),
-         across(delta, ~.-beta),
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:prob) %>% 
-  group_by(task, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"))
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = task,
-             shape = task)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 23)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "Task",
-       shape = "Task") +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = task,
-             shape = task)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 23)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "Task",
-       shape = "Task") +
-  theme(axis.title.y = element_blank()) +
-    guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = task,
-             shape = task)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  geom_errorbar(width = 0, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 23)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "Task",
-       shape = "Task") +
-  theme(axis.title.y = element_blank()) +
-    guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-# LIFT
-
-## Methods
-
-LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
-
-## Model comparisons
-
-### Fit to data
-
-```{r predictionslift, fig.cap="LIFT data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-files <- list.files("../stanout/lift", pattern = "sims", full.names = T)
-sims <- read_csv(files) %>% 
-  pivot_longer(cols = c(y_obs, y_tilde))  %>% 
-  mutate(across(name, recode_factor, 
-                y_tilde = "Simulated data",
-                y_obs = "Observed data",
-                .ordered = T)) %>% 
-  filter(name ==  "Simulated data" |
-         (name == "Observed data" & sim_idx == 1))
-
-max_iki <- 2000
-
-plot_mog <- sims %>%
-  filter(model == "mog", value < max_iki) %>% 
-  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmm <- sims %>%
-  filter(model == "lmm", value < max_iki) %>% 
-  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmgaus <- sims %>%
-  filter(model == "lmmgaus", value < max_iki) %>% 
-  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plot_lmmuneqvar <- sims %>%
-  filter(model == "lmmuneqvar", value < max_iki) %>% 
-  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
-  ggplot(aes(x = value, 
-             group = interaction(sim_idx,name), 
-             colour = name)) +
-  geom_density() +
-  facet_wrap(~model) +
-  scale_colour_manual(values = c("firebrick3", "black")) +
-  scale_x_continuous(labels = scales::comma) +
-  labs(colour = "") +
-  theme(axis.title = element_blank())
-
-plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
-  plot_layout(guides = "collect") +
-  plot_annotation(tag_levels = 'A')
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Out-of-samples cross-validation
-
-```{r}
-files <- list.files("../stanout/lift", pattern = "^modelcomp", full.names = T)
-table <- purrr::map_dfr(files, read_csv) %>% 
-  mutate(across(location, replace_na, "overall")) %>% 
-  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
-  mutate(across(where(is.numeric), round, 0),
-         across(where(is.numeric), format, big.mark = ","),
-         across(everything(), str_trim)) %>%
-  transmute(across(location, ~str_c("Text location: ",.)),
-            across(model, recode, lmm = "Unimodal log-normal",
-                                  lmmgaus = "Unimodal normal",
-                                  mog = "Bimodal log-normal",
-                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
-            across(elpd_diff, str_c, " (", se_diff, ")"),
-            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
-  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
-  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
-  select(Model = model, 
-         ends_with("overall"), 
-         ends_with("sent"), 
-         ends_with("beforeword"), 
-         ends_with("withinword"))
-```
-
-```{r lift, results='asis'}
-names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)
-
-kable(table, caption = "LIFT data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 8))) %>%
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 11) %>%
-  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
-  add_header_above(c(" " = 1, 
-                     "All transitions" = 2,
-                     "Before sentence" = 2, 
-                     "Before word" = 2,
-                     "Within word" = 2)) %>%
-  column_spec(1, width = "8em")
-
-```
-
-## Posterior parameter estimates of mixture model
-
-```{r}
-file <- list.files(path = "../stanout/lift", pattern = "mog.csv$",full.names = T)
-ps <- read_csv(file) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(across(delta, ~.+beta),
-         across(c(delta, beta), exp),
-         across(delta, ~.-beta),
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:prob) %>% 
-  group_by(genre, topic, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}")) %>% 
-  ungroup() %>% 
-  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre))
-```
-
-```{r liftpost, fig.cap="LIFT. Posterior parameter distribution"}
-posd <- position_dodge(.65)
-dotsize <- 2.5
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = group,
-             shape = group)) +
-  geom_point(size = dotsize, position = posd) +
-  geom_errorbar(width = 0, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(1:8)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "",
-       shape = "")
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = group,
-             shape = group)) +
-  geom_point(size = dotsize, position = posd) +
-  geom_errorbar(width = 0, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(1:8)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             colour = group,
-             shape = group)) +
-  geom_point(size = dotsize, position = posd) +
-  geom_errorbar(width = 0, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(1:8)) +
-  scale_colour_viridis_d(end = .7) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-# Cross-task / data set comparisons
+# Cross-data set comparisons
 
@@
 
-# Also look at by ppt estimates?
-
 ```
@@
 ```{r}
-files <- str_c("../stanout/", 
-               c("lift", "spl2", "plantra", "cato", "c2l1"), 
-               "/mog.csv")
+files <- list.files(str_c("../stanout/", 
+                          c("lift", "spl2", "plantra", "cato", "c2l1")), 
+                    pattern = "mogbetaun.+.csv", 
+                    full.names = T)
+
 ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
   mutate(data = .x)) %>% 
-  mutate(across(data, str_remove_all, 
-                "\\.|\\/|stanout|\\mog.csv"),
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
          across(data, recode_factor, 
@@
                 "spl2" = "SPL2 (L1)", 
+#                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
                 "plantra" = "PLanTra",
@@
          !(str_detect(data, "CATO") & task == "masked")) %>% 
-#  filter(str_detect(data, "CATO"))
   pivot_wider(names_from = param, 
               values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(across(delta, ~.+beta),
-         across(c(delta, beta), exp),
-         across(delta, ~.-beta),
+  unnest(cols = beta:theta) %>% 
+  mutate(across(c(beta, beta2), exp),
+         delta = beta2 - beta,
          across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:prob) %>% 
-  group_by(data, location, name) %>% 
+  pivot_longer(beta:theta) %>% 
   summarise(across(value, 
@@
                         upper = upper),
-                   .names = "{.fn}")) %>% 
-  ungroup()
-```
+                   .names = "{.fn}"), .by = c(data, location, name)) %>% 
+  filter(name %in% c("beta", "delta", "prob"))
+```
+
+## Cross-data set visualisation
+
+The model estimates for the mixture-model with the highest predictive performance are shown in Figure \@ref(fig:crossstudypost).
+In this visualisation we ignore dataset specific conditions that are presented in detail below.
 
@@
              x = location,
-             colour = data,
              shape = data,
              group = interaction(data))) +
-  geom_line(size = .5, alpha = .35, position = posd) +
-  geom_point(size = dotsize, position = posd) +
-  geom_errorbar(width = 0, position = posd) +
+  geom_line(size = .5, 
+            alpha = .35, 
+            position = posd) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = data), 
+            size = dotsize, 
+            position = posd) +
   facet_grid(~name) +
-  scale_shape_manual(values = c(1, 4, 7, 8, 9)) +
-  scale_colour_viridis_d(end = .7) +
+  scale_shape_manual(values = c(1, 4, 7, 8, 9, 11)) +
+  scale_colour_colorblind() +
   scale_y_log10(labels = scales::comma) +
@@
              x = location,
-             colour = data,
              shape = data,
@@
   geom_line(size = .5, alpha = .35, position = posd) +
-  geom_point(size = dotsize, position = posd) +
-  geom_errorbar(width = 0, position = posd) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = data), 
+            size = dotsize, 
+            position = posd) +
   facet_grid(~name) +
-  scale_shape_manual(values = c(1, 4, 7, 8, 9)) +
-  scale_colour_viridis_d(end = .7) +
+  scale_shape_manual(values = c(1, 4, 7, 8, 9, 11)) +
+  scale_colour_colorblind() +
   scale_y_log10(labels = scales::comma) +
@@
   geom_line(size = .5, alpha = .35, position = posd) +
-  geom_point(size = dotsize, position = posd) +
-  geom_errorbar(width = 0, position = posd) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = data), 
+            size = dotsize, 
+            position = posd) +
   facet_grid(~name) +
-  scale_shape_manual(values = c(1, 4, 7, 8, 9)) +
-  scale_colour_viridis_d(end = .7) +
+  scale_shape_manual(values = c(1, 4, 7, 8, 9, 11)) +
+  scale_colour_colorblind() +
   scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
@@
 
+## Effect of transition location
+
+It's generally believed that pausing is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges, i.e.
+before sentence > before word > within word.
+We have evaluated the differences between transition locations for all data sets.
+The results are shown in Table \@ref(tab:loceffect).
+
+Results are largely consistent across data sets (with caveats) but differ, to some extent, from what the literature would predict.
+In line with the literature hesitations are more frequent before words than within words.
+Also hesitations are longer at before-sentence transitions compared to before-word transitions (except dataset C2L1) compared to within-word transitions (except dataset LIFT).
+However, our results do not support that writers pause more frequently at before-sentence locations compared to before-word locations (except for dataset SPL2; this also shows that more pauses at before-sentence locations can not be explained on the basis of multi-key combinations for sentence-initial capitalisation).
+Also, we observe that even fluent key-transitions are slower at before-word locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for dataset SPL2).
+
+The datasets differ to the extent that sentence-initial key transitions do (PLanTra, LIFT) or do not (CATO, C2L1, SPL2) include the character following the shift key for capitalisation.
+In other words, the pause before sentences may sum across two key intervals, namely `_^[shift]^C` but only involves one keyintervals, namely `_^[shift]`.
+For the SPL2 dataset, we calculated location effects for sentence-initial transitions that do and do not involve the shift-to-key transition.
+The results were the same for the transition location effects.
+A comparison that is untangling the effects of the multi-keycombination on the mixture-model estimates can be found in Table \@ref(tab:shiftcellmeans).
+In short, the duration of fluent transitions and the hesitation slowdown are affected but not the hesitation probability.
+
+In conclusion, while pauses tend to be longer before sentences they are not more frequent than before words.
+
+```{r loceffect, results='asis'}
+files <- list.files(str_c("../stanout/", 
+                          c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
+                    pattern = "mogbetaun.+.csv", 
+                    full.names = T)
+
+ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
+  mutate(data = .x)) %>% 
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
+         across(data, recode_factor, 
+                "cato" = "CATO (non-dyslexic\nunmasked)",
+                "spl2" = "SPL2 (L1)", 
+                "spl2_shift" = "SPL2 (L1; shift + C)", 
+                "plantra" = "PLanTra",
+                "lift" = "LIFT",
+                "c2l1" = "C2L1",
+                .ordered = TRUE)) %>% 
+  filter(!(str_detect(data, "SPL2") & lang == "ES"),
+         !(str_detect(data, "CATO") & group == "dyslexic"),
+         !(str_detect(data, "CATO") & task == "masked"),
+         param %in% c("beta", "delta", "theta")) %>%
+  select(param, location, data, value) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
+  pivot_wider(names_from = location, values_from = value) %>%
+  unnest(-c(data, param)) %>% 
+  mutate(diff.1 = `before sentence` - `before word`,
+         diff.2 = `before word` - `within word`) %>% 
+  summarise(across(starts_with("diff"), 
+                   list(mean = mean, lower = lower, upper = upper, BF = BF)),
+            .by = c(data, param))
+
+tmp <- ps_loc_diffs %>% 
+  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
+  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
+         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
+  select(-lower, -upper) %>% 
+  pivot_wider(names_from = param, values_from = c(mean, BF)) %>% 
+  mutate(across(diffid, ~recode(., diff.1 = "before sentence vs word",
+                                   diff.2 = "before vs within word"))) %>% 
+  select(data, diffid, ends_with("beta"), ends_with("delta"), ends_with("theta"))
+
+write_csv(tmp, "tables/location_effect_alldata.csv")
+
+tmp %>% knitr::kable(align =c("l", "l", rep("r", 6)), 
+                     col.names = c("Data set", "Difference", rep(c("Est. with 95% PIs", "BF"), 3)),
+                     caption = "Effect of transition location on keystroke intervals.
+Differences between transition locations are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95% PIs in brackets.") %>%
+  kableExtra::collapse_rows(columns = 1) %>%
+  kable_styling(position = "center", full_width = T, font_size = 11) %>%
+  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", threeparttable = T) %>% 
+  add_header_above(c(" " = 2, 
+                     "Fluent transitions" = 2, 
+                     "Slowdown for hesitations" = 2, 
+                     "Probability of hesitations" = 2))
+
+```
+
 ```{r eval = F}
+
+#
+# Do I really need all this?
+#
+#
+#
+#
+
 # use posterior to calculate interactions, cell means and dataset differences
-files <- str_c("../stanout/", c("lift", "spl2", "plantra", "cato", "c2l1"), "/mog.csv")
+files <- list.files(str_c("../stanout/", 
+                          c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
+                    pattern = "mogbetaun.+.csv", 
+                    full.names = T)
 
@@
 
-```{r}
+```{r eval = F}
 ps_effects <- read_csv("tables/ps_effects_acrossdata.csv")
@@
 
-```{r effects, results='asis'}
+```{r effects, results='asis', eval = F}
 ps_effects_summary %>%
@@
 
-```{r cellmeans, results = 'asis'}
+```{r cellmeans, results = 'asis', eval = F}
 tmp <- read_csv("tables/cellmeans_transdur.csv")
@@
 
+# Posterior by data set
+
+## C2L1
+
+The C2L1 data set comprises data Norwegian 6th graders--*N*=126, mean age 11 years 10 months--published in @ronneberg2022process.
+The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
+
+```{r eval = F}
+d <- read_csv("../data/c2l1.csv") %>%
+  filter(!is.na(iki), 
+         iki > 50, 
+         iki < 30000) %>%
+  mutate(ppt = as.numeric(factor(subno)),
+         condition = factor(str_c(location, sep = "_")),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt, iki, condition, cond_num)
+
+count(d, ppt) %>% 
+  filter(n < 100)
+pull(d, ppt) %>% unique() %>% length()
+```
+
+TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
+
+### Fit to data
+
+```{r predictionsc2l1, fig.cap="C2L1 data.
+Comparison of 100 simulated (predicted) sets of data to observed data illustated by model.
+For illustration the x-axis was truncated at 2,000 msecs.", eval = T}
+file <- list.files("../stanout/c2l1", 
+                   pattern = "sims", 
+                   full.names = T)
+
+plots <- get_pdens_plot(file, max_iki = 2000)
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Posterior parameter estimates of mixture model
+
+```{r c2l1post, fig.cap="C2l1. Posterior parameter distribution"}
+file <- list.files(path = "../stanout/c2l1", 
+                   pattern = "mogbetaun.*.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% 
+  filter(param %in% c("beta", "beta2", "prob")) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, beta2, prob)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(beta:delta) %>% 
+  group_by(location, name) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"))
+
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location)) +
+  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_y_continuous(labels = scales::comma) +
+  labs(y = "Posterior estimate with 95% PIs",
+       x = "",
+       colour = "",
+       shape = "") +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location)) +
+  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location)) +
+  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
+  geom_point(size = 3, position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+beta + delta + theta +
+  plot_layout(guides = "collect")
+
+```
+
+## CATO
+
+Data are published in @torrance2016adolescent. Norwegian upper
+secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
+
+### Fit to data
+
+```{r predictionscato, fig.cap="CATO data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+file <- list.files("../stanout/cato", 
+                    pattern = "sims", 
+                    full.names = T)
+
+plots <- get_pdens_plot(file, max_iki = 2000)
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Posterior parameter estimates of mixture model
+
+```{r catopost, fig.cap="CATO. Posterior parameter distribution"}
+file <- list.files(path = "../stanout/cato", 
+                   pattern = "mogbetaun.+.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% 
+  filter(param %in% c("beta", "beta2", "prob")) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, beta2, prob)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(beta:delta) %>% 
+  group_by(task, group, location, name) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}")) %>% 
+  mutate(task = str_c("Task: ", task, "; Group: ", group))
+
+pdodge <- position_dodge(.35)
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = task)) +
+  geom_errorbar(width = .2,
+                alpha = .75, 
+                position = pdodge) +
+  geom_point(aes(colour = task), 
+            size = 3, 
+            position = pdodge) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_colorblind() +
+  scale_y_continuous(labels = scales::comma) +
+  labs(y = "Posterior estimate with 95% PIs",
+       x = "",
+       colour = "",
+       shape = "") +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = task)) +
+  geom_errorbar(width = .2, alpha = .75, 
+                position = pdodge) +
+  geom_point(aes(colour = task), 
+             size = 3, 
+             position = pdodge) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_colorblind() +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = task)) +
+  geom_errorbar(width = .2, alpha = .75, 
+                position = pdodge) +
+  geom_point(aes(colour = task), 
+             size = 3, 
+             position = pdodge) +
+  facet_grid(~name) +
+  scale_shape_manual(values = 1:4) +
+  scale_colour_colorblind() +
+  scale_y_continuous(labels = dezero_plot, 
+                     limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+beta + delta + theta +
+  plot_layout(guides = "collect")
+
+```
+
+## PLanTra
+
+### Methods
+
+The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
+In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
+Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
+During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
+making it easier to read for a lay audience--by applying what they had learned from their respective training.
+The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
+It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
+
+### Fit to data
+
+```{r predictionsplantra, fig.cap="PLanTra data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+file <- list.files("../stanout/plantra", 
+                   pattern = "sims", 
+                   full.names = T)
+
+plots <- get_pdens_plot(file, max_iki = 2000)
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Posterior parameter estimates of mixture model
+
+```{r plantrapost, fig.cap="PLanTra. Posterior parameter distribution"}
+file <- list.files(path = "../stanout/plantra", 
+                   pattern = "mogbetaun.+.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% 
+  mutate(across(task, factor, 
+                levels = c("pre-test", "post-test"), 
+                ordered = T)) %>% 
+  filter(param %in% c("beta", "beta2", "prob")) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, beta2, prob)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(beta:delta) %>% 
+  group_by(task, location, name) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"))
+
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = task)) +
+  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
+  geom_point(aes(colour = task), 
+             size = 3, 
+             position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(21, 8)) +
+  scale_colour_colorblind() +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "Posterior estimate with 95% PIs",
+       x = "",
+       colour = "Task",
+       shape = "Task") +
+  guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = task)) +
+  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
+  geom_point(aes(colour = task), 
+             size = 3, 
+             position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(21, 8)) +
+  scale_colour_colorblind() +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "Task",
+       shape = "Task") +
+  theme(axis.title.y = element_blank()) +
+    guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = task)) +
+  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
+  geom_point(aes(colour = task), 
+             size = 3,
+             position = position_dodge(.25)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(21, 8)) +
+  scale_colour_colorblind() +
+  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "Task",
+       shape = "Task") +
+  theme(axis.title.y = element_blank()) +
+    guides(colour=guide_legend(nrow=2,byrow=TRUE),
+         shape = guide_legend(nrow=2,byrow=TRUE))
+
+beta + delta + theta +
+  plot_layout(guides = "collect")
+
+```
+
+## LIFT
+
+### Methods
+
+LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
+
+### Fit to data
+
+```{r predictionslift, fig.cap="LIFT data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+file <- list.files("../stanout/lift", 
+                    pattern = "sims", 
+                    full.names = T)
+
+plots <- get_pdens_plot(file, max_iki = 2000)
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Posterior parameter estimates of mixture model
+
+```{r}
+file <- list.files(path = "../stanout/lift", 
+                   pattern = "mogbetaun.+.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% 
+  filter(param %in% c("beta", "beta2", "prob")) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, beta2, prob)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(beta:delta) %>% 
+  group_by(genre, topic, location, name) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}")) %>% 
+  ungroup() %>% 
+  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre))
+```
+
+```{r liftpost, fig.cap="LIFT. Posterior parameter distribution"}
+posd <- position_dodge(.65)
+dotsize <- 2.5
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, 
+             position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(1:8)) +
+  scale_colour_colorblind() +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "Posterior estimate with 95% PIs",
+       x = "",
+       colour = "",
+       shape = "")
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, 
+             position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(1:8)) +
+  scale_colour_colorblind() +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank())
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, 
+             position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(1:8)) +
+  scale_colour_colorblind() +
+  scale_y_continuous(labels = dezero_plot, 
+                     limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank())
+
+beta + delta + theta +
+  plot_layout(guides = "collect")
+
+```
+
+## SPL2
+
+Data are going to be published in @torrance.
+
+### Methods
+
+Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
+CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
+We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
+
+Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
+
+### Fit to data
+
+```{r predictionsspl2, fig.cap="SPL2 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
+
+file <- list.files("../stanout/spl2", pattern = "sims", full.names = T)
+
+plots <- get_pdens_plot(file, max_iki = 2000)
+
+grid.arrange(patchworkGrob(plots), 
+             left = "Density",
+             bottom = "Transition duration in msecs")
+```
+
+### Posterior parameter estimates of mixture model
+
+Can slowdowns for sentence-pauses be explained on the basis of a complex keystrokes that were summed across? -- No
+
+The data sets used differ to the extent that the keystroke interval before sentences does (PLanTra, LIFT) or does not (CATO, C2L1, SPL2) scope over the character following Shift.
+In other words, the pause before sentences sums across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involves one keyintervals, namely `_^[shift]` for the remaining data sets.
+
+For the SPL2 dataset we compared whether the different patterns for sentences pauses can be explain but the keycombination.
+We analysed the SPL2 data including and excluding the keystroke after shift.
+Model estimates are presented in Figure \ref(fig:spl2post).
+The results of this comparison can be found in Table \@ref(tab:shiftcellmeans).
+Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the interval between shift and the first character but not the hesitation probability.
+Fluent keytransitions were substantially longer when including the interval following the shift-key.
+The slowdown for hesitations was affected too but the difference is numerically small.
+There was no conclusive evidence for an increased hesitation probability.
+Taken together, including the character following shift affects the duration of fluent transitions more than it affects pause duration and frequency.
+
+```{r spl2post, fig.cap="SPL2. Posterior parameter distribution"}
+file1 <- list.files(path = "../stanout/spl2", 
+                   pattern = "mogbetaun.+.csv$",
+                   full.names = T)
+
+file2 <- list.files(path = "../stanout/spl2_shift", 
+                   pattern = "mogbetaun.+.csv$",
+                   full.names = T)
+
+ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
+      mutate(data = .x)) %>% 
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
+         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
+                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
+  filter(param %in% c("beta", "delta", "prob")) %>% 
+  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2")) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, delta, prob)) %>% 
+  mutate(beta2 = beta + delta,
+         across(c(beta2, beta), ~exp(.)),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(c(beta, delta, prob)) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper),
+                   .names = "{.fn}"),
+            .by = c(lang, location, name, data))
+
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = data,
+             group = interaction(lang, data))) +
+  geom_errorbar(width = .2, alpha = .75, 
+                position = position_dodge(.5)) +
+  geom_point(aes(colour = lang), 
+             size = 3, 
+             position = position_dodge(.5)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(21, 8)) +
+  scale_color_colorblind() +
+  scale_y_continuous(labels = scales::comma) +
+  labs(y = "Posterior estimate with 95% PIs",
+       x = "",
+       colour = "Language",
+       shape = "Before-sentence\ntransitions")
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = data,
+             group = interaction(lang, data))) +
+  geom_errorbar(width = .2, alpha = .75,
+                position = position_dodge(.5)) +
+  geom_point(aes(colour = lang), 
+             size = 3, 
+             position = position_dodge(.5)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(21, 8)) +
+  scale_color_colorblind() +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "Language",
+       shape = "Before-sentence\ntransitions") +
+  theme(axis.title.y = element_blank())
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = data,
+             group = interaction(lang, data))) +
+  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.5)) +
+  geom_point(aes(colour = lang), 
+             size = 3, 
+             position = position_dodge(.5)) +
+  facet_grid(~name) +
+  scale_shape_manual(values = c(21, 8)) +
+  scale_colour_colorblind() +
+#  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "Language",
+       shape = "Before-sentence\ntransitions") +
+  theme(axis.title.y = element_blank())
+
+beta + delta + theta +
+  plot_layout(guides = "collect")
+
+```
+
+```{r shiftcellmeans, results = 'asis'}
+# cell means and data set difference
+cellmeans <- ps %>% 
+  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
+  pivot_wider(names_from = name, 
+              values_from = c(mean, lower, upper)) %>% 
+  mutate(across(ends_with("prob"), ~dezero(., 2)),
+         across(c(ends_with("beta"),
+                  ends_with("delta")), 
+                ~as.character(scales::comma(round(.))))) %>% 
+  pivot_longer(-c(lang, location, data), 
+               names_to = c(".value", "param"),
+               names_sep = "_") %>% 
+  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
+  select(-mean:-upper) %>% 
+  pivot_wider(names_from = data, 
+              values_from = est)
+
+file1 <- list.files(path = "../stanout/spl2", 
+                   pattern = "mogbetaun.+.csv$",
+                   full.names = T)
+
+file2 <- list.files(path = "../stanout/spl2_shift", 
+                   pattern = "mogbetaun.+.csv$",
+                   full.names = T)
+
+tmp <- map_dfr(c(file1, file2), 
+               ~read_csv(.x) %>% 
+          mutate(data = .x)) %>% 
+  filter(param %in% c("beta", "delta", "theta")) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
+  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
+         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
+  pivot_wider(names_from = data, values_from = value) %>% 
+  unnest(c(spl2, spl2_shift)) %>% 
+  mutate(diff = spl2_shift - spl2) %>% 
+  summarise(across(diff, list(mean = mean, 
+                              lower = lower, 
+                              upper = upper,
+                              BF = BF), 
+                   .names = "{.fn}"),
+            .by = c(lang, location, param)) %>% 
+  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
+         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
+         across(param, recode, theta = "prob")) %>% 
+  left_join(cellmeans) %>% 
+  select(param, Language = lang,
+         `Transition location` = location, 
+         `_\\^[shift] + C` = ends_with(" + C)"),
+         `_\\^[shift]` = ends_with("FT])"),
+         "Difference" = mean, BF)
+
+write_csv(tmp, "tables/shift_effect_cellmeans.csv")
+
+tmp[,-1] %>% knitr::kable(align = c("l", "l", "r", "r", "r", "r"), 
+                          caption = "Mixture model estimates for key transitions.
+Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of disfluent transitions.
+The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of disfluent transitions.
+95% PIs in brackets.") %>%
+  #  cell_spec(1, 3:4) %>% 
+  kableExtra::collapse_rows(columns = 1) %>%
+  kable_styling(position = "center", full_width = T) %>%
+  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
+           threeparttable = T) %>% 
+  group_rows("Fluent transitions", 1, 6) %>%
+  group_rows("Disfluencies", 7, 12) %>%
+  group_rows("Probability of disfluencies", 13, 18)
+
+```
+
 # References

### 2023-05-05 tutorial/tutorial.Rmd
Add tutorial and draft results

@@
     toc_depth: 3
-    use_bookdown: false
+    use_bookdown: true
 
@@
 # Turn off scientific notation
-options(scipen = 999)
+options(scipen = 999, width = 200)
 
 # Chunk defaults
-knitr::opts_knit$set(width=90)
+knitr::opts_knit$set(width=100)
 knitr::opts_chunk$set(fig.width = 8, 
                       fig.height = 4.5,
-                      size = "small",
                       echo = TRUE,
@@
                       message = FALSE)
+
+# Reduce font size in chunks
+def.chunk.hook  <- knitr::knit_hooks$get("chunk")
+knitr::knit_hooks$set(chunk = function(x, options) {
+  x <- def.chunk.hook(x, options)
+  ifelse(options$size != "small", paste0("\n \\", options$size,"\n\n", x, "\n\n \\small"), x)
+})
 
@@
 
-```{r}
+```{r eval = F}
 # Load data
@@
 
+```{r echo = F}
+# Load data
+data <- read_csv("data/sampledata.csv") 
+```
+
 We reduced the data to three variables: a participant identifier `participant`, the inter-keystroke intervals `iki`, and the transition location `location` (levels: within-word, before-word, before-sentence).
@@
 
+We start with two uni-modal models -- models that assume that the data come from a single underlying process that can be described as as Gaussian in the first example or a log-Gaussian in the second example.
+We will compare the models performance.
+
 ## Gaussian mixed-effects model
@@
 
+We start with a Gaussian mixed-effects models similar to a standard analysis of variance.
+We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
+The mean can be decomposed into $\beta$ and $\text{participant}_i$.
+$\beta$ will be allowed to take on a different value for each transition location.
+Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.
+
 $$
-\text{iki}_i \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\mu_i = \beta_\text{location[i]} + u_i\\
-u \sim \text{N}(0, \sigma_u) 
+\begin{align}
+\text{iki}_i \sim & \text{N}(\mu_i, \sigma_\text{e}^2)\\
+\mu_i = & \beta_\text{location[i]} + \text{participant}_i\\
+\text{participant} \sim & \text{N}(0, \sigma_\text{p}^2) 
+\end{align}
 $$
+This model can be described in R using the following formula wrapped in the `brms` formula function `bf`.
 
@@
 
+`brms` is using sensible default priors for most model parameters which can be checked using `get_prior`.
+
 ```{r}
@@
 ```
+
+We want to make sure that `(flat)` is replaced by weakly informative priors because uninformative or flat priors are never a good idea [@mcelreath2016statistical].
 
@@
 For the model to converge we need to run a sufficient number of iterations.
-30,000 iterations, as below, are a lot but does not guarantee convergence.
+10,000 iterations, as below, are a lot but does not guarantee convergence.
 We need to test convergence using the model's posterior (see blow).
@@
 $$
-\text{iki}_i \sim \text{logN}(\mu_i, \sigma_\text{e}^2)\\
-\mu_i = \beta_\text{location[i]} + u_i\\
-u \sim \text{N}(0, \sigma_u) 
+\begin{align}
+\text{iki}_i \sim & \text{logN}(\mu_i, \sigma_\text{e}^2)\\
+\mu_i = & \beta_\text{location[i]} + \text{participant}_i\\
+\text{participant} \sim & \text{N}(0, \sigma_\text{p}^2) 
+\end{align}
 $$
@@
 
-## TODO Model comparison
-
-ADD here info from slides on how to calculate elpd
+## Model comparison
 
@@
 Cross-validation, in contrast to more conventional model-comparison metrics such as $R^2$, prevent overfitting by penalising models with more parameters [see @mcelreath2016statistical; @lambert2018student].
-Predictive quality was summarised as the sum of the expected log predictive density (`elpd_loo`; $\widehat{elpd}$).
-The difference in `elpd_loo` is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
+
+Cross validation generally is measuring the performance of a model, meaning its ability to capture new data.
+Overfitting happens when a model is capturing nuances of a sample, rather than patterns one would expect to repeat in new samples.
+The leave-one-out information criterion (LOO-IC) is determined as follows: train a model on $N-1$ observations, predict the remaining data points from the training model, repeat process $N$ times to predict every observation from a model of th remaining data.
+Adding up the prediction results gives an estimate of the expected log predictive density (`elpd_loo`; $\widehat{elpd}$), i.e.
+an approximation of the results that would be expected for new data.
+
+The function `loo` can be used to access the expected log predictive density which uses probability calculations for approximate LOO-IC vio Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical].
+
+Approximation involved in `loo` uses the log posterior predictive densities: how likely is each data point given the distribution parameter estimates?
+This is shown for each data point in Figure \@ref(fig:loglik).
+Extremely small values indicate that an observation is very unlike for a given set of parameter values.
+
+```{r loglik, fig.cap="Log-posterior densities of each data point"}
+log_lik(fit_gaus) %>% 
+  as_tibble() %>% 
+  mutate(sample = 1:n()) %>%
+  pivot_longer(-sample, 
+               names_to = "obs", 
+               values_to = "loglik") %>%
+  group_by(obs) %>%
+  summarise(M = mean(loglik), var = var(loglik)) %>%
+  filter(M > -10) %>% # there are some terrible ones for this model :)
+  mutate(obs = as.numeric(str_remove(obs, "V"))) %>%
+  ggplot(aes(x = obs, y = M, ymin = M-var, ymax = M+var)) +
+  geom_pointrange(fatten = .5) +
+  labs(x = "Observations", 
+       y = "log-posterior predictive density (with variance)")
+```
+
+`elpd_loo` is the sum of the means and returns the expected log predictive density.
+`p_loo` is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters).
+`looic`, or LOO-IC, corresponds to $-2 \cdot ($`elpd_loo`$-$`p_loo`$)$ (for deviance scale).
+This can be manually estimated
+
+```{r}
+log_lik(fit_gaus) %>% 
+  as_tibble() %>% 
+  mutate(sample = 1:n()) %>%
+  pivot_longer(-sample, 
+               names_to = "obs", 
+               values_to = "loglik") %>%
+  group_by(obs) %>%
+  summarise(mean_loglik = mean(exp(loglik)),
+            var_loglik = var(loglik)) %>% 
+  ungroup() %>%
+  summarise(elpd_loo = sum(log(mean_loglik)), # sum of mean log lik
+            p_loo = sum(var_loglik), # effective number of parameters
+            looic = -2 * (elpd_loo - p_loo)) %>% # Bayesian deviance
+  pivot_longer(everything(), names_to = "Estimate") %>% 
+  mutate(across(value, round, 1))
+```
+
+or directly calculated using `loo`. The Pareto k diagnostic indicates that there are some observations that are very unusual under the uni-modal Gaussian model.
+
+```{r}
+loo(fit_gaus)
+```
+
+The difference in `elpd_loo` for both models can be obtained using `loo` and is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
 
@@
 
+Aside, `elpd_loo` and `looic` have rarely got a direct interpretation.
+Important is the difference between models.
+The size of the values is increasing as a function of the number of observations.
+Generally `elpd_loo` and `looic` measures how many possible futures the model can rule out.
+
+In contrast, `loo_R2` provides a Bayesian $R^2$ that measures the variability in the data that the model can captures.
+This quantity is not ideal for model comparisons, as normal $R^2$, because it is subject to overfitting: models with more parameters tend to have a larger $R^2$.
+A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; chapter 6 in @mcelreath2016statistical].
+The is an example for Bayesian $R^2$ for the uni-modal log-Gaussian model.
+
+```{r}
+loo_R2(fit_lgaus)
+```
+
 # Bi-modal mixed-effects log-Gaussian mixture model
@@
 $$
-\text{iki}_{i} \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_i, \sigma_{e'_\text{location[i]}}^2) +\\
-  (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_i, \sigma_{e_\text{location[i]}}^2)\\
-		\text{where}\\
-		\delta \sim \text{N}(0,1)\\
-		\text{constraint: } \delta > 0
+\begin{align}
+\text{iki}_{i} \sim & \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
+		&\text{where}\\
+		&\delta \sim \text{N}(0,1)\\
+		&\text{constraint: } \delta > 0
+\end{align}
 $$
 
-## TODO Mixture model simulation
+The parameters of this equation are visualised in Figure \@ref(fig:mixturemodel).
+The illustrate shows the dissociation of two underlying processes that the model is looking for in the data.
+There are two distributions of which the left one in grey is the distribution of fluent keystroke intervals; the right distribution in green is wider and captures generally longer keystroke intervals (with some overlap between fluent and disfluent transitions).
+
+```{r mixturemodel, echo = F, fig.cap="What's a mixture model for keystroke intervals?
+The parameters of the mixture model explained visually.
+Each of the three parameters is determined for each transition location."}
+
+plot_mix_comps <- function(x, mu, sigma, lam) lam * dnorm(x, mu, sigma)
+
+ps <- read_csv("data/posterior_for_densityplot.csv") %>%
+  mutate(edit = str_to_sentence(edit)) %>%
+  filter(pos == "before word", edit == "No editing", lang == "L1") %>% 
+  select(-pos, -edit) %>%
+  pivot_wider(names_from = param, values_from = est)
+
+alpha <- pull(ps, beta)
+alpha2 <- pull(ps, betalong)
+theta <- pull(ps, prob)
+sigma <- pull(ps, sigma_e)
+sigmap <- pull(ps, sigmap_e)
+
+data_mm <- data.frame(x = c(0, 12))
+
+ggplot(data_mm, aes(x)) +
+  scale_x_continuous(labels = exp, breaks = log(c(0, 10, 100, 1000, 10000)), limits = log(c(30, 50000))) +
+  scale_y_continuous(limits = c(0, .7)) +
+  stat_function(geom = "line", fun = plot_mix_comps, colour = "#000000",
+                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
+  stat_function(geom = "line", fun = plot_mix_comps, colour = "#009E73",
+                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
+  stat_function(geom = "area", fun = plot_mix_comps, fill = "#000000",
+                show.legend = F, alpha = .25, 
+                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
+  stat_function(geom = "area", fun = plot_mix_comps, fill = "#009E73",
+                show.legend = F, alpha = .25, 
+                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
+  annotate("label", x=4,y=.6, label="beta*':'~'Quick typing'", size = 4, parse = T) +
+  annotate("label", x=7.25,y=.35, label="atop(theta*':'~'Hesitation probability', '\"Pausing\"')", parse = T, size = 4) +
+  annotate("label", x=9,y=.15, parse = T, label="delta*':'~'\"Slowdown\"'", size = 4) +
+  geom_segment(x = alpha, y = 0.05, xend = alpha2, yend = 0.05, 
+               colour = "blue", alpha = .5,
+               arrow = arrow(length = unit(0.2, "cm"), 
+                             ends = "both", 
+                             type = "closed")) +
+  geom_segment(x = alpha, y = 0, xend = alpha, yend = Inf, colour = "darkred", alpha = .5) +
+  geom_segment(x = alpha2, y = 0, xend = alpha2, yend = Inf, colour = "darkred", alpha = .5) +
+  geom_curve(x = 4, y = .57, xend = alpha - .75, yend = 0.1, 
+  arrow = arrow(length = unit(0.03, "npc"))) +
+  geom_curve(x = 7.5, y = .3, xend = alpha2 + .85, yend = 0.1,
+             curvature=-.3,
+  arrow = arrow(length = unit(0.03, "npc"))) +
+  geom_curve(x = 9, y = .125, xend = 5.75, yend = 0.0255,
+             curvature=-.3,
+   arrow = arrow(length = unit(0.03, "npc"))) +
+  labs(x = "iki [in msecs]", 
+       y = "Density",
+       caption = "see Roeser et al. (2021)")
+
+```
+
+The average typing speed of fluent keystroke transitions is captured by the $\beta$ parameter.
+The distance between fluent transitions and disfluent transitions is captured by $\delta$ (i.e.
+the distance between the grey and the green distribution).
+The hesitation probability or the "pausing frequency" is the relative height of the grey and the green distribution and captured by $\theta$.
+Most keystrokes are, typically, fluent transitions and hence in the grey distribution.
+The frequency of the minority of disfluent transitions is captured by the height of the green distribution.
+As this model is bi-modal, the probability of fluent transitions is just $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions was $.6$.
+If the pausing probability decreases to $.3$ the proportion of fluent transitions had increased to $.7$.
 
@@
   int<lower=1> nS;                   // number of ppts
-  int<lower=1, upper=nS> ppts[N];    // ppts identifier
-  int<lower=1> nB;                   // number of bigrams
-  int<lower=1, upper=nB> bigrams[N]; // ppts identifier
+  int<lower=1, upper=nS> ppt[N];     // ppts identifier
   int<lower=1> K;                    // number of conditions
-  int condition[N];                  // condition identifiers
+  int<lower=1> condition[N];                  // condition identifiers
   vector[N] y;                       // outcome
@@
   scale_color_brewer("Transition\nlocation", palette = "Dark2") +
-  labs(x = "Posterior parameter estimate")
+  labs(x = "Posterior parameter estimate") +
+  theme(legend.position = "top")
 ```
@@
 
-## TODO Bayes Factor
-
-Add info from Savage Dickey Bayes Factor slide
-
-## Note on using more than a single IV
+## Bayes Factor
+
+Frequentist null-hypothesis testing is based on the *p*-value.
+Conceptually this test is evaluates whether the data allow us to reject the null hypothesis (i.e.
+the true parameter value being some difference, is zero).
+More specifically it tests whether the probability of observing a test statistic (e.g.
+*t*, *F*, $\chi^2$) as or more extreme if the null hypothesis were true.
+Therefore, *p*-values have two possible outcomes: reject the null (i.e.
+$p<.05$); inconclusive evidence (i.e.
+$p>.05$).
+And that's it.
+It does not provide direct evidence for the alternative hypothesis; nor does it provide evidence for the null hypothesis.
+
+Bayes Factors (BFs) provide a continuum of three possible outcomes [@dienes2014using; @dienes2016bayes; @dienes2018four]: (1) there is evidence to reject the null / alternative hypothesis; (2) the evidence is inconclusive (get more data!); (3) accept the null / alternative hypothesis.
+
+For example, if our BF is the probably of the alternative hypothesis $H1$ over the null hypothesis $H_0$, values $>1$ indicate that there is more evidence for $H1$ and $<1$ indicates more evidence for $H_0$.
+This can be turned around by swapping nominator and denominator
+
+$$
+\text{BF}_{10} = \frac{p(H_1 \mid y)}{p(H_0 \mid y)} 
+$$
+or by taking the inverse of $\text{BF}_{10}$, meaning $\text{BF}_{01} = \frac{1}{\text{BF}_{10}}$.
+
+The strength of evidence for $H_0$ and $H_1$ is continuous rather than binary (i.e.
+evidence vs no evidence).
+Authors have proposed different soft thresholds of what constitutes evidence [see, e.g., @bag12; @jeffreys1961theory; @lee2014bayesian] but generally a BF larger than 3 is weak evidence, larger than 5 / 6 indicate moderate evidence and 10 / 32 is strong evidence for a statistically meaningful effect.
+For example BF$_{10}$=2 reflects that the alternative hypothesis is two times more likely than the null hypothesis given the evidence.
+BF$_{10}$=0.33 is weak evidence in favour of the null hypothesis because $\frac{1}{0.33} \approx 3$.
+Notably BFs can be used to obtain evidence in favour of the null [for discussion see @dienes2014using; @dienes2018four; @dienes2016bayes].
+
+The BF can be calculated using the Savage-Dickey method [see, e.g., @dickey1970weighted; @jeffreys1961theory; @wagenmakers2010bayesian].
+The BF is determined on the basis of the height of the posterior density at zero compared to the height of the prior density at zero.
+Important to note here is that the resulting BF is sensitive to the prior in situations when the sample is small and / or prior information play a big role.
+It is therefore a good idea to carefully think about the prior in advance of analysis (hence "prior") and to be mindful of the role of priors in the calculation of BFs.
+
+Figure \@ref(fig:sdbf) shows an example for the calculation of BFs based on a posterior with an effect difference that is distributed according to $\text{N}(2.5, 1)$ compared against three priors with a wide SD $\text{N}(0, 10)$, a SD equal to the posterior $\text{N}(0, 2.5)$, and relatively small SD $\text{N}(0, 1)$.
+
+```{r sdbf, fig.cap="Example for the BF calculated using the Savage-Dickey method for 3 differently wide priors.", echo=F}
+bf_labs <- function(pr, po, rd = 3) {
+  bquote(paste("BF ="~frac(.(round(pr,rd)),.(round(po,rd)))~"="~.(round(pr/po,rd))))
+}
+
+pd <- tibble(x = seq(-5, 5, by = .1),
+             Prior_1 = dnorm(x, 0, 10),
+             Prior_2 = dnorm(x, 0, 2.5),
+             Prior_3 = dnorm(x, 0, 1),
+             Posterior = dnorm(x, 2.5, 1)) %>%
+  pivot_longer(Prior_1:Posterior)
+
+data_at_zero <- filter(pd, x == 0)
+pr_1 <- filter(data_at_zero, name == "Prior_1") %>% pull(value)
+pr_2 <- filter(data_at_zero, name == "Prior_2") %>% pull(value)
+pr_3 <- filter(data_at_zero, name == "Prior_3") %>% pull(value)
+po <- filter(data_at_zero, name == "Posterior") %>% pull(value)
+
+data_at_zero <- data_at_zero %>% 
+  mutate(is_prior = as.numeric(!name == "Posterior"),
+         po = po,
+         pr = NA_real_,
+         pr = ifelse(name == "Prior_1", pr_1, pr),
+         pr = ifelse(name == "Prior_2", pr_2, pr),
+         pr = ifelse(name == "Prior_3", pr_3, pr)) %>% 
+  rowwise() %>% 
+  mutate(pr_label = list(bf_labs(pr, po, rd = 2))) %>% 
+  ungroup()
+
+data_at_zero[data_at_zero$name == "Posterior",'pr_label'] <- NA
+
+pd %>% mutate(across(name, recode_factor,
+                     Posterior = "Posterior",
+                     Prior_1 = "Prior: N(0,10)",
+                     Prior_2 = "Prior: N(0,2.5)",
+                     Prior_3 = "Prior: N(0,1)", .ordered = TRUE)) %>% 
+  ggplot(aes(x = x, y = value, colour = name)) +
+  geom_line() +
+  geom_point(data = data_at_zero, 
+             inherit.aes = F,
+             aes(x = x, y = value), 
+             size = 2) +
+  geom_label(data = data_at_zero, 
+             inherit.aes = F,
+             aes(x = x, 
+                 y = value, 
+                 label = pr_label, 
+                 alpha = is_prior),
+             size = 2.75, 
+             parse = T, 
+             nudge_y = .04, 
+             show.legend = F) +
+  ggthemes::scale_color_colorblind("") +
+  scale_y_continuous(limits = c(0, .45)) +
+  labs(y = "density", x = "difference") 
+```
+
+To calculate the BF on the basis of posterior samples, the Savage-Dickey method can be implemented like so [taken from @nicenboim2016statistical] which requires an appropriate prior distribution.
+We will fix the mean of the prior to 0 but allow to change the SD as function argument.
+
+```{r}
+BF <- function(posterior_difference, prior_sd = 1){
+  library(polspline)
+  fit <- logspline(posterior_difference) 
+  posterior <- dlogspline(0, fit) # Height of the posterior at 0 
+  prior <- dnorm(0, 0, prior_sd) # Height of the prior at 0
+  return(prior/posterior)
+}
+```
+
+For demonstration purposes we will calculate the BF with different priors SDs for the hesitation probability $\theta$, or in particular the posterior difference in hesitation $\theta_\text{diff}$ for before-sentence vs before-word transitons and before-word vs within-word transitions.
+
+First, we extract vectors of differences
+
+```{r}
+# For sentence vs word
+theta_diff_sent <- filter(posterior_diffs, parameter == "theta[diff]") %>% pull(diff_sent)
+
+# For before word vs within word
+theta_diff_word <- filter(posterior_diffs, parameter == "theta[diff]") %>% pull(diff_word)
+```
+
+and then apply the `BF` function from above first to the difference between before-sentence and before-word transitions.
+For interpretation it is important to remember that $\theta$ is a probability; hence the difference between two probabilities can only range between -1 and 1.
+In other words, a prior SD of 1 is saying that all differences are as likely as no difference (which makes very little sense here).
+Similar to the calculations for ROPE above, it is useful to contemplate which prior distribution appropriately reflects the null hypothesis for the scale the difference is on (e.g.
+probability, log msecs, msecs).
+
+```{r}
+# Prior with sd of 1
+BF(theta_diff_sent, prior_sd = 1)
+# Prior with sd of .5
+BF(theta_diff_sent, prior_sd = .5)
+# Prior with sd of .1
+BF(theta_diff_sent, prior_sd = .1)
+```
+
+For `brms` fits the `hypothesis` function calculates the BFs indicated as evidence ratio.
+For a model such as `y ~ 1 + condition`, `hypothesis` can calculate the evidence for the hypothesis that the effect for `condition` is, for example, equal to 0.
+`hypothesis` can then be applied to the model with a statement that describes the hypothesis that should be evaluated.
+
+```{r eval = F, echo = F}
+diff <- posterior_samples(fit_lgaus) %>% 
+  mutate(diff = b_locationbeforeMsentence - b_locationbeforeMword)
+
+hypothesis(diff, "diff = 0")
+
+```
+
+```{r eval = F}
+hypothesis(brm_model, "conditionb = 0")
+```
+
+To calculate the evidence ratio, which is based on the height of prior and posterior at 0, the argument `sample_prior = TRUE` (or `sample_prior = 'yes'`) needs to be specified in `brm`.
+
+## Note on using more than a single grouping variable
 
@@
   labs(x = "Participant id", 
-       y = "Parameter estimates with 95% PIs") +
+       y = "Hesitation probability with 95% PIs") +
   scale_colour_brewer("Transition\nlocation", palette = "Dark2") +
@@
   labs(x = "Participant id", 
-       y = "Difference between transition locations") 
+       y = "Difference in hesitation probabilities with 95% PIs") 
 ```

### 2023-05-05 scripts/c2l1/get_data.R

@@ -1,0 +1,22 @@
+get_data <- function(file, n_samples){
+  d <- read_csv(file) %>%
+    filter(!is.na(iki), 
+           iki > 50, 
+           iki < 30000) %>%
+    mutate(ppt = as.numeric(factor(subno)),
+           condition = factor(str_c(location, sep = "_")),
+           cond_num = as.integer(condition),  
+           location = factor(location),
+           loc_num = as.integer(location)) %>% 
+    select(ppt, iki, condition, location, cond_num, loc_num) 
+  
+  # Sample within each category random data points
+  set.seed(365)
+  d <- d %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples,
+           .by = c(ppt, condition)) %>% 
+    filter(keep) %>% 
+    select(-keep)
+}

### 2023-05-05 scripts/c2l1/get_ytilde.R

@@ -1,82 +1,42 @@
 # Extract simulated data and compare to read data
-# for all four models
+# for all models
 
 # Load packages
 library(tidyverse)
 library(rstan)
 
-# Extract model simulations
-get_ytilde <- function(files, idx) {
-  # Load model
-  m <- readRDS(files)
-  
-  # Get model name
-  model_name <- str_remove_all(files, "stanout\\/|c2l1\\/|\\.rda")
-  
-  # Extract simulations
-  y_tilde <- as.matrix(m, pars = "y_tilde")
-    
-  # Extract 100 simulations
-  y_tilde_sims <- y_tilde[idx,]
+# Get functions
+source("scripts/c2l1/get_data.R")
+source("scripts/get_ytilde.R")
 
-  # Make data frame
-  sims <- y_tilde_sims %>% 
-    as_tibble() %>% 
-    rownames_to_column("sim_idx") %>% 
-    pivot_longer(-sim_idx) %>% 
-    mutate(data_idx = parse_number(name)) %>% 
-    arrange(data_idx) %>% 
-    mutate(model = model_name) %>% 
-    select(-name)
-  
-  return(sims)
-}
-
-# Load data
-n_sample <- 50 # number of random data points
+# File location and number of random samples
+n_samples <- 100 # number of random data points
+file <- "data/c2l1.csv"
 
 # Load df
-d <- read_csv("data/c2l1.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000) %>%
-  mutate(ppt = as.numeric(factor(subno)),
-         condition = factor(str_c(location, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num) 
-
-# Sample within each category random data points
-set.seed(365)
-d <- d %>% group_by(ppt, condition) %>%
-  mutate(keep = 1:n(),
-         keep = sample(keep),
-         keep = keep <= n_sample) %>% 
-  ungroup() %>% 
-  filter(keep) %>% 
-  select(-keep) %>% 
+d <- get_data(file, n_samples) %>% 
   mutate(data_idx = 1:n())
-
-count(d, ppt, condition)
 
 # Number of simulations per model
 nsims <- 100
 
 # Max number of iterations
-maxiter <- 15000
+maxiter <- 20000
 
 # Sample random idx for simulations
 idx <- sample(maxiter, nsims)
 
 # File names
[…]

### 2023-05-05 scripts/c2l1/lmmgaus.R

@@ -1,35 +1,21 @@
 # Load packages
 library(tidyverse)
 library(rstan)
+source("scripts/c2l1/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
 # Sampling parameters
 n_cores <- 3
 n_chain <- 3
-iterations <- 10000
-n_sample <- 50 # number of random data points
+iterations <- 20000
+n_samples <- 100 # number of random data points
+file <- "data/c2l1.csv"
 
 # Load df
-d <- read_csv("data/c2l1.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000) %>%
-  mutate(ppt = as.numeric(factor(subno)),
-         condition = factor(str_c(location, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num) 
+d <- get_data(file, n_samples)
 
-# Sample within each category random data points
-set.seed(365)
-d <- d %>% group_by(ppt, condition) %>%
-  mutate(keep = 1:n(),
-         keep = sample(keep),
-         keep = keep <= n_sample) %>% 
-  ungroup() %>% 
-  filter(keep) %>% 
-  select(-keep)
-
+# Count observations
 count(d, ppt, condition)
 
 # Data as list
@@ -42,18 +28,12 @@ dat <- within( list(), {
   N <- nrow(d)
 } );str(dat)
 
-# Check lmer for starting values
-m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
-
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_mu = 770
-          , beta_sigma = 100
-          , beta_raw = rep(0, dat$K)
-          , sigma_mu = 1580
-          , sigma_sigma = 10
-          , sigma_raw = rep(100, dat$K)
-          , sigma_u = 300)}
+  list(    alpha = 1400
+          , beta_e = rep(0, dat$K)
+          , sigma = 1580
+          , sigma_u = 250)}
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 
@@ -61,7 +41,7 @@ start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 lmm <- stan_model(file = "stan/lmmgaus.stan")
 
 # Parameters to omit in output
[…]

### 2023-05-05 scripts/c2l1/mogbetaconstr.R

@@ -1,0 +1,104 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+source("scripts/c2l1/get_data.R")
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 20000
+n_samples <- 100 # number of random data points
+file <- "data/c2l1.csv"
+
+# Load df
+d <- get_data(file, n_samples)
+
+# Count observations
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  K_loc <- length(unique(d$loc_num))
+  location <- as.numeric(d$loc_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta = 4
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta = rep(0, dat$K)
+          , tau = .1
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+mog <- stan_model(file = "stan/mogbetaconstr.stan")
+
+# Parameters to omit in output
+omit <- c("prob_tilde", "mu", "theta_s")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 16,
+                             adapt_delta = 0.99))
+
+# Save model
+saveRDS(m, 
+        file = "stanout/c2l1/mogbetaconstr.rda",
+        compress = "xz")
+
+#m <- readRDS(file = "stanout/c2l1/mogbetaconstr.rda")
+
[…]

### 2023-05-05 scripts/cato/lmm.R

@@ -1,35 +1,21 @@
 # Load packages
 library(tidyverse)
 library(rstan)
+source("scripts/cato/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
 # Sampling parameters
 n_cores <- 3
 n_chain <- 3
-iterations <- 10000
-n_sample <- 50 # number of random data points
+iterations <- 20000
+n_samples <- 100 # number of random data points
+file <- "data/cato.csv"
 
 # Load df
-d <- read_csv("data/cato.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000) %>%
-  mutate(ppt = as.numeric(factor(subno)),
-         condition = factor(str_c(location, xn, dystyp, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num) 
+d <- get_data(file, n_samples)
 
-# Sample within each category random data points
-set.seed(365)
-d <- d %>% group_by(ppt, condition) %>%
-  mutate(keep = 1:n(),
-         keep = sample(keep),
-         keep = keep <= n_sample) %>% 
-  ungroup() %>% 
-  filter(keep) %>% 
-  select(-keep)
-
+# Count sampled observations
 count(d, ppt, condition)
 
 # Data as list
@@ -44,12 +30,9 @@ dat <- within( list(), {
 
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_mu = 5
-          , beta_sigma = .1
-          , beta_raw = rep(0, dat$K)
-          , sigma_mu = 1
-          , sigma_sigma = .1
-          , sigma_raw = rep(0.1, dat$K)
+  list(   alpha = 5
+          , beta_e = rep(0, dat$K)
+          , sigma = 1
           , sigma_u = 0.1)}
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
@@ -62,7 +45,7 @@ start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 lmm <- stan_model(file = "stan/lmm.stan")
 
 # Parameters to omit in output
-omit <- c("mu",  "_mu", "_raw", "_sigma", "z_u", "L_u")
+omit <- c("mu")
 
 # Fit model
 m <- sampling(lmm, 
@@ -77,9 +60,8 @@ m <- sampling(lmm,
               include = FALSE, # Don't include the following parameters in the output
[…]

### 2023-05-05 scripts/cato/lmmgaus.R

@@ -1,35 +1,21 @@
 # Load packages
 library(tidyverse)
 library(rstan)
+source("scripts/cato/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
 # Sampling parameters
 n_cores <- 3
 n_chain <- 3
-iterations <- 10000
-n_sample <- 50 # number of random data points
+iterations <- 20000
+n_samples <- 100 # number of random data points
+file <- "data/cato.csv"
 
 # Load df
-d <- read_csv("data/cato.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000) %>%
-  mutate(ppt = as.numeric(factor(subno)),
-         condition = factor(str_c(location, xn, dystyp, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num) 
+d <- get_data(file, n_samples)
 
-# Sample within each category random data points
-set.seed(365)
-d <- d %>% group_by(ppt, condition) %>%
-  mutate(keep = 1:n(),
-         keep = sample(keep),
-         keep = keep <= n_sample) %>% 
-  ungroup() %>% 
-  filter(keep) %>% 
-  select(-keep)
-
+# Count sampled observations
 count(d, ppt, condition)
 
 # Data as list
@@ -42,18 +28,12 @@ dat <- within( list(), {
   N <- nrow(d)
 } );str(dat)
 
-# Check lmer for starting values
-m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
-
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_mu = 615
-          , beta_sigma = 100
-          , beta_raw = rep(0, dat$K)
-          , sigma_mu = 1850
-          , sigma_sigma = 10
-          , sigma_raw = rep(100, dat$K)
-          , sigma_u = 165)}
+  list(    beta_e = rep(0, dat$K)
+          , alpha = 5
+          , sigma = 1150
+          , sigma_u = 195)}
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 
@@ -61,7 +41,7 @@ start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 lmm <- stan_model(file = "stan/lmmgaus.stan")
 
 # Parameters to omit in output
[…]

### 2023-05-05 scripts/cato/modelcomparison.R

@@ -1,11 +1,11 @@
 library(loo)
 library(tidyverse)
 
-path <- "stanout/plantra"
+path <- "stanout/cato"
 (files <- list.files(path, pattern = "^.[^\\_]*.rda$", full.names = T))
 
 for(file in files){
-    varname <- str_remove_all(file, ".rda|stanout/plantra/")
+    varname <- str_remove_all(file, ".rda|stanout/cato/")
     m <- readRDS(file)
     log_lik <- extract_log_lik(m, merge_chains = F) 
     r_eff <- relative_eff(exp(log_lik)) 
@@ -22,8 +22,9 @@ mcs <- mcs %>% as.data.frame() %>%
   mutate(model = recode(model, model1 = loos[1],
                                model2 = loos[2],
                                model3 = loos[3],
-                               model4 = loos[4]),
+                               model4 = loos[4],
+                               model5 = loos[5]),
          model = str_remove(model, "^loo_"));mcs
 
-file_out <- paste0("stanout/plantra/modelcomparison_overall.csv")
+file_out <- paste0("stanout/cato/modelcomparison.csv")
 write_csv(mcs, file_out)

### 2023-05-05 scripts/cato/mogbetaunconstr.R

@@ -1,0 +1,101 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+source("scripts/cato/get_data.R")
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 20000
+n_samples <- 100 # number of random data points
+file <- "data/cato.csv"
+
+# Load df
+d <- get_data(file, n_samples)
+
+# Count sampled observations
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  K_loc <- length(unique(d$loc_num))
+  location <- as.numeric(d$loc_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(    beta = rep(4, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta = rep(0, dat$K)
+          , tau = .1
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K_loc)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+mog <- stan_model(file = "stan/mogbetaunconstr.stan")
+
+# Parameters to omit in output
+omit <- c("prob_tilde", "mu", "theta_s")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 16,
+                             adapt_delta = 0.99))
+
+# Save model
+saveRDS(m, 
+        file = "stanout/cato/mogbetaunconstr.rda",
+        compress = "xz")
+
+#m <- readRDS(file = "stanout/cato/mog.rda")
+
[…]

### 2023-05-05 scripts/lift/get_ytilde.R

@@ -1,86 +1,35 @@
 # Extract simulated data and compare to read data
-# for all four models
+# for all models
 
 # Load packages
 library(tidyverse)
 library(rstan)
 
-# Extract model simulations
-get_ytilde <- function(files, idx) {
-  # Load model
-  m <- readRDS(files)
-  
-  # Get model name
-  model_name <- str_remove_all(files, "stanout\\/|lift\\/|\\.rda")
-  
-  # Extract simulations
-  y_tilde <- as.matrix(m, pars = "y_tilde")
-    
-  # Extract 100 simulations
-  y_tilde_sims <- y_tilde[idx,]
+# Get functions
+source("scripts/lift/get_data.R")
+source("scripts/get_ytilde.R")
 
-  # Make data frame
-  sims <- y_tilde_sims %>% 
-    as_tibble() %>% 
-    rownames_to_column("sim_idx") %>% 
-    pivot_longer(-sim_idx) %>% 
-    mutate(data_idx = parse_number(name)) %>% 
-    arrange(data_idx) %>% 
-    mutate(model = model_name) %>% 
-    select(-name)
-  
-  return(sims)
-}
-
-# Load data
+# File location and number of random samples
 n_samples <- 50 # number of random data points
 n_ppts <- 100 # number of random ppts
+file <- "data/lift.csv"
 
-ppt_keep <- read_csv("data/lift.csv") %>% 
-  count(ppt, topic, genre) %>% 
-  group_by(ppt) %>% 
-  summarise(n = n()) %>% 
-  ungroup() %>% 
-  filter(n == 4) %>% 
-  pull(ppt)
-
-# Load df
-d <- read_csv("data/lift.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000,
-         enough_sentences,
-         ppt %in% ppt_keep) %>%
-  drop_na() %>% 
-  mutate(across(ppt, ~as.numeric(factor(.))),
-         condition = factor(str_c(location, genre, topic, sep = "_"))) %>% 
-  select(ppt, iki, condition) 
-
-# Sample within each category random data points
-set.seed(365)
-d <- d %>% group_by(ppt, condition) %>%
-  mutate(keep = 1:n(),
-         keep = sample(keep),
-         keep = keep <= n_samples) %>% 
-  ungroup() %>% 
-  filter(keep,
-         ppt %in% sample(unique(ppt), n_ppts)) %>% select(-keep) %>% 
[…]

### 2023-05-05 scripts/lift/lmmgaus.R

@@ -1,48 +1,22 @@
 # Load packages
 library(tidyverse)
 library(rstan)
+source("scripts/lift/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
 # Sampling parameters
 n_cores <- 3
 n_chain <- 3
-iterations <- 10000
+iterations <- 20000
 n_samples <- 50 # number of random data points
 n_ppts <- 100 # number of random ppts
+file <- "data/lift.csv"
 
-ppt_keep <- read_csv("data/lift.csv") %>% 
-  count(ppt, topic, genre) %>% 
-  group_by(ppt) %>% 
-  summarise(n = n()) %>% 
-  ungroup() %>% 
-  filter(n == 4) %>% 
-  pull(ppt)
+# Load data
+d <- get_data(file, n_samples, n_ppts)
 
-# Load df
-d <- read_csv("data/lift.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000,
-         enough_sentences,
-         ppt %in% ppt_keep) %>%
-  drop_na() %>% 
-  mutate(across(ppt, ~as.numeric(factor(.))),
-         condition = factor(str_c(location, genre, topic, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num) 
-
-# Sample within each category random data points
-set.seed(365)
-d <- d %>% group_by(ppt, condition) %>%
-  mutate(keep = 1:n(),
-         keep = sample(keep),
-         keep = keep <= n_samples) %>% 
-  ungroup() %>% 
-  filter(keep,
-         ppt %in% sample(unique(ppt), n_ppts)) %>% select(-keep) %>% 
-  mutate(across(ppt, ~as.numeric(factor(.))))
-
+# Check number of observations
 count(d, ppt, condition)
 
 # Data as list
@@ -55,18 +29,12 @@ dat <- within( list(), {
   N <- nrow(d)
 } );str(dat)
 
-# Check lmer for starting values
-m <- lme4::lmer(iki ~ 1 + (1|ppt), data = d);m
-
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_mu = 350
-          , beta_sigma = 100
-          , beta_raw = rep(0, dat$K)
-          , sigma_mu = 890
-          , sigma_sigma = 10
-          , sigma_raw = rep(100, dat$K)
-          , sigma_u = 80)}
[…]

### 2023-05-05 scripts/lift/zzz/get_loos.R

@@ -1,0 +1,20 @@
+library(loo)
+library(tidyverse)
+options(mc.cores = 1)
+
+path <- "~/prowrite-mixture-models/stanout/lift"
+
+# Iterate over log liklihood and get loo-ic
+logliks <- list.files(str_c(path, "/loglik"), 
+                      pattern = "^.[^\\_]*.rda$", 
+                      full.names = T)
+
+# Get variable names
+#varnames <- str_remove_all(logliks, ".rda|prowrite-mixture-models/stanout/lift/loglik/")
+
+log_lik <- readRDS(logliks[1]);gc
+r_eff <- relative_eff(exp(log_lik));gc()
+loo <- loo(log_lik, r_eff = r_eff, cores = 1);gc
+saveRDS(loo, 
+        str_c(path, "/loos/", "lmm", ".rda"),
+        compress = "xz")

### 2023-05-05 scripts/plantra/get_ytilde.R

@@ -1,80 +1,40 @@
 # Extract simulated data and compare to read data
-# for all four models
+# for all models
 
 # Load packages
 library(tidyverse)
 library(rstan)
 
-# Extract model simulations
-get_ytilde <- function(files, idx) {
-  # Load model
-  m <- readRDS(files)
-  
-  # Get model name
-  model_name <- str_remove_all(files, "stanout\\/|plantra\\/|\\.rda")
-  
-  # Extract simulations
-  y_tilde <- as.matrix(m, pars = "y_tilde")
-    
-  # Extract 100 simulations
-  y_tilde_sims <- y_tilde[idx,]
-
-  # Make data frame
-  sims <- y_tilde_sims %>% 
-    as_tibble() %>% 
-    rownames_to_column("sim_idx") %>% 
-    pivot_longer(-sim_idx) %>% 
-    mutate(data_idx = parse_number(name)) %>% 
-    arrange(data_idx) %>% 
-    mutate(model = model_name) %>% 
-    select(-name)
-  
-  return(sims)
-}
+# Get functions
+source("scripts/get_ytilde.R")
+source("scripts/plantra/get_data.R")
 
 # Load data
-n_sample <- 50 # number of random data points
+n_sample <- 100 # number of random data points
+file <- "data/plantra.csv"
 
-# Load df
-d <- read_csv("data/plantra.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000,
-         enough_sentences) %>%
-  mutate(across(ppt, ~as.numeric(factor(.))),
-         condition = factor(str_c(location, task, sep = "_"))) %>% 
-  select(ppt, iki, condition) 
-
-# Sample within each category random data points
-set.seed(365)
-d <- d %>% group_by(ppt, condition) %>%
-  mutate(keep = 1:n(),
-         keep = sample(keep),
-         keep = keep <= n_sample) %>% 
-  ungroup() %>% 
-  filter(keep) %>% 
-  select(-keep) %>% 
+# Load data
+d <- get_data(file, n_sample) %>% 
   mutate(data_idx = 1:n())
 
 # Number of simulations per model
 nsims <- 100
 
 # Max number of iterations
-maxiter <- 15000
+maxiter <- 20000
 
 # Sample random idx for simulations
 idx <- sample(maxiter, nsims)
 
 # File names
-(files <- str_c("stanout/plantra/", 
[…]

### 2023-05-05 scripts/plantra/modelcomparison.R

@@ -1,11 +1,11 @@
 library(loo)
 library(tidyverse)
 
-path <- "stanout/cato"
+path <- "stanout/plantra"
 (files <- list.files(path, pattern = "^.[^\\_]*.rda$", full.names = T))
 
 for(file in files){
-    varname <- str_remove_all(file, ".rda|stanout/cato/")
+    varname <- str_remove_all(file, ".rda|stanout/plantra/")
     m <- readRDS(file)
     log_lik <- extract_log_lik(m, merge_chains = F) 
     r_eff <- relative_eff(exp(log_lik)) 
@@ -22,8 +22,9 @@ mcs <- mcs %>% as.data.frame() %>%
   mutate(model = recode(model, model1 = loos[1],
                                model2 = loos[2],
                                model3 = loos[3],
-                               model4 = loos[4]),
+                               model4 = loos[4],
+                               model5 = loos[5]),
          model = str_remove(model, "^loo_"));mcs
 
-file_out <- paste0("stanout/cato/modelcomparison_overall.csv")
+file_out <- paste0("stanout/plantra/modelcomparison.csv")
 write_csv(mcs, file_out)

### 2023-05-05 scripts/plantra/mogbetaconstr.R

@@ -1,0 +1,103 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+source("scripts/plantra/get_data.R")
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 20000
+n_sample <- 100 # number of random data points
+file <- "data/plantra.csv"
+
+# Load data
+d <- get_data(file)
+
+# Check counts
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  K_loc <- length(unique(d$loc_num))
+  location <- as.numeric(d$loc_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta = 4
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta = rep(0, dat$K)
+          , tau = .1
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K_loc)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+mog <- stan_model(file = "stan/mogbetaconstr.stan")
+
+# Parameters to omit in output
+omit <- c("prob_tilde", "mu", "theta_s")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 16,
+                             adapt_delta = 0.99))
+
+# Save model
+saveRDS(m, 
+        file = "stanout/plantra/mogbetacontr.rda",
+        compress = "xz")
+
+#m <- readRDS(file = "stanout/plantra/mogbetacontr.rda")
+
+# Select relevant parameters
[…]

### 2023-05-05 scripts/plantra/mogbetaunconstr.R

@@ -1,0 +1,101 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+source("scripts/plantra/get_data.R")
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 20000
+n_sample <- 100 # number of random data points
+file <- "data/plantra.csv"
+
+# Load data
+d <- get_data(file)
+
+# Check counts
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  K_loc <- max(d$loc_num)
+  location <- d$loc_num
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta = rep(5, dat$K)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta = rep(0, dat$K)
+          , tau = .1
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K_loc)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+mog <- stan_model(file = "stan/mogbetaunconstr.stan")
+
+# Parameters to omit in output
+omit <- c("prob_tilde", "mu", "theta_s")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 2000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 16,
+                             adapt_delta = 0.99))
+
+# Save model
+saveRDS(m, 
+        file = "stanout/plantra/mogbetaunconstr.rda",
+        compress = "xz")
+
+#m <- readRDS(file = "stanout/plantra/mogbetaunconstr.rda")
+
[…]

### 2023-05-05 scripts/spl2/get_data.R

@@ -1,0 +1,31 @@
+get_data <- function(file, n_samples){
+  # Load df
+  d <- read_csv(file) %>%
+    select(-transition_dur) %>% 
+    rename(transition_dur = transition_dur_to_mod) %>% 
+    filter(!is.na(transition_dur), 
+           transition_dur > 50, 
+           transition_dur < 30000,
+           edit == "noedit") %>%
+    mutate(location_count = n(),
+           .by = c(SubNo, Lang, transition_type)) %>% 
+    mutate(enough_sentences = min(location_count) > 10,
+           .by = SubNo) %>% # at least 10 sentences
+    filter(enough_sentences) %>% 
+    mutate(SubNo = as.numeric(factor(SubNo)),
+           condition = factor(str_c(Lang, transition_type, sep =  "_")),
+           cond_num = as.integer(condition),
+           location = factor(transition_type),
+           loc_num = as.integer(location)) %>% 
+    select(ppt = SubNo, iki = transition_dur, condition, cond_num, location, loc_num) 
+  
+  # Sample within each category 100 random data points per loc and ppt
+  set.seed(365)
+  d <- d %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples,
+           .by = c(ppt, condition)) %>% 
+    filter(keep) %>% 
+    select(-keep)
+}

### 2023-05-05 scripts/spl2/modelcomparison.R

@@ -1,9 +1,12 @@
+# Load packages
 library(loo)
 library(tidyverse)
 
+# Get model file names
 path <- "stanout/spl2"
 (files <- list.files(path, pattern = "^.[^\\_]*.rda$", full.names = T))
 
+# Load models and extract loo 
 for(file in files){
     varname <- str_remove_all(file, ".rda|stanout/spl2/")
     m <- readRDS(file)
@@ -13,19 +16,22 @@ for(file in files){
     print(varname); rm(list = "m"); gc()
 }
 
+# Compare all models
 (loos <- ls(pattern = "loo_.*"))
 mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
 
+# Process model summary
 mcs <- mcs %>% as.data.frame() %>% 
   rownames_to_column("model") %>%
   as_tibble() %>%
   mutate(model = recode(model, model1 = loos[1],
                                model2 = loos[2],
                                model3 = loos[3],
-                               model4 = loos[4]),
+                               model4 = loos[4],
+                               model5 = loos[5]),
          model = str_remove(model, "^loo_"));mcs
 
-file_out <- paste0("stanout/spl2/modelcomparison_overall.csv")
+file_out <- paste0("stanout/spl2/modelcomparison.csv")
 write_csv(mcs, file_out)
 
 

### 2023-05-05 scripts/spl2/zzz/mog.R

@@ -11,6 +11,8 @@ iterations = 10000
 
 # Load df
 d <- read_csv("data/spl2.csv") %>%
+  select(-transition_dur) %>% 
+  rename(transition_dur = transition_dur_to_mod) %>% 
   filter(!is.na(transition_dur), 
          transition_dur > 50, 
          transition_dur < 30000,
@@ -20,10 +22,9 @@ d <- read_csv("data/spl2.csv") %>%
   group_by(SubNo) %>% 
   mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
   ungroup() %>%
-  filter(enough_sentences,
-         transition_type == "word_before") %>% 
+  filter(enough_sentences) %>% 
   mutate(SubNo = as.numeric(factor(SubNo)),
-         condition = factor(Lang),
+         condition = factor(str_c(Lang, transition_type, sep =  "_")),
          cond_num = as.integer(condition)) %>% 
   select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
 
@@ -37,8 +38,6 @@ d <- d %>% group_by(ppt, condition) %>%
   filter(keep) %>% 
   select(-keep)
 
-count(d, ppt, condition)
-
 # Data as list
 dat <- within( list(), {
   nS <- length(unique(d$ppt))
@@ -49,10 +48,9 @@ dat <- within( list(), {
   N <- nrow(d)
 } );str(dat)
 
-
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_mu = 6
+  list(   beta_mu = 5
           , beta_sigma = .1
           , beta_raw = rep(0, dat$K)
           , delta = rep(.1, dat$K)
@@ -67,8 +65,7 @@ start <- function(chain_id = 1){
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 
-# Stan models
-
+# Stan model
 # Load model
 mog <- stan_model(file = "stan/mog.stan")
 
@@ -83,7 +80,7 @@ m <- sampling(mog,
               warmup = iterations/2,
               chains = n_chain, 
               cores = n_cores,
-              refresh = 2000,
+              refresh = 5000,
               save_warmup = FALSE, # Don't save the warmup
               include = FALSE, # Don't include the following parameters in the output
               pars = omit,
@@ -94,19 +91,40 @@ m <- sampling(mog,
 
 # Save model
 saveRDS(m, 
[…]

### 2023-05-29 report/results.Rmd
add alternative spl2 analysis and gunnexp2

@@
 library(gtable)
-library(gridExtra)
 library(scales)
@@
 $$
-This models takes into account two sources of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
+This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
 
@@
 
-# Datasets in overview
-
-Five dataset with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).
+# Datasets
+
+## General overview
+
+Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).
 
@@
                                  "Synthesis"),
-                n_ppt = c(126, 26*2, 39, 47, 658),
+                 `N (ppts)` = c(126, 26*2, 39, 47, 658),
 #                n_texts = c(1, 2, 2, 2, NA),
@@
 
+##  Transition types
+
+The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g.
+@torrance; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \@ref(tab:keyloc).
+Keytransitions that terminated in an editing operation were excluded from the analysis.
+Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions.
+
+```{r keyloc, results='asis'}
+tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
+       Description = c("Transitions between any letter",
+                       "Keypress after space followed by any letter",
+                       "Keypress following a space preceding any letter"),
+       Example = c("T^h^e c^a^t m^e^o^w^e^d. T^h^a^t[bsp][bsp]e^n i^t s^l^e^p^t.", 
+                   "The ^cat ^meowed. That[bsp][bsp]en ^it ^slept.", 
+                   "The cat meowed. ^That[bsp][bsp]en it slept.")) %>% 
+  kable(caption = "Transition location classification.") %>%
+  footnote("'^' marks transition location, [bsp] represents backspace. IKIs were timed to the shift keypress.") %>% 
+  kable_styling(position = "center", 
+                full_width = T, 
+                font_size = 12)
+
+```
+
+## Data reduction
+
+```{r reductionfunctions}
+# Function to get data reduction infos
+get_lift <- function(file, n_samples, n_ppts){
+#  file <- "../data/lift.csv"
+# n_samples = 50
+#  n_ppts = 100
+  result <- list()
+
+  d <- read_csv(file)
+
+  full_set <- d %>% 
+    count(ppt, topic, genre) %>% 
+    summarise(n = n(), .by = ppt) %>% 
+    mutate(has_full_set = n == 4)
+
+  result[[3]] <- full_set %>% count(has_full_set)
+
+  ppt_keep <- full_set %>% 
+    filter(has_full_set) %>% 
+    pull(ppt)
+
+  result[[4]] <- d %>% 
+    filter(ppt %in% ppt_keep) %>% 
+    select(ppt, enough_sentences) %>% 
+    unique() %>% 
+    count(enough_sentences)
+
+  result[[1]] <- d %>%
+    filter(!is.na(iki), 
+           !is_edit,
+           ppt %in% ppt_keep,
+           enough_sentences) %>%
+    select(ppt, iki, location) %>% 
+    drop_na() %>% 
+    mutate(too_fast = iki <= 50, 
+           too_slow = iki >= 30000,
+           across(ppt, ~as.numeric(factor(.)))) %>% 
+    summarise(across(starts_with("too"), 
+                     list(mean = mean, 
+                          se = se_bin)),
+              .by = c(ppt)) %>% 
+    summarise(across(starts_with("too"), 
+                     list(mean = mean), 
+                     .names = "{.col}"))
+
+  # Sample within each category random data points
+  set.seed(365)
+  result[[2]] <- d %>% 
+    filter(!is.na(iki), 
+           !is_edit,
+           ppt %in% ppt_keep,
+           enough_sentences,
+           iki > 50,
+           iki < 30000) %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples,
+           .by = c(ppt, location, genre, topic)) %>% 
+   summarise(across(keep, 
+             list(mean = mean, 
+                  se = se_bin)),
+              .by = c(ppt, location)) %>% 
+   summarise(across(starts_with("keep"), 
+                     list(mean = mean), 
+                     .names = "{.col}"), 
+              .by = location)
+
+   set.seed(365)
+   result[[5]] <- d %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples, 
+           .by = c(ppt, location, genre, topic)) %>% 
+    filter(keep) %>% 
+    mutate(ppt_keep = ppt %in% sample(unique(ppt), n_ppts)) %>% 
+    select(ppt, ppt_keep) %>% 
+    unique() %>% 
+    count(ppt_keep)
+
+   return(result)
+}
+get_c2l1 <- function(file, n_samples){
+  #file <- "../data/c2l1.csv"
+  #n_samples <- 100
+  d <- read_csv(file) %>% 
+    filter(!is.na(iki)) %>% 
+    mutate(ppt = as.numeric(factor(subno))) %>% 
+    select(ppt, iki, location)
+
+  result <- list()
+
+  result[[1]] <- d %>% 
+    mutate(too_fast = iki <= 50,
+           too_slow = iki >= 30000) %>%
+    summarise(across(starts_with("too"), 
+                     list(mean = mean, se = se_bin)),
+              .by = c(ppt)) %>% 
+    summarise(across(starts_with("too"), 
+                     list(mean = mean), 
+                     .names = "{.col}"))
+
+  set.seed(365)
+  result[[2]] <- d %>% 
+    filter(iki > 50,
+           iki < 30000) %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples,
+           .by = c(ppt, location)) %>% 
+   summarise(across(keep, 
+             list(mean = mean, se = se_bin)),
+              .by = c(ppt, location)) %>% 
+    summarise(across(starts_with("keep"), 
+                     list(mean = mean), 
+                     .names = "{.col}"), 
+              .by = location)
+
+  return(result)
+}
+get_cato <- function(file, n_samples){
+  d <- read_csv(file) %>% 
+    filter(!is.na(iki)) %>% 
+    mutate(ppt = as.numeric(factor(subno))) %>% 
+    select(ppt, iki, location, xn, dystyp)
+
+  result <- list()
+
+  result[[1]] <- d %>% 
+    mutate(too_fast = iki <= 50,
+           too_slow = iki >= 30000) %>%
+    summarise(across(starts_with("too"), 
+                     list(mean = mean, 
+                          se = se_bin)),
+              .by = c(ppt)) %>% 
+    summarise(across(starts_with("too"), 
+                     list(mean = mean), 
+                     .names = "{.col}"))
+
+  set.seed(365)
+  result[[2]] <- d %>% 
+    filter(iki > 50,
+           iki < 30000) %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples,
+           .by = c(ppt, location, xn, dystyp)) %>% 
+   summarise(across(keep, 
+             list(mean = mean, 
+                  se = se_bin)),
+              .by = c(ppt, location)) %>% 
+   summarise(across(starts_with("keep"), 
+                     list(mean = mean), 
+                     .names = "{.col}"), 
+              .by = location)
+
+  return(result)
+}
+get_plantra <- function(file, n_samples){
+  d <- read_csv(file) %>% 
+    filter(!is.na(iki), !is_edit) %>% 
+    mutate(ppt = as.numeric(factor(ppt))) %>% 
+    select(ppt, iki, location, task)
+
+  result <- list()
+
+  result[[1]] <- d %>% 
+    mutate(too_fast = iki <= 50,
+           too_slow = iki >= 30000) %>%
+    summarise(across(starts_with("too"), 
+                     list(mean = mean, 
+                          se = se_bin)),
+              .by = c(ppt, task)) %>% 
+    summarise(across(starts_with("too"), 
+                     list(mean = mean), 
+                     .names = "{.col}"))
+
+  # keep only participants that did both the pre and post task.
+  keep_ppts <- select(d, ppt, task) %>% 
+    unique() %>% 
+    count(ppt) %>% 
+    mutate(both_tasks = n == 2)
+
+  result[[3]] <- count(keep_ppts, both_tasks)
+
+  keep_ppts <- keep_ppts %>% filter(both_tasks) %>% pull(ppt)
+  d <- filter(d, ppt %in% keep_ppts)
+
+  # Filter ppts with too few sentences
+  keep_ppts <- count(d, ppt, location) %>% 
+    filter(location == "before sentence") %>% 
+    mutate(enough_sentences = n >= 10)
+
+  result[[4]] <- count(keep_ppts, enough_sentences) %>% 
+    rename(more_than_10_sentences = enough_sentences)
+
+  keep_ppts <- keep_ppts %>% 
+    filter(enough_sentences) %>% 
+    pull(ppt)
+
+  d <- filter(d, ppt %in% keep_ppts)
+
+  set.seed(365)
+  result[[2]] <- d %>% 
+    filter(iki > 50,
+           iki < 30000) %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples,
+           .by = c(ppt, location, task)) %>% 
+   summarise(across(keep, 
+             list(mean = mean, 
+                  se = se_bin)),
+              .by = c(ppt, location)) %>% 
+   summarise(across(starts_with("keep"), 
+                     list(mean = mean), 
+                     .names = "{.col}"), 
+              .by = location)
+  return(result)
+}
+get_spl2 <- function(file, n_samples){
+  result <- list()
+#  file <- "../data/spl2.csv"
+  d <- read_csv(file) %>% 
+    select(-transition_dur) %>% 
+    rename(iki = transition_dur_to_mod,
+           location = transition_type) %>% 
+    filter(!is.na(iki), 
+           edit == "noedit") %>% 
+    select(ppt = SubNo, iki, location, Lang) %>% 
+    drop_na()
+
+  result[[1]] <- d %>% 
+    mutate(too_short = iki <= 50,
+           too_fast = iki >= 30000,
+           across(ppt, ~as.numeric(factor(.)))) %>% 
+    summarise(across(starts_with("too"), 
+                     list(mean = mean, 
+                          se = se_bin)),
+              .by = c(ppt)) %>% 
+    summarise(across(starts_with("too"), 
+                     list(mean = mean), 
+                     .names = "{.col}"))
+
+  d <- d %>% filter(iki > 50, iki < 30000)
+
+  # Filter ppts with too few sentences
+  keep_ppts <- count(d, ppt, location) %>% 
+    filter(location == "sentence_before") %>% 
+    mutate(enough_sentences = n >= 10)
+
+  result[[3]] <- count(keep_ppts, enough_sentences) %>% 
+    rename(more_than_10_sentences = enough_sentences)
+
+  keep_ppts <- keep_ppts %>% 
+    filter(enough_sentences) %>% 
+    pull(ppt)
+
+  d <- filter(d, ppt %in% keep_ppts)
+
+  set.seed(365)
+  result[[2]] <- d %>% 
+    mutate(keep = 1:n(),
+           keep = sample(keep),
+           keep = keep <= n_samples,
+           .by = c(ppt, location, Lang)) %>% 
+   summarise(across(keep, 
+             list(mean = mean, 
+                  se = se_bin)),
+              .by = c(ppt, location)) %>% 
+   summarise(across(starts_with("keep"), 
+                     list(mean = mean), 
+                     .names = "{.col}"), 
+              .by = location)
+
+  return(result)
+}
+```
+
+```{r}
+# Get info about data reduction
+n_samples <- 100
+c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
+cato <- get_cato("../data/cato.csv", n_samples = n_samples)
+lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
+plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
+spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples)
+
+```
+
 ```{r eval = F}
@@
 
-Locations with example from Marks
-
 Removed transitions that were followed by an editing operation
@@
 Add a description with general data filtering and a table with how many data were removed from the analysis by study.
+
+Add final sample number of ppt, and transitions by location
 
@@
 
+## GUNNEXP2
+
+### Fit to data
+
+### Posterior parameter estimates of mixture model
+
 ## C2L1

### 2023-05-29 scripts/c2l1/mogbetaconstr.R

@@ -38,7 +38,7 @@ start <- function(chain_id = 1){
           , theta = rep(0, dat$K)
           , tau = .1
           , sigma = 1.25
-          , sigma_diff = rep(.1, dat$K)
+          , sigma_diff = rep(.1, dat$K_loc)
           , sigma_u = 0.41)}
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )

### 2023-05-29 scripts/gunnexp2/lmm.R

@@ -1,7 +1,7 @@
 # Load packages
 library(tidyverse)
 library(rstan)
-source("scripts/plantra/get_data.R")
+source("scripts/gunnexp2/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
@@ -10,12 +10,12 @@ n_cores <- 3
 n_chain <- 3
 iterations <- 20000
 n_samples <- 100 # number of random data points
-file <- "data/plantra.csv"
+file <- "data/gunnexp2.csv"
 
-# Load data
+# Load df
 d <- get_data(file, n_samples)
 
-# Check counts
+# Count observations
 count(d, ppt, condition)
 
 # Data as list
@@ -30,18 +30,15 @@ dat <- within( list(), {
 
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_e= rep(0, dat$K)
-          , alpha = 600
-          , sigma = 1850
-          , sigma_u = 165)}
+  list(   alpha = 6 
+          , beta_e = rep(0, dat$K)
+          , sigma = .7
+          , sigma_u = 0.3)}
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 
 # Load model
-lmm <- stan_model(file = "stan/lmmgaus.stan")
-
-# Parameters to omit in output
-omit <- c("mu")
+lmm <- stan_model(file = "stan/lmm.stan")
 
 # Fit model
 m <- sampling(lmm, 
@@ -54,14 +51,14 @@ m <- sampling(lmm,
               refresh = 2000,
               save_warmup = FALSE, # Don't save the warmup
               include = FALSE, # Don't include the following parameters in the output
-              pars = omit,
+              pars = "mu",
               seed = 81,
               control = list(max_treedepth = 16,
                              adapt_delta = 0.99))
 
 # Save model
 saveRDS(m, 
-        file = "stanout/plantra/lmmgaus.rda",
+        file = "stanout/gunnexp2/lmm.rda",
         compress = "xz")
 
 # Select relevant parameters
@@ -72,3 +69,4 @@ summary(print(m, pars = param, probs = c(.025,.975)))
 
 # Traceplots
 traceplot(m, param, inc_warmup = F)
+

### 2023-05-29 scripts/gunnexp2/mogbetaconstr.R

@@ -1,7 +1,7 @@
 # Load packages
 library(tidyverse)
 library(rstan)
-source("scripts/plantra/get_data.R")
+source("scripts/gunnexp2/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
@@ -10,12 +10,12 @@ n_cores <- 3
 n_chain <- 3
 iterations <- 20000
 n_samples <- 100 # number of random data points
-file <- "data/plantra.csv"
+file <- "data/gunnexp2.csv"
 
-# Load data
+# Load df
 d <- get_data(file, n_samples)
 
-# Check counts
+# Count observations
 count(d, ppt, condition)
 
 # Data as list
@@ -24,15 +24,15 @@ dat <- within( list(), {
   subj <- d$ppt
   K <- length(unique(d$cond_num))
   condition <- as.numeric(d$cond_num)
-  K_loc <- max(d$loc_num)
-  location <- d$loc_num
+  K_loc <- length(unique(d$loc_num))
+  location <- as.numeric(d$loc_num)
   y <- d$iki
   N <- nrow(d)
 } );str(dat)
 
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta = rep(5, dat$K)
+  list(   beta = 4
           , delta = rep(.1, dat$K)
           , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
           , theta = rep(0, dat$K)
@@ -44,7 +44,7 @@ start <- function(chain_id = 1){
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 
 # Load model
-mog <- stan_model(file = "stan/mogbetaunconstr.stan")
+mog <- stan_model(file = "stan/mogbetaconstr.stan")
 
 # Parameters to omit in output
 omit <- c("prob_tilde", "mu", "theta_s")
@@ -67,13 +67,13 @@ m <- sampling(mog,
 
 # Save model
 saveRDS(m, 
-        file = "stanout/plantra/mogbetaunconstr.rda",
+        file = "stanout/gunnexp2/mogbetaconstr.rda",
         compress = "xz")
 
-#m <- readRDS(file = "stanout/plantra/mogbetaunconstr.rda")
+#m <- readRDS(file = "stanout/c2l1/mogbetaconstr.rda")
 
 # Select relevant parameters
-param <- c("beta", "delta", "prob", "sigma") 
[…]

### 2023-05-29 scripts/gunnexp2/mogbetaunconstr.R

@@ -1,7 +1,7 @@
 # Load packages
 library(tidyverse)
 library(rstan)
-source("scripts/plantra/get_data.R")
+source("scripts/gunnexp2/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
@@ -10,12 +10,12 @@ n_cores <- 3
 n_chain <- 3
 iterations <- 20000
 n_samples <- 100 # number of random data points
-file <- "data/plantra.csv"
+file <- "data/gunnexp2.csv"
 
-# Load data
+# Load df
 d <- get_data(file, n_samples)
 
-# Check counts
+# Count observations
 count(d, ppt, condition)
 
 # Data as list
@@ -32,7 +32,7 @@ dat <- within( list(), {
 
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta = rep(5, dat$K)
+  list(   beta = rep(4, dat$K)
           , delta = rep(.1, dat$K)
           , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
           , theta = rep(0, dat$K)
@@ -61,19 +61,19 @@ m <- sampling(mog,
               save_warmup = FALSE, # Don't save the warmup
               include = FALSE, # Don't include the following parameters in the output
               pars = omit,
-              seed = 81,
+              seed = 83,
               control = list(max_treedepth = 16,
                              adapt_delta = 0.99))
 
 # Save model
 saveRDS(m, 
-        file = "stanout/plantra/mogbetaunconstr.rda",
+        file = "stanout/gunnexp2/mogbetaunconstr.rda",
         compress = "xz")
 
-#m <- readRDS(file = "stanout/plantra/mogbetaunconstr.rda")
+#m <- readRDS(file = "stanout/cato/mogbetaunconstr.rda")
 
 # Select relevant parameters
-param <- c("beta", "delta", "prob", "sigma") 
+param <- c("beta", "delta", "prob", "sigma", "sigma_u") 
 
 # Traceplots
 traceplot(m, param, inc_warmup = F)
@@ -82,19 +82,20 @@ traceplot(m, param, inc_warmup = F)
 summary(print(m, pars = param, probs = c(.025,.975)))
 
 # Extract posterior
-ps <- as.data.frame(m, c("beta", "delta", "theta", "beta2", "prob")) %>% as_tibble()
[…]

### 2023-05-29 scripts/gunnexp2/processing.R

@@ -1,0 +1,28 @@
+# load libraries
+library(tidyverse)
+library(janitor)
+
+# get file name
+(files <- list.files("data/MarksData", pattern = "Gunn", full.names = T))
+
+# load data
+readRDS(files) %>% 
+  as_tibble() %>% 
+  clean_names() %>% 
+  filter(!is.na(ke_location), 
+         ke_location %in% c("a_^a", "a^a", "a._^a")) %>% 
+  select(word, iki = iki_mod, ppt = subno, xn, topic, 
+         location = ke_location,
+         isfluent = pwd_isfluent) %>% 
+  mutate(across(xn, recode, 
+                X = "masked", 
+                N = "unmasked"),
+         across(location, recode,
+                `a_^a` = "before word",
+                `a^a` = "within word",
+                `a._^a` = "before sentence"),
+         across(ppt, ~as.numeric(factor(.))),
+         condition = str_c(location, xn, topic, sep = "_")) %>% 
+  write_csv("data/gunnexp2.csv")
+
+  

### 2023-05-29 scripts/lift/get_data.R

@@ -1,15 +1,17 @@
 get_data <- function(file, n_samples, n_ppts){
-  ppt_keep <- read_csv(file) %>% 
-    count(ppt, topic, genre) %>% 
+   d <- read_csv(file) 
+   
+   ppt_keep <- d %>% count(ppt, topic, genre) %>% 
     summarise(n = n(), .by = ppt) %>% 
     filter(n == 4) %>% 
     pull(ppt)
   
   # Load df
-  d <- read_csv(file) %>%
+  d <- d %>%
     filter(!is.na(iki), 
            iki > 50, 
            iki < 30000,
+           !is_edit,
            enough_sentences,
            ppt %in% ppt_keep) %>%
     drop_na() %>% 

### 2023-05-29 scripts/plantra/preprocessing.R

@@ -1,0 +1,46 @@
+library(tidyverse)
+library(janitor)
+source("scripts/functions.R")
+
+# plantra: translation and rewrite text to plain language; professional translators
+
+# Get log files
+files <- list.files(path = "data/WritingPhases/PlanTra_Data/data_out", 
+                    full.names = TRUE);files
+
+# Process data: get only typing events
+plantra <- map_dfr(files, read_csv) %>% 
+  remove_constant() %>% 
+  remove_empty() %>% 
+  clean_names() %>% 
+  #count(participant, folder) %>% count(participant) %>% 
+  #filter(n == 2)
+  indicate_dels_and_inserts() %>% 
+  filter(type == "keyboard", 
+         !arrowkeys,
+         !output %in% c("CAPS LOCK", "RCTRL", "PAUSE"), 
+         !str_detect(output, "OEM|CTRL|SHIFT|ALT")) %>%
+  select(-arrowkeys, -type, -starts_with("deleted_")) %>% 
+  rename(iki = pause_time,
+         location = pause_location_full) %>% 
+  mutate(output = ifelse(output %in% c("SPACE", "TAB", "RETURN", "BACK", "DELETE"), "_", output),
+         ppt = as.numeric(factor(participant)),
+         across(c(location, task), str_to_lower),
+         across(location, ~ifelse(. != "revision", str_sub(., 1, nchar(.)-1), .))) %>% 
+  filter(str_detect(location, "before|within") | location %in% c("revision")) %>% 
+  select(ppt, task, output, iki, location) %>% 
+  filter(location != "before paragraph") 
+
+# Key transition ends in editing operation
+plantra <- plantra %>% 
+  mutate(is_edit = lead(location) == "revision", .by = c(ppt, task)) %>% 
+  mutate(across(is_edit, ~replace_na(., FALSE))) %>% 
+  filter(location != "revision") %>% 
+  mutate(streak = sequence(rle(is_edit)$lengths),
+         is_end_of_streak = lead(streak) < streak,
+         streak_id = lag(cumsum(is_end_of_streak) + 1),
+         across(streak_id, ~replace_na(., 1)),
+         .by = c(ppt, task)) %>% 
+  select(-is_end_of_streak, -streak, -streak_id) 
+
+write_csv(plantra, "data/plantra.csv")

### 2023-05-29 scripts/plantra/zzz/preprocessing.R

@@ -2,8 +2,6 @@ library(tidyverse)
 library(janitor)
 #source("scripts/functions.R")
 
-# plantra: translation and rewrite text to plain language; professional translators
-
 # Get log files
 files <- list.files(path = "data/WritingPhases/PlanTra_Data/data_out", 
                     full.names = TRUE);files

### 2023-05-29 scripts/spl2 (shift + C)/get_ytilde.R

@@ -1,0 +1,44 @@
+# Extract simulated data and compare to read data
+# for all four models
+
+# Load packages
+library(tidyverse)
+library(rstan)
+
+# Get functions
+source("scripts/get_ytilde.R")
+source("scripts/spl2 (shift + C)/get_data.R")
+
+# Load data
+n_sample <- 100 # number of random data points
+file <- "data/spl2.csv"
+
+# Load data
+d <- get_data(file, n_sample) %>% 
+  mutate(data_idx = 1:n())
+
+# Number of simulations per model
+nsims <- 100
+
+# Max number of iterations
+maxiter <- 20000
+
+# Sample random idx for simulations
+idx <- sample(maxiter, nsims)
+
+# File names
+(files <- list.files('stanout/spl2 (shift + C)', 
+                     pattern = ".rda$", 
+                     full.names = T))
+
+# Apply get_ytilde function
+sims <- map_dfr(files, get_ytilde, idx = idx)
+
+# Combine model predictions and data
+allsims <- left_join(d, sims, by = c("data_idx")) %>% 
+  arrange(sim_idx, data_idx) %>% 
+  rename(y_obs = iki, y_tilde = value) 
+
+# Save simulations
+write_csv(allsims, "stanout/spl2 (shift + C)/all_sims.csv")
+

### 2023-05-29 scripts/spl2 (shift + C)/lmm.R

@@ -1,22 +1,19 @@
 # Load packages
 library(tidyverse)
 library(rstan)
-source("scripts/plantra/get_data.R")
+source("scripts/spl2 (shift + C)/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
 # Sampling parameters
-n_cores <- 3
-n_chain <- 3
-iterations <- 20000
-n_samples <- 100 # number of random data points
-file <- "data/plantra.csv"
+n_cores = 3
+n_chain = 3
+iterations = 20000
+nsamples = 100
+file = "data/spl2.csv"
 
-# Load data
-d <- get_data(file, n_samples)
-
-# Check counts
-count(d, ppt, condition)
+# Load df
+d <- get_data(file, nsamples)
 
 # Data as list
 dat <- within( list(), {
@@ -30,15 +27,19 @@ dat <- within( list(), {
 
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_e= rep(0, dat$K)
-          , alpha = 600
-          , sigma = 1850
-          , sigma_u = 165)}
+  list(   alpha = 6
+          , beta_e = rep(0, dat$K)
+          , sigma = 1
+          , sigma_u = 0.1)}
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
 
+# --------------
+# Stan model ##
+# --------------
+
 # Load model
-lmm <- stan_model(file = "stan/lmmgaus.stan")
+lmm <- stan_model(file = "stan/lmm.stan")
 
 # Parameters to omit in output
 omit <- c("mu")
@@ -61,8 +62,10 @@ m <- sampling(lmm,
 
 # Save model
 saveRDS(m, 
-        file = "stanout/plantra/lmmgaus.rda",
+        file = "stanout/spl2 (shift + C)/lmm.rda",
         compress = "xz")
+
+#m <- readRDS(file = "stanout/spl2  (shift + C)/lmm.rda")
 
 # Select relevant parameters
 (param <- c("beta", "sigma", "sigma_u"))

### 2023-05-29 scripts/spl2 (shift + C)/zzz/lmm_beforeword.R

@@ -1,0 +1,105 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores = 3
+n_chain = 3
+iterations = 10000
+
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit") %>%
+  group_by(SubNo, Lang, transition_type) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences,
+         transition_type == "word_before") %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(Lang),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
+
+# Sample within each category 100 random data points per loc and ppt
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= 100) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmm.stan")
+
+# Parameters to omit in output
[…]

### 2023-05-29 scripts/spl2 (shift + C)/zzz/lmmgaus_beforesent.R

@@ -1,21 +1,32 @@
 # Load packages
 library(tidyverse)
 library(rstan)
-source("scripts/plantra/get_data.R")
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
 # Sampling parameters
-n_cores <- 3
-n_chain <- 3
-iterations <- 20000
-n_samples <- 100 # number of random data points
-file <- "data/plantra.csv"
+n_cores = 3
+n_chain = 3
+iterations = 10000
 
-# Load data
-d <- get_data(file, n_samples)
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit",
+         transition_type == "sentence_before") %>%
+  group_by(SubNo, Lang) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences) %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(Lang),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
 
-# Check counts
 count(d, ppt, condition)
 
 # Data as list
@@ -28,20 +39,28 @@ dat <- within( list(), {
   N <- nrow(d)
 } );str(dat)
 
+
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta_e= rep(0, dat$K)
-          , alpha = 600
-          , sigma = 1850
-          , sigma_u = 165)}
+  list(   beta_mu = 250
+          , beta_sigma = 100
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 50
+          , sigma_sigma = 10
+          , sigma_raw = rep(100, dat$K)
+          , sigma_u = 10)}
 
 start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
 
 # Load model
 lmm <- stan_model(file = "stan/lmmgaus.stan")
 
 # Parameters to omit in output
-omit <- c("mu")
[…]

### 2023-05-29 scripts/spl2 (shift + C)/zzz/lmmuneqvar_beforeword.R

@@ -1,0 +1,104 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores = 3
+n_chain = 3
+iterations = 10000
+
+# Load df
+d <- read_csv("data/spl2.csv") %>%
+  filter(!is.na(transition_dur), 
+         transition_dur > 50, 
+         transition_dur < 30000,
+         edit == "noedit") %>%
+  group_by(SubNo, Lang, transition_type) %>% 
+  mutate(location_count = n()) %>% 
+  group_by(SubNo) %>% 
+  mutate(enough_sentences = min(location_count) > 10) %>% # at least 10 sentences
+  ungroup() %>%
+  filter(enough_sentences,
+         transition_type == "word_before") %>% 
+  mutate(SubNo = as.numeric(factor(SubNo)),
+         condition = factor(Lang),
+         cond_num = as.integer(condition)) %>% 
+  select(ppt = SubNo, iki = transition_dur, condition, cond_num) 
+
+# Sample within each category 100 random data points per loc and ppt
+set.seed(365)
+d <- d %>% group_by(ppt, condition) %>%
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= 100) %>% 
+  ungroup() %>% 
+  filter(keep) %>% 
+  select(-keep)
+
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta_mu = 5
+          , beta_sigma = .1
+          , beta_raw = rep(0, dat$K)
+          , sigma_mu = 1
+          , sigma_sigma = .1
+          , sigma_raw = rep(0.1, dat$K)
+          , sigma_u = 0.1)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# --------------
+# Stan model  ##
+# --------------
+
+# Load model
+lmm <- stan_model(file = "stan/lmmuneqvar.stan")
+
+# Parameters to omit in output
[…]

### 2023-06-06 report/results.Rmd
add gunnexp2

@@
 \text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e}^2)\\
+  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
 \text{where: }
 		&\delta \sim \text{N}(0,1)\\
+ 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta > 0\\
@@
 	\text{where: } & \delta \sim \text{N}(0,1)\\
+	 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta > 0\\
@@
                        "Keypress following a space preceding any letter"),
-       Example = c("T^h^e c^a^t m^e^o^w^e^d. T^h^a^t[bsp][bsp]e^n i^t s^l^e^p^t.", 
-                   "The ^cat ^meowed. That[bsp][bsp]en ^it ^slept.", 
-                   "The cat meowed. ^That[bsp][bsp]en it slept.")) %>% 
+       Example = c("T\\^h\\^e c\\^a\\^t m\\^e\\^o\\^w\\^e\\^d. T\\^h\\^a\\^t[bsp][bsp]e\\^n i\\^t s\\^l\\^e\\^p\\^t.", 
+                   "The \\^cat \\^meowed. That[bsp][bsp]en \\^it \\^slept.", 
+                   "The cat meowed. \\^That[bsp][bsp]en it slept.")) %>% 
   kable(caption = "Transition location classification.") %>%
-  footnote("'^' marks transition location, [bsp] represents backspace. IKIs were timed to the shift keypress.") %>% 
+  footnote("'\\^' marks transition location, [bsp] represents backspace. IKIs were timed to the shift keypress.") %>% 
   kable_styling(position = "center", 
@@
   str_c("../stanout/", 
-  c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
+  c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
   pattern = "model.+.csv", 
@@
                 "lift" = "LIFT",
-                "c2l1" = "C2L1"),
+                "c2l1" = "C2L1",
+                "gunnexp2" = "GUNNEXP2"),
     across(model, recode, lmm = "Unimodal log-normal",
@@
   pivot_wider(names_from = dataset, values_from = contains("elpd")) %>% 
-  select(Model = model, ends_with("CATO"), ends_with("C2L1"), ends_with("LIFT"), ends_with("PLanTra"), ends_with("SPL2"), ends_with("+ C)"))
+  select(Model = model, ends_with("GUNNEXP2"), ends_with("CATO"), ends_with("C2L1"), ends_with("LIFT"), ends_with("PLanTra"), ends_with("SPL2"), ends_with("+ C)"))
 
@@
 ```{r loos, results='asis'}
-names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 6)
+names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 8)
 
@@
       caption = "Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 12))) %>%
+        align =c("l", rep("r", 14))) %>%
   kable_styling(position = "center", 
@@
   add_header_above(c(" " = 1, 
+                     "GUNNEXP2" = 2,
                      "CATO" = 2,
@@
 files <- list.files(str_c("../stanout/", 
-                          c("lift", "spl2", "plantra", "cato", "c2l1")), 
+                          c("lift", "spl2", "plantra", "cato", "c2l1", "gunnexp2")), 
                     pattern = "mogbetaun.+.csv", 
@@
                 "cato" = "CATO (non-dyslexic\nunmasked)",
+                "gunnexp2" = "GUNNEXP2 (unmasked)",
                 "spl2" = "SPL2 (L1)", 
@@
          !(str_detect(data, "CATO") & group == "dyslexic"),
-         !(str_detect(data, "CATO") & task == "masked")) %>% 
+         !(str_detect(data, "CATO") & task == "masked"),
+         !(str_detect(data, "GUNNEXP2") & xn == "masked")) %>% 
   pivot_wider(names_from = param, 
@@
                         upper = upper),
-                   .names = "{.fn}"), .by = c(data, location, name)) %>% 
+                   .names = "{.fn}"), 
+            .by = c(data, location, name)) %>% 
   filter(name %in% c("beta", "delta", "prob"))
@@
 grouplabel <- "Data set"
+shapes <- c(1, 4, 6, 7, 8, 9, 11)
 beta <- filter(ps, name == "beta") %>% 
@@
   facet_grid(~name) +
-  scale_shape_manual(values = c(1, 4, 7, 8, 9, 11)) +
+  scale_shape_manual(values = shapes) +
   scale_colour_colorblind() +
@@
 files <- list.files(str_c("../stanout/", 
-                          c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
+                          c("lift", "spl2", "spl2_shift", "gunnexp2", "plantra", "cato", "c2l1")), 
                     pattern = "mogbetaun.+.csv", 
@@
                 "spl2" = "SPL2 (L1)", 
+                "gunnexp2" = "GUNNEXP2 (unmasked)",
                 "spl2_shift" = "SPL2 (L1; shift + C)", 
@@
          !(str_detect(data, "CATO") & task == "masked"),
+         !(str_detect(data, "GUNNEXP2") & xn == "masked"),
          param %in% c("beta", "delta", "theta")) %>%
@@
                      "Slowdown for hesitations" = 2, 
-                     "Probability of hesitations" = 2))
+                     "Probability of hesitations" = 2)) %>% 
+  column_spec(1, width = "3cm")
 
@@
 
-### Posterior parameter estimates of mixture model
+### Unconstrained mixture model
+
+#### Posterior parameter estimates
+
+```{r gunnpostun, fig.cap="GUNNEXP2 (unconstrained model). Posterior parameter distribution"}
+file <- list.files(path = "../stanout/gunnexp2", 
+                   pattern = "mogbetaun.*.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% 
+  filter(param %in% c("beta", "beta2", "prob")) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta, beta2, prob)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(location, xn, name)) %>% 
+  mutate(#group = str_c("Topic: ",topic, "; ", xn))
+         group = str_c(xn))
+
+posd <- position_dodge(.35)
+dotsize <- 3
+#shapes <- 1:length(unique(ps$group))
+#shapes <- c(2, 7, 8, 9)
+shapes <- c(8, 9)
+
+beta <- filter(ps, name == "beta") %>% 
+  mutate(name = beta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank())
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank())
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(y = "",
+       x = "",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank())
+
+beta + delta + theta + plot_layout(guides = "collect")
+
+```
+
+#### Masking effect
+
+```{r shiftcellmeansgunn, results = 'asis'}
+# cell means and data set difference
+cellmeans <- ps %>% 
+  select(-group) %>% 
+  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
+  pivot_wider(names_from = name, 
+              values_from = c(mean, lower, upper)) %>% 
+  mutate(across(ends_with("prob"), ~dezero(., 2)),
+         across(c(ends_with("beta"),
+                  ends_with("delta")), 
+                ~as.character(scales::comma(round(.))))) %>% 
+  pivot_longer(-c(xn, location), 
+               names_to = c(".value", "param"),
+               names_sep = "_") %>% 
+  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
+  select(-mean:-upper) %>% 
+  pivot_wider(names_from = xn, 
+              values_from = est)
+
+tmp <- read_csv(file) %>% 
+  filter(param %in% c("beta", "delta", "theta")) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
+  pivot_wider(names_from = xn, values_from = value) %>% 
+  unnest(c(masked, unmasked)) %>% 
+  mutate(diff = masked - unmasked) %>% 
+  summarise(across(diff, list(mean = mean, 
+                              lower = lower, 
+                              upper = upper,
+                              BF = BF), 
+                   .names = "{.fn}"),
+            .by = c(location, param)) %>% 
+  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
+         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
+         across(param, recode, theta = "prob")) %>% 
+  left_join(cellmeans) %>% 
+  arrange(param, location) %>% 
+  select(param, 
+         `Transition location` = location, 
+         Unmasked = unmasked,
+         Masked = masked,
+         "Difference" = mean, BF)
+
+write_csv(tmp, "tables/masking_effect_cellmeans.csv")
+
+tmp[,-1] %>% knitr::kable(align = c("l", "r", "r", "r", "r"), 
+                          caption = "Mixture model estimates for key transitions.
+Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of disfluent transitions.
+The effect for masking is shown on log scale (for transition durations) and logit scale for probability of disfluent transitions.
+95% PIs in brackets.") %>%
+  #  cell_spec(1, 3:4) %>% 
+  kableExtra::collapse_rows(columns = 1) %>%
+  kable_styling(position = "center", full_width = T) %>%
+  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
+           threeparttable = T) %>% 
+  group_rows("Fluent transitions", 1, 3) %>%
+  group_rows("Disfluencies", 4, 6) %>%
+  group_rows("Probability of disfluencies", 7, 9)
+
+```
+
+### Cononstrained mixture model
+
+#### Posterior parameter estimates
+
+```{r}
+file <- list.files(path = "../stanout/gunnexp2", 
+                   pattern = "mogbetac.*.csv$",
+                   full.names = T)
+
+ps_beta <- read_csv(file) %>% 
+  filter(param %in% c("beta")) %>% 
+  nest(beta = value) %>% 
+  select(beta)
+
+beta <- ps_beta %>% 
+   unnest(beta) %>% 
+   mutate(across(value, exp)) %>% 
+   summarise(across(value, list(mean = mean, 
+                                lower = lower, 
+                                upper = upper), .names = "{.fn}")) %>% 
+  mutate(across(everything(), round, 0)) %>% 
+  transmute(result = str_c(mean, " msecs, PI: (", lower, ", ", upper, ")" )) %>% 
+  pull(result)
+```
+
+The posterior of the constrained model is shown in Figure \@ref(fig:gunnpost) showing the posterior slowdown for disfluent keystrokes (left panel) and the probability of disfluent keystrokes (right panel).
+Fluent keystroke transitions are distributed around a posterior mean of `r beta`.
+
+```{r gunnpost, fig.cap="GUNNEXP2 (constrained model). Posterior parameter distribution."}
+
+ps <- read_csv(file) %>% 
+  filter(param %in% c("beta2", "prob")) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  bind_cols(ps_beta) %>% 
+  unnest(cols = c(beta, beta2, prob)) %>% 
+  rename(beta = value) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>%
+  filter(name != "beta") %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(location, xn, name)) %>% 
+  mutate(#group = str_c("Topic: ",topic, "; ", xn))
+         group = str_c(xn))
+
+posd <- position_dodge(.35)
+dotsize <- 3
+#shapes <- 1:length(unique(ps$group))
+shapes <- c(8, 9)
+
+delta <- filter(ps, name == "delta") %>% 
+  mutate(name = delta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_y_log10(labels = scales::comma) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank()) 
+#  guides(colour=guide_legend(nrow=2,byrow=F),
+#         shape = guide_legend(nrow=2,byrow=F))
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = location,
+             shape = group)) +
+  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_point(aes(colour = group), 
+             size = dotsize, position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(y = "",
+       x = "Transition location",
+       colour = "",
+       shape = "") +
+  theme(axis.title.y = element_blank())
+
+delta + theta + plot_layout(guides = "collect")
+
+```
+
+#### Masking effect
+
+```{r shiftcellmeansgunn2, results = 'asis'}
+# cell means and data set difference
+cellmeans <- ps %>% 
+  select(-group) %>% 
+  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
+  pivot_wider(names_from = name, 
+              values_from = c(mean, lower, upper)) %>% 
+  mutate(across(ends_with("prob"), ~dezero(., 2)),
+         across(ends_with("delta"), ~as.character(scales::comma(round(.))))) %>% 
+  pivot_longer(-c(xn, location), 
+               names_to = c(".value", "param"),
+               names_sep = "_") %>% 
+  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
+  select(-mean:-upper) %>% 
+  pivot_wider(names_from = xn, 
+              values_from = est)
+
+tmp <- read_csv(file) %>% 
+  filter(param %in% c("delta", "theta")) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
+  pivot_wider(names_from = xn, values_from = value) %>% 
+  unnest(c(masked, unmasked)) %>% 
+  mutate(diff = masked - unmasked) %>% 
+  summarise(across(diff, list(mean = mean, 
+                              lower = lower, 
+                              upper = upper,
+                              BF = BF), 
+                   .names = "{.fn}"),
+            .by = c(location, param)) %>% 
+  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
+         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
+         across(param, recode, theta = "prob")) %>% 
+  left_join(cellmeans) %>% 
+  arrange(param, location) %>% 
+  select(param, 
+         `Transition location` = location, 
+         Unmasked = unmasked,
+         Masked = masked,
+         "Difference" = mean, BF)
+
+write_csv(tmp, "tables/masking_effect_cellmeans_constrained.csv")
+
+tmp[,-1] %>% knitr::kable(align = c("l", "r", "r", "r", "r"), 
+                          caption = "Mixture model estimates for key transitions.
+Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of disfluent transitions.
+The effect for masking is shown on log scale (for transition durations) and logit scale for probability of disfluent transitions.
+95% PIs in brackets.") %>%
+  #  cell_spec(1, 3:4) %>% 
+  kableExtra::collapse_rows(columns = 1) %>%
+  kable_styling(position = "center", full_width = T) %>%
+  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
+           threeparttable = T) %>% 
+  group_rows("Disfluencies", 1, 3) %>%
+  group_rows("Probability of disfluencies", 4, 6)
+
+```
 
@@
   pivot_longer(beta:delta) %>% 
-  group_by(location, name) %>% 
   summarise(across(value, list(mean = mean, 
                                lower = lower, 
-                               upper = upper), .names = "{.fn}"))
+                               upper = upper), .names = "{.fn}"),
+            .by = c(location, name))
 

### 2023-06-06 scripts/c2l1/mogbetaconstr.R

@@ -57,7 +57,7 @@ m <- sampling(mog,
               warmup = iterations/2,
               chains = n_chain, 
               cores = n_cores,
-              refresh = 2000,
+              refresh = 5000,
               save_warmup = FALSE, # Don't save the warmup
               include = FALSE, # Don't include the following parameters in the output
               pars = omit,

### 2023-07-17 report/manuscript.Rmd
Add manuscript and simulation

--- 2023-06-06 report/results.Rmd
@@
 ---
-title: "Modelling writing hesitations in text writing as finite mixture process"
-author: "Jens Roeser"
-date: "Compiled `r format(Sys.Date(), '%b %d %Y')`"
-output: 
-  rmdformats::downcute:
-    keep_md: true
-    self_contained: true
-    thumbnails: false # for images
-    lightbox: true
-    gallery: false
-    highlight: tango
-    use_bookdown: true # for cross references
-bibliography: references.bib
-csl: apa.csl
-link-citations: yes
+title             : "Modelling writing hesitations in text writing as finite mixture process"
+shorttitle        : "Modelling writing hesitations"
+
+csl               : "apa.csl"
+
+author: 
+  - name          : "Jens Roeser"
+    affiliation   : "1"
+    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
+    corresponding : yes 
+    email         : "jens.roeser@ntu.ac.uk"
+  - name          : "Mark Torrance"
+    affiliation   : "1"
+  - name          : "Rianne Conijn"
+    affiliation   : "2"
+  - name          : "Evgeny Chukharev-Hudilainen"
+    affiliation   : "3"
+
+affiliation:
+  - id            : "1"
+    institution   : "Department of Psychology, Nottingham Trent University, United Kingdom"
+
+  - id            : "2"
+    institution   : "Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands"
+
+  - id            : "3"
+    institution   : "Department of English, Iowa State University, Iowa"
+
+abstract: |
+
+keywords: "Keystroke modelling; finite mixture models; Bayesian models; text composition"
+
+bibliography      : ["references.bib"]
+
+documentclass     : "apa7"
+classoption       : "man"
+output            : 
+  papaja::apa6_pdf:
+    keep_tex: TRUE
+
+figsintext        : yes
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+lineno            : no
+mask              : no
+csquotes          : true
+
+header-includes:
+  - \usepackage{booktabs}
+  - \usepackage{longtable}
+  - \usepackage{graphicx}
+  - \usepackage{array}
+  - \usepackage{multirow}
+  - \usepackage{float}
+  - \usepackage{colortbl}
+  - \usepackage{threeparttable}
+  - \usepackage[normalem]{ulem}
+  - \usepackage[utf8]{inputenc}
+  - \usepackage{icomma}
+  - \usepackage{pdflscape}
+  - \newcommand{\blandscape}{\begin{landscape}}
+  - \newcommand{\elandscape}{\end{landscape}}
 ---
 
-```{r include = F}
+```{r packages, include=FALSE}
 library(tidyverse)
@@
 
-theme_set(theme_bw(base_size = 12) +
+theme_set(theme_few(base_size = 9) +
             theme(strip.background = element_blank(),
@@
 theta_label <- "Probability of hesitant\ntransitions"
+
 ```
@@
                       comment = NA,
-                      #                      dev = c("pdf", "png"), 
-                      #                     fig.keep = "all", 
-                      #                    fig.path = "figures/",
+                      cache.extra = R.version,
+                      dev = "cairo_pdf",
                       fig.align = 'center', 
-                      #                   pdf.options(encoding = "ISOLatin9.enc"),
                       fig.width = 8, 
@@
                       width=90)
+
+options(kableExtra.auto_format = FALSE)
+dev.args <- list(pdf = list(type = "cairo"))
 ```
@@
 
-```{r eval = F}
-Take intro from Mark's draft of SPL2 paper
-
-Different views on pausing in text production, largely representing typing as a unimodal or bimodal process
-
-- pauses are represented in the right tail of the distribution (unimodal) explaining the long positive skew and therefore possibly making a log transformation redundant (there are people arguing against log transformation of keystrokes but I don't understand their argument)
-- as pauses are more likely at linguistic edges, the variance of the distribution is going to be larger
-- pauses are represented as an independent distribution / process
-  - short distribution: fluent key transitions only limited by the ability to move fingers (same across text locations)
-  - short distribution: processes involved in cascade are different when lexical and syntactic planning is involved (differences across text locations)
-
-```
-
-# Model definitions
+# Methods
+
+## Statistical models
 
@@
 
-## Uni-modal Gaussian
+### Uni-modal Gaussian
 
@@
 $$
-\begin{align}
+\begin{align}(\#eq:unimodgaus)
 \text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
@@
 
-## Uni-modal log-Gaussian
+### Uni-modal log-Gaussian
 
@@
 $$
-\begin{align}
+\begin{align}(\#eq:unimodloggaus)
 \text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
@@
 
-## Uni-modal unequal-variance log-Gaussian
-
-$$
-\begin{align}
+### Uni-modal unequal-variance log-Gaussian
+
+$$
+\begin{align}(\#eq:unimoduv)
 \text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
@@
 
-## Bi-modal log-Gaussian (constrained)
+### Bi-modal log-Gaussian (constrained)
 
@@
 $$
-\begin{align}
+\begin{align}(\#eq:bimodcon)
 \text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
@@
 
-## Bi-modal log-Gaussian (unconstrained)
+### Bi-modal log-Gaussian (unconstrained)
 
@@
 $$
-\begin{align}
+\begin{align}(\#eq:bimoduncon)
 \text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
@@
 
-# Analysis
-
-We reanalysed data sets including process information from participants writing text.
-For all data sets we fit a series of four models each with random effects for participants.
-Probability functions used were normal and log-normal in line with typically treatments used in the literature, a log-normal distribution with unequal variances for model predictions, and a bimodal mixed effects model.
-Stan code for mixture models was based on @roeser2021modelling.
-Text locations (levels: before sentence, before word, within word) was included as predictor in all models.
-
-Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
-The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
-
-# Datasets
-
-## General overview
+## Data sets
 
@@
 ```{r datasets, results='asis'}
-tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT"),
+tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
                 Source = c("@ronneberg2022process",
@@
                               "@rossetti2022s",
-                              "@nina_vandermeulen_2020_3893538"),
+                              "@nina_vandermeulen_2020_3893538", 
+                           ""),
                 Keylogger = c("EyeWrite", 
@@
                            "InputLog", 
-                           "InputLog"),
+                           "InputLog",
+                           "EyeWrite"),
                 `Writing task` = c(
@@
                                  "Text simplification",
-                                 "Synthesis"),
-                 `N (ppts)` = c(126, 26*2, 39, 47, 658),
+                                 "Synthesis", 
+                                 ""),
+                 `N (ppts)` = c(126, 26*2, 39, 47, 658, 45),
 #                n_texts = c(1, 2, 2, 2, NA),
@@
                                "pre / post test trained in plain language principles and control", 
-                               "Various topics and genres"),
-                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95),
+                               "Various topics and genres",
+                               "masked / unmasked"),
+                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
                  Language = c( "Norwegian", 
@@
                                "English (L2)", 
-                               "Dutch")) %>% 
+                               "Dutch",
+                               "Norwegian")) %>% 
   kable(caption = "Datasets in brief.") %>%
@@
 
-##  Transition types
+### GUNNEXP2
+
+### C2L1
+
+The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process.
+The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
+
+TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
+
+### CATO
+
+Data are published in @torrance2016adolescent. Norwegian upper
+secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
+
+### PLanTra
+
+The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
+In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
+Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
+During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
+making it easier to read for a lay audience--by applying what they had learned from their respective training.
+The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
+It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
+
+### LIFT
+
+LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
+
+### SPL2
+
+Data are going to be published in @torrance.
+
+Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
+CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
+We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
+
+Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
+
+## Data analysis
+
+### Transition types
 
@@
 
-## Data reduction
+### Data reduction
 
 ```{r reductionfunctions}
-# Function to get data reduction infos
-get_lift <- function(file, n_samples, n_ppts){
-#  file <- "../data/lift.csv"
-# n_samples = 50
-#  n_ppts = 100
-  result <- list()
-
-  d <- read_csv(file)
-
-  full_set <- d %>% 
-    count(ppt, topic, genre) %>% 
-    summarise(n = n(), .by = ppt) %>% 
-    mutate(has_full_set = n == 4)
-
-  result[[3]] <- full_set %>% count(has_full_set)
-
-  ppt_keep <- full_set %>% 
-    filter(has_full_set) %>% 
-    pull(ppt)
-
-  result[[4]] <- d %>% 
-    filter(ppt %in% ppt_keep) %>% 
-    select(ppt, enough_sentences) %>% 
-    unique() %>% 
-    count(enough_sentences)
-
-  result[[1]] <- d %>%
-    filter(!is.na(iki), 
-           !is_edit,
-           ppt %in% ppt_keep,
-           enough_sentences) %>%
-    select(ppt, iki, location) %>% 
-    drop_na() %>% 
-    mutate(too_fast = iki <= 50, 
-           too_slow = iki >= 30000,
-           across(ppt, ~as.numeric(factor(.)))) %>% 
-    summarise(across(starts_with("too"), 
-                     list(mean = mean, 
-                          se = se_bin)),
-              .by = c(ppt)) %>% 
-    summarise(across(starts_with("too"), 
-                     list(mean = mean), 
-                     .names = "{.col}"))
-
-  # Sample within each category random data points
-  set.seed(365)
-  result[[2]] <- d %>% 
-    filter(!is.na(iki), 
-           !is_edit,
-           ppt %in% ppt_keep,
-           enough_sentences,
-           iki > 50,
-           iki < 30000) %>% 
-    mutate(keep = 1:n(),
-           keep = sample(keep),
-           keep = keep <= n_samples,
-           .by = c(ppt, location, genre, topic)) %>% 
-   summarise(across(keep, 
-             list(mean = mean, 
-                  se = se_bin)),
-              .by = c(ppt, location)) %>% 
-   summarise(across(starts_with("keep"), 
-                     list(mean = mean), 
-                     .names = "{.col}"), 
-              .by = location)
-
-   set.seed(365)
-   result[[5]] <- d %>% 
-    mutate(keep = 1:n(),
-           keep = sample(keep),
-           keep = keep <= n_samples, 
-           .by = c(ppt, location, genre, topic)) %>% 
-    filter(keep) %>% 
-    mutate(ppt_keep = ppt %in% sample(unique(ppt), n_ppts)) %>% 
-    select(ppt, ppt_keep) %>% 
-    unique() %>% 
-    count(ppt_keep)
-
-   return(result)
-}
-get_c2l1 <- function(file, n_samples){
-  #file <- "../data/c2l1.csv"
-  #n_samples <- 100
-  d <- read_csv(file) %>% 
-    filter(!is.na(iki)) %>% 
-    mutate(ppt = as.numeric(factor(subno))) %>% 
-    select(ppt, iki, location)
-
-  result <- list()
-
-  result[[1]] <- d %>% 
-    mutate(too_fast = iki <= 50,
-           too_slow = iki >= 30000) %>%
-    summarise(across(starts_with("too"), 
-                     list(mean = mean, se = se_bin)),
-              .by = c(ppt)) %>% 
-    summarise(across(starts_with("too"), 
-                     list(mean = mean), 
-                     .names = "{.col}"))
-
-  set.seed(365)
-  result[[2]] <- d %>% 
-    filter(iki > 50,
-           iki < 30000) %>% 
-    mutate(keep = 1:n(),
-           keep = sample(keep),
-           keep = keep <= n_samples,
-           .by = c(ppt, location)) %>% 
-   summarise(across(keep, 
-             list(mean = mean, se = se_bin)),
-              .by = c(ppt, location)) %>% 
-    summarise(across(starts_with("keep"), 
-                     list(mean = mean), 
-                     .names = "{.col}"), 
-              .by = location)
-
-  return(result)
-}
-get_cato <- function(file, n_samples){
-  d <- read_csv(file) %>% 
-    filter(!is.na(iki)) %>% 
-    mutate(ppt = as.numeric(factor(subno))) %>% 
-    select(ppt, iki, location, xn, dystyp)
-
-  result <- list()
-
-  result[[1]] <- d %>% 
-    mutate(too_fast = iki <= 50,
-           too_slow = iki >= 30000) %>%
-    summarise(across(starts_with("too"), 
-                     list(mean = mean, 
-                          se = se_bin)),
-              .by = c(ppt)) %>% 
-    summarise(across(starts_with("too"), 
-                     list(mean = mean), 
-                     .names = "{.col}"))
-
-  set.seed(365)
-  result[[2]] <- d %>% 
-    filter(iki > 50,
-           iki < 30000) %>% 
-    mutate(keep = 1:n(),
-           keep = sample(keep),
-           keep = keep <= n_samples,
-           .by = c(ppt, location, xn, dystyp)) %>% 
-   summarise(across(keep, 
-             list(mean = mean, 
-                  se = se_bin)),
-              .by = c(ppt, location)) %>% 
-   summarise(across(starts_with("keep"), 
-                     list(mean = mean), 
-                     .names = "{.col}"), 
-              .by = location)
-
-  return(result)
-}
-get_plantra <- function(file, n_samples){
-  d <- read_csv(file) %>% 
-    filter(!is.na(iki), !is_edit) %>% 
-    mutate(ppt = as.numeric(factor(ppt))) %>% 
-    select(ppt, iki, location, task)
-
-  result <- list()
-
-  result[[1]] <- d %>% 
-    mutate(too_fast = iki <= 50,
-           too_slow = iki >= 30000) %>%
-    summarise(across(starts_with("too"), 
-                     list(mean = mean, 
-                          se = se_bin)),
-              .by = c(ppt, task)) %>% 
-    summarise(across(starts_with("too"), 
-                     list(mean = mean), 
-                     .names = "{.col}"))
-
-  # keep only participants that did both the pre and post task.
-  keep_ppts <- select(d, ppt, task) %>% 
-    unique() %>% 
-    count(ppt) %>% 
-    mutate(both_tasks = n == 2)
-
-  result[[3]] <- count(keep_ppts, both_tasks)
-
-  keep_ppts <- keep_ppts %>% filter(both_tasks) %>% pull(ppt)
-  d <- filter(d, ppt %in% keep_ppts)
-
-  # Filter ppts with too few sentences
-  keep_ppts <- count(d, ppt, location) %>% 
-    filter(location == "before sentence") %>% 
-    mutate(enough_sentences = n >= 10)
-
-  result[[4]] <- count(keep_ppts, enough_sentences) %>% 
-    rename(more_than_10_sentences = enough_sentences)
-
-  keep_ppts <- keep_ppts %>% 
-    filter(enough_sentences) %>% 
-    pull(ppt)
-
-  d <- filter(d, ppt %in% keep_ppts)
-
-  set.seed(365)
-  result[[2]] <- d %>% 
-    filter(iki > 50,
-           iki < 30000) %>% 
-    mutate(keep = 1:n(),
-           keep = sample(keep),
-           keep = keep <= n_samples,
-           .by = c(ppt, location, task)) %>% 
-   summarise(across(keep, 
-             list(mean = mean, 
-                  se = se_bin)),
-              .by = c(ppt, location)) %>% 
-   summarise(across(starts_with("keep"), 
-                     list(mean = mean), 
-                     .names = "{.col}"), 
-              .by = location)
-  return(result)
-}
-get_spl2 <- function(file, n_samples){
-  result <- list()
-#  file <- "../data/spl2.csv"
-  d <- read_csv(file) %>% 
-    select(-transition_dur) %>% 
-    rename(iki = transition_dur_to_mod,
-           location = transition_type) %>% 
-    filter(!is.na(iki), 
-           edit == "noedit") %>% 
-    select(ppt = SubNo, iki, location, Lang) %>% 
-    drop_na()
-
-  result[[1]] <- d %>% 
-    mutate(too_short = iki <= 50,
-           too_fast = iki >= 30000,
-           across(ppt, ~as.numeric(factor(.)))) %>% 
-    summarise(across(starts_with("too"), 
-                     list(mean = mean, 
-                          se = se_bin)),
-              .by = c(ppt)) %>% 
-    summarise(across(starts_with("too"), 
-                     list(mean = mean), 
-                     .names = "{.col}"))
-
-  d <- d %>% filter(iki > 50, iki < 30000)
-
-  # Filter ppts with too few sentences
-  keep_ppts <- count(d, ppt, location) %>% 
-    filter(location == "sentence_before") %>% 
-    mutate(enough_sentences = n >= 10)
-
-  result[[3]] <- count(keep_ppts, enough_sentences) %>% 
-    rename(more_than_10_sentences = enough_sentences)
-
-  keep_ppts <- keep_ppts %>% 
-    filter(enough_sentences) %>% 
-    pull(ppt)
-
-  d <- filter(d, ppt %in% keep_ppts)
-
-  set.seed(365)
-  result[[2]] <- d %>% 
-    mutate(keep = 1:n(),
-           keep = sample(keep),
-           keep = keep <= n_samples,
-           .by = c(ppt, location, Lang)) %>% 
-   summarise(across(keep, 
-             list(mean = mean, 
-                  se = se_bin)),
-              .by = c(ppt, location)) %>% 
-   summarise(across(starts_with("keep"), 
-                     list(mean = mean), 
-                     .names = "{.col}"), 
-              .by = location)
-
-  return(result)
-}
+source("scripts/get_data_summary.R")
 ```
@@
 spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
-
-```
-
-```{r eval = F}
-Data reduction using random samples of subsets of participants for larger data sets
-
-Also for larger data sets, only used participants that provided a min number of sentences
-
-Removed transitions that were followed by an editing operation
-
-Extremely short and long intervals
-
-Add a description with general data filtering and a table with how many data were removed from the analysis by study.
-
-Add final sample number of ppt, and transitions by location
-
-```
-
-# Out-of-samples cross-validation
+gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
+```
+
+For all datasets we only used transitions that were not followed by an editing operation.
+
+We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
+We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant)
+
+We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \@ref(tab:datareduction).
+
+From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
+This was done for computational reasons to reduce the time the Bayesian models need to complete.
+For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis.
+Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant.
+The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \@ref(tab:datareduction).
+
+```{r datareduction, results='asis'}
+c2l1$ds <- "c2l1"
+cato$ds <- "cato"
+lift$ds <- "lift"
+plantra$ds <- "plantra"
+spl2$ds <- "spl2"
+gunnexp2$ds <- "gunnexp2"
+
+extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
+          mutate(ds = .x$ds)) %>% 
+  mutate(across(where(is.numeric), ~round(.*100, 2)),
+         across(starts_with("too"), ~str_c(., "%"))) %>% 
+  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
+  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
+  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
+  rename(`$\\le$ 50 msecs` = `<50ms`,
+         `$\\ge$ 30 secs` = `>30,000ms`)
+
+random_sample <- 
+  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
+          mutate(ds = .x$ds)) %>% 
+    mutate(across(where(is.numeric), ~round(.*100, 1)),
+           across(starts_with("keep"), ~str_c(., "%"))) %>%
+    unite("keep", starts_with("keep"), sep = " (") %>% 
+    mutate(across(c(keep), ~str_c(., ")"))) %>%
+  pivot_wider(names_from = location, values_from = keep)
+
+datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
+  relocate(ds) %>% 
+  mutate(across(ds, recode, 
+                "cato" = "CATO",
+                "spl2" = "SPL2", 
+                "spl2_shift" = "SPL2 (shift + C)", 
+                "plantra" = "PLanTra",
+                "lift" = "LIFT",
+                "c2l1" = "C2L1",
+                "gunnexp2" = "GUNNEXP2")) %>% 
+  rename(`Dataset` = ds)
+
+kable(datareduction, 
+      caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition locattion. Standard error is shown in parentheses.",
+        align =c("l", rep("r", 14))) %>%
+  kable_styling(position = "center", 
+                full_width = T, 
+                font_size = 11) %>%
+  add_header_above(c(" " = 1, 
+                     "Extreme values" = 2,
+                     "Randomly sampled data" = 3)) %>%
+  column_spec(1, width = "8em")
+
+```
+
+### Bayesian modelling
+
+Keystroke data from text composition tasks were modelled in a series of five models. An overview of all models can be found in Table \ref{tab:models}.
+
+```{r models, results = 'asis'}
+models <- tibble(Models = paste0("M",1:4),
+       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
+       Equation = paste0("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "}"),
+       Description = c("Gaussian distribution",
+                       "Log-normal distribution",
+                       "Log-normal distribution and unequal variances",
+                       "Fluent transition are not allowed to vary by transition location",
+                       "Fluent transition are allowed to vary by transition location"))
+
+papaja::apa_table(models, align = c("l", "l", "l"), escape = FALSE, digits = 0, placement = "bp!",
+                  caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
+                  note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
+```
+
+each with random effects for participants.
+Probability functions used were normal and log-normal in line with typically treatments used in the literature, a log-normal distribution with unequal variances for model predictions, and a bimodal mixed effects model.
+Stan code for mixture models was based on @roeser2021modelling.
+Text locations (levels: before sentence, before word, within word) was included as predictor in all models.
+
+Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
+The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
+Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
+Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
+
+# Results
+
+## Out-of-samples cross-validation
 
@@
 
-# Cross-data set comparisons
+## Cross-data set comparisons
 
@@
 
-## Cross-data set visualisation
-
 The model estimates for the mixture-model with the highest predictive performance are shown in Figure \@ref(fig:crossstudypost).
@@
 
-## Effect of transition location
-
-It's generally believed that pausing is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges, i.e.
+It's generally agreed that pausing is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges, i.e.
 before sentence > before word > within word.
@@
 
-```{r eval = F}
-
-#
-# Do I really need all this?
-#
-#
-#
-#
-
-# use posterior to calculate interactions, cell means and dataset differences
-files <- list.files(str_c("../stanout/", 
-                          c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
-                    pattern = "mogbetaun.+.csv", 
-                    full.names = T)
-
-plantra <- read_csv(files[1], id = "data") %>% 
-  pivot_wider(names_from = c(genre, topic), values_from = value) %>% 
-  unnest(cols = c(ARG_A, ARG_B, ARG_C, ARG_D, INF_A, INF_B, INF_C, INF_D)) %>% 
-  rowwise() %>% 
-  mutate(value = mean(c_across(contains("_")))) %>% 
-  ungroup() %>% 
-  select(param, location, value, data)
-
-spl2 <- read_csv(files[2], id = "data") %>% 
-  pivot_wider(names_from = c(lang), values_from = value) %>% 
-  unnest(cols = c(EN, ES)) %>% 
-  rowwise() %>% 
-  mutate(value = mean(c_across(c(EN, ES)))) %>% 
-  ungroup() %>% 
-  select(param, location, value, data)
-
-lift <- read_csv(files[3], id = "data") %>% 
-  pivot_wider(names_from = c(task), values_from = value) %>% 
-  unnest(cols = c(`pre-test`, `post-test`)) %>% 
-  rowwise() %>% 
-  mutate(value = mean(c_across(contains("-test")))) %>% 
-  ungroup() %>% 
-  select(param, location, value, data)
-
-cato <- read_csv(files[4], id = "data") %>% 
-  pivot_wider(names_from = c(task, group), values_from = value) %>% 
-  unnest() %>% 
-  rowwise() %>% 
-  mutate(value = mean(c_across(contains("_")))) %>% 
-  ungroup() %>% 
-  select(param, location, value, data)
-
-c2l1 <- read_csv(files[5], id = "data") %>% 
-  select(param, location, value, data)
-
-ps <- bind_rows(plantra, spl2, lift, cato, c2l1) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(across(data, 
-                str_remove_all, 
-                "\\.|\\/|stanout|\\mog.csv"),
-         theta = logit_scaled(prob)) %>% 
-  select(-prob) %>%
-  pivot_longer(beta:theta) %>%
-  pivot_wider(names_from = c(data, location), 
-              values_from = value, 
-              names_sep = "_") %>% 
-  unnest(contains("_"))
-
-# Get main effects, interactions, nested effects
-ps_effects <- ps %>%
-  rowwise() %>%
-  mutate(# Main effects
-         DS_1 = mean(c_across(starts_with("lift_"))) - 
-                mean(c_across(starts_with("spl2_"))),
-         DS_2 = mean(c_across(starts_with("lift_"))) - 
-                mean(c_across(starts_with("plantra_"))),
-         DS_3 = mean(c_across(starts_with("spl2_"))) - 
-                mean(c_across(starts_with("plantra_"))),
-         DS_4 = mean(c_across(starts_with("lift_"))) - 
-                mean(c_across(starts_with("cato_"))),
-         DS_5 = mean(c_across(starts_with("plantra_"))) - 
-                mean(c_across(starts_with("cato_"))),
-         DS_6 = mean(c_across(starts_with("spl2_"))) - 
-                mean(c_across(starts_with("cato_"))),
-         DS_7 = mean(c_across(starts_with("c2l1_"))) - 
-                mean(c_across(starts_with("plantra_"))),
-         DS_8 = mean(c_across(starts_with("c2l1_"))) - 
-                mean(c_across(starts_with("cato_"))),
-         DS_9 = mean(c_across(starts_with("c2l1_"))) - 
-                mean(c_across(starts_with("spl2_"))),
-         DS_10 = mean(c_across(starts_with("c2l1_"))) - 
-                mean(c_across(starts_with("lift_"))),
-
-         TL_1 =  mean(c_across(contains(" sentence"))) - 
-                mean(c_across(contains("_before word"))),
-         TL_2 =  mean(c_across(contains("_before"))) - 
-                mean(c_across(contains("_within word"))),
-         # Two-way interactions
-         `DS_1:TL_1` = (`lift_before sentence` - `lift_before word`) -
-                       (`spl2_before sentence` - `spl2_before word`),
-
-         `DS_2:TL_1` = (`lift_before sentence` - `lift_before word`) -
-                       (`plantra_before sentence` - `plantra_before word`),
-
-         `DS_3:TL_1` = (`spl2_before sentence` - `spl2_before word`) -
-                       (`plantra_before sentence` - `plantra_before word`),
-
-         `DS_4:TL_1` = (`lift_before sentence` - `lift_before word`) -
-                       (`cato_before sentence` - `cato_before word`),
-
-         `DS_5:TL_1` = (`plantra_before sentence` - `plantra_before word`) - 
-                       (`cato_before sentence` - `cato_before word`),
-
-         `DS_6:TL_1` = (`spl2_before sentence` - `spl2_before word`) -
-                       (`cato_before sentence` - `cato_before word`),
-
-         `DS_7:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
-                       (`plantra_before sentence` - `plantra_before word`),
-
-         `DS_8:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
-                       (`cato_before sentence` - `cato_before word`),
-
-         `DS_9:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
-                       (`spl2_before sentence` - `spl2_before word`),
-
-         `DS_10:TL_1` = (`c2l1_before sentence` - `c2l1_before word`) -
-                        (`lift_before sentence` - `lift_before word`),
-
-         `DS_1:TL_2` = (mean(c_across(starts_with("lift_") & contains("_before"))) - 
-                        mean(c_across(starts_with("spl2_") & contains("_before")))) -
-                       (`lift_within word` - `spl2_within word`),
-
-         `DS_2:TL_2` = (mean(c_across(starts_with("lift_") & contains("_before"))) - 
-                        mean(c_across(starts_with("plantra_") & contains("_before")))) -
-                       (`lift_within word` - `plantra_within word`),
-
-         `DS_3:TL_2` = (mean(c_across(starts_with("spl2_") & contains("_before"))) - 
-                        mean(c_across(starts_with("plantra_") & contains("_before")))) -
-                       (`spl2_within word` - `plantra_within word`),
-
-         `DS_4:TL_2` = (mean(c_across(starts_with("lift_") & contains("_before"))) - 
-                        mean(c_across(starts_with("cato_") & contains("_before")))) -
-                       (`lift_within word` - `cato_within word`),
-
-         `DS_5:TL_2` = (mean(c_across(starts_with("plantra_") & contains("_before"))) - 
-                        mean(c_across(starts_with("cato_") & contains("_before")))) -
-                       (`plantra_within word` - `cato_within word`),
-
-         `DS_6:TL_2` = (mean(c_across(starts_with("spl2_") & contains("_before"))) - 
-                        mean(c_across(starts_with("cato_") & contains("_before")))) -
-                       (`spl2_within word` - `cato_within word`),
-
-         `DS_7:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
-                        mean(c_across(starts_with("plantra_") & contains("_before")))) -
-                       (`c2l1_within word` - `plantra_within word`),
-         `DS_8:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
-                        mean(c_across(starts_with("cato_") & contains("_before")))) -
-                       (`c2l1_within word` - `cato_within word`),
-         `DS_9:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
-                        mean(c_across(starts_with("spl2_") & contains("_before")))) -
-                       (`c2l1_within word` - `spl2_within word`),
-         `DS_10:TL_2` = (mean(c_across(starts_with("c2l1_") & contains("_before"))) - 
-                        mean(c_across(starts_with("lift_") & contains("_before")))) -
-                       (`c2l1_within word` - `lift_within word`)
-         ) %>%
-  ungroup() %>%
+## Simulation
+
+A general concern with mixture models is that in principle, as the mixture model has more parameters it might simply always lead to a better fit, even though cross-validation is addressing potential problems with overfitting models.
+
+To address this concern we simulated two sets of data.
+Both data sets have two conditions and 1,000 observations each.
+The first set of data has as underlying data generating process a mixture model with two mixture components similar to the process described above.
+The difference between the two conditions is that the mixing proportion is larger for condition 2 than for condition 1, hence long observations are more likely in condition 2.
+This model can be summarised as followed:
+
+$$
+\begin{align}(\#eq:simmog)
+\text{y}_i \sim\text{ } & \theta_\text{condition[i]} \cdot \text{logN}(\mu_1, \sigma^2_1) +\\
+& (1 - \theta_\text{condition[i]}) \cdot \text{logN}(\mu_2, \sigma^2_2)
+\end{align}
+$$
+
+The second data set was generated with an unequal variance unimodal model as data generating process.
+Condition 2 has a larger mean and standard deviation than condition 1.
+The model can be summaried as followed:
+
+$$
+\begin{align}(\#eq:simuv)
+\text{y}_i \sim\text{ }& \text{logN}(\mu_\text{condition[i]}, \sigma^2_\text{condition[i]})
+\end{align}
+$$
+
+The true parameter values used for each of the two data simulations can be found in Table \@ref(tab:simparam).
+The simulated data are visualised in Figure \@ref(fig:simdata).
+The data are simulated to be similarly distributed to keystroke transitions.
+
+```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=4}
+files <- list.files("../mixture-model-sim/data", full.names = T)
+
+map_dfr(files, ~read_csv(.) %>% 
+          mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
+  mutate(across(condition, factor),
+         across(condition, ~str_c("Condition: ", condition)),
+         across(Dataset, ~recode(., mogdata = "bimodal", uvlmdata = "unimodal"))) %>% 
+  ggplot(aes(x = value, colour = condition, fill = condition)) +
+  geom_density(alpha = .25) +
+  facet_grid(~Dataset, labeller = label_both) +
+  scale_colour_colorblind() + 
+  scale_fill_colorblind() +
+  labs(colour = "", fill = "", x = "Simulated data", y = "Density")
+
+```
+
+We run 4 models: 2 mixture models, one on the data generated with a mixture process and one on the data generated with the unimodal unequal variance process.
+We repeated the same using an unimodal unequal variance model.
+Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
+The Stan models uncovered the model parameters of their respective data sets successfully, as shown in Table \@ref(tab:simparam), not less so when the model was applied to data generated with the other underlying process.
+
+```{r}
+mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
+  mutate(data = "mog")
+mog2 <- read_csv("../mixture-model-sim/stanout/mog_uvlmdata.csv") %>% 
+  mutate(data = "uvlm")
+uvlm1 <- read_csv("../mixture-model-sim/stanout/uvlm_uvlmdata.csv") %>% 
+  mutate(data = "uvlm")
+uvlm2 <- read_csv("../mixture-model-sim/stanout/uvlm_mogdata.csv") %>% 
+  mutate(data = "mog")
+
+mog_sim <- bind_rows(mog1, mog2) %>% 
+  summarise(across(value, list(est = mean,
+                               lo = lower,
+                               up = upper),
+                   .names = "{.fn}"),
+            .by = c(name, data)) %>% 
+  mutate(across(where(is.numeric), round, 2),
+         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
+         across(est, ~str_c(., " [", lo, ", ", up, "]")),
+         across(name, ~str_c("\\", .)),
+         across(name, ~str_replace(., "a_e", "a^2_1")),
+         across(name, ~str_replace(., "ap_e", "a^2_2")),
+         across(name, ~str_replace(., "\\[1\\]", "_\\\\text{condition=1}")),
+         across(name, ~str_replace(., "\\[2\\]", "_\\\\text{condition=2}"))) %>%
+  select(-lo, -up) %>% 
+  pivot_wider(names_from = data, values_from = est) %>% 
+  mutate(true = c(5, 1, .1, .4, .25, .5),
+         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
+         across(name, ~str_replace(., "prob", "theta"))) %>% 
+  relocate(name, true)
+
+uvlm_sim <- bind_rows(uvlm1, uvlm2) %>% 
+  summarise(across(value, list(est = mean,
+                               lo = lower,
+                               up = upper),
+                   .names = "{.fn}"),
+            .by = c(name, data)) %>% 
+  mutate(across(where(is.numeric), round, 2),
+         across(est, ~str_c(., " [", lo, ", ", up, "]")),
+         across(name, ~str_c("\\", .)),
+         across(name, ~str_replace(., "a_e", "a^2_1")),
+         across(name, ~str_replace(., "ap_e", "a^2_2")),
+         across(name, ~str_replace(., "\\[1\\]", "_\\\\text{condition=1}")),
+         across(name, ~str_replace(., "\\[2\\]", "_\\\\text{condition=2}"))) %>% 
+  select(-lo, -up) %>%
+  pivot_wider(names_from = data, values_from = est) %>% 
+  mutate(true = as.character(c(5, 6, .25, .5))) %>% 
+  relocate(name, true)
+
+```
+
+```{r simparam, results='asis'}
+table <- bind_rows(
+  mog_sim %>% mutate(model = "mog"),
+  uvlm_sim %>% mutate(model = "uvlm")) %>% 
+  relocate(model) %>% 
+  mutate(across(name, ~str_c("$", ., "$")))
+
+names(table) <- c("Model", "Parameter", "True value", "Bimodal data", "Unimodal data")
+
+kable(table[,-1], 
+      caption = "Uncovered parameter estimates with 95% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
+      align =c("l", rep("r", 3))) %>%
+  kable_styling(position = "center", 
+                full_width = F, 
+                font_size = 11) %>% 
+  add_header_above(c(" " = 1, 
+                     " " = 1,
+                     "Estimate with 95\\% PI" = 2)) %>%
+  pack_rows("Model: Bimodal mixture model", 1, 6) %>% 
+  pack_rows("Model: Unimodal process", 7, 10)
+
+```
+
+We used LOO-CV to compare the fit of the two models for each data set.
+The model comparisons can be found for each data generating process in Table \ref@(tab:loossim).
+The results show that the mixture model does not always lead to higher predictive performance.
+Indeed, the mixture model showed a lower predictive performance for the data that were generated with a unimodal process.
+However, for the data generated with a bimodal process, the mixture model model shows a higher predictive performance.
+In fact, the ratio of $\Delta\widehat{elpd}$ and its SE, as metric for the strength of evidence, shows that the mixture model performs 11 times better than the unimodal model for the data generated with a bimodal process.
+In comparison, for the unimodal data, the unimodal model shows only 3 times better than the bimodal mixture model.
+Thus, even though the mixture model does not necessarily perform better for non-bimodal data but it also doesn't necessarily perform much worse.
+This contrast is likely a reflection of the increased number of parameters in the mixture model.
+
+```{r loossim, results='asis'}
+table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
+  select(1:6) %>% 
   mutate(
-    # Locations effecs
-    lift_sent = `lift_before sentence` - `lift_before word`,
-    lift_word = `lift_before word` - `lift_within word`,
-    spl2_sent = `spl2_before sentence` - `spl2_before word`,
-    spl2_word = `spl2_before word` - `spl2_within word`,
-    cato_sent = `cato_before sentence` - `cato_before word`,
-    cato_word = `cato_before word` - `cato_within word`,
-    c2l1_sent = `c2l1_before sentence` - `c2l1_before word`,
-    c2l1_word = `c2l1_before word` - `c2l1_within word`,
-    plantra_sent = `plantra_before sentence` - `plantra_before word`,
-    plantra_word = `plantra_before word` - `plantra_within word`
-    # data set effects
-         # before_word_lift_spl2 = `lift_before word` - `spl2_before word`,
-         # before_word_lift_plantra = `lift_before word` - `plantra_before word`,
-         # before_word_spl2_plantra = `spl2_before word` - `plantra_before word`,
-         # before_word_cato_spl2 = `cato_before word` - `spl2_before word`,
-         # before_word_cato_plantra = `cato_before word` - `plantra_before word`,
-         # before_word_cato_lift = `cato_before word` - `lift_before word`,
-         # before_word_c2l1_plantra = `c2l1_before word` - `plantra_before word`,
-         # before_word_c2l1_cato = `c2l1_before word` - `cato_before word`,
-         # before_word_c2l1_spl2 = `c2l1_before word` - `spl2_before word`,
-         # before_word_c2l1_lift = `c2l1_before word` - `lift_before word`,
-         # 
-         # before_sentence_lift_spl2 = `lift_before sentence` - `spl2_before sentence`,
-         # before_sentence_lift_plantra = `lift_before sentence` - `plantra_before sentence`,
-         # before_sentence_spl2_plantra = `spl2_before sentence` - `plantra_before sentence`,
-         # before_sentence_cato_spl2 = `cato_before sentence` - `spl2_before sentence`,
-         # before_sentence_cato_plantra = `cato_before sentence` - `plantra_before sentence`,
-         # before_sentence_cato_lift = `cato_before sentence` - `lift_before sentence`,
-         # before_sentence_c2l1_plantra = `c2l1_before sentence` - `plantra_before sentence`,
-         # before_sentence_c2l1_cato = `c2l1_before sentence` - `cato_before sentence`,
-         # before_sentence_c2l1_spl2 = `c2l1_before sentence` - `spl2_before sentence`,
-         # before_sentence_c2l1_lift = `c2l1_before sentence` - `lift_before sentence`,
-         # 
-         # within_word_lift_spl2 = `lift_within word` - `spl2_within word`,
-         # within_word_lift_plantra = `lift_within word` - `plantra_within word`,
-         # within_word_spl2_plantra = `spl2_within word` - `plantra_within word`,
-         # within_word_cato_spl2 = `cato_within word` - `spl2_within word`,
-         # within_word_cato_plantra = `cato_within word` - `plantra_within word`,
-         # within_word_cato_lift = `cato_within word` - `lift_within word`,
-         # within_word_c2l1_plantra = `c2l1_within word` - `plantra_within word`,
-         # within_word_c2l1_cato = `c2l1_within word` - `cato_within word`,
-         # within_word_c2l1_spl2 = `c2l1_within word` - `spl2_within word`,
-         # within_word_c2l1_lift = `c2l1_within word` - `lift_within word`
-    ) %>%
-  rename(parameter = name)
-
-write_csv(ps_effects, "tables/ps_effects_acrossdata.csv")
-
-#conda <- c(2, 3, 4, 1)
-#condb <- c(4, 5, 6, 1)
-#condc <- c(1, 2, 3, 4)
-#condd <- c(4, 3, 2, 1)
-
-#me1 <- (conda + condb)/2 - (condc + condd)/2
-#me2 <- (conda + condc)/2 - (condb + condd)/2
-#int <- (conda - condc) - (condb - condd)
-
-```
-
-```{r eval = F}
-ps_effects <- read_csv("tables/ps_effects_acrossdata.csv")
-
-# Main effects and interaction summary
-ps_effects_summary <- ps_effects %>% 
-#  mutate(across(-parameter, scale)) %>% 
-  select(parameter,
-         starts_with("DS"), 
-         starts_with("TL")) %>%
-  pivot_longer(-parameter) %>%
-  group_by(parameter, name) %>%
-  summarise(
-    across(value, list(est = mean,
-                       lower = lower,
-                       upper = upper,
-                       BF = BF), .names = "{.fn}")) %>%
-  ungroup() %>%
-  mutate(across(where(is.numeric), round, 2),
-#         across(est, str_c, " [", lower, " -- ", upper, "]"),
-         across(BF, ~ifelse( . > 100, "> 100", as.character(.))),
-         across(name, recode_factor, 
-                DS_1 = "Dataset 1 (LIFT, SPL2)",
-                DS_2 = "Dataset 2 (LIFT, PLanTra)",
-                DS_3 = "Dataset 3 (SPL2, PLanTra)",
-                DS_4 = "Dataset 4 (CATO, SPL2)",
-                DS_5 = "Dataset 5 (CATO, PLanTra)",
-                DS_6 = "Dataset 6 (CATO, LIFT)",
-                DS_7 = "Dataset 7 (C2L1, PLanTra)",
-                DS_8 = "Dataset 8 (C2L1, CATO)",
-                DS_9 = "Dataset 9 (C2L1, SPL2)",
-                DS_10 = "Dataset 10 (C2L1, LIFT)",
-
-                TL_1 = "Location 1 (before sentence, before word)",
-                TL_2 = "Location 2 (before word / sentence, within word)",
-               `DS_1:TL_1` = "Dataset 1 : Location 1",
-               `DS_1:TL_2` = "Dataset 1 : Location 2",
-               `DS_2:TL_1` = "Dataset 2 : Location 1",
-               `DS_2:TL_2` = "Dataset 2 : Location 2",
-               `DS_3:TL_1` = "Dataset 3 : Location 1",
-               `DS_3:TL_2` = "Dataset 3 : Location 2",
-               `DS_4:TL_1` = "Dataset 4 : Location 1",
-               `DS_4:TL_2` = "Dataset 4 : Location 2",
-               `DS_5:TL_1` = "Dataset 5 : Location 1",
-               `DS_5:TL_2` = "Dataset 5 : Location 2",
-               `DS_6:TL_1` = "Dataset 6 : Location 1",
-               `DS_6:TL_2` = "Dataset 6 : Location 2",
-               `DS_7:TL_1` = "Dataset 7 : Location 1",
-               `DS_7:TL_2` = "Dataset 7 : Location 2",
-               `DS_8:TL_1` = "Dataset 8 : Location 1",
-               `DS_8:TL_2` = "Dataset 8 : Location 2",
-               `DS_9:TL_1` = "Dataset 9 : Location 1",
-               `DS_9:TL_2` = "Dataset 9 : Location 2",
-               `DS_10:TL_1` = "Dataset 10 : Location 1",
-               `DS_10:TL_2` = "Dataset 10 : Location 2",
-               .ordered = T)) %>%
-#  select(-lower,-upper) %>%
-  pivot_wider(names_from = parameter, values_from = c(est,lower,upper,BF)) %>%
-  arrange(name) %>% 
-  select(Predictor = name, ends_with("beta"), ends_with("delta"), ends_with("theta"))
-
-```
-
-```{r effects, results='asis', eval = F}
-ps_effects_summary %>%
-  mutate(est_1 = "",
-         est_2 = "",
-         est_3 = "") %>%
-  select(Predictor, est_1, BF_beta, est_2, BF_delta, est_3, BF_theta) %>%
-  knitr::kable(align =c("l", rep("c", 6)), 
-               col.names = c("Predictor", rep(c("Est.", "BF"), 3)),
-               caption = "Mixture model results of transition duration with predictor estimates (main effects, interactions) for the distribution of short and hesitant transition durations (on log scale) and the probability of hesitant transitions (on logit scale).
-Estimates are shown with 95% PI.
-Dotted vertical line indicates 0.") %>%
-  kable_styling(position = "center", full_width = T, font_size = 12) %>%
-  column_spec(2, image = spec_pointrange(
-    x = ps_effects_summary$est_beta, 
-    xmin = ps_effects_summary$lower_beta, 
-    xmax = ps_effects_summary$upper_beta, 
-    vline = 0)
-    ) %>% 
-  column_spec(4, image = spec_pointrange(
-    x = ps_effects_summary$est_delta, 
-    xmin = ps_effects_summary$lower_delta, 
-    xmax = ps_effects_summary$upper_delta, 
-    vline = 0)
-  ) %>% 
-  column_spec(6, image = spec_pointrange(
-    x = ps_effects_summary$est_theta, 
-    xmin = ps_effects_summary$lower_theta, 
-    xmax = ps_effects_summary$upper_theta, 
-    vline = 0)
-  ) %>% 
-  footnote("Colon indicates interactions. PI is the probability interval. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
-           threeparttable = T) %>%
-  add_header_above(c("", 
-                     "Short transition duration" = 2, 
-                     "Slowdown for hesitant transitions" = 2, 
-                     "Probability of hesitant transitions" = 2))  %>%
-  group_rows("Main effects", 1, 12) %>%
-  group_rows("Two-way interactions", 13, 32)  %>% 
-  column_spec(1, width = "5.5cm") %>% 
-  column_spec(2:7, width = "3cm")
-
-```
-
-```{r eval = F}
-ps_effects <- read_csv("tables/ps_effects_acrossdata.csv")
-
-cellmeans <- ps_effects %>% 
-  select(parameter, contains(" ")) %>%
-  pivot_longer(-parameter) %>%
-  pivot_wider(names_from = parameter, values_from = value) %>%
-  unnest() %>%
-  mutate(delta = exp(beta + delta) - exp(beta),
-         across(beta, exp),
-         across(theta, ilogit)) %>%
-  pivot_longer(cols = beta:theta, names_to = "param") %>% 
-  group_by(name, param) %>%
-  summarise(across(value, 
-                   list(mean = mean, 
-                        lower = lower, 
-                        upper = upper),
-                   .names = "{.fn}")) %>%
-  ungroup() %>% 
-  separate(name, into = c("data", "loc"), sep = "_") %>%
-  mutate(across(data, recode, 
-                "lift" = "LIFT", 
-                "spl2" = "SPL2", 
-                "plantra" = "PLanTra",
-                "cato" = "CATO",
-                "c2l1" = "C2L1")) %>% 
-  pivot_wider(names_from = param, values_from = c(mean, lower, upper)) %>%
-  mutate(across(c(ends_with("beta"), ends_with("delta")), round, 0),
-         across(c(ends_with("beta"), ends_with("delta")), scales::comma, accuracy = 1),
-         across(contains("theta"), dezero, 2),
-         across(everything(), as.character)) %>%
-  pivot_longer(contains("_"), 
-               names_to = c(".value", "parameter"), 
-               names_sep ="_") %>%
-#  mutate(across(mean, str_c, " [", lower, " -- ", upper, "]")) %>%
-#  select(-lower,-upper) %>% 
-  pivot_wider(names_from = loc, values_from = c(mean, lower, upper))
-
-tmp <- ps_effects %>% 
-  select(parameter, ends_with("_sent"), ends_with("_word")) %>%
-  pivot_longer(-parameter) %>%
-  group_by(parameter, name) %>%
-  summarise(across(value, 
-                   list(mean = mean, 
-                        lower = lower, 
-                        upper = upper,
-                        BF = BF),
-                   .names = "{.fn}")) %>%
-  ungroup() %>%
-  mutate(across(where(is.numeric), round, 2),
-#         across(mean, str_c, " [", lower, " -- ", upper, "]"),
-         across(BF, ~ifelse(.>100, "> 100", as.character(.)))) %>%
-  rename(diff = mean) %>% 
-#  select(-lower,-upper) %>% 
-  separate(name, into = c("data", "loc")) %>% 
-  mutate(across(data, recode, 
-                "lift" = "LIFT", 
-                "spl2" = "SPL2", 
-                "plantra" = "PLanTra",
-                "cato" = "CATO",
-                "c2l1" = "C2L1"),
-         across(loc, recode, 
-                "sent" = "before sentence - before word",
-                "word" = "before word - within word")) %>% 
-  right_join(cellmeans) %>% 
-  arrange(parameter, data, loc) %>% 
-  select(parameter, data, #`Transition location` = loc, 
-         ends_with("before sentence"), 
-         ends_with("before word"), 
-         ends_with("within word"),
-         Contrasts = loc, diff, lower, upper, BF)
-
-write_csv(tmp, "tables/cellmeans_transdur.csv")
-```
-
-```{r cellmeans, results = 'asis', eval = F}
-tmp <- read_csv("tables/cellmeans_transdur.csv")
-
-cellmeans <- tmp %>% select(1:2, 
-                            starts_with("mean_"), 
-                            starts_with("lower_"),
-                            starts_with("upper_")) %>% 
-  pivot_longer(-1:-2) %>% 
-  unique() %>% 
-  arrange(parameter, data) %>% 
-  select(-parameter, -data)
-
-table <- tmp %>% select(-contains("_")) %>% 
-  arrange(parameter, Contrasts, lower) %>% 
-  pivot_wider(names_from = parameter, 
-              values_from = c(diff, lower, upper, BF))
-
-table %>% 
-  mutate(est_1 = "",
-         est_2 = "",
-         est_3 = "") %>%  
-  select(data, #Contrasts, 
-         est_1, BF_beta,
-         est_2, BF_delta,
-         est_3, BF_theta) %>% 
-  knitr::kable(
-  align =c("l", rep("c", 6)),
-  col.names = c("Data set", rep(c("Est.", "BF"),3)),
-  caption = "By-data set differences between transition locations for transition duration estimates inferred from mixture model.
-Differences are shown on the log scale (for durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.
-Dotted line indicates 0.") %>%
-#  collapse_rows(columns = 1, valign = "middle") %>% 
+#    elpd_se_ratio = abs(elpd_diff / se_diff),
+    across(where(is.numeric), round, 0),
+    across(where(is.numeric), format, big.mark = ","),
+    across(everything(), as.character),
+    across(everything(), str_trim),
+    across(everything(), str_replace_all, "NaN", "--")) %>% 
+  mutate(across(elpd_diff, str_c, " (", se_diff, ")"),
+         across(elpd_loo, str_c, " (", se_elpd_loo, ")"),
+         across(elpd_diff, str_replace_all, "0 \\( 0\\)", "--"),
+         across(c(data, model), ~recode(., mog = "Bimodal mixture model",
+                                        uvlm = "Unimodal unequal-variance model"))) %>% 
+  select(data, model, starts_with("elpd"))
+
+names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
+
+kable(table[,-1], 
+      caption = "Model comparisons by data set. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+        align =c("l", rep("r", 2))) %>%
   kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 9) %>%
-    column_spec(2, image = spec_pointrange(
-    x = table$diff_beta, 
-    xmin = table$lower_beta, 
-    xmax = table$upper_beta, 
-    vline = 0)
-    ) %>% 
-  column_spec(4, image = spec_pointrange(
-    x = table$diff_delta, 
-    xmin = table$lower_delta, 
-    xmax = table$upper_delta, 
-    vline = 0)
-  ) %>% 
-  column_spec(6, image = spec_pointrange(
-    x = table$diff_theta, 
-    xmin = table$lower_theta, 
-    xmax = table$upper_theta, 
-    vline = 0)
-  ) %>% 
-  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
-           threeparttable = T) %>% 
-  add_header_above(c(" " = 1, "Short interval durations" = 2, "Slowdown for hesitant transitions" = 2, "Probability of hesitant transitions" = 2)) %>% 
-#  group_rows("Short transition duration", 1, 9) %>%
-#  group_rows("Slowdown for hesitant transitions", 10, 18) %>%
-#  group_rows("Probability of hesitant transitions", 19, 27) #%>%
-#  column_spec(1, "1.75cm") %>%
-#  column_spec(2, "4cm") %>% 
-  group_rows("Contrast: before sentence -- before word", 1, 5) %>%
-  group_rows("Contrast: before word -- within word", 6, 10)
-
-#  column_spec(2:4, "1.75cm") %>% 
-#  column_spec(5:6, "1.75cm") %>% 
-#  column_spec(7, "1cm") 
-```
-
-# Posterior by data set
-
-## GUNNEXP2
-
-### Fit to data
-
-### Unconstrained mixture model
-
-#### Posterior parameter estimates
-
-```{r gunnpostun, fig.cap="GUNNEXP2 (unconstrained model). Posterior parameter distribution"}
-file <- list.files(path = "../stanout/gunnexp2", 
-                   pattern = "mogbetaun.*.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% 
-  filter(param %in% c("beta", "beta2", "prob")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, beta2, prob)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>%
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(location, xn, name)) %>% 
-  mutate(#group = str_c("Topic: ",topic, "; ", xn))
-         group = str_c(xn))
-
-posd <- position_dodge(.35)
-dotsize <- 3
-#shapes <- 1:length(unique(ps$group))
-#shapes <- c(2, 7, 8, 9)
-shapes <- c(8, 9)
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-beta + delta + theta + plot_layout(guides = "collect")
-
-```
-
-#### Masking effect
-
-```{r shiftcellmeansgunn, results = 'asis'}
-# cell means and data set difference
-cellmeans <- ps %>% 
-  select(-group) %>% 
-  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
-  pivot_wider(names_from = name, 
-              values_from = c(mean, lower, upper)) %>% 
-  mutate(across(ends_with("prob"), ~dezero(., 2)),
-         across(c(ends_with("beta"),
-                  ends_with("delta")), 
-                ~as.character(scales::comma(round(.))))) %>% 
-  pivot_longer(-c(xn, location), 
-               names_to = c(".value", "param"),
-               names_sep = "_") %>% 
-  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
-  select(-mean:-upper) %>% 
-  pivot_wider(names_from = xn, 
-              values_from = est)
-
-tmp <- read_csv(file) %>% 
-  filter(param %in% c("beta", "delta", "theta")) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
-  pivot_wider(names_from = xn, values_from = value) %>% 
-  unnest(c(masked, unmasked)) %>% 
-  mutate(diff = masked - unmasked) %>% 
-  summarise(across(diff, list(mean = mean, 
-                              lower = lower, 
-                              upper = upper,
-                              BF = BF), 
-                   .names = "{.fn}"),
-            .by = c(location, param)) %>% 
-  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
-         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
-         across(param, recode, theta = "prob")) %>% 
-  left_join(cellmeans) %>% 
-  arrange(param, location) %>% 
-  select(param, 
-         `Transition location` = location, 
-         Unmasked = unmasked,
-         Masked = masked,
-         "Difference" = mean, BF)
-
-write_csv(tmp, "tables/masking_effect_cellmeans.csv")
-
-tmp[,-1] %>% knitr::kable(align = c("l", "r", "r", "r", "r"), 
-                          caption = "Mixture model estimates for key transitions.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of disfluent transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of disfluent transitions.
-95% PIs in brackets.") %>%
-  #  cell_spec(1, 3:4) %>% 
-  kableExtra::collapse_rows(columns = 1) %>%
-  kable_styling(position = "center", full_width = T) %>%
-  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
-           threeparttable = T) %>% 
-  group_rows("Fluent transitions", 1, 3) %>%
-  group_rows("Disfluencies", 4, 6) %>%
-  group_rows("Probability of disfluencies", 7, 9)
-
-```
-
-### Cononstrained mixture model
-
-#### Posterior parameter estimates
-
-```{r}
-file <- list.files(path = "../stanout/gunnexp2", 
-                   pattern = "mogbetac.*.csv$",
-                   full.names = T)
-
-ps_beta <- read_csv(file) %>% 
-  filter(param %in% c("beta")) %>% 
-  nest(beta = value) %>% 
-  select(beta)
-
-beta <- ps_beta %>% 
-   unnest(beta) %>% 
-   mutate(across(value, exp)) %>% 
-   summarise(across(value, list(mean = mean, 
-                                lower = lower, 
-                                upper = upper), .names = "{.fn}")) %>% 
-  mutate(across(everything(), round, 0)) %>% 
-  transmute(result = str_c(mean, " msecs, PI: (", lower, ", ", upper, ")" )) %>% 
-  pull(result)
-```
-
-The posterior of the constrained model is shown in Figure \@ref(fig:gunnpost) showing the posterior slowdown for disfluent keystrokes (left panel) and the probability of disfluent keystrokes (right panel).
-Fluent keystroke transitions are distributed around a posterior mean of `r beta`.
-
-```{r gunnpost, fig.cap="GUNNEXP2 (constrained model). Posterior parameter distribution."}
-
-ps <- read_csv(file) %>% 
-  filter(param %in% c("beta2", "prob")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  bind_cols(ps_beta) %>% 
-  unnest(cols = c(beta, beta2, prob)) %>% 
-  rename(beta = value) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>%
-  filter(name != "beta") %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(location, xn, name)) %>% 
-  mutate(#group = str_c("Topic: ",topic, "; ", xn))
-         group = str_c(xn))
-
-posd <- position_dodge(.35)
-dotsize <- 3
-#shapes <- 1:length(unique(ps$group))
-shapes <- c(8, 9)
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) 
-#  guides(colour=guide_legend(nrow=2,byrow=F),
-#         shape = guide_legend(nrow=2,byrow=F))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-delta + theta + plot_layout(guides = "collect")
-
-```
-
-#### Masking effect
-
-```{r shiftcellmeansgunn2, results = 'asis'}
-# cell means and data set difference
-cellmeans <- ps %>% 
-  select(-group) %>% 
-  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
-  pivot_wider(names_from = name, 
-              values_from = c(mean, lower, upper)) %>% 
-  mutate(across(ends_with("prob"), ~dezero(., 2)),
-         across(ends_with("delta"), ~as.character(scales::comma(round(.))))) %>% 
-  pivot_longer(-c(xn, location), 
-               names_to = c(".value", "param"),
-               names_sep = "_") %>% 
-  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
-  select(-mean:-upper) %>% 
-  pivot_wider(names_from = xn, 
-              values_from = est)
-
-tmp <- read_csv(file) %>% 
-  filter(param %in% c("delta", "theta")) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
-  pivot_wider(names_from = xn, values_from = value) %>% 
-  unnest(c(masked, unmasked)) %>% 
-  mutate(diff = masked - unmasked) %>% 
-  summarise(across(diff, list(mean = mean, 
-                              lower = lower, 
-                              upper = upper,
-                              BF = BF), 
-                   .names = "{.fn}"),
-            .by = c(location, param)) %>% 
-  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
-         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
-         across(param, recode, theta = "prob")) %>% 
-  left_join(cellmeans) %>% 
-  arrange(param, location) %>% 
-  select(param, 
-         `Transition location` = location, 
-         Unmasked = unmasked,
-         Masked = masked,
-         "Difference" = mean, BF)
-
-write_csv(tmp, "tables/masking_effect_cellmeans_constrained.csv")
-
-tmp[,-1] %>% knitr::kable(align = c("l", "r", "r", "r", "r"), 
-                          caption = "Mixture model estimates for key transitions.
-Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of disfluent transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of disfluent transitions.
-95% PIs in brackets.") %>%
-  #  cell_spec(1, 3:4) %>% 
-  kableExtra::collapse_rows(columns = 1) %>%
-  kable_styling(position = "center", full_width = T) %>%
-  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
-           threeparttable = T) %>% 
-  group_rows("Disfluencies", 1, 3) %>%
-  group_rows("Probability of disfluencies", 4, 6)
-
-```
-
-## C2L1
-
-The C2L1 data set comprises data Norwegian 6th graders--*N*=126, mean age 11 years 10 months--published in @ronneberg2022process.
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-
-```{r eval = F}
-d <- read_csv("../data/c2l1.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000) %>%
-  mutate(ppt = as.numeric(factor(subno)),
-         condition = factor(str_c(location, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num)
-
-count(d, ppt) %>% 
-  filter(n < 100)
-pull(d, ppt) %>% unique() %>% length()
-```
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
-
-### Fit to data
-
-```{r predictionsc2l1, fig.cap="C2L1 data.
+                full_width = F, 
+                font_size = 11) %>%
+  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
+  pack_rows("Data: Bimodal mixture process", 1, 2) %>% 
+  pack_rows("Data: Unimodal process", 3, 4)
+
+```
+
+# Discussion
+
+# References
+```{r create_r-references, echo=FALSE, include=FALSE}
+r_refs(file = "references.bib")
+```
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+<div id = "references"></div>
+\endgroup
+
+# Appendix
+
+## Fit to data
+
+```{r prediction, fig.cap="Fit to data.
 Comparison of 100 simulated (predicted) sets of data to observed data illustated by model.
-For illustration the x-axis was truncated at 2,000 msecs.", eval = T}
-file <- list.files("../stanout/c2l1", 
-                   pattern = "sims", 
-                   full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
-
-```{r c2l1post, fig.cap="C2l1. Posterior parameter distribution"}
-file <- list.files(path = "../stanout/c2l1", 
-                   pattern = "mogbetaun.*.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% 
-  filter(param %in% c("beta", "beta2", "prob")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, beta2, prob)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(beta:delta) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(location, name))
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location)) +
-  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_y_continuous(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "",
-       shape = "") +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location)) +
-  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location)) +
-  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
-  geom_point(size = 3, position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-## CATO
-
-Data are published in @torrance2016adolescent. Norwegian upper
-secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-
-### Fit to data
-
-```{r predictionscato, fig.cap="CATO data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/cato", 
-                    pattern = "sims", 
-                    full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
-
-```{r catopost, fig.cap="CATO. Posterior parameter distribution"}
-file <- list.files(path = "../stanout/cato", 
-                   pattern = "mogbetaun.+.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% 
-  filter(param %in% c("beta", "beta2", "prob")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, beta2, prob)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:delta) %>% 
-  group_by(task, group, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}")) %>% 
-  mutate(task = str_c("Task: ", task, "; Group: ", group))
-
-pdodge <- position_dodge(.35)
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = task)) +
-  geom_errorbar(width = .2,
-                alpha = .75, 
-                position = pdodge) +
-  geom_point(aes(colour = task), 
-            size = 3, 
-            position = pdodge) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "",
-       shape = "") +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = task)) +
-  geom_errorbar(width = .2, alpha = .75, 
-                position = pdodge) +
-  geom_point(aes(colour = task), 
-             size = 3, 
-             position = pdodge) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = task)) +
-  geom_errorbar(width = .2, alpha = .75, 
-                position = pdodge) +
-  geom_point(aes(colour = task), 
-             size = 3, 
-             position = pdodge) +
-  facet_grid(~name) +
-  scale_shape_manual(values = 1:4) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = dezero_plot, 
-                     limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-## PLanTra
-
-### Methods
-
-The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
-In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
-Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
-During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
-making it easier to read for a lay audience--by applying what they had learned from their respective training.
-The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
-It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
-
-### Fit to data
-
-```{r predictionsplantra, fig.cap="PLanTra data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/plantra", 
-                   pattern = "sims", 
-                   full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
-
-```{r plantrapost, fig.cap="PLanTra. Posterior parameter distribution"}
-file <- list.files(path = "../stanout/plantra", 
-                   pattern = "mogbetaun.+.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% 
-  mutate(across(task, factor, 
-                levels = c("pre-test", "post-test"), 
-                ordered = T)) %>% 
-  filter(param %in% c("beta", "beta2", "prob")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, beta2, prob)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:delta) %>% 
-  group_by(task, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"))
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = task)) +
-  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
-  geom_point(aes(colour = task), 
-             size = 3, 
-             position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 8)) +
-  scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "Task",
-       shape = "Task") +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = task)) +
-  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
-  geom_point(aes(colour = task), 
-             size = 3, 
-             position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 8)) +
-  scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "Task",
-       shape = "Task") +
-  theme(axis.title.y = element_blank()) +
-    guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = task)) +
-  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.25)) +
-  geom_point(aes(colour = task), 
-             size = 3,
-             position = position_dodge(.25)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 8)) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "Task",
-       shape = "Task") +
-  theme(axis.title.y = element_blank()) +
-    guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-## LIFT
-
-### Methods
-
-LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
-
-### Fit to data
-
-```{r predictionslift, fig.cap="LIFT data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/lift", 
-                    pattern = "sims", 
-                    full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
-
-```{r}
-file <- list.files(path = "../stanout/lift", 
-                   pattern = "mogbetaun.+.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% 
-  filter(param %in% c("beta", "beta2", "prob")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, beta2, prob)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(beta:delta) %>% 
-  group_by(genre, topic, location, name) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}")) %>% 
-  ungroup() %>% 
-  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre))
-```
-
-```{r liftpost, fig.cap="LIFT. Posterior parameter distribution"}
-posd <- position_dodge(.65)
-dotsize <- 2.5
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, 
-             position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(1:8)) +
-  scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "",
-       shape = "")
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, 
-             position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(1:8)) +
-  scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = group)) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
-  geom_point(aes(colour = group), 
-             size = dotsize, 
-             position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(1:8)) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = dezero_plot, 
-                     limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "",
-       shape = "") +
-  theme(axis.title.y = element_blank())
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-## SPL2
-
-Data are going to be published in @torrance.
-
-### Methods
-
-Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
-CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
-We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
-
-Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-### Fit to data
-
-```{r predictionsspl2, fig.cap="SPL2 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/spl2", pattern = "sims", full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
-
-Can slowdowns for sentence-pauses be explained on the basis of a complex keystrokes that were summed across? -- No
-
-The data sets used differ to the extent that the keystroke interval before sentences does (PLanTra, LIFT) or does not (CATO, C2L1, SPL2) scope over the character following Shift.
-In other words, the pause before sentences sums across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involves one keyintervals, namely `_^[shift]` for the remaining data sets.
-
-For the SPL2 dataset we compared whether the different patterns for sentences pauses can be explain but the keycombination.
-We analysed the SPL2 data including and excluding the keystroke after shift.
-Model estimates are presented in Figure \ref(fig:spl2post).
-The results of this comparison can be found in Table \@ref(tab:shiftcellmeans).
-Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the interval between shift and the first character but not the hesitation probability.
-Fluent keytransitions were substantially longer when including the interval following the shift-key.
-The slowdown for hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation probability.
-Taken together, including the character following shift affects the duration of fluent transitions more than it affects pause duration and frequency.
-
-```{r spl2post, fig.cap="SPL2. Posterior parameter distribution"}
-file1 <- list.files(path = "../stanout/spl2", 
-                   pattern = "mogbetaun.+.csv$",
-                   full.names = T)
-
-file2 <- list.files(path = "../stanout/spl2_shift", 
-                   pattern = "mogbetaun.+.csv$",
-                   full.names = T)
-
-ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
-      mutate(data = .x)) %>% 
-  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
-                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
-  filter(param %in% c("beta", "delta", "prob")) %>% 
-  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2")) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta, delta, prob)) %>% 
-  mutate(beta2 = beta + delta,
-         across(c(beta2, beta), ~exp(.)),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  pivot_longer(c(beta, delta, prob)) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper),
-                   .names = "{.fn}"),
-            .by = c(lang, location, name, data))
-
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(lang, data))) +
-  geom_errorbar(width = .2, alpha = .75, 
-                position = position_dodge(.5)) +
-  geom_point(aes(colour = lang), 
-             size = 3, 
-             position = position_dodge(.5)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 8)) +
-  scale_color_colorblind() +
-  scale_y_continuous(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = "Language",
-       shape = "Before-sentence\ntransitions")
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(lang, data))) +
-  geom_errorbar(width = .2, alpha = .75,
-                position = position_dodge(.5)) +
-  geom_point(aes(colour = lang), 
-             size = 3, 
-             position = position_dodge(.5)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 8)) +
-  scale_color_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = "Language",
-       shape = "Before-sentence\ntransitions") +
-  theme(axis.title.y = element_blank())
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(lang, data))) +
-  geom_errorbar(width = .2, alpha = .75, position = position_dodge(.5)) +
-  geom_point(aes(colour = lang), 
-             size = 3, 
-             position = position_dodge(.5)) +
-  facet_grid(~name) +
-  scale_shape_manual(values = c(21, 8)) +
-  scale_colour_colorblind() +
-#  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = "Language",
-       shape = "Before-sentence\ntransitions") +
-  theme(axis.title.y = element_blank())
-
-beta + delta + theta +
-  plot_layout(guides = "collect")
-
-```
-
-```{r shiftcellmeans, results = 'asis'}
-# cell means and data set difference
-cellmeans <- ps %>% 
-  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
-  pivot_wider(names_from = name, 
-              values_from = c(mean, lower, upper)) %>% 
-  mutate(across(ends_with("prob"), ~dezero(., 2)),
-         across(c(ends_with("beta"),
-                  ends_with("delta")), 
-                ~as.character(scales::comma(round(.))))) %>% 
-  pivot_longer(-c(lang, location, data), 
-               names_to = c(".value", "param"),
-               names_sep = "_") %>% 
-  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
-  select(-mean:-upper) %>% 
-  pivot_wider(names_from = data, 
-              values_from = est)
-
-file1 <- list.files(path = "../stanout/spl2", 
-                   pattern = "mogbetaun.+.csv$",
-                   full.names = T)
-
-file2 <- list.files(path = "../stanout/spl2_shift", 
-                   pattern = "mogbetaun.+.csv$",
-                   full.names = T)
-
-tmp <- map_dfr(c(file1, file2), 
-               ~read_csv(.x) %>% 
-          mutate(data = .x)) %>% 
-  filter(param %in% c("beta", "delta", "theta")) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
-  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
-         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
-  pivot_wider(names_from = data, values_from = value) %>% 
-  unnest(c(spl2, spl2_shift)) %>% 
-  mutate(diff = spl2_shift - spl2) %>% 
-  summarise(across(diff, list(mean = mean, 
-                              lower = lower, 
-                              upper = upper,
-                              BF = BF), 
-                   .names = "{.fn}"),
-            .by = c(lang, location, param)) %>% 
-  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
-         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
-         across(param, recode, theta = "prob")) %>% 
-  left_join(cellmeans) %>% 
-  select(param, Language = lang,
-         `Transition location` = location, 
-         `_\\^[shift] + C` = ends_with(" + C)"),
-         `_\\^[shift]` = ends_with("FT])"),
-         "Difference" = mean, BF)
-
-write_csv(tmp, "tables/shift_effect_cellmeans.csv")
-
-tmp[,-1] %>% knitr::kable(align = c("l", "l", "r", "r", "r", "r"), 
-                          caption = "Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of disfluent transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of disfluent transitions.
-95% PIs in brackets.") %>%
-  #  cell_spec(1, 3:4) %>% 
-  kableExtra::collapse_rows(columns = 1) %>%
-  kable_styling(position = "center", full_width = T) %>%
-  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", 
-           threeparttable = T) %>% 
-  group_rows("Fluent transitions", 1, 6) %>%
-  group_rows("Disfluencies", 7, 12) %>%
-  group_rows("Probability of disfluencies", 13, 18)
-
-```
-
-# References
+For illustration the x-axis was truncated at 2,000 msecs.", eval = T, fig.height=10}
+
+run <- FALSE
+
+if(run){
+  file <- list.files("../stanout", pattern = "sims", full.names = T, recursive = T)
+  ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
+    str_to_upper() %>% 
+    str_replace(., "PLANTRA", "PLanTra")
+  ds <- ds[-6]
+  plots <- list()
+  plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
+  #plots <- map(1, ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
+  plots_fin <- plots[[1]] / plots[[2]] / plots[[3]] / plots[[4]] / plots[[5]] / plots[[6]] +
+    plot_layout(guides = "collect") &
+  #  plot_annotation(tag_levels = 'A') & 
+    theme(legend.position = 'bottom')
+  saveRDS(plots_fin, "objects/fitplots.rda", compress = "xz")
+}
+
+plots <- readRDS("objects/fitplots.rda")
+```
+
+## Posterior parameter estimates

### 2023-07-17 report/results.Rmd
Add manuscript and simulation

@@
 ```
+### GUNNEXP2
+
+### C2L1
+
+The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process.
+The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
+
+TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
+
+### CATO
+
+Data are published in @torrance2016adolescent. Norwegian upper
+secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
+
+### PLanTra
+
+The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
+In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
+Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
+During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
+making it easier to read for a lay audience--by applying what they had learned from their respective training.
+The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
+It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
+
+### LIFT
+
+LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
+
+### SPL2
+
+Data are going to be published in @torrance.
+
+Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
+CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
+We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
+
+Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
 
@@
 
+  result[[4]] <- d %>% 
+    select(ppt, enough_sentences) %>% 
+    unique() %>%
+    filter(!enough_sentences) %>%
+    drop_na() %>% 
+    summarise(less_than_10_sentences = length(ppt))
+
+  ppt_remove <- d %>% 
+    select(ppt, enough_sentences) %>% 
+    unique() %>%
+    filter(!enough_sentences) %>% 
+    drop_na() %>% 
+    pull(ppt)
+
   full_set <- d %>% 
+    filter(!ppt %in% ppt_remove) %>% 
     count(ppt, topic, genre) %>% 
@@
     pull(ppt)
-
-  result[[4]] <- d %>% 
-    filter(ppt %in% ppt_keep) %>% 
-    select(ppt, enough_sentences) %>% 
-    unique() %>% 
-    count(enough_sentences)
 
@@
            ppt %in% ppt_keep,
+           !ppt %in% ppt_remove,
            enough_sentences) %>%
@@
            ppt %in% ppt_keep,
+           !ppt %in% ppt_remove,
            enough_sentences,
@@
               .by = location)
-
-   set.seed(365)
-   result[[5]] <- d %>% 
-    mutate(keep = 1:n(),
-           keep = sample(keep),
-           keep = keep <= n_samples, 
-           .by = c(ppt, location, genre, topic)) %>% 
-    filter(keep) %>% 
-    mutate(ppt_keep = ppt %in% sample(unique(ppt), n_ppts)) %>% 
-    select(ppt, ppt_keep) %>% 
-    unique() %>% 
-    count(ppt_keep)
 
@@
   result[[1]] <- d %>% 
-    mutate(too_short = iki <= 50,
-           too_fast = iki >= 30000,
+    mutate(too_fast = iki <= 50,
+           too_slow = iki >= 30000,
            across(ppt, ~as.numeric(factor(.)))) %>% 
@@
                      .names = "{.col}"), 
+              .by = location) %>% 
+    mutate(across(location, ~str_replace(., "_", " ")),
+           across(location, ~recode(., `sentence before` = "before sentence",
+                                  `word before` = "before word")))
+
+  return(result)
+}
+get_gunnexp2 <- function(file, n_samples){
+  d <- read_csv(file) %>%
+    filter(!is.na(iki),
+           !is.na(isfluent),
+           isfluent == 1) %>%
+    mutate(ppt = as.numeric(factor(ppt))) %>% 
+    select(ppt, iki, xn, location)
+
+  result <- list()
+
+  result[[1]] <- d %>% 
+  mutate(too_fast = iki <= 50,
+         too_slow = iki >= 30000) %>%
+  summarise(across(starts_with("too"), 
+                   list(mean = mean, 
+                        se = se_bin)),
+            .by = c(ppt)) %>% 
+  summarise(across(starts_with("too"), 
+                   list(mean = mean), 
+                   .names = "{.col}"))
+  # Sample within each category random data points
+  set.seed(365)
+  result[[2]] <- d %>% 
+  filter(iki > 50,
+         iki < 30000) %>% 
+  mutate(keep = 1:n(),
+         keep = sample(keep),
+         keep = keep <= n_samples,
+         .by = c(ppt, location, xn)) %>% 
+  summarise(across(keep, 
+                   list(mean = mean, 
+                        se = se_bin)),
+            .by = c(ppt, location)) %>% 
+  summarise(across(starts_with("keep"), 
+                   list(mean = mean), 
+                   .names = "{.col}"), 
             .by = location)
-
   return(result)
@@
 spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
-
-```
-
-```{r eval = F}
-Data reduction using random samples of subsets of participants for larger data sets
-
-Also for larger data sets, only used participants that provided a min number of sentences
-
-Removed transitions that were followed by an editing operation
-
-Extremely short and long intervals
-
-Add a description with general data filtering and a table with how many data were removed from the analysis by study.
-
-Add final sample number of ppt, and transitions by location
+gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
+
+#lift[[3]]
+#lift[[4]]
+#plantra[[3]]
+#plantra[[4]]
+#spl2[[3]] 
+```
+
+For all datasets we only used transitions that were not followed by an editing operation.
+
+We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
+We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant)
+
+We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \@ref(tab:datareduction).
+
+From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
+This was done for computational reasons to reduce the time the Bayesian models need to complete.
+For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis.
+Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant.
+The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \@ref(tab:datareduction).
+
+```{r datareduction, results='asis'}
+c2l1$ds <- "c2l1"
+cato$ds <- "cato"
+lift$ds <- "lift"
+plantra$ds <- "plantra"
+spl2$ds <- "spl2"
+gunnexp2$ds <- "gunnexp2"
+
+extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
+          mutate(ds = .x$ds)) %>% 
+  mutate(across(where(is.numeric), ~round(.*100, 2)),
+         across(starts_with("too"), ~str_c(., "%"))) %>% 
+  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
+  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
+  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
+  rename(`$\\le$ 50 msecs` = `<50ms`,
+         `$\\ge$ 30 secs` = `>30,000ms`)
+
+random_sample <- 
+  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
+          mutate(ds = .x$ds)) %>% 
+    mutate(across(where(is.numeric), ~round(.*100, 1)),
+           across(starts_with("keep"), ~str_c(., "%"))) %>%
+    unite("keep", starts_with("keep"), sep = " (") %>% 
+    mutate(across(c(keep), ~str_c(., ")"))) %>%
+  pivot_wider(names_from = location, values_from = keep)
+
+datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
+  relocate(ds) %>% 
+  mutate(across(ds, recode, 
+                "cato" = "CATO",
+                "spl2" = "SPL2", 
+                "spl2_shift" = "SPL2 (shift + C)", 
+                "plantra" = "PLanTra",
+                "lift" = "LIFT",
+                "c2l1" = "C2L1",
+                "gunnexp2" = "GUNNEXP2")) %>% 
+  rename(`Dataset` = ds)
+
+kable(datareduction, 
+      caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition locattion. Standard error is shown in parentheses.",
+        align =c("l", rep("r", 14))) %>%
+  kable_styling(position = "center", 
+                full_width = T, 
+                font_size = 11) %>%
+  add_header_above(c(" " = 1, 
+                     "Extreme values" = 2,
+                     "Randomly sampled data" = 3)) %>%
+  column_spec(1, width = "8em")
 
@@
 
-# Posterior by data set
-
-## GUNNEXP2
-
-### Fit to data
-
-### Unconstrained mixture model
-
-#### Posterior parameter estimates
+# Model comparison with simulated data
+
+A general concern with mixture models is that in principle, as the mixture model has more parameters it might simply always lead to a better fit, even though cross-validation is addressing potential problems with overfitting models.
+
+To address this concern we simulated two sets of data.
+Both data sets have two conditions and 1,000 observations each.
+The first set of data has as underlying data generating process a mixture model with two mixture components similar to the process described above.
+The difference between the two conditions is that the mixing proportion is larger for condition 2 than for condition 1, hence long observations are more likely in condition 2.
+This model can be summarised as followed:
+
+$$
+\text{y}_i \sim \theta_\text{condition[i]} \cdot \text{logN}(\mu_1, \sigma^2_1) +\\
+(1 - \theta_\text{condition[i]}) \cdot \text{logN}(\mu_2, \sigma^2_2)
+$$
+
+The second data set was generated with an unequal variance unimodal model as data generating process.
+Condition 2 has a larger mean and standard deviation than condition 1.
+The model can be summaried as followed:
+
+$$
+\text{y}_i \sim \text{logN}(\mu_\text{condition[i]}, \sigma^2_\text{condition[i]})
+$$
+
+The true parameter values used for each of the two data simulations can be found in Table \@ref(tab:simparam).
+The simulated data are visualised in Figure \@ref(fig:simdata).
+The data are simulated to be similarly distributed to keystroke transitions.
+
+```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=4}
+files <- list.files("../mixture-model-sim/data", full.names = T)
+
+map_dfr(files, ~read_csv(.) %>% 
+          mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
+  mutate(across(condition, factor),
+         across(condition, ~str_c("Condition: ", condition)),
+         across(Dataset, ~recode(., mogdata = "bimodal", uvlmdata = "unimodal"))) %>% 
+  ggplot(aes(x = value, colour = condition, fill = condition)) +
+  geom_density(alpha = .25) +
+  facet_grid(~Dataset, labeller = label_both) +
+  scale_colour_colorblind() + 
+  scale_fill_colorblind() +
+  labs(colour = "", fill = "", x = "Simulated data", y = "Density")
+
+```
+
+We run 4 models: 2 mixture models, one on the data generated with a mixture process and one on the data generated with the unimodal unequal variance process.
+We repeated the same using an unimodal unequal variance model.
+Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
+The Stan models uncovered the model parameters of their respective data sets successfully, as shown in Table \@ref(tab:simparam), not less so when the model was applied to data generated with the other underlying process.
+
+```{r}
+mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
+  mutate(data = "mog")
+mog2 <- read_csv("../mixture-model-sim/stanout/mog_uvlmdata.csv") %>% 
+  mutate(data = "uvlm")
+uvlm1 <- read_csv("../mixture-model-sim/stanout/uvlm_uvlmdata.csv") %>% 
+  mutate(data = "uvlm")
+uvlm2 <- read_csv("../mixture-model-sim/stanout/uvlm_mogdata.csv") %>% 
+  mutate(data = "mog")
+
+mog_sim <- bind_rows(mog1, mog2) %>% 
+  summarise(across(value, list(est = mean,
+                               lo = lower,
+                               up = upper),
+                   .names = "{.fn}"),
+            .by = c(name, data)) %>% 
+  mutate(across(where(is.numeric), round, 2),
+         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
+         across(est, ~str_c(., " [", lo, ", ", up, "]")),
+         across(name, ~str_c("\\", .)),
+         across(name, ~str_replace(., "a_e", "a^2_1")),
+         across(name, ~str_replace(., "ap_e", "a^2_2")),
+         across(name, ~str_replace(., "\\[1\\]", "_\\\\text{condition=1}")),
+         across(name, ~str_replace(., "\\[2\\]", "_\\\\text{condition=2}"))) %>%
+  select(-lo, -up) %>% 
+  pivot_wider(names_from = data, values_from = est) %>% 
+  mutate(true = c(5, 1, .1, .4, .25, .5),
+         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
+         across(name, ~str_replace(., "prob", "theta"))) %>% 
+  relocate(name, true)
+
+uvlm_sim <- bind_rows(uvlm1, uvlm2) %>% 
+  summarise(across(value, list(est = mean,
+                               lo = lower,
+                               up = upper),
+                   .names = "{.fn}"),
+            .by = c(name, data)) %>% 
+  mutate(across(where(is.numeric), round, 2),
+         across(est, ~str_c(., " [", lo, ", ", up, "]")),
+         across(name, ~str_c("\\", .)),
+         across(name, ~str_replace(., "a_e", "a^2_1")),
+         across(name, ~str_replace(., "ap_e", "a^2_2")),
+         across(name, ~str_replace(., "\\[1\\]", "_\\\\text{condition=1}")),
+         across(name, ~str_replace(., "\\[2\\]", "_\\\\text{condition=2}"))) %>% 
+  select(-lo, -up) %>%
+  pivot_wider(names_from = data, values_from = est) %>% 
+  mutate(true = as.character(c(5, 6, .25, .5))) %>% 
+  relocate(name, true)
+
+```
+
+```{r simparam, results='asis'}
+table <- bind_rows(
+  mog_sim %>% mutate(model = "mog"),
+  uvlm_sim %>% mutate(model = "uvlm")) %>% 
+  relocate(model) %>% 
+  mutate(across(name, ~str_c("$", ., "$")))
+
+names(table) <- c("Model", "Parameter", "True value", "Bimodal data", "Unimodal data")
+
+kable(table[,-1], 
+      caption = "Uncovered parameter estimates with 95% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
+      align =c("l", rep("r", 3))) %>%
+  kable_styling(position = "center", 
+                full_width = F, 
+                font_size = 11) %>% 
+  add_header_above(c(" " = 1, 
+                     " " = 1,
+                     "Estimate with 95\\% PI" = 2)) %>%
+  pack_rows("Model: Bimodal mixture model", 1, 6) %>% 
+  pack_rows("Model: Unimodal process", 7, 10)
+
+```
+
+We used LOO-CV to compare the fit of the two models for each data set.
+The model comparisons can be found for each data generating process in Table \ref@(tab:loossim).
+The results show that the mixture model does not always lead to higher predictive performance.
+Indeed, the mixture model showed a lower predictive performance for the data that were generated with a unimodal process.
+However, for the data generated with a bimodal process, the mixture model model shows a higher predictive performance.
+In fact, the ratio of $\Delta\widehat{elpd}$ and its SE, as metric for the strength of evidence, shows that the mixture model performs 11 times better than the unimodal model for the data generated with a bimodal process.
+In comparison, for the unimodal data, the unimodal model shows only 3 times better than the bimodal mixture model.
+Thus, even though the mixture model does not necessarily perform better for non-bimodal data but it also doesn't necessarily perform much worse.
+This contrast is likely a reflection of the increased number of parameters in the mixture model.
+
+```{r loossim, results='asis'}
+table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
+  select(1:6) %>% 
+  mutate(
+#    elpd_se_ratio = abs(elpd_diff / se_diff),
+    across(where(is.numeric), round, 0),
+    across(where(is.numeric), format, big.mark = ","),
+    across(everything(), as.character),
+    across(everything(), str_trim),
+    across(everything(), str_replace_all, "NaN", "--")) %>% 
+  mutate(across(elpd_diff, str_c, " (", se_diff, ")"),
+         across(elpd_loo, str_c, " (", se_elpd_loo, ")"),
+         across(elpd_diff, str_replace_all, "0 \\( 0\\)", "--"),
+         across(c(data, model), ~recode(., mog = "Bimodal mixture model",
+                                        uvlm = "Unimodal unequal-variance model"))) %>% 
+  select(data, model, starts_with("elpd"))
+
+names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
+
+kable(table[,-1], 
+      caption = "Model comparisons by data set. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+        align =c("l", rep("r", 2))) %>%
+  kable_styling(position = "center", 
+                full_width = F, 
+                font_size = 11) %>%
+  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
+  pack_rows("Data: Bimodal mixture process", 1, 2) %>% 
+  pack_rows("Data: Unimodal process", 3, 4)
+
+```
+
+## Posterior parameter estimates of mixture model
+
+### GUNNEXP2
+
+#### Unconstrained mixture model
+
+##### Posterior parameter estimates
 
@@
 
-#### Masking effect
+##### Masking effect
 
@@
 
-### Cononstrained mixture model
-
-#### Posterior parameter estimates
+#### Constrained mixture model
+
+##### Posterior parameter estimates
 
@@
 
-## C2L1
-
-The C2L1 data set comprises data Norwegian 6th graders--*N*=126, mean age 11 years 10 months--published in @ronneberg2022process.
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-
-```{r eval = F}
-d <- read_csv("../data/c2l1.csv") %>%
-  filter(!is.na(iki), 
-         iki > 50, 
-         iki < 30000) %>%
-  mutate(ppt = as.numeric(factor(subno)),
-         condition = factor(str_c(location, sep = "_")),
-         cond_num = as.integer(condition)) %>% 
-  select(ppt, iki, condition, cond_num)
-
-count(d, ppt) %>% 
-  filter(n < 100)
-pull(d, ppt) %>% unique() %>% length()
-```
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
-
-### Fit to data
-
-```{r predictionsc2l1, fig.cap="C2L1 data.
-Comparison of 100 simulated (predicted) sets of data to observed data illustated by model.
-For illustration the x-axis was truncated at 2,000 msecs.", eval = T}
-file <- list.files("../stanout/c2l1", 
-                   pattern = "sims", 
-                   full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
+### C2L1
 
@@
 
-## CATO
-
-Data are published in @torrance2016adolescent. Norwegian upper
-secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-
-### Fit to data
-
-```{r predictionscato, fig.cap="CATO data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/cato", 
-                    pattern = "sims", 
-                    full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
+### CATO
 
@@
 
-## PLanTra
-
-### Methods
-
-The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
-In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
-Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
-During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
-making it easier to read for a lay audience--by applying what they had learned from their respective training.
-The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
-It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
-
-### Fit to data
-
-```{r predictionsplantra, fig.cap="PLanTra data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/plantra", 
-                   pattern = "sims", 
-                   full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
+### PLanTra
 
@@
 
-## LIFT
-
-### Methods
-
-LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
-
-### Fit to data
-
-```{r predictionslift, fig.cap="LIFT data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/lift", 
-                    pattern = "sims", 
-                    full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
+### LIFT
 
@@
 
-## SPL2
-
-Data are going to be published in @torrance.
-
-### Methods
-
-Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
-CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
-We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
-
-Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-### Fit to data
-
-```{r predictionsspl2, fig.cap="SPL2 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}
-
-file <- list.files("../stanout/spl2", pattern = "sims", full.names = T)
-
-plots <- get_pdens_plot(file, max_iki = 2000)
-
-grid.arrange(patchworkGrob(plots), 
-             left = "Density",
-             bottom = "Transition duration in msecs")
-```
-
-### Posterior parameter estimates of mixture model
+### SPL2
 
@@
 
+## Fit to data
+
+```{r prediction, fig.cap="Fit to data.
+Comparison of 100 simulated (predicted) sets of data to observed data illustated by model.
+For illustration the x-axis was truncated at 2,000 msecs.", eval = T, fig.height=10}
+
+run <- FALSE
+
+if(run){
+  file <- list.files("../stanout", pattern = "sims", full.names = T, recursive = T)
+  ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
+    str_to_upper() %>% 
+    str_replace(., "PLANTRA", "PLanTra")
+  ds <- ds[-6]
+  plots <- list()
+  plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
+  #plots <- map(1, ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
+  plots_fin <- plots[[1]] / plots[[2]] / plots[[3]] / plots[[4]] / plots[[5]] / plots[[6]] +
+    plot_layout(guides = "collect") &
+  #  plot_annotation(tag_levels = 'A') & 
+    theme(legend.position = 'bottom')
+  saveRDS(plots_fin, "objects/fitplots.rda", compress = "xz")
+}
+
+readRDS("objects/fitplots.rda")
+```
+
 # References

### 2023-07-17 mixture-model-sim/scripts/densityplot.R

@@ -1,0 +1,15 @@
+library(tidyverse)
+theme_set(theme_bw()+
+            theme(legend.justification = "top",
+                  panel.grid = element_blank()))
+
+files <- list.files("data", full.names = T)
+
+map_dfr(files, ~read_csv(.) %>% 
+          mutate(data = str_remove_all(.x, "data|/|.csv" ))) %>% 
+  mutate(across(condition, factor),
+         across(data, ~recode(., mog = "bimodal", uvlm = "unimodal"))) %>% 
+  ggplot(aes(x = value, colour = condition, fill = condition)) +
+  geom_density(alpha = .25) +
+  facet_grid(~data, labeller = label_both) 
+  

### 2023-07-17 mixture-model-sim/scripts/get_uvlm_data.R

@@ -1,0 +1,16 @@
+library(tidyverse)
+
+set.seed(123)
+N <- 1000 # number of subjects
+beta <- c(5, 6) # Population parameter
+sigma <- c(.25, .5) # trial-by-trial error
+
+# iterate over subject to generate data for each one
+data <- map_dfr(1:2, 
+                ~tibble(condition = ., 
+                        value = rlnorm(n = N, mean = beta[.], sd = sigma[.]))) 
+
+ggplot(data, aes(x = value, colour = factor(condition))) +
+  geom_density() 
+
+write_csv(data, "data/uvlmdata.csv")

### 2023-07-17 mixture-model-sim/scripts/mog_mogdata.R

@@ -1,0 +1,64 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+n_chain = n_core = 3 # number of cores/chains
+iterations = 6e3
+
+# Get some fake data (see script for parameter values)
+data <- read_csv("data/mogdata.csv")
+
+# Data as list
+dat <- within(list(), {
+    N <- nrow(data)
+    y <- data$value
+    K <- max(data$condition)
+    condition <- data$condition # or whatever grouping variable
+  }  ); str(dat)
+
+# Initialise start values
+start <-  function(chain_id = 1){
+    list(   beta = 5
+          , delta = 1
+          , theta = c(-.5, -.5)
+          , sigma = .5
+          , sigma_diff = .2 ) }
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Check compiling
+mog <- stan(file = "stan/mog.stan", data=dat, chains=0)
+
+# Fit model
+m <- stan(fit = mog, 
+          data = dat, 
+          init = start_ll,
+          iter = iterations, 
+          warmup= iterations/2,
+          chains = n_chain, 
+          cores = n_core, 
+          refresh = 2000, 
+          seed = 365,
+          control = list(max_treedepth = 16,
+                        adapt_delta = 0.99))
+
+# Save posterior samples
+saveRDS(m,
+        file="stanout/mog_mogdata.rda",
+        compress="xz")
+
+param <- c("beta", "delta", "prob", "sigma")
+traceplot(m, param)
+summary(m, param, prob = c(.025, .975))$summary %>% round(2)
+
+# Get posterior
+m <- readRDS("stanout/mog_mogdata.rda")
+
+# Select relevant parameters
+param <- c("beta", "delta", "prob", "sigma_e", "sigmap_e") 
+
+# Extract posterior
+as.data.frame(m, param) %>% 
+  pivot_longer(everything()) %>% 
+  # Save posterior
+  write_csv("stanout/mog_mogdata.csv")

### 2023-07-17 mixture-model-sim/scripts/mog_uvlmdata.R

@@ -1,0 +1,60 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+n_chain = n_core = 3 # number of cores/chains
+iterations = 6e3
+
+# Get some fake data (see script for parameter values)
+data <- read_csv("data/uvlmdata.csv")
+
+# Data as list
+dat <- within(list(), {
+    N <- nrow(data)
+    y <- data$value
+    K <- max(data$condition)
+    condition <- data$condition 
+  }  ); str(dat)
+
+# Initialise start values
+start <-  function(chain_id = 1){
+    list(   beta = 5
+          , delta = 1
+          , theta = c(-.5, -.5)
+          , sigma = .5
+          , sigma_diff = .1 ) }
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Check compiling
+mog <- stan(file = "stan/mog.stan", data=dat, chains=0)
+
+# Fit model
+m <- stan(fit = mog, 
+          data = dat, 
+          init = start_ll,
+          iter = iterations, 
+          warmup= iterations/2,
+          chains = n_chain, 
+          cores = n_core, 
+          refresh = 2000, 
+          seed = 365,
+          control = list(max_treedepth = 16,
+                        adapt_delta = 0.99))
+
+# Save posterior samples
+saveRDS(m,
+        file="stanout/mog_uvlmdata.rda",
+        compress="xz")
+
+param <- c("beta", "delta", "prob", "sigma_e", "sigmap_e")
+traceplot(m, param)
+summary(m, param, prob = c(.025, .975))$summary %>% round(2)
+
+# Get posterior
+m <- readRDS("stanout/mog_uvlmdata.rda")
+
+# Extract posterior
+as.data.frame(m, param) %>% 
+  pivot_longer(everything()) %>% 
+  write_csv("stanout/mog_uvlmdata.csv")

### 2023-07-17 mixture-model-sim/scripts/uvlm_uvlmdata.R

@@ -1,0 +1,56 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+rstan_options(auto_write = TRUE)
+n_chain = n_core = 3 # number of cores/chains
+iterations = 6e3
+
+# Get some fake data (see script for parameter values)
+data <- read_csv("data/uvlmdata.csv")
+
+# Data as list
+dat <- within(list(), {
+    N <- nrow(data)
+    y <- data$value
+    K <- max(data$condition) 
+    condition <- data$condition 
+  }  ); str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(beta = c(5, 6), sigma = c(.25, .5))}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Check compiling
+uvlm <- stan(file = "stan/uvlm.stan", data=dat, chains=0)
+
+# Fit model
+m <- stan(fit = uvlm, 
+          data = dat, 
+          init = start_ll,
+          iter = iterations, 
+          warmup= iterations/2,
+          chains = n_chain, 
+          cores = n_core, 
+          refresh = 2000, 
+          seed = 365,
+          control = list(max_treedepth = 16,
+                        adapt_delta = 0.99))
+
+# Save posterior samples
+saveRDS(m,
+        file="stanout/uvlm_uvlmdata.rda",
+        compress="xz")
+
+param <- c("beta", "sigma")
+traceplot(m, param)
+summary(m, param, prob = c(.025, .975))$summary %>% round(2)
+
+# Get posterior
+m <- readRDS("stanout/uvlm_uvlmdata.rda")
+
+# Extract posterior
+as.data.frame(m, param) %>% 
+  pivot_longer(everything()) %>% 
+  write_csv("stanout/uvlm_uvlmdata.csv")

### 2023-07-17 mixture-model-sim/scripts/zzz/get_lmm_data.R

@@ -1,0 +1,32 @@
+library(tidyverse)
+
+set.seed(123)
+Nsubj <- 100 # number of subjects
+K <- 50 # number of observations per subject per condition
+# assumed data were transformed to proportions
+# Population parameters
+beta_mean <- 1000
+beta_sd <- 50
+beta2_mean <- 1100
+sigma <- log(c(2, 3)) # trial-by-trial error
+SubjBeta <- log(rnorm(Nsubj, beta_mean, beta_sd))
+SubjBeta2 <- log(rnorm(Nsubj, beta2_mean, beta_sd))
+
+# iterate over subject to generate data for each one
+data <- map_dfr(1:Nsubj, 
+                ~bind_rows(tibble(
+                          id = ., condition = 1, 
+                          value = rnorm(n = K, 
+                                        mean = SubjBeta[.], 
+                                        sd = sigma[1])),
+                          tibble(
+                          id = ., condition = 2, 
+                          value = rnorm(n = K, 
+                                        mean = SubjBeta2[.], 
+                                        sd = sigma[2])))) %>% 
+  mutate(across(value, exp))
+
+ggplot(data, aes(x = value, colour = factor(condition))) +
+  geom_density() 
+
+write_csv(data, "data/lmmdata.csv")

### 2023-07-17 report/scripts/get_pdens_plot.R

@@ -1,4 +1,4 @@
-get_pdens_plot <- function(file, max_iki){
+get_pdens_plot <- function(file, max_iki, ds){
   sims <- read_csv(file) %>% 
     pivot_longer(cols = c(y_obs, y_tilde)) %>% 
     mutate(across(name, recode_factor, 
@@ -6,79 +6,31 @@ get_pdens_plot <- function(file, max_iki){
                   y_obs = "Observed data",
                   .ordered = T)) %>% 
     filter(name ==  "Simulated data" |
-             (name == "Observed data" & sim_idx == 1)) 
+          (name == "Observed data" & sim_idx == 1),
+           value < max_iki) %>% 
+    mutate(across(model, ~case_when(
+      . == "lmmgaus" ~ "Unimodal normal",
+      . == "lmm" ~ "Unimodal log-normal",
+      . == "lmmuneqvar" ~ "Unimodal log-normal (unequal variance)",
+      . == "mogbetaunconstr" ~ "Bimodal log-normal (unconstrained)",
+       str_detect(., "mogbetacon.+") ~ "Bimodal log-normal (constrained)")),
+      across(model, ~str_wrap(., 30)),
+      across(model, ~factor(., levels = unique(model)[c(2, 1, 3, 4, 5)], ordered = T)))
   
-  plot_mog <- sims %>%
-    filter(model == "mogbetaunconstr", value < max_iki) %>% 
-    mutate(model = "Bimodal log-normal (unconstrained)") %>% 
+  plot <- sims %>%
     ggplot(aes(x = value, 
                group = interaction(sim_idx,name), 
                colour = name)) +
     geom_density() +
-    facet_wrap(~model) +
+    facet_wrap(~model, scales = "free_x", nrow = 1) +
     scale_colour_manual(values = c("firebrick3", "black")) +
     scale_x_continuous(labels = scales::comma) +
-    labs(colour = "") +
-    theme(axis.title = element_blank())
+    labs(colour = "", 
+         y = "Density", 
+         x = "Transition duration in msecs",
+         title = ds) +
+    theme(axis.text.y = element_blank(),
+          axis.ticks.y = element_blank())
   
-  
-  plot_mog2 <- sims %>%
-    filter(str_detect(model, "mogbetacon.+"), value < max_iki) %>% 
-    mutate(model = "Bimodal log-normal (constrained)") %>% 
-    ggplot(aes(x = value, 
-               group = interaction(sim_idx,name), 
[…]

### 2023-07-17 scripts/c2l1/mogbetaconstrloc.R

@@ -1,0 +1,104 @@
+# Load packages
+library(tidyverse)
+library(rstan)
+source("scripts/c2l1/get_data.R")
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+# Sampling parameters
+n_cores <- 3
+n_chain <- 3
+iterations <- 20000
+n_samples <- 100 # number of random data points
+file <- "data/c2l1.csv"
+
+# Load df
+d <- get_data(file, n_samples)
+
+# Count observations
+count(d, ppt, condition)
+
+# Data as list
+dat <- within( list(), {
+  nS <- length(unique(d$ppt))
+  subj <- d$ppt
+  K <- length(unique(d$cond_num))
+  condition <- as.numeric(d$cond_num)
+  K_loc <- length(unique(d$loc_num))
+  location <- as.numeric(d$loc_num)
+  y <- d$iki
+  N <- nrow(d)
+} );str(dat)
+
+# Initialise start values
+start <- function(chain_id = 1){
+  list(   beta = c(7,6,5)
+          , delta = rep(.1, dat$K)
+          , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
+          , theta = rep(0, dat$K)
+          , tau = .1
+          , sigma = 1.25
+          , sigma_diff = rep(.1, dat$K_loc)
+          , sigma_u = 0.41)}
+
+start_ll <- lapply(1:n_chain, function(id) start(chain_id = id) )
+
+# Load model
+mog <- stan_model(file = "stan/mogbetaconstrloc.stan")
+
+# Parameters to omit in output
+omit <- c("prob_tilde", "mu", "theta_s")
+
+# Fit model
+m <- sampling(mog, 
+              data = dat,
+              init = start_ll,
+              iter = iterations,
+              warmup = iterations/2,
+              chains = n_chain, 
+              cores = n_cores,
+              refresh = 5000,
+              save_warmup = FALSE, # Don't save the warmup
+              include = FALSE, # Don't include the following parameters in the output
+              pars = omit,
+              seed = 81,
+              control = list(max_treedepth = 16,
+                             adapt_delta = 0.99))
+
+# Save model
+saveRDS(m, 
+        file = "stanout/c2l1/mogbetaconstrloc.rda",
+        compress = "xz")
+
+#m <- readRDS(file = "stanout/c2l1/mogbetaconstr.rda")
+
[…]

### 2023-07-17 scripts/gunnexp2/get_ytilde.R

@@ -6,12 +6,12 @@ library(tidyverse)
 library(rstan)
 
 # Get functions
-source("scripts/c2l1/get_data.R")
+source("scripts/gunnexp2/get_data.R")
 source("scripts/get_ytilde.R")
 
 # File location and number of random samples
 n_samples <- 100 # number of random data points
-file <- "data/c2l1.csv"
+file <- "data/gunnexp2.csv"
 
 # Load df
 d <- get_data(file, n_samples) %>% 
@@ -27,7 +27,7 @@ maxiter <- 20000
 idx <- sample(maxiter, nsims)
 
 # File names
-path <- "stanout/c2l1"
+path <- "stanout/gunnexp2"
 files <- list.files(path = path, pattern = ".rda", full.names = T)
 
 # Apply get_ytilde function
@@ -41,5 +41,5 @@ allsims <- left_join(d, sims,
   rename(y_obs = iki, y_tilde = value) 
 
 # Save simulations
-write_csv(allsims, "stanout/c2l1/all_sims.csv")
+write_csv(allsims, "stanout/gunnexp2/all_sims.csv")
 

### 2023-07-17 scripts/spl2 (shift + C)/mogbetaconstr.R

@@ -29,7 +29,7 @@ dat <- within( list(), {
 
 # Initialise start values
 start <- function(chain_id = 1){
-  list(   beta = 4
+  list(   beta = 5
           , delta = rep(.1, dat$K)
           , theta_s = matrix(0, nrow = dat$K, ncol = dat$nS)
           , theta = rep(0, dat$K)
@@ -64,7 +64,7 @@ m <- sampling(mog,
 
 # Save model
 saveRDS(m, 
-        file = "stanout/spl2 (shift + C)/mogbetaconstr.rda",
+        file = "stanout/spl2_shift/mogbetaconstr.rda",
         compress = "xz")
 
 # Load model
@@ -102,5 +102,5 @@ ps_fin <- ps %>%
          across(location, ~str_replace(., "NA_NA", "overall")))
 
 # Save posterior
-write_csv(ps_fin, "stanout/spl2 (shift + C)/mogbetaconstr.csv")
+write_csv(ps_fin, "stanout/spl2_shift/mogbetaconstr.csv")
 

### 2023-07-24 report/appendix.tex
Manuscript with appendices separately

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\hypertarget{fit-to-data}{%
\subsection{Fit to data}\label{fit-to-data}}

Model fit was assessed by testing to what extent simulations obtained
from the model fit the observed data. From each model we obtained 100
simulations (predictions). These can be found in Figure
\ref{fig:prediction}. The model predictions fitted the data
substantially better for the mixture models. The worse fit was observed
for the unimodal Gaussian model which predicted large amounts of
negative keystroke intervals which are implausible. Using a log-normal
distribution removed these negative keystroke intervals.

\begin{figure}

{\centering \includegraphics{figures/fitplots}

}

\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
\end{figure}

\newpage

\hypertarget{posterior-parameter-estimates}{%
\subsection{Posterior parameter
estimates}\label{posterior-parameter-estimates}}

\begin{figure}[!htb]
\centering
\includegraphics{figures/psplots1.pdf}
\caption{Posterior parameter distribution}
\end{figure}
\newpage
\begin{figure}[!htb]
\ContinuedFloat
\captionsetup{list=off,format=cont}
\centering
\includegraphics{figures/psplots2.pdf}
\label{fig:fullps1}
\caption{Posterior parameter distribution}
\end{figure}

\hypertarget{posterior-differences}{%
\subsection{Posterior differences}\label{posterior-differences}}

\hypertarget{key-combination-effect-spl2}{%
\subsubsection{Key-combination effect
(SPL2)}\label{key-combination-effect-spl2}}

The analysed datasets differ to the extent that keystroke intervals at
before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
In other words, the pause before sentences sumed across two key
intervals in the PLanTra and LIFT data, namely
\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
longer or more frequent pauses at before-sentence locations compared to
before-word locations can be explained without reference to linguistic
edges. Also there is a possibility that some inconsistencies in our
findings can be explained on the basis of including the keystroke
following shift.

Therefore we compared whether the different patterns can be explain on
the basis of the additional keystroked involved in before-sentence
transitions. We compared the SPL2 data including and excluding the
keystroke after shift. Although we modelled all transition locations, we
present only before-sentence transitions below as there was, as one
would expect, no difference at word locations. The results of this
comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
fluent transition duration and the hesitation duration were affected by
whether or not the sentence-initial transition include the character
following shift. Fluent key transitions were substantially longer when
including the interval following the shift key. The slowdown for
hesitations was affected too but the difference is numerically small.
There was no conclusive evidence for an increased hesitation
probability.

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions. Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
\toprule
Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  & \\
\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
Hesitation duration &  &  &  & \\
\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
Hesitation probability &  &  &  & \\
\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}

\hypertarget{l2-effect-spl2}{%
\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}

For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
sentence-transitions) we calculated the L2 effect (i.e.~the difference
between writing in L2 and L1). The results can be found in Table
\ref{tab:l2effect}. The results show longer keystrokes and more pauses
across all transition locations. However, the pause duration only
differed at before-word locations, not at within-word or before-sentence
transition locations.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:l2effect}Mixture model estimates for language effect. Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions. The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}

\small{

\begin{tabular}{lrrrr}
\toprule
Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
Fluent transitions &  &  &  & \\
\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
Hesitation duration &  &  &  & \\
\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
Hesitation probability &  &  &  & \\
\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
\bottomrule
\addlinespace
\end{tabular}

}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{tablenotes}

\end{threeparttable}
\end{center}

\end{table}

\hypertarget{masking-effect-cato-gunnexp2}{%
\subsubsection{Masking effect (CATO,
GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}

Studies associated with two datasets (CATO, GUNNEXP2) investigated to
what extent masking the previously written text affects keystroke
behaviour. The mixture model results for the effect of masking is shown
in Table \ref{tab:maskingeffect}. Key transition were longer for masked
text than for unmasked text at before and within word transitions but
not before sentences. Also dyslexics showed no masking difference at
before word transitions. There was no indication of longer or more
frequent pauses as a result of masked text.

\blandscape

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect. Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
\toprule
Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
Hesitation duration &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
Hesitattion probability &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\elandscape

\hypertarget{pre-post-test-plantra}{%
\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}

The pre-post test effect for the PLanTra dataset is reported in Table
\ref{tab:retesteffect}. Transition durations were shorter at before-word
locations for the post-test. Evidence was negligible for any other
comparisons.

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect. Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
\toprule
Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  & \\
\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
Hesitation duration &  &  &  & \\
\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
Hesitation probability &  &  &  & \\
\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}

\hypertarget{genre-effect-lift}{%
\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}

From the LIFT data, we compared the difference between genres, i.e.~when
writing an informative text as opposed to writing an argumentative text.
The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
differences were average across writing topic. Evidence for a difference
between genres was negligible.

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:genreeffect}Mixture model estimates for genre effect. Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
\toprule
Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  & \\
\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
Hesitation duration &  &  &  & \\
\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
Hesitation probability &  &  &  & \\
\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\end{appendix}

### 2023-07-24 report/appendix_a.Rmd
Manuscript with appendices separately

## Fit to data

Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data. From each model we obtained 100 simulations (predictions). These can be found in Figure \ref{fig:prediction}. The model predictions fitted the data substantially better for the mixture models. The worse fit was observed for the unimodal Gaussian model which predicted large amounts of negative keystroke intervals which are implausible. Using a log-normal distribution removed these negative keystroke intervals.

```{r prediction, fig.cap="Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8, fig.width=6.5}

run <- F

if(run){
  theme_set(theme_few(base_size = 9) +
              theme(strip.background = element_blank(),
                    legend.position = "top",
                    legend.justification = "right",
                    panel.grid = element_blank(),
                    panel.background = element_rect(fill = "transparent"), # bg of the panel
                    plot.background = element_rect(fill = "transparent", color = NA),
                    axis.text.y = element_blank(),
                    axis.ticks.y = element_blank(),
                    axis.title = element_blank()))
  file <- list.files("../stanout", pattern = "sims", full.names = T, recursive = T)
  ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
    str_to_upper() %>% 
    str_replace(., "PLANTRA", "PLanTra")
  ds <- ds[-6]
  plots <- list()
  plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
  plots_fin <- plots[[1]] / plots[[2]] / plots[[3]] / plots[[4]] / plots[[5]] / plots[[6]] + plot_layout(guides = "collect") & theme(legend.position = 'bottom')

  plots_fin <- grid.arrange(patchworkGrob(plots_fin), 
                            left = "Density",
                            bottom = "Transition duration in secs")

  ggsave("figures/fitplots.pdf", plots_fin, height = 8.5, width = 6.25)
}

knitr::include_graphics("figures/fitplots.pdf")
```

### 2023-07-24 report/appendix_a.tex
Manuscript with appendices separately

--- 2023-07-24 report/appendix.tex
@@
 \begin{appendix}
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
+\section{}
 \hypertarget{fit-to-data}{%
@@
 \end{figure}
-
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-
-\hypertarget{posterior-differences}{%
-\subsection{Posterior differences}\label{posterior-differences}}
-
-\hypertarget{key-combination-effect-spl2}{%
-\subsubsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{l2-effect-spl2}{%
-\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsubsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-
-\hypertarget{pre-post-test-plantra}{%
-\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{genre-effect-lift}{%
-\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
 \end{appendix}

### 2023-07-24 report/appendix_b.Rmd
Manuscript with appendices separately

## Posterior parameter estimates

```{r}
# get posterior
file <- list.files(path = "../stanout/cato", 
                   pattern = "mogbetaun.+.csv$",
                   full.names = T)

ps_cato <- read_csv(file) %>% 
  filter(param %in% c("beta", "beta2", "prob")) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, beta2, prob)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, group, location, name)) %>% 
  mutate(group = str_c("Task: ", task, "; Group: ", group),
         ds = "CATO") %>% 
  select(-task)

file <- list.files(path = "../stanout/c2l1", 
                   pattern = "mogbetaun.*.csv$",
                   full.names = T)

ps_c2l1 <- read_csv(file) %>% 
  filter(param %in% c("beta", "beta2", "prob")) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, beta2, prob)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, name)) %>% 
  mutate(ds = "C2L1",
         group = "1")

file <- list.files(path = "../stanout/gunnexp2", 
                   pattern = "mogbetaun.*.csv$",
                   full.names = T)

ps_gunn <- read_csv(file) %>% 
  filter(param %in% c("beta", "beta2", "prob")) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, beta2, prob)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, xn, name)) %>% 
  mutate(ds = "GUNNEXP2") %>% 
  rename(group = xn)

file <- list.files(path = "../stanout/lift", 
                   pattern = "mogbetaun.+.csv$",
                   full.names = T)

ps_lift <- read_csv(file) %>% 
  filter(param %in% c("beta", "beta2", "prob")) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, beta2, prob)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(genre, topic, location, name)) %>% 
  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre),
         ds = "LIFT") %>% 
  select(-topic, -genre)

file <- list.files(path = "../stanout/plantra", 
                   pattern = "mogbetaun.+.csv$",
                   full.names = T)

ps_plantra <- read_csv(file) %>% 
  mutate(across(task, factor, 
                levels = c("pre-test", "post-test"), 
                ordered = T)) %>% 
  filter(param %in% c("beta", "beta2", "prob")) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, beta2, prob)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, location, name)) %>% 
  mutate(ds = "PLanTra") %>% 
  rename(group = task)

file1 <- list.files(path = "../stanout/spl2", 
                    pattern = "mogbetaun.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../stanout/spl2_shift", 
                    pattern = "mogbetaun.+.csv$",
                    full.names = T)

ps_spl2 <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
                     mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
  filter(param %in% c("beta", "delta", "prob")) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2")) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(beta2 = beta + delta,
         across(c(beta2, beta), ~exp(.)),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper),
                   .names = "{.fn}"),
            .by = c(lang, location, name, data)) %>% 
  mutate(ds = "SPL2",
         group = str_c("Data: ", data, "; Language: ", lang)) %>% 
  select(-lang)

# Combine all
ps <- bind_rows(ps_cato, ps_c2l1, ps_gunn, ps_lift, ps_plantra, ps_spl2)
```

```{r eval = F}
theme_set(theme_few(base_size = 11) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))

pdodge <- position_dodge(-.5)
beta_plots <- list()
delta_plots <- list()
theta_plots <- list()
datasets <- unique(ps$ds)
size <- 2
width <- 0
linewidth <- .35

beta_plots <- map(datasets, ~filter(ps, 
                                    name == "beta", 
                                    ds == .x) %>% 
                    mutate(across(c(ds,group), ~factor(.)),
                           name = beta_label) %>% 
                    ggplot(aes(y = mean, 
                               ymin = lower, 
                               ymax = upper,
                               x = location,
                               shape = group)) +
                    geom_errorbar(width = width,
                                  linewidth = linewidth,
                                  alpha = .75, 
                                  position = pdodge) +
                    geom_point(aes(colour = group), 
                               size = size, 
                               position = pdodge) +
                    facet_grid(~name) +
                    scale_shape_manual(values = 1:length(filter(ps, ds == .x) %>% 
                                                           pull(group) %>% 
                                                           factor())) +
                    scale_colour_colorblind() +
                    scale_y_continuous(labels = scales::comma) +
                    labs(subtitle = .x) +
                    theme(axis.title = element_blank(),
                          legend.title = element_blank()) +
                    guides(colour=guide_legend(nrow=2,byrow=TRUE),
                           shape = guide_legend(nrow=2,byrow=TRUE)))

delta_plots <- map(datasets, ~filter(ps, 
                                     name == "delta", 
                                     ds == .x) %>% 
                     mutate(across(c(ds,group), ~factor(.)),
                            across(c(mean, lower, upper), ~ . /1000),
                            name = delta_label) %>% 
                     ggplot(aes(y = mean, 
                                ymin = lower, 
                                ymax = upper,
                                x = location,
                                shape = group)) +
                     geom_errorbar(width = width, 
                                   alpha = .75, 
                                   linewidth = linewidth,
                                   position = pdodge) +
                     geom_point(aes(colour = group), 
                                size = size, 
                                position = pdodge) +
                     facet_grid(~name) +
                     scale_shape_manual(values = 1:length(filter(ps, ds == .x) %>% 
                                                            pull(group) %>% 
                                                            factor())) +
                     scale_colour_colorblind() +
                     scale_y_log10(labels = scales::comma) +
                     theme(axis.title = element_blank(),
                           legend.title = element_blank()) +
                     guides(colour=guide_legend(nrow=2,byrow=TRUE),
                            shape = guide_legend(nrow=2,byrow=TRUE)))

theta_plots <- map(datasets, ~filter(ps, 
                                     name == "prob", 
                                     ds == .x) %>% 
                     mutate(across(c(ds,group), ~factor(.)),
                            name = theta_label) %>% 
                     ggplot(aes(y = mean, 
                                ymin = lower, 
                                ymax = upper,
                                x = location,
                                shape = group)) +
                     geom_errorbar(width = width,
                                   linewidth = linewidth,
                                   alpha = .75, 
                                   position = pdodge) +
                     geom_point(aes(colour = group), 
                                size = size, 
                                position = pdodge) +
                     facet_grid(~name) +
                     scale_shape_manual(values = 1:length(filter(ps, ds == .x) %>% 
                                                            pull(group) %>% 
                                                            factor())) +
                     scale_colour_colorblind() +
                     scale_y_continuous(labels = dezero_plot, 
                                        limits = c(0, 1)) +
                     theme(axis.title = element_blank(),
                           legend.title = element_blank()) +
                     guides(colour=guide_legend(nrow=2,byrow=TRUE),
                            shape = guide_legend(nrow=2,byrow=TRUE)))

# Combine parameter plots 
plot_param <- list()
plot_param <- map(1:6, ~beta_plots[[.]] + delta_plots[[.]] + theta_plots[[.]] +
                    plot_layout(guides = "collect") &
                    theme(legend.position = "bottom"))

plot_param[[2]] <- plot_param[[2]] & theme(legend.position = "none")
plot_param[[3]] <- plot_param[[3]] & guides(colour=guide_legend(nrow=1,byrow=TRUE),
                                            shape = guide_legend(nrow=1,byrow=TRUE))

plot_param[[5]] <- plot_param[[5]] & guides(colour=guide_legend(nrow=1,
                                                                byrow=TRUE,
                                                                reverse=TRUE),
                                            shape = guide_legend(nrow=1,
                                                                 byrow=TRUE,
                                                                 reverse=TRUE))

plot_fin_1 <- plot_param[[1]] / plot_param[[2]] / plot_param[[3]]
plot_fin_1 <- grid.arrange(patchworkGrob(plot_fin_1), 
                           left = "Posterior estimate with 95% PIs",
                           bottom = "Transition location")
ggsave("figures/psplots1.pdf", plot_fin_1, height = 8, width = 7)

plot_fin_2 <- plot_param[[4]] / plot_param[[5]] / plot_param[[6]]
plot_fin_2 <- grid.arrange(patchworkGrob(plot_fin_2), 
                           left = "Posterior estimate with 95% PIs",
                           bottom = "Transition location")
ggsave("figures/psplots2.pdf", plot_fin_2, height = 8, width = 7)
```

\begin{figure}[!htb]
\centering
\includegraphics{figures/psplots1.pdf}
\caption{Posterior parameter distribution}
\end{figure}
\newpage
\begin{figure}[!htb]
\ContinuedFloat
\captionsetup{list=off,format=cont}
\centering
\includegraphics{figures/psplots2.pdf}
\label{fig:fullps1}
\caption{Posterior parameter distribution}
\end{figure}

### 2023-07-24 report/appendix_c.Rmd
Manuscript with appendices separately

--- 2023-07-24 report/appendix.tex
@@
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
+## Key-combination effect (SPL2)
 
-\begin{appendix}
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
+The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+In other words, the pause before sentences sumed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets.
+Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
+Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.
 
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
+Therefore we compared whether the different patterns can be explain on the basis of the additional keystroked involved in before-sentence transitions.
+We compared the SPL2 data including and excluding the keystroke after shift.
+Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.
+Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift.
+Fluent key transitions were substantially longer when including the interval following the shift key.
+The slowdown for hesitations was affected too but the difference is numerically small.
+There was no conclusive evidence for an increased hesitation probability.
 
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
+```{r shiftcellmeans, results = 'asis'}
+# cell means and data set difference
+cellmeans <- ps_spl2 %>% 
+  separate(group, into = c("group", "lang"), sep = "; ") %>%
+  select(-ds, -group) %>%
+  mutate(across(lang, ~str_remove(., "Language: ")),
+         across(location, ~str_replace(., "\\n", " "))) %>% 
+  pivot_wider(names_from = name, 
+              values_from = c(mean, lower, upper)) %>% 
+  mutate(across(ends_with("prob"), ~dezero(., 2)),
+         across(c(ends_with("beta"),
+                  ends_with("delta")), 
+                ~as.character(scales::comma(round(.))))) %>% 
+  pivot_longer(-c(lang, location, data), 
+               names_to = c(".value", "param"),
+               names_sep = "_") %>% 
+  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
+  select(-mean:-upper) %>% 
+  pivot_wider(names_from = data, 
+              values_from = est)
 
-\begin{figure}
+file1 <- list.files(path = "../stanout/spl2", 
+                    pattern = "mogbetaun.+.csv$",
+                    full.names = T)
 
-{\centering \includegraphics{figures/fitplots}
+file2 <- list.files(path = "../stanout/spl2_shift", 
+                    pattern = "mogbetaun.+.csv$",
+                    full.names = T)
 
-}
+tmp <- map_dfr(c(file1, file2), ~read_csv(.x) %>% 
+                 mutate(data = .x)) %>% 
+  filter(param %in% c("beta", "delta", "theta")) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
+  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
+         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
+  pivot_wider(names_from = data, values_from = value) %>% 
+  unnest(c(spl2, spl2_shift)) %>% 
+  mutate(diff = spl2_shift - spl2) %>% 
+  summarise(across(diff, list(mean = mean, 
+                              lower = lower, 
+                              upper = upper,
+                              BF = BF), 
+                   .names = "{.fn}"),
+            .by = c(lang, location, param)) %>% 
+  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
+         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
+         across(param, recode, theta = "prob")) %>% 
+  left_join(cellmeans) %>% 
+  arrange(param, location, lang) %>% 
+  filter(location == "before sentence") %>% 
+  select(param, 
+         Language = lang,
+         `_^[shift] + C` = ends_with(" + C)"),
+         `_^[shift]` = ends_with("FT])"),
+         "Difference" = mean, BF)
 
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
-\end{figure}
+write_csv(tmp, "tables/shift_effect_cellmeans.csv")
 
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-
-\hypertarget{posterior-differences}{%
-\subsection{Posterior differences}\label{posterior-differences}}
-
-\hypertarget{key-combination-effect-spl2}{%
-\subsubsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
+apa_table(tmp[,-1],
+          align = c("l", "r", "r", "r", "r"), 
+          caption = "Mixture model estimates for key transitions.
 Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
 The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
+95% PIs in brackets.",
+          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
+          font_size = "footnotesize",
+          longtable = T,
+          stub_indents = list("Fluent transitions" = 1:2,
+                              "Hesitation duration" = 3:4,
+                              "Hesitation probability" = 5:6) )
 
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{l2-effect-spl2}{%
-\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsubsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-
-\hypertarget{pre-post-test-plantra}{%
-\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{genre-effect-lift}{%
-\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
+```

### 2023-07-24 report/appendix_c.tex
Manuscript with appendices separately

--- 2023-07-24 report/appendix.tex
@@
 \begin{appendix}
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
-\end{figure}
-
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-
-\hypertarget{posterior-differences}{%
-\subsection{Posterior differences}\label{posterior-differences}}
-
+\section{}
 \hypertarget{key-combination-effect-spl2}{%
-\subsubsection{Key-combination effect
+\subsection{Key-combination effect
 (SPL2)}\label{key-combination-effect-spl2}}
@@
 \end{center}
-
-\hypertarget{l2-effect-spl2}{%
-\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsubsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-
-\hypertarget{pre-post-test-plantra}{%
-\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{genre-effect-lift}{%
-\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
 \end{appendix}

### 2023-07-24 report/appendix_d.Rmd
Manuscript with appendices separately

## L2 effect (SPL2)

For the SPL2 data (only the `_^[shift]` sentence-transitions) we calculated the L2 effect (i.e. the difference between writing in L2 and L1). The results can be found in Table \ref{tab:l2effect}. The results show longer keystrokes and more pauses across all transition locations. However, the pause duration only differed at before-word locations, not at within-word or before-sentence transition locations.

```{r l2effect, results = 'asis'}
# cell means and data set difference
cellmeans <- cellmeans %>% 
  select(-`SPL2 (_^[shift] + C)`) %>% 
  pivot_wider(names_from = lang, values_from = `SPL2 (_^[shift])`)

file1 <- list.files(path = "../stanout/spl2", 
                    pattern = "mogbetaun.+.csv$",
                    full.names = T)

tmp <- map_dfr(file1, read_csv) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2")) %>% 
  pivot_wider(names_from = lang, values_from = value) %>% 
  unnest(c(L1, L2)) %>% 
  mutate(diff = L2 - L1) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, recode, theta = "prob")) %>% 
  left_join(cellmeans) %>% 
  arrange(param, location) %>% 
  select(param, `Transition location` = location, L1, L2,
         "Difference" = mean, BF)

write_csv(tmp, "tables/l2_effect_cellmeans.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions. The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          stub_indents = list("Fluent transitions" = 1:3,
                              "Hesitation duration" = 4:6,
                              "Hesitation probability" = 7:9) )

```

### 2023-07-24 report/appendix_d.tex
Manuscript with appendices separately

--- 2023-07-24 report/appendix.tex
@@
 \begin{appendix}
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
-\end{figure}
-
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-
-\hypertarget{posterior-differences}{%
-\subsection{Posterior differences}\label{posterior-differences}}
-
-\hypertarget{key-combination-effect-spl2}{%
-\subsubsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
+\section{}
 \hypertarget{l2-effect-spl2}{%
-\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
+\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
 
@@
 \end{table}
-
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsubsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-
-\hypertarget{pre-post-test-plantra}{%
-\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{genre-effect-lift}{%
-\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
 \end{appendix}

### 2023-07-24 report/appendix_e.Rmd
Manuscript with appendices separately

## Masking effect (CATO, GUNNEXP2)

Studies associated with two datasets (CATO, GUNNEXP2) investigated to what extent masking the previously written text affects keystroke behaviour. The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}. Key transition were longer for masked text than for unmasked text at before and within word transitions but not before sentences. Also dyslexics showed no masking difference at before word transitions. There was no indication of longer or more frequent pauses as a result of masked text.

\blandscape
```{r maskingeffect, results = 'asis'}
ps_cato_2 <- ps_cato %>% 
  separate(group, into = c("mask","group"), sep = "; ") %>% 
  mutate(across(c(mask, group), ~str_remove(., "Task: |Group: ")))

cellmeans_mask <- ps_gunn %>% 
  rename(mask = group) %>%
  mutate(group = "non dyslexic") %>% 
  bind_rows(ps_cato_2) %>% 
  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, group, mask, ds), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = mask, 
              values_from = est)

file <- list.files(path = "../stanout/gunnexp2", 
                   pattern = "mogbetaun.*.csv$",
                   full.names = T)

d_gunn <- read_csv(file) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value)),
         ds = "gunnexp2",
         across(param, ~recode(., "prob" = "theta")),
         group = "non dyslexic") %>% 
  rename(mask = xn)

# get posterior
file <- list.files(path = "../stanout/cato", 
                   pattern = "mogbetaun.+.csv$",
                   full.names = T)

d_cato <- read_csv(file) %>% 
  filter(param %in% c("beta", "delta", "prob")) %>% 
  mutate(ds = "cato") %>% 
  rename(mask = task)

tmp <- bind_rows(d_gunn, d_cato) %>% 
  pivot_wider(names_from = mask, values_from = value) %>% 
  unnest(c(masked, unmasked)) %>% 
  mutate(diff = masked - unmasked) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(location, param, group, ds)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, ~recode(., theta = "prob")),
         across(ds, ~str_to_upper(.))) %>% 
  left_join(cellmeans_mask) %>% 
  arrange(param, location, ds, group) %>% 
  select(param, `Transition location` = location,
         `Dataset` = ds, Group = group, 
         Unmasked = unmasked,
         Masked = masked,
         "Difference" = mean, BF)

write_csv(tmp, "tables/masking_effect_cellmeans.csv")

apa_table(tmp[,-1],
          align = c("l", "l", "l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for masking effect. Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Fluent transitions" = 1:9, 
            "Hesitation duration" = 10:18,
            "Hesitattion probability" = 19:27) )

```
\elandscape

### 2023-07-24 report/appendix_e.tex
Manuscript with appendices separately

--- 2023-07-24 report/appendix.tex
@@
 \begin{appendix}
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
-\end{figure}
-
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-
-\hypertarget{posterior-differences}{%
-\subsection{Posterior differences}\label{posterior-differences}}
-
-\hypertarget{key-combination-effect-spl2}{%
-\subsubsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{l2-effect-spl2}{%
-\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
+\section{}
 \hypertarget{masking-effect-cato-gunnexp2}{%
-\subsubsection{Masking effect (CATO,
+\subsection{Masking effect (CATO,
 GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
@@
 \elandscape
-
-\hypertarget{pre-post-test-plantra}{%
-\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{genre-effect-lift}{%
-\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
 \end{appendix}

### 2023-07-24 report/appendix_f.tex
Manuscript with appendices separately

--- 2023-07-24 report/appendix.tex
@@
 \begin{appendix}
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
-\end{figure}
-
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-
-\hypertarget{posterior-differences}{%
-\subsection{Posterior differences}\label{posterior-differences}}
-
-\hypertarget{key-combination-effect-spl2}{%
-\subsubsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{l2-effect-spl2}{%
-\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsubsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-
-\hypertarget{pre-post-test-plantra}{%
-\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
+\section{}
 \hypertarget{genre-effect-lift}{%
-\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}
+\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
 
@@
 \caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
 The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.

### 2023-07-24 report/appendix_g.Rmd
Manuscript with appendices separately

## Pre-post test (PLanTra)

The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}. Transition durations were shorter at before-word locations for the post-test. Evidence was negligible for any other comparisons.

```{r retesteffect, results = 'asis'}
cellmeans <- ps_plantra %>% 
  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, group, ds), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper, -ds) %>% 
  pivot_wider(names_from = group, values_from = est)

file <- list.files(path = "../stanout/plantra", 
                   pattern = "mogbetaun.*.csv$",
                   full.names = T)

tmp <- read_csv(file) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  pivot_wider(names_from = task, values_from = value) %>% 
  unnest(c(`post-test`,`pre-test`)) %>% 
  mutate(diff = `pre-test` - `post-test`) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, ~recode(., theta = "prob"))) %>% 
  left_join(cellmeans) %>% 
  arrange(param, location) %>% 
  select(param, 
         `Transition location` = location,
         `Pre-test` = `pre-test`,
         `Post-test` = `post-test`, 
         "Difference" = mean, BF)

write_csv(tmp, "tables/posttest_effect_cellmeans.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for post-test effect. Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Fluent transitions" = 1:3, 
            "Hesitation duration" = 4:6,
            "Hesitation probability" = 7:9))

```

### 2023-07-24 report/appendix_g.tex
Manuscript with appendices separately

--- 2023-07-24 report/appendix.tex
@@
 \begin{appendix}
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
-\end{figure}
-
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-
-\hypertarget{posterior-differences}{%
-\subsection{Posterior differences}\label{posterior-differences}}
-
-\hypertarget{key-combination-effect-spl2}{%
-\subsubsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\hypertarget{l2-effect-spl2}{%
-\subsubsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsubsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-
+\section{}
 \hypertarget{pre-post-test-plantra}{%
-\subsubsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
+\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
 
@@
 \end{center}
-
-\hypertarget{genre-effect-lift}{%
-\subsubsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for the argumentative and informative texts in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
 \end{appendix}

### 2023-07-24 report/manuscript.Rmd
Manuscript with appendices separately

@@
     keep_tex: TRUE
+    includes:
+      after_body: 
+        - "appendix_a.tex"
+        - "appendix_b.tex"
+        - "appendix_c.tex"
+        - "appendix_d.tex"
+        - "appendix_e.tex"
+        - "appendix_g.tex"
+        - "appendix_f.tex"
 
@@
   - \newcommand{\elandscape}{\end{landscape}}
+  - \DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
 ---
-
-```{r packages, include=FALSE}
-library(tidyverse)
-library(kableExtra)
-library(janitor)
-library(brms)
-library(polspline)
-library(patchwork)
-library(rmdformats)
-library(gridExtra)
-library(grid)
-library(gtable)
-library(scales)
-library(ggthemes)
-source("../scripts/functions.R")
-source("scripts/get_pdens_plot.R")
-
-theme_set(theme_few(base_size = 9) +
-            theme(strip.background = element_blank(),
-                  legend.position = "top",
-                  legend.justification = "right",
-                  panel.grid = element_blank(),
-                  panel.background = element_rect(fill = "transparent"), # bg of the panel
-                  plot.background = element_rect(fill = "transparent", color = NA)))
-
-options(scipen = 999)
-
-beta_label <- "Short key transition\nduration"
-delta_label <- "Slowdown for hesitant key\ntransitions"
-theta_label <- "Probability of hesitant\ntransitions"
-
-```
 
@@
 
+```{r libs, include=FALSE}
+library(tidyverse)
+library(janitor)
+library(brms)
+library(papaja)
+library(polspline)
+library(patchwork)
+library(gridExtra)
+library(grid)
+library(gtable)
+library(scales)
+library(ggthemes)
+source("../scripts/functions.R")
+source("scripts/get_pdens_plot.R")
+
+theme_set(theme_few(base_size = 12) +
+            theme(strip.background = element_blank(),
+                  legend.position = "top",
+                  legend.justification = "right",
+                  panel.grid = element_blank(),
+                  panel.background = element_rect(fill = "transparent"), # bg of the panel
+                  plot.background = element_rect(fill = "transparent", color = NA)))
+
+options(scipen = 999)
+
+beta_label <- "Short key transition\nduration (in msecs)"
+delta_label <- "Slowdown for hesitant key\ntransitions (in secs)"
+theta_label <- "Probability of hesitant\ntransitions"
+```
+
 # Introduction
@@
 
-### Uni-modal Gaussian
+### Unimodal Gaussian
 
@@
 
-$$
-\begin{align}(\#eq:unimodgaus)
+\begin{equation}
+\begin{aligned}
+(\#eq:unimodgaus)
 \text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
 \text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
-& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
-\end{align}
-$$
+& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
 
-### Uni-modal log-Gaussian
+### Unimodal log-Gaussian
 
@@
 
-$$
-\begin{align}(\#eq:unimodloggaus)
+\begin{equation}
+\begin{aligned}
+(\#eq:unimodloggaus)
 \text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
@@
 \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
-& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
-\end{align}
-$$
+& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
 
-### Uni-modal unequal-variance log-Gaussian
+### Unimodal unequal-variance log-Gaussian
 
-$$
-\begin{align}(\#eq:unimoduv)
+\begin{equation}
+\begin{aligned}
+(\#eq:unimoduv)
 \text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
 \text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
- & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
-\end{align}
-$$
+ & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+ \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
 
-### Bi-modal log-Gaussian (constrained)
+### Bimodal log-Gaussian (constrained)
 
@@
 
-$$
-\begin{align}(\#eq:bimodcon)
+\begin{equation}
+\begin{aligned}
+(\#eq:bimodcon)
 \text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
   & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: }
-		&\delta \sim \text{N}(0,1)\\
- 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta > 0\\
+\text{where: } & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
 		& \sigma_{e'}^2 > \sigma_{e}^2
-\end{align}
-$$
+\end{aligned}
+\end{equation}
+
+```{r}
+#		<!-- &\delta \sim \text{N}(0,1)\\ -->
+
+```
+
 This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
 
-### Bi-modal log-Gaussian (unconstrained)
+### Bimodal log-Gaussian (unconstrained)
 
@@
 
-$$
-\begin{align}(\#eq:bimoduncon)
+\begin{equation}
+\begin{aligned}
+(\#eq:bimoduncon)
 \text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
   & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-	\text{where: } & \delta \sim \text{N}(0,1)\\
-	 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta > 0\\
+	\text{where: }  & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
 		& \sigma_{e'}^2 > \sigma_{e}^2
-\end{align}
-$$
+\end{aligned}
+\end{equation}
+
+```{r}
+#<!-- & \delta \sim \text{N}(0,1)\\ -->
+
+```
 
@@
 
-Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).
+Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \ref{tab:datasets}.
 
+\blandscape
 ```{r datasets, results='asis'}
 tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
-                Source = c("@ronneberg2022process",
-                              "@torrance2016adolescent",
-                              "@torrance",
-                              "@rossetti2022s",
-                              "@nina_vandermeulen_2020_3893538", 
-                           ""),
+                Source = c("Rønneberg et al. (2022)",
+                           "Torrance et al. (2016)",
+                           "Torrance et al. (n.d.)",
+                           "Rossetti and Van Waes (2022)",
+                           "Vandermeulen et al. (2020)", 
+                           "Torrance and Ofstad (n.d)"),
                 Keylogger = c("EyeWrite", 
@@
                            "EyeWrite"),
-                `Writing task` = c(
-                                 "Argumentative essays", 
+                `Writing task` = c("Argumentative essays", 
                                  "Expository texts",
@@
                                  ""),
-                 `N (ppts)` = c(126, 26*2, 39, 47, 658, 45),
+                 `N (ppts)` = as.character(c(126, 26*2, 39, 47, 658, 45)),
 #                n_texts = c(1, 2, 2, 2, NA),
@@
 #                n_words = c(),
-                conditions = c(" ", 
-                               "weak decoders / control; masked / unmasked",
-                               "write in L1 / L2",
-                               "pre / post test trained in plain language principles and control", 
-                               "Various topics and genres",
-                               "masked / unmasked"),
                 `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
@@
                                "Dutch",
-                               "Norwegian")) %>% 
-  kable(caption = "Datasets in brief.") %>%
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 12)
+                               "Norwegian"),
+                Manipulation = c("--", 
+                               "weak decoders / control; masked / unmasked",
+                               "write in L1 / L2",
+                               "pre / post test trained in plain language principles and control", 
+                               "Various topics and genres",
+                               "masked / unmasked")) %>% 
+  arrange(Dataset) %>% 
+  apa_table(align = c(rep("l", 4), rep("r", 2), "p{2cm}", "p{3cm}"), 
+            escape = FALSE, 
+            font_size = "footnotesize",
+            caption = "Datasets in brief.")
 
 ```
+\elandscape
 
@@
 
+TODO: Mark, can you add information here? I've added a provisional bib entry that needs changing or we just say "unpublished"
+
+The GUNNEXP2 dataset published in @torranceb includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
+In this dataset participants wrote text either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
+Keystroke data were captured using EyeWrite [@sim07; XXXX].
+
 ### C2L1
+
+TODO: Mark, can you add information here?
+
+TODO: might need to remove kids that don't speak Norwegian at home (see github issue)?
 
@@
 The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue).
+Keystroke data were captured using EyeWrite [@sim07; XXXX].
 
@@
 
-Data are published in @torrance2016adolescent. Norwegian upper
-secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
+TODO: Mark, can you add information here?
+
+Data are published in @torrance2016adolescent.
+Norwegian upper secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
+Keystroke data were captured using EyeWrite [@sim07; XXXX].
 
 ### PLanTra
+
+TODO: Rianne, can you add information here?
 
@@
 It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
+Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
 
@@
 
-LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].
+TODO: Rianne, can you add information here?
+
+The LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) data are published in @vandermeulen2020.
+Students produced text for 4 topics in two different writing genres (levels: argumentative text, informative text).
+Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
 
@@
 
-Data are going to be published in @torrance.
+TODO: Mark, can you add information here?
 
-Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
+The SPL2 dataset is published in @torrancea.
+The data come from a textcomposition task of undergraduate university students -- *N* = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
 CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
 We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
-
-Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
+Participants were given a 40 minute time limit.
+They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
 
 ## Data analysis
+
+### Bayesian modelling
+
+We reanalysed keystroke data from 6 datasets in Bayesian implementations of 5 statistical models. 
+Bayesian models, as used in this paper, are ideal for the estimation of parameter values: Bayesian parameter estimates are expresses the uncertainty associated with parameter values as probability distribution [@farrell2018computational; @gelman2014; @lee2014bayesian].
+To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
+existing knowledge about parameter values, for regulating the parameter space.
+For small data sets priors influence the inferred parameter estimates (known as posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
+automatic Ockham's razor, @jefferys1992ockham].
+In other words the choice of priors values has less impact on the posterior.
+In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
+@lambert2018student; @mcelreath2016statistical].
+The predictive performance (i.e.
+fit) of these models is compared in the Results section using leave-one-out cross-validation.
+
+Keystroke data from text composition tasks were modelled in a series of five models. An overview of all models can be found in Table \ref{tab:models}.
+
+```{r models, results = 'asis'}
+models <- tibble(Models = paste0("M",1:5),
+       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
+       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
+       Description = c("Gaussian distribution",
+                       "Log-normal distribution",
+                       "Log-normal distribution and unequal variances",
+                       "Fluent transition are not allowed to vary by transition location",
+                       "Fluent transition are allowed to vary by transition location"))
+
+apa_table(models, 
+          align = c(rep("l", 2), "r", "p{9cm}"), 
+          escape = FALSE, 
+          digits = 0,
+          caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
+          note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
+```
+
+For all models we modelled transition locations (levels: before sentence, before word, within word) was included as predictor in all models.
+Also, we included the dataset specific experimental conditions as fixed effects and included random intercepts for participants as described in the equations above.
+We used normal and log-normal probability function in line with typically treatments fouund in the literature.
+Stan code for mixture models was based on @roeser2021modelling  [see also @vasishth2017; @vasishth2017feature].
+
+Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
+The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
+Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
+Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
 
@@
 The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g.
-@torrance; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \@ref(tab:keyloc).
+@torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
 Keytransitions that terminated in an editing operation were excluded from the analysis.
@@
 ```{r keyloc, results='asis'}
-tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
+
+table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
        Description = c("Transitions between any letter",
@@
                        "Keypress following a space preceding any letter"),
-       Example = c("T\\^h\\^e c\\^a\\^t m\\^e\\^o\\^w\\^e\\^d. T\\^h\\^a\\^t[bsp][bsp]e\\^n i\\^t s\\^l\\^e\\^p\\^t.", 
-                   "The \\^cat \\^meowed. That[bsp][bsp]en \\^it \\^slept.", 
-                   "The cat meowed. \\^That[bsp][bsp]en it slept.")) %>% 
-  kable(caption = "Transition location classification.") %>%
-  footnote("'\\^' marks transition location, [bsp] represents backspace. IKIs were timed to the shift keypress.") %>% 
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 12)
+       Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d.
+T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.",
+                   "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
+                   "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept."))
+
+apa_table(table, caption = "Transition location classification.",
+            align = c("p{3cm}", "p{3cm}", "p{8cm}"), 
+            escape = FALSE, 
+            font_size = "footnotesize",
+            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace. At before-sentence locations, IKIs were timed to the shift keypress.")
 
@@
 source("scripts/get_data_summary.R")
-```
 
-```{r}
 # Get info about data reduction
@@
 
-We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \@ref(tab:datareduction).
+We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
 
@@
 Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \@ref(tab:datareduction).
+The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
 
@@
           mutate(ds = .x$ds)) %>% 
-  mutate(across(where(is.numeric), ~round(.*100, 2)),
-         across(starts_with("too"), ~str_c(., "%"))) %>% 
+  mutate(across(where(is.numeric), ~round(.*100, 2))) %>% 
   unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
@@
           mutate(ds = .x$ds)) %>% 
-    mutate(across(where(is.numeric), ~round(.*100, 1)),
-           across(starts_with("keep"), ~str_c(., "%"))) %>%
+    mutate(across(where(is.numeric), ~round(.*100, 1))) %>%
     unite("keep", starts_with("keep"), sep = " (") %>% 
@@
   relocate(ds) %>% 
-  mutate(across(ds, recode, 
+  mutate(across(ds, ~recode(., 
                 "cato" = "CATO",
@@
                 "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2")) %>% 
+                "gunnexp2" = "GUNNEXP2"))) %>% 
   rename(`Dataset` = ds)
 
-kable(datareduction, 
-      caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition locattion. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 14))) %>%
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 11) %>%
-  add_header_above(c(" " = 1, 
-                     "Extreme values" = 2,
-                     "Randomly sampled data" = 3)) %>%
-  column_spec(1, width = "8em")
+apa_table(datareduction, 
+          escape = FALSE,  
+          placement = "bp!",
+          caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location. Standard error is shown in parentheses.",
+          align =c("l", rep("r", 5)),
+          col_spanners = list(" " = 1,
+                              "Extreme values in \\%" = c(2, 3),
+                              "Randomly sampled data in \\%" = c(4, 6)))
 
 ```
-
-### Bayesian modelling
-
-Keystroke data from text composition tasks were modelled in a series of five models. An overview of all models can be found in Table \ref{tab:models}.
-
-```{r models, results = 'asis'}
-models <- tibble(Models = paste0("M",1:4),
-       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
-       Equation = paste0("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "}"),
-       Description = c("Gaussian distribution",
-                       "Log-normal distribution",
-                       "Log-normal distribution and unequal variances",
-                       "Fluent transition are not allowed to vary by transition location",
-                       "Fluent transition are allowed to vary by transition location"))
-
-papaja::apa_table(models, align = c("l", "l", "l"), escape = FALSE, digits = 0, placement = "bp!",
-                  caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
-                  note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
-```
-
-each with random effects for participants.
-Probability functions used were normal and log-normal in line with typically treatments used in the literature, a log-normal distribution with unequal variances for model predictions, and a bimodal mixed effects model.
-Stan code for mixture models was based on @roeser2021modelling.
-Text locations (levels: before sentence, before word, within word) was included as predictor in all models.
-
-Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
-The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
 
@@
 
-For model comparisons we used out-of-sample predictions estimated using Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
-Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and the difference $\Delta\widehat{elpd}$ between models.
+To compare the out-of-sample predictive performance of the models we used Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
+Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and its  difference $\Delta\widehat{elpd}$ between models.
 The advantage of using leave-one-out cross-validation is that models with more parameters are penalised to prevent overfit.
 
-Results for all data sets are shown in Table \@ref(tab:loos).
+Results for all data sets are shown in Table \ref{tab:loos}.
 For all data sets we found the same pattern.
-The mixture of log-normal distributions provided a substantially better fit than uni-modal distribution models.
-The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across conditions.
+The mixture of log-normal distribution models provided a substantially better fit than any of the uni-modal distribution models.
+The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across transition locations.
+
+We also evaluated to what extent model predictions fit observed data.
+These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
 
@@
   mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(where(is.numeric), round, 0),
-         across(where(is.numeric), format, big.mark = ","),
-         across(everything(), str_trim)) %>%
+         across(where(is.numeric), ~round(., 0)),
+         across(where(is.numeric), ~format(., big.mark = ",")),
+         across(everything(), ~str_trim(.))) %>%
+  filter(dataset != "spl2_shift") %>% 
   transmute(
-     across(dataset, recode, 
+     across(dataset, ~recode(., 
                 "cato" = "CATO",
                 "spl2" = "SPL2", 
-                "spl2_shift" = "SPL2 (shift + C)", 
+#                "spl2_shift" = "SPL2 (shift + C)", 
                 "plantra" = "PLanTra",
@@
                 "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2"),
-    across(model, recode, lmm = "Unimodal log-normal",
+                "gunnexp2" = "GUNNEXP2")),
+    m = recode(model, 
+               lmmgaus = "M1",
+               lmm = "M2",
+               lmmuneqvar = "M3",
+               mogbetaconstr = "M4",
+               mogbetacontr = "M4",
+               mogbetaunconstr = "M5"),
+    across(model, ~recode(., lmm = "Unimodal log-normal",
                           lmmgaus = "Unimodal normal",
-                          mogbetaconstr = "Bimodal log-normal (constrained)",
-                          mogbetacontr = "Bimodal log-normal (constrained)",
-                          mogbetaunconstr = "Bimodal log-normal (unconstrained)",
-                          lmmuneqvar = "Unimodal log-normal (unequal variance)"),
-            across(elpd_diff, str_c, " (", se_diff, ")"),
-            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
-  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
-  pivot_wider(names_from = dataset, values_from = contains("elpd")) %>% 
-  select(Model = model, ends_with("GUNNEXP2"), ends_with("CATO"), ends_with("C2L1"), ends_with("LIFT"), ends_with("PLanTra"), ends_with("SPL2"), ends_with("+ C)"))
-
+                          mogbetaconstr = "Bimodal (constrained)",
+                          mogbetacontr = "Bimodal (constrained)",
+                          mogbetaunconstr = "Bimodal (unconstrained)",
+                          lmmuneqvar = "Unimodal (unequal variance)")),
+            across(elpd_diff, ~str_c(., " (", se_diff, ")")),
+            across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")"))) %>% 
+  mutate(across(elpd_diff, ~str_replace(., "0 \\(0\\)", "--"))) %>% 
+  arrange(dataset)
 ```
@@
 ```{r loos, results='asis'}
-names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 8)
+names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
 
-kable(table, 
+apa_table(table[,-1], 
       caption = "Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 14))) %>%
-  kable_styling(position = "center", 
-                full_width = T, 
-                font_size = 11) %>%
-  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
-  add_header_above(c(" " = 1, 
-                     "GUNNEXP2" = 2,
-                     "CATO" = 2,
-                     "CL21" = 2, 
-                     "LIFT" = 2,
-                     "PLanTra" = 2,
-                     "SPL2" = 2,
-                     "SPL2 (shift + C)" = 2)) %>%
-  column_spec(1, width = "8em")
+      align =c("p{3.5cm}", "p{6cm}", rep("r", 2)), 
+      escape = FALSE,
+      longtable = T,
+      font_size = "small",
+      note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
+      stub_indents = list(
+                     "\\textbf{CL21}" = 1:5,
+                     "\\textbf{CATO}" = 6:10,
+                     "\\textbf{GUNNEXP2}" = 11:15, 
+                     "\\textbf{LIFT}" = 16:20,
+                     "\\textbf{PLanTra}" = 21:25,
+                     "\\textbf{SPL2}" = 26:30))
+#                     "\\textbf{SPL2 (shift + C)}" = 31:35))
 
@@
 ## Cross-data set comparisons
-
-```{r}
-# Get the differences for tasks / across data set
-# use effect size = beta / sd
-# averaged across grouping variable
-
-```
 
@@
   mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, recode_factor, 
+         across(data, ~recode_factor(., 
                 "cato" = "CATO (non-dyslexic\nunmasked)",
@@
                 "c2l1" = "C2L1",
-                .ordered = TRUE)) %>% 
+                .ordered = TRUE))) %>% 
   filter(!(str_detect(data, "SPL2") & lang == "ES"),
@@
   unnest(cols = beta:theta) %>% 
-  mutate(across(c(beta, beta2), exp),
+  mutate(across(c(beta, beta2), ~exp(.)),
          delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
+         across(location, ~str_replace(., " ", "\n"))) %>% 
   pivot_longer(beta:theta) %>% 
@@
 
-The model estimates for the mixture-model with the highest predictive performance are shown in Figure \@ref(fig:crossstudypost).
-In this visualisation we ignore dataset specific conditions that are presented in detail below.
+As detailed above the mixture model captures the writing process in three conceptually important parameter values: (1) the transition duration associated with fluent writing (which we indicated as $\beta$), (2) the size of the slowdown for hesitant transitions (indicated as $\delta$), and (3) the mixing proportion which represents the probability of hesitant transition durations (indicated as $\theta$).
 
-```{r crossstudypost, fig.cap="Across studies. Posterior parameter distribution"}
+We present the mixture-model estimates for each of the three parameter values in Figure \ref{fig:crossstudypost}.
+Although models were fitted with all dataset-specific condition, we aggregated across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
+We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence for differences between these conditions], and removed conditions that might conflate comparisons^[We removed masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO dataset, and L2 writing in the SPL2 data.
+Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that evidence for differences in parameter values for the involved manipulations.].
+For posteriors of all conditions within dataset see Appendix \ref{posterior-parameter-estimates}.
+The posterior of the remaining conditions allows us to compare differences between transition locations within dataset associated with each of the three mixture-model parameters.
+
+```{r crossstudypost, fig.cap="Estimates across studies. Posterior parameter distribution represented as estimate and 95\\% probability interval (PI)."}
+
 posd <- position_dodge(.65)
@@
 shapes <- c(1, 4, 6, 7, 8, 9, 11)
+width <- 0
+linewidth <- .75
 beta <- filter(ps, name == "beta") %>% 
@@
              group = interaction(data))) +
-  geom_line(size = .5, 
+  geom_line(linewidth = linewidth, 
             alpha = .35, 
             position = posd) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_errorbar(width = width, alpha = .75, position = posd) +
   geom_point(aes(colour = data), 
@@
   scale_y_log10(labels = scales::comma) +
-  labs(y = "Posterior estimate with 95% PIs",
-       x = "",
-       colour = grouplabel,
+  theme(axis.title = element_blank()) +
+  labs(colour = grouplabel,
        shape = grouplabel) +
@@
 delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label) %>% 
+  mutate(name = delta_label,
+         across(c(mean, lower, upper), ~./1000)) %>% 
   ggplot(aes(y = mean, 
@@
              group = interaction(data))) +
-  geom_line(size = .5, alpha = .35, position = posd) +
-  geom_errorbar(width = .2, alpha = .75, position = posd) +
+  geom_line(linewidth = linewidth, alpha = .35, position = posd) +
+  geom_errorbar(width = width, alpha = .75, position = posd) +
   geom_point(aes(colour = data), 
@@
   scale_y_log10(labels = scales::comma) +
-  labs(y = "",
-       x = "Transition location",
-       colour = grouplabel,
+  labs(colour = grouplabel,
        shape = grouplabel) +
-  theme(axis.title.y = element_blank()) +
+  theme(axis.title = element_blank()) +
   guides(colour=guide_legend(nrow=2,byrow=TRUE),
@@
   scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(y = "",
-       x = "",
-       colour = grouplabel,
+  labs(colour = grouplabel,
        shape = grouplabel) +
-  theme(axis.title.y = element_blank()) +
+  theme(axis.title = element_blank()) +
   guides(colour=guide_legend(nrow=2,byrow=TRUE),
@@
 
-beta + delta + theta +
+plot <- beta + delta + theta +
   plot_layout(guides = "collect")
+
+grid.arrange(patchworkGrob(plot), 
+             left = "Posterior estimate with 95% PIs",
+             bottom = "Transition location")
 
@@
 
-It's generally agreed that pausing is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges, i.e.
-before sentence > before word > within word.
-We have evaluated the differences between transition locations for all data sets.
-The results are shown in Table \@ref(tab:loceffect).
+Figure \ref{fig:crossstudypost} shows largely the same patterns for keystroke interval estimates by transition location across datasets for all three mixture model parameters.
+keystroke transitions were substantially longer for the C2L1 data presumably reflecting that the sample of in this dataset was the youngest.
+Pause durations and frequencies were relatively similar across datasets.
 
-Results are largely consistent across data sets (with caveats) but differ, to some extent, from what the literature would predict.
-In line with the literature hesitations are more frequent before words than within words.
+In the following we will evaluate differences between transition locations for all three mixture model parameters.
+This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence $>$ before-word $>$ within-word.
+We evaluated the differences between transition locations for all data sets.
+The results are shown in Table \ref{tab:loceffect}.
+
+Comparing transition locations rendered patterns that are largely consistent across data sets (with caveats).
+We found that hesitations appear more frequently at before-word transitions than within words across datasets.
 Also hesitations are longer at before-sentence transitions compared to before-word transitions (except dataset C2L1) compared to within-word transitions (except dataset LIFT).
-However, our results do not support that writers pause more frequently at before-sentence locations compared to before-word locations (except for dataset SPL2; this also shows that more pauses at before-sentence locations can not be explained on the basis of multi-key combinations for sentence-initial capitalisation).
-Also, we observe that even fluent key-transitions are slower at before-word locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for dataset SPL2).
+However, for most dataset there is negligble evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
+In fact, the dataset LIFT showed less pausing before sentences compared to before words.
+Also, we observe that even fluent key-transitions are slower at before-word locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for datasets SPL2 and GUNNEXP2).
+Hesitation durations tend to be longer at before-sentence locations compared to before-word locations (except for datasets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT dataset).
 
-The datasets differ to the extent that sentence-initial key transitions do (PLanTra, LIFT) or do not (CATO, C2L1, SPL2) include the character following the shift key for capitalisation.
-In other words, the pause before sentences may sum across two key intervals, namely `_^[shift]^C` but only involves one keyintervals, namely `_^[shift]`.
-For the SPL2 dataset, we calculated location effects for sentence-initial transitions that do and do not involve the shift-to-key transition.
-The results were the same for the transition location effects.
-A comparison that is untangling the effects of the multi-keycombination on the mixture-model estimates can be found in Table \@ref(tab:shiftcellmeans).
-In short, the duration of fluent transitions and the hesitation slowdown are affected but not the hesitation probability.
+In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
 
-In conclusion, while pauses tend to be longer before sentences they are not more frequent than before words.
+The inconsistencies for before-sentence transitions could, to some extent, be explained on the basis that datasets differ as to whether before-sentence transitions should be analysed as complex key combination.
+In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+Notice though that for the datasets PLanTra and LIFT, there were no consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
 
+We test to what extent this difference may affect the modelling results for the SPL2 dataset.
+The results are shown in Appendix \ref{key-combination-effect-spl2}.
+We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
+However, this is in conflict with the absence of differences in datasets that included the character following shift at before-sentence transitions.
+In other words, it is unlikely that patterns in our results can be explained on the basis of complex key combinations at before-sentence locations.
+
+Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
+
+\blandscape
 ```{r loceffect, results='asis'}
 files <- list.files(str_c("../stanout/", 
-                          c("lift", "spl2", "spl2_shift", "gunnexp2", "plantra", "cato", "c2l1")), 
-                    pattern = "mogbetaun.+.csv", 
-                    full.names = T)
+c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetaun.+.csv", full.names = T) # "spl2_shift",
 
@@
   mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, recode_factor, 
-                "cato" = "CATO (non-dyslexic\nunmasked)",
-                "spl2" = "SPL2 (L1)", 
-                "gunnexp2" = "GUNNEXP2 (unmasked)",
-                "spl2_shift" = "SPL2 (L1; shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                .ordered = TRUE)) %>% 
+         across(data, ~recode(., 
+                "cato" = "*CATO* (non-dyslexic\nunmasked)",
+                "spl2" = "*SPL2* (L1)", 
+                "gunnexp2" = "*GUNNEXP2* (unmasked)",
+#                "spl2_shift" = "*SPL2* (L1; shift + C)", 
+                "plantra" = "*PLanTra*",
+                "lift" = "*LIFT*",
+                "c2l1" = "*C2L1*"))) %>% 
   filter(!(str_detect(data, "SPL2") & lang == "ES"),
@@
   summarise(across(starts_with("diff"), 
-                   list(mean = mean, lower = lower, upper = upper, BF = BF)),
+                   list(mean = mean, 
+                        lower = lower, 
+                        upper = upper, 
+                        BF = BF)),
             .by = c(data, param))
@@
                                    diff.2 = "before vs within word"))) %>% 
-  select(data, diffid, ends_with("beta"), ends_with("delta"), ends_with("theta"))
+  select(data, diffid, ends_with("beta"), ends_with("delta"), ends_with("theta")) %>% 
+  arrange(data)
 
@@
 
-tmp %>% knitr::kable(align =c("l", "l", rep("r", 6)), 
-                     col.names = c("Data set", "Difference", rep(c("Est. with 95% PIs", "BF"), 3)),
+apa_table(tmp[,-1],
+          align =c("l", rep("r", 6)), 
+          longtable = T,
+          escape = FALSE,
+          row.names = T,
+          font_size = "footnotesize",
+          col.names = c("Comparison", rep(c("Est. [95\\% PIs]", "BF"), 3)),
+          stub_indents = list(`\\textbf{C2L1}` = 1:2, 
+                              `\\textbf{CATO (non-dyslexic, unmasked)}`= 3:4,
+                              `\\textbf{GUNNEXP2 (unmasked)}` = 5:6,
+                              `\\textbf{LIFT}` = 7:8,
+                              `\\textbf{PLanTra}` = 9:10,
+                              `\\textbf{SPL2 (L1)}` = 11:12),
+#                              `\\textbf{SPL2 (L1; shift + C)}` = 13:14),
+          col_spanners = list(" " = c(1),
+                     "Fluent transitions" = c(2, 3), 
+                     "Hesitation slowdown" = c(4, 5), 
+                     "Hesitation probability" = c(6, 7)),
+          note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
          caption = "Effect of transition location on keystroke intervals.
-Differences between transition locations are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.") %>%
-  kableExtra::collapse_rows(columns = 1) %>%
-  kable_styling(position = "center", full_width = T, font_size = 11) %>%
-  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", threeparttable = T) %>% 
-  add_header_above(c(" " = 2, 
-                     "Fluent transitions" = 2, 
-                     "Slowdown for hesitations" = 2, 
-                     "Probability of hesitations" = 2)) %>% 
-  column_spec(1, width = "3cm")
+Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\\% PIs in brackets.")
 
 ```
+\elandscape
 
@@
 
-A general concern with mixture models is that in principle, as the mixture model has more parameters it might simply always lead to a better fit, even though cross-validation is addressing potential problems with overfitting models.
+A possible concern with the results -- substantially better predictive performance for mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
 
-To address this concern we simulated two sets of data.
-Both data sets have two conditions and 1,000 observations each.
+To address this concern we repeated a comparison for an unimodal model and a mixture model for two sets of simulated data. For each dataset we simulated 1,000 observations.
+
+These two datasets differ with regards to their underlying data generating process.
 The first set of data has as underlying data generating process a mixture model with two mixture components similar to the process described above.
-The difference between the two conditions is that the mixing proportion is larger for condition 2 than for condition 1, hence long observations are more likely in condition 2.
-This model can be summarised as followed:
+This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
 
-$$
-\begin{align}(\#eq:simmog)
-\text{y}_i \sim\text{ } & \theta_\text{condition[i]} \cdot \text{logN}(\mu_1, \sigma^2_1) +\\
-& (1 - \theta_\text{condition[i]}) \cdot \text{logN}(\mu_2, \sigma^2_2)
-\end{align}
-$$
+\begin{equation}
+\begin{aligned}
+(\#eq:simmog)
+\text{y}_i \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
+& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
+\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
+		& \sigma_{1}^2 > \sigma_{2}^2
+\end{aligned}
+\end{equation}
 
-The second data set was generated with an unequal variance unimodal model as data generating process.
-Condition 2 has a larger mean and standard deviation than condition 1.
-The model can be summaried as followed:
+This model is largely identical with the models before but reduced to its main parameters.
+There are two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.
 
-$$
-\begin{align}(\#eq:simuv)
-\text{y}_i \sim\text{ }& \text{logN}(\mu_\text{condition[i]}, \sigma^2_\text{condition[i]})
-\end{align}
-$$
+The second dataset was generated with an unimodal model as data generating process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
 
-The true parameter values used for each of the two data simulations can be found in Table \@ref(tab:simparam).
-The simulated data are visualised in Figure \@ref(fig:simdata).
+\begin{equation}
+\begin{aligned}
+(\#eq:simuv)
+\text{y}_i \sim\text{ }& \text{logN}(\beta, \sigma^2)
+\end{aligned}
+\end{equation}
+
+Again, this model is a simplified version of the unimodal models used in the main analysis.
+The model assumes a log-normal distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
+
+The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
+The simulated data are visualised in Figure \ref{fig:simdata}.
 The data are simulated to be similarly distributed to keystroke transitions.
 
-```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=4}
+```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=3.5, fig.width=5}
 files <- list.files("../mixture-model-sim/data", full.names = T)
@@
         mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
-  mutate(across(condition, factor),
-         across(condition, ~str_c("Condition: ", condition)),
-         across(Dataset, ~recode(., mogdata = "bimodal", uvlmdata = "unimodal"))) %>% 
-  ggplot(aes(x = value, colour = condition, fill = condition)) +
+  mutate(across(Dataset, ~recode(., mogdata = "bimodal", lmdata = "unimodal"))) %>% 
+  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
   geom_density(alpha = .25) +
-  facet_grid(~Dataset, labeller = label_both) +
   scale_colour_colorblind() + 
   scale_fill_colorblind() +
-  labs(colour = "", fill = "", x = "Simulated data", y = "Density")
-
+  scale_x_continuous(labels = ~format(., big.mark = ",")) +
+  labs(colour = "Data process", 
+       fill = "Data process", 
+       x = "Simulated data", 
+       y = "Density")
 ```
 
-We run 4 models: 2 mixture models, one on the data generated with a mixture process and one on the data generated with the unimodal unequal variance process.
-We repeated the same using an unimodal unequal variance model.
+We run 4 models, 2 on each of the two datasets: 2 mixture models, one for the data generated with a bimodal mixture process and one for the data generated with the unimodal process.
+We repeated the same using an unimodal model.
 Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
-The Stan models uncovered the model parameters of their respective data sets successfully, as shown in Table \@ref(tab:simparam), not less so when the model was applied to data generated with the other underlying process.
+Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}.
+The parameters are shown by data generating process along with with true parameter values.
+Parameter estimates are shown for the dataset generated with a unimodal process and the dataest generated with a bimodal process.
+The Bayesian models uncovered the model parameters of the dataset with the matching data generating process successfully, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
 
@@
   mutate(data = "mog")
-mog2 <- read_csv("../mixture-model-sim/stanout/mog_uvlmdata.csv") %>% 
-  mutate(data = "uvlm")
-uvlm1 <- read_csv("../mixture-model-sim/stanout/uvlm_uvlmdata.csv") %>% 
-  mutate(data = "uvlm")
-uvlm2 <- read_csv("../mixture-model-sim/stanout/uvlm_mogdata.csv") %>% 
+mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
+  mutate(data = "lm")
+lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
+  mutate(data = "lm")
+lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
   mutate(data = "mog")
@@
             .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), round, 2),
+  mutate(across(where(is.numeric), ~round(., 2)),
          across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
@@
          across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2")),
-         across(name, ~str_replace(., "\\[1\\]", "_\\\\text{condition=1}")),
-         across(name, ~str_replace(., "\\[2\\]", "_\\\\text{condition=2}"))) %>%
+         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>%
   select(-lo, -up) %>% 
   pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = c(5, 1, .1, .4, .25, .5),
+  mutate(true = c(5, 1, .35, .25, .5),
          across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
          across(name, ~str_replace(., "prob", "theta"))) %>% 
-  relocate(name, true)
+  relocate(name, true) %>% 
+  unite("name", name:true, sep = " = ")
 
-uvlm_sim <- bind_rows(uvlm1, uvlm2) %>% 
+lm_sim <- bind_rows(lm1, lm2) %>% 
   summarise(across(value, list(est = mean,
@@
             .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), round, 2),
+  mutate(across(where(is.numeric), ~round(., 2)),
          across(est, ~str_c(., " [", lo, ", ", up, "]")),
@@
          across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2")),
-         across(name, ~str_replace(., "\\[1\\]", "_\\\\text{condition=1}")),
-         across(name, ~str_replace(., "\\[2\\]", "_\\\\text{condition=2}"))) %>% 
+         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>% 
   select(-lo, -up) %>%
   pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = as.character(c(5, 6, .25, .5))) %>% 
-  relocate(name, true)
+  mutate(true = as.character(c(5, .25))) %>% 
+  relocate(name, true) %>% 
+  unite("name", name:true, sep = " = ")
 
@@
 ```{r simparam, results='asis'}
-table <- bind_rows(
-  mog_sim %>% mutate(model = "mog"),
-  uvlm_sim %>% mutate(model = "uvlm")) %>% 
+table <- bind_rows(mog_sim %>% mutate(model = "mog"),
+                   lm_sim %>% mutate(model = "lm")) %>% 
   relocate(model) %>% 
@@
 
-names(table) <- c("Model", "Parameter", "True value", "Bimodal data", "Unimodal data")
+names(table) <- c("Model", "Parameter with true value", "Bimodal data", "Unimodal data")
 
-kable(table[,-1], 
-      caption = "Uncovered parameter estimates with 95% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
-      align =c("l", rep("r", 3))) %>%
-  kable_styling(position = "center", 
-                full_width = F, 
-                font_size = 11) %>% 
-  add_header_above(c(" " = 1, 
-                     " " = 1,
-                     "Estimate with 95\\% PI" = 2)) %>%
-  pack_rows("Model: Bimodal mixture model", 1, 6) %>% 
-  pack_rows("Model: Unimodal process", 7, 10)
+apa_table(table[,-1], 
+          escape = FALSE,
+      col_spanners = list("Estimate with 95\\% PI" = 2:3),
+      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
+      align = c("l", rep("r", 2)),
+      stub_indents = list(`Bimodal mixture process` = 1:5, 
+                          `Unimodal process` = 6:7))
 
@@
 We used LOO-CV to compare the fit of the two models for each data set.
-The model comparisons can be found for each data generating process in Table \ref@(tab:loossim).
-The results show that the mixture model does not always lead to higher predictive performance.
-Indeed, the mixture model showed a lower predictive performance for the data that were generated with a unimodal process.
-However, for the data generated with a bimodal process, the mixture model model shows a higher predictive performance.
-In fact, the ratio of $\Delta\widehat{elpd}$ and its SE, as metric for the strength of evidence, shows that the mixture model performs 11 times better than the unimodal model for the data generated with a bimodal process.
-In comparison, for the unimodal data, the unimodal model shows only 3 times better than the bimodal mixture model.
-Thus, even though the mixture model does not necessarily perform better for non-bimodal data but it also doesn't necessarily perform much worse.
-This contrast is likely a reflection of the increased number of parameters in the mixture model.
+The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
+The results rules out the possibility that the mixture model does always lead to higher predictive performance.
+Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a unimodal process.
+However, for the data generated with a bimodal process, the mixture model shows a substantially higher predictive performance.
+
+In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence [@sivula2020uncertainty], shows that the mixture model performs 11.6 times better than the unimodal model for the data generated with a bimodal process.
+In comparison, for the unimodal data, the unimodal model performs only 0.77 times better than the bimodal mixture model.
+Thus, even though the mixture model does not necessarily perform better for non-bimodal data, it also does not necessarily perform worse.
 
@@
 #    elpd_se_ratio = abs(elpd_diff / se_diff),
-    across(where(is.numeric), round, 0),
-    across(where(is.numeric), format, big.mark = ","),
-    across(everything(), as.character),
-    across(everything(), str_trim),
-    across(everything(), str_replace_all, "NaN", "--")) %>% 
-  mutate(across(elpd_diff, str_c, " (", se_diff, ")"),
-         across(elpd_loo, str_c, " (", se_elpd_loo, ")"),
-         across(elpd_diff, str_replace_all, "0 \\( 0\\)", "--"),
+    across(ends_with("diff"), ~round(., 1)),
+    across(ends_with("loo"), ~round(., 0)),
+    across(where(is.numeric), ~format(., big.mark = ",")),
+    across(everything(), ~as.character(.)),
+    across(everything(), ~str_trim(.)),
+    across(everything(), ~str_replace_all(., "NaN", "--")),
+    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
+    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
+    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
     across(c(data, model), ~recode(., mog = "Bimodal mixture model",
-                                        uvlm = "Unimodal unequal-variance model"))) %>% 
+                                      lm = "Unimodal model"))) %>% 
   select(data, model, starts_with("elpd"))
@@
 
-kable(table[,-1], 
-      caption = "Model comparisons by data set. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-        align =c("l", rep("r", 2))) %>%
-  kable_styling(position = "center", 
-                full_width = F, 
-                font_size = 11) %>%
-  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
-  pack_rows("Data: Bimodal mixture process", 1, 2) %>% 
-  pack_rows("Data: Unimodal process", 3, 4)
+apa_table(table[,-1], 
+          escape = FALSE,
+      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
+      align =c("l", rep("r", 2)),
+      stub_indents = list(`Data: Bimodal mixture process` = 1:2, 
+                          `Data: Unimodal process` = 3:4))
 
@@
 
-# Appendix
-
-## Fit to data
-
-```{r prediction, fig.cap="Fit to data.
-Comparison of 100 simulated (predicted) sets of data to observed data illustated by model.
-For illustration the x-axis was truncated at 2,000 msecs.", eval = T, fig.height=10}
-
-run <- FALSE
-
-if(run){
-  file <- list.files("../stanout", pattern = "sims", full.names = T, recursive = T)
-  ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
-    str_to_upper() %>% 
-    str_replace(., "PLANTRA", "PLanTra")
-  ds <- ds[-6]
-  plots <- list()
-  plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
-  #plots <- map(1, ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
-  plots_fin <- plots[[1]] / plots[[2]] / plots[[3]] / plots[[4]] / plots[[5]] / plots[[6]] +
-    plot_layout(guides = "collect") &
-  #  plot_annotation(tag_levels = 'A') & 
-    theme(legend.position = 'bottom')
-  saveRDS(plots_fin, "objects/fitplots.rda", compress = "xz")
-}
-
-plots <- readRDS("objects/fitplots.rda")
+```{r render_appendix, include=FALSE}
+render_appendix("appendix_a.Rmd")
+render_appendix("appendix_b.Rmd")
+render_appendix("appendix_c.Rmd")
+render_appendix("appendix_d.Rmd")
+render_appendix("appendix_e.Rmd")
+render_appendix("appendix_g.Rmd")
+render_appendix("appendix_f.Rmd")
 ```
-
-## Posterior parameter estimates

### 2023-07-24 report/results.Rmd
Manuscript with appendices separately

@@
 library(rmdformats)
+library(rmarkdown)
 library(gridExtra)
 library(grid)
+library(knitr)
 library(gtable)
@@
 run <- FALSE
-
 if(run){
@@
     theme(legend.position = 'bottom')
-  saveRDS(plots_fin, "objects/fitplots.rda", compress = "xz")
+  ggsave("objects/fitplots.pdf", plots_fin, height = 10, width = 6)
+#  saveRDS(plots_fin, "objects/fitplots.rda", compress = "xz")
 }
 
-readRDS("objects/fitplots.rda")
-```
+include_graphics("objects/fitplots.pdf")
+```
+
+\begin{figure}
+\includegraphics{objects/fitplots.pdf}
+\end{figure}
 

### 2023-09-15 report/appendix_h.Rmd
Add simulation

## Simulation

A possible concern with these results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit. We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models. To address this concern we repeated a comparison for a uni-modal model and a bi-modal mixture model for two sets of simulated data that were simulated either with a uni-modal or bi-modal process at heart. In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.

In particular, the first data set was simulated using a bi-modal mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.

\begin{equation}
\begin{aligned}
(\#eq:simmog)
\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
		& \sigma_{1}^2 > \sigma_{2}^2
\end{aligned}
\end{equation}

This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants). The model includes two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.

The second data set was generated with a uni-modal log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.

\begin{equation}
\begin{aligned}
(\#eq:simuv)
\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
\text{constraint: } & \sigma^2>0
\end{aligned}
\end{equation}

Again, this model is a simplified version of the uni-modal models used in the main analysis. The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.

The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}. The simulated data are visualised in Figure \ref{fig:simdata}. Parameter values were chosen so that the simulated data are roughly similarly distributed to keystroke transitions.

```{r simdata, fig.cap= "Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
files <- list.files("../mixture-model-sim/data", full.names = T)

map_dfr(files, ~read_csv(.) %>% 
        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
  mutate(across(Dataset, ~recode(., mogdata = "bi-modal", lmdata = "uni-modal"))) %>% 
  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
  geom_density(alpha = .25) +
  scale_colour_colorblind() + 
  scale_fill_colorblind() +
  scale_x_log10(labels = ~format(., big.mark = ",")) +
  theme(text = element_text(size = 10)) +
  labs(colour = "Data process", 
       fill = "Data process", 
       x = "y", 
       y = "Density")
```

For each of these two data sets we simulated 1,000 observations. We run 2 models -- a bi-modal mixture model and a uni-modal model -- each for both data sets. Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup. Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}. The parameters are shown by type of data generating process along with the true parameter values. Parameter value estimates are shown by type of Bayesian model. The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process. Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they uni-modal model.

```{r}
mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
  mutate(data = "mog")
mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
  mutate(data = "lm")
lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
  mutate(data = "lm")
lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
  mutate(data = "mog")

mog_sim <- bind_rows(mog1, mog2) %>% 
  summarise(across(value, list(est = mean,
                               lo = lower,
                               up = upper),
                   .names = "{.fn}"),
            .by = c(name, data)) %>% 
  mutate(across(where(is.numeric), ~round(., 2)),
         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
         across(est, ~str_c(., " [", lo, ", ", up, "]")),
         across(name, ~str_c("\\", .)),
         across(name, ~str_replace(., "a_e", "a^2_1")),
         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>%
  select(-lo, -up) %>% 
  pivot_wider(names_from = data, values_from = est) %>% 
  mutate(true = c(5, 1, .35, .25, .5),
         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
         across(name, ~str_replace(., "prob", "theta"))) %>% 
  relocate(name, true) %>% 
  unite("name", name:true, sep = " = ")

lm_sim <- bind_rows(lm1, lm2) %>% 
  summarise(across(value, list(est = mean,
                               lo = lower,
                               up = upper),
                   .names = "{.fn}"),
            .by = c(name, data)) %>% 
  mutate(across(where(is.numeric), ~round(., 2)),
         across(est, ~str_c(., " [", lo, ", ", up, "]")),
         across(name, ~str_c("\\", .)),
         across(name, ~str_replace(., "a_e", "a^2_1")),
         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>% 
  select(-lo, -up) %>%
  pivot_wider(names_from = data, values_from = est) %>% 
  mutate(true = as.character(c(5, .25))) %>% 
  relocate(name, true) %>% 
  unite("name", name:true, sep = " = ")

```

```{r simparam, results='asis'}
table <- bind_rows(mog_sim %>% mutate(model = "mog"),
                   lm_sim %>% mutate(model = "lm")) %>% 
  relocate(model) %>% 
  mutate(across(name, ~str_c("$", ., "$")))

names(table) <- c("Model", "Parameter with true value", "Bi-modal model", "Uni-modal model")

apa_table(table[,-1], 
          escape = FALSE,
      col_spanners = list("Estimate with 95\\% PI" = 2:3),
      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
      align = c("l", rep("r", 2)),
      stub_indents = list(`Bi-modal data` = 1:5, 
                          `Uni-modal data` = 6:7))

```

We used LOO-CV to compare the predictive performance of the two models for each data generating process. The model comparisons can be found for each data generating process in Table \ref{tab:loossim}. The results rule out the possibility that the mixture model does always lead to higher predictive performance. Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a uni-modal process. However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance. In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence [@sivula2020uncertainty], shows that the mixture model performs 11.6 times better than the uni-modal model. In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model. Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the uni-modal is preferred by the law of parsimony.

```{r loossim, results='asis'}
table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
  select(1:6) %>% 
  mutate(
#    elpd_se_ratio = abs(elpd_diff / se_diff),
    across(ends_with("diff"), ~round(., 1)),
    across(ends_with("loo"), ~round(., 0)),
    across(where(is.numeric), ~format(., big.mark = ",")),
    across(everything(), ~as.character(.)),
    across(everything(), ~str_trim(.)),
    across(everything(), ~str_replace_all(., "NaN", "--")),
    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
    across(c(data, model), ~recode(., mog = "Bi-modal mixture model",
                                      lm = "Uni-modal model"))) %>% 
  select(data, model, starts_with("elpd"))

names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")

apa_table(table[,-1], 
          escape = FALSE,
      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
      align =c("l", rep("r", 2)),
      stub_indents = list(`Data: Bi-modal mixture process` = 1:2, 
                          `Data: Uni-modal process` = 3:4))

```

### 2023-09-15 report/manuscript.Rmd
Add simulation

@@
 ---
-title             : "Modelling writing hesitations in text writing as finite mixture process"
+title             : "Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process"
 shorttitle        : "Modelling writing hesitations"
@@
 
-abstract: |
+abstract: Classical serial models view the writing process as a chain of pauses and writing burst.
+In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
+We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
+We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
+Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
+Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
+Our results cannot be explain by the serial but are in line with the cascading view of writing.
 
@@
         - "appendix_f.tex"
+        - "appendix_g.tex"
 
@@
 
-beta_label <- "Short key transition\nduration (in msecs)"
-delta_label <- "Slowdown for hesitant key\ntransitions (in secs)"
+beta_label <- "Short transition\nduration (in msecs)"
+delta_label <- "Slowdown for hesitant\ntransitions (in secs)"
 theta_label <- "Probability of hesitant\ntransitions"
@@
 
-# Introduction
+Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing [@bock2014syntactically; @wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].
+
+Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
+Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
+Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking [@roeser2018advance; @torrance2012written].
+This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader [e.g.
+@Clark2002].
+Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
+Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].
+
+The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
+On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06].
+For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses].
+The same is generally observed for spoken utterances [@lee13; @meyer1996; @wheeldon2013] with some exceptions [@gri01].
+We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
+Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
+
+<!-- a statistical framework for the analysis of keystroke data that captures not just the multilevel-structure of the data but also presents a principled way to capture the probabilistic nature of pauses in text composition.
+Hesitations are probabilistic in the sense that we cannot be predicted their occurrence deterministically.
+In principle we can produce several words and even sentences without pausing.
+However, hesitations are more of less likely to occur in certain contexts, for example, at sentence boundaries.
+-->
+
+Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements [@van1991handwriting; @olive2014toward].
+The intervals between adjacent keystrokes during typing vary as a function of various factors.
+Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
+Processing may be serial.
+This means that the processing cascade operates on one unit at a time.
+Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
+Therefore longer IKIs represent periods when the writer is planning what to write next.
+The subsequent burst of fluent output is then the result this planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing].
+This is particularly reflected in the model proposed by @flower1980ynamics that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts [also @chenoweth2001fluency; @chenoweth2003inner].
+
+This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second].
+However this account is not consistent with what we know about writing.
+Consider these two observations:
+
+First, utterances are not fully formed at production onset.
+Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
+This is well known for spoken production [e.g.
+@bock2014syntactically].
+For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases [@not10; dam09; @roeser2018advance].
+For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e.
+*A and the B moved*) or a simple noun (i.e., *A moved above the B and the C*).
+Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
+This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
+In fact, eye movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e.
+*the B*) was more likely for conjoined noun phrases.
+
+Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
+For example, @medimorec2017pauses found that, in undergraduate students writing on familiar topics 71% of sentence-initial IKIs were less than 1 sec in duration.
+Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
+For comparison, this is less than mean written naming response latency for single objects [@torrance2018timed] and short sentences when describing arrays of images [@roeser2018advance] in a similar population.
+Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
+The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
+
+If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
+The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
+Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction].
+@van1991handwriting, for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now].
+Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
+In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
+
+This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
+IKIs result from one of two data-generating processes.
+When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.
+just by processing below the last buffer in the processing cascade).
+However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
+IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling].
+
+For example, @roeser2021modelling demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
+In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
+Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
+As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
+Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
+For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.
+
+The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
+This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
+
+Hesitations in writing have potential disadvantages.
+Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
+However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck.
+Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
+If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
+@ronneberg2022process coined this possibility the "process-disruption hypothesis".
+
+In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts.
+On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].
+
+Data form spontaneous text production [see @gernsbacher1995coherence] are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
+Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy [@dux2022automating;@dux2021effect;@vandermeulen2023writing].
+
+Second, since the emergence of software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing [@conijn2021timecourse].
+Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
+These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning [@conijn2019understanding; @conijn2020keys].
+
+Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments [@not10; @dam09; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual].
+In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
+Instead text composition provides data from both inter and intra-sentence transitions.
+Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
+
+The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
+In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
+This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
+The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
+In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
+Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
+We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
 
@@
 
+We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
+The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
+In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component [@wagenmakers2007linear; @wing1973response; @schoner2002timing].
+
+The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in @roeser2021modelling.
+Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
+Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
+This view was implemented as a constrained and an unconstrained model.
+
 ## Statistical models
 
-The models presented in the following can be divided into two general groups.
-The first three models are largely akin to models typically used in the literature.
-By this we mean models that assume a uni-modal process that generates keystroke data as is incorporated in statistical models such as analysis of variance and linear mixed-effects models.
-Second, the last two models model keystroke intervals as a combination of two weighted processes of which one presents a smooth information flow from mind into finger; the other component is more important as it represents moments at which the information flow was interrupted leading to longer latencies.
-The latter two models map directly on the idea of a cascading model of writing.
-
-### Unimodal Gaussian
-
-We start with a Gaussian mixed-effects models similar to a standard analysis of variance.
-We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
-The mean can be decomposed into $\beta$ and $\text{participant}_i$.
-$\beta$ will be allowed to take on a different value for each transition location.
-Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.
+We are using the Bayesian framework [e.g.
+@farrell2018computational; @gelman2014; @lee2014bayesian] to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
+In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
+The models presented in this section build on one another so that later model include assumptions of earlier models.
+
+### Serial model of writing
+
+#### Unimodal Gaussian
+
+Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
+The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
+For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
+There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
+We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average IKI associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the variability associated with factors that we did not further specify in the model.
+This can be expressed as $\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)$.
+Of importance is the estimated posterior value of $\mu$ as this value captures that time it takes to mentally plan a word.
+
+We can extend this simple model of word-planning to other linguistic location.
+We introduce earlier that larger linguistic edges are associated with planning on higher levels.
+For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07].
+Regression formulas can, and commonly are, used to decompose $\mu$ and capture that there is a change in the outcome variable associated with another factors.
+We can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is the average IKI for word boundaries.
+However when $\text{x}_\text{sentence}$ takes on the value 1, the average IKI for word boundaries $\alpha$ is incremented by a sentence-related slowdown of $\beta$ msecs.
+Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning.
+The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.
+whether there is a statistically meaningful difference between IKI associated with words and sentences).
+
+For computational ease, we implemented the differences associated with transition locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
+Therefore, the model will return one $\beta$ per transition location that capture the posterior distribution of average IKIs.
+Also the decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts $u_\text{participant}$.
+The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$.
+This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.
+a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
+
+The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
 
@@
 (\#eq:unimodgaus)
-\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
-& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
+\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
@@
 
-### Unimodal log-Gaussian
-
-This model is largely identical to the previous model, except we use a log-normal distribution instead of a normal distribution.
-The advantage of using a log-normal distribution is two-fold: (1) the log-normal distribution has a lower bound of zero.
-For our data we generally consider negative ikis as mistakes which occasionally occur when the next key was pressed before the current key.
-Other than that keystroke intervals are constrained by a persons ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour, in particular motor responses.
-In particular, a normal distribution assumes that units are linearly scaled.
-For example, a 50 msecs difference is the same between 100 msecs and 150 msecs than it is between 5 secs and 5 secs 50 msecs (i.e.
-5,050 msecs).
-This is not necessarily plausible for keystroke data.
-We would assume that differences that are due to motor activity (typing a high frequency bigram vs a low frequency bigram) are smaller than difference that are due to high levels of cognitive activity (retrieving a word in your L1 or L2).
-Log-normal distributions are a natural way of scaling units so that a 50 msecs difference on the lower end of the iki scale (motor activity) is more meaningful than a 50 msecs difference on the upper end of the iki scale (retrieving words, planning sentences).
-
-The model can be described like this:
+Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
+
+#### Unimodal log-Gaussian
+
+The previous model assumes that the data-generating process is a Gaussian distribution.
+The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
+There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
+IKIs, as the distance between two subsequent key-down events, must be positive.
+The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
+(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
+In particular, a Gaussian distribution assumes that units are scaled linearly.
+For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
+This does not necessarily map onto the psychological interpretation for short and long keyintervals.
+For example effects that result on the motor level within words (e.g.
+typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
+In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
+Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
+
+The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{N}()$ was replaced by $\text{logN}()$.
 
@@
 (\#eq:unimodloggaus)
-\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
+\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
 \text{where: } &
-\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
-& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
@@
 
-### Unimodal unequal-variance log-Gaussian
+#### Unimodal unequal-variance log-Gaussian
+
+The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
+The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
+The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
+This assumption is however not in line with what we know about data from human behaviour.
+Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing].
+For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
+Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation $\sigma_{e_\text{location}}^2)$ varies as a function of transition location.
 
@@
 (\#eq:unimoduv)
-\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
- & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+ & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
  \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
@@
 
-### Bimodal log-Gaussian (constrained)
-
-This models extends the intuition from the previous model that higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that underlies the generation of ikis, we assume there are two.
-(1) activation can flow into keystrokes without interrupts.
-These fluent keystroke transitions are merely limited by a person's ability to move their finger and will be captured but the parameter $\beta$.
-In principle, there are no differences for fluent key-transitions between transition location.
-The next model with loosen this assumption (hence, "unconstrained model").
-(2) difficulty in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when spelling, words, or contents couldn't be retrieved in time.
-The size of these pauses will depend on the reason for delays which is typically associated with transition locations (contents are typically planned before sentences, words are retrieved before they are typed and spelling difficulty typically occurs when typing a word).
-Pauses will be captured by two model parameters: (1) the slowdown for these hesitant transitions will be captured by $\delta$ which is the deviation compared to normal typing intervals (constrained to be positive).
-(2) the frequency of hesitant transitions will be captured by $\theta$ for each level of a categorical predictor.
+### Parallel cascading model of writing
+
+The following two models are are extensions of the models introduced for the serial view.
+Crucially, the cascading view allows planning to happen in parallel to production.
+Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
+This is done by assuming that IKIs come from a weighted combination of two distributions.
+
+#### Bimodal log-Gaussian (constrained)
+
+This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
+Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
+This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+
+1.
+Activation can flow into keystrokes without interruption.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter in equation \ref{eq:bimodcon}.
+Note that $\beta$ in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
+
+2.
+Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
+The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}.
+The magnitude of the slowdown is associated with transition location.
+This is because delays at larger linguistic units are likely to be associated with higher level planning.
+By constraining $\delta$ to be positive, it captures how much longer hesitant IKIs in addition to $\beta$.
 
@@
 (\#eq:bimodcon)
-\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
+		& \sigma_{e'}^2 > \sigma_{e}^2\\
+		& 0 > \theta > 1
+\end{aligned}
+\end{equation}
+
+The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
+Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
+$\theta$ is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
+This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$.
+In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
+The weights of both distributions must sum to 1.
+We will call this parameter the probability of hesitant transitions.
+
+The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
+In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
+As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
+This is akin to what is known as a random by-participant slopes model.
+
+Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location.
+In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].
+This was achieved by introducing a parameter $\sigma_\text{diff}$.
+The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
+
+#### Bimodal log-Gaussian (unconstrained)
+
+In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
+In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions.
+This is what we called a constrained model.
+However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
+This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
+Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
+We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
+
+In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
+
+\begin{equation}
+\begin{aligned}
+(\#eq:bimoduncon)
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+	\text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
@@
 
-```{r}
-#		<!-- &\delta \sim \text{N}(0,1)\\ -->
-
-```
-
-This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
-
-### Bimodal log-Gaussian (unconstrained)
-
-This model is identical to the previous model with one exception.
-The distribution of fluent keystroke transitions captured by $\beta$ was fixed to be the same across transition locations is the previous model.
-In other words, the mean $\beta$ and it's standard deviation $\sigma_{e}^2$ was the same for before-sentence, before-word, and within-word transitions.
-This means, because of the naturally larger number of within-word transitions, the posterior is dominated by within-word transitions.
-
-In this model we will loosen this constraint and allow $\beta$ and $\sigma_{e}^2$ to vary by transition-location.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:bimoduncon)
-\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-	\text{where: }  & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-```{r}
-#<!-- & \delta \sim \text{N}(0,1)\\ -->
-
-```
-
 ## Data sets
 
-Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \ref{tab:datasets}.
+Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
 
@@
                            "Torrance et al. (n.d.)",
-                           "Rossetti and Van Waes (2022)",
-                           "Vandermeulen et al. (2020)", 
+                           "Rossetti and Van Waes (2022b)",
+                           "Vandermeulen, Steendam, et al. (2020)", 
                            "Torrance and Ofstad (n.d)"),
@@
 
+### C2L1
+
+TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
+
+TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
+
+The C2L1 data set comprises data Norwegian 6th graders -- *N* = 126, mean age 11 years 10 months -- published in @ronneberg2022process.
+The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
+Keystroke data were captured using EyeWrite [@sim07; @torrance201203].
+
+### CATO
+
+TODO: Mark, can you add information here?
+
+Data are published in @torrance2016adolescent.
+Norwegian upper secondary students -- *N* = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
+Keystroke data were captured using EyeWrite [@sim07; @torrance201203].
+
 ### GUNNEXP2
 
-TODO: Mark, can you add information here? I've added a provisional bib entry that needs changing or we just say "unpublished"
+TODO: Mark, can you add information here?
+I've added a provisional bib entry that needs changing or we just say "unpublished".
+I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
+Was the sample Norwegian?
 
 The GUNNEXP2 dataset published in @torranceb includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote text either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### C2L1
+In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
+Keystroke data were captured using EyeWrite [@sim07; @torrance201203].
+
+### LIFT
+
+The LIFT data are published in @vandermeulen2020 and described in @vandermeulen2020mapping.
+The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
+Within this national survey, a representative sample of Dutch students (*N* = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
+The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
+In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
+The students received 50 minutes for each tasks.
+Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
+During the synthesis tasks keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
+
+### PLanTra
+
+<!-- It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch. -->
+
+The PLanTra (Plain Language Training for business content) data are published in @rossetti2022text and described in @rossetti2022s.
+The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
+A total of *N* = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
+The study adopted a pre-test post-test design.
+As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
+Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
+Participants were asked to spend at least 45 minutes on the instruction module.
+As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
+Both reports were written in English (second language) and similar in length (274-278 words) and readability.
+Participants received as much time as needed for each task.
+During the task, keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
+
+### SPL2
 
@@
 
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)?
-
-The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process.
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### CATO
-
-TODO: Mark, can you add information here?
-
-Data are published in @torrance2016adolescent.
-Norwegian upper secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### PLanTra
-
-TODO: Rianne, can you add information here?
-
-The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
-In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
-Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
-During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
-making it easier to read for a lay audience--by applying what they had learned from their respective training.
-The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
-It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
-Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### LIFT
-
-TODO: Rianne, can you add information here?
-
-The LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) data are published in @vandermeulen2020.
-Students produced text for 4 topics in two different writing genres (levels: argumentative text, informative text).
-Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### SPL2
-
-TODO: Mark, can you add information here?
-
 The SPL2 dataset is published in @torrancea.
-The data come from a textcomposition task of undergraduate university students -- *N* = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
-CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
-We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
-Participants were given a 40 minute time limit.
-They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
+The data come from a text composition task of undergraduate university students -- *N* = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
+Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
 
@@
 
-We reanalysed keystroke data from 6 datasets in Bayesian implementations of 5 statistical models. 
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values: Bayesian parameter estimates are expresses the uncertainty associated with parameter values as probability distribution [@farrell2018computational; @gelman2014; @lee2014bayesian].
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
-existing knowledge about parameter values, for regulating the parameter space.
-For small data sets priors influence the inferred parameter estimates (known as posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
-automatic Ockham's razor, @jefferys1992ockham].
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
-@lambert2018student; @mcelreath2016statistical].
-The predictive performance (i.e.
-fit) of these models is compared in the Results section using leave-one-out cross-validation.
-
-Keystroke data from text composition tasks were modelled in a series of five models. An overview of all models can be found in Table \ref{tab:models}.
+We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
 
@@
        Description = c("Gaussian distribution",
-                       "Log-normal distribution",
-                       "Log-normal distribution and unequal variances",
-                       "Fluent transition are not allowed to vary by transition location",
-                       "Fluent transition are allowed to vary by transition location"))
+                       "Log-Gaussian distribution",
+                       "Log-Gaussian distribution with unequal variances",
+                       "Fluent transition do not vary by transition location",
+                       "Fluent transition vary by transition location"))
 
@@
 
-For all models we modelled transition locations (levels: before sentence, before word, within word) was included as predictor in all models.
-Also, we included the dataset specific experimental conditions as fixed effects and included random intercepts for participants as described in the equations above.
-We used normal and log-normal probability function in line with typically treatments fouund in the literature.
-Stan code for mixture models was based on @roeser2021modelling  [see also @vasishth2017; @vasishth2017feature].
-
-Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
+Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
+This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates [@farrell2018computational; @gelman2014; @lee2014bayesian].
+To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
+existing knowledge about parameter values.
+For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
+automatic Ockham's razor, @jefferys1992ockham].
+In other words the choice of priors values has less impact on the posterior.
+In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
+@lambert2018student; @mcelreath2016statistical].
+Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
+
+For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
+Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
+Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
+
+Data were analysed in Bayesian mixed effects models [@gelman2014; @mcelreath2016statistical].
 The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
+Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
 Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
+The predictive performance of our models was compared using leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical; @sivula2020uncertainty].
 
@@
 
-The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g.
+The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g.
 @torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
-Keytransitions that terminated in an editing operation were excluded from the analysis.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions.
-
-```{r keyloc, results='asis'}
+In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
+At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
+Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
+
+```{r keyloc, results='asis', fig.pos="b"}
 
@@
             font_size = "footnotesize",
-            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace. At before-sentence locations, IKIs were timed to the shift keypress.")
+            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace.")
 
@@
 For all datasets we only used transitions that were not followed by an editing operation.
-
+Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
 We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant)
-
-We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-
+We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
+
+We further removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
 From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis.
-Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant.
+This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
+For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
+Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
 The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
@@
     mutate(across(c(keep), ~str_c(., ")"))) %>%
-  pivot_wider(names_from = location, values_from = keep)
+  pivot_wider(names_from = location, values_from = keep) %>% 
+  relocate(`within word`)
 
@@
   relocate(ds) %>% 
-  mutate(across(ds, ~recode(., 
+  mutate(across(ds, ~recode_factor(., 
+                "c2l1" = "C2L1",
                 "cato" = "CATO",
+                "gunnexp2" = "GUNNEXP2",
+                "lift" = "LIFT",
+                "plantra" = "PLanTra",
+#                "spl2_shift" = "SPL2 (shift + C)", 
                 "spl2" = "SPL2",
-                "spl2_shift" = "SPL2 (shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2"))) %>% 
+                .ordered = T
+                ))) %>% 
+  arrange(ds) %>% 
   rename(`Dataset` = ds)
@@
 
-To compare the out-of-sample predictive performance of the models we used Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
-Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and its  difference $\Delta\widehat{elpd}$ between models.
-The advantage of using leave-one-out cross-validation is that models with more parameters are penalised to prevent overfit.
-
-Results for all data sets are shown in Table \ref{tab:loos}.
+To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
+Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared via its difference $\Delta\widehat{elpd}$ between models.
+Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
+
+Model comparison results for all data sets are shown in Table \ref{tab:loos}.
 For all data sets we found the same pattern.
-The mixture of log-normal distribution models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across transition locations.
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
+Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
+The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
+Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
+The weakest model was the uni-modal Gaussian model.
 
@@
                mogbetaunconstr = "M5"),
-    across(model, ~recode(., lmm = "Unimodal log-normal",
-                          lmmgaus = "Unimodal normal",
+    across(model, ~recode(., lmm = "Unimodal log-Gaussian",
+                          lmmgaus = "Unimodal Gaussian",
                           mogbetaconstr = "Bimodal (constrained)",
@@
 apa_table(table[,-1], 
-      caption = "Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
+      caption = "Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.",
       align =c("p{3.5cm}", "p{6cm}", rep("r", 2)), 
@@
 
+We also evaluated to what extent model predictions fit observed data.
+These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
+Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
+
 ## Cross-data set comparisons
@@
          across(data, ~recode_factor(., 
-                "cato" = "CATO (non-dyslexic\nunmasked)",
-                "gunnexp2" = "GUNNEXP2 (unmasked)",
-                "spl2" = "SPL2 (L1)", 
+                "cato" = "CATO",# (non-dyslexic, unmasked)",
+                "gunnexp2" = "GUNNEXP2",# (unmasked)",
+                "spl2" = "SPL2", # (L1)", 
 #                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
@@
 
-As detailed above the mixture model captures the writing process in three conceptually important parameter values: (1) the transition duration associated with fluent writing (which we indicated as $\beta$), (2) the size of the slowdown for hesitant transitions (indicated as $\delta$), and (3) the mixing proportion which represents the probability of hesitant transition durations (indicated as $\theta$).
-
-We present the mixture-model estimates for each of the three parameter values in Figure \ref{fig:crossstudypost}.
-Although models were fitted with all dataset-specific condition, we aggregated across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence for differences between these conditions], and removed conditions that might conflate comparisons^[We removed masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO dataset, and L2 writing in the SPL2 data.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that evidence for differences in parameter values for the involved manipulations.].
-For posteriors of all conditions within dataset see Appendix \ref{posterior-parameter-estimates}.
-The posterior of the remaining conditions allows us to compare differences between transition locations within dataset associated with each of the three mixture-model parameters.
-
-```{r crossstudypost, fig.cap="Estimates across studies. Posterior parameter distribution represented as estimate and 95\\% probability interval (PI)."}
-
-posd <- position_dodge(.65)
-dotsize <- 2.5
-grouplabel <- "Data set"
+As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
+Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
+There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as $\beta$ in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as $\delta$), and (3) the probability of hesitant transition duration (represented as $\theta$).
+
+We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
+Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
+We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.], and removed conditions that might conflate comparisons^[We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
+Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.].
+For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
+The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
+
+```{r crossstudypost, fig.cap="Mixture model parameter estimates across studies.
+Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
+Axes for transitions durations are log-scaled for visability."}
+
+posd <- position_dodge(.5)
+dotsize <- 2.75
+grouplabel <- "Data set:"
 shapes <- c(1, 4, 6, 7, 8, 9, 11)
 width <- 0
-linewidth <- .75
+linewidth <- .35
+nrow <- 1
 beta <- filter(ps, name == "beta") %>% 
@@
   geom_line(linewidth = linewidth, 
-            alpha = .35, 
+            alpha = .75, 
             position = posd) +
@@
   scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
+  scale_y_log10(labels = scales::comma, breaks = c(50, 150, 200, 300, 400, 500)) +
   theme(axis.title = element_blank()) +
@@
        shape = grouplabel) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
+  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
+         shape = guide_legend(nrow=nrow,byrow=TRUE))
 
@@
              group = interaction(data))) +
-  geom_line(linewidth = linewidth, alpha = .35, position = posd) +
+  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
   geom_errorbar(width = width, alpha = .75, position = posd) +
@@
   theme(axis.title = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
+  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
+         shape = guide_legend(nrow=nrow,byrow=TRUE))
 
@@
   theme(axis.title = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-plot <- beta + delta + theta +
-  plot_layout(guides = "collect")
+  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
+         shape = guide_legend(nrow=nrow,byrow=TRUE))
+
+plot <- beta + delta + theta + plot_layout(guides = "collect")
 
@@
 
-Figure \ref{fig:crossstudypost} shows largely the same patterns for keystroke interval estimates by transition location across datasets for all three mixture model parameters.
-keystroke transitions were substantially longer for the C2L1 data presumably reflecting that the sample of in this dataset was the youngest.
-Pause durations and frequencies were relatively similar across datasets.
-
-In the following we will evaluate differences between transition locations for all three mixture model parameters.
+In the following we evaluate differences between transition locations for all three mixture model parameters.
+Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
+We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
+We found that hesitations appear more frequently at before-word transitions than within words across data sets.
+Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
+Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
+
+Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
+However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
 This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence $>$ before-word $>$ within-word.
-We evaluated the differences between transition locations for all data sets.
-The results are shown in Table \ref{tab:loceffect}.
-
-Comparing transition locations rendered patterns that are largely consistent across data sets (with caveats).
-We found that hesitations appear more frequently at before-word transitions than within words across datasets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except dataset C2L1) compared to within-word transitions (except dataset LIFT).
-However, for most dataset there is negligble evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
-In fact, the dataset LIFT showed less pausing before sentences compared to before words.
-Also, we observe that even fluent key-transitions are slower at before-word locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for datasets SPL2 and GUNNEXP2).
-Hesitation durations tend to be longer at before-sentence locations compared to before-word locations (except for datasets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT dataset).
+In fact, the data set LIFT showed less pausing before sentences compared to before words.
 
@@
 
-The inconsistencies for before-sentence transitions could, to some extent, be explained on the basis that datasets differ as to whether before-sentence transitions should be analysed as complex key combination.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-Notice though that for the datasets PLanTra and LIFT, there were no consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-We test to what extent this difference may affect the modelling results for the SPL2 dataset.
+Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
+This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
+Hesitation duration and frequencies were similar to the other data sets.
+
+There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
+Some data sets show before-sentence slowdowns for fluent transition compared to words.
+These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
+In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
+Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
+
+To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
 The results are shown in Appendix \ref{key-combination-effect-spl2}.
 We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this is in conflict with the absence of differences in datasets that included the character following shift at before-sentence transitions.
-In other words, it is unlikely that patterns in our results can be explained on the basis of complex key combinations at before-sentence locations.
+However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
+In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
 
@@
 
-## Simulation
-
-A possible concern with the results -- substantially better predictive performance for mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-
-To address this concern we repeated a comparison for an unimodal model and a mixture model for two sets of simulated data. For each dataset we simulated 1,000 observations.
-
-These two datasets differ with regards to their underlying data generating process.
-The first set of data has as underlying data generating process a mixture model with two mixture components similar to the process described above.
-This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:simmog)
-\text{y}_i \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-		& \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical with the models before but reduced to its main parameters.
-There are two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.
-
-The second dataset was generated with an unimodal model as data generating process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:simuv)
-\text{y}_i \sim\text{ }& \text{logN}(\beta, \sigma^2)
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the unimodal models used in the main analysis.
-The model assumes a log-normal distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
-
-The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
-The simulated data are visualised in Figure \ref{fig:simdata}.
-The data are simulated to be similarly distributed to keystroke transitions.
-
-```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=3.5, fig.width=5}
-files <- list.files("../mixture-model-sim/data", full.names = T)
-
-map_dfr(files, ~read_csv(.) %>% 
-        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
-  mutate(across(Dataset, ~recode(., mogdata = "bimodal", lmdata = "unimodal"))) %>% 
-  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
-  geom_density(alpha = .25) +
-  scale_colour_colorblind() + 
-  scale_fill_colorblind() +
-  scale_x_continuous(labels = ~format(., big.mark = ",")) +
-  labs(colour = "Data process", 
-       fill = "Data process", 
-       x = "Simulated data", 
-       y = "Density")
-```
-
-We run 4 models, 2 on each of the two datasets: 2 mixture models, one for the data generated with a bimodal mixture process and one for the data generated with the unimodal process.
-We repeated the same using an unimodal model.
-Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
-Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}.
-The parameters are shown by data generating process along with with true parameter values.
-Parameter estimates are shown for the dataset generated with a unimodal process and the dataest generated with a bimodal process.
-The Bayesian models uncovered the model parameters of the dataset with the matching data generating process successfully, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-
-```{r}
-mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
-  mutate(data = "mog")
-mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
-  mutate(data = "mog")
-
-mog_sim <- bind_rows(mog1, mog2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>%
-  select(-lo, -up) %>% 
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = c(5, 1, .35, .25, .5),
-         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(name, ~str_replace(., "prob", "theta"))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
-
-lm_sim <- bind_rows(lm1, lm2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>% 
-  select(-lo, -up) %>%
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = as.character(c(5, .25))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
-
-```
-
-```{r simparam, results='asis'}
-table <- bind_rows(mog_sim %>% mutate(model = "mog"),
-                   lm_sim %>% mutate(model = "lm")) %>% 
-  relocate(model) %>% 
-  mutate(across(name, ~str_c("$", ., "$")))
-
-names(table) <- c("Model", "Parameter with true value", "Bimodal data", "Unimodal data")
-
-apa_table(table[,-1], 
-          escape = FALSE,
-      col_spanners = list("Estimate with 95\\% PI" = 2:3),
-      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
-      align = c("l", rep("r", 2)),
-      stub_indents = list(`Bimodal mixture process` = 1:5, 
-                          `Unimodal process` = 6:7))
-
-```
-
-We used LOO-CV to compare the fit of the two models for each data set.
-The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
-The results rules out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a unimodal process.
-However, for the data generated with a bimodal process, the mixture model shows a substantially higher predictive performance.
-
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence [@sivula2020uncertainty], shows that the mixture model performs 11.6 times better than the unimodal model for the data generated with a bimodal process.
-In comparison, for the unimodal data, the unimodal model performs only 0.77 times better than the bimodal mixture model.
-Thus, even though the mixture model does not necessarily perform better for non-bimodal data, it also does not necessarily perform worse.
-
-```{r loossim, results='asis'}
-table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
-  select(1:6) %>% 
-  mutate(
-#    elpd_se_ratio = abs(elpd_diff / se_diff),
-    across(ends_with("diff"), ~round(., 1)),
-    across(ends_with("loo"), ~round(., 0)),
-    across(where(is.numeric), ~format(., big.mark = ",")),
-    across(everything(), ~as.character(.)),
-    across(everything(), ~str_trim(.)),
-    across(everything(), ~str_replace_all(., "NaN", "--")),
-    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
-    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
-    across(c(data, model), ~recode(., mog = "Bimodal mixture model",
-                                      lm = "Unimodal model"))) %>% 
-  select(data, model, starts_with("elpd"))
-
-names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
-
-apa_table(table[,-1], 
-          escape = FALSE,
-      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      align =c("l", rep("r", 2)),
-      stub_indents = list(`Data: Bimodal mixture process` = 1:2, 
-                          `Data: Unimodal process` = 3:4))
-
-```
-
 # Discussion
+
+The classical serial view [@flower1980ynamics] characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
+Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
+We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
+On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
+
+Patters observed in the parameter estimates are generally similar across data sets (with caveats).
+Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
+On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
+We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
+Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
+In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
+However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
+This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed,  they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
+
+Interestingly even the youngest sample in our data pool (C2L1)
+
+showed evidence that pauses at before sentence locations were not obligatory and similar to pauses before words.
+This is interesting because previous research suggested that encoding of subsequent planning units is more separated -- and therefore operates in a serial fashion -- in inexperienced writers [@olive2014toward].
+
+We did observe four difference between data sets that are worth highlighting.
+These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre [see also @conijn2019understanding]:
+
+First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
+Yet the overall pausing probability showed a patterns comparable to most other data sets.
+In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
+The second and third exceptions concern the GUNNEXP2 and SPL2 data sets:  pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
+**TODO: I don't have an explanation for this.** Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
+This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
+Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
+(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
+The absence of a difference could be explained as specific to the synthesis writing task [**REFERENCE**].
+
+Using cross-validation and a simulation we ruled out that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
+Modelling details can be found in Appendix \ref{simulation}.
+
+A practical advantage of our mixture model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
+Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent].
+There is one central problems with this approach: A threshold requires a definition of what passes as a pause [@wen06;@van2016keystroke], i.e.
+a pause criterion threshold often set to 2 secs [@chanquoy1996writing; @kaufer1986composing; @sullivan2002self; @wen02] or some other lower bound [@chukharev2014pauses; @connelly2012predicting; @leijten2013keystroke].
+Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause [see e.g.
+@wen06] they would be arbitrary.
+For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
+Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
+So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
+However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose [@chenu2014interword].
+Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [ see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying].
+
+<!--
+
+All plantra results are L2 but not the other studies
+
+to what extent do pauses reflect reading behaviour?
+
+Say how the results don't contradic the reported studies but speaks better to their conclusions 
+but the modelling frameword doesn't require us to impose pause thresholds
+
+(2) dichotomising latencies into pauses and non-pauses discards much of the information that is captured within a keystroke log.
+
+One strategy for handling the complex distribution of latency from spontaneous text production is to avoid reporting central tendency altogether and to just count latencies that exceed a specific threshold – traditionally set at 2 seconds.
+This practice, inherited from early research in speech production (see Rochester, 1973 for a review) has been widely adopted by writing researchers, with studies dating back to at least the mid 1990s (Foulin, 1995, 1998).
+Pause counting has two disadvantages.
+First, researchers have to make an a priori decision about where to set the pause threshold.
+Current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
+So while a 2 s threshold undoubtedly captures an interesting distinction – processing that occurs the range zero to 2 s is very likely to be qualitatively different from processing that takes more than 2 s.
+However the same could be argued for any threshold between perhaps 250 ms and 10 that a researcher might care to choose (Chenu et al., 2014).
+Moreover, as we have already discussed, interpretation of latencies depends on location within the text.
+A sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
+A second disadvantage of pause counting is that dichotomizing latencies discards much of the information that is captured within a keystroke log or digitized handwriting trace.
+We therefore require statistical methods that model the variance across the full range keystroke latencies (or similar measures from handwriting).
+
+Open: to what extent are even short latencies longer at sentences.
+This is obvious for when multiple keystrokes are summed across at sentences boundaries as there are typically three key transitions necessary to start a sentence: (1) .^_ , (2) _^[shift], (3) [shift]^C
+We believe the second one is the linguistically meaningful one
+I think give Davids papers as examples, if he uses sums instead of means (other studies?)
+
+I just read the abstract of the first one, but the neural oscillations that they describe are occur every 153 ms (6.5 hz). Which is kind of the magic number for the short-duration IKI component.
+
+Duprez, J., Stokkermans, M., Drijvers, L., & Cohen, M.
+X.
+(2021).
+Synchronization between keyboard typing and neural oscillations.
+Journal of Cognitive Neuroscience, 33(5), 887–901.
+https://doi.org/10.1162/jocn_a_01692
+
+González, N., & Calot, E.
+P.
+(2023).
+Dataset of human-written and synthesized samples of keystroke dynamics features for free-text inputs.
+Data in Brief, 48, 109125.
+https://doi.org/10.1016/J.DIB.2023.109125
+
+I think the serial account isn't all inconsistent with the mixture model.
+I don't think the serial account says that pauses must occure on linguistic edges but between bursts which can in principlpe happen anywhere (but why would they be random).
+Obviously the problem here is that one needs to postulate what a pause is instead of leaving it to the model to work out.
+The serial account and the cascading account make very similar predictions for inexperienced writers.
+
+To what extent do pauses reflect reading back. Cite literature from Marks first SPL2 draft on reading during writing.
+
+- Reading is necessary for revision: check if writing goals have been met.
+- Planning what to say next: reinstating ideas after interruption of writing process.
+- Error evaluation: e.g. spelling, grammar
+
+- we removed transitions that terminate in editiing so lookbacks are not associaed with revisions.
+
+- Adults writing argumentative essays
+- @torrance2016reading: 
+  - before sentences 45% lookbacks
+  - before words 12% lookbacks
+  - within word 5% lookbacks 
+- 36% sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]
+- Limited work: @alamargot2010using; @beers2010adolescent; @de2018exploring; @van2010reading
+
+-->
+
+# Conclusion
+
+In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
+Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
+Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
+This pattern was found to be largely consistent different levels of writing experience and languages (e.g.
+young / L2 writers, students) and writing tasks (e.g.
+essays, syntheses) included in our data pool.
 
@@
 render_appendix("appendix_f.Rmd")
-```
+render_appendix("appendix_h.Rmd")
+```

### 2023-09-15 report/appendix_h.Rmd
Add first rough draft

@@
 However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance.
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence [@sivula2020uncertainty], shows that the mixture model performs 11.6 times better than the uni-modal model.
+In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence (Sivula et al., 2020), shows that the mixture model performs 11.6 times better than the uni-modal model.
 In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model.

### 2023-09-15 report/appendix_h.tex
Add first rough draft

--- 2023-09-15 report/appendix_h.Rmd
@@
-## Simulation
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
 
-A possible concern with these results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-To address this concern we repeated a comparison for a uni-modal model and a bi-modal mixture model for two sets of simulated data that were simulated either with a uni-modal or bi-modal process at heart.
-In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
+\begin{appendix}
+\section{}
+\hypertarget{simulation}{%
+\subsection{Simulation}\label{simulation}}
 
-In particular, the first data set was simulated using a bi-modal mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}).
-This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
+A possible concern with these results -- substantially better predictive
+performance for bi-modal mixture models -- is that, in principle, as the
+mixture model has more parameters it might always lead to a better fit.
+We addressed this concern before by using cross-validation techniques
+for model comparison which is preventing overfitting models. To address
+this concern we repeated a comparison for a uni-modal model and a
+bi-modal mixture model for two sets of simulated data that were
+simulated either with a uni-modal or bi-modal process at heart. In other
+words, this allows us to test the predictive performance of our models
+in a context where we know the true underlying data generating process
+-- uni-modal vs bi-modal -- and we can test whether these models can
+successfully uncover the true parameter value.
+
+In particular, the first data set was simulated using a bi-modal mixture
+model with two mixture components similar to the process described above
+(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
+and the corresponding Bayesian model is summarised in equation
+\ref{eq:simmog}.
 
@@
 
-This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants).
-The model includes two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.
+This model is largely identical to the models before but reduced to its
+main parameters (but not mixed effects for participants). The model
+includes two log-normal distributions with a mixing proportion
+\(\theta\) of which the distribution of shorter values has a mean of
+\(\beta\) and a standard deviation \(\sigma^2_2\); the second
+distribution of longer values is constrained to have a mean that is
+larger by a factor of \(\delta\) and has a larger standard deviation
+\(\sigma^2_1\).
 
-The second data set was generated with a uni-modal log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
+The second data set was generated with a uni-modal log-Gaussian process.
+The model and its corresponding Bayesian model is summarised in equation
+\ref{eq:simuv}.
 
@@
 
-Again, this model is a simplified version of the uni-modal models used in the main analysis.
-The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
+Again, this model is a simplified version of the uni-modal models used
+in the main analysis. The model assumes a log-Gaussian distribution with
+a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
 
-The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
-The simulated data are visualised in Figure \ref{fig:simdata}.
-Parameter values were chosen so that the simulated data are roughly similarly distributed to keystroke transitions.
+The parameter values used for each of the two data simulations can be
+found in Table \ref{tab:simparam}. The simulated data are visualised in
+Figure \ref{fig:simdata}. Parameter values were chosen so that the
+simulated data are roughly similarly distributed to keystroke
+transitions.
 
-```{r simdata, fig.cap= "Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process.
-The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
-files <- list.files("../mixture-model-sim/data", full.names = T)
+\begin{figure}
 
-map_dfr(files, ~read_csv(.) %>% 
-        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
-  mutate(across(Dataset, ~recode(., mogdata = "bi-modal", lmdata = "uni-modal"))) %>% 
-  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
-  geom_density(alpha = .25) +
-  scale_colour_colorblind() + 
-  scale_fill_colorblind() +
-  scale_x_log10(labels = ~format(., big.mark = ",")) +
-  theme(text = element_text(size = 10)) +
-  labs(colour = "Data process", 
-       fill = "Data process", 
-       x = "y", 
-       y = "Density")
-```
+{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
 
-For each of these two data sets we simulated 1,000 observations.
-We run 2 models -- a bi-modal mixture model and a uni-modal model -- each for both data sets.
-Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
-Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}.
-The parameters are shown by type of data generating process along with the true parameter values.
-Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they uni-modal model.
+}
 
-```{r}
-mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
-  mutate(data = "mog")
-mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
-  mutate(data = "mog")
+\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}(\#fig:simdata)
+\end{figure}
 
-mog_sim <- bind_rows(mog1, mog2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>%
-  select(-lo, -up) %>% 
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = c(5, 1, .35, .25, .5),
-         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(name, ~str_replace(., "prob", "theta"))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
+For each of these two data sets we simulated 1,000 observations. We run
+2 models -- a bi-modal mixture model and a uni-modal model -- each for
+both data sets. Models were run with 3 chains, with each 6,000
+iterations of which 3,000 were warmup. Estimates with 95\% probability
+intervals are shown in Table \ref{tab:simparam}. The parameters are
+shown by type of data generating process along with the true parameter
+values. Parameter value estimates are shown by type of Bayesian model.
+The results show that either of the two Bayesian models succesfully
+uncovered the model parameters of the data with its corresponding data
+generating process, as shown in Table \ref{tab:simparam}, but less so
+when the model was applied to data generated with the incorrect
+underlying process. Particularly the mixing proportion \(\theta\) and
+the slowdown parameter \(\delta\) were not uncovered at all by they
+uni-modal model.
 
-lm_sim <- bind_rows(lm1, lm2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>% 
-  select(-lo, -up) %>%
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = as.character(c(5, .25))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
+\begin{table}[tbp]
 
-```
+\begin{center}
+\begin{threeparttable}
 
-```{r simparam, results='asis'}
-table <- bind_rows(mog_sim %>% mutate(model = "mog"),
-                   lm_sim %>% mutate(model = "lm")) %>% 
-  relocate(model) %>% 
-  mutate(across(name, ~str_c("$", ., "$")))
+\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
 
-names(table) <- c("Model", "Parameter with true value", "Bi-modal model", "Uni-modal model")
+\begin{tabular}{lrr}
+\toprule
+ & \multicolumn{2}{c}{Estimate with 95\% PI} \\
+\cmidrule(r){2-3}
+Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
+\midrule
+Bi-modal data &  & \\
+\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
+\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
+\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
+\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
+\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
+Uni-modal data &  & \\
+\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
+\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
+\bottomrule
+\end{tabular}
 
-apa_table(table[,-1], 
-          escape = FALSE,
-      col_spanners = list("Estimate with 95\\% PI" = 2:3),
-      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
-      align = c("l", rep("r", 2)),
-      stub_indents = list(`Bi-modal data` = 1:5, 
-                          `Uni-modal data` = 6:7))
+\end{threeparttable}
+\end{center}
 
-```
+\end{table}
 
-We used LOO-CV to compare the predictive performance of the two models for each data generating process.
-The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
-The results rule out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a uni-modal process.
-However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance.
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence (Sivula et al., 2020), shows that the mixture model performs 11.6 times better than the uni-modal model.
-In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model.
-Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the uni-modal is preferred by the law of parsimony.
+We used LOO-CV to compare the predictive performance of the two models
+for each data generating process. The model comparisons can be found for
+each data generating process in Table \ref{tab:loossim}. The results
+rule out the possibility that the mixture model does always lead to
+higher predictive performance. Indeed, the mixture model showed a
+slightly lower predictive performance for the data that were generated
+with a uni-modal process. However, for the data generated with a
+bi-modal process, the mixture model shows a substantially higher
+predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
+and its standard error, as metric for the strength of evidence (Sivula
+et al., 2020), shows that the mixture model performs 11.6 times better
+than the uni-modal model. In comparison, for the uni-modal data, the
+uni-modal model performs only 0.77 times better than the bi-modal
+mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
+the model is negligible, the uni-modal is preferred by the law of
+parsimony.
 
-```{r loossim, results='asis'}
-table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
-  select(1:6) %>% 
-  mutate(
-#    elpd_se_ratio = abs(elpd_diff / se_diff),
-    across(ends_with("diff"), ~round(., 1)),
-    across(ends_with("loo"), ~round(., 0)),
-    across(where(is.numeric), ~format(., big.mark = ",")),
-    across(everything(), ~as.character(.)),
-    across(everything(), ~str_trim(.)),
-    across(everything(), ~str_replace_all(., "NaN", "--")),
-    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
-    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
-    across(c(data, model), ~recode(., mog = "Bi-modal mixture model",
-                                      lm = "Uni-modal model"))) %>% 
-  select(data, model, starts_with("elpd"))
+\begin{table}[tbp]
 
-names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
+\begin{center}
+\begin{threeparttable}
 
-apa_table(table[,-1], 
-          escape = FALSE,
-      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
-     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      align =c("l", rep("r", 2)),
-      stub_indents = list(`Data: Bi-modal mixture process` = 1:2, 
-                          `Data: Uni-modal process` = 3:4))
+\caption{\label{tab:loossim}Model comparisons by data set.
+The top row shows the models with the highest predictive performance for each data generating process.
+Standard error is shown in parentheses.}
 
-```
+\begin{tabular}{lrr}
+\toprule
+Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
+\midrule
+Data: Bi-modal mixture process &  & \\
+\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
+\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
+Data: Uni-modal process &  & \\
+\ \ \ Uni-modal model & -- & -5,030 (24)\\
+\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
+\bottomrule
+\addlinespace
+\end{tabular}
+
+\begin{tablenotes}[para]
+\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
+\end{tablenotes}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+\end{appendix}

### 2023-09-15 report/manuscript.Rmd
Add first rough draft

@@
         - "appendix_f.tex"
-        - "appendix_g.tex"
+        - "appendix_h.tex"
 
@@
 
-Interestingly even the youngest sample in our data pool (C2L1)
-
-showed evidence that pauses at before sentence locations were not obligatory and similar to pauses before words.
-This is interesting because previous research suggested that encoding of subsequent planning units is more separated -- and therefore operates in a serial fashion -- in inexperienced writers [@olive2014toward].
+Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
+For example, @olive2014toward described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
+This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
+However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
+The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
+Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
 
@@
 
-Using cross-validation and a simulation we ruled out that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
+A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
+In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
+We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
+In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
 Modelling details can be found in Appendix \ref{simulation}.
+As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
+Parameters were not successfully uncovered when we switched model type and data set.
+Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model,  when applied to the uni-modal data (but for the bi-modal data).
+Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
+
+Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
+Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
+There is a limited amount of work that has used eye tracking to investiatge regressive eye movement behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
+There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.
+writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.
+spelling, grammar.
+We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
+However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
+This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
+Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
+In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
+In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
 
@@
 
-All plantra results are L2 but not the other studies
-
-to what extent do pauses reflect reading behaviour?
-
+~~~~~
 Say how the results don't contradic the reported studies but speaks better to their conclusions 
 but the modelling frameword doesn't require us to impose pause thresholds
-
-(2) dichotomising latencies into pauses and non-pauses discards much of the information that is captured within a keystroke log.
-
-One strategy for handling the complex distribution of latency from spontaneous text production is to avoid reporting central tendency altogether and to just count latencies that exceed a specific threshold – traditionally set at 2 seconds.
-This practice, inherited from early research in speech production (see Rochester, 1973 for a review) has been widely adopted by writing researchers, with studies dating back to at least the mid 1990s (Foulin, 1995, 1998).
-Pause counting has two disadvantages.
-First, researchers have to make an a priori decision about where to set the pause threshold.
-Current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 s threshold undoubtedly captures an interesting distinction – processing that occurs the range zero to 2 s is very likely to be qualitatively different from processing that takes more than 2 s.
-However the same could be argued for any threshold between perhaps 250 ms and 10 that a researcher might care to choose (Chenu et al., 2014).
-Moreover, as we have already discussed, interpretation of latencies depends on location within the text.
-A sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-A second disadvantage of pause counting is that dichotomizing latencies discards much of the information that is captured within a keystroke log or digitized handwriting trace.
-We therefore require statistical methods that model the variance across the full range keystroke latencies (or similar measures from handwriting).
-
+~~~~~
+
+~~~~
+Move from results to discussion?
 Open: to what extent are even short latencies longer at sentences.
@@
 
-I just read the abstract of the first one, but the neural oscillations that they describe are occur every 153 ms (6.5 hz). Which is kind of the magic number for the short-duration IKI component.
+~~~
+
+neural oscillationsoccur every 153 ms (6.5 hz). Which is kind of the magic number for the short-duration IKI component.
 
@@
 
+~~~~
 I think the serial account isn't all inconsistent with the mixture model.
@@
 
-To what extent do pauses reflect reading back. Cite literature from Marks first SPL2 draft on reading during writing.
-
-- Reading is necessary for revision: check if writing goals have been met.
-- Planning what to say next: reinstating ideas after interruption of writing process.
-- Error evaluation: e.g. spelling, grammar
-
-- we removed transitions that terminate in editiing so lookbacks are not associaed with revisions.
-
-- Adults writing argumentative essays
-- @torrance2016reading: 
-  - before sentences 45% lookbacks
-  - before words 12% lookbacks
-  - within word 5% lookbacks 
-- 36% sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]
-- Limited work: @alamargot2010using; @beers2010adolescent; @de2018exploring; @van2010reading
+Maybe distinguih between a strong version of serial model and a weak version that is basically a chunky cacading model
 

### 2023-09-15 report/manuscript.tex
Add first rough draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,floatsintext]{apa7}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
  pdflang={en-EN},
  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Modelling writing hesitations}
\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{colortbl}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}
\usepackage[utf8]{inputenc}
\usepackage{icomma}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
\date{}

\authornote{

Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}

\abstract{
Classical serial models view the writing process as a chain of pauses and writing burst. In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution. We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production. We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions. Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets. Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations. Our results cannot be explain by the serial but are in line with the cascading view of writing.
}

\begin{document}
\maketitle

Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).

Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech. Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language. Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012). This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g. Clark \& Fox Tree, 2002). Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication. Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).

The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models. On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006). For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019). The same is generally observed for spoken utterances (E.-K. Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001). We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted. Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.

Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991). The intervals between adjacent keystrokes during typing vary as a function of various factors. Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured. Processing may be serial. This means that the processing cascade operates on one unit at a time. Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output. Therefore longer IKIs represent periods when the writer is planning what to write next. The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002). This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).

This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002). However this account is not consistent with what we know about writing. Consider these two observations:

First, utterances are not fully formed at production onset. Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process. This is well known for spoken production (e.g. Bock \& Ferreira, 2014). For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al. (2019){]}. For example, in Roeser et al. (2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}). Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset. This is not what one would expect if writers plan sentences (or even phrases) in full before production onset. In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al. (2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.

Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid. For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration. Also Rønneberg et al. (2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses. For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population. Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent. The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.

If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations. The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control. Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991). Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016). Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes. In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.

This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals. IKIs result from one of two data-generating processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement. IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).

For example, Roeser et al. (2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling. In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach. Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions. As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition. Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes. For similar approaches see Almond et al. (2012), Baaijen et al. (2012), Chenu et al. (2014), Guo et al. (2018), Hall et al. (2022), Li (2021), and Van Waes et al. (2021).

The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. Rønneberg et al. (2022) coined this possibility the \enquote{process-disruption hypothesis.}

In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al. (2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts. On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).

Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production. Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).

Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023). Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016. These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).

Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021). In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks. Instead text composition provides data from both inter and intra-sentence transitions. Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).

The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location. In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations. This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation. The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing. In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production. Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability. We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models. The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models. In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).

The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al. (2021). Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow. Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur. This view was implemented as a constrained and an unconstrained model.

\hypertarget{statistical-models}{%
\subsection{Statistical models}\label{statistical-models}}

We are using the Bayesian framework (e.g. Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M. D. Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing. In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models. The models presented in this section build on one another so that later model include assumptions of earlier models.

\hypertarget{serial-model-of-writing}{%
\subsubsection{Serial model of writing}\label{serial-model-of-writing}}

\hypertarget{unimodal-gaussian}{%
\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}

Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit. The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors. For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes. There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis. We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model. This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\). Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.

We can extend this simple model of word-planning to other linguistic location. We introduce earlier that larger linguistic edges are associated with planning on higher levels. For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019). Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors. We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries. However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs. Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning. The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).

For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location. Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs. Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\). The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\). This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).

The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.

\begin{equation}
\begin{aligned}
\label{eq:unimodgaus}
\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.

\hypertarget{unimodal-log-gaussian}{%
\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}

The previous model assumes that the data-generating process is a Gaussian distribution. The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution. There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values. IKIs, as the distance between two subsequent key-down events, must be positive. The lower bound is delimited by a person's ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour and motor responses. In particular, a Gaussian distribution assumes that units are scaled linearly. For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs. This does not necessarily map onto the psychological interpretation for short and long keyintervals. For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2). In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals. Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).

The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).

\begin{equation}
\begin{aligned}
\label{eq:unimodloggaus}
\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

\hypertarget{unimodal-unequal-variance-log-gaussian}{%
\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}

The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section. The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text. The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance). This assumption is however not in line with what we know about data from human behaviour. Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973). For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes. Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.

\begin{equation}
\begin{aligned}
\label{eq:unimoduv}
\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
 & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

\hypertarget{parallel-cascading-model-of-writing}{%
\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}

The following two models are are extensions of the models introduced for the serial view. Crucially, the cascading view allows planning to happen in parallel to production. Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset. This is done by assuming that IKIs come from a weighted combination of two distributions.

\hypertarget{bimodal-log-gaussian-constrained}{%
\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}

This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses. Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory. This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Activation can flow into keystrokes without interruption. These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}. Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
\item
  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time. The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}. The magnitude of the slowdown is associated with transition location. This is because delays at larger linguistic units are likely to be associated with higher level planning. By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
\end{enumerate}

\begin{equation}
\begin{aligned}
\label{eq:bimodcon}
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
        & \sigma_{e'}^2 > \sigma_{e}^2\\
        & 0 > \theta > 1
\end{aligned}
\end{equation}

The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions. Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1. \(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs. This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\). In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution. The weights of both distributions must sum to 1. We will call this parameter the probability of hesitant transitions.

The probability of hesitant transitions is assumed to vary as a function of both transition location and participants. In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges. As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less. This is akin to what is known as a random by-participant slopes model.

Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location. In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017). This was achieved by introducing a parameter \(\sigma_\text{diff}\). The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.

\hypertarget{bimodal-log-gaussian-unconstrained}{%
\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}

In principle, we do not need to assume that the size of a fluent key-transition varies by transition location. In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions. This is what we called a constrained model. However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section). This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words. Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data. We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.

In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.

\begin{equation}
\begin{aligned}
\label{eq:bimoduncon}
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
        & \sigma_{e'}^2 > \sigma_{e}^2
\end{aligned}
\end{equation}

\hypertarget{data-sets}{%
\subsection{Data sets}\label{data-sets}}

Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.

\begin{landscape}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:datasets}Datasets in brief.}

\footnotesize{

\begin{tabular}{llllrrp{2cm}p{3cm}}
\toprule
Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
\midrule
C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
\bottomrule
\end{tabular}

}

\end{threeparttable}
\end{center}

\end{table}
\end{landscape}

\hypertarget{c2l1}{%
\subsubsection{C2L1}\label{c2l1}}

TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.

TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.

The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al. (2022). The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography. Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).

\hypertarget{cato}{%
\subsubsection{CATO}\label{cato}}

TODO: Mark, can you add information here?

Data are published in Torrance, Rønneberg, et al. (2016). Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing. Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).

\hypertarget{gunnexp2}{%
\subsubsection{GUNNEXP2}\label{gunnexp2}}

TODO: Mark, can you add information here? I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk). Was the sample Norwegian?

The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX). In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal. Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).

\hypertarget{lift}{%
\subsubsection{LIFT}\label{lift}}

The LIFT data are published in Vandermeulen, Steendam, et al. (2020) and described in Vandermeulen, De Maeyer, et al. (2020). The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing. Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools. The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between. In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school. The students received 50 minutes for each tasks. Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts. During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).

\hypertarget{plantra}{%
\subsubsection{PLanTra}\label{plantra}}

The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a). The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts. A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated. The study adopted a pre-test post-test design. As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience. Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability. Participants were asked to spend at least 45 minutes on the instruction module. As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability. Both reports were written in English (second language) and similar in length (274-278 words) and readability. Participants received as much time as needed for each task. During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).

\hypertarget{spl2}{%
\subsubsection{SPL2}\label{spl2}}

TODO: Mark, can you add information here?

The SPL2 dataset is published in Torrance et al. (n.d.). The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019). Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

\hypertarget{bayesian-modelling}{%
\subsubsection{Bayesian modelling}\label{bayesian-modelling}}

We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}

\begin{tabular}{llrp{9cm}}
\toprule
Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
\midrule
M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
\bottomrule
\addlinespace
\end{tabular}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
\end{tablenotes}

\end{threeparttable}
\end{center}

\end{table}

Bayesian models, as used in this paper, are ideal for the estimation of parameter values. This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M. D. Lee \& Wagenmakers, 2014). To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values. For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992). In other words the choice of priors values has less impact on the posterior. In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g. Lambert, 2018; McElreath, 2016). Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.

For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}. Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location. Stan code for mixture models was based on Roeser et al. (2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.

Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016). The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models. Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning. Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains. The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).

\hypertarget{transition-types}{%
\subsubsection{Transition types}\label{transition-types}}

The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g. Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}. In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions. At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section. Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:keyloc}Transition location classification.}

\footnotesize{

\begin{tabular}{p{3cm}p{3cm}p{8cm}}
\toprule
Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
\midrule
Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d. T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
\bottomrule
\addlinespace
\end{tabular}

}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
\end{tablenotes}

\end{threeparttable}
\end{center}

\end{table}

\hypertarget{data-reduction}{%
\subsubsection{Data reduction}\label{data-reduction}}

For all datasets we only used transitions that were not followed by an editing operation. Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed. We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set). We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).

We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}. From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set. This was done for computational reasons to reduce the time the Bayesian models need to complete sampling. For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool. Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources. The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.

\begin{table}[bp!]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:datareduction}Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location. Standard error is shown in parentheses.}

\begin{tabular}{lrrrrr}
\toprule
 \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
\midrule
C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{out-of-samples-cross-validation}{%
\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}

To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017). Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models. Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.

Model comparison results for all data sets are shown in Table \ref{tab:loos}. For all data sets we found the same pattern. Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models. The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations). Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model. The weakest model was the uni-modal Gaussian model.

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
\end{TableNotes}

\small{

\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
\toprule
Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
\toprule
Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
\midrule
\endhead
\textbf{CL21} &  &  & \\
\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
\textbf{CATO} &  &  & \\
\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
\textbf{GUNNEXP2} &  &  & \\
\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
\textbf{LIFT} &  &  & \\
\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
\textbf{PLanTra} &  &  & \\
\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
\textbf{SPL2} &  &  & \\
\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}

We also evaluated to what extent model predictions fit observed data. These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}. Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.

\hypertarget{cross-data-set-comparisons}{%
\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}

As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models. Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates. There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).

We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}. Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set. We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set. Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}. For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}. The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}

}

\caption{Mixture model parameter estimates across studies. Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI). Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group. Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
\end{figure}

In the following we evaluate differences between transition locations for all three mixture model parameters. Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters. We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}. We found that hesitations appear more frequently at before-word transitions than within words across data sets. Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT). Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).

Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set). However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2). This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word. In fact, the data set LIFT showed less pausing before sentences compared to before words.

In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.

Outstanding are the overall substantially longer fluent transitions for the C2L1 data. This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool. Hesitation duration and frequencies were similar to the other data sets.

There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions. Some data sets show before-sentence slowdowns for fluent transition compared to words. These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters. In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2). Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).

To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set. The results are shown in Appendix \ref{key-combination-effect-spl2}. We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency. However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT). In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).

Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.

\begin{landscape}

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals. Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
 \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
\toprule
 \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
\midrule
\endhead
\textbf{C2L1} &  &  &  &  &  & \\
\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
\textbf{LIFT} &  &  &  &  &  & \\
\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
\textbf{PLanTra} &  &  &  &  &  & \\
\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
\textbf{SPL2 (L1)} &  &  &  &  &  & \\
\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\end{landscape}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses. Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production. We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively. On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.

Patters observed in the parameter estimates are generally similar across data sets (with caveats). Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing. On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit. We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare. Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses). In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing. However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view. This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.

Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning. For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers. This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion. However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause. The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample. Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.

We did observe four difference between data sets that are worth highlighting. These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):

First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words. Yet the overall pausing probability showed a patterns comparable to most other data sets. In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions. The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations). \textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions. This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets. Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production. (2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets. The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.

A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit. We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models. In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model. We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process. In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value. Modelling details can be found in Appendix \ref{simulation}. As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process. Parameters were not successfully uncovered when we switched model type and data set. Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data). Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.

Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units. Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text. There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010). There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar. We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions. However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al. (2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019). This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words. Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not). In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}. In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.

A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production. Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998). There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013). Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g. Wengelin, 2006) they would be arbitrary. For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word. Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision. So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs. However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014). Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution. Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view. Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view. This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.

\hypertarget{references}{%
\section{References}\label{references}}

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

\hypertarget{references}{}

\endgroup

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}

\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
Ailhaud, E., Chenu, F., \& Jisa, H. (2016). A developmental perspective on the units of written {F}rench. In J. Perera, M. Aparici, E. Rosado, \& N. Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp. 287--305). Springer. \url{https://doi.org/10.1007/978-3-319-21136-7_17}

\leavevmode\hypertarget{ref-alamargot2010using}{}%
Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D. (2010). Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer. \emph{Reading and Writing}, \emph{23}(7), 853--888.

\leavevmode\hypertarget{ref-almond2012preliminary}{}%
Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T. (2012). \emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No. RR-12-23). Educational Testing Service.

\leavevmode\hypertarget{ref-alves2015progress}{}%
Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.

\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.

\leavevmode\hypertarget{ref-beers2010adolescent}{}%
Beers, S. F., Quinlan, T., \& Harbaugh, A. G. (2010). Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study. \emph{Reading and Writing}, \emph{23}(7), 743--775.

\leavevmode\hypertarget{ref-bock2014syntactically}{}%
Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.

\leavevmode\hypertarget{ref-bonin2012evidence}{}%
Bonin, P., Roux, S., Barry, C., \& Canell, L. (2012). Evidence for a limited-cascading account of written word naming. \emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758. \url{https://doi.org/10.1037/a0028471}

\leavevmode\hypertarget{ref-carpenter2016stan}{}%
Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M. A., Guo, J., Li, P., \& Riddell, A. (2016). Stan: A probabilistic programming language. \emph{Journal of Statistical Software}, \emph{20}.

\leavevmode\hypertarget{ref-chang2006}{}%
Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.

\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
Chanquoy, L., Foulin, J.-N., \& Fayol, M. (1996). Writing in adults: {A} real-time approach. In G. Rijlaarsdam, H. Van den Bergh, \& M. Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp. 36--44). Amsterdam University Press.

\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.

\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.

\leavevmode\hypertarget{ref-chenu2014interword}{}%
Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}

\leavevmode\hypertarget{ref-christiansen2016now}{}%
Christiansen, M. H., \& Chater, N. (2016). The now-or-never bottleneck: A fundamental constraint on language. \emph{Behavioral and Brain Sciences}, \emph{39}, e62. \url{https://doi.org/10.1017/S0140525X1500031X}

\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.

\leavevmode\hypertarget{ref-chukharev2019combined}{}%
Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H. (2019). Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency. \emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.

\leavevmode\hypertarget{ref-Clark2002}{}%
Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.

\leavevmode\hypertarget{ref-conijn2020keys}{}%
Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.

\leavevmode\hypertarget{ref-conijn2019understanding}{}%
Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.

\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}

\leavevmode\hypertarget{ref-connelly2012predicting}{}%
Connelly, V., Dockrell, J. E., Walter, K., \& Critten, S. (2012). Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment. \emph{Written Communication}, \emph{29}(3), 278--302.

\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
Crump, M. J. C., \& Logan, G. D. (2010). Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes. \emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380. \url{https://doi.org/10.1037/a0020696}

\leavevmode\hypertarget{ref-dam09}{}%
Damian, M. F., \& Stadthagen-Gonzalez, H. (2009). Advance planning of form properties in the written production of single and multiple words. \emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.

\leavevmode\hypertarget{ref-de2018exploring}{}%
De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.

\leavevmode\hypertarget{ref-dell2007case}{}%
Dell, G. S., Martin, N., \& Schwartz, M. F. (2007). A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming. \emph{Journal of Memory and Language}, \emph{56}(4), 490--520.

\leavevmode\hypertarget{ref-dux2021effect}{}%
Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.

\leavevmode\hypertarget{ref-dux2022automating}{}%
Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E. (2022). Automating individualized, process-focused writing instruction: A design-based research study. \emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878. \url{https://doi.org/0.3389/fcomm.2022.933878}

\leavevmode\hypertarget{ref-farrell2018computational}{}%
Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.

\leavevmode\hypertarget{ref-flower1980ynamics}{}%
Flower, L. S., \& Hayes, J. R. (1980). The dynamics of composing: Making plans and juggling constraints. In L. W. Gregg \& E. R. Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp. 31--50). Lawrence Erlbaum.

\leavevmode\hypertarget{ref-flower1981pregnant}{}%
Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.

\leavevmode\hypertarget{ref-foulin1998extent}{}%
Foulin, J.-N. (1998). To what extent does pause location predict pause duration in adults' and children's writing? \emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.

\leavevmode\hypertarget{ref-gelman2014}{}%
Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.

\leavevmode\hypertarget{ref-gelman1992}{}%
Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.

\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.

\leavevmode\hypertarget{ref-gri01}{}%
Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.

\leavevmode\hypertarget{ref-griffin2003reversed}{}%
Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.

\leavevmode\hypertarget{ref-guo2018modeling}{}%
Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.

\leavevmode\hypertarget{ref-hall2022constructing}{}%
Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.

\leavevmode\hypertarget{ref-hayes2012evidence}{}%
Hayes, J. R. (2012). Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes. In M. Fayol, D. Alamargot, \& V. Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp. 15--25)). Psychology Press.

\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.

\leavevmode\hypertarget{ref-kaufer1986composing}{}%
Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.

\leavevmode\hypertarget{ref-lambert2018student}{}%
Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.

\leavevmode\hypertarget{ref-lee13}{}%
Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.

\leavevmode\hypertarget{ref-lee2014bayesian}{}%
Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.

\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.

\leavevmode\hypertarget{ref-li2021identifying}{}%
Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.

\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.

\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.

\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.

\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.

\leavevmode\hypertarget{ref-meyer1996}{}%
Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.

\leavevmode\hypertarget{ref-mohsen2021second}{}%
Mohsen, M. (2021). Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program. \emph{Psycholinguistics}, \emph{30}(1), 180--202. \url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}

\leavevmode\hypertarget{ref-mohsen2021second}{}%
Mohsen, M. (2021). Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program. \emph{Psycholinguistics}, \emph{30}(1), 180--202. \url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}

\leavevmode\hypertarget{ref-not10}{}%
Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.

\leavevmode\hypertarget{ref-not07}{}%
Nottbusch, G., Weingarten, R., \& Sahel, S. (2007). From written word to written sentence production. In M. Torrance, L. van Waes, \& D. W. Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol. 20, pp. 31--53). Elsevier.

\leavevmode\hypertarget{ref-olive2014toward}{}%
Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.

\leavevmode\hypertarget{ref-R-base}{}%
R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}

\leavevmode\hypertarget{ref-rochester1973significance}{}%
Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.

\leavevmode\hypertarget{ref-roeser2021modelling}{}%
Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L. (2021). Modelling typing disfluencies as finite mixture process. \emph{Reading and Writing}, 1--26. \url{https://doi.org/10.1007/s11145-021-10203-z}

\leavevmode\hypertarget{ref-roeser2018advance}{}%
Roeser, J., Torrance, M., \& Baguley, T. (2019). Advance planning in written and spoken sentence production. \emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009. \url{https://doi.org/10.1037/xlm0000685}

\leavevmode\hypertarget{ref-rossetti2022s}{}%
Rossetti, A., \& Van Waes, L. (2022a). It's not just a phase: Investigating text simplification in a second language from a process and product perspective. \emph{Frontiers in Artificial Intelligence}, \emph{5}. \url{https://doi.org/10.3389/frai.2022.983008}

\leavevmode\hypertarget{ref-rossetti2022text}{}%
Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}

\leavevmode\hypertarget{ref-roux2013interaction}{}%
Roux, S., McKeeff, T. J., Grosjacques, G., Afonso, O., \& Kandel, S. (2013). The interaction between central and peripheral processes in handwriting production. \emph{Cognition}, \emph{127}(2), 235--241.

\leavevmode\hypertarget{ref-ronneberg2022process}{}%
Rønneberg, V., Torrance, M., Uppstad, P. H., \& Johansson, C. (2022). The process-disruption hypothesis: How spelling and typing skill affects written composition process and product. \emph{Psychological Research}, \emph{86}(7), 2239--2255.

\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
Schilperoord, J. (2002). On the cognitive status of pauses in discourse production. In T. Olive \& C. M. Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp. 61--87). Springer.

\leavevmode\hypertarget{ref-schoner2002timing}{}%
Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.

\leavevmode\hypertarget{ref-sim07}{}%
Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.

\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.

\leavevmode\hypertarget{ref-rstan}{}%
Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}

\leavevmode\hypertarget{ref-sullivan2002self}{}%
Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.

\leavevmode\hypertarget{ref-torrance201203}{}%
Torrance, M. (2012). EyeWrite -- a tool for recording writers' eye movements. In M. Torrance, D. Alamargot, M. Castelló, F. Ganier, O. Kruse, A. Mangen, L. Tolchinsky, \& L. V. Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp. 355--357). Brill. \url{https://doi.org/doi.org/10.1163/9781780529295_082}

\leavevmode\hypertarget{ref-torrance2016reading}{}%
Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å. (2016). Reading during the composition of multi-sentence texts: An eye-movement study. \emph{Psychological Research}, \emph{80}(5), 729--743.

\leavevmode\hypertarget{ref-torrance2012written}{}%
Torrance, M., \& Nottbusch, G. (2012). Written production of single words and simple sentences. In V. W. Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp. 403--422). Psychology Press.

\leavevmode\hypertarget{ref-torrance2018timed}{}%
Torrance, M., Nottbusch, G., Alves, R. A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó. I., \& others. (2018). Timed written picture naming in 14 {E}uropean languages. \emph{Behavior Research Methods}, \emph{50}, 744--758.

\leavevmode\hypertarget{ref-torranceb}{}%
Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.

\leavevmode\hypertarget{ref-torrancea}{}%
Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.

\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P. H. (2016). Adolescent weak decoders writing in a shallow orthography: Process and product. \emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.

\leavevmode\hypertarget{ref-van1991handwriting}{}%
Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.

\leavevmode\hypertarget{ref-van2016keystroke}{}%
Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.

\leavevmode\hypertarget{ref-van2019multilingual}{}%
Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E. (2019). A multilingual copy task: Measuring typing and motor skills in writing with inputlog. \emph{Journal of Open Research Software}, \emph{7}(30), 1--8.

\leavevmode\hypertarget{ref-van2010reading}{}%
Van Waes, L., Leijten, M., \& Quinlan, T. (2010). Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity. \emph{Reading and Writing}, \emph{23}(7), 803--834.

\leavevmode\hypertarget{ref-waes2019}{}%
Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J. (2021). Measuring and assessing typing skills in writing research. \emph{Journal of Writing Research}, \emph{13}(1), 107--153. \url{https://doi.org/10.17239/jowr-2021.13.01.04}

\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G. (2020). Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing. \emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.

\leavevmode\hypertarget{ref-vandermeulen2020}{}%
Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}

\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G. (2023). Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts. \emph{Written Communication}, \emph{40}(1), 90--144.

\leavevmode\hypertarget{ref-vasishth2017}{}%
Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B. (2017). Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses. \emph{ArXiv e-Prints}.

\leavevmode\hypertarget{ref-vasishth2017feature}{}%
Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.

\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.

\leavevmode\hypertarget{ref-vehtari2017practical}{}%
Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.

\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
Wagenmakers, E.-J., \& Brown, S. (2007). On the linear relation between the mean and the standard deviation of a response time distribution. \emph{Psychological Review}, \emph{114}(3), 830--841. \url{https://doi.org/10.1037/0033-295X.114.3.830}

\leavevmode\hypertarget{ref-wen02}{}%
Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.

\leavevmode\hypertarget{ref-wen06}{}%
Wengelin, Å. (2006). Examining pauses in writing: {T}heory, methods and empirical data. In K. P. H. Sullivan \& E. Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol. 18, pp. 107--130). Elsevier.

\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}

\leavevmode\hypertarget{ref-wheeldon2013}{}%
Wheeldon, L. R., Ohlson, N., Ashby, A., \& Gator, S. (2013). Lexical availability and grammatical encoding scope during spoken sentence production. \emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.

\leavevmode\hypertarget{ref-wing1973response}{}%
Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.

\end{CSLReferences}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{fit-to-data}{%
\subsection{Fit to data}\label{fit-to-data}}

Model fit was assessed by testing to what extent simulations obtained
from the model fit the observed data. From each model we obtained 100
simulations (predictions). These can be found in Figure
\ref{fig:prediction}. The model predictions fitted the data
substantially better for the mixture models. The worse fit was observed
for the unimodal Gaussian model which predicted large amounts of
negative keystroke intervals which are implausible. Using a log-normal
distribution removed these negative keystroke intervals.

\begin{figure}

{\centering \includegraphics{figures/fitplots}

}

\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
\end{figure}
\end{appendix}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{posterior-parameter-estimates}{%
\subsection{Posterior parameter
estimates}\label{posterior-parameter-estimates}}

\begin{figure}[!htb]
\centering
\includegraphics{figures/psplots1.pdf}
\caption{Posterior parameter distribution}
\end{figure}
\newpage
\begin{figure}[!htb]
\ContinuedFloat
\captionsetup{list=off,format=cont}
\centering
\includegraphics{figures/psplots2.pdf}
\label{fig:fullps1}
\caption{Posterior parameter distribution}
\end{figure}
\end{appendix}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{key-combination-effect-spl2}{%
\subsection{Key-combination effect
(SPL2)}\label{key-combination-effect-spl2}}

The analysed datasets differ to the extent that keystroke intervals at
before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
In other words, the pause before sentences sumed across two key
intervals in the PLanTra and LIFT data, namely
\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
longer or more frequent pauses at before-sentence locations compared to
before-word locations can be explained without reference to linguistic
edges. Also there is a possibility that some inconsistencies in our
findings can be explained on the basis of including the keystroke
following shift.

Therefore we compared whether the different patterns can be explain on
the basis of the additional keystroked involved in before-sentence
transitions. We compared the SPL2 data including and excluding the
keystroke after shift. Although we modelled all transition locations, we
present only before-sentence transitions below as there was, as one
would expect, no difference at word locations. The results of this
comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
fluent transition duration and the hesitation duration were affected by
whether or not the sentence-initial transition include the character
following shift. Fluent key transitions were substantially longer when
including the interval following the shift key. The slowdown for
hesitations was affected too but the difference is numerically small.
There was no conclusive evidence for an increased hesitation
probability.

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions. Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
\toprule
Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  & \\
\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
Hesitation duration &  &  &  & \\
\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
Hesitation probability &  &  &  & \\
\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\end{appendix}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{l2-effect-spl2}{%
\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}

For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
sentence-transitions) we calculated the L2 effect (i.e.~the difference
between writing in L2 and L1). The results can be found in Table
\ref{tab:l2effect}. The results show longer keystrokes and more pauses
across all transition locations. However, the pause duration only
differed at before-word locations, not at within-word or before-sentence
transition locations.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:l2effect}Mixture model estimates for language effect. Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions. The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}

\small{

\begin{tabular}{lrrrr}
\toprule
Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
Fluent transitions &  &  &  & \\
\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
Hesitation duration &  &  &  & \\
\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
Hesitation probability &  &  &  & \\
\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
\bottomrule
\addlinespace
\end{tabular}

}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{tablenotes}

\end{threeparttable}
\end{center}

\end{table}
\end{appendix}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{masking-effect-cato-gunnexp2}{%
\subsection{Masking effect (CATO,
GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}

Studies associated with two datasets (CATO, GUNNEXP2) investigated to
what extent masking the previously written text affects keystroke
behaviour. The mixture model results for the effect of masking is shown
in Table \ref{tab:maskingeffect}. Key transition were longer for masked
text than for unmasked text at before and within word transitions but
not before sentences. Also dyslexics showed no masking difference at
before word transitions. There was no indication of longer or more
frequent pauses as a result of masked text.

\blandscape

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect. Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
\toprule
Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
Hesitation duration &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
Hesitattion probability &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\elandscape
\end{appendix}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{pre-post-test-plantra}{%
\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}

The pre-post test effect for the PLanTra dataset is reported in Table
\ref{tab:retesteffect}. Transition durations were shorter at before-word
locations for the post-test. Evidence was negligible for any other
comparisons.

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect. Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
\toprule
Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  & \\
\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
Hesitation duration &  &  &  & \\
\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
Hesitation probability &  &  &  & \\
\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\end{appendix}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{genre-effect-lift}{%
\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}

From the LIFT data, we compared the difference between genres, i.e.~when
writing an informative text as opposed to writing an argumentative text.
The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
differences were average across writing topic. Evidence for a difference
between genres was negligible.

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:genreeffect}Mixture model estimates for genre effect. Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
\toprule
Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  & \\
\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
Hesitation duration &  &  &  & \\
\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
Hesitation probability &  &  &  & \\
\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\end{appendix}

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{simulation}{%
\subsection{Simulation}\label{simulation}}

A possible concern with these results -- substantially better predictive
performance for bi-modal mixture models -- is that, in principle, as the
mixture model has more parameters it might always lead to a better fit.
We addressed this concern before by using cross-validation techniques
for model comparison which is preventing overfitting models. To address
this concern we repeated a comparison for a uni-modal model and a
bi-modal mixture model for two sets of simulated data that were
simulated either with a uni-modal or bi-modal process at heart. In other
words, this allows us to test the predictive performance of our models
in a context where we know the true underlying data generating process
-- uni-modal vs bi-modal -- and we can test whether these models can
successfully uncover the true parameter value.

In particular, the first data set was simulated using a bi-modal mixture
model with two mixture components similar to the process described above
(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
and the corresponding Bayesian model is summarised in equation
\ref{eq:simmog}.

\begin{equation}
\begin{aligned}
\label{eq:simmog}
\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
        & \sigma_{1}^2 > \sigma_{2}^2
\end{aligned}
\end{equation}

This model is largely identical to the models before but reduced to its
main parameters (but not mixed effects for participants). The model
includes two log-normal distributions with a mixing proportion
\(\theta\) of which the distribution of shorter values has a mean of
\(\beta\) and a standard deviation \(\sigma^2_2\); the second
distribution of longer values is constrained to have a mean that is
larger by a factor of \(\delta\) and has a larger standard deviation
\(\sigma^2_1\).

The second data set was generated with a uni-modal log-Gaussian process.
The model and its corresponding Bayesian model is summarised in equation
\ref{eq:simuv}.

\begin{equation}
\begin{aligned}
\label{eq:simuv}
\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
\text{constraint: } & \sigma^2>0
\end{aligned}
\end{equation}

Again, this model is a simplified version of the uni-modal models used
in the main analysis. The model assumes a log-Gaussian distribution with
a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).

The parameter values used for each of the two data simulations can be
found in Table \ref{tab:simparam}. The simulated data are visualised in
Figure \ref{fig:simdata}. Parameter values were chosen so that the
simulated data are roughly similarly distributed to keystroke
transitions.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}

}

\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
\end{figure}

For each of these two data sets we simulated 1,000 observations. We run
2 models -- a bi-modal mixture model and a uni-modal model -- each for
both data sets. Models were run with 3 chains, with each 6,000
iterations of which 3,000 were warmup. Estimates with 95\% probability
intervals are shown in Table \ref{tab:simparam}. The parameters are
shown by type of data generating process along with the true parameter
values. Parameter value estimates are shown by type of Bayesian model.
The results show that either of the two Bayesian models succesfully
uncovered the model parameters of the data with its corresponding data
generating process, as shown in Table \ref{tab:simparam}, but less so
when the model was applied to data generated with the incorrect
underlying process. Particularly the mixing proportion \(\theta\) and
the slowdown parameter \(\delta\) were not uncovered at all by they
uni-modal model.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}

\begin{tabular}{lrr}
\toprule
 & \multicolumn{2}{c}{Estimate with 95\% PI} \\
\cmidrule(r){2-3}
Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
\midrule
Bi-modal data &  & \\
\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
Uni-modal data &  & \\
\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

We used LOO-CV to compare the predictive performance of the two models
for each data generating process. The model comparisons can be found for
each data generating process in Table \ref{tab:loossim}. The results
rule out the possibility that the mixture model does always lead to
higher predictive performance. Indeed, the mixture model showed a
slightly lower predictive performance for the data that were generated
with a uni-modal process. However, for the data generated with a
bi-modal process, the mixture model shows a substantially higher
predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
and its standard error, as metric for the strength of evidence (Sivula
et al., 2020), shows that the mixture model performs 11.6 times better
than the uni-modal model. In comparison, for the uni-modal data, the
uni-modal model performs only 0.77 times better than the bi-modal
mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
the model is negligible, the uni-modal is preferred by the law of
parsimony.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:loossim}Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.}

\begin{tabular}{lrr}
\toprule
Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
\midrule
Data: Bi-modal mixture process &  & \\
\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
Data: Uni-modal process &  & \\
\ \ \ Uni-modal model & -- & -5,030 (24)\\
\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
\bottomrule
\addlinespace
\end{tabular}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
\end{tablenotes}

\end{threeparttable}
\end{center}

\end{table}
\end{appendix}

\end{document}

### 2023-12-09 report/appendix_fit2data.Rmd
Reorganised manuscript

## Fit to data

Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data. From each model we obtained 100 simulations (predictions). These can be found in Figure \ref{fig:prediction}. The model predictions fitted the data substantially better for the mixture models. The worse fit was observed for the single distribution Gaussian model which predicted large amounts of negative keystroke intervals which are implausible. Using a log-normal distribution removed these negative keystroke intervals.

```{r eval = F}
theme_set(theme_few(base_size = 9) +
              theme(strip.background = element_blank(),
                    legend.position = "top",
                    legend.justification = "right",
                    panel.grid = element_blank(),
                    panel.background = element_rect(fill = "transparent"), # bg of the panel
                    plot.background = element_rect(fill = "transparent", color = NA),
                    axis.text.y = element_blank(),
                    axis.ticks.y = element_blank(),
                    axis.title = element_blank()))

file <- list.files("../stanout", pattern = "sims", full.names = T, recursive = T)
ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
    str_to_upper() %>% 
    str_replace(., "PLANTRA", "PLanTra")
ds <- ds[-6]
plots <- list()
plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
plots_fin <- plots[[1]] / plots[[2]] / plots[[3]] / plots[[4]] / plots[[5]] / plots[[6]] + plot_layout(guides = "collect") & theme(legend.position = 'bottom')

plots_fin <- grid.arrange(patchworkGrob(plots_fin), 
                            left = "Density",
                            bottom = "Transition duration in secs")

ggsave("figures/fitplots.pdf", plots_fin, height = 8.5, width = 6.25)

```

```{r prediction, fig.cap="Fit to data with models (distribution type) above the plots. Comparison of 50 simulations to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8, fig.width=6.5}
knitr::include_graphics("figures/fitplots.pdf")
```

### 2023-12-09 report/appendix_fit2data.tex
Reorganised manuscript

--- 2023-07-24 report/appendix_a.tex
@@
 substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
+for the single distribution Gaussian model which predicted large amounts
+of negative keystroke intervals which are implausible. Using a
+log-normal distribution removed these negative keystroke intervals.
 
@@
 
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
+\caption{Fit to data with models (distribution type) above the plots.
+Comparison of 50 simulations to observed data summarised by model.
+For illustration the x-axis was truncated at 2 secs.}(\#fig:prediction)
 \end{figure}

### 2023-12-09 report/appendix_genre.Rmd
Reorganised manuscript

## Genre effect (LIFT)

From the LIFT data, we compared the difference between genres, i.e. when writing an informative text as opposed to writing an argumentative text. The results are shown in Table \ref{tab:genreeffect}. Cellmeans and differences were average across writing topic. Evidence for a difference between genres was negligible.

```{r genreeffect, results = 'asis'}
cellmeans <- ps_lift %>% 
  separate(group, into = c("topic","genre"), sep = "; ") %>% 
  mutate(across(c(topic, genre), ~str_remove(., "Topic: |Genre: "))) %>% 
  mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
  summarise(across(c(mean, lower, upper), list(mean = mean), 
                   .names = "{.col}"),
            .by = c(location, name, genre)) %>% 
  pivot_wider(names_from = name, values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"), ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, genre), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = genre, values_from = est)

file <- list.files(path = "../stanout/lift", pattern = "mogbetaun.*.csv$",
                   full.names = T)

tmp <- read_csv(file) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  pivot_wider(names_from = genre, values_from = value) %>% 
  unnest(c(ARG, INF)) %>% 
  mutate(diff = ARG - INF) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, ~recode(., theta = "prob"))) %>% 
  left_join(cellmeans) %>% 
  arrange(param, location) %>% 
  select(param, 
         `Transition location` = location,
         `Argumentative` = ARG,
         `Informative` = INF, 
         "Difference" = mean, BF)

write_csv(tmp, "tables/genre_effect_cellmeans.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for genre effect. Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Fluent transitions" = 1:3, 
            "Hesitation duration" = 4:6,
            "Hesitation probability" = 7:9))

```

### 2023-12-09 report/appendix_genre_constr.Rmd
Reorganised manuscript

--- 2023-12-09 report/appendix_genre.Rmd
@@
 Cellmeans and differences were average across writing topic.
-Evidence for a difference between genres was negligible.
+Evidence for a genre comparisons was negligible.
 
@@
 
-file <- list.files(path = "../stanout/lift", pattern = "mogbetaun.*.csv$",
+file <- list.files(path = "../stanout/lift", pattern = "mogbetaco.*.csv$",
                    full.names = T)
@@
   filter(param %in% c("beta", "delta", "theta")) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, value)),
+         across(genre, ~ifelse(is.na(genre) & param != "beta", "ARG", .)),
+         across(genre, ~ifelse(. == "Inf", "INF", .))) %>% 
   pivot_wider(names_from = genre, values_from = value) %>% 
@@
          `Transition location` = location,
-         `Argumentative` = ARG,
-         `Informative` = INF, 
+         `Argumentative` = Arg,
+         `Informative` = `Inf`, 
          "Difference" = mean, BF)
 
-write_csv(tmp, "tables/genre_effect_cellmeans.csv")
+write_csv(tmp, "tables/genre_effect_cellmeans_constr.csv")
 
@@
           caption = "Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for argumentative and informative texts in msecs for the hesitation slowdown and the probability of hesitant transitions.
 The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
@@
           stub_indents = list(
-            "Fluent transitions" = 1:3, 
-            "Hesitation duration" = 4:6,
-            "Hesitation probability" = 7:9))
+            "Hesitation duration" = 1:3,
+            "Hesitation probability" = 4:6))
 

### 2023-12-09 report/appendix_genre_constr.tex
Reorganised manuscript

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
 \clearpage
@@
 The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
+differences were average across writing topic. Evidence for a genre
+comparisons was negligible.
 
@@
 \caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for argumentative and informative texts in msecs for the hesitation slowdown and the probability of hesitant transitions.
 The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
@@
 \endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
 Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
+\ \ \ before sentence & 373 [301, 454] & 411 [330, 504] & -0.07 [-0.33, 0.19] & 0.16\\
+\ \ \ before word & 287 [250, 326] & 284 [248, 323] & 0.01 [-0.15, 0.16] & 0.09\\
+\ \ \ within word & 199 [118, 306] & 246 [155, 361] & -0.12 [-0.6, 0.36] & 0.29\\
 Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
+\ \ \ before sentence & .35 [.28, .43] & .34 [.27, .42] & 0.05 [-0.57, 0.73] & 0.36\\
+\ \ \ before word & .46 [.39, .54] & .48 [.41, .55] & -0.05 [-0.51, 0.44] & 0.25\\
+\ \ \ within word & .07 [.05, .10] & .07 [.05, .10] & -0.02 [-0.81, 0.76] & 0.46\\
 \bottomrule
@@
 \end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}

### 2023-12-09 report/appendix_l2_constr.Rmd
Reorganised manuscript

## L2 effect (SPL2)

For the SPL2 data (only the `_^[shift]` sentence-transitions) we calculated the L2 effect (i.e. the difference between writing in L2 and L1). The results can be found in Table \ref{tab:l2effect}. The results show longer hesitation duration and more pauses across all transition locations when writing in L2. Only within-word transitions did show negligible evidence for an increased hesitation duration in L2.

```{r l2effect, results = 'asis'}
# cell means and data set difference
cellmeans <- cellmeans %>% 
  select(-`SPL2 (_^[shift] + C)`) %>% 
  pivot_wider(names_from = lang, values_from = `SPL2 (_^[shift])`)

file1 <- list.files(path = "../stanout/spl2", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

tmp <- map_dfr(file1, read_csv) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2")) %>% 
  pivot_wider(names_from = lang, values_from = value) %>% 
  unnest(c(L1, L2)) %>% 
  mutate(diff = L2 - L1) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, recode, theta = "prob")) %>% 
  left_join(cellmeans, by = join_by(location, param)) %>% 
  arrange(param, location) %>% 
  select(param, `Transition location` = location, L1, L2,
         "Difference" = mean, BF)

write_csv(tmp, "tables/l2_effect_cellmeans_constr.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cell means are shown the slowdown for hesitant transitions when writing in L1 and L2 and the probability of hesitant transitions. The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          placement = "b",
          stub_indents = list("Hesitation duration" = 1:3,
                              "Hesitation probability" = 4:6) )

```

### 2023-12-09 report/appendix_l2_constr.tex
Reorganised manuscript

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
 \clearpage
@@
 between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
+\ref{tab:l2effect}. The results show longer hesitation duration and more
+pauses across all transition locations when writing in L2. Only
+within-word transitions did show negligible evidence for an increased
+hesitation duration in L2.
 
-\begin{table}[tbp]
+\begin{table}[b]
 
@@
 \caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown the slowdown for hesitant transitions when writing in L1 and L2 and the probability of hesitant transitions.
 The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
@@
 \midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
 Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
+\ \ \ before sentence & 1,281 [1,082, 1,496] & 1,802 [1,557, 2,073] & 0.31 [0.17, 0.45] & > 100\\
+\ \ \ before word & 289 [254, 327] & 560 [507, 617] & 0.47 [0.39, 0.55] & > 100\\
+\ \ \ within word & 120 [53, 222] & 202 [156, 258] & 0.27 [-0.07, 0.55] & 0.68\\
 Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
+\ \ \ before sentence & .78 [.69, .86] & .92 [.87, .96] & 1.23 [0.52, 1.97] & > 100\\
+\ \ \ before word & .51 [.41, .62] & .89 [.83, .93] & 2.02 [1.4, 2.66] & > 100\\
+\ \ \ within word & .04 [.02, .07] & .15 [.10, .22] & 1.52 [0.78, 2.29] & > 100\\
 \bottomrule
@@
 \end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}

### 2023-12-09 report/appendix_masking_constr.Rmd
Reorganised manuscript

--- 2023-07-24 report/appendix_e.Rmd
@@
 The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}.
-Key transition were longer for masked text than for unmasked text at before and within word transitions but not before sentences.
-Also dyslexics showed no masking difference at before word transitions.
-There was no indication of longer or more frequent pauses as a result of masked text.
+There was some evidence for longer before-sentence hesitation durations when the text was masked in non-dyslexic writers.
+Evidence for all other comparisons was negligible.
 
@@
 file <- list.files(path = "../stanout/gunnexp2", 
-                   pattern = "mogbetaun.*.csv$",
+                   pattern = "mogbetaco.*.csv$",
                    full.names = T)
@@
 file <- list.files(path = "../stanout/cato", 
-                   pattern = "mogbetaun.+.csv$",
+                   pattern = "mogbetaco.+.csv$",
                    full.names = T)
@@
          across(ds, ~str_to_upper(.))) %>% 
-  left_join(cellmeans_mask) %>% 
+  left_join(cellmeans_mask, by = join_by(location, param, group, ds)) %>% 
   arrange(param, location, ds, group) %>% 
@@
 
-write_csv(tmp, "tables/masking_effect_cellmeans.csv")
+write_csv(tmp, "tables/masking_effect_cellmeans_constr.csv")
 
@@
           caption = "Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of hesitant transitions.
 The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
@@
           stub_indents = list(
-            "Fluent transitions" = 1:9, 
-            "Hesitation duration" = 10:18,
-            "Hesitattion probability" = 19:27) )
+            "Hesitation duration" = 1:9,
+            "Hesitattion probability" = 10:18) )
 

### 2023-12-09 report/appendix_masking_constr.tex
Reorganised manuscript

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{masking-effect-cato-gunnexp2}{%
\subsection{Masking effect (CATO,
GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}

Studies associated with two datasets (CATO, GUNNEXP2) investigated to
what extent masking the previously written text affects keystroke
behaviour. The mixture model results for the effect of masking is shown
in Table \ref{tab:maskingeffect}. There was some evidence for longer
before-sentence hesitation durations when the text was masked in
non-dyslexic writers. Evidence for all other comparisons was negligible.

\blandscape

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect. Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of hesitant transitions. The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
\toprule
Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Hesitation duration &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & 2,231 [1,543, 3,111] & 1,680 [1,239, 2,205] & -0.25 [-0.64, 0.13] & 0.44\\
\ \ \ before sentence & CATO & non dyslexic & 1,733 [1,241, 2,325] & 1,502 [1,134, 1,942] & -0.12 [-0.47, 0.22] & 0.23\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,124 [954, 1,307] & 1,418 [1,218, 1,634] & 0.21 [0.05, 0.36] & 2.59\\
\ \ \ before word & CATO & dyslexic & 444 [405, 485] & 407 [369, 447] & -0.06 [-0.13, 0.01] & 0.14\\
\ \ \ before word & CATO & non dyslexic & 345 [312, 381] & 326 [295, 361] & -0.04 [-0.11, 0.04] & 0.06\\
\ \ \ before word & GUNNEXP2 & non dyslexic & 363 [325, 403] & 379 [341, 420] & 0.03 [-0.05, 0.11] & 0.05\\
\ \ \ within word & CATO & dyslexic & 278 [166, 400] & 177 [116, 252] & -0.24 [-0.52, 0.06] & 0.57\\
\ \ \ within word & CATO & non dyslexic & 409 [244, 593] & 392 [230, 591] & -0.03 [-0.42, 0.38] & 0.2\\
\ \ \ within word & GUNNEXP2 & non dyslexic & 182 [105, 279] & 209 [135, 306] & 0.08 [-0.24, 0.4] & 0.19\\
Hesitattion probability &  &  &  &  &  & \\
\ \ \ before sentence & CATO & dyslexic & .60 [.38, .79] & .80 [.60, .93] & 0.2 [-0.06, 0.46] & 0.45\\
\ \ \ before sentence & CATO & non dyslexic & .55 [.34, .76] & .68 [.47, .84] & 0.13 [-0.16, 0.4] & 0.21\\
\ \ \ before sentence & GUNNEXP2 & non dyslexic & .88 [.82, .93] & .90 [.85, .94] & 0.16 [-0.46, 0.79] & 0.36\\
\ \ \ before word & CATO & dyslexic & .75 [.64, .85] & .72 [.59, .82] & -0.04 [-0.19, 0.12] & 0.09\\
\ \ \ before word & CATO & non dyslexic & .64 [.51, .75] & .62 [.49, .74] & -0.02 [-0.19, 0.16] & 0.09\\
\ \ \ before word & GUNNEXP2 & non dyslexic & .47 [.40, .54] & .45 [.38, .52] & -0.06 [-0.46, 0.35] & 0.21\\
\ \ \ within word & CATO & dyslexic & .07 [.03, .12] & .11 [.06, .19] & 0.05 [-0.02, 0.13] & 0.08\\
\ \ \ within word & CATO & non dyslexic & .03 [.02, .06] & .03 [.02, .06] & 0 [-0.03, 0.03] & 0.01\\
\ \ \ within word & GUNNEXP2 & non dyslexic & .05 [.03, .08] & .04 [.03, .06] & -0.29 [-0.88, 0.28] & 0.48\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}
\elandscape
\end{appendix}

### 2023-12-09 report/appendix_models.Rmd
Reorganised manuscript

--- 2023-07-24 report/manuscript.Rmd
@@
----
-title             : "Modelling writing hesitations in text writing as finite mixture process"
-shorttitle        : "Modelling writing hesitations"
-
-csl               : "apa.csl"
-
-author: 
-  - name          : "Jens Roeser"
-    affiliation   : "1"
-    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
-    corresponding : yes 
-    email         : "jens.roeser@ntu.ac.uk"
-  - name          : "Mark Torrance"
-    affiliation   : "1"
-  - name          : "Rianne Conijn"
-    affiliation   : "2"
-  - name          : "Evgeny Chukharev-Hudilainen"
-    affiliation   : "3"
-
-affiliation:
-  - id            : "1"
-    institution   : "Department of Psychology, Nottingham Trent University, United Kingdom"
-
-  - id            : "2"
-    institution   : "Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands"
-
-  - id            : "3"
-    institution   : "Department of English, Iowa State University, Iowa"
-
-abstract: |
-
-keywords: "Keystroke modelling; finite mixture models; Bayesian models; text composition"
-
-bibliography      : ["references.bib"]
-
-documentclass     : "apa7"
-classoption       : "man"
-output            : 
-  papaja::apa6_pdf:
-    keep_tex: TRUE
-    includes:
-      after_body: 
-        - "appendix_a.tex"
-        - "appendix_b.tex"
-        - "appendix_c.tex"
-        - "appendix_d.tex"
-        - "appendix_e.tex"
-        - "appendix_g.tex"
-        - "appendix_f.tex"
-
-figsintext        : yes
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-lineno            : no
-mask              : no
-csquotes          : true
-
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{graphicx}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{threeparttable}
-  - \usepackage[normalem]{ulem}
-  - \usepackage[utf8]{inputenc}
-  - \usepackage{icomma}
-  - \usepackage{pdflscape}
-  - \newcommand{\blandscape}{\begin{landscape}}
-  - \newcommand{\elandscape}{\end{landscape}}
-  - \DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, 
-                      warning = FALSE, 
-                      message = FALSE, 
-                      comment = NA,
-                      cache.extra = R.version,
-                      dev = "cairo_pdf",
-                      fig.align = 'center', 
-                      fig.width = 8, 
-                      fig.height = 5,
-                      width=90)
-
-options(kableExtra.auto_format = FALSE)
-dev.args <- list(pdf = list(type = "cairo"))
-```
-
-```{r libs, include=FALSE}
-library(tidyverse)
-library(janitor)
-library(brms)
-library(papaja)
-library(polspline)
-library(patchwork)
-library(gridExtra)
-library(grid)
-library(gtable)
-library(scales)
-library(ggthemes)
-source("../scripts/functions.R")
-source("scripts/get_pdens_plot.R")
-
-theme_set(theme_few(base_size = 12) +
-            theme(strip.background = element_blank(),
-                  legend.position = "top",
-                  legend.justification = "right",
-                  panel.grid = element_blank(),
-                  panel.background = element_rect(fill = "transparent"), # bg of the panel
-                  plot.background = element_rect(fill = "transparent", color = NA)))
-
-options(scipen = 999)
-
-beta_label <- "Short key transition\nduration (in msecs)"
-delta_label <- "Slowdown for hesitant key\ntransitions (in secs)"
-theta_label <- "Probability of hesitant\ntransitions"
-```
-
-# Introduction
-
-# Methods
-
 ## Statistical models
 
-The models presented in the following can be divided into two general groups.
-The first three models are largely akin to models typically used in the literature.
-By this we mean models that assume a uni-modal process that generates keystroke data as is incorporated in statistical models such as analysis of variance and linear mixed-effects models.
-Second, the last two models model keystroke intervals as a combination of two weighted processes of which one presents a smooth information flow from mind into finger; the other component is more important as it represents moments at which the information flow was interrupted leading to longer latencies.
-The latter two models map directly on the idea of a cascading model of writing.
+We are using the Bayesian framework [e.g.
+@farrell2018computational; @gelman2014; @lee2014bayesian] to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
+In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
+The models presented in this section build on one another so that later model include assumptions of earlier models.
 
-### Unimodal Gaussian
+### Serial model of writing
 
-We start with a Gaussian mixed-effects models similar to a standard analysis of variance.
-We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
-The mean can be decomposed into $\beta$ and $\text{participant}_i$.
-$\beta$ will be allowed to take on a different value for each transition location.
-Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.
+#### Single distribution Gaussian
+
+Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
+The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
+For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes.
+There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
+We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average IKI associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the variability associated with factors that we did not further specify in the model.
+This can be expressed as $\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)$.
+Of importance is the estimated posterior value of $\mu$ as this value captures that time it takes to mentally plan a word.
+
+We can extend this simple model of word-planning to other linguistic location.
+We introduce earlier that larger linguistic edges are associated with planning on higher levels.
+For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07].
+Regression formulas can, and commonly are, used to decompose $\mu$ and capture that there is a change in the outcome variable associated with another factors.
+We can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is the average IKI for word boundaries.
+However when $\text{x}_\text{sentence}$ takes on the value 1, the average IKI for word boundaries $\alpha$ is incremented by a sentence-related slowdown of $\beta$ msecs.
+Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning.
+The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.
+whether there is a statistically meaningful difference between IKI associated with words and sentences).
+
+For computational ease, we implemented the differences associated with transition locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
+Therefore, the model will return one $\beta$ per transition location that capture the posterior distribution of average IKIs.
+Also the decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts $u_\text{participant}$.
+The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$.
+This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.
+a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
+
+The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
 
@@
 
-### Unimodal log-Gaussian
+Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
 
-This model is largely identical to the previous model, except we use a log-normal distribution instead of a normal distribution.
-The advantage of using a log-normal distribution is two-fold: (1) the log-normal distribution has a lower bound of zero.
-For our data we generally consider negative ikis as mistakes which occasionally occur when the next key was pressed before the current key.
-Other than that keystroke intervals are constrained by a persons ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour, in particular motor responses.
-In particular, a normal distribution assumes that units are linearly scaled.
-For example, a 50 msecs difference is the same between 100 msecs and 150 msecs than it is between 5 secs and 5 secs 50 msecs (i.e.
-5,050 msecs).
-This is not necessarily plausible for keystroke data.
-We would assume that differences that are due to motor activity (typing a high frequency bigram vs a low frequency bigram) are smaller than difference that are due to high levels of cognitive activity (retrieving a word in your L1 or L2).
-Log-normal distributions are a natural way of scaling units so that a 50 msecs difference on the lower end of the iki scale (motor activity) is more meaningful than a 50 msecs difference on the upper end of the iki scale (retrieving words, planning sentences).
+#### Single distribution log-Gaussian
 
-The model can be described like this:
+The previous model assumes that the data-generating process is a Gaussian distribution.
+The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
+There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
+IKIs, as the distance between two subsequent key-down events, must be positive.
+The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
+(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
+In particular, a Gaussian distribution assumes that units are scaled linearly.
+For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
+This does not necessarily map onto the psychological interpretation for short and long keyintervals.
+For example effects that result on the motor level within words (e.g.
+typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
+In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
+Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
+
+The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{N}()$ was replaced by $\text{logN}()$.
 
@@
 
-### Unimodal unequal-variance log-Gaussian
+#### Single distribution unequal-variance log-Gaussian
+
+The third model representing the serial view of writing is a single distribution unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
+The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
+The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
+This assumption is however not in line with what we know about data from human behaviour.
+Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing].
+For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
+Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation $\sigma_{e_\text{location}}^2)$ varies by transition location.
 
@@
 
-### Bimodal log-Gaussian (constrained)
+### Parallel cascading model of writing
 
-This models extends the intuition from the previous model that higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that underlies the generation of ikis, we assume there are two.
-(1) activation can flow into keystrokes without interrupts.
-These fluent keystroke transitions are merely limited by a person's ability to move their finger and will be captured but the parameter $\beta$.
-In principle, there are no differences for fluent key-transitions between transition location.
-The next model with loosen this assumption (hence, "unconstrained model").
-(2) difficulty in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when spelling, words, or contents couldn't be retrieved in time.
-The size of these pauses will depend on the reason for delays which is typically associated with transition locations (contents are typically planned before sentences, words are retrieved before they are typed and spelling difficulty typically occurs when typing a word).
-Pauses will be captured by two model parameters: (1) the slowdown for these hesitant transitions will be captured by $\delta$ which is the deviation compared to normal typing intervals (constrained to be positive).
-(2) the frequency of hesitant transitions will be captured by $\theta$ for each level of a categorical predictor.
+The following two models are are extensions of the models introduced for the serial view.
+Crucially, the cascading view allows planning to happen in parallel to production.
+Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
+This is done by assuming that IKIs come from a weighted combination of two distributions.
+
+#### Two-distributions log-Gaussian mixture model (constrained)
+
+This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
+Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
+This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+
+1.
+Activation can flow into keystrokes without interruption.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter in equation \ref{eq:bimodcon}.
+Note that $\beta$ is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.
+
+2.
+Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
+The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}.
+The magnitude of the slowdown is associated with transition location.
+This is because delays at larger linguistic units are likely to be associated with higher level planning.
+By constraining $\delta$ to be positive, it captures how much longer hesitant IKIs in addition to $\beta$.
 
@@
 (\#eq:bimodcon)
-\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2
+		& \sigma_{e'}^2 > \sigma_{e}^2\\
+		& 0 < \theta < 1
 \end{aligned}
@@
 
-```{r}
-#		<!-- &\delta \sim \text{N}(0,1)\\ -->
+The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
+Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
+$\theta$ is here parameterised to represent the probability that an IKI is associated with the distribution of long IKIs.
+This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$ .
+In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
+The weights of both distributions must sum to 1.
+We will call this parameter the probability of hesitant transitions.
 
-```
+The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
+In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
+As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
+This is akin to what is known as a random by-participant slopes model.
 
-This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
+Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location.
+In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].
+This was achieved by introducing a parameter $\sigma_\text{diff}$.
+The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
 
-### Bimodal log-Gaussian (unconstrained)
+#### Two-distributions log-Gaussian mixture model (unconstrained)
 
-This model is identical to the previous model with one exception.
-The distribution of fluent keystroke transitions captured by $\beta$ was fixed to be the same across transition locations is the previous model.
-In other words, the mean $\beta$ and it's standard deviation $\sigma_{e}^2$ was the same for before-sentence, before-word, and within-word transitions.
-This means, because of the naturally larger number of within-word transitions, the posterior is dominated by within-word transitions.
+The size of a fluent key-transition does not necessarily vary by transition location.
+In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions.
+This is what we called a constrained model.
+However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
+This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter.
+Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
+We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
 
-In this model we will loosen this constraint and allow $\beta$ and $\sigma_{e}^2$ to vary by transition-location.
+In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
 
@@
 (\#eq:bimoduncon)
-\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-	\text{where: }  & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+	\text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2
+		& \sigma_{e'}^2 > \sigma_{e}^2\\
+		& 0 < \theta < 1
 \end{aligned}
 \end{equation}
-
-```{r}
-#<!-- & \delta \sim \text{N}(0,1)\\ -->
-
-```
-
-## Data sets
-
-Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \ref{tab:datasets}.
-
-\blandscape
-```{r datasets, results='asis'}
-tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
-                Source = c("Rønneberg et al. (2022)",
-                           "Torrance et al. (2016)",
-                           "Torrance et al. (n.d.)",
-                           "Rossetti and Van Waes (2022)",
-                           "Vandermeulen et al. (2020)", 
-                           "Torrance and Ofstad (n.d)"),
-                Keylogger = c("EyeWrite", 
-                           "EyeWrite", 
-                           "CyWrite", 
-                           "InputLog", 
-                           "InputLog",
-                           "EyeWrite"),
-                `Writing task` = c("Argumentative essays", 
-                                 "Expository texts",
-                                 "Argumentative essays",
-                                 "Text simplification",
-                                 "Synthesis", 
-                                 ""),
-                 `N (ppts)` = as.character(c(126, 26*2, 39, 47, 658, 45)),
-#                n_texts = c(1, 2, 2, 2, NA),
-#                n_sentences = c(),
-#                n_words = c(),
-                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
-                 Language = c( "Norwegian", 
-                               "Norwegian", 
-                               "English (L1) / Spanish (L2)", 
-                               "English (L2)", 
-                               "Dutch",
-                               "Norwegian"),
-                Manipulation = c("--", 
-                               "weak decoders / control; masked / unmasked",
-                               "write in L1 / L2",
-                               "pre / post test trained in plain language principles and control", 
-                               "Various topics and genres",
-                               "masked / unmasked")) %>% 
-  arrange(Dataset) %>% 
-  apa_table(align = c(rep("l", 4), rep("r", 2), "p{2cm}", "p{3cm}"), 
-            escape = FALSE, 
-            font_size = "footnotesize",
-            caption = "Datasets in brief.")
-
-```
-\elandscape
-
-### GUNNEXP2
-
-TODO: Mark, can you add information here? I've added a provisional bib entry that needs changing or we just say "unpublished"
-
-The GUNNEXP2 dataset published in @torranceb includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote text either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### C2L1
-
-TODO: Mark, can you add information here?
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)?
-
-The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process.
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### CATO
-
-TODO: Mark, can you add information here?
-
-Data are published in @torrance2016adolescent.
-Norwegian upper secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### PLanTra
-
-TODO: Rianne, can you add information here?
-
-The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
-In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
-Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
-During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
-making it easier to read for a lay audience--by applying what they had learned from their respective training.
-The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
-It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
-Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### LIFT
-
-TODO: Rianne, can you add information here?
-
-The LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) data are published in @vandermeulen2020.
-Students produced text for 4 topics in two different writing genres (levels: argumentative text, informative text).
-Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### SPL2
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in @torrancea.
-The data come from a textcomposition task of undergraduate university students -- *N* = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
-CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
-We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
-Participants were given a 40 minute time limit.
-They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-## Data analysis
-
-### Bayesian modelling
-
-We reanalysed keystroke data from 6 datasets in Bayesian implementations of 5 statistical models. 
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values: Bayesian parameter estimates are expresses the uncertainty associated with parameter values as probability distribution [@farrell2018computational; @gelman2014; @lee2014bayesian].
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
-existing knowledge about parameter values, for regulating the parameter space.
-For small data sets priors influence the inferred parameter estimates (known as posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
-automatic Ockham's razor, @jefferys1992ockham].
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
-@lambert2018student; @mcelreath2016statistical].
-The predictive performance (i.e.
-fit) of these models is compared in the Results section using leave-one-out cross-validation.
-
-Keystroke data from text composition tasks were modelled in a series of five models. An overview of all models can be found in Table \ref{tab:models}.
-
-```{r models, results = 'asis'}
-models <- tibble(Models = paste0("M",1:5),
-       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
-       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
-       Description = c("Gaussian distribution",
-                       "Log-normal distribution",
-                       "Log-normal distribution and unequal variances",
-                       "Fluent transition are not allowed to vary by transition location",
-                       "Fluent transition are allowed to vary by transition location"))
-
-apa_table(models, 
-          align = c(rep("l", 2), "r", "p{9cm}"), 
-          escape = FALSE, 
-          digits = 0,
-          caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
-          note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
-```
-
-For all models we modelled transition locations (levels: before sentence, before word, within word) was included as predictor in all models.
-Also, we included the dataset specific experimental conditions as fixed effects and included random intercepts for participants as described in the equations above.
-We used normal and log-normal probability function in line with typically treatments fouund in the literature.
-Stan code for mixture models was based on @roeser2021modelling  [see also @vasishth2017; @vasishth2017feature].
-
-Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
-The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
-
-### Transition types
-
-The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g.
-@torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
-Keytransitions that terminated in an editing operation were excluded from the analysis.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions.
-
-```{r keyloc, results='asis'}
-
-table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
-       Description = c("Transitions between any letter",
-                       "Keypress after space followed by any letter",
-                       "Keypress following a space preceding any letter"),
-       Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d.
-T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.",
-                   "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
-                   "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept."))
-
-apa_table(table, caption = "Transition location classification.",
-            align = c("p{3cm}", "p{3cm}", "p{8cm}"), 
-            escape = FALSE, 
-            font_size = "footnotesize",
-            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace. At before-sentence locations, IKIs were timed to the shift keypress.")
-
-```
-
-### Data reduction
-
-```{r reductionfunctions}
-source("scripts/get_data_summary.R")
-
-# Get info about data reduction
-n_samples <- 100
-c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
-cato <- get_cato("../data/cato.csv", n_samples = n_samples)
-lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
-plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
-spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
-gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
-```
-
-For all datasets we only used transitions that were not followed by an editing operation.
-
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant)
-
-We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-
-From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis.
-Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-```{r datareduction, results='asis'}
-c2l1$ds <- "c2l1"
-cato$ds <- "cato"
-lift$ds <- "lift"
-plantra$ds <- "plantra"
-spl2$ds <- "spl2"
-gunnexp2$ds <- "gunnexp2"
-
-extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
-          mutate(ds = .x$ds)) %>% 
-  mutate(across(where(is.numeric), ~round(.*100, 2))) %>% 
-  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
-  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
-  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
-  rename(`$\\le$ 50 msecs` = `<50ms`,
-         `$\\ge$ 30 secs` = `>30,000ms`)
-
-random_sample <- 
-  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
-          mutate(ds = .x$ds)) %>% 
-    mutate(across(where(is.numeric), ~round(.*100, 1))) %>%
-    unite("keep", starts_with("keep"), sep = " (") %>% 
-    mutate(across(c(keep), ~str_c(., ")"))) %>%
-  pivot_wider(names_from = location, values_from = keep)
-
-datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
-  relocate(ds) %>% 
-  mutate(across(ds, ~recode(., 
-                "cato" = "CATO",
-                "spl2" = "SPL2", 
-                "spl2_shift" = "SPL2 (shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2"))) %>% 
-  rename(`Dataset` = ds)
-
-apa_table(datareduction, 
-          escape = FALSE,  
-          placement = "bp!",
-          caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location. Standard error is shown in parentheses.",
-          align =c("l", rep("r", 5)),
-          col_spanners = list(" " = 1,
-                              "Extreme values in \\%" = c(2, 3),
-                              "Randomly sampled data in \\%" = c(4, 6)))
-
-```
-
-# Results
-
-## Out-of-samples cross-validation
-
-To compare the out-of-sample predictive performance of the models we used Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
-Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and its  difference $\Delta\widehat{elpd}$ between models.
-The advantage of using leave-one-out cross-validation is that models with more parameters are penalised to prevent overfit.
-
-Results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-The mixture of log-normal distribution models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across transition locations.
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-
-```{r}
-files <- list.files(
-  str_c("../stanout/", 
-  c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
-  pattern = "model.+.csv", 
-  full.names = T)
-
-table <- purrr::map_dfr(files, ~read_csv(.) %>% 
-                          mutate(dataset = .x)) %>% 
-  select(dataset, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
-  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(where(is.numeric), ~round(., 0)),
-         across(where(is.numeric), ~format(., big.mark = ",")),
-         across(everything(), ~str_trim(.))) %>%
-  filter(dataset != "spl2_shift") %>% 
-  transmute(
-     across(dataset, ~recode(., 
-                "cato" = "CATO",
-                "spl2" = "SPL2", 
-#                "spl2_shift" = "SPL2 (shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2")),
-    m = recode(model, 
-               lmmgaus = "M1",
-               lmm = "M2",
-               lmmuneqvar = "M3",
-               mogbetaconstr = "M4",
-               mogbetacontr = "M4",
-               mogbetaunconstr = "M5"),
-    across(model, ~recode(., lmm = "Unimodal log-normal",
-                          lmmgaus = "Unimodal normal",
-                          mogbetaconstr = "Bimodal (constrained)",
-                          mogbetacontr = "Bimodal (constrained)",
-                          mogbetaunconstr = "Bimodal (unconstrained)",
-                          lmmuneqvar = "Unimodal (unequal variance)")),
-            across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-            across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")"))) %>% 
-  mutate(across(elpd_diff, ~str_replace(., "0 \\(0\\)", "--"))) %>% 
-  arrange(dataset)
-```
-
-```{r loos, results='asis'}
-names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
-
-apa_table(table[,-1], 
-      caption = "Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-      align =c("p{3.5cm}", "p{6cm}", rep("r", 2)), 
-      escape = FALSE,
-      longtable = T,
-      font_size = "small",
-      note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      stub_indents = list(
-                     "\\textbf{CL21}" = 1:5,
-                     "\\textbf{CATO}" = 6:10,
-                     "\\textbf{GUNNEXP2}" = 11:15, 
-                     "\\textbf{LIFT}" = 16:20,
-                     "\\textbf{PLanTra}" = 21:25,
-                     "\\textbf{SPL2}" = 26:30))
-#                     "\\textbf{SPL2 (shift + C)}" = 31:35))
-
-```
-
-## Cross-data set comparisons
-
-```{r}
-files <- list.files(str_c("../stanout/", 
-                          c("lift", "spl2", "plantra", "cato", "c2l1", "gunnexp2")), 
-                    pattern = "mogbetaun.+.csv", 
-                    full.names = T)
-
-ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
-  mutate(data = .x)) %>% 
-  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, ~recode_factor(., 
-                "cato" = "CATO (non-dyslexic\nunmasked)",
-                "gunnexp2" = "GUNNEXP2 (unmasked)",
-                "spl2" = "SPL2 (L1)", 
-#                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                .ordered = TRUE))) %>% 
-  filter(!(str_detect(data, "SPL2") & lang == "ES"),
-         !(str_detect(data, "CATO") & group == "dyslexic"),
-         !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked")) %>% 
-  pivot_wider(names_from = param, 
-              values_from = value) %>% 
-  unnest(cols = beta:theta) %>% 
-  mutate(across(c(beta, beta2), ~exp(.)),
-         delta = beta2 - beta,
-         across(location, ~str_replace(., " ", "\n"))) %>% 
-  pivot_longer(beta:theta) %>% 
-  summarise(across(value, 
-                   list(mean = mean, 
-                        lower = lower, 
-                        upper = upper),
-                   .names = "{.fn}"), 
-            .by = c(data, location, name)) %>% 
-  filter(name %in% c("beta", "delta", "prob"))
-```
-
-As detailed above the mixture model captures the writing process in three conceptually important parameter values: (1) the transition duration associated with fluent writing (which we indicated as $\beta$), (2) the size of the slowdown for hesitant transitions (indicated as $\delta$), and (3) the mixing proportion which represents the probability of hesitant transition durations (indicated as $\theta$).
-
-We present the mixture-model estimates for each of the three parameter values in Figure \ref{fig:crossstudypost}.
-Although models were fitted with all dataset-specific condition, we aggregated across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence for differences between these conditions], and removed conditions that might conflate comparisons^[We removed masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO dataset, and L2 writing in the SPL2 data.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that evidence for differences in parameter values for the involved manipulations.].
-For posteriors of all conditions within dataset see Appendix \ref{posterior-parameter-estimates}.
-The posterior of the remaining conditions allows us to compare differences between transition locations within dataset associated with each of the three mixture-model parameters.
-
-```{r crossstudypost, fig.cap="Estimates across studies. Posterior parameter distribution represented as estimate and 95\\% probability interval (PI)."}
-
-posd <- position_dodge(.65)
-dotsize <- 2.5
-grouplabel <- "Data set"
-shapes <- c(1, 4, 6, 7, 8, 9, 11)
-width <- 0
-linewidth <- .75
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(data))) +
-  geom_line(linewidth = linewidth, 
-            alpha = .35, 
-            position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  theme(axis.title = element_blank()) +
-  labs(colour = grouplabel,
-       shape = grouplabel) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label,
-         across(c(mean, lower, upper), ~./1000)) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(data))) +
-  geom_line(linewidth = linewidth, alpha = .35, position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
-  labs(colour = grouplabel,
-       shape = grouplabel) +
-  theme(axis.title = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(data))) +
-  geom_line(linewidth = linewidth, alpha = .35, position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(colour = grouplabel,
-       shape = grouplabel) +
-  theme(axis.title = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
-
-plot <- beta + delta + theta +
-  plot_layout(guides = "collect")
-
-grid.arrange(patchworkGrob(plot), 
-             left = "Posterior estimate with 95% PIs",
-             bottom = "Transition location")
-
-```
-
-Figure \ref{fig:crossstudypost} shows largely the same patterns for keystroke interval estimates by transition location across datasets for all three mixture model parameters.
-keystroke transitions were substantially longer for the C2L1 data presumably reflecting that the sample of in this dataset was the youngest.
-Pause durations and frequencies were relatively similar across datasets.
-
-In the following we will evaluate differences between transition locations for all three mixture model parameters.
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence $>$ before-word $>$ within-word.
-We evaluated the differences between transition locations for all data sets.
-The results are shown in Table \ref{tab:loceffect}.
-
-Comparing transition locations rendered patterns that are largely consistent across data sets (with caveats).
-We found that hesitations appear more frequently at before-word transitions than within words across datasets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except dataset C2L1) compared to within-word transitions (except dataset LIFT).
-However, for most dataset there is negligble evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
-In fact, the dataset LIFT showed less pausing before sentences compared to before words.
-Also, we observe that even fluent key-transitions are slower at before-word locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for datasets SPL2 and GUNNEXP2).
-Hesitation durations tend to be longer at before-sentence locations compared to before-word locations (except for datasets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT dataset).
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-The inconsistencies for before-sentence transitions could, to some extent, be explained on the basis that datasets differ as to whether before-sentence transitions should be analysed as complex key combination.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-Notice though that for the datasets PLanTra and LIFT, there were no consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-We test to what extent this difference may affect the modelling results for the SPL2 dataset.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this is in conflict with the absence of differences in datasets that included the character following shift at before-sentence transitions.
-In other words, it is unlikely that patterns in our results can be explained on the basis of complex key combinations at before-sentence locations.
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\blandscape
-```{r loceffect, results='asis'}
-files <- list.files(str_c("../stanout/", 
-c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetaun.+.csv", full.names = T) # "spl2_shift",
-
-ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
-  mutate(data = .x)) %>% 
-  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, ~recode(., 
-                "cato" = "*CATO* (non-dyslexic\nunmasked)",
-                "spl2" = "*SPL2* (L1)", 
-                "gunnexp2" = "*GUNNEXP2* (unmasked)",
-#                "spl2_shift" = "*SPL2* (L1; shift + C)", 
-                "plantra" = "*PLanTra*",
-                "lift" = "*LIFT*",
-                "c2l1" = "*C2L1*"))) %>% 
-  filter(!(str_detect(data, "SPL2") & lang == "ES"),
-         !(str_detect(data, "CATO") & group == "dyslexic"),
-         !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked"),
-         param %in% c("beta", "delta", "theta")) %>%
-  select(param, location, data, value) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
-  pivot_wider(names_from = location, values_from = value) %>%
-  unnest(-c(data, param)) %>% 
-  mutate(diff.1 = `before sentence` - `before word`,
-         diff.2 = `before word` - `within word`) %>% 
-  summarise(across(starts_with("diff"), 
-                   list(mean = mean, 
-                        lower = lower, 
-                        upper = upper, 
-                        BF = BF)),
-            .by = c(data, param))
-
-tmp <- ps_loc_diffs %>% 
-  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
-  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
-         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
-  select(-lower, -upper) %>% 
-  pivot_wider(names_from = param, values_from = c(mean, BF)) %>% 
-  mutate(across(diffid, ~recode(., diff.1 = "before sentence vs word",
-                                   diff.2 = "before vs within word"))) %>% 
-  select(data, diffid, ends_with("beta"), ends_with("delta"), ends_with("theta")) %>% 
-  arrange(data)
-
-write_csv(tmp, "tables/location_effect_alldata.csv")
-
-apa_table(tmp[,-1],
-          align =c("l", rep("r", 6)), 
-          longtable = T,
-          escape = FALSE,
-          row.names = T,
-          font_size = "footnotesize",
-          col.names = c("Comparison", rep(c("Est. [95\\% PIs]", "BF"), 3)),
-          stub_indents = list(`\\textbf{C2L1}` = 1:2, 
-                              `\\textbf{CATO (non-dyslexic, unmasked)}`= 3:4,
-                              `\\textbf{GUNNEXP2 (unmasked)}` = 5:6,
-                              `\\textbf{LIFT}` = 7:8,
-                              `\\textbf{PLanTra}` = 9:10,
-                              `\\textbf{SPL2 (L1)}` = 11:12),
-#                              `\\textbf{SPL2 (L1; shift + C)}` = 13:14),
-          col_spanners = list(" " = c(1),
-                     "Fluent transitions" = c(2, 3), 
-                     "Hesitation slowdown" = c(4, 5), 
-                     "Hesitation probability" = c(6, 7)),
-          note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
-         caption = "Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\\% PIs in brackets.")
-
-```
-\elandscape
-
-## Simulation
-
-A possible concern with the results -- substantially better predictive performance for mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-
-To address this concern we repeated a comparison for an unimodal model and a mixture model for two sets of simulated data. For each dataset we simulated 1,000 observations.
-
-These two datasets differ with regards to their underlying data generating process.
-The first set of data has as underlying data generating process a mixture model with two mixture components similar to the process described above.
-This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:simmog)
-\text{y}_i \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-		& \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical with the models before but reduced to its main parameters.
-There are two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.
-
-The second dataset was generated with an unimodal model as data generating process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:simuv)
-\text{y}_i \sim\text{ }& \text{logN}(\beta, \sigma^2)
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the unimodal models used in the main analysis.
-The model assumes a log-normal distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
-
-The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
-The simulated data are visualised in Figure \ref{fig:simdata}.
-The data are simulated to be similarly distributed to keystroke transitions.
-
-```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=3.5, fig.width=5}
-files <- list.files("../mixture-model-sim/data", full.names = T)
-
-map_dfr(files, ~read_csv(.) %>% 
-        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
-  mutate(across(Dataset, ~recode(., mogdata = "bimodal", lmdata = "unimodal"))) %>% 
-  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
-  geom_density(alpha = .25) +
-  scale_colour_colorblind() + 
-  scale_fill_colorblind() +
-  scale_x_continuous(labels = ~format(., big.mark = ",")) +
-  labs(colour = "Data process", 
-       fill = "Data process", 
-       x = "Simulated data", 
-       y = "Density")
-```
-
-We run 4 models, 2 on each of the two datasets: 2 mixture models, one for the data generated with a bimodal mixture process and one for the data generated with the unimodal process.
-We repeated the same using an unimodal model.
-Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
-Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}.
-The parameters are shown by data generating process along with with true parameter values.
-Parameter estimates are shown for the dataset generated with a unimodal process and the dataest generated with a bimodal process.
-The Bayesian models uncovered the model parameters of the dataset with the matching data generating process successfully, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-
-```{r}
-mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
-  mutate(data = "mog")
-mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
-  mutate(data = "mog")
-
-mog_sim <- bind_rows(mog1, mog2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>%
-  select(-lo, -up) %>% 
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = c(5, 1, .35, .25, .5),
-         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(name, ~str_replace(., "prob", "theta"))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
-
-lm_sim <- bind_rows(lm1, lm2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>% 
-  select(-lo, -up) %>%
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = as.character(c(5, .25))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
-
-```
-
-```{r simparam, results='asis'}
-table <- bind_rows(mog_sim %>% mutate(model = "mog"),
-                   lm_sim %>% mutate(model = "lm")) %>% 
-  relocate(model) %>% 
-  mutate(across(name, ~str_c("$", ., "$")))
-
-names(table) <- c("Model", "Parameter with true value", "Bimodal data", "Unimodal data")
-
-apa_table(table[,-1], 
-          escape = FALSE,
-      col_spanners = list("Estimate with 95\\% PI" = 2:3),
-      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
-      align = c("l", rep("r", 2)),
-      stub_indents = list(`Bimodal mixture process` = 1:5, 
-                          `Unimodal process` = 6:7))
-
-```
-
-We used LOO-CV to compare the fit of the two models for each data set.
-The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
-The results rules out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a unimodal process.
-However, for the data generated with a bimodal process, the mixture model shows a substantially higher predictive performance.
-
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence [@sivula2020uncertainty], shows that the mixture model performs 11.6 times better than the unimodal model for the data generated with a bimodal process.
-In comparison, for the unimodal data, the unimodal model performs only 0.77 times better than the bimodal mixture model.
-Thus, even though the mixture model does not necessarily perform better for non-bimodal data, it also does not necessarily perform worse.
-
-```{r loossim, results='asis'}
-table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
-  select(1:6) %>% 
-  mutate(
-#    elpd_se_ratio = abs(elpd_diff / se_diff),
-    across(ends_with("diff"), ~round(., 1)),
-    across(ends_with("loo"), ~round(., 0)),
-    across(where(is.numeric), ~format(., big.mark = ",")),
-    across(everything(), ~as.character(.)),
-    across(everything(), ~str_trim(.)),
-    across(everything(), ~str_replace_all(., "NaN", "--")),
-    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
-    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
-    across(c(data, model), ~recode(., mog = "Bimodal mixture model",
-                                      lm = "Unimodal model"))) %>% 
-  select(data, model, starts_with("elpd"))
-
-names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
-
-apa_table(table[,-1], 
-          escape = FALSE,
-      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      align =c("l", rep("r", 2)),
-      stub_indents = list(`Data: Bimodal mixture process` = 1:2, 
-                          `Data: Unimodal process` = 3:4))
-
-```
-
-# Discussion
-
-# References
-```{r create_r-references, echo=FALSE, include=FALSE}
-r_refs(file = "references.bib")
-```
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-<div id = "references"></div>
-\endgroup
-
-```{r render_appendix, include=FALSE}
-render_appendix("appendix_a.Rmd")
-render_appendix("appendix_b.Rmd")
-render_appendix("appendix_c.Rmd")
-render_appendix("appendix_d.Rmd")
-render_appendix("appendix_e.Rmd")
-render_appendix("appendix_g.Rmd")
-render_appendix("appendix_f.Rmd")
-```

### 2023-12-09 report/appendix_models.tex
Reorganised manuscript

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
+\clearpage
 \makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
+\efloat@restorefloats
 \makeatother
 
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
+\begin{appendix}
+\section{}
 \hypertarget{statistical-models}{%
@@
 
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
+We are using the Bayesian framework (e.g. Farrell \& Lewandowsky, 2018;
+Gelman et al., 2014; Lee \& Wagenmakers, 2014) to implement 5
+statistical models of writing and to evaluate which -- the serial or the
+cascading view -- captures more successfully the data that arise during
+keyboard typing. In other words, we use statistical models to map
+between the data and the theoretically assumed process that generates
+the data to then compare the predictive power of those models. The
+models presented in this section build on one another so that later
+model include assumptions of earlier models.
 
@@
 
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
+\hypertarget{single-distribution-gaussian}{%
+\paragraph{Single distribution
+Gaussian}\label{single-distribution-gaussian}}
 
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
+Under the serial view, all planning must be completed priors to the
+production onset of the corresponding planning unit. The resulting IKI
+is sometimes faster or slower depending on, among others,
+psycholinguistic factors. For example, the interval before a word is
+shorter for easily retrievable high-frequency words, or longer for low
+frequency words, shorter for words with fewer graphemes, syllables, and
+morphemes. There are word-specific factors that influence the IKI that
+precedes a word but these are beyond the scope of our analysis. We will
+capture the variability associated with word-features by assuming that
+before-word IKIs can be described as coming from a distribution that is
+normal (Gaussian) with two parameters, an unknown mean \(\mu\), that
+describes the average IKI associated with word-level planning, and a
+standard deviation \(\sigma_\text{e}^2\), that captures the variability
+associated with factors that we did not further specify in the model.
+This can be expressed as
+\(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
+Of importance is the estimated posterior value of \(\mu\) as this value
+captures that time it takes to mentally plan a word.
 
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
+We can extend this simple model of word-planning to other linguistic
+location. We introduce earlier that larger linguistic edges are
+associated with planning on higher levels. For example, at sentence
+boundaries, planning needs to happen for word-level properties -- which
+was captured above as average IKI \(\mu\) -- but also higher linguistic
+planning such as clause-level meaning, and dependencies of the
+sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
+Regression formulas can, and commonly are, used to decompose \(\mu\) and
+capture that there is a change in the outcome variable associated with
+another factors. We can decompose \(\mu\) as
+\(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that
+when \(\text{x}_\text{sentence}\) takes on the value 0, the equation
+reduces to \(\mu = \alpha\) which is the average IKI for word
+boundaries. However when \(\text{x}_\text{sentence}\) takes on the value
+1, the average IKI for word boundaries \(\alpha\) is incremented by a
+sentence-related slowdown of \(\beta\) msecs. Therefore the value of the
+\(\beta\) parameter represents the additional cognitive demand
+associated with sentence-initial planning. The application of such a
+statistical model to the data will then provide us with an estimate of
+the parameter value that can be used for statistical inference
+(e.g.~whether there is a statistically meaningful difference between IKI
+associated with words and sentences).
 
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
+For computational ease, we implemented the differences associated with
+transition locations as \(\beta_\text{location}\) in equation
+\ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for
+each transition location. Therefore, the model will return one \(\beta\)
+per transition location that capture the posterior distribution of
+average IKIs. Also the decomposition of \(\mu\) allows us to address the
+fact that some writers are faster than other writers by introducing a
+parameter for what is typically called random intercepts
+\(u_\text{participant}\). The random intercepts term
+\(u_\text{participant}\) is constrained so that the value it takes on
+comes from a normal distribution with a mean of 0 and a standard
+deviation of \(\sigma_\text{p}^2\). This value is therefore the
+difference between the overall posterior estimate and estimate average
+IKI of a particular participant (i.e.~a positive value indicates that a
+participant is slower than average, a negative value indicates that a
+participant is faster than average).
 
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
+The final model is shown in equation \ref{eq:unimodgaus} and represents
+a Gaussian mixed effects model.
 
@@
 \begin{aligned}
-\label{eq:unimodgaus}
+(\#eq:unimodgaus)
 \text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
@@
 
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
+Note aside that standard-deviation parameters were constrained to be
+positive because standard deviations can never be negative.
 
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
+\hypertarget{single-distribution-log-gaussian}{%
+\paragraph{Single distribution
+log-Gaussian}\label{single-distribution-log-gaussian}}
 
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
+The previous model assumes that the data-generating process is a
+Gaussian distribution. The next model is largely identical to the
+previous model but instead of assuming a Gaussian, we assume that the
+data come from a log-normal (log-Gaussian) distribution. There are, at
+least, two arguments for using a log-Gaussian distribution: (1)
+log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian
+does not allow negative values. IKIs, as the distance between two
+subsequent key-down events, must be positive. The lower bound is
+delimited by a person's ability to move their fingers and keyboard
+polling. (2) the log-scale is known to be a better match for data from
+human behaviour and motor responses. In particular, a Gaussian
+distribution assumes that units are scaled linearly. For example, a
+difference of 25 msecs is the same between 100 and 125 msecs as between
+5 secs and 5,025 msecs. This does not necessarily map onto the
+psychological interpretation for short and long keyintervals. For
+example effects that result on the motor level within words (e.g.~typing
+a high vs a low frequency bigram) are smaller than differences that are
+due to high levels of processing (retrieving a word in an L1 or L2). In
+other words, although an effect of 25 msecs is large in the context
+overall fast keyintervals, it is small in the context of overall slow
+intervals. Log-Gaussian distributions are a natural way of translating a
+linear scale to an exponential scale so that a 25 msecs difference on
+the lower end of the IKI scale (motor activity) is more meaningful than
+a 25 msecs difference on the upper end of the IKI scale (retrieving
+words, planning sentences).
 
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
+The model can be described as in equation \ref{eq:unimodloggaus} in
+which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
 
@@
 \begin{aligned}
-\label{eq:unimodloggaus}
+(\#eq:unimodloggaus)
 \text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
@@
 
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
+\hypertarget{single-distribution-unequal-variance-log-gaussian}{%
+\paragraph{Single distribution unequal-variance
+log-Gaussian}\label{single-distribution-unequal-variance-log-gaussian}}
 
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
+The third model representing the serial view of writing is a single
+distribution unequal-variance model that, except for the
+unequal-variance assumption is identical to the model presented in the
+previous section. The previous models modelled IKIs as a function of
+transition location so that the estimated average IKI depends on the
+position of an IKI in the text. The variance associated with the
+estimated IKIs for each transition location was assumed to be identical
+(equal variance). This assumption is however not in line with what we
+know about data from human behaviour. Longer latencies are known to be
+associated with a larger variances for both response-time data in
+particular (Wagenmakers \& Brown, 2007) and human motor behaviour in
+general (Schöner, 2002; Wing \& Kristofferson, 1973). For IKIs pauses at
+larger linguistic edges are plausibly associated with a larger variance
+because of the larger number of associated processes. Therefore, in
+equation \ref{eq:unimoduv}, we introduce the assumption that standard
+deviation \(\sigma_{e_\text{location}}^2)\) varies by transition
+location.
 
@@
 \begin{aligned}
-\label{eq:unimoduv}
+(\#eq:unimoduv)
 \text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
@@
 \hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
+\subsubsection{Parallel cascading model of
+writing}\label{parallel-cascading-model-of-writing}}
 
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
+The following two models are are extensions of the models introduced for
+the serial view. Crucially, the cascading view allows planning to happen
+in parallel to production. Therefore, we will reduce the constrain of
+the serial models that requires all planning to be completed before
+writing onset. This is done by assuming that IKIs come from a weighted
+combination of two distributions.
 
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
+\hypertarget{two-distributions-log-gaussian-mixture-model-constrained}{%
+\paragraph{Two-distributions log-Gaussian mixture model
+(constrained)}\label{two-distributions-log-gaussian-mixture-model-constrained}}
 
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+This model extends the assumption of the previous model that processing
+that involves higher levels of activation lead to longer pauses. Instead
+of assuming that there is one process that is shifted for IKIs of larger
+linguistic edges, we introduce the assumption that pauses at larger
+linguistic edges are more likely but not obligatory. This is achieved by
+modelling IKIs as coming from a weighted mixture of two distributions
+associated with two different states:
 
@@
 \item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
+  Activation can flow into keystrokes without interruption. These fluent
+  keystroke transitions are merely constrained by a person's ability to
+  move their finger and will be captured by the \(\beta\) parameter in
+  equation \ref{eq:bimodcon}. Note that \(\beta\) is represented in both
+  log-Gaussian distributions in equation \ref{eq:bimodcon} referring to
+  the same unknown parameter.
 \item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
+  Interruptions in the activation flow from higher to lower levels
+  result in longer keystroke intervals, when words or their spelling
+  could not be retrieved in time. The slowdown for these hesitant
+  transitions is captured by \(\delta\) in the first line of equation
+  \ref{eq:bimodcon}. The magnitude of the slowdown is associated with
+  transition location. This is because delays at larger linguistic units
+  are likely to be associated with higher level planning. By
+  constraining \(\delta\) to be positive, it captures how much longer
+  hesitant IKIs in addition to \(\beta\).
 \end{enumerate}
@@
 \begin{aligned}
-\label{eq:bimodcon}
+(\#eq:bimodcon)
 \text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
@@
         & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
+        & 0 < \theta < 1
 \end{aligned}
@@
 
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
+The first line of equation \ref{eq:bimodcon} represents the distribution
+of hesitant key transitions, and the second line represents fluent key
+transitions. Each of these two distributions is associated with the
+mixing weight \(\theta\) which is a proportion that is constrained to be
+larger than 0 and smaller than 1. \(\theta\) is here parameterised to
+represent the probability that an IKI is associated with the
+distribution of long IKIs. This probability is inversely related to the
+mixing weight of the distribution of short IKIs by \(1-\theta\) . In
+other words, a larger weight of one distribution inevitably means a
+lower weight for the other distribution. The weights of both
+distributions must sum to 1. We will call this parameter the probability
+of hesitant transitions.
 
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
+The probability of hesitant transitions is assumed to vary as a function
+of both transition location and participants. In line with the
+literature discussed in the introduction, we assume that pauses are more
+likely at larger linguistic edges. As pausing is subject to individual
+differences and writing style (skills), we also assume that some
+participants pause more at certain transition locations and other
+participants pause less. This is akin to what is known as a random
+by-participant slopes model.
 
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
+Lastly, we carried over the unequal variance assumption and let the
+standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by
+transition location. In addition we constrained the variances so that
+\(\sigma_{e'}\) associated with the distribution of typing disfluencies
+is larger than the variance associated with fluent transitions
+\(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et
+al., 2017). This was achieved by introducing a parameter
+\(\sigma_\text{diff}\). The consequence is that fluent transitions are
+assumed to come from a narrower distribution than hesitant transitions.
 
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
+\hypertarget{two-distributions-log-gaussian-mixture-model-unconstrained}{%
+\paragraph{Two-distributions log-Gaussian mixture model
+(unconstrained)}\label{two-distributions-log-gaussian-mixture-model-unconstrained}}
 
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
+The size of a fluent key-transition does not necessarily vary by
+transition location. In other words, the parameter \(\beta\) is the same
+for before-sentence, before-word, and within-word transitions. This is
+what we called a constrained model. However, letter bigrams and trigrams
+may be systematically executed faster than transitions between between
+space and a letter (REFERENCE?) or complex keystrokes that comprise
+space and shift-letter combinations for upper case characters before
+sentences (we will address the latter possibility in the results
+section). This is because bigrams / trigrams might be stored, retrieved
+and executed as motor codes but not transitions to a space following a
+word or between a space or shift key press preceding a letter. Also,
+because of the necessarily larger number of within-word transitions, as
+opposed to before-word and sentence transitions, the posterior of the
+constrained model is dominated by within-word transition data. We
+therefore also implemented an unconstrained model that assumes that the
+size of fluent transitions varies across transition locations.
 
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
+In this model we assume that \(\beta\) varies by transition-location as
+illustrated in equation \ref{eq:bimoduncon}.
 
@@
 \begin{aligned}
-\label{eq:bimoduncon}
+(\#eq:bimoduncon)
 \text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
@@
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
+        & \sigma_{e'}^2 > \sigma_{e}^2\\
+        & 0 < \theta < 1
 \end{aligned}
@@
 
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
 \hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
 \leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
+Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of
+cognition and behavior}. Cambridge University Press.
 
 \leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
+Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \&
+Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman;
+Hall/CRC.
 
 \leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
+Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive
+modeling: A practical course}. Cambridge University Press.
 
 \leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
+Nottbusch, G., Weingarten, R., \& Sahel, S. (2007). From written word to
+written sentence production. In M. Torrance, L. van Waes, \& D. W.
+Galbraith (Eds.), \emph{Writing and cognition: Research and
+applications} (Vol. 20, pp. 31--53). Elsevier.
 
 \leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
+Roeser, J., Torrance, M., \& Baguley, T. (2019). Advance planning in
+written and spoken sentence production. \emph{Journal of Experimental
+Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
 \url{https://doi.org/10.1037/xlm0000685}
 
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
 \leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
+Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain
+and Cognition}, \emph{48}(1), 31--51.
 
 \leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
+Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B. (2017). Modelling
+dependency completion in sentence comprehension as a {B}ayesian
+hierarchical mixture process: {A} case study involving {C}hinese
+relative clauses. \emph{ArXiv e-Prints}.
 
 \leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
+Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting
+as a finite mixture process: Evidence from comprehension data.
+\emph{arXiv Preprint arXiv:1703.04081}.
 
 \leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
+Wagenmakers, E.-J., \& Brown, S. (2007). On the linear relation between
+the mean and the standard deviation of a response time distribution.
 \emph{Psychological Review}, \emph{114}(3), 830--841.
@@
 
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
 \leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
+Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the
+timing of discrete motor responses. \emph{Perception \& Psychophysics},
+\emph{14}(1), 5--12.
 
 \end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
 \end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}

### 2023-12-09 report/appendix_postparam_constr.Rmd
Reorganised manuscript

## Posterior parameter estimates

```{r}
# get posterior
file <- list.files(path = "../stanout/cato", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n())

ps_cato <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, group, location, name)) %>% 
  mutate(group = str_c("Task: ", task, "; Group: ", group),
         ds = "CATO") %>% 
  select(-task)

file <- list.files(path = "../stanout/c2l1", 
                   pattern = "mogbetacon.*.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n())

ps_c2l1 <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta,by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, name)) %>% 
  mutate(ds = "C2L1",
         group = "1")

file <- list.files(path = "../stanout/gunnexp2", 
                   pattern = "mogbetacon.*.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n())

ps_gunn <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, xn, name)) %>% 
  mutate(ds = "GUNNEXP2") %>% 
  rename(group = xn)

file <- list.files(path = "../stanout/lift", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n())

ps_lift <- ps %>% filter(param != "beta") %>% 
  mutate(across(genre, ~ifelse(is.na(genre), "Arg", .))) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(genre, topic, location, name)) %>% 
  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre),
         ds = "LIFT") %>% 
  select(-topic, -genre)

file <- list.files(path = "../stanout/plantra", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n())

ps_plantra <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, location, name)) %>% 
  mutate(ds = "PLanTra") %>% 
  rename(group = task)

file1 <- list.files(path = "../stanout/spl2", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../stanout/spl2_shift", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
                     mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
  filter(param %in% c("beta", "delta", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value, data) %>% 
  mutate(idx = 1:n(), .by = data)

ps_spl2_both <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(delta, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx, data)) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
         beta2 = beta + delta,
         across(c(beta2, beta), ~exp(.)),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper),
                   .names = "{.fn}"),
            .by = c(lang, location, name, data)) %>% 
  mutate(ds = "SPL2",
         group = str_c("Data: ", data, "; Language: ", lang)) %>% 
  select(-lang)

ps_spl2 <- ps_spl2_both %>% filter(data == "SPL2 (_^[shift])")

# Combine all
ps <- bind_rows(ps_cato, ps_c2l1, ps_gunn, ps_lift, ps_plantra, ps_spl2)
```

```{r eval = F}

theme_set(theme_few(base_size = 11) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))

pdodge <- position_dodge(-.5)
beta_plots <- list()
delta_plots <- list()
theta_plots <- list()
datasets <- unique(ps$ds)
size <- 2
width <- 0
linewidth <- .35
beta_label <- "Fluent transitions\n(in msecs)"

beta_plots <- map(datasets, ~filter(ps, 
                                    name == "beta", 
                                    location == location[1],
                                    ds == .x) %>% 
                    mutate(name = beta_label,
                           group == "overall",
                           location = "overall") %>% 
                    unique() %>% 
                    ggplot(aes(y = mean, 
                               ymin = lower, 
                               ymax = upper,
                               x = location)) +
                    geom_errorbar(width = width,
                                  linewidth = linewidth,
                                  alpha = .75,
                                  position = pdodge) +
                    geom_point(size = size,
                               position = pdodge) +
                    facet_grid(~name) +
                    scale_colour_colorblind() +
                    scale_y_continuous(labels = scales::comma) +
                    labs(subtitle = .x) +
                    theme(axis.title = element_blank(),
                          legend.title = element_blank()) +
                    guides(colour=guide_legend(nrow=1,byrow=TRUE),
                           shape = guide_legend(nrow=1,byrow=TRUE)))

delta_plots <- map(datasets, ~filter(ps, 
                                     name == "delta", 
                                     ds == .x) %>% 
                     mutate(across(c(ds,group), ~factor(.)),
                            across(c(mean, lower, upper), ~ . /1000),
                            name = delta_label) %>% 
                     ggplot(aes(y = mean, 
                                ymin = lower, 
                                ymax = upper,
                                x = location,
                                shape = group)) +
                     geom_errorbar(width = width, 
                                   alpha = .75, 
                                   linewidth = linewidth,
                                   position = pdodge) +
                     geom_point(aes(colour = group), 
                                size = size, 
                                position = pdodge) +
                     facet_grid(~name) +
                     scale_shape_manual(values = 1:length(filter(ps, ds == .x) %>% 
                                                            pull(group) %>% 
                                                            factor())) +
                     scale_colour_colorblind() +
                     scale_y_log10(labels = scales::comma) +
                     theme(axis.title = element_blank(),
                           legend.title = element_blank()) +
                     guides(colour=guide_legend(nrow=2,byrow=TRUE),
                            shape = guide_legend(nrow=2,byrow=TRUE)))

theta_plots <- map(datasets, ~filter(ps, 
                                     name == "prob", 
                                     ds == .x) %>% 
                     mutate(across(c(ds,group), ~factor(.)),
                            name = theta_label) %>% 
                     ggplot(aes(y = mean, 
                                ymin = lower, 
                                ymax = upper,
                                x = location,
                                shape = group)) +
                     geom_errorbar(width = width,
                                   linewidth = linewidth,
                                   alpha = .75, 
                                   position = pdodge) +
                     geom_point(aes(colour = group), 
                                size = size, 
                                position = pdodge) +
                     facet_grid(~name) +
                     scale_shape_manual(values = 1:length(filter(ps, ds == .x) %>% 
                                                            pull(group) %>% 
                                                            factor())) +
                     scale_colour_colorblind() +
                     scale_y_continuous(labels = dezero_plot, 
                                        limits = c(0, 1)) +
                     theme(axis.title = element_blank(),
                           legend.title = element_blank()) +
                     guides(colour=guide_legend(nrow=2,byrow=TRUE),
                            shape = guide_legend(nrow=2,byrow=TRUE)))

# Combine parameter plots 
plot_param <- list()
plot_param <- map(1:6, ~beta_plots[[.]] + delta_plots[[.]] + theta_plots[[.]] +
                    plot_layout(guides = "collect", widths = c(.45, 1, 1)) &
                    theme(legend.position = "bottom"))

plot_param[[2]] <- plot_param[[2]] & theme(legend.position = "none")
plot_param[[3]] <- plot_param[[3]] & guides(colour=guide_legend(nrow=1,byrow=TRUE),
                                            shape = guide_legend(nrow=1,byrow=TRUE))

plot_param[[5]] <- plot_param[[5]] & guides(colour=guide_legend(nrow=1,
                                                                byrow=TRUE,
                                                                reverse=TRUE),
                                            shape = guide_legend(nrow=1,
                                                                 byrow=TRUE,
                                                                 reverse=TRUE))

plot_fin_1 <- plot_param[[1]] / plot_param[[2]] / plot_param[[3]]
plot_fin_1 <- grid.arrange(patchworkGrob(plot_fin_1), 
                           left = "Posterior estimate with 95% PIs",
                           bottom = "Transition location")
ggsave("figures/psplots1constr.pdf", plot_fin_1, height = 8, width = 7)

plot_fin_2 <- plot_param[[4]] / plot_param[[5]] / plot_param[[6]]
plot_fin_2 <- grid.arrange(patchworkGrob(plot_fin_2), 
                           left = "Posterior estimate with 95% PIs",
                           bottom = "Transition location")
ggsave("figures/psplots2constr.pdf", plot_fin_2, height = 8, width = 7)
```

\begin{figure}[!htb]
\centering
\includegraphics{figures/psplots1constr.pdf}
\caption{Posterior parameter distribution}
\end{figure}
\newpage
\begin{figure}[!htb]
\ContinuedFloat
\captionsetup{list=off,format=cont}
\centering
\includegraphics{figures/psplots2constr.pdf}
\label{fig:fullps1}
\caption{Posterior parameter distribution}
\end{figure}

### 2023-12-09 report/appendix_prepost_constr.Rmd
Reorganised manuscript

--- 2023-07-24 report/appendix_g.Rmd
@@
 
-The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}.
-Transition durations were shorter at before-word locations for the post-test.
-Evidence was negligible for any other comparisons.
+The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}. Evidence for all comparisons was negligible.
 
@@
 file <- list.files(path = "../stanout/plantra", 
-                   pattern = "mogbetaun.*.csv$",
+                   pattern = "mogbetaco.*.csv$",
                    full.names = T)
@@
 
-write_csv(tmp, "tables/posttest_effect_cellmeans.csv")
+write_csv(tmp, "tables/posttest_effect_cellmeans_constr.csv")
 
@@
           caption = "Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant transitions and the probability of hesitant transitions.
 The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.

### 2023-12-09 report/appendix_prepost_constr.tex
Reorganised manuscript

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
 \clearpage
@@
 The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
+\ref{tab:retesteffect}. Evidence for all comparisons was negligible.
 
@@
 \caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant transitions and the probability of hesitant transitions.
 The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
@@
 \endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
 Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
+\ \ \ before sentence & 879 [723, 1,052] & 909 [740, 1,100] & -0.03 [-0.23, 0.17] & 0.1\\
+\ \ \ before word & 438 [395, 485] & 434 [389, 483] & 0.01 [-0.08, 0.1] & 0.05\\
+\ \ \ within word & 245 [180, 323] & 269 [201, 353] & -0.06 [-0.3, 0.18] & 0.14\\
 Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
+\ \ \ before sentence & .58 [.50, .66] & .54 [.47, .62] & 0.15 [-0.26, 0.58] & 0.27\\
+\ \ \ before word & .56 [.50, .61] & .49 [.43, .55] & 0.27 [-0.04, 0.59] & 0.72\\
+\ \ \ within word & .11 [.08, .14] & .09 [.07, .12] & 0.13 [-0.27, 0.54] & 0.26\\
 \bottomrule
@@
 \end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}

### 2023-12-09 report/appendix_shift.Rmd
Reorganised manuscript

--- 2023-07-24 report/appendix_c.Rmd
@@
-## Key-combination effect (SPL2)
+## Key-combination effect (unconstrained mixture model)
 
@@
 
-Therefore we compared whether the different patterns can be explain on the basis of the additional keystroked involved in before-sentence transitions.
+Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
 We compared the SPL2 data including and excluding the keystroke after shift.
 Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
-The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.
-Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift.
-Fluent key transitions were substantially longer when including the interval following the shift key.
-The slowdown for hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation probability.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans2}.
 
-```{r shiftcellmeans, results = 'asis'}
+```{r shiftcellmeans2, results = 'asis'}
 # cell means and data set difference
@@
 ```
+
+Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift.
+Fluent key transitions were substantially longer when including the interval following the shift key.
+The slowdown for hesitations was affected too but the difference is numerically small.
+There was no conclusive evidence for an increased hesitation probability.

### 2023-12-09 report/appendix_shift.tex
Reorganised manuscript

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
 \clearpage
@@
 \section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
+\hypertarget{key-combination-effect-unconstrained-mixture-model}{%
+\subsection{Key-combination effect (unconstrained mixture
+model)}\label{key-combination-effect-unconstrained-mixture-model}}
 
@@
 Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
+the basis of the additional keystroke involved in before-sentence
 transitions. We compared the SPL2 data including and excluding the
@@
 would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
+comparison can be found in Table \ref{tab:shiftcellmeans2}.
 
@@
 
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
+\begin{longtable}{lrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:shiftcellmeans2}Mixture model estimates for key transitions.
 Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
@@
 \toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
 \midrule
 \endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
+\caption*{\normalfont{Table \ref{tab:shiftcellmeans2} continued}}\\
 \toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
 \midrule
 \endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
+Fluent transitions &  &  & \\
+\ \ \ L1 & 159 [147, 172] & 0.48 [0.34, 0.63] & > 100\\
+\ \ \ L2 & 159 [147, 172] & 0.41 [0.19, 0.63] & 46.08\\
+Hesitation duration &  &  & \\
+\ \ \ L1 & 1,281 [1,082, 1,496] & -0.46 [-0.61, -0.3] & > 100\\
+\ \ \ L2 & 1,802 [1,557, 2,073] & -0.34 [-0.51, -0.17] & > 100\\
+Hesitation probability &  &  & \\
+\ \ \ L1 & .78 [.69, .86] & 0.49 [-0.04, 1.03] & 1.43\\
+\ \ \ L2 & .92 [.87, .96] & 0.48 [-0.16, 1.13] & 0.94\\
 \bottomrule
@@
 \end{center}
+
+Overall, fluent transition duration and the hesitation duration were
+affected by whether or not the sentence-initial transition include the
+character following shift. Fluent key transitions were substantially
+longer when including the interval following the shift key. The slowdown
+for hesitations was affected too but the difference is numerically
+small. There was no conclusive evidence for an increased hesitation
+probability.
 \end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}

### 2023-12-09 report/appendix_shift_constr.Rmd
Reorganised manuscript

## Key-combination effect (constrained mixture model)

The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key. In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets. Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.

<!-- Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.  -->

Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions. We compared the SPL2 data including and excluding the keystroke after shift. Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations. The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.

```{r shiftcellmeans, results = 'asis'}
# cell means and data set difference
cellmeans <- ps_spl2_both %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  select(-ds, -group) %>%
  mutate(across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(lang, location, data), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est) %>% 
  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
  unique()

file1 <- list.files(path = "../stanout/spl2", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../stanout/spl2_shift", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

tmp <- map_dfr(c(file1, file2), ~read_csv(.x) %>% 
                 mutate(data = .x)) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
  pivot_wider(names_from = data, values_from = value) %>% 
  unnest(c(spl2, spl2_shift)) %>% 
  mutate(diff = spl2_shift - spl2) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(lang, location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, recode, theta = "prob")) %>% 
  left_join(cellmeans) %>% 
  arrange(param, location, lang) %>% 
  filter(location %in% c("overall", "before sentence")) %>% 
  select(param, 
         Language = lang,
         `_^[shift] + C` = ends_with(" + C)"),
         `_^[shift]` = ends_with("FT])"),
         "Difference" = mean, BF)

write_csv(tmp, "tables/shift_effect_cellmeans_constr.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for key transitions immediately preceding a sentence. Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions. The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "footnotesize",
          longtable = T,
          stub_indents = list("Fluent transitions" = 1,
                              "Hesitation duration" = 2:3,
                              "Hesitation probability" = 4:5) )

```

The fluent transition duration and hesitation duration were not affected by whether or not the sentence-initial transition include the character following shift for neither language. However, we found strong evidence for an increased hesitation probability -- in both -- languages when the before-sentence key transition included the character following the shift key. Notably, the hesitation probability increased ceiling. It is not entirely surprised that the constrained mixture model identifies essentially all before-sentence transition as pauses: the constrained model does not distinguish between transition locations for the duration of fluent transitions. Therefore essentially all before-sentence transitions are identified as hesitation as they always include two keystrokes while all other transition location (which is the majority of the data) do not.

### 2023-12-09 report/appendix_shift_constr.tex
Reorganised manuscript

--- 2023-12-09 report/appendix_shift_constr.Rmd
@@
-## Key-combination effect (constrained mixture model)
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
 
-The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets.
-Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
+\begin{appendix}
+\section{}
+\hypertarget{key-combination-effect-constrained-mixture-model}{%
+\subsection{Key-combination effect (constrained mixture
+model)}\label{key-combination-effect-constrained-mixture-model}}
 
-<!-- Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.  -->
+The analysed datasets differ to the extent that keystroke intervals at
+before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
+C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+In other words, the pause before sentences summed across two key
+intervals in the PLanTra and LIFT data, namely
+\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
+namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
+longer or more frequent pauses at before-sentence locations compared to
+before-word locations can be explained without reference to linguistic
+edges.
 
-Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
-We compared the SPL2 data including and excluding the keystroke after shift.
-Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
-The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.
+Therefore we compared whether the different patterns can be explain on
+the basis of the additional keystroke involved in before-sentence
+transitions. We compared the SPL2 data including and excluding the
+keystroke after shift. Although we modelled all transition locations, we
+present only before-sentence transitions below as there was, as one
+would expect, no difference at word locations. The results of this
+comparison can be found in Table \ref{tab:shiftcellmeans}.
 
-```{r shiftcellmeans, results = 'asis'}
-# cell means and data set difference
-cellmeans <- ps_spl2_both %>% 
-  separate(group, into = c("group", "lang"), sep = "; ") %>%
-  select(-ds, -group) %>%
-  mutate(across(lang, ~str_remove(., "Language: ")),
-         across(location, ~str_replace(., "\\n", " "))) %>% 
-  pivot_wider(names_from = name, 
-              values_from = c(mean, lower, upper)) %>% 
-  mutate(across(ends_with("prob"), ~dezero(., 2)),
-         across(c(ends_with("beta"),
-                  ends_with("delta")), 
-                ~as.character(scales::comma(round(.))))) %>% 
-  pivot_longer(-c(lang, location, data), 
-               names_to = c(".value", "param"),
-               names_sep = "_") %>% 
-  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
-  select(-mean:-upper) %>% 
-  pivot_wider(names_from = data, 
-              values_from = est) %>% 
-  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
-  unique()
+\begin{center}
+\begin{ThreePartTable}
 
-file1 <- list.files(path = "../stanout/spl2", 
-                    pattern = "mogbetacon.+.csv$",
-                    full.names = T)
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
 
-file2 <- list.files(path = "../stanout/spl2_shift", 
-                    pattern = "mogbetacon.+.csv$",
-                    full.names = T)
+\footnotesize{
 
-tmp <- map_dfr(c(file1, file2), ~read_csv(.x) %>% 
-                 mutate(data = .x)) %>% 
-  filter(param %in% c("beta", "delta", "theta")) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
-  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
-         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
-  pivot_wider(names_from = data, values_from = value) %>% 
-  unnest(c(spl2, spl2_shift)) %>% 
-  mutate(diff = spl2_shift - spl2) %>% 
-  summarise(across(diff, list(mean = mean, 
-                              lower = lower, 
-                              upper = upper,
-                              BF = BF), 
-                   .names = "{.fn}"),
-            .by = c(lang, location, param)) %>% 
-  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
-         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
-         across(param, recode, theta = "prob")) %>% 
-  left_join(cellmeans) %>% 
-  arrange(param, location, lang) %>% 
-  filter(location %in% c("overall", "before sentence")) %>% 
-  select(param, 
-         Language = lang,
-         `_^[shift] + C` = ends_with(" + C)"),
-         `_^[shift]` = ends_with("FT])"),
-         "Difference" = mean, BF)
-
-write_csv(tmp, "tables/shift_effect_cellmeans_constr.csv")
-
-apa_table(tmp[,-1],
-          align = c("l", "r", "r", "r", "r"), 
-          caption = "Mixture model estimates for key transitions immediately preceding a sentence.
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions immediately preceding a sentence.
 Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions.
 The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.",
-          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
-          font_size = "footnotesize",
-          longtable = T,
-          stub_indents = list("Fluent transitions" = 1,
-                              "Hesitation duration" = 2:3,
-                              "Hesitation probability" = 4:5) )
+95\% PIs in brackets.}\\
+\toprule
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
+\toprule
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Fluent transitions &  &  &  & \\
+\ \ \ overall & 156 [145, 169] & 159 [147, 172] & -0.02 [-0.13, 0.09] & 0.06\\
+Hesitation duration &  &  &  & \\
+\ \ \ L1 & 1,167 [1,037, 1,305] & 1,281 [1,082, 1,496] & -0.06 [-0.21, 0.09] & 0.11\\
+\ \ \ L2 & 2,019 [1,772, 2,292] & 1,802 [1,557, 2,073] & 0.12 [-0.02, 0.27] & 0.29\\
+Hesitation probability &  &  &  & \\
+\ \ \ L1 & 1.00 [.99, 1.00] & .78 [.69, .86] & 4.91 [3.68, 6.35] & > 100\\
+\ \ \ L2 & 1.00 [.99, 1.00] & .92 [.87, .96] & 3.4 [2.12, 4.85] & > 100\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
 
-```
+}
 
-The fluent transition duration and hesitation duration were not affected by whether or not the sentence-initial transition include the character following shift for neither language.
-However, we found strong evidence for an increased hesitation probability -- in both -- languages when the before-sentence key transition included the character following the shift key.
-Notably, the hesitation probability increased ceiling.
-It is not entirely surprised that the constrained mixture model identifies essentially all before-sentence transition as pauses: the constrained model does not distinguish between transition locations for the duration of fluent transitions.
-Therefore essentially all before-sentence transitions are identified as hesitation as they always include two keystrokes while all other transition location (which is the majority of the data) do not.
+\end{ThreePartTable}
+\end{center}
+
+The fluent transition duration and hesitation duration were not affected
+by whether or not the sentence-initial transition include the character
+following shift for neither language. However, we found strong evidence
+for an increased hesitation probability -- in both -- languages when the
+before-sentence key transition included the character following the
+shift key. Notably, the hesitation probability increased ceiling. It is
+not entirely surprised that the constrained mixture model identifies
+essentially all before-sentence transition as pauses: the constrained
+model does not distinguish between transition locations for the duration
+of fluent transitions. Therefore essentially all before-sentence
+transitions are identified as hesitation as they always include two
+keystrokes while all other transition location (which is the majority of
+the data) do not.
+\end{appendix}

### 2023-12-09 report/appendix_sim.Rmd
Reorganised manuscript

--- 2023-09-15 report/appendix_h.Rmd
@@
 
-A possible concern with these results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+A possible concern with these results -- substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
 We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-To address this concern we repeated a comparison for a uni-modal model and a bi-modal mixture model for two sets of simulated data that were simulated either with a uni-modal or bi-modal process at heart.
-In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
+To address this concern we repeated a comparison for a single distribution model and a two-distrobutionsl mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart.
+In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.
 
-In particular, the first data set was simulated using a bi-modal mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}).
+In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}).
 This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
@@
 
-The second data set was generated with a uni-modal log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
+The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
 
@@
 
-Again, this model is a simplified version of the uni-modal models used in the main analysis.
+Again, this model is a simplified version of the single distribution models used in the main analysis.
 The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
@@
 
-```{r simdata, fig.cap= "Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process.
+```{r simdata, fig.cap= "Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process.
 The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
@@
         mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
-  mutate(across(Dataset, ~recode(., mogdata = "bi-modal", lmdata = "uni-modal"))) %>% 
+  mutate(across(Dataset, ~recode(., mogdata = "two distributions", lmdata = "single distributions"))) %>% 
   ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
-  geom_density(alpha = .25) +
+  geom_density(alpha = .25, 
+               linewidth = .25) +
   scale_colour_colorblind() + 
@@
 For each of these two data sets we simulated 1,000 observations.
-We run 2 models -- a bi-modal mixture model and a uni-modal model -- each for both data sets.
+We run 2 models -- a two-distribution mixture model and a single-distribution model -- each for both data sets.
 Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
@@
 The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they uni-modal model.
+Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they single-distribution model.
 
@@
 
-names(table) <- c("Model", "Parameter with true value", "Bi-modal model", "Uni-modal model")
+names(table) <- c("Model", "Parameter with true value", "Two-distributions model", "Single-distributions model")
 
@@
       align = c("l", rep("r", 2)),
-      stub_indents = list(`Bi-modal data` = 1:5, 
-                          `Uni-modal data` = 6:7))
+      stub_indents = list(`Two-distributions data` = 1:5, 
+                          `Single-distribution data` = 6:7))
 
@@
 The results rule out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a uni-modal process.
-However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance.
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence (Sivula et al., 2020), shows that the mixture model performs 11.6 times better than the uni-modal model.
-In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model.
-Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the uni-modal is preferred by the law of parsimony.
+Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a single-distribution process.
+However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
+In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error -- a *z*-score approximation -- reveals a value of 11.6 which is strong evidence in in favour of the mixture model compared to the single-distribution model (Sivula et al., 2020).
+In comparison, for the single-distribution data, the single-distribution model was negligibly better than the two-distribution mixture model which was evidence by a value of 0.77 for the ratio of $\Delta\widehat{elpd}$ and its standard error.
+Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the single distribution is preferred by the law of parsimony.
 
@@
     across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
-    across(c(data, model), ~recode(., mog = "Bi-modal mixture model",
-                                      lm = "Uni-modal model"))) %>% 
+    across(c(data, model), ~recode(., mog = "Two-distribution mixture model",
+                                      lm = "Single-distribution model"))) %>% 
   select(data, model, starts_with("elpd"))
@@
       align =c("l", rep("r", 2)),
-      stub_indents = list(`Data: Bi-modal mixture process` = 1:2, 
-                          `Data: Uni-modal process` = 3:4))
+      stub_indents = list(`Data: Two-distribution mixture process` = 1:2, 
+                          `Data: Single-distribution process` = 3:4))
 

### 2023-12-09 report/appendix_sim.tex
Reorganised manuscript

--- 2023-09-15 report/appendix_h.tex
@@
 A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
+performance for two-distribution mixture models -- is that, in
+principle, as the mixture model has more parameters it might always lead
+to a better fit. We addressed this concern before by using
+cross-validation techniques for model comparison which is preventing
+overfitting models. To address this concern we repeated a comparison for
+a single distribution model and a two-distrobutionsl mixture model for
+two sets of simulated data that were simulated either with a single
+distribution or two-distributional process at heart. In other words,
+this allows us to test the predictive performance of our models in a
+context where we know the true underlying data generating process --
+single distribution vs two distributions -- and we can test whether
+these models can successfully uncover the true parameter value.
 
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
+In particular, the first data set was simulated using a
+two-distributions mixture model with two mixture components similar to
+the process described above (equations in \ref{eq:bimodcon} and
+\ref{eq:bimoduncon}). This process and the corresponding Bayesian model
+is summarised in equation \ref{eq:simmog}.
 
@@
 
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
+The second data set was generated with a single distribution
+log-Gaussian process. The model and its corresponding Bayesian model is
+summarised in equation \ref{eq:simuv}.
 
@@
 
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
+Again, this model is a simplified version of the single distribution
+models used in the main analysis. The model assumes a log-Gaussian
+distribution with a mean parameter \(\beta\) and a standard deviation
+\(\sigma^2\).
 
@@
 
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}(\#fig:simdata)
+\caption{Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process.
+The x-axis showing the outcome y was log-scaled for visability.}(\#fig:simdata)
 \end{figure}
@@
 For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
+2 models -- a two-distribution mixture model and a single-distribution
+model -- each for both data sets. Models were run with 3 chains, with
+each 6,000 iterations of which 3,000 were warmup. Estimates with 95\%
+probability intervals are shown in Table \ref{tab:simparam}. The
+parameters are shown by type of data generating process along with the
+true parameter values. Parameter value estimates are shown by type of
+Bayesian model. The results show that either of the two Bayesian models
+succesfully uncovered the model parameters of the data with its
+corresponding data generating process, as shown in Table
+\ref{tab:simparam}, but less so when the model was applied to data
+generated with the incorrect underlying process. Particularly the mixing
+proportion \(\theta\) and the slowdown parameter \(\delta\) were not
+uncovered at all by they single-distribution model.
 
@@
 \cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
+Parameter with true value & \multicolumn{1}{c}{Two-distributions model} & \multicolumn{1}{c}{Single-distributions model}\\
 \midrule
-Bi-modal data &  & \\
+Two-distributions data &  & \\
 \ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
@@
 \ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
+Single-distribution data &  & \\
 \ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
@@
 slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
+with a single-distribution process. However, for the data generated with
+a two-distribution mixture process, the mixture model shows a
+substantially higher predictive performance. In fact, the ratio of
+\(\Delta\widehat{elpd}\) and its standard error -- a \emph{z}-score
+approximation -- reveals a value of 11.6 which is strong evidence in in
+favour of the mixture model compared to the single-distribution model
+(Sivula et al., 2020). In comparison, for the single-distribution data,
+the single-distribution model was negligibly better than the
+two-distribution mixture model which was evidence by a value of 0.77 for
+the ratio of \(\Delta\widehat{elpd}\) and its standard error. Thus, as
+the difference \(\Delta\widehat{elpd}\) between the model is negligible,
+the single distribution is preferred by the law of parsimony.
 
@@
 \midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
+Data: Two-distribution mixture process &  & \\
+\ \ \ Two-distribution mixture model & -- & -6,068 (41)\\
+\ \ \ Single-distribution model & -191.3 (16.5) & -6,259 (38)\\
+Data: Single-distribution process &  & \\
+\ \ \ Single-distribution model & -- & -5,030 (24)\\
+\ \ \ Two-distribution mixture model & -0.5 (0.7) & -5,030 (24)\\
 \bottomrule

### 2023-12-09 report/appendix_transloc.Rmd
Reorganised manuscript

--- 2023-07-24 report/manuscript.Rmd
@@
----
-title             : "Modelling writing hesitations in text writing as finite mixture process"
-shorttitle        : "Modelling writing hesitations"
-
-csl               : "apa.csl"
-
-author: 
-  - name          : "Jens Roeser"
-    affiliation   : "1"
-    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
-    corresponding : yes 
-    email         : "jens.roeser@ntu.ac.uk"
-  - name          : "Mark Torrance"
-    affiliation   : "1"
-  - name          : "Rianne Conijn"
-    affiliation   : "2"
-  - name          : "Evgeny Chukharev-Hudilainen"
-    affiliation   : "3"
-
-affiliation:
-  - id            : "1"
-    institution   : "Department of Psychology, Nottingham Trent University, United Kingdom"
-
-  - id            : "2"
-    institution   : "Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands"
-
-  - id            : "3"
-    institution   : "Department of English, Iowa State University, Iowa"
-
-abstract: |
-
-keywords: "Keystroke modelling; finite mixture models; Bayesian models; text composition"
-
-bibliography      : ["references.bib"]
-
-documentclass     : "apa7"
-classoption       : "man"
-output            : 
-  papaja::apa6_pdf:
-    keep_tex: TRUE
-    includes:
-      after_body: 
-        - "appendix_a.tex"
-        - "appendix_b.tex"
-        - "appendix_c.tex"
-        - "appendix_d.tex"
-        - "appendix_e.tex"
-        - "appendix_g.tex"
-        - "appendix_f.tex"
-
-figsintext        : yes
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-lineno            : no
-mask              : no
-csquotes          : true
-
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{graphicx}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{threeparttable}
-  - \usepackage[normalem]{ulem}
-  - \usepackage[utf8]{inputenc}
-  - \usepackage{icomma}
-  - \usepackage{pdflscape}
-  - \newcommand{\blandscape}{\begin{landscape}}
-  - \newcommand{\elandscape}{\end{landscape}}
-  - \DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, 
-                      warning = FALSE, 
-                      message = FALSE, 
-                      comment = NA,
-                      cache.extra = R.version,
-                      dev = "cairo_pdf",
-                      fig.align = 'center', 
-                      fig.width = 8, 
-                      fig.height = 5,
-                      width=90)
-
-options(kableExtra.auto_format = FALSE)
-dev.args <- list(pdf = list(type = "cairo"))
-```
-
-```{r libs, include=FALSE}
-library(tidyverse)
-library(janitor)
-library(brms)
-library(papaja)
-library(polspline)
-library(patchwork)
-library(gridExtra)
-library(grid)
-library(gtable)
-library(scales)
-library(ggthemes)
-source("../scripts/functions.R")
-source("scripts/get_pdens_plot.R")
-
-theme_set(theme_few(base_size = 12) +
-            theme(strip.background = element_blank(),
-                  legend.position = "top",
-                  legend.justification = "right",
-                  panel.grid = element_blank(),
-                  panel.background = element_rect(fill = "transparent"), # bg of the panel
-                  plot.background = element_rect(fill = "transparent", color = NA)))
-
-options(scipen = 999)
-
-beta_label <- "Short key transition\nduration (in msecs)"
-delta_label <- "Slowdown for hesitant key\ntransitions (in secs)"
-theta_label <- "Probability of hesitant\ntransitions"
-```
-
-# Introduction
-
-# Methods
-
-## Statistical models
-
-The models presented in the following can be divided into two general groups.
-The first three models are largely akin to models typically used in the literature.
-By this we mean models that assume a uni-modal process that generates keystroke data as is incorporated in statistical models such as analysis of variance and linear mixed-effects models.
-Second, the last two models model keystroke intervals as a combination of two weighted processes of which one presents a smooth information flow from mind into finger; the other component is more important as it represents moments at which the information flow was interrupted leading to longer latencies.
-The latter two models map directly on the idea of a cascading model of writing.
-
-### Unimodal Gaussian
-
-We start with a Gaussian mixed-effects models similar to a standard analysis of variance.
-We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
-The mean can be decomposed into $\beta$ and $\text{participant}_i$.
-$\beta$ will be allowed to take on a different value for each transition location.
-Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:unimodgaus)
-\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
-& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-### Unimodal log-Gaussian
-
-This model is largely identical to the previous model, except we use a log-normal distribution instead of a normal distribution.
-The advantage of using a log-normal distribution is two-fold: (1) the log-normal distribution has a lower bound of zero.
-For our data we generally consider negative ikis as mistakes which occasionally occur when the next key was pressed before the current key.
-Other than that keystroke intervals are constrained by a persons ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour, in particular motor responses.
-In particular, a normal distribution assumes that units are linearly scaled.
-For example, a 50 msecs difference is the same between 100 msecs and 150 msecs than it is between 5 secs and 5 secs 50 msecs (i.e.
-5,050 msecs).
-This is not necessarily plausible for keystroke data.
-We would assume that differences that are due to motor activity (typing a high frequency bigram vs a low frequency bigram) are smaller than difference that are due to high levels of cognitive activity (retrieving a word in your L1 or L2).
-Log-normal distributions are a natural way of scaling units so that a 50 msecs difference on the lower end of the iki scale (motor activity) is more meaningful than a 50 msecs difference on the upper end of the iki scale (retrieving words, planning sentences).
-
-The model can be described like this:
-
-\begin{equation}
-\begin{aligned}
-(\#eq:unimodloggaus)
-\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
-& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-### Unimodal unequal-variance log-Gaussian
-
-\begin{equation}
-\begin{aligned}
-(\#eq:unimoduv)
-\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
- & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-### Bimodal log-Gaussian (constrained)
-
-This models extends the intuition from the previous model that higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that underlies the generation of ikis, we assume there are two.
-(1) activation can flow into keystrokes without interrupts.
-These fluent keystroke transitions are merely limited by a person's ability to move their finger and will be captured but the parameter $\beta$.
-In principle, there are no differences for fluent key-transitions between transition location.
-The next model with loosen this assumption (hence, "unconstrained model").
-(2) difficulty in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when spelling, words, or contents couldn't be retrieved in time.
-The size of these pauses will depend on the reason for delays which is typically associated with transition locations (contents are typically planned before sentences, words are retrieved before they are typed and spelling difficulty typically occurs when typing a word).
-Pauses will be captured by two model parameters: (1) the slowdown for these hesitant transitions will be captured by $\delta$ which is the deviation compared to normal typing intervals (constrained to be positive).
-(2) the frequency of hesitant transitions will be captured by $\theta$ for each level of a categorical predictor.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:bimodcon)
-\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-```{r}
-#		<!-- &\delta \sim \text{N}(0,1)\\ -->
-
-```
-
-This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
-
-### Bimodal log-Gaussian (unconstrained)
-
-This model is identical to the previous model with one exception.
-The distribution of fluent keystroke transitions captured by $\beta$ was fixed to be the same across transition locations is the previous model.
-In other words, the mean $\beta$ and it's standard deviation $\sigma_{e}^2$ was the same for before-sentence, before-word, and within-word transitions.
-This means, because of the naturally larger number of within-word transitions, the posterior is dominated by within-word transitions.
-
-In this model we will loosen this constraint and allow $\beta$ and $\sigma_{e}^2$ to vary by transition-location.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:bimoduncon)
-\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
-	\text{where: }  & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-```{r}
-#<!-- & \delta \sim \text{N}(0,1)\\ -->
-
-```
-
-## Data sets
-
-Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \ref{tab:datasets}.
-
-\blandscape
-```{r datasets, results='asis'}
-tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
-                Source = c("Rønneberg et al. (2022)",
-                           "Torrance et al. (2016)",
-                           "Torrance et al. (n.d.)",
-                           "Rossetti and Van Waes (2022)",
-                           "Vandermeulen et al. (2020)", 
-                           "Torrance and Ofstad (n.d)"),
-                Keylogger = c("EyeWrite", 
-                           "EyeWrite", 
-                           "CyWrite", 
-                           "InputLog", 
-                           "InputLog",
-                           "EyeWrite"),
-                `Writing task` = c("Argumentative essays", 
-                                 "Expository texts",
-                                 "Argumentative essays",
-                                 "Text simplification",
-                                 "Synthesis", 
-                                 ""),
-                 `N (ppts)` = as.character(c(126, 26*2, 39, 47, 658, 45)),
-#                n_texts = c(1, 2, 2, 2, NA),
-#                n_sentences = c(),
-#                n_words = c(),
-                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
-                 Language = c( "Norwegian", 
-                               "Norwegian", 
-                               "English (L1) / Spanish (L2)", 
-                               "English (L2)", 
-                               "Dutch",
-                               "Norwegian"),
-                Manipulation = c("--", 
-                               "weak decoders / control; masked / unmasked",
-                               "write in L1 / L2",
-                               "pre / post test trained in plain language principles and control", 
-                               "Various topics and genres",
-                               "masked / unmasked")) %>% 
-  arrange(Dataset) %>% 
-  apa_table(align = c(rep("l", 4), rep("r", 2), "p{2cm}", "p{3cm}"), 
-            escape = FALSE, 
-            font_size = "footnotesize",
-            caption = "Datasets in brief.")
-
-```
-\elandscape
-
-### GUNNEXP2
-
-TODO: Mark, can you add information here? I've added a provisional bib entry that needs changing or we just say "unpublished"
-
-The GUNNEXP2 dataset published in @torranceb includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote text either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### C2L1
-
-TODO: Mark, can you add information here?
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)?
-
-The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process.
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### CATO
-
-TODO: Mark, can you add information here?
-
-Data are published in @torrance2016adolescent.
-Norwegian upper secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite [@sim07; XXXX].
-
-### PLanTra
-
-TODO: Rianne, can you add information here?
-
-The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group.
-In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience.
-Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability.
-During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e.
-making it easier to read for a lay audience--by applying what they had learned from their respective training.
-The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language.
-It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.
-Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### LIFT
-
-TODO: Rianne, can you add information here?
-
-The LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) data are published in @vandermeulen2020.
-Students produced text for 4 topics in two different writing genres (levels: argumentative text, informative text).
-Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### SPL2
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in @torrancea.
-The data come from a textcomposition task of undergraduate university students -- *N* = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
-CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste.
-We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.
-Participants were given a 40 minute time limit.
-They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-## Data analysis
-
-### Bayesian modelling
-
-We reanalysed keystroke data from 6 datasets in Bayesian implementations of 5 statistical models. 
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values: Bayesian parameter estimates are expresses the uncertainty associated with parameter values as probability distribution [@farrell2018computational; @gelman2014; @lee2014bayesian].
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
-existing knowledge about parameter values, for regulating the parameter space.
-For small data sets priors influence the inferred parameter estimates (known as posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
-automatic Ockham's razor, @jefferys1992ockham].
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
-@lambert2018student; @mcelreath2016statistical].
-The predictive performance (i.e.
-fit) of these models is compared in the Results section using leave-one-out cross-validation.
-
-Keystroke data from text composition tasks were modelled in a series of five models. An overview of all models can be found in Table \ref{tab:models}.
-
-```{r models, results = 'asis'}
-models <- tibble(Models = paste0("M",1:5),
-       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
-       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
-       Description = c("Gaussian distribution",
-                       "Log-normal distribution",
-                       "Log-normal distribution and unequal variances",
-                       "Fluent transition are not allowed to vary by transition location",
-                       "Fluent transition are allowed to vary by transition location"))
-
-apa_table(models, 
-          align = c(rep("l", 2), "r", "p{9cm}"), 
-          escape = FALSE, 
-          digits = 0,
-          caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
-          note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
-```
-
-For all models we modelled transition locations (levels: before sentence, before word, within word) was included as predictor in all models.
-Also, we included the dataset specific experimental conditions as fixed effects and included random intercepts for participants as described in the equations above.
-We used normal and log-normal probability function in line with typically treatments fouund in the literature.
-Stan code for mixture models was based on @roeser2021modelling  [see also @vasishth2017; @vasishth2017feature].
-
-Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical].
-The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
-
-### Transition types
-
-The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g.
-@torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
-Keytransitions that terminated in an editing operation were excluded from the analysis.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions.
-
-```{r keyloc, results='asis'}
-
-table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
-       Description = c("Transitions between any letter",
-                       "Keypress after space followed by any letter",
-                       "Keypress following a space preceding any letter"),
-       Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d.
-T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.",
-                   "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
-                   "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept."))
-
-apa_table(table, caption = "Transition location classification.",
-            align = c("p{3cm}", "p{3cm}", "p{8cm}"), 
-            escape = FALSE, 
-            font_size = "footnotesize",
-            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace. At before-sentence locations, IKIs were timed to the shift keypress.")
-
-```
-
-### Data reduction
-
-```{r reductionfunctions}
-source("scripts/get_data_summary.R")
-
-# Get info about data reduction
-n_samples <- 100
-c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
-cato <- get_cato("../data/cato.csv", n_samples = n_samples)
-lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
-plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
-spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
-gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
-```
-
-For all datasets we only used transitions that were not followed by an editing operation.
-
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant)
-
-We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-
-From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis.
-Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-```{r datareduction, results='asis'}
-c2l1$ds <- "c2l1"
-cato$ds <- "cato"
-lift$ds <- "lift"
-plantra$ds <- "plantra"
-spl2$ds <- "spl2"
-gunnexp2$ds <- "gunnexp2"
-
-extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
-          mutate(ds = .x$ds)) %>% 
-  mutate(across(where(is.numeric), ~round(.*100, 2))) %>% 
-  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
-  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
-  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
-  rename(`$\\le$ 50 msecs` = `<50ms`,
-         `$\\ge$ 30 secs` = `>30,000ms`)
-
-random_sample <- 
-  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
-          mutate(ds = .x$ds)) %>% 
-    mutate(across(where(is.numeric), ~round(.*100, 1))) %>%
-    unite("keep", starts_with("keep"), sep = " (") %>% 
-    mutate(across(c(keep), ~str_c(., ")"))) %>%
-  pivot_wider(names_from = location, values_from = keep)
-
-datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
-  relocate(ds) %>% 
-  mutate(across(ds, ~recode(., 
-                "cato" = "CATO",
-                "spl2" = "SPL2", 
-                "spl2_shift" = "SPL2 (shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2"))) %>% 
-  rename(`Dataset` = ds)
-
-apa_table(datareduction, 
-          escape = FALSE,  
-          placement = "bp!",
-          caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location. Standard error is shown in parentheses.",
-          align =c("l", rep("r", 5)),
-          col_spanners = list(" " = 1,
-                              "Extreme values in \\%" = c(2, 3),
-                              "Randomly sampled data in \\%" = c(4, 6)))
-
-```
-
-# Results
-
-## Out-of-samples cross-validation
-
-To compare the out-of-sample predictive performance of the models we used Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
-Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and its  difference $\Delta\widehat{elpd}$ between models.
-The advantage of using leave-one-out cross-validation is that models with more parameters are penalised to prevent overfit.
-
-Results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-The mixture of log-normal distribution models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across transition locations.
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-
-```{r}
-files <- list.files(
-  str_c("../stanout/", 
-  c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
-  pattern = "model.+.csv", 
-  full.names = T)
-
-table <- purrr::map_dfr(files, ~read_csv(.) %>% 
-                          mutate(dataset = .x)) %>% 
-  select(dataset, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
-  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(where(is.numeric), ~round(., 0)),
-         across(where(is.numeric), ~format(., big.mark = ",")),
-         across(everything(), ~str_trim(.))) %>%
-  filter(dataset != "spl2_shift") %>% 
-  transmute(
-     across(dataset, ~recode(., 
-                "cato" = "CATO",
-                "spl2" = "SPL2", 
-#                "spl2_shift" = "SPL2 (shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2")),
-    m = recode(model, 
-               lmmgaus = "M1",
-               lmm = "M2",
-               lmmuneqvar = "M3",
-               mogbetaconstr = "M4",
-               mogbetacontr = "M4",
-               mogbetaunconstr = "M5"),
-    across(model, ~recode(., lmm = "Unimodal log-normal",
-                          lmmgaus = "Unimodal normal",
-                          mogbetaconstr = "Bimodal (constrained)",
-                          mogbetacontr = "Bimodal (constrained)",
-                          mogbetaunconstr = "Bimodal (unconstrained)",
-                          lmmuneqvar = "Unimodal (unequal variance)")),
-            across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-            across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")"))) %>% 
-  mutate(across(elpd_diff, ~str_replace(., "0 \\(0\\)", "--"))) %>% 
-  arrange(dataset)
-```
-
-```{r loos, results='asis'}
-names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
-
-apa_table(table[,-1], 
-      caption = "Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-      align =c("p{3.5cm}", "p{6cm}", rep("r", 2)), 
-      escape = FALSE,
-      longtable = T,
-      font_size = "small",
-      note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      stub_indents = list(
-                     "\\textbf{CL21}" = 1:5,
-                     "\\textbf{CATO}" = 6:10,
-                     "\\textbf{GUNNEXP2}" = 11:15, 
-                     "\\textbf{LIFT}" = 16:20,
-                     "\\textbf{PLanTra}" = 21:25,
-                     "\\textbf{SPL2}" = 26:30))
-#                     "\\textbf{SPL2 (shift + C)}" = 31:35))
-
-```
-
-## Cross-data set comparisons
+## Transition location effect (unconstrained mixture model)
 
@@
 
-As detailed above the mixture model captures the writing process in three conceptually important parameter values: (1) the transition duration associated with fluent writing (which we indicated as $\beta$), (2) the size of the slowdown for hesitant transitions (indicated as $\delta$), and (3) the mixing proportion which represents the probability of hesitant transition durations (indicated as $\theta$).
+We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
+Similar to the constrained mixture model, we aggregated the posterior across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.] and removed conditions that might conflate comparisons^[We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.].
+The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
 
-We present the mixture-model estimates for each of the three parameter values in Figure \ref{fig:crossstudypost}.
-Although models were fitted with all dataset-specific condition, we aggregated across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence for differences between these conditions], and removed conditions that might conflate comparisons^[We removed masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO dataset, and L2 writing in the SPL2 data.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that evidence for differences in parameter values for the involved manipulations.].
-For posteriors of all conditions within dataset see Appendix \ref{posterior-parameter-estimates}.
-The posterior of the remaining conditions allows us to compare differences between transition locations within dataset associated with each of the three mixture-model parameters.
+<!-- We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.],  -->
 
-```{r crossstudypost, fig.cap="Estimates across studies. Posterior parameter distribution represented as estimate and 95\\% probability interval (PI)."}
+<!-- ^[We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
+Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.].
+For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
+-->
 
-posd <- position_dodge(.65)
-dotsize <- 2.5
-grouplabel <- "Data set"
+```{r crossstudypost, fig.cap="Mixture model parameter estimates across studies.
+Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
+Axes for transitions durations are log-scaled for visability."}
+
+posd <- position_dodge(.5)
+dotsize <- 2.75
+grouplabel <- "Data set:"
 shapes <- c(1, 4, 6, 7, 8, 9, 11)
 width <- 0
-linewidth <- .75
+linewidth <- .35
+nrow <- 1
 beta <- filter(ps, name == "beta") %>% 
@@
   scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
+#  scale_y_log10(labels = scales::comma, breaks = c(50, 150, 200, 300, 400, 500)) +
   theme(axis.title = element_blank()) +
@@
   scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
+  #scale_y_(labels = scales::comma) +
   labs(colour = grouplabel,
@@
   theme(axis.title = element_blank()) +
-  guides(colour=guide_legend(nrow=2,byrow=TRUE),
-         shape = guide_legend(nrow=2,byrow=TRUE))
+  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
+         shape = guide_legend(nrow=nrow,byrow=TRUE))
 
-plot <- beta + delta + theta +
-  plot_layout(guides = "collect")
+plot <- beta + delta + theta + plot_layout(guides = "collect")
 
@@
 
-Figure \ref{fig:crossstudypost} shows largely the same patterns for keystroke interval estimates by transition location across datasets for all three mixture model parameters.
-keystroke transitions were substantially longer for the C2L1 data presumably reflecting that the sample of in this dataset was the youngest.
-Pause durations and frequencies were relatively similar across datasets.
+In the following we evaluate differences between transition locations for all three mixture model parameters.
+Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
+We found that hesitations appear more frequently at before-word transitions than within words across data sets (all BFs > 100).
+Also hesitations are longer at before-sentence transitions compared to before-word transitions (all BFs > 100; except for C2L1: BF = 0.35 and LIFT = 2.14) and before-word transitions compared to within-word transitions (all BF > 10; except for LIFT: BF = 0.71).
+Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations (all BFs > 100) but there is generally no evidence for a difference for fluent transitions for before-sentence transitions compared to before-word transitions (BFs < 0.08; except for SPL2: BF > 100 and GUNNEXP2: BF > 100).
+The full results of these pairwise comparisons can be found in Table \ref{tab:loceffect}.
 
-In the following we will evaluate differences between transition locations for all three mixture model parameters.
+Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
+However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
 This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence $>$ before-word $>$ within-word.
-We evaluated the differences between transition locations for all data sets.
-The results are shown in Table \ref{tab:loceffect}.
-
-Comparing transition locations rendered patterns that are largely consistent across data sets (with caveats).
-We found that hesitations appear more frequently at before-word transitions than within words across datasets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except dataset C2L1) compared to within-word transitions (except dataset LIFT).
-However, for most dataset there is negligble evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
-In fact, the dataset LIFT showed less pausing before sentences compared to before words.
-Also, we observe that even fluent key-transitions are slower at before-word locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for datasets SPL2 and GUNNEXP2).
-Hesitation durations tend to be longer at before-sentence locations compared to before-word locations (except for datasets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT dataset).
-
+In fact, the data set LIFT showed less pausing before sentences compared to before words.
 In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
 
-The inconsistencies for before-sentence transitions could, to some extent, be explained on the basis that datasets differ as to whether before-sentence transitions should be analysed as complex key combination.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-Notice though that for the datasets PLanTra and LIFT, there were no consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
+Outstanding is the overall substantially longer fluent transitions for the C2L1 data.
+This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
+Hesitation duration and frequencies were similar to the other data sets.
+There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
+Some data sets show before-sentence slowdowns for fluent transition compared to words.
+These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
+In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
+Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
 
-We test to what extent this difference may affect the modelling results for the SPL2 dataset.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
+To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
+The results are shown in Appendix \ref{key-combination-effect-unconstrained-mixture-model}.
 We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this is in conflict with the absence of differences in datasets that included the character following shift at before-sentence transitions.
-In other words, it is unlikely that patterns in our results can be explained on the basis of complex key combinations at before-sentence locations.
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
+However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
+In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
 
@@
 \elandscape
-
-## Simulation
-
-A possible concern with the results -- substantially better predictive performance for mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-
-To address this concern we repeated a comparison for an unimodal model and a mixture model for two sets of simulated data. For each dataset we simulated 1,000 observations.
-
-These two datasets differ with regards to their underlying data generating process.
-The first set of data has as underlying data generating process a mixture model with two mixture components similar to the process described above.
-This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:simmog)
-\text{y}_i \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-		& \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical with the models before but reduced to its main parameters.
-There are two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.
-
-The second dataset was generated with an unimodal model as data generating process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:simuv)
-\text{y}_i \sim\text{ }& \text{logN}(\beta, \sigma^2)
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the unimodal models used in the main analysis.
-The model assumes a log-normal distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
-
-The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
-The simulated data are visualised in Figure \ref{fig:simdata}.
-The data are simulated to be similarly distributed to keystroke transitions.
-
-```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=3.5, fig.width=5}
-files <- list.files("../mixture-model-sim/data", full.names = T)
-
-map_dfr(files, ~read_csv(.) %>% 
-        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
-  mutate(across(Dataset, ~recode(., mogdata = "bimodal", lmdata = "unimodal"))) %>% 
-  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
-  geom_density(alpha = .25) +
-  scale_colour_colorblind() + 
-  scale_fill_colorblind() +
-  scale_x_continuous(labels = ~format(., big.mark = ",")) +
-  labs(colour = "Data process", 
-       fill = "Data process", 
-       x = "Simulated data", 
-       y = "Density")
-```
-
-We run 4 models, 2 on each of the two datasets: 2 mixture models, one for the data generated with a bimodal mixture process and one for the data generated with the unimodal process.
-We repeated the same using an unimodal model.
-Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
-Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}.
-The parameters are shown by data generating process along with with true parameter values.
-Parameter estimates are shown for the dataset generated with a unimodal process and the dataest generated with a bimodal process.
-The Bayesian models uncovered the model parameters of the dataset with the matching data generating process successfully, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-
-```{r}
-mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
-  mutate(data = "mog")
-mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
-  mutate(data = "lm")
-lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
-  mutate(data = "mog")
-
-mog_sim <- bind_rows(mog1, mog2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>%
-  select(-lo, -up) %>% 
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = c(5, 1, .35, .25, .5),
-         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
-         across(name, ~str_replace(., "prob", "theta"))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
-
-lm_sim <- bind_rows(lm1, lm2) %>% 
-  summarise(across(value, list(est = mean,
-                               lo = lower,
-                               up = upper),
-                   .names = "{.fn}"),
-            .by = c(name, data)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(est, ~str_c(., " [", lo, ", ", up, "]")),
-         across(name, ~str_c("\\", .)),
-         across(name, ~str_replace(., "a_e", "a^2_1")),
-         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>% 
-  select(-lo, -up) %>%
-  pivot_wider(names_from = data, values_from = est) %>% 
-  mutate(true = as.character(c(5, .25))) %>% 
-  relocate(name, true) %>% 
-  unite("name", name:true, sep = " = ")
-
-```
-
-```{r simparam, results='asis'}
-table <- bind_rows(mog_sim %>% mutate(model = "mog"),
-                   lm_sim %>% mutate(model = "lm")) %>% 
-  relocate(model) %>% 
-  mutate(across(name, ~str_c("$", ., "$")))
-
-names(table) <- c("Model", "Parameter with true value", "Bimodal data", "Unimodal data")
-
-apa_table(table[,-1], 
-          escape = FALSE,
-      col_spanners = list("Estimate with 95\\% PI" = 2:3),
-      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
-      align = c("l", rep("r", 2)),
-      stub_indents = list(`Bimodal mixture process` = 1:5, 
-                          `Unimodal process` = 6:7))
-
-```
-
-We used LOO-CV to compare the fit of the two models for each data set.
-The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
-The results rules out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a unimodal process.
-However, for the data generated with a bimodal process, the mixture model shows a substantially higher predictive performance.
-
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence [@sivula2020uncertainty], shows that the mixture model performs 11.6 times better than the unimodal model for the data generated with a bimodal process.
-In comparison, for the unimodal data, the unimodal model performs only 0.77 times better than the bimodal mixture model.
-Thus, even though the mixture model does not necessarily perform better for non-bimodal data, it also does not necessarily perform worse.
-
-```{r loossim, results='asis'}
-table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
-  select(1:6) %>% 
-  mutate(
-#    elpd_se_ratio = abs(elpd_diff / se_diff),
-    across(ends_with("diff"), ~round(., 1)),
-    across(ends_with("loo"), ~round(., 0)),
-    across(where(is.numeric), ~format(., big.mark = ",")),
-    across(everything(), ~as.character(.)),
-    across(everything(), ~str_trim(.)),
-    across(everything(), ~str_replace_all(., "NaN", "--")),
-    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
-    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
-    across(c(data, model), ~recode(., mog = "Bimodal mixture model",
-                                      lm = "Unimodal model"))) %>% 
-  select(data, model, starts_with("elpd"))
-
-names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
-
-apa_table(table[,-1], 
-          escape = FALSE,
-      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      align =c("l", rep("r", 2)),
-      stub_indents = list(`Data: Bimodal mixture process` = 1:2, 
-                          `Data: Unimodal process` = 3:4))
-
-```
-
-# Discussion
-
-# References
-```{r create_r-references, echo=FALSE, include=FALSE}
-r_refs(file = "references.bib")
-```
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-<div id = "references"></div>
-\endgroup
-
-```{r render_appendix, include=FALSE}
-render_appendix("appendix_a.Rmd")
-render_appendix("appendix_b.Rmd")
-render_appendix("appendix_c.Rmd")
-render_appendix("appendix_d.Rmd")
-render_appendix("appendix_e.Rmd")
-render_appendix("appendix_g.Rmd")
-render_appendix("appendix_f.Rmd")
-```

### 2023-12-09 report/appendix_transloc.tex
Reorganised manuscript

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
+\clearpage
 \makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
+\efloat@restorefloats
 \makeatother
 
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\begin{appendix}
+\section{}
+\hypertarget{transition-location-effect-unconstrained-mixture-model}{%
+\subsection{Transition location effect (unconstrained mixture
+model)}\label{transition-location-effect-unconstrained-mixture-model}}
 
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
+We present the mixture-model posterior estimates for each of the three
+parameter values in the three facets of Figure \ref{fig:crossstudypost}.
+Similar to the constrained mixture model, we aggregated the posterior
+across conditions\footnote{We aggregated across pre-post test for the
+  PLanTra dataset as well as genre and topic of the LIFT data set.} and
+removed conditions that might conflate comparisons\footnote{We removed
+  the masked writing condition in the GUNNEXP2 and CATO, the dyslexic
+  group in the CATO data set, and L2 writing in the SPL2 data set.}. The
+resulting posterior allows us to examine differences between transition
+locations for each data set associated with each of the three
+mixture-model parameters.
 
@@
 Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
+Axes for transitions durations are log-scaled for visability.}(\#fig:crossstudypost)
 \end{figure}
 
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
+In the following we evaluate differences between transition locations
+for all three mixture model parameters. Figure \ref{fig:crossstudypost}
+shows largely the same patterns (with caveats) for keystroke interval
+estimates by transition location across data sets for all three mixture
+model parameters. We found that hesitations appear more frequently at
+before-word transitions than within words across data sets (all BFs
+\textgreater{} 100). Also hesitations are longer at before-sentence
+transitions compared to before-word transitions (all BFs \textgreater{}
+100; except for C2L1: BF = 0.35 and LIFT = 2.14) and before-word
+transitions compared to within-word transitions (all BF \textgreater{}
+10; except for LIFT: BF = 0.71). Further, we observed that fluent
+key-transitions are slower at before-word and before-sentence locations
+compared to within-word locations (all BFs \textgreater{} 100) but there
+is generally no evidence for a difference for fluent transitions for
+before-sentence transitions compared to before-word transitions (BFs
+\textless{} 0.08; except for SPL2: BF \textgreater{} 100 and GUNNEXP2:
+BF \textgreater{} 100). The full results of these pairwise comparisons
+can be found in Table \ref{tab:loceffect}.
 
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
+Hesitation duration tends to be longer at before-sentence locations
+compared to before-word locations (except for data sets C2L1 and LIFT)
+and longer for before-word locations compared to within-word locations
+(except for the LIFT data set). However, for most data set there is
+negligible evidence for the idea that writers pause more frequently at
+before-sentence locations compared to before-word locations (except for
+data sets SPL2 and GUNNEXP2). This is interesting because it is
+generally believed that pausing behaviour is associated with syntactic
+edges such that more and longer pauses are predicted for key transitions
+at larger syntactic edges following the pattern before-sentence \(>\)
+before-word \(>\) within-word. In fact, the data set LIFT showed less
+pausing before sentences compared to before words. In brief, while
+keystroke transitions and pauses tend to be longer and more frequent at
+before-word locations compared to within-word transitions, it is not
+clear in which contexts keystroke transitions are before-sentence
+locations are slower, and their hesitations are longer and more
+frequent.
 
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
+Outstanding is the overall substantially longer fluent transitions for
+the C2L1 data. This is presumably reflecting that the population that
+this sample is from was the youngest among our data sets presumably
+involving the least experienced writers in our data pool. Hesitation
+duration and frequencies were similar to the other data sets. There are
+some inconsistencies for fluent before-sentence transitions compared to
+before-word transitions. Some data sets show before-sentence slowdowns
+for fluent transition compared to words. These inconsistencies could, to
+some extent, be explained on the basis that data sets differ as to
+whether before-sentence transitions involve complex key combination
+involve the mean or sum of transitions between space, shift and / or the
+sentence-initial letters. In particular, some data include the character
+key following the shift key at before sentence location (PLanTra, LIFT)
+but others did not scope over the character following the shift key
+(CATO, C2L1, SPL2, GUNNEXP2). Notice though that for the data sets
+PLanTra and LIFT, there was no evidence for a consistent differences
+between before-sentence and before-word transitions (except for longer
+hesitations in the PLanTra dataset).
 
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
+To address this finding, and inconsistency in how before-sentence
+transitions were timed, we test to what extent this difference affected
+the modelling results for the SPL2 data set. The results are shown in
+Appendix \ref{key-combination-effect-unconstrained-mixture-model}. We
+found that including the character following the shift key substantially
+extends both the transition duration and the hesitation duration but not
+the hesitation frequency. However, this conflicts with the absence of
+differences in data sets that included the character following shift at
+before-sentence transitions (PLanTra, LIFT). In other words, it is
+unlikely that patterns in our results can be explained on the basis of
+how before-keystroke transitions were operationalised (complex
+key-combinations at before-sentence locations).
 
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
+\blandscape
 
@@
 \end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
 \elandscape
 \end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}

### 2023-12-09 report/appendix_transloc_constr.Rmd
Reorganised manuscript

## Transition location effect (constrained mixture model)

```{r loceffectcon, results='asis'}

files <- list.files(str_c("../stanout/", 
                          c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetacon.+.csv", full.names = T) # "spl2_shift",

ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
                                 mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, ~recode(., 
                              "cato" = "*CATO* (non-dyslexic\nunmasked)",
                              "spl2" = "*SPL2* (L1)", 
                              "gunnexp2" = "*GUNNEXP2* (unmasked)",
                              #                "spl2_shift" = "*SPL2* (L1; shift + C)", 
                              "plantra" = "*PLanTra*",
                              "lift" = "*LIFT*",
                              "c2l1" = "*C2L1*"))) %>% 
  filter(!(str_detect(data, "SPL2") & lang == "ES"),
         !(str_detect(data, "CATO") & group == "dyslexic"),
         !(str_detect(data, "CATO") & task == "masked"),
         !(str_detect(data, "GUNNEXP2") & xn == "masked"),
         param %in% c("delta", "theta")) %>%
  select(param, location, data, value) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
  pivot_wider(names_from = location, values_from = value) %>%
  unnest(-c(data, param)) %>% 
  mutate(diff.1 = `before sentence` - `before word`,
         diff.2 = `before word` - `within word`) %>% 
  summarise(across(starts_with("diff"), 
                   list(mean = mean, 
                        lower = lower, 
                        upper = upper, 
                        BF = BF)),
            .by = c(data, param))

tmp <- ps_loc_diffs %>% 
  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
  select(-lower, -upper) %>% 
  pivot_wider(names_from = param, values_from = c(mean, BF)) %>% 
  mutate(across(diffid, ~recode(., diff.1 = "before sentence vs word",
                                diff.2 = "before vs within word"))) %>% 
  select(data, diffid, ends_with("delta"), ends_with("theta")) %>% 
  arrange(data)

write_csv(tmp, "tables/location_effect_constr_alldata.csv")

apa_table(tmp[,-1],
          align =c("l", rep("r", 4)), 
          longtable = T,
          escape = FALSE,
          row.names = T,
          font_size = "footnotesize",
          col.names = c("Comparison", rep(c("Est. [95\\% PIs]", "BF"), 2)),
          stub_indents = list(`\\textbf{C2L1}` = 1:2, 
                              `\\textbf{CATO (non-dyslexic, unmasked)}`= 3:4,
                              `\\textbf{GUNNEXP2 (unmasked)}` = 5:6,
                              `\\textbf{LIFT}` = 7:8,
                              `\\textbf{PLanTra}` = 9:10,
                              `\\textbf{SPL2 (L1)}` = 11:12),
          #                              `\\textbf{SPL2 (L1; shift + C)}` = 13:14),
          col_spanners = list(" " = c(1),
                          #    "Fluent transitions" = c(2, 3), 
                              "Hesitation slowdown" = c(2, 3), 
                              "Hesitation probability" = c(4, 5)),
          note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
          caption = "Effect of transition location on keystroke intervals. Differences are shown on log scale (for durations) and logit scale for hesitation probability. 95\\% PIs in brackets.")

```

### 2023-12-09 report/manuscript.Rmd
Reorganised manuscript

@@
     affiliation   : "2"
-  - name          : "Evgeny Chukharev-Hudilainen"
+  - name          : "Evgeny Chukharev"
     affiliation   : "3"
@@
 
-abstract: Classical serial models view the writing process as a chain of pauses and writing burst.
+abstract: Classical serial models view the process of producing a text as a descrete chain of pauses and writing bursts.
 In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
@@
 Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
+Our results cannot be explained by the serial but are in line with the cascading view of writing.
 
@@
       after_body: 
-        - "appendix_a.tex"
-        - "appendix_b.tex"
-        - "appendix_c.tex"
-        - "appendix_d.tex"
-        - "appendix_e.tex"
-        - "appendix_g.tex"
-        - "appendix_f.tex"
-        - "appendix_h.tex"
+        - "appendix_models.tex"
+        - "appendix_fit2data.tex"
+#        - "appendix_postparam.tex"
+        - "appendix_postparam_constr.tex"
+        - "appendix_shift_constr.tex"
+#        - "appendix_l2.tex"
+        - "appendix_l2_constr.tex"
+#        - "appendix_masking.tex"
+        - "appendix_masking_constr.tex"
+#        - "appendix_prepost.tex"
+        - "appendix_prepost_constr.tex"
+#        - "appendix_genre.tex"
+        - "appendix_genre_constr.tex"
+        - "appendix_transloc_constr.tex"
+        - "appendix_transloc.tex"
+        - "appendix_shift.tex"
+        - "appendix_sim.tex"
 
@@
                       fig.width = 8, 
-                      fig.height = 5,
+                      fig.height = 4.5,
                       width=90)
@@
 
-beta_label <- "Short transition\nduration (in msecs)"
+beta_label <- "Fluent transition\nduration (in msecs)"
 delta_label <- "Slowdown for hesitant\ntransitions (in secs)"
@@
 We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
+The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models.
 In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component [@wagenmakers2007linear; @wing1973response; @schoner2002timing].
 
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in @roeser2021modelling.
+The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in @roeser2021modelling.
 Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
@@
 
-## Statistical models
-
-We are using the Bayesian framework [e.g.
-@farrell2018computational; @gelman2014; @lee2014bayesian] to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-### Serial model of writing
-
-#### Unimodal Gaussian
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average IKI associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as $\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)$.
-Of importance is the estimated posterior value of $\mu$ as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07].
-Regression formulas can, and commonly are, used to decompose $\mu$ and capture that there is a change in the outcome variable associated with another factors.
-We can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is the average IKI for word boundaries.
-However when $\text{x}_\text{sentence}$ takes on the value 1, the average IKI for word boundaries $\alpha$ is incremented by a sentence-related slowdown of $\beta$ msecs.
-Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.
-whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
-Therefore, the model will return one $\beta$ per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts $u_\text{participant}$.
-The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$.
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.
-a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:unimodgaus)
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-#### Unimodal log-Gaussian
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.
-typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{N}()$ was replaced by $\text{logN}()$.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:unimodloggaus)
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-#### Unimodal unequal-variance log-Gaussian
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing].
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation $\sigma_{e_\text{location}}^2)$ varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:unimoduv)
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-### Parallel cascading model of writing
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-#### Bimodal log-Gaussian (constrained)
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-1.
-Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter in equation \ref{eq:bimodcon}.
-Note that $\beta$ in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-
-2.
-Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining $\delta$ to be positive, it captures how much longer hesitant IKIs in addition to $\beta$.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:bimodcon)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2\\
-		& 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
-$\theta$ is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$.
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location.
-In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].
-This was achieved by introducing a parameter $\sigma_\text{diff}$.
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-#### Bimodal log-Gaussian (unconstrained)
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:bimoduncon)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-	\text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
+We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts.
+We discussed advantages with this approach in the introduction section.
+These datasets include samples from populations with different writing experience and languages (e.g.
+young / L2 writers, students) and writing tasks (e.g.
+essays, syntheses).
+Datasets from a variety of different writers and writing tasks were deliberately chosen to challange our modelling approach and to test to what extent resulting pausing patterns generalise across writing contexts.
 
@@
 
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
+Six datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
 
@@
 
-### C2L1
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
+TODO: Mark, can you add information for C2L1 and CATO as needed
 
@@
 
-The C2L1 data set comprises data Norwegian 6th graders -- *N* = 126, mean age 11 years 10 months -- published in @ronneberg2022process.
+For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite [@sim07; @torrance201203].
+
+The C2L1 dataset comprises data Norwegian 6th graders -- *N* = 126, mean age 11 years 10 months -- published in @ronneberg2022process.
 The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite [@sim07; @torrance201203].
-
-### CATO
-
-TODO: Mark, can you add information here?
-
-Data are published in @torrance2016adolescent.
+
+The CATO dataset is published in @torrance2016adolescent.
 Norwegian upper secondary students -- *N* = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite [@sim07; @torrance201203].
-
-### GUNNEXP2
 
@@
 In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite [@sim07; @torrance201203].
-
-### LIFT
+
+The SPL2 data set involves undergraduate university students in the US -- *N* = 39, 2mean age = 20.6 years, SD = 1.51 -- composing argumentative essays in English (L1) and Spanish (L2).
+Students wrote within the CyWrite word-processing environment [@chukharev2019combined].
+Participants wrote, with a 40 minute time limit, to two writing prompts.
+Order and writing prompt were counterbalanced across language.
+
+LIFT and PLanTra data data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
 
 The LIFT data are published in @vandermeulen2020 and described in @vandermeulen2020mapping.
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (*N* = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
+LIFT is a representative sample of Dutch students (*N* = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education collected from 43 schools.
+The students were asked to conduct four synthesis writing tasks, two argumentative and two informative texts, about each of four topics, with order randomized per school.
 The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### PLanTra
 
@@
 Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-### SPL2
 
@@
 
-## Data analysis
-
-### Bayesian modelling
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-```{r models, results = 'asis'}
-models <- tibble(Models = paste0("M",1:5),
-       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
-       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
-       Description = c("Gaussian distribution",
-                       "Log-Gaussian distribution",
-                       "Log-Gaussian distribution with unequal variances",
-                       "Fluent transition do not vary by transition location",
-                       "Fluent transition vary by transition location"))
-
-apa_table(models, 
-          align = c(rep("l", 2), "r", "p{9cm}"), 
-          escape = FALSE, 
-          digits = 0,
-          caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
-          note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
-```
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates [@farrell2018computational; @gelman2014; @lee2014bayesian].
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
-existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
-automatic Ockham's razor, @jefferys1992ockham].
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
-@lambert2018student; @mcelreath2016statistical].
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models [@gelman2014; @mcelreath2016statistical].
-The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical; @sivula2020uncertainty].
-
-### Transition types
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g.
-@torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
+## Transition types
+
+The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful [@torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
 In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
@@
 
-### Data reduction
+## Data reduction
 
@@
 From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
+This was done for computational reasons to reduce the time the Bayesian models -- decribed in the next section -- require to complete sampling.
 For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
@@
 
+## Bayesian statitical modelling
+
+We reanalysed keystroke data -- transition durations measured as the time between first and second keypress -- from six datasets in a series of five Bayesian models.
+An overview of all models can be found in Table \ref{tab:models}.
+More detailed model descriptions and motivations can be found in Appendix \ref{statistical-models}.
+
+```{r models, results = 'asis'}
+models <- tibble(Models = paste0("M",1:5),
+       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
+       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
+       Description = c("Standard single Gaussian model with effects of transition location by study-specific manipulations and by-participant random effect",
+                       "Single distribution log-Gaussian model equivialent to M1",
+                       "Equivialent to M2 but with different variance components for each transition location (unequal variance)",
+                       "Two-distributions mixture of a log-Gaussian of fluent intervals and another of hesitant; distribution of hesitant keystroke intervals is equivialent to M3 but distribution of fluent intervals has a mean across all design cell.
+Mixing proportion captures the relative number of disfluent transitions",
+                       "Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location")) %>% 
+  select(Models, Description)
+
+apa_table(models, 
+          align = c(rep("l", 1), "p{13cm}"), 
+          escape = FALSE, 
+          digits = 0,
+          caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.")
+#          note = "LMM = Single distribution mixed-effects models; MoG = two distributions mixture of log-Gaussians") 
+```
+
+Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset specific manipulations and random effects for participants.
+In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
+For model M3 we relaxed the equal variance assumption for transition locations, thus allowing each transition location to assume a different standard deviation (see Appendix \ref{statistical-models} for rationale).
+
+Models M4 and M5 and extensions of the single distribution models.
+Importantly models M4 and M5 assume that keystroke data are the result of a combination of two data generating processes -- consistent with the cascading account -- rather than one.
+These are called finite mixture models in the literature [@gelman2014; @roeser2021modelling; @peel2000finite].
+
+These models extend the assumption that processing at higher levels of activation lead to longer pauses.
+Instead of assuming that there is one process that shifts the distribution of IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
+This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+
+1.
+Activation can flow into keystrokes without interruption.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter illustrated in equation \ref{eq:bimodcon2}.
+
+2.
+Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
+The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon2}.
+
+\begin{equation}
+\begin{aligned}
+(\#eq:bimodcon2)
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
+		& \sigma_{e'}^2 > \sigma_{e}^2\\
+		& 0 < \theta < 1
+\end{aligned}
+\end{equation}
+
+The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
+Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
+$\theta$ is here parameterised to represent the probability that an IKI is associated with the distribution of long IKIs.
+This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$ .
+In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
+The weights of both distributions must sum to 1.
+We call this parameter the probability of hesitant transitions.
+
+The size of a fluent key-transition does not necessarily vary by transition location which is what our constrained model M4 assumes.
+In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions.
+However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
+This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter.
+We therefore also implemented an unconstrained model M5 that assumes that the distribution over durations of fluent transitions -- represented as $\beta$ -- varies across transition locations.
+
+All models were implemented in the Bayesian framework.
+The Bayesian framework, as used in this paper, is ideal for the estimation of parameter values.
+This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates [@farrell2018computational; @gelman2014; @lee2014bayesian].
+To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
+existing knowledge about parameter values.
+For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
+automatic Ockham's razor, @jefferys1992ockham].
+In other words the choice of priors values has less impact on the posterior.
+In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
+@lambert2018student; @mcelreath2016statistical].
+Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
+
+For all models we included transition location (levels: before sentence, before word, within word) as predictor; for detailed descriptions of parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon} in Appendix \ref{statistical-models}.
+Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
+Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.^[Data were analysed in Bayesian mixed effects models [@gelman2014; @mcelreath2016statistical].
+The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
+Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
+Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
+The predictive performance of our models was compared using leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical; @sivula2020uncertainty].]
+
 # Results
 
-## Out-of-samples cross-validation
+## Model comparisons
 
@@
 For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
+Both two-distributions mixture models provided a substantially better fit than any of the single distribution models.
 The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
+Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
+The weakest model was the single distribution Gaussian model.
 
@@
                mogbetaunconstr = "M5"),
-    across(model, ~recode(., lmm = "Unimodal log-Gaussian",
-                          lmmgaus = "Unimodal Gaussian",
-                          mogbetaconstr = "Bimodal (constrained)",
-                          mogbetacontr = "Bimodal (constrained)",
-                          mogbetaunconstr = "Bimodal (unconstrained)",
-                          lmmuneqvar = "Unimodal (unequal variance)")),
+    across(model, ~recode(., lmm = "Single distribution log-Gaussian",
+                          lmmgaus = "Single distribution Gaussian",
+                          mogbetaconstr = "Two distributions mixture of log-Gaussians (constrained)",
+                          mogbetacontr = "Two distributions mixture of log-Gaussians (constrained)",
+                          mogbetaunconstr = "Two distributions mixture of log-Gaussians (unconstrained)",
+                          lmmuneqvar = "Single distribution log-Gaussian (unequal variance)")),
             across(elpd_diff, ~str_c(., " (", se_diff, ")")),
@@
   arrange(dataset)
-```
-
+
+table <- table %>% select(-elpd_loo) %>% 
+  pivot_wider(names_from = dataset, values_from = elpd_diff) %>% 
+  rename(Model = m, Description = model)
+
+```
+
+\blandscape
 ```{r loos, results='asis'}
-names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
-
-apa_table(table[,-1], 
-      caption = "Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.",
-      align =c("p{3.5cm}", "p{6cm}", rep("r", 2)), 
+#names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
+#names(table) <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$"), 1))
+
+apa_table(table, 
+      caption = "Model comparisons. The top row shows the models with the highest predictive performance compared to all other models. Standard error is shown in parentheses.",
+      align =c("p{1.5cm}", "p{5cm}", rep("r", 6)), 
       escape = FALSE,
@@
       font_size = "small",
-      note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      stub_indents = list(
-                     "\\textbf{CL21}" = 1:5,
-                     "\\textbf{CATO}" = 6:10,
-                     "\\textbf{GUNNEXP2}" = 11:15, 
-                     "\\textbf{LIFT}" = 16:20,
-                     "\\textbf{PLanTra}" = 21:25,
-                     "\\textbf{SPL2}" = 26:30))
+      note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density -- relative to the model with the highest predictive performance in the top row.")
+#                col_spanners = list(" " = 1:2,
+#                     "\\textbf{CL21}" = 1:5,
+#                     "\\textbf{CATO}" = 6:10,
+#                     "\\textbf{GUNNEXP2}" = 11:15, 
+#                     "\\textbf{LIFT}" = 16:20,
+#                     "\\textbf{PLanTra}" = 21:25,
+#                     "\\textbf{SPL2}" = 26:30))
+#      stub_indents = list(
+#                     "\\textbf{CL21}" = 1:5,
+#                     "\\textbf{CATO}" = 6:10,
+#                     "\\textbf{GUNNEXP2}" = 11:15, 
+#                     "\\textbf{LIFT}" = 16:20,
+#                    "\\textbf{PLanTra}" = 21:25,
+#                     "\\textbf{SPL2}" = 26:30))
 #                     "\\textbf{SPL2 (shift + C)}" = 31:35))
 
-```
+# $\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density;
+```
+\elandscape
 
@@
 These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-## Cross-data set comparisons
+Model predictions fit the data well in the case of two-distributions mixture models and poorest for the single distribution Gaussian model.
+
+## Transition location effect
+
+As demonstrated in the model comparisons, the two-distributions mixture-model capture the writing-process data better than single-distribution models.
+Therefore, writing process is captured by the posterior estimates of the model parameter values.
+There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as $\beta$ in equation \ref{eq:bimodcon2}), (2) the magnitude of the slowdown for hesitant transitions (represented as $\delta$), and (3) the probability of hesitant transition duration (represented as $\theta$).
+
+Figure \ref{fig:mixmodel} illustrates the mixture model approach: for each transition durations we obtain two posterior distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in longer durations.
+The relative weighting of these two distributions was captured as the probability of hesitant transitions, the parameter that captures the height of the distribution of slow keystrokes.
+
+```{r mixmodel, fig.cap='Mixture model plot for each transition location.
+Shown are the distributions for both mixture components (fluent and hesitant transition durations) and their respective weighting.
+Posterior of SPL2 data (English sessions only).', fig.height=6}
+
+ps <- read_csv("../stanout/spl2/mog_constr_posterior_for_densityplot.csv") 
+# Extract parameters
+alpha <- filter(ps, param == 'beta') %>% pull(est)
+alpha2_bs <- filter(ps, param == 'delta', loc == 'before sentence') %>% pull(est)
+alpha2_bw <- filter(ps, param == 'delta', loc == 'before word') %>% pull(est)
+alpha2_ww <- filter(ps, param == 'delta', loc == 'within word') %>% pull(est)
+
+theta_bs <- filter(ps, param == 'prob',  loc == 'before sentence') %>% pull(est)
+theta_bw <- filter(ps, param == 'prob',  loc == 'before word') %>% pull(est)
+theta_ww <- filter(ps, param == 'prob',  loc == 'within word') %>% pull(est)
+
+sigma <- filter(ps, param == 'sigma') %>% pull(est)
+sigmap_bs <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'before sentence')) %>% pull(est) %>% sum()
+sigmap_bw <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'before word')) %>% pull(est) %>% sum()
+sigmap_ww <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'within word')) %>% pull(est) %>% sum()
+
+# Data grid
+data <- tibble(x = c(0, 20))
+
+colours <- c("Fluent transitions" = "#000000", 
+             #"Fluent transitions" = "#56B4E9",
+             "Hesitant transitions" = "#E69F00")
+
+# Create density plot
+plot_bs <- ggplot(data, aes(x)) +
+  stat_function(aes(colour = "Fluent transitions"),
+                geom = "line", 
+                linewidth = .25,
+                fun = plot_mix_comps,
+                args = list(alpha, sigma, lam = 1 - theta_bs)) +
+  stat_function(aes(colour = "Hesitant transitions"),
+                geom = "line", 
+                linewidth = .25,
+                fun = plot_mix_comps, 
+                args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
+  stat_function(aes(fill = "Fluent transitions"),
+                geom = "area", 
+                fun = plot_mix_comps, 
+                alpha = .25, 
+                args = list(alpha, sigma, lam = 1 - theta_bs)) +
+  stat_function(aes(fill = "Hesitant transitions"),
+                geom = "area", 
+                fun = plot_mix_comps, 
+                alpha = .25, 
+                args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
+  scale_x_continuous(labels = exp, 
+                     breaks = log(c(1, 10, 100, 1000, 10000)), 
+                     limits = log(c(10, 50000))) +
+  scale_colour_manual(values = colours) +
+  scale_fill_manual(values = colours) +
+  labs(y = "", x = "", 
+       subtitle = "Before-sentence transitions",
+       colour = "Mixture\ncomponent",
+       fill = "Mixture\ncomponent") +
+  theme(legend.position = "right",
+        legend.justification = "top")
+
+plot_bw <- ggplot(data, aes(x)) +
+  stat_function(aes(colour = "Fluent transitions"),
+                linewidth = .25,
+                geom = "line", 
+                fun = plot_mix_comps,
+                args = list(alpha, sigma, lam = 1 - theta_bw)) +
+  stat_function(aes(colour = "Hesitant transitions"),
+                geom = "line", 
+                linewidth = .25,
+                fun = plot_mix_comps, 
+                args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
+  stat_function(aes(fill = "Fluent transitions"),
+                geom = "area", 
+                linewidth = .25,
+                fun = plot_mix_comps, 
+                alpha = .25, 
+                args = list(alpha, sigma, lam = 1 - theta_bw)) +
+  stat_function(aes(fill = "Hesitant transitions"),
+                geom = "area", 
+                fun = plot_mix_comps, 
+                alpha = .25, 
+                args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
+  scale_x_continuous(labels = exp, 
+                     breaks = log(c(1, 10, 100, 1000, 10000)), 
+                     limits = log(c(10, 50000))) +
+  scale_colour_manual(values = colours) +
+  scale_fill_manual(values = colours) +
+  labs(y = "Density", x = "", 
+       subtitle = "Before-word transitions",
+       colour = "Mixture\ncomponent",
+       fill = "Mixture\ncomponent") +
+  theme(legend.position = "none")
+
+plot_ww <- ggplot(data, aes(x)) +
+  stat_function(aes(colour = "Fluent transitions"),
+                geom = "line", 
+                fun = plot_mix_comps,
+                linewidth = .25,
+                args = list(alpha, sigma, lam = 1 - theta_ww)) +
+  stat_function(aes(colour = "Hesitant transitions"),
+                geom = "line", 
+                fun = plot_mix_comps, 
+                linewidth = .25,
+                args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
+  stat_function(aes(fill = "Fluent transitions"),
+                geom = "area", 
+                fun = plot_mix_comps, 
+                alpha = .25, 
+                args = list(alpha, sigma, lam = 1 - theta_ww)) +
+  stat_function(aes(fill = "Hesitant transitions"),
+                geom = "area", 
+                fun = plot_mix_comps, 
+                alpha = .25, 
+                args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
+  scale_x_continuous(labels = exp, 
+                     breaks = log(c(1, 10, 100, 1000, 10000)), 
+                     limits = log(c(10, 50000))) +
+  scale_colour_manual(values = colours) +
+  scale_fill_manual(values = colours) +
+  labs(subtitle = "Within-sentence transitions", 
+       y = "", x = "Transition duration [in msecs]",
+       colour = "Mixture\ncomponent",
+       fill = "Mixture\ncomponent") +
+  theme(legend.position = "none")
+
+plot_bs / plot_bw / plot_ww
+
+```
+
+The predictive performance of the unconstrained mixture model is substantially higher than the predictive performance of the constrained mixture model.
+However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
+We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following with reference to the unconstrained implementation.
 
@@
                           c("lift", "spl2", "plantra", "cato", "c2l1", "gunnexp2")), 
-                    pattern = "mogbetaun.+.csv", 
+                    pattern = "mogbetaco.+.csv", 
                     full.names = T)
@@
                 .ordered = TRUE))) %>% 
-  filter(!(str_detect(data, "SPL2") & lang == "ES"),
-         !(str_detect(data, "CATO") & group == "dyslexic"),
-         !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked")) %>% 
+  filter(!(str_detect(data, "SPL2") & lang == "ES" & !is.na(lang)),
+         !(str_detect(data, "CATO") & group == "dyslexic" & !is.na(group)),
+         !(str_detect(data, "CATO") & task == "masked" & !is.na(task)),
+         !(str_detect(data, "GUNNEXP2") & xn == "masked" & !is.na(xn)))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  rename(beta = value) %>%
+  mutate(idx = 1:n(), .by = data) %>% 
+  select(idx, beta, data)
+
+ps <- filter(ps, param != "beta")  %>% 
   pivot_wider(names_from = param, 
               values_from = value) %>% 
-  unnest(cols = beta:theta) %>% 
+  unnest(cols = beta2:theta) %>%
+  mutate(idx = 1:n(), .by = c(data, location, task, xn, genre, topic, lang)) %>% 
+#  summarise(across(idx, list(max = max)), .by = c(location, data, task))
+  left_join(ps_beta, by = c("data", "idx")) %>%
   mutate(across(c(beta, beta2), ~exp(.)),
@@
          across(location, ~str_replace(., " ", "\n"))) %>% 
-  pivot_longer(beta:theta) %>% 
+  pivot_longer(c(beta, beta2, delta, prob)) %>% 
   summarise(across(value, 
@@
             .by = c(data, location, name)) %>% 
-  filter(name %in% c("beta", "delta", "prob"))
-```
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as $\beta$ in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as $\delta$), and (3) the probability of hesitant transition duration (represented as $\theta$).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
+  filter(name %in% c("beta", "beta2", "delta", "prob"))
+```
+
+We present the mixture-model posterior estimates for each of the three parameters in the three facets of Figure \ref{fig:crossstudypost2}.
 Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
 We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.], and removed conditions that might conflate comparisons^[We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.].
+While there was evidence for language-specific effects in Appendix \ref{l2-effect-spl2}, evidence for marking effects was weak as shown in Appendix \ref{masking-effect-cato-gunnexp2}.].
 For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-```{r crossstudypost, fig.cap="Mixture model parameter estimates across studies.
+The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the slowdown duration for hesitant transitions and the hesitation probability.
+Note that the estimate for fluent transitions was estimated by dataset.
+
+```{r crossstudypost2, fig.cap="Mixture model parameter estimates across studies.
 Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability."}
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group."}
 
@@
 nrow <- 1
-beta <- filter(ps, name == "beta") %>% 
-  mutate(name = beta_label) %>% 
+beta <- filter(ps, name == "beta", 
+               location == "before\nword") %>%
+  mutate(name = beta_label,
+         location = "overall") %>% 
   ggplot(aes(y = mean, 
@@
             position = posd) +
-  facet_grid(~name) +
+  facet_grid(~name, scales = "free_x") +
   scale_shape_manual(values = shapes) +
   scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma, breaks = c(50, 150, 200, 300, 400, 500)) +
+  scale_y_continuous(labels = scales::comma) +
   theme(axis.title = element_blank()) +
@@
   scale_colour_colorblind() +
-  scale_y_log10(labels = scales::comma) +
+  scale_y_continuous(labels = scales::comma) +
   labs(colour = grouplabel,
@@
 
-plot <- beta + delta + theta + plot_layout(guides = "collect")
+plot <- beta + delta + theta + plot_layout(guides = "collect",
+                                           widths = c(.5, 1, 1))
 
@@
 In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
+Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for keystroke-interval estimates by transition location across data sets for all three mixture model parameters.
+Full results of these pairwise comparisons can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
+We found that hesitations appear more frequently -- higher hesitation probability -- at before-word transitions than within words across data sets (all BFs > 100).
+Hesitations were more likely at before-sentence locations compared to before-word locations only for 2 out of 6 datasets (GUNNEXP2 and SPL2, both BFs > 100; but C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6).
+Hesitations were longer at before-sentence transitions compared to before-word transitions (all BFs > 100, except for C2L1: BF = 1.11 and LIFT = 0.5); before-word transitions were longer compared to within-word transitions for half of the data sets (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1) but not for the other (C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
+
+To reiterate, for most datasets we found negligible evidence supporting the idea that writers pause more frequently at before-sentence boundaries compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
 This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence $>$ before-word $>$ within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
+In fact, LIFT showed some indication that pauses were less likely before sentences than words.
+In brief, while pauses tend to be longer before sentences and less likely at mid-word locations, it is less clear in which contexts pauses are (1) longer before than within words and (2) more likely to appear before sentences than before words.
+
+The duration of fluent durations was not distinguished by transition location in the constrained version of the two distribution model.
+Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transition durations.
+This is presumably reflecting that the population that this sample is from was the youngest and least experienced writers in our data pool.
+Hesitation duration and frequencies were similar to other data sets.
+
+Inconsistencies between datasets could, to some extent, be explained on the basis that before-sentence transitions were sometimes defined as involving complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
+In particular, before-sentence transitions in some  datasets (PLanTra, LIFT) included the character key following the shift key but others did not (CATO, C2L1, SPL2, GUNNEXP2).
+Appendix \ref{key-combination-effect-constrained-mixture-model} addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increases the hesitation probability (BFs > 100) but it did not affect the duration of fluent transitions or hesitations (BFs < 0.3).
+However this is in conflict with the earlier finding that before-sentence hesitations are not more likely than before-word hesitations for the datasets that included the character following shift (PLanTra, LIFT).
 In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
 
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\blandscape
-```{r loceffect, results='asis'}
-files <- list.files(str_c("../stanout/", 
-c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetaun.+.csv", full.names = T) # "spl2_shift",
-
-ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
-  mutate(data = .x)) %>% 
-  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, ~recode(., 
-                "cato" = "*CATO* (non-dyslexic\nunmasked)",
-                "spl2" = "*SPL2* (L1)", 
-                "gunnexp2" = "*GUNNEXP2* (unmasked)",
-#                "spl2_shift" = "*SPL2* (L1; shift + C)", 
-                "plantra" = "*PLanTra*",
-                "lift" = "*LIFT*",
-                "c2l1" = "*C2L1*"))) %>% 
-  filter(!(str_detect(data, "SPL2") & lang == "ES"),
-         !(str_detect(data, "CATO") & group == "dyslexic"),
-         !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked"),
-         param %in% c("beta", "delta", "theta")) %>%
-  select(param, location, data, value) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
-  pivot_wider(names_from = location, values_from = value) %>%
-  unnest(-c(data, param)) %>% 
-  mutate(diff.1 = `before sentence` - `before word`,
-         diff.2 = `before word` - `within word`) %>% 
-  summarise(across(starts_with("diff"), 
-                   list(mean = mean, 
-                        lower = lower, 
-                        upper = upper, 
-                        BF = BF)),
-            .by = c(data, param))
-
-tmp <- ps_loc_diffs %>% 
-  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
-  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
-         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
-  select(-lower, -upper) %>% 
-  pivot_wider(names_from = param, values_from = c(mean, BF)) %>% 
-  mutate(across(diffid, ~recode(., diff.1 = "before sentence vs word",
-                                   diff.2 = "before vs within word"))) %>% 
-  select(data, diffid, ends_with("beta"), ends_with("delta"), ends_with("theta")) %>% 
-  arrange(data)
-
-write_csv(tmp, "tables/location_effect_alldata.csv")
-
-apa_table(tmp[,-1],
-          align =c("l", rep("r", 6)), 
-          longtable = T,
-          escape = FALSE,
-          row.names = T,
-          font_size = "footnotesize",
-          col.names = c("Comparison", rep(c("Est. [95\\% PIs]", "BF"), 3)),
-          stub_indents = list(`\\textbf{C2L1}` = 1:2, 
-                              `\\textbf{CATO (non-dyslexic, unmasked)}`= 3:4,
-                              `\\textbf{GUNNEXP2 (unmasked)}` = 5:6,
-                              `\\textbf{LIFT}` = 7:8,
-                              `\\textbf{PLanTra}` = 9:10,
-                              `\\textbf{SPL2 (L1)}` = 11:12),
-#                              `\\textbf{SPL2 (L1; shift + C)}` = 13:14),
-          col_spanners = list(" " = c(1),
-                     "Fluent transitions" = c(2, 3), 
-                     "Hesitation slowdown" = c(4, 5), 
-                     "Hesitation probability" = c(6, 7)),
-          note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
-         caption = "Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\\% PIs in brackets.")
-
-```
-\elandscape
+The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
+We, therefore, summarize briefly differences in the results for the transition-location effects.
+Results for the unconstrained version of the mixture model can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
+The main difference between the unconstrained and the constrained model is that the latter does not allow the parameter that captures fluent transition durations to vary by transition location or otherwise.
+Instead it represents the lower bound of keystroke activity for uninhibited activation flow.
+Higher predictive performance for the unconstrained model suggests variability across transition locations for fluent transition durations, in particular for before-word transitions compared to mid-word transitions (all BFs > 100) but not for before-sentence transitions compared to before-word transitions (all BFs < 0.08, but SPL2 and GUNNEXP2: BFs > 100).
+The pattern for the hesitation probability was generally similar in both the constrained and unconstrained model.
+However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs > 10, except for LIFT: BF = 0.7); the hesitation duration pattern for before-sentence transitions compared to before-word transitions remained the same.
+
+<!-- TODO: address difference between posteriors of shift analysis for constrained and unconstrained model 
+- possibly not relevant: there are difference between the posteriors but that's not important for the differences between data sets
+-->
 
@@
 Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
+We used Bayesian statistical models, in particular single-distribution and two-distributions hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
 On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
@@
 Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
+On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
 We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
@@
 
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+A possible concern with our results -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
 We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
+In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
+We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
+In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two-distributions -- and we can test whether these models can successfully uncover the true parameter value.
 Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
+As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
 Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model,  when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
+Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a two-distributions process (but for the two-distributions data).
+Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of model overfitting.
 
@@
 Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to investiatge regressive eye movement behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
+There is a limited amount of work that has used eye tracking to investigate regressive eye movement behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
 There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.
@@
 
+----
+Add to OSF
+Work on tutorial and add as link to methods
+----
+
 ~~~~~
@@
 ```{r render_appendix, include=FALSE}
-render_appendix("appendix_a.Rmd")
-render_appendix("appendix_b.Rmd")
-render_appendix("appendix_c.Rmd")
-render_appendix("appendix_d.Rmd")
-render_appendix("appendix_e.Rmd")
-render_appendix("appendix_g.Rmd")
-render_appendix("appendix_f.Rmd")
-render_appendix("appendix_h.Rmd")
-```
+render_appendix("appendix_models.Rmd")
+render_appendix("appendix_fit2data.Rmd")
+#render_appendix("appendix_postparam.Rmd")
+render_appendix("appendix_postparam_constr.Rmd")
+render_appendix("appendix_shift_constr.Rmd")
+#render_appendix("appendix_l2.Rmd")
+render_appendix("appendix_l2_constr.Rmd")
+#render_appendix("appendix_masking.Rmd")
+render_appendix("appendix_masking_constr.Rmd")
+#render_appendix("appendix_prepost.Rmd")
+render_appendix("appendix_prepost_constr.Rmd")
+#render_appendix("appendix_genre.Rmd")
+render_appendix("appendix_genre_constr.Rmd")
+render_appendix("appendix_transloc.Rmd")
+render_appendix("appendix_transloc_constr.Rmd")
+render_appendix("appendix_shift.Rmd")
+render_appendix("appendix_sim.Rmd")
+
+```

### 2023-12-09 report/manuscript.tex
Reorganised manuscript

@@
   pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
+  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev3},
   pdflang={en-EN},
@@
 \title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
+\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev\textsuperscript{3}}
 \date{}
@@
 \abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
+Classical serial models view the process of producing a text as a descrete chain of pauses and writing bursts.
 In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
@@
 Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
+Our results cannot be explained by the serial but are in line with the cascading view of writing.
 }
@@
 We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
+The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models.
 In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
 
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
+The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in Roeser et al.
 (2021).
@@
 
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
+We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts.
+We discussed advantages with this approach in the introduction section.
+These datasets include samples from populations with different writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses).
+Datasets from a variety of different writers and writing tasks were deliberately chosen to challange our modelling approach and to test to what extent resulting pausing patterns generalise across writing contexts.
 
@@
 
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
+TODO: Mark, can you add information for C2L1 and CATO as needed
 
@@
 
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
+For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
+
+The C2L1 dataset comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
 (2022).
 The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
+
+The CATO dataset is published in Torrance, Rønneberg, et al.
 (2016).
 Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
 
@@
 In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
+
+The SPL2 data set involves undergraduate university students in the US -- \emph{N} = 39, 2mean age = 20.6 years, SD = 1.51 -- composing argumentative essays in English (L1) and Spanish (L2).
+Students wrote within the CyWrite word-processing environment (Chukharev-Hudilainen et al., 2019).
+Participants wrote, with a 40 minute time limit, to two writing prompts.
+Order and writing prompt were counterbalanced across language.
+
+LIFT and PLanTra data data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
 
@@
 (2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
+LIFT is a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education collected from 43 schools.
+The students were asked to conduct four synthesis writing tasks, two argumentative and two informative texts, about each of four topics, with order randomized per school.
 The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
 
@@
 Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
 
@@
 
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
 \hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
+\subsection{Transition types}\label{transition-types}}
+
+The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
 In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
@@
 \hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
+\subsection{Data reduction}\label{data-reduction}}
 
@@
 From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
+This was done for computational reasons to reduce the time the Bayesian models -- decribed in the next section -- require to complete sampling.
 For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
@@
 
+\hypertarget{bayesian-statitical-modelling}{%
+\subsection{Bayesian statitical modelling}\label{bayesian-statitical-modelling}}
+
+We reanalysed keystroke data -- transition durations measured as the time between first and second keypress -- from six datasets in a series of five Bayesian models.
+An overview of all models can be found in Table \ref{tab:models}.
+More detailed model descriptions and motivations can be found in Appendix \ref{statistical-models}.
+
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
+
+\begin{tabular}{lp{13cm}}
+\toprule
+Models & \multicolumn{1}{c}{Description}\\
+\midrule
+M1 & Standard single Gaussian model with effects of transition location by study-specific manipulations and by-participant random effect\\
+M2 & Single distribution log-Gaussian model equivialent to M1\\
+M3 & Equivialent to M2 but with different variance components for each transition location (unequal variance)\\
+M4 & Two-distributions mixture of a log-Gaussian of fluent intervals and another of hesitant; distribution of hesitant keystroke intervals is equivialent to M3 but distribution of fluent intervals has a mean across all design cell.
+Mixing proportion captures the relative number of disfluent transitions\\
+M5 & Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location\\
+\bottomrule
+\end{tabular}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset specific manipulations and random effects for participants.
+In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
+For model M3 we relaxed the equal variance assumption for transition locations, thus allowing each transition location to assume a different standard deviation (see Appendix \ref{statistical-models} for rationale).
+
+Models M4 and M5 and extensions of the single distribution models.
+Importantly models M4 and M5 assume that keystroke data are the result of a combination of two data generating processes -- consistent with the cascading account -- rather than one.
+These are called finite mixture models in the literature (Gelman et al., 2014; MacLahlan \& Peel, 2000; Roeser et al., 2021).
+
+These models extend the assumption that processing at higher levels of activation lead to longer pauses.
+Instead of assuming that there is one process that shifts the distribution of IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
+This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Activation can flow into keystrokes without interruption.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter illustrated in equation \ref{eq:bimodcon2}.
+\item
+  Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
+The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon2}.
+\end{enumerate}
+
+\begin{equation}
+\begin{aligned}
+\label{eq:bimodcon2}
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
+        & \sigma_{e'}^2 > \sigma_{e}^2\\
+        & 0 < \theta < 1
+\end{aligned}
+\end{equation}
+
+The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
+Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
+\(\theta\) is here parameterised to represent the probability that an IKI is associated with the distribution of long IKIs.
+This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\) .
+In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
+The weights of both distributions must sum to 1.
+We call this parameter the probability of hesitant transitions.
+
+The size of a fluent key-transition does not necessarily vary by transition location which is what our constrained model M4 assumes.
+In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
+However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
+This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter.
+We therefore also implemented an unconstrained model M5 that assumes that the distribution over durations of fluent transitions -- represented as \(\beta\) -- varies across transition locations.
+
+All models were implemented in the Bayesian framework.
+The Bayesian framework, as used in this paper, is ideal for the estimation of parameter values.
+This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
+D.
+Lee \& Wagenmakers, 2014).
+To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
+For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
+In other words the choice of priors values has less impact on the posterior.
+In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
+Lambert, 2018; McElreath, 2016).
+Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
+
+For all models we included transition location (levels: before sentence, before word, within word) as predictor; for detailed descriptions of parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon} in Appendix \ref{statistical-models}.
+Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
+Stan code for mixture models was based on Roeser et al.
+(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.\footnote{Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
+The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
+Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
+Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
+The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).}
+
 \hypertarget{results}{%
@@
 
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
+\hypertarget{model-comparisons}{%
+\subsection{Model comparisons}\label{model-comparisons}}
 
@@
 For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
+Both two-distributions mixture models provided a substantially better fit than any of the single distribution models.
 The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
+Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
+The weakest model was the single distribution Gaussian model.
+
+\begin{landscape}
 
@@
 \begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
+\normalsize{\textit{Note.} $\Delta\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density -- relative to the model with the highest predictive performance in the top row.}
 \end{TableNotes}
@@
 
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
+\begin{longtable}{p{1.5cm}p{5cm}rrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:loos}Model comparisons. The top row shows the models with the highest predictive performance compared to all other models. Standard error is shown in parentheses.}\\
 \toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
+Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
 \midrule
@@
 \toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
+Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
 \midrule
 \endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
+M5 & Two distributions mixture of log-Gaussians (unconstrained) & -- & -- & -- & -- & -- & --\\
+M4 & Two distributions mixture of log-Gaussians (constrained) & -546 (51) & -607 (43) & -554 (38) & -500 (37) & -130 (18) & -564 (41)\\
+M3 & Single distribution log-Gaussian (unequal variance) & -2,184 (83) & -2,389 (79) & -2,981 (94) & -5,605 (131) & -2,514 (92) & -1,744 (68)\\
+M2 & Single distribution log-Gaussian & -2,968 (98) & -3,689 (100) & -5,293 (111) & -7,716 (147) & -3,895 (94) & -3,457 (81)\\
+M1 & Single distribution Gaussian & -40,485 (856) & -41,362 (927) & -44,136 (660) & -82,567 (1,691) & -37,718 (572) & -32,174 (450)\\
 \bottomrule
@@
 \end{center}
+\end{landscape}
 
@@
 These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
+Model predictions fit the data well in the case of two-distributions mixture models and poorest for the single distribution Gaussian model.
+
+\hypertarget{transition-location-effect}{%
+\subsection{Transition location effect}\label{transition-location-effect}}
+
+As demonstrated in the model comparisons, the two-distributions mixture-model capture the writing-process data better than single-distribution models.
+Therefore, writing process is captured by the posterior estimates of the model parameter values.
+There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimodcon2}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
+
+Figure \ref{fig:mixmodel} illustrates the mixture model approach: for each transition durations we obtain two posterior distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in longer durations.
+The relative weighting of these two distributions was captured as the probability of hesitant transitions, the parameter that captures the height of the distribution of slow keystrokes.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/mixmodel-1}
+
+}
+
+\caption{Mixture model plot for each transition location.
+Shown are the distributions for both mixture components (fluent and hesitant transition durations) and their respective weighting.
+Posterior of SPL2 data (English sessions only).}\label{fig:mixmodel}
+\end{figure}
+
+The predictive performance of the unconstrained mixture model is substantially higher than the predictive performance of the constrained mixture model.
+However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
+We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following with reference to the unconstrained implementation.
+
+We present the mixture-model posterior estimates for each of the three parameters in the three facets of Figure \ref{fig:crossstudypost2}.
 Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
 We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
+While there was evidence for language-specific effects in Appendix \ref{l2-effect-spl2}, evidence for marking effects was weak as shown in Appendix \ref{masking-effect-cato-gunnexp2}.}.
 For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
+The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the slowdown duration for hesitant transitions and the hesitation probability.
+Note that the estimate for fluent transitions was estimated by dataset.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost2-1}
+
+}
+
+\caption{Mixture model parameter estimates across studies.
+Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.}\label{fig:crossstudypost2}
+\end{figure}
+
+In the following we evaluate differences between transition locations for all three mixture model parameters.
+Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for keystroke-interval estimates by transition location across data sets for all three mixture model parameters.
+Full results of these pairwise comparisons can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
+We found that hesitations appear more frequently -- higher hesitation probability -- at before-word transitions than within words across data sets (all BFs \textgreater{} 100).
+Hesitations were more likely at before-sentence locations compared to before-word locations only for 2 out of 6 datasets (GUNNEXP2 and SPL2, both BFs \textgreater{} 100; but C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6).
+Hesitations were longer at before-sentence transitions compared to before-word transitions (all BFs \textgreater{} 100, except for C2L1: BF = 1.11 and LIFT = 0.5); before-word transitions were longer compared to within-word transitions for half of the data sets (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1) but not for the other (C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
+
+To reiterate, for most datasets we found negligible evidence supporting the idea that writers pause more frequently at before-sentence boundaries compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
+This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
+In fact, LIFT showed some indication that pauses were less likely before sentences than words.
+In brief, while pauses tend to be longer before sentences and less likely at mid-word locations, it is less clear in which contexts pauses are (1) longer before than within words and (2) more likely to appear before sentences than before words.
+
+The duration of fluent durations was not distinguished by transition location in the constrained version of the two distribution model.
+Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transition durations.
+This is presumably reflecting that the population that this sample is from was the youngest and least experienced writers in our data pool.
+Hesitation duration and frequencies were similar to other data sets.
+
+Inconsistencies between datasets could, to some extent, be explained on the basis that before-sentence transitions were sometimes defined as involving complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
+In particular, before-sentence transitions in some datasets (PLanTra, LIFT) included the character key following the shift key but others did not (CATO, C2L1, SPL2, GUNNEXP2).
+Appendix \ref{key-combination-effect-constrained-mixture-model} addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increases the hesitation probability (BFs \textgreater{} 100) but it did not affect the duration of fluent transitions or hesitations (BFs \textless{} 0.3).
+However this is in conflict with the earlier finding that before-sentence hesitations are not more likely than before-word hesitations for the datasets that included the character following shift (PLanTra, LIFT).
+In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
+
+The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
+We, therefore, summarize briefly differences in the results for the transition-location effects.
+Results for the unconstrained version of the mixture model can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
+The main difference between the unconstrained and the constrained model is that the latter does not allow the parameter that captures fluent transition durations to vary by transition location or otherwise.
+Instead it represents the lower bound of keystroke activity for uninhibited activation flow.
+Higher predictive performance for the unconstrained model suggests variability across transition locations for fluent transition durations, in particular for before-word transitions compared to mid-word transitions (all BFs \textgreater{} 100) but not for before-sentence transitions compared to before-word transitions (all BFs \textless{} 0.08, but SPL2 and GUNNEXP2: BFs \textgreater{} 100).
+The pattern for the hesitation probability was generally similar in both the constrained and unconstrained model.
+However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs \textgreater{} 10, except for LIFT: BF = 0.7); the hesitation duration pattern for before-sentence transitions compared to before-word transitions remained the same.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
+Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
+We used Bayesian statistical models, in particular single-distribution and two-distributions hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
+On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
+
+Patters observed in the parameter estimates are generally similar across data sets (with caveats).
+Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
+On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
+We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
+Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
+In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
+However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
+This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
+
+Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
+For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
+This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
+However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
+The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
+Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
+
+We did observe four difference between data sets that are worth highlighting.
+These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
+
+First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
+Yet the overall pausing probability showed a patterns comparable to most other data sets.
+In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
+The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
+\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
+This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
+Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
+(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
+The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
+
+A possible concern with our results -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
+In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
+We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
+In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two-distributions -- and we can test whether these models can successfully uncover the true parameter value.
+Modelling details can be found in Appendix \ref{simulation}.
+As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
+Parameters were not successfully uncovered when we switched model type and data set.
+Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a two-distributions process (but for the two-distributions data).
+Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of model overfitting.
+
+Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
+Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
+There is a limited amount of work that has used eye tracking to investigate regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
+There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
+We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
+However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
+(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
+This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
+Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
+In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
+In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
+
+A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
+Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
+There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
+Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
+Wengelin, 2006) they would be arbitrary.
+For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
+Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
+So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
+However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
+Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
+Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
+Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
+This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+\hypertarget{references}{}
+
+\endgroup
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
+Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
+
+\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
+Ailhaud, E., Chenu, F., \& Jisa, H.
+(2016).
+A developmental perspective on the units of written {F}rench.
+In J.
+Perera, M.
+Aparici, E.
+Rosado, \& N.
+Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
+287--305).
+Springer.
+\url{https://doi.org/10.1007/978-3-319-21136-7_17}
+
+\leavevmode\hypertarget{ref-alamargot2010using}{}%
+Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
+(2010).
+Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
+\emph{Reading and Writing}, \emph{23}(7), 853--888.
+
+\leavevmode\hypertarget{ref-almond2012preliminary}{}%
+Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
+(2012).
+\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
+RR-12-23).
+Educational Testing Service.
+
+\leavevmode\hypertarget{ref-alves2015progress}{}%
+Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
+
+\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
+Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
+
+\leavevmode\hypertarget{ref-beers2010adolescent}{}%
+Beers, S.
+F., Quinlan, T., \& Harbaugh, A.
+G.
+(2010).
+Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
+\emph{Reading and Writing}, \emph{23}(7), 743--775.
+
+\leavevmode\hypertarget{ref-bock2014syntactically}{}%
+Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
+
+\leavevmode\hypertarget{ref-bonin2012evidence}{}%
+Bonin, P., Roux, S., Barry, C., \& Canell, L.
+(2012).
+Evidence for a limited-cascading account of written word naming.
+\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
+\url{https://doi.org/10.1037/a0028471}
+
+\leavevmode\hypertarget{ref-carpenter2016stan}{}%
+Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
+A., Guo, J., Li, P., \& Riddell, A.
+(2016).
+Stan: A probabilistic programming language.
+\emph{Journal of Statistical Software}, \emph{20}.
+
+\leavevmode\hypertarget{ref-chang2006}{}%
+Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
+
+\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
+Chanquoy, L., Foulin, J.-N., \& Fayol, M.
+(1996).
+Writing in adults: {A} real-time approach.
+In G.
+Rijlaarsdam, H.
+Van den Bergh, \& M.
+Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
+36--44).
+Amsterdam University Press.
+
+\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
+Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
+
+\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
+Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
+
+\leavevmode\hypertarget{ref-chenu2014interword}{}%
+Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
+
+\leavevmode\hypertarget{ref-christiansen2016now}{}%
+Christiansen, M.
+H., \& Chater, N.
+(2016).
+The now-or-never bottleneck: A fundamental constraint on language.
+\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
+\url{https://doi.org/10.1017/S0140525X1500031X}
+
+\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
+Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
+
+\leavevmode\hypertarget{ref-chukharev2019combined}{}%
+Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
+(2019).
+Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
+\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
+
+\leavevmode\hypertarget{ref-Clark2002}{}%
+Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
+
+\leavevmode\hypertarget{ref-conijn2020keys}{}%
+Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
+
+\leavevmode\hypertarget{ref-conijn2019understanding}{}%
+Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
+
+\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
+Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
+
+\leavevmode\hypertarget{ref-connelly2012predicting}{}%
+Connelly, V., Dockrell, J.
+E., Walter, K., \& Critten, S.
+(2012).
+Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
+\emph{Written Communication}, \emph{29}(3), 278--302.
+
+\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
+Crump, M.
+J.
+C., \& Logan, G.
+D.
+(2010).
+Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
+\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
+\url{https://doi.org/10.1037/a0020696}
+
+\leavevmode\hypertarget{ref-dam09}{}%
+Damian, M.
+F., \& Stadthagen-Gonzalez, H.
+(2009).
+Advance planning of form properties in the written production of single and multiple words.
+\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
+
+\leavevmode\hypertarget{ref-de2018exploring}{}%
+De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
+
+\leavevmode\hypertarget{ref-dell2007case}{}%
+Dell, G.
+S., Martin, N., \& Schwartz, M.
+F.
+(2007).
+A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
+\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
+
+\leavevmode\hypertarget{ref-dux2021effect}{}%
+Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
+
+\leavevmode\hypertarget{ref-dux2022automating}{}%
+Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
+(2022).
+Automating individualized, process-focused writing instruction: A design-based research study.
+\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
+\url{https://doi.org/0.3389/fcomm.2022.933878}
+
+\leavevmode\hypertarget{ref-farrell2018computational}{}%
+Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
+
+\leavevmode\hypertarget{ref-flower1980ynamics}{}%
+Flower, L.
+S., \& Hayes, J.
+R.
+(1980).
+The dynamics of composing: Making plans and juggling constraints.
+In L.
+W.
+Gregg \& E.
+R.
+Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
+31--50).
+Lawrence Erlbaum.
+
+\leavevmode\hypertarget{ref-flower1981pregnant}{}%
+Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
+
+\leavevmode\hypertarget{ref-foulin1998extent}{}%
+Foulin, J.-N.
+(1998).
+To what extent does pause location predict pause duration in adults' and children's writing?
+\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
+
+\leavevmode\hypertarget{ref-gelman2014}{}%
+Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
+
+\leavevmode\hypertarget{ref-gelman1992}{}%
+Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
+
+\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
+Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
+
+\leavevmode\hypertarget{ref-gri01}{}%
+Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
+
+\leavevmode\hypertarget{ref-griffin2003reversed}{}%
+Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
+
+\leavevmode\hypertarget{ref-guo2018modeling}{}%
+Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
+
+\leavevmode\hypertarget{ref-hall2022constructing}{}%
+Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
+
+\leavevmode\hypertarget{ref-hayes2012evidence}{}%
+Hayes, J.
+R.
+(2012).
+Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
+In M.
+Fayol, D.
+Alamargot, \& V.
+Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
+15--25).
+Psychology Press.
+
+\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
+Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
+
+\leavevmode\hypertarget{ref-kaufer1986composing}{}%
+Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
+
+\leavevmode\hypertarget{ref-lambert2018student}{}%
+Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
+
+\leavevmode\hypertarget{ref-lee13}{}%
+Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
+
+\leavevmode\hypertarget{ref-lee2014bayesian}{}%
+Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
+
+\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
+Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
+
+\leavevmode\hypertarget{ref-li2021identifying}{}%
+Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
+
+\leavevmode\hypertarget{ref-peel2000finite}{}%
+MacLahlan, G., \& Peel, D. (2000). \emph{Finite mixture models}. John \& Sons.
+
+\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
+Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
+
+\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
+McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
+
+\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
+Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
+
+\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
+Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
+
+\leavevmode\hypertarget{ref-meyer1996}{}%
+Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
+
+\leavevmode\hypertarget{ref-mohsen2021second}{}%
+Mohsen, M.
+(2021).
+Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
+\emph{Psycholinguistics}, \emph{30}(1), 180--202.
+\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
+
+\leavevmode\hypertarget{ref-mohsen2021second}{}%
+Mohsen, M.
+(2021).
+Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
+\emph{Psycholinguistics}, \emph{30}(1), 180--202.
+\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
+
+\leavevmode\hypertarget{ref-not10}{}%
+Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
+
+\leavevmode\hypertarget{ref-olive2014toward}{}%
+Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
+
+\leavevmode\hypertarget{ref-R-base}{}%
+R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
+
+\leavevmode\hypertarget{ref-rochester1973significance}{}%
+Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
+
+\leavevmode\hypertarget{ref-roeser2021modelling}{}%
+Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
+(2021).
+Modelling typing disfluencies as finite mixture process.
+\emph{Reading and Writing}, 1--26.
+\url{https://doi.org/10.1007/s11145-021-10203-z}
+
+\leavevmode\hypertarget{ref-roeser2018advance}{}%
+Roeser, J., Torrance, M., \& Baguley, T.
+(2019).
+Advance planning in written and spoken sentence production.
+\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
+\url{https://doi.org/10.1037/xlm0000685}
+
+\leavevmode\hypertarget{ref-rossetti2022s}{}%
+Rossetti, A., \& Van Waes, L.
+(2022a).
+It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
+\emph{Frontiers in Artificial Intelligence}, \emph{5}.
+\url{https://doi.org/10.3389/frai.2022.983008}
+
+\leavevmode\hypertarget{ref-rossetti2022text}{}%
+Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
+
+\leavevmode\hypertarget{ref-roux2013interaction}{}%
+Roux, S., McKeeff, T.
+J., Grosjacques, G., Afonso, O., \& Kandel, S.
+(2013).
+The interaction between central and peripheral processes in handwriting production.
+\emph{Cognition}, \emph{127}(2), 235--241.
+
+\leavevmode\hypertarget{ref-ronneberg2022process}{}%
+Rønneberg, V., Torrance, M., Uppstad, P.
+H., \& Johansson, C.
+(2022).
+The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
+\emph{Psychological Research}, \emph{86}(7), 2239--2255.
+
+\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
+Schilperoord, J.
+(2002).
+On the cognitive status of pauses in discourse production.
+In T.
+Olive \& C.
+M.
+Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
+61--87).
+Springer.
+
+\leavevmode\hypertarget{ref-schoner2002timing}{}%
+Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
+
+\leavevmode\hypertarget{ref-sim07}{}%
+Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
+
+\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
+Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
+
+\leavevmode\hypertarget{ref-rstan}{}%
+Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
+
+\leavevmode\hypertarget{ref-sullivan2002self}{}%
+Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
+
+\leavevmode\hypertarget{ref-torrance201203}{}%
+Torrance, M.
+(2012).
+EyeWrite -- a tool for recording writers' eye movements.
+In M.
+Torrance, D.
+Alamargot, M.
+Castelló, F.
+Ganier, O.
+Kruse, A.
+Mangen, L.
+Tolchinsky, \& L.
+V.
+Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
+355--357).
+Brill.
+\url{https://doi.org/doi.org/10.1163/9781780529295_082}
+
+\leavevmode\hypertarget{ref-torrance2016reading}{}%
+Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
+(2016).
+Reading during the composition of multi-sentence texts: An eye-movement study.
+\emph{Psychological Research}, \emph{80}(5), 729--743.
+
+\leavevmode\hypertarget{ref-torrance2012written}{}%
+Torrance, M., \& Nottbusch, G.
+(2012).
+Written production of single words and simple sentences.
+In V.
+W.
+Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
+403--422).
+Psychology Press.
+
+\leavevmode\hypertarget{ref-torrance2018timed}{}%
+Torrance, M., Nottbusch, G., Alves, R.
+A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
+I., \& others.
+(2018).
+Timed written picture naming in 14 {E}uropean languages.
+\emph{Behavior Research Methods}, \emph{50}, 744--758.
+
+\leavevmode\hypertarget{ref-torranceb}{}%
+Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
+
+\leavevmode\hypertarget{ref-torrancea}{}%
+Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
+
+\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
+Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
+H.
+(2016).
+Adolescent weak decoders writing in a shallow orthography: Process and product.
+\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
+
+\leavevmode\hypertarget{ref-van1991handwriting}{}%
+Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
+
+\leavevmode\hypertarget{ref-van2016keystroke}{}%
+Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
+
+\leavevmode\hypertarget{ref-van2019multilingual}{}%
+Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
+(2019).
+A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
+\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
+
+\leavevmode\hypertarget{ref-van2010reading}{}%
+Van Waes, L., Leijten, M., \& Quinlan, T.
+(2010).
+Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
+\emph{Reading and Writing}, \emph{23}(7), 803--834.
+
+\leavevmode\hypertarget{ref-waes2019}{}%
+Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
+(2021).
+Measuring and assessing typing skills in writing research.
+\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
+\url{https://doi.org/10.17239/jowr-2021.13.01.04}
+
+\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
+Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
+(2020).
+Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
+\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
+
+\leavevmode\hypertarget{ref-vandermeulen2020}{}%
+Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
+
+\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
+Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
+(2023).
+Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
+\emph{Written Communication}, \emph{40}(1), 90--144.
+
+\leavevmode\hypertarget{ref-vasishth2017}{}%
+Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
+(2017).
+Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
+\emph{ArXiv e-Prints}.
+
+\leavevmode\hypertarget{ref-vasishth2017feature}{}%
+Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
+
+\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
+Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
+
+\leavevmode\hypertarget{ref-vehtari2017practical}{}%
+Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
+
+\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
+Wagenmakers, E.-J., \& Brown, S.
+(2007).
+On the linear relation between the mean and the standard deviation of a response time distribution.
+\emph{Psychological Review}, \emph{114}(3), 830--841.
+\url{https://doi.org/10.1037/0033-295X.114.3.830}
+
+\leavevmode\hypertarget{ref-wen02}{}%
+Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
+
+\leavevmode\hypertarget{ref-wen06}{}%
+Wengelin, Å.
+(2006).
+Examining pauses in writing: {T}heory, methods and empirical data.
+In K.
+P.
+H.
+Sullivan \& E.
+Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
+18, pp.
+107--130).
+Elsevier.
+
+\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
+Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
+
+\leavevmode\hypertarget{ref-wheeldon2013}{}%
+Wheeldon, L.
+R., Ohlson, N., Ashby, A., \& Gator, S.
+(2013).
+Lexical availability and grammatical encoding scope during spoken sentence production.
+\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
+
+\leavevmode\hypertarget{ref-wing1973response}{}%
+Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
+
+\end{CSLReferences}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{statistical-models}{%
+\subsection{Statistical models}\label{statistical-models}}
+
+We are using the Bayesian framework (e.g. Farrell \& Lewandowsky, 2018;
+Gelman et al., 2014; Lee \& Wagenmakers, 2014) to implement 5
+statistical models of writing and to evaluate which -- the serial or the
+cascading view -- captures more successfully the data that arise during
+keyboard typing. In other words, we use statistical models to map
+between the data and the theoretically assumed process that generates
+the data to then compare the predictive power of those models. The
+models presented in this section build on one another so that later
+model include assumptions of earlier models.
+
+\hypertarget{serial-model-of-writing}{%
+\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
+
+\hypertarget{single-distribution-gaussian}{%
+\paragraph{Single distribution
+Gaussian}\label{single-distribution-gaussian}}
+
+Under the serial view, all planning must be completed priors to the
+production onset of the corresponding planning unit. The resulting IKI
+is sometimes faster or slower depending on, among others,
+psycholinguistic factors. For example, the interval before a word is
+shorter for easily retrievable high-frequency words, or longer for low
+frequency words, shorter for words with fewer graphemes, syllables, and
+morphemes. There are word-specific factors that influence the IKI that
+precedes a word but these are beyond the scope of our analysis. We will
+capture the variability associated with word-features by assuming that
+before-word IKIs can be described as coming from a distribution that is
+normal (Gaussian) with two parameters, an unknown mean \(\mu\), that
+describes the average IKI associated with word-level planning, and a
+standard deviation \(\sigma_\text{e}^2\), that captures the variability
+associated with factors that we did not further specify in the model.
+This can be expressed as
+\(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
+Of importance is the estimated posterior value of \(\mu\) as this value
+captures that time it takes to mentally plan a word.
+
+We can extend this simple model of word-planning to other linguistic
+location. We introduce earlier that larger linguistic edges are
+associated with planning on higher levels. For example, at sentence
+boundaries, planning needs to happen for word-level properties -- which
+was captured above as average IKI \(\mu\) -- but also higher linguistic
+planning such as clause-level meaning, and dependencies of the
+sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
+Regression formulas can, and commonly are, used to decompose \(\mu\) and
+capture that there is a change in the outcome variable associated with
+another factors. We can decompose \(\mu\) as
+\(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that
+when \(\text{x}_\text{sentence}\) takes on the value 0, the equation
+reduces to \(\mu = \alpha\) which is the average IKI for word
+boundaries. However when \(\text{x}_\text{sentence}\) takes on the value
+1, the average IKI for word boundaries \(\alpha\) is incremented by a
+sentence-related slowdown of \(\beta\) msecs. Therefore the value of the
+\(\beta\) parameter represents the additional cognitive demand
+associated with sentence-initial planning. The application of such a
+statistical model to the data will then provide us with an estimate of
+the parameter value that can be used for statistical inference
+(e.g.~whether there is a statistically meaningful difference between IKI
+associated with words and sentences).
+
+For computational ease, we implemented the differences associated with
+transition locations as \(\beta_\text{location}\) in equation
+\ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for
+each transition location. Therefore, the model will return one \(\beta\)
+per transition location that capture the posterior distribution of
+average IKIs. Also the decomposition of \(\mu\) allows us to address the
+fact that some writers are faster than other writers by introducing a
+parameter for what is typically called random intercepts
+\(u_\text{participant}\). The random intercepts term
+\(u_\text{participant}\) is constrained so that the value it takes on
+comes from a normal distribution with a mean of 0 and a standard
+deviation of \(\sigma_\text{p}^2\). This value is therefore the
+difference between the overall posterior estimate and estimate average
+IKI of a particular participant (i.e.~a positive value indicates that a
+participant is slower than average, a negative value indicates that a
+participant is faster than average).
+
+The final model is shown in equation \ref{eq:unimodgaus} and represents
+a Gaussian mixed effects model.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimodgaus}
+\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+Note aside that standard-deviation parameters were constrained to be
+positive because standard deviations can never be negative.
+
+\hypertarget{single-distribution-log-gaussian}{%
+\paragraph{Single distribution
+log-Gaussian}\label{single-distribution-log-gaussian}}
+
+The previous model assumes that the data-generating process is a
+Gaussian distribution. The next model is largely identical to the
+previous model but instead of assuming a Gaussian, we assume that the
+data come from a log-normal (log-Gaussian) distribution. There are, at
+least, two arguments for using a log-Gaussian distribution: (1)
+log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian
+does not allow negative values. IKIs, as the distance between two
+subsequent key-down events, must be positive. The lower bound is
+delimited by a person's ability to move their fingers and keyboard
+polling. (2) the log-scale is known to be a better match for data from
+human behaviour and motor responses. In particular, a Gaussian
+distribution assumes that units are scaled linearly. For example, a
+difference of 25 msecs is the same between 100 and 125 msecs as between
+5 secs and 5,025 msecs. This does not necessarily map onto the
+psychological interpretation for short and long keyintervals. For
+example effects that result on the motor level within words (e.g.~typing
+a high vs a low frequency bigram) are smaller than differences that are
+due to high levels of processing (retrieving a word in an L1 or L2). In
+other words, although an effect of 25 msecs is large in the context
+overall fast keyintervals, it is small in the context of overall slow
+intervals. Log-Gaussian distributions are a natural way of translating a
+linear scale to an exponential scale so that a 25 msecs difference on
+the lower end of the IKI scale (motor activity) is more meaningful than
+a 25 msecs difference on the upper end of the IKI scale (retrieving
+words, planning sentences).
+
+The model can be described as in equation \ref{eq:unimodloggaus} in
+which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimodloggaus}
+\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
+\text{where: } &
+\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+\hypertarget{single-distribution-unequal-variance-log-gaussian}{%
+\paragraph{Single distribution unequal-variance
+log-Gaussian}\label{single-distribution-unequal-variance-log-gaussian}}
+
+The third model representing the serial view of writing is a single
+distribution unequal-variance model that, except for the
+unequal-variance assumption is identical to the model presented in the
+previous section. The previous models modelled IKIs as a function of
+transition location so that the estimated average IKI depends on the
+position of an IKI in the text. The variance associated with the
+estimated IKIs for each transition location was assumed to be identical
+(equal variance). This assumption is however not in line with what we
+know about data from human behaviour. Longer latencies are known to be
+associated with a larger variances for both response-time data in
+particular (Wagenmakers \& Brown, 2007) and human motor behaviour in
+general (Schöner, 2002; Wing \& Kristofferson, 1973). For IKIs pauses at
+larger linguistic edges are plausibly associated with a larger variance
+because of the larger number of associated processes. Therefore, in
+equation \ref{eq:unimoduv}, we introduce the assumption that standard
+deviation \(\sigma_{e_\text{location}}^2)\) varies by transition
+location.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimoduv}
+\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+ & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+ \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+\hypertarget{parallel-cascading-model-of-writing}{%
+\subsubsection{Parallel cascading model of
+writing}\label{parallel-cascading-model-of-writing}}
+
+The following two models are are extensions of the models introduced for
+the serial view. Crucially, the cascading view allows planning to happen
+in parallel to production. Therefore, we will reduce the constrain of
+the serial models that requires all planning to be completed before
+writing onset. This is done by assuming that IKIs come from a weighted
+combination of two distributions.
+
+\hypertarget{two-distributions-log-gaussian-mixture-model-constrained}{%
+\paragraph{Two-distributions log-Gaussian mixture model
+(constrained)}\label{two-distributions-log-gaussian-mixture-model-constrained}}
+
+This model extends the assumption of the previous model that processing
+that involves higher levels of activation lead to longer pauses. Instead
+of assuming that there is one process that is shifted for IKIs of larger
+linguistic edges, we introduce the assumption that pauses at larger
+linguistic edges are more likely but not obligatory. This is achieved by
+modelling IKIs as coming from a weighted mixture of two distributions
+associated with two different states:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Activation can flow into keystrokes without interruption. These fluent
+  keystroke transitions are merely constrained by a person's ability to
+  move their finger and will be captured by the \(\beta\) parameter in
+  equation \ref{eq:bimodcon}. Note that \(\beta\) is represented in both
+  log-Gaussian distributions in equation \ref{eq:bimodcon} referring to
+  the same unknown parameter.
+\item
+  Interruptions in the activation flow from higher to lower levels
+  result in longer keystroke intervals, when words or their spelling
+  could not be retrieved in time. The slowdown for these hesitant
+  transitions is captured by \(\delta\) in the first line of equation
+  \ref{eq:bimodcon}. The magnitude of the slowdown is associated with
+  transition location. This is because delays at larger linguistic units
+  are likely to be associated with higher level planning. By
+  constraining \(\delta\) to be positive, it captures how much longer
+  hesitant IKIs in addition to \(\beta\).
+\end{enumerate}
+
+\begin{equation}
+\begin{aligned}
+\label{eq:bimodcon}
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
+        & \sigma_{e'}^2 > \sigma_{e}^2\\
+        & 0 < \theta < 1
+\end{aligned}
+\end{equation}
+
+The first line of equation \ref{eq:bimodcon} represents the distribution
+of hesitant key transitions, and the second line represents fluent key
+transitions. Each of these two distributions is associated with the
+mixing weight \(\theta\) which is a proportion that is constrained to be
+larger than 0 and smaller than 1. \(\theta\) is here parameterised to
+represent the probability that an IKI is associated with the
+distribution of long IKIs. This probability is inversely related to the
+mixing weight of the distribution of short IKIs by \(1-\theta\) . In
+other words, a larger weight of one distribution inevitably means a
+lower weight for the other distribution. The weights of both
+distributions must sum to 1. We will call this parameter the probability
+of hesitant transitions.
+
+The probability of hesitant transitions is assumed to vary as a function
+of both transition location and participants. In line with the
+literature discussed in the introduction, we assume that pauses are more
+likely at larger linguistic edges. As pausing is subject to individual
+differences and writing style (skills), we also assume that some
+participants pause more at certain transition locations and other
+participants pause less. This is akin to what is known as a random
+by-participant slopes model.
+
+Lastly, we carried over the unequal variance assumption and let the
+standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by
+transition location. In addition we constrained the variances so that
+\(\sigma_{e'}\) associated with the distribution of typing disfluencies
+is larger than the variance associated with fluent transitions
+\(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et
+al., 2017). This was achieved by introducing a parameter
+\(\sigma_\text{diff}\). The consequence is that fluent transitions are
+assumed to come from a narrower distribution than hesitant transitions.
+
+\hypertarget{two-distributions-log-gaussian-mixture-model-unconstrained}{%
+\paragraph{Two-distributions log-Gaussian mixture model
+(unconstrained)}\label{two-distributions-log-gaussian-mixture-model-unconstrained}}
+
+The size of a fluent key-transition does not necessarily vary by
+transition location. In other words, the parameter \(\beta\) is the same
+for before-sentence, before-word, and within-word transitions. This is
+what we called a constrained model. However, letter bigrams and trigrams
+may be systematically executed faster than transitions between between
+space and a letter (REFERENCE?) or complex keystrokes that comprise
+space and shift-letter combinations for upper case characters before
+sentences (we will address the latter possibility in the results
+section). This is because bigrams / trigrams might be stored, retrieved
+and executed as motor codes but not transitions to a space following a
+word or between a space or shift key press preceding a letter. Also,
+because of the necessarily larger number of within-word transitions, as
+opposed to before-word and sentence transitions, the posterior of the
+constrained model is dominated by within-word transition data. We
+therefore also implemented an unconstrained model that assumes that the
+size of fluent transitions varies across transition locations.
+
+In this model we assume that \(\beta\) varies by transition-location as
+illustrated in equation \ref{eq:bimoduncon}.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:bimoduncon}
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
+        & \sigma_{e'}^2 > \sigma_{e}^2\\
+        & 0 < \theta < 1
+\end{aligned}
+\end{equation}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-farrell2018computational}{}%
+Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of
+cognition and behavior}. Cambridge University Press.
+
+\leavevmode\hypertarget{ref-gelman2014}{}%
+Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \&
+Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman;
+Hall/CRC.
+
+\leavevmode\hypertarget{ref-lee2014bayesian}{}%
+Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive
+modeling: A practical course}. Cambridge University Press.
+
+\leavevmode\hypertarget{ref-not07}{}%
+Nottbusch, G., Weingarten, R., \& Sahel, S. (2007). From written word to
+written sentence production. In M. Torrance, L. van Waes, \& D. W.
+Galbraith (Eds.), \emph{Writing and cognition: Research and
+applications} (Vol. 20, pp. 31--53). Elsevier.
+
+\leavevmode\hypertarget{ref-roeser2018advance}{}%
+Roeser, J., Torrance, M., \& Baguley, T. (2019). Advance planning in
+written and spoken sentence production. \emph{Journal of Experimental
+Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
+\url{https://doi.org/10.1037/xlm0000685}
+
+\leavevmode\hypertarget{ref-schoner2002timing}{}%
+Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain
+and Cognition}, \emph{48}(1), 31--51.
+
+\leavevmode\hypertarget{ref-vasishth2017}{}%
+Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B. (2017). Modelling
+dependency completion in sentence comprehension as a {B}ayesian
+hierarchical mixture process: {A} case study involving {C}hinese
+relative clauses. \emph{ArXiv e-Prints}.
+
+\leavevmode\hypertarget{ref-vasishth2017feature}{}%
+Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting
+as a finite mixture process: Evidence from comprehension data.
+\emph{arXiv Preprint arXiv:1703.04081}.
+
+\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
+Wagenmakers, E.-J., \& Brown, S. (2007). On the linear relation between
+the mean and the standard deviation of a response time distribution.
+\emph{Psychological Review}, \emph{114}(3), 830--841.
+\url{https://doi.org/10.1037/0033-295X.114.3.830}
+
+\leavevmode\hypertarget{ref-wing1973response}{}%
+Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the
+timing of discrete motor responses. \emph{Perception \& Psychophysics},
+\emph{14}(1), 5--12.
+
+\end{CSLReferences}
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{fit-to-data}{%
+\subsection{Fit to data}\label{fit-to-data}}
+
+Model fit was assessed by testing to what extent simulations obtained
+from the model fit the observed data. From each model we obtained 100
+simulations (predictions). These can be found in Figure
+\ref{fig:prediction}. The model predictions fitted the data
+substantially better for the mixture models. The worse fit was observed
+for the single distribution Gaussian model which predicted large amounts
+of negative keystroke intervals which are implausible. Using a
+log-normal distribution removed these negative keystroke intervals.
+
+\begin{figure}
+
+{\centering \includegraphics{figures/fitplots}
+
+}
+
+\caption{Fit to data with models (distribution type) above the plots.
+Comparison of 50 simulations to observed data summarised by model.
+For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
+\end{figure}
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{posterior-parameter-estimates}{%
+\subsection{Posterior parameter
+estimates}\label{posterior-parameter-estimates}}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics{figures/psplots1constr.pdf}
+\caption{Posterior parameter distribution}
+\end{figure}
+\newpage
+\begin{figure}[!htb]
+\ContinuedFloat
+\captionsetup{list=off,format=cont}
+\centering
+\includegraphics{figures/psplots2constr.pdf}
+\label{fig:fullps1}
+\caption{Posterior parameter distribution}
+\end{figure}
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{key-combination-effect-constrained-mixture-model}{%
+\subsection{Key-combination effect (constrained mixture
+model)}\label{key-combination-effect-constrained-mixture-model}}
+
+The analysed datasets differ to the extent that keystroke intervals at
+before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
+C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+In other words, the pause before sentences summed across two key
+intervals in the PLanTra and LIFT data, namely
+\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
+namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
+longer or more frequent pauses at before-sentence locations compared to
+before-word locations can be explained without reference to linguistic
+edges.
+
+Therefore we compared whether the different patterns can be explain on
+the basis of the additional keystroke involved in before-sentence
+transitions. We compared the SPL2 data including and excluding the
+keystroke after shift. Although we modelled all transition locations, we
+present only before-sentence transitions below as there was, as one
+would expect, no difference at word locations. The results of this
+comparison can be found in Table \ref{tab:shiftcellmeans}.
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions immediately preceding a sentence.
+Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions.
+The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\% PIs in brackets.}\\
+\toprule
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
+\toprule
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Fluent transitions &  &  &  & \\
+\ \ \ overall & 156 [145, 169] & 159 [147, 172] & -0.02 [-0.13, 0.09] & 0.06\\
+Hesitation duration &  &  &  & \\
+\ \ \ L1 & 1,167 [1,037, 1,305] & 1,281 [1,082, 1,496] & -0.06 [-0.21, 0.09] & 0.11\\
+\ \ \ L2 & 2,019 [1,772, 2,292] & 1,802 [1,557, 2,073] & 0.12 [-0.02, 0.27] & 0.29\\
+Hesitation probability &  &  &  & \\
+\ \ \ L1 & 1.00 [.99, 1.00] & .78 [.69, .86] & 4.91 [3.68, 6.35] & > 100\\
+\ \ \ L2 & 1.00 [.99, 1.00] & .92 [.87, .96] & 3.4 [2.12, 4.85] & > 100\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+
+The fluent transition duration and hesitation duration were not affected
+by whether or not the sentence-initial transition include the character
+following shift for neither language. However, we found strong evidence
+for an increased hesitation probability -- in both -- languages when the
+before-sentence key transition included the character following the
+shift key. Notably, the hesitation probability increased ceiling. It is
+not entirely surprised that the constrained mixture model identifies
+essentially all before-sentence transition as pauses: the constrained
+model does not distinguish between transition locations for the duration
+of fluent transitions. Therefore essentially all before-sentence
+transitions are identified as hesitation as they always include two
+keystrokes while all other transition location (which is the majority of
+the data) do not.
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{l2-effect-spl2}{%
+\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
+
+For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
+sentence-transitions) we calculated the L2 effect (i.e.~the difference
+between writing in L2 and L1). The results can be found in Table
+\ref{tab:l2effect}. The results show longer hesitation duration and more
+pauses across all transition locations when writing in L2. Only
+within-word transitions did show negligible evidence for an increased
+hesitation duration in L2.
+
+\begin{table}[b]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:l2effect}Mixture model estimates for language effect.
+Cell means are shown the slowdown for hesitant transitions when writing in L1 and L2 and the probability of hesitant transitions.
+The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\% PIs in brackets.}
+
+\small{
+
+\begin{tabular}{lrrrr}
+\toprule
+Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+Hesitation duration &  &  &  & \\
+\ \ \ before sentence & 1,281 [1,082, 1,496] & 1,802 [1,557, 2,073] & 0.31 [0.17, 0.45] & > 100\\
+\ \ \ before word & 289 [254, 327] & 560 [507, 617] & 0.47 [0.39, 0.55] & > 100\\
+\ \ \ within word & 120 [53, 222] & 202 [156, 258] & 0.27 [-0.07, 0.55] & 0.68\\
+Hesitation probability &  &  &  & \\
+\ \ \ before sentence & .78 [.69, .86] & .92 [.87, .96] & 1.23 [0.52, 1.97] & > 100\\
+\ \ \ before word & .51 [.41, .62] & .89 [.83, .93] & 2.02 [1.4, 2.66] & > 100\\
+\ \ \ within word & .04 [.02, .07] & .15 [.10, .22] & 1.52 [0.78, 2.29] & > 100\\
+\bottomrule
+\addlinespace
+\end{tabular}
+
+}
+
+\begin{tablenotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{tablenotes}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{masking-effect-cato-gunnexp2}{%
+\subsection{Masking effect (CATO,
+GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
+
+Studies associated with two datasets (CATO, GUNNEXP2) investigated to
+what extent masking the previously written text affects keystroke
+behaviour. The mixture model results for the effect of masking is shown
+in Table \ref{tab:maskingeffect}. There was some evidence for longer
+before-sentence hesitation durations when the text was masked in
+non-dyslexic writers. Evidence for all other comparisons was negligible.
+
+\blandscape
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
+Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of hesitant transitions.
+The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\% PIs in brackets.}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Hesitation duration &  &  &  &  &  & \\
+\ \ \ before sentence & CATO & dyslexic & 2,231 [1,543, 3,111] & 1,680 [1,239, 2,205] & -0.25 [-0.64, 0.13] & 0.44\\
+\ \ \ before sentence & CATO & non dyslexic & 1,733 [1,241, 2,325] & 1,502 [1,134, 1,942] & -0.12 [-0.47, 0.22] & 0.23\\
+\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,124 [954, 1,307] & 1,418 [1,218, 1,634] & 0.21 [0.05, 0.36] & 2.59\\
+\ \ \ before word & CATO & dyslexic & 444 [405, 485] & 407 [369, 447] & -0.06 [-0.13, 0.01] & 0.14\\
+\ \ \ before word & CATO & non dyslexic & 345 [312, 381] & 326 [295, 361] & -0.04 [-0.11, 0.04] & 0.06\\
+\ \ \ before word & GUNNEXP2 & non dyslexic & 363 [325, 403] & 379 [341, 420] & 0.03 [-0.05, 0.11] & 0.05\\
+\ \ \ within word & CATO & dyslexic & 278 [166, 400] & 177 [116, 252] & -0.24 [-0.52, 0.06] & 0.57\\
+\ \ \ within word & CATO & non dyslexic & 409 [244, 593] & 392 [230, 591] & -0.03 [-0.42, 0.38] & 0.2\\
+\ \ \ within word & GUNNEXP2 & non dyslexic & 182 [105, 279] & 209 [135, 306] & 0.08 [-0.24, 0.4] & 0.19\\
+Hesitattion probability &  &  &  &  &  & \\
+\ \ \ before sentence & CATO & dyslexic & .60 [.38, .79] & .80 [.60, .93] & 0.2 [-0.06, 0.46] & 0.45\\
+\ \ \ before sentence & CATO & non dyslexic & .55 [.34, .76] & .68 [.47, .84] & 0.13 [-0.16, 0.4] & 0.21\\
+\ \ \ before sentence & GUNNEXP2 & non dyslexic & .88 [.82, .93] & .90 [.85, .94] & 0.16 [-0.46, 0.79] & 0.36\\
+\ \ \ before word & CATO & dyslexic & .75 [.64, .85] & .72 [.59, .82] & -0.04 [-0.19, 0.12] & 0.09\\
+\ \ \ before word & CATO & non dyslexic & .64 [.51, .75] & .62 [.49, .74] & -0.02 [-0.19, 0.16] & 0.09\\
+\ \ \ before word & GUNNEXP2 & non dyslexic & .47 [.40, .54] & .45 [.38, .52] & -0.06 [-0.46, 0.35] & 0.21\\
+\ \ \ within word & CATO & dyslexic & .07 [.03, .12] & .11 [.06, .19] & 0.05 [-0.02, 0.13] & 0.08\\
+\ \ \ within word & CATO & non dyslexic & .03 [.02, .06] & .03 [.02, .06] & 0 [-0.03, 0.03] & 0.01\\
+\ \ \ within word & GUNNEXP2 & non dyslexic & .05 [.03, .08] & .04 [.03, .06] & -0.29 [-0.88, 0.28] & 0.48\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+\elandscape
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{pre-post-test-plantra}{%
+\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
+
+The pre-post test effect for the PLanTra dataset is reported in Table
+\ref{tab:retesteffect}. Evidence for all comparisons was negligible.
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
+Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant transitions and the probability of hesitant transitions.
+The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\% PIs in brackets.}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Hesitation duration &  &  &  & \\
+\ \ \ before sentence & 879 [723, 1,052] & 909 [740, 1,100] & -0.03 [-0.23, 0.17] & 0.1\\
+\ \ \ before word & 438 [395, 485] & 434 [389, 483] & 0.01 [-0.08, 0.1] & 0.05\\
+\ \ \ within word & 245 [180, 323] & 269 [201, 353] & -0.06 [-0.3, 0.18] & 0.14\\
+Hesitation probability &  &  &  & \\
+\ \ \ before sentence & .58 [.50, .66] & .54 [.47, .62] & 0.15 [-0.26, 0.58] & 0.27\\
+\ \ \ before word & .56 [.50, .61] & .49 [.43, .55] & 0.27 [-0.04, 0.59] & 0.72\\
+\ \ \ within word & .11 [.08, .14] & .09 [.07, .12] & 0.13 [-0.27, 0.54] & 0.26\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{genre-effect-lift}{%
+\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
+
+From the LIFT data, we compared the difference between genres, i.e.~when
+writing an informative text as opposed to writing an argumentative text.
+The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
+differences were average across writing topic. Evidence for a genre
+comparisons was negligible.
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
+Cell means are shown for argumentative and informative texts in msecs for the hesitation slowdown and the probability of hesitant transitions.
+The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\% PIs in brackets.}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Hesitation duration &  &  &  & \\
+\ \ \ before sentence & 373 [301, 454] & 411 [330, 504] & -0.07 [-0.33, 0.19] & 0.16\\
+\ \ \ before word & 287 [250, 326] & 284 [248, 323] & 0.01 [-0.15, 0.16] & 0.09\\
+\ \ \ within word & 199 [118, 306] & 246 [155, 361] & -0.12 [-0.6, 0.36] & 0.29\\
+Hesitation probability &  &  &  & \\
+\ \ \ before sentence & .35 [.28, .43] & .34 [.27, .42] & 0.05 [-0.57, 0.73] & 0.36\\
+\ \ \ before word & .46 [.39, .54] & .48 [.41, .55] & -0.05 [-0.51, 0.44] & 0.25\\
+\ \ \ within word & .07 [.05, .10] & .07 [.05, .10] & -0.02 [-0.81, 0.76] & 0.46\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{transition-location-effect-constrained-mixture-model}{%
+\subsection{Transition location effect (constrained mixture
+model)}\label{transition-location-effect-constrained-mixture-model}}
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:loceffectcon}Effect of transition location on keystroke intervals.
+Differences are shown on log scale (for durations) and logit scale for hesitation probability.
+95\% PIs in brackets.}\\
+\toprule
+ \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
+\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5}
+Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:loceffectcon} continued}}\\
+\toprule
+ \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
+\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5}
+Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
+\midrule
+\endhead
+\textbf{C2L1} &  &  &  & \\
+\ \ \ before sentence vs word & 0.24 [0.02, 0.48] & 1.11 & 0.44 [-0.56, 1.66] & 0.68\\
+\ \ \ before vs within word & -0.03 [-0.14, 0.09] & 0.07 & 3.95 [3.54, 4.38] & > 100\\
+\textbf{CATO (non-dyslexic, unmasked)} &  &  &  & \\
+\ \ \ before sentence vs word & 1.27 [0.98, 1.55] & > 100 & -0.36 [-1.41, 0.69] & 0.67\\
+\ \ \ before vs within word & -0.1 [-0.38, 0.21] & 0.2 & 4.05 [3.2, 4.91] & > 100\\
+\textbf{GUNNEXP2 (unmasked)} &  &  &  & \\
+\ \ \ before sentence vs word & 0.9 [0.76, 1.04] & > 100 & 2.15 [1.6, 2.75] & > 100\\
+\ \ \ before vs within word & 0.44 [0.17, 0.69] & 26.78 & 2.76 [2.24, 3.28] & > 100\\
+\textbf{LIFT} &  &  &  & \\
+\ \ \ before sentence vs word & 0.21 [-0.05, 0.48] & 0.5 & -0.53 [-1.04, 0.08] & 1.61\\
+\ \ \ before vs within word & 0.17 [-0.32, 0.53] & 0.41 & 2.48 [1.81, 3.22] & > 100\\
+\textbf{PLanTra} &  &  &  & \\
+\ \ \ before sentence vs word & 0.57 [0.4, 0.74] & > 100 & 0.16 [-0.25, 0.57] & 0.28\\
+\ \ \ before vs within word & 0.36 [0.16, 0.56] & 55.53 & 2.3 [1.9, 2.7] & > 100\\
+\textbf{SPL2 (L1)} &  &  &  & \\
+\ \ \ before sentence vs word & 1.17 [1.02, 1.3] & > 100 & 1.23 [0.59, 1.88] & > 100\\
+\ \ \ before vs within word & 0.48 [0.16, 0.76] & 9.07 & 3.3 [2.56, 4.08] & > 100\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+\end{appendix}
+
+\clearpage
+\makeatletter
+\efloat@restorefloats
+\makeatother
+
+\begin{appendix}
+\section{}
+\hypertarget{transition-location-effect-unconstrained-mixture-model}{%
+\subsection{Transition location effect (unconstrained mixture
+model)}\label{transition-location-effect-unconstrained-mixture-model}}
+
+We present the mixture-model posterior estimates for each of the three
+parameter values in the three facets of Figure \ref{fig:crossstudypost}.
+Similar to the constrained mixture model, we aggregated the posterior
+across conditions\footnote{We aggregated across pre-post test for the
+  PLanTra dataset as well as genre and topic of the LIFT data set.} and
+removed conditions that might conflate comparisons\footnote{We removed
+  the masked writing condition in the GUNNEXP2 and CATO, the dyslexic
+  group in the CATO data set, and L2 writing in the SPL2 data set.}. The
+resulting posterior allows us to examine differences between transition
+locations for each data set associated with each of the three
+mixture-model parameters.
 
@@
 
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
+In the following we evaluate differences between transition locations
+for all three mixture model parameters. Figure \ref{fig:crossstudypost}
+shows largely the same patterns (with caveats) for keystroke interval
+estimates by transition location across data sets for all three mixture
+model parameters. We found that hesitations appear more frequently at
+before-word transitions than within words across data sets (all BFs
+\textgreater{} 100). Also hesitations are longer at before-sentence
+transitions compared to before-word transitions (all BFs \textgreater{}
+100; except for C2L1: BF = 0.35 and LIFT = 2.14) and before-word
+transitions compared to within-word transitions (all BF \textgreater{}
+10; except for LIFT: BF = 0.71). Further, we observed that fluent
+key-transitions are slower at before-word and before-sentence locations
+compared to within-word locations (all BFs \textgreater{} 100) but there
+is generally no evidence for a difference for fluent transitions for
+before-sentence transitions compared to before-word transitions (BFs
+\textless{} 0.08; except for SPL2: BF \textgreater{} 100 and GUNNEXP2:
+BF \textgreater{} 100). The full results of these pairwise comparisons
+can be found in Table \ref{tab:loceffect}.
+
+Hesitation duration tends to be longer at before-sentence locations
+compared to before-word locations (except for data sets C2L1 and LIFT)
+and longer for before-word locations compared to within-word locations
+(except for the LIFT data set). However, for most data set there is
+negligible evidence for the idea that writers pause more frequently at
+before-sentence locations compared to before-word locations (except for
+data sets SPL2 and GUNNEXP2). This is interesting because it is
+generally believed that pausing behaviour is associated with syntactic
+edges such that more and longer pauses are predicted for key transitions
+at larger syntactic edges following the pattern before-sentence \(>\)
+before-word \(>\) within-word. In fact, the data set LIFT showed less
+pausing before sentences compared to before words. In brief, while
+keystroke transitions and pauses tend to be longer and more frequent at
+before-word locations compared to within-word transitions, it is not
+clear in which contexts keystroke transitions are before-sentence
+locations are slower, and their hesitations are longer and more
+frequent.
+
+Outstanding is the overall substantially longer fluent transitions for
+the C2L1 data. This is presumably reflecting that the population that
+this sample is from was the youngest among our data sets presumably
+involving the least experienced writers in our data pool. Hesitation
+duration and frequencies were similar to the other data sets. There are
+some inconsistencies for fluent before-sentence transitions compared to
+before-word transitions. Some data sets show before-sentence slowdowns
+for fluent transition compared to words. These inconsistencies could, to
+some extent, be explained on the basis that data sets differ as to
+whether before-sentence transitions involve complex key combination
+involve the mean or sum of transitions between space, shift and / or the
+sentence-initial letters. In particular, some data include the character
+key following the shift key at before sentence location (PLanTra, LIFT)
+but others did not scope over the character following the shift key
+(CATO, C2L1, SPL2, GUNNEXP2). Notice though that for the data sets
+PLanTra and LIFT, there was no evidence for a consistent differences
+between before-sentence and before-word transitions (except for longer
+hesitations in the PLanTra dataset).
+
+To address this finding, and inconsistency in how before-sentence
+transitions were timed, we test to what extent this difference affected
+the modelling results for the SPL2 data set. The results are shown in
+Appendix \ref{key-combination-effect-unconstrained-mixture-model}. We
+found that including the character following the shift key substantially
+extends both the transition duration and the hesitation duration but not
+the hesitation frequency. However, this conflicts with the absence of
+differences in data sets that included the character following shift at
+before-sentence transitions (PLanTra, LIFT). In other words, it is
+unlikely that patterns in our results can be explained on the basis of
+how before-keystroke transitions were operationalised (complex
+key-combinations at before-sentence locations).
+
+\blandscape
 
@@
 \end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
+\elandscape
+\end{appendix}
 
@@
 \end{center}
+
+Overall, fluent transition duration and the hesitation duration were
+affected by whether or not the sentence-initial transition include the
+character following shift. Fluent key transitions were substantially
+longer when including the interval following the shift key. The slowdown
+for hesitations was affected too but the difference is numerically
+small. There was no conclusive evidence for an increased hesitation
+probability.
 \end{appendix}
@@
 \section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
 \hypertarget{simulation}{%
@@
 A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
+performance for two-distribution mixture models -- is that, in
+principle, as the mixture model has more parameters it might always lead
+to a better fit. We addressed this concern before by using
+cross-validation techniques for model comparison which is preventing
+overfitting models. To address this concern we repeated a comparison for
+a single distribution model and a two-distrobutionsl mixture model for
+two sets of simulated data that were simulated either with a single
+distribution or two-distributional process at heart. In other words,
+this allows us to test the predictive performance of our models in a
+context where we know the true underlying data generating process --
+single distribution vs two distributions -- and we can test whether
+these models can successfully uncover the true parameter value.
+
+In particular, the first data set was simulated using a
+two-distributions mixture model with two mixture components similar to
+the process described above (equations in \ref{eq:bimodcon} and
+\ref{eq:bimoduncon}). This process and the corresponding Bayesian model
+is summarised in equation \ref{eq:simmog}.
 
@@
 
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
+\caption{Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process.
+The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
 \end{figure}

### 2023-12-10 report/manuscript.Rmd
Update model comparison using stepwise difference

@@
 Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared via its difference $\Delta\widehat{elpd}$ between models.
+We also summarise the comparison between models as the absolute value of the ratio of the difference $\Delta\widehat{elpd}$ and its standard error -- $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
 Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both two-distributions mixture models provided a substantially better fit than any of the single distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the single distribution Gaussian model.
 
@@
                           mutate(dataset = .x)) %>% 
-  select(dataset, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
+  select(dataset, model, contains("elpd_diff"), se_diff) %>%
+  arrange(dataset, desc(elpd_diff)) %>% 
   mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
@@
   transmute(
+    Comparison = model,
+    elpd_diff_ratio,
      across(dataset, ~recode(., 
@@
                 "gunnexp2" = "GUNNEXP2")),
-    m = recode(model, 
-               lmmgaus = "M1",
-               lmm = "M2",
-               lmmuneqvar = "M3",
-               mogbetaconstr = "M4",
-               mogbetacontr = "M4",
-               mogbetaunconstr = "M5"),
-    across(model, ~recode(., lmm = "Single distribution log-Gaussian",
-                          lmmgaus = "Single distribution Gaussian",
-                          mogbetaconstr = "Two distributions mixture of log-Gaussians (constrained)",
-                          mogbetacontr = "Two distributions mixture of log-Gaussians (constrained)",
-                          mogbetaunconstr = "Two distributions mixture of log-Gaussians (unconstrained)",
-                          lmmuneqvar = "Single distribution log-Gaussian (unequal variance)")),
-            across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-            across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")"))) %>% 
-  mutate(across(elpd_diff, ~str_replace(., "0 \\(0\\)", "--"))) %>% 
-  arrange(dataset)
-
-table <- table %>% select(-elpd_loo) %>% 
-  pivot_wider(names_from = dataset, values_from = elpd_diff) %>% 
-  rename(Model = m, Description = model)
-
-```
+    Description = case_when(str_detect(model, "M2$") ~ "Single distribution log-Gaussian (M2)",
+                            str_detect(model, "M3$") ~ "Single distribution log-Gaussian (unequal variance; M3)",
+                            str_detect(model, "M4$") ~ "Two distributions mixture of log-Gaussians (constrained; M4)",
+                            str_detect(model, "M5$") ~ "Two distributions mixture of log-Gaussians (unconstrained; M5)"),
+            across(elpd_diff, ~str_c(., " (", se_diff, ")")))
+
+table_ratios <- table %>% select(dataset, Comparison, elpd_diff_ratio) %>% 
+  arrange(Comparison)
+
+table <- table %>% select(-elpd_diff_ratio) %>% 
+  pivot_wider(names_from = dataset, values_from = elpd_diff)
+
+table_ratios <- table_ratios %>% 
+  mutate(across(elpd_diff_ratio, as.numeric)) %>% 
+  summarise(across(elpd_diff_ratio, list(min = min, max = max), .names = "{.fn}"), .by = Comparison) %>% 
+  mutate(range = str_c("[", min, ", ", max, "]"))
+
+tr_m2 <- filter(table_ratios, str_detect(Comparison, "M2$")) %>% pull(range)
+tr_m3 <- filter(table_ratios, str_detect(Comparison, "M3$")) %>% pull(range)
+tr_m4 <- filter(table_ratios, str_detect(Comparison, "M4$")) %>% pull(range)
+tr_m5 <- filter(table_ratios, str_detect(Comparison, "M5$")) %>% pull(range)
+
+```
+
+Model differences for all data sets are shown in Table \ref{tab:loos}.
+For all data sets we found the same pattern.
+Overall both two-distributions mixture models provided substantially better fit than any of the single distribution models.
+Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single distribution model; across datasets the difference $\Delta\widehat{elpd}$ ranged within `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model.
+The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which did not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations); difference in predictive performance was `r tr_m5` standard errors higher for the unconstrained model.
+Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; $\Delta\widehat{elpd}$ was in a range of `r tr_m3` standard errors.
+The weakest model was the single distribution Gaussian model which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of `r tr_m2` standard errors.
 
@@
 apa_table(table, 
-      caption = "Model comparisons. The top row shows the models with the highest predictive performance compared to all other models. Standard error is shown in parentheses.",
+      caption = "Model comparisons.
+Models were compared stepwise from the simples to the most complex model.
+The comparison is indicated along with the description of the more complex model in each comparison.
+A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model.
+Standard error is shown in parentheses.",
       align =c("p{1.5cm}", "p{5cm}", rep("r", 6)), 
@@
 Shown are the distributions for both mixture components (fluent and hesitant transition durations) and their respective weighting.
-Posterior of SPL2 data (English sessions only).', fig.height=6}
-
-ps <- read_csv("../stanout/spl2/mog_constr_posterior_for_densityplot.csv") 
+Posterior of CATO data.', fig.height=6}
+
+ps <- read_csv("../stanout/cato/mog_constr_posterior_for_densityplot.csv") 
 # Extract parameters

### 2023-12-10 scripts/c2l1/modelcomparison.R

@@ -14,17 +14,17 @@ for(file in files){
 }
 
 (loos <- ls(pattern = "loo_.*"))
-mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
 
-mcs <- mcs %>% as.data.frame() %>% 
+mc1 <- do.call(what = loo_compare, args = lapply(loos[2:1], as.name))
+mc2 <- do.call(what = loo_compare, args = lapply(loos[2:3], as.name))
+mc3 <- do.call(what = loo_compare, args = lapply(loos[3:4], as.name))
+mc4 <- do.call(what = loo_compare, args = lapply(loos[4:5], as.name))
+
+mcs <- map_dfr(.x = list(mc1, mc2, mc3, mc4), ~as.data.frame(.x) %>% 
   rownames_to_column("model") %>%
-  as_tibble() %>%
-  mutate(model = recode(model, model1 = loos[1],
-                               model2 = loos[2],
-                               model3 = loos[3],
-                               model4 = loos[4],
-                               model5 = loos[5]),
-         model = str_remove(model, "^loo_"));mcs
+  as_tibble() %>% 
+  slice(2)) %>%
+  mutate(model = c("M1 -- M2", "M2 -- M3", "M3 -- M4", "M4 -- M5"),
+         elpd_diff_ratio = abs(elpd_diff / se_diff))
 
-file_out <- paste0("stanout/c2l1/modelcomparison.csv")
-write_csv(mcs, file_out)
+write_csv(mcs, "stanout/c2l1/modelcomparison.csv")

### 2023-12-10 scripts/cato/modelcomparison.R

@@ -14,17 +14,19 @@ for(file in files){
 }
 
 (loos <- ls(pattern = "loo_.*"))
-mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
 
-mcs <- mcs %>% as.data.frame() %>% 
-  rownames_to_column("model") %>%
-  as_tibble() %>%
-  mutate(model = recode(model, model1 = loos[1],
-                               model2 = loos[2],
-                               model3 = loos[3],
-                               model4 = loos[4],
-                               model5 = loos[5]),
-         model = str_remove(model, "^loo_"));mcs
+mc1 <- do.call(what = loo_compare, args = lapply(loos[2:1], as.name))
+mc2 <- do.call(what = loo_compare, args = lapply(loos[2:3], as.name))
+mc3 <- do.call(what = loo_compare, args = lapply(loos[3:4], as.name))
+mc4 <- do.call(what = loo_compare, args = lapply(loos[4:5], as.name))
 
-file_out <- paste0("stanout/cato/modelcomparison.csv")
-write_csv(mcs, file_out)
+mcs <- map_dfr(.x = list(mc1, mc2, mc3, mc4), ~as.data.frame(.x) %>% 
+                 rownames_to_column("model") %>%
+                 as_tibble() %>% 
+                 slice(2)) %>%
+  mutate(model = c("M1 -- M2", "M2 -- M3", "M3 -- M4", "M4 -- M5"),
+         elpd_diff_ratio = abs(elpd_diff / se_diff))
+
+glimpse(mcs)
+
+write_csv(mcs, "stanout/cato/modelcomparison.csv")

### 2023-12-10 scripts/lift/modelcomparison.R

@@ -14,17 +14,18 @@ for(file in files){
 }
 
 (loos <- ls(pattern = "loo_.*"))
-mcs <- do.call(what = loo_compare, args = lapply(loos, as.name))
+mc1 <- do.call(what = loo_compare, args = lapply(loos[2:1], as.name))
+mc2 <- do.call(what = loo_compare, args = lapply(loos[2:3], as.name))
+mc3 <- do.call(what = loo_compare, args = lapply(loos[3:4], as.name))
+mc4 <- do.call(what = loo_compare, args = lapply(loos[4:5], as.name))
 
-mcs <- mcs %>% as.data.frame() %>% 
-  rownames_to_column("model") %>%
-  as_tibble() %>%
-  mutate(model = recode(model, model1 = loos[1],
-                               model2 = loos[2],
-                               model3 = loos[3],
-                               model4 = loos[4],
-                               model5 = loos[5]),
-         model = str_remove(model, "^loo_"));mcs
+mcs <- map_dfr(.x = list(mc1, mc2, mc3, mc4), ~as.data.frame(.x) %>% 
+                 rownames_to_column("model") %>%
+                 as_tibble() %>% 
+                 slice(2)) %>%
+  mutate(model = c("M1 -- M2", "M2 -- M3", "M3 -- M4", "M4 -- M5"),
+         elpd_diff_ratio = abs(elpd_diff / se_diff))
 
-file_out <- paste0("stanout/lift/modelcomparison.csv")
-write_csv(mcs, file_out)
+glimpse(mcs)
+
+write_csv(mcs, "stanout/lift/modelcomparison.csv")

### 2023-12-10 scripts/spl2_shift/modelcomparison.R

@@ -1,11 +1,11 @@
 library(loo)
 library(tidyverse)
 
-path <- "stanout/plantra"
+path <- "stanout/spl2_shift"
 (files <- list.files(path, pattern = "^.[^\\_]*.rda$", full.names = T))
 
 for(file in files){
-    varname <- str_remove_all(file, ".rda|stanout/plantra/")
+    varname <- str_remove_all(file, ".rda|stanout/spl2_shift/")
     m <- readRDS(file)
     log_lik <- extract_log_lik(m, merge_chains = F) 
     r_eff <- relative_eff(exp(log_lik)) 
@@ -28,4 +28,9 @@ mcs <- map_dfr(.x = list(mc1, mc2, mc3, mc4), ~as.data.frame(.x) %>%
 
 glimpse(mcs)
 
-write_csv(mcs, "stanout/plantra/modelcomparison.csv")
+write_csv(mcs, "stanout/spl2_shift/modelcomparison.csv")
+
+
+
+
+

### 2023-12-20 report/manuscript.Rmd
minor fix

@@
 Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared via its difference $\Delta\widehat{elpd}$ between models.
-We also summarise the comparison between models as the absolute value of the ratio of the difference $\Delta\widehat{elpd}$ and its standard error -- $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
+We also summarised the comparison between models as the absolute value of the ratio of the difference $\Delta\widehat{elpd}$ and its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
 Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
@@
                 "gunnexp2" = "GUNNEXP2")),
-    Description = case_when(str_detect(model, "M2$") ~ "Single distribution log-Gaussian (M2)",
-                            str_detect(model, "M3$") ~ "Single distribution log-Gaussian (unequal variance; M3)",
-                            str_detect(model, "M4$") ~ "Two distributions mixture of log-Gaussians (constrained; M4)",
-                            str_detect(model, "M5$") ~ "Two distributions mixture of log-Gaussians (unconstrained; M5)"),
+    Distribution = case_when(str_detect(model, "M2$") ~ "Single log-Gaussian (M2)",
+                            str_detect(model, "M3$") ~ "Single log-Gaussian (unequal variance; M3)",
+                            str_detect(model, "M4$") ~ "Mixture of two log-Gaussians (constrained; M4)",
+                            str_detect(model, "M5$") ~ "Mixture of two log-Gaussians (unconstrained; M5)"),
             across(elpd_diff, ~str_c(., " (", se_diff, ")")))
 
-table_ratios <- table %>% select(dataset, Comparison, elpd_diff_ratio) %>% 
-  arrange(Comparison)
+table_ratios <- table %>% select(dataset, Comparison, elpd_diff_ratio)
 
@@
 
-apa_table(table, 
-      caption = "Model comparisons.
-Models were compared stepwise from the simples to the most complex model.
-The comparison is indicated along with the description of the more complex model in each comparison.
+table %>% 
+  arrange(desc(Comparison)) %>% 
+  apa_table(caption = "Model comparisons.
+Models were compared incrementally from the simples to the most complex model.
+The comparison is indicated along with the distribution of the more complex model in each comparison.
 A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model.
@@
       longtable = T,
-      font_size = "small",
-      note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density -- relative to the model with the highest predictive performance in the top row.")
+      font_size = "footnotesize",
+      note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density -- relative to the model with the highest predictive performance in the top row.
+Model details can be found in Appendix \ref{statistical-models}.")
 #                col_spanners = list(" " = 1:2,

### 2023-12-25 report/appendix_shift_constr.Rmd
Last draft before feedback on docs

@@
           caption = "Mixture model estimates for key transitions immediately preceding a sentence.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown and the probability of hesitant transitions.
 The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.

### 2023-12-25 report/appendix_shift_constr.tex
Last draft before feedback on docs

@@
 \caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions immediately preceding a sentence.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown and the probability of hesitant transitions.
 The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.

### 2023-12-25 report/manuscript.Rmd
Last draft before feedback on docs

@@
 
-abstract: Classical serial models view the process of producing a text as a descrete chain of pauses and writing bursts.
+abstract: Classical serial models view the process of producing a text as a discrete chain of pauses and writing bursts.
 In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
+We implemented these two views in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
 We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
 Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
+Further we found -- across populations -- that transitions across word and even sentence boundaries are often fluent and uninterrupted by planning of upcoming linguistic material; when pausing occurred pauses were longer before sentence than before word which is consistent with the idea larger linguistic edges are associated with higher level planning.
 Our results cannot be explained by the serial but are in line with the cascading view of writing.
@@
 #        - "appendix_postparam.tex"
-        - "appendix_postparam_constr.tex"
+        - "appendix_transloc.tex"
+        - "appendix_shift.tex"
         - "appendix_shift_constr.tex"
@@
         - "appendix_masking_constr.tex"
+        - "appendix_postparam_constr.tex"
 #        - "appendix_prepost.tex"
@@
         - "appendix_transloc_constr.tex"
-        - "appendix_transloc.tex"
-        - "appendix_shift.tex"
         - "appendix_sim.tex"
@@
 
-<!-- a statistical framework for the analysis of keystroke data that captures not just the multilevel-structure of the data but also presents a principled way to capture the probabilistic nature of pauses in text composition.
-Hesitations are probabilistic in the sense that we cannot be predicted their occurrence deterministically.
-In principle we can produce several words and even sentences without pausing.
-However, hesitations are more of less likely to occur in certain contexts, for example, at sentence boundaries.
--->
-
 Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements [@van1991handwriting; @olive2014toward].
@@
 
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component [@wagenmakers2007linear; @wing1973response; @schoner2002timing].
+In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts.
+Full modelling details can be found in Appendix \ref{statistical-models; we summarise the most important model properties in the following.
+The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data.
+In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.
 
 The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in @roeser2021modelling.
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
+Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes.
+These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs.
+This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.
 
@@
 We discussed advantages with this approach in the introduction section.
-These datasets include samples from populations with different writing experience and languages (e.g.
-young / L2 writers, students) and writing tasks (e.g.
+These datasets include samples from populations with various writing experience and languages (e.g.
+young / L2 writers, students) performing different writing tasks (e.g.
 essays, syntheses).
-Datasets from a variety of different writers and writing tasks were deliberately chosen to challange our modelling approach and to test to what extent resulting pausing patterns generalise across writing contexts.
+Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.
 
@@
 
-Six datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
+TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2.
+For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say "unpublished".
+I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk).
+Was the GUNNEXP2 sample Norwegian?
+Should we include information like time limit or text length in the table?
+
+Six datasets with keystroke data from free text production were used for analysis.
+An overview can be found in Table \ref{tab:datasets}; descriptions below.
+Dataset used are C2l1 [@ronneberg2022process], CATO [@torrance2016adolescent], GUNNEXP2 [@torranceb], SPL2 [@torrancea], LIFT published in @vandermeulen2020 and described in @vandermeulen2020mapping, PLanTra data are published in @rossetti2022text and described in @rossetti2022s.
+In C2L1 Norwegian 6th graders composed argumentative essays.
+In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text.
+In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced.
+The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each.
+The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability.
+The GUNNEXP2 data ....
+[MARK, please add a sentence here to describe the production context].
+
+For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite [@sim07; @torrance201203].
+LIFT and PLanTra data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019] and SPL2 data were collected using CyWrite [@chukharev2019combined].
 
@@
                                  "Synthesis", 
-                                 ""),
-                 `N (ppts)` = as.character(c(126, 26*2, 39, 47, 658, 45)),
+                                 "?????"),
+                 N = as.character(c(126, 26*2, 39, 47, 658, 45)),
 #                n_texts = c(1, 2, 2, 2, NA),
@@
 #                n_words = c(),
-                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
+                `Mean age` = c(11.8, 16.9, 20.6, 23, 17.0, "????"),
+                 Sample = c("6th graders", "Secondary school students", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
+                 Country = c("Norway", "Norway", "USA", "Belgium", "The Netherlands", "???? Norway"),
                  Language = c( "Norwegian", 
@@
                                "Dutch",
-                               "Norwegian"),
-                Manipulation = c("--", 
-                               "weak decoders / control; masked / unmasked",
-                               "write in L1 / L2",
-                               "pre / post test trained in plain language principles and control", 
-                               "Various topics and genres",
-                               "masked / unmasked")) %>% 
+                               "???? Norwegian")) %>% 
   arrange(Dataset) %>% 
-  apa_table(align = c(rep("l", 4), rep("r", 2), "p{2cm}", "p{3cm}"), 
+  apa_table(align = c("l", "p{3cm}", "p{2cm}", "p{2.5cm}", "r", "r", "p{2.5cm}", "p{2.25cm}", "p{2.5cm}"), 
             escape = FALSE, 
@@
 
+#                Manipulation = c("--", 
+#                               "Weak decoders / control; texted masked / unmasked",
+#                               "Text produced in L1 / L2",
+#                               "Pre / post test with training on plain language principles", 
+#                              "Various topics and genres",
+#                               "masked / unmasked")
 ```
@@
 
-TODO: Mark, can you add information for C2L1 and CATO as needed
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite [@sim07; @torrance201203].
-
-The C2L1 dataset comprises data Norwegian 6th graders -- *N* = 126, mean age 11 years 10 months -- published in @ronneberg2022process.
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-
-The CATO dataset is published in @torrance2016adolescent.
-Norwegian upper secondary students -- *N* = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say "unpublished".
-I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in @torranceb includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-
-The SPL2 data set involves undergraduate university students in the US -- *N* = 39, 2mean age = 20.6 years, SD = 1.51 -- composing argumentative essays in English (L1) and Spanish (L2).
-Students wrote within the CyWrite word-processing environment [@chukharev2019combined].
-Participants wrote, with a 40 minute time limit, to two writing prompts.
-Order and writing prompt were counterbalanced across language.
-
-LIFT and PLanTra data data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].
-
-The LIFT data are published in @vandermeulen2020 and described in @vandermeulen2020mapping.
-LIFT is a representative sample of Dutch students (*N* = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education collected from 43 schools.
-The students were asked to conduct four synthesis writing tasks, two argumentative and two informative texts, about each of four topics, with order randomized per school.
-The students received 50 minutes for each tasks.
-
-<!-- It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch. -->
-
-The PLanTra (Plain Language Training for business content) data are published in @rossetti2022text and described in @rossetti2022s.
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of *N* = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in @torrancea.
-The data come from a text composition task of undergraduate university students -- *N* = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined].
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-## Transition types
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful [@torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
+## Data extraction
+
+From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful [@chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
 In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
+At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section.
 Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-```{r keyloc, results='asis', fig.pos="b"}
+Also we removed transitions that were followed by an editing operation.
+
+```{r keyloc, results='asis'}
 
@@
 apa_table(table, caption = "Transition location classification.",
-            align = c("p{3cm}", "p{3cm}", "p{8cm}"), 
+            align = c("p{3cm}", "p{5cm}", "p{6cm}"), 
             escape = FALSE, 
@@
 ```
-
-## Data reduction
 
@@
 
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
+We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data).
 We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models -- decribed in the next section -- require to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
+We further removed keystroke intervals that were too short to represent a behavioural response ($\le$ 50 msecs) or too long ($\ge$ 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}.
+From the remaining data we randomly sampled a maximum of `r n_samples` observations per participant, condition, and transition location (when more than `r n_samples` were available).
+This was done to reduce the computation time of the Bayesian models require to complete sampling.
+For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets.
+Because the LIFT data set included the largest number of writing tasks, we sampled 50 observations per condition, location and participant which would otherwise exceed the computational resources available to us.
+The number of keystroke data used in the analysis can be found in Table \ref{tab:datareduction}.
 
@@
           mutate(ds = .x$ds)) %>% 
-  mutate(across(where(is.numeric), ~round(.*100, 2))) %>% 
+  mutate(across(where(is.numeric), ~signif(.*100, 2))) %>% 
   unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
   unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
-  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
+  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")")),
+         across(c(`<50ms`, `>30,000ms`), ~str_replace(., "0 \\(0\\)", "--"))) %>% 
   rename(`$\\le$ 50 msecs` = `<50ms`,
@@
           mutate(ds = .x$ds)) %>% 
-    mutate(across(where(is.numeric), ~round(.*100, 1))) %>%
+    mutate(across(where(is.numeric), ~signif(.*100, 2))) %>%
     unite("keep", starts_with("keep"), sep = " (") %>% 
-    mutate(across(c(keep), ~str_c(., ")"))) %>%
+    mutate(across(keep, ~str_c(., ")")),
+          across(keep, ~str_replace(., "100 \\(0\\)", "--"))) %>%
   pivot_wider(names_from = location, values_from = keep) %>% 
@@
           placement = "bp!",
-          caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location. Standard error is shown in parentheses.",
+          caption = "Data reduction.
+Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location (`--` if all data were used).
+Standard error in parentheses.",
           align =c("l", rep("r", 5)),
@@
 An overview of all models can be found in Table \ref{tab:models}.
-More detailed model descriptions and motivations can be found in Appendix \ref{statistical-models}.
+More detailed model descriptions and motivation can be found in Appendix \ref{statistical-models}.
 
 ```{r models, results = 'asis'}
-models <- tibble(Models = paste0("M",1:5),
+models <- tibble(Models = str_c("M",1:5),
        Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
        `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
-       Description = c("Standard single Gaussian model with effects of transition location by study-specific manipulations and by-participant random effect",
-                       "Single distribution log-Gaussian model equivialent to M1",
-                       "Equivialent to M2 but with different variance components for each transition location (unequal variance)",
-                       "Two-distributions mixture of a log-Gaussian of fluent intervals and another of hesitant; distribution of hesitant keystroke intervals is equivialent to M3 but distribution of fluent intervals has a mean across all design cell.
-Mixing proportion captures the relative number of disfluent transitions",
-                       "Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location")) %>% 
+#       group = c(rep("Serial", 3), rep("Cascading", 2)),
+       Description = c("Single Gaussian model with effects of transition location (and experiment-specific manipulation).",
+                       "Single distribution log-Gaussian model equivialent to M1.",
+                       "Equivalent to M2 but with different variance components for each transition location (unequal variance).",
+                       "Two-distributions mixture of a log-Gaussian for fluent transition intervals and another log-Gaussian that is wider than the previous distribution hesitant / disfluent key transitions; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition location but the distribution of fluent intervals has the same mean across all design cell.
+Mixing proportions capture the relative number of disfluent transitions for the distribution of hesitant transitions for each combination of transition location.",
+                       "Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location.")) %>% 
   select(Models, Description)
@@
           digits = 0,
-          caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.")
-#          note = "LMM = Single distribution mixed-effects models; MoG = two distributions mixture of log-Gaussians") 
-```
-
-Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset specific manipulations and random effects for participants.
+          caption = "Overview of cognitive models of writing. All models included by-participant random effect.",
+          stub_indents = list(
+                     "\\textbf{Serial}" = 1:3,
+                     "\\textbf{Cascading}" = 4:5))
+
+```
+
+Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset-specific manipulations.
 In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
-For model M3 we relaxed the equal variance assumption for transition locations, thus allowing each transition location to assume a different standard deviation (see Appendix \ref{statistical-models} for rationale).
-
-Models M4 and M5 and extensions of the single distribution models.
-Importantly models M4 and M5 assume that keystroke data are the result of a combination of two data generating processes -- consistent with the cascading account -- rather than one.
-These are called finite mixture models in the literature [@gelman2014; @roeser2021modelling; @peel2000finite].
-
-These models extend the assumption that processing at higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that shifts the distribution of IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; this is because slower human behaviour is generally known to be associated with a larger variability [@wagenmakers2007linear; @wing1973response; @schoner2002timing]; see Appendix \ref{statistical-models} for rationale.
+
+Models M4 and M5 are extensions of the unequal-variance single distribution model M3.
+Importantly models M4 and M5 assume that keystroke data result from a combination of two data generating processes -- consistent with the cascading account -- rather than one.
+These model are called finite mixture models in the literature [@gelman2014; @roeser2021modelling; @peel2000finite] and capture the assumption that processing at higher levels of activation leads to longer pauses which can in principle happen at transition before sentence, before words or even mid-word.
+In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that pauses at larger linguistic edges are more likely but not obligated by the cognitive system that generates keystrokes.
+This is achieved by modelling key transitions as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
 
 1.
-Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter illustrated in equation \ref{eq:bimodcon2}.
+Information from upstream mental processes can flow into keystrokes without interruption at intermediate levels.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger.
+The population estimate for fluent key transitions (i.e.
+typing speed) is captured by the $\beta$ parameter.
 
 2.
-Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon2}.
+Any interruptions at upstream levels of mental representation result in a delay in the information flow and therefore create a lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
+The population estimate of the hesitation slowdown is captured by $\delta$ and its frequency is captured by the mixing weight (proportion) $\theta$.
 
@@
 
-The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
+The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents the distribution of fluent -- uninterrupted -- key transitions.
 Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
-$\theta$ is here parameterised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$ .
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
+$\theta$ is here parameterised to represent the probability that a key transition is associated with the distribution of hesitant transitions.
+This probability is inversely related to the mixing weight of the distribution of fluent transitions by $1-\theta$ .
+In other words, a larger weight for either distribution inevitably means a lower weight for the other.
 We call this parameter the probability of hesitant transitions.
-
-The size of a fluent key-transition does not necessarily vary by transition location which is what our constrained model M4 assumes.
-In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions.
-However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter.
-We therefore also implemented an unconstrained model M5 that assumes that the distribution over durations of fluent transitions -- represented as $\beta$ -- varies across transition locations.
-
-All models were implemented in the Bayesian framework.
-The Bayesian framework, as used in this paper, is ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates [@farrell2018computational; @gelman2014; @lee2014bayesian].
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.
+When a parameter was assumed to vary by transition location (levels: before sentence, before word, within word) this was indicated as subscript; similar for participants.
+The hesitation probability was allowed to vary by participants constrained by a hyper-parameter for each transition location.
+
+The constrained implementation M4 assume that the size of a fluent key-transition does not vary by transition location.
+In other words, the value that the $\beta$ parameter is taking on is the same for transitions at before-sentence, before-word, and within-word locations.
+However, letter bigrams may be executed systematically faster than transitions between space and either a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
+This is for two reasons: first, and importantly, highly frequent bi- and even trigrams are stored, retrieved and executed as one motor code [@behmer2016crunching; @grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants] because of the strong association between letters.
+This association is reducing the length of transition durations.
+It is less likely that the same association exists for space-letter and space-shift pairs.
+Second, before-sentence transition in two datasets (PLanTra, LIFT) are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transition; we return to this in the results section.
+We therefore implemented an unconstrained mixture model M5 that assumes that the distribution over durations of fluent transitions -- represented as $\beta$ -- may vary across transition locations.
+
+All models were implemented in the Bayesian framework [@gelman2014; @mcelreath2016statistical].
+The Bayesian framework is ideal for the estimation of parameter values.
+This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty [@farrell2018computational; @gelman2014; @lee2014bayesian].
+To achieve this, Bayesian models require the inclusion of prior information, i.e.
 existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e.
+For small datasets priors influence the inferred parameter values; for larger datasets weakly informative and vague priors are quickly overcome by the data [i.e.
 automatic Ockham's razor, @jefferys1992ockham].
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g.
+In other words the choice of priors has less impact on the posterior, certainly for weakly informative priors.
+In the present analysis, we used weakly informative priors to aid model convergence by constraining the parameter space to plausible values [see e.g.
 @lambert2018student; @mcelreath2016statistical].
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for detailed descriptions of parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon} in Appendix \ref{statistical-models}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.^[Data were analysed in Bayesian mixed effects models [@gelman2014; @mcelreath2016statistical].
-The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
+Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
+
+Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); for a tutorial see ADD URL HERE.^[The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
 Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
@@
 
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical].
-Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared via its difference $\Delta\widehat{elpd}$ between models.
-We also summarised the comparison between models as the absolute value of the ratio of the difference $\Delta\widehat{elpd}$ and its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
+To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
+Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared by  the difference between models $\Delta\widehat{elpd}$.
+We also summarised this difference as normalised over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
+Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.
 
@@
 
-Model differences for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Overall both two-distributions mixture models provided substantially better fit than any of the single distribution models.
-Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single distribution model; across datasets the difference $\Delta\widehat{elpd}$ ranged within `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which did not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations); difference in predictive performance was `r tr_m5` standard errors higher for the unconstrained model.
-Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; $\Delta\widehat{elpd}$ was in a range of `r tr_m3` standard errors.
-The weakest model was the single distribution Gaussian model which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of `r tr_m2` standard errors.
+Model differences are shown in Table \ref{tab:loos}.
+All datasets revealed the same pattern.
+Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single-distribution model; the difference $\Delta\widehat{elpd}$ between models ranged within `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model.
+The unconstrained mixture model M5 rendered a higher predictive performance than the constrained model M4 which did not allow the distribution of fluent keystroke-intervals to vary by transition location; the difference in predictive performance was between `r tr_m5` standard errors higher for the unconstrained model.
+Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; the difference $\Delta\widehat{elpd}$ ranged within `r tr_m3` standard errors.
+The weakest model was the single distribution Gaussian which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of `r tr_m2` standard errors.
 
@@
   apa_table(caption = "Model comparisons.
-Models were compared incrementally from the simples to the most complex model.
-The comparison is indicated along with the distribution of the more complex model in each comparison.
+Models were compared incrementally from the simplest to the most complex model.
+The comparison is shown along with the distribution type of the more complex model.
 A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model.
@@
       font_size = "footnotesize",
-      note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density -- relative to the model with the highest predictive performance in the top row.
-Model details can be found in Appendix \ref{statistical-models}.")
+      note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density")
 #                col_spanners = list(" " = 1:2,
@@
 
-# $\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density;
 ```
@@
 
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of two-distributions mixture models and poorest for the single distribution Gaussian model.
+We also tested to what extent model predictions fit observed data.
+Comparisons can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
+Data predicted by the two-distribution mixture models are similar to observed data; data predicted by single distribution Gaussian models showed a poor fit to the data.
 
@@
 
-As demonstrated in the model comparisons, the two-distributions mixture-model capture the writing-process data better than single-distribution models.
-Therefore, writing process is captured by the posterior estimates of the model parameter values.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as $\beta$ in equation \ref{eq:bimodcon2}), (2) the magnitude of the slowdown for hesitant transitions (represented as $\delta$), and (3) the probability of hesitant transition duration (represented as $\theta$).
-
-Figure \ref{fig:mixmodel} illustrates the mixture model approach: for each transition durations we obtain two posterior distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in longer durations.
-The relative weighting of these two distributions was captured as the probability of hesitant transitions, the parameter that captures the height of the distribution of slow keystrokes.
+As demonstrated in the model comparisons, the two-distributions mixture-models is a better statistical representation on the underlying process generating keystroke data than compared to any of the single-distribution models.
+Information about this process is captured by the posterior estimates of the models' parameter values.
+To reiterate, there are three conceptually important parameters: (1) the magnitude of the slowdown for transition hesitations (represented as $\delta$ in equation \ref{eq:bimodcon2}), (2) the probability of hesitant transitions (represented as $\theta$ in equation \ref{eq:bimodcon2}), and (3) the duration of fluent transitions (the typing speed, which was indicated as $\beta$ in equation \ref{eq:bimodcon2}).
+
+Figure \ref{fig:mixmodel} illustrates the mixture model approach on the basis of the posterior of the constrained implementation applied to the CATO data: for each transition durations we obtain two posterior probability distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in hesitant transitions.
+The hesitation probability is indicated as the relative height of either distribution.
+The transition locations are characterized by the hesitation parameters: as we illustrate the constrained mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location but their height differs as the probability of fluent transition is the inverse of the height of hesitant transitions; i.e.
+$1 - \theta$.
+More interesting are the distributions associated with hesitations; indicated in yellow: Within word transitions have a negligibly small hesitation probability barely visible in the visualisation.
+Hesitations before words and sentences are both roughly equally likely than fluent transitions.
+In other words, half of the time participants did not pause before a word or sentence; we explore this below.
+However when hesitations occurred, these are longer at before-sentence transitions compared to before word transitions.
 
 ```{r mixmodel, fig.cap='Mixture model plot for each transition location.
-Shown are the distributions for both mixture components (fluent and hesitant transition durations) and their respective weighting.
+Shown are the distributions for both mixture components (fluent and hesitant transitions) and their weighting.
 Posterior of CATO data.', fig.height=6}
@@
 
-colours <- c("Fluent transitions" = "#000000", 
-             #"Fluent transitions" = "#56B4E9",
-             "Hesitant transitions" = "#E69F00")
+colours <- c("Fluent" = "#000000", "Hesitant" = "#E69F00")
 
@@
 plot_bs <- ggplot(data, aes(x)) +
-  stat_function(aes(colour = "Fluent transitions"),
+  stat_function(aes(colour = "Fluent"),
                 geom = "line", 
@@
                 args = list(alpha, sigma, lam = 1 - theta_bs)) +
-  stat_function(aes(colour = "Hesitant transitions"),
+  stat_function(aes(colour = "Hesitant"),
                 geom = "line", 
@@
                 args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
-  stat_function(aes(fill = "Fluent transitions"),
+  stat_function(aes(fill = "Fluent"),
                 geom = "area", 
@@
                 args = list(alpha, sigma, lam = 1 - theta_bs)) +
-  stat_function(aes(fill = "Hesitant transitions"),
+  stat_function(aes(fill = "Hesitant"),
                 geom = "area", 
@@
 plot_bw <- ggplot(data, aes(x)) +
-  stat_function(aes(colour = "Fluent transitions"),
+  stat_function(aes(colour = "Fluent"),
                 linewidth = .25,
@@
                 args = list(alpha, sigma, lam = 1 - theta_bw)) +
-  stat_function(aes(colour = "Hesitant transitions"),
+  stat_function(aes(colour = "Hesitant"),
                 geom = "line", 
@@
                 args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
-  stat_function(aes(fill = "Fluent transitions"),
+  stat_function(aes(fill = "Fluent"),
                 geom = "area", 
@@
                 args = list(alpha, sigma, lam = 1 - theta_bw)) +
-  stat_function(aes(fill = "Hesitant transitions"),
+  stat_function(aes(fill = "Hesitant"),
                 geom = "area", 
@@
 plot_ww <- ggplot(data, aes(x)) +
-  stat_function(aes(colour = "Fluent transitions"),
+  stat_function(aes(colour = "Fluent"),
                 geom = "line", 
@@
                 args = list(alpha, sigma, lam = 1 - theta_ww)) +
-  stat_function(aes(colour = "Hesitant transitions"),
+  stat_function(aes(colour = "Hesitant"),
                 geom = "line", 
@@
                 args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
-  stat_function(aes(fill = "Fluent transitions"),
+  stat_function(aes(fill = "Fluent"),
                 geom = "area", 
@@
                 args = list(alpha, sigma, lam = 1 - theta_ww)) +
-  stat_function(aes(fill = "Hesitant transitions"),
+  stat_function(aes(fill = "Hesitant"),
                 geom = "area", 
@@
   scale_fill_manual(values = colours) +
-  labs(subtitle = "Within-sentence transitions", 
+  labs(subtitle = "Within-word transitions", 
        y = "", x = "Transition duration [in msecs]",
@@
 
-The predictive performance of the unconstrained mixture model is substantially higher than the predictive performance of the constrained mixture model.
+We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
+We demonstrated above that the predictive performance of the unconstrained mixture model was substantially higher than the predictive performance of the constrained mixture model.
 However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
-We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following with reference to the unconstrained implementation.
+We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following.
 
@@
 
-We present the mixture-model posterior estimates for each of the three parameters in the three facets of Figure \ref{fig:crossstudypost2}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.], and removed conditions that might conflate comparisons^[We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-While there was evidence for language-specific effects in Appendix \ref{l2-effect-spl2}, evidence for marking effects was weak as shown in Appendix \ref{masking-effect-cato-gunnexp2}.].
+The mixture-model posterior parameter estimates for all datasets is visualised in Figure \ref{fig:crossstudypost2}.
+Although models were fitted with all data set-specific conditions, we aggregated the posterior across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
+We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.], and excluded conditions that might conflate comparisons^[We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
+There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-gunnexp2}).].
 For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the slowdown duration for hesitant transitions and the hesitation probability.
-Note that the estimate for fluent transitions was estimated by dataset.
+The resulting posterior allows us to examine differences and similarities associated with transition locations for all dataset.
+In particular we can see how the linguistic location in the text is associated with hesitation duration and probability.
+The estimate for fluent transitions is shown by dataset, across transition location.
 
@@
 
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for keystroke-interval estimates by transition location across data sets for all three mixture model parameters.
-Full results of these pairwise comparisons can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
-We found that hesitations appear more frequently -- higher hesitation probability -- at before-word transitions than within words across data sets (all BFs > 100).
-Hesitations were more likely at before-sentence locations compared to before-word locations only for 2 out of 6 datasets (GUNNEXP2 and SPL2, both BFs > 100; but C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6).
-Hesitations were longer at before-sentence transitions compared to before-word transitions (all BFs > 100, except for C2L1: BF = 1.11 and LIFT = 0.5); before-word transitions were longer compared to within-word transitions for half of the data sets (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1) but not for the other (C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
-
-To reiterate, for most datasets we found negligible evidence supporting the idea that writers pause more frequently at before-sentence boundaries compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence $>$ before-word $>$ within-word.
+Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets.
+We summarise the results of the pairwise comparisons for transitions locations; full results and parameter estimates can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
+We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs > 100 for all datasets) but only for half of the data sets also longer (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
+Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GUNNEXP2 and SPL2, both BFs > 100) but longer (all BFs > 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
+
+To reiterate, for most datasets we found that pauses are before sentences are longer but not necessarily more likely than before words (except for datasets SPL2 and GUNNEXP2).
 In fact, LIFT showed some indication that pauses were less likely before sentences than words.
-In brief, while pauses tend to be longer before sentences and less likely at mid-word locations, it is less clear in which contexts pauses are (1) longer before than within words and (2) more likely to appear before sentences than before words.
-
-The duration of fluent durations was not distinguished by transition location in the constrained version of the two distribution model.
-Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transition durations.
-This is presumably reflecting that the population that this sample is from was the youngest and least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to other data sets.
-
-Inconsistencies between datasets could, to some extent, be explained on the basis that before-sentence transitions were sometimes defined as involving complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, before-sentence transitions in some  datasets (PLanTra, LIFT) included the character key following the shift key but others did not (CATO, C2L1, SPL2, GUNNEXP2).
-Appendix \ref{key-combination-effect-constrained-mixture-model} addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increases the hesitation probability (BFs > 100) but it did not affect the duration of fluent transitions or hesitations (BFs < 0.3).
-However this is in conflict with the earlier finding that before-sentence hesitations are not more likely than before-word hesitations for the datasets that included the character following shift (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
+This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for key transitions at larger linguistic edges.
+It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.
+
+Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character while the remaining data sets measured before-sentence transitions as the difference between space key and shift.
+To control for the use of complex key combinations, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability (BFs > 100) but it did not affect the duration of fluent or hesitant transitions (BFs < 0.3); see Appendix \ref{key-combination-effect-constrained-mixture-model}.
+This conflicts the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did included the character following shift (PLanTra, LIFT).
+In other words, patterns in our results cannot be explained on the basis of how before-sentence key-transitions were operationalised.
+
+Finally, the duration of fluent transitions was not distinguished by location in the constrained distribution model.
+Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transitions.
+This is presumably reflecting that these data were sampled from was the youngest and least experienced population of writers in our data pool.
+Hesitation duration and probability followed the same patterns as the majority of datasets.
 
 The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
-We, therefore, summarize briefly differences in the results for the transition-location effects.
-Results for the unconstrained version of the mixture model can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-The main difference between the unconstrained and the constrained model is that the latter does not allow the parameter that captures fluent transition durations to vary by transition location or otherwise.
-Instead it represents the lower bound of keystroke activity for uninhibited activation flow.
-Higher predictive performance for the unconstrained model suggests variability across transition locations for fluent transition durations, in particular for before-word transitions compared to mid-word transitions (all BFs > 100) but not for before-sentence transitions compared to before-word transitions (all BFs < 0.08, but SPL2 and GUNNEXP2: BFs > 100).
-The pattern for the hesitation probability was generally similar in both the constrained and unconstrained model.
-However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs > 10, except for LIFT: BF = 0.7); the hesitation duration pattern for before-sentence transitions compared to before-word transitions remained the same.
-
-<!-- TODO: address difference between posteriors of shift analysis for constrained and unconstrained model 
-- possibly not relevant: there are difference between the posteriors but that's not important for the differences between data sets
+We, therefore, summarise briefly differences in the results for the transition-location effects; full results can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
+Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs > 100) but not before-sentence compared to before-word transitions (all BFs < 0.08, but SPL2 and GUNNEXP2: BFs > 100).
+Patterns for hesitation probability was generally similar in both the constrained and unconstrained model.
+However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs > 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.
+
+<!-- TODO: 
+- address difference between posteriors of shift analysis for constrained and unconstrained model?
+- pausing frequency for words and sentences are equal but more importantly pauses occure less than 80% of the time indicating parallel planing.
+- would be really good, on some point, to distinguish between open and closed class words
 -->
@@
 
-The classical serial view [@flower1980ynamics] characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular single-distribution and two-distributions hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
+The classical serial view [@flower1980ynamics] characterizes writing as a sequence of planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
+The cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production [@olive2014toward].
+We used Bayesian statistical models, in particular hierarchical single-distribution and two-distribution mixture models, to directly compare the serial and the cascading model of writing, respectively.
+On the basis of six datasets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
 
 Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
+Under the cascading view of writing, key transitions that directly precede sentences or words are not necessarily associated with a pause because writers can plan in parallel to writing.
 On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
+We found that in the majority of datasets, pauses were as likely to occur before sentences and words, in both experienced and novice writers, while mid-word pauses are generally rare.
+Also, with few exceptions, pauses were typically longer before sentences than before words (which were longer than mid-word pauses).
+In other words, when writers do pause, the involvement of higher-level processing at sentence boundaries is typically leading to longer pauses.
 However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed,  they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, @olive2014toward described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre [see also @conijn2019understanding]:
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
+This is because our results for the hesitation probability suggest that writers often do not pause before sentences; indeed, hesitation probabilities are largely identical before sentences and before words.
+This suggest that writers frequently plan the next word and even sentence in parallel to production which mitigates the need to pause to plan up-coming linguistic information.
+
+Interestingly even the youngest writers in our data pool (C2L1) showed evidence of parallel planning expressed in fluent transitions between words and even sentences.
+Key-transitions were substantially longer compared to all other datasets with a larger slowdown before words.
 Yet the overall pausing probability showed a patterns comparable to most other data sets.
 In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets:  pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-**TODO: I don't have an explanation for this.** Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task [**REFERENCE**].
-
-A possible concern with our results -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
+This is interesting because, for example, @olive2014toward described a cascading model that is operating in a serial fashion for inexperienced or struggling writers explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
+However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers but pauses were more likely before words and sentences but similarly long indicating a tendency to prepare utterances in planning units possibly smaller than the clause.
+The main difference for these young writers was that their writing speed (fluent transitions) was substantially slower compared to more experienced writers in the sample.
+Also for L2 writers (PlanTra) we observe pausing patterns that do not resemble a serialised writing process as indicated by a larger number of fluent transitions across word and sentence boundaries.
+
+We observed two more differences between data sets -- other than the patterns observed for C2L1 summarised above -- that are worth highlighting.
+These can largely be explained on the basis of differences associated with the population (writing competence / experience) and text genre [see also @conijn2019understanding] that these samples were obtained from: The first exception concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations.
+**TODO: I don't have an explanation for this.** Second, pauses were not always longer before sentence compared to before words compared to mid-word paues; there were two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
+(2) pause durations did not vary by transition location in the LIFT dataset although the posterior showed the same general patterns as observed in most other datasets.
+The absence of a difference could be explained as specific to the synthesis writing task because there is no need for content creation [**REFERENCE?**] but the results for the PlanTra dataset rule out this possibility.
+
+A possible concern with our result -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+This concern is preempted by using cross-validation techniques for model comparison which is preventing model overfit.
 In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
 We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two-distributions -- and we can test whether these models can successfully uncover the true parameter value.
+This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
 Modelling details can be found in Appendix \ref{simulation}.
 As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a two-distributions process (but for the two-distributions data).
-Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of model overfitting.
+Parameters were not successfully uncovered when we switched model type and dataset.
+Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a single distributions process but a substantially higher predictive performance for data simulated from a two-distributions mixture process.
+Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of overfitting.
 
@@
 Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to investigate regressive eye movement behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.
-writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.
-spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
+There is limited research that used eye tracking to investigate regressive gaze behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
+There are at least four reasons why eye movements away from the inscription point might occure during writing: First, reading is necessary for revision (i.e.
+writers must check if writing goals have been met).
+Second, writers might use text to cue or reinstating ideas after interruption of writing process.
+Third, writers might look for or correct spelling or grammatical errors.
+We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text.
+Fourth, writers might look at the source their are writing from.
+For two datasets (PlanTra, LIFT) which involved writing from sources, we removed transitions that were made when the sources (rather than the writing document) was in focus.
+
+Frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
+This is except, lookbacks are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
+Finally even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
+This can be seen in the result of two datasets in our pool (CATO, GUNNEXP2): these experiments included a manipulation where the produced text was masked (or not).
 In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
+This suggest that pausing patterns do not change if reading previous text is not possible.
+
+A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
 Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent].
@@
 @wen06] they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
+For example, a sentence-initial pause of 2 secs has a very different interpretation from a 2 secs pause that occurs before or within a word.
 Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
@@
 However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose [@chenu2014interword].
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [ see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying].
+Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying].
+
+# Conclusion
+
+In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
+Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
+Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
+We found that key transitions across word and even sentences boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
+This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.
+young / L2 writers, students) and writing tasks (e.g.
+essays, syntheses).
 
@@
 -->
-
-# Conclusion
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.
-young / L2 writers, students) and writing tasks (e.g.
-essays, syntheses) included in our data pool.
 

### 2024-04-27 report/appendix/fit2data.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_fit2data.Rmd
@@
-## Fit to data
+# Fit to data
 
@@
 
-file <- list.files("../stanout", pattern = "sims", full.names = T, recursive = T)
+file <- list.files("../../stanout", pattern = "sims", full.names = T, recursive = T)
 ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
@@
 
-ggsave("figures/fitplots.pdf", plots_fin, height = 8.5, width = 6.25)
+ggsave("../figures/fitplots.pdf", plots_fin, height = 8.5, width = 6.25)
 

### 2024-04-27 report/appendix/genre_constr.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_genre_constr.Rmd
@@
-## Genre effect (LIFT)
+# Genre effect (LIFT)
 
@@
 
-file <- list.files(path = "../stanout/lift", pattern = "mogbetaco.*.csv$",
+file <- list.files(path = "../../stanout/lift", pattern = "mogbetaco.*.csv$",
                    full.names = T)
@@
 
-write_csv(tmp, "tables/genre_effect_cellmeans_constr.csv")
+write_csv(tmp, "../tables/genre_effect_cellmeans_constr.csv")
 

### 2024-04-27 report/appendix/masking.Rmd
rearrange markdown sections

--- 2023-07-24 report/appendix_e.Rmd
@@
-## Masking effect (CATO, GUNNEXP2)
+# Masking effect (CATO, GUNNEXP2)
 
@@
 
-file <- list.files(path = "../stanout/gunnexp2", 
+file <- list.files(path = "../../stanout/gunnexp2", 
                    pattern = "mogbetaun.*.csv$",
@@
 # get posterior
-file <- list.files(path = "../stanout/cato", 
+file <- list.files(path = "../../stanout/cato", 
                    pattern = "mogbetaun.+.csv$",
@@
 
-write_csv(tmp, "tables/masking_effect_cellmeans.csv")
+write_csv(tmp, "../tables/masking_effect_cellmeans.csv")
 

### 2024-04-27 report/appendix/masking_constr.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_masking_constr.Rmd
@@
 
-file <- list.files(path = "../stanout/gunnexp2", 
+file <- list.files(path = "../../stanout/gunnexp2", 
                    pattern = "mogbetaco.*.csv$",
@@
 # get posterior
-file <- list.files(path = "../stanout/cato", 
+file <- list.files(path = "../../stanout/cato", 
                    pattern = "mogbetaco.+.csv$",
@@
 
-write_csv(tmp, "tables/masking_effect_cellmeans_constr.csv")
+write_csv(tmp, "../tables/masking_effect_cellmeans_constr.csv")
 

### 2024-04-27 report/appendix/models.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_models.Rmd
@@
-## Statistical models
+# Statistical models
 
@@
 
-### Serial model of writing
+## Serial model of writing
 
-#### Single distribution Gaussian
+### Single distribution Gaussian
 
@@
 We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average IKI associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as $\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)$.
+This can be expressed as $\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)$.
 Of importance is the estimated posterior value of $\mu$ as this value captures that time it takes to mentally plan a word.
@@
 
-For computational ease, we implemented the differences associated with transition locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
+We implemented the average keytransition associated with a given text locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
 Therefore, the model will return one $\beta$ per transition location that capture the posterior distribution of average IKIs.
@@
 
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
+The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model
 
@@
 (\#eq:unimodgaus)
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
+\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
 \text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
+& u_\text{participant[i]} \sim \text{N}(0, \sigma_\text{p}^2)\\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
@@
 
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
+where $i$ iterates over every key-transition $i \in 1 \dots N$ with $N$ being the total number of keystroke intervals, and thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.
 
-#### Single distribution log-Gaussian
+Standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
+
+### Single distribution log-Gaussian
 
@@
 
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{N}()$ was replaced by $\text{logN}()$.
+The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{log}\mathcal{N}()$ was replaced by $\text{logN}()$.
 
@@
 (\#eq:unimodloggaus)
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
+\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
 \text{where: } &
 \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
@@
 
-#### Single distribution unequal-variance log-Gaussian
+### Single distribution unequal-variance log-Gaussian
 
@@
 (\#eq:unimoduv)
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
+\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
 \text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+ & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
  \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
@@
 
-### Parallel cascading model of writing
+## Parallel cascading model of writing
 
@@
 
-#### Two-distributions log-Gaussian mixture model (constrained)
+### Two-distributions log-Gaussian mixture model (constrained)
 
@@
 (\#eq:bimodcon)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
@@
 
-#### Two-distributions log-Gaussian mixture model (unconstrained)
+### Two-distributions log-Gaussian mixture model (unconstrained)
 
@@
 (\#eq:bimoduncon)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-	\text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+	\text{where: }  & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\

### 2024-04-27 report/appendix/postparam_constr.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_postparam_constr.Rmd
@@
-## Posterior parameter estimates
-
-```{r}
-# get posterior
-file <- list.files(path = "../stanout/cato", 
-                   pattern = "mogbetaco.+.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
-
-ps_beta <- filter(ps, param == "beta") %>% 
-  select(beta = value) %>% 
-  mutate(idx = 1:n())
-
-ps_cato <- ps %>% filter(param != "beta") %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta2, prob)) %>% 
-  mutate(idx = 1:n(), .by = where(is.character)) %>% 
-  left_join(ps_beta, by = join_by(idx)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(task, group, location, name)) %>% 
-  mutate(group = str_c("Task: ", task, "; Group: ", group),
-         ds = "CATO") %>% 
-  select(-task)
-
-file <- list.files(path = "../stanout/c2l1", 
-                   pattern = "mogbetacon.*.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
-
-ps_beta <- filter(ps, param == "beta") %>% 
-  select(beta = value) %>% 
-  mutate(idx = 1:n())
-
-ps_c2l1 <- ps %>% filter(param != "beta") %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta2, prob)) %>% 
-  mutate(idx = 1:n(), .by = where(is.character)) %>% 
-  left_join(ps_beta,by = join_by(idx)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>% 
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(location, name)) %>% 
-  mutate(ds = "C2L1",
-         group = "1")
-
-file <- list.files(path = "../stanout/gunnexp2", 
-                   pattern = "mogbetacon.*.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
-
-ps_beta <- filter(ps, param == "beta") %>% 
-  select(beta = value) %>% 
-  mutate(idx = 1:n())
-
-ps_gunn <- ps %>% filter(param != "beta") %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta2, prob)) %>% 
-  mutate(idx = 1:n(), .by = where(is.character)) %>% 
-  left_join(ps_beta, by = join_by(idx)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>%
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(location, xn, name)) %>% 
-  mutate(ds = "GUNNEXP2") %>% 
-  rename(group = xn)
-
-file <- list.files(path = "../stanout/lift", 
-                   pattern = "mogbetaco.+.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
-
-ps_beta <- filter(ps, param == "beta") %>% 
-  select(beta = value) %>% 
-  mutate(idx = 1:n())
-
-ps_lift <- ps %>% filter(param != "beta") %>% 
-  mutate(across(genre, ~ifelse(is.na(genre), "Arg", .))) %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta2, prob)) %>% 
-  mutate(idx = 1:n(), .by = where(is.character)) %>% 
-  left_join(ps_beta, by = join_by(idx)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>%
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(genre, topic, location, name)) %>% 
-  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre),
-         ds = "LIFT") %>% 
-  select(-topic, -genre)
-
-file <- list.files(path = "../stanout/plantra", 
-                   pattern = "mogbetaco.+.csv$",
-                   full.names = T)
-
-ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
-
-ps_beta <- filter(ps, param == "beta") %>% 
-  select(beta = value) %>% 
-  mutate(idx = 1:n())
-
-ps_plantra <- ps %>% filter(param != "beta") %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(beta2, prob)) %>% 
-  mutate(idx = 1:n(), .by = where(is.character)) %>% 
-  left_join(ps_beta, by = join_by(idx)) %>% 
-  mutate(across(c(beta2, beta), exp),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>%
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper), .names = "{.fn}"),
-            .by = c(task, location, name)) %>% 
-  mutate(ds = "PLanTra") %>% 
-  rename(group = task)
-
-file1 <- list.files(path = "../stanout/spl2", 
-                    pattern = "mogbetaco.+.csv$",
-                    full.names = T)
-
-file2 <- list.files(path = "../stanout/spl2_shift", 
-                    pattern = "mogbetaco.+.csv$",
-                    full.names = T)
-
-ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
-                     mutate(data = .x)) %>% 
-  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
-                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
-  filter(param %in% c("beta", "delta", "prob"))
-
-ps_beta <- filter(ps, param == "beta") %>% 
-  select(beta = value, data) %>% 
-  mutate(idx = 1:n(), .by = data)
-
-ps_spl2_both <- ps %>% filter(param != "beta") %>% 
-  pivot_wider(names_from = param, values_from = value) %>% 
-  unnest(cols = c(delta, prob)) %>% 
-  mutate(idx = 1:n(), .by = where(is.character)) %>% 
-  left_join(ps_beta, by = join_by(idx, data)) %>% 
-  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
-         beta2 = beta + delta,
-         across(c(beta2, beta), ~exp(.)),
-         delta = beta2 - beta,
-         across(location, str_replace, " ", "\n")) %>% 
-  select(-beta2) %>% 
-  pivot_longer(c(beta, prob, delta)) %>%
-  summarise(across(value, list(mean = mean, 
-                               lower = lower, 
-                               upper = upper),
-                   .names = "{.fn}"),
-            .by = c(lang, location, name, data)) %>% 
-  mutate(ds = "SPL2",
-         group = str_c("Data: ", data, "; Language: ", lang)) %>% 
-  select(-lang)
-
-ps_spl2 <- ps_spl2_both %>% filter(data == "SPL2 (_^[shift])")
-
-# Combine all
-ps <- bind_rows(ps_cato, ps_c2l1, ps_gunn, ps_lift, ps_plantra, ps_spl2)
-```
+# Posterior parameter estimates
 
@@
                            bottom = "Transition location")
-ggsave("figures/psplots1constr.pdf", plot_fin_1, height = 8, width = 7)
+ggsave("../figures/psplots1constr.pdf", plot_fin_1, height = 8, width = 7)
 
@@
                            bottom = "Transition location")
-ggsave("figures/psplots2constr.pdf", plot_fin_2, height = 8, width = 7)
+ggsave("../figures/psplots2constr.pdf", plot_fin_2, height = 8, width = 7)
 ```

### 2024-04-27 report/appendix/prepost.Rmd
rearrange markdown sections

--- 2023-07-24 report/appendix_g.Rmd
@@
-## Pre-post test (PLanTra)
+# Pre-post test (PLanTra)
 
@@
 
-file <- list.files(path = "../stanout/plantra", 
+file <- list.files(path = "../../stanout/plantra", 
                    pattern = "mogbetaun.*.csv$",
@@
 
-write_csv(tmp, "tables/posttest_effect_cellmeans.csv")
+write_csv(tmp, "../tables/posttest_effect_cellmeans.csv")
 

### 2024-04-27 report/appendix/prepost_constr.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_prepost_constr.Rmd
@@
 
-file <- list.files(path = "../stanout/plantra", 
+file <- list.files(path = "../../stanout/plantra", 
                    pattern = "mogbetaco.*.csv$",
@@
 
-write_csv(tmp, "tables/posttest_effect_cellmeans_constr.csv")
+write_csv(tmp, "../tables/posttest_effect_cellmeans_constr.csv")
 

### 2024-04-27 report/appendix/shift.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_postparam_constr.Rmd
@@
-## Posterior parameter estimates
-
 ```{r}
 # get posterior
-file <- list.files(path = "../stanout/cato", 
+file <- list.files(path = "../../stanout/cato", 
                    pattern = "mogbetaco.+.csv$",
@@
 
-file <- list.files(path = "../stanout/c2l1", 
+file <- list.files(path = "../../stanout/c2l1", 
                    pattern = "mogbetacon.*.csv$",
@@
 
-file <- list.files(path = "../stanout/gunnexp2", 
+file <- list.files(path = "../../stanout/gunnexp2", 
                    pattern = "mogbetacon.*.csv$",
@@
 
-file <- list.files(path = "../stanout/lift", 
+file <- list.files(path = "../../stanout/lift", 
                    pattern = "mogbetaco.+.csv$",
@@
 
-file <- list.files(path = "../stanout/plantra", 
+file <- list.files(path = "../../stanout/plantra", 
                    pattern = "mogbetaco.+.csv$",
@@
 
-file1 <- list.files(path = "../stanout/spl2", 
+file1 <- list.files(path = "../../stanout/spl2", 
                     pattern = "mogbetaco.+.csv$",
@@
 
-file2 <- list.files(path = "../stanout/spl2_shift", 
+file2 <- list.files(path = "../../stanout/spl2_shift", 
                     pattern = "mogbetaco.+.csv$",
@@
 
-```{r eval = F}
+# Key-combination effect (unconstrained mixture model)
 
-theme_set(theme_few(base_size = 11) +
-            theme(strip.background = element_blank(),
-                  legend.position = "top",
-                  legend.justification = "right",
-                  panel.grid = element_blank(),
-                  panel.background = element_rect(fill = "transparent"), # bg of the panel
-                  plot.background = element_rect(fill = "transparent", color = NA)))
+The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+In other words, the pause before sentences sumed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets.
+Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
+Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.
 
-pdodge <- position_dodge(-.5)
-beta_plots <- list()
-delta_plots <- list()
-theta_plots <- list()
-datasets <- unique(ps$ds)
-size <- 2
-width <- 0
-linewidth <- .35
-beta_label <- "Fluent transitions\n(in msecs)"
+Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
+We compared the SPL2 data including and excluding the keystroke after shift.
+Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans2}.
 
-beta_plots <- map(datasets, ~filter(ps, 
-                                    name == "beta", 
-                                    location == location[1],
-                                    ds == .x) %>% 
-                    mutate(name = beta_label,
-                           group == "overall",
-                           location = "overall") %>% 
-                    unique() %>% 
-                    ggplot(aes(y = mean, 
-                               ymin = lower, 
-                               ymax = upper,
-                               x = location)) +
-                    geom_errorbar(width = width,
-                                  linewidth = linewidth,
-                                  alpha = .75,
-                                  position = pdodge) +
-                    geom_point(size = size,
-                               position = pdodge) +
-                    facet_grid(~name) +
-                    scale_colour_colorblind() +
-                    scale_y_continuous(labels = scales::comma) +
-                    labs(subtitle = .x) +
-                    theme(axis.title = element_blank(),
-                          legend.title = element_blank()) +
-                    guides(colour=guide_legend(nrow=1,byrow=TRUE),
-                           shape = guide_legend(nrow=1,byrow=TRUE)))
+```{r shiftcellmeans2, results = 'asis'}
+# cell means and data set difference
+cellmeans <- ps_spl2 %>% 
+  separate(group, into = c("group", "lang"), sep = "; ") %>%
+  select(-ds, -group) %>%
+  mutate(across(lang, ~str_remove(., "Language: ")),
+         across(location, ~str_replace(., "\\n", " "))) %>% 
+  pivot_wider(names_from = name, 
+              values_from = c(mean, lower, upper)) %>% 
+  mutate(across(ends_with("prob"), ~dezero(., 2)),
+         across(c(ends_with("beta"),
+                  ends_with("delta")), 
+                ~as.character(scales::comma(round(.))))) %>% 
+  pivot_longer(-c(lang, location, data), 
+               names_to = c(".value", "param"),
+               names_sep = "_") %>% 
+  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
+  select(-mean:-upper) %>% 
+  pivot_wider(names_from = data, 
+              values_from = est)
 
-delta_plots <- map(datasets, ~filter(ps, 
-                                     name == "delta", 
-                                     ds == .x) %>% 
-                     mutate(across(c(ds,group), ~factor(.)),
-                            across(c(mean, lower, upper), ~ . /1000),
-                            name = delta_label) %>% 
-                     ggplot(aes(y = mean, 
-                                ymin = lower, 
-                                ymax = upper,
-                                x = location,
-                                shape = group)) +
-                     geom_errorbar(width = width, 
-                                   alpha = .75, 
-                                   linewidth = linewidth,
-                                   position = pdodge) +
-                     geom_point(aes(colour = group), 
-                                size = size, 
-                                position = pdodge) +
-                     facet_grid(~name) +
-                     scale_shape_manual(values = 1:length(filter(ps, ds == .x) %>% 
-                                                            pull(group) %>% 
-                                                            factor())) +
-                     scale_colour_colorblind() +
-                     scale_y_log10(labels = scales::comma) +
-                     theme(axis.title = element_blank(),
-                           legend.title = element_blank()) +
-                     guides(colour=guide_legend(nrow=2,byrow=TRUE),
-                            shape = guide_legend(nrow=2,byrow=TRUE)))
+file1 <- list.files(path = "../../stanout/spl2", 
+                    pattern = "mogbetaun.+.csv$",
+                    full.names = T)
 
-theta_plots <- map(datasets, ~filter(ps, 
-                                     name == "prob", 
-                                     ds == .x) %>% 
-                     mutate(across(c(ds,group), ~factor(.)),
-                            name = theta_label) %>% 
-                     ggplot(aes(y = mean, 
-                                ymin = lower, 
-                                ymax = upper,
-                                x = location,
-                                shape = group)) +
-                     geom_errorbar(width = width,
-                                   linewidth = linewidth,
-                                   alpha = .75, 
-                                   position = pdodge) +
-                     geom_point(aes(colour = group), 
-                                size = size, 
-                                position = pdodge) +
-                     facet_grid(~name) +
-                     scale_shape_manual(values = 1:length(filter(ps, ds == .x) %>% 
-                                                            pull(group) %>% 
-                                                            factor())) +
-                     scale_colour_colorblind() +
-                     scale_y_continuous(labels = dezero_plot, 
-                                        limits = c(0, 1)) +
-                     theme(axis.title = element_blank(),
-                           legend.title = element_blank()) +
-                     guides(colour=guide_legend(nrow=2,byrow=TRUE),
-                            shape = guide_legend(nrow=2,byrow=TRUE)))
+file2 <- list.files(path = "../../stanout/spl2_shift", 
+                    pattern = "mogbetaun.+.csv$",
+                    full.names = T)
 
-# Combine parameter plots 
-plot_param <- list()
-plot_param <- map(1:6, ~beta_plots[[.]] + delta_plots[[.]] + theta_plots[[.]] +
-                    plot_layout(guides = "collect", widths = c(.45, 1, 1)) &
-                    theme(legend.position = "bottom"))
+tmp <- map_dfr(c(file1, file2), ~read_csv(.x) %>% 
+                 mutate(data = .x)) %>% 
+  filter(param %in% c("beta", "delta", "theta")) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
+  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
+         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
+  pivot_wider(names_from = data, values_from = value) %>% 
+  unnest(c(spl2, spl2_shift)) %>% 
+  mutate(diff = spl2_shift - spl2) %>% 
+  summarise(across(diff, list(mean = mean, 
+                              lower = lower, 
+                              upper = upper,
+                              BF = BF), 
+                   .names = "{.fn}"),
+            .by = c(lang, location, param)) %>% 
+  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
+         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
+         across(param, recode, theta = "prob")) %>% 
+  left_join(cellmeans) %>% 
+  arrange(param, location, lang) %>% 
+  filter(location == "before sentence") %>% 
+  select(param, 
+         Language = lang,
+         `_^[shift] + C` = ends_with(" + C)"),
+         `_^[shift]` = ends_with("FT])"),
+         "Difference" = mean, BF)
 
-plot_param[[2]] <- plot_param[[2]] & theme(legend.position = "none")
-plot_param[[3]] <- plot_param[[3]] & guides(colour=guide_legend(nrow=1,byrow=TRUE),
-                                            shape = guide_legend(nrow=1,byrow=TRUE))
+write_csv(tmp, "../tables/shift_effect_cellmeans.csv")
 
-plot_param[[5]] <- plot_param[[5]] & guides(colour=guide_legend(nrow=1,
-                                                                byrow=TRUE,
-                                                                reverse=TRUE),
-                                            shape = guide_legend(nrow=1,
-                                                                 byrow=TRUE,
-                                                                 reverse=TRUE))
+apa_table(tmp[,-1],
+          align = c("l", "r", "r", "r", "r"), 
+          caption = "Mixture model estimates for key transitions.
+Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95% PIs in brackets.",
+          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
+          font_size = "footnotesize",
+          longtable = T,
+          stub_indents = list("Fluent transitions" = 1:2,
+                              "Hesitation duration" = 3:4,
+                              "Hesitation probability" = 5:6) )
 
-plot_fin_1 <- plot_param[[1]] / plot_param[[2]] / plot_param[[3]]
-plot_fin_1 <- grid.arrange(patchworkGrob(plot_fin_1), 
-                           left = "Posterior estimate with 95% PIs",
-                           bottom = "Transition location")
-ggsave("figures/psplots1constr.pdf", plot_fin_1, height = 8, width = 7)
-
-plot_fin_2 <- plot_param[[4]] / plot_param[[5]] / plot_param[[6]]
-plot_fin_2 <- grid.arrange(patchworkGrob(plot_fin_2), 
-                           left = "Posterior estimate with 95% PIs",
-                           bottom = "Transition location")
-ggsave("figures/psplots2constr.pdf", plot_fin_2, height = 8, width = 7)
 ```
 
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1constr.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2constr.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
+Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift.
+Fluent key transitions were substantially longer when including the interval following the shift key.
+The slowdown for hesitations was affected too but the difference is numerically small.
+There was no conclusive evidence for an increased hesitation probability.

### 2024-04-27 report/appendix/shift_constr.Rmd
rearrange markdown sections

--- 2023-12-25 report/appendix_shift_constr.Rmd
@@
-## Key-combination effect (constrained mixture model)
+# Key-combination effect (constrained mixture model)
 
@@
 
-file1 <- list.files(path = "../stanout/spl2", 
+file1 <- list.files(path = "../../stanout/spl2", 
                     pattern = "mogbetacon.+.csv$",
@@
 
-file2 <- list.files(path = "../stanout/spl2_shift", 
+file2 <- list.files(path = "../../stanout/spl2_shift", 
                     pattern = "mogbetacon.+.csv$",
@@
 
-write_csv(tmp, "tables/shift_effect_cellmeans_constr.csv")
+write_csv(tmp, "../tables/shift_effect_cellmeans_constr.csv")
 

### 2024-04-27 report/appendix/sim.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_sim.Rmd
@@
-## Simulation
+# Simulation
 
@@
 We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-To address this concern we repeated a comparison for a single distribution model and a two-distrobutionsl mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart.
+To address this concern we repeated a comparison for a single distribution model and a two-distributions mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart.
 In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.
@@
 The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
-files <- list.files("../mixture-model-sim/data", full.names = T)
+files <- list.files("../../mixture-model-sim/data", full.names = T)
 
@@
 ```{r}
-mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
+mog1 <- read_csv("../../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
   mutate(data = "mog")
-mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
+mog2 <- read_csv("../../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
   mutate(data = "lm")
-lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
+lm1 <- read_csv("../../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
   mutate(data = "lm")
-lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
+lm2 <- read_csv("../../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
   mutate(data = "mog")
@@
 ```{r loossim, results='asis'}
-table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
+table <- read_csv("../../mixture-model-sim/stanout/modelcomparison.csv") %>% 
   select(1:6) %>% 

### 2024-04-27 report/appendix/transloc.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_transloc.Rmd
@@
-## Transition location effect (unconstrained mixture model)
+# Transition location effect (unconstrained mixture model)
 
 ```{r}
-files <- list.files(str_c("../stanout/", 
+files <- list.files(str_c("../../stanout/", 
                           c("lift", "spl2", "plantra", "cato", "c2l1", "gunnexp2")), 
@@
 ```{r loceffect, results='asis'}
-files <- list.files(str_c("../stanout/", 
+files <- list.files(str_c("../../stanout/", 
                           c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetaun.+.csv", full.names = T) # "spl2_shift",
@@
 
-write_csv(tmp, "tables/location_effect_alldata.csv")
+write_csv(tmp, "../tables/location_effect_alldata.csv")
 

### 2024-04-27 report/appendix/transloc_constr.Rmd
rearrange markdown sections

--- 2023-12-09 report/appendix_transloc_constr.Rmd
@@
-## Transition location effect (constrained mixture model)
+# Transition location effect (constrained mixture model)
 
@@
 
-files <- list.files(str_c("../stanout/", 
+files <- list.files(str_c("../../stanout/", 
                           c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetacon.+.csv", full.names = T) # "spl2_shift",
@@
 
-write_csv(tmp, "tables/location_effect_constr_alldata.csv")
+write_csv(tmp, "../tables/location_effect_constr_alldata.csv")
 

### 2024-04-27 report/manuscript.Rmd
rearrange markdown sections

@@
     keep_tex: TRUE
-    includes:
-      after_body: 
-        - "appendix_models.tex"
-        - "appendix_fit2data.tex"
-#        - "appendix_postparam.tex"
-        - "appendix_transloc.tex"
-        - "appendix_shift.tex"
-        - "appendix_shift_constr.tex"
-#        - "appendix_l2.tex"
-        - "appendix_l2_constr.tex"
-#        - "appendix_masking.tex"
-        - "appendix_masking_constr.tex"
-        - "appendix_postparam_constr.tex"
-#        - "appendix_prepost.tex"
-        - "appendix_prepost_constr.tex"
-#        - "appendix_genre.tex"
-        - "appendix_genre_constr.tex"
-        - "appendix_transloc_constr.tex"
-        - "appendix_sim.tex"
 
@@
 
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing [@bock2014syntactically; @wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].
+# Introduction
 
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking [@roeser2018advance; @torrance2012written].
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader [e.g.
-@Clark2002].
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06].
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses].
-The same is generally observed for spoken utterances [@lee13; @meyer1996; @wheeldon2013] with some exceptions [@gri01].
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements [@van1991handwriting; @olive2014toward].
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing].
-This is particularly reflected in the model proposed by @flower1980ynamics that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts [also @chenoweth2001fluency; @chenoweth2003inner].
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second].
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production [e.g.
-@bock2014syntactically].
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases [@not10; dam09; @roeser2018advance].
-For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e.
-*A and the B moved*) or a simple noun (i.e., *A moved above the B and the C*).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e.
-*the B*) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, @medimorec2017pauses found that, in undergraduate students writing on familiar topics 71% of sentence-initial IKIs were less than 1 sec in duration.
-Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects [@torrance2018timed] and short sentences when describing arrays of images [@roeser2018advance] in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction].
-@van1991handwriting, for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now].
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.
-just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling].
-
-For example, @roeser2021modelling demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-@ronneberg2022process coined this possibility the "process-disruption hypothesis".
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].
-
-Data form spontaneous text production [see @gernsbacher1995coherence] are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy [@dux2022automating;@dux2021effect;@vandermeulen2023writing].
-
-Second, since the emergence of software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing [@conijn2021timecourse].
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning [@conijn2019understanding; @conijn2020keys].
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments [@not10; @dam09; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual].
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
+```{r child = "sections/introduction.Rmd"}
+```
 
@@
 
-In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts.
-Full modelling details can be found in Appendix \ref{statistical-models; we summarise the most important model properties in the following.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data.
-In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.
-
-The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in @roeser2021modelling.
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes.
-These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs.
-This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.
-
-We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts.
-We discussed advantages with this approach in the introduction section.
-These datasets include samples from populations with various writing experience and languages (e.g.
-young / L2 writers, students) performing different writing tasks (e.g.
-essays, syntheses).
-Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.
-
-## Data sets
-
-TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2.
-For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say "unpublished".
-I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk).
-Was the GUNNEXP2 sample Norwegian?
-Should we include information like time limit or text length in the table?
-
-Six datasets with keystroke data from free text production were used for analysis.
-An overview can be found in Table \ref{tab:datasets}; descriptions below.
-Dataset used are C2l1 [@ronneberg2022process], CATO [@torrance2016adolescent], GUNNEXP2 [@torranceb], SPL2 [@torrancea], LIFT published in @vandermeulen2020 and described in @vandermeulen2020mapping, PLanTra data are published in @rossetti2022text and described in @rossetti2022s.
-In C2L1 Norwegian 6th graders composed argumentative essays.
-In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text.
-In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced.
-The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each.
-The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability.
-The GUNNEXP2 data ....
-[MARK, please add a sentence here to describe the production context].
-
-For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite [@sim07; @torrance201203].
-LIFT and PLanTra data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019] and SPL2 data were collected using CyWrite [@chukharev2019combined].
-
-\blandscape
-```{r datasets, results='asis'}
-tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
-                Source = c("Rønneberg et al. (2022)",
-                           "Torrance et al. (2016)",
-                           "Torrance et al. (n.d.)",
-                           "Rossetti and Van Waes (2022b)",
-                           "Vandermeulen, Steendam, et al. (2020)", 
-                           "Torrance and Ofstad (n.d)"),
-                Keylogger = c("EyeWrite", 
-                           "EyeWrite", 
-                           "CyWrite", 
-                           "InputLog", 
-                           "InputLog",
-                           "EyeWrite"),
-                `Writing task` = c("Argumentative essays", 
-                                 "Expository texts",
-                                 "Argumentative essays",
-                                 "Text simplification",
-                                 "Synthesis", 
-                                 "?????"),
-                 N = as.character(c(126, 26*2, 39, 47, 658, 45)),
-#                n_texts = c(1, 2, 2, 2, NA),
-#                n_sentences = c(),
-#                n_words = c(),
-                `Mean age` = c(11.8, 16.9, 20.6, 23, 17.0, "????"),
-                 Sample = c("6th graders", "Secondary school students", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
-                 Country = c("Norway", "Norway", "USA", "Belgium", "The Netherlands", "???? Norway"),
-                 Language = c( "Norwegian", 
-                               "Norwegian", 
-                               "English (L1) / Spanish (L2)", 
-                               "English (L2)", 
-                               "Dutch",
-                               "???? Norwegian")) %>% 
-  arrange(Dataset) %>% 
-  apa_table(align = c("l", "p{3cm}", "p{2cm}", "p{2.5cm}", "r", "r", "p{2.5cm}", "p{2.25cm}", "p{2.5cm}"), 
-            escape = FALSE, 
-            font_size = "footnotesize",
-            caption = "Datasets in brief.")
-
-#                Manipulation = c("--", 
-#                               "Weak decoders / control; texted masked / unmasked",
-#                               "Text produced in L1 / L2",
-#                               "Pre / post test with training on plain language principles", 
-#                              "Various topics and genres",
-#                               "masked / unmasked")
+```{r child = "sections/methods.Rmd"}
 ```
-\elandscape
-
-## Data extraction
-
-From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful [@chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-Also we removed transitions that were followed by an editing operation.
-
-```{r keyloc, results='asis'}
-
-table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
-       Description = c("Transitions between any letter",
-                       "Keypress after space followed by any letter",
-                       "Keypress following a space preceding any letter"),
-       Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d.
-T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.",
-                   "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
-                   "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept."))
-
-apa_table(table, caption = "Transition location classification.",
-            align = c("p{3cm}", "p{5cm}", "p{6cm}"), 
-            escape = FALSE, 
-            font_size = "footnotesize",
-            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace.")
-
-```
-
-```{r reductionfunctions}
-source("scripts/get_data_summary.R")
-
-# Get info about data reduction
-n_samples <- 100
-c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
-cato <- get_cato("../data/cato.csv", n_samples = n_samples)
-lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
-plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
-spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
-gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
-```
-
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-We further removed keystroke intervals that were too short to represent a behavioural response ($\le$ 50 msecs) or too long ($\ge$ 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled a maximum of `r n_samples` observations per participant, condition, and transition location (when more than `r n_samples` were available).
-This was done to reduce the computation time of the Bayesian models require to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets.
-Because the LIFT data set included the largest number of writing tasks, we sampled 50 observations per condition, location and participant which would otherwise exceed the computational resources available to us.
-The number of keystroke data used in the analysis can be found in Table \ref{tab:datareduction}.
-
-```{r datareduction, results='asis'}
-c2l1$ds <- "c2l1"
-cato$ds <- "cato"
-lift$ds <- "lift"
-plantra$ds <- "plantra"
-spl2$ds <- "spl2"
-gunnexp2$ds <- "gunnexp2"
-
-extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
-          mutate(ds = .x$ds)) %>% 
-  mutate(across(where(is.numeric), ~signif(.*100, 2))) %>% 
-  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
-  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
-  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")")),
-         across(c(`<50ms`, `>30,000ms`), ~str_replace(., "0 \\(0\\)", "--"))) %>% 
-  rename(`$\\le$ 50 msecs` = `<50ms`,
-         `$\\ge$ 30 secs` = `>30,000ms`)
-
-random_sample <- 
-  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
-          mutate(ds = .x$ds)) %>% 
-    mutate(across(where(is.numeric), ~signif(.*100, 2))) %>%
-    unite("keep", starts_with("keep"), sep = " (") %>% 
-    mutate(across(keep, ~str_c(., ")")),
-          across(keep, ~str_replace(., "100 \\(0\\)", "--"))) %>%
-  pivot_wider(names_from = location, values_from = keep) %>% 
-  relocate(`within word`)
-
-datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
-  relocate(ds) %>% 
-  mutate(across(ds, ~recode_factor(., 
-                "c2l1" = "C2L1",
-                "cato" = "CATO",
-                "gunnexp2" = "GUNNEXP2",
-                "lift" = "LIFT",
-                "plantra" = "PLanTra",
-#                "spl2_shift" = "SPL2 (shift + C)", 
-                "spl2" = "SPL2",
-                .ordered = T
-                ))) %>% 
-  arrange(ds) %>% 
-  rename(`Dataset` = ds)
-
-apa_table(datareduction, 
-          escape = FALSE,  
-          placement = "bp!",
-          caption = "Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location (`--` if all data were used).
-Standard error in parentheses.",
-          align =c("l", rep("r", 5)),
-          col_spanners = list(" " = 1,
-                              "Extreme values in \\%" = c(2, 3),
-                              "Randomly sampled data in \\%" = c(4, 6)))
-
-```
-
-## Bayesian statitical modelling
-
-We reanalysed keystroke data -- transition durations measured as the time between first and second keypress -- from six datasets in a series of five Bayesian models.
-An overview of all models can be found in Table \ref{tab:models}.
-More detailed model descriptions and motivation can be found in Appendix \ref{statistical-models}.
-
-```{r models, results = 'asis'}
-models <- tibble(Models = str_c("M",1:5),
-       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
-       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
-#       group = c(rep("Serial", 3), rep("Cascading", 2)),
-       Description = c("Single Gaussian model with effects of transition location (and experiment-specific manipulation).",
-                       "Single distribution log-Gaussian model equivialent to M1.",
-                       "Equivalent to M2 but with different variance components for each transition location (unequal variance).",
-                       "Two-distributions mixture of a log-Gaussian for fluent transition intervals and another log-Gaussian that is wider than the previous distribution hesitant / disfluent key transitions; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition location but the distribution of fluent intervals has the same mean across all design cell.
-Mixing proportions capture the relative number of disfluent transitions for the distribution of hesitant transitions for each combination of transition location.",
-                       "Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location.")) %>% 
-  select(Models, Description)
-
-apa_table(models, 
-          align = c(rep("l", 1), "p{13cm}"), 
-          escape = FALSE, 
-          digits = 0,
-          caption = "Overview of cognitive models of writing. All models included by-participant random effect.",
-          stub_indents = list(
-                     "\\textbf{Serial}" = 1:3,
-                     "\\textbf{Cascading}" = 4:5))
-
-```
-
-Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset-specific manipulations.
-In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
-For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; this is because slower human behaviour is generally known to be associated with a larger variability [@wagenmakers2007linear; @wing1973response; @schoner2002timing]; see Appendix \ref{statistical-models} for rationale.
-
-Models M4 and M5 are extensions of the unequal-variance single distribution model M3.
-Importantly models M4 and M5 assume that keystroke data result from a combination of two data generating processes -- consistent with the cascading account -- rather than one.
-These model are called finite mixture models in the literature [@gelman2014; @roeser2021modelling; @peel2000finite] and capture the assumption that processing at higher levels of activation leads to longer pauses which can in principle happen at transition before sentence, before words or even mid-word.
-In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that pauses at larger linguistic edges are more likely but not obligated by the cognitive system that generates keystrokes.
-This is achieved by modelling key transitions as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
-
-1.
-Information from upstream mental processes can flow into keystrokes without interruption at intermediate levels.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger.
-The population estimate for fluent key transitions (i.e.
-typing speed) is captured by the $\beta$ parameter.
-
-2.
-Any interruptions at upstream levels of mental representation result in a delay in the information flow and therefore create a lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
-The population estimate of the hesitation slowdown is captured by $\delta$ and its frequency is captured by the mixing weight (proportion) $\theta$.
-
-\begin{equation}
-\begin{aligned}
-(\#eq:bimodcon2)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2\\
-		& 0 < \theta < 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents the distribution of fluent -- uninterrupted -- key transitions.
-Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
-$\theta$ is here parameterised to represent the probability that a key transition is associated with the distribution of hesitant transitions.
-This probability is inversely related to the mixing weight of the distribution of fluent transitions by $1-\theta$ .
-In other words, a larger weight for either distribution inevitably means a lower weight for the other.
-We call this parameter the probability of hesitant transitions.
-When a parameter was assumed to vary by transition location (levels: before sentence, before word, within word) this was indicated as subscript; similar for participants.
-The hesitation probability was allowed to vary by participants constrained by a hyper-parameter for each transition location.
-
-The constrained implementation M4 assume that the size of a fluent key-transition does not vary by transition location.
-In other words, the value that the $\beta$ parameter is taking on is the same for transitions at before-sentence, before-word, and within-word locations.
-However, letter bigrams may be executed systematically faster than transitions between space and either a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
-This is for two reasons: first, and importantly, highly frequent bi- and even trigrams are stored, retrieved and executed as one motor code [@behmer2016crunching; @grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants] because of the strong association between letters.
-This association is reducing the length of transition durations.
-It is less likely that the same association exists for space-letter and space-shift pairs.
-Second, before-sentence transition in two datasets (PLanTra, LIFT) are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transition; we return to this in the results section.
-We therefore implemented an unconstrained mixture model M5 that assumes that the distribution over durations of fluent transitions -- represented as $\beta$ -- may vary across transition locations.
-
-All models were implemented in the Bayesian framework [@gelman2014; @mcelreath2016statistical].
-The Bayesian framework is ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty [@farrell2018computational; @gelman2014; @lee2014bayesian].
-To achieve this, Bayesian models require the inclusion of prior information, i.e.
-existing knowledge about parameter values.
-For small datasets priors influence the inferred parameter values; for larger datasets weakly informative and vague priors are quickly overcome by the data [i.e.
-automatic Ockham's razor, @jefferys1992ockham].
-In other words the choice of priors has less impact on the posterior, certainly for weakly informative priors.
-In the present analysis, we used weakly informative priors to aid model convergence by constraining the parameter space to plausible values [see e.g.
-@lambert2018student; @mcelreath2016statistical].
-Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
-
-Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); for a tutorial see ADD URL HERE.^[The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical; @sivula2020uncertainty].]
 
@@
 
-## Model comparisons
-
-To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
-Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared by  the difference between models $\Delta\widehat{elpd}$.
-We also summarised this difference as normalised over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
-Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.
-
-```{r}
-files <- list.files(
-  str_c("../stanout/", 
-  c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
-  pattern = "model.+.csv", 
-  full.names = T)
-
-table <- purrr::map_dfr(files, ~read_csv(.) %>% 
-                          mutate(dataset = .x)) %>% 
-  select(dataset, model, contains("elpd_diff"), se_diff) %>%
-  arrange(dataset, desc(elpd_diff)) %>% 
-  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(where(is.numeric), ~round(., 0)),
-         across(where(is.numeric), ~format(., big.mark = ",")),
-         across(everything(), ~str_trim(.))) %>%
-  filter(dataset != "spl2_shift") %>% 
-  transmute(
-    Comparison = model,
-    elpd_diff_ratio,
-     across(dataset, ~recode(., 
-                "cato" = "CATO",
-                "spl2" = "SPL2", 
-#                "spl2_shift" = "SPL2 (shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                "gunnexp2" = "GUNNEXP2")),
-    Distribution = case_when(str_detect(model, "M2$") ~ "Single log-Gaussian (M2)",
-                            str_detect(model, "M3$") ~ "Single log-Gaussian (unequal variance; M3)",
-                            str_detect(model, "M4$") ~ "Mixture of two log-Gaussians (constrained; M4)",
-                            str_detect(model, "M5$") ~ "Mixture of two log-Gaussians (unconstrained; M5)"),
-            across(elpd_diff, ~str_c(., " (", se_diff, ")")))
-
-table_ratios <- table %>% select(dataset, Comparison, elpd_diff_ratio)
-
-table <- table %>% select(-elpd_diff_ratio) %>% 
-  pivot_wider(names_from = dataset, values_from = elpd_diff)
-
-table_ratios <- table_ratios %>% 
-  mutate(across(elpd_diff_ratio, as.numeric)) %>% 
-  summarise(across(elpd_diff_ratio, list(min = min, max = max), .names = "{.fn}"), .by = Comparison) %>% 
-  mutate(range = str_c("[", min, ", ", max, "]"))
-
-tr_m2 <- filter(table_ratios, str_detect(Comparison, "M2$")) %>% pull(range)
-tr_m3 <- filter(table_ratios, str_detect(Comparison, "M3$")) %>% pull(range)
-tr_m4 <- filter(table_ratios, str_detect(Comparison, "M4$")) %>% pull(range)
-tr_m5 <- filter(table_ratios, str_detect(Comparison, "M5$")) %>% pull(range)
-
+```{r child = "sections/results.Rmd"}
 ```
 
-Model differences are shown in Table \ref{tab:loos}.
-All datasets revealed the same pattern.
-Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single-distribution model; the difference $\Delta\widehat{elpd}$ between models ranged within `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model.
-The unconstrained mixture model M5 rendered a higher predictive performance than the constrained model M4 which did not allow the distribution of fluent keystroke-intervals to vary by transition location; the difference in predictive performance was between `r tr_m5` standard errors higher for the unconstrained model.
-Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; the difference $\Delta\widehat{elpd}$ ranged within `r tr_m3` standard errors.
-The weakest model was the single distribution Gaussian which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of `r tr_m2` standard errors.
-
-\blandscape
-```{r loos, results='asis'}
-#names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
-#names(table) <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$"), 1))
-
-table %>% 
-  arrange(desc(Comparison)) %>% 
-  apa_table(caption = "Model comparisons.
-Models were compared incrementally from the simplest to the most complex model.
-The comparison is shown along with the distribution type of the more complex model.
-A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model.
-Standard error is shown in parentheses.",
-      align =c("p{1.5cm}", "p{5cm}", rep("r", 6)), 
-      escape = FALSE,
-      longtable = T,
-      font_size = "footnotesize",
-      note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density")
-#                col_spanners = list(" " = 1:2,
-#                     "\\textbf{CL21}" = 1:5,
-#                     "\\textbf{CATO}" = 6:10,
-#                     "\\textbf{GUNNEXP2}" = 11:15, 
-#                     "\\textbf{LIFT}" = 16:20,
-#                     "\\textbf{PLanTra}" = 21:25,
-#                     "\\textbf{SPL2}" = 26:30))
-#      stub_indents = list(
-#                     "\\textbf{CL21}" = 1:5,
-#                     "\\textbf{CATO}" = 6:10,
-#                     "\\textbf{GUNNEXP2}" = 11:15, 
-#                     "\\textbf{LIFT}" = 16:20,
-#                    "\\textbf{PLanTra}" = 21:25,
-#                     "\\textbf{SPL2}" = 26:30))
-#                     "\\textbf{SPL2 (shift + C)}" = 31:35))
-
-```
-\elandscape
-
-We also tested to what extent model predictions fit observed data.
-Comparisons can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
-Data predicted by the two-distribution mixture models are similar to observed data; data predicted by single distribution Gaussian models showed a poor fit to the data.
-
-## Transition location effect
-
-As demonstrated in the model comparisons, the two-distributions mixture-models is a better statistical representation on the underlying process generating keystroke data than compared to any of the single-distribution models.
-Information about this process is captured by the posterior estimates of the models' parameter values.
-To reiterate, there are three conceptually important parameters: (1) the magnitude of the slowdown for transition hesitations (represented as $\delta$ in equation \ref{eq:bimodcon2}), (2) the probability of hesitant transitions (represented as $\theta$ in equation \ref{eq:bimodcon2}), and (3) the duration of fluent transitions (the typing speed, which was indicated as $\beta$ in equation \ref{eq:bimodcon2}).
-
-Figure \ref{fig:mixmodel} illustrates the mixture model approach on the basis of the posterior of the constrained implementation applied to the CATO data: for each transition durations we obtain two posterior probability distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in hesitant transitions.
-The hesitation probability is indicated as the relative height of either distribution.
-The transition locations are characterized by the hesitation parameters: as we illustrate the constrained mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location but their height differs as the probability of fluent transition is the inverse of the height of hesitant transitions; i.e.
-$1 - \theta$.
-More interesting are the distributions associated with hesitations; indicated in yellow: Within word transitions have a negligibly small hesitation probability barely visible in the visualisation.
-Hesitations before words and sentences are both roughly equally likely than fluent transitions.
-In other words, half of the time participants did not pause before a word or sentence; we explore this below.
-However when hesitations occurred, these are longer at before-sentence transitions compared to before word transitions.
-
-```{r mixmodel, fig.cap='Mixture model plot for each transition location.
-Shown are the distributions for both mixture components (fluent and hesitant transitions) and their weighting.
-Posterior of CATO data.', fig.height=6}
-
-ps <- read_csv("../stanout/cato/mog_constr_posterior_for_densityplot.csv") 
-# Extract parameters
-alpha <- filter(ps, param == 'beta') %>% pull(est)
-alpha2_bs <- filter(ps, param == 'delta', loc == 'before sentence') %>% pull(est)
-alpha2_bw <- filter(ps, param == 'delta', loc == 'before word') %>% pull(est)
-alpha2_ww <- filter(ps, param == 'delta', loc == 'within word') %>% pull(est)
-
-theta_bs <- filter(ps, param == 'prob',  loc == 'before sentence') %>% pull(est)
-theta_bw <- filter(ps, param == 'prob',  loc == 'before word') %>% pull(est)
-theta_ww <- filter(ps, param == 'prob',  loc == 'within word') %>% pull(est)
-
-sigma <- filter(ps, param == 'sigma') %>% pull(est)
-sigmap_bs <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'before sentence')) %>% pull(est) %>% sum()
-sigmap_bw <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'before word')) %>% pull(est) %>% sum()
-sigmap_ww <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'within word')) %>% pull(est) %>% sum()
-
-# Data grid
-data <- tibble(x = c(0, 20))
-
-colours <- c("Fluent" = "#000000", "Hesitant" = "#E69F00")
-
-# Create density plot
-plot_bs <- ggplot(data, aes(x)) +
-  stat_function(aes(colour = "Fluent"),
-                geom = "line", 
-                linewidth = .25,
-                fun = plot_mix_comps,
-                args = list(alpha, sigma, lam = 1 - theta_bs)) +
-  stat_function(aes(colour = "Hesitant"),
-                geom = "line", 
-                linewidth = .25,
-                fun = plot_mix_comps, 
-                args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
-  stat_function(aes(fill = "Fluent"),
-                geom = "area", 
-                fun = plot_mix_comps, 
-                alpha = .25, 
-                args = list(alpha, sigma, lam = 1 - theta_bs)) +
-  stat_function(aes(fill = "Hesitant"),
-                geom = "area", 
-                fun = plot_mix_comps, 
-                alpha = .25, 
-                args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
-  scale_x_continuous(labels = exp, 
-                     breaks = log(c(1, 10, 100, 1000, 10000)), 
-                     limits = log(c(10, 50000))) +
-  scale_colour_manual(values = colours) +
-  scale_fill_manual(values = colours) +
-  labs(y = "", x = "", 
-       subtitle = "Before-sentence transitions",
-       colour = "Mixture\ncomponent",
-       fill = "Mixture\ncomponent") +
-  theme(legend.position = "right",
-        legend.justification = "top")
-
-plot_bw <- ggplot(data, aes(x)) +
-  stat_function(aes(colour = "Fluent"),
-                linewidth = .25,
-                geom = "line", 
-                fun = plot_mix_comps,
-                args = list(alpha, sigma, lam = 1 - theta_bw)) +
-  stat_function(aes(colour = "Hesitant"),
-                geom = "line", 
-                linewidth = .25,
-                fun = plot_mix_comps, 
-                args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
-  stat_function(aes(fill = "Fluent"),
-                geom = "area", 
-                linewidth = .25,
-                fun = plot_mix_comps, 
-                alpha = .25, 
-                args = list(alpha, sigma, lam = 1 - theta_bw)) +
-  stat_function(aes(fill = "Hesitant"),
-                geom = "area", 
-                fun = plot_mix_comps, 
-                alpha = .25, 
-                args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
-  scale_x_continuous(labels = exp, 
-                     breaks = log(c(1, 10, 100, 1000, 10000)), 
-                     limits = log(c(10, 50000))) +
-  scale_colour_manual(values = colours) +
-  scale_fill_manual(values = colours) +
-  labs(y = "Density", x = "", 
-       subtitle = "Before-word transitions",
-       colour = "Mixture\ncomponent",
-       fill = "Mixture\ncomponent") +
-  theme(legend.position = "none")
-
-plot_ww <- ggplot(data, aes(x)) +
-  stat_function(aes(colour = "Fluent"),
-                geom = "line", 
-                fun = plot_mix_comps,
-                linewidth = .25,
-                args = list(alpha, sigma, lam = 1 - theta_ww)) +
-  stat_function(aes(colour = "Hesitant"),
-                geom = "line", 
-                fun = plot_mix_comps, 
-                linewidth = .25,
-                args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
-  stat_function(aes(fill = "Fluent"),
-                geom = "area", 
-                fun = plot_mix_comps, 
-                alpha = .25, 
-                args = list(alpha, sigma, lam = 1 - theta_ww)) +
-  stat_function(aes(fill = "Hesitant"),
-                geom = "area", 
-                fun = plot_mix_comps, 
-                alpha = .25, 
-                args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
-  scale_x_continuous(labels = exp, 
-                     breaks = log(c(1, 10, 100, 1000, 10000)), 
-                     limits = log(c(10, 50000))) +
-  scale_colour_manual(values = colours) +
-  scale_fill_manual(values = colours) +
-  labs(subtitle = "Within-word transitions", 
-       y = "", x = "Transition duration [in msecs]",
-       colour = "Mixture\ncomponent",
-       fill = "Mixture\ncomponent") +
-  theme(legend.position = "none")
-
-plot_bs / plot_bw / plot_ww
-
-```
-
-We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-We demonstrated above that the predictive performance of the unconstrained mixture model was substantially higher than the predictive performance of the constrained mixture model.
-However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
-We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following.
-
-```{r}
-files <- list.files(str_c("../stanout/", 
-                          c("lift", "spl2", "plantra", "cato", "c2l1", "gunnexp2")), 
-                    pattern = "mogbetaco.+.csv", 
-                    full.names = T)
-
-ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
-  mutate(data = .x)) %>% 
-  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, ~recode_factor(., 
-                "cato" = "CATO",# (non-dyslexic, unmasked)",
-                "gunnexp2" = "GUNNEXP2",# (unmasked)",
-                "spl2" = "SPL2", # (L1)", 
-#                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
-                .ordered = TRUE))) %>% 
-  filter(!(str_detect(data, "SPL2") & lang == "ES" & !is.na(lang)),
-         !(str_detect(data, "CATO") & group == "dyslexic" & !is.na(group)),
-         !(str_detect(data, "CATO") & task == "masked" & !is.na(task)),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked" & !is.na(xn)))
-
-ps_beta <- filter(ps, param == "beta") %>% 
-  rename(beta = value) %>%
-  mutate(idx = 1:n(), .by = data) %>% 
-  select(idx, beta, data)
-
-ps <- filter(ps, param != "beta")  %>% 
-  pivot_wider(names_from = param, 
-              values_from = value) %>% 
-  unnest(cols = beta2:theta) %>%
-  mutate(idx = 1:n(), .by = c(data, location, task, xn, genre, topic, lang)) %>% 
-#  summarise(across(idx, list(max = max)), .by = c(location, data, task))
-  left_join(ps_beta, by = c("data", "idx")) %>%
-  mutate(across(c(beta, beta2), ~exp(.)),
-         delta = beta2 - beta,
-         across(location, ~str_replace(., " ", "\n"))) %>% 
-  pivot_longer(c(beta, beta2, delta, prob)) %>% 
-  summarise(across(value, 
-                   list(mean = mean, 
-                        lower = lower, 
-                        upper = upper),
-                   .names = "{.fn}"), 
-            .by = c(data, location, name)) %>% 
-  filter(name %in% c("beta", "beta2", "delta", "prob"))
-```
-
-The mixture-model posterior parameter estimates for all datasets is visualised in Figure \ref{fig:crossstudypost2}.
-Although models were fitted with all data set-specific conditions, we aggregated the posterior across conditions^[We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.], and excluded conditions that might conflate comparisons^[We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-gunnexp2}).].
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences and similarities associated with transition locations for all dataset.
-In particular we can see how the linguistic location in the text is associated with hesitation duration and probability.
-The estimate for fluent transitions is shown by dataset, across transition location.
-
-```{r crossstudypost2, fig.cap="Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group."}
-
-posd <- position_dodge(.5)
-dotsize <- 2.75
-grouplabel <- "Data set:"
-shapes <- c(1, 4, 6, 7, 8, 9, 11)
-width <- 0
-linewidth <- .35
-nrow <- 1
-beta <- filter(ps, name == "beta", 
-               location == "before\nword") %>%
-  mutate(name = beta_label,
-         location = "overall") %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(data))) +
-  geom_line(linewidth = linewidth, 
-            alpha = .75, 
-            position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
-  facet_grid(~name, scales = "free_x") +
-  scale_shape_manual(values = shapes) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = scales::comma) +
-  theme(axis.title = element_blank()) +
-  labs(colour = grouplabel,
-       shape = grouplabel) +
-  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
-         shape = guide_legend(nrow=nrow,byrow=TRUE))
-
-delta <- filter(ps, name == "delta") %>% 
-  mutate(name = delta_label,
-         across(c(mean, lower, upper), ~./1000)) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(data))) +
-  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = scales::comma) +
-  labs(colour = grouplabel,
-       shape = grouplabel) +
-  theme(axis.title = element_blank()) +
-  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
-         shape = guide_legend(nrow=nrow,byrow=TRUE))
-
-theta <- filter(ps, name == "prob") %>% 
-  mutate(name = theta_label) %>% 
-  ggplot(aes(y = mean, 
-             ymin = lower, 
-             ymax = upper,
-             x = location,
-             shape = data,
-             group = interaction(data))) +
-  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
-  facet_grid(~name) +
-  scale_shape_manual(values = shapes) +
-  scale_colour_colorblind() +
-  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
-  labs(colour = grouplabel,
-       shape = grouplabel) +
-  theme(axis.title = element_blank()) +
-  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
-         shape = guide_legend(nrow=nrow,byrow=TRUE))
-
-plot <- beta + delta + theta + plot_layout(guides = "collect",
-                                           widths = c(.5, 1, 1))
-
-grid.arrange(patchworkGrob(plot), 
-             left = "Posterior estimate with 95% PIs",
-             bottom = "Transition location")
-
-```
-
-Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets.
-We summarise the results of the pairwise comparisons for transitions locations; full results and parameter estimates can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
-We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs > 100 for all datasets) but only for half of the data sets also longer (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
-Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GUNNEXP2 and SPL2, both BFs > 100) but longer (all BFs > 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
-
-To reiterate, for most datasets we found that pauses are before sentences are longer but not necessarily more likely than before words (except for datasets SPL2 and GUNNEXP2).
-In fact, LIFT showed some indication that pauses were less likely before sentences than words.
-This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for key transitions at larger linguistic edges.
-It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.
-
-Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character while the remaining data sets measured before-sentence transitions as the difference between space key and shift.
-To control for the use of complex key combinations, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability (BFs > 100) but it did not affect the duration of fluent or hesitant transitions (BFs < 0.3); see Appendix \ref{key-combination-effect-constrained-mixture-model}.
-This conflicts the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did included the character following shift (PLanTra, LIFT).
-In other words, patterns in our results cannot be explained on the basis of how before-sentence key-transitions were operationalised.
-
-Finally, the duration of fluent transitions was not distinguished by location in the constrained distribution model.
-Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transitions.
-This is presumably reflecting that these data were sampled from was the youngest and least experienced population of writers in our data pool.
-Hesitation duration and probability followed the same patterns as the majority of datasets.
-
-The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
-We, therefore, summarise briefly differences in the results for the transition-location effects; full results can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs > 100) but not before-sentence compared to before-word transitions (all BFs < 0.08, but SPL2 and GUNNEXP2: BFs > 100).
-Patterns for hesitation probability was generally similar in both the constrained and unconstrained model.
-However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs > 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.
-
+```{=html}
 <!-- TODO: 
@@
 -->
+```
 
@@
 
-The classical serial view [@flower1980ynamics] characterizes writing as a sequence of planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-The cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production [@olive2014toward].
-We used Bayesian statistical models, in particular hierarchical single-distribution and two-distribution mixture models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of six datasets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key transitions that directly precede sentences or words are not necessarily associated with a pause because writers can plan in parallel to writing.
-On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in the majority of datasets, pauses were as likely to occur before sentences and words, in both experienced and novice writers, while mid-word pauses are generally rare.
-Also, with few exceptions, pauses were typically longer before sentences than before words (which were longer than mid-word pauses).
-In other words, when writers do pause, the involvement of higher-level processing at sentence boundaries is typically leading to longer pauses.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because our results for the hesitation probability suggest that writers often do not pause before sentences; indeed, hesitation probabilities are largely identical before sentences and before words.
-This suggest that writers frequently plan the next word and even sentence in parallel to production which mitigates the need to pause to plan up-coming linguistic information.
-
-Interestingly even the youngest writers in our data pool (C2L1) showed evidence of parallel planning expressed in fluent transitions between words and even sentences.
-Key-transitions were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-This is interesting because, for example, @olive2014toward described a cascading model that is operating in a serial fashion for inexperienced or struggling writers explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers but pauses were more likely before words and sentences but similarly long indicating a tendency to prepare utterances in planning units possibly smaller than the clause.
-The main difference for these young writers was that their writing speed (fluent transitions) was substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra) we observe pausing patterns that do not resemble a serialised writing process as indicated by a larger number of fluent transitions across word and sentence boundaries.
-
-We observed two more differences between data sets -- other than the patterns observed for C2L1 summarised above -- that are worth highlighting.
-These can largely be explained on the basis of differences associated with the population (writing competence / experience) and text genre [see also @conijn2019understanding] that these samples were obtained from: The first exception concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations.
-**TODO: I don't have an explanation for this.** Second, pauses were not always longer before sentence compared to before words compared to mid-word paues; there were two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT dataset although the posterior showed the same general patterns as observed in most other datasets.
-The absence of a difference could be explained as specific to the synthesis writing task because there is no need for content creation [**REFERENCE?**] but the results for the PlanTra dataset rule out this possibility.
-
-A possible concern with our result -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-This concern is preempted by using cross-validation techniques for model comparison which is preventing model overfit.
-In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
-We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
-This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
-Parameters were not successfully uncovered when we switched model type and dataset.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a single distributions process but a substantially higher predictive performance for data simulated from a two-distributions mixture process.
-Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is limited research that used eye tracking to investigate regressive gaze behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
-There are at least four reasons why eye movements away from the inscription point might occure during writing: First, reading is necessary for revision (i.e.
-writers must check if writing goals have been met).
-Second, writers might use text to cue or reinstating ideas after interruption of writing process.
-Third, writers might look for or correct spelling or grammatical errors.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text.
-Fourth, writers might look at the source their are writing from.
-For two datasets (PlanTra, LIFT) which involved writing from sources, we removed transitions that were made when the sources (rather than the writing document) was in focus.
-
-Frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
-This is except, lookbacks are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-This can be seen in the result of two datasets in our pool (CATO, GUNNEXP2): these experiments included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-This suggest that pausing patterns do not change if reading previous text is not possible.
-
-A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent].
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause [@wen06;@van2016keystroke], i.e.
-a pause criterion threshold often set to 2 secs [@chanquoy1996writing; @kaufer1986composing; @sullivan2002self; @wen02] or some other lower bound [@chukharev2014pauses; @connelly2012predicting; @leijten2013keystroke].
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause [see e.g.
-@wen06] they would be arbitrary.
-For example, a sentence-initial pause of 2 secs has a very different interpretation from a 2 secs pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose [@chenu2014interword].
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying].
+```{r child = "sections/discussion.Rmd"}
+```
 
@@
 
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-We found that key transitions across word and even sentences boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
-This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.
-young / L2 writers, students) and writing tasks (e.g.
-essays, syntheses).
+```{r child = "sections/conclusion.Rmd"}
+```
 
+```{=html}
 <!--
@@
 -->
+```
+
+\newpage
 
 # References
-```{r create_r-references, echo=FALSE, include=FALSE}
-r_refs(file = "references.bib")
-```
 
 \begingroup
+
 \setlength{\parindent}{-0.5in}
+
 \setlength{\leftskip}{0.5in}
@@
 <div id = "references"></div>
+
 \endgroup
 
-```{r render_appendix, include=FALSE}
-render_appendix("appendix_models.Rmd")
-render_appendix("appendix_fit2data.Rmd")
+\newpage
+
+# (APPENDIX) Appendix {.unnumbered}
+
+```{r child = 'appendix/models.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/fit2data.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/transloc.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/shift.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/shift_constr.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/l2_constr.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/masking_constr.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/postparam_constr.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/prepost_constr.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/genre_constr.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/transloc_constr.Rmd'}
+```
+
+\newpage
+
+```{r child = 'appendix/sim.Rmd'}
+```
+
+\newpage
+
+```{r}
 #render_appendix("appendix_postparam.Rmd")
-render_appendix("appendix_postparam_constr.Rmd")
-render_appendix("appendix_shift_constr.Rmd")
+#render_appendix("appendix_genre.Rmd")
+#render_appendix("appendix_prepost.Rmd")
+#render_appendix("appendix_masking.Rmd")
 #render_appendix("appendix_l2.Rmd")
-render_appendix("appendix_l2_constr.Rmd")
-#render_appendix("appendix_masking.Rmd")
-render_appendix("appendix_masking_constr.Rmd")
-#render_appendix("appendix_prepost.Rmd")
-render_appendix("appendix_prepost_constr.Rmd")
-#render_appendix("appendix_genre.Rmd")
-render_appendix("appendix_genre_constr.Rmd")
-render_appendix("appendix_transloc.Rmd")
-render_appendix("appendix_transloc_constr.Rmd")
-render_appendix("appendix_shift.Rmd")
-render_appendix("appendix_sim.Rmd")
-
 ```

### 2024-04-27 report/manuscript.tex
rearrange markdown sections

@@
 \documentclass[
-  english,
   man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
   \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage{graphicx}
@@
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\ifLuaTeX
+\usepackage[bidi=basic]{babel}
+\else
+\usepackage[bidi=default]{babel}
+\fi
+\babelprovide[main,import]{english}
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
 % Manuscript styling
@@
 % \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
+%   {\begin{landscape}\centering\begin{threeparttable}}
+%   {\end{threeparttable}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
 
@@
 
-% \usepackage{etoolbox}
 \makeatletter
-\patchcmd{\HyOrg@maketitle}
+\usepackage{etoolbox}
+\patchcmd{\maketitle}
   {\section{\normalfont\normalsize\abstractname}}
@@
   {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
+\patchcmd{\maketitle}
   {\section{\protect\normalfont{\@title}}}
@@
 \makeatother
-\shorttitle{Modelling writing hesitations}
+
+\usepackage{xpatch}
+\makeatletter
+\xapptocmd\appendix
+  {\xapptocmd\section
+    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
+    {}{\InnerPatchFailed}%
+  }
+{}{\PatchFailed}
 \keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
@@
 \DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
+\ifLuaTeX
   \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage{csquotes}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
+  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev3},
+  pdflang={en-EN},
+  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 
+\shorttitle{Modelling writing hesitations}
+
 \authornote{
@@
 
-\abstract{
-Classical serial models view the process of producing a text as a descrete chain of pauses and writing bursts.
+\abstract{%
+Classical serial models view the process of producing a text as a discrete chain of pauses and writing bursts.
 In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
+We implemented these two views in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
 We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
 Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
+Further we found -- across populations -- that transitions across word and even sentence boundaries are often fluent and uninterrupted by planning of upcoming linguistic material; when pausing occurred pauses were longer before sentence than before word which is consistent with the idea larger linguistic edges are associated with higher level planning.
 Our results cannot be explained by the serial but are in line with the cascading view of writing.
@@
 \maketitle
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
+(2022) coined this possibility the \enquote{process-disruption hypothesis}.
 
@@
 
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
+In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts.
+Full modelling details can be found in Appendix \textbackslash ref\{statistical-models; we summarise the most important model properties in the following.
+The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data.
+In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.
 
@@
 (2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
+Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes.
+These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs.
+This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.
 
@@
 We discussed advantages with this approach in the introduction section.
-These datasets include samples from populations with different writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses).
-Datasets from a variety of different writers and writing tasks were deliberately chosen to challange our modelling approach and to test to what extent resulting pausing patterns generalise across writing contexts.
+These datasets include samples from populations with various writing experience and languages (e.g.~young / L2 writers, students) performing different writing tasks (e.g.~essays, syntheses).
+Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.
 
@@
 
-Six datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
+TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2.
+For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say \enquote{unpublished}.
+I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk).
+Was the GUNNEXP2 sample Norwegian?
+Should we include information like time limit or text length in the table?
+
+Six datasets with keystroke data from free text production were used for analysis.
+An overview can be found in Table \ref{tab:datasets}; descriptions below.
+Dataset used are C2l1 (Rønneberg et al., 2022), CATO (Torrance, Rønneberg, et al., 2016), GUNNEXP2 (Torrance \& Ofstad, n.d.), SPL2 (Torrance et al., n.d.), LIFT published in Vandermeulen, Steendam, et al.
+(2020) and described in Vandermeulen, De Maeyer, et al.
+(2020), PLanTra data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
+In C2L1 Norwegian 6th graders composed argumentative essays.
+In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text.
+In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced.
+The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each.
+The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability.
+The GUNNEXP2 data \ldots.
+{[}MARK, please add a sentence here to describe the production context{]}.
+
+For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
+LIFT and PLanTra data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021) and SPL2 data were collected using CyWrite (Chukharev-Hudilainen et al., 2019).
 
@@
 
-\begin{tabular}{llllrrp{2cm}p{3cm}}
+\begin{tabular}{lp{3cm}p{2cm}p{2.5cm}rrp{2.5cm}p{2.25cm}p{2.5cm}}
 \toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
+Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Sample} & \multicolumn{1}{c}{Country} & \multicolumn{1}{c}{Language}\\
 \midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
+C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.8 & 6th graders & Norway & Norwegian\\
+CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.9 & Secondary school students & Norway & Norwegian\\
+GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite & ????? & 45 & ???? & Undergraduate students & ???? Norway & ???? Norwegian\\
+LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 17 & Secondary school students & The Netherlands & Dutch\\
+PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23 & Master students & Belgium & English (L2)\\
+SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.6 & Undergraduate students & USA & English (L1) / Spanish (L2)\\
 \bottomrule
@@
 \end{table}
+
 \end{landscape}
 
-TODO: Mark, can you add information for C2L1 and CATO as needed
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-The C2L1 dataset comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-
-The CATO dataset is published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-
-The SPL2 data set involves undergraduate university students in the US -- \emph{N} = 39, 2mean age = 20.6 years, SD = 1.51 -- composing argumentative essays in English (L1) and Spanish (L2).
-Students wrote within the CyWrite word-processing environment (Chukharev-Hudilainen et al., 2019).
-Participants wrote, with a 40 minute time limit, to two writing prompts.
-Order and writing prompt were counterbalanced across language.
-
-LIFT and PLanTra data data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-LIFT is a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education collected from 43 schools.
-The students were asked to conduct four synthesis writing tasks, two argumentative and two informative texts, about each of four topics, with order randomized per school.
-The students received 50 minutes for each tasks.
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{transition-types}{%
-\subsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
+\hypertarget{data-extraction}{%
+\subsection{Data extraction}\label{data-extraction}}
+
+From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
 In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
+At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section.
 Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
+Also we removed transitions that were followed by an editing operation.
 
@@
 
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
+\begin{tabular}{p{3cm}p{5cm}p{6cm}}
 \toprule
@@
 
-\hypertarget{data-reduction}{%
-\subsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
+We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data).
 We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models -- decribed in the next section -- require to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
+We further removed keystroke intervals that were too short to represent a behavioural response (\(\le\) 50 msecs) or too long (\(\ge\) 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}.
+From the remaining data we randomly sampled a maximum of 100 observations per participant, condition, and transition location (when more than 100 were available).
+This was done to reduce the computation time of the Bayesian models require to complete sampling.
+For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets.
+Because the LIFT data set included the largest number of writing tasks, we sampled 50 observations per condition, location and participant which would otherwise exceed the computational resources available to us.
+The number of keystroke data used in the analysis can be found in Table \ref{tab:datareduction}.
 
@@
 \caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
+Mean percentage of extreme data removed and the mean percentage of randomly sampled data included by transition location (`--` if all data were used).
+Standard error in parentheses.}
 
@@
 \midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
+C2L1 & 0.19 (0.1) & 0.074 (0.063) & 35 (2.6) & 85 (1.8) & --\\
+CATO & 0.65 (0.15) & 0.023 (0.016) & 15 (0.95) & 49 (2.2) & --\\
+GUNNEXP2 & 2.2 (0.17) & 0.0072 (0.0055) & 6.2 (0.43) & 23 (1.4) & --\\
+LIFT & 2.6 (0.16) & 0.0048 (0.0038) & 3.2 (0.22) & 13 (0.86) & 99 (0.15)\\
+PLanTra & 2.5 (0.41) & 0.04 (0.031) & 9.7 (0.64) & 37 (1.9) & --\\
+SPL2 & 2.3 (0.2) & 0.028 (0.016) & 5.7 (0.39) & 23 (1.4) & --\\
 \bottomrule
@@
 An overview of all models can be found in Table \ref{tab:models}.
-More detailed model descriptions and motivations can be found in Appendix \ref{statistical-models}.
+More detailed model descriptions and motivation can be found in Appendix \ref{statistical-models}.
 
@@
 
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
+\caption{\label{tab:models}Overview of cognitive models of writing. All models included by-participant random effect.}
 
@@
 \midrule
-M1 & Standard single Gaussian model with effects of transition location by study-specific manipulations and by-participant random effect\\
-M2 & Single distribution log-Gaussian model equivialent to M1\\
-M3 & Equivialent to M2 but with different variance components for each transition location (unequal variance)\\
-M4 & Two-distributions mixture of a log-Gaussian of fluent intervals and another of hesitant; distribution of hesitant keystroke intervals is equivialent to M3 but distribution of fluent intervals has a mean across all design cell.
-Mixing proportion captures the relative number of disfluent transitions\\
-M5 & Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location\\
+\textbf{Serial} & \\
+\ \ \ M1 & Single Gaussian model with effects of transition location (and experiment-specific manipulation).\\
+\ \ \ M2 & Single distribution log-Gaussian model equivialent to M1.\\
+\ \ \ M3 & Equivalent to M2 but with different variance components for each transition location (unequal variance).\\
+\textbf{Cascading} & \\
+\ \ \ M4 & Two-distributions mixture of a log-Gaussian for fluent transition intervals and another log-Gaussian that is wider than the previous distribution hesitant / disfluent key transitions; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition location but the distribution of fluent intervals has the same mean across all design cell.
+Mixing proportions capture the relative number of disfluent transitions for the distribution of hesitant transitions for each combination of transition location.\\
+\ \ \ M5 & Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location.\\
 \bottomrule
@@
 
-Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset specific manipulations and random effects for participants.
+Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset-specific manipulations.
 In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
-For model M3 we relaxed the equal variance assumption for transition locations, thus allowing each transition location to assume a different standard deviation (see Appendix \ref{statistical-models} for rationale).
-
-Models M4 and M5 and extensions of the single distribution models.
-Importantly models M4 and M5 assume that keystroke data are the result of a combination of two data generating processes -- consistent with the cascading account -- rather than one.
-These are called finite mixture models in the literature (Gelman et al., 2014; MacLahlan \& Peel, 2000; Roeser et al., 2021).
-
-These models extend the assumption that processing at higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that shifts the distribution of IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; this is because slower human behaviour is generally known to be associated with a larger variability (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973); see Appendix \ref{statistical-models} for rationale.
+
+Models M4 and M5 are extensions of the unequal-variance single distribution model M3.
+Importantly models M4 and M5 assume that keystroke data result from a combination of two data generating processes -- consistent with the cascading account -- rather than one.
+These model are called finite mixture models in the literature (Gelman et al., 2014; MacLahlan \& Peel, 2000; Roeser et al., 2021) and capture the assumption that processing at higher levels of activation leads to longer pauses which can in principle happen at transition before sentence, before words or even mid-word.
+In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that pauses at larger linguistic edges are more likely but not obligated by the cognitive system that generates keystrokes.
+This is achieved by modelling key transitions as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
 
@@
 \item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter illustrated in equation \ref{eq:bimodcon2}.
+  Information from upstream mental processes can flow into keystrokes without interruption at intermediate levels.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger.
+The population estimate for fluent key transitions (i.e.~typing speed) is captured by the \(\beta\) parameter.
 \item
-  Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon2}.
+  Any interruptions at upstream levels of mental representation result in a delay in the information flow and therefore create a lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
+The population estimate of the hesitation slowdown is captured by \(\delta\) and its frequency is captured by the mixing weight (proportion) \(\theta\).
 \end{enumerate}
@@
 
-The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
+The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents the distribution of fluent -- uninterrupted -- key transitions.
+Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
+\(\theta\) is here parameterised to represent the probability that a key transition is associated with the distribution of hesitant transitions.
+This probability is inversely related to the mixing weight of the distribution of fluent transitions by \(1-\theta\) .
+In other words, a larger weight for either distribution inevitably means a lower weight for the other.
+We call this parameter the probability of hesitant transitions.
+When a parameter was assumed to vary by transition location (levels: before sentence, before word, within word) this was indicated as subscript; similar for participants.
+The hesitation probability was allowed to vary by participants constrained by a hyper-parameter for each transition location.
+
+The constrained implementation M4 assume that the size of a fluent key-transition does not vary by transition location.
+In other words, the value that the \(\beta\) parameter is taking on is the same for transitions at before-sentence, before-word, and within-word locations.
+However, letter bigrams may be executed systematically faster than transitions between space and either a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
+This is for two reasons: first, and importantly, highly frequent bi- and even trigrams are stored, retrieved and executed as one motor code (Behmer Jr.
+\& Crump, 2016; Grudin \& Larochelle, 1982; Ostry, 1983; Terzuolo \& Viviani, 1980) because of the strong association between letters.
+This association is reducing the length of transition durations.
+It is less likely that the same association exists for space-letter and space-shift pairs.
+Second, before-sentence transition in two datasets (PLanTra, LIFT) are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transition; we return to this in the results section.
+We therefore implemented an unconstrained mixture model M5 that assumes that the distribution over durations of fluent transitions -- represented as \(\beta\) -- may vary across transition locations.
+
+All models were implemented in the Bayesian framework (Gelman et al., 2014; McElreath, 2016).
+The Bayesian framework is ideal for the estimation of parameter values.
+This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
+D.
+Lee \& Wagenmakers, 2014).
+To achieve this, Bayesian models require the inclusion of prior information, i.e.~existing knowledge about parameter values.
+For small datasets priors influence the inferred parameter values; for larger datasets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
+In other words the choice of priors has less impact on the posterior, certainly for weakly informative priors.
+In the present analysis, we used weakly informative priors to aid model convergence by constraining the parameter space to plausible values (see e.g.
+Lambert, 2018; McElreath, 2016).
+Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
+
+Stan code for mixture models was based on (Roeser et al., 2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); for a tutorial see \href{rpubs.com/jensroes/1000459}{https://rpubs.com/jensroes/1000459}.\footnote{The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
+Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
+Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
+The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{model-comparisons}{%
+\subsection{Model comparisons}\label{model-comparisons}}
+
+To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling (Vehtari et al., 2015, 2017).
+Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared by the difference between models \(\Delta\widehat{elpd}\).
+We also summarised this difference as normalised over its standard error \(\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid\), hence the \emph{z}-score of the difference between models (Sivula et al., 2020).
+Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.
+
+Model differences are shown in Table \ref{tab:loos}.
+All datasets revealed the same pattern.
+Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single-distribution model; the difference \(\Delta\widehat{elpd}\) between models ranged within {[}18, 40{]} standard errors, indicating a substantially higher predictive performance for the two-distributions model.
+The unconstrained mixture model M5 rendered a higher predictive performance than the constrained model M4 which did not allow the distribution of fluent keystroke-intervals to vary by transition location; the difference in predictive performance was between {[}7, 15{]} standard errors higher for the unconstrained model.
+Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; the difference \(\Delta\widehat{elpd}\) ranged within {[}43, 68{]} standard errors.
+The weakest model was the single distribution Gaussian which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of {[}43, 69{]} standard errors.
+
+\begin{landscape}
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} $\Delta\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{p{1.5cm}p{5cm}rrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:loos}Model comparisons.
+Models were compared incrementally from the simplest to the most complex model.
+The comparison is shown along with the distribution type of the more complex model.
+A negative difference in $\Delta\widehat{elpd}$ indicates higher predictive performance for the more complex model.
+Standard error is shown in parentheses.}\\
+\toprule
+Comparison & \multicolumn{1}{c}{Distribution} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
+\toprule
+Comparison & \multicolumn{1}{c}{Distribution} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
+\midrule
+\endhead
+M4 -- M5 & Mixture of two log-Gaussians (unconstrained; M5) & -546 (51) & -607 (43) & -554 (38) & -500 (37) & -130 (18) & -564 (41)\\
+M3 -- M4 & Mixture of two log-Gaussians (constrained; M4) & -1,637 (72) & -1,782 (75) & -2,427 (92) & -5,105 (127) & -2,384 (90) & -1,181 (64)\\
+M2 -- M3 & Single log-Gaussian (unequal variance; M3) & -38,301 (823) & -38,972 (913) & -41,155 (653) & -76,961 (1,656) & -35,204 (556) & -30,430 (450)\\
+M1 -- M2 & Single log-Gaussian (M2) & -37,517 (798) & -37,673 (881) & -38,843 (616) & -74,851 (1,624) & -33,823 (526) & -28,717 (418)\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+
+\end{landscape}
+
+We also tested to what extent model predictions fit observed data.
+Comparisons can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
+Data predicted by the two-distribution mixture models are similar to observed data; data predicted by single distribution Gaussian models showed a poor fit to the data.
+
+\hypertarget{transition-location-effect}{%
+\subsection{Transition location effect}\label{transition-location-effect}}
+
+As demonstrated in the model comparisons, the two-distributions mixture-models is a better statistical representation on the underlying process generating keystroke data than compared to any of the single-distribution models.
+Information about this process is captured by the posterior estimates of the models' parameter values.
+To reiterate, there are three conceptually important parameters: (1) the magnitude of the slowdown for transition hesitations (represented as \(\delta\) in equation \ref{eq:bimodcon2}), (2) the probability of hesitant transitions (represented as \(\theta\) in equation \ref{eq:bimodcon2}), and (3) the duration of fluent transitions (the typing speed, which was indicated as \(\beta\) in equation \ref{eq:bimodcon2}).
+
+Figure \ref{fig:mixmodel} illustrates the mixture model approach on the basis of the posterior of the constrained implementation applied to the CATO data: for each transition durations we obtain two posterior probability distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in hesitant transitions.
+The hesitation probability is indicated as the relative height of either distribution.
+The transition locations are characterized by the hesitation parameters: as we illustrate the constrained mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location but their height differs as the probability of fluent transition is the inverse of the height of hesitant transitions; i.e.~\(1 - \theta\).
+More interesting are the distributions associated with hesitations; indicated in yellow: Within word transitions have a negligibly small hesitation probability barely visible in the visualisation.
+Hesitations before words and sentences are both roughly equally likely than fluent transitions.
+In other words, half of the time participants did not pause before a word or sentence; we explore this below.
+However when hesitations occurred, these are longer at before-sentence transitions compared to before word transitions.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/mixmodel-1}
+
+}
+
+\caption{Mixture model plot for each transition location.
+Shown are the distributions for both mixture components (fluent and hesitant transitions) and their weighting.
+Posterior of CATO data.}\label{fig:mixmodel}
+\end{figure}
+
+We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
+We demonstrated above that the predictive performance of the unconstrained mixture model was substantially higher than the predictive performance of the constrained mixture model.
+However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
+We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following.
+
+The mixture-model posterior parameter estimates for all datasets is visualised in Figure \ref{fig:crossstudypost2}.
+Although models were fitted with all data set-specific conditions, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
+We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.}, and excluded conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
+There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-gunnexp2}).}.
+For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
+The resulting posterior allows us to examine differences and similarities associated with transition locations for all dataset.
+In particular we can see how the linguistic location in the text is associated with hesitation duration and probability.
+The estimate for fluent transitions is shown by dataset, across transition location.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost2-1}
+
+}
+
+\caption{Mixture model parameter estimates across studies.
+Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.}\label{fig:crossstudypost2}
+\end{figure}
+
+Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets.
+We summarise the results of the pairwise comparisons for transitions locations; full results and parameter estimates can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
+We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs \textgreater{} 100 for all datasets) but only for half of the data sets also longer (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
+Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GUNNEXP2 and SPL2, both BFs \textgreater{} 100) but longer (all BFs \textgreater{} 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
+
+To reiterate, for most datasets we found that pauses are before sentences are longer but not necessarily more likely than before words (except for datasets SPL2 and GUNNEXP2).
+In fact, LIFT showed some indication that pauses were less likely before sentences than words.
+This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for key transitions at larger linguistic edges.
+It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.
+
+Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character while the remaining data sets measured before-sentence transitions as the difference between space key and shift.
+To control for the use of complex key combinations, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability (BFs \textgreater{} 100) but it did not affect the duration of fluent or hesitant transitions (BFs \textless{} 0.3); see Appendix \ref{key-combination-effect-constrained-mixture-model}.
+This conflicts the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did included the character following shift (PLanTra, LIFT).
+In other words, patterns in our results cannot be explained on the basis of how before-sentence key-transitions were operationalised.
+
+Finally, the duration of fluent transitions was not distinguished by location in the constrained distribution model.
+Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transitions.
+This is presumably reflecting that these data were sampled from was the youngest and least experienced population of writers in our data pool.
+Hesitation duration and probability followed the same patterns as the majority of datasets.
+
+The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
+We, therefore, summarise briefly differences in the results for the transition-location effects; full results can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
+Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs \textgreater{} 100) but not before-sentence compared to before-word transitions (all BFs \textless{} 0.08, but SPL2 and GUNNEXP2: BFs \textgreater{} 100).
+Patterns for hesitation probability was generally similar in both the constrained and unconstrained model.
+However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs \textgreater{} 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence of planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
+The cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production (Olive, 2014).
+We used Bayesian statistical models, in particular hierarchical single-distribution and two-distribution mixture models, to directly compare the serial and the cascading model of writing, respectively.
+On the basis of six datasets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
+
+Patters observed in the parameter estimates are generally similar across data sets (with caveats).
+Under the cascading view of writing, key transitions that directly precede sentences or words are not necessarily associated with a pause because writers can plan in parallel to writing.
+On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
+We found that in the majority of datasets, pauses were as likely to occur before sentences and words, in both experienced and novice writers, while mid-word pauses are generally rare.
+Also, with few exceptions, pauses were typically longer before sentences than before words (which were longer than mid-word pauses).
+In other words, when writers do pause, the involvement of higher-level processing at sentence boundaries is typically leading to longer pauses.
+However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
+This is because our results for the hesitation probability suggest that writers often do not pause before sentences; indeed, hesitation probabilities are largely identical before sentences and before words.
+This suggest that writers frequently plan the next word and even sentence in parallel to production which mitigates the need to pause to plan up-coming linguistic information.
+
+Interestingly even the youngest writers in our data pool (C2L1) showed evidence of parallel planning expressed in fluent transitions between words and even sentences.
+Key-transitions were substantially longer compared to all other datasets with a larger slowdown before words.
+Yet the overall pausing probability showed a patterns comparable to most other data sets.
+In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
+This is interesting because, for example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
+However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers but pauses were more likely before words and sentences but similarly long indicating a tendency to prepare utterances in planning units possibly smaller than the clause.
+The main difference for these young writers was that their writing speed (fluent transitions) was substantially slower compared to more experienced writers in the sample.
+Also for L2 writers (PlanTra) we observe pausing patterns that do not resemble a serialised writing process as indicated by a larger number of fluent transitions across word and sentence boundaries.
+
+We observed two more differences between data sets -- other than the patterns observed for C2L1 summarised above -- that are worth highlighting.
+These can largely be explained on the basis of differences associated with the population (writing competence / experience) and text genre (see also Conijn et al., 2019) that these samples were obtained from: The first exception concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations.
+\textbf{TODO: I don't have an explanation for this.} Second, pauses were not always longer before sentence compared to before words compared to mid-word paues; there were two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
+(2) pause durations did not vary by transition location in the LIFT dataset although the posterior showed the same general patterns as observed in most other datasets.
+The absence of a difference could be explained as specific to the synthesis writing task because there is no need for content creation {[}\textbf{REFERENCE?}{]} but the results for the PlanTra dataset rule out this possibility.
+
+A possible concern with our result -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+This concern is preempted by using cross-validation techniques for model comparison which is preventing model overfit.
+In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
+We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
+This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
+Modelling details can be found in Appendix \ref{simulation}.
+As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
+Parameters were not successfully uncovered when we switched model type and dataset.
+Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a single distributions process but a substantially higher predictive performance for data simulated from a two-distributions mixture process.
+Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of overfitting.
+
+Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
+Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
+There is limited research that used eye tracking to investigate regressive gaze behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
+There are at least four reasons why eye movements away from the inscription point might occure during writing: First, reading is necessary for revision (i.e.~writers must check if writing goals have been met).
+Second, writers might use text to cue or reinstating ideas after interruption of writing process.
+Third, writers might look for or correct spelling or grammatical errors.
+We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text.
+Fourth, writers might look at the source their are writing from.
+For two datasets (PlanTra, LIFT) which involved writing from sources, we removed transitions that were made when the sources (rather than the writing document) was in focus.
+
+Frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example Torrance, Johansson, et al.
+(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words (\enquote{hopping,} see also Chukharev-Hudilainen et al., 2019).
+This is except, lookbacks are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
+Finally even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
+This can be seen in the result of two datasets in our pool (CATO, GUNNEXP2): these experiments included a manipulation where the produced text was masked (or not).
+In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
+This suggest that pausing patterns do not change if reading previous text is not possible.
+
+A practical advantage of the mixture-model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
+Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
+There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
+Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
+Wengelin, 2006) they would be arbitrary.
+For example, a sentence-initial pause of 2 secs has a very different interpretation from a 2 secs pause that occurs before or within a word.
+Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
+So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
+However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
+Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
+Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
+Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
+We found that key transitions across word and even sentences boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
+This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses).
+
+\newpage
+
+\hypertarget{appendix-appendix}{%
+\appendix}
+
+\hypertarget{statistical-models}{%
+\section{Statistical models}\label{statistical-models}}
+
+We are using the Bayesian framework (e.g.
+Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
+D.
+Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
+In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
+The models presented in this section build on one another so that later model include assumptions of earlier models.
+
+\hypertarget{serial-model-of-writing}{%
+\subsection{Serial model of writing}\label{serial-model-of-writing}}
+
+\hypertarget{single-distribution-gaussian}{%
+\subsubsection{Single distribution Gaussian}\label{single-distribution-gaussian}}
+
+Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
+The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
+For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes.
+There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
+We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
+This can be expressed as \(\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)\).
+Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
+
+We can extend this simple model of word-planning to other linguistic location.
+We introduce earlier that larger linguistic edges are associated with planning on higher levels.
+For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
+Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
+We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
+However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
+Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
+The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
+
+We implemented the average keytransition associated with a given text locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
+Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
+Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
+The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
+This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
+
+The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimodgaus}
+\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant[i]} \sim \text{N}(0, \sigma_\text{p}^2)\\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+where \(i\) iterates over every key-transition \(i \in 1 \dots N\) with \(N\) being the total number of keystroke intervals, and thus \(\text{location[i]}\) is the transition location associated with the \(i\)\^{}\text{th} keystroke and \(\text{participant[i]}\) is the participant associated with the \(i\)\^{}\text{th} keystroke.
+
+Standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
+
+\hypertarget{single-distribution-log-gaussian}{%
+\subsubsection{Single distribution log-Gaussian}\label{single-distribution-log-gaussian}}
+
+The previous model assumes that the data-generating process is a Gaussian distribution.
+The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
+There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
+IKIs, as the distance between two subsequent key-down events, must be positive.
+The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
+(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
+In particular, a Gaussian distribution assumes that units are scaled linearly.
+For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
+This does not necessarily map onto the psychological interpretation for short and long keyintervals.
+For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
+In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
+Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
+
+The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{log}\mathcal{N}()\) was replaced by \(\text{logN}()\).
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimodloggaus}
+\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
+\text{where: } &
+\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+\hypertarget{single-distribution-unequal-variance-log-gaussian}{%
+\subsubsection{Single distribution unequal-variance log-Gaussian}\label{single-distribution-unequal-variance-log-gaussian}}
+
+The third model representing the serial view of writing is a single distribution unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
+The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
+The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
+This assumption is however not in line with what we know about data from human behaviour.
+Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
+For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
+Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies by transition location.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimoduv}
+\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+ & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
+ \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+\hypertarget{parallel-cascading-model-of-writing}{%
+\subsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
+
+The following two models are are extensions of the models introduced for the serial view.
+Crucially, the cascading view allows planning to happen in parallel to production.
+Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
+This is done by assuming that IKIs come from a weighted combination of two distributions.
+
+\hypertarget{two-distributions-log-gaussian-mixture-model-constrained}{%
+\subsubsection{Two-distributions log-Gaussian mixture model (constrained)}\label{two-distributions-log-gaussian-mixture-model-constrained}}
+
+This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
+Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
+This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Activation can flow into keystrokes without interruption.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
+Note that \(\beta\) is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.
+\item
+  Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
+The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
+The magnitude of the slowdown is associated with transition location.
+This is because delays at larger linguistic units are likely to be associated with higher level planning.
+By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
+\end{enumerate}
+
+\begin{equation}
+\begin{aligned}
+\label{eq:bimodcon}
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
+        & \sigma_{e'}^2 > \sigma_{e}^2\\
+        & 0 < \theta < 1
+\end{aligned}
+\end{equation}
+
+The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
 Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
@@
 The weights of both distributions must sum to 1.
-We call this parameter the probability of hesitant transitions.
-
-The size of a fluent key-transition does not necessarily vary by transition location which is what our constrained model M4 assumes.
+We will call this parameter the probability of hesitant transitions.
+
+The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
+In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
+As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
+This is akin to what is known as a random by-participant slopes model.
+
+Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
+In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
+This was achieved by introducing a parameter \(\sigma_\text{diff}\).
+The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
+
+\hypertarget{two-distributions-log-gaussian-mixture-model-unconstrained}{%
+\subsubsection{Two-distributions log-Gaussian mixture model (unconstrained)}\label{two-distributions-log-gaussian-mixture-model-unconstrained}}
+
+The size of a fluent key-transition does not necessarily vary by transition location.
 In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
+This is what we called a constrained model.
 However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
 This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter.
-We therefore also implemented an unconstrained model M5 that assumes that the distribution over durations of fluent transitions -- represented as \(\beta\) -- varies across transition locations.
-
-All models were implemented in the Bayesian framework.
-The Bayesian framework, as used in this paper, is ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for detailed descriptions of parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon} in Appendix \ref{statistical-models}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.\footnote{Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{model-comparisons}{%
-\subsection{Model comparisons}\label{model-comparisons}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both two-distributions mixture models provided a substantially better fit than any of the single distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the single distribution Gaussian model.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\Delta\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density -- relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{1.5cm}p{5cm}rrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row shows the models with the highest predictive performance compared to all other models. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
-\midrule
-\endhead
-M5 & Two distributions mixture of log-Gaussians (unconstrained) & -- & -- & -- & -- & -- & --\\
-M4 & Two distributions mixture of log-Gaussians (constrained) & -546 (51) & -607 (43) & -554 (38) & -500 (37) & -130 (18) & -564 (41)\\
-M3 & Single distribution log-Gaussian (unequal variance) & -2,184 (83) & -2,389 (79) & -2,981 (94) & -5,605 (131) & -2,514 (92) & -1,744 (68)\\
-M2 & Single distribution log-Gaussian & -2,968 (98) & -3,689 (100) & -5,293 (111) & -7,716 (147) & -3,895 (94) & -3,457 (81)\\
-M1 & Single distribution Gaussian & -40,485 (856) & -41,362 (927) & -44,136 (660) & -82,567 (1,691) & -37,718 (572) & -32,174 (450)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of two-distributions mixture models and poorest for the single distribution Gaussian model.
-
-\hypertarget{transition-location-effect}{%
-\subsection{Transition location effect}\label{transition-location-effect}}
-
-As demonstrated in the model comparisons, the two-distributions mixture-model capture the writing-process data better than single-distribution models.
-Therefore, writing process is captured by the posterior estimates of the model parameter values.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimodcon2}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-Figure \ref{fig:mixmodel} illustrates the mixture model approach: for each transition durations we obtain two posterior distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in longer durations.
-The relative weighting of these two distributions was captured as the probability of hesitant transitions, the parameter that captures the height of the distribution of slow keystrokes.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/mixmodel-1}
-
-}
-
-\caption{Mixture model plot for each transition location.
-Shown are the distributions for both mixture components (fluent and hesitant transition durations) and their respective weighting.
-Posterior of SPL2 data (English sessions only).}\label{fig:mixmodel}
-\end{figure}
-
-The predictive performance of the unconstrained mixture model is substantially higher than the predictive performance of the constrained mixture model.
-However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
-We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following with reference to the unconstrained implementation.
-
-We present the mixture-model posterior estimates for each of the three parameters in the three facets of Figure \ref{fig:crossstudypost2}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-While there was evidence for language-specific effects in Appendix \ref{l2-effect-spl2}, evidence for marking effects was weak as shown in Appendix \ref{masking-effect-cato-gunnexp2}.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the slowdown duration for hesitant transitions and the hesitation probability.
-Note that the estimate for fluent transitions was estimated by dataset.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost2-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.}\label{fig:crossstudypost2}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for keystroke-interval estimates by transition location across data sets for all three mixture model parameters.
-Full results of these pairwise comparisons can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
-We found that hesitations appear more frequently -- higher hesitation probability -- at before-word transitions than within words across data sets (all BFs \textgreater{} 100).
-Hesitations were more likely at before-sentence locations compared to before-word locations only for 2 out of 6 datasets (GUNNEXP2 and SPL2, both BFs \textgreater{} 100; but C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6).
-Hesitations were longer at before-sentence transitions compared to before-word transitions (all BFs \textgreater{} 100, except for C2L1: BF = 1.11 and LIFT = 0.5); before-word transitions were longer compared to within-word transitions for half of the data sets (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1) but not for the other (C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
-
-To reiterate, for most datasets we found negligible evidence supporting the idea that writers pause more frequently at before-sentence boundaries compared to before-word locations (except for datasets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, LIFT showed some indication that pauses were less likely before sentences than words.
-In brief, while pauses tend to be longer before sentences and less likely at mid-word locations, it is less clear in which contexts pauses are (1) longer before than within words and (2) more likely to appear before sentences than before words.
-
-The duration of fluent durations was not distinguished by transition location in the constrained version of the two distribution model.
-Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transition durations.
-This is presumably reflecting that the population that this sample is from was the youngest and least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to other data sets.
-
-Inconsistencies between datasets could, to some extent, be explained on the basis that before-sentence transitions were sometimes defined as involving complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, before-sentence transitions in some datasets (PLanTra, LIFT) included the character key following the shift key but others did not (CATO, C2L1, SPL2, GUNNEXP2).
-Appendix \ref{key-combination-effect-constrained-mixture-model} addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increases the hesitation probability (BFs \textgreater{} 100) but it did not affect the duration of fluent transitions or hesitations (BFs \textless{} 0.3).
-However this is in conflict with the earlier finding that before-sentence hesitations are not more likely than before-word hesitations for the datasets that included the character following shift (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
-We, therefore, summarize briefly differences in the results for the transition-location effects.
-Results for the unconstrained version of the mixture model can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-The main difference between the unconstrained and the constrained model is that the latter does not allow the parameter that captures fluent transition durations to vary by transition location or otherwise.
-Instead it represents the lower bound of keystroke activity for uninhibited activation flow.
-Higher predictive performance for the unconstrained model suggests variability across transition locations for fluent transition durations, in particular for before-word transitions compared to mid-word transitions (all BFs \textgreater{} 100) but not for before-sentence transitions compared to before-word transitions (all BFs \textless{} 0.08, but SPL2 and GUNNEXP2: BFs \textgreater{} 100).
-The pattern for the hesitation probability was generally similar in both the constrained and unconstrained model.
-However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs \textgreater{} 10, except for LIFT: BF = 0.7); the hesitation duration pattern for before-sentence transitions compared to before-word transitions remained the same.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular single-distribution and two-distributions hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
-We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two-distributions -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a two-distributions process (but for the two-distributions data).
-Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to investigate regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-peel2000finite}{}%
-MacLahlan, G., \& Peel, D. (2000). \emph{Finite mixture models}. John \& Sons.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g. Farrell \& Lewandowsky, 2018;
-Gelman et al., 2014; Lee \& Wagenmakers, 2014) to implement 5
-statistical models of writing and to evaluate which -- the serial or the
-cascading view -- captures more successfully the data that arise during
-keyboard typing. In other words, we use statistical models to map
-between the data and the theoretically assumed process that generates
-the data to then compare the predictive power of those models. The
-models presented in this section build on one another so that later
-model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{single-distribution-gaussian}{%
-\paragraph{Single distribution
-Gaussian}\label{single-distribution-gaussian}}
-
-Under the serial view, all planning must be completed priors to the
-production onset of the corresponding planning unit. The resulting IKI
-is sometimes faster or slower depending on, among others,
-psycholinguistic factors. For example, the interval before a word is
-shorter for easily retrievable high-frequency words, or longer for low
-frequency words, shorter for words with fewer graphemes, syllables, and
-morphemes. There are word-specific factors that influence the IKI that
-precedes a word but these are beyond the scope of our analysis. We will
-capture the variability associated with word-features by assuming that
-before-word IKIs can be described as coming from a distribution that is
-normal (Gaussian) with two parameters, an unknown mean \(\mu\), that
-describes the average IKI associated with word-level planning, and a
-standard deviation \(\sigma_\text{e}^2\), that captures the variability
-associated with factors that we did not further specify in the model.
-This can be expressed as
-\(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value
-captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic
-location. We introduce earlier that larger linguistic edges are
-associated with planning on higher levels. For example, at sentence
-boundaries, planning needs to happen for word-level properties -- which
-was captured above as average IKI \(\mu\) -- but also higher linguistic
-planning such as clause-level meaning, and dependencies of the
-sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and
-capture that there is a change in the outcome variable associated with
-another factors. We can decompose \(\mu\) as
-\(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that
-when \(\text{x}_\text{sentence}\) takes on the value 0, the equation
-reduces to \(\mu = \alpha\) which is the average IKI for word
-boundaries. However when \(\text{x}_\text{sentence}\) takes on the value
-1, the average IKI for word boundaries \(\alpha\) is incremented by a
-sentence-related slowdown of \(\beta\) msecs. Therefore the value of the
-\(\beta\) parameter represents the additional cognitive demand
-associated with sentence-initial planning. The application of such a
-statistical model to the data will then provide us with an estimate of
-the parameter value that can be used for statistical inference
-(e.g.~whether there is a statistically meaningful difference between IKI
-associated with words and sentences).
-
-For computational ease, we implemented the differences associated with
-transition locations as \(\beta_\text{location}\) in equation
-\ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for
-each transition location. Therefore, the model will return one \(\beta\)
-per transition location that capture the posterior distribution of
-average IKIs. Also the decomposition of \(\mu\) allows us to address the
-fact that some writers are faster than other writers by introducing a
-parameter for what is typically called random intercepts
-\(u_\text{participant}\). The random intercepts term
-\(u_\text{participant}\) is constrained so that the value it takes on
-comes from a normal distribution with a mean of 0 and a standard
-deviation of \(\sigma_\text{p}^2\). This value is therefore the
-difference between the overall posterior estimate and estimate average
-IKI of a particular participant (i.e.~a positive value indicates that a
-participant is slower than average, a negative value indicates that a
-participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents
-a Gaussian mixed effects model.
+Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
+We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
+
+In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
 
@@
 \begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be
-positive because standard deviations can never be negative.
-
-\hypertarget{single-distribution-log-gaussian}{%
-\paragraph{Single distribution
-log-Gaussian}\label{single-distribution-log-gaussian}}
-
-The previous model assumes that the data-generating process is a
-Gaussian distribution. The next model is largely identical to the
-previous model but instead of assuming a Gaussian, we assume that the
-data come from a log-normal (log-Gaussian) distribution. There are, at
-least, two arguments for using a log-Gaussian distribution: (1)
-log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian
-does not allow negative values. IKIs, as the distance between two
-subsequent key-down events, must be positive. The lower bound is
-delimited by a person's ability to move their fingers and keyboard
-polling. (2) the log-scale is known to be a better match for data from
-human behaviour and motor responses. In particular, a Gaussian
-distribution assumes that units are scaled linearly. For example, a
-difference of 25 msecs is the same between 100 and 125 msecs as between
-5 secs and 5,025 msecs. This does not necessarily map onto the
-psychological interpretation for short and long keyintervals. For
-example effects that result on the motor level within words (e.g.~typing
-a high vs a low frequency bigram) are smaller than differences that are
-due to high levels of processing (retrieving a word in an L1 or L2). In
-other words, although an effect of 25 msecs is large in the context
-overall fast keyintervals, it is small in the context of overall slow
-intervals. Log-Gaussian distributions are a natural way of translating a
-linear scale to an exponential scale so that a 25 msecs difference on
-the lower end of the IKI scale (motor activity) is more meaningful than
-a 25 msecs difference on the upper end of the IKI scale (retrieving
-words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in
-which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{single-distribution-unequal-variance-log-gaussian}{%
-\paragraph{Single distribution unequal-variance
-log-Gaussian}\label{single-distribution-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is a single
-distribution unequal-variance model that, except for the
-unequal-variance assumption is identical to the model presented in the
-previous section. The previous models modelled IKIs as a function of
-transition location so that the estimated average IKI depends on the
-position of an IKI in the text. The variance associated with the
-estimated IKIs for each transition location was assumed to be identical
-(equal variance). This assumption is however not in line with what we
-know about data from human behaviour. Longer latencies are known to be
-associated with a larger variances for both response-time data in
-particular (Wagenmakers \& Brown, 2007) and human motor behaviour in
-general (Schöner, 2002; Wing \& Kristofferson, 1973). For IKIs pauses at
-larger linguistic edges are plausibly associated with a larger variance
-because of the larger number of associated processes. Therefore, in
-equation \ref{eq:unimoduv}, we introduce the assumption that standard
-deviation \(\sigma_{e_\text{location}}^2)\) varies by transition
-location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of
-writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for
-the serial view. Crucially, the cascading view allows planning to happen
-in parallel to production. Therefore, we will reduce the constrain of
-the serial models that requires all planning to be completed before
-writing onset. This is done by assuming that IKIs come from a weighted
-combination of two distributions.
-
-\hypertarget{two-distributions-log-gaussian-mixture-model-constrained}{%
-\paragraph{Two-distributions log-Gaussian mixture model
-(constrained)}\label{two-distributions-log-gaussian-mixture-model-constrained}}
-
-This model extends the assumption of the previous model that processing
-that involves higher levels of activation lead to longer pauses. Instead
-of assuming that there is one process that is shifted for IKIs of larger
-linguistic edges, we introduce the assumption that pauses at larger
-linguistic edges are more likely but not obligatory. This is achieved by
-modelling IKIs as coming from a weighted mixture of two distributions
-associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption. These fluent
-  keystroke transitions are merely constrained by a person's ability to
-  move their finger and will be captured by the \(\beta\) parameter in
-  equation \ref{eq:bimodcon}. Note that \(\beta\) is represented in both
-  log-Gaussian distributions in equation \ref{eq:bimodcon} referring to
-  the same unknown parameter.
-\item
-  Interruptions in the activation flow from higher to lower levels
-  result in longer keystroke intervals, when words or their spelling
-  could not be retrieved in time. The slowdown for these hesitant
-  transitions is captured by \(\delta\) in the first line of equation
-  \ref{eq:bimodcon}. The magnitude of the slowdown is associated with
-  transition location. This is because delays at larger linguistic units
-  are likely to be associated with higher level planning. By
-  constraining \(\delta\) to be positive, it captures how much longer
-  hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\label{eq:bimoduncon}
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+    \text{where: }  & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
@@
 
-The first line of equation \ref{eq:bimodcon} represents the distribution
-of hesitant key transitions, and the second line represents fluent key
-transitions. Each of these two distributions is associated with the
-mixing weight \(\theta\) which is a proportion that is constrained to be
-larger than 0 and smaller than 1. \(\theta\) is here parameterised to
-represent the probability that an IKI is associated with the
-distribution of long IKIs. This probability is inversely related to the
-mixing weight of the distribution of short IKIs by \(1-\theta\) . In
-other words, a larger weight of one distribution inevitably means a
-lower weight for the other distribution. The weights of both
-distributions must sum to 1. We will call this parameter the probability
-of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function
-of both transition location and participants. In line with the
-literature discussed in the introduction, we assume that pauses are more
-likely at larger linguistic edges. As pausing is subject to individual
-differences and writing style (skills), we also assume that some
-participants pause more at certain transition locations and other
-participants pause less. This is akin to what is known as a random
-by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the
-standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by
-transition location. In addition we constrained the variances so that
-\(\sigma_{e'}\) associated with the distribution of typing disfluencies
-is larger than the variance associated with fluent transitions
-\(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et
-al., 2017). This was achieved by introducing a parameter
-\(\sigma_\text{diff}\). The consequence is that fluent transitions are
-assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{two-distributions-log-gaussian-mixture-model-unconstrained}{%
-\paragraph{Two-distributions log-Gaussian mixture model
-(unconstrained)}\label{two-distributions-log-gaussian-mixture-model-unconstrained}}
-
-The size of a fluent key-transition does not necessarily vary by
-transition location. In other words, the parameter \(\beta\) is the same
-for before-sentence, before-word, and within-word transitions. This is
-what we called a constrained model. However, letter bigrams and trigrams
-may be systematically executed faster than transitions between between
-space and a letter (REFERENCE?) or complex keystrokes that comprise
-space and shift-letter combinations for upper case characters before
-sentences (we will address the latter possibility in the results
-section). This is because bigrams / trigrams might be stored, retrieved
-and executed as motor codes but not transitions to a space following a
-word or between a space or shift key press preceding a letter. Also,
-because of the necessarily larger number of within-word transitions, as
-opposed to before-word and sentence transitions, the posterior of the
-constrained model is dominated by within-word transition data. We
-therefore also implemented an unconstrained model that assumes that the
-size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as
-illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 < \theta < 1
-\end{aligned}
-\end{equation}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of
-cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \&
-Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman;
-Hall/CRC.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive
-modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S. (2007). From written word to
-written sentence production. In M. Torrance, L. van Waes, \& D. W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and
-applications} (Vol. 20, pp. 31--53). Elsevier.
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T. (2019). Advance planning in
-written and spoken sentence production. \emph{Journal of Experimental
-Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain
-and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B. (2017). Modelling
-dependency completion in sentence comprehension as a {B}ayesian
-hierarchical mixture process: {A} case study involving {C}hinese
-relative clauses. \emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting
-as a finite mixture process: Evidence from comprehension data.
-\emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S. (2007). On the linear relation between
-the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the
-timing of discrete motor responses. \emph{Perception \& Psychophysics},
-\emph{14}(1), 5--12.
-
-\end{CSLReferences}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
+\newpage
+
 \hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the single distribution Gaussian model which predicted large amounts
-of negative keystroke intervals which are implausible. Using a
-log-normal distribution removed these negative keystroke intervals.
+\section{Fit to data}\label{fit-to-data}}
+
+Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data.
+From each model we obtained 100 simulations (predictions).
+These can be found in Figure \ref{fig:prediction}.
+The model predictions fitted the data substantially better for the mixture models.
+The worse fit was observed for the single distribution Gaussian model which predicted large amounts of negative keystroke intervals which are implausible.
+Using a log-normal distribution removed these negative keystroke intervals.
 
@@
 \end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1constr.pdf}
-\caption{Posterior parameter distribution}
+
+\newpage
+
+\hypertarget{transition-location-effect-unconstrained-mixture-model}{%
+\section{Transition location effect (unconstrained mixture model)}\label{transition-location-effect-unconstrained-mixture-model}}
+
+We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
+Similar to the constrained mixture model, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.} and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.}.
+The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
+
+}
+
+\caption{Mixture model parameter estimates across studies.
+Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
+Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
 \end{figure}
+
+In the following we evaluate differences between transition locations for all three mixture model parameters.
+Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
+We found that hesitations appear more frequently at before-word transitions than within words across data sets (all BFs \textgreater{} 100).
+Also hesitations are longer at before-sentence transitions compared to before-word transitions (all BFs \textgreater{} 100; except for C2L1: BF = 0.35 and LIFT = 2.14) and before-word transitions compared to within-word transitions (all BF \textgreater{} 10; except for LIFT: BF = 0.71).
+Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations (all BFs \textgreater{} 100) but there is generally no evidence for a difference for fluent transitions for before-sentence transitions compared to before-word transitions (BFs \textless{} 0.08; except for SPL2: BF \textgreater{} 100 and GUNNEXP2: BF \textgreater{} 100).
+The full results of these pairwise comparisons can be found in Table \ref{tab:loceffect}.
+
+Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
+However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
+This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
+In fact, the data set LIFT showed less pausing before sentences compared to before words.
+In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
+
+Outstanding is the overall substantially longer fluent transitions for the C2L1 data.
+This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
+Hesitation duration and frequencies were similar to the other data sets.
+There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
+Some data sets show before-sentence slowdowns for fluent transition compared to words.
+These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
+In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
+Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
+
+To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
+The results are shown in Appendix \ref{key-combination-effect-unconstrained-mixture-model}.
+We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
+However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
+In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
+
+\begin{landscape}
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
+Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\% PIs in brackets.}\\
+\toprule
+ \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
+\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
+Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
+\toprule
+ \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
+\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
+Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
+\midrule
+\endhead
+\textbf{C2L1} &  &  &  &  &  & \\
+\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
+\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
+\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
+\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
+\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
+\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
+\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
+\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
+\textbf{LIFT} &  &  &  &  &  & \\
+\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
+\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
+\textbf{PLanTra} &  &  &  &  &  & \\
+\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
+\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
+\textbf{SPL2 (L1)} &  &  &  &  &  & \\
+\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
+\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+\end{landscape}
+
 \newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2constr.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-constrained-mixture-model}{%
-\subsection{Key-combination effect (constrained mixture
-model)}\label{key-combination-effect-constrained-mixture-model}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences summed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroke involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}.
+
+\hypertarget{key-combination-effect-unconstrained-mixture-model}{%
+\section{Key-combination effect (unconstrained mixture model)}\label{key-combination-effect-unconstrained-mixture-model}}
+
+The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+In other words, the pause before sentences sumed across two key intervals in the PLanTra and LIFT data, namely \texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval, namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets.
+Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
+Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.
+
+Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
+We compared the SPL2 data including and excluding the keystroke after shift.
+Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans2}.
 
@@
 
+\begin{longtable}{lrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:shiftcellmeans2}Mixture model estimates for key transitions.
+Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
+The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+95\% PIs in brackets.}\\
+\toprule
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:shiftcellmeans2} continued}}\\
+\toprule
+Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Fluent transitions &  &  & \\
+\ \ \ L1 & 159 [147, 172] & 0.48 [0.34, 0.63] & > 100\\
+\ \ \ L2 & 159 [147, 172] & 0.41 [0.19, 0.63] & 46.08\\
+Hesitation duration &  &  & \\
+\ \ \ L1 & 1,281 [1,082, 1,496] & -0.46 [-0.61, -0.3] & > 100\\
+\ \ \ L2 & 1,802 [1,557, 2,073] & -0.34 [-0.51, -0.17] & > 100\\
+Hesitation probability &  &  & \\
+\ \ \ L1 & .78 [.69, .86] & 0.49 [-0.04, 1.03] & 1.43\\
+\ \ \ L2 & .92 [.87, .96] & 0.48 [-0.16, 1.13] & 0.94\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+
+Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift.
+Fluent key transitions were substantially longer when including the interval following the shift key.
+The slowdown for hesitations was affected too but the difference is numerically small.
+There was no conclusive evidence for an increased hesitation probability.
+
+\newpage
+
+\hypertarget{key-combination-effect-constrained-mixture-model}{%
+\section{Key-combination effect (constrained mixture model)}\label{key-combination-effect-constrained-mixture-model}}
+
+The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely \texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval, namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets.
+Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
+
+Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
+We compared the SPL2 data including and excluding the keystroke after shift.
+Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
 \begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
 \caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions immediately preceding a sentence.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions.
+Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown and the probability of hesitant transitions.
 The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
@@
 
-The fluent transition duration and hesitation duration were not affected
-by whether or not the sentence-initial transition include the character
-following shift for neither language. However, we found strong evidence
-for an increased hesitation probability -- in both -- languages when the
-before-sentence key transition included the character following the
-shift key. Notably, the hesitation probability increased ceiling. It is
-not entirely surprised that the constrained mixture model identifies
-essentially all before-sentence transition as pauses: the constrained
-model does not distinguish between transition locations for the duration
-of fluent transitions. Therefore essentially all before-sentence
-transitions are identified as hesitation as they always include two
-keystrokes while all other transition location (which is the majority of
-the data) do not.
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
+The fluent transition duration and hesitation duration were not affected by whether or not the sentence-initial transition include the character following shift for neither language.
+However, we found strong evidence for an increased hesitation probability -- in both -- languages when the before-sentence key transition included the character following the shift key.
+Notably, the hesitation probability increased ceiling.
+It is not entirely surprised that the constrained mixture model identifies essentially all before-sentence transition as pauses: the constrained model does not distinguish between transition locations for the duration of fluent transitions.
+Therefore essentially all before-sentence transitions are identified as hesitation as they always include two keystrokes while all other transition location (which is the majority of the data) do not.
+
+\newpage
+
 \hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer hesitation duration and more
-pauses across all transition locations when writing in L2. Only
-within-word transitions did show negligible evidence for an increased
-hesitation duration in L2.
+\section{L2 effect (SPL2)}\label{l2-effect-spl2}}
+
+For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}} sentence-transitions) we calculated the L2 effect (i.e.~the difference between writing in L2 and L1).
+The results can be found in Table \ref{tab:l2effect}.
+The results show longer hesitation duration and more pauses across all transition locations when writing in L2.
+Only within-word transitions did show negligible evidence for an increased hesitation duration in L2.
 
@@
 \end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
+
+\newpage
+
 \hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. There was some evidence for longer
-before-sentence hesitation durations when the text was masked in
-non-dyslexic writers. Evidence for all other comparisons was negligible.
-
-\blandscape
+\section{Masking effect (CATO, GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
+
+Studies associated with two datasets (CATO, GUNNEXP2) investigated to what extent masking the previously written text affects keystroke behaviour.
+The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}.
+There was some evidence for longer before-sentence hesitation durations when the text was masked in non-dyslexic writers.
+Evidence for all other comparisons was negligible.
+
+\begin{landscape}
 
@@
 \end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
+\end{landscape}
+
+\newpage
+
+\hypertarget{posterior-parameter-estimates}{%
+\section{Posterior parameter estimates}\label{posterior-parameter-estimates}}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics{figures/psplots1constr.pdf}
+\caption{Posterior parameter distribution}
+\end{figure}
+\newpage
+\begin{figure}[!htb]
+\ContinuedFloat
+\captionsetup{list=off,format=cont}
+\centering
+\includegraphics{figures/psplots2constr.pdf}
+\label{fig:fullps1}
+\caption{Posterior parameter distribution}
+\end{figure}
+
+\newpage
+
 \hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Evidence for all comparisons was negligible.
+\section{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
+
+The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}. Evidence for all comparisons was negligible.
 
@@
 \end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
+
+\newpage
+
 \hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a genre
-comparisons was negligible.
+\section{Genre effect (LIFT)}\label{genre-effect-lift}}
+
+From the LIFT data, we compared the difference between genres, i.e.~when writing an informative text as opposed to writing an argumentative text.
+The results are shown in Table \ref{tab:genreeffect}.
+Cellmeans and differences were average across writing topic.
+Evidence for a genre comparisons was negligible.
 
@@
 \end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
+
+\newpage
+
 \hypertarget{transition-location-effect-constrained-mixture-model}{%
-\subsection{Transition location effect (constrained mixture
-model)}\label{transition-location-effect-constrained-mixture-model}}
+\section{Transition location effect (constrained mixture model)}\label{transition-location-effect-constrained-mixture-model}}
 
@@
 \end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{transition-location-effect-unconstrained-mixture-model}{%
-\subsection{Transition location effect (unconstrained mixture
-model)}\label{transition-location-effect-unconstrained-mixture-model}}
-
-We present the mixture-model posterior estimates for each of the three
-parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Similar to the constrained mixture model, we aggregated the posterior
-across conditions\footnote{We aggregated across pre-post test for the
-  PLanTra dataset as well as genre and topic of the LIFT data set.} and
-removed conditions that might conflate comparisons\footnote{We removed
-  the masked writing condition in the GUNNEXP2 and CATO, the dyslexic
-  group in the CATO data set, and L2 writing in the SPL2 data set.}. The
-resulting posterior allows us to examine differences between transition
-locations for each data set associated with each of the three
-mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations
-for all three mixture model parameters. Figure \ref{fig:crossstudypost}
-shows largely the same patterns (with caveats) for keystroke interval
-estimates by transition location across data sets for all three mixture
-model parameters. We found that hesitations appear more frequently at
-before-word transitions than within words across data sets (all BFs
-\textgreater{} 100). Also hesitations are longer at before-sentence
-transitions compared to before-word transitions (all BFs \textgreater{}
-100; except for C2L1: BF = 0.35 and LIFT = 2.14) and before-word
-transitions compared to within-word transitions (all BF \textgreater{}
-10; except for LIFT: BF = 0.71). Further, we observed that fluent
-key-transitions are slower at before-word and before-sentence locations
-compared to within-word locations (all BFs \textgreater{} 100) but there
-is generally no evidence for a difference for fluent transitions for
-before-sentence transitions compared to before-word transitions (BFs
-\textless{} 0.08; except for SPL2: BF \textgreater{} 100 and GUNNEXP2:
-BF \textgreater{} 100). The full results of these pairwise comparisons
-can be found in Table \ref{tab:loceffect}.
-
-Hesitation duration tends to be longer at before-sentence locations
-compared to before-word locations (except for data sets C2L1 and LIFT)
-and longer for before-word locations compared to within-word locations
-(except for the LIFT data set). However, for most data set there is
-negligible evidence for the idea that writers pause more frequently at
-before-sentence locations compared to before-word locations (except for
-data sets SPL2 and GUNNEXP2). This is interesting because it is
-generally believed that pausing behaviour is associated with syntactic
-edges such that more and longer pauses are predicted for key transitions
-at larger syntactic edges following the pattern before-sentence \(>\)
-before-word \(>\) within-word. In fact, the data set LIFT showed less
-pausing before sentences compared to before words. In brief, while
-keystroke transitions and pauses tend to be longer and more frequent at
-before-word locations compared to within-word transitions, it is not
-clear in which contexts keystroke transitions are before-sentence
-locations are slower, and their hesitations are longer and more
-frequent.
-
-Outstanding is the overall substantially longer fluent transitions for
-the C2L1 data. This is presumably reflecting that the population that
-this sample is from was the youngest among our data sets presumably
-involving the least experienced writers in our data pool. Hesitation
-duration and frequencies were similar to the other data sets. There are
-some inconsistencies for fluent before-sentence transitions compared to
-before-word transitions. Some data sets show before-sentence slowdowns
-for fluent transition compared to words. These inconsistencies could, to
-some extent, be explained on the basis that data sets differ as to
-whether before-sentence transitions involve complex key combination
-involve the mean or sum of transitions between space, shift and / or the
-sentence-initial letters. In particular, some data include the character
-key following the shift key at before sentence location (PLanTra, LIFT)
-but others did not scope over the character following the shift key
-(CATO, C2L1, SPL2, GUNNEXP2). Notice though that for the data sets
-PLanTra and LIFT, there was no evidence for a consistent differences
-between before-sentence and before-word transitions (except for longer
-hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence
-transitions were timed, we test to what extent this difference affected
-the modelling results for the SPL2 data set. The results are shown in
-Appendix \ref{key-combination-effect-unconstrained-mixture-model}. We
-found that including the character following the shift key substantially
-extends both the transition duration and the hesitation duration but not
-the hesitation frequency. However, this conflicts with the absence of
-differences in data sets that included the character following shift at
-before-sentence transitions (PLanTra, LIFT). In other words, it is
-unlikely that patterns in our results can be explained on the basis of
-how before-keystroke transitions were operationalised (complex
-key-combinations at before-sentence locations).
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-unconstrained-mixture-model}{%
-\subsection{Key-combination effect (unconstrained mixture
-model)}\label{key-combination-effect-unconstrained-mixture-model}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroke involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans2}.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans2}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans2} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  & \\
-\ \ \ L1 & 159 [147, 172] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 159 [147, 172] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  & \\
-\ \ \ L1 & 1,281 [1,082, 1,496] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 1,802 [1,557, 2,073] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  & \\
-\ \ \ L1 & .78 [.69, .86] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .92 [.87, .96] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-Overall, fluent transition duration and the hesitation duration were
-affected by whether or not the sentence-initial transition include the
-character following shift. Fluent key transitions were substantially
-longer when including the interval following the shift key. The slowdown
-for hesitations was affected too but the difference is numerically
-small. There was no conclusive evidence for an increased hesitation
-probability.
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
+
+\newpage
+
 \hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for two-distribution mixture models -- is that, in
-principle, as the mixture model has more parameters it might always lead
-to a better fit. We addressed this concern before by using
-cross-validation techniques for model comparison which is preventing
-overfitting models. To address this concern we repeated a comparison for
-a single distribution model and a two-distrobutionsl mixture model for
-two sets of simulated data that were simulated either with a single
-distribution or two-distributional process at heart. In other words,
-this allows us to test the predictive performance of our models in a
-context where we know the true underlying data generating process --
-single distribution vs two distributions -- and we can test whether
-these models can successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a
-two-distributions mixture model with two mixture components similar to
-the process described above (equations in \ref{eq:bimodcon} and
-\ref{eq:bimoduncon}). This process and the corresponding Bayesian model
-is summarised in equation \ref{eq:simmog}.
+\section{Simulation}\label{simulation}}
+
+A possible concern with these results -- substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
+To address this concern we repeated a comparison for a single distribution model and a two-distributions mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart.
+In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.
+
+In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}).
+This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
 
@@
 
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a single distribution
-log-Gaussian process. The model and its corresponding Bayesian model is
-summarised in equation \ref{eq:simuv}.
+This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants).
+The model includes two log-normal distributions with a mixing proportion \(\theta\) of which the distribution of shorter values has a mean of \(\beta\) and a standard deviation \(\sigma^2_2\); the second distribution of longer values is constrained to have a mean that is larger by a factor of \(\delta\) and has a larger standard deviation \(\sigma^2_1\).
+
+The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
 
@@
 
-Again, this model is a simplified version of the single distribution
-models used in the main analysis. The model assumes a log-Gaussian
-distribution with a mean parameter \(\beta\) and a standard deviation
-\(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
+Again, this model is a simplified version of the single distribution models used in the main analysis.
+The model assumes a log-Gaussian distribution with a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
+
+The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
+The simulated data are visualised in Figure \ref{fig:simdata}.
+Parameter values were chosen so that the simulated data are roughly similarly distributed to keystroke transitions.
 
@@
 
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a two-distribution mixture model and a single-distribution
-model -- each for both data sets. Models were run with 3 chains, with
-each 6,000 iterations of which 3,000 were warmup. Estimates with 95\%
-probability intervals are shown in Table \ref{tab:simparam}. The
-parameters are shown by type of data generating process along with the
-true parameter values. Parameter value estimates are shown by type of
-Bayesian model. The results show that either of the two Bayesian models
-succesfully uncovered the model parameters of the data with its
-corresponding data generating process, as shown in Table
-\ref{tab:simparam}, but less so when the model was applied to data
-generated with the incorrect underlying process. Particularly the mixing
-proportion \(\theta\) and the slowdown parameter \(\delta\) were not
-uncovered at all by they single-distribution model.
+For each of these two data sets we simulated 1,000 observations.
+We run 2 models -- a two-distribution mixture model and a single-distribution model -- each for both data sets.
+Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
+Estimates with 95\% probability intervals are shown in Table \ref{tab:simparam}.
+The parameters are shown by type of data generating process along with the true parameter values.
+Parameter value estimates are shown by type of Bayesian model.
+The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
+Particularly the mixing proportion \(\theta\) and the slowdown parameter \(\delta\) were not uncovered at all by they single-distribution model.
 
@@
 
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a single-distribution process. However, for the data generated with
-a two-distribution mixture process, the mixture model shows a
-substantially higher predictive performance. In fact, the ratio of
-\(\Delta\widehat{elpd}\) and its standard error -- a \emph{z}-score
-approximation -- reveals a value of 11.6 which is strong evidence in in
-favour of the mixture model compared to the single-distribution model
-(Sivula et al., 2020). In comparison, for the single-distribution data,
-the single-distribution model was negligibly better than the
-two-distribution mixture model which was evidence by a value of 0.77 for
-the ratio of \(\Delta\widehat{elpd}\) and its standard error. Thus, as
-the difference \(\Delta\widehat{elpd}\) between the model is negligible,
-the single distribution is preferred by the law of parsimony.
+We used LOO-CV to compare the predictive performance of the two models for each data generating process.
+The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
+The results rule out the possibility that the mixture model does always lead to higher predictive performance.
+Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a single-distribution process.
+However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
+In fact, the ratio of \(\Delta\widehat{elpd}\) and its standard error -- a \emph{z}-score approximation -- reveals a value of 11.6 which is strong evidence in in favour of the mixture model compared to the single-distribution model (Sivula et al., 2020).
+In comparison, for the single-distribution data, the single-distribution model was negligibly better than the two-distribution mixture model which was evidence by a value of 0.77 for the ratio of \(\Delta\widehat{elpd}\) and its standard error.
+Thus, as the difference \(\Delta\widehat{elpd}\) between the model is negligible, the single distribution is preferred by the law of parsimony.
 
@@
 \end{table}
-\end{appendix}
+
+\newpage
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\begingroup
+
+\setlength{\parindent}{-0.5in}
+
+\setlength{\leftskip}{0.5in}
+
+\hypertarget{references}{}
+
+\endgroup
+
+\newpage
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-ailhaud2018variations}{}}%
+Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ailhaud2016developmental}{}}%
+Ailhaud, E., Chenu, F., \& Jisa, H.
+(2016).
+A developmental perspective on the units of written {F}rench.
+In J.
+Perera, M.
+Aparici, E.
+Rosado, \& N.
+Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
+287--305).
+Springer.
+\url{https://doi.org/10.1007/978-3-319-21136-7_17}
+
+\leavevmode\vadjust pre{\hypertarget{ref-alamargot2010using}{}}%
+Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
+(2010).
+Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
+\emph{Reading and Writing}, \emph{23}(7), 853--888.
+
+\leavevmode\vadjust pre{\hypertarget{ref-almond2012preliminary}{}}%
+Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
+(2012).
+\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
+RR-12-23).
+Educational Testing Service.
+
+\leavevmode\vadjust pre{\hypertarget{ref-alves2015progress}{}}%
+Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
+
+\leavevmode\vadjust pre{\hypertarget{ref-baaijen2012keystroke}{}}%
+Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
+
+\leavevmode\vadjust pre{\hypertarget{ref-beers2010adolescent}{}}%
+Beers, S.
+F., Quinlan, T., \& Harbaugh, A.
+G.
+(2010).
+Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
+\emph{Reading and Writing}, \emph{23}(7), 743--775.
+
+\leavevmode\vadjust pre{\hypertarget{ref-behmer2016crunching}{}}%
+Behmer Jr., L.
+P., \& Crump, M.
+J.
+C.
+(2016).
+Crunching big data with fingertips: How typists tune their performance toward the statistics of natural language.
+In M.
+N.
+Jones (Ed.), \emph{Big data in cognitive science} (Vol.
+1, pp.
+329--345).
+Psychology Press.
+\url{https://doi.org/10.4324/9781315413570-27}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bock2014syntactically}{}}%
+Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bonin2012evidence}{}}%
+Bonin, P., Roux, S., Barry, C., \& Canell, L.
+(2012).
+Evidence for a limited-cascading account of written word naming.
+\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
+\url{https://doi.org/10.1037/a0028471}
+
+\leavevmode\vadjust pre{\hypertarget{ref-carpenter2016stan}{}}%
+Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
+A., Guo, J., Li, P., \& Riddell, A.
+(2016).
+Stan: A probabilistic programming language.
+\emph{Journal of Statistical Software}, \emph{20}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chang2006}{}}%
+Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chanquoy1996writing}{}}%
+Chanquoy, L., Foulin, J.-N., \& Fayol, M.
+(1996).
+Writing in adults: {A} real-time approach.
+In G.
+Rijlaarsdam, H.
+Van den Bergh, \& M.
+Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
+36--44).
+Amsterdam University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chenoweth2001fluency}{}}%
+Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chenoweth2003inner}{}}%
+Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chenu2014interword}{}}%
+Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
+
+\leavevmode\vadjust pre{\hypertarget{ref-christiansen2016now}{}}%
+Christiansen, M.
+H., \& Chater, N.
+(2016).
+The now-or-never bottleneck: A fundamental constraint on language.
+\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
+\url{https://doi.org/10.1017/S0140525X1500031X}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chukharev2014pauses}{}}%
+Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chukharev2019combined}{}}%
+Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
+(2019).
+Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
+\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Clark2002}{}}%
+Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
+
+\leavevmode\vadjust pre{\hypertarget{ref-conijn2020keys}{}}%
+Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
+
+\leavevmode\vadjust pre{\hypertarget{ref-conijn2019understanding}{}}%
+Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
+
+\leavevmode\vadjust pre{\hypertarget{ref-conijn2021timecourse}{}}%
+Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
+
+\leavevmode\vadjust pre{\hypertarget{ref-connelly2012predicting}{}}%
+Connelly, V., Dockrell, J.
+E., Walter, K., \& Critten, S.
+(2012).
+Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
+\emph{Written Communication}, \emph{29}(3), 278--302.
+
+\leavevmode\vadjust pre{\hypertarget{ref-crump2010hierarchical}{}}%
+Crump, M.
+J.
+C., \& Logan, G.
+D.
+(2010).
+Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
+\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
+\url{https://doi.org/10.1037/a0020696}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dam09}{}}%
+Damian, M.
+F., \& Stadthagen-Gonzalez, H.
+(2009).
+Advance planning of form properties in the written production of single and multiple words.
+\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
+
+\leavevmode\vadjust pre{\hypertarget{ref-de2018exploring}{}}%
+De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dell2007case}{}}%
+Dell, G.
+S., Martin, N., \& Schwartz, M.
+F.
+(2007).
+A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
+\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dux2021effect}{}}%
+Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dux2022automating}{}}%
+Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
+(2022).
+Automating individualized, process-focused writing instruction: A design-based research study.
+\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
+\url{https://doi.org/0.3389/fcomm.2022.933878}
+
+\leavevmode\vadjust pre{\hypertarget{ref-farrell2018computational}{}}%
+Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-flower1980ynamics}{}}%
+Flower, L.
+S., \& Hayes, J.
+R.
+(1980).
+The dynamics of composing: Making plans and juggling constraints.
+In L.
+W.
+Gregg \& E.
+R.
+Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
+31--50).
+Lawrence Erlbaum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-flower1981pregnant}{}}%
+Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
+
+\leavevmode\vadjust pre{\hypertarget{ref-foulin1998extent}{}}%
+Foulin, J.-N.
+(1998).
+To what extent does pause location predict pause duration in adults' and children's writing?
+\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gelman2014}{}}%
+Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gelman1992}{}}%
+Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gernsbacher1995coherence}{}}%
+Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gri01}{}}%
+Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
+
+\leavevmode\vadjust pre{\hypertarget{ref-griffin2003reversed}{}}%
+Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
+
+\leavevmode\vadjust pre{\hypertarget{ref-grudin1982digraph}{}}%
+Grudin, J. T., \& Larochelle, S. (1982). \emph{Digraph frequency effects in skilled typing} (ONR-8201). Office of Naval Research.
+
+\leavevmode\vadjust pre{\hypertarget{ref-guo2018modeling}{}}%
+Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hall2022constructing}{}}%
+Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hayes2012evidence}{}}%
+Hayes, J.
+R.
+(2012).
+Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
+In M.
+Fayol, D.
+Alamargot, \& V.
+Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
+15--25).
+Psychology Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jefferys1992ockham}{}}%
+Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kaufer1986composing}{}}%
+Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lambert2018student}{}}%
+Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lee13}{}}%
+Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lee2014bayesian}{}}%
+Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-leijten2013keystroke}{}}%
+Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
+
+\leavevmode\vadjust pre{\hypertarget{ref-li2021identifying}{}}%
+Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
+
+\leavevmode\vadjust pre{\hypertarget{ref-peel2000finite}{}}%
+MacLahlan, G., \& Peel, D. (2000). \emph{Finite mixture models}. John \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-matsuhashi1981pausing}{}}%
+Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mcelreath2016statistical}{}}%
+McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-medimorec2017pauses}{}}%
+Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
+
+\leavevmode\vadjust pre{\hypertarget{ref-medimorec2017disfluency}{}}%
+Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
+
+\leavevmode\vadjust pre{\hypertarget{ref-meyer1996}{}}%
+Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mohsen2021second}{}}%
+Mohsen, M.
+(2021).
+Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
+\emph{Psycholinguistics}, \emph{30}(1), 180--202.
+\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
+
+\leavevmode\vadjust pre{\hypertarget{ref-not10}{}}%
+Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
+
+\leavevmode\vadjust pre{\hypertarget{ref-not07}{}}%
+Nottbusch, G., Weingarten, R., \& Sahel, S.
+(2007).
+From written word to written sentence production.
+In M.
+Torrance, L.
+van Waes, \& D.
+W.
+Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
+20, pp.
+31--53).
+Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-olive2014toward}{}}%
+Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ostry1983determinants}{}}%
+Ostry, D.
+J.
+(1983).
+Determinants of interkey times in typing.
+In W.
+E.
+Cooper (Ed.), \emph{Cognitive aspects of skilled typewriting} (pp.
+225--246).
+Springer.
+\url{https://doi.org/10.1007/978-1-4612-5470-6_9}
+
+\leavevmode\vadjust pre{\hypertarget{ref-R-base}{}}%
+R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rochester1973significance}{}}%
+Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
+
+\leavevmode\vadjust pre{\hypertarget{ref-roeser2021modelling}{}}%
+Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
+(2021).
+Modelling typing disfluencies as finite mixture process.
+\emph{Reading and Writing}, 1--26.
+\url{https://doi.org/10.1007/s11145-021-10203-z}
+
+\leavevmode\vadjust pre{\hypertarget{ref-roeser2018advance}{}}%
+Roeser, J., Torrance, M., \& Baguley, T.
+(2019).
+Advance planning in written and spoken sentence production.
+\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
+\url{https://doi.org/10.1037/xlm0000685}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ronneberg2022process}{}}%
+Rønneberg, V., Torrance, M., Uppstad, P.
+H., \& Johansson, C.
+(2022).
+The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
+\emph{Psychological Research}, \emph{86}(7), 2239--2255.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rossetti2022s}{}}%
+Rossetti, A., \& Van Waes, L.
+(2022a).
+It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
+\emph{Frontiers in Artificial Intelligence}, \emph{5}.
+\url{https://doi.org/10.3389/frai.2022.983008}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rossetti2022text}{}}%
+Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
+
+\leavevmode\vadjust pre{\hypertarget{ref-roux2013interaction}{}}%
+Roux, S., McKeeff, T.
+J., Grosjacques, G., Afonso, O., \& Kandel, S.
+(2013).
+The interaction between central and peripheral processes in handwriting production.
+\emph{Cognition}, \emph{127}(2), 235--241.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schilperoord2002cognitive}{}}%
+Schilperoord, J.
+(2002).
+On the cognitive status of pauses in discourse production.
+In T.
+Olive \& C.
+M.
+Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
+61--87).
+Springer.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schoner2002timing}{}}%
+Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sim07}{}}%
+Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sivula2020uncertainty}{}}%
+Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rstan}{}}%
+Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sullivan2002self}{}}%
+Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
+
+\leavevmode\vadjust pre{\hypertarget{ref-terzuolo1980determinants}{}}%
+Terzuolo, C.
+A., \& Viviani, P.
+(1980).
+Determinants and characteristics of motor patterns used for typing.
+\emph{Neuroscience}, \emph{5}(6), 1085--1103.
+\url{https://doi.org/10.1016/0306-4522(80)90188-8}
+
+\leavevmode\vadjust pre{\hypertarget{ref-torrance201203}{}}%
+Torrance, M.
+(2012).
+EyeWrite -- a tool for recording writers' eye movements.
+In M.
+Torrance, D.
+Alamargot, M.
+Castelló, F.
+Ganier, O.
+Kruse, A.
+Mangen, L.
+Tolchinsky, \& L.
+V.
+Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
+355--357).
+Brill.
+\url{https://doi.org/doi.org/10.1163/9781780529295_082}
+
+\leavevmode\vadjust pre{\hypertarget{ref-torrance2016reading}{}}%
+Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
+(2016).
+Reading during the composition of multi-sentence texts: An eye-movement study.
+\emph{Psychological Research}, \emph{80}(5), 729--743.
+
+\leavevmode\vadjust pre{\hypertarget{ref-torrance2012written}{}}%
+Torrance, M., \& Nottbusch, G.
+(2012).
+Written production of single words and simple sentences.
+In V.
+W.
+Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
+403--422).
+Psychology Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-torrance2018timed}{}}%
+Torrance, M., Nottbusch, G., Alves, R.
+A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
+I., et al.
+(2018).
+Timed written picture naming in 14 {E}uropean languages.
+\emph{Behavior Research Methods}, \emph{50}, 744--758.
+
+\leavevmode\vadjust pre{\hypertarget{ref-torranceb}{}}%
+Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-torrancea}{}}%
+Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-torrance2016adolescent}{}}%
+Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
+H.
+(2016).
+Adolescent weak decoders writing in a shallow orthography: Process and product.
+\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
+
+\leavevmode\vadjust pre{\hypertarget{ref-van1991handwriting}{}}%
+Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
+
+\leavevmode\vadjust pre{\hypertarget{ref-van2016keystroke}{}}%
+Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
+
+\leavevmode\vadjust pre{\hypertarget{ref-van2019multilingual}{}}%
+Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
+(2019).
+A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
+\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
+
+\leavevmode\vadjust pre{\hypertarget{ref-van2010reading}{}}%
+Van Waes, L., Leijten, M., \& Quinlan, T.
+(2010).
+Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
+\emph{Reading and Writing}, \emph{23}(7), 803--834.
+
+\leavevmode\vadjust pre{\hypertarget{ref-waes2019}{}}%
+Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
+(2021).
+Measuring and assessing typing skills in writing research.
+\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
+\url{https://doi.org/10.17239/jowr-2021.13.01.04}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vandermeulen2020mapping}{}}%
+Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
+(2020).
+Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
+\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vandermeulen2020}{}}%
+Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vandermeulen2023writing}{}}%
+Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
+(2023).
+Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
+\emph{Written Communication}, \emph{40}(1), 90--144.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vasishth2017}{}}%
+Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
+(2017).
+Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
+\emph{ArXiv e-Prints}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vasishth2017feature}{}}%
+Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vehtari2015pareto}{}}%
+Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vehtari2017practical}{}}%
+Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers2007linear}{}}%
+Wagenmakers, E.-J., \& Brown, S.
+(2007).
+On the linear relation between the mean and the standard deviation of a response time distribution.
+\emph{Psychological Review}, \emph{114}(3), 830--841.
+\url{https://doi.org/10.1037/0033-295X.114.3.830}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wen02}{}}%
+Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wen06}{}}%
+Wengelin, Å.
+(2006).
+Examining pauses in writing: {T}heory, methods and empirical data.
+In K.
+P.
+H.
+Sullivan \& E.
+Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
+18, pp.
+107--130).
+Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wheeldon_konopka_2023}{}}%
+Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wheeldon2013}{}}%
+Wheeldon, L.
+R., Ohlson, N., Ashby, A., \& Gator, S.
+(2013).
+Lexical availability and grammatical encoding scope during spoken sentence production.
+\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wing1973response}{}}%
+Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
+
+\end{CSLReferences}
 

### 2024-04-27 report/sections/conclusion.Rmd
rearrange markdown sections

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
 In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
@@
 Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}
+We found that key transitions across word and even sentences boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
+This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.
+young / L2 writers, students) and writing tasks (e.g.
+essays, syntheses).

### 2024-04-27 report/sections/discussion.Rmd
rearrange markdown sections

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
-
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
+The classical serial view [@flower1980ynamics] characterizes writing as a sequence of planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
+The cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production [@olive2014toward].
+We used Bayesian statistical models, in particular hierarchical single-distribution and two-distribution mixture models, to directly compare the serial and the cascading model of writing, respectively.
+On the basis of six datasets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
 
 Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
+Under the cascading view of writing, key transitions that directly precede sentences or words are not necessarily associated with a pause because writers can plan in parallel to writing.
+On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
+We found that in the majority of datasets, pauses were as likely to occur before sentences and words, in both experienced and novice writers, while mid-word pauses are generally rare.
+Also, with few exceptions, pauses were typically longer before sentences than before words (which were longer than mid-word pauses).
+In other words, when writers do pause, the involvement of higher-level processing at sentence boundaries is typically leading to longer pauses.
 However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
+This is because our results for the hesitation probability suggest that writers often do not pause before sentences; indeed, hesitation probabilities are largely identical before sentences and before words.
+This suggest that writers frequently plan the next word and even sentence in parallel to production which mitigates the need to pause to plan up-coming linguistic information.
 
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
+Interestingly even the youngest writers in our data pool (C2L1) showed evidence of parallel planning expressed in fluent transitions between words and even sentences.
+Key-transitions were substantially longer compared to all other datasets with a larger slowdown before words.
 Yet the overall pausing probability showed a patterns comparable to most other data sets.
 In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
+This is interesting because, for example, @olive2014toward described a cascading model that is operating in a serial fashion for inexperienced or struggling writers explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
+However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers but pauses were more likely before words and sentences but similarly long indicating a tendency to prepare utterances in planning units possibly smaller than the clause.
+The main difference for these young writers was that their writing speed (fluent transitions) was substantially slower compared to more experienced writers in the sample.
+Also for L2 writers (PlanTra) we observe pausing patterns that do not resemble a serialised writing process as indicated by a larger number of fluent transitions across word and sentence boundaries.
 
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
+We observed two more differences between data sets -- other than the patterns observed for C2L1 summarised above -- that are worth highlighting.
+These can largely be explained on the basis of differences associated with the population (writing competence / experience) and text genre [see also @conijn2019understanding] that these samples were obtained from: The first exception concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations.
+**TODO: I don't have an explanation for this.** Second, pauses were not always longer before sentence compared to before words compared to mid-word paues; there were two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
+(2) pause durations did not vary by transition location in the LIFT dataset although the posterior showed the same general patterns as observed in most other datasets.
+The absence of a difference could be explained as specific to the synthesis writing task because there is no need for content creation [**REFERENCE?**] but the results for the PlanTra dataset rule out this possibility.
+
+A possible concern with our result -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
+This concern is preempted by using cross-validation techniques for model comparison which is preventing model overfit.
+In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
+We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
+This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
 Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
+As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
+Parameters were not successfully uncovered when we switched model type and dataset.
+Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a single distributions process but a substantially higher predictive performance for data simulated from a two-distributions mixture process.
+Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of overfitting.
 
@@
 Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
+There is limited research that used eye tracking to investigate regressive gaze behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
+There are at least four reasons why eye movements away from the inscription point might occure during writing: First, reading is necessary for revision (i.e.
+writers must check if writing goals have been met).
+Second, writers might use text to cue or reinstating ideas after interruption of writing process.
+Third, writers might look for or correct spelling or grammatical errors.
+We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text.
+Fourth, writers might look at the source their are writing from.
+For two datasets (PlanTra, LIFT) which involved writing from sources, we removed transitions that were made when the sources (rather than the writing document) was in focus.
+
+Frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
+This is except, lookbacks are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
+Finally even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
+This can be seen in the result of two datasets in our pool (CATO, GUNNEXP2): these experiments included a manipulation where the produced text was masked (or not).
 In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
+This suggest that pausing patterns do not change if reading previous text is not possible.
 
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
+A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
+Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent].
+There is one central problems with this approach: A threshold requires a definition of what passes as a pause [@wen06; @van2016keystroke], i.e.
+a pause criterion threshold often set to 2 secs [@chanquoy1996writing; @kaufer1986composing; @sullivan2002self; @wen02] or some other lower bound [@chukharev2014pauses; @connelly2012predicting; @leijten2013keystroke].
+Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause [see e.g.
+@wen06] they would be arbitrary.
+For example, a sentence-initial pause of 2 secs has a very different interpretation from a 2 secs pause that occurs before or within a word.
 Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
 So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}
+However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose [@chenu2014interword].
+Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying].

### 2024-04-27 report/sections/introduction.Rmd
rearrange markdown sections

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
-
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
-
-\begin{document}
-\maketitle
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
+Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing [@bock2014syntactically; @wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].
 
@@
 Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
+Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking [@roeser2018advance; @torrance2012written].
+This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader [e.g.
+@Clark2002].
 Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
+Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].
 
 The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
+On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06].
+For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses].
+The same is generally observed for spoken utterances [@lee13; @meyer1996; @wheeldon2013] with some exceptions [@gri01].
 We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
@@
 
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
+Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements [@van1991handwriting; @olive2014toward].
 The intervals between adjacent keystrokes during typing vary as a function of various factors.
@@
 Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
+The subsequent burst of fluent output is then the result this planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing].
+This is particularly reflected in the model proposed by @flower1980ynamics that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts [also @chenoweth2001fluency; @chenoweth2003inner].
 
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
+This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second].
 However this account is not consistent with what we know about writing.
@@
 Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
+This is well known for spoken production [e.g.
+@bock2014syntactically].
+For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases [@not10; dam09; @roeser2018advance].
+For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e.
+*A and the B moved*) or a simple noun (i.e., *A moved above the B and the C*).
 Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
 This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
+In fact, eye movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e.
+*the B*) was more likely for conjoined noun phrases.
 
 Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
+For example, @medimorec2017pauses found that, in undergraduate students writing on familiar topics 71% of sentence-initial IKIs were less than 1 sec in duration.
+Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
+For comparison, this is less than mean written naming response latency for single objects [@torrance2018timed] and short sentences when describing arrays of images [@roeser2018advance] in a similar population.
 Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
@@
 The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
+Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction].
+@van1991handwriting, for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now].
 Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
@@
 IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
+When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.
+just by processing below the last buffer in the processing cascade).
 However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
+IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling].
 
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
+For example, @roeser2021modelling demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
 In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
@@
 Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
+For similar approaches see @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.
 
@@
 Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
+However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck.
 Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
 If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
+@ronneberg2022process coined this possibility the "process-disruption hypothesis".
 
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
+In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts.
+On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].
 
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
+Data form spontaneous text production [see @gernsbacher1995coherence] are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
+Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy [@dux2022automating; @dux2021effect; @vandermeulen2023writing].
 
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
+Second, since the emergence of software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing [@conijn2021timecourse].
 Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
+These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning [@conijn2019understanding; @conijn2020keys].
 
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
+Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments [@not10; @dam09; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual].
 In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
@@
 We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
-
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
-
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}

### 2024-04-27 report/sections/methods.Rmd
rearrange markdown sections

--- 2023-09-15 report/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man,floatsintext]{apa7}
-\usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
-  pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
+In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts.
+Full modelling details can be found in Appendix \\ref{statistical-models; we summarise the most important model properties in the following.
+The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data.
+In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.
 
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in @roeser2021modelling.
+Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes.
+These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs.
+This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.
 
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
+We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts.
+We discussed advantages with this approach in the introduction section.
+These datasets include samples from populations with various writing experience and languages (e.g.
+young / L2 writers, students) performing different writing tasks (e.g.
+essays, syntheses).
+Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.
 
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
+## Data sets
 
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
+TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2.
+For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say "unpublished".
+I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk).
+Was the GUNNEXP2 sample Norwegian?
+Should we include information like time limit or text length in the table?
 
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
+Six datasets with keystroke data from free text production were used for analysis.
+An overview can be found in Table \ref{tab:datasets}; descriptions below.
+Dataset used are C2l1 [@ronneberg2022process], CATO [@torrance2016adolescent], GUNNEXP2 [@torranceb], SPL2 [@torrancea], LIFT published in @vandermeulen2020 and described in @vandermeulen2020mapping, PLanTra data are published in @rossetti2022text and described in @rossetti2022s.
+In C2L1 Norwegian 6th graders composed argumentative essays.
+In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text.
+In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced.
+The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each.
+The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability.
+The GUNNEXP2 data ....
+[MARK, please add a sentence here to describe the production context].
 
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
+For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite [@sim07; @torrance201203].
+LIFT and PLanTra data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019] and SPL2 data were collected using CyWrite [@chukharev2019combined].
 
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{Modelling writing hesitations}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
-\usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{threeparttable}
-\usepackage[normalem]{ulem}
-\usepackage[utf8]{inputenc}
-\usepackage{icomma}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\blandscape
 
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
-\date{}
+```{r datasets, results='asis'}
+tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
+                Source = c("Rønneberg et al. (2022)",
+                           "Torrance et al. (2016)",
+                           "Torrance et al. (n.d.)",
+                           "Rossetti and Van Waes (2022b)",
+                           "Vandermeulen, Steendam, et al. (2020)", 
+                           "Torrance and Ofstad (n.d)"),
+                Keylogger = c("EyeWrite", 
+                           "EyeWrite", 
+                           "CyWrite", 
+                           "InputLog", 
+                           "InputLog",
+                           "EyeWrite"),
+                `Writing task` = c("Argumentative essays", 
+                                 "Expository texts",
+                                 "Argumentative essays",
+                                 "Text simplification",
+                                 "Synthesis", 
+                                 "?????"),
+                 N = as.character(c(126, 26*2, 39, 47, 658, 45)),
+#                n_texts = c(1, 2, 2, 2, NA),
+#                n_sentences = c(),
+#                n_words = c(),
+                `Mean age` = c(11.8, 16.9, 20.6, 23, 17.0, "????"),
+                 Sample = c("6th graders", "Secondary school students", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
+                 Country = c("Norway", "Norway", "USA", "Belgium", "The Netherlands", "???? Norway"),
+                 Language = c( "Norwegian", 
+                               "Norwegian", 
+                               "English (L1) / Spanish (L2)", 
+                               "English (L2)", 
+                               "Dutch",
+                               "???? Norwegian")) %>% 
+  arrange(Dataset) %>% 
+  apa_table(align = c("l", "p{3cm}", "p{2cm}", "p{2.5cm}", "r", "r", "p{2.5cm}", "p{2.25cm}", "p{2.5cm}"), 
+            escape = FALSE, 
+            font_size = "footnotesize",
+            caption = "Datasets in brief.")
 
-\authornote{
+#                Manipulation = c("--", 
+#                               "Weak decoders / control; texted masked / unmasked",
+#                               "Text produced in L1 / L2",
+#                               "Pre / post test with training on plain language principles", 
+#                              "Various topics and genres",
+#                               "masked / unmasked")
+```
 
-Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
+\elandscape
 
-}
+## Data extraction
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
+From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful [@chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
+In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
+At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section.
+Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
+Also we removed transitions that were followed by an editing operation.
 
-\abstract{
-Classical serial models view the writing process as a chain of pauses and writing burst.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations.
-Our results cannot be explain by the serial but are in line with the cascading view of writing.
-}
+```{r keyloc, results='asis'}
 
-\begin{document}
-\maketitle
+table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
+       Description = c("Transitions between any letter",
+                       "Keypress after space followed by any letter",
+                       "Keypress following a space preceding any letter"),
+       Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d.
+T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.",
+                   "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
+                   "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept."))
 
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
+apa_table(table, caption = "Transition location classification.",
+            align = c("p{3cm}", "p{5cm}", "p{6cm}"), 
+            escape = FALSE, 
+            font_size = "footnotesize",
+            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace.")
 
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
-Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
+```
 
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
+```{r reductionfunctions}
+source("../scripts/get_data_summary.R")
 
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
+# Get info about data reduction
+n_samples <- 100
+c2l1 <- get_c2l1("../../data/c2l1.csv", n_samples = n_samples)
+cato <- get_cato("../../data/cato.csv", n_samples = n_samples)
+lift <- get_lift("../../data/lift.csv", n_samples = 50, n_ppts = 100)
+plantra <- get_plantra("../../data/plantra.csv", n_samples = n_samples)
+spl2 <- get_spl2("../../data/spl2.csv", n_samples = n_samples) 
+gunnexp2 <- get_gunnexp2("../../data/gunnexp2.csv", n_samples = n_samples)
+```
 
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
+We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data).
+We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
+We further removed keystroke intervals that were too short to represent a behavioural response ($\le$ 50 msecs) or too long ($\ge$ 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}.
+From the remaining data we randomly sampled a maximum of `r n_samples` observations per participant, condition, and transition location (when more than `r n_samples` were available).
+This was done to reduce the computation time of the Bayesian models require to complete sampling.
+For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets.
+Because the LIFT data set included the largest number of writing tasks, we sampled 50 observations per condition, location and participant which would otherwise exceed the computational resources available to us.
+The number of keystroke data used in the analysis can be found in Table \ref{tab:datareduction}.
 
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
-For example, in Roeser et al.
-(2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
+```{r datareduction, results='asis'}
+c2l1$ds <- "c2l1"
+cato$ds <- "cato"
+lift$ds <- "lift"
+plantra$ds <- "plantra"
+spl2$ds <- "spl2"
+gunnexp2$ds <- "gunnexp2"
 
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
-Also Rønneberg et al.
-(2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
-Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
+extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
+          mutate(ds = .x$ds)) %>% 
+  mutate(across(where(is.numeric), ~signif(.*100, 2))) %>% 
+  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
+  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
+  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")")),
+         across(c(`<50ms`, `>30,000ms`), ~str_replace(., "0 \\(0\\)", "--"))) %>% 
+  rename(`$\\le$ 50 msecs` = `<50ms`,
+         `$\\ge$ 30 secs` = `>30,000ms`)
 
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
-Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
+random_sample <- 
+  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
+          mutate(ds = .x$ds)) %>% 
+    mutate(across(where(is.numeric), ~signif(.*100, 2))) %>%
+    unite("keep", starts_with("keep"), sep = " (") %>% 
+    mutate(across(keep, ~str_c(., ")")),
+          across(keep, ~str_replace(., "100 \\(0\\)", "--"))) %>%
+  pivot_wider(names_from = location, values_from = keep) %>% 
+  relocate(`within word`)
 
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
+datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
+  relocate(ds) %>% 
+  mutate(across(ds, ~recode_factor(., 
+                "c2l1" = "C2L1",
+                "cato" = "CATO",
+                "gunnexp2" = "GUNNEXP2",
+                "lift" = "LIFT",
+                "plantra" = "PLanTra",
+#                "spl2_shift" = "SPL2 (shift + C)", 
+                "spl2" = "SPL2",
+                .ordered = T
+                ))) %>% 
+  arrange(ds) %>% 
+  rename(`Dataset` = ds)
 
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
-(2012), Baaijen et al.
-(2012), Chenu et al.
-(2014), Guo et al.
-(2018), Hall et al.
-(2022), Li (2021), and Van Waes et al.
-(2021).
+apa_table(datareduction, 
+          escape = FALSE,  
+          placement = "bp!",
+          caption = "Data reduction.
+Mean percentage of extreme data removed and the mean percentage of randomly sampled data included by transition location (`--` if all data were used).
+Standard error in parentheses.",
+          align =c("l", rep("r", 5)),
+          col_spanners = list(" " = 1,
+                              "Extreme values in \\%" = c(2, 3),
+                              "Randomly sampled data in \\%" = c(4, 6)))
 
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
+```
 
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis.}
+## Bayesian statitical modelling
 
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
+We reanalysed keystroke data -- transition durations measured as the time between first and second keypress -- from six datasets in a series of five Bayesian models.
+An overview of all models can be found in Table \ref{tab:models}.
+More detailed model descriptions and motivation can be found in Appendix \ref{statistical-models}.
 
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
+```{r models, results = 'asis'}
+models <- tibble(Models = str_c("M",1:5),
+       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
+       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
+#       group = c(rep("Serial", 3), rep("Cascading", 2)),
+       Description = c("Single Gaussian model with effects of transition location (and experiment-specific manipulation).",
+                       "Single distribution log-Gaussian model equivialent to M1.",
+                       "Equivalent to M2 but with different variance components for each transition location (unequal variance).",
+                       "Two-distributions mixture of a log-Gaussian for fluent transition intervals and another log-Gaussian that is wider than the previous distribution hesitant / disfluent key transitions; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition location but the distribution of fluent intervals has the same mean across all design cell.
+Mixing proportions capture the relative number of disfluent transitions for the distribution of hesitant transitions for each combination of transition location.",
+                       "Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location.")) %>% 
+  select(Models, Description)
 
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
+apa_table(models, 
+          align = c(rep("l", 1), "p{13cm}"), 
+          escape = FALSE, 
+          digits = 0,
+          caption = "Overview of cognitive models of writing. All models included by-participant random effect.",
+          stub_indents = list(
+                     "\\textbf{Serial}" = 1:3,
+                     "\\textbf{Cascading}" = 4:5))
 
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
+```
 
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
+Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset-specific manipulations.
+In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
+For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; this is because slower human behaviour is generally known to be associated with a larger variability [@wagenmakers2007linear; @wing1973response; @schoner2002timing]; see Appendix \ref{statistical-models} for rationale.
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+Models M4 and M5 are extensions of the unequal-variance single distribution model M3.
+Importantly models M4 and M5 assume that keystroke data result from a combination of two data generating processes -- consistent with the cascading account -- rather than one.
+These model are called finite mixture models in the literature [@gelman2014; @roeser2021modelling; @peel2000finite] and capture the assumption that processing at higher levels of activation leads to longer pauses which can in principle happen at transition before sentence, before words or even mid-word.
+In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that pauses at larger linguistic edges are more likely but not obligated by the cognitive system that generates keystrokes.
+This is achieved by modelling key transitions as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
 
-We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models.
-In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
+1.
+Information from upstream mental processes can flow into keystrokes without interruption at intermediate levels.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger.
+The population estimate for fluent key transitions (i.e.
+typing speed) is captured by the $\beta$ parameter.
 
-The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow.
-Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur.
-This view was implemented as a constrained and an unconstrained model.
+  2.
+Any interruptions at upstream levels of mental representation result in a delay in the information flow and therefore create a lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
+The population estimate of the hesitation slowdown is captured by $\delta$ and its frequency is captured by the mixing weight (proportion) $\theta$.
 
-\hypertarget{statistical-models}{%
-\subsection{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsubsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{unimodal-gaussian}{%
-\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.
-
+  ```{=tex}
   \begin{equation}
   \begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{unimodal-log-gaussian}{%
-\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{unimodal-unequal-variance-log-gaussian}{%
-\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{bimodal-log-gaussian-constrained}{%
-\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
-\item
-  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
+  (\#eq:bimodcon2)
     \text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
@@
 & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 > \theta > 1
+        & 0 < \theta < 1
 \end{aligned}
 \end{equation}
+```
+The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents the distribution of fluent -- uninterrupted -- key transitions.
+Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
+$\theta$ is here parameterised to represent the probability that a key transition is associated with the distribution of hesitant transitions.
+This probability is inversely related to the mixing weight of the distribution of fluent transitions by $1-\theta$ .
+In other words, a larger weight for either distribution inevitably means a lower weight for the other.
+We call this parameter the probability of hesitant transitions.
+When a parameter was assumed to vary by transition location (levels: before sentence, before word, within word) this was indicated as subscript; similar for participants.
+The hesitation probability was allowed to vary by participants constrained by a hyper-parameter for each transition location.
 
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\).
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
+The constrained implementation M4 assume that the size of a fluent key-transition does not vary by transition location.
+In other words, the value that the $\beta$ parameter is taking on is the same for transitions at before-sentence, before-word, and within-word locations.
+However, letter bigrams may be executed systematically faster than transitions between space and either a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
+This is for two reasons: first, and importantly, highly frequent bi- and even trigrams are stored, retrieved and executed as one motor code [@behmer2016crunching; @grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants] because of the strong association between letters.
+This association is reducing the length of transition durations.
+It is less likely that the same association exists for space-letter and space-shift pairs.
+Second, before-sentence transition in two datasets (PLanTra, LIFT) are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transition; we return to this in the results section.
+We therefore implemented an unconstrained mixture model M5 that assumes that the distribution over durations of fluent transitions -- represented as $\beta$ -- may vary across transition locations.
 
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
+All models were implemented in the Bayesian framework [@gelman2014; @mcelreath2016statistical].
+The Bayesian framework is ideal for the estimation of parameter values.
+This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty [@farrell2018computational; @gelman2014; @lee2014bayesian].
+To achieve this, Bayesian models require the inclusion of prior information, i.e.
+existing knowledge about parameter values.
+For small datasets priors influence the inferred parameter values; for larger datasets weakly informative and vague priors are quickly overcome by the data [i.e.
+automatic Ockham's razor, @jefferys1992ockham].
+In other words the choice of priors has less impact on the posterior, certainly for weakly informative priors.
+In the present analysis, we used weakly informative priors to aid model convergence by constraining the parameter space to plausible values [see e.g.
+@lambert2018student; @mcelreath2016statistical].
+Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
 
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
+Stan code for mixture models was based on [@roeser2021modelling; see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); for a tutorial see [https://rpubs.com/jensroes/1000459](rpubs.com/jensroes/1000459).[^1]
 
-\hypertarget{bimodal-log-gaussian-unconstrained}{%
-\paragraph{Bimodal log-Gaussian (unconstrained)}\label{bimodal-log-gaussian-unconstrained}}
-
-In principle, we do not need to assume that the size of a fluent key-transition varies by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2
-\end{aligned}
-\end{equation}
-
-\hypertarget{data-sets}{%
-\subsection{Data sets}\label{data-sets}}
-
-Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
-
-\begin{landscape}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datasets}Datasets in brief.}
-
-\footnotesize{
-
-\begin{tabular}{llllrrp{2cm}p{3cm}}
-\toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
-\midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{landscape}
-
-\hypertarget{c2l1}{%
-\subsubsection{C2L1}\label{c2l1}}
-
-TODO: Mark, can you add information here if needed? I was thinking about adding the aim of the study but it feels unnessary.
-
-TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.
-
-The C2L1 data set comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al.
-(2022).
-The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{cato}{%
-\subsubsection{CATO}\label{cato}}
-
-TODO: Mark, can you add information here?
-
-Data are published in Torrance, Rønneberg, et al.
-(2016).
-Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{gunnexp2}{%
-\subsubsection{GUNNEXP2}\label{gunnexp2}}
-
-TODO: Mark, can you add information here?
-I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk).
-Was the sample Norwegian?
-
-The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX).
-In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.
-Keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-
-\hypertarget{lift}{%
-\subsubsection{LIFT}\label{lift}}
-
-The LIFT data are published in Vandermeulen, Steendam, et al.
-(2020) and described in Vandermeulen, De Maeyer, et al.
-(2020).
-The primary aim of this data set was to create a national baseline on synthesis writing in Dutch secondary education, including student's text quality, writing process, and perspectives on writing.
-Within this national survey, a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education (grades 10, 11, and 12) in the Netherlands was collected from 43 schools.
-The students first received instruction on synthesis writing, after which they were asked to conduct two synthesis tasks, with small breaks in between, thereafter students had a longer lunch break, followed by a survey on writing perspectives and again two synthesis tasks with a small break in between.
-In the four tasks, students were asked to write two argumentative and two informative texts on laptop, about each of four topics (food additives, self-driving cars, the human-wildlife conflict in Africa, and the pay gap), with order randomized per school.
-The students received 50 minutes for each tasks.
-Not all students conducted all four tasks, resulting in a final sample of 2310 synthesis texts.
-During the synthesis tasks keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{plantra}{%
-\subsubsection{PLanTra}\label{plantra}}
-
-The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts.
-A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated.
-The study adopted a pre-test post-test design.
-As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience.
-Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability.
-Participants were asked to spend at least 45 minutes on the instruction module.
-As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability.
-Both reports were written in English (second language) and similar in length (274-278 words) and readability.
-Participants received as much time as needed for each task.
-During the task, keystroke data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).
-
-\hypertarget{spl2}{%
-\subsubsection{SPL2}\label{spl2}}
-
-TODO: Mark, can you add information here?
-
-The SPL2 dataset is published in Torrance et al.
-(n.d.).
-The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019).
-Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-\hypertarget{bayesian-modelling}{%
-\subsubsection{Bayesian modelling}\label{bayesian-modelling}}
-
-We reanalysed keystroke data from 6 datasets in a series of 5 Bayesian models. An overview of all models can be found in Table \ref{tab:models}.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:models}Overview of typing-process models. All models were fitted with random intercepts for participants.}
-
-\begin{tabular}{llrp{9cm}}
-\toprule
-Models & \multicolumn{1}{c}{Type} & \multicolumn{1}{c}{Eq.} & \multicolumn{1}{c}{Description}\\
-\midrule
-M1 & LMM & \ref{eq:unimodgaus} & Gaussian distribution\\
-M2 & LMM & \ref{eq:unimodloggaus} & Log-Gaussian distribution\\
-M3 & LMM & \ref{eq:unimoduv} & Log-Gaussian distribution with unequal variances\\
-M4 & MoG & \ref{eq:bimodcon} & Fluent transition do not vary by transition location\\
-M5 & MoG & \ref{eq:bimoduncon} & Fluent transition vary by transition location\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-Bayesian models, as used in this paper, are ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions of the uncertainty associated with parameter value estimates (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the explicit inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small data sets priors influence the inferred parameter value estimates (known as the posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors values has less impact on the posterior.
-In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space (see e.g.
-Lambert, 2018; McElreath, 2016).
-Also, as the sample size of the reanalysed datasets is large, our weakly informative priors will not affect the posterior in any meaningful way.
-
-For all models we included transition location (levels: before sentence, before word, within word) as predictor; for the parameters associated with transition location see equations \ref{eq:unimodgaus} -- \ref{eq:bimoduncon}.
-Also, we included dataset-specific experimental manipulations as fixed effects for the same model parameters as transition location.
-Stan code for mixture models was based on Roeser et al.
-(2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); also for a tutorial how to fit Bayesian mixture models see ADD URL HERE.
-
-Data were analysed in Bayesian mixed effects models (Gelman et al., 2014; McElreath, 2016).
-The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
+[^1]: The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
 Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic (\(\hat{R}\) = 1) (Gelman \& Rubin, 1992) and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation (Sivula et al., 2020; Vehtari et al., 2015, 2017).
-
-\hypertarget{transition-types}{%
-\subsubsection{Transition types}\label{transition-types}}
-
-The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; e.g.
-Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:keyloc}Transition location classification.}
-
-\footnotesize{
-
-\begin{tabular}{p{3cm}p{3cm}p{8cm}}
-\toprule
-Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
-\midrule
-Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d.
-T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
-Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $'^{\wedge}$' marks transition location; [bsp] represents backspace.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{data-reduction}{%
-\subsubsection{Data reduction}\label{data-reduction}}
-
-For all datasets we only used transitions that were not followed by an editing operation.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed.
-We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-
-We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.
-From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set.
-This was done for computational reasons to reduce the time the Bayesian models need to complete sampling.
-For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool.
-Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources.
-The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.
-
-\begin{table}[bp!]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrrrrr}
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
-Dataset & \multicolumn{1}{c}{$\le$ 50 msecs} & \multicolumn{1}{c}{$\ge$ 30 secs} & \multicolumn{1}{c}{within word} & \multicolumn{1}{c}{before word} & \multicolumn{1}{c}{before sentence}\\
-\midrule
-C2L1 & 0.19 (0.1) & 0.07 (0.06) & 35.1 (2.6) & 84.5 (1.8) & 100 (0)\\
-CATO & 0.65 (0.15) & 0.02 (0.02) & 14.9 (0.9) & 48.6 (2.2) & 100 (0)\\
-GUNNEXP2 & 2.16 (0.17) & 0.01 (0.01) & 6.2 (0.4) & 22.5 (1.4) & 100 (0)\\
-LIFT & 2.65 (0.16) & 0 (0) & 3.2 (0.2) & 13.1 (0.9) & 99.4 (0.1)\\
-PLanTra & 2.49 (0.41) & 0.04 (0.03) & 9.7 (0.6) & 36.6 (1.9) & 100 (0)\\
-SPL2 & 2.29 (0.2) & 0.03 (0.02) & 5.7 (0.4) & 22.6 (1.4) & 100 (0)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{out-of-samples-cross-validation}{%
-\subsection{Out-of-samples cross-validation}\label{out-of-samples-cross-validation}}
-
-To compare the out-of-sample predictive performance of our models we used Pareto smoothed importance-sampling leave-one-out cross-validation (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared via its difference \(\Delta\widehat{elpd}\) between models.
-Similar to other cross-validation techniques, the advantage of using leave-one-out cross-validation is that more complicated models -- models with more parameters -- are penalised to prevent overfit.
-
-Model comparison results for all data sets are shown in Table \ref{tab:loos}.
-For all data sets we found the same pattern.
-Both bi-modal mixture models provided a substantially better fit than any of the uni-modal distribution models.
-The unconstrained version of the mixture model rendered a higher predictive performance than the constrained version which does not allow the distribution of short keystroke-intervals to vary across transition locations (and data set specific manipulations).
-Among the uni-modal models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model.
-The weakest model was the uni-modal Gaussian model.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{TableNotes}
-
-\small{
-
-\begin{longtable}{p{3.5cm}p{6cm}rr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loos}Model comparisons. The top row of each dataset shows the models with the highest predictive performance. Standard error is shown in parentheses.}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Model & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-\endhead
-\textbf{CL21} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -165,434 (231)\\
-\ \ \ M4 & Bimodal (constrained) & -546 (51) & -165,980 (237)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,184 (83) & -167,617 (256)\\
-\ \ \ M2 & Unimodal log-Gaussian & -2,968 (98) & -168,402 (267)\\
-\ \ \ M1 & Unimodal Gaussian & -40,485 (856) & -205,919 (935)\\
-\textbf{CATO} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -139,691 (230)\\
-\ \ \ M4 & Bimodal (constrained) & -607 (43) & -140,298 (230)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,389 (79) & -142,080 (246)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,689 (100) & -143,379 (258)\\
-\ \ \ M1 & Unimodal Gaussian & -41,362 (927) & -181,052 (992)\\
-\textbf{GUNNEXP2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -120,675 (237)\\
-\ \ \ M4 & Bimodal (constrained) & -554 (38) & -121,228 (235)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,981 (94) & -123,656 (258)\\
-\ \ \ M2 & Unimodal log-Gaussian & -5,293 (111) & -125,968 (263)\\
-\ \ \ M1 & Unimodal Gaussian & -44,136 (660) & -164,811 (713)\\
-\textbf{LIFT} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -272,696 (327)\\
-\ \ \ M4 & Bimodal (constrained) & -500 (37) & -273,197 (327)\\
-\ \ \ M3 & Unimodal (unequal variance) & -5,605 (131) & -278,301 (365)\\
-\ \ \ M2 & Unimodal log-Gaussian & -7,716 (147) & -280,412 (375)\\
-\ \ \ M1 & Unimodal Gaussian & -82,567 (1,691) & -355,263 (1,779)\\
-\textbf{PLanTra} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -107,324 (228)\\
-\ \ \ M4 & Bimodal (constrained) & -130 (18) & -107,454 (228)\\
-\ \ \ M3 & Unimodal (unequal variance) & -2,514 (92) & -109,837 (248)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,895 (94) & -111,219 (246)\\
-\ \ \ M1 & Unimodal Gaussian & -37,718 (572) & -145,042 (637)\\
-\textbf{SPL2} &  &  & \\
-\ \ \ M5 & Bimodal (unconstrained) & -- & -96,671 (217)\\
-\ \ \ M4 & Bimodal (constrained) & -564 (41) & -97,234 (216)\\
-\ \ \ M3 & Unimodal (unequal variance) & -1,744 (68) & -98,415 (224)\\
-\ \ \ M2 & Unimodal log-Gaussian & -3,457 (81) & -100,128 (221)\\
-\ \ \ M1 & Unimodal Gaussian & -32,174 (450) & -128,845 (493)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-We also evaluated to what extent model predictions fit observed data.
-These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.
-Model predictions fit the data well in the case of bi-modal mixture models and poorest for the uni-modal Gaussian model.
-
-\hypertarget{cross-data-set-comparisons}{%
-\subsection{Cross-data set comparisons}\label{cross-data-set-comparisons}}
-
-As demonstrated in the model comparisons, the unconstrained bi-modal mixture-model captures the writing-process data better than uni-modal models.
-Conclusions about the the writing process are captured by the model parameters and the posterior parameter value estimates.
-There are three conceptually important parameters: (1) the average duration of fluent transitions (which was indicated as \(\beta\) in equation \ref{eq:bimoduncon}), (2) the magnitude of the slowdown for hesitant transitions (represented as \(\delta\)), and (3) the probability of hesitant transition duration (represented as \(\theta\)).
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Although models were fitted with all data set-specific condition, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift} respectively that there is negligible evidence of differences between these conditions.}, and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-Appendix \ref{masking-effect-cato-gunnexp2} and \ref{l2-effect-spl2} demonstrate that there is evidence for differences in parameter values for these manipulations.}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We evaluated the evidence for differences between transition locations for all data sets: Results are shown in Table \ref{tab:loceffect}.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets.
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (except data set C2L1) compared to within-word transitions (except data set LIFT).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for data sets SPL2 and GUNNEXP2).
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding are the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-spl2}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-Further comparisons can be found in the Appendix: L2 effect (SPL2) in Appendix \ref{l2-effect-spl2}; masking effect (CATO, GUNNEXP2) in Appendix \ref{masking-effect-cato-gunnexp2}; pre-post test effect (PLanTra) in Appendix \ref{pre-post-test-plantra}; genre effect (LIFT) in Appendix \ref{genre-effect-lift}.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-Our cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production.
-We used Bayesian statistical models, in particular uni-modal and bi-modal hierarchical models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of 6 data sets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key-transitions that directly precede sentences or words are not necessarily associated with a pause but writers might plan in parallel to writing.
-On the basis of parameter estimates of the bi-modal mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in 4 of 6 data sets, pauses were as likely to occur before sentences as words, in experienced and novice writers, while mid-word pauses were rare.
-Also, with few exceptions, pauses at before-sentence locations were associated were longer compared to pauses at before-word transitions (which were longer than the few mid-word pauses).
-In other words, when writers pause, the duration of their pause generally suggest that larger linguistic edges are associated with higher-level processing.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because the hesitation probability results suggest that writers often do not pause before sentences; indeed, they frequently plan the next sentence in parallel production resulting in roughly identical hesitation probabilities at before-sentence and before-word locations.
-
-Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning.
-For example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers.
-This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause.
-The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.
-
-We did observe four difference between data sets that are worth highlighting.
-These can largely be explained on the basis of difference in sample population (writing competence / experience) and text genre (see also Conijn et al., 2019):
-
-First, key-transitions of the youngest sample (C2L1) were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-The second and third exceptions concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations (compared to mid-word locations).
-\textbf{TODO: I don't have an explanation for this.} Also, fluent transitions were longer at before-sentence locations compared to before-word locations while in the remaining data sets fluent key-transitions were always shorter at mid-word compared to before-words transitions.
-This shows a general tendency to slow down tying at sentence boundaries for the GUNNEXP2 and SPL2 samples which was pronounced in the other data sets.
-Fourth, pauses were longer before sentence compared to before words compared to mid-word paues with two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT data set although the posterior showed the same general patterns as observed in most data sets.
-The absence of a difference could be explained as specific to the synthesis writing task {[}\textbf{REFERENCE}{]}.
-
-A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models.
-In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model.
-We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process.
-In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process.
-Parameters were not successfully uncovered when we switched model type and data set.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model, when applied to the uni-modal data (but for the bi-modal data).
-Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is a limited amount of work that has used eye tracking to invesitage regressive eye movement behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e.~writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g.~spelling, grammar.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions.
-However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words ({``hopping,''} see also Chukharev-Hudilainen et al., 2019).
-This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-
-A practical advantage of our mixture model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-This pattern was found to be largely consistent different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses) included in our data pool.
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-ailhaud2018variations}{}%
-Ailhaud, E., \& Chenu, F. (2017). Variations of chronometric measures of written production depending on clause packaging. \emph{CogniTextes}, \emph{17}. \url{https://doi.org/10.4000/cognitextes.992}
-
-\leavevmode\hypertarget{ref-ailhaud2016developmental}{}%
-Ailhaud, E., Chenu, F., \& Jisa, H.
-(2016).
-A developmental perspective on the units of written {F}rench.
-In J.
-Perera, M.
-Aparici, E.
-Rosado, \& N.
-Salas (Eds.), \emph{Written and spoken language development across the lifespan: Essays in honour of {Liliana Tolchinsky}} (pp.
-287--305).
-Springer.
-\url{https://doi.org/10.1007/978-3-319-21136-7_17}
-
-\leavevmode\hypertarget{ref-alamargot2010using}{}%
-Alamargot, D., Plane, S., Lambert, E., \& Chesnet, D.
-(2010).
-Using eye and pen movements to trace the development of writing expertise: Case studies of a 7th, 9th and 12th grader, graduate student, and professional writer.
-\emph{Reading and Writing}, \emph{23}(7), 853--888.
-
-\leavevmode\hypertarget{ref-almond2012preliminary}{}%
-Almond, R., Deane, P., Quinlan, T., Wagner, M., \& Sydorenko, T.
-(2012).
-\emph{A preliminary analysis of keystroke log data from a timed writing task} (Research Report No.
-RR-12-23).
-Educational Testing Service.
-
-\leavevmode\hypertarget{ref-alves2015progress}{}%
-Alves, R. A., \& Limpo, T. (2015). Progress in written language bursts, pauses, transcription, and written composition across schooling. \emph{Scientific Studies of Reading}, \emph{19}(5), 374--391.
-
-\leavevmode\hypertarget{ref-baaijen2012keystroke}{}%
-Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
-
-\leavevmode\hypertarget{ref-beers2010adolescent}{}%
-Beers, S.
-F., Quinlan, T., \& Harbaugh, A.
-G.
-(2010).
-Adolescent students' reading during writing behaviors and relationships with text quality: An eyetracking study.
-\emph{Reading and Writing}, \emph{23}(7), 743--775.
-
-\leavevmode\hypertarget{ref-bock2014syntactically}{}%
-Bock, J. K., \& Ferreira, V. S. (2014). In M. Goldrick, V. S. Ferreira, \& M. Miozzo (Eds.), \emph{{The Oxford Handbook of Language Production}} (pp. 21--46). Oxford University Press.
-
-\leavevmode\hypertarget{ref-bonin2012evidence}{}%
-Bonin, P., Roux, S., Barry, C., \& Canell, L.
-(2012).
-Evidence for a limited-cascading account of written word naming.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{38}(6), 1741--1758.
-\url{https://doi.org/10.1037/a0028471}
-
-\leavevmode\hypertarget{ref-carpenter2016stan}{}%
-Carpenter, B., Gelman, A., Hoffman, M., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M.
-A., Guo, J., Li, P., \& Riddell, A.
-(2016).
-Stan: A probabilistic programming language.
-\emph{Journal of Statistical Software}, \emph{20}.
-
-\leavevmode\hypertarget{ref-chang2006}{}%
-Chang, F., Dell, G. S., \& Bock, J. K. (2006). Becoming syntactic. \emph{Psychological Review}, \emph{113}(2), 234--272.
-
-\leavevmode\hypertarget{ref-chanquoy1996writing}{}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
-\leavevmode\hypertarget{ref-chenoweth2001fluency}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\hypertarget{ref-chenoweth2003inner}{}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
-
-\leavevmode\hypertarget{ref-chenu2014interword}{}%
-Chenu, F., Pellegrino, F., Jisa, H., \& Fayol, M. (2014). Interword and intraword pause threshold in writing. \emph{Frontiers in Psychology}, \emph{5}. \url{https://doi.org/10.3389/fpsyg.2014.00182}
-
-\leavevmode\hypertarget{ref-christiansen2016now}{}%
-Christiansen, M.
-H., \& Chater, N.
-(2016).
-The now-or-never bottleneck: A fundamental constraint on language.
-\emph{Behavioral and Brain Sciences}, \emph{39}, e62.
-\url{https://doi.org/10.1017/S0140525X1500031X}
-
-\leavevmode\hypertarget{ref-chukharev2014pauses}{}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
-\leavevmode\hypertarget{ref-chukharev2019combined}{}%
-Chukharev-Hudilainen, E., Saricaoglu, A., Torrance, M., \& Feng, H.-H.
-(2019).
-Combined deployable keystroke logging and eyetracking for investigating {L2} writing fluency.
-\emph{Studies in Second Language Acquisition}, \emph{41}(3), 583--604.
-
-\leavevmode\hypertarget{ref-Clark2002}{}%
-Clark, H. H., \& Fox Tree, J. E. (2002). Using \emph{uh} and \emph{um} in spontaneous speaking. \emph{Cognition}, \emph{84}, 73--111.
-
-\leavevmode\hypertarget{ref-conijn2020keys}{}%
-Conijn, R. (2020). \emph{The keys to writing: A writing analytics approach to studying writing processes using keystroke logging} {[}PhD thesis{]}. Tilburg University.
-
-\leavevmode\hypertarget{ref-conijn2019understanding}{}%
-Conijn, R., Roeser, J., \& van Zaanen, M. (2019). Understanding the keystroke log: The effect of writing task on keystroke features. \emph{Reading and Writing}, \emph{32}(9), 2353--2374.
-
-\leavevmode\hypertarget{ref-conijn2021timecourse}{}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\hypertarget{ref-connelly2012predicting}{}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
-
-\leavevmode\hypertarget{ref-crump2010hierarchical}{}%
-Crump, M.
-J.
-C., \& Logan, G.
-D.
-(2010).
-Hierarchical control and skilled typing: Evidence for word-level control over the execution of individual keystrokes.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{36}(6), 1369--1380.
-\url{https://doi.org/10.1037/a0020696}
-
-\leavevmode\hypertarget{ref-dam09}{}%
-Damian, M.
-F., \& Stadthagen-Gonzalez, H.
-(2009).
-Advance planning of form properties in the written production of single and multiple words.
-\emph{Language and Cognitive Processes}, \emph{24}(4), 555--579.
-
-\leavevmode\hypertarget{ref-de2018exploring}{}%
-De Smet, M. J. R., Leijten, M., \& Van Waes, L. (2018). Exploring the process of reading during writing using eye tracking and keystroke logging. \emph{Written Communication}, \emph{35}(4), 411--447.
-
-\leavevmode\hypertarget{ref-dell2007case}{}%
-Dell, G.
-S., Martin, N., \& Schwartz, M.
-F.
-(2007).
-A case-series test of the interactive two-step model of lexical access: Predicting word repetition from picture naming.
-\emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
-
-\leavevmode\hypertarget{ref-dux2021effect}{}%
-Dux Speltz, E., \& Chukharev-Hudilainen, E. (2021). The effect of automated fluency-focused feedback on text production. \emph{Journal of Writing Research}, \emph{13}(2), 231--255.
-
-\leavevmode\hypertarget{ref-dux2022automating}{}%
-Dux Speltz, E., Roeser, J., \& Chukharev-Hudilainen, E.
-(2022).
-Automating individualized, process-focused writing instruction: A design-based research study.
-\emph{Frontiers in Communication: Emerging Technologies and Writing: Pedagogy and Research}, \emph{7}, 933878.
-\url{https://doi.org/0.3389/fcomm.2022.933878}
-
-\leavevmode\hypertarget{ref-farrell2018computational}{}%
-Farrell, S., \& Lewandowsky, S. (2018). \emph{Computational modeling of cognition and behavior}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-flower1980ynamics}{}%
-Flower, L.
-S., \& Hayes, J.
-R.
-(1980).
-The dynamics of composing: Making plans and juggling constraints.
-In L.
-W.
-Gregg \& E.
-R.
-Steinberg (Eds.), \emph{Cognitive processes in writing: An interdisciplinary approach} (pp.
-31--50).
-Lawrence Erlbaum.
-
-\leavevmode\hypertarget{ref-flower1981pregnant}{}%
-Flower, L. S., \& Hayes, J. R. (1981). The pregnant pause: An inquiry into the nature of planning. \emph{Research in the Teaching of English}, 229--243.
-
-\leavevmode\hypertarget{ref-foulin1998extent}{}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
-
-\leavevmode\hypertarget{ref-gelman2014}{}%
-Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., \& Rubin, D. B. (2014). \emph{Bayesian data analysis} (3rd ed.). Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-gelman1992}{}%
-Gelman, A., \& Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. \emph{Statistical Science}, \emph{7}(4), 457--472.
-
-\leavevmode\hypertarget{ref-gernsbacher1995coherence}{}%
-Gernsbacher, M. A., \& Givón, T. (1995). \emph{Coherence in spontaneous text} (Vol. 31). John Benjamins Publishing.
-
-\leavevmode\hypertarget{ref-gri01}{}%
-Griffin, Z. M. (2001). Gaze durations during speech reflect word selection and phonological encoding. \emph{Cognition}, \emph{82}(1), B1--B14.
-
-\leavevmode\hypertarget{ref-griffin2003reversed}{}%
-Griffin, Z. M. (2003). A reversed word length effect in coordinating the preparation and articulation of words in speaking. \emph{Psychonomic Bulletin \& Review}, \emph{10}(3), 603--609.
-
-\leavevmode\hypertarget{ref-guo2018modeling}{}%
-Guo, H., Deane, P. D., van Rijn, P. W., Zhang, M., \& Bennett, R. E. (2018). Modeling basic writing processes from keystroke logs. \emph{Journal of Educational Measurement}, \emph{55}(2), 194--216.
-
-\leavevmode\hypertarget{ref-hall2022constructing}{}%
-Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
-
-\leavevmode\hypertarget{ref-hayes2012evidence}{}%
-Hayes, J.
-R.
-(2012).
-Evidence from language bursts, revision, and transcription for translation and its relation to other writing processes.
-In M.
-Fayol, D.
-Alamargot, \& V.
-Berninger (Eds.), \emph{Translation of thought to written text while composing: Advancing theory, knowledge, methods, and applications} (pp.
-15--25)).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-jefferys1992ockham}{}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
-
-\leavevmode\hypertarget{ref-kaufer1986composing}{}%
-Kaufer, D. S., Hayes, J. R., \& Flower, L. S. (1986). Composing written sentences. \emph{Research in the Teaching of English}, 121--140.
-
-\leavevmode\hypertarget{ref-lambert2018student}{}%
-Lambert, B. (2018). \emph{A student's guide to {B}ayesian statistics}. Sage.
-
-\leavevmode\hypertarget{ref-lee13}{}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
-\leavevmode\hypertarget{ref-lee2014bayesian}{}%
-Lee, M. D., \& Wagenmakers, E.-J. (2014). \emph{Bayesian cognitive modeling: A practical course}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-leijten2013keystroke}{}%
-Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
-
-\leavevmode\hypertarget{ref-li2021identifying}{}%
-Li, T. (2021). Identifying mixture components from large-scale keystroke log data. \emph{Frontiers in Psychology}, \emph{12}, 628660.
-
-\leavevmode\hypertarget{ref-matsuhashi1981pausing}{}%
-Matsuhashi, A. (1981). Pausing and planning: The tempo of written discourse production. \emph{Research in the Teaching of English}, 113--134.
-
-\leavevmode\hypertarget{ref-mcelreath2016statistical}{}%
-McElreath, R. (2016). \emph{Statistical rethinking: {A} bayesian course with examples in {R} and {Stan}}. CRC Press.
-
-\leavevmode\hypertarget{ref-medimorec2017pauses}{}%
-Medimorec, S., \& Risko, E. F. (2017). Pauses in written composition: On the importance of where writers pause. \emph{Reading and Writing}, \emph{30}, 1267--1285.
-
-\leavevmode\hypertarget{ref-medimorec2017disfluency}{}%
-Medimorec, S., Young, T. P., \& Risko, E. F. (2017). Disfluency effects on lexical selection. \emph{Cognition}, \emph{158}, 28--32.
-
-\leavevmode\hypertarget{ref-meyer1996}{}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-mohsen2021second}{}%
-Mohsen, M.
-(2021).
-Second language learners' pauses over different times intervals in {L2} writing essays: Evidence from a keystroke logging program.
-\emph{Psycholinguistics}, \emph{30}(1), 180--202.
-\url{https://doi.org/10.31470/2309-1797-2021-30-1-180-202}
-
-\leavevmode\hypertarget{ref-not10}{}%
-Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
-
-\leavevmode\hypertarget{ref-not07}{}%
-Nottbusch, G., Weingarten, R., \& Sahel, S.
-(2007).
-From written word to written sentence production.
-In M.
-Torrance, L.
-van Waes, \& D.
-W.
-Galbraith (Eds.), \emph{Writing and cognition: Research and applications} (Vol.
-20, pp.
-31--53).
-Elsevier.
-
-\leavevmode\hypertarget{ref-olive2014toward}{}%
-Olive, T. (2014). Toward a parallel and cascading model of the writing system: {A} review of research on writing processes coordination. \emph{Journal of Writing Research}, \emph{6}(2), 173--194.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing}. R Foundation for Statistical Computing. \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-rochester1973significance}{}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
-
-\leavevmode\hypertarget{ref-roeser2021modelling}{}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
-
-\leavevmode\hypertarget{ref-roeser2018advance}{}%
-Roeser, J., Torrance, M., \& Baguley, T.
-(2019).
-Advance planning in written and spoken sentence production.
-\emph{Journal of Experimental Psychology: Learning, Memory, and Cognition}, \emph{45}(11), 1983--2009.
-\url{https://doi.org/10.1037/xlm0000685}
-
-\leavevmode\hypertarget{ref-rossetti2022s}{}%
-Rossetti, A., \& Van Waes, L.
-(2022a).
-It's not just a phase: Investigating text simplification in a second language from a process and product perspective.
-\emph{Frontiers in Artificial Intelligence}, \emph{5}.
-\url{https://doi.org/10.3389/frai.2022.983008}
-
-\leavevmode\hypertarget{ref-rossetti2022text}{}%
-Rossetti, A., \& Van Waes, L. (2022b). \emph{Text simplification in second language: Process and product data} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.6720290}
-
-\leavevmode\hypertarget{ref-roux2013interaction}{}%
-Roux, S., McKeeff, T.
-J., Grosjacques, G., Afonso, O., \& Kandel, S.
-(2013).
-The interaction between central and peripheral processes in handwriting production.
-\emph{Cognition}, \emph{127}(2), 235--241.
-
-\leavevmode\hypertarget{ref-ronneberg2022process}{}%
-Rønneberg, V., Torrance, M., Uppstad, P.
-H., \& Johansson, C.
-(2022).
-The process-disruption hypothesis: How spelling and typing skill affects written composition process and product.
-\emph{Psychological Research}, \emph{86}(7), 2239--2255.
-
-\leavevmode\hypertarget{ref-schilperoord2002cognitive}{}%
-Schilperoord, J.
-(2002).
-On the cognitive status of pauses in discourse production.
-In T.
-Olive \& C.
-M.
-Levy (Eds.), \emph{Contemporary tools and techniques for studying writing} (pp.
-61--87).
-Springer.
-
-\leavevmode\hypertarget{ref-schoner2002timing}{}%
-Schöner, G. (2002). Timing, clocks, and dynamical systems. \emph{Brain and Cognition}, \emph{48}(1), 31--51.
-
-\leavevmode\hypertarget{ref-sim07}{}%
-Simpson, S., \& Torrance, M. (2007). \emph{EyeWrite}. SR Research; Nottingham Trent University.
-
-\leavevmode\hypertarget{ref-sivula2020uncertainty}{}%
-Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
-
-\leavevmode\hypertarget{ref-rstan}{}%
-Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\hypertarget{ref-sullivan2002self}{}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
-
-\leavevmode\hypertarget{ref-torrance201203}{}%
-Torrance, M.
-(2012).
-EyeWrite -- a tool for recording writers' eye movements.
-In M.
-Torrance, D.
-Alamargot, M.
-Castelló, F.
-Ganier, O.
-Kruse, A.
-Mangen, L.
-Tolchinsky, \& L.
-V.
-Waes (Eds.), \emph{Learning to write effectively: Current trends in {E}uropean research} (pp.
-355--357).
-Brill.
-\url{https://doi.org/doi.org/10.1163/9781780529295_082}
-
-\leavevmode\hypertarget{ref-torrance2016reading}{}%
-Torrance, M., Johansson, R., Johansson, V., \& Wengelin, Å.
-(2016).
-Reading during the composition of multi-sentence texts: An eye-movement study.
-\emph{Psychological Research}, \emph{80}(5), 729--743.
-
-\leavevmode\hypertarget{ref-torrance2012written}{}%
-Torrance, M., \& Nottbusch, G.
-(2012).
-Written production of single words and simple sentences.
-In V.
-W.
-Berninger (Ed.), \emph{Past, present, and future contributions of cognitive writing research to cognitive psychology} (pp.
-403--422).
-Psychology Press.
-
-\leavevmode\hypertarget{ref-torrance2018timed}{}%
-Torrance, M., Nottbusch, G., Alves, R.
-A., Arfé, B., Chanquoy, L., Chukharev-Hudilainen, E., Dimakos, I., Fidalgo, R., Hyönä, J., Jóhannesson, Ó.
-I., \& others.
-(2018).
-Timed written picture naming in 14 {E}uropean languages.
-\emph{Behavior Research Methods}, \emph{50}, 744--758.
-
-\leavevmode\hypertarget{ref-torranceb}{}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
-\leavevmode\hypertarget{ref-torrancea}{}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
-
-\leavevmode\hypertarget{ref-torrance2016adolescent}{}%
-Torrance, M., Rønneberg, V., Johansson, C., \& Uppstad, P.
-H.
-(2016).
-Adolescent weak decoders writing in a shallow orthography: Process and product.
-\emph{Scientific Studies of Reading}, \emph{20}(5), 375--388.
-
-\leavevmode\hypertarget{ref-van1991handwriting}{}%
-Van Galen, G. P. (1991). Handwriting: Issues for a psychomotor theory. \emph{Human Movement Science}, \emph{10}(2), 165--191.
-
-\leavevmode\hypertarget{ref-van2016keystroke}{}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
-\leavevmode\hypertarget{ref-van2019multilingual}{}%
-Van Waes, L., Leijten, M., Pauwaert, T., \& Van Horenbeeck, E.
-(2019).
-A multilingual copy task: Measuring typing and motor skills in writing with inputlog.
-\emph{Journal of Open Research Software}, \emph{7}(30), 1--8.
-
-\leavevmode\hypertarget{ref-van2010reading}{}%
-Van Waes, L., Leijten, M., \& Quinlan, T.
-(2010).
-Reading during sentence composing and error correction: A multilevel analysis of the influences of task complexity.
-\emph{Reading and Writing}, \emph{23}(7), 803--834.
-
-\leavevmode\hypertarget{ref-waes2019}{}%
-Van Waes, L., Leijten, M., Roeser, J., Olive, T., \& Grabowski, J.
-(2021).
-Measuring and assessing typing skills in writing research.
-\emph{Journal of Writing Research}, \emph{13}(1), 107--153.
-\url{https://doi.org/10.17239/jowr-2021.13.01.04}
-
-\leavevmode\hypertarget{ref-vandermeulen2020mapping}{}%
-Vandermeulen, N., De Maeyer, S., Van Steendam, E., Lesterhuis, M., Van den Bergh, H., \& Rijlaarsdam, G.
-(2020).
-Mapping synthesis writing in various levels of {Dutch} upper-secondary education: A national baseline study on text quality, writing process and students' perspectives on writing.
-\emph{Pedagogische Studiën: Tijdschrift Voor Onderwijskunde En Opvoedkunde}, \emph{97}(3), 187--236.
-
-\leavevmode\hypertarget{ref-vandermeulen2020}{}%
-Vandermeulen, N., Steendam, E. V., \& Rijlaarsdam, G. (2020). \emph{{DATASET} -- baseline data {LIFT} synthesis writing project} {[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3893538}
-
-\leavevmode\hypertarget{ref-vandermeulen2023writing}{}%
-Vandermeulen, N., Van Steendam, E., De Maeyer, S., \& Rijlaarsdam, G.
-(2023).
-Writing process feedback based on keystroke logging and comparison with exemplars: Effects on the quality and process of synthesis texts.
-\emph{Written Communication}, \emph{40}(1), 90--144.
-
-\leavevmode\hypertarget{ref-vasishth2017}{}%
-Vasishth, S., Chopin, N., Ryder, R., \& Nicenboim, B.
-(2017).
-Modelling dependency completion in sentence comprehension as a {B}ayesian hierarchical mixture process: {A} case study involving {C}hinese relative clauses.
-\emph{ArXiv e-Prints}.
-
-\leavevmode\hypertarget{ref-vasishth2017feature}{}%
-Vasishth, S., Jäger, L. A., \& Nicenboim, B. (2017). Feature overwriting as a finite mixture process: Evidence from comprehension data. \emph{arXiv Preprint arXiv:1703.04081}.
-
-\leavevmode\hypertarget{ref-vehtari2015pareto}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2015). Pareto smoothed importance sampling. \emph{arXiv Preprint arXiv:1507.02646}.
-
-\leavevmode\hypertarget{ref-vehtari2017practical}{}%
-Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model evaluation using leave-one-out cross-validation and {WAIC}. \emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
-
-\leavevmode\hypertarget{ref-wagenmakers2007linear}{}%
-Wagenmakers, E.-J., \& Brown, S.
-(2007).
-On the linear relation between the mean and the standard deviation of a response time distribution.
-\emph{Psychological Review}, \emph{114}(3), 830--841.
-\url{https://doi.org/10.1037/0033-295X.114.3.830}
-
-\leavevmode\hypertarget{ref-wen02}{}%
-Wengelin, Å. (2002). \emph{Text production in adults with reading and writing difficulties} {[}PhD thesis{]}. Göteborg University.
-
-\leavevmode\hypertarget{ref-wen06}{}%
-Wengelin, Å.
-(2006).
-Examining pauses in writing: {T}heory, methods and empirical data.
-In K.
-P.
-H.
-Sullivan \& E.
-Lindgren (Eds.), \emph{Computer keystroke logging and writing: Methods and applications} (Vol.
-18, pp.
-107--130).
-Elsevier.
-
-\leavevmode\hypertarget{ref-wheeldon_konopka_2023}{}%
-Wheeldon, L. R., \& Konopka, A. (2023). \emph{Grammatical encoding for speech production}. Cambridge University Press. \url{https://doi.org/10.1017/9781009264518}
-
-\leavevmode\hypertarget{ref-wheeldon2013}{}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
-\leavevmode\hypertarget{ref-wing1973response}{}%
-Wing, A. M., \& Kristofferson, A. B. (1973). Response delays and the timing of discrete motor responses. \emph{Perception \& Psychophysics}, \emph{14}(1), 5--12.
-
-\end{CSLReferences}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{fit-to-data}{%
-\subsection{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained
-from the model fit the observed data. From each model we obtained 100
-simulations (predictions). These can be found in Figure
-\ref{fig:prediction}. The model predictions fitted the data
-substantially better for the mixture models. The worse fit was observed
-for the unimodal Gaussian model which predicted large amounts of
-negative keystroke intervals which are implausible. Using a log-normal
-distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data. Comparison of 100 simulated (predicted) sets of data to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{posterior-parameter-estimates}{%
-\subsection{Posterior parameter
-estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1.pdf}
-\caption{Posterior parameter distribution}
-\end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
-\end{figure}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{key-combination-effect-spl2}{%
-\subsection{Key-combination effect
-(SPL2)}\label{key-combination-effect-spl2}}
-
-The analysed datasets differ to the extent that keystroke intervals at
-before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
-C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key
-intervals in the PLanTra and LIFT data, namely
-\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
-namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
-longer or more frequent pauses at before-sentence locations compared to
-before-word locations can be explained without reference to linguistic
-edges. Also there is a possibility that some inconsistencies in our
-findings can be explained on the basis of including the keystroke
-following shift.
-
-Therefore we compared whether the different patterns can be explain on
-the basis of the additional keystroked involved in before-sentence
-transitions. We compared the SPL2 data including and excluding the
-keystroke after shift. Although we modelled all transition locations, we
-present only before-sentence transitions below as there was, as one
-would expect, no difference at word locations. The results of this
-comparison can be found in Table \ref{tab:shiftcellmeans}. Overall,
-fluent transition duration and the hesitation duration were affected by
-whether or not the sentence-initial transition include the character
-following shift. Fluent key transitions were substantially longer when
-including the interval following the shift key. The slowdown for
-hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation
-probability.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ L1 & 390 [350, 434] & 240 [216, 266] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 448 [379, 521] & 296 [253, 343] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 2,398 [2,001, 2,836] & 2,469 [2,119, 2,855] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 2,859 [2,407, 3,368] & 2,769 [2,348, 3,236] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & .62 [.53, .71] & .50 [.41, .59] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .81 [.73, .88] & .72 [.63, .80] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{l2-effect-spl2}{%
-\subsection{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}}
-sentence-transitions) we calculated the L2 effect (i.e.~the difference
-between writing in L2 and L1). The results can be found in Table
-\ref{tab:l2effect}. The results show longer keystrokes and more pauses
-across all transition locations. However, the pause duration only
-differed at before-word locations, not at within-word or before-sentence
-transition locations.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown for transitions for writing in L1 and L2, the slowdown for long transitions and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 240 [216, 266] & 296 [253, 343] & 0.21 [0.08, 0.33] & 11.81\\
-\ \ \ before word & 188 [173, 205] & 259 [236, 284] & 0.32 [0.28, 0.36] & > 100\\
-\ \ \ within word & 138 [127, 150] & 156 [143, 169] & 0.12 [0.1, 0.14] & > 100\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 2,469 [2,119, 2,855] & 2,769 [2,348, 3,236] & -0.08 [-0.24, 0.07] & 0.14\\
-\ \ \ before word & 343 [289, 404] & 759 [667, 862] & 0.33 [0.22, 0.44] & > 100\\
-\ \ \ within word & 138 [93, 196] & 171 [132, 217] & 0.05 [-0.16, 0.26] & 0.12\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .50 [.41, .59] & .72 [.63, .80] & 0.97 [0.44, 1.5] & > 100\\
-\ \ \ before word & .34 [.26, .42] & .59 [.51, .68] & 1.06 [0.58, 1.54] & > 100\\
-\ \ \ within word & .07 [.05, .10] & .18 [.13, .24] & 1.08 [0.56, 1.63] & > 100\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\subsection{Masking effect (CATO,
-GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to
-what extent masking the previously written text affects keystroke
-behaviour. The mixture model results for the effect of masking is shown
-in Table \ref{tab:maskingeffect}. Key transition were longer for masked
-text than for unmasked text at before and within word transitions but
-not before sentences. Also dyslexics showed no masking difference at
-before word transitions. There was no indication of longer or more
-frequent pauses as a result of masked text.
-
-\blandscape
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 282 [240, 329] & 285 [228, 354] & 0.01 [-0.23, 0.25] & 0.12\\
-\ \ \ before sentence & CATO & non dyslexic & 207 [176, 243] & 263 [230, 300] & 0.24 [0.07, 0.41] & 4.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 220 [195, 248] & 250 [224, 279] & 0.13 [0.02, 0.23] & 1.01\\
-\ \ \ before word & CATO & dyslexic & 259 [238, 281] & 267 [246, 289] & 0.03 [-0.01, 0.07] & 0.07\\
-\ \ \ before word & CATO & non dyslexic & 214 [197, 231] & 229 [211, 247] & 0.07 [0.03, 0.1] & 13.25\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 177 [165, 189] & 185 [173, 198] & 0.05 [0.03, 0.07] & 70.05\\
-\ \ \ within word & CATO & dyslexic & 188 [173, 202] & 198 [183, 214] & 0.06 [0.03, 0.08] & 13.22\\
-\ \ \ within word & CATO & non dyslexic & 151 [140, 163] & 163 [151, 176] & 0.08 [0.05, 0.1] & > 100\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 140 [131, 150] & 146 [137, 156] & 0.04 [0.03, 0.06] & > 100\\
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,986 [2,070, 4,118] & 2,161 [1,515, 2,987] & -0.3 [-0.72, 0.13] & 0.57\\
-\ \ \ before sentence & CATO & non dyslexic & 1,767 [1,244, 2,415] & 1,954 [1,417, 2,612] & -0.12 [-0.5, 0.26] & 0.24\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,602 [1,336, 1,908] & 2,310 [1,968, 2,703] & 0.21 [0.05, 0.38] & 1.97\\
-\ \ \ before word & CATO & dyslexic & 536 [469, 610] & 531 [462, 608] & -0.03 [-0.13, 0.07] & 0.06\\
-\ \ \ before word & CATO & non dyslexic & 402 [344, 468] & 452 [377, 538] & 0.03 [-0.09, 0.16] & 0.07\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 401 [355, 451] & 472 [415, 535] & 0.08 [-0.02, 0.18] & 0.18\\
-\ \ \ within word & CATO & dyslexic & 188 [140, 245] & 153 [113, 201] & -0.12 [-0.3, 0.05] & 0.23\\
-\ \ \ within word & CATO & non dyslexic & 157 [107, 222] & 168 [107, 250] & -0.01 [-0.27, 0.27] & 0.14\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 173 [129, 227] & 172 [124, 230] & -0.03 [-0.24, 0.19] & 0.11\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .50 [.33, .67] & .66 [.47, .82] & 0.16 [-0.08, 0.4] & 0.3\\
-\ \ \ before sentence & CATO & non dyslexic & .51 [.34, .69] & .52 [.34, .69] & 0 [-0.24, 0.25] & 0.12\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .69 [.61, .77] & .69 [.62, .76] & 0 [-0.45, 0.45] & 0.23\\
-\ \ \ before word & CATO & dyslexic & .55 [.44, .66] & .45 [.35, .56] & -0.1 [-0.25, 0.05] & 0.19\\
-\ \ \ before word & CATO & non dyslexic & .41 [.31, .52] & .31 [.22, .40] & -0.1 [-0.24, 0.03] & 0.23\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .38 [.32, .44] & .32 [.27, .38] & -0.24 [-0.57, 0.09] & 0.46\\
-\ \ \ within word & CATO & dyslexic & .15 [.09, .21] & .17 [.11, .25] & 0.03 [-0.06, 0.12] & 0.05\\
-\ \ \ within word & CATO & non dyslexic & .10 [.06, .15] & .08 [.05, .13] & -0.02 [-0.08, 0.04] & 0.04\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .08 [.06, .10] & .06 [.05, .09] & -0.22 [-0.65, 0.21] & 0.36\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\elandscape
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{pre-post-test-plantra}{%
-\subsection{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table
-\ref{tab:retesteffect}. Transition durations were shorter at before-word
-locations for the post-test. Evidence was negligible for any other
-comparisons.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 170 [156, 185] & 175 [161, 189] & -0.03 [-0.11, 0.05] & 0.05\\
-\ \ \ before word & 176 [166, 186] & 166 [156, 176] & 0.06 [0.03, 0.09] & 21.33\\
-\ \ \ within word & 145 [137, 154] & 146 [138, 154] & 0 [-0.03, 0.02] & 0.01\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,029 [827, 1,260] & 1,107 [877, 1,364] & -0.04 [-0.25, 0.17] & 0.12\\
-\ \ \ before word & 480 [426, 540] & 463 [410, 520] & -0.01 [-0.12, 0.09] & 0.05\\
-\ \ \ within word & 224 [174, 284] & 242 [188, 306] & -0.05 [-0.25, 0.16] & 0.11\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .52 [.44, .60] & .48 [.40, .55] & 0.17 [-0.23, 0.58] & 0.29\\
-\ \ \ before word & .48 [.43, .54] & .45 [.39, .50] & 0.15 [-0.16, 0.45] & 0.24\\
-\ \ \ within word & .13 [.10, .16] & .12 [.09, .14] & 0.13 [-0.24, 0.49] & 0.24\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{genre-effect-lift}{%
-\subsection{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when
-writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}. Cellmeans and
-differences were average across writing topic. Evidence for a difference
-between genres was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ before sentence & 164 [155, 173] & 166 [158, 175] & -0.01 [-0.15, 0.1] & 0.1\\
-\ \ \ before word & 174 [166, 181] & 171 [164, 179] & 0.01 [-0.05, 0.09] & 0.03\\
-\ \ \ within word & 140 [134, 145] & 140 [135, 146] & 0 [-0.06, 0.05] & 0.03\\
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 495 [379, 631] & 558 [431, 707] & -0.08 [-0.41, 0.25] & 0.2\\
-\ \ \ before word & 338 [285, 399] & 327 [277, 384] & 0.01 [-0.19, 0.23] & 0.11\\
-\ \ \ within word & 165 [113, 230] & 188 [131, 258] & -0.07 [-0.42, 0.26] & 0.18\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .26 [.20, .33] & .25 [.19, .32] & 0.04 [-0.5, 0.61] & 0.29\\
-\ \ \ before word & .35 [.29, .41] & .37 [.31, .43] & -0.1 [-0.53, 0.35] & 0.25\\
-\ \ \ within word & .10 [.08, .14] & .11 [.08, .14] & -0.03 [-0.62, 0.56] & 0.32\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{appendix}
-
-\clearpage
-\makeatletter
-\efloat@restorefloats
-\makeatother
-
-\begin{appendix}
-\section{}
-\hypertarget{simulation}{%
-\subsection{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive
-performance for bi-modal mixture models -- is that, in principle, as the
-mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques
-for model comparison which is preventing overfitting models. To address
-this concern we repeated a comparison for a uni-modal model and a
-bi-modal mixture model for two sets of simulated data that were
-simulated either with a uni-modal or bi-modal process at heart. In other
-words, this allows us to test the predictive performance of our models
-in a context where we know the true underlying data generating process
--- uni-modal vs bi-modal -- and we can test whether these models can
-successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a bi-modal mixture
-model with two mixture components similar to the process described above
-(equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process
-and the corresponding Bayesian model is summarised in equation
-\ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its
-main parameters (but not mixed effects for participants). The model
-includes two log-normal distributions with a mixing proportion
-\(\theta\) of which the distribution of shorter values has a mean of
-\(\beta\) and a standard deviation \(\sigma^2_2\); the second
-distribution of longer values is constrained to have a mean that is
-larger by a factor of \(\delta\) and has a larger standard deviation
-\(\sigma^2_1\).
-
-The second data set was generated with a uni-modal log-Gaussian process.
-The model and its corresponding Bayesian model is summarised in equation
-\ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the uni-modal models used
-in the main analysis. The model assumes a log-Gaussian distribution with
-a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be
-found in Table \ref{tab:simparam}. The simulated data are visualised in
-Figure \ref{fig:simdata}. Parameter values were chosen so that the
-simulated data are roughly similarly distributed to keystroke
-transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations. We run
-2 models -- a bi-modal mixture model and a uni-modal model -- each for
-both data sets. Models were run with 3 chains, with each 6,000
-iterations of which 3,000 were warmup. Estimates with 95\% probability
-intervals are shown in Table \ref{tab:simparam}. The parameters are
-shown by type of data generating process along with the true parameter
-values. Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully
-uncovered the model parameters of the data with its corresponding data
-generating process, as shown in Table \ref{tab:simparam}, but less so
-when the model was applied to data generated with the incorrect
-underlying process. Particularly the mixing proportion \(\theta\) and
-the slowdown parameter \(\delta\) were not uncovered at all by they
-uni-modal model.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
-\toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
-\midrule
-Bi-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Uni-modal data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models
-for each data generating process. The model comparisons can be found for
-each data generating process in Table \ref{tab:loossim}. The results
-rule out the possibility that the mixture model does always lead to
-higher predictive performance. Indeed, the mixture model showed a
-slightly lower predictive performance for the data that were generated
-with a uni-modal process. However, for the data generated with a
-bi-modal process, the mixture model shows a substantially higher
-predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
-and its standard error, as metric for the strength of evidence (Sivula
-et al., 2020), shows that the mixture model performs 11.6 times better
-than the uni-modal model. In comparison, for the uni-modal data, the
-uni-modal model performs only 0.77 times better than the bi-modal
-mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
-the model is negligible, the uni-modal is preferred by the law of
-parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Bi-modal mixture process &  & \\
-\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
-\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
-Data: Uni-modal process &  & \\
-\ \ \ Uni-modal model & -- & -5,030 (24)\\
-\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
-\bottomrule
-\addlinespace
-\end{tabular}
-
-\begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
-\end{tablenotes}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-\end{appendix}
-
-\end{document}
+Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
+The predictive performance of our models was compared using leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical; @sivula2020uncertainty].

### 2024-04-27 report/sections/results.Rmd
rearrange markdown sections

## Model comparisons

To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical]. Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared by the difference between models $\Delta\widehat{elpd}$. We also summarised this difference as normalised over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty]. Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.

```{r}
files <- list.files(
  str_c("../../stanout/", 
        c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
  pattern = "model.+.csv", 
  full.names = T)

table <- purrr::map_dfr(files, ~read_csv(.) %>% 
                          mutate(dataset = .x)) %>% 
  select(dataset, model, contains("elpd_diff"), se_diff) %>%
  arrange(dataset, desc(elpd_diff)) %>% 
  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(where(is.numeric), ~round(., 0)),
         across(where(is.numeric), ~format(., big.mark = ",")),
         across(everything(), ~str_trim(.))) %>%
  filter(dataset != "spl2_shift") %>% 
  transmute(
    Comparison = model,
    elpd_diff_ratio,
    across(dataset, ~recode(., 
                            "cato" = "CATO",
                            "spl2" = "SPL2", 
                            #                "spl2_shift" = "SPL2 (shift + C)", 
                            "plantra" = "PLanTra",
                            "lift" = "LIFT",
                            "c2l1" = "C2L1",
                            "gunnexp2" = "GUNNEXP2")),
    Distribution = case_when(str_detect(model, "M2$") ~ "Single log-Gaussian (M2)",
                             str_detect(model, "M3$") ~ "Single log-Gaussian (unequal variance; M3)",
                             str_detect(model, "M4$") ~ "Mixture of two log-Gaussians (constrained; M4)",
                             str_detect(model, "M5$") ~ "Mixture of two log-Gaussians (unconstrained; M5)"),
    across(elpd_diff, ~str_c(., " (", se_diff, ")")))

table_ratios <- table %>% select(dataset, Comparison, elpd_diff_ratio)

table <- table %>% select(-elpd_diff_ratio) %>% 
  pivot_wider(names_from = dataset, values_from = elpd_diff)

table_ratios <- table_ratios %>% 
  mutate(across(elpd_diff_ratio, as.numeric)) %>% 
  summarise(across(elpd_diff_ratio, list(min = min, max = max), .names = "{.fn}"), .by = Comparison) %>% 
  mutate(range = str_c("[", min, ", ", max, "]"))

tr_m2 <- filter(table_ratios, str_detect(Comparison, "M2$")) %>% pull(range)
tr_m3 <- filter(table_ratios, str_detect(Comparison, "M3$")) %>% pull(range)
tr_m4 <- filter(table_ratios, str_detect(Comparison, "M4$")) %>% pull(range)
tr_m5 <- filter(table_ratios, str_detect(Comparison, "M5$")) %>% pull(range)

```

Model differences are shown in Table \ref{tab:loos}. All datasets revealed the same pattern. Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single-distribution model; the difference $\Delta\widehat{elpd}$ between models ranged within `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model. The unconstrained mixture model M5 rendered a higher predictive performance than the constrained model M4 which did not allow the distribution of fluent keystroke-intervals to vary by transition location; the difference in predictive performance was between `r tr_m5` standard errors higher for the unconstrained model. Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; the difference $\Delta\widehat{elpd}$ ranged within `r tr_m3` standard errors. The weakest model was the single distribution Gaussian which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of `r tr_m2` standard errors.

\blandscape

```{r loos, results='asis'}
#names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
#names(table) <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$"), 1))

table %>% 
  arrange(desc(Comparison)) %>% 
  apa_table(caption = "Model comparisons. Models were compared incrementally from the simplest to the most complex model. The comparison is shown along with the distribution type of the more complex model. A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model.  Standard error is shown in parentheses.",
            align =c("p{1.5cm}", "p{5cm}", rep("r", 6)), 
            escape = FALSE,
            longtable = T,
            font_size = "footnotesize",
            note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density")
#                col_spanners = list(" " = 1:2,
#                     "\\textbf{CL21}" = 1:5,
#                     "\\textbf{CATO}" = 6:10,
#                     "\\textbf{GUNNEXP2}" = 11:15, 
#                     "\\textbf{LIFT}" = 16:20,
#                     "\\textbf{PLanTra}" = 21:25,
#                     "\\textbf{SPL2}" = 26:30))
#      stub_indents = list(
#                     "\\textbf{CL21}" = 1:5,
#                     "\\textbf{CATO}" = 6:10,
#                     "\\textbf{GUNNEXP2}" = 11:15, 
#                     "\\textbf{LIFT}" = 16:20,
#                    "\\textbf{PLanTra}" = 21:25,
#                     "\\textbf{SPL2}" = 26:30))
#                     "\\textbf{SPL2 (shift + C)}" = 31:35))

```

\elandscape

We also tested to what extent model predictions fit observed data. Comparisons can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}. Data predicted by the two-distribution mixture models are similar to observed data; data predicted by single distribution Gaussian models showed a poor fit to the data.

## Transition location effect

As demonstrated in the model comparisons, the two-distributions mixture-models is a better statistical representation on the underlying process generating keystroke data than compared to any of the single-distribution models. Information about this process is captured by the posterior estimates of the models' parameter values. To reiterate, there are three conceptually important parameters: (1) the magnitude of the slowdown for transition hesitations (represented as $\delta$ in equation \ref{eq:bimodcon2}), (2) the probability of hesitant transitions (represented as $\theta$ in equation \ref{eq:bimodcon2}), and (3) the duration of fluent transitions (the typing speed, which was indicated as $\beta$ in equation \ref{eq:bimodcon2}).

Figure \ref{fig:mixmodel} illustrates the mixture model approach on the basis of the posterior of the constrained implementation applied to the CATO data: for each transition durations we obtain two posterior probability distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in hesitant transitions. The hesitation probability is indicated as the relative height of either distribution. The transition locations are characterized by the hesitation parameters: as we illustrate the constrained mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location but their height differs as the probability of fluent transition is the inverse of the height of hesitant transitions; i.e. $1 - \theta$. More interesting are the distributions associated with hesitations; indicated in yellow: Within word transitions have a negligibly small hesitation probability barely visible in the visualisation. Hesitations before words and sentences are both roughly equally likely than fluent transitions. In other words, half of the time participants did not pause before a word or sentence; we explore this below. However when hesitations occurred, these are longer at before-sentence transitions compared to before word transitions.

```{r mixmodel, fig.cap='Mixture model plot for each transition location. Shown are the distributions for both mixture components (fluent and hesitant transitions) and their weighting. Posterior of CATO data.', fig.height=6}

ps <- read_csv("../../stanout/cato/mog_constr_posterior_for_densityplot.csv") 
# Extract parameters
alpha <- filter(ps, param == 'beta') %>% pull(est)
alpha2_bs <- filter(ps, param == 'delta', loc == 'before sentence') %>% pull(est)
alpha2_bw <- filter(ps, param == 'delta', loc == 'before word') %>% pull(est)
alpha2_ww <- filter(ps, param == 'delta', loc == 'within word') %>% pull(est)

theta_bs <- filter(ps, param == 'prob',  loc == 'before sentence') %>% pull(est)
theta_bw <- filter(ps, param == 'prob',  loc == 'before word') %>% pull(est)
theta_ww <- filter(ps, param == 'prob',  loc == 'within word') %>% pull(est)

sigma <- filter(ps, param == 'sigma') %>% pull(est)
sigmap_bs <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'before sentence')) %>% pull(est) %>% sum()
sigmap_bw <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'before word')) %>% pull(est) %>% sum()
sigmap_ww <- filter(ps, str_detect(param, 'sigma'), loc %in% c('overall', 'within word')) %>% pull(est) %>% sum()

# Data grid
data <- tibble(x = c(0, 20))

colours <- c("Fluent" = "#000000", "Hesitant" = "#E69F00")

# Create density plot
plot_bs <- ggplot(data, aes(x)) +
  stat_function(aes(colour = "Fluent"),
                geom = "line", 
                linewidth = .25,
                fun = plot_mix_comps,
                args = list(alpha, sigma, lam = 1 - theta_bs)) +
  stat_function(aes(colour = "Hesitant"),
                geom = "line", 
                linewidth = .25,
                fun = plot_mix_comps, 
                args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
  stat_function(aes(fill = "Fluent"),
                geom = "area", 
                fun = plot_mix_comps, 
                alpha = .25, 
                args = list(alpha, sigma, lam = 1 - theta_bs)) +
  stat_function(aes(fill = "Hesitant"),
                geom = "area", 
                fun = plot_mix_comps, 
                alpha = .25, 
                args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
  scale_x_continuous(labels = exp, 
                     breaks = log(c(1, 10, 100, 1000, 10000)), 
                     limits = log(c(10, 50000))) +
  scale_colour_manual(values = colours) +
  scale_fill_manual(values = colours) +
  labs(y = "", x = "", 
       subtitle = "Before-sentence transitions",
       colour = "Mixture\ncomponent",
       fill = "Mixture\ncomponent") +
  theme(legend.position = "right",
        legend.justification = "top")

plot_bw <- ggplot(data, aes(x)) +
  stat_function(aes(colour = "Fluent"),
                linewidth = .25,
                geom = "line", 
                fun = plot_mix_comps,
                args = list(alpha, sigma, lam = 1 - theta_bw)) +
  stat_function(aes(colour = "Hesitant"),
                geom = "line", 
                linewidth = .25,
                fun = plot_mix_comps, 
                args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
  stat_function(aes(fill = "Fluent"),
                geom = "area", 
                linewidth = .25,
                fun = plot_mix_comps, 
                alpha = .25, 
                args = list(alpha, sigma, lam = 1 - theta_bw)) +
  stat_function(aes(fill = "Hesitant"),
                geom = "area", 
                fun = plot_mix_comps, 
                alpha = .25, 
                args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
  scale_x_continuous(labels = exp, 
                     breaks = log(c(1, 10, 100, 1000, 10000)), 
                     limits = log(c(10, 50000))) +
  scale_colour_manual(values = colours) +
  scale_fill_manual(values = colours) +
  labs(y = "Density", x = "", 
       subtitle = "Before-word transitions",
       colour = "Mixture\ncomponent",
       fill = "Mixture\ncomponent") +
  theme(legend.position = "none")

plot_ww <- ggplot(data, aes(x)) +
  stat_function(aes(colour = "Fluent"),
                geom = "line", 
                fun = plot_mix_comps,
                linewidth = .25,
                args = list(alpha, sigma, lam = 1 - theta_ww)) +
  stat_function(aes(colour = "Hesitant"),
                geom = "line", 
                fun = plot_mix_comps, 
                linewidth = .25,
                args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
  stat_function(aes(fill = "Fluent"),
                geom = "area", 
                fun = plot_mix_comps, 
                alpha = .25, 
                args = list(alpha, sigma, lam = 1 - theta_ww)) +
  stat_function(aes(fill = "Hesitant"),
                geom = "area", 
                fun = plot_mix_comps, 
                alpha = .25, 
                args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
  scale_x_continuous(labels = exp, 
                     breaks = log(c(1, 10, 100, 1000, 10000)), 
                     limits = log(c(10, 50000))) +
  scale_colour_manual(values = colours) +
  scale_fill_manual(values = colours) +
  labs(subtitle = "Within-word transitions", 
       y = "", x = "Transition duration [in msecs]",
       colour = "Mixture\ncomponent",
       fill = "Mixture\ncomponent") +
  theme(legend.position = "none")

plot_bs / plot_bw / plot_ww

```

We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}. We demonstrated above that the predictive performance of the unconstrained mixture model was substantially higher than the predictive performance of the constrained mixture model. However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected. We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following.

```{r}
files <- list.files(str_c("../../stanout/", 
                          c("lift", "spl2", "plantra", "cato", "c2l1", "gunnexp2")), 
                    pattern = "mogbetaco.+.csv", 
                    full.names = T)

ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
  mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, ~recode_factor(., 
                "cato" = "CATO",# (non-dyslexic, unmasked)",
                "gunnexp2" = "GUNNEXP2",# (unmasked)",
                "spl2" = "SPL2", # (L1)", 
#                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1",
                .ordered = TRUE))) %>% 
  filter(!(str_detect(data, "SPL2") & lang == "ES" & !is.na(lang)),
         !(str_detect(data, "CATO") & group == "dyslexic" & !is.na(group)),
         !(str_detect(data, "CATO") & task == "masked" & !is.na(task)),
         !(str_detect(data, "GUNNEXP2") & xn == "masked" & !is.na(xn)))

ps_beta <- filter(ps, param == "beta") %>% 
  rename(beta = value) %>%
  mutate(idx = 1:n(), .by = data) %>% 
  select(idx, beta, data)

ps <- filter(ps, param != "beta")  %>% 
  pivot_wider(names_from = param, 
              values_from = value) %>% 
  unnest(cols = beta2:theta) %>%
  mutate(idx = 1:n(), .by = c(data, location, task, xn, genre, topic, lang)) %>% 
#  summarise(across(idx, list(max = max)), .by = c(location, data, task))
  left_join(ps_beta, by = c("data", "idx")) %>%
  mutate(across(c(beta, beta2), ~exp(.)),
         delta = beta2 - beta,
         across(location, ~str_replace(., " ", "\n"))) %>% 
  pivot_longer(c(beta, beta2, delta, prob)) %>% 
  summarise(across(value, 
                   list(mean = mean, 
                        lower = lower, 
                        upper = upper),
                   .names = "{.fn}"), 
            .by = c(data, location, name)) %>% 
  filter(name %in% c("beta", "beta2", "delta", "prob"))
```

The mixture-model posterior parameter estimates for all datasets is visualised in Figure \ref{fig:crossstudypost2}. Although models were fitted with all data set-specific conditions, we aggregated the posterior across conditions[^2], and excluded conditions that might conflate comparisons[^3]. For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}. The resulting posterior allows us to examine differences and similarities associated with transition locations for all dataset. In particular we can see how the linguistic location in the text is associated with hesitation duration and probability. The estimate for fluent transitions is shown by dataset, across transition location.

[^2]: We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set. We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.

[^3]: We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set. There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-gunnexp2}).

```{r crossstudypost2, fig.cap="Mixture model parameter estimates across studies. Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI). Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group."}

posd <- position_dodge(.5)
dotsize <- 2.75
grouplabel <- "Data set:"
shapes <- c(1, 4, 6, 7, 8, 9, 11)
width <- 0
linewidth <- .35
nrow <- 1
beta <- filter(ps, name == "beta", 
               location == "before\nword") %>%
  mutate(name = beta_label,
         location = "overall") %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             shape = data,
             group = interaction(data))) +
  geom_line(linewidth = linewidth, 
            alpha = .75, 
            position = posd) +
  geom_errorbar(width = width, alpha = .75, position = posd) +
  geom_point(aes(colour = data), 
            size = dotsize, 
            position = posd) +
  facet_grid(~name, scales = "free_x") +
  scale_shape_manual(values = shapes) +
  scale_colour_colorblind() +
  scale_y_continuous(labels = scales::comma) +
  theme(axis.title = element_blank()) +
  labs(colour = grouplabel,
       shape = grouplabel) +
  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
         shape = guide_legend(nrow=nrow,byrow=TRUE))

delta <- filter(ps, name == "delta") %>% 
  mutate(name = delta_label,
         across(c(mean, lower, upper), ~./1000)) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             shape = data,
             group = interaction(data))) +
  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
  geom_errorbar(width = width, alpha = .75, position = posd) +
  geom_point(aes(colour = data), 
            size = dotsize, 
            position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = shapes) +
  scale_colour_colorblind() +
  scale_y_continuous(labels = scales::comma) +
  labs(colour = grouplabel,
       shape = grouplabel) +
  theme(axis.title = element_blank()) +
  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
         shape = guide_legend(nrow=nrow,byrow=TRUE))

theta <- filter(ps, name == "prob") %>% 
  mutate(name = theta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             shape = data,
             group = interaction(data))) +
  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
  geom_errorbar(width = width, alpha = .75, position = posd) +
  geom_point(aes(colour = data), 
            size = dotsize, 
            position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = shapes) +
  scale_colour_colorblind() +
  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
  labs(colour = grouplabel,
       shape = grouplabel) +
  theme(axis.title = element_blank()) +
  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
         shape = guide_legend(nrow=nrow,byrow=TRUE))

plot <- beta + delta + theta + plot_layout(guides = "collect",
                                           widths = c(.5, 1, 1))

grid.arrange(patchworkGrob(plot), 
             left = "Posterior estimate with 95% PIs",
             bottom = "Transition location")

```

Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets. We summarise the results of the pairwise comparisons for transitions locations; full results and parameter estimates can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}. We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs \> 100 for all datasets) but only for half of the data sets also longer (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4). Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GUNNEXP2 and SPL2, both BFs \> 100) but longer (all BFs \> 100, except for C2L1: BF = 1.11 and LIFT = 0.5).

To reiterate, for most datasets we found that pauses are before sentences are longer but not necessarily more likely than before words (except for datasets SPL2 and GUNNEXP2). In fact, LIFT showed some indication that pauses were less likely before sentences than words. This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for key transitions at larger linguistic edges. It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.

Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character while the remaining data sets measured before-sentence transitions as the difference between space key and shift. To control for the use of complex key combinations, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability (BFs \> 100) but it did not affect the duration of fluent or hesitant transitions (BFs \< 0.3); see Appendix \ref{key-combination-effect-constrained-mixture-model}. This conflicts the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did included the character following shift (PLanTra, LIFT). In other words, patterns in our results cannot be explained on the basis of how before-sentence key-transitions were operationalised.

Finally, the duration of fluent transitions was not distinguished by location in the constrained distribution model. Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transitions. This is presumably reflecting that these data were sampled from was the youngest and least experienced population of writers in our data pool. Hesitation duration and probability followed the same patterns as the majority of datasets.

The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}). We, therefore, summarise briefly differences in the results for the transition-location effects; full results can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}. Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs \> 100) but not before-sentence compared to before-word transitions (all BFs \< 0.08, but SPL2 and GUNNEXP2: BFs \> 100). Patterns for hesitation probability was generally similar in both the constrained and unconstrained model. However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs \> 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.

### 2024-04-27 tutorial/tutorial.Rmd
rearrange markdown sections

@@
 ---
-title: "Tutorial: fitting mixed-effects and log-Gaussian mixture models on text-writing data"
+title: "Tutorial: Fitting log-Gaussian mixed-effects mixture models for keystroke data"
 author: "Jens Roeser"
@@
 
-This is a step-by-step walk-through that describes how to fit a finite mixture model of two log-Gaussian distributions using the statistical program R and the `rstan` package to interface with the probabilistic programming language Stan [@carpenter2016stan; @hoffman2014no; @rstan; @rstan2].
-We will compare this mixture model to standard Gaussian and log-Gaussian models implemented in `brms` [@burkner2017brms; @R-brms_b].^[An alternative to writing models in Stan is the R package `brms` which provides a flexible framework to implement mixture models and many other types of probability models [@burkner2017brms; @R-brms_b].
-In particular, `brms` has a `mixture` function to specify mixtures of various types of distributions.
+This is a step-by-step tutorial for fitting mixed-effects mixture models with two log-Gaussian distributions [@roeser2021] using R [@R-base] and the `rstan` package to interface with the probabilistic programming language Stan [@carpenter2016stan; @hoffman2014no; @rstan; @rstan2].
+We will compare the mixture-model analysis to standard Gaussian and log-Gaussian mixed-effect models fitted using `brms` [@burkner2017brms; @R-brms_b].^[The R package `brms` provides a flexible framework to implement mixture models and many other types of probability models [@burkner2017brms; @R-brms_b].
+`brms` has a `mixture` function to specify mixtures of various types of distributions.
 These could be Gaussian, skewed-Normal, shifted-Normal, ex-Gaussian etc.
@@
 There is a large number of probability models for continuous data that are plausible candidates [for reaction time data see e.g.
-@matzke2009psychological].] For a rationale of using mixture models for writing process data see @roeser2021.
-
-This guide shows how to fit a mixture model written in Stan to keystroke data to then calculate the differences between transition locations in the text.
+@matzke2009psychological].
+In this tutorial we use Stan for mixture modelling to demonstrate the process and because the publication associated with this tutorial is fully using Stan instead of `brms`.]
+
+In this guide we show how to fit a mixture model written in Stan to keystroke data.
+Then we demonstrate how the posterior of this model can be used to calculate differences between transition locations in the text.
+Also we show how to compare model using leave-one-out cross-validation.
 This tutorial is largely self-contained.
 To run the code below, the reader only needs to install the required packages.
-Data and Stan code can be loaded from the code presented below.
-
-We require three packages: (1) `rstan` to use R to interface with Stan for fitting Bayesian models^[Instructions on how to install `rstan` can be found [here]( https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).]; (2) `tidyverse` for data processing and visualisation.^[Run `install.packages("tidyverse")` for installation.]; (3) `brms` for simple Bayesian models.^[Run `install.packages("brms")` for installation.]
+Data and the Stan programme can be loaded using the code below.
+
+Three packages need to be installed before (see footnotes for instructions): (1) `rstan` to use R to interface with Stan for fitting Bayesian models^[Instructions on how to install `rstan` can be found [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).]; (2) `tidyverse` for data wrangling and visualisation.^[Run `install.packages("tidyverse")` for installation.]; (3) `brms` for fitting Bayesian models.^[Run `install.packages("brms")` for installation.]
 
@@
 First we load a sample of keystroke data.
-Data are loaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/) using the URL as below and the `read_csv` function.
-For brevity we use a random subset.
-Data are from participants that completed an essay writing task in the context of the Prowrite project.
+Data are loaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/) using the URL as below.
+For simplicity we use a random subset of data from participants who completed an essay-writing task in the context of the ProWrite project.
 
@@
 
-# Uni-modal mixed-effects models
-
-We start with two uni-modal models -- models that assume that the data come from a single underlying process that can be described as as Gaussian in the first example or a log-Gaussian in the second example.
-We will compare the models performance.
+# Mixed-effects models
+
+We start with two standard single distribution mixed-effects models [as in @pinheiro2007linear; @baa08], i.e.
+models that assume that data come from a single underlying process that can be described as Gaussian or log-Gaussian.
+We will compare these models.
 
@@
 
-We start with a Gaussian mixed-effects models similar to a standard analysis of variance.
-We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
-The mean can be decomposed into $\beta$ and $\text{participant}_i$.
-$\beta$ will be allowed to take on a different value for each transition location.
-Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.
+We start with a Gaussian mixed-effects model.
+We describe the process that generates the $i^\text{th}$ keystroke, where $i \in 1\dots N$ and $N$ is the total number of keystroke, as a single Gaussian (normal) distribution $\mathcal{N}()$ which is characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
+The mean can be decomposed into $\beta$ and $u_\text{participant[i]}$.
+$\beta$ will be allowed to take on a different value for each transition location associated with the $i^\text{th}$ keystroke.
+By-participant ikis are assumed to vary around the average which is achieved by placing a normal distribution on the difference between average and by-participant ikis, indicated as $u_\text{participant[i]}$, with a mean of 0 and a standard deviation $\sigma_\text{p}^2$.
 
@@
 \begin{align}
-\text{iki}_i \sim & \text{N}(\mu_i, \sigma_\text{e}^2)\\
-\mu_i = & \beta_\text{location[i]} + \text{participant}_i\\
-\text{participant} \sim & \text{N}(0, \sigma_\text{p}^2) 
+\text{iki}_i \sim & \mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
+\mu_i = & \beta_\text{location[i]} + u_\text{participant[i]}\\
+u_\text{participant[i]} \sim & \mathcal{N}(0, \sigma_\text{p}^2) 
 \end{align}
@@
 
-`brms` is using sensible default priors for most model parameters which can be checked using `get_prior`.
+`brms` is using sensible regulating default priors for most model parameters which can be inspected using `get_prior` which needs the model formula and the data as input.
 
@@
 
-We want to make sure that `(flat)` is replaced by weakly informative priors because uninformative or flat priors are never a good idea [@mcelreath2016statistical].
-
-```{r}
-prior <- set_prior('student_t(3, 350, 100)', class = 'b')
-```
-
-```{r}
+We want to make sure that `(flat)` is replaced by more (i.e.
+weakly or regulating) informative priors [see @mcelreath2016statistical; @kruschke2010doing; @lambert2018student].
+For example, we might consider a Student-*t* distribution as prior over the slope parameters (class `b`) which in our model are parameterised as the cell means of the three transition locations (levels: within word, before word, before sentence) with a mean of 350 msecs and a standard deviation of 75 msecs.
+
+```{r}
+prior <- set_prior('student_t(3, 350, 75)', class = 'b')
+```
+
+Student-*t* distributions, in comparison to normal distributions, have a degrees of freedom parameter (first argument) in addition to the mean and the standard deviation (second and third parameter, respectively).
+The degrees of freedom control the tails of the distribution with smaller values assigning more probability mass to the tails; the shape of the Student-*t* distribution is moving closer to a normal distribution when the degrees of freedom approach $\infty$.
+This is illustrated in Figure \@ref(fig:prior) for the mean and the standard deviation of the prior above.
+
+```{r prior, echo = F, fig.cap="Student-t prior with varying degrees of freedom."}
 mean <- 350
-sd <- 100
+sd <- 75
 tibble(x = 0:1000) %>% 
@@
 
-### Initiate sampling
+### Sampling parameters
 
@@
 10,000 iterations, as below, are a lot but does not guarantee convergence.
-We need to test convergence using the model's posterior (see blow).
-To test whether the model has converged, we need to run the model more than one time (i.e.
+To be able to test whether converged has reached, we need to run the model several time (i.e.
 different chains).
-These can be run at the same time (in parallel) using more than one core of your computer (three cores below).^[If you want to use R to check how many cores are available on your machine, run `parallel::detectCores()` (you might need to install the package `parallel`).] There is no need to use more cores and chains and using less cores would mean that at least one process has to wait until after other processes are completed.
-When the model has settled on a parameter values, we should observe that all three streams (chains) overlap.
-Running long chains (many iterations) is useful to get more accurate parameter estimates.
-
-```{r}
-iterations <- 10000      # No of iterations
+This can be done at the same time (in parallel) when using more than one core of your computer (three cores below).^[If you want to check how many cores are available on your machine, run `parallel::detectCores()` (you might need to install the package `parallel`).] There is no need to use more cores than chains; using less cores would mean that at least one process has to wait until after another processes has been completed.
+When the model has settled on a distribution of parameter values, we should observe, after the model has finished sampling, that all chains mixed nicely.
+Running long chains (many iterations) is useful to get more accurate parameter estimates but will take more time.
+
+```{r}
+iterations <- 10000      # No. of iterations
 warmup <- iterations / 2 # Warmup samples to be discarded
-nchains <- ncores <- 3   # No of chains / cores used (one chain per core)
+nchains <- ncores <- 3   # No. of chains / cores used (one chain per core)
 ```
@@
 
-The following code combines the model, the data, and the prior and returns the posterior. This will take a couple of minutes.
+The following code combines the model, the data, and the prior and returns the posterior.
+This will take a few of minutes to complete sampling from the posterior after `brms` has compiled the model code.
 
@@
 
-Because Bayesian models tend to take more time to run, it makes sense to save the results using `saveRDS` for later re-use.
-
-The output of the model (i.e.
-the posterior) can be saved, so we don't need to run the model again (which can take a while).
+Because Bayesian models tend to take more time to run, it makes sense to save the results using `saveRDS` for re-use.
 The function requires the name of the fitted model `fit_gaus`.
-We prefer to keep the name of the output similar to the name of the Stan code used and ideally the R script that is running the model so it is easy to find the script that is running which model.
 The model is stored as compressed ".rda" file.
@@
 
-To read the fitted model back into your working environment used `readRDS`.
+To read the fitted model back into R's working environment use `readRDS`.
 
@@
 
-The function `fixef` provide a summary of the model inference showing the posterior estimates as average and 95% probability interval by transition location in msecs.
+The function `fixef` provide a summary of the posterior showing parameter value estimates as average and 95% probability interval by transition location in msecs.
 
@@
 ### Model
+
+This model is largely identical to the Gaussian mixed-effect model in the previous section, except instead of a normal distribution this model assumes the ikis come from a single underlying process that follows a log-normal distribution.
+Assuming a log-normal distribution is useful as ikis and other response times are zero-bound (keystroke intervals cannot be shorter than 0).
+A log-normal distribution forces responses to be positive.
+Also the log scale is not linear, meaning the distance between 100 msecs and 110 msecs is not the same as the distance between 500 and 510 msecs; instead the distance between adjacent values is reduced for larger values while the distances between adjacent values on the lower end is increased.
+For example, for a log with the base 10, the distance between 10 msecs and 100 msecs is identical to the distance between 100 msecs and 1000 msecs.
 
@@
 \begin{align}
-\text{iki}_i \sim & \text{logN}(\mu_i, \sigma_\text{e}^2)\\
-\mu_i = & \beta_\text{location[i]} + \text{participant}_i\\
-\text{participant} \sim & \text{N}(0, \sigma_\text{p}^2) 
+\text{iki}_i \sim & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
+\mu_i = & \beta_\text{location[i]} + u_\text{participant[i]}\\
+u_\text{participant[i]} \sim & \mathcal{N}(0, \sigma_\text{p}^2) 
 \end{align}
 $$
+For the model formula we use the distribution family `lognormal`.
 
@@
 
+The priors default priors and the priors we are going to set are now no longer on the msecs scale but on the log scale.
+
 ```{r}
@@
 
-```{r}
-prior <- set_prior('student_t(3, 5, .5)', class = 'b')
-```
-
-```{r}
-mean <- 5
+For example, a mean of 5 msecs on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)`. This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).
+
+```{r}
+prior <- set_prior('student_t(3, 5.85, .5)', class = 'b')
+```
+
+```{r logprior, fig.cap="Student-t prior with varying degrees of freedom on log scale.", echo = F}
+mean <- 5.85
 sd <- .5
-tibble(x = seq(3, 7, .01)) %>% 
+tibble(x = seq(3, 9, .01)) %>% 
   mutate(df_3 = dstudent_t(x, 3, mean, sd),
@@
 ### Fitting the model
+
+We fit the model as before with the same sampling parameters.
 
@@
 
-or directly calculated using `loo`. The Pareto k diagnostic indicates that there are some observations that are very unusual under the uni-modal Gaussian model.
+or directly calculated using `loo`. The Pareto k diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model.
 
@@
 A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; chapter 6 in @mcelreath2016statistical].
-The is an example for Bayesian $R^2$ for the uni-modal log-Gaussian model.
+The is an example for Bayesian $R^2$ for the single-distribution log-Gaussian model.
 
@@
 
-# Bi-modal mixed-effects log-Gaussian mixture model
+# Two-distributions log-Gaussian mixed-effects mixture model
 
@@
 \begin{align}
-\text{iki}_{i} \sim & \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
+\text{iki}_{i} \sim & \theta_\text{location[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
 		&\text{where}\\
@@
 The frequency of the minority of disfluent transitions is captured by the height of the green distribution.
-As this model is bi-modal, the probability of fluent transitions is just $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions was $.6$.
+As this model assumes a finite mixture of two distributions, the probability of fluent transitions is just $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions was $.6$.
 If the pausing probability decreases to $.3$ the proportion of fluent transitions had increased to $.7$.
@@
 
-# Uni-modal log-Gaussian
+# Single distribution log-Gaussian
 loo_lgaus <- loo(fit_lgaus)
 
-# Uni-modal Gaussian
+# Single distribution Gaussian
 loo_gaus <- loo(fit_gaus)
@@
 
-As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to each the uni-modal Gaussian and log-Gaussian model.
+As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to each the single distribution Gaussian and log-Gaussian model.
 Their difference is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.

### 2024-04-27 report/scripts/get_pdens_plot.R

@@ -17,7 +17,7 @@ get_pdens_plot <- function(file, max_iki, ds){
       . == "lmm" ~ "Single log-Gaussian",
       . == "lmmgaus" ~ "Single Gaussian")),
       across(model, ~str_wrap(., 20)),
-      across(model, ~factor(., levels = sort(unique(model))[c(5, 4, 3, 1, 2)], ordered = T)))
+      across(model, ~factor(., levels = sort(unique(model))[c(5, 4, 3, 2, 1)], ordered = T)))
   
   plot <- sims %>%
     ggplot(aes(x = value / 1000, 

### 2024-04-28 report/manuscript.Rmd
manuscript update

@@
 
-abstract: Classical serial models view the process of producing a text as a discrete chain of pauses and writing bursts.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two views in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found -- across populations -- that transitions across word and even sentence boundaries are often fluent and uninterrupted by planning of upcoming linguistic material; when pausing occurred pauses were longer before sentence than before word which is consistent with the idea larger linguistic edges are associated with higher level planning.
-Our results cannot be explained by the serial but are in line with the cascading view of writing.
+abstract: "Classical serial models view the process of producing a text as a chain of discrete pauses and writing bursts.
+In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel with execution.
+These two theories make different assumptions about the underlying process that generates inter-keystroke intervals.
+Across six datasets comprising multi-sentence text composition we instantiated these assumptions as two sets of Bayesian mixed-effects models: (1) single-distribution models as assumed by the serial processing account, and (2) finite two-distribution mixture models as assumed by the parallel processing account.
+We analysed keystroke intervals at before-sentence, before word, and within word transitions.
+Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view across all datasets.
+When pausing occurred, sentence initial pauses were longer than word initial pauses which is consistent with the idea that larger linguistic edges are associated with higher level planning.
+However, we found -- across populations -- that keystroke latencies at word and even at sentence boundaries were often too rapid to permit planning of what was written next.
+Our results cannot be explained by the serial but are in line with the cascading, semi-parallel view of writing."
 
-keywords: "Keystroke modelling; finite mixture models; Bayesian models; text composition"
+keywords: "Cascading parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"
 
@@
 ```
-
-# Introduction
 
@@
 
+\newpage
+
 ```{r child = 'appendix/models.Rmd'}

### 2024-04-28 report/manuscript.tex
manuscript update

@@
 {}{\PatchFailed}
-\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
+\keywords{Cascading parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes}
 \usepackage{csquotes}
@@
   pdflang={en-EN},
-  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
+  pdfkeywords={Cascading parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes},
   hidelinks,
@@
 \abstract{%
-Classical serial models view the process of producing a text as a discrete chain of pauses and writing bursts.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution.
-We implemented these two views in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production.
-We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets.
-Further we found -- across populations -- that transitions across word and even sentence boundaries are often fluent and uninterrupted by planning of upcoming linguistic material; when pausing occurred pauses were longer before sentence than before word which is consistent with the idea larger linguistic edges are associated with higher level planning.
-Our results cannot be explained by the serial but are in line with the cascading view of writing.
+Classical serial models view the process of producing a text as a chain of discrete pauses and writing bursts.
+In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel with execution.
+These two theories make different assumptions about the underlying process that generates inter-keystroke intervals.
+Across six datasets comprising multi-sentence text composition we instantiated these assumptions as two sets of Bayesian mixed-effects models: (1) single-distribution models as assumed by the serial processing account, and (2) finite two-distribution mixture models as assumed by the parallel processing account.
+We analysed keystroke intervals at before-sentence, before word, and within word transitions.
+Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view across all datasets.
+When pausing occurred, sentence initial pauses were longer than word initial pauses which is consistent with the idea that larger linguistic edges are associated with higher level planning.
+However, we found -- across populations -- that keystroke latencies at word and even at sentence boundaries were often too rapid to permit planning of what was written next.
+Our results cannot be explained by the serial but are in line with the cascading, semi-parallel view of writing.
 }
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
-
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g.
+Translating ideas into language involves a sequence of processes starting with communicative intent, proceeding through generation of a message, retrieving appropriate lexical items, deciding with which part of the message to start the utterance which, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), is determined by syntactic requirements of our language, then retrieving orthographic and / or phonological codes, and that are finally submitted to the motor programs that output speech, key or pen-strokes or signs (Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
+
+The coordination of these processes in writing is arguably more demanding than in spoken language.
+This is because writing requires spelling retrieval which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen are later-developing and more effortful than those required to articulate language sounds in speech leading to a reduced output speed compared to speech (see for a recent review Kandel, 2023).
+Consequently when writing, more than in speech, we have to mentally buffer linguistic plans before we can express them in language.
+Written communication is also less time constrained: speaking underlies fluency requirements, because hesitating during speech has a communicative effect (e.g.
 Clark \& Fox Tree, 2002).
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
-
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-The same is generally observed for spoken utterances (E.-K.
-Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001).
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
-
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991).
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
+There is therefore stronger pressure to plan ahead in speech than in writing (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
+Indeed writers can pause when they wish -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
+Written text production -- for example writing argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
+
+The study reported in this paper compares statistical models that instantiate two different theories of how cognitive planning processes are coordinated in writing.
+Through the analysis of 6 different datasets from writers of different ages and competencies composing multi-sentence texts, we compare these statistical models and therefore the cognitive model of how keystrokes in writing are being generated they are mapping on.
+Also, we evaluate competing predictions of these models by revisiting the often-reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
+In what follows we first summarise previous research that has explored pause patterns in spontaneous text production.
+We then introduce two possible cognitive frameworks in which these can be interpreted.
+Finally we argue that analysis of pause-patterns at linguistic edges are important to permit differentiation between these two theories.
+
+We focus on data form spontaneous text production (see Gernsbacher \& Givón, 1995) for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production and have, therefore, real-world relevance (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
+Second, since the emergence of keystroke-logging software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; for a recent special issue see Conijn and Torrance (2023).\footnote{Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.} These data are therefore important for contemporary theories of writing and freely available for statistical modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
+Third, spontaneous text production -- unlike picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Meulemans et al., 2022; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021) -- content generation is not artificially constrained to individual sentence units (or less) by one or more images and linguistic form is not constrained to a sequence of words, respectively.
+Instead text composition provides data from both inter and intra-sentence transitions.\footnote{Also, pauses in text production do not reflect visual encoding of the stimulus (although possibly reflect looking back during writing which we will return to in the discussion).}
+
+Time intervals between adjacent keystrokes during typing vary as a function of various factors.
+Explaining the factors that influence the duration of inter-keystroke intervals (IKIs) requires a theory of how the various processes that transform intent into keypresses are structured.
+Processing -- planning what to say and how to say it -- may be serial.
+This means that the processing sequence operates one unit at a time.
+Processing of the next language unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
 Therefore longer IKIs represent periods when the writer is planning what to write next.
@@
 
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
+This serial account is consistent with the frequent finding that average IKI duration immediately before larger linguistic units compared to smaller linguistic units.
+For example, in spontaneous text the mean pause duration before a writer starts a sentence has typically been reported to be longer than before a mid-sentence word and these are, in turn, longer than between mid-word key presses (Deane \& Zhang, 2015; Mohsen, 2021; Spelman Miller, 2006; Spelman Miller et al., 2008; Torrance, Rønneberg, et al., 2016; Wengelin, 2002; Xu \& Qi, 2017).
+This is understood as being due to time needed to plan message and syntax for the next sentence, which is added to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke of that word (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
 However this account is not consistent with what we know about writing.
@@
 
-First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production (e.g.
-Bock \& Ferreira, 2014).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al.
+First, writers and speakers can in principle plan utterances in full for one or multiple sentences before beginning to output words.
+However this is not what people typically do: utterances are not fully formed at production onset.
+Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of a real-time process.
+For speech it is well known that the pre-planning duration depends -- among other factors -- on the need to avoid intra-sentence hesitation that are needed to plan upcoming words in parallel to production (e.g.
+Levelt \& Meyer, 2000; Wheeldon \& Konopka, 2023).
+For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech {[}Nottbusch (2010); dam09; Roeser et al.
 (2019){]}.
@@
 (2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
-(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.
-
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration.
+Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset (and speech onset).
+This is not what one would expect if writers plan sentences in full before production onset.
+In fact, eye-movement data reported in Torrance and Nottbusch (2012) and Roeser et al.
+(2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases, i.e.~when the second noun was part of the same phrase as the first noun.
+
+Second, even though the average sentence-initial IKI tends to be longer, they still tend to be very rapid.
+For example, Medimorec and Risko (2017) found that, in undergraduate students' writing on familiar topics 71\% of sentence-initial IKIs were less than 1 seconds in duration.
 Also Rønneberg et al.
 (2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population.
+For comparison, these short pauses are less than mean written naming response latency for single objects (Torrance et al., 2018) in a similar population.
+Longer pauses in the remainder were similar to response latencies for short sentences when describing arrays of images (Roeser et al., 2019), also in a similar population.
 Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
+The fact that these very short sentence initial pauses exist, and in fact are quite common, is not consistent with the assumption that planning of the upcoming sentence is started and finished in the period between completion of the previous sentence and starting to write the next.
 
 If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
+The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that operate partly in parallel and largely without executive control.
 Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
@@
 Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
+In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing the output of the previous sentence.
 
 This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
+IKIs result from one of two underlying mental processes.
+When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then IKIs are determined just by the time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
+However if one or more upstream processes provide output more slowly, then IKIs become dependent directly on upstream processing times and not on finger movement.
 IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-
-For example, Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see Almond et al.
+The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
+This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
+
+In language comprehension, Vasishth and colleagues (Vasishth, Jäger, et al., 2017; Vasishth, Chopin, et al., 2017) demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental.
+Similar for keystroke data Roeser et al.
+(2021) demonstrated -- using a similar approach as used in the present paper\footnote{For similar approaches see Almond et al.
 (2012), Baaijen et al.
@@
 (2022), Li (2021), and Van Waes et al.
-(2021).
-
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
-This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
-
-Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-Rønneberg et al.
-(2022) coined this possibility the \enquote{process-disruption hypothesis}.
-
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al.
+(2021).} -- how models that assume this mixture of two underlying data generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g.~\enquote{dkdkdkdk}).
+As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
+However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes.
+Typing hesitations in the context of copy-typing may -- to some extent -- reflect not just production but also, and possibly largely, visual encoding.
+Text composition, in contrast, provides data from the free generation of words and full sentences that is not dependent on a visually presented text stimulus.
+For these reasons our research extends the analysis of keystroke data using a mixture model approach to data from spontaneous text production.
+
+In the study that is presented in this paper we test using Bayesian hierarchical models of keystroke intervals -- similar to Roeser et al.
 (2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
@@
 
-Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-
-Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023).
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021).
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
 The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
+In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word locations compared to within-word locations.
 This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
+The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not (always) pause to plan their text but plan in parallel to writing.
 In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
+Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occur to some extent probabilistically across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
+We hypothesise that the statistical model of the parallel cascading view provides a better out-of-sample generalisation than the implementation of the serial model of writing in the context of unconstrained spontaneous text composition.
 
@@
 The Bayesian framework is ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014).
+This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; Lee \& Wagenmakers, 2014).
 To achieve this, Bayesian models require the inclusion of prior information, i.e.~existing knowledge about parameter values.
@@
 
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\begingroup
+
+\setlength{\parindent}{-0.5in}
+
+\setlength{\leftskip}{0.5in}
+
+\hypertarget{references}{}
+
+\endgroup
+
+\newpage
+
 \hypertarget{appendix-appendix}{%
@@
 
+\newpage
+
 \hypertarget{statistical-models}{%
@@
 We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M.
-D.
-Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
+Farrell \& Lewandowsky, 2018; Gelman et al., 2014; Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
 In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-
-\setlength{\parindent}{-0.5in}
-
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\newpage
-
 \hypertarget{refs}{}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-deane2015exploring}{}}%
+Deane, P., \& Zhang, M. (2015). \emph{Exploring the feasibility of using writing process features to assess text production skills} (RR-15-02). Educational Testing Service.
+
 \leavevmode\vadjust pre{\hypertarget{ref-dell2007case}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kandel2023written}{}}%
+Kandel, S. (2023). Written production: The {APOMI} model of word writing: Anticipatory processing of orthographic and motor information. In \emph{Language production} (pp. 209--232). Routledge.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kaufer1986composing}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lee13}{}}%
-Lee, E.-K., Brown-Schmidt, S., \& Watson, D. G. (2013). Ways of looking ahead: Hierarchical planning in language production. \emph{Cognition}, \emph{129}(3), 544--562.
-
 \leavevmode\vadjust pre{\hypertarget{ref-lee2014bayesian}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-levelt2000word}{}}%
+Levelt, W. J. M., \& Meyer, A. S. (2000). Word for word: Multiple lexical access in speech production. \emph{European Journal of Cognitive Psychology}, \emph{12}(4), 433--452.
+
 \leavevmode\vadjust pre{\hypertarget{ref-li2021identifying}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-meyer1996}{}}%
-Meyer, A. S. (1996). Lexical access in phrase and sentence production: Results from picture--word interference experiments. \emph{Journal of Memory and Language}, \emph{35}(4), 477--496.
+\leavevmode\vadjust pre{\hypertarget{ref-meulemans2021}{}}%
+Meulemans, C., De Maeyer, S., \& Leijten, M.
+(2022).
+Generalizability of pause times in sentence production to distinguish between adult writers.
+\emph{Behavior Research Methods}, \emph{54}(4), 1976--1988.
+\url{https://doi.org/10.3758/s13428-021-01707-1}
 
@@
 Sivula, T., Magnusson, M., Matamoros, A. A., \& Vehtari, A. (2020). Uncertainty in {B}ayesian leave-one-out cross-validation based model comparison. \emph{arXiv Preprint arXiv:2008.10296}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-miller2006pausing}{}}%
+Spelman Miller, K.
+(2006).
+Pausing, productivity and the processing of topic in online writing.
+In K.
+P.
+H.
+Sullivan \& E.
+Lindgren (Eds.), \emph{Computer key-stroke logging and writing} (Vol.
+18, pp.
+131--156).
+Brill.
+\url{https://doi.org/10.1163/9780080460932_009}
+
+\leavevmode\vadjust pre{\hypertarget{ref-miller2008psycholinguistic}{}}%
+Spelman Miller, K., Lindgren, E., \& Sullivan, K.
+P.
+(2008).
+The psycholinguistic dimension in second language writing: Opportunities for research and pedagogy using computer keystroke logging.
+\emph{Tesol Quarterly}, \emph{42}(3), 433--454.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wheeldon2013}{}}%
-Wheeldon, L.
-R., Ohlson, N., Ashby, A., \& Gator, S.
-(2013).
-Lexical availability and grammatical encoding scope during spoken sentence production.
-\emph{The Quarterly Journal of Experimental Psychology}, \emph{66}(8), 1653--1673.
-
 \leavevmode\vadjust pre{\hypertarget{ref-wing1973response}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-xu2017analyzing}{}}%
+Xu, C., \& Qi, Y. (2017). Analyzing pauses in computer-assisted EFL writing---a computer-keystroke-log perspective. \emph{Journal of Educational Technology \& Society}, \emph{20}(4), 24--34.
+
 \end{CSLReferences}

### 2024-04-28 report/sections/introduction.Rmd
manuscript update

@@
-Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing [@bock2014syntactically; @wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].
+Translating ideas into language involves a sequence of processes starting with communicative intent, proceeding through generation of a message, retrieving appropriate lexical items, deciding with which part of the message to start the utterance which, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], is determined by syntactic requirements of our language, then retrieving orthographic and / or phonological codes, and that are finally submitted to the motor programs that output speech, key or pen-strokes or signs [@wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].
 
-Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech.
-Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language.
-Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking [@roeser2018advance; @torrance2012written].
-This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader [e.g.
+The coordination of these processes in writing is arguably more demanding than in spoken language.
+This is because writing requires spelling retrieval which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen are later-developing and more effortful than those required to articulate language sounds in speech leading to a reduced output speed compared to speech [see for a recent review @kandel2023written].
+Consequently when writing, more than in speech, we have to mentally buffer linguistic plans before we can express them in language.
+Written communication is also less time constrained: speaking underlies fluency requirements, because hesitating during speech has a communicative effect [e.g.
 @Clark2002].
-Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].
+There is therefore stronger pressure to plan ahead in speech than in writing [@roeser2018advance; @torrance2012written].
+Indeed writers can pause when they wish -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
+Written text production -- for example writing argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].
 
-The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models.
-On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06].
-For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses].
-The same is generally observed for spoken utterances [@lee13; @meyer1996; @wheeldon2013] with some exceptions [@gri01].
-We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted.
-Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.
+The study reported in this paper compares statistical models that instantiate two different theories of how cognitive planning processes are coordinated in writing.
+Through the analysis of 6 different datasets from writers of different ages and competencies composing multi-sentence texts, we compare these statistical models and therefore the cognitive model of how keystrokes in writing are being generated they are mapping on.
+Also, we evaluate competing predictions of these models by revisiting the often-reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06].
+In what follows we first summarise previous research that has explored pause patterns in spontaneous text production.
+We then introduce two possible cognitive frameworks in which these can be interpreted.
+Finally we argue that analysis of pause-patterns at linguistic edges are important to permit differentiation between these two theories.
 
-Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements [@van1991handwriting; @olive2014toward].
-The intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing may be serial.
-This means that the processing cascade operates on one unit at a time.
-Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
+We focus on data form spontaneous text production [see @gernsbacher1995coherence] for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production and have, therefore, real-world relevance [@dux2022automating; @dux2021effect; @vandermeulen2023writing].
+Second, since the emergence of keystroke-logging software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; for a recent special issue see @conijn2021timecourse.^[Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.] These data are therefore important for contemporary theories of writing and freely available for statistical modelling and machine learning [@conijn2019understanding; @conijn2020keys].
+Third, spontaneous text production -- unlike picture description experiments [@not10; @dam09; @meulemans2021; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual] -- content generation is not artificially constrained to individual sentence units (or less) by one or more images and linguistic form is not constrained to a sequence of words, respectively.
+Instead text composition provides data from both inter and intra-sentence transitions.^[Also, pauses in text production do not reflect visual encoding of the stimulus (although possibly reflect looking back during writing which we will return to in the discussion).]
+
+Time intervals between adjacent keystrokes during typing vary as a function of various factors.
+Explaining the factors that influence the duration of inter-keystroke intervals (IKIs) requires a theory of how the various processes that transform intent into keypresses are structured.
+Processing -- planning what to say and how to say it -- may be serial.
+This means that the processing sequence operates one unit at a time.
+Processing of the next language unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
 Therefore longer IKIs represent periods when the writer is planning what to write next.
@@
 
-This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second].
+This serial account is consistent with the frequent finding that average IKI duration immediately before larger linguistic units compared to smaller linguistic units.
+For example, in spontaneous text the mean pause duration before a writer starts a sentence has typically been reported to be longer than before a mid-sentence word and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second; @miller2006pausing; @xu2017analyzing; @miller2008psycholinguistic;@deane2015exploring].
+This is understood as being due to time needed to plan message and syntax for the next sentence, which is added to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke of that word [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses].
 However this account is not consistent with what we know about writing.
@@
 
-  First, utterances are not fully formed at production onset.
-Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process.
-This is well known for spoken production [e.g.
-@bock2014syntactically].
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases [@not10; dam09; @roeser2018advance].
+First, writers and speakers can in principle plan utterances in full for one or multiple sentences before beginning to output words.
+However this is not what people typically do: utterances are not fully formed at production onset.
+Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of a real-time process.
+For speech it is well known that the pre-planning duration depends -- among other factors -- on the need to avoid intra-sentence hesitation that are needed to plan upcoming words in parallel to production [e.g.
+@levelt2000word; @wheeldon_konopka_2023].
+For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech [@not10; dam09; @roeser2018advance].
 For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e.
 *A and the B moved*) or a simple noun (i.e., *A moved above the B and the C*).
-Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset.
-This is not what one would expect if writers plan sentences (or even phrases) in full before production onset.
-In fact, eye movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e.
-*the B*) was more likely for conjoined noun phrases.
+Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset (and speech onset).
+This is not what one would expect if writers plan sentences in full before production onset.
+In fact, eye-movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e.
+*the B*) was more likely for conjoined noun phrases, i.e.
+when the second noun was part of the same phrase as the first noun.
 
-Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid.
-For example, @medimorec2017pauses found that, in undergraduate students writing on familiar topics 71% of sentence-initial IKIs were less than 1 sec in duration.
+Second, even though the average sentence-initial IKI tends to be longer, they still tend to be very rapid.
+For example, @medimorec2017pauses found that, in undergraduate students’ writing on familiar topics 71% of sentence-initial IKIs were less than 1 seconds in duration.
 Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
-For comparison, this is less than mean written naming response latency for single objects [@torrance2018timed] and short sentences when describing arrays of images [@roeser2018advance] in a similar population.
+For comparison, these short pauses are less than mean written naming response latency for single objects [@torrance2018timed] in a similar population.
+Longer pauses in the remainder were similar to response latencies for short sentences when describing arrays of images [@roeser2018advance], also in a similar population.
 Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.
+The fact that these very short sentence initial pauses exist, and in fact are quite common, is not consistent with the assumption that planning of the upcoming sentence is started and finished in the period between completion of the previous sentence and starting to write the next.
 
 If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control.
+The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that operate partly in parallel and largely without executive control.
 Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction].
@@
 Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.
+In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing the output of the previous sentence.
 
 This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two data-generating processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.
+IKIs result from one of two underlying mental processes.
+When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then IKIs are determined just by the time needed for executing finger movements (i.e.
 just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement.
+However if one or more upstream processes provide output more slowly, then IKIs become dependent directly on upstream processing times and not on finger movement.
 IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling].
-
-For example, @roeser2021modelling demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling.
-In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach.
-Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions.
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-For similar approaches see @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.
-
 The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
@@
 
-Hesitations in writing have potential disadvantages.
+```{r eval = F}
+#Hesitations in writing have potential disadvantages.
 Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
@@
 @ronneberg2022process coined this possibility the "process-disruption hypothesis".
+```
 
-In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts.
+In language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental.
+Similar for keystroke data @roeser2021modelling demonstrated -- using a similar approach as used in the present paper^[For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.] -- how models that assume this mixture of two underlying data generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g.
+“dkdkdkdk”).
+As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
+However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes.
+Typing hesitations in the context of copy-typing may – to some extent – reflect not just production but also, and possibly largely, visual encoding.
+Text composition, in contrast, provides data from the free generation of words and full sentences that is not dependent on a visually presented text stimulus.
+For these reasons our research extends the analysis of keystroke data using a mixture model approach to data from spontaneous text production.
+
+In the study that is presented in this paper we test using Bayesian hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts.
 On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].
 
-Data form spontaneous text production [see @gernsbacher1995coherence] are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production.
-Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy [@dux2022automating; @dux2021effect; @vandermeulen2023writing].
-
-Second, since the emergence of software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing [@conijn2021timecourse].
-Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.
-These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning [@conijn2019understanding; @conijn2020keys].
-
-Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments [@not10; @dam09; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual].
-In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks.
-Instead text composition provides data from both inter and intra-sentence transitions.
-Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).
-
 The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations.
+In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word locations compared to within-word locations.
 This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing.
+The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not (always) pause to plan their text but plan in parallel to writing.
 In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.
+Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occur to some extent probabilistically across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
+We hypothesise that the statistical model of the parallel cascading view provides a better out-of-sample generalisation than the implementation of the serial model of writing in the context of unconstrained spontaneous text composition.

### 2024-04-28 tutorial/tutorial.Rmd
manuscript update

@@
 
-For example, a mean of 5 msecs on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)`. This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).
+For example, a mean of 5 msecs on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)` msecs on a linear scale.
+This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).
 
@@
                values_to = "loglik") %>%
-  group_by(obs) %>%
-  summarise(M = mean(loglik), var = var(loglik)) %>%
-  filter(M > -10) %>% # there are some terrible ones for this model :)
+  summarise(M = mean(loglik), var = var(loglik), .by = obs) %>%
+  filter(M > -10) %>% # there are some terrible ones for this model 
   mutate(obs = as.numeric(str_remove(obs, "V"))) %>%
-  ggplot(aes(x = obs, y = M, ymin = M-var, ymax = M+var)) +
+  ggplot(aes(x = obs, y = M, ymin = M - var, ymax = M + var)) +
   geom_pointrange(fatten = .5) +
@@
                values_to = "loglik") %>%
-  group_by(obs) %>%
   summarise(mean_loglik = mean(exp(loglik)),
-            var_loglik = var(loglik)) %>% 
-  ungroup() %>%
+            var_loglik = var(loglik),
+            .by = obs) %>% 
   summarise(elpd_loo = sum(log(mean_loglik)), # sum of mean log lik
@@
   pivot_longer(beta:theta, names_to = "parameter") %>%
-  group_by(location) %>% 
-  mutate(id = row_number()) %>%
+  mutate(id = row_number(), .by = location) %>%
   pivot_wider(names_from = location, values_from = value) %>%
@@
   pivot_longer(starts_with("diff")) %>% 
-  group_by(parameter, name) %>%      # group by parameters, comparison
   summarise(across(value, 
-            list(mean = base::mean,  # mean difference
+            list(mean = base::mean,  # mean of difference
                  lower = lower,      # 2.5% lower bound of difference
@@
                  p_diff = p_diff),   # prob that difference is negative
-            .names = "{.fn}")) %>% 
-  ungroup() %>% 
+            .names = "{.fn}"),
+            .by = c(parameter, name)) %>%   # group by parameters, comparison 
   mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
@@
   pivot_longer(-parameter) %>% 
-  group_by(parameter, name) %>%
   summarise(across(value, 
@@
                  p_diff = p_diff),   
-            .names = "{.fn}")) %>% 
-  ungroup() %>% 
-  mutate(across(where(is.numeric), round, 2),
-         parameter = gsub("\\[diff\\]", "", parameter)) 
+            .names = "{.fn}"),
+            .by = c(parameter, name)) %>% 
+  mutate(across(where(is.numeric), ~round(., 2)),
+         across(parameter, ~str_remove(., "\\[diff\\]"))) 
 ```
@@
   # create different columns for location
-  group_by(location) %>% 
-  mutate(id = row_number()) %>%
+  mutate(id = row_number(), .by = location) %>%
   pivot_wider(names_from = location, values_from = value) %>% 
@@
   pivot_longer(-participant, names_to = "location") %>%
-  group_by(location, participant) %>%
   summarise(across(value, list(mean = base::mean, 
                                lower = lower, 
-                               upper = upper), .names = '{.fn}')) %>% 
-  ungroup()
+                               upper = upper),
+                   .names = '{.fn}'),
+            .by = c(location, participant))
 
@@
                values_to = 'diff') %>% 
-  group_by(participant, difference) %>%
   summarise(across(diff, list(mean = base::mean, 
                                lower = lower, 
-                               upper = upper), .names = '{.fn}')) %>% 
-  ungroup()
+                               upper = upper),
+                   .names = '{.fn}'),
+            .by = c(participant, difference))
 

### 2024-05-02 report/manuscript.Rmd
update tutorial final for now

@@
 
-csl               : "apa.csl"
+csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
+documentclass     : "apa7"
+output            : 
+  papaja::apa6_pdf:
+    keep_tex: FALSE
 
@@
     affiliation   : "2"
+  - name          : "Gunn Helen Ofstad"
+    affiliation   : "3"
   - name          : "Evgeny Chukharev"
-    affiliation   : "3"
+    affiliation   : "4"
 
@@
     institution   : "Department of Psychology, Nottingham Trent University, United Kingdom"
-
   - id            : "2"
     institution   : "Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands"
-
   - id            : "3"
+    institution   : "Educational Sciences and Humanities, University of Stavanger, Norway"
+  - id            : "4"
     institution   : "Department of English, Iowa State University, Iowa"
@@
 
-keywords: "Cascading parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"
+keywords: "Cascading semi-parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"
 
 bibliography      : ["references.bib"]
-
-documentclass     : "apa7"
-classoption       : "man"
-output            : 
-  papaja::apa6_pdf:
-    keep_tex: TRUE
 
@@
   - \DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
+  - |
+    \makeatletter
+    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
+      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+      {-1em}%
+      {\normalfont\normalsize\bfseries\typesectitle}}
+
+    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
+      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+      {-\z@\relax}%
+      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
+    \makeatother
+
 ---
@@
 
-```{=html}
 <!-- TODO: 
-- address difference between posteriors of shift analysis for constrained and unconstrained model?
-- pausing frequency for words and sentences are equal but more importantly pauses occure less than 80% of the time indicating parallel planing.
 - would be really good, on some point, to distinguish between open and closed class words
 -->
-```
 
@@
 
-```{=html}
 <!--
 
-----
 Add to OSF
 Work on tutorial and add as link to methods
-----
-
-~~~~~
 Say how the results don't contradic the reported studies but speaks better to their conclusions
-but the modelling frameword doesn't require us to impose pause thresholds
-~~~~~
-
-~~~~
-Move from results to discussion?
-Open: to what extent are even short latencies longer at sentences.
-This is obvious for when multiple keystrokes are summed across at sentences boundaries as there are typically three key transitions necessary to start a sentence: (1) .^_ , (2) _^[shift], (3) [shift]^C
-We believe the second one is the linguistically meaningful one
-I think give Davids papers as examples, if he uses sums instead of means (other studies?)
-
-~~~
 
@@
 
-~~~~
-I think the serial account isn't all inconsistent with the mixture model.
-I don't think the serial account says that pauses must occure on linguistic edges but between bursts which can in principlpe happen anywhere (but why would they be random).
-Obviously the problem here is that one needs to postulate what a pause is instead of leaving it to the model to work out.
-The serial account and the cascading account make very similar predictions for inexperienced writers.
-
-Maybe distinguih between a strong version of serial model and a weak version that is basically a chunky cacading model
-
 -->
-```
 
@@
 
-\setlength{\parindent}{-0.5in}
+<!-- \setlength{\parindent}{-0.5in} -->
 
-\setlength{\leftskip}{0.5in}
+<!-- \setlength{\leftskip}{0.5in} -->
 
-<div id = "references"></div>
+<div id = "refs"></div>
 

### 2024-05-02 tutorial/tutorial.Rmd
update tutorial final for now

@@
 library(knitr)
+library(patchwork)
 library(tidyverse)
@@
 
+The choice of priors here is weakly informative.
+The intention is to use knowledge that we generally have about how long key intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values.
+Even if we know very little about the duration of keystroke interval, we do know that they shouldn't be longer than a day, or event an hour, or several minutes, because such long intervals are unlikely to reflect active writing.
+
 ### Sampling parameters
@@
 
+Before you save the posterior, create a new directory called "stan" where we will save all Stan outputs in this tutorial.
+
+```{r eval = F}
+# Create an empty directory called "stan"
+dir.create("stan")
+```
+
 ```{r eval = F}
@@
 
-The priors default priors and the priors we are going to set are now no longer on the msecs scale but on the log scale.
+The default priors and the priors that we need to specity are now no longer on the msecs scale but on the log scale.
 
@@
 
-For example, a mean of 5 msecs on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)` msecs on a linear scale.
+For example, a mean of 5 on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)` msecs on a linear scale.
 This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).
@@
 
-## Model comparison
-
-To evaluate the predictive performance of the models we can use leave-one-out cross-validation.
-The out-of-sample predictive performance was determined via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
-Cross-validation, in contrast to more conventional model-comparison metrics such as $R^2$, prevent overfitting by penalising models with more parameters [see @mcelreath2016statistical; @lambert2018student].
-
-Cross validation generally is measuring the performance of a model, meaning its ability to capture new data.
+These can be converted into msecs using the exponential function `exp`.
+
+```{r}
+fixef(fit_lgaus)[,-2] %>% 
+  exp() %>% round(0)
+```
+
+The posterior estimates obtained from the log-Gaussian model are radically different (larger) from the posterior estimates of the Gaussian model which is likely due.
+This difference is likely due to the missing lower bound at zero in the Gaussian model as we as its inability to give less emphasis to values in the right tail.
+
+```{r}
+fixef(fit_gaus)[,-2] %>% 
+  round(0)
+```
+
+# Model comparison I
+
+To evaluate the out-of-sample predictive performance of the models we can use leave-one-out cross-validation via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
+Cross-validation is a alternative to more conventional model-comparison metrics such as $R^2$ as it penalises models with more parameters to prevent overfitting [see @mcelreath2016statistical; @lambert2018student].
 Overfitting happens when a model is capturing nuances of a sample, rather than patterns one would expect to repeat in new samples.
+
+In contrast, cross-validation is measuring the performance of a model, meaning its ability to generalise to new data.
 The leave-one-out information criterion (LOO-IC) is determined as follows: train a model on $N-1$ observations, predict the remaining data points from the training model, repeat process $N$ times to predict every observation from a model of th remaining data.
-Adding up the prediction results gives an estimate of the expected log predictive density (`elpd_loo`; $\widehat{elpd}$), i.e.
+Adding up the prediction results gives an estimate of the expected log-predictive density (`elpd_loo`; $\widehat{elpd}$), i.e.
 an approximation of the results that would be expected for new data.
 
-The function `loo` can be used to access the expected log predictive density which uses probability calculations for approximate LOO-IC vio Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical].
-
-Approximation involved in `loo` uses the log posterior predictive densities: how likely is each data point given the distribution parameter estimates?
+The function `loo` can be used to access the expected log-predictive density which uses probability calculations to approximate LOO-IC via Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical].
+LOO can be approximated with the log posterior predictive densities as a measure of how likely a data point is, given the parameter estimates.
 This is shown for each data point in Figure \@ref(fig:loglik).
-Extremely small values indicate that an observation is very unlike for a given set of parameter values.
-
-```{r loglik, fig.cap="Log-posterior densities of each data point"}
+Extremely small values indicate that an observation is very unlike under the model's parameter values and we are unlike to observe them in a new sample.
+
+```{r loglik, fig.cap="Log-posterior densities of each data point (errorbars show variance) of the Gaussian model."}
 log_lik(fit_gaus) %>% 
   as_tibble() %>% 
-  mutate(sample = 1:n()) %>%
-  pivot_longer(-sample, 
+  pivot_longer(cols = everything(),
                names_to = "obs", 
                values_to = "loglik") %>%
-  summarise(M = mean(loglik), var = var(loglik), .by = obs) %>%
-  filter(M > -10) %>% # there are some terrible ones for this model 
+  summarise(mean = mean(loglik), var = var(loglik), .by = obs) %>%
+  filter(mean > -10) %>% # there are some terrible ones for this model 
   mutate(obs = as.numeric(str_remove(obs, "V"))) %>%
-  ggplot(aes(x = obs, y = M, ymin = M - var, ymax = M + var)) +
+  ggplot(aes(x = obs, y = mean, ymin = mean - var, ymax = mean + var)) +
   geom_pointrange(fatten = .5) +
-  labs(x = "Observations", 
-       y = "log-posterior predictive density (with variance)")
-```
-
-`elpd_loo` is the sum of the means and returns the expected log predictive density.
-`p_loo` is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters).
-`looic`, or LOO-IC, corresponds to $-2 \cdot ($`elpd_loo`$-$`p_loo`$)$ (for deviance scale).
-This can be manually estimated
+  labs(x = "observations", 
+       y = "log-posterior predictive density")
+```
+
+If we sum over those means we obtain the $\widehat{elpd}$, the expected log predictive density, indicated as `elpd_loo`.
+$p_\text{loo}$ (`p_loo`) is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters).
+`looic`, or LOO-IC, corresponds to $-2 \times ($`elpd_loo`$-$`p_loo`$)$ (for the deviance scale).
+To illustrate these, we can calculate them manually
 
@@
   as_tibble() %>% 
-  mutate(sample = 1:n()) %>%
-  pivot_longer(-sample, 
+  pivot_longer(cols = everything(),
                names_to = "obs", 
                values_to = "loglik") %>%
-  summarise(mean_loglik = mean(exp(loglik)),
-            var_loglik = var(loglik),
-            .by = obs) %>% 
-  summarise(elpd_loo = sum(log(mean_loglik)), # sum of mean log lik
-            p_loo = sum(var_loglik), # effective number of parameters
+  summarise(mean = mean(loglik), var = var(loglik), .by = obs) %>% 
+  summarise(elpd_loo = sum(mean), # sum of mean log lik
+            p_loo = sum(var), # effective number of parameters
             looic = -2 * (elpd_loo - p_loo)) %>% # Bayesian deviance
-  pivot_longer(everything(), names_to = "Estimate") %>% 
+  pivot_longer(everything()) %>% 
   mutate(across(value, round, 1))
@@
 
-or directly calculated using `loo`. The Pareto k diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model.
+or directly using `loo` (a small difference isn't unexpected).
 
@@
 ```
+
+The Pareto *k* diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model.
 
@@
 ```
-The model comparison shows that the model `fit_lgaus` (the model with the log-Gaussian likelihood) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` times better than `fit_gaus` (try `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])`).
-
-Aside, `elpd_loo` and `looic` have rarely got a direct interpretation.
+
+The model comparison shows that the model `fit_lgaus` (the model with the log-Gaussian likelihood) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors better than `fit_gaus`.
+This can be calculated by `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])` by normalising the difference over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, which returns the *z* or *t* value of the difference between models [@sivula2020uncertainty].
+A difference of `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors is substantial.
+By comparison, a *t*-value of more extreme than $\mid2\mid$ corresponds to a difference that is significant at a $\alpha$-level of 0.05 (i.e.
+$p<0.05$) [e.g.
+@baa08book].
+
+Aside, unlike $R^2$, `elpd_loo` and `looic` have no direct interpretation.
+This is because the size of `elpd_loo` increases as a function of the number of observations.
+Generally you can think of `elpd_loo` and `looic` as how many possible futures the model can rule out.
 Important is the difference between models.
-The size of the values is increasing as a function of the number of observations.
-Generally `elpd_loo` and `looic` measures how many possible futures the model can rule out.
-
-In contrast, `loo_R2` provides a Bayesian $R^2$ that measures the variability in the data that the model can captures.
-This quantity is not ideal for model comparisons, as normal $R^2$, because it is subject to overfitting: models with more parameters tend to have a larger $R^2$.
-A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; chapter 6 in @mcelreath2016statistical].
+
+In contrast, `loo_R2` provides a Bayesian $R^2$ that measures the variance in the data that the model can capture.
+This quantity is not ideal for model comparisons because, similar to the frequentist $R^2$, it is subject to overfitting: a model with more parameters can never have a smaller $R^2$, and $R^2$ will innevitabily increase when we add more parameters to a model.
+A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; also chapter 6 in @mcelreath2016statistical].
 The is an example for Bayesian $R^2$ for the single-distribution log-Gaussian model.
@@
 
-The model returns population estimates for the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the frequency of hesitant transitions ($\theta$) for each level of a categorical predictor after taking into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.
+The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the probability of hesitant transitions ($\theta$).
+In the model below, these are estimated each level of a categorical predictor `location` for the $i^\text{th}$ keystroke interval.
 
@@
 \begin{align}
-\text{iki}_{i} \sim & \theta_\text{location[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{iki}_{i} \sim & \theta_\text{location[i],participant[i]} \times \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i],participant[i]}) \times \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
 		&\text{where}\\
-		&\delta \sim \text{N}(0,1)\\
+		&\delta \sim \mathcal{N}(0,1)\\
 		&\text{constraint: } \delta > 0
@@
 
+Model takes into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor and is distribution $u_\text{participant} \sim \mathcal{N}(0, \sigma^2_p)$; (2) each participant has in individual hesitation probability $\theta_\text{participant}$ that differs across levels of the categorical predictor.
+
 The parameters of this equation are visualised in Figure \@ref(fig:mixturemodel).
-The illustrate shows the dissociation of two underlying processes that the model is looking for in the data.
-There are two distributions of which the left one in grey is the distribution of fluent keystroke intervals; the right distribution in green is wider and captures generally longer keystroke intervals (with some overlap between fluent and disfluent transitions).
+The illustrate shows the dissociation of two underlying processes that the model is estimating on the basis of the data.
+There are two distributions of which the left one in grey is the distribution of fluent keystroke intervals; the right distribution in green is wider and captures longer keystroke intervals (with some overlap between fluent and disfluent transitions).
 
@@
 The parameters of the mixture model explained visually.
-Each of the three parameters is determined for each transition location."}
+Each of the three parameters can determined for each transition location."}
 
@@
                 args = list(alpha2[1], sigmap[1], lam = theta[1])) +
-  annotate("label", x=4,y=.6, label="beta*':'~'Quick typing'", size = 4, parse = T) +
+  annotate("label", x=4,y=.6, label="beta*':'~'Typing speed'", size = 4, parse = T) +
   annotate("label", x=7.25,y=.35, label="atop(theta*':'~'Hesitation probability', '\"Pausing\"')", parse = T, size = 4) +
@@
 
-The average typing speed of fluent keystroke transitions is captured by the $\beta$ parameter.
-The distance between fluent transitions and disfluent transitions is captured by $\delta$ (i.e.
-the distance between the grey and the green distribution).
-The hesitation probability or the "pausing frequency" is the relative height of the grey and the green distribution and captured by $\theta$.
-Most keystrokes are, typically, fluent transitions and hence in the grey distribution.
+Importantly Figure \@ref(fig:mixturemodel) illustrates the average typing speed of fluent keystroke transitions captured by the $\beta$ parameter; the distance between fluent transitions and disfluent transitions captured by $\delta$ (i.e.
+the distance between the grey and the green distribution); the hesitation probability or the "pausing frequency" which is the height of the green distribution captured by $\theta$.
+Most keystrokes are, typically, fluent transitions and hence part of the grey distribution.
 The frequency of the minority of disfluent transitions is captured by the height of the green distribution.
-As this model assumes a finite mixture of two distributions, the probability of fluent transitions is just $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions was $.6$.
-If the pausing probability decreases to $.3$ the proportion of fluent transitions had increased to $.7$.
-
-## Model fit
+As this model assumes a finite mixture of two distributions, the probability of fluent transitions is the inverse $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions is $.6$.
+If the pausing probability decreases to $.3$ the proportion of fluent transitions had increased to $.7$ etc.
+
+## Model input values
 
@@
 The information required from the data is determined in the Stan code and has to do with what we need to estimate the model parameters and how the model is implemented.
-The exert from the Stan code below shows which information is expected as input (e.g.
-int is integer), how they are named, and what the smallest (i.e.
-lower) and largest possible value (i.e.
-upper) are.
+The exert from the Stan code below shows the information that is expected as input, which data type is expected (e.g.
+`int` means integer), their names, and what the smallest (i.e.
+`lower`) and largest possible value (i.e.
+`upper`) are.
 The full Stan code can be found at the end of this tutorial.
@@
 
-Depending on the data we might have different numbers of participants and conditions.
-The Stan code is using indices to be able to fit varying numbers for each.
-In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices.
-We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `location` for each transition locations (levels: before-sentence = 1, before-word = 2, within-word = 3).
-The returned values are indices for the parameters in the model: For example, for the location identifier, `beta[1]` is the population estimate for non-hesitant typing intervals at before-sentence transition locations, `beta[2]` is same for before-word transition locations; `theta[1]` and `theta[2]` are the disfluency probability on the logit scale for before-sentence and before-word transition locations, respectively.
+Participants and conditions are expected to represent indices in the Stan code.
+In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices for participants and conditions.
+We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before-sentence = 1, before-word = 2, within-word = 3).
+The returned values are indices for the parameters in the model: For example, for the location identifier, `beta[1]` is the population estimate for non-hesitant typing intervals at before-sentence transition locations, `beta[2]` for before-word transition locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.
 
@@
 
-## Load model
-
-The Stan code can be downloaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/blob/main/tutorial/stan/molg.stan) using `readLines` and `writeLines`.
-The code will be saved as "molg.stan" (mixture of log-Gaussians) in a directory called "stan" which needs to be available in your current working directory.
+## Loading the model
+
+The Stan code can be downloaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/blob/main/tutorial/stan/molg.stan) using `readLines` and saved locally using `writeLines` as "molg.stan" (mixture of log-Gaussians) in the directory "stan".
 
 ```{r eval = F}
-# Create an empty directory called "stan"
-dir.create("stan")
 # Download Stan code
 file <- readLines("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/stan/molg.stan")
-# Save Stan code to current working directory
+# Save Stan code in "stan" directory
 writeLines(file, "stan/molg.stan")
@@
 
-The Stan model can be loaded and translated into an S4 class object using the `stan_model` function and assign it to the object `molg`.
+The Stan model can be loaded and translated into an S4 class object using the `stan_model` function and assigned to `molg`.
 The "molg.stan" file needs to be available in the directory "stan" located inside the current working directory.
@@
 This code fits the posterior of a categorical predictor with any number of levels.
-In other words, the posterior that model returns can be used to calculate simple effects, main effects and interactions of a factorial design.
-We demonstrate below how to calculate a simple effect.
+In other words, the posterior that the model returns can be used to calculate simple differences, main effects and interactions of factorial designs.
+We demonstrate below how to calculate a simple difference.
 The same logic can be used to calculate main effects and interactions.
-Also, participant estimates are calculated for both mixture components and the mixing proportions by condition.
-For including more predictors, or removing, e.g., random error terms, or adjusting priors, the user will have to work directly in the Stan code.
-
-This Stan code used is largely based on @sorensen2016bayesian and @vasishth2017. @sorensen2016bayesian presents a detailed tutorial on how to fit Stan models [see also @lambert2018student].
-
-The model is fitting a number of parameters that are not relevant for our inference.
-To reduce the size of the R object that contains the posterior we can select the parameters of interest and disregard all parameters that are not relevant for our inference.
+Also, participant estimates are calculated for both mixture components and the hesitation probability by location.
+For including more predictors, or removing, e.g., random error terms, or adjusting priors, the Stan code can be changed by opening it in R or a text editor.
+The Stan code used is largely based on @sorensen2016bayesian and @vasishth2017.
+@sorensen2016bayesian presents a detailed tutorial on how to write models in Stan [see also @lambert2018student].
+
+The model is fitting a number of parameters of which not all are interesting for inference.
+To reduce the size of the model object that contains the posterior we can determine which parameters should be saved and disregard all parameters.
 
@@
 
-The `sampling` function applies the model to the data using the information specified above (iterations, warmup, chains, cores, parameters `pars` to be saved).
-`save_warmup` is set to `FALSE` to discard the warmup samples (which are not used for inference anyway) to reduce the size of the posterior.
-To allow reproducibility (and because Bayesian models involves random number generation) we set the seed.
-The seed can be any number but using the same number ensures the use of the same random number.
-Lastly, the control argument was specified with higher values for `adapt_delta` and `max_treedepth`: using higher values here mean the model runs slower but supports more careful parameter estimation.
-The model is assigned to the variable `m`.
+The `sampling` function applies the model to the data using the information introduced before (iterations, warmup, chains, cores, parameters `pars` to be saved).
+`save_warmup` is set to `FALSE` to discard the warmup samples (which are not used for inference anyway) to reduce the size of the model object.
+To allow reproducibility (and because Bayesian models involves random number generation) we set the seed using an arbitrary number.
+Lastly, the control argument was specified with higher values for `adapt_delta` and `max_treedepth`: using higher values here means the model runs slower but means more careful parameter estimation.
+The model is assigned to `fit_molg`.
 
@@
 
-Save the results using `saveRDS`.
+As above we save the results using `saveRDS`.
 
@@
 Running this model will take a while to complete sampling depending on your hardware specifications.
-The time it took my machine to complete this job can me viewed using `get_elapsed_time`.
-Therefore it is worth to not use all cores for this to run or to use a dedicated high performance machine.
-
-```{r}
-get_elapsed_time(fit_molg)/60 # in mins
+The time it took my laptop to complete this job can me viewed using `get_elapsed_time`.
+It is worth to not use all cores to run this model or to use a dedicated high performance machine.
+
+```{r}
+get_elapsed_time(fit_molg) / 60 # in mins
 ```
@@
 Inferred posterior estimates shown as average and 95% probability interval can be extracted using the `print` function for each transition location and the three parameter values of interest.
+`prob` is the $\theta$ parameter on the probability scale rather than on the logit scale.
+$\beta$ and $\delta$ are on the log scale.
 
@@
 
-## Model comparsion
-
-Similar to above we will compare the three models -- the mixture of log-Gaussians, and the uni-model Gaussian and log-Gaussian models -- using leave-one-out cross-validation.
-Because the mixture model was fitted directly in Stan and the uni-model models were fitted in `brms`, we need to extract the fit statistics before comparing the models using functions provided by the package `loo`.
+We will return to the interpretation of these estimates below.
+
+## Model diagnostics
+
+### Convergence
+
+Before we interpret these result we need to ensure that the model reached convergence.
+Convergence can be established using two simple techniques.
+First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain in different colours.
+We will look at the population-level parameters saved in `pars`.
+
+```{r}
+pars <- c("beta", "delta", "theta")
+```
+
+To create trace plots, we can apply the `stan_trace` function to the model `fit_molg` and extract the MCMC chains for the parameters in `pars`.
+The `alpha` argument makes the colours slightly more transparent otherwise it is difficult to see the chains hiding in the back.
+If the chains overlap and look like "fat hairy caterpillars", chains have converged on the same target distribution.
+The chains below look good except for one divergent transition for `beta[3]`.
+
+```{r}
+# Check convergence
+stan_trace(fit_molg, pars = pars, alpha = .5)
+```
+
+Second, we can calculate the $\hat{R}$ statistic.
+Successful convergence is reflected in $\hat{R}$ values smaller than 1.05 [@gelman1992].
+$\hat{R}$ is similar to the *F* statistic in standard analysis of variance: it tells us how much bigger the variability between chains is compared to the variability within all chains.
+A value of $\approx 1$ indicates that the variability is essentially identical signifying convergence.
+Larger $\hat{R}$ values indicate that the chains are diverging.
+The `rhat` function applied to the model and the parameters of interest extracts the $\hat{R}$ statistic.
+There is no evidence for convergence problems.
+
+```{r}
+summary(fit_molg, pars=pars)$summary %>% 
+  as.data.frame() %>% 
+  rownames_to_column("parameter") %>% 
+  mutate(across(Rhat, round, 3)) %>% 
+  select(parameter, Rhat)
+```
+
+Convergence problems can have many reasons and therefore many solutions: for example, running longer chains (in particular longer warmups), increasing the maximum treedepth and the average acceptance probability (`adapt_delta`), specifying starting values, using sensible regulating priors, constraining parameters, or changing the parametrisation of the model.
+Severe convergence problems indicate model misspecifications.
+
+### Posterior predictive check
+
+```{r echo = F, include = F}
+(total_samples <- (iterations - warmup) * nchains)
+```
+
+The model is simulating hypothetical data sets on the basis of the estimated posterior parameter values.
+This happened for every iteration for every chains.
+In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) available to us because
+
+```{r}
+(total_samples <- (iterations - warmup) * nchains)
+```
+
+We can draw predicted data from the model output and compare these to the observed data. A model that makes reasonable predictions should fit the observed data.
+
+Using the `as.matrix` function we extract a matrix of `y_tilde`, the simulated data.
+This returns a matrix of the size `total_samples` $\times$ `nrow(data)` (so `r scales::comma(total_samples)` $\times$ `r scales::comma(nrow(data))`).
+We use the `sample` function to draw `N` randomly sampled hypothetical data sets.
+
+```{r}
+y_tilde <- as.matrix(fit_molg, pars = "y_tilde")   # extract simulated data sets
+N <- 50                                            # number of simulations to use
+rnd_sims <- sample(total_samples, N)               # created random indices
+y_tilde_sample <- y_tilde[rnd_sims,]               # draw N random simulations 
+```
+
+The `ppc_dens_oberlay` function from the `bayesplot` package is then mapping the observed data ($y$; thick blue line) to the simulated data ($y_{rep}$; thin light blue lines).
+We find that the simulated data sets closely align with the observed data.
+
+```{r}
+library(bayesplot)
+ppc_dens_overlay(data_list$y, y_tilde_sample) +
+  scale_x_continuous(limits = c(0, 3000))
+```
+
+For comparison we also visualised the fit for the simulated data of the Gaussian and the log-Gaussian model in Figure \@ref(fig:fitother).
+Both models show a substantially poorer fit to the keystroke data than the mixture model-based simulations.
+Note that the Gaussian model does not rule out negative values which are conceptually implausible.
+
+```{r fitother, echo=FALSE, fig.cap="Fit of simulated datasets compared to observed data for the Gaussian and the log-Gaussian model. x-axes were truncated for visability."}
+y_tilde <- posterior_predict(fit_gaus, ndraws = N)
+p_gaus <- ppc_dens_overlay(fit_gaus$data$iki, y_tilde) +
+  labs(subtitle = "Gaussian model") +
+  scale_x_continuous(limits = c(-1000, 3000))
+
+y_tilde <- posterior_predict(fit_lgaus, ndraws = N)
+p_lgaus <- ppc_dens_overlay(fit_lgaus$data$iki, y_tilde) +
+  labs(subtitle = "log-Gaussian model") +
+  scale_x_continuous(limits = c(0, 3000))
+
+p_gaus + p_lgaus +
+  plot_layout(guides = "collect") 
+```
+
+# Model comparsion II
+
+Similar to the section "Model comparison I" we will use leave-one-out cross-validation to compare the three models, the mixture of log-Gaussians, and the single distribution Gaussian and log-Gaussian models.
+Because the mixture model was fitted directly in Stan and the single-distribution models were fitted in `brms`, we need to extract LOO statistics before for which we need the log-likelihood and the relative effective sample size for the Stan model.
+We can then go ahead and compare the models using functions provided by the `loo` package .
 
@@
 
-As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to each the single distribution Gaussian and log-Gaussian model.
-Their difference is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
-Models are ordered from the model with the highest predictive performance to the lowest (i.e.
-`elpd_loo`).
+As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to the single distribution Gaussian and log-Gaussian models.
+This time we need to use the `loo_compare` function because we had to extract the LOO statistics.
 
@@
 
-To extract the model comparisons and fit statistics:
-
-```{r}
-loo_comp %>% as.data.frame() %>% 
+We can then extract the model comparisons and fit statistics:
+
+```{r}
+loo_comp %>% 
+  as.data.frame() %>% 
   rownames_to_column("models") %>% 
@@
 
-The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian ) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` times better than `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`).
-
-## Model diagnostics
-
-### Convergence
-
-Model convergence can be established in two practically simple techniques.
-First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain.
-We will look at the population-level parameters (names are determined in the Stan code) and assign those to `pars`.
-
-```{r}
-pars <- c("beta", "delta", "theta")
-```
-
-To create trace plots, we can apply the `stan_trace` function to the model `m` and extract the MCMC chains for the parameters in `pars`.
-The `alpha` argument makes the colours slightly more transparent.
-If the chains overlap and look like "fat hairy caterpillars", chains have converged on the same target distribution.
-
-```{r}
-# Check convergence
-stan_trace(fit_molg, pars = pars, alpha = .5)
-```
-
-Second, we can calculate the $\hat{R}$ statistic.
-Successful convergence is reflected in $\hat{R}$ values smaller than 1.05 [@gelman1992].
-$\hat{R}$ is similar to the F statistic: it tells us how much bigger the variability between chains is compared to the variability within chains.
-A value of $\approx 1$ indicates that the variability is essentially identical signifying convergence.
-For this we can use the `rhat` function applied to the model and the population parameters.
-
-```{r}
-summary(fit_molg, pars=pars)$summary %>% 
-  as.data.frame() %>% 
-  rownames_to_column("parameter") %>% 
-  mutate(across(Rhat, round, 3)) %>% 
-  select(parameter, Rhat)
-```
-
-Convergence problems can have many reasons and therefore many solutions: running longer chains (in particular longer warmups), increasing the maximum treedepth and the average acceptance probability (`adapt_delta`), specifying starting values, adjusting priors, constraining parameters, or changing the parametrisation of the model.
-Severe convergence problems are indicative of a misspecification of the model.
-
-### Posterior predictive check
-
-```{r echo = F, include = F}
-(total_samples <- (iterations - warmup) * nchains)
-```
-
-The model used the posterior parameter values to simulate hypothetical data sets.
-This happened for every iteration for every chains.
-In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) available to us because
-
-```{r}
-(total_samples <- (iterations - warmup) * nchains)
-```
-
-We can draw predicted data from the model output and compare these to the observed data. A model that makes reasonable predictions should fit the observed data.
-
-Using the `as.matrix` function we extract a matrix of `y_tilde`, the simulated data.
-This returns a matrix with the size `total_samples` $\times$ `nrow(data)` (so `r scales::comma(total_samples)` $\times$ `r scales::comma(nrow(data))`).
-We use the `sample` function to draw `N` randomly sampled hypothetical data sets.
-The `ppc_dens_oberlay` function from the `bayesplot` package is then mapping the observed data ($y$; thick blue line) to the simulated data ($y_{rep}$; thin lightblue lines).
-
-```{r}
-y_tilde <- as.matrix(fit_molg, pars = "y_tilde")   # extract simulated data sets
-N <- 50                                            # number of simulations to use
-rnd_sims <- sample(total_samples, N)               # created random indices
-y_tilde_sample <- y_tilde[rnd_sims,]               # draw N random simulations 
-```
-
-```{r}
-library(bayesplot)
-ppc_dens_overlay(data_list$y, y_tilde_sample) +
-  scale_x_continuous(limits = c(0, 3000))
-```
-
-# Posterior probability distributions
+The difference between the model in the top row and the two other models is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
+Models are ordered from the model with the highest predictive performance to the lowest (i.e.
+`elpd_loo`).
+The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian ) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` standard error better than the single distribution log-Gaussian `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`).
+This constitutes substantial support for the mixture model over the single distribution log-Gaussian.
+
+# Posterior probability distribution
 
 At the core of Bayesian inference is the posterior probability distribution.
-For each model parameter we have `r scales::comma(total_samples)` posterior samples that form a posterior probability distribution.
-These samples are the results of the number of iterations after warmup for all chains.
+For each model parameter we have `r scales::comma(total_samples)` posterior samples that render the posterior probability distribution.
+These samples are the result of the posterior sample from the target distribution after warmup across all chains.
+If the target distribution was not found by the end of the warmup, inference will not be reliable.
+We know whether the target distribution was determined if the chains converged.
 
@@
 
-The `names` function can be used to remind us of the parameter names used in the model (we reduced the output to the first 15 parameter names).
+The `names` function can be used to remind us of the parameter names available in the model (we omitted the log-likelihood `log_lik`, and the stimulated data sets `y_tilde`).
 The meaning of the parameters is described above and is determined in the Stan code.
@@
 
-## Parameter-value estimates
-
-The parameter estimates can be visualised using the `stan_hist` function with the model variable `m` and the to-be visualised parameters `pars` as arguments.
+The posterior probability distribution of the parameter estimates can be visualised using the `stan_hist` function with the model variable `fit_molg` and the to-be visualised parameters `pars` as arguments.
 The function returns a histogram for each other parameter values.
-Each histogram is the probability distribution of the parameter values indicating which values are more or less probable to be the true parameter value.
+Each histogram is the probability distribution of the parameter values indicating which values are more or less probable estimates for the true parameter value.
 
@@
 The posterior samples of the parameter values can be summarised using the `print` function.
-The `probs` argument requires the lower and upper bound of the probability interval that we are interested in.
+The `probs` argument can be used to extract the lower and upper bound of the probability interval.
 A lower bound of .025 and an upper bound of .975 gives the 95% probability interval (PI; also called "credible intervals"), i.e.
@@
 
+```{r}
+print(fit_molg, pars = pars, probs = c(.025,.975))
+```
+
 The output summaries the most probable parameter value as mean with its standard error (`se_mean`) and standard deviation (`sd`), the effective sample size (`n_eff`) indicating sampling efficiency and the convergence metric $\hat{R}$ (`Rhat`) we introduced above.
 
-```{r}
-print(fit_molg, pars = pars, probs = c(.025,.975))
-```
-
-For the following steps, we will focus on the three parameters that have conceptually interesting interpretations: (1) the average fluent typing speed $\beta$, (2) the disfluency slowdown $\delta$, and (3) the disfluency probability $\theta$.
-
-```{r}
-pars <- c("beta", "delta", "theta")
-```
+For the following steps, we will focus on the three parameters that have conceptually interesting interpretations: (1) the average fluent typing speed $\beta$, (2) the hesitation slowdown $\delta$, and (3) the hesitation probability $\theta$ (which at the moment is still on the logit scale).
 
@@
 The values for $\beta$ and $\delta$ are shown on a log-scale.
-To transform their values back to msecs we can extract the posterior samples using the `as.data.frame` function.
-We prefer the use of tibble objects.
+To transform their values back to msecs we can extract the posterior samples using the `as.data.frame` function (we prefer the use of tibbles).
 
@@
 
-The `pivot_longer` function is transforming the data above to a long format with an additional column for component and the model parameters kept as columns.
+The `pivot_longer` function transforms the data to a long format with an additional column for component and the model parameters as columns.
 The `names_pattern` argument is using a regular expression to extract the number in the squared brackets.
@@
 
-This code is then transforming `beta` and `delta` to msecs using the exponential function `exp` to un-log the values.
+This code is then transforming `beta` and `delta` to msecs using the exponential function `exp`.
 In order to transform the slowdown `delta` into msecs we need to add `beta` before using the exponential function; we can then subtract beta again.
-`theta` samples are currently on the logit scale and can be transformed to probabilities using `brms::inv_logit_scaled`.
-The `recode` function changes the component indices from 1 to "before-sentence", 2 to "before-word", 3 to "within-word".
+`theta` samples are currently on the logit scale and can be transformed to probabilities using `brms::inv_logit_scaled` (the model also contained a `prob` parameter which is `theta` but on the probability scale).
+The `recode` function changes the location indices from 1 to "before-sentence", 2 to "before-word", 3 to "within-word".
 
@@
 The posterior distribution of the parameter estimates can then be visualised in, for example, histograms.
-
-```{r}
-pivot_longer(posterior_in_msecs, beta:theta, names_to = "parameter") %>%
+The histograms show that especially the hesitation parameter is sensitive to transition location with more hesitations before sentences than before words than within words.
+
+```{r}
+pivot_longer(posterior_in_msecs, beta:theta) %>%
   ggplot(aes(x = value, colour = location, fill = location)) +
   geom_histogram(position = "identity", alpha = .25) +
-  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
+  facet_wrap(~name, scales = "free", labeller = label_parsed) +
   scale_fill_brewer("Transition\nlocation", palette = "Dark2") +
@@
 ```
-
-## Difference between transition locations
-
-We can calculate the differences between the transition locations for each parameter.
-This is giving an indication of whether fluent typing ($\beta$) is influenced by linguistic edges, maybe the size of the disfluency slowdown ($\delta$), or the disfluency probability ($\theta$).
-
-To determine these differences we change the data format above and create a variable that indicates the parameter and one column for the corresponding values for each transition location.
+```{r echo = F}
+rm("mean")
+```
+
+To illustrate differences to the estimated cells means compared to the single distribution models, here are the means for all parameters and their location.
+The central tendencies of the single distribution models are in principle a combination of `beta` and `beta + delta` weighted by `theta`.
+
+```{r}
+pivot_longer(posterior_in_msecs, beta:theta) %>% 
+  summarise(across(value, list(mean = mean), 
+                   .names = "{.col}"), 
+            .by = c(name, location)) %>% 
+  arrange(name)
+```
+
+These are the estimated cell means from the log-Gaussian model
+
+```{r}
+fixef(fit_lgaus)[,1] %>% exp() %>% round(0)
+```
+
+and the estimated cell means of the Gaussian model
+
+```{r}
+fixef(fit_gaus)[,1] %>% round(0)
+```
+
+Differences are especially noticeable for before-sentence transitions.
+
+# Hypothesis testing
+
+## Differences between transition locations
+
+We can calculate the differences between the transition locations for each parameter to determine to what extent each of the parameters are sensitive to transition locations.
+This will show us to what extent fluent typing ($\beta$), the size of the hesitation slowdown ($\delta$), and the hesitation probability ($\theta$) are sensitive to the linguistic location of a key pair.
+
+To determine these differences we change the data format above and create a column that indicates the parameter name and a column for the corresponding values for each transition location.
 The last line calculates the difference between (1) transitions before sentences and words, and (2) transitions before and within words for each of the three parameters.
@@
   select(-id) %>% # drop id column
-  mutate(diff_sent = `before-sentence` - `before-word`,
-         diff_word = `before-word` - `within-word`,
+  mutate(diff_sent = `before-sentence` - `before-word`, # compare sentences to word
+         diff_word = `before-word` - `within-word`, # compare within to before word
          parameter = paste0(parameter, "[diff]")) %>% 
@@
 
-The difference between the conditions can be summarised using the mean, the 95% PIs and the probability that the difference between the components is small 0 (indicated as e.g.
-$P(\hat{\beta} <0)$ for the parameter $\beta$; the hat $\hat{.}$ symbol indicates that the value is an estimate of the population parameter value).
-This summary tells us whether the difference between the consonants task and the LF-bigrams task is different from zero and the most probable value for the difference between tasks.
-
-```{r}
-# Create some summary functions
-lower <- function(x) quantile(x, .025)
-upper <- function(x) quantile(x, .975)
-p_diff <- function(x) mean(x < 0)
+The difference between the conditions can be summarised using the mean, the 95% PIs and the probability that the difference between the components is smaller than 0 (indicated as e.g.
+$P(\hat{\beta} <0)$ for the parameter $\beta$; the hat symbol indicates $\hat{.}$ that the value is an estimate of the population parameter value).
+This summary tells us how certainly we can rule out a null difference between transition locations the most probable value for the differences.
+
+```{r}
+# Create summary functions for convenience
+lower <- function(x) quantile(x, .025) # lower bound of 95% PI
+upper <- function(x) quantile(x, .975) # upper bound of 95% PI
+p_diff <- function(x) mean(x < 0)      # prob that difference is negative
 
@@
   summarise(across(value, 
-            list(mean = base::mean,  # mean of difference
+            list(mean = mean,        # mean of difference
                  lower = lower,      # 2.5% lower bound of difference
@@
 
-```{r echo=F}
+```{r echo = F}
+ests <- posterior_diffs %>%
+  pivot_longer(starts_with("diff")) %>% 
+  summarise(across(value, 
+            list(mean = mean,        # mean of difference
+                 lower = lower,      # 2.5% lower bound of difference
+                 upper = upper,      # 97.5% upper bound of difference
+                 p_diff = p_diff),   # prob that difference is negative
+            .names = "{.fn}"),
+            .by = c(parameter, name)) %>%   # group by parameters, comparison 
+  mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
+  arrange(name)
+
+theta <- ests %>% filter(str_detect(parameter, "theta"))
+delta <- ests %>% filter(str_detect(parameter, "delta"))
+```
+
+For example, form these results we can infer that there is a $P(\hat{\theta} <0)$ = `r theta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` chance that hesitations do not occur more frequently before sentence; in fact the results show hesitations are more probable my `r theta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(2)` to occur before sentences compare to before words (95% PI: `r theta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(2)`, `r theta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(2)`).
+Also, the slowdown associated with hesitations is `r delta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(0)` msecs longer for sentences (95% PI: `r delta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(0)`, `r delta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(0)`) with a probability of $P(\hat{\delta} <0)$ = `r delta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` to be smaller than zero.
+
+```{r echo = F}
 ps <- posterior_diffs %>%
@@
 These differences can be viewed in histograms.
-The vertical line indicates a difference between tasks corresponding to a value of zero.
-The area of the histogram to the left of this line corresponds to the probability `p` to observe a value smaller than zero as in the output above.
+The vertical line indicates a difference between transition locations corresponding to a value of zero.
+The area of the histogram to the left of this line corresponds to the probability `p_diff` to observe a negative difference.
 
@@
 
-We observe the following differences that can be written up in a report like this: We observe evidence for a higher probability to pause before sentences which is `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(upper)`) times more likely than before words, which was is `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(upper)`) more likely than within words.
-
-Similarly, differences were found for the hesitation difference.
-Hesitations before sentences are `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(upper)`) longer than before words, which were `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(upper)`) longer than within words.
-
-Interestingly though fluent transitions (i.e.
-planning and executing motor plans) were only faster within words by `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper)`) compared to before-word transitions but there was no difference for before-sentence transitions compared to before-word transitions ($\hat{\beta}$ = `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean)` msecs, 95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper)`).
+We observe the following differences that can be reported as followed: Evidence was found for a higher probability to hesitate before sentences with a difference of `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(upper)`) compared to before-word transitions; before-word transitions show an increased hesitation probability of `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(upper)`) compared to within-word transitions.
+
+Similarly, differences were found for the hesitation slowdown.
+Hesitations before sentences were `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(mean) %>% round(0)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(upper) %>% round(0)`) longer than before words.
+Hesitations before words were `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer than within words, although the probability interval (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(upper) %>% round(0)`) did not rule out a negative difference, or no difference, which however had a low probability $P(\hat{\delta} <0)$ = `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(p_diff) %>% round(2)`.
+
+For fluent transitions there was evidence that within-word transitions were `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer (95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper) %>% round(0)`) compared to before-word transitions.
+There was negligible evidence for longer transitions for before-sentence transitions compared to before-word transitions ($\hat{\beta}$ = `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(mean) %>% round(0)` msecs, 95% PI: `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(upper) %>% round(0)`).
 
@@
 
-In practice, we rarely believe that only an effect of exactly null is indicating that there is no difference between our groups (which is what a null hypothesis assumes).
-Instead there are values that are non-different from null; for example a slowdown of 1 msecs is meaningless in most, if not all, contexts.
-To take this into account we can determine the *region of practical equivalence* [ROPE, @kruschke2010believe; @kruschke2011bayesian; @kruschke2014doing].
+One problem with null-hypothesis significance testing it that, in practice, we rarely believe that only a difference of exactly null is indicating that there is no difference between groups.
+Instead there are values that are non-different from null; for example a difference in keystroke intervals of 1 msecs is meaningless in most, if not all, contexts.
+A hesitation probability increase of 0.001 is too small to be meaningful.
+To take this into account we can determine a *region of practical equivalence* [ROPE, @kruschke2010believe; @kruschke2011bayesian; @kruschke2014doing].
 We can define the ROPE as the range of values that is _a priori_ considered non-different from a null effect.
 The ROPE value indicates to what extent the posterior can not rule out a negligible effect.
-A meaningful difference between groups (or effects more generally) should have a small proportion of posterior samples within the ROPE.
-
-To calculate the ROPE for out differences between two transition locations, we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference.
-We can create such a vector from our posterior differences.
-Let's focus on $\beta_\text{diff}$, the average typing speed for fluent transitions.
+A meaningful difference between groups should have a small proportion of posterior samples within the ROPE.
+
+To calculate the ROPE for our differences between two transition locations, we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference.
+We can create such a vector from our posterior differences that we calculated above.
+Let's focus on $\delta_\text{diff}$, the slowdown difference for hesitant transitions.
 
@@
 # For sentence vs word
-beta_diff_sent <- filter(posterior_diffs, parameter == "beta[diff]") %>% pull(diff_sent)
+delta_diff_sent <- filter(posterior_diffs, parameter == "delta[diff]") %>% pull(diff_sent)
 
 # For before word vs within word
-beta_diff_word <- filter(posterior_diffs, parameter == "beta[diff]") %>% pull(diff_word)
-```
-
-Also, we need to define the `range` argument which specifies the range of values are not considerably different from null.
-Of course this range will depend on what the parameter represents and researcher intuition.
-For example, for fluent typing we would consider small differences, of say 10 msecs, more meaningful than for disfluencies.
+delta_diff_word <- filter(posterior_diffs, parameter == "delta[diff]") %>% pull(diff_word)
+```
+
+Also, we need to define the `range` argument which specifies the range of values are considered indifferent from a null difference.
+Of course this range will depend on what the parameter represents and researcher's intuition as well as prior knowledge of the domain.
+For example, for fluent transitions we would consider small differences, of say 10 msecs, more meaningful than for hesitations.
 If we consider differences of 10 msecs or smaller equivalent to a null difference, we can define the `range` argument with a lower bound of -10 and the upper bound of 10.
@@
 library(bayestestR)
-rope(beta_diff_sent, range = c(-10, 10))
-```
-
-The ROPE for fluent typing speed for the difference between before-sentence and before-word transitions contains `r round(pull(rope(beta_diff_sent, range = c(-10, 10)))*100,2)`% of the posterior samples.
-In other words, there is a `r round(pull(rope(beta_diff_sent, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.
-
-```{r}
-rope(beta_diff_word, range = c(-10, 10))
-```
-
-However, the ROPE for fluent typing speed for the difference between before-word and within-word transitions contains `r round(pull(rope(beta_diff_word, range = c(-10, 10)))*100,2)`% of the posterior samples.
-In other words, there is a `r round(pull(rope(beta_diff_word, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.
-
-Let's see what happens if we determine a wider ROPE for negligible differences.
-What happens if we define a ROPE of -50 through 50 msecs for the difference between before-word and within-word transitions.
-
-```{r}
-rope(beta_diff_word, range = c(-50, 50))
-```
-
-The ROPE contains `r round(pull(rope(beta_diff_word, range = c(-50, 50)))*100,2)`% of the posterior samples.
-In other words, there is a `r round(pull(rope(beta_diff_word, range = c(-50, 50)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -50 and 50 msecs which we considered negligible.
+rope(delta_diff_word, range = c(-10, 10))
+```
+
+The ROPE for hesitation durations for the difference between before-word and within-word transitions contains `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% of the posterior samples.
+In other words, there is a `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% chance that the true difference $\beta_\text{diff}$ is between -10 and 10 msecs and therefore negligible.
+
+Let's see what happens if we determine a wider ROPE for negligible differences which can be understood as a more conservative test for a non-zero difference.
+What happens if we define a ROPE of -50 through 50 msecs for the difference between before-word and within-word hesitations
+
+```{r}
+rope(delta_diff_word, range = c(-50, 50))
+```
+
+The ROPE contains `r round(pull(rope(delta_diff_word, range = c(-50, 50)))*100,2)`% of the posterior samples.
+In other words, there is a `r round(pull(rope(delta_diff_word, range = c(-50, 50)))*100,2)`% probability that the true difference $\beta_\text{diff}$ is between -50 and 50 msecs which would therefore be a negligible effect.
 
@@
 Frequentist null-hypothesis testing is based on the *p*-value.
-Conceptually this test is evaluates whether the data allow us to reject the null hypothesis (i.e.
+Conceptually this probability represents how likely the data are if there was no effect and we can, if this probability is low enough (typically smaller than .05) reject the null hypothesis (i.e.
 the true parameter value being some difference, is zero).
+In other words, if the data are very unlikely, the null-hypothesis becomes implausible.
 More specifically it tests whether the probability of observing a test statistic (e.g.
-*t*, *F*, $\chi^2$) as or more extreme if the null hypothesis were true.
-Therefore, *p*-values have two possible outcomes: reject the null (i.e.
+*t*, *F*, $\chi^2$) as or more extreme if the null hypothesis is true.
+There are two possible outcomes for *p*-values: reject the null (i.e.
 $p<.05$); inconclusive evidence (i.e.
@@
 And that's it.
-It does not provide direct evidence for the alternative hypothesis; nor does it provide evidence for the null hypothesis.
-
-Bayes Factors (BFs) provide a continuum of three possible outcomes [@dienes2014using; @dienes2016bayes; @dienes2018four]: (1) there is evidence to reject the null / alternative hypothesis; (2) the evidence is inconclusive (get more data!); (3) accept the null / alternative hypothesis.
-
-For example, if our BF is the probably of the alternative hypothesis $H1$ over the null hypothesis $H_0$, values $>1$ indicate that there is more evidence for $H1$ and $<1$ indicates more evidence for $H_0$.
+*p*-values do neither provide direct evidence for the alternative hypothesis (only evidence against the null), nor do they provide evidence for the null hypothesis (that data are not inconsistent with the null hypothesis is not evidence that there is no difference; or absentence of evidence is not evidence of absence).
+
+In contrast, Bayes Factors (BFs) provide a continuum of three possible outcomes [@dienes2014using; @dienes2016bayes; @dienes2018four]: (1) there is evidence to reject the null / alternative hypothesis; (2) the evidence is inconclusive (get more data!); (3) accept the null / alternative hypothesis.
+
+For example, if our BF is the probability of the alternative hypothesis $H_1$ over the null hypothesis $H_0$, given the data $y$, values $>1$ indicate that there is more evidence for $H_1$ than for $H_0$, given the data, and $<1$ indicates that there is more evidence for $H_0$.
 This can be turned around by swapping nominator and denominator
@@
 $$
-or by taking the inverse of $\text{BF}_{10}$, meaning $\text{BF}_{01} = \frac{1}{\text{BF}_{10}}$.
+or by taking the inverse of $\text{BF}_{10}$ which is $\text{BF}_{01} = \frac{1}{\text{BF}_{10}}$.
 
@@
 evidence vs no evidence).
-Authors have proposed different soft thresholds of what constitutes evidence [see, e.g., @bag12; @jeffreys1961theory; @lee2014bayesian] but generally a BF larger than 3 is weak evidence, larger than 5 / 6 indicate moderate evidence and 10 / 32 is strong evidence for a statistically meaningful effect.
+Authors have proposed different soft thresholds of what constitutes meaningful evidence [see, e.g., @bag12; @jeffreys1961theory; @lee2014bayesian] but generally a BF larger than 3 is weak evidence, larger than 5 / 6 indicate moderate evidence and 10 / 32 is strong evidence for a statistically meaningful effect.
 For example BF$_{10}$=2 reflects that the alternative hypothesis is two times more likely than the null hypothesis given the evidence.
 BF$_{10}$=0.33 is weak evidence in favour of the null hypothesis because $\frac{1}{0.33} \approx 3$.
-Notably BFs can be used to obtain evidence in favour of the null [for discussion see @dienes2014using; @dienes2018four; @dienes2016bayes].
+Notably, BFs can be used to obtain evidence in favour of the null; for discussion see @dienes2014using, @dienes2018four, @dienes2016bayes and for a practical application in writing research see @spilling2022handwriting and @spilling2023writing.
 
@@
 
-Figure \@ref(fig:sdbf) shows an example for the calculation of BFs based on a posterior with an effect difference that is distributed according to $\text{N}(2.5, 1)$ compared against three priors with a wide SD $\text{N}(0, 10)$, a SD equal to the posterior $\text{N}(0, 2.5)$, and relatively small SD $\text{N}(0, 1)$.
+Figure \@ref(fig:sdbf) shows an example for the calculation of BFs based on a posterior with an effect difference that is normally distributed with a mean of 2.5 and a SD of 1, expressed as $\mathcal{N}(2.5, 1)$, compared against three priors with a wide SD $\mathcal{N}(0, 10)$, a SD equal to the posterior $\mathcal{N}(0, 2.5)$, and relatively small SD $\mathcal{N}(0, 1)$.
+The resulting BFs are shown in the boxes as well as their calculation.
+The denominator shows the height of the posterior (black line) at zero and the nominator shows the height of the prior at zero (indicates as black dots).
+The BF is larger for the more informative prior.
+Of course in practice, the shape of the posterior density is not independent of the prior; for more informative priors the posterior will move closer to the prior.
 
@@
 
-To calculate the BF on the basis of posterior samples, the Savage-Dickey method can be implemented like so [taken from @nicenboim2016statistical] which requires an appropriate prior distribution.
-We will fix the mean of the prior to 0 but allow to change the SD as function argument.
+To calculate the BF on the basis of posterior samples, the Savage-Dickey method can be implemented as below [taken from @nicenboim2016statistical] which requires an appropriate prior distribution.
 
@@
 
-For demonstration purposes we will calculate the BF with different priors SDs for the hesitation probability $\theta$, or in particular the posterior difference in hesitation $\theta_\text{diff}$ for before-sentence vs before-word transitons and before-word vs within-word transitions.
+The `logspline` function estimates the density of the posterior and the `dlogspline` and the `dnorm` functions return the height of the density at 0.
+We fixed the mean of the prior to 0 but allow to change the SD as function argument.
+Aside, the mean of the prior does not need to be zero but can take on any value that can be justified on the basis of domain knowledge.
+We are illustrating here how BFs can be used to test the evidence again a null hypothesis.
+
+For demonstration purposes we will calculate the BF with different priors SDs for the hesitation probability $\theta$, or in particular the posterior difference in hesitation $\theta_\text{diff}$ for before-sentence vs before-word transitions and before-word vs within-word transitions.
 
@@
 and then apply the `BF` function from above first to the difference between before-sentence and before-word transitions.
-For interpretation it is important to remember that $\theta$ is a probability; hence the difference between two probabilities can only range between -1 and 1.
-In other words, a prior SD of 1 is saying that all differences are as likely as no difference (which makes very little sense here).
-Similar to the calculations for ROPE above, it is useful to contemplate which prior distribution appropriately reflects the null hypothesis for the scale the difference is on (e.g.
+For interpretation it is important to remember that $\theta$ is a probability; hence the difference between two probabilities can only range between -1 and 1 (in the most extreme case).
+In other words, a prior SD of 1 is saying that all differences are as likely as no difference (which makes very little sense).
+Similar to the calculations of the ROPE above, it is useful to contemplate which prior distribution appropriately reflects the null hypothesis for the scale the difference is on (e.g.
 probability, log msecs, msecs).
+An SD of 1 and .5 are not useful prior distributions are they assign some prior probability to differences across the full range of differences.
+An SD of .1 is a more realistic representation for the distribution of null differences
 
@@
 
-The Stan code presented in this walk through can be used to estimate the posterior for two and more conditions: We can calculate the difference between conditions from the posterior as well as main effects and interactions for more complex factorial designs as any factorial design can be reduced to a single variable.^[For example, say we have a 2 $\times$ 2 factorial design with factor 1 having two levels AB and CD and factor 2 having the corresponding levels AC and BD.
+The Stan code presented in this tutorial can be used to estimate the posterior for two and more conditions represented in one predictor variable (called `condition` above): We can calculate the difference between conditions from the posterior as well as main effects and interactions for more complex factorial designs as any factorial design can be reduced to a single variable.
+
+For example, say we have a 2 $\times$ 2 factorial design with factor 1 having two levels AB and CD and factor 2 having the corresponding levels AC and BD.
 These two factors render four conditions A, B, C, D.
-From posterior samples for each level A through D we can calculate main effect 1 as $\text{ME1}=(A+B) - (C+D)$, main effect 2 as $\text{ME2}=(A+C) - (B+D)$, and their interaction as $\text{Interaction}=(A-B) - (C-D)$ (or $\text{Interaction}=(A-C) - (B-D)$) and summary statistics as shown above.]
+From posterior samples for each level A through D we can calculate main effect 1 as $\text{ME1}=(A+B) - (C+D)$, main effect 2 as $\text{ME2}=(A+C) - (B+D)$, and their interaction as $\text{Interaction}=(A-B) - (C-D)$ (or $\text{Interaction}=(A-C) - (B-D)$) and summary statistics as shown above.
 
@@
 
-In some contexts we would like to obtain by-participant estimates of the probability of a participant to exhibit a disfluency (i.e.
-$\theta_s$).
-These estimates inform us about the prevalence of typing hesitations.
+In some contexts we would like to obtain by-participant estimates of the probability of a participant to exhibit hesitations (i.e. $\theta_s$) reflecting how often a participant is pausing.
 
@@
 
-This posterior tibble has the format
-
-| "$<\text{parameter name}>\text{_s}[<\text{transition location id}>, <\text{participant id}>]$"
-
-where indices indicate the transition-location id and the participant identifier of the parameters.
-The following code creates a long format with columns `name` with the names above as levels and a `value` column with their respective values (both `name` and `value` are `pivot_longer` defaults).
-The `name` column is then separated into `name` (levels: `prob`), `location` (levels: `1`, `2`, `3`), and participant (an index for each participant) using "," to separate the three variables.
+The column names have the format `<parameter>_s[<location>, <participant>]` where indices indicate the transition-location id and the participant identifier of the parameters.
+The following code creates a long format with columns `location` and `participant` and a `value` column with their respective values (`value` is `pivot_longer` defaults).
+The previous column names were separated into `location` (levels: `1`, `2`, `3`), and `participant` (an index for each participant) using "," to separate the two variables and extracting the numbers inside the squared brackets (the remainder of the name was discarded).
 
@@
 posterior_ppts_long <- posterior_ppts %>%
-  pivot_longer(everything()) %>%
-  # separate into location and ppt using "," as separator
-  separate(name, into = c("location", "participant"), sep = ",") %>% 
-  # extract number from new variables and recode location
-  mutate(across(c(location, participant), parse_number),
-         across(location, recode, 
+  pivot_longer(everything(), 
+               names_to = c("location", "participant"), 
+               # extract indice for location and ppt using "," as separator
+               names_pattern = "\\[(.+),(.+)\\]") %>%
+  mutate(across(location, recode, 
                 `1` = "before-sentence",
@@
 
+```{r echo = F}
+rm("mean")
+```
+
 ```{r}
@@
   pivot_longer(-participant, names_to = "location") %>%
-  summarise(across(value, list(mean = base::mean, 
+  summarise(across(value, list(mean = mean, 
                                lower = lower, 
@@
 posterior_ppts_long_summary %>%
+  mutate(across(participant, as.numeric)) %>% 
   ggplot(aes(x = participant, 
-             y = mean, ymin = lower, ymax = upper,
+             y = mean, 
+             ymin = lower, 
+             ymax = upper,
              colour = location,
@@
                values_to = 'diff') %>% 
-  summarise(across(diff, list(mean = base::mean, 
+  summarise(across(diff, list(mean = mean, 
                               lower = lower, 
@@
 
-This calculation and visualisation allows researchers to identify participants that do or do not differ in terms of hesitation frequency, in our example depending on transition locations.
+This calculation and visualisation allows us to identify participants that do or do not differ in terms of hesitation frequency depending on transition locations.
 
@@
                 diff_sent = "Comparison: before-sentence vs before-word",
-                diff_word = "Comparison: before-word vs within-word")) %>% 
-  ggplot(aes(x = factor(as.numeric(participant)), 
-             y = mean, ymin = lower, ymax = upper)) +
+                diff_word = "Comparison: before-word vs within-word"),
+         across(participant, as.numeric)) %>% 
+  ggplot(aes(x = participant, 
+             y = mean, 
+             ymin = lower, 
+             ymax = upper)) +
   geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
-  geom_pointrange(position = position_dodge(.75)) +
+  geom_pointrange() +
   facet_wrap(~difference) +
   coord_flip() +
+  scale_x_continuous(breaks = 1:10) +
   labs(x = "Participant id", 

### 2024-05-19 report/appendix/fit2data.Rmd
change tutorial to contrained molg

@@
 Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data.
-From each model we obtained 100 simulations (predictions).
-These can be found in Figure \ref{fig:prediction}.
+From each model we obtained 50 simulations (predictions); see Figure \ref{fig:prediction}.
 The model predictions fitted the data substantially better for the mixture models.
-The worse fit was observed for the single distribution Gaussian model which predicted large amounts of negative keystroke intervals which are implausible.
-Using a log-normal distribution removed these negative keystroke intervals.
+The worse fit was observed for the single distribution Gaussian model which predicted a larger number of negative keystroke intervals; a log-normal distribution removed these negative predictions.
 
 ```{r eval = F}
-theme_set(theme_few(base_size = 9) +
+theme_set(theme_few(base_size = 10) +
               theme(strip.background = element_blank(),
@@
     str_to_upper() %>% 
-    str_replace(., "PLANTRA", "PLanTra")
-ds <- ds[-6]
-plots <- list()
-plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
-plots_fin <- plots[[1]] / plots[[2]] / plots[[3]] / plots[[4]] / plots[[5]] / plots[[6]] + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
+    str_replace(., "PLANTRA", "PLanTra") %>%
+    str_replace(., "GUNNEXP2", "GE2")
 
-plots_fin <- grid.arrange(patchworkGrob(plots_fin), 
-                            left = "Density",
-                            bottom = "Transition duration in secs")
+# remove the shift model
+ds <- ds[!str_detect(ds, "SHIFT")]
 
-ggsave("../figures/fitplots.pdf", plots_fin, height = 8.5, width = 6.25)
+# get all density plots
+plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000))
+
+# remove strips after first row
+plots[2:length(ds)] <- map(2:length(ds), ~plots[[.]] + theme(strip.text = element_blank()))
+
+# create labels for datasets
+labs <- map(ds, ~ggplot(mapping = aes(x = 0, y = 0, label = .x)) + 
+  geom_text(angle = 270, size = 2.5) +
+  theme_void())
+
+# add names for data sets on the right
+plots_labs <- map(1:length(ds), ~plots[[.]] + labs[[.]] + plot_layout(widths = c(.95, .05)))
+
+# combine legends
+plots_ledg <- wrap_plots(plots_labs, guides = "collect", ncol = 1)
+
+# add axe titles
+plots_fin <- grid.arrange(patchworkGrob(plots_ledg), 
+                            left = "Density", bottom = "Interkey interval in secs")
+
+# save plot 
+ggsave("../figures/fitplots.pdf", plots_fin, height = 7.25, width = 6.25)
 
@@
 
-```{r prediction, fig.cap="Fit to data with models (distribution type) above the plots.
-Comparison of 50 simulations to observed data summarised by model.
-For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8, fig.width=6.5}
+```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red in comparison to the interkey interval data shown in black by model types and datasets.
+Model fit to the data was order from the best (left) to the weakest (right).
+For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
 knitr::include_graphics("figures/fitplots.pdf")

### 2024-05-19 report/appendix/l2_constr.Rmd
rearrange markdown sections
change tutorial to contrained molg

--- 2023-12-09 report/appendix_l2_constr.Rmd
@@
-## L2 effect (SPL2)
+# L2 effect (SPL2)
 
-For the SPL2 data (only the `_^[shift]` sentence-transitions) we calculated the L2 effect (i.e.
+For the SPL2 data we calculated the L2 effect (i.e.
 the difference between writing in L2 and L1).
@@
 ```{r l2effect, results = 'asis'}
+cellmeans <- ps_spl2_both %>% 
+  separate(group, into = c("group", "lang"), sep = "; ") %>%
+  select(-ds, -group) %>%
+  mutate(across(lang, ~str_remove(., "Language: ")),
+         across(location, ~str_replace(., "\\n", " "))) %>% 
+  pivot_wider(names_from = name, 
+              values_from = c(mean, lower, upper)) %>% 
+  mutate(across(ends_with("prob"), ~dezero(., 2)),
+         across(c(ends_with("beta"),
+                  ends_with("delta")), 
+                ~as.character(scales::comma(round(.))))) %>% 
+  pivot_longer(-c(location, data, lang), 
+               names_to = c(".value", "param"),
+               names_sep = "_") %>% 
+  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
+  select(-mean:-upper) %>% 
+  pivot_wider(names_from = data, 
+              values_from = est) %>% 
+  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
+  unique()
+
 # cell means and data set difference
@@
 
-write_csv(tmp, "tables/l2_effect_cellmeans_constr.csv")
+write_csv(tmp, "../tables/l2_effect_cellmeans_constr.csv")
 
@@
           caption = "Mixture model estimates for language effect.
-Cell means are shown the slowdown for hesitant transitions when writing in L1 and L2 and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+Cell means are shown the slowdown for hesitant interkey intervals when writing in L1 and L2 and the probability of hesitant keyintervals.
+The language difference is shown on log scale (for durations) and logit scale for probability of hesitant key intervals.
 95% PIs in brackets.",
@@
           placement = "b",
-          stub_indents = list("Hesitation duration" = 1:3,
+          stub_indents = list("Hesitation showdown" = 1:3,
                               "Hesitation probability" = 4:6) )

### 2024-05-19 report/appendix/masking.Rmd
change tutorial to contrained molg

@@
-# Masking effect (CATO, GUNNEXP2)
+# Masking effect (CATO, GE2)
 
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to what extent masking the previously written text affects keystroke behaviour.
+Studies associated with two datasets (CATO, GE2) investigated to what extent masking the previously written text affects keystroke behaviour.
 The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}.
@@
   mutate(across(value, ~ifelse(param == "theta", .*-1, value)),
-         ds = "gunnexp2",
+         ds = "GE2",
          across(param, ~recode(., "prob" = "theta")),

### 2024-05-19 report/appendix/masking_constr.Rmd
change tutorial to contrained molg

@@
           caption = "Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+Cell means are shown for the masked and unmasked writing task for the slowdown (in msecs) and the probability of hesitant transitions.
+The effect for masking is shown on log scale (for durations) and logit scale for probability of hesitant transitions.
 95% PIs in brackets.",
@@
           stub_indents = list(
-            "Hesitation duration" = 1:9,
+            "Hesitation slowdown" = 1:9,
             "Hesitattion probability" = 10:18) )

### 2024-05-19 report/appendix/models.Rmd
change tutorial to contrained molg

@@
 
-We are using the Bayesian framework [e.g.
-@farrell2018computational; @gelman2014; @lee2014bayesian] to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
+We used the Bayesian framework [e.g.
+@farrell2018computational; @gelman2014; @lee2014bayesian] to implement four statistical models to evaluate which -- the serial or the parallel view -- captures interkey data more successfully.
+In other words, we use statistical models to map between the keystroke intervals and the theoretically assumed process that generates theses data to then compare the predictive power of those models.
+These models presented in this section are building on one another so that later models include assumptions of earlier models.
 
-## Serial model of writing
+## Single-distribution Gaussian
 
-### Single distribution Gaussian
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average IKI associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as $\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)$.
-Of importance is the estimated posterior value of $\mu$ as this value captures that time it takes to mentally plan a word.
+Under the serial view, all planning must be completed prior to typing onset of the corresponding planning unit.
+The resulting interkey interval is sometimes faster or slower depending on, among others, psycholinguistic factors; i.e.
+the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes.
+For example, we can capture the variability associated with word-features by assuming that before-word interkey intervals can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average interkey interval associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the remaining variability associated with factors that we did not further specify in the model.
+This can be expressed as $\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)$ where $\text{iki}_\text{before word}$ are all interkey intervals immediately proceeding a word.
+Of importance is the estimated posterior value of $\mu$ as this value captures average time needed to mentally prepare a word.
 
 We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07].
-Regression formulas can, and commonly are, used to decompose $\mu$ and capture that there is a change in the outcome variable associated with another factors.
-We can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is the average IKI for word boundaries.
-However when $\text{x}_\text{sentence}$ takes on the value 1, the average IKI for word boundaries $\alpha$ is incremented by a sentence-related slowdown of $\beta$ msecs.
+Larger linguistic edges are typically understood as being associated with planning on higher levels.
+For example, at sentence boundaries, planning needs to happen for word-level properties -- which we captured before as average interkey interval $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07].
+Using regression formulae we can capture changes in the outcome variable that are associated with other factors.
+More specifically we can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is then the average interkey interval for word boundaries but, when $\text{x}_\text{sentence}$ takes on the value 1, the average interkey interval for word boundaries $\alpha$ is incremented by a slowdown associated with sentence boundaries of $\beta$ msecs.
 Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.
-whether there is a statistically meaningful difference between IKI associated with words and sentences).
+The application of such a statistical model to the data will therefore provide us with an estimate of the parameter value that can be used for statistical inference (e.g.
+whether there is a statistically meaningful difference associated with word and sentence boundaries).
 
-We implemented the average keytransition associated with a given text locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
-Therefore, the model will return one $\beta$ per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts $u_\text{participant}$.
+We implemented the average interkey interval associated with a given text locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
+Therefore, the model returns three distributions of $\beta$ values that capture the posterior distribution of average interkey intervals for each transition location.
+The decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is called random intercepts $u_\text{participant}$.
 The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$.
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.
-a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
+The $u_\text{participant}$ value is therefore the difference between the overall posterior estimate and estimate average interkey interval of a particular participant (i.e.
+a positive value indicates that a participant is slower than average; a negative value indicates that a participant is faster than average).
 
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model
+The resulting model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed-effects model
 
@@
 
-where $i$ iterates over every key-transition $i \in 1 \dots N$ with $N$ being the total number of keystroke intervals, and thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.
+where $i$ iterates over every interkey interval $i \in 1 \dots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.
 
@@
 
-### Single distribution log-Gaussian
+## Single distribution log-Gaussian
 
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
+The previous model assumes that the data-generating process follows a Gaussian distribution.
+As a consequence the Gaussian model generates values that are symmetrically distributed and, in principle, allows values to be negative.
+The model presented in this section is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
+
 There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
+This is a desirable property because the distance between two subsequent key events must be positive.
+Instead the lower bound is delimited by a person's ability to move their fingers and keyboard polling.
 (2) the log-scale is known to be a better match for data from human behaviour and motor responses.
@@
 For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.
+This does not map onto the psychological interpretation for short and long interkey intervals.
+For example effects that result from the motor level within words (e.g.
 typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
+In other words, an effect of 25 msecs can be large in the context of fast inerkey intervals but it is small in the context of overall slow intervals.
+Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of interkey intervals (motor activity) is more psychologically meaningful than a 25 msecs difference on the upper end (retrieving words, planning sentences).
 
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{log}\mathcal{N}()$ was replaced by $\text{logN}()$.
+This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{log}\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.
 
@@
 
-### Single distribution unequal-variance log-Gaussian
+## Single distribution unequal-variance log-Gaussian
 
-The third model representing the serial view of writing is a single distribution unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
+The third model is identical to the model presented in the previous section but removed the equal-variance assumption for the width of the distributions assocated with transition locations.
+The previous models modelled interkey intervals as a function of transition location meaning that the average interkey interval depends on its position in the text.
+The variance associated with the estimated interkey intervals for each transition location was assumed to be identical (equal variance).
+This assumption is however not in line with what we know about data from human motor behaviour.
 Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing].
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation $\sigma_{e_\text{location}}^2)$ varies by transition location.
+For interkey intervals pauses at larger linguistic edges can therefore be assumed to be associated with a larger variance also because of the larger number of associated processes.
+Therefore, in equation \ref{eq:unimoduv}, we introduced the assumption that standard deviations $\sigma_{e_\text{location}}^2)$ vary by transition location.
 
@@
 
-## Parallel cascading model of writing
+## Two-distributions log-Gaussian mixture model
 
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-### Two-distributions log-Gaussian mixture model (constrained)
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
+The following model is an extension of the last single-distribution model.
+In contrast to the models introduced for the serial view, the cascading view allows planning to unfold in parallel to production.
+Therefore, we removed for the model presented in this section the constraint that all planning must be completed before writing onset.
+Instead of assuming that larger linguistic edges shift the distribution of interkey intervals as in the previous models, we modelled pauses in the distribution of interkey intervals, under the parallel view, to be more likely to occur at larger linguistic edges.
+This was achieved by introducing the model assumption that interkey intervals come from a weighted combination (mixure) of two distributions associated with two different states:
 
@@
 Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter in equation \ref{eq:bimodcon}.
+These fluent keystroke transitions are merely constrained by a person's ability to move their finger and were captured by $\beta$ in equation \ref{eq:bimodcon}.
 Note that $\beta$ is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.
@@
 2.
-Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining $\delta$ to be positive, it captures how much longer hesitant IKIs in addition to $\beta$.
+Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when information, for example words or their spelling, could not be retrieved in time.
+The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon} which represents the magnitude of the slowdown associated with transition location.
+The slowdown $\delta$ was allowed to vary by transition locations because hesitations at larger linguistic units are likely to be associated with higher level planning which causes longer delays in the output.
+The $\delta$ parameter was constrained to be positive, so that it captures how much longer hesitant interkey intervals are in addition to $\beta$.
 
@@
 (\#eq:bimodcon)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \times \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
 \text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
@@
 
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
+The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant interkey intervals; the second line represents fluent interkey intervals.
 Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
-$\theta$ is here parameterised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$ .
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
+$\theta$ is here parameterised to represent the probability that an interkey interval is associated with the distribution of long interkey intervals, hence representing the hesitation probability.
+This probability is inversely related to the mixing weight of the distribution of short interkey intervals by $1-\theta$ because the weights of both distributions must sum to 1.
+In other words, a larger weight for one distribution inevitably means a lower weight for the other distribution.
 
 The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
+In line with the literature discussed in the introduction, we assumed that pauses are likely to vary across linguistic edges.
+As pausing is subject to individual differences and writing style (skills), we also assumed that some participants pause more at certain transition locations and other participants pause less.
+
+<!-- This is akin to what is known as a random by-participant slopes model.  -->
 
 Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location.
-In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].
-This was achieved by introducing a parameter $\sigma_\text{diff}$.
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
+In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing hesitations is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].
 
-### Two-distributions log-Gaussian mixture model (unconstrained)
+<!-- ### Two-distributions log-Gaussian mixture model (unconstrained) -->
 
-The size of a fluent key-transition does not necessarily vary by transition location.
+<!-- The size of a fluent key-transition does not necessarily vary by transition location.
 In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions.
 This is what we called a constrained model.
-However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
+However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
 This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter.
@@
 We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
+-->
 
-In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
+<!-- In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}. -->
 
-\begin{equation}
-\begin{aligned}
-(\#eq:bimoduncon)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-	\text{where: }  & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-		& \sigma_{e'}^2 > \sigma_{e}^2\\
-		& 0 < \theta < 1
-\end{aligned}
-\end{equation}
+<!-- \begin{equation} -->
+<!-- \begin{aligned} -->
+<!-- (\#eq:bimoduncon) -->
+<!-- \text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\ -->
+<!--   & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\ -->
+<!-- 	\text{where: }  & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\ -->
+<!-- \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\ -->
+<!-- 		& \sigma_{e'}^2 > \sigma_{e}^2\\ -->
+<!-- 		& 0 < \theta < 1 -->
+<!-- \end{aligned} -->
+<!-- \end{equation} -->

### 2024-05-19 report/appendix/postparam.Rmd
rearrange markdown sections
change tutorial to contrained molg

--- 2023-07-24 report/appendix_b.Rmd
@@
-## Posterior parameter estimates
+# Posterior parameter estimates
 
@@
 
-file <- list.files(path = "../stanout/c2l1", 
+file <- list.files(path = "../../stanout/c2l1", 
                    pattern = "mogbetaun.*.csv$",
@@
             .by = c(location, xn, name)) %>% 
-  mutate(ds = "GUNNEXP2") %>% 
+  mutate(ds = "GE2") %>% 
   rename(group = xn)
 
-file <- list.files(path = "../stanout/lift", 
+file <- list.files(path = "../../stanout/lift", 
                    pattern = "mogbetaun.+.csv$",
@@
 
-file <- list.files(path = "../stanout/plantra", 
+file <- list.files(path = "../../stanout/plantra", 
                    pattern = "mogbetaun.+.csv$",
@@
 
-file1 <- list.files(path = "../stanout/spl2", 
+file1 <- list.files(path = "../../stanout/spl2", 
                     pattern = "mogbetaun.+.csv$",
@@
 
-file2 <- list.files(path = "../stanout/spl2_shift", 
+file2 <- list.files(path = "../../stanout/spl2_shift", 
                     pattern = "mogbetaun.+.csv$",
@@
                            bottom = "Transition location")
-ggsave("figures/psplots1.pdf", plot_fin_1, height = 8, width = 7)
+ggsave("../figures/psplots1.pdf", plot_fin_1, height = 8, width = 7)
 
@@
                            bottom = "Transition location")
-ggsave("figures/psplots2.pdf", plot_fin_2, height = 8, width = 7)
+ggsave("../figures/psplots2.pdf", plot_fin_2, height = 8, width = 7)
 ```

### 2024-05-19 report/appendix/postparam_constr.Rmd
change tutorial to contrained molg

@@
 ```{r eval = F}
-
-theme_set(theme_few(base_size = 11) +
+theme_set(theme_few(base_size = 10) +
             theme(strip.background = element_blank(),
@@
 linewidth <- .35
-beta_label <- "Fluent transitions\n(in msecs)"
+beta_label <- "Fluent interkey\nintervals"
+delta_label <- "Slowdown for hesitations"
+theta_label <- "Probability of hesitations"
+
+ps <- ps %>% mutate(across(group, ~case_when(str_detect(., "L1") ~ "L1",
+                                             str_detect(., "L2") ~ "L2",
+                                             TRUE ~ .)))
 
@@
                     scale_y_continuous(labels = scales::comma) +
-                    labs(subtitle = .x) +
                     theme(axis.title = element_blank(),
@@
 
+labs <- list()
+labs <- map(datasets, ~filter(ps, ds == .x) %>% 
+      select(ds) %>% 
+      unique() %>% 
+  ggplot(mapping = aes(x = 0, y = 0, label = ds)) + 
+  geom_text(angle = 270, size = 2.5) +
+  theme_void())
+
 # Combine parameter plots 
 plot_param <- list()
-plot_param <- map(1:6, ~beta_plots[[.]] + delta_plots[[.]] + theta_plots[[.]] +
-                    plot_layout(guides = "collect", widths = c(.45, 1, 1)) &
+
+beta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~beta_plots[[.]] + theme(strip.text = element_blank()))
+delta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~delta_plots[[.]] + theme(strip.text = element_blank()))
+theta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~theta_plots[[.]] + theme(strip.text = element_blank()))
+
+beta_plots[c(1,2,4,5)] <- map(c(1,2,4,5), ~beta_plots[[.]] + theme(axis.text.x = element_blank(),
+                                                                   axis.ticks.x = element_blank()))
+delta_plots[c(1,2,4,5)] <- map(c(1,2,4,5), ~delta_plots[[.]] + theme(axis.text.x = element_blank(),
+                                                                   axis.ticks.x = element_blank()))
+theta_plots[c(1,2,4,5)] <- map(c(1,2,4,5), ~theta_plots[[.]] + theme(axis.text.x = element_blank(),
+                                                                   axis.ticks.x = element_blank()))
+
+plot_param <- map(1:6, ~beta_plots[[.]] + delta_plots[[.]] + theta_plots[[.]] + labs[[.]] +
+                    plot_layout(guides = "collect", widths = c(.45, 1, 1, .1)) &
                     theme(legend.position = "bottom"))
@@
 
+plot_param[[6]] <- plot_param[[6]] & guides(colour=guide_legend(nrow=1,byrow=TRUE),
+                                            shape = guide_legend(nrow=1,byrow=TRUE))
+
 plot_fin_1 <- plot_param[[1]] / plot_param[[2]] / plot_param[[3]]
@@
                            bottom = "Transition location")
-ggsave("../figures/psplots1constr.pdf", plot_fin_1, height = 8, width = 7)
+ggsave("../figures/psplots1constr.pdf", plot_fin_1, height = 6, width = 6)
 
@@
                            bottom = "Transition location")
-ggsave("../figures/psplots2constr.pdf", plot_fin_2, height = 8, width = 7)
+ggsave("../figures/psplots2constr.pdf", plot_fin_2, height = 6.75, width = 6.5)
 ```
@@
 \includegraphics{figures/psplots1constr.pdf}
-\caption{Posterior parameter distribution}
+\caption{Distribution of posterior parameter estimates with 95\% probability intervals (PI).
+Fluent interkey intervals are shown in msecs and the slowdown for hesitant key intervals in secs; probability of hesitations is the proportion of hesitant interkey intervals.}
 \end{figure}
@@
 \label{fig:fullps1}
-\caption{Posterior parameter distribution}
+\caption{Distribution of posterior parameter estimates with 95\% probability intervals (PI).
+Fluent interkey intervals are shown in msecs and the slowdown for hesitant key intervals in secs; probability of hesitations is the proportion of hesitant interkey intervals.}
+
 \end{figure}

### 2024-05-19 report/appendix/prepost.Rmd
change tutorial to contrained molg

@@
           caption = "Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+Cell means are shown for the pre-test and post-test in msecs for fluent interkey intervals, the slowdown for long interkey intervals and the probability of hesitant transitions.
+The effect for post-test is shown on log scale (for interkey intervals) and logit scale for probability of hesitant interkey intervals.
 95% PIs in brackets.",
@@
           stub_indents = list(
-            "Fluent transitions" = 1:3, 
+            "Fluent interkey intervals" = 1:3, 
             "Hesitation duration" = 4:6,

### 2024-05-19 report/appendix/prepost_constr.Rmd
change tutorial to contrained molg

@@
           caption = "Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant interkey intervals (hesitation duration) and the probability of hesitant interkey intervals.
+The effect for post-test is shown on log scale (for interkey intervals) and logit scale for probability of hesitant interkey intervals.
 95% PIs in brackets.",

### 2024-05-19 report/appendix/shift.Rmd
change tutorial to contrained molg

@@
             .by = c(location, xn, name)) %>% 
-  mutate(ds = "GUNNEXP2") %>% 
+  mutate(ds = "GE2") %>% 
   rename(group = xn)
@@
 
-The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
+The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GE2) scope over the character following the shift key.
 In other words, the pause before sentences sumed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets.
@@
           align = c("l", "r", "r", "r", "r"), 
-          caption = "Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
+          caption = "Mixture model estimates for interkey intervals.
+Cell means are shown for interkey intervals that do and do not involve the transition to the character following shift in msecs for fluent interkey intervals, the slowdown for long interkey intervals and the probability of hesitant interkey intervals.
+The difference for including the interkey intervals to the character after shift is shown on log scale (for interkey intervals) and logit scale for probability of hesitant interkey intervals.
 95% PIs in brackets.",
@@
           longtable = T,
-          stub_indents = list("Fluent transitions" = 1:2,
+          stub_indents = list("Fluent interkey intervals" = 1:2,
                               "Hesitation duration" = 3:4,

### 2024-05-19 report/appendix/shift_constr.Rmd
change tutorial to contrained molg

@@
-# Key-combination effect (constrained mixture model)
+```{r}
+# get posterior
+file <- list.files(path = "../../stanout/cato", 
+                   pattern = "mogbetaco.+.csv$",
+                   full.names = T)
 
-The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets.
-Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
 
-<!-- Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.  -->
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
 
-Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
-We compared the SPL2 data including and excluding the keystroke after shift.
-Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
-The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.
+ps_cato <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(task, group, location, name)) %>% 
+  mutate(group = str_c("Task: ", task, "; Group: ", group),
+         ds = "CATO") %>% 
+  select(-task)
+
+file <- list.files(path = "../../stanout/c2l1", 
+                   pattern = "mogbetacon.*.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_c2l1 <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta,by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(location, name)) %>% 
+  mutate(ds = "C2L1",
+         group = "1")
+
+file <- list.files(path = "../../stanout/gunnexp2", 
+                   pattern = "mogbetacon.*.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_gunn <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(location, xn, name)) %>% 
+  mutate(ds = "GE2") %>% 
+  rename(group = xn)
+
+file <- list.files(path = "../../stanout/lift", 
+                   pattern = "mogbetaco.+.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_lift <- ps %>% filter(param != "beta") %>% 
+  mutate(across(genre, ~ifelse(is.na(genre), "Arg", .))) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(genre, topic, location, name)) %>% 
+  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre),
+         ds = "LIFT") %>% 
+  select(-topic, -genre)
+
+file <- list.files(path = "../../stanout/plantra", 
+                   pattern = "mogbetaco.+.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_plantra <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(task, location, name)) %>% 
+  mutate(ds = "PLanTra") %>% 
+  rename(group = task)
+
+file1 <- list.files(path = "../../stanout/spl2", 
+                    pattern = "mogbetaco.+.csv$",
+                    full.names = T)
+
+file2 <- list.files(path = "../../stanout/spl2_shift", 
+                    pattern = "mogbetaco.+.csv$",
+                    full.names = T)
+
+ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
+                     mutate(data = .x)) %>% 
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
+         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
+                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
+  filter(param %in% c("beta", "delta", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value, data) %>% 
+  mutate(idx = 1:n(), .by = data)
+
+ps_spl2_both <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(delta, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx, data)) %>% 
+  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
+         beta2 = beta + delta,
+         across(c(beta2, beta), ~exp(.)),
+         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  select(-beta2) %>% 
+  pivot_longer(c(beta, prob, delta)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper),
+                   .names = "{.fn}"),
+            .by = c(lang, location, name, data)) %>% 
+  mutate(ds = "SPL2",
+         group = str_c("Data: ", data, "; Language: ", lang)) %>% 
+  select(-lang)
+
+ps_spl2 <- ps_spl2_both %>% filter(data == "SPL2 (_^[shift])")
+
+# Combine all
+ps <- bind_rows(ps_cato, ps_c2l1, ps_gunn, ps_lift, ps_plantra, ps_spl2)
+```
+
+# Key-combination effect
+
+The analysed datasets differ to the extent that interkey intervals at before sentence location did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GE2) scope over both the transition to the shift key and the the transition to the character following the shift key.
+In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C`, where `^` represents a transition between keys, but only one key interval, namely `_^[shift]`, for the remaining datasets.
+Therefore, longer or more frequent pauses for transitions immediately proceeding a sentence could be explained without reference to linguistic edges.
+
+Therefore we compared whether any difference can be explain on the basis of the additional keystroke involved in before-sentence transitions.
+We used the SPL2 data for this comparison and calculated intervals including and excluding the keystroke after shift.
+Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference for before and within-word key-intervals.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans} (only for the data from participants writing in their first language).
 
@@
   separate(group, into = c("group", "lang"), sep = "; ") %>%
-  select(-ds, -group) %>%
-  mutate(across(lang, ~str_remove(., "Language: ")),
+  filter(str_detect(lang, "L1")) %>% 
+  select(-ds, -group, -lang) %>%
+  mutate(#across(lang, ~str_remove(., "Language: ")),
          across(location, ~str_replace(., "\\n", " "))) %>% 
@@
                 ~as.character(scales::comma(round(.))))) %>% 
-  pivot_longer(-c(lang, location, data), 
+  pivot_longer(-c(location, data), 
                names_to = c(".value", "param"),
@@
               values_from = est) %>% 
-  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
+  mutate(across(c(location), ~ifelse(param == "beta", "overall", .))) %>% 
   unique()
@@
   mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
-  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
+  filter(lang %in% c("EN", "overall")) %>% 
+  mutate(#across(lang, recode, "EN" = "L1", "ES" = "L2"),
          across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
@@
                    .names = "{.fn}"),
-            .by = c(lang, location, param)) %>% 
+            .by = c(location, param)) %>% 
   mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
@@
   left_join(cellmeans) %>% 
-  arrange(param, location, lang) %>% 
   filter(location %in% c("overall", "before sentence")) %>% 
-  select(param, 
-         Language = lang,
-         `_^[shift] + C` = ends_with(" + C)"),
+  mutate(across(param, ~recode_factor(., 
+                                `beta` = "Fluent interkey intervals",
+                                `delta` = "Hesitation slowdown",
+                                `prob` = "Hesitation probability",
+                                .ordered = T))) %>% 
+  select(`Model parameter` = param, 
+         `_^[shift]^C` = ends_with(" + C)"),
          `_^[shift]` = ends_with("FT])"),
@@
 
-apa_table(tmp[,-1],
-          align = c("l", "r", "r", "r", "r"), 
-          caption = "Mixture model estimates for key transitions immediately preceding a sentence.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.",
+apa_table(tmp,
+          align = c("l", "l", "r", "r", "r", "r"), 
+          caption = "Mixture model estimates for interkey interval immediately preceding a sentence.
+Cell means are shown for interkey intervals that do and do not involve the interkey intervals to the character following shift in msecs for the slowdown and the probability of hesitant interkey intervals.
+The difference for including the interkey intervals to the character after shift is shown on log scale (for durations) and logit scale for probability of hesitant interkey intervals; 95% PIs are shown in brackets.",
           note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
           font_size = "footnotesize",
-          longtable = T,
-          stub_indents = list("Fluent transitions" = 1,
-                              "Hesitation duration" = 2:3,
-                              "Hesitation probability" = 4:5) )
+          longtable = T)
+          # stub_indents = list("Fluent interkey intervals" = 1,
+          #                     "Hesitation duration" = 2:3,
+          #                     "Hesitation probability" = 4:5) )
 
@@
 
-The fluent transition duration and hesitation duration were not affected by whether or not the sentence-initial transition include the character following shift for neither language.
-However, we found strong evidence for an increased hesitation probability -- in both -- languages when the before-sentence key transition included the character following the shift key.
-Notably, the hesitation probability increased ceiling.
-It is not entirely surprised that the constrained mixture model identifies essentially all before-sentence transition as pauses: the constrained model does not distinguish between transition locations for the duration of fluent transitions.
-Therefore essentially all before-sentence transitions are identified as hesitation as they always include two keystrokes while all other transition location (which is the majority of the data) do not.
+The duration of fluent and hesitant interkey intervals was not affected by whether or not the sentence-initial transition include the character following shift.
+However, we found strong evidence for an increased hesitation probability when the before-sentence key transition included the character following the shift key.
+Notably, the hesitation probability increased closes to ceiling when the transition following the shift key was added.
+In other words the mixture model identifies almost all before-sentence transition as being hesitant as they always include two keystrokes while interkey intervals from all other transition location -- the majority of the data -- do not.

### 2024-05-19 report/appendix/sim.Rmd
update tutorial final for now
change tutorial to contrained molg

@@
 
-In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}).
+In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equation \ref{eq:bimodcon}).
 This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
@@
 (\#eq:simmog)
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
+\text{y} \sim\text{ } & \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma^2_1) +\\
+& (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma^2_2)\\
 \text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
@@
 
+In R, data that follow the model in equation \ref{eq:simmog} can be simulate using a helper function called `molg` (mixture of log Gaussians).
+
+```{r eval = F, echo = T}
+# Function for a mixture of log Gaussians
+molg <- function(n, theta, mu1, mu2, sigma1, sigma2) {
+  y0 <- rlnorm(n, mean = mu1, sd = sigma1)
+  y1 <- rlnorm(n, mean = mu2, sd = sigma2)
+  mix <- rbinom(n, size = 1, prob = theta)
+  y <- y0 * (1 - mix) + y1 * mix 
+}
+
+N <- 1000 # number of subjects
+beta <- 5 # mean of fluent key intervals (in log)
+delta <- 1 # slowdown for hesitant interkey intervals (in log)
+theta <- .35 # propotion of hesitations
+sigma <- c(.25, .5) # trial-by-trial error
+
+# Generate data from a mixture of two log-Gaussians
+y <- molg(n = N, 
+          theta = theta,
+          mu1 = beta,
+          mu2 = beta + delta,
+          sigma1 = sigma[1],
+          sigma2 = sigma[2])
+
+```
+
 The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
@@
 (\#eq:simuv)
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
+\text{y} \sim\text{ }& \text{log}\mathcal{N}(\beta, \sigma^2)\\
 \text{constraint: } & \sigma^2>0
@@
 \end{equation}
+
+Data following the distirbution in equation \ref{eq:simuv} can be simulated using a few lines of R code:
+
+```{r echo = T, eval=F}
+N <- 1000 # number of subjects
+beta <- 5 # Population mean
+sigma <- .25 # trial-by-trial error
+
+# Generate data from a log normal distribution
+y <- rlnorm(n = N, mean = beta, sd = sigma)
+```
 
@@
 However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error -- a *z*-score approximation -- reveals a value of 11.6 which is strong evidence in in favour of the mixture model compared to the single-distribution model (Sivula et al., 2020).
-In comparison, for the single-distribution data, the single-distribution model was negligibly better than the two-distribution mixture model which was evidence by a value of 0.77 for the ratio of $\Delta\widehat{elpd}$ and its standard error.
-Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the single distribution is preferred by the law of parsimony.
+In fact, the mixture model's predictive performance is 11.6 standard errors higher compared to the single-distribution model.
+In comparison, for the single-distribution data, the two-distributions mixture model was negligibly different from the single-distribution model with a difference of 0.77 standard errors.
+Thus, the single distribution is more parsimonious.
 

### 2024-05-19 report/appendix/transloc.Rmd
change tutorial to contrained molg

@@
                 "cato" = "CATO",# (non-dyslexic, unmasked)",
-                "gunnexp2" = "GUNNEXP2",# (unmasked)",
+                "gunnexp2" = "GE2",# (unmasked)",
                 "spl2" = "SPL2", # (L1)", 
@@
          !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked")) %>% 
+         !(str_detect(data, "GE2") & xn == "masked")) %>% 
   pivot_wider(names_from = param, 
@@
 Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability."}
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
+Axes for interkey intervals are log-scaled for visability."}
 
@@
                               "spl2" = "*SPL2* (L1)", 
-                              "gunnexp2" = "*GUNNEXP2* (unmasked)",
+                              "gunnexp2" = "*GE2* (unmasked)",
                               #                "spl2_shift" = "*SPL2* (L1; shift + C)", 
@@
          !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked"),
+         !(str_detect(data, "GE2") & xn == "masked"),
          param %in% c("beta", "delta", "theta")) %>%
@@
                               `\\textbf{CATO (non-dyslexic, unmasked)}`= 3:4,
-                              `\\textbf{GUNNEXP2 (unmasked)}` = 5:6,
+                              `\\textbf{GE2 (unmasked)}` = 5:6,
                               `\\textbf{LIFT}` = 7:8,

### 2024-05-19 report/appendix/transloc_constr.Rmd
change tutorial to contrained molg

@@
-# Transition location effect (constrained mixture model)
+# Transition location effect
 
@@
          !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked"),
+         !(str_detect(data, "GE2") & xn == "masked"),
          param %in% c("delta", "theta")) %>%
@@
           note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
-          caption = "Effect of transition location on keystroke intervals. Differences are shown on log scale (for durations) and logit scale for hesitation probability. 95\\% PIs in brackets.")
+          caption = "Effect of transition location on interkey intervals. Differences are shown on log scale (for durations) and logit scale for hesitation probability. 95\\% PIs in brackets.")
 

### 2024-05-19 report/manuscript.Rmd
change tutorial to contrained molg

@@
 ---
-title             : "Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process"
-shorttitle        : "Modelling writing hesitations"
+title             : "Ideas cascade in parallel into keystrokes -- modelling writing hesitations as finite mixture process"
+shorttitle        : "Ideas cascade in parallel into keystrokes"
 
@@
 
-abstract: "Classical serial models view the process of producing a text as a chain of discrete pauses and writing bursts.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel with execution.
-These two theories make different assumptions about the underlying process that generates inter-keystroke intervals.
-Across six datasets comprising multi-sentence text composition we instantiated these assumptions as two sets of Bayesian mixed-effects models: (1) single-distribution models as assumed by the serial processing account, and (2) finite two-distribution mixture models as assumed by the parallel processing account.
-We analysed keystroke intervals at before-sentence, before word, and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view across all datasets.
+abstract: "Classical serial models view the process of producing a text as a chain of discrete pauses, during which the next span of text is planned, and bursts of activity during which this text is output onto the page or computer screen.
+In contrast, semi-parallel models of writing assume that planning is not complete at production onset and operates in parallel with execution.
+Across six datasets from participants composing multi-sentence texts we instantiated these assumptions in two sets of Bayesian mixed effects models: (1) single-distribution models as assumed by the serial processing account, and (2) a finite (two-distribution) mixture model as assumed by the parallel processing account.
+We analysed interkey intervals at before-sentence, before word, and within word transitions.
+Model comparisons demonstrated strong evidence in favour of the parallel view across all datasets.
 When pausing occurred, sentence initial pauses were longer than word initial pauses which is consistent with the idea that larger linguistic edges are associated with higher level planning.
-However, we found -- across populations -- that keystroke latencies at word and even at sentence boundaries were often too rapid to permit planning of what was written next.
-Our results cannot be explained by the serial but are in line with the cascading, semi-parallel view of writing."
+However, we found -- across populations -- that interkey intervals at word and even at sentence boundaries were often too rapid to permit planning of what was written next.
+Our results cannot be explained by the serial processing but are in line with the semi-parallel view of writing."
 
@@
 
-beta_label <- "Fluent transition\nduration (in msecs)"
-delta_label <- "Slowdown for hesitant\ntransitions (in secs)"
-theta_label <- "Probability of hesitant\ntransitions"
+beta_label <- "Fluent interkey\nintervals (in msecs)"
+delta_label <- "Slowdown for hesitant\ninterkey intervals (in secs)"
+theta_label <- "Probability of hesitant\ninterkey intervals"
 ```
@@
 
-```{r child = 'appendix/transloc.Rmd'}
+```{r child = 'appendix/transloc.Rmd', eval=F}
 ```
@@
 
-```{r child = 'appendix/shift.Rmd'}
+```{r child = 'appendix/shift.Rmd', eval = F}
 ```

### 2024-05-19 report/sections/conclusion.Rmd
change tutorial to contrained molg

@@
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-We found that key transitions across word and even sentences boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
+In contrast to the serial pause-and-burst view, the parallel view emphasises that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
+Using the Bayesian framework we implemented models of both the serial model of writing and the parallel view.
+Model comparisons and the inspection of posterior parameter estimates support the parallel view of writing but provide strong evidence against the serial view.
+We found that interkey intervals across word and even sentence boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
 This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.

### 2024-05-19 report/sections/discussion.Rmd
change tutorial to contrained molg

@@
-The classical serial view [@flower1980ynamics] characterizes writing as a sequence of planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-The cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production [@olive2014toward].
-We used Bayesian statistical models, in particular hierarchical single-distribution and two-distribution mixture models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of six datasets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
+The classical serial view [@flower1980ynamics] characterises composing spontaneous multi-sentence text as a sequence of planning and execution cycles.
+Text is therefore produced as a series of production bursts bounded by pauses.
+The parallel model of writing, in contrast, sees  planning, at least in competent writers, as occurring to a large extent in parallel with output [@olive2014toward].
+The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is the pause duration doesn’t necessarily correlated with planning necessary for next output.
+We implemented  these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing.
+Fitting these models with six sets of keystroke data from free text production tasks -- with a total of 953 participants producing 5,223,280 interkey intervals -- we have found compelling and consistent statistical evidence in favour of the parallel view.
+In all cases, the predictive performance of the two-distribution mixture-models (consistent with the parallel view of writing) outperformed the single-distribution models.
 
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key transitions that directly precede sentences or words are not necessarily associated with a pause because writers can plan in parallel to writing.
-On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in the majority of datasets, pauses were as likely to occur before sentences and words, in both experienced and novice writers, while mid-word pauses are generally rare.
-Also, with few exceptions, pauses were typically longer before sentences than before words (which were longer than mid-word pauses).
-In other words, when writers do pause, the involvement of higher-level processing at sentence boundaries is typically leading to longer pauses.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because our results for the hesitation probability suggest that writers often do not pause before sentences; indeed, hesitation probabilities are largely identical before sentences and before words.
-This suggest that writers frequently plan the next word and even sentence in parallel to production which mitigates the need to pause to plan up-coming linguistic information.
+The two distribution model assumes two data generating processes, one associated with the immediate demands of motor planning required for the production of a keystroke, and one that captures all upstream processing from message to abstract letter representation.
+A direct implication of parallel, cascaded processing is that the distribution over interkey intervals is not additive -- it is not a combination of time to motor plan plus time for upstream processing.
+The interval between keypresses is *either* determined by motor execution *or* by processing that occurred upstream of motor execution (in cases where provision of output from the upstream processes lags behind the rate at which it can be output by the fingers).
+Under a serial account -- where all planning necessary for production of next-text occurs in the interkey interval immediately before output of -- interkey interval will be the sum of times for each processing step required for generation of that text.
+Both accounts predict longer mean interkey intervals at higher-order text edges (interkey intervals greater before sentences than before mid-sentence words, and before mid-sentence words than before mid-word characters).
+This has been consistently observed in a number of previous studies [@torrance2016adolescent; @wen02; @mohsen2021second; @miller2006pausing; @xu2017analyzing; @miller2008psycholinguistic;@deane2015exploring].
+However, the serial account does not permit the possibility of a distribution of short interkey intervals at higher-order edges: It strongly predicts that *all* sentence boundaries will be associated with interkey intervals that are longer than those required just for the motor planning of the next character.
+Under the parallel-processing account it is very likely that some -- perhaps a large majority -- of interkey intervals before sentences will be longer.
+However there will be a distinct subset that is drawn from a distribution associated just with interkey intervals necessary for motor execution.
 
-Interestingly even the youngest writers in our data pool (C2L1) showed evidence of parallel planning expressed in fluent transitions between words and even sentences.
-Key-transitions were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-This is interesting because, for example, @olive2014toward described a cascading model that is operating in a serial fashion for inexperienced or struggling writers explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers but pauses were more likely before words and sentences but similarly long indicating a tendency to prepare utterances in planning units possibly smaller than the clause.
-The main difference for these young writers was that their writing speed (fluent transitions) was substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra) we observe pausing patterns that do not resemble a serialised writing process as indicated by a larger number of fluent transitions across word and sentence boundaries.
+This is what we found.
+At all linguistic edges we found a distribution of interkey intervals with durations that, in previous research, have been attributed just to motor execution (refs).
+For mid-word interkey intervals, as might be expected, the mixture model predicted that the vast majority of interkey intervals fall within this distribution.
+Across the six datasets that we sampled, the central tendency for the duration of mid-word interkey intervals was broadly similar to mean mid-word interkey intervals observed in previous studies.
+Importantly, however, these short interkey intervals were also present at higher-order text edges.
+Across all datasets we found, at minimum, a non-trivial minority of sentence-initial interkey intervals were sampled from the short distribution (i.e.
+with durations that could only be associated with motor execution).
+In two cases -- for adult writers sampled in the PLanTra and Lift datasets -- the majority of interkey intervals fell within this distribution.
+Thus, although as might be expected there was a considerably increased tendency for longer interkey intervals at higher order boundaries: Planning associated with planning a new sentence is, probabilistically, much more likely to spill beyond output of the previous sentence and therefore affect duration of the sentence-initial interkey interval.
+However the presence of a non-trivial proportion of interkey intervals with a duration that is inconsistent with anything other than motor execution can only be explained with recourse to theories that permit preparation in parallel with preceding output.
 
-We observed two more differences between data sets -- other than the patterns observed for C2L1 summarised above -- that are worth highlighting.
-These can largely be explained on the basis of differences associated with the population (writing competence / experience) and text genre [see also @conijn2019understanding] that these samples were obtained from: The first exception concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations.
-**TODO: I don't have an explanation for this.** Second, pauses were not always longer before sentence compared to before words compared to mid-word paues; there were two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT dataset although the posterior showed the same general patterns as observed in most other datasets.
-The absence of a difference could be explained as specific to the synthesis writing task because there is no need for content creation [**REFERENCE?**] but the results for the PlanTra dataset rule out this possibility.
+Our results showed that the hesitation probabilities are largely identical before sentences and before words, which suggests that writers frequently plan the next word and even sentence in parallel to production -- consistent with the parallel view of writing -- which mitigates the need to pause to plan up-coming linguistic information.
 
-A possible concern with our result -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-This concern is preempted by using cross-validation techniques for model comparison which is preventing model overfit.
+This parallelism was evident even in the youngest writers in our data pool (C2L1; mean age: 11.8 years).
+As might be expected for children who compose predominantly by handwriting and whose typing skills will still be developing, motor execution rate (mean duration of fluent transitions) was substantially slower compared to the more experienced typists in other datasets.
+In all other respects, however, findings for this dataset were similar to that of other datasets and consistent with parallel processing.
+A two-distribution mixture model provided best fit to these data.
+At both before-word and before-sentence locations a non-trivial number of interkey intervals were within the short distribution - i.e.
+at durations too short to permit lexical retrieval or syntactic planning.
+Probability of longer intervals was greater before-sentence and before-word transitions compared to mid-word transitions.
+This is interesting because, for example, @olive2014toward described a parallel model that operates  in a serial fashion for inexperienced or struggling writers, explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
+However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers: Pauses were both equally long and likely at before-word and before-sentence transitions but more likely compared to mid-word transitions.
+Also for L2 writers (PLanTra) we observe pausing patterns that do not resemble a serialised writing process as evidenced by a large number of fluent transitions at word and sentence boundaries.
+
+We observed two more differences across datasets -- other than the patterns summarised above for C2L1 -- that are worth highlighting.
+Both can largely be explained on the basis of differences associated with the population (writing competence / experience) that these samples were obtained from and text genre they were asked to produce [see also @conijn2019understanding]: The first exception concern the GE2 and SPL2 datasets: pauses were more likely to occur before sentences than before words; this difference was absent in the remaining data sets.
+**TODO: I don't have an explanation for this.** Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production.
+(2) pause durations did not vary across transition locations in the LIFT dataset.
+The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning content and text structure: the structure is often kept similar to the source in synthesis writing [“structural isomorphism”’, @hidi1986producing].
+However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents or generating new sentence structures.
+
+An observation that was not strictly predicted by the parallel view is that the unconstrained mixture model M5 that allowed the distribution over durations of fluent transitions to vary between transition locations showed a higher predictive performance.
+This can be explained on the basis of low-level keyboarding activity: It is possible that letter bigrams are systematically produced faster than transitions between space and a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
+This is for two reasons: first, and importantly, highly frequent bi- and even trigrams may be stored, retrieved and executed as one motor code [@behmer2016crunching; @grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants] because of the strong association between letters.
+This association reduces the length of transition durations.
+It is less likely that the same association exists for space-letter and space-shift pairs.
+Second, in two datasets (PLanTra, LIFT) before-sentence transitions are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transitions.
+We found no evidence for the latter possibility (see Results section and Appendix \ref{key-combination-effect}).
+
+A possible concern with our result -- that two-distributions mixture models showed substantially better predictive performance -- is that as mixture models have more parameters they might always outperform single-distribution models.
+This concern was preempted by using cross-validation techniques for model comparison which is preventing model overfit.
 In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
-We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
-This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
+We simulated two datasets that were either based on a single distribution data-generating process or a process with a two-distributions process.
+This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- a single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
+Importantly, we can test the performance of a mixture model for data that come from a single process.
 Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
+As might be expected, we found that the two-distribution mixture model was successful at uncovering the parameter values of the data generated with a two-distributions mixture process; we observed the same for the single-distribution model applied to the data generated with a single distribution process.
 Parameters were not successfully uncovered when we switched model type and dataset.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a single distributions process but a substantially higher predictive performance for data simulated from a two-distributions mixture process.
-Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of overfitting.
+Importantly though, cross-validation rendered a substantially higher predictive performance for the mixture model compared to the single-distribution model for data simulated from a two-distributions mixture process, but not when applied to data based on a single-distribution process.
+Therefore we can rule out the possibility that the statistical support for the two-distributions mixture-models can be explained on the basis of their larger number of parameters (i.e.model overfitting).
 
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
+Throughout this paper we assumed that hesitations reflect mental planning of upcoming ideas and encoding of linguistic units.
+Alternatively, at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
 There is limited research that used eye tracking to investigate regressive gaze behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
-There are at least four reasons why eye movements away from the inscription point might occure during writing: First, reading is necessary for revision (i.e.
+There are at least four reasons why eye movements away from the inscription point might occur during writing: First, reading is necessary for revision (i.e.
 writers must check if writing goals have been met).
-Second, writers might use text to cue or reinstating ideas after interruption of writing process.
-Third, writers might look for or correct spelling or grammatical errors.
+Second, writers might look for or correct spelling or grammatical errors.
 We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text.
-Fourth, writers might look at the source their are writing from.
-For two datasets (PlanTra, LIFT) which involved writing from sources, we removed transitions that were made when the sources (rather than the writing document) was in focus.
+Third, writers might look at the source they are writing from.
+For two datasets (PLanTra, LIFT) which involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.^[As for PlanTra, most participants copy-pasted the source text into their word log.
+In other words, pauses might to some extent result from reading the source text.] Fourth, writers might use text to cue or reinstate ideas after interruption of the writing process [@johansson2010looking; @hayes2013new].
 
-Frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
-This is except, lookbacks are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-This can be seen in the result of two datasets in our pool (CATO, GUNNEXP2): these experiments included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-This suggest that pausing patterns do not change if reading previous text is not possible.
+As for the fourth reason for why writers might look back in their text: frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
+In contrast to this observation -- that lookbacks are more frequent before sentences compared to words -- we found that pauses appear equally often before sentences and words under our two-distributions mixture-model.
+Finally, even though reading is likely to explain some proportion of our results, lookbacks in previously written text are overall rare, certainly in samples of university students [@torrance2016reading; @chukharev2019combined].
+Also, pause frequency did not change when the produced text was masked (see comparisons for the CATO and GE2 data in Appendix \ref{masking-effect-cato-ge2}), and therefore if reading previously produced text was not possible.
 
-A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent].
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause [@wen06; @van2016keystroke], i.e.
+A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidates for lower pause bound to handle the complex distribution of latency from spontaneous text production.
+Such thresholds are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent].
+There is one central problems with this approach: A threshold requires a definition of what passes as a pause [@wen06;@van2016keystroke], i.e.
 a pause criterion threshold often set to 2 secs [@chanquoy1996writing; @kaufer1986composing; @sullivan2002self; @wen02] or some other lower bound [@chukharev2014pauses; @connelly2012predicting; @leijten2013keystroke].
@@
 Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
+So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs in the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
 However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose [@chenu2014interword].

### 2024-05-19 report/sections/introduction.Rmd
change tutorial to contrained molg

@@
-Translating ideas into language involves a sequence of processes starting with communicative intent, proceeding through generation of a message, retrieving appropriate lexical items, deciding with which part of the message to start the utterance which, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], is determined by syntactic requirements of our language, then retrieving orthographic and / or phonological codes, and that are finally submitted to the motor programs that output speech, key or pen-strokes or signs [@wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].
+Translating ideas into language involves a sequence of processes starting with communicative intent, proceeding through generation of a message, retrieving appropriate lexical items, deciding with which part of the message to start the utterance (which, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], is determined by syntactic requirements of our language) and then retrieving orthographic and / or phonological codes, these are finally submitted to the motor programs that output speech, key or pen-strokes or signs [@wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].
 
 The coordination of these processes in writing is arguably more demanding than in spoken language.
-This is because writing requires spelling retrieval which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen are later-developing and more effortful than those required to articulate language sounds in speech leading to a reduced output speed compared to speech [see for a recent review @kandel2023written].
-Consequently when writing, more than in speech, we have to mentally buffer linguistic plans before we can express them in language.
-Written communication is also less time constrained: speaking underlies fluency requirements, because hesitating during speech has a communicative effect [e.g.
+This is because writing requires spelling retrieval, which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen.
+Pen control and typing skills are later-developing and more effortful than those required to articulate language sounds in speech and, possibly as a consequence, written output rate is typically slower [see for a recent review @kandel2023written].
+Consequently mentally buffering of linguistic plans prior to output is likely to play a greater role in writing than in speech.
+
+Written communication is also less time constrained: speaking comes with fluency requirements because hesitating during speech has a communicative effect [e.g.
 @Clark2002].
 There is therefore stronger pressure to plan ahead in speech than in writing [@roeser2018advance; @torrance2012written].
-Indeed writers can pause when they wish -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example writing argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].
+Indeed writers can pause when they wish -- even in the middle of a word -- and for how long they want without compromising communication.
+Written production of multi-sentence texts -- composing an argumentative essay, for example --  therefore involves a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].
 
 The study reported in this paper compares statistical models that instantiate two different theories of how cognitive planning processes are coordinated in writing.
-Through the analysis of 6 different datasets from writers of different ages and competencies composing multi-sentence texts, we compare these statistical models and therefore the cognitive model of how keystrokes in writing are being generated they are mapping on.
+Through the analysis of 6 different datasets from writers of different ages and competencies composing multi-sentence texts, we compare these statistical models and therefore the cognitive models of text production onto which they map.
 Also, we evaluate competing predictions of these models by revisiting the often-reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06].
@@
 We then introduce two possible cognitive frameworks in which these can be interpreted.
-Finally we argue that analysis of pause-patterns at linguistic edges are important to permit differentiation between these two theories.
+Finally we argue that analysis of pause-patterns at linguistic edges is necessary to permit differentiation between these two theories.
 
-We focus on data form spontaneous text production [see @gernsbacher1995coherence] for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production and have, therefore, real-world relevance [@dux2022automating; @dux2021effect; @vandermeulen2023writing].
-Second, since the emergence of keystroke-logging software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; for a recent special issue see @conijn2021timecourse.^[Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.] These data are therefore important for contemporary theories of writing and freely available for statistical modelling and machine learning [@conijn2019understanding; @conijn2020keys].
-Third, spontaneous text production -- unlike picture description experiments [@not10; @dam09; @meulemans2021; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual] -- content generation is not artificially constrained to individual sentence units (or less) by one or more images and linguistic form is not constrained to a sequence of words, respectively.
-Instead text composition provides data from both inter and intra-sentence transitions.^[Also, pauses in text production do not reflect visual encoding of the stimulus (although possibly reflect looking back during writing which we will return to in the discussion).]
+We focus on data form spontaneous text production [see @gernsbacher1995coherence] for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production [@dux2022automating; @dux2021effect; @vandermeulen2023writing].
+Second, since the emergence of keystroke-logging software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature; for a recent special issue see @conijn2021timecourse.^[Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.] These data are therefore important for contemporary theories of writing and freely available for statistical modelling and machine learning [@conijn2019understanding; @conijn2020keys].
+Third, spontaneous text production is not constrained in the same way as it would be in experiments in which text is elicited in response to picture stimuli [@not10; @dam09; @meulemans2021; @roeser2018advance] or is copied [@roeser2021modelling; @waes2019; @van2019multilingual].
+Instead the content-determination processes are free to run entirely internally, without the need to consult an image or a text that must be copied.^[Also, pauses in text production do not reflect visual encoding of the stimulus (although possibly reflect looking back during writing which we will return to in the discussion).]
 
 Time intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKIs) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing -- planning what to say and how to say it -- may be serial.
+Explaining the factors that influence the duration of inter-keystroke intervals (interkey intervals) requires a theory of how the various processes that transform intent into keypresses are structured.
+Planning what to say and how to say it may be serial.
 This means that the processing sequence operates one unit at a time.
-Processing of the next language unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing].
+Under this theory, planning of the next language unit can only start when the previous unit has been fully processed, from communicative intention to completion of the written output.
+Therefore interkey intervals represent periods when the writer is planning what to write next: interkey interval will be longer at linguistic edges where planning scopes over larger text spans and their duration represents the time necessary to complete that planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing].
+
+<!--
  This is particularly reflected in the model proposed by @flower1980ynamics that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts [also @chenoweth2001fluency; @chenoweth2003inner].
+-->
 
-This serial account is consistent with the frequent finding that average IKI duration immediately before larger linguistic units compared to smaller linguistic units.
+This serial account is consistent with the frequent finding that the average interkey interval duration immediately before larger linguistic units is longer compared to smaller linguistic units.
 For example, in spontaneous text the mean pause duration before a writer starts a sentence has typically been reported to be longer than before a mid-sentence word and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second; @miller2006pausing; @xu2017analyzing; @miller2008psycholinguistic;@deane2015exploring].
 This is understood as being due to time needed to plan message and syntax for the next sentence, which is added to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke of that word [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses].
-However this account is not consistent with what we know about writing.
-Consider these two observations:
+However this account is not consistent with two other sets for findings:
 
@@
 Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of a real-time process.
-For speech it is well known that the pre-planning duration depends -- among other factors -- on the need to avoid intra-sentence hesitation that are needed to plan upcoming words in parallel to production [e.g.
-@levelt2000word; @wheeldon_konopka_2023].
+For speech it is well known that the pre-planning duration depends -- among other factors -- on the need to avoid intra-sentential hesitations; speakers have a preference towards minimising language pre-planning depending on the extent to which upcoming words can be planned in parallel to production [e.g.
+@levelt2000word; @griffin2003reversed; @wheeldon_konopka_2023].
 For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech [@not10; dam09; @roeser2018advance].
@@
 
-Second, even though the average sentence-initial IKI tends to be longer, they still tend to be very rapid.
-For example, @medimorec2017pauses found that, in undergraduate students’ writing on familiar topics 71% of sentence-initial IKIs were less than 1 seconds in duration.
+Second, even though the average sentence-initial interkey interval tends to be longer, they still tend to be very rapid.
+For example, @medimorec2017pauses found that, in undergraduate students’ writing on familiar topics 71% of sentence-initial interkey intervals were less than 1 seconds in duration.
 Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses.
@@
 Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The fact that these very short sentence initial pauses exist, and in fact are quite common, is not consistent with the assumption that planning of the upcoming sentence is started and finished in the period between completion of the previous sentence and starting to write the next.
+The fact that these very short sentence-initial pauses exist, and in fact are quite common, is not consistent with the assumption that planning of the upcoming sentence is started and finished in the period between completion of the previous sentence and starting to write the next.
 
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that operate partly in parallel and largely without executive control.
+If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers composting multi-sentence texts typically occurs remarkably fluently, with very few hesitations.
+The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that operates partly in parallel and largely without executive control.
 Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction].
-@van1991handwriting, for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now].
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing the output of the previous sentence.
+@van1991handwriting, for example, argued for a cascade of modular processes each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now].
+Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes buffers allow fingers to catch up with message and language processing.
+Under this account when writers move from one sentence to another without hesitation this is due to then having completed required planning for the next sentence in parallel with output of the the sentence that they have just completed.
 
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two underlying mental processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then IKIs are determined just by the time needed for executing finger movements (i.e.
+This semi-parallel cascaded account of the composition process therefore gives a rather different understanding of inter-keystroke intervals.
+Interkey intervals result from one of two sets of underlying mental processes.
+When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then interkey intervals are determined just by the time needed for executing finger movements (i.e.
 just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly, then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling].
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
+However if one or more upstream processes provide output more slowly, then interkey intervals are determined just by time taken to complete upstream processes and not on finger movement.
+These two distinct data-generating processes mean that interkey intervals are a mixture of two distributions, one associated with the very rapid processing necessary for actioning keystrokes, and another that forms as a result of delays resulting from some combination of processing at the semantic, syntactic, lexical, or orthographic level not being complete at the time of the immediately preceding keystroke.
+[@roeser2021modelling].
+The parallel account provides an explanation for why writers sometimes hesitate before sentence (or word) onset, but often do not.
 This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
+
+<!--
+Hesitations in writing have potential disadvantages.
+Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
+However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck.
+Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
+If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
+@ronneberg2022process coined this possibility the "process-disruption hypothesis".
+-->
+
+Capturing this understanding of interkey intervals requires statistical models that are premised on more than one data-generating process.
+These models have been adopted in a number of psycholinguistic studies.
+For example, in language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental.
+In the context of written production @roeser2021modelling demonstrated^[For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.]  how models that allow the possibility of two underlying data-generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g.
+“dkdkdkdk”).
+As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition: copy-typing provides a typing context that is constrained to visual encoding of the target string and the retrieval of motor codes for letter combinations.
+For relatively experienced writers, visual encoding is likely to happen in parallel to writing [@de2018exploring]; hesitations occur when motor codes, for example for low frequency bigrams, were not retrieved in time [@roeser2021modelling].
+However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes.
+Typing hesitations in the context of copy-typing may – to some extent – reflect not just production but also, and possibly largely, visual encoding.
+Text composition, in contrast, provides data from the free generation of words and full sentences that is not dependent on a visually presented text stimulus.
+For these reasons our research extends the analysis of keystroke data using a mixture model approach to data from spontaneous text production.
+
+## Present Study
+
+In the study that is presented in this paper we use Bayesian hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- to identify whether the semi-parallel view of writing generalises to contexts in which participants compose texts.
+On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].
+
+The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
+In other words, we expect interkey intervals to be proportionally longer at before-sentence locations compared to before-word locations compared to within-word locations.
+This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
+The semi-parallel cascading view, in contrast, assumes that although interkey intervals at larger linguistic edges might be longer, it is in principle possible that writers do not (always) pause to plan their text but plan in parallel to writing.
+In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
+Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occur to some extent probabilistically across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
+We hypothesise that the statistical model of the parallel view provides a better out-of-sample generalisation than the implementation of the serial model of writing in the context of unconstrained spontaneous text composition.
+We tested these models across 6 datasets.
+
+For each dataset  we evaluated four Bayesian statistical models.
+The first three models were premised on the (traditional) single-distribution understanding of interkey intervals.
+These models are consistent with the assumption that processing leading to a keystroke is fundamentally serial.
+In the first model we assumed the interkey intervals were drawn from a Gaussian (normal)distribution.
+In the second we assumed a log-Gaussian distribution.
+In both cases, consistent with previous studies that have modelled inter-keystroke intervals, we assumed equal variance  across different linguistic edges (within words, at word boundaries, and at sentence boundaries).
+To provide a fairer comparison with the two-distribution model that followed we tested a third unequal variance model that allowed intervals associated with higher-order linguistic edges to have greater variability.
+
+The fourth model was premised on the assumption that the processing that results in a keystroke runs, in part, in parallel with preceding output.
+As we have discussed, this model assumes two data-generating processes.
+We therefore followed the finite (two-distribution) mixture-modelling approach described by@roeser2021modelling.
+This model assumes that interkey intervals result from a mixture of: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels that cause delays in this information flow.
+Therefore interkey intervals are not directly and sufficiently determined by transition location -- as in the serial account -- but are also associated with different probabilities that a hesitation occurs.
 
@@
 ```
-
-In language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental.
-Similar for keystroke data @roeser2021modelling demonstrated -- using a similar approach as used in the present paper^[For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.] -- how models that assume this mixture of two underlying data generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g.
-“dkdkdkdk”).
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
-However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes.
-Typing hesitations in the context of copy-typing may – to some extent – reflect not just production but also, and possibly largely, visual encoding.
-Text composition, in contrast, provides data from the free generation of words and full sentences that is not dependent on a visually presented text stimulus.
-For these reasons our research extends the analysis of keystroke data using a mixture model approach to data from spontaneous text production.
-
-In the study that is presented in this paper we test using Bayesian hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts.
-On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].
-
-The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word locations compared to within-word locations.
-This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not (always) pause to plan their text but plan in parallel to writing.
-In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
-Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occur to some extent probabilistically across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesise that the statistical model of the parallel cascading view provides a better out-of-sample generalisation than the implementation of the serial model of writing in the context of unconstrained spontaneous text composition.

### 2024-05-19 report/sections/methods.Rmd
change tutorial to contrained molg

@@
-In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts.
-Full modelling details can be found in Appendix \\ref{statistical-models; we summarise the most important model properties in the following.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data.
-In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.
-
-The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in @roeser2021modelling.
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes.
-These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs.
-This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.
-
-We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts.
-We discussed advantages with this approach in the introduction section.
-These datasets include samples from populations with various writing experience and languages (e.g.
+We analysed interkey intervals from six existing studies in which participants composed spontaneous, multi-sentence texts in response to writing prompts.
+The datasets include samples from populations with various writing experience and languages (e.g.
 young / L2 writers, students) performing different writing tasks (e.g.
@@
 
-TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2.
-For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say "unpublished".
-I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk).
-Was the GUNNEXP2 sample Norwegian?
-Should we include information like time limit or text length in the table?
-
 Six datasets with keystroke data from free text production were used for analysis.
 An overview can be found in Table \ref{tab:datasets}; descriptions below.
-Dataset used are C2l1 [@ronneberg2022process], CATO [@torrance2016adolescent], GUNNEXP2 [@torranceb], SPL2 [@torrancea], LIFT published in @vandermeulen2020 and described in @vandermeulen2020mapping, PLanTra data are published in @rossetti2022text and described in @rossetti2022s.
-In C2L1 Norwegian 6th graders composed argumentative essays.
-In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text.
-In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced.
-The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each.
-The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability.
-The GUNNEXP2 data ....
-[MARK, please add a sentence here to describe the production context].
-
-For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite [@sim07; @torrance201203].
-LIFT and PLanTra data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019] and SPL2 data were collected using CyWrite [@chukharev2019combined].
+Datasets used are C2l1 [@ronneberg2022process], CATO [@torrance2016adolescent], GE2 [@torranceb], SPL2 [@torrancea], LIFT published in @vandermeulen2020 and described in @vandermeulen2020mapping, and PLanTra published in @rossetti2022text and described in @rossetti2022s.
 
@@
 ```{r datasets, results='asis'}
-tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
+tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GE2"),
                 Source = c("Rønneberg et al. (2022)",
@@
                            "EyeWrite"),
-                `Writing task` = c("Argumentative essays", 
-                                 "Expository texts",
-                                 "Argumentative essays",
+                `Writing task` = c("Argumentative", 
+                                 "Expository",
+                                 "Argumentative",
                                  "Text simplification",
                                  "Synthesis", 
-                                 "?????"),
+                                 "Argumentative"),
                  N = as.character(c(126, 26*2, 39, 47, 658, 45)),
@@
 #                n_words = c(),
-                `Mean age` = c(11.8, 16.9, 20.6, 23, 17.0, "????"),
-                 Sample = c("6th graders", "Secondary school students", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
-                 Country = c("Norway", "Norway", "USA", "Belgium", "The Netherlands", "???? Norway"),
+                `Mean age` = c(11.8, 16.9, 20.6, 23, 17.0, 19),
+                 Sample = c("6th graders", "Secondary school students (dyslexic, non dyslexic)", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
+                 Country = c("Norway", "Norway", "USA", "Belgium", "The Netherlands", "UK"),
                  Language = c( "Norwegian", 
@@
                                "Dutch",
-                               "???? Norwegian")) %>% 
+                               "English")) %>% 
   arrange(Dataset) %>% 
@@
 
+In C2L1 Norwegian 6th graders composed argumentative essays.
+In CATO upper Norwegian secondary students with and without dyslexia composed expository texts either normally or with masked letters to prevent writers reading their unfolding text.
+In SPL2 undergraduate students produced argumentative essays in their first language (L1; English) and in a second language L2 (Spanish) in which they were able to compose text but in which their mastery fell well behind their first language.
+Order of language (L1 / L2) and two writing prompts were counterbalanced.
+The LIFT data are from pre-university students producing argumentative and informative text syntheses on four topics each.
+The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability, before and after receiving either online instruction on how to apply plain language principles to sustainability content or an online instruction exclusively on the topic of sustainability.
+The GE2 data are from Undergraduate students producing two argumentative essays on general-knowledge topics in either masked text or normal writing conditions, with masking, topic and order counterbalanced.
+
+For C2L1, CATO, and GE2, keystroke data were captured using EyeWrite [@sim07; @torrance201203].
+LIFT and PLanTra data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019] and SPL2 data were collected using CyWrite [@chukharev2019combined].
+
 ## Data extraction
 
-From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful [@chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-Also we removed transitions that were followed by an editing operation.
+From the keystroke data we extracted interkey intervals between adjacent keys at locations that have been identified repeatedly in previous research as locations where observed mean interkey intervals (or counts of interkey intervals over a predetermined threshold) have tended to vary substantially [e.g.
+@chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}.
+In particular we analysed the interkey intervals that resulted in the insertion of a character that started a new sentence (before-sentence transition); interkey intervals that started a new word (other than those at the beginning of a sentence;  before-word transitions); and transitions between characters within words all transitions.
+At before-sentence locations, interkey intervals were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GE2) but were timed to the character key press in the  PLanTra and LIFT datasets (i.e.
+the duration included the time to perform the capitalisting shift keypress); we return to this difference in the Results section.
+Transitions that occurred at the very start of the text or at the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
+Importantly we also removed transitions that were followed by an editing operation.
+We therefore just modelled the times between keypresses in ongoing production.
 
@@
 We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-We further removed keystroke intervals that were too short to represent a behavioural response ($\le$ 50 msecs) or too long ($\ge$ 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}.
+We removed participants that produced fewer than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
+We further removed keystroke intervals that were too short to represent a behavioural response ($\le$ 50 msecs) or were of a length where such that they were unlikely to be associated with ongoing text production ($\ge$ 30 secs); percentages can be found in Table \ref{tab:datareduction}.
 From the remaining data we randomly sampled a maximum of `r n_samples` observations per participant, condition, and transition location (when more than `r n_samples` were available).
-This was done to reduce the computation time of the Bayesian models require to complete sampling.
+This was done to reduce the computation time of the Bayesian models required to complete sampling.
 For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets.
@@
                 "cato" = "CATO",
-                "gunnexp2" = "GUNNEXP2",
+                "gunnexp2" = "GE2",
                 "lift" = "LIFT",
@@
           caption = "Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data included by transition location (`--` if all data were used).
+Mean percentage of extreme data removed and the mean percentage of randomly sampled data analysed by transition location (`--` if all data were used).
 Standard error in parentheses.",
@@
                               "Extreme values in \\%" = c(2, 3),
-                              "Randomly sampled data in \\%" = c(4, 6)))
+                              "Data analysed in \\%" = c(4, 6)))
 
@@
 
-## Bayesian statitical modelling
+## Statistical modelling
 
-We reanalysed keystroke data -- transition durations measured as the time between first and second keypress -- from six datasets in a series of five Bayesian models.
+We analysed keystroke data in a series of four Bayesian mixed-effects models.
 An overview of all models can be found in Table \ref{tab:models}.
-More detailed model descriptions and motivation can be found in Appendix \ref{statistical-models}.
+Full modelling details can be found in Appendix \ref{statistical-models} but we summarised the most relevant properties in this section.
 
-```{r models, results = 'asis'}
+The first three models -- models M1, M2, and M3 -- were single-distribution models consistent with the serial account.
+They were single distribution models, with fixed effects for each combination of transition location and dataset-specific manipulations.
+Models M1 (Guassian) and M2 (log-Guassian) are consistent with standard models used in the literature and therefore serve as baseline models.
+For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; see Appendix \ref{statistical-models} for rationale.
+
+```{r eval=F}
 models <- tibble(Models = str_c("M",1:5),
@@
   select(Models, Description)
+```
+
+```{r models, results = 'asis'}
+models <- tibble(Models = str_c("M",1:4),
+       Type = c("LMM", "LMM", "LMM", "MoG"),
+       Description = c("Single Gaussian model with effects of transition location (and experiment-specific manipulation).",
+                       "Single distribution log-Gaussian model equivialent to M1.",
+                       "Equivalent to M2 but with different variance components for each transition location (unequal variance).",
+                       "Two-distributions mixture of log-Gaussians for fluent interkey intervals and hesitant / disfluent interkey intervals; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition locations but the distribution of fluent intervals does not.")) %>%
+  select(Models, Description)
 
@@
           digits = 0,
-          caption = "Overview of cognitive models of writing. All models included by-participant random effect.",
+          caption = "Model overview. All models included by-participant random effect.",
           stub_indents = list(
                      "\\textbf{Serial}" = 1:3,
-                     "\\textbf{Cascading}" = 4:5))
+                     "\\textbf{Parallel}" = 4))
 
 ```
-
-Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset-specific manipulations.
-In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
-For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; this is because slower human behaviour is generally known to be associated with a larger variability [@wagenmakers2007linear; @wing1973response; @schoner2002timing]; see Appendix \ref{statistical-models} for rationale.
-
-Models M4 and M5 are extensions of the unequal-variance single distribution model M3.
-Importantly models M4 and M5 assume that keystroke data result from a combination of two data generating processes -- consistent with the cascading account -- rather than one.
-These model are called finite mixture models in the literature [@gelman2014; @roeser2021modelling; @peel2000finite] and capture the assumption that processing at higher levels of activation leads to longer pauses which can in principle happen at transition before sentence, before words or even mid-word.
-In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that pauses at larger linguistic edges are more likely but not obligated by the cognitive system that generates keystrokes.
-This is achieved by modelling key transitions as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
+Model M4 is two-distribution mixture models consistent with the parallel cascading account.
+This model assumes that keystroke data result from a combination of two data generating processes; these models are referred to as finite mixture models in the literature [@gelman2014; @peel2000finite].
+Both single-distribution models and the two-distribution mixture model capture that processing difficulty at higher levels of activation leads to longer pauses but only the two-distribution model captures that planning-related pauses may occur at any interkey interval, even mid-word, and that planning may not be reflected in sentence or word-initial pauses when planning happens in parallel to output.
+In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that key-transitions at larger linguistic edges are more likely to reflect processing delays but the cognitive system does not obligate planning to occur before sentences or words.
+This is achieved by modelling interkey intervals as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
 
@@
 These fluent keystroke transitions are merely constrained by a person's ability to move their finger.
-The population estimate for fluent key transitions (i.e.
-typing speed) is captured by the $\beta$ parameter.
+Fluent interkey intervals (i.e.
+typing speed) are captured by the $\beta$ parameter.
+$\beta$ is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon2} referring to the same unknown parameter.
 
 2.
-Any interruptions at upstream levels of mental representation result in a delay in the information flow and therefore create a lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
-The population estimate of the hesitation slowdown is captured by $\delta$ and its frequency is captured by the mixing weight (proportion) $\theta$.
+Any interruption at upstream levels of mental representation delays the information flow reflected in a resulting lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
+The slowdown for such hesitations is captured by $\delta$ and its frequency by the mixing proportion $\theta$.
+The slowdown $\delta$ was allowed to vary by transition locations because hesitations at larger linguistic units are likely to be associated with higher level planning which causes longer delays in the output.
+The $\delta$ parameter was constrained to be positive, so that it captures how much longer hesitant interkey intervals are in addition to $\beta$.
 
-  ```{=tex}
 \begin{equation}
@@
 (\#eq:bimodcon2)
-    \text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \times \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
 \text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
@@
 \end{equation}
-```
-The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents the distribution of fluent -- uninterrupted -- key transitions.
-Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
-$\theta$ is here parameterised to represent the probability that a key transition is associated with the distribution of hesitant transitions.
+
+The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant interkey intervals that include the $\delta$ parameter presenting the slowdown for hesitant transitions; the second line represents the distribution of fluent -- uninterrupted -- interkey intervals that is characterised by $\beta$ which is also present in the first file.
+Either of these two distributions is associated with the mixing proportion $\theta$ that ranges between 0 and 1.
+$\theta$ was parameterised here to represent the probability that an interkey interval is associated with the distribution of hesitant transitions.
 This probability is inversely related to the mixing weight of the distribution of fluent transitions by $1-\theta$ .
 In other words, a larger weight for either distribution inevitably means a lower weight for the other.
-We call this parameter the probability of hesitant transitions.
-When a parameter was assumed to vary by transition location (levels: before sentence, before word, within word) this was indicated as subscript; similar for participants.
-The hesitation probability was allowed to vary by participants constrained by a hyper-parameter for each transition location.
+We call this parameter the probability of hesitant interkey intervals.
+When a parameter is allowed to vary by transition location (levels: before sentence, before word, within word), this was indicated as subscript; similar for participants.
+For example, $\theta_\text{location[i], participant[i]}$ means that the mixing proportion $\theta$ is allowed to take on a different value for both the transition location and participant associated with the $i^\text{th}$ keystroke interval $i \dots \in N$ where $N$ is the total number of keystroke intervals.
+Because the hesitation probability was allowed to vary by participants as well as location type, we included a hyper-parameter for the mixing proportion $\theta$ for each transition location.
 
-The constrained implementation M4 assume that the size of a fluent key-transition does not vary by transition location.
-In other words, the value that the $\beta$ parameter is taking on is the same for transitions at before-sentence, before-word, and within-word locations.
-However, letter bigrams may be executed systematically faster than transitions between space and either a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
-This is for two reasons: first, and importantly, highly frequent bi- and even trigrams are stored, retrieved and executed as one motor code [@behmer2016crunching; @grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants] because of the strong association between letters.
-This association is reducing the length of transition durations.
-It is less likely that the same association exists for space-letter and space-shift pairs.
-Second, before-sentence transition in two datasets (PLanTra, LIFT) are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transition; we return to this in the results section.
-We therefore implemented an unconstrained mixture model M5 that assumes that the distribution over durations of fluent transitions -- represented as $\beta$ -- may vary across transition locations.
+As with model M3, M4 allowed the variance associated with the key-intervals indicated as $\sigma_{e'}^2$ to vary by transition location.
+Further each of the two distributions in M4 had its own variance constrained such that the variance associated with the distribution of hesitant keystrokes $\sigma_{e'}^2$ was larger than the distribution of fluent keystroke intervals $\sigma_{e}^2$.
+This is important because slower human behaviour similar to hesitant keystroke transitions is generally known to be associated with a larger variability [@wagenmakers2007linear; @wing1973response; @schoner2002timing].
+Model M4 and all other models included a random intercepts term $u_\text{participants}$ for participants which is constrained to be distributed around 0 with an unknown standard deviation, i.e.
+$\text{N}(0, \sigma_\text{p}^2)$, hence indicating participant specific deviations from the average typing speed $\beta$.
 
 All models were implemented in the Bayesian framework [@gelman2014; @mcelreath2016statistical].
-The Bayesian framework is ideal for the estimation of parameter values.
+In the present analysis, we used weakly informative priors to aid model convergence by constraining the parameter space to plausible values [see e.g.
+@lambert2018student; @mcelreath2016statistical].
+Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
+Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (https://osf.io/z65dw/); for a tutorial see [https://rpubs.com/jensroes/mixture-models-tutorial](rpubs.com/jensroes/mixture-models-tutorial).^[The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
+Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
+Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
+The predictive performance of our models was compared using leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical; @sivula2020uncertainty].]
+
+To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
+Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared by  the difference between models $\Delta\widehat{elpd}$.
+We also summarised this difference as normalised over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
+Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.
+
+<!-- The Bayesian framework is ideal for the estimation of parameter values.
 This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty [@farrell2018computational; @gelman2014; @lee2014bayesian].
@@
 In other words the choice of priors has less impact on the posterior, certainly for weakly informative priors.
-In the present analysis, we used weakly informative priors to aid model convergence by constraining the parameter space to plausible values [see e.g.
-@lambert2018student; @mcelreath2016statistical].
-Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
-
-Stan code for mixture models was based on [@roeser2021modelling; see also @vasishth2017; @vasishth2017feature] and can be found on OSF (ADD URL HERE); for a tutorial see [https://rpubs.com/jensroes/1000459](rpubs.com/jensroes/1000459).[^1]
-
-[^1]: The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
-Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
-Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.
-The predictive performance of our models was compared using leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical; @sivula2020uncertainty].
+-->

### 2024-05-19 report/sections/results.Rmd
change tutorial to contrained molg

@@
 ## Model comparisons
-
-To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
-Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared by the difference between models $\Delta\widehat{elpd}$.
-We also summarised this difference as normalised over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
-Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.
 
@@
          across(everything(), ~str_trim(.))) %>%
-  filter(dataset != "spl2_shift") %>% 
+  filter(dataset != "spl2_shift",
+         model != "M4 -- M5") %>% 
   transmute(
-    Comparison = model,
     elpd_diff_ratio,
@@
                             "c2l1" = "C2L1",
-                            "gunnexp2" = "GUNNEXP2")),
-    Distribution = case_when(str_detect(model, "M2$") ~ "Single log-Gaussian (M2)",
-                             str_detect(model, "M3$") ~ "Single log-Gaussian (unequal variance; M3)",
-                             str_detect(model, "M4$") ~ "Mixture of two log-Gaussians (constrained; M4)",
-                             str_detect(model, "M5$") ~ "Mixture of two log-Gaussians (unconstrained; M5)"),
+                            "gunnexp2" = "GE2")),
+    Comparison = case_when(str_detect(model, "M2$") ~ "Single log-Gaussian (M1 -- M2)",
+                               str_detect(model, "M3$") ~ "Single log-Gaussian (unequal variance; M2 -- M3)",
+                               str_detect(model, "M4$") ~ "Mixture of two log-Gaussians (M3 -- M4)",
+                             #str_detect(model, "M5$") ~ "Mixture of two log-Gaussians (unconstrained; M5)"
+                             ),
     across(elpd_diff, ~str_c(., " (", se_diff, ")")))
@@
 
-tr_m2 <- filter(table_ratios, str_detect(Comparison, "M2$")) %>% pull(range)
-tr_m3 <- filter(table_ratios, str_detect(Comparison, "M3$")) %>% pull(range)
-tr_m4 <- filter(table_ratios, str_detect(Comparison, "M4$")) %>% pull(range)
-tr_m5 <- filter(table_ratios, str_detect(Comparison, "M5$")) %>% pull(range)
+tr_m2 <- filter(table_ratios, str_detect(Comparison, "M2\\)$")) %>% pull(range)
+tr_m3 <- filter(table_ratios, str_detect(Comparison, "M3\\)$")) %>% pull(range)
+tr_m4 <- filter(table_ratios, str_detect(Comparison, "M4\\)$")) %>% pull(range)
+tr_m5 <- filter(table_ratios, str_detect(Comparison, "M5\\)$")) %>% pull(range)
 
@@
 
-Model differences are shown in Table \ref{tab:loos}.
-All datasets revealed the same pattern.
-Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single-distribution model; the difference $\Delta\widehat{elpd}$ between models ranged within `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model.
-The unconstrained mixture model M5 rendered a higher predictive performance than the constrained model M4 which did not allow the distribution of fluent keystroke-intervals to vary by transition location; the difference in predictive performance was between `r tr_m5` standard errors higher for the unconstrained model.
-Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; the difference $\Delta\widehat{elpd}$ ranged within `r tr_m3` standard errors.
-The weakest model was the single distribution Gaussian which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of `r tr_m2` standard errors.
-
-\blandscape
+The differences in the predictive performance of models for all datasets are shown in Table \ref{tab:loos}.
+For all six dataset the two-distributions mixture model (M4) provided greater predictive performance than all three single-distribution models.
+Differences in predictive performance ($\Delta\widehat{elpd}$) between models M4 and the nearest competing single-distribution model (M3) ranged between `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model.
+Among the single distribution models we found higher predictive performance for the unequal variance model and the lowest predictive performance for the single distribution Gaussian.
+Comparisons showing the fit of model predictions to the data can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
+Data predicted by the two-distribution mixture model closely align with the observed data; data predicted by single distribution Gaussian models showed an inferior fit to the data.
 
 ```{r loos, results='asis'}
-#names(table)[-1] <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 1))
-#names(table) <- c("Model", "Description", rep(c("$\\Delta\\widehat{elpd}$"), 1))
-
 table %>% 
   arrange(desc(Comparison)) %>% 
+  pivot_longer(-Comparison, names_to = "Dataset") %>% 
+  pivot_wider(names_from = Comparison, values_from = value) %>% 
+  relocate(Dataset, ends_with("M4)")) %>% 
   apa_table(caption = "Model comparisons.
 Models were compared incrementally from the simplest to the most complex model.
-The comparison is shown along with the distribution type of the more complex model.
+The comparison is shown along with the distribution type of model with the higher predictive performance.
 A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model.
 Standard error is shown in parentheses.",
-            align =c("p{1.5cm}", "p{5cm}", rep("r", 6)), 
+            align =c("p{1.75cm}", rep("p{4cm}", 3)), 
             escape = FALSE,
+            row.names = T,
+            col.names = c("Dataset", "Mixture of two log-Gaussians (M3 -- M4)", "Single log-Gaussian (unequal variance; M2 -- M3)", "Single log-Gaussian (M1 -- M2)"),
             longtable = T,
-            font_size = "footnotesize",
+            font_size = "small",
             note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density")
-#                col_spanners = list(" " = 1:2,
-#                     "\\textbf{CL21}" = 1:5,
-#                     "\\textbf{CATO}" = 6:10,
-#                     "\\textbf{GUNNEXP2}" = 11:15, 
-#                     "\\textbf{LIFT}" = 16:20,
-#                     "\\textbf{PLanTra}" = 21:25,
-#                     "\\textbf{SPL2}" = 26:30))
-#      stub_indents = list(
-#                     "\\textbf{CL21}" = 1:5,
-#                     "\\textbf{CATO}" = 6:10,
-#                     "\\textbf{GUNNEXP2}" = 11:15, 
-#                     "\\textbf{LIFT}" = 16:20,
-#                    "\\textbf{PLanTra}" = 21:25,
-#                     "\\textbf{SPL2}" = 26:30))
-#                     "\\textbf{SPL2 (shift + C)}" = 31:35))
 
@@
 
-\elandscape
-
-We also tested to what extent model predictions fit observed data.
-Comparisons can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
-Data predicted by the two-distribution mixture models are similar to observed data; data predicted by single distribution Gaussian models showed a poor fit to the data.
-
 ## Transition location effect
 
-As demonstrated in the model comparisons, the two-distributions mixture-models is a better statistical representation on the underlying process generating keystroke data than compared to any of the single-distribution models.
-Information about this process is captured by the posterior estimates of the models' parameter values.
-To reiterate, there are three conceptually important parameters: (1) the magnitude of the slowdown for transition hesitations (represented as $\delta$ in equation \ref{eq:bimodcon2}), (2) the probability of hesitant transitions (represented as $\theta$ in equation \ref{eq:bimodcon2}), and (3) the duration of fluent transitions (the typing speed, which was indicated as $\beta$ in equation \ref{eq:bimodcon2}).
-
-Figure \ref{fig:mixmodel} illustrates the mixture model approach on the basis of the posterior of the constrained implementation applied to the CATO data: for each transition durations we obtain two posterior probability distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in hesitant transitions.
-The hesitation probability is indicated as the relative height of either distribution.
-The transition locations are characterized by the hesitation parameters: as we illustrate the constrained mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location but their height differs as the probability of fluent transition is the inverse of the height of hesitant transitions; i.e.
+Figure \ref{fig:mixmodel} illustrates the mixture model results for every dataset: for each interkey interval we obtained two posterior probability distributions, one associated with fluent transitions between keys and another for interkey intervals where upstream difficulty resulted in hesitant interkey intervals.
+The hesitation probability -- the probability of  interkey intervals associated with hesitations -- is indicated as the height of the distribution shown in yellow (on the right of each panel).
+The transition locations are characterised by the hesitation parameters: as we illustrate the mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location (within each dataset) but their height differs as the probability of fluent interkey intervals is the inverse of the height of hesitant interkey intervals; i.e.
 $1 - \theta$.
-More interesting are the distributions associated with hesitations; indicated in yellow: Within word transitions have a negligibly small hesitation probability barely visible in the visualisation.
+Within-word interkey intervals have a negligibly small hesitation probability barely visible in the visualisation.
 Hesitations before words and sentences are both roughly equally likely than fluent transitions.
 In other words, half of the time participants did not pause before a word or sentence; we explore this below.
-However when hesitations occurred, these are longer at before-sentence transitions compared to before word transitions.
+However, when hesitations occurred, these are longer at before-sentence interkey intervals compared to before word interkey intervals.
+The slowdown for hesitant interkey intervals -- represented earlier as $\delta$ -- is the distance between the distribution of fluent interkey intervals ($\beta$) shown in grey and the distribution of hesitant interkey intervals shown in yellow.
 
 ```{r mixmodel, fig.cap='Mixture model plot for each transition location.
-Shown are the distributions for both mixture components (fluent and hesitant transitions) and their weighting.
-Posterior of CATO data.', fig.height=6}
+Shown are the distributions for both mixture components (fluent and hesitant interkey intervals) and their weighting.
+Interkey intervals are shown in seconds on log scaled intervals.', fig.height=8, fig.width=7}
 
-ps <- read_csv("../../stanout/cato/mog_constr_posterior_for_densityplot.csv") 
+# Data grid
+data <- tibble(x = c(0, 100))
+
+# plotting details
+colours <- c("Fluent" = "#000000", "Hesitant" = "#E69F00")
+ylims <- c(0, .7)
+xbreaks <- log(c(1, 10, 100, 1000, 10000))
+xlims <- log(c(10, 50000))
+subtitle_bs <- "Before sentence"
+subtitle_bw <- "Before word"
+subtitle_ww <- "Within word"
+
+# get files
+files <- list.files("../../stanout", pattern = "mog_constr_posterior_for_densityplot.csv", recursive = T, full.names = T)
+
+# load all posteriors
+ps_all <- map_dfr(files, ~read_csv(.) %>% mutate(dataset = .x)) %>% 
+  mutate(dataname = case_when(str_detect(dataset, "spl2") ~ "SPL2",
+                              str_detect(dataset, "lift") ~ "LIFT",
+                              str_detect(dataset, "plantra") ~ "PLanTra",
+                              str_detect(dataset, "gunnexp2") ~ "GE2",
+                              str_detect(dataset, "cato") ~ "CATO",
+                              str_detect(dataset, "c2l1") ~ "C2L1"))
+
+# plot counter
+plot_id <- 0
+
+for(i in files){
+
+  if(plot_id == 0){
+    # template for output
+    plot_bs <- list()
+    plot_bw <- list()
+    plot_ww <- list()
+  }
+
+  plot_id <- plot_id + 1
+
+  # select dataset
+  ps <- filter(ps_all, dataset == i)
+
+  # add data name
+  data <- mutate(data, dataname = unique(pull(ps, dataname)))
+
   # Extract parameters
@@
 
-# Data grid
-data <- tibble(x = c(0, 20))
-
-colours <- c("Fluent" = "#000000", "Hesitant" = "#E69F00")
-
   # Create density plot
-plot_bs <- ggplot(data, aes(x)) +
+  plot_bs[[plot_id]] <- ggplot(data, aes(x)) +
     stat_function(aes(colour = "Fluent"),
@@
                   args = list(alpha2_bs, sigmap_bs, lam = theta_bs)) +
-  scale_x_continuous(labels = exp, 
-                     breaks = log(c(1, 10, 100, 1000, 10000)), 
-                     limits = log(c(10, 50000))) +
+    scale_x_continuous(labels = ~exp(.) / 1000, 
+                       breaks = xbreaks, 
+                       limits = xlims) +
+    scale_y_continuous(limits = ylims) +
     scale_colour_manual(values = colours) +
     scale_fill_manual(values = colours) +
-  labs(y = "", x = "", 
-       subtitle = "Before-sentence transitions",
-       colour = "Mixture\ncomponent",
-       fill = "Mixture\ncomponent") +
-  theme(legend.position = "right",
-        legend.justification = "top")
+    labs(colour = "Mixture component",
+         fill = "Mixture component")
 
-plot_bw <- ggplot(data, aes(x)) +
+  plot_bw[[plot_id]] <- ggplot(data, aes(x)) +
     stat_function(aes(colour = "Fluent"),
@@
                   args = list(alpha2_bw, sigmap_bw, lam = theta_bw)) +
-  scale_x_continuous(labels = exp, 
-                     breaks = log(c(1, 10, 100, 1000, 10000)), 
-                     limits = log(c(10, 50000))) +
+    scale_x_continuous(labels = ~exp(.) / 1000, 
+                       breaks = xbreaks, 
+                       limits = xlims) +
+    scale_y_continuous(limits = ylims) +
     scale_colour_manual(values = colours) +
     scale_fill_manual(values = colours) +
-  labs(y = "Density", x = "", 
-       subtitle = "Before-word transitions",
-       colour = "Mixture\ncomponent",
-       fill = "Mixture\ncomponent") +
-  theme(legend.position = "none")
+    labs(colour = "Mixture component",
+         fill = "Mixture component")
 
-plot_ww <- ggplot(data, aes(x)) +
+  if(plot_id != 6){
+    plot_bs[[plot_id]] <- plot_bs[[plot_id]] + 
+      theme(legend.position = "none",
+            axis.title = element_blank(),
+            axis.text = element_blank(),
+            axis.ticks = element_blank())
+    plot_bw[[plot_id]] <- plot_bw[[plot_id]] + 
+      theme(legend.position = "none",
+            axis.title = element_blank(),
+            axis.text = element_blank(),
+            axis.ticks = element_blank())
+  } else if(plot_id == 6){
+    plot_bs[[plot_id]] <- plot_bs[[plot_id]] + 
+      theme(legend.position = "none",
+            axis.title = element_blank(),
+            axis.text.y = element_blank(),
+            axis.ticks.y = element_blank())
+    plot_bw[[plot_id]] <- plot_bw[[plot_id]] + 
+      theme(legend.position = "none",
+            axis.title = element_blank(),
+            axis.text.y = element_blank(),
+            axis.ticks.y = element_blank())
+  }
+
+  plot_ww[[plot_id]] <- ggplot(data, aes(x)) +
     stat_function(aes(colour = "Fluent"),
@@
                   args = list(alpha2_ww, sigmap_ww, lam = theta_ww)) +
-  scale_x_continuous(labels = exp, 
-                     breaks = log(c(1, 10, 100, 1000, 10000)), 
-                     limits = log(c(10, 50000))) +
+    scale_x_continuous(labels = ~exp(.) / 1000, 
+                       breaks = xbreaks, 
+                       limits = xlims) +
+    scale_y_continuous(limits = ylims) +
     scale_colour_manual(values = colours) +
     scale_fill_manual(values = colours) +
-  labs(subtitle = "Within-word transitions", 
-       y = "", x = "Transition duration [in msecs]",
-       colour = "Mixture\ncomponent",
-       fill = "Mixture\ncomponent") +
-  theme(legend.position = "none")
+    facet_grid(dataname~.) +
+    labs(colour = "Mixture component",
+         fill = "Mixture component")
 
-plot_bs / plot_bw / plot_ww
+   if(plot_id != 6){
+    plot_ww[[plot_id]] <- plot_ww[[plot_id]] + 
+      theme(legend.position = "top",
+            axis.title = element_blank(),
+            axis.text = element_blank(),
+            axis.ticks = element_blank())
+  } else{
+    plot_ww[[plot_id]] <- plot_ww[[plot_id]] + 
+      theme(legend.position = "top",
+            axis.title = element_blank(), 
+            axis.text.y = element_blank(),
+            axis.ticks.y = element_blank())
+  }
+}
+
+bs_lab <- ggplot(mapping = aes(x = 0, y = 0, label = subtitle_bs)) + 
+  geom_text() +
+  theme_void()
+
+bw_lab <- ggplot(mapping = aes(x = 0, y = 0, label = subtitle_bw)) + 
+  geom_text() +
+  theme_void()
+
+ww_lab <- ggplot(mapping = aes(x = 0, y = 0, label = subtitle_ww)) + 
+  geom_text() +
+  theme_void()
+
+# combine plots
+plots_fin <- wrap_plots(
+  wrap_plots(bs_lab, wrap_plots(plot_bs, ncol = 1), heights = c(.025, .975)),
+  wrap_plots(bw_lab, wrap_plots(plot_bw, ncol = 1), heights = c(.025, .95)),
+  wrap_plots(ww_lab, wrap_plots(plot_ww, ncol = 1), heights = c(.025, .975)), guides =  "collect")
+
+# add axes titles
+grid.arrange(patchworkGrob(plots_fin), 
+                          left = "Density",
+                          bottom = "Interkey interval in secs")
+#ggsave("mixdists.png", plots_fin, width = 12, height = 6)
 
@@
 
-We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
+<!-- We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
 We demonstrated above that the predictive performance of the unconstrained mixture model was substantially higher than the predictive performance of the constrained mixture model.
@@
 We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following.
+-->
 
@@
          !(str_detect(data, "CATO") & task == "masked" & !is.na(task)),
-         !(str_detect(data, "GUNNEXP2") & xn == "masked" & !is.na(xn)))
+         !(str_detect(data, "GE2") & xn == "masked" & !is.na(xn)))
 
@@
 
-The mixture-model posterior parameter estimates for all datasets is visualised in Figure \ref{fig:crossstudypost2}.
-Although models were fitted with all data set-specific conditions, we aggregated the posterior across conditions[^2], and excluded conditions that might conflate comparisons[^3].
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences and similarities associated with transition locations for all dataset.
-In particular we can see how the linguistic location in the text is associated with hesitation duration and probability.
-The estimate for fluent transitions is shown by dataset, across transition location.
-
-[^2]: We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.
-
-[^3]: We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-gunnexp2}).
+Figure \ref{fig:mixmodel} highlighted how model parameters relate to each other in pairs of distributions but to compare the effect of transition location on model parameters we need to compare the posterior distribution of the three parameters of interest individually.
+Therefore Figure \ref{fig:crossstudypost2} visualises the same posterior as Figure \ref{fig:mixmodel} but expressed directly in terms of posterior probability distributions of the individual mixture-model parameters.
+These are shown for all datasets by transition location.
+Although models were fitted with all dataset-specific conditions, we aggregated the posterior across conditions^[We aggregated across pre and post tests for the PLanTra dataset as well as genre and topic of the LIFT data set.
+We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.], and excluded conditions that might confound comparisons^[We removed the masked writing condition in the GE2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
+There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-ge2}).].
+For posteriors of all conditions within datasets see Appendix \ref{posterior-parameter-estimates}.
+The resulting posterior allows us to examine differences and similarities associated with transition locations for all datasets.
+In particular we can see how transition location in the text is associated with hesitant interkey intervals and hesitation probability.
+For completeness we also report the estimate for fluent interkey intervals.
 
@@
 Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group."}
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group."}
 
@@
 dotsize <- 2.75
-grouplabel <- "Data set:"
+grouplabel <- "Dataset:"
 shapes <- c(1, 4, 6, 7, 8, 9, 11)
@@
 Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets.
-We summarise the results of the pairwise comparisons for transitions locations; full results and parameter estimates can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
-We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs \> 100 for all datasets) but only for half of the data sets also longer (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
-Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GUNNEXP2 and SPL2, both BFs \> 100) but longer (all BFs \> 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
+We summarise the evidence for differences between transitions locations in Table \ref{tab:loceffectconbf}.
+We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs > 100 for all datasets) but only for half of the datasets hesitation durations were longer (GE2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
+Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GE2 and SPL2; both BFs > 100) but longer (all BFs > 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
+Full results and parameter estimates can be found in Appendix \ref{transition-location-effect}.
 
-To reiterate, for most datasets we found that pauses are before sentences are longer but not necessarily more likely than before words (except for datasets SPL2 and GUNNEXP2).
+```{r loceffectconbf, results='asis'}
+
+files <- list.files(str_c("../../stanout/", 
+                          c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetacon.+.csv", full.names = T) # "spl2_shift",
+
+ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
+                                 mutate(data = .x)) %>% 
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
+         across(data, ~recode(., 
+                              "cato" = "CATO",
+                              "spl2" = "SPL2", 
+                              "gunnexp2" = "GE2",
+                              #                "spl2_shift" = "*SPL2* (L1; shift + C)", 
+                              "plantra" = "PLanTra",
+                              "lift" = "LIFT",
+                              "c2l1" = "C2L1"))) %>% 
+  filter(!(str_detect(data, "SPL2") & lang == "ES"),
+         !(str_detect(data, "CATO") & group == "dyslexic"),
+         !(str_detect(data, "CATO") & task == "masked"),
+         !(str_detect(data, "GE2") & xn == "masked"),
+         param %in% c("delta", "theta")) %>%
+  select(param, location, data, value) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
+  pivot_wider(names_from = location, values_from = value) %>%
+  unnest(-c(data, param)) %>% 
+  mutate(diff.1 = `before sentence` - `before word`,
+         diff.2 = `before word` - `within word`) %>% 
+  summarise(across(starts_with("diff"), 
+                   list(mean = mean, 
+                        lower = lower, 
+                        upper = upper, 
+                        BF = BF)),
+            .by = c(data, param))
+
+tmp <- ps_loc_diffs %>% 
+  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
+  mutate(across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
+  select(-mean, -lower, -upper) %>% 
+  pivot_wider(names_from = c(diffid, param), values_from = BF) %>% 
+  select(data, contains("delta"), contains("theta")) %>% 
+  arrange(data)
+
+apa_table(tmp,
+          align = c("l", rep("p{3cm}", 4)), 
+#          font_size = "small",
+          longtable = T,
+          escape = FALSE,
+          row.names = T,
+          col.names = c("Dataset", rep(c("before sentence vs word", "before vs within word"), 2)),
+          col_spanners = list(" " = c(1),
+                              "Hesitation slowdown" = c(2, 3), 
+                              "Hesitation probability" = c(4, 5)),
+          note = "BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
+          caption = "BFs for the effect of transition location on interkey intervals shown for the hesitation slowdown and hesitation probability.
+Comparisons for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.")
+
+```
+
+To reiterate, for most datasets we found that pauses before sentences are longer but not always more likely than before words (except for SPL2 and GE2).
 In fact, LIFT showed some indication that pauses were less likely before sentences than words.
-This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for key transitions at larger linguistic edges.
+This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for interkey intervals at larger linguistic edges (see Introduction section).
 It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.
 
-Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character while the remaining data sets measured before-sentence transitions as the difference between space key and shift.
-To control for the use of complex key combinations, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability (BFs \> 100) but it did not affect the duration of fluent or hesitant transitions (BFs \< 0.3); see Appendix \ref{key-combination-effect-constrained-mixture-model}.
-This conflicts the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did included the character following shift (PLanTra, LIFT).
-In other words, patterns in our results cannot be explained on the basis of how before-sentence key-transitions were operationalised.
+Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character, while the remaining data sets measured before-sentence transitions as the interval between space key and shift.
+To test to what extent complex key combinations affects the parameter estimates, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability for sentence-initial transitions (BFs > 100) but it did not affect the duration hesitant interkey intervals (BFs < 0.3); see Appendix \ref{key-combination-effect}.
+This conflicts with the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did include the character following shift (PLanTra, LIFT).
+In other words, patterns in our results cannot be explained on the basis of how sentence-initial key transitions were operationalised.
 
-Finally, the duration of fluent transitions was not distinguished by location in the constrained distribution model.
-Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transitions.
-This is presumably reflecting that these data were sampled from was the youngest and least experienced population of writers in our data pool.
-Hesitation duration and probability followed the same patterns as the majority of datasets.
+Finally, Figure \ref{fig:crossstudypost2} highlights that fluent interkey intervals in two datasets (C2L1, CATO) were substantially longer than for other datasets.
+This is presumably reflecting that these data were sampled from the youngest and least experienced population of writers in our data pool.
+Interestingly, however, hesitation duration and hesitation probability followed the same patterns as the majority of datasets.
 
-The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
+<!-- The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
 We, therefore, summarise briefly differences in the results for the transition-location effects; full results can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs \> 100) but not before-sentence compared to before-word transitions (all BFs \< 0.08, but SPL2 and GUNNEXP2: BFs \> 100).
-Patterns for hesitation probability was generally similar in both the constrained and unconstrained model.
-However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs \> 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.
+Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs > 100) but not before-sentence compared to before-word transitions (all BFs < 0.08, but SPL2 and GE2: BFs > 100).
+Patterns for hesitation probability were generally similar in both the constrained and unconstrained model.
+However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs > 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.
+-->

### 2024-05-19 tutorial/tutorial.Rmd
change tutorial to contrained molg

@@
 
-We reduced the data to three variables: a participant identifier `participant`, the inter-keystroke intervals `iki`, and the transition location `location` (levels: within-word, before-word, before-sentence).
+We reduced the data to three variables: a participant identifier `participant`, the interkey intervals `iki`, and the transition location `location` (levels: within-word, before-word, before-sentence).
 
@@
 
-We reduced the number of keystroke transitions to `n` observations per participant and per location.
+We reduced the number of interkey intervals to `n` observations per participant and per location.
 
@@
 We start with a Gaussian mixed-effects model.
-We describe the process that generates the $i^\text{th}$ keystroke, where $i \in 1\dots N$ and $N$ is the total number of keystroke, as a single Gaussian (normal) distribution $\mathcal{N}()$ which is characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
+We describe the process that generates the $i^\text{th}$ interkey interval, where $i \in 1\dots N$ and $N$ is the total number of keystroke, as a single Gaussian (normal) distribution $\mathcal{N}()$ which is characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
 The mean can be decomposed into $\beta$ and $u_\text{participant[i]}$.
 $\beta$ will be allowed to take on a different value for each transition location associated with the $i^\text{th}$ keystroke.
-By-participant ikis are assumed to vary around the average which is achieved by placing a normal distribution on the difference between average and by-participant ikis, indicated as $u_\text{participant[i]}$, with a mean of 0 and a standard deviation $\sigma_\text{p}^2$.
+By-participant interkey intervals (ikis) are assumed to vary around the average which is achieved by placing a normal distribution on the difference between average and by-participant ikis, indicated as $u_\text{participant[i]}$, with a mean of 0 and a standard deviation $\sigma_\text{p}^2$.
 
@@
 The choice of priors here is weakly informative.
-The intention is to use knowledge that we generally have about how long key intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values.
-Even if we know very little about the duration of keystroke interval, we do know that they shouldn't be longer than a day, or event an hour, or several minutes, because such long intervals are unlikely to reflect active writing.
+The intention is to use knowledge that we generally have about how long interkey intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values.
+Even if we know very little about the duration of interkey intervals, we do know that they shouldn't be longer than a day, or event an hour, or several minutes, because such long intervals are unlikely to reflect active writing.
 
@@
 This model is largely identical to the Gaussian mixed-effect model in the previous section, except instead of a normal distribution this model assumes the ikis come from a single underlying process that follows a log-normal distribution.
-Assuming a log-normal distribution is useful as ikis and other response times are zero-bound (keystroke intervals cannot be shorter than 0).
+Assuming a log-normal distribution is useful as ikis and other response times are zero-bound (interkey intervals cannot be shorter than 0).
 A log-normal distribution forces responses to be positive.
@@
 
-The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent keystroke transitions ($\beta$), the slowdown for hesitant transitions ($\delta$) and the probability of hesitant transitions ($\theta$).
-In the model below, these are estimated each level of a categorical predictor `location` for the $i^\text{th}$ keystroke interval.
+The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$).
+In the model below, these are estimated each level of a categorical predictor `location` for the $i^\text{th}$ interkey intervals.
 
@@
 \begin{align}
-\text{iki}_{i} \sim & \theta_\text{location[i],participant[i]} \times \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i],participant[i]}) \times \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{iki}_{i} \sim & \theta_\text{location[i],participant[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i],participant[i]}) \times \text{log}\mathcal{N}(\beta+ u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
 		&\text{where}\\
@@
 The illustrate shows the dissociation of two underlying processes that the model is estimating on the basis of the data.
-There are two distributions of which the left one in grey is the distribution of fluent keystroke intervals; the right distribution in green is wider and captures longer keystroke intervals (with some overlap between fluent and disfluent transitions).
-
-```{r mixturemodel, echo = F, fig.cap="What's a mixture model for keystroke intervals?
+There are two distributions of which the left one in grey is the distribution of fluent interkey intervals; the right distribution in green is wider and captures longer interkey intervals (with some overlap between fluent and disfluent interkey intervals).
+
+```{r mixturemodel, echo = F, fig.cap="What's a mixture model for interkey intervals?
 The parameters of the mixture model explained visually.
@@
 
-Importantly Figure \@ref(fig:mixturemodel) illustrates the average typing speed of fluent keystroke transitions captured by the $\beta$ parameter; the distance between fluent transitions and disfluent transitions captured by $\delta$ (i.e.
+Importantly Figure \@ref(fig:mixturemodel) illustrates the average typing speed of fluent interkey intervals captured by the $\beta$ parameter; the distance between fluent transitions and disfluent transitions captured by $\delta$ (i.e.
 the distance between the grey and the green distribution); the hesitation probability or the "pausing frequency" which is the height of the green distribution captured by $\theta$.
-Most keystrokes are, typically, fluent transitions and hence part of the grey distribution.
-The frequency of the minority of disfluent transitions is captured by the height of the green distribution.
-As this model assumes a finite mixture of two distributions, the probability of fluent transitions is the inverse $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent transitions is $.6$.
-If the pausing probability decreases to $.3$ the proportion of fluent transitions had increased to $.7$ etc.
+Most interkey intervals are, typically, fluent and hence part of the grey distribution.
+The frequency of the minority of disfluent interkey intervals is captured by the height of the green distribution.
+As this model assumes a finite mixture of two distributions, the probability of fluent interkey intervals is the inverse $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent interkey intervals is $.6$.
+If the pausing probability decreases to $.3$ the proportion of fluent interkey intervals had increased to $.7$ etc.
 
@@
 We assign the participant and location identifiers created above and their maximum values `nS` and `K`, respectively.
-The keystroke data `iki` are assigned to `y` and the total number of observations is assigned to `N` (number of rows in the data `nrow(data`).
+The interkey interval data `iki` are assigned to `y` and the total number of observations is assigned to `N` (number of rows in the data `nrow(data`).
 
@@
 
-We will return to the interpretation of these estimates below.
+We return to the interpretation of these estimates below.
 
@@
 If the chains overlap and look like "fat hairy caterpillars", chains have converged on the same target distribution.
-The chains below look good except for one divergent transition for `beta[3]`.
+The chains below look good except for a divergent transition for `theta[3]`.
 
@@
 ```{r}
-posterior <- as.data.frame(fit_molg, pars) %>% as_tibble() 
+posterior <- as.data.frame(fit_molg, pars) %>% as_tibble() %>% rownames_to_column("idx")
 posterior
@@
 ```{r}
-posterior_long <- pivot_longer(posterior, everything(), 
+posterior_long <- pivot_longer(posterior, 
+                               cols = -c(idx, beta), 
                                names_to = c(".value", "location"),
@@
 `theta` samples are currently on the logit scale and can be transformed to probabilities using `brms::inv_logit_scaled` (the model also contained a `prob` parameter which is `theta` but on the probability scale).
-The `recode` function changes the location indices from 1 to "before-sentence", 2 to "before-word", 3 to "within-word".
+The `recode` function changes the location indices from 1 to "before sentence", 2 to "before word", 3 to "within word".
 
@@
                              across(theta, inv_logit_scaled),
-                             across(location, recode, `1` = "before-sentence",
-                                                      `2` = "before-word",
-                                                      `3` = "within-word"))
+                             across(location, recode, `1` = "before sentence",
+                                                      `2` = "before word",
+                                                      `3` = "within word"))
 posterior_in_msecs
@@
 ```{r}
-pivot_longer(posterior_in_msecs, beta:theta) %>%
+pivot_longer(posterior_in_msecs, c(beta, delta, theta)) %>%
+  mutate(across(location, ~case_when(str_detect(name, "beta") ~ "overall",
+                                     TRUE ~ .)), # correct location for beta
+         across(location, ~factor(.,             # order locations for plotting
+                                  levels = c("before sentence", 
+                                             "before word", 
+                                             "within word",
+                                             "overall"), 
+                                  ordered = TRUE))) %>% 
+  unique() %>%  # remove duplicated values for beta 
   ggplot(aes(x = value, colour = location, fill = location)) +
@@
 ```{r}
-pivot_longer(posterior_in_msecs, beta:theta) %>% 
+pivot_longer(posterior_in_msecs, c(beta, delta, theta)) %>% 
+  mutate(across(location, ~case_when(str_detect(name, "beta") ~ "overall",
+                                     TRUE ~ .))) %>% 
+  unique() %>% 
   summarise(across(value, list(mean = mean), 
@@
 We can calculate the differences between the transition locations for each parameter to determine to what extent each of the parameters are sensitive to transition locations.
-This will show us to what extent fluent typing ($\beta$), the size of the hesitation slowdown ($\delta$), and the hesitation probability ($\theta$) are sensitive to the linguistic location of a key pair.
+This will show us to what extent fluent interkey intervals ($\beta$), the size of the hesitation slowdown ($\delta$), and the hesitation probability ($\theta$) are sensitive to the linguistic location of a key pair.
 
@@
 posterior_diffs <- posterior_in_msecs %>%
-  pivot_longer(beta:theta, names_to = "parameter") %>%
+  select(-idx) %>% 
+  pivot_longer(delta:theta, names_to = "parameter") %>%
   mutate(id = row_number(), .by = location) %>%
@@
   select(-id) %>% # drop id column
-  mutate(diff_sent = `before-sentence` - `before-word`, # compare sentences to word
-         diff_word = `before-word` - `within-word`, # compare within to before word
+  mutate(diff_sent = `before sentence` - `before word`, # compare sentences to word
+         diff_word = `before word` - `within word`,     # compare within to before word
          parameter = paste0(parameter, "[diff]")) %>% 
@@
 
-The difference between the conditions can be summarised using the mean, the 95% PIs and the probability that the difference between the components is smaller than 0 (indicated as e.g.
-$P(\hat{\beta} <0)$ for the parameter $\beta$; the hat symbol indicates $\hat{.}$ that the value is an estimate of the population parameter value).
+The difference between the conditions can be summarised using the mean, the 95% PIs and the probability that the difference between transition locations is smaller than 0 (indicated as e.g.
+$P(\hat{\theta} < 0)$ for the parameter $\theta$; the hat symbol indicates $\hat{.}$ that the value is an estimate of the population parameter value).
 This summary tells us how certainly we can rule out a null difference between transition locations the most probable value for the differences.
@@
 
-For example, form these results we can infer that there is a $P(\hat{\theta} <0)$ = `r theta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` chance that hesitations do not occur more frequently before sentence; in fact the results show hesitations are more probable my `r theta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(2)` to occur before sentences compare to before words (95% PI: `r theta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(2)`, `r theta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(2)`).
-Also, the slowdown associated with hesitations is `r delta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(0)` msecs longer for sentences (95% PI: `r delta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(0)`, `r delta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(0)`) with a probability of $P(\hat{\delta} <0)$ = `r delta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` to be smaller than zero.
+For example, form these results we can infer that there is a $P(\hat{\theta} < 0)$ = `r theta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` chance that hesitations do not occur more frequently before sentence; in fact the results show hesitations are more probable by `r theta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(2)` to occur before sentences compare to before words (95% PI: `r theta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(2)`, `r theta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(2)`).
+Also, the slowdown associated with hesitations is `r delta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(0)` msecs longer for sentences (95% PI: `r delta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(0)`, `r delta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(0)`) with a probability of $P(\hat{\delta} < 0)$ = `r delta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` to be smaller than zero.
 
@@
   ggplot(aes(x = value, colour = name, fill = name)) +
+  geom_vline(xintercept = 0, colour = "black", linetype = "dotted", size = .5) +
   geom_histogram(position = "identity", alpha = .25) +
@@
   scale_color_brewer("Difference", palette = "Dark2") +
-  geom_vline(xintercept = 0, colour = "red", linetype = "dashed", size = .5) +
   labs(x = "Difference between transition locations") +
@@
 Hesitations before sentences were `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(mean) %>% round(0)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(upper) %>% round(0)`) longer than before words.
-Hesitations before words were `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer than within words, although the probability interval (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(upper) %>% round(0)`) did not rule out a negative difference, or no difference, which however had a low probability $P(\hat{\delta} <0)$ = `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(p_diff) %>% round(2)`.
-
-For fluent transitions there was evidence that within-word transitions were `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer (95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper) %>% round(0)`) compared to before-word transitions.
+Hesitations before words were `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer than within words, although the probability interval (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(upper) %>% round(0)`) did not rule out a negative difference, or no difference, which however had a low probability $P(\hat{\delta} < 0)$ = `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(p_diff) %>% round(2)`.
+
+<!-- For fluent transitions there was evidence that within-word transitions were `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer (95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper) %>% round(0)`) compared to before-word transitions.
 There was negligible evidence for longer transitions for before-sentence transitions compared to before-word transitions ($\hat{\beta}$ = `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(mean) %>% round(0)` msecs, 95% PI: `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(upper) %>% round(0)`).
+-->
 
@@
 
-The ROPE for hesitation durations for the difference between before-word and within-word transitions contains `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% of the posterior samples.
+The ROPE for the hesitation duration of the difference between before-word and within-word transitions contains `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% of the posterior samples.
 In other words, there is a `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% chance that the true difference $\beta_\text{diff}$ is between -10 and 10 msecs and therefore negligible.

### 2024-05-19 report/scripts/get_pdens_plot.R

@@ -1,6 +1,7 @@
-get_pdens_plot <- function(file, max_iki, ds){
-#  file <- "stanout/c2l1/all_sims.csv"
-  sims <- read_csv(file) %>% 
+get_pdens_plot <- function(file, max_iki){
+  #file <- "stanout/c2l1/all_sims.csv
+  #max_iki <- 2000
+  read_csv(file) %>% 
     filter(sim_idx <= 50) %>% 
     pivot_longer(cols = c(y_obs, y_tilde)) %>% 
     mutate(across(name, recode_factor, 
@@ -9,17 +10,16 @@ get_pdens_plot <- function(file, max_iki, ds){
                   .ordered = T)) %>% 
     filter(name ==  "Simulated data" |
           (name == "Observed data" & sim_idx == 1),
-           value < max_iki) %>% 
+          value <= max_iki,
+          model != "mogbetaunconstr") %>% 
     mutate(across(model, ~case_when(
-      str_detect(., "mogbetacon.+") ~ "Two log-Gaussians (constrained)",
-      . == "mogbetaunconstr" ~ "Two log-Gaussians (unconstrained)",
+      str_detect(., "mogbetacon.+") ~ "Two log-Gaussians",
+#      . == "mogbetaunconstr" ~ "Two log-Gaussians (unconstrained)",
       . == "lmmuneqvar" ~ "Single log-Gaussian (unequal variance)",
       . == "lmm" ~ "Single log-Gaussian",
       . == "lmmgaus" ~ "Single Gaussian")),
       across(model, ~str_wrap(., 20)),
-      across(model, ~factor(., levels = sort(unique(model))[c(5, 4, 3, 2, 1)], ordered = T)))
-  
-  plot <- sims %>%
+      across(model, ~factor(., levels = sort(unique(model))[c(4, 3, 2, 1)], ordered = T))) %>% 
     ggplot(aes(x = value / 1000, 
                group = interaction(sim_idx,name), 
                colour = name)) +
@@ -27,7 +27,5 @@ get_pdens_plot <- function(file, max_iki, ds){
     facet_wrap(~model, scales = "free_x", nrow = 1) +
     scale_colour_manual(values = c("firebrick3", "black")) +
     scale_x_continuous(labels = scales::comma) +
-    labs(colour = "", subtitle = ds) 
-  
-  return(plot)
+    labs(colour = "") 
 }

### 2024-05-19 scripts/gunnexp2/get_posterior_for_density_plot.R

@@ -1,0 +1,41 @@
+library(tidyverse)
+library(rstan)
+source("scripts/functions.R")
+source("scripts/gunnexp2/get_data.R")
+
+# Read model
+m <- readRDS(file = "stanout/gunnexp2/mogbetaconstr.rda")
+
+# Load df
+file <- "data/gunnexp2.csv"
+n_samples <- 100 # number of random data points
+d <- get_data(file, n_samples)
+
+# Get conditions
+cond_codes <- select(d, starts_with('cond'), starts_with("loc")) %>% unique()
+glimpse(cond_codes)
+
+# Select relevant parameters
+param <- c("beta", "delta", "prob", "sigma", "sigma_diff") 
+
+# Extract posterior
+as.data.frame(m, param) %>%
+  as_tibble() %>%
+  mutate(across(starts_with('delta'), ~beta + .)) %>% 
+  pivot_longer(everything()) %>% 
+  summarise(across(value, list(est = mean), .names = "{.fn}"), .by = name) %>% 
+  mutate(cond_num = parse_number(name),
+         across(name, ~str_remove_all(., "\\s*\\[[^\\)]+\\]"))) %>%
+  left_join(select(cond_codes, cond_num, condition), by = "cond_num", multiple = "all") %>%
+  separate(condition, into = c("loc", "xn"), sep = "_") %>%
+  mutate(across(loc, ~case_when(str_detect(name, "sigma_diff") & cond_num == 3 ~ "within word", 
+                                str_detect(name, "sigma_diff") & cond_num == 2 ~ "before word",
+                                str_detect(name, "sigma_diff") & cond_num == 1 ~ "before sentence",
+                                TRUE ~ .)),
+         across(c(xn), ~ifelse(str_detect(name, "sigma"), NA, .)),
+         across(loc, ~replace_na(., "overall"))) %>% 
+  filter(is.na(xn) | xn == "unmasked") %>%
+  select(param = name, est, loc) %>% 
+  # Save posterior
+  write_csv("stanout/gunnexp2/mog_constr_posterior_for_densityplot.csv")
+

### 2024-05-19 scripts/plantra/get_posterior_for_density_plot.R

@@ -1,0 +1,40 @@
+library(tidyverse)
+library(rstan)
+source("scripts/functions.R")
+source("scripts/plantra/get_data.R")
+
+# Read model
+m <- readRDS(file = "stanout/plantra/mogbetacontr.rda")
+
+# Load df
+file <- "data/plantra.csv"
+n_samples <- 100 # number of random data points
+d <- get_data(file, n_samples)
+
+# Get conditions
+cond_codes <- select(d, starts_with('cond'), starts_with("loc")) %>% unique()
+glimpse(cond_codes)
+
+# Select relevant parameters
+param <- c("beta", "delta", "prob", "sigma", "sigma_diff") 
+
+# Extract posterior
+as.data.frame(m, param) %>%
+  as_tibble() %>%
+  mutate(across(starts_with('delta'), ~beta + .)) %>% 
+  pivot_longer(everything()) %>% 
+  summarise(across(value, list(est = mean), .names = "{.fn}"), .by = name) %>% 
+  mutate(cond_num = parse_number(name),
+         across(name, ~str_remove_all(., "\\s*\\[[^\\)]+\\]"))) %>%
+  left_join(select(cond_codes, cond_num, condition), by = "cond_num", multiple = "all") %>%
+  separate(condition, into = c("loc", "group"), sep = "_") %>%
+  mutate(across(loc, ~case_when(str_detect(name, "sigma_diff") & cond_num == 3 ~ "within word", 
+                                str_detect(name, "sigma_diff") & cond_num == 2 ~ "before word",
+                                str_detect(name, "sigma_diff") & cond_num == 1 ~ "before sentence",
+                                TRUE ~ .)),
+         across(loc, ~replace_na(., "overall"))) %>% 
+  summarise(across(est, list(mean = mean), .names = "{.col}"), .by = c("name", "loc")) %>% 
+  select(param = name, loc, est) %>% 
+  # Save posterior
+  write_csv("stanout/plantra/mog_constr_posterior_for_densityplot.csv")
+

### 2024-05-19 scripts/spl2/get_posterior_for_density_plot.R

@@ -1,0 +1,43 @@
+library(tidyverse)
+library(rstan)
+source("scripts/functions.R")
+source("scripts/spl2/get_data.R")
+
+# Read model
+m <- readRDS(file = "stanout/spl2/mogbetaconstr.rda")
+
+# Load df
+file <- "data/spl2.csv"
+n_samples <- 100 # number of random data points
+d <- get_data(file, n_samples)
+
+# Get conditions
+cond_codes <- select(d, starts_with('cond'), starts_with("loc")) %>% unique()
+glimpse(cond_codes)
+
+# Select relevant parameters
+param <- c("beta", "delta", "prob", "sigma", "sigma_diff") 
+
+# Extract posterior
+as.data.frame(m, param) %>%
+  as_tibble() %>%
+  mutate(across(starts_with('delta'), ~beta + .)) %>% 
+  pivot_longer(everything()) %>% 
+  summarise(across(value, list(est = mean), .names = "{.fn}"), .by = name) %>% 
+  mutate(cond_num = parse_number(name),
+         across(name, ~str_remove_all(., "\\s*\\[[^\\)]+\\]"))) %>%
+  left_join(select(cond_codes, cond_num, condition), by = "cond_num", multiple = "all") %>%
+  separate(condition, into = c("lang", "loc1", "loc2"), sep = "_") %>%
+  unite("loc", loc2:loc1, sep = " ") %>%  
+  mutate(across(loc, ~case_when(str_detect(name, "sigma_diff") & cond_num == 2 ~ "within word", 
+                                str_detect(name, "sigma_diff") & cond_num == 3 ~ "before word",
+                                str_detect(name, "sigma_diff") & cond_num == 1 ~ "before sentence",
+                                . == "word within" ~ "within word",
+                                TRUE ~ .)),
+         across(c(lang,loc), ~replace_na(., "overall")),
+         across(loc, ~str_replace(., "NA NA", "overall"))) %>% 
+  filter(lang %in% c("EN", "overall")) %>%
+  select(param = name, est, loc) %>% 
+  # Save posterior
+  write_csv("stanout/spl2/mog_constr_posterior_for_densityplot.csv")
+

### 2024-05-19 tutorial/tutorial.Rmd
remove noncentreing

@@
 The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$).
-In the model below, these are estimated each level of a categorical predictor `location` for the $i^\text{th}$ interkey intervals.
+In the model below, the latter two are estimated separately for each level of the categorical prediction `location` (i.e.
+transition location) for the $i^\text{th}$ observation.
 
@@
 
-```{r mixturemodel, echo = F, fig.cap="What's a mixture model for interkey intervals?
-The parameters of the mixture model explained visually.
-Each of the three parameters can determined for each transition location."}
+```{r mixturemodel, echo = F, fig.cap="What's a mixture model for interkey intervals? The parameters of the mixture model explained visually."}
 
@@
 In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices for participants and conditions.
-We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before-sentence = 1, before-word = 2, within-word = 3).
-The returned values are indices for the parameters in the model: For example, for the location identifier, `beta[1]` is the population estimate for non-hesitant typing intervals at before-sentence transition locations, `beta[2]` for before-word transition locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.
+We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before sentence = 1, before word = 2, within word = 3).
+The returned values are indices for the parameters in the model: For example, for the location identifier, `delta[1]` is the population estimate for the slowdown for hesitant interkey intervals at before sentence transition locations, `delta[2]` for before word locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.
 
@@
              ymax = upper)) +
-  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
+  geom_hline(yintercept = 0, colour = "black", linetype = "dotted") +
   geom_pointrange() +

### 2024-05-19 tutorial/tutorial.Rmd
minor fix

@@
           "sigma",                       # variance component
-          "sigma_diff", "sigmap_e", "sigma_e",  # variance by mixture component
+          "sigmap_e", "sigma_e",         # variance by mixture component
+          "sigma_diff",                  # difference between variances
           "sigma_u",                     # variance for random ppt 
@@
                                     max_treedepth = 12)) # default: 10
-```
-
-```{r echo = F}
-#sampler_params <- get_sampler_params(m, inc_warmup = FALSE)
-#sampler_params_chain1 <- sampler_params[[3]]#colnames(sampler_params_chain1)
-#sapply(sampler_params, function(x) mean(x[, "accept_stat__"]))
-#sapply(sampler_params, function(x) max(x[, "treedepth__"]))
-#inits <- get_inits(m)
-#inits_chain1 <- inits[[1]]
-#print(inits_chain1)
 ```

### 2024-05-28 report/appendix/l2_constr.Rmd
Submitted paper

@@
           font_size = "small",
-          placement = "b",
           stub_indents = list("Hesitation showdown" = 1:3,

### 2024-05-28 report/appendix/models.Rmd
Submitted paper

@@
 
-This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{log}\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.
+This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.
 

### 2024-05-28 report/appendix/postparam_constr.Rmd
Submitted paper

@@
 
-beta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~beta_plots[[.]] + theme(strip.text = element_blank()))
-delta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~delta_plots[[.]] + theme(strip.text = element_blank()))
-theta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~theta_plots[[.]] + theme(strip.text = element_blank()))
+beta_plots[c(1,3,5,6)] <- map(c(1,3,5,6), ~beta_plots[[.]] + theme(strip.text = element_blank()))
+delta_plots[c(1,3,5,6)] <- map(c(1,3,5,6), ~delta_plots[[.]] + theme(strip.text = element_blank()))
+theta_plots[c(1,3,5,6)] <- map(c(1,3,5,6), ~theta_plots[[.]] + theme(strip.text = element_blank()))
 
@@
 
-plot_fin_1 <- plot_param[[1]] / plot_param[[2]] / plot_param[[3]]
+plot_fin_1 <-  plot_param[[2]] / plot_param[[1]] / plot_param[[3]]
 plot_fin_1 <- grid.arrange(patchworkGrob(plot_fin_1), 

### 2024-05-28 report/manuscript.Rmd
Submitted paper

@@
 ---
-title             : "Ideas cascade in parallel into keystrokes -- modelling writing hesitations as finite mixture process"
-shorttitle        : "Ideas cascade in parallel into keystrokes"
+title             : "Typing in tandem: finite mixture models reveal fundamental parallelism in multi-sentence text composition"
+shorttitle        : "Typing in tandem"
 
@@
   papaja::apa6_pdf:
-    keep_tex: FALSE
+    keep_tex: TRUE
 
@@
     email         : "jens.roeser@ntu.ac.uk"
+  - name          : "Rianne Conijn"
+    affiliation   : "2"
+  - name          : "Evgeny Chukharev"
+    affiliation   : "3"
+  - name          : "Gunn Helen Ofstad"
+    affiliation   : "4"
   - name          : "Mark Torrance"
     affiliation   : "1"
-  - name          : "Rianne Conijn"
-    affiliation   : "2"
-  - name          : "Gunn Helen Ofstad"
-    affiliation   : "3"
-  - name          : "Evgeny Chukharev"
-    affiliation   : "4"
 
@@
   - id            : "2"
-    institution   : "Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands"
+    institution   : "Human-Technology Interaction Group, Eindhoven University of Technology, The Netherlands"
   - id            : "3"
+    institution   : "Department of English, Iowa State University, Iowa"
+  - id            : "4"
     institution   : "Educational Sciences and Humanities, University of Stavanger, Norway"
-  - id            : "4"
-    institution   : "Department of English, Iowa State University, Iowa"
 
-abstract: "Classical serial models view the process of producing a text as a chain of discrete pauses, during which the next span of text is planned, and bursts of activity during which this text is output onto the page or computer screen.
+# We are grateful for all authors who made their available (to us), in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues.
+This work was supported by the National Science Foundation under Grant No.
+2016868: "ProWrite: Biometric feedback for improving college students' writing processes." and UKRI ESRC under Grant No.
+ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production difficulties shape text formation."
+# 
+# Data and analysis scripts written in R and Stan can be found on OSF (https://osf.io/z65dw/; Roeser et al., 2024).
+For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see https://rpubs.com/jensroes/mixture-models-tutorial.
+
+abstract: 'Classical serial models view the process of producing a text as a chain of discrete pauses, during which the next span of text is planned, and bursts of activity during which this text is output onto the page or computer screen.
 In contrast, semi-parallel models of writing assume that planning is not complete at production onset and operates in parallel with execution.
@@
 However, we found -- across populations -- that interkey intervals at word and even at sentence boundaries were often too rapid to permit planning of what was written next.
-Our results cannot be explained by the serial processing but are in line with the semi-parallel view of writing."
+Our results cannot be explained by the serial processing but are in line with the semi-parallel view of writing.'
 
-keywords: "Cascading semi-parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"
+keywords: "Cascade; parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"
 

### 2024-05-28 report/sections/conclusion.Rmd
Submitted paper

@@
-In contrast to the serial pause-and-burst view, the parallel view emphasises that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the parallel view.
-Model comparisons and the inspection of posterior parameter estimates support the parallel view of writing but provide strong evidence against the serial view.
-We found that interkey intervals across word and even sentence boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
+The parallel view of writing, in contrast to the serial view, claims that writers do not necessarily pause to plan the upcoming language unit but plan in parallel with output.
+In this paper we have demonstrated, across six datasets, that statistical models premised on a parallel-processing understanding of planning in written composition outperform models premised on serial processing.
+We also found that interkey intervals across word and even sentence boundaries are frequently at timescales that are not consistent with planning of upcoming linguistic information.
 This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.
-young / L2 writers, students) and writing tasks (e.g.
-essays, syntheses).
+young / L2 writers, students) and across different composition tasks.
+Our argument that written composition -- a very common form of language production -- is a largely parallel process is probably uncontroversial.
+However, the findings we presented in this paper constitute the first robust evidence in support of this claim.

### 2024-05-28 report/sections/discussion.Rmd
Submitted paper

@@
 The parallel model of writing, in contrast, sees  planning, at least in competent writers, as occurring to a large extent in parallel with output [@olive2014toward].
-The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is the pause duration doesn’t necessarily correlated with planning necessary for next output.
+The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is that pause duration doesn’t necessarily correlate with planning necessary for next output.
 We implemented  these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing.
@@
 This is what we found.
-At all linguistic edges we found a distribution of interkey intervals with durations that, in previous research, have been attributed just to motor execution (refs).
+At all linguistic edges we found a distribution of interkey intervals at durations to motor execution at durations substantially below a threshold that might plausibly be necessary for, for example, lexical / orthographic retrieval [@grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants; @waes2019].
 For mid-word interkey intervals, as might be expected, the mixture model predicted that the vast majority of interkey intervals fall within this distribution.
@@
 with durations that could only be associated with motor execution).
-In two cases -- for adult writers sampled in the PLanTra and Lift datasets -- the majority of interkey intervals fell within this distribution.
+In two cases -- for writers sampled in the PLanTra and Lift datasets -- the majority of interkey intervals fell within this distribution.
 Thus, although as might be expected there was a considerably increased tendency for longer interkey intervals at higher order boundaries: Planning associated with planning a new sentence is, probabilistically, much more likely to spill beyond output of the previous sentence and therefore affect duration of the sentence-initial interkey interval.
@@
 
-We observed two more differences across datasets -- other than the patterns summarised above for C2L1 -- that are worth highlighting.
-Both can largely be explained on the basis of differences associated with the population (writing competence / experience) that these samples were obtained from and text genre they were asked to produce [see also @conijn2019understanding]: The first exception concern the GE2 and SPL2 datasets: pauses were more likely to occur before sentences than before words; this difference was absent in the remaining data sets.
-**TODO: I don't have an explanation for this.** Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production.
+We observed two more differences among datasets that are worth highlighting.
+Writers sampled in the GE2 and SPL2 datasets were more likely to pause before sentences than before words [see also @medimorec2017pauses; @wen02].
+In the remaining datasets from writers who were either younger (and not educationally-selected) or writing in second language,  hesitation was equally probable pre-word and pre-sentence.
+This is most readily attributed simply to increased demands of lexical and / or orthographic processing in these samples.
+
+ Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production.
 (2) pause durations did not vary across transition locations in the LIFT dataset.
-The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning content and text structure: the structure is often kept similar to the source in synthesis writing [“structural isomorphism”’, @hidi1986producing].
-However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents or generating new sentence structures.
+The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning novel contents and text structure: the structure of one the source texts might have been used during synthesis [“structural isomorphism”, @hidi1986producing].
+However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents.
 
-An observation that was not strictly predicted by the parallel view is that the unconstrained mixture model M5 that allowed the distribution over durations of fluent transitions to vary between transition locations showed a higher predictive performance.
-This can be explained on the basis of low-level keyboarding activity: It is possible that letter bigrams are systematically produced faster than transitions between space and a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
-This is for two reasons: first, and importantly, highly frequent bi- and even trigrams may be stored, retrieved and executed as one motor code [@behmer2016crunching; @grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants] because of the strong association between letters.
-This association reduces the length of transition durations.
-It is less likely that the same association exists for space-letter and space-shift pairs.
-Second, in two datasets (PLanTra, LIFT) before-sentence transitions are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transitions.
-We found no evidence for the latter possibility (see Results section and Appendix \ref{key-combination-effect}).
-
-A possible concern with our result -- that two-distributions mixture models showed substantially better predictive performance -- is that as mixture models have more parameters they might always outperform single-distribution models.
-This concern was preempted by using cross-validation techniques for model comparison which is preventing model overfit.
-In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
-We simulated two datasets that were either based on a single distribution data-generating process or a process with a two-distributions process.
-This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- a single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
-Importantly, we can test the performance of a mixture model for data that come from a single process.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found that the two-distribution mixture model was successful at uncovering the parameter values of the data generated with a two-distributions mixture process; we observed the same for the single-distribution model applied to the data generated with a single distribution process.
+ We are, therefore, interpreting the superior predictive performance of the two-distribution mixture model as strong evidence for parallelism in written composition.
+An alternative, more prosaic explanation might be that two-distribution models outperform single distributions simply because they estimate more parameters; two distributions will always provide better fit than one distribution models due just to overfitting.
+There are two related reasons not to prefer this explanation.
+First, the cross-validation methods that we used to compare model predictive performance favour more parsimonious models and are therefore robust to overfitting.
+Second, in the analysis described in Appendix \ref{simulation} we directly tested the hypothesis that modelling data randomly sampled from a single log-normal distribution as a two-distribution mixture will provide better fit than modelling as a single distribution.
+We found that the two-distribution mixture model was successful at uncovering the parameter values of the data generated with a two-distributions mixture process; we observed the same for the single-distribution model applied to the data generated with a single distribution process.
 Parameters were not successfully uncovered when we switched model type and dataset.
-Importantly though, cross-validation rendered a substantially higher predictive performance for the mixture model compared to the single-distribution model for data simulated from a two-distributions mixture process, but not when applied to data based on a single-distribution process.
+Cross-validation rendered a substantially higher predictive performance for the mixture model compared to the single-distribution model for data sampled from a two-distributions mixture process, but not when applied to data based on a single-distribution process.
 Therefore we can rule out the possibility that the statistical support for the two-distributions mixture-models can be explained on the basis of their larger number of parameters (i.e.model overfitting).
 
-Throughout this paper we assumed that hesitations reflect mental planning of upcoming ideas and encoding of linguistic units.
-Alternatively, at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is limited research that used eye tracking to investigate regressive gaze behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
-There are at least four reasons why eye movements away from the inscription point might occur during writing: First, reading is necessary for revision (i.e.
-writers must check if writing goals have been met).
-Second, writers might look for or correct spelling or grammatical errors.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text.
-Third, writers might look at the source they are writing from.
-For two datasets (PLanTra, LIFT) which involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.^[As for PlanTra, most participants copy-pasted the source text into their word log.
-In other words, pauses might to some extent result from reading the source text.] Fourth, writers might use text to cue or reinstate ideas after interruption of the writing process [@johansson2010looking; @hayes2013new].
+Throughout this paper we have been attributing longer (second-distribution) interkey intervals to spill-over from upstream pre-motor processing that could not be completed in its entirety in parallel with previous output.
+These longer inter-key intervals may also, on occasion, be associated with activity that is initiated after the preceding process and is entirely bounded by the duration of the pause.
+We are thinking here specifically of the tendency, particularly at higher order text boundaries, for writers to glance back into their existing text [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading].
+@torrance2016reading found that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
+There are broadly two reasons why writers might look back into their existing text.^[Eye movement between key presses in writing-from-sources tasks will also be associated with reading source texts.
+Note, however, that for  two datasets (PLanTra, LIFT) that involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.]  Lookback may be associated with monitoring just-produced text for errors.
+In the present study we excluded interkey intervals that preceded deletion or cursor movement, and also intervals at timescales that might be associated with strategic reviewing of large spans of text (e.g., reading text-already-written from top to bottom).
+However it may be that some longer intervals were associated with monitoring that failed to find errors that needed correcting.
+A second function for lookback may be to support planning next-text [@johansson2010looking; @hayes2013new; @torrance2016reading].
+The preceding text, and particularly the preceding sentence, is a rich source of cues to support retrieval of both content and lexical items necessary to prepare what to write next.
+This will become particularly important where upstream processing lags behind the rate at which it can be output, and one or more of the buffers necessary to resolve this “temporal friction” [@van1991handwriting] becomes overloaded.
+Comparison of conditions unmasked and masked text conditions in the CATO and GE2 datasets -- conditions in which writers could and could not read the text that they had already produced -- suggested that, if anything, proportion of longer interkey intervals at sentence boundaries was greater when text was masked (details in Appendix \ref{masking-effect-cato-ge2}).
+This finding is consistent both with lookback being only a minor contributor to interkey interval and, where it does occur, serving a rapid memory-refresh function.
+There is also some evidence, at least in the context of handwritten production, that eye movements back into already-written text themselves occasionally occur in parallel with output [@Alamargot2007].
 
-As for the fourth reason for why writers might look back in their text: frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined].
-In contrast to this observation -- that lookbacks are more frequent before sentences compared to words -- we found that pauses appear equally often before sentences and words under our two-distributions mixture-model.
-Finally, even though reading is likely to explain some proportion of our results, lookbacks in previously written text are overall rare, certainly in samples of university students [@torrance2016reading; @chukharev2019combined].
-Also, pause frequency did not change when the produced text was masked (see comparisons for the CATO and GE2 data in Appendix \ref{masking-effect-cato-ge2}), and therefore if reading previously produced text was not possible.
+The findings we report in this paper are, we argue, the first robust evidence for parallelism in written composition of multi-sentence texts.
+The conclusion that text production processes are  fundamentally parallel will not come as a surprise to most researchers.
+Parallel processing has a long history in theories of speech production [@gar75; @levelt1989speaking; @dell1986; @chang2006; @dell1999connectionist; @dell1992stages].
+The transitory nature of activation and storage during fluent spoken production prevent, for example, generating phonological code a long time before it is required to drive articulation [@christiansen2016now].
+Spoken production is “just-in-time”.
+Perhaps because, at least in principle, there is not the same push to fluency in written production -- in principle writers can pause at any point with no detriment to communication -- writing extended text has tended to be seen as fundamentally serial, with pauses to prepare next-text followed by bursts of output during which this is executed [@flower1980ynamics].
+Our findings suggest that this is not the case.
+The language production processes that underlie text production follow the same just-in-time principle as speech.
 
-A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidates for lower pause bound to handle the complex distribution of latency from spontaneous text production.
+<!-- A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidates for lower pause bound to handle the complex distribution of latency from spontaneous text production.
 Such thresholds are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent].
@@
 Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying].
+-->

### 2024-05-28 report/sections/introduction.Rmd
Submitted paper

@@
 However if one or more upstream processes provide output more slowly, then interkey intervals are determined just by time taken to complete upstream processes and not on finger movement.
-These two distinct data-generating processes mean that interkey intervals are a mixture of two distributions, one associated with the very rapid processing necessary for actioning keystrokes, and another that forms as a result of delays resulting from some combination of processing at the semantic, syntactic, lexical, or orthographic level not being complete at the time of the immediately preceding keystroke.
-[@roeser2021modelling].
+These two distinct data-generating processes mean that interkey intervals are a mixture of two distributions, one associated with the very rapid processing necessary for actioning keystrokes, and another that forms as a result of delays resulting from some combination of processing at the semantic, syntactic, lexical, or orthographic level not being complete at the time of the immediately preceding keystroke [@roeser2021modelling].
 The parallel account provides an explanation for why writers sometimes hesitate before sentence (or word) onset, but often do not.
@@
 As we have discussed, this model assumes two data-generating processes.
-We therefore followed the finite (two-distribution) mixture-modelling approach described by@roeser2021modelling.
+We therefore followed the finite (two-distribution) mixture-modelling approach described by @roeser2021modelling.
 This model assumes that interkey intervals result from a mixture of: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels that cause delays in this information flow.
 Therefore interkey intervals are not directly and sufficiently determined by transition location -- as in the serial account -- but are also associated with different probabilities that a hesitation occurs.
-
-```{r eval = F}
-#Hesitations in writing have potential disadvantages.
-Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text.
-However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck.
-Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions.
-If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost.
-@ronneberg2022process coined this possibility the "process-disruption hypothesis".
-```

### 2024-05-28 report/sections/methods.Rmd
Submitted paper

@@
 Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.
+
+This study was not preregistered.
 
@@
                                "Norwegian", 
-                               "English (L1) / Spanish (L2)", 
+                               "English", 
                                "English (L2)", 
@@
 We removed participants that produced fewer than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-We further removed keystroke intervals that were too short to represent a behavioural response ($\le$ 50 msecs) or were of a length where such that they were unlikely to be associated with ongoing text production ($\ge$ 30 secs); percentages can be found in Table \ref{tab:datareduction}.
+We further removed keystroke intervals that were too short to represent intentional typing [$\le$ 50 msecs, see @gentner1980finger; @rumelhart1982simulating] or were of a length where such that they were unlikely to be associated with ongoing text production ($\ge$ 30 secs); percentages can be found in Table \ref{tab:datareduction}.
 From the remaining data we randomly sampled a maximum of `r n_samples` observations per participant, condition, and transition location (when more than `r n_samples` were available).
@@
 They were single distribution models, with fixed effects for each combination of transition location and dataset-specific manipulations.
-Models M1 (Guassian) and M2 (log-Guassian) are consistent with standard models used in the literature and therefore serve as baseline models.
+Models M1 (Gaussian) and M2 (log-Gaussian) are consistent with standard models used in the literature and therefore serve as baseline models.
 For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; see Appendix \ref{statistical-models} for rationale.
@@
 #       group = c(rep("Serial", 3), rep("Cascading", 2)),
-       Description = c("Single Gaussian model with effects of transition location (and experiment-specific manipulation).",
+       Description = c("Single Gaussian model with effects of transition location.",
                        "Single distribution log-Gaussian model equivialent to M1.",
                        "Equivalent to M2 but with different variance components for each transition location (unequal variance).",
-                       "Two-distributions mixture of a log-Gaussian for fluent transition intervals and another log-Gaussian that is wider than the previous distribution hesitant / disfluent key transitions; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition location but the distribution of fluent intervals has the same mean across all design cell.
-Mixing proportions capture the relative number of disfluent transitions for the distribution of hesitant transitions for each combination of transition location.",
+                       "Two-distributions mixture of a log-Gaussian for fluent interkey intervals and another wider log-Gaussian for hesitant / disfluent interkey intervals; the distribution of hesitant interkey intervals assumes -- similar to M3 -- different values for transition locations.
+The mixing weight of these distributions capture the relative number of disfluent transitions for each transition location.",
                        "Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location.")) %>% 
@@
        Type = c("LMM", "LMM", "LMM", "MoG"),
-       Description = c("Single Gaussian model with effects of transition location (and experiment-specific manipulation).",
+       Description = c("Single Gaussian model with effects of transition location.",
                        "Single distribution log-Gaussian model equivialent to M1.",
                        "Equivalent to M2 but with different variance components for each transition location (unequal variance).",
-                       "Two-distributions mixture of log-Gaussians for fluent interkey intervals and hesitant / disfluent interkey intervals; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition locations but the distribution of fluent intervals does not.")) %>%
+                       "Two-distributions mixture of a log-Gaussian for fluent interkey intervals and another wider log-Gaussian for hesitant / disfluent interkey intervals; the distribution of hesitant interkey intervals assumes -- similar to M3 -- different values for transition locations.
+The mixing weight of these distributions capture the relative number of disfluent transitions for each transition location.")) %>%
   select(Models, Description)
@@
           digits = 0,
-          caption = "Model overview. All models included by-participant random effect.",
+          caption = "Model overview. All models included by-participant random effect and study-specific manipulations.",
           stub_indents = list(
@@
 ```
-Model M4 is two-distribution mixture models consistent with the parallel cascading account.
+
+Model M4 is a two-distribution mixture model consistent with the parallel cascading account.
 This model assumes that keystroke data result from a combination of two data generating processes; these models are referred to as finite mixture models in the literature [@gelman2014; @peel2000finite].
@@
 (\#eq:bimodcon2)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \times \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \times \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
@@
 Model M4 and all other models included a random intercepts term $u_\text{participants}$ for participants which is constrained to be distributed around 0 with an unknown standard deviation, i.e.
-$\text{N}(0, \sigma_\text{p}^2)$, hence indicating participant specific deviations from the average typing speed $\beta$.
+$\mathcal{N}(0, \sigma_\text{p}^2)$, hence indicating participant specific deviations from the average typing speed $\beta$.
 
@@
 Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
-Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF (https://osf.io/z65dw/); for a tutorial see [https://rpubs.com/jensroes/mixture-models-tutorial](rpubs.com/jensroes/mixture-models-tutorial).^[The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
+Stan code for mixture models was based on @roeser2021modelling [see also @vasishth2017; @vasishth2017feature] and can be found on OSF [@roeser2024osf]; for a tutorial see [https://rpubs.com/jensroes/mixture-models-tutorial](rpubs.com/jensroes/mixture-models-tutorial).^[The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models.
 Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
@@
 Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and compared by  the difference between models $\Delta\widehat{elpd}$.
-We also summarised this difference as normalised over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, hence the *z*-score of the difference between models [@sivula2020uncertainty].
+We also summarised this difference as normalised over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, the *z*-score of the difference between models [@sivula2020uncertainty].
 Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.

### 2024-05-28 report/sections/results.Rmd
Submitted paper

@@
   relocate(Dataset, ends_with("M4)")) %>% 
+  arrange(Dataset) %>% 
   apa_table(caption = "Model comparisons.
 Models were compared incrementally from the simplest to the most complex model.
-The comparison is shown along with the distribution type of model with the higher predictive performance.
-A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model.
-Standard error is shown in parentheses.",
-            align =c("p{1.75cm}", rep("p{4cm}", 3)), 
+Comparison are shown with the distribution type of the model with the higher predictive performance.
+A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model; standard error shown in parentheses.",
+            align =c("p{1.75cm}", "p{4cm}", "p{4.25cm}", "p{3.5cm}"), 
             escape = FALSE,
@@
             col.names = c("Dataset", "Mixture of two log-Gaussians (M3 -- M4)", "Single log-Gaussian (unequal variance; M2 -- M3)", "Single log-Gaussian (M1 -- M2)"),
-            longtable = T,
+            longtable = FALSE,
             font_size = "small",
@@
          across(data, ~recode_factor(., 
+                "c2l1" = "C2L1",
                 "cato" = "CATO",# (non-dyslexic, unmasked)",
                 "gunnexp2" = "GE2",# (unmasked)",
+                "lift" = "LIFT",
+                "plantra" = "PLanTra",
                 "spl2" = "SPL2", # (L1)", 
 #                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
-                "plantra" = "PLanTra",
-                "lift" = "LIFT",
-                "c2l1" = "C2L1",
                 .ordered = TRUE))) %>% 
@@
 Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GE2 and SPL2; both BFs > 100) but longer (all BFs > 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
-Full results and parameter estimates can be found in Appendix \ref{transition-location-effect}.
+Full results and parameter estimates can be found in Appendix \ref{transition-location-effect-1}.
 
@@
 #          font_size = "small",
-          longtable = T,
+          longtable = FALSE,
           escape = FALSE,

### 2024-05-28 tutorial/tutorial.Rmd
minor fix
Submitted paper

@@
 
-This is a step-by-step tutorial for fitting mixed-effects mixture models with two log-Gaussian distributions [@roeser2021] using R [@R-base] and the `rstan` package to interface with the probabilistic programming language Stan [@carpenter2016stan; @hoffman2014no; @rstan; @rstan2].
+This is a step-by-step tutorial for fitting mixed-effects mixture models with two log-Gaussian distributions [@peel2000finite] using R [@R-base] and the `rstan` package to interface with the probabilistic programming language Stan [@carpenter2016stan; @hoffman2014no; @rstan; @rstan2].
 We will compare the mixture-model analysis to standard Gaussian and log-Gaussian mixed-effect models fitted using `brms` [@burkner2017brms; @R-brms_b].^[The R package `brms` provides a flexible framework to implement mixture models and many other types of probability models [@burkner2017brms; @R-brms_b].
@@
 
-In this guide we show how to fit a mixture model written in Stan to keystroke data.
+In this tutorial we show how to fit a mixture model written in Stan to keystroke data [similar to @roeser2021] and how to compare this to model mainstream mixed-effects models using leave-one-out cross-validation.
 Then we demonstrate how the posterior of this model can be used to calculate differences between transition locations in the text.
-Also we show how to compare model using leave-one-out cross-validation.
 This tutorial is largely self-contained.
@@
 
-Three packages need to be installed before (see footnotes for instructions): (1) `rstan` to use R to interface with Stan for fitting Bayesian models^[Instructions on how to install `rstan` can be found [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).]; (2) `tidyverse` for data wrangling and visualisation.^[Run `install.packages("tidyverse")` for installation.]; (3) `brms` for fitting Bayesian models.^[Run `install.packages("brms")` for installation.]
+Three packages need to be installed before (see footnotes for instructions): (1) `rstan` to interface with Stan^[Instructions on how to install `rstan` can be found [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).]; (2) `tidyverse` for general data wrangling and visualisation.^[Run `install.packages("tidyverse")` for installation.]; (3) `brms` to for Bayesian models without directly working in Stan.^[Run `install.packages("brms")` for installation.]
 
@@
 First we load a sample of keystroke data.
-Data are loaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/) using the URL as below.
-For simplicity we use a random subset of data from participants who completed an essay-writing task in the context of the ProWrite project.
+Data are loaded from our [OSF](https://osf.io/z65dw/) repository using the URL as below.
+For simplicity we use a random subset of data from participants who completed an essay-writing task in the context of the ProWrite project [e.g.
+@dux2022automating].
 
@@
 # Load data
-data <- read_csv("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/data/sampledata.csv") 
+data <- read_csv("https://osf.io/tjms9/download")
 ```
@@
 # Load data
-data <- read_csv("data/sampledata.csv") 
-```
-
-We reduced the data to three variables: a participant identifier `participant`, the interkey intervals `iki`, and the transition location `location` (levels: within-word, before-word, before-sentence).
+data <- read_csv("https://osf.io/tjms9/download")
+```
+
+```{r echo = F}
+n_ppts <- data %>% pull(participant) %>% unique() %>% length()
+ns <- count(data, participant, location) %>% pull(n) %>% unique()
+```
+
+We reduced the data to `r n_ppts` participants and three variables: a participant identifier `participant`, the interkey intervals `iki`, and the transition location `location` (levels: within word, before word, before sentence).
 
@@
 
-We reduced the number of interkey intervals to `n` observations per participant and per location.
+We reduced the number of interkey intervals to `r ns` observations per participant and per transition location by randomly sampling interkey intervals. See for example:
 
@@
 
-We start with two standard single distribution mixed-effects models [as in @pinheiro2007linear; @baa08], i.e.
-models that assume that data come from a single underlying process that can be described as Gaussian or log-Gaussian.
-We will compare these models.
+We start with standard single distribution mixed-effects models [as in e.g.
+@baa08; @pinheiro2007linear], that the data come from a single underlying process that, the first model, can be described as Gaussian and log-Gaussian for the other single distribution model.
+We compare these models below.
 
@@
 We start with a Gaussian mixed-effects model.
-We describe the process that generates the $i^\text{th}$ interkey interval, where $i \in 1\dots N$ and $N$ is the total number of keystroke, as a single Gaussian (normal) distribution $\mathcal{N}()$ which is characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
-The mean can be decomposed into $\beta$ and $u_\text{participant[i]}$.
-$\beta$ will be allowed to take on a different value for each transition location associated with the $i^\text{th}$ keystroke.
-By-participant interkey intervals (ikis) are assumed to vary around the average which is achieved by placing a normal distribution on the difference between average and by-participant ikis, indicated as $u_\text{participant[i]}$, with a mean of 0 and a standard deviation $\sigma_\text{p}^2$.
+The process that generates the $i^\text{th}$ interkey interval (iki below), where $i \in 1\dots N$ and $N$ is the total number of keystroke, is a single Gaussian (normal) distribution $\mathcal{N}()$ which is characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$.
+This model can be expressed mathematically using the following equation:
 
@@
 $$
+The mean can be decomposed into $\beta$ and $u_\text{participant[i]}$.
+$\beta$ was allowed to take on a different value for each transition location associated with the $i^\text{th}$ interkey interval.
+By-participant interkey intervals are assumed to vary around the average which is achieved by placing a normal distribution on the difference between average and by-participant interkey intervals, indicated as $u_\text{participant[i]}$, with a mean of 0 and a standard deviation $\sigma_\text{p}^2$.
+
 This model can be described in R using the following formula wrapped in the `brms` formula function `bf`.
+The `family` argument can be used to specify distributions where `gaussian` is the default can could be omitted below.
 
@@
 
+The syntax used in `brms` is generally similar to `lme4` syntax that many readers will be familiar with.
+Like in `lm`, `glm`, `glmer`, `lmer` and other modelling functions `0 + ` can be used to drop the intercept parameter which means that the model is parametrised to return cell means and standard errors for each level of `location`.
+
 ### Priors
 
-`brms` is using sensible regulating default priors for most model parameters which can be inspected using `get_prior` which needs the model formula and the data as input.
+As we are modelling data in the Bayesian framework, we need to define priors before model fit.
+`brms` is using sensible regulating default priors for most model parameters which can be inspected using `get_prior` which needs the model formula defined above and the data as input.
 
@@
 We want to make sure that `(flat)` is replaced by more (i.e.
-weakly or regulating) informative priors [see @mcelreath2016statistical; @kruschke2010doing; @lambert2018student].
+weakly or regulating) informative priors [see @kruschke2010doing; @lambert2018student; @mcelreath2016statistical].
 For example, we might consider a Student-*t* distribution as prior over the slope parameters (class `b`) which in our model are parameterised as the cell means of the three transition locations (levels: within word, before word, before sentence) with a mean of 350 msecs and a standard deviation of 75 msecs.
@@
 Student-*t* distributions, in comparison to normal distributions, have a degrees of freedom parameter (first argument) in addition to the mean and the standard deviation (second and third parameter, respectively).
-The degrees of freedom control the tails of the distribution with smaller values assigning more probability mass to the tails; the shape of the Student-*t* distribution is moving closer to a normal distribution when the degrees of freedom approach $\infty$.
+The degrees of freedom control the thickness of the tails of the distribution with smaller values assigning more probability mass to the tails; the shape of the Student-*t* distribution is moving closer to a normal distribution when the degrees of freedom approach $\infty$.
 This is illustrated in Figure \@ref(fig:prior) for the mean and the standard deviation of the prior above.
@@
 
-The choice of priors here is weakly informative.
-The intention is to use knowledge that we generally have about how long interkey intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values.
-Even if we know very little about the duration of interkey intervals, we do know that they shouldn't be longer than a day, or event an hour, or several minutes, because such long intervals are unlikely to reflect active writing.
+These priors are weakly informative.
+The intention is to use general knowledge that we have about how long interkey intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values.
+Even if we know very little about the duration of interkey intervals, we do know that they shouldn't be longer than a day, or an hour, or even several minutes, because such long intervals are too extreme to reflect active writing.
 
@@
 
-This model is largely identical to the Gaussian mixed-effect model in the previous section, except instead of a normal distribution this model assumes the ikis come from a single underlying process that follows a log-normal distribution.
-Assuming a log-normal distribution is useful as ikis and other response times are zero-bound (interkey intervals cannot be shorter than 0).
-A log-normal distribution forces responses to be positive.
-Also the log scale is not linear, meaning the distance between 100 msecs and 110 msecs is not the same as the distance between 500 and 510 msecs; instead the distance between adjacent values is reduced for larger values while the distances between adjacent values on the lower end is increased.
-For example, for a log with the base 10, the distance between 10 msecs and 100 msecs is identical to the distance between 100 msecs and 1000 msecs.
+This model is largely identical to the Gaussian mixed-effect model in the previous section, except instead of a normal distribution this model assumes the interkey intervals come from a single underlying process that follows a log-normal distribution indicated as $\text{log}\mathcal{N}()$.
 
@@
 $$
-For the model formula we use the distribution family `lognormal`.
+
+Assuming a log-normal distribution is useful as interkey intervals and other response times are zero-bound (interkey intervals cannot be shorter than 0).
+A log-normal distribution forces responses to be positive.
+Also the log scale is not linear, meaning the distance between 100 msecs and 110 msecs is not the same as the distance between 500 and 510 msecs; instead the distance between adjacent values is reduced for larger values while the distances between adjacent values on the lower end is increased.
+For example, for a logarithm with the base 10, the distance between 10 msecs and 100 msecs is identical to the distance between 100 msecs and 1000 msecs.
+
+For the model formula we need to change the distribution family from `gaussian` (default) to `lognormal`.
 
@@
 
-The default priors and the priors that we need to specity are now no longer on the msecs scale but on the log scale.
+The default priors and the priors that we need to specify are now no longer on the msecs scale but on the log scale.
 
@@
 ```
+
+Here is again an example for different degrees of freedom for the prior above.
 
@@
 
-The function `fixef` provide a summary of the inferred posterior estimates as average and 95% probability interval by transition location in log msecs.
+The `fixef` function provides a summary of the inferred posterior estimates as average and 95% probability interval by transition location in log msecs.
 
@@
 
-The posterior estimates obtained from the log-Gaussian model are radically different (larger) from the posterior estimates of the Gaussian model which is likely due.
-This difference is likely due to the missing lower bound at zero in the Gaussian model as we as its inability to give less emphasis to values in the right tail.
+The posterior estimates obtained from the log-Gaussian model are radically different from the posterior estimates of the Gaussian model.
+This difference is likely due to the missing lower bound at zero in the Gaussian model as we as its inability to de-emphasize values in the right tail.
 
@@
 To evaluate the out-of-sample predictive performance of the models we can use leave-one-out cross-validation via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical].
-Cross-validation is a alternative to more conventional model-comparison metrics such as $R^2$ as it penalises models with more parameters to prevent overfitting [see @mcelreath2016statistical; @lambert2018student].
-Overfitting happens when a model is capturing nuances of a sample, rather than patterns one would expect to repeat in new samples.
+Cross-validation is an alternative to more conventional model-comparison metrics such as $R^2$ as it penalises models with more parameters to prevent overfitting [see @lambert2018student; @mcelreath2016statistical].
+Overfitting happens when a model captures nuances of a sample, rather than those patterns that one would expect to observe in new samples.
 
@@
 
-The function `loo` can be used to access the expected log-predictive density which uses probability calculations to approximate LOO-IC via Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical].
+The function `loo` can be used to access the expected log-predictive density which uses probability calculations to approximate the LOO-IC via Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical].
 LOO can be approximated with the log posterior predictive densities as a measure of how likely a data point is, given the parameter estimates.
-This is shown for each data point in Figure \@ref(fig:loglik).
-Extremely small values indicate that an observation is very unlike under the model's parameter values and we are unlike to observe them in a new sample.
+This is shown in Figure \@ref(fig:loglik) for each data point in the sample.
+Extremely small values indicate that an observation is very unlike to repeat in a new sample under the given model and the estimated posterior parameter values.
 
@@
 
-If we sum over those means we obtain the $\widehat{elpd}$, the expected log predictive density, indicated as `elpd_loo`.
+If we sum over the means shown in Figure \@ref(fig:loglik), we obtain $\widehat{elpd}$, the expected log predictive density, indicated as `elpd_loo`.
 $p_\text{loo}$ (`p_loo`) is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters).
-`looic`, or LOO-IC, corresponds to $-2 \times ($`elpd_loo`$-$`p_loo`$)$ (for the deviance scale).
+`looic`, or LOO-IC, corresponds to $-2 \times ($`elpd_loo`$-$`p_loo`$)$ (for deviance scale).
 To illustrate these, we can calculate them manually
@@
 
-The Pareto *k* diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model.
-
-The difference in `elpd_loo` for both models can be obtained using `loo` and is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
+The Pareto *k* diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model and the obtained parameter estimates.
+
+The difference in `elpd_loo` for the two single distribution models is returned by `loo` and is shown in the `elpd_diff` column ($\Delta\widehat{elpd}$) with standard error shown in the `se_diff` column.
 
@@
 
-The model comparison shows that the model `fit_lgaus` (the model with the log-Gaussian likelihood) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors better than `fit_gaus`.
-This can be calculated by `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])` by normalising the difference over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, which returns the *z* or *t* value of the difference between models [@sivula2020uncertainty].
+The model comparison shows that the model `fit_lgaus` (the log-Gaussian model) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors better than `fit_gaus`.
+This can be calculated by `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])` by normalising the difference over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, which returns the difference between models expressed in standard errors [@sivula2020uncertainty].
 A difference of `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors is substantial.
-By comparison, a *t*-value of more extreme than $\mid2\mid$ corresponds to a difference that is significant at a $\alpha$-level of 0.05 (i.e.
+By comparison, a *t*-value more extreme than $\mid2\mid$ corresponds to an effect that is significant at an $\alpha$-level of 0.05 (i.e.
 $p<0.05$) [e.g.
@@
 This is because the size of `elpd_loo` increases as a function of the number of observations.
-Generally you can think of `elpd_loo` and `looic` as how many possible futures the model can rule out.
-Important is the difference between models.
-
-In contrast, `loo_R2` provides a Bayesian $R^2$ that measures the variance in the data that the model can capture.
-This quantity is not ideal for model comparisons because, similar to the frequentist $R^2$, it is subject to overfitting: a model with more parameters can never have a smaller $R^2$, and $R^2$ will innevitabily increase when we add more parameters to a model.
-A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; also chapter 6 in @mcelreath2016statistical].
+Generally `elpd_loo` and `looic` can be understood as a quantification of how many possible futures the model can rule out.
+However, what is of importance is the difference between models.
+
+If a measure of variance explained is desired, `loo_R2` provides a Bayesian $R^2$.
+This quantity is not ideal for model comparisons because, similar to the frequentist $R^2$, it is subject to overfitting: $R^2$ will inevitabily increase when more parameters are added to the model.
+A large $R^2$ means that a model captures the data well but might mean that the model makes predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; also chapter 6 in @mcelreath2016statistical].
 The is an example for Bayesian $R^2$ for the single-distribution log-Gaussian model.
@@
 
-The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$).
+The mixture model captures the writing process as a combination of three parameters: the average typing speed or fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$).
 In the model below, the latter two are estimated separately for each level of the categorical prediction `location` (i.e.
@@
 
-Model takes into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor and is distribution $u_\text{participant} \sim \mathcal{N}(0, \sigma^2_p)$; (2) each participant has in individual hesitation probability $\theta_\text{participant}$ that differs across levels of the categorical predictor.
+This model takes into account two sources of random error: (1) each participant has an individual typing speed modelled as deviations from the average $u_\text{participant} \sim \mathcal{N}(0, \sigma^2_p)$ as above; (2) each participant has in individual hesitation probability $\theta_\text{participant}$ that differs across levels of the categorical predictor `location`.
 
 The parameters of this equation are visualised in Figure \@ref(fig:mixturemodel).
-The illustrate shows the dissociation of two underlying processes that the model is estimating on the basis of the data.
+This illustration shows the dissociation of two underlying processes that the model is estimating on the basis of the data.
 There are two distributions of which the left one in grey is the distribution of fluent interkey intervals; the right distribution in green is wider and captures longer interkey intervals (with some overlap between fluent and disfluent interkey intervals).
@@
   labs(x = "iki [in msecs]", 
-       y = "Density",
-       caption = "see Roeser et al. (2021)")
+       y = "Density")
 
@@
 the distance between the grey and the green distribution); the hesitation probability or the "pausing frequency" which is the height of the green distribution captured by $\theta$.
+
 Most interkey intervals are, typically, fluent and hence part of the grey distribution.
@@
 As this model assumes a finite mixture of two distributions, the probability of fluent interkey intervals is the inverse $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent interkey intervals is $.6$.
-If the pausing probability decreases to $.3$ the proportion of fluent interkey intervals had increased to $.7$ etc.
+If the pausing probability decreases to $.3$ the proportion of fluent interkey intervals had increased to $.7$.
 
@@
 We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before sentence = 1, before word = 2, within word = 3).
-The returned values are indices for the parameters in the model: For example, for the location identifier, `delta[1]` is the population estimate for the slowdown for hesitant interkey intervals at before sentence transition locations, `delta[2]` for before word locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.
+The returned values are indices for the parameters in the model: For example, for the location identifier, `delta[1]` is the population estimate for the slowdown for hesitant interkey intervals at before sentence locations, `delta[2]` for before word locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transitions, respectively.
 
@@
 In the below code, the names on the left side of the arrow must correspond to the names expected in the Stan code (above); names on the right side do not.
-We assign the participant and location identifiers created above and their maximum values `nS` and `K`, respectively.
-The interkey interval data `iki` are assigned to `y` and the total number of observations is assigned to `N` (number of rows in the data `nrow(data`).
+We assigned the participant and location identifiers created above and their maximum values `nS` and `K`, respectively.
+The interkey interval data `iki` were assigned to `y` and the total number of observations (number of rows in the data `nrow(data)`) was assigned to `N`.
 
@@
 
-The Stan code can be downloaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/blob/main/tutorial/stan/molg.stan) using `readLines` and saved locally using `writeLines` as "molg.stan" (mixture of log-Gaussians) in the directory "stan".
+The Stan code can be downloaded from our [OSF](https://osf.io/z65dw/) repository using `readLines` and saved locally using `writeLines` as "molg.stan" (mixture of log-Gaussians) in the directory "stan".
 
@@
 # Download Stan code
-file <- readLines("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/stan/molg.stan")
+file <- readLines("https://osf.io/r7aqx/download")
 # Save Stan code in "stan" directory
@@
 
-This code fits the posterior of a categorical predictor with any number of levels.
+This code can fit the posterior of a categorical predictor with any number of levels.
 In other words, the posterior that the model returns can be used to calculate simple differences, main effects and interactions of factorial designs.
@@
 The same logic can be used to calculate main effects and interactions.
-Also, participant estimates are calculated for both mixture components and the hesitation probability by location.
-For including more predictors, or removing, e.g., random error terms, or adjusting priors, the Stan code can be changed by opening it in R or a text editor.
-The Stan code used is largely based on @sorensen2016bayesian and @vasishth2017.
+For more predictors, or removing or adding, e.g., random error terms, or adjusting priors, the Stan code can be opened and changed in R or any text editor.
+The Stan code used is largely based on @vasishth2017.
 @sorensen2016bayesian presents a detailed tutorial on how to write models in Stan [see also @lambert2018student].
 
-The model is fitting a number of parameters of which not all are interesting for inference.
-To reduce the size of the model object that contains the posterior we can determine which parameters should be saved and disregard all parameters.
+This model estimates a number of parameters of which not all are interesting for inference.
+To reduce the physical size of the model object that contains the posterior we can determine which parameters we want to save.
 
@@
                      # Improve sampling efficiency
-                     control = list(adapt_delta = 0.94,  # default: 0.9
-                                    max_treedepth = 12)) # default: 10
+                     control = list(adapt_delta = 0.96,  # default: 0.9
+                                    max_treedepth = 14)) # default: 10
 ```
@@
 
-Running this model will take a while to complete sampling depending on your hardware specifications.
-The time it took my laptop to complete this job can me viewed using `get_elapsed_time`.
-It is worth to not use all cores to run this model or to use a dedicated high performance machine.
+Running this model will take a moment to complete sampling depending on your hardware specifications. The time it took my laptop to complete this job can be viewed using `get_elapsed_time`.
 
@@
 
-Inferred posterior estimates shown as average and 95% probability interval can be extracted using the `print` function for each transition location and the three parameter values of interest.
+Inferred posterior estimates shown as averages and 95% probability intervals can be extracted using the `print` function for each transition location and the three parameter values of interest.
 `prob` is the $\theta$ parameter on the probability scale rather than on the logit scale.
@@
 print(fit_molg, 
-      pars = c("beta", "delta", "prob"), 
+      pars = c("beta", "delta", "prob", "theta"), 
       probs = c(.025, .975))
@@
 
-Before we interpret these result we need to ensure that the model reached convergence.
-Convergence can be established using two simple techniques.
+Before we interpret these results we need to ensure that the model reached convergence.
+Convergence can be established using two simple methods.
 First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain in different colours.
@@
 If the chains overlap and look like "fat hairy caterpillars", chains have converged on the same target distribution.
-The chains below look good except for a divergent transition for `theta[3]`.
+The chains below look good (i.e.
+well mixed).
 
@@
 Successful convergence is reflected in $\hat{R}$ values smaller than 1.05 [@gelman1992].
-$\hat{R}$ is similar to the *F* statistic in standard analysis of variance: it tells us how much bigger the variability between chains is compared to the variability within all chains.
+$\hat{R}$ is similar to the *F* statistic in standard analysis of variance: it tells us how much bigger the variability between chains is compared to the variability within chains.
 A value of $\approx 1$ indicates that the variability is essentially identical signifying convergence.
@@
 Convergence problems can have many reasons and therefore many solutions: for example, running longer chains (in particular longer warmups), increasing the maximum treedepth and the average acceptance probability (`adapt_delta`), specifying starting values, using sensible regulating priors, constraining parameters, or changing the parametrisation of the model.
-Severe convergence problems indicate model misspecifications.
+Severe convergence problems indicate model misspecification.
 
@@
 
-The model is simulating hypothetical data sets on the basis of the estimated posterior parameter values.
+The model is simulating hypothetical datasets on the basis of the estimated posterior parameter values.
 This happened for every iteration for every chains.
-In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) available to us because
+In other words there is a total of `r scales::comma(total_samples)` simulated data sets (and samples per parameter) because
 
@@
 For comparison we also visualised the fit for the simulated data of the Gaussian and the log-Gaussian model in Figure \@ref(fig:fitother).
-Both models show a substantially poorer fit to the keystroke data than the mixture model-based simulations.
+Both models show a substantially poorer fit to the keystroke data than the mixture-model based simulations.
 Note that the Gaussian model does not rule out negative values which are conceptually implausible.
@@
 
-Similar to the section "Model comparison I" we will use leave-one-out cross-validation to compare the three models, the mixture of log-Gaussians, and the single distribution Gaussian and log-Gaussian models.
-Because the mixture model was fitted directly in Stan and the single-distribution models were fitted in `brms`, we need to extract LOO statistics before for which we need the log-likelihood and the relative effective sample size for the Stan model.
-We can then go ahead and compare the models using functions provided by the `loo` package .
+Similar to the section "Model comparison I" we use leave-one-out cross-validation to compare now all three models, the mixture of log-Gaussians, the single distribution Gaussian and log-Gaussian models.
+Because the mixture model was fitted directly in Stan and the single-distribution models were fitted in `brms`, we need to extract LOO statistics for which we need the log-likelihood and the relative effective sample size for the Stan model.
+We can then compare the models using functions provided by the `loo` package.
 
@@
 
-As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to the single distribution Gaussian and log-Gaussian models.
-This time we need to use the `loo_compare` function because we had to extract the LOO statistics.
+In "Model comparison I" we used `loo` to compare two models directly. This time we need the `loo_compare` function because we had to extract the LOO statistics for comparison.
 
@@
 
-The difference between the model in the top row and the two other models is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.
+The difference between the model in the top row and the two other models is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in the `se_diff` column.
 Models are ordered from the model with the highest predictive performance to the lowest (i.e.
 `elpd_loo`).
-The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian ) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` standard error better than the single distribution log-Gaussian `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`).
+The model comparison shows that the model `fit_molg` (the mixture of log-Gaussian) is approximately `r round(abs(loo_comp[2,1] / loo_comp[2,2]),0)` standard error better than the single distribution log-Gaussian `fit_lgaus` (try `abs(loo_comp[2,1] / loo_comp[2,2])`).
 This constitutes substantial support for the mixture model over the single distribution log-Gaussian.
@@
 At the core of Bayesian inference is the posterior probability distribution.
-For each model parameter we have `r scales::comma(total_samples)` posterior samples that render the posterior probability distribution.
-These samples are the result of the posterior sample from the target distribution after warmup across all chains.
-If the target distribution was not found by the end of the warmup, inference will not be reliable.
-We know whether the target distribution was determined if the chains converged.
+For each model parameter we have `r scales::comma(total_samples)` posterior samples which constitute the posterior probability distribution.
+These samples are the result of the posterior samples taken from the target distribution after warmup collapsed across all chains.
+If the target distribution was not identified by the end of the warmup, inference will not be reliable.
+We know that the target distribution was determined because chains converged.
 
@@
 There is a large range of things one can do with a posterior.
-Below we will focus on summarising parameter estimates, comparing conditions, and extracting by-participant estimates.
-
-The `names` function can be used to remind us of the parameter names available in the model (we omitted the log-likelihood `log_lik`, and the stimulated data sets `y_tilde`).
+Below we focus on summarising parameter estimates, comparing conditions, and extracting by-participant estimates.
+
+The `names` function can be used to check which parameters are available in the model (we omitted the log-likelihood `log_lik`, and the stimulated datasets `y_tilde`).
 The meaning of the parameters is described above and is determined in the Stan code.
-Indices refer to the transition locations (1 = before-sentence; 2 = before-word; 3 = within-word) and to the participant identifier for by-participant parameters (indicated by `_s`).
+Indices refer to the transition locations (1 = before sentence; 2 = before word; 3 = within word) and to the participant identifier for by-participant parameters (indicated by `_s`).
 
@@
 
-The posterior probability distribution of the parameter estimates can be visualised using the `stan_hist` function with the model variable `fit_molg` and the to-be visualised parameters `pars` as arguments.
-The function returns a histogram for each other parameter values.
-Each histogram is the probability distribution of the parameter values indicating which values are more or less probable estimates for the true parameter value.
+The posterior probability distribution of the parameter estimates can be visualised using the `stan_hist` function applied to the model `fit_molg` and the to-be-visualised parameters `pars` as arguments.
+The function returns a histogram for each parameter.
+Each histogram is a probability distribution indicating which values are more (or less) probable estimates for the true parameter value and the uncertaintly associated with it.
 
@@
 ```{r}
-print(fit_molg, pars = pars, probs = c(.025,.975))
-```
-
-The output summaries the most probable parameter value as mean with its standard error (`se_mean`) and standard deviation (`sd`), the effective sample size (`n_eff`) indicating sampling efficiency and the convergence metric $\hat{R}$ (`Rhat`) we introduced above.
-
-For the following steps, we will focus on the three parameters that have conceptually interesting interpretations: (1) the average fluent typing speed $\beta$, (2) the hesitation slowdown $\delta$, and (3) the hesitation probability $\theta$ (which at the moment is still on the logit scale).
+print(fit_molg, pars = pars, probs = c(.025, .975))
+```
+
+This output summarises the most probable parameter value as mean and its standard error (`se_mean`), standard deviation (`sd`), the effective sample size (`n_eff`) indicating sampling efficiency and the convergence metric $\hat{R}$ (`Rhat`).
+
+For the following steps, we focus on the three parameters that have conceptually interesting interpretations: (1) the typing speed $\beta$, (2) the hesitation slowdown $\delta$, and (3) the hesitation probability $\theta$ (shown on the logit scale).
 
@@
 
-The values for $\beta$ and $\delta$ are shown on a log-scale.
-To transform their values back to msecs we can extract the posterior samples using the `as.data.frame` function (we prefer the use of tibbles).
+The values for $\beta$ and $\delta$ are shown on a log-scale. To transform their values back to msecs we can extract the posterior samples using the `as.data.frame` function.
 
@@
 
-The `pivot_longer` function transforms the data to a long format with an additional column for component and the model parameters as columns.
+The `pivot_longer` function transforms the data to a long format with an additional column for location and the `delta` and `theta` model parameters as columns.
 The `names_pattern` argument is using a regular expression to extract the number in the squared brackets.
@@
 
-This code is then transforming `beta` and `delta` to msecs using the exponential function `exp`.
-In order to transform the slowdown `delta` into msecs we need to add `beta` before using the exponential function; we can then subtract beta again.
+Note that the values for `beta` are now repeated for each location a a consequence. We had to do this for the next step.
+
+The next code transforms `beta` and `delta` to msecs using the exponential function `exp`.
+In order to transform the slowdown `delta` into msecs we had to add `beta` before using the exponential function; we can then subtract `beta` again.
 `theta` samples are currently on the logit scale and can be transformed to probabilities using `brms::inv_logit_scaled` (the model also contained a `prob` parameter which is `theta` but on the probability scale).
-The `recode` function changes the location indices from 1 to "before sentence", 2 to "before word", 3 to "within word".
+The `recode` function changes the location indices from 1 to "before sentence", 2 to "before word", and 3 to "within word".
 
@@
 The posterior distribution of the parameter estimates can then be visualised in, for example, histograms.
-The histograms show that especially the hesitation parameter is sensitive to transition location with more hesitations before sentences than before words than within words.
+The histograms show the probability distribution of each parameter with the hesitation slowdown $\delta$ and the hesitation probability $\theta$ shown for each transition location.
+The latter two illustrate that their posterior parameter values are sensitive to the interkey intervals location in the text with more and longer hesitations before sentences than before words than within words which we evaluate in the next section.
 
@@
 
-To illustrate differences to the estimated cells means compared to the single distribution models, here are the means for all parameters and their location.
-The central tendencies of the single distribution models are in principle a combination of `beta` and `beta + delta` weighted by `theta`.
+To illustrate differences for the estimated cells means compared to the single distribution models, here are the means for all parameters and their location.
+The central tendencies of the single distribution models are in principle a combination of `beta` and `beta + delta` weighted by `prob`.
 
@@
 
-We can calculate the differences between the transition locations for each parameter to determine to what extent each of the parameters are sensitive to transition locations.
-This will show us to what extent fluent interkey intervals ($\beta$), the size of the hesitation slowdown ($\delta$), and the hesitation probability ($\theta$) are sensitive to the linguistic location of a key pair.
+We calculate the differences between transition locations for both hesitation parameters to determine to what extent each of the parameters are sensitive to transition locations.
+This shows us to what extent the size of the hesitation slowdown $\delta$, and the hesitation probability $\theta$ are sensitive to the linguistic location of a key pair.
 
 To determine these differences we change the data format above and create a column that indicates the parameter name and a column for the corresponding values for each transition location.
-The last line calculates the difference between (1) transitions before sentences and words, and (2) transitions before and within words for each of the three parameters.
+The last line calculates the difference between (1) transitions before sentences and words, and (2) transitions before and within words for each of the two hesitation parameters.
 
@@
   select(-id) %>% # drop id column
-  mutate(diff_sent = `before sentence` - `before word`, # compare sentences to word
-         diff_word = `before word` - `within word`,     # compare within to before word
-         parameter = paste0(parameter, "[diff]")) %>% 
-  select(parameter, starts_with("diff"))
+  transmute(parameter = str_c(parameter, "[diff]"),
+            diff_sent = `before sentence` - `before word`, # compare sentences to word
+            diff_word = `before word` - `within word`)     # compare within to before word
 
@@
 
-The difference between the conditions can be summarised using the mean, the 95% PIs and the probability that the difference between transition locations is smaller than 0 (indicated as e.g.
-$P(\hat{\theta} < 0)$ for the parameter $\theta$; the hat symbol indicates $\hat{.}$ that the value is an estimate of the population parameter value).
-This summary tells us how certainly we can rule out a null difference between transition locations the most probable value for the differences.
+The differences between locations can be summarised as the mean, the 95% PIs and the probability that the difference between transition locations is smaller than 0 (notated as e.g.
+$P(\hat{\theta}_\text{diff} < 0)$ for the parameter $\theta$; the hat symbol $\hat{.}$ indicates that the value is an estimate of the population parameter value).
+This summary tells us to what extent we can rule out a null difference between transition locations.
 
@@
             .by = c(parameter, name)) %>%   # group by parameters, comparison 
-  mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
-  arrange(name)
+  mutate(across(where(is.numeric), round, 2)) # round after 2nd decimal place
 ```
@@
             .by = c(parameter, name)) %>%   # group by parameters, comparison 
-  mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
-  arrange(name)
+  mutate(across(where(is.numeric), round, 2))  # round after 2nd decimal place
 
@@
 
-For example, form these results we can infer that there is a $P(\hat{\theta} < 0)$ = `r theta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` chance that hesitations do not occur more frequently before sentence; in fact the results show hesitations are more probable by `r theta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(2)` to occur before sentences compare to before words (95% PI: `r theta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(2)`, `r theta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(2)`).
-Also, the slowdown associated with hesitations is `r delta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(0)` msecs longer for sentences (95% PI: `r delta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(0)`, `r delta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(0)`) with a probability of $P(\hat{\delta} < 0)$ = `r delta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` to be smaller than zero.
-
-```{r echo = F}
-ps <- posterior_diffs %>%
-  pivot_longer(-parameter) %>% 
-  summarise(across(value, 
-            list(mean = base::mean,  
-                 lower = lower,      
-                 upper = upper,      
-                 p_diff = p_diff),   
-            .names = "{.fn}"),
-            .by = c(parameter, name)) %>% 
-  mutate(across(where(is.numeric), ~round(., 2)),
-         across(parameter, ~str_remove(., "\\[diff\\]"))) 
-```
+Form these results we can infer that there is a $P(\hat{\theta}_\text{diff} < 0)$ = `r theta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` chance that hesitations do not occur more frequently before sentence; the results show that hesitations are `r theta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(2)` more probable to occur sentence initially than word initially (95% PI: `r theta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(2)`, `r theta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(2)`); before-word transitions show an increased hesitation probability of `r theta %>% filter(name == "diff_word") %>% pull(mean)` (95% PI: `r theta %>% filter(name == "diff_word") %>% pull(lower)`, `r theta %>% filter(name == "diff_word") %>% pull(upper)`) compared to within-word transitions.
+
+Also, the slowdown associated with hesitations is `r delta %>% filter(name == "diff_sent") %>% pull(mean) %>% round(0)` msecs longer for sentences (95% PI: `r delta %>% filter(name == "diff_sent") %>% pull(lower) %>% round(0)`, `r delta %>% filter(name == "diff_sent") %>% pull(upper) %>% round(0)`) with a probability of $P(\hat{\delta}_\text{diff} < 0)$ = `r delta %>% filter(name == "diff_sent") %>% pull(p_diff) %>% round(2)` (not exactly 0 but smaller than 0.01) to be smaller than zero.
+Hesitations before words were `r delta %>% filter(name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer than within words, although the probability interval (95% PI: `r delta %>% filter(name == "diff_word") %>% pull(lower) %>% round(0)`, `r delta %>% filter(name == "diff_word") %>% pull(upper) %>% round(0)`) did not rule out a negative difference, or no difference, which however had a low probability $P(\hat{\delta}_\text{diff} < 0)$ = `r delta %>% filter(name == "diff_word") %>% pull(p_diff) %>% round(2)`.
 
 These differences can be viewed in histograms.
-The vertical line indicates a difference between transition locations corresponding to a value of zero.
+The vertical dotted lines indicates a zero difference between transition locations.
 The area of the histogram to the left of this line corresponds to the probability `p_diff` to observe a negative difference.
+Sentence boundaries seem to have a stronger effect on the hesitation slowdown while both sentence and word boundaries increase the hesitation probability.
 
@@
   pivot_longer(starts_with("diff")) %>% 
-  mutate(across(name, recode, diff_sent = "before-sentence vs before-word",
-                              diff_word = "before-word vs within word")) %>% 
+  mutate(across(name, recode, diff_sent = "before sentence vs before word",
+                              diff_word = "before word vs within word")) %>% 
   ggplot(aes(x = value, colour = name, fill = name)) +
@@
 
-We observe the following differences that can be reported as followed: Evidence was found for a higher probability to hesitate before sentences with a difference of `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(upper)`) compared to before-word transitions; before-word transitions show an increased hesitation probability of `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(upper)`) compared to within-word transitions.
-
-Similarly, differences were found for the hesitation slowdown.
-Hesitations before sentences were `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(mean) %>% round(0)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(upper) %>% round(0)`) longer than before words.
-Hesitations before words were `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer than within words, although the probability interval (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(upper) %>% round(0)`) did not rule out a negative difference, or no difference, which however had a low probability $P(\hat{\delta} < 0)$ = `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(p_diff) %>% round(2)`.
-
-<!-- For fluent transitions there was evidence that within-word transitions were `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean) %>% round(0)` msecs longer (95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper) %>% round(0)`) compared to before-word transitions.
-There was negligible evidence for longer transitions for before-sentence transitions compared to before-word transitions ($\hat{\beta}$ = `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(mean) %>% round(0)` msecs, 95% PI: `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(lower) %>% round(0)`, `r ps %>% filter(parameter == "beta", name == "diff_sent") %>% pull(upper) %>% round(0)`).
--->
-
 ## ROPE
@@
 To take this into account we can determine a *region of practical equivalence* [ROPE, @kruschke2010believe; @kruschke2011bayesian; @kruschke2014doing].
-We can define the ROPE as the range of values that is _a priori_ considered non-different from a null effect.
-The ROPE value indicates to what extent the posterior can not rule out a negligible effect.
+
+The ROPE can be defined as the range of values that are _a priori_ considered non-different from a null effect.
+The result can be understood as the extent to which the posterior cannot rule out a negligible effect.
 A meaningful difference between groups should have a small proportion of posterior samples within the ROPE.
 
-To calculate the ROPE for our differences between two transition locations, we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference.
-We can create such a vector from our posterior differences that we calculated above.
+To calculate the ROPE for differences between transition locations, we use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference.
+We can create this from the posterior differences calculated above.
 Let's focus on $\delta_\text{diff}$, the slowdown difference for hesitant transitions.
@@
 # For sentence vs word
-delta_diff_sent <- filter(posterior_diffs, parameter == "delta[diff]") %>% pull(diff_sent)
+delta_diff_sent <- filter(posterior_diffs, str_detect(parameter, "delta")) %>% pull(diff_sent)
 
 # For before word vs within word
-delta_diff_word <- filter(posterior_diffs, parameter == "delta[diff]") %>% pull(diff_word)
-```
-
-Also, we need to define the `range` argument which specifies the range of values are considered indifferent from a null difference.
+delta_diff_word <- filter(posterior_diffs, str_detect(parameter, "delta")) %>% pull(diff_word)
+```
+
+Also, we need to define the `range` argument which specifies the range of values that are considered indifferent from null.
 Of course this range will depend on what the parameter represents and researcher's intuition as well as prior knowledge of the domain.
@@
 The ROPE for the hesitation duration of the difference between before-word and within-word transitions contains `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% of the posterior samples.
-In other words, there is a `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% chance that the true difference $\beta_\text{diff}$ is between -10 and 10 msecs and therefore negligible.
-
-Let's see what happens if we determine a wider ROPE for negligible differences which can be understood as a more conservative test for a non-zero difference.
-What happens if we define a ROPE of -50 through 50 msecs for the difference between before-word and within-word hesitations
+In other words, there is a `r round(pull(rope(delta_diff_word, range = c(-10, 10)))*100,2)`% chance that the true difference $\delta_\text{diff}$ is between -10 and 10 msecs and therefore negligible.
+
+Let's see what happens if we determine a wider ROPE between -50 through 50 msecs which can be understood as a more conservative test for a non-zero difference.
 
@@
 The ROPE contains `r round(pull(rope(delta_diff_word, range = c(-50, 50)))*100,2)`% of the posterior samples.
-In other words, there is a `r round(pull(rope(delta_diff_word, range = c(-50, 50)))*100,2)`% probability that the true difference $\beta_\text{diff}$ is between -50 and 50 msecs which would therefore be a negligible effect.
+In other words, there is a `r round(pull(rope(delta_diff_word, range = c(-50, 50)))*100,2)`% probability that the true difference $\delta_\text{diff}$ is between -50 and 50 msecs which would therefore be a negligible effect.
 
@@
 Conceptually this probability represents how likely the data are if there was no effect and we can, if this probability is low enough (typically smaller than .05) reject the null hypothesis (i.e.
-the true parameter value being some difference, is zero).
-In other words, if the data are very unlikely, the null-hypothesis becomes implausible.
-More specifically it tests whether the probability of observing a test statistic (e.g.
-*t*, *F*, $\chi^2$) as or more extreme if the null hypothesis is true.
+the true parameter value being zero).
+In other words, if the data are very unlikely, the null-hypothesis is implausible.
+More specifically it tests whether the probability of the obtained test statistic (e.g.
+*t*, *F*, $\chi^2$) is as or more extreme than what one would expect if the null hypothesis is true.
 There are two possible outcomes for *p*-values: reject the null (i.e.
@@
 And that's it.
-*p*-values do neither provide direct evidence for the alternative hypothesis (only evidence against the null), nor do they provide evidence for the null hypothesis (that data are not inconsistent with the null hypothesis is not evidence that there is no difference; or absentence of evidence is not evidence of absence).
+*p*-values do neither provide direct evidence for the alternative hypothesis (but evidence against the null), nor do they provide evidence for the null hypothesis (that data are not inconsistent with the null hypothesis is not evidence that there is no difference; or absence of evidence is not evidence of absence).
 
@@
 
-For example, if our BF is the probability of the alternative hypothesis $H_1$ over the null hypothesis $H_0$, given the data $y$, values $>1$ indicate that there is more evidence for $H_1$ than for $H_0$, given the data, and $<1$ indicates that there is more evidence for $H_0$.
+For example, if the BF is the probability of the alternative hypothesis $H_1$ over the null hypothesis $H_0$, given the data $y$, values $>1$ indicate that there is more evidence for $H_1$ than for $H_0$, given the data, and $<1$ indicates that there is more evidence for $H_0$.
 This can be turned around by swapping nominator and denominator
@@
 evidence vs no evidence).
-Authors have proposed different soft thresholds of what constitutes meaningful evidence [see, e.g., @bag12; @jeffreys1961theory; @lee2014bayesian] but generally a BF larger than 3 is weak evidence, larger than 5 / 6 indicate moderate evidence and 10 / 32 is strong evidence for a statistically meaningful effect.
-For example BF$_{10}$=2 reflects that the alternative hypothesis is two times more likely than the null hypothesis given the evidence.
-BF$_{10}$=0.33 is weak evidence in favour of the null hypothesis because $\frac{1}{0.33} \approx 3$.
+Authors have proposed different soft thresholds for meaningful evidence [see, e.g., @bag12; @jeffreys1961theory; @lee2014bayesian] but generally a BF larger than 3 is weak evidence, larger than 5 / 6 indicate moderate evidence and 10 / 32 is strong evidence for a statistically meaningful effect.
+For example BF$_{10}$ = 2 reflects that the alternative hypothesis is two times more likely than the null hypothesis given the evidence.
+BF$_{10}$ = 0.33 is weak evidence in favour of the null hypothesis because $\frac{1}{0.33} \approx 3$.
 Notably, BFs can be used to obtain evidence in favour of the null; for discussion see @dienes2014using, @dienes2018four, @dienes2016bayes and for a practical application in writing research see @spilling2022handwriting and @spilling2023writing.
 
-The BF can be calculated using the Savage-Dickey method [see, e.g., @dickey1970weighted; @jeffreys1961theory; @wagenmakers2010bayesian].
-The BF is determined on the basis of the height of the posterior density at zero compared to the height of the prior density at zero.
-Important to note here is that the resulting BF is sensitive to the prior in situations when the sample is small and / or prior information play a big role.
+BFs can be calculated using the Savage-Dickey method [see, e.g., @dickey1970weighted; @jeffreys1961theory; @wagenmakers2010bayesian] where the BF is determined on the basis of the height of the posterior density at zero compared to the height of the prior density at zero.
+Important to note here is that the BF is sensitive to the prior especially in situations when the sample was small and / or prior information played a big role.
 It is therefore a good idea to carefully think about the prior in advance of analysis (hence "prior") and to be mindful of the role of priors in the calculation of BFs.
 
-Figure \@ref(fig:sdbf) shows an example for the calculation of BFs based on a posterior with an effect difference that is normally distributed with a mean of 2.5 and a SD of 1, expressed as $\mathcal{N}(2.5, 1)$, compared against three priors with a wide SD $\mathcal{N}(0, 10)$, a SD equal to the posterior $\mathcal{N}(0, 2.5)$, and relatively small SD $\mathcal{N}(0, 1)$.
+Figure \@ref(fig:sdbf) shows an example for the calculation of BFs based on a posterior with an effect difference that is normally distributed with a mean of 2.5 and a SD of 1, expressed as $\mathcal{N}(2.5, 1)$, compared against three priors with (1) a wide SD $\mathcal{N}(0, 10)$, (2) a SD equal to the posterior $\mathcal{N}(0, 2.5)$, and (3) a relatively small SD $\mathcal{N}(0, 1)$.
 The resulting BFs are shown in the boxes as well as their calculation.
 The denominator shows the height of the posterior (black line) at zero and the nominator shows the height of the prior at zero (indicates as black dots).
-The BF is larger for the more informative prior.
+The BF was larger for more informative priors.
 Of course in practice, the shape of the posterior density is not independent of the prior; for more informative priors the posterior will move closer to the prior.
 
-```{r sdbf, fig.cap="Example for the BF calculated using the Savage-Dickey method for 3 differently wide priors.", echo=F}
+```{r sdbf, fig.cap="Example for the BF calculated using the Savage-Dickey method for three differently wide priors.", echo=F}
 bf_labs <- function(pr, po, rd = 3) {
@@
 The `logspline` function estimates the density of the posterior and the `dlogspline` and the `dnorm` functions return the height of the density at 0.
-We fixed the mean of the prior to 0 but allow to change the SD as function argument.
+We fixed the mean of the prior at 0 but allow to change the SD as function argument.
 Aside, the mean of the prior does not need to be zero but can take on any value that can be justified on the basis of domain knowledge.
-We are illustrating here how BFs can be used to test the evidence again a null hypothesis.
-
-For demonstration purposes we will calculate the BF with different priors SDs for the hesitation probability $\theta$, or in particular the posterior difference in hesitation $\theta_\text{diff}$ for before-sentence vs before-word transitions and before-word vs within-word transitions.
+We illustrate here how BFs can be used to test the evidence again the null hypothesis.
+
+For demonstration purposes we calculate the BF with different prior SDs for the hesitation probability $\theta$, or in particular the posterior difference in hesitation $\theta_\text{diff}$ for before-sentence vs before-word transitions and before-word vs within-word transitions.
 
@@
 # For sentence vs word
-theta_diff_sent <- filter(posterior_diffs, parameter == "theta[diff]") %>% pull(diff_sent)
+theta_diff_sent <- filter(posterior_diffs, str_detect(parameter, "theta")) %>% pull(diff_sent)
 
 # For before word vs within word
-theta_diff_word <- filter(posterior_diffs, parameter == "theta[diff]") %>% pull(diff_word)
+theta_diff_word <- filter(posterior_diffs, str_detect(parameter, "theta")) %>% pull(diff_word)
 ```
@@
 and then apply the `BF` function from above first to the difference between before-sentence and before-word transitions.
-For interpretation it is important to remember that $\theta$ is a probability; hence the difference between two probabilities can only range between -1 and 1 (in the most extreme case).
-In other words, a prior SD of 1 is saying that all differences are as likely as no difference (which makes very little sense).
+For interpretation it is important to remember that $\theta$ is a probability (which we transformed from logit to probability above); hence the difference between two probabilities can only range between -1 and 1 (in the most extreme case).
+In other words, a prior SD of 1 is saying anything goes (which is not helpful).
 Similar to the calculations of the ROPE above, it is useful to contemplate which prior distribution appropriately reflects the null hypothesis for the scale the difference is on (e.g.
-probability, log msecs, msecs).
-An SD of 1 and .5 are not useful prior distributions are they assign some prior probability to differences across the full range of differences.
-An SD of .1 is a more realistic representation for the distribution of null differences
+logit, probability, log msecs, msecs).
+SDs of 1 and .5 render very weakly informative prior distributions (flat, really) as they assign about the same prior probability to differences across the board.
+An SD of .1 is a more realistic representation for the distribution of null differences.^[For `brms` fits the `hypothesis` function calculates the BFs indicated as evidence ratio in the output.
+For a model such as `y ~ 1 + condition`, `hypothesis` can calculate the evidence for the hypothesis that the effect for `condition` is, for example, equal to 0.
+`hypothesis` can then be applied to the model with a statement that describes the hypothesis that should be evaluated.
+
+```{r eval = F}
+hypothesis(brm_model, "conditionb = 0")
+```
+
+To calculate the evidence ratio the argument `sample_prior = TRUE` (or `sample_prior = 'yes'`) needs to be specified in `brm`.]
 
@@
 
-For `brms` fits the `hypothesis` function calculates the BFs indicated as evidence ratio.
-For a model such as `y ~ 1 + condition`, `hypothesis` can calculate the evidence for the hypothesis that the effect for `condition` is, for example, equal to 0.
-`hypothesis` can then be applied to the model with a statement that describes the hypothesis that should be evaluated.
-
-```{r eval = F, echo = F}
-diff <- posterior_samples(fit_lgaus) %>% 
-  mutate(diff = b_locationbeforeMsentence - b_locationbeforeMword)
-
-hypothesis(diff, "diff = 0")
-
-```
-
-```{r eval = F}
-hypothesis(brm_model, "conditionb = 0")
-```
-
-To calculate the evidence ratio, which is based on the height of prior and posterior at 0, the argument `sample_prior = TRUE` (or `sample_prior = 'yes'`) needs to be specified in `brm`.
-
 ## Note on using more than a single grouping variable
 
-The Stan code presented in this tutorial can be used to estimate the posterior for two and more conditions represented in one predictor variable (called `condition` above): We can calculate the difference between conditions from the posterior as well as main effects and interactions for more complex factorial designs as any factorial design can be reduced to a single variable.
-
+The Stan code presented in this tutorial can be used to estimate the posterior for two and more conditions represented in one predictor variable (called `condition` in the Stan code): We can calculate the difference between conditions from the posterior as well as main effects and interactions for more complex factorial designs as any factorial design can be reduced to a single vector.
 For example, say we have a 2 $\times$ 2 factorial design with factor 1 having two levels AB and CD and factor 2 having the corresponding levels AC and BD.
@@
 
-## By-participant estimates
-
-In some contexts we would like to obtain by-participant estimates of the probability of a participant to exhibit hesitations (i.e. $\theta_s$) reflecting how often a participant is pausing.
-
+# By-participant estimates
+
+In some contexts we would like to obtain by-participant estimates of the probability of a participant to exhibit hesitations (i.e.
+$\theta_s$) reflecting how often a participant is pausing.
 As before we can extract the posterior from the model `fit_molg`.
@@
 The column names have the format `<parameter>_s[<location>, <participant>]` where indices indicate the transition-location id and the participant identifier of the parameters.
-The following code creates a long format with columns `location` and `participant` and a `value` column with their respective values (`value` is `pivot_longer` defaults).
+The following code creates a long format with columns `location` and `participant` and a `value` column with their respective values (`value` is `pivot_longer` default).
 The previous column names were separated into `location` (levels: `1`, `2`, `3`), and `participant` (an index for each participant) using "," to separate the two variables and extracting the numbers inside the squared brackets (the remainder of the name was discarded).
@@
   mutate(across(location, recode, 
-                `1` = "before-sentence",
-                `2` = "before-word",
-                `3` = "within-word")) %>%
+                `1` = "before sentence",
+                `2` = "before word",
+                `3` = "within word")) %>%
   # create different columns for location
-  mutate(id = row_number(), .by = location) %>%
-  pivot_wider(names_from = location, values_from = value) %>% 
+  mutate(id = row_number(), 
+         .by = location) %>%
+  pivot_wider(names_from = location, 
+              values_from = value) %>% 
   select(-id) # drop id
@@
 We observe in general that hesitations are more frequent before sentences and words than within words.
-However, participants vary to the extend they hesitate before words and sentences suggesting that some participant plan sentences more globally and others plan sentences word-by-word (incrementally).
+However, participants vary to the extend they hesitate before words and sentences.
 
@@
        y = "Hesitation probability with 95% PIs") +
-  scale_colour_brewer("Transition\nlocation", palette = "Dark2") +
-  scale_shape_manual("Transition\nlocation", values = c(21, 3, 8)) +
+  scale_colour_brewer("Transition location", palette = "Dark2") +
+  scale_shape_manual("Transition location", values = c(21, 3, 8)) +
   scale_x_continuous(breaks = 1:10) +
@@
 
-Finally, we can calculate and visualise the by-participant differences.
+Finally, we calculated and visualised the by-participant differences.
 
@@
 posterior_ppts_long_diff_summary <- posterior_ppts_long %>% 
-  mutate(diff_sent = `before-sentence` - `before-word`,
-         diff_word = `before-word` - `within-word`) %>% 
+  mutate(diff_sent = `before sentence` - `before word`,
+         diff_word = `before word` - `within word`) %>% 
   select(participant, starts_with('diff')) %>%
@@
 
-This calculation and visualisation allows us to identify participants that do or do not differ in terms of hesitation frequency depending on transition locations.
+These allow us to identify participants that do or do not differ in terms of hesitation frequency depending on transition locations.
 
@@
   mutate(across(difference, recode, 
-                diff_sent = "Comparison: before-sentence vs before-word",
-                diff_word = "Comparison: before-word vs within-word"),
+                diff_sent = "Comparison: before sentence vs before word",
+                diff_word = "Comparison: before word vs within word"),
          across(participant, as.numeric)) %>% 

### 2024-06-13 report/appendix/models.Rmd
Add slides and model comparisons update

@@
 
-where $i$ iterates over every interkey interval $i \in 1 \dots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.
+where $i$ iterates over every interkey interval $i \in 1 \ldots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.
 

### 2024-06-13 report/appendix/sim.Rmd
Add slides and model comparisons update

@@
 map_dfr(files, ~read_csv(.) %>% 
-        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
-  mutate(across(Dataset, ~recode(., mogdata = "two distributions", lmdata = "single distributions"))) %>% 
+        mutate(Dataset = str_remove_all(.x, "\\../|/|mixture-model-sim/data|/|\\.|csv"))) %>%
+  mutate(across(Dataset, ~recode(., mogdata = "two distributions", lmdata = "single distribution"))) %>% 
   ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +

### 2024-06-13 report/sections/discussion.Rmd
Add slides and model comparisons update

@@
 The parallel model of writing, in contrast, sees  planning, at least in competent writers, as occurring to a large extent in parallel with output [@olive2014toward].
-The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is that pause duration doesn’t necessarily correlate with planning necessary for next output.
+The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is that pause duration doesn't necessarily correlate with planning necessary for next output.
 We implemented these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing.
@@
 (2) pause durations did not vary across transition locations in the LIFT dataset.
-The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning novel contents and text structure: the structure of one the source texts might have been used during synthesis [“structural isomorphism”, @hidi1986producing].
+The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning novel contents and text structure: the structure of one the source texts might have been used during synthesis ["structural isomorphism", @hidi1986producing].
 However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents.
@@
 The preceding text, and particularly the preceding sentence, is a rich source of cues to support retrieval of both content and lexical items necessary to prepare what to write next.
-This will become particularly important where upstream processing lags behind the rate at which it can be output, and one or more of the buffers necessary to resolve this “temporal friction” [@van1991handwriting] becomes overloaded.
+This will become particularly important where upstream processing lags behind the rate at which it can be output, and one or more of the buffers necessary to resolve this "temporal friction" [@van1991handwriting] becomes overloaded.
 Comparison of conditions unmasked and masked text conditions in the CATO and GE2 datasets -- conditions in which writers could and could not read the text that they had already produced -- suggested that, if anything, proportion of longer interkey intervals at sentence boundaries was greater when text was masked (details in Appendix \ref{masking-effect-cato-ge2}).
@@
 The transitory nature of activation and storage during fluent spoken production prevent, for example, generating phonological code a long time before it is required to drive articulation [@christiansen2016now].
-Spoken production is “just-in-time”.
+Spoken production is "just-in-time".
 Perhaps because, at least in principle, there is not the same push to fluency in written production -- in principle writers can pause at any point with no detriment to communication -- writing extended text has tended to be seen as fundamentally serial, with pauses to prepare next-text followed by bursts of output during which this is executed [@flower1980ynamics].

### 2024-06-13 report/sections/introduction.Rmd
Add slides and model comparisons update

@@
 This is because writing requires spelling retrieval, which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen.
-Pen control and typing skills are later-developing and more effortful than those required to articulate language sounds in speech and, possibly as a consequence, written output rate is typically slower [see for a recent review @kandel2023written].
+Pen control and typing skills are later-developing and more effortful than those required to articulate language sounds in speech and, possibly as a consequence, written output rate is typically slower [for a recent review see @kandel2023written].
 Consequently mentally buffering of linguistic plans prior to output is likely to play a greater role in writing than in speech.
@@
 These models have been adopted in a number of psycholinguistic studies.
-For example, in language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental.
+For example, in language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implementation.
 In the context of written production @roeser2021modelling demonstrated^[For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.] how models that allow the possibility of two underlying data-generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g.

### 2024-06-13 report/sections/methods.Rmd
Add slides and model comparisons update

@@
 When a parameter is allowed to vary by transition location (levels: before sentence, before word, within word), this was indicated as subscript; similar for participants.
-For example, $\theta_\text{location[i], participant[i]}$ means that the mixing proportion $\theta$ is allowed to take on a different value for both the transition location and participant associated with the $i^\text{th}$ keystroke interval $i \dots \in N$ where $N$ is the total number of keystroke intervals.
+For example, $\theta_\text{location[i], participant[i]}$ means that the mixing proportion $\theta$ is allowed to take on a different value for both the transition location and participant associated with the $i^\text{th}$ keystroke interval $i \in 1 \ldots N$ where $N$ is the total number of keystroke intervals.
 Because the hesitation probability was allowed to vary by participants as well as location type, we included a hyper-parameter for the mixing proportion $\theta$ for each transition location.

### 2024-06-13 report/sections/results.Rmd
Add slides and model comparisons update

@@
 
-```{r}
+```{r message=FALSE}
 files <- list.files(

### 2024-06-13 slides/sections/datasets.Rmd
Add slides and model comparisons update

## Data sets

```{r}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GE2"),
                Source = c("@ronneberg2022process", # "Rønneberg et al. (2022)",
                            "@torrance2016adolescent", # "Torrance et al. (2016)",
                           "@torrancea",#  "Torrance et al. (n.d.)",
                          "@rossetti2022text",#   "Rossetti and Van Waes (2022b)",
                          "@vandermeulen2020mapping",#  "Vandermeulen et al. (2020)", 
                           "@torranceb"), # "Torrance and Ofstad (n.d)"),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog",
                           "EyeWrite"),
                Task = c("Argumentative", 
                                 "Expository",
                                 "Argumentative",
                                 "Text simplification",
                                 "Synthesis", 
                                 "Argumentative"),
                 N = as.character(c(126, 26*2, 39, 47, 658, 45)),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                 Age = c(11.8, 16.9, 20.6, 23, 17.0, 19),
                 Sample = c("6th graders", "Secondary school students (dyslexic, non dyslexic)", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
                 Country = c("Norway", "Norway", "USA", "Belgium", "The Netherlands", "UK"),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English", 
                               "English (L2)", 
                               "Dutch",
                               "English")) %>% 
  arrange(Dataset) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = T, 
                position = "right",
                font_size = 12) %>% 
  row_spec(0:6, extra_css = "line-height: 2.25;")

```

### 2024-06-13 slides/sections/discussion.Rmd
Add slides and model comparisons update

## Planning text unfolds in parallel to production!

- Writers do not necessarily pause but plan utterances in parallel to writing.
- Evidenced by 
  - stronger predictive performance of mixture models [@roeser2021modelling].
  - pauses are generally equally likely before before sentences compare to before words (but often longer).
- Largely consistent for populations with different levels of writing experience, languages (e.g. young / L2 writers, students), composition tasks. 
- First robust evidence that written composition is largely a parallel process.

### 2024-06-13 slides/sections/evidence.Rmd
Add slides and model comparisons update

## Serial view

- Interkey intervals are longer at larger linguistic edges following relatively fluent production bursts [@chenoweth2001fluency; @hayes2012evidence; @kaufer1986composing].
- Mean pause duration before a writer starts a sentence is typically longer than before a mid-sentence word and these are, in turn, longer than between mid-word key presses [e.g. @conijn2019understanding]. 
- Planning message and syntax for the next sentence adds to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke [@baaijen2012keystroke; @roeser2018advance].

## Parallel view

- Writers often do not hesitate before sentence (or word) onset.
- In some cases sentences are not planned, planning was postponed until after sentence onset, or planning was completed in parallel with previous output [@not10; dam09; @roeser2018advance].
- Sentence initial durations are often shorter than what one would expect if sentence-initial pauses reflect planning.
  - @medimorec2017pauses: 71% of sentence-initial interkey intervals shorter than 1 sec. 
  - @ronneberg2022process: 50% sentence-initial intervals were around 430 msecs but 1.2 secs for the remainder. 
  <!-- - For comparison, these short pauses are less than mean written naming response latency for single objects [@torrance2018timed] and     longer pauses were similar to response latencies for short sentences when describing arrays of images [@roeser2018advance].  -->
- Fluent output is maintained by a cascade of processes that run in parallel [@olive2014toward;@van1991handwriting].

## Two distinct data-generating processes leading to a mixture of two distributions of interkey intervals [@roeser2021modelling].

- Minimum interkey intervals for executing motor movements; 100-150 msecs [@conijn2019understanding;@waes2019].
- If upstream processes provide output more slowly, then interkey intervals are determined by time taken to complete upstream processes and not on finger movement. 
- Neither process (or distribution of keystrokes) can be explained on the basis of the keystroke location in the text.

### 2024-06-13 slides/sections/implementation.Rmd
Add slides and model comparisons update

## Implentation

- Bayesian inference [e.g. @lambert2018student]
- Random effects for participants [see @roeser2021modelling] with code based on @vasishth2017
- Stan [@carpenter2016stan] using `rstan` [@rstan] to interface with $R$
- Leave-one-out cross-validation estimated as the sum of the expected log predictive density $\widehat{elpd}$ and compared by the difference between models $\Delta\widehat{elpd}$ [@vehtari2015pareto; @vehtari2017practical]
- Difference between models was summarised as $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, the *z*-score of the difference between models [@sivula2020uncertainty].

### 2024-06-13 slides/sections/locationeffects.Rmd
Add slides and model comparisons update

--- 2024-05-28 report/sections/results.Rmd
@@
-## Model comparisons
+## Mixture distributions
 
-```{r}
-files <- list.files(
-  str_c("../../stanout/", 
-        c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
-  pattern = "model.+.csv", 
-  full.names = T)
-
-table <- purrr::map_dfr(files, ~read_csv(.) %>% 
-                          mutate(dataset = .x)) %>% 
-  select(dataset, model, contains("elpd_diff"), se_diff) %>%
-  arrange(dataset, desc(elpd_diff)) %>% 
-  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(where(is.numeric), ~round(., 0)),
-         across(where(is.numeric), ~format(., big.mark = ",")),
-         across(everything(), ~str_trim(.))) %>%
-  filter(dataset != "spl2_shift",
-         model != "M4 -- M5") %>% 
-  transmute(
-    elpd_diff_ratio,
-    across(dataset, ~recode(., 
-                            "cato" = "CATO",
-                            "spl2" = "SPL2", 
-                            #                "spl2_shift" = "SPL2 (shift + C)", 
-                            "plantra" = "PLanTra",
-                            "lift" = "LIFT",
-                            "c2l1" = "C2L1",
-                            "gunnexp2" = "GE2")),
-    Comparison = case_when(str_detect(model, "M2$") ~ "Single log-Gaussian (M1 -- M2)",
-                               str_detect(model, "M3$") ~ "Single log-Gaussian (unequal variance; M2 -- M3)",
-                               str_detect(model, "M4$") ~ "Mixture of two log-Gaussians (M3 -- M4)",
-                             #str_detect(model, "M5$") ~ "Mixture of two log-Gaussians (unconstrained; M5)"
-                             ),
-    across(elpd_diff, ~str_c(., " (", se_diff, ")")))
-
-table_ratios <- table %>% select(dataset, Comparison, elpd_diff_ratio)
-
-table <- table %>% select(-elpd_diff_ratio) %>% 
-  pivot_wider(names_from = dataset, values_from = elpd_diff)
-
-table_ratios <- table_ratios %>% 
-  mutate(across(elpd_diff_ratio, as.numeric)) %>% 
-  summarise(across(elpd_diff_ratio, list(min = min, max = max), .names = "{.fn}"), .by = Comparison) %>% 
-  mutate(range = str_c("[", min, ", ", max, "]"))
-
-tr_m2 <- filter(table_ratios, str_detect(Comparison, "M2\\)$")) %>% pull(range)
-tr_m3 <- filter(table_ratios, str_detect(Comparison, "M3\\)$")) %>% pull(range)
-tr_m4 <- filter(table_ratios, str_detect(Comparison, "M4\\)$")) %>% pull(range)
-tr_m5 <- filter(table_ratios, str_detect(Comparison, "M5\\)$")) %>% pull(range)
-
-```
-
-The differences in the predictive performance of models for all datasets are shown in Table \ref{tab:loos}.
-For all six dataset the two-distributions mixture model (M4) provided greater predictive performance than all three single-distribution models.
-Differences in predictive performance ($\Delta\widehat{elpd}$) between models M4 and the nearest competing single-distribution model (M3) ranged between `r tr_m4` standard errors, indicating a substantially higher predictive performance for the two-distributions model.
-Among the single distribution models we found higher predictive performance for the unequal variance model and the lowest predictive performance for the single distribution Gaussian.
-Comparisons showing the fit of model predictions to the data can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
-Data predicted by the two-distribution mixture model closely align with the observed data; data predicted by single distribution Gaussian models showed an inferior fit to the data.
-
-```{r loos, results='asis'}
-table %>% 
-  arrange(desc(Comparison)) %>% 
-  pivot_longer(-Comparison, names_to = "Dataset") %>% 
-  pivot_wider(names_from = Comparison, values_from = value) %>% 
-  relocate(Dataset, ends_with("M4)")) %>% 
-  arrange(Dataset) %>% 
-  apa_table(caption = "Model comparisons.
-Models were compared incrementally from the simplest to the most complex model.
-Comparison are shown with the distribution type of the model with the higher predictive performance.
-A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model; standard error shown in parentheses.",
-            align =c("p{1.75cm}", "p{4cm}", "p{4.25cm}", "p{3.5cm}"), 
-            escape = FALSE,
-            row.names = T,
-            col.names = c("Dataset", "Mixture of two log-Gaussians (M3 -- M4)", "Single log-Gaussian (unequal variance; M2 -- M3)", "Single log-Gaussian (M1 -- M2)"),
-            longtable = FALSE,
-            font_size = "small",
-            note = "$\\Delta\\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density")
-
-```
-
-## Transition location effect
-
-Figure \ref{fig:mixmodel} illustrates the mixture model results for every dataset: for each interkey interval we obtained two posterior probability distributions, one associated with fluent transitions between keys and another for interkey intervals where upstream difficulty resulted in hesitant interkey intervals.
-The hesitation probability -- the probability of  interkey intervals associated with hesitations -- is indicated as the height of the distribution shown in yellow (on the right of each panel).
-The transition locations are characterised by the hesitation parameters: as we illustrate the mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location (within each dataset) but their height differs as the probability of fluent interkey intervals is the inverse of the height of hesitant interkey intervals; i.e.
-$1 - \theta$.
-Within-word interkey intervals have a negligibly small hesitation probability barely visible in the visualisation.
-Hesitations before words and sentences are both roughly equally likely than fluent transitions.
-In other words, half of the time participants did not pause before a word or sentence; we explore this below.
-However, when hesitations occurred, these are longer at before-sentence interkey intervals compared to before word interkey intervals.
-The slowdown for hesitant interkey intervals -- represented earlier as $\delta$ -- is the distance between the distribution of fluent interkey intervals ($\beta$) shown in grey and the distribution of hesitant interkey intervals shown in yellow.
-
-```{r mixmodel, fig.cap='Mixture model plot for each transition location.
-Shown are the distributions for both mixture components (fluent and hesitant interkey intervals) and their weighting.
-Interkey intervals are shown in seconds on log scaled intervals.', fig.height=8, fig.width=7}
-
+```{r mixmodel, fig.cap='', fig.height=12, fig.width=10, message=FALSE}
 # Data grid
@@
 
-<!-- We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-We demonstrated above that the predictive performance of the unconstrained mixture model was substantially higher than the predictive performance of the constrained mixture model.
-However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
-We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following.
--->
+## Location estimates by model parameter
 
-```{r}
+```{r message=FALSE}
 files <- list.files(str_c("../../stanout/", 
@@
 
-Figure \ref{fig:mixmodel} highlighted how model parameters relate to each other in pairs of distributions but to compare the effect of transition location on model parameters we need to compare the posterior distribution of the three parameters of interest individually.
-Therefore Figure \ref{fig:crossstudypost2} visualises the same posterior as Figure \ref{fig:mixmodel} but expressed directly in terms of posterior probability distributions of the individual mixture-model parameters.
-These are shown for all datasets by transition location.
-Although models were fitted with all dataset-specific conditions, we aggregated the posterior across conditions^[We aggregated across pre and post tests for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.], and excluded conditions that might confound comparisons^[We removed the masked writing condition in the GE2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-ge2}).].
-For posteriors of all conditions within datasets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences and similarities associated with transition locations for all datasets.
-In particular we can see how transition location in the text is associated with hesitant interkey intervals and hesitation probability.
-For completeness we also report the estimate for fluent interkey intervals.
-
-```{r crossstudypost2, fig.cap="Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group."}
+```{r crossstudypost2, fig.cap="", fig.height=4.5, fig.width=10}
 
@@
              group = interaction(data))) +
-  geom_line(linewidth = linewidth, 
-            alpha = .75, 
-            position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
   facet_grid(~name, scales = "free_x") +
@@
              group = interaction(data))) +
-  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
-  geom_errorbar(width = width, alpha = .75, position = posd) +
-  geom_point(aes(colour = data), 
-            size = dotsize, 
-            position = posd) +
   facet_grid(~name) +
@@
        shape = grouplabel) +
-  theme(axis.title = element_blank()) +
+  theme(axis.title = element_blank(),
+        legend.position = "top") +
   guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
@@
              left = "Posterior estimate with 95% PIs",
-             bottom = "Transition location")
+             bottom = "Key location in text")
 
@@
 
-Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets.
-We summarise the evidence for differences between transitions locations in Table \ref{tab:loceffectconbf}.
-We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs > 100 for all datasets) but only for half of the datasets hesitation durations were longer (GE2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
-Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GE2 and SPL2; both BFs > 100) but longer (all BFs > 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
-Full results and parameter estimates can be found in Appendix \ref{transition-location-effect-1}.
+## Location estimates by model parameter
+
+```{r crossstudypost3, fig.cap="", fig.height=4.5, fig.width=10}
+beta <- beta +
+  geom_line(linewidth = linewidth, 
+            alpha = .75, 
+            position = posd) +
+  geom_errorbar(width = width, alpha = .75, position = posd) +
+  geom_point(aes(colour = data), 
+            size = dotsize, 
+            position = posd)
+
+delta <- delta + 
+  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
+  geom_errorbar(width = width, alpha = .75, position = posd) +
+  geom_point(aes(colour = data), 
+            size = dotsize, 
+            position = posd)
+
+theta <- theta + 
+  geom_line(linewidth = linewidth, alpha = .75, position = posd) +
+  geom_errorbar(width = width, alpha = .75, position = posd) +
+  geom_point(aes(colour = data), 
+            size = dotsize, 
+            position = posd)
+
+plot <- beta + delta + theta + plot_layout(guides = "collect",
+                                           widths = c(.5, 1, 1))
+
+grid.arrange(patchworkGrob(plot), 
+             left = "Posterior estimate with 95% PIs",
+             bottom = "Key location in text")
+
+```
+
+## Location estimates by model parameter
 
@@
 
-apa_table(tmp,
-          align = c("l", rep("p{3cm}", 4)), 
-#          font_size = "small",
-          longtable = FALSE,
-          escape = FALSE,
-          row.names = T,
-          col.names = c("Dataset", rep(c("before sentence vs word", "before vs within word"), 2)),
-          col_spanners = list(" " = c(1),
-                              "Hesitation slowdown" = c(2, 3), 
-                              "Hesitation probability" = c(4, 5)),
-          note = "BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
-          caption = "BFs for the effect of transition location on interkey intervals shown for the hesitation slowdown and hesitation probability.
-Comparisons for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.")
+tmp %>% 
+  kable(col.names = c("Dataset", rep(c("before sentence vs word", "before vs within word"), 2)),
+        caption = "Values indicate BFs in support of the alternative hypothesis over the null hypothesis.", align = c("l", rep("r", 4))) %>% 
+  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
+                full_width = T, 
+                position = "right",
+                font_size = 14) %>% 
+  add_header_above(c(" " = 1,
+                     "Hesitation slowdown" = 2,
+                     "Hesitation probability" = 2))
 
 ```
-
-To reiterate, for most datasets we found that pauses before sentences are longer but not always more likely than before words (except for SPL2 and GE2).
-In fact, LIFT showed some indication that pauses were less likely before sentences than words.
-This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for interkey intervals at larger linguistic edges (see Introduction section).
-It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.
-
-Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character, while the remaining data sets measured before-sentence transitions as the interval between space key and shift.
-To test to what extent complex key combinations affects the parameter estimates, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability for sentence-initial transitions (BFs > 100) but it did not affect the duration hesitant interkey intervals (BFs < 0.3); see Appendix \ref{key-combination-effect}.
-This conflicts with the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did include the character following shift (PLanTra, LIFT).
-In other words, patterns in our results cannot be explained on the basis of how sentence-initial key transitions were operationalised.
-
-Finally, Figure \ref{fig:crossstudypost2} highlights that fluent interkey intervals in two datasets (C2L1, CATO) were substantially longer than for other datasets.
-This is presumably reflecting that these data were sampled from the youngest and least experienced population of writers in our data pool.
-Interestingly, however, hesitation duration and hesitation probability followed the same patterns as the majority of datasets.
-
-<!-- The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
-We, therefore, summarise briefly differences in the results for the transition-location effects; full results can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs > 100) but not before-sentence compared to before-word transitions (all BFs < 0.08, but SPL2 and GE2: BFs > 100).
-Patterns for hesitation probability were generally similar in both the constrained and unconstrained model.
-However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs > 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.
--->

### 2024-06-13 slides/sections/modelcomparisons.Rmd
Add slides and model comparisons update

<!-- ## Model comparisons -->

```{r loadloos, warning=FALSE, message=FALSE}
files <- list.files(
  str_c("../../stanout/",
        c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")),
  pattern = "model.+.csv",
  full.names = T)

mc <- purrr::map_dfr(files, ~read_csv(.) %>%
                          mutate(dataset = .x)) %>%
  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>%
  filter(dataset != "spl2_shift",
         model != "M4 -- M5") %>%
  mutate(across(dataset, ~recode(.,
                            "cato" = "CATO",
                            "spl2" = "SPL2",
                            #                "spl2_shift" = "SPL2 (shift + C)",
                            "plantra" = "PLanTra",
                            "lift" = "LIFT",
                            "c2l1" = "C2L1",
                            "gunnexp2" = "GE2")),
         across(model, ~case_when(str_detect(., "M2$") ~ "Single process (M1 vs M2)",
                               str_detect(., "M3$") ~ "Single process (unequal var.; M2 vs M3)",
                               str_detect(., "M4$") ~ "Mixture process (M3 vs M4)",
                             )),
         across(model, ~factor(., levels = unique(model), ordered = T)))
```

<!-- ```{r modelcomp, fig.cap="Shown is the difference in predictive performance of models."} -->
<!-- plot <- mc %>%  -->
<!--   mutate(value = elpd_diff, -->
<!--          se = se_diff, -->
<!--          lower = value - 2 * se, -->
<!--          upper = value + 2 * se) %>%  -->
<!--   select(model, dataset, value, lower, upper) %>%  -->
<!--   ggplot(aes(x = reorder(dataset, value), -->
<!--              y = value,  -->
<!--              ymin  = lower,  -->
<!--              ymax = upper, -->
<!--              colour = model)) + -->
<!--   facet_wrap(~model, scales = "free_x", nrow = 2) + -->
<!--   coord_flip() + -->
<!--   theme_bw() + -->
<!--   scale_y_continuous(labels = scales::comma) + -->
<!--   ggthemes::scale_color_colorblind() + -->
<!--   labs(y = bquote(Delta*widehat(elpd)~"with 2 \u00D7 SE")) + -->
<!--   theme(axis.title.y = element_blank(), -->
<!--         axis.ticks.y = element_blank()); plot -->
<!-- ``` -->
<!-- ## Model comparisons -->

<!-- ```{r modelcomp2, fig.cap="Shown is the difference in predictive performance of models."} -->
<!-- plot + geom_pointrange(show.legend = F) -->
<!-- ``` -->

## Model comparisons

```{r loos, results='asis'}
mc %>% 
  select(dataset, Comparison = model, elpd_diff_ratio) %>%
  pivot_wider(names_from = Comparison, values_from = elpd_diff_ratio) %>%
  mutate(across(-dataset, ~" ")) %>% 
  rename(`Data set` = dataset) %>% 
  select(1, 4, 3, 2) %>% 
  kable(align = c("l", rep("r", 3)),
        caption = "Vales are the absolute ratio of the difference in predictive performance measured as $\\widehat{elpd}$ and its standard error $\\mid\\frac{\\Delta\\widehat{elpd}}{\\text{SE}}\\mid$ which corresponds to the $z$-score of the difference between models [@sivula2020uncertainty].") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = T, 
                position = "right",
                font_size = 14) %>% 
  row_spec(0:6, extra_css = "line-height: 2.25;")  
```

## Model comparisons

```{r loos2, results='asis'}
mc %>% 
  select(dataset, Comparison = model, elpd_diff_ratio) %>%
  pivot_wider(names_from = Comparison, values_from = elpd_diff_ratio) %>%
  rename(`Data set` = dataset) %>% 
  select(1, 4, 3, 2) %>% 
  kable(align = c("l", rep("r", 3)),
        caption = "Vales are the absolute ratio of the difference in predictive performance measured as $\\widehat{elpd}$ and its standard error $\\mid\\frac{\\Delta\\widehat{elpd}}{\\text{SE}}\\mid$ which corresponds to the $z$-score of the difference between models [@sivula2020uncertainty].") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = T, 
                position = "right",
                font_size = 14) %>% 
  row_spec(0:6, extra_css = "line-height: 2.25;")  
```

```{r eval = F}
mc %>% 
  select(dataset, model, elpd = elpd_diff_ratio) %>% 
  ggplot(aes(y = elpd, x = model, colour = dataset, group = dataset)) +
  geom_point() +
  geom_line()
```

### 2024-06-13 slides/sections/modelfit.Rmd
Add slides and model comparisons update

--- 2024-05-19 report/appendix/fit2data.Rmd
@@
-# Fit to data
+## Model fit to data
 
-Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data.
-From each model we obtained 50 simulations (predictions); see Figure \ref{fig:prediction}.
-The model predictions fitted the data substantially better for the mixture models.
-The worse fit was observed for the single distribution Gaussian model which predicted a larger number of negative keystroke intervals; a log-normal distribution removed these negative predictions.
-
-```{r eval = F}
-theme_set(theme_few(base_size = 10) +
+```{r getdensity, eval = F}
+theme_set(ggthemes::theme_few(base_size = 8) +
               theme(strip.background = element_blank(),
                     legend.position = "top",
-                    legend.justification = "right",
                     panel.grid = element_blank(),
@@
 # get all density plots
-plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000))
+plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 5000))
 
@@
 
-# save plot 
-ggsave("../figures/fitplots.pdf", plots_fin, height = 7.25, width = 6.25)
+ggsave("../misc/gfx/fit.png", plot = plots_fin, width = 6.25, height = 8.5)
 
@@
 
-```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red in comparison to the interkey interval data shown in black by model types and datasets.
-Model fit to the data was order from the best (left) to the weakest (right).
-For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
-knitr::include_graphics("figures/fitplots.pdf")
+```{r fitplot, fig.cap="", eval = T}
+include_graphics("../misc/gfx/fit.png")
 ```

### 2024-06-13 slides/sections/modeloverview.Rmd
Add slides and model comparisons update

## Model overview

```{r models, results = 'asis'}
tibble(Models = str_c("M",1:4),
       Type = c("LMM", "LMM", "LMM", "MoG"),
       Description = c("Single distribution Gaussian",
                       "Single distribution log-Gaussian",
                       "Single distribution log-Gaussian with different variance components per text location",
                       "Two-distributions mixture of log-Gaussians.")) %>% 
  select(Models, Description) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = T, 
                position = "right",
                font_size = 18) %>% 
  group_rows("Serial", 1, 3) %>% 
  group_rows("Parallel", 4, 4) %>% 
  row_spec(0:4, extra_css = "line-height: 2.25;")  %>% 
  column_spec(1, width = "3cm")

```

### 2024-06-13 slides/sections/mog_app.Rmd
Add slides and model comparisons update

---

::: {style="margin-top:-20px; width: 96%;"}
```{r contpred}
inputPanel(
  sliderInput("beta_0", "Fluent typing", min = 100, max = 500, value = 150, ticks = FALSE),
  sliderInput("delta", "Hesitation slowdown", min = 50, max = 300, value = 100, ticks = FALSE),
  sliderInput("theta", "Hesitation probability", min = .001, max = .999, value = .3, ticks = FALSE),  
  switchInput("params", 
              size = "mini", 
              label = "Show means",
              onLabel = "on", 
              offLabel = "off", 
              labelWidth = "50%",
              width = "50%")
)

renderPlot({
  theme_set(theme_bw(base_size = 16) +
              theme(legend.position = "bottom",
                    legend.justification = "right",
                    legend.key.width = unit(.75, "cm"),
                    panel.grid = element_blank(),
                    axis.title = element_text(hjust = 0)))

  plot_mix_comps <- function(x, mu, sigma, lam) lam * dlnorm(x, mu, sigma)
  size <- .25
  alpha <- .25

  beta_0 <- log(200)
  beta_0 <- log(input$beta_0)
  delta <- 100
  delta <- input$delta
  theta <- .01
  theta <- input$theta
  beta_1 <- log(exp(beta_0) + delta)
  sigma <- .3
  sigma_diff <- .01
  sigmae <- sigma
  sigmae_p <- sigma + sigma_diff
  data <- data.frame(x = c(0, 1500))

  # Formula shown in plot
  formula <- list(~'iki'[i]~"~"~theta~paste("\u00B7")~  "logN("*beta~+~delta*","~sigma[2]^2*")",
                  ~+~"(1" - theta*")"~paste("\u00B7")~"logN("*beta*","~sigma[1]^2*")")

  p2 <- ggplot() +
    annotate("text", 
             x = 1.5, 
             y = seq(1, .25, 
                     length = length(formula)),
             label = formula, 
             parse = TRUE, 
             size = 6) +
    geom_rect(aes(xmin = -Inf, 
                  xmax = Inf, 
                  ymin = -Inf, 
                  ymax = Inf), 
              fill = "blue", 
              alpha = 0.05, 
              color = "black") +
    theme_void() +
    coord_cartesian(xlim = c(0, 3), ylim = c(-2.25, 2.4))

  if(input$params){
    beta_value <- round(exp(beta_0),0)
    theta_value <- dezero(theta, 2)
    delta_value <- round(delta, 2)
    beta_label <- substitute(beta==beta0, list(
      beta0 = beta_value))
    theta_label <- substitute(theta==theta_value, list(
      theta_value = theta_value))
    delta_label <- substitute(delta==delta_value, list(
      delta_value = delta_value))

    p2 <- p2 + 
      annotate("text",
               x = 2.25,
               y = -1,
               label = deparse(beta_label), 
               parse = TRUE, 
               size = 6) +
      annotate("text",
               x = 2.25,
               y = -1.5,
               label = deparse(theta_label), 
               parse = TRUE, 
               size = 6) +
      annotate("text",
               x = 2.25,
               y = -2,
               label = deparse(delta_label), 
               parse = TRUE, 
               size = 6)

  }

  p1 <- ggplot(data, aes(x)) +
    stat_function(geom = "line", fun = plot_mix_comps,
                  size = size,
                  args = list(beta_0, sigmae, lam = 1 - theta)) +
    stat_function(geom = "line", fun = plot_mix_comps,
                  size = size,
                  args = list(beta_1, sigmae_p, lam = theta)) +
    geom_area(stat = "function", 
              fun = plot_mix_comps, 
              show.legend = T, 
              alpha = alpha, 
              size = size, 
              aes(fill = "Fluent typing"), 
              args = list(beta_0, sigmae, lam = 1 - theta)) +
    geom_area(stat = "function", 
              fun = plot_mix_comps, 
              show.legend = T, 
              alpha = alpha, 
              size = size,
              aes(fill = 'Hesitation ("Pause")'), 
              args = list(beta_1, sigmae_p, lam = theta)) +
    geom_segment(x = exp(beta_0), 
                 y = 0, 
                 xend = exp(beta_0), 
                 yend = Inf, 
                 colour = "darkred", 
                 size = 1,
                 alpha = alpha) +
    geom_segment(x = exp(beta_1), 
                 y = 0, 
                 xend = exp(beta_1), 
                 yend = Inf, 
                 colour = "darkred", 
                 size = 1,
                 alpha = alpha) +
    geom_segment(x = exp(beta_0),
                 y = 0.0065, 
                 xend = exp(beta_1), 
                 yend = 0.0065, 
                 colour = "blue", 
                 alpha = alpha,
                 size = 1,
                 arrow = arrow(length = unit(0.3, "cm"), 
                               ends = "both", 
                               type = "closed")) +
    scale_fill_manual(breaks = c("Fluent typing", 'Hesitation ("Pause")'),
                      values = c("#000000","#009E73")) +
    scale_x_continuous(labels = scales::comma, breaks = seq(0, 2500, 200)) +
    coord_cartesian(ylim = c(0, .007), 
                    xlim = c(0, 800)) +
    labs(title = "",
         fill = "Mixture components",
         x = "Inter-keystroke interval [in msecs]",
         y = "Density")

  p1 + p2 + 
    plot_layout(widths = c(3, 1.25)) + 
    plot_layout(guides = 'collect')

})

```
:::

### 2024-06-13 slides/sections/sampledata.Rmd
Add slides and model comparisons update

## Some sample data

Essay-writing task from ProWrite project [e.g. @dux2022automating] without deletions and spaces.

```{r eval = F}
read_csv('tutorial/data/process_data_by_event.csv') %>% 
  filter(location != "post-paragraph", 
         !is.na(location),
         next_event_type == "production",
         event_type == "production",
         event_duration > 50,
         event_duration < 30000,
         session_no == 1) %>%
  drop_na(text) %>%
  mutate(across(text, str_to_lower)) %>% 
  filter(text %in% letters) %>% 
  mutate(.by = c(token, location), 
         idx = sample(n())) %>% 
  filter(idx <= 50) %>% 
  select(participant, iki = event_duration, location, pretext, text) %>% 
  mutate(across(location, ~recode(., 
                                  "within-word" = "within word",
                                  "post-word" = "before word",
                                  "post-sentence" = "before sentence"))) %>% 
  filter(participant %in% sample(unique(participant), 10)) %>% 
  mutate(recent_text = substr(pretext, 
                              start = nchar(pretext) - 40,
                              stop = nchar(pretext))) %>% 
  select(-pretext) %>% 
  rename(new_letter = text) %>% 
  relocate(participant, location, recent_text, new_letter, iki) %>% 
  write_csv("slides/data/sampledata.csv")
```

```{r}
data <- read_csv("slides/data/sampledata.csv")
slice(data, sample(nrow(data), size = 10))
```

## How are these data distributed?

```{r fig.width=10, fig.height=6}
ggplot(data, aes(x = iki, colour = location, fill = location)) +
  geom_density(alpha = .25) +
  scale_colour_colorblind() +
  scale_fill_colorblind() +
  scale_x_log10() 
```

### 2024-06-13 slides/sections/simulation.Rmd
Add slides and model comparisons update

--- 2024-04-27 report/appendix/sim.Rmd
@@
-# Simulation
+```{r loossim, results='asis'}
+table <- read_csv("../../mixture-model-sim/stanout/modelcomparison.csv") %>% 
+  select(1:6) %>% 
+  mutate(
+#    elpd_se_ratio = abs(elpd_diff / se_diff),
+    across(ends_with("diff"), ~round(., 1)),
+    across(ends_with("loo"), ~round(., 0)),
+    across(where(is.numeric), ~format(., big.mark = ",")),
+    across(everything(), ~as.character(.)),
+    across(everything(), ~str_trim(.)),
+    across(everything(), ~str_replace_all(., "NaN", "--")),
+    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
+    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
+    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
+    across(c(data, model), ~recode(., mog = "Two-distribution mixture model",
+                                      lm = "Single-distribution model"))) %>% 
+  select(data, model, starts_with("elpd"))
 
-A possible concern with these results -- substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-To address this concern we repeated a comparison for a single distribution model and a two-distributions mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart.
-In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.
+# names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
+# 
+# table %>% 
+#   select(-Data) %>% 
+#   kable(align = c("l", "r", "r")) %>% 
+#   kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
+#                 full_width = T, 
+#                 position = "right",
+#                 font_size = 16) %>% 
+#   group_rows("Two-distributions data", 1, 2) %>% 
+#   group_rows("Single-distribution data", 3, 4) 
+mixturedata <- table %>% filter(str_detect(data, "Two"), elpd_diff != "--") %>% pull(elpd_diff)
+singledata <- table %>% filter(str_detect(data, "Single"), elpd_diff != "--") %>% pull(elpd_diff)
+```
 
-In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}).
-This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
+## Simulation {.smaller}
 
-\begin{equation}
-\begin{aligned}
-(\#eq:simmog)
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-		& \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
+<div style="float: left; width: 43%;">
+\
+\
 
-This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants).
-The model includes two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.
+- Simulate data from a single log-normal distribution
+- Simulate data from a mixture of to log-normal distribution
+- Data simulated with mixture process: 
+  - Advantage for mixture model over single process model: $\Delta\widehat{elpd} =$ `r mixturedata`
+- Data simulated with single process: 
+  - Negligible difference between models: $\Delta\widehat{elpd} =$ `r singledata`
 
-The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
+</div>
 
-\begin{equation}
-\begin{aligned}
-(\#eq:simuv)
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the single distribution models used in the main analysis.
-The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
-
-The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
-The simulated data are visualised in Figure \ref{fig:simdata}.
-Parameter values were chosen so that the simulated data are roughly similarly distributed to keystroke transitions.
-
-```{r simdata, fig.cap= "Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process.
-The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
+<div style="float: right; width: 50%;">
+```{r simdata, fig.cap= "", fig.width=4, fig.height=4, out.width = "90%"}
 files <- list.files("../../mixture-model-sim/data", full.names = T)
@@
   scale_x_log10(labels = ~format(., big.mark = ",")) +
-  theme(text = element_text(size = 10)) +
-  labs(colour = "Data process", 
-       fill = "Data process", 
-       x = "y", 
-       y = "Density")
+  theme(text = element_text(size = 10),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        axis.title.y = element_blank()) +
+  labs(colour = "Data", 
+       fill = "Data", 
+       x = "y")
 ```
+</div>
 
-For each of these two data sets we simulated 1,000 observations.
-We run 2 models -- a two-distribution mixture model and a single-distribution model -- each for both data sets.
-Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
-Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}.
-The parameters are shown by type of data generating process along with the true parameter values.
-Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they single-distribution model.
+```{r eval = F}
+## Parameter estimates
 
-```{r}
 mog1 <- read_csv("../../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
@@
 
-```{r simparam, results='asis'}
+```{r simparam, results='asis', eval = F}
 table <- bind_rows(mog_sim %>% mutate(model = "mog"),
@@
 
-names(table) <- c("Model", "Parameter with true value", "Two-distributions model", "Single-distributions model")
+names(table) <- c("Model", "Parameter values used in simulation", "Two-distributions model", "Single-distributions model")
 
-apa_table(table[,-1], 
-          escape = FALSE,
-      col_spanners = list("Estimate with 95\\% PI" = 2:3),
-      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
-      align = c("l", rep("r", 2)),
-      stub_indents = list(`Two-distributions data` = 1:5, 
-                          `Single-distribution data` = 6:7))
+table %>% 
+  select(-Model) %>% 
+  kable(align = c("l", "r", "r")) %>% 
+  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
+                full_width = T, 
+                position = "right",
+                font_size = 16) %>% 
+  group_rows("Two-distributions data", 1, 5) %>% 
+  group_rows("Single-distribution data", 6, 7) %>% 
+  add_header_above(c(" " = 1,
+                     "Estimate with 95\\% PI" = 2))
 
 ```
-
-We used LOO-CV to compare the predictive performance of the two models for each data generating process.
-The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
-The results rule out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a single-distribution process.
-However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
-In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error -- a *z*-score approximation -- reveals a value of 11.6 which is strong evidence in in favour of the mixture model compared to the single-distribution model (Sivula et al., 2020).
-In comparison, for the single-distribution data, the single-distribution model was negligibly better than the two-distribution mixture model which was evidence by a value of 0.77 for the ratio of $\Delta\widehat{elpd}$ and its standard error.
-Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the single distribution is preferred by the law of parsimony.
-
-```{r loossim, results='asis'}
-table <- read_csv("../../mixture-model-sim/stanout/modelcomparison.csv") %>% 
-  select(1:6) %>% 
-  mutate(
-#    elpd_se_ratio = abs(elpd_diff / se_diff),
-    across(ends_with("diff"), ~round(., 1)),
-    across(ends_with("loo"), ~round(., 0)),
-    across(where(is.numeric), ~format(., big.mark = ",")),
-    across(everything(), ~as.character(.)),
-    across(everything(), ~str_trim(.)),
-    across(everything(), ~str_replace_all(., "NaN", "--")),
-    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
-    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
-    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
-    across(c(data, model), ~recode(., mog = "Two-distribution mixture model",
-                                      lm = "Single-distribution model"))) %>% 
-  select(data, model, starts_with("elpd"))
-
-names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")
-
-apa_table(table[,-1], 
-          escape = FALSE,
-      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
-     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
-      align =c("l", rep("r", 2)),
-      stub_indents = list(`Data: Two-distribution mixture process` = 1:2, 
-                          `Data: Single-distribution process` = 3:4))
-
-```

### 2024-06-13 slides/sections/statisticalmodels.Rmd
Add slides and model comparisons update

## Serial view: single-process model {.smaller}

<div style="float: left; width: 40%;">

```{r fig.width=5}
```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
$$

</div>

## Serial view: single-process model  {.smaller}

<div style="float: left; width: 40%;">

```{r fig.width=5}
linewidth <- .35
data <- data.frame(x = c(200, 1500))
word_mean <- 800
sent_mean <- 950
sd <- c(40, 60)
word_lbl <- "word initial"
sent_lbl <- "sentence initial"

ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
$$

</div>

## Serial view: single-process model  {.smaller}

<div style="float: left; width: 40%;">

```{r fig.width=5}
ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial")) +
  stat_function(fun = dnorm, n = 101, args = list(mean = sent_mean, sd = sd[2]), 
                linewidth = linewidth,
                aes(colour = "sentence initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  geom_area(stat = "function", fun = dnorm, aes(fill = sent_lbl), alpha = .25, 
            args = list(mean = sent_mean, sd = sd[2])) +
  geom_curve(aes(x = word_mean, xend = sent_mean, y = 0.01, yend = 0.00675),
             curvature = -0.2, 
             size = .25,
             arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x=(sent_mean+word_mean)/2, y=.0105, label=paste0("beta[diff]"), size = 4, parse=TRUE) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta_\text{textlocation[i]}, \sigma_{e_\text{textlocation[i]}}^2)
$$

</div>

## Parallel view: mixture of two processes {.smaller}

<div style="float: left; width: 40%;">

</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma_{e'}^2) + \\
   (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
$$

</div>

## Parallel view: mixture of two processes {.smaller}

<div style="float: left; width: 40%;">

```{r fig.width=5}
# function for mixture models
plot_mix_comps <- function(x, mu, sigma, lam) lam * dnorm(x, mu, sigma)

# Parameters
data <- data.frame(x = c(50, 1500))
theta_conj <- .3
theta_simp <- .1
alpha <- 250
delta <- c(250, 500)
alpha2_w <- alpha + delta[1]
alpha2_s <- alpha + delta[2]
sigma <- 40
sigma_diff <- c(30, 60)
sigmae <- sigma
sigmae_p_w <- sigma + sigma_diff[1]
sigmae_p_s <- sigma + sigma_diff[2]

ggplot(data, aes(x)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_x_continuous(labels = scales::comma, 
                     limits = c(50, 1000), 
                     breaks = seq(0, 1500, 200)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma_{e'}^2) + \\
   (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
$$

</div>

## Parallel view: mixture of two processes {.smaller}

<div style="float: left; width: 40%;">

```{r fig.width=5}
ggplot(data, aes(x)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = sent_lbl),
                linewidth = linewidth,
                args = list(alpha, sigmae, lam = 1 - theta_conj)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = sent_lbl),
                linewidth = linewidth,
                args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
            args = list(alpha, sigmae, lam = 1 - theta_conj)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
            args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
  geom_curve(aes(x = alpha2_w, 
                 xend = alpha2_s, y = 0.0008, yend = 0.0015),
             curvature = -0.2, 
             size = .15,
             arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", 
           x = (alpha2_w + alpha2_s)/2,
           y=.0022, 
           label=paste0("theta[diff]", "~and~", "delta[diff]"), size = 4, parse=TRUE) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_x_continuous(labels = scales::comma, 
                     limits = c(50, 1000), 
                     breaks = seq(0, 1500, 200)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_{i} \sim\text{ } \theta_\text{location[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
   (1 - \theta_\text{location[i]}) \times \text{log}\mathcal{N}(\beta, \sigma_{e_\text{location[i]}}^2)
$$

</div>

### 2024-06-13 slides/sections/textlocations.Rmd
Add slides and model comparisons update

## Text location classifications

```{r keyloc, results='asis'}
tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
                Description = c("Transitions between any letter",
                                "Keypress after space followed by any letter",
                                "Keypress following a space preceding any letter"),
                Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d. T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.", 
                            "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
                            "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept.")) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = T, 
                position = "right",
                font_size = 16) %>% 
  add_footnote("Note: Key intervals that terminated in a revision were removed.") %>% 
  row_spec(0:3, extra_css = "line-height: 3.25;")  # Adjust the number for more spacing

```

### 2024-06-13 slides/sig-paris-2024-keynote/main.Rmd
Add slides and model comparisons update

---
title: '<span style="font-size: 50px; font-face: bold">Modelling cascaded planning in written production as Bayesian hierarchical mixture process</span>'
author: '<span style="font-size: 30px; font-face: bold">Jens Roeser (Nottingham Trent University)</span>'
output: 
  ioslides_presentation:
    incremental: false
    transition: slower
    widescreen: true
    css: ../misc/slides.css
    self_contained: true
#    logo: ../misc/gfx/ntu.png
runtime: shiny
bibliography      : ["../../report/references.bib"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      eval = TRUE, 
                      message = FALSE, 
                      comment=NA,
                      fig.align = "right")
options("kableExtra.html.bsTable" = T, digits = 2)
options(pillar.print_min = 5, pillar.print_max = 10)
library(shiny)
library(shinyWidgets)
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
library(ggthemes)
library(polspline)
source("../../scripts/functions.R")
source("../scripts/get_pdens_plot.R")

beta_label <- "Fluent interkey\nintervals (in msecs)"
delta_label <- "Slowdown for hesitant\ninterkey intervals (in secs)"
theta_label <- "Probability of hesitant\ninterkey intervals"

theme_set(theme_few(base_size = 12) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))
```

## Keystroke interval data

```{r}
data_pseq_1 <- read_csv('tutorial/data/process_data_by_event.csv') %>% 
  filter(location != "post-paragraph", 
         !is.na(location),
         next_event_type == "production",
         event_type == "production",
         event_duration > 50,
         session_no == 1) %>% 
  filter(participant == unique(participant)[1], 
         pseq_id == 10, 
         session_no == 1) %>% 
  select(text, iki = event_duration, location) %>% 
  rowid_to_column("id")
```

```{r fig.width=10, fig.height=6}
d <- data_pseq_1 %>% 
  mutate(
    across(location, ~str_replace(., "post-", "before ")),
    across(location, ~str_replace(., "-", " ")),
    across(text, ~replace_na(., " ")))

ggplot(d, aes(y = iki, x = id, group = 1)) +
  geom_path() +
  geom_point(aes(colour = location, shape = location), size = 3.5) +
  theme_classic(base_size = 14) +
  labs(y = "Interkey intervals (in msecs)") +
  scale_x_continuous(breaks = d$id, labels = d$text) +
  scale_colour_colorblind() +
  scale_shape_manual(values = c(8, 17, 16)) +
  scale_y_log10() +
  labs(colour = "text locations",
       shape = "text locations") +
#  scale_y_continuous(limits = c(0, 1100)) +
  theme(axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "bottom")
```

# How are the processes that generate and sequences keystrokes mentally coordinated?

```{r intro, child = "../sections/introduction.Rmd"}
```

```{r intro, child = "../sections/evidence.Rmd"}
```

```{r intro, child = "../sections/sampledata.Rmd"}
```

## Usual practice

```{r echo = T}
data_with_pause <- data %>% 
  mutate(is_pause_500 = iki > 1000,
         is_pause_1k = iki > 1000,
         is_pause_2k = iki > 2000)
data_with_pause
```

```{r echo = T}
data_with_pause %>% 
  summarise(across(starts_with("is_pause"), list(avg = mean)), .by = location)
```

## Usual practice

```{r echo = T}
data_with_pause %>% 
  filter(!is_pause_500) %>% 
  summarise(iki_avg = mean(iki),
            iki_sd = sd(iki), 
            .by = location)
```

# How can we represent these models statistcally?

## Problems

- Pause thresholds are in principle arbitrary.
- Separation of distributions does not capture the process that generates keystroke data.

```{r data, child = "../sections/statisticalmodels.Rmd"}
```

```{r intro, child = "../sections/mog_app.Rmd"}
```

# How can we apply these formulae to data?

## Single process models

## Why Bayesian models?

```{r}

# mixture models map directly on parallel cascaded architecture of typing

# how did people usually deal with pauses -> log, threshold

# Why Bayesian framework?

# start with normal models and lm, lmer, brms

# show Bayesian inference

```

## Thank you!

This work was supported by UKRI ESRC under Grant No. ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production
difficulties shape text formation." (Nottingham Trent University)

Data and analysis scripts written in R and Stan can be found on OSF https://osf.io/z65dw/ [@roeser2024osf].

For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see
https://rpubs.com/jensroes/mixture-models-tutorial.

"Ideas cascading into keystrokes -- Modelling writing hesitations as Bayesian mixture process" June 27 Session D7 (D 321) 10.30am-12.00pm

Roeser, J., De Maeyer, S., Leijten, M., & Van Waes, L. (2024). Modelling typing disfluencies as finite mixture process. *Reading and Writing*, 37(2), 359-384.

## References {.smaller}

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):after {
  content: '';
}
</style>

### 2024-06-13 slides/sig-paris-2024-talk/main.Rmd
Add slides and model comparisons update

---
title: '<span style="font-size: 50px; font-face: bold">Ideas cascading into keystrokes -- Modelling writing hesitations as Bayesian mixture process</span>'
author:  '<span style="font-size: 25px; font-face: bold">Jens Roeser, Mark Torrance (both Nottingham Trent University), Rianne Conijn (Eindhoven University of Technology), Evgeny Chukharev (Iowa State University)</span>'
output: 
  ioslides_presentation:
    incremental: false
    transition: slower
    widescreen: true
    css: ../misc/slides.css
    self_contained: true
#    logo: ../misc/gfx/ntu.png
bibliography      : ["../../report/references.bib"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      eval = TRUE, 
                      message = FALSE, 
                      comment=NA,
                      fig.align = "right")
options("kableExtra.html.bsTable" = T, digits = 2)
options(pillar.print_min = 5, pillar.print_max = 6)
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
library(ggthemes)
library(polspline)
source("../../scripts/functions.R")
source("../scripts/get_pdens_plot.R")

beta_label <- "Fluent interkey\nintervals (in msecs)"
delta_label <- "Slowdown for hesitant\ninterkey intervals (in secs)"
theta_label <- "Probability of hesitant\ninterkey intervals"

theme_set(theme_few(base_size = 12) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))
```

##

<div style="float: left; width: 70%;">
\
\

```{r fig.width=7}
target <- str_split_1("These are keystroke intervals", pattern = "")
x <- 1:length(target)
y <- c(900,
       rnorm(9, 200, 50), 500, 
       rnorm(9, 200, 50), 400, 
       rnorm(8, 200, 50))

ggplot(data = NULL,aes(y = y, x = x)) +
  geom_path() +
  geom_point(size = 5, shape = 21) +
  theme_classic(base_size = 14) +
  labs(x= "Produced key", 
       y = "Interkey intervals (in msecs)") +
  scale_x_continuous(breaks = x, labels = target) +
  scale_y_continuous(limits = c(0, 1100)) +
  theme(axis.ticks = element_blank())
```

</div>

<div style="float: right; width: 25%;">
\
\

- Understanding factors that influence interkey intervals requires a theory of how keystrokes are sequenced.
- Which processes are required and how are they coordinated?

</div>

# How do we produce texts?

```{r intro, child = "../sections/introduction.Rmd"}
```

```{r intro, child = "../sections/evidence.Rmd"}
```

# How can we represent these models statistcally?

```{r data, child = "../sections/statisticalmodels.Rmd"}
```

# How can we evaluate these models?

```{r data, child = "../sections/datasets.Rmd"}
```

```{r textloc, child = "../sections/textlocations.Rmd"}
```

```{r modeloverview, child = "../sections/modeloverview.Rmd"}
```

```{r implementation, child = "../sections/implementation.Rmd"}
```

# Which model showed better performance?

```{r modelcompresults, child = "../sections/modelcomparisons.Rmd"}
```

```{r fittodata, child = "../sections/modelfit.Rmd"}
```

# How do text locations affect the mixture process?

```{r locresults, child = "../sections/locationeffects.Rmd"}
```

# Planning text unfolds in parallel to production!

```{r sim, child = "../sections/discussion.Rmd"}
```

# "Maybe mixture models just always fit bettter?"

```{r sim, child = "../sections/simulation.Rmd"}
```

## Thank you!

We are grateful for all authors who made their available (to us), in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues. This work was supported by

- National Science Foundation under Grant No. 2016868: "ProWrite: Biometric feedback for improving college students' writing processes." (Iowa State University)
- UKRI ESRC under Grant No. ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production
difficulties shape text formation." (Nottingham Trent University)

Data and analysis scripts written in R and Stan can be found on OSF https://osf.io/z65dw/ [@roeser2024osf].

For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see
https://rpubs.com/jensroes/mixture-models-tutorial.

## References {.smaller}

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):after {
  content: '';
}
</style>

### 2024-06-13 scripts/c2l1/modelcomparison.R

@@ -1,3 +1,4 @@
+rm(list = ls())
 library(loo)
 library(tidyverse)
 
@@ -16,7 +17,7 @@ for(file in files){
 (loos <- ls(pattern = "loo_.*"))
 
 mc1 <- do.call(what = loo_compare, args = lapply(loos[2:1], as.name))
-mc2 <- do.call(what = loo_compare, args = lapply(loos[2:3], as.name))
+mc2 <- do.call(what = loo_compare, args = lapply(loos[c(1,3)], as.name))
 mc3 <- do.call(what = loo_compare, args = lapply(loos[3:4], as.name))
 mc4 <- do.call(what = loo_compare, args = lapply(loos[4:5], as.name))
 
@@ -25,6 +26,6 @@ mcs <- map_dfr(.x = list(mc1, mc2, mc3, mc4), ~as.data.frame(.x) %>%
   as_tibble() %>% 
   slice(2)) %>%
   mutate(model = c("M1 -- M2", "M2 -- M3", "M3 -- M4", "M4 -- M5"),
-         elpd_diff_ratio = abs(elpd_diff / se_diff))
+         elpd_diff_ratio = abs(elpd_diff / se_diff));mcs
 
 write_csv(mcs, "stanout/c2l1/modelcomparison.csv")

### 2024-06-13 scripts/spl2_shift/modelcomparison.R

@@ -1,3 +1,4 @@
+rm(list = ls())
 library(loo)
 library(tidyverse)
 
@@ -15,7 +16,7 @@ for(file in files){
 
 (loos <- ls(pattern = "loo_.*"))
 mc1 <- do.call(what = loo_compare, args = lapply(loos[2:1], as.name))
-mc2 <- do.call(what = loo_compare, args = lapply(loos[2:3], as.name))
+mc2 <- do.call(what = loo_compare, args = lapply(loos[c(1,3)], as.name))
 mc3 <- do.call(what = loo_compare, args = lapply(loos[3:4], as.name))
 mc4 <- do.call(what = loo_compare, args = lapply(loos[4:5], as.name))
 
@@ -25,8 +26,6 @@ mcs <- map_dfr(.x = list(mc1, mc2, mc3, mc4), ~as.data.frame(.x) %>%
                  slice(2)) %>%
   mutate(model = c("M1 -- M2", "M2 -- M3", "M3 -- M4", "M4 -- M5"),
          elpd_diff_ratio = abs(elpd_diff / se_diff))
-
-glimpse(mcs)
 
 write_csv(mcs, "stanout/spl2_shift/modelcomparison.csv")
 

### 2024-06-13 slides/scripts/get_pdens_plot.R

@@ -1,0 +1,31 @@
+get_pdens_plot <- function(file, max_iki){
+  #file <- "stanout/c2l1/all_sims.csv
+  #max_iki <- 2000
+  read_csv(file) %>% 
+    filter(sim_idx <= 50) %>% 
+    pivot_longer(cols = c(y_obs, y_tilde)) %>% 
+    mutate(across(name, recode_factor, 
+                  y_tilde = "Simulated",
+                  y_obs = "Observed",
+                  .ordered = T)) %>% 
+    filter(name ==  "Simulated" |
+          (name == "Observed" & sim_idx == 1),
+          value <= max_iki,
+          model != "mogbetaunconstr") %>% 
+    mutate(across(model, ~case_when(
+      str_detect(., "mogbetacon.+") ~ "Two log-Gaussians",
+#      . == "mogbetaunconstr" ~ "Two log-Gaussians (unconstrained)",
+      . == "lmmuneqvar" ~ "Single log-Gaussian (unequal variance)",
+      . == "lmm" ~ "Single log-Gaussian",
+      . == "lmmgaus" ~ "Single Gaussian")),
+      across(model, ~str_wrap(., 20)),
+      across(model, ~factor(., levels = sort(unique(model))[c(4, 3, 2, 1)], ordered = T))) %>% 
+    ggplot(aes(x = value / 1000, 
+               group = interaction(sim_idx,name), 
+               colour = name)) +
+    geom_density() +
+    facet_wrap(~model, scales = "free_x", nrow = 1) +
+    scale_colour_manual(values = c("firebrick3", "black")) +
+    scale_x_continuous(labels = scales::comma) +
+    labs(colour = "") 
+}

### 2024-10-26 report/appendix/fit2data.Rmd
After review update

@@
 
-Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data.
-From each model we obtained 50 simulations (predictions); see Figure \ref{fig:prediction}.
-The model predictions fitted the data substantially better for the mixture models.
-The worse fit was observed for the single distribution Gaussian model which predicted a larger number of negative keystroke intervals; a log-normal distribution removed these negative predictions.
+Model fit was assessed by testing to what extent data simulated from the model match the observed data.
+From each model we obtained 50 simulations (predictions).
+Simulated and observed data are visualised in Figure \ref{fig:prediction}.
+Data predicted by the mixture model were substantially more similar to the observed data compared to any of the single-distribution models.
+The worse fit was observed for the single-distribution Gaussian model which predicted a larger number of negative keystroke intervals.
 
@@
 
-```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red in comparison to the interkey interval data shown in black by model types and datasets.
-Model fit to the data was order from the best (left) to the weakest (right).
-For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
+```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red compared to interkey interval data shown in black shown by model and datasets.
+Better fit i shown on the left and weaker fit is shown on the right.
+X-axes were truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
 knitr::include_graphics("figures/fitplots.pdf")

### 2024-10-26 report/appendix/genre_constr.Rmd
After review update

@@
 
-From the LIFT data, we compared the difference between genres, i.e.
-when writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}.
+From the LIFT data, we assessed differences between genres, i.e.
+writing an informative as opposed to argumentative text.
+Results are shown in Table \ref{tab:genreeffect}.
 Cellmeans and differences were average across writing topic.
-Evidence for a genre comparisons was negligible.
+Evidence for all comparisons was negligible.
 
@@
 cellmeans <- ps_lift %>% 
+  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
   separate(group, into = c("topic","genre"), sep = "; ") %>% 
@@
   mutate(across(ends_with("prob"), ~dezero(., 2)),
-         across(c(ends_with("beta"), ends_with("delta")), 
+         across(c(ends_with("beta"), 
+                  ends_with("delta")), 
                 ~as.character(scales::comma(round(.))))) %>% 
@@
          across(genre, ~ifelse(. == "Inf", "INF", .))) %>% 
-  pivot_wider(names_from = genre, values_from = value) %>% 
+  pivot_wider(names_from = genre, 
+              values_from = value) %>% 
   unnest(c(ARG, INF)) %>% 
@@
           align = c("l", "r", "r", "r", "r"), 
-          caption = "Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for the hesitation slowdown and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.",
+          caption = "Mixture-model estimates for genre effect.
+Cellmeans for hesitation duration (in msecs) and the probability of hesitant transitions are shown for argumentative and informative texts.
+The genre effect is shown on the log scale for hesitationn durations and on the logit scale for the probability of hesitant transitions; 95% PIs in brackets.",
           note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",

### 2024-10-26 report/appendix/l2_constr.Rmd
After review update

@@
 For the SPL2 data we calculated the L2 effect (i.e.
-the difference between writing in L2 and L1).
+the effect of writing in L2 or L1).
 The results can be found in Table \ref{tab:l2effect}.
-The results show longer hesitation duration and more pauses across all transition locations when writing in L2.
-Only within-word transitions did show negligible evidence for an increased hesitation duration in L2.
+The results show longer hesitation duration and more hesitations across all transition locations when writing in L2.
+Only within-word transitions showed negligible evidence for more or longer hesitations.
 
@@
 cellmeans <- ps_spl2_both %>% 
+  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
   separate(group, into = c("group", "lang"), sep = "; ") %>%
@@
           caption = "Mixture model estimates for language effect.
-Cell means are shown the slowdown for hesitant interkey intervals when writing in L1 and L2 and the probability of hesitant keyintervals.
-The language difference is shown on log scale (for durations) and logit scale for probability of hesitant key intervals.
-95% PIs in brackets.",
+Cellmeans are shown for the hesitation duration and for the probability of hesitant interkey intervals for texts produced in the writer's L1 and L2.
+Language differences are shown for the hesitation duration (log scale) and the hesitation probability (logit scale); 95% PIs in brackets.",
           note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
           font_size = "small",
-          stub_indents = list("Hesitation showdown" = 1:3,
+          stub_indents = list("Hesitation duration" = 1:3,
                               "Hesitation probability" = 4:6) )

### 2024-10-26 report/appendix/masking_constr.Rmd
After review update

@@
 
-Studies associated with two datasets (CATO, GE2) investigated to what extent masking the previously written text affects keystroke behaviour.
-The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}.
-There was some evidence for longer before-sentence hesitation durations when the text was masked in non-dyslexic writers.
+Studies associated with the CATO and GE2 datasets investigated to what extent hiding (masking) previously written text from the reader affects keystroke behaviour.
+Mixture model results for the masking effect are shown in Table \ref{tab:maskingeffect}.
+There is some evidence that when the text was masked dylexic writers hesitated longer before starting to type sentences.
 Evidence for all other comparisons was negligible.
@@
   bind_rows(ps_cato_2) %>%
+  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
   mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
@@
           align = c("l", "l", "l", "r", "r", "r", "r"), 
-          caption = "Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task for the slowdown (in msecs) and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.",
+          caption = "Mixture-model estimates for masking effect.
+Cellmeans are shown for the masked and unmasked writing task for hesitation duration (in msecs) and hesitation probability.
+The masking effect is shown on the log scale for durations and on the logit scale for probabilities; 95% PIs in brackets.",
           note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
@@
           stub_indents = list(
-            "Hesitation slowdown" = 1:9,
-            "Hesitattion probability" = 10:18) )
+            "Hesitation duration" = 1:9,
+            "Hesitation probability" = 10:18) )
 

### 2024-10-26 report/appendix/models.Rmd
After review update

--- 2024-05-28 report/appendix/models.Rmd
@@
 
-We used the Bayesian framework [e.g.
-@farrell2018computational; @gelman2014; @lee2014bayesian] to implement four statistical models to evaluate which -- the serial or the parallel view -- captures interkey data more successfully.
-In other words, we use statistical models to map between the keystroke intervals and the theoretically assumed process that generates theses data to then compare the predictive power of those models.
-These models presented in this section are building on one another so that later models include assumptions of earlier models.
+We implemented four statistical models using the Bayesian framework [e.g.
+@farrell2018computational; @gelman2014; @lee2014bayesian] to evaluate the serial or the parallel view.
+In other words, we used statistical models to map between keystroke data and the theoretically assumed cognitive process that underlies the generation of theses data.
+These models are presented in more detail in this section.
 
-## Single-distribution Gaussian
+## Single-distribution Gaussian model
 
-Under the serial view, all planning must be completed prior to typing onset of the corresponding planning unit.
-The resulting interkey interval is sometimes faster or slower depending on, among others, psycholinguistic factors; i.e.
-the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes.
-For example, we can capture the variability associated with word-features by assuming that before-word interkey intervals can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average interkey interval associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the remaining variability associated with factors that we did not further specify in the model.
-This can be expressed as $\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)$ where $\text{iki}_\text{before word}$ are all interkey intervals immediately proceeding a word.
-Of importance is the estimated posterior value of $\mu$ as this value captures average time needed to mentally prepare a word.
+Under the serial view, all planning relevant for the to-be-produced utterance must be completed prior to typing onset.
+The duration of the resulting interkey interval depends on, among others, psycholinguistic factors; i.e.
+interkey intervals that are located before easily retrievable high-frequency words tend to be short; similar for words with fewer graphemes, syllables, and morphemes [@torrance2018timed; @roeser2018advance; @nottbusch2005syllabic].
+We can capture the variability associated with word-features by assuming that before-word interkey intervals can be described as coming from a normal (Gaussian) distribution that can be fully characterised by an unknown central tendency $\mu$ of interkey intervals associated with word-level planning and a dispersion $\sigma_\text{e}^2$, that captures the residual variance in the data that is not captured by the model.
+This can be expressed as $\text{iki}_\text{before word} \sim \mathcal{N}(\mu, \sigma_\text{e}^2)$ where $\text{iki}_\text{before word}$ are all interkey intervals immediately proceeding a word.
+Of importance is the estimated value of the central tendency parameter $\mu$ representing in this case the average time needed to mentally prepare a word.
 
-We can extend this simple model of word-planning to other linguistic location.
-Larger linguistic edges are typically understood as being associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which we captured before as average interkey interval $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07].
-Using regression formulae we can capture changes in the outcome variable that are associated with other factors.
-More specifically we can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is then the average interkey interval for word boundaries but, when $\text{x}_\text{sentence}$ takes on the value 1, the average interkey interval for word boundaries $\alpha$ is incremented by a slowdown associated with sentence boundaries of $\beta$ msecs.
-Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will therefore provide us with an estimate of the parameter value that can be used for statistical inference (e.g.
-whether there is a statistically meaningful difference associated with word and sentence boundaries).
+We can extend this simple model of word planning to other linguistic locations.
+Edges of larger linguistic units are typically understood as being associated with planning on higher levels.
+For example, at sentence boundaries, planning needs to happen for word-level properties -- which we captured before with the parameter $\mu$ -- but also for higher level information such as what meaning it is that should be conveyed, which word to plan and output first, and potentially even dependencies of the sentence-initial noun [@roeser2018advance; @not07].
+We can capture factors that we assume to impact interkey intervals by decomposing $\mu$ into the general form $\mu = \alpha + \beta \times \text{x}$.
+For example, to capture the additional planning needed at sentence boundaries we can add the predictor $\text{x}_\text{sentence[0,1]}$ so that $\mu = \alpha + \beta \times \text{x}_\text{sentence[0,1]}$.
+This states that when the value of $\text{x}_\text{sentence}$ takes on 0, the equation reduces to $\mu = \alpha$ which is then the average interkey interval for word boundaries but, when $\text{x}_\text{sentence}$ takes on the value 1, the average interkey interval for word boundaries $\alpha$ is incremented by a changed in the outcome variable of $\beta$ msecs.
+The value of the $\beta$ parameter represents, therefore, the additional planning necessary for sentences.
+The application of such a statistical model to the data therefore provides us with an estimate of the parameter value that can be used for statistical inference (e.g.
+to determine whether there is evidence for a statistically meaningful difference for interkey intervals immediately proceeding words vs sentences).
 
-We implemented the average interkey interval associated with a given text locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location.
-Therefore, the model returns three distributions of $\beta$ values that capture the posterior distribution of average interkey intervals for each transition location.
-The decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is called random intercepts $u_\text{participant}$.
-The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$.
-The $u_\text{participant}$ value is therefore the difference between the overall posterior estimate and estimate average interkey interval of a particular participant (i.e.
-a positive value indicates that a participant is slower than average; a negative value indicates that a participant is faster than average).
-
-The resulting model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed-effects model
+We implemented such a model -- a standard Gaussian mixed-effects model -- as described in equation \ref{eq:unimodgaus}.
+This model assumes that interkey intervals $\text{iki}_i$ with $i \in 1 \ldots N$ where $N$ is the total number of interkey intervals come from a Gaussian distribution with a mean $\mu$,
 
@@
 (\#eq:unimodgaus)
-\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
+\text{iki}_i \sim\text{ } & \mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
 \text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant[i]} \sim \text{N}(0, \sigma_\text{p}^2)\\
+& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2)\\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
@@
 
-where $i$ iterates over every interkey interval $i \in 1 \dots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.
+which is a linear function of the $i^\text{th}$ transition location (we used there levels: before sentence, before word, within word) captured by $\beta_\text{location}$, and the random-intercept term $u_\text{participant}$ which is constrained to come from a normal distribution with a mean of 0 and the standard deviation $\sigma_\text{p}^2$.
+For the random-intercept term, $\text{participant[i]}$ is the participant associated with the $i^\text{th}$ interkey interval.
+The posterior of $u_\text{participant}$ is therefore the difference between the posterior interkey-interval estimate of each participant (i.e.
+a positive value indicates that a participant is slower than average; a negative value indicates that a participant is faster than average) and allows to capture the fact that some writers are faster and other writers are slower.
+Importantly this model returns posterior probability distributions with interkey-interval estimates for each $\beta$, one for each transition location.
 
-Standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
+Standard-deviation parameters were constrained to be positive because standard deviations can, by definition, never be negative.
 
-## Single distribution log-Gaussian
+## Single-distribution log-Gaussian model
 
-The previous model assumes that the data-generating process follows a Gaussian distribution.
-As a consequence the Gaussian model generates values that are symmetrically distributed and, in principle, allows values to be negative.
-The model presented in this section is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
+The previous model assumes a Gaussian probability function as the underlying data-generating process.
+The model presented in this section is identical to the previous model but instead of assuming a Gaussian probability function, we assume that the process that generates the follows a log-normal (log-Gaussian) distribution.
 
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-This is a desirable property because the distance between two subsequent key events must be positive.
-Instead the lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
+There are, at least, two reason for why a log-Gaussian distribution is more appropriate than a standard Gaussian: (1) the log-Gaussian have a natural lower bound.
+This is a desirable property because the distance between two subsequent keystroke events is by definition positive.
+The lower bound of the distribution of interkey intervals is delimited by a writer's ability to move their fingers and keyboard polling.
+(2) the log-scale is known to be a better match for data from human motor behaviour and response time data than the normal distribution [@wagenmakers2007linear; @baayen2008analyzing].
+This is because in a Gaussian distribution the distance between adjacent units is linear.
+In other words a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
 This does not map onto the psychological interpretation for short and long interkey intervals.
-For example effects that result from the motor level within words (e.g.
-typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, an effect of 25 msecs can be large in the context of fast inerkey intervals but it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of interkey intervals (motor activity) is more psychologically meaningful than a 25 msecs difference on the upper end (retrieving words, planning sentences).
+For example effects that result from difficulty on the motor level (e.g.
+executing familiar key combinations such as *ng* compared to *gn*) are typically smaller than differences that are due to higher levels of processing (e.g.
+struggling to retrieve a word in an L1 or L2).
+In other words, an effect of 25 msecs is relatively large in the context of lower-level motor events but small in the context of higher-level cognitive activity such as planning what to say next.
+Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end (motor activity) is more psychologically meaningful than a 25 msecs difference on the upper end of the distribution interkey intervals (retrieving words, planning contents).
 
-This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.
+This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\mathcal{N}()$ was replaced by $\log\mathcal{N}()$ to represent a log-normal distribution.
 
@@
 (\#eq:unimodloggaus)
-\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
+\text{iki}_i \sim\text{ } & \log\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
 \text{where: } &
@@
 
-## Single distribution unequal-variance log-Gaussian
+## Single-distribution unequal-variance log-Gaussian model
 
-The third model is identical to the model presented in the previous section but removed the equal-variance assumption for the width of the distributions assocated with transition locations.
-The previous models modelled interkey intervals as a function of transition location meaning that the average interkey interval depends on its position in the text.
-The variance associated with the estimated interkey intervals for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human motor behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing].
-For interkey intervals pauses at larger linguistic edges can therefore be assumed to be associated with a larger variance also because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduced the assumption that standard deviations $\sigma_{e_\text{location}}^2)$ vary by transition location.
+The third model is identical to the model presented in the previous section but relaxed the equal-variance assumption associated with transition location.
+In particular the previous models assumed that the variances associated with each transition location were identical.
+This assumption does however not fit with what we know about data from human motor behaviour.
+Longer latencies are known to be associated with a larger variance for human motor behaviour [@wagenmakers2007linear; @wing1973response; @schoner2002timing].
+For interkey intervals pauses at edges of larger linguistic units can therefore be assumed to be associated with a larger variance.
+In equation \ref{eq:unimoduv}, we introduced this assumption by allowing the standard deviations $\sigma_{e_\text{location}}^2)$ to vary by transition location.
 
@@
 (\#eq:unimoduv)
-\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
+\text{iki}_i \sim\text{ } & \log\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
 \text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
@@
 
-## Two-distributions log-Gaussian mixture model
+## Two-distribution log-Gaussian mixture model
 
-The following model is an extension of the last single-distribution model.
-In contrast to the models introduced for the serial view, the cascading view allows planning to unfold in parallel to production.
-Therefore, we removed for the model presented in this section the constraint that all planning must be completed before writing onset.
-Instead of assuming that larger linguistic edges shift the distribution of interkey intervals as in the previous models, we modelled pauses in the distribution of interkey intervals, under the parallel view, to be more likely to occur at larger linguistic edges.
-This was achieved by introducing the model assumption that interkey intervals come from a weighted combination (mixure) of two distributions associated with two different states:
+The following model is an extension of the previous single-distribution model.
+In contrast to the models introduced for the serial view of writing, the parallel view assumes that planning occurs while previously planned language units are being executed in writing.
+For the model presented in this section we removed the constraint that all planning must be completed before writing onset.
+In other words, instead of assuming that different linguistic edges shift the distribution over average interkey intervals towards larger values -- as in the previous models -- the parallel view assumes the frequency of observing a hesitant intekey interval and the size of this hesitation depends on text location of the key transition.
+This was achieved by introducing the assumption that interkey intervals come from a weighted combination (i.e.
+mixture) of two distributions associated with two different states:
 
 1.
-Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and were captured by $\beta$ in equation \ref{eq:bimodcon}.
-Note that $\beta$ is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.
+Activation flows into keystrokes without interruption.
+These fluent interkey intervals are merely constrained by a writer's ability to move their fingers and were captured by $\beta$ in equation \ref{eq:bimodcon}.
+In other words, $\beta$ represents the average typing speed for fluent transitions between keys.
+Note that the $\beta$ parameter is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon}.
 
 2.
-Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when information, for example words or their spelling, could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon} which represents the magnitude of the slowdown associated with transition location.
-The slowdown $\delta$ was allowed to vary by transition locations because hesitations at larger linguistic units are likely to be associated with higher level planning which causes longer delays in the output.
-The $\delta$ parameter was constrained to be positive, so that it captures how much longer hesitant interkey intervals are in addition to $\beta$.
+Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals when information was not available in time, for example when competition occurs during lexical retrieval or when its orthographic representation was not easily available.
+The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}.
+This $\delta$ parameter was constrained to be positive and added to the distribution of fluent key transitions $\beta$.
+It therefore represents the magnitude of the slowdown associated with hesitant transitions.
+The slowdown $\delta$ was allowed to vary by transition locations because hesitations at larger linguistic units are more likely to be associated with higher level planning which may delay output.
+
+The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant interkey intervals; the second line represents fluent interkey intervals.
 
@@
 (\#eq:bimodcon)
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \times \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \log\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \times \log\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
 \text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
@@
 
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant interkey intervals; the second line represents fluent interkey intervals.
-Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1.
-$\theta$ is here parameterised to represent the probability that an interkey interval is associated with the distribution of long interkey intervals, hence representing the hesitation probability.
-This probability is inversely related to the mixing weight of the distribution of short interkey intervals by $1-\theta$ because the weights of both distributions must sum to 1.
-In other words, a larger weight for one distribution inevitably means a lower weight for the other distribution.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assumed that pauses are likely to vary across linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assumed that some participants pause more at certain transition locations and other participants pause less.
-
-<!-- This is akin to what is known as a random by-participant slopes model.  -->
+These two distributions are associated with the mixing weight $\theta$ which must be larger than 0 and smaller than 1.
+$\theta$ is parameterised to represent the weighting of the distribution in the first line, hence representing the *hesitation probability*.
+This probability is inversely related to the mixing weight of the distribution of short interkey intervals by $1-\theta$ as the weights of both distributions must sum to 1.
+In line with the literature discussed in the introduction, we assume that the hesitation probability is likely to vary across linguistic locations.
+As hesitation frequency is subject to individual differences and writing style (and skills), we also assumed that some participants are more and others are less likely to hesitate at certain transition locations [@waes2019].
 
 Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location.
-In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing hesitations is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].
+We constrained the standard deviations so that $\sigma_{e'}$ which is associated with the distribution of hesitant interkey intervals is larger than the standard deviation associated with fluent transitions $\sigma_e$ for reasons discussed above.
 

### 2024-10-26 report/appendix/parallel.Rmd
After review update

# Interkey-interval predictions under a parallel theory of text production

Figure \ref{fig:model} shows how a parallel understanding of language production [following @van1991handwriting; @olive2014toward] predicts interkey intervals. It demonstrates a fundamental property, in which upstream processes provide downstream processes with information as it becomes available (i.e. on a just-in-time basis), and there is the possibility of buffering. For the purposes of illustration, all processing above the final buffer is lumped together as "message and language processing" and all processing below this buffer as "motor processing and execution". We make no claims as to exactly how these processes are divided.

\begin{figure}[!htb]
  \centering
  \includegraphics{figures/cascade_figure.pdf}
  \caption{Illustration of a parallel theory of text production.}
  \label{fig:model}
\end{figure}

In both cases the specific processing that is responsible for determining interkey interval is shown in black. Figure \ref{fig:model}A shows the case where upstream processing runs faster than, and therefore ahead of, motor execution. Message and language processing provides necessary information to the execution processes in advance of (or at exactly the same time as) when it can be used. In this case interkey intervals -- both IKI1 and IKI2 -- are determined just by the time taken to prepare and execute the keystroke. It is entirely unaffected by the duration of upstream processing. Figure \ref{fig:model}B shows the case where upstream processes deliver information to motor processes at a rate that lags behind that at which it could be potentially motor-planned and executed (i.e. where "just in time" processing fails). This results in a delay in output -- a "hesitation" in our terminology. IKI1 in Figure \ref{fig:model}B is therefore longer than IKI1 in Figure \ref{fig:model}A. Just-in-time processing is then resumed, and IKI2 in Figure \ref{fig:model}B is of the same duration as the interkey intervals in Figure \ref{fig:model}A.

In the case where upstream processes run ahead of motor processing, the interkey interval is obviously determined just by the speed at which the writer can move their finger (can motor-plan and execute). The essential point that Figure \ref{fig:model} illustrates is that in the case where upstream processes lag behind motor output (Figure \ref{fig:model}B), the interkey interval is determined _just_ by time taken for upstream processing. Effects on the interkey interval are not additive. Interkey intervals are therefore determined by one of two entirely separate data-generating processes. This is the basis for our claim that a finite mixture of two distributions represents a statistical model of parallel processing.

### 2024-10-26 report/appendix/postparam_constr.Rmd
After review update

@@
 datasets <- unique(ps$ds)
-size <- 2
+size <- 1.5
 width <- 0
@@
 beta_label <- "Fluent interkey\nintervals"
-delta_label <- "Slowdown for hesitations"
-theta_label <- "Probability of hesitations"
+delta_label <- "Hesitation duration"
+theta_label <- "Hesitation probability"
 
@@
 delta_plots <- map(datasets, ~filter(ps, 
-                                     name == "delta", 
+                                     name == "beta2", 
                                      ds == .x) %>% 
                      mutate(across(c(ds,group), ~factor(.)),
-                            across(c(mean, lower, upper), ~ . /1000),
+#                            across(c(mean, lower, upper), ~ . /1000),
                             name = delta_label) %>% 
@@
 plot_param <- map(1:6, ~beta_plots[[.]] + delta_plots[[.]] + theta_plots[[.]] + labs[[.]] +
-                    plot_layout(guides = "collect", widths = c(.45, 1, 1, .1)) &
+                    plot_layout(guides = "collect", widths = c(.4, 1, 1, .1)) &
                     theme(legend.position = "bottom"))
@@
 plot_fin_1 <- grid.arrange(patchworkGrob(plot_fin_1), 
-                           left = "Posterior estimate with 95% PIs",
+                           left = "Posterior estimates with 95% PIs",
                            bottom = "Transition location")
@@
 plot_fin_2 <- grid.arrange(patchworkGrob(plot_fin_2), 
-                           left = "Posterior estimate with 95% PIs",
+                           left = "Posterior estimates with 95% PIs",
                            bottom = "Transition location")
@@
 \caption{Distribution of posterior parameter estimates with 95\% probability intervals (PI).
-Fluent interkey intervals are shown in msecs and the slowdown for hesitant key intervals in secs; probability of hesitations is the proportion of hesitant interkey intervals.}
+Fluent interkey intervals and hesitation durations are shown in msecs; probability of hesitations is the proportion of hesitant interkey intervals.}
 \end{figure}
@@
 \caption{Distribution of posterior parameter estimates with 95\% probability intervals (PI).
-Fluent interkey intervals are shown in msecs and the slowdown for hesitant key intervals in secs; probability of hesitations is the proportion of hesitant interkey intervals.}
+Fluent interkey intervals and hesitation durations are shown in msecs; probability of hesitations is the proportion of hesitant interkey intervals.}
 

### 2024-10-26 report/appendix/prepost_constr.Rmd
After review update

--- 2024-04-27 report/appendix/prepost_constr.Rmd
@@
 cellmeans <- ps_plantra %>% 
+  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
   mutate(across(location, ~str_replace(., "\\n", " "))) %>% 
@@
           caption = "Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.",
+Cellmeans are shown for pre and post-test for the duration of hesitant interkey intervals in msecs and its probability.
+Differences are shown on the log scale for hesitation duration and on the logit scale for the hesitation probability; 95% PIs in brackets.",
           note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",

### 2024-10-26 report/appendix/shift_constr.Rmd
After review update

--- 2024-04-27 report/appendix/shift_constr.Rmd
@@
-# Key-combination effect (constrained mixture model)
+```{r}
+# get posterior
+file <- list.files(path = "../../stanout/cato", 
+                   pattern = "mogbetaco.+.csv$",
+                   full.names = T)
 
-The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets.
-Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
 
-<!-- Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.  -->
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
 
-Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
-We compared the SPL2 data including and excluding the keystroke after shift.
-Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
-The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.
+ps_cato <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+#         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(c(beta, prob, beta2)) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(task, group, location, name)) %>% 
+  mutate(group = str_c("Task: ", task, "; Group: ", group),
+         ds = "CATO") %>% 
+  select(-task)
+
+file <- list.files(path = "../../stanout/c2l1", 
+                   pattern = "mogbetacon.*.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_c2l1 <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta,by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+#         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(c(beta, prob, beta2)) %>% 
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(location, name)) %>% 
+  mutate(ds = "C2L1",
+         group = "1")
+
+file <- list.files(path = "../../stanout/gunnexp2", 
+                   pattern = "mogbetacon.*.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_gunn <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+#         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(c(beta, prob, beta2)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(location, xn, name)) %>% 
+  mutate(ds = "GE2") %>% 
+  rename(group = xn)
+
+file <- list.files(path = "../../stanout/lift", 
+                   pattern = "mogbetaco.+.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_lift <- ps %>% filter(param != "beta") %>% 
+  mutate(across(genre, ~ifelse(is.na(genre), "Arg", .))) %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+#         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(c(beta, prob, beta2)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(genre, topic, location, name)) %>% 
+  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre),
+         ds = "LIFT") %>% 
+  select(-topic, -genre)
+
+file <- list.files(path = "../../stanout/plantra", 
+                   pattern = "mogbetaco.+.csv$",
+                   full.names = T)
+
+ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value) %>% 
+  mutate(idx = 1:n())
+
+ps_plantra <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(beta2, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx)) %>% 
+  mutate(across(c(beta2, beta), exp),
+#         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(c(beta, prob, beta2)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper), .names = "{.fn}"),
+            .by = c(task, location, name)) %>% 
+  mutate(ds = "PLanTra") %>% 
+  rename(group = task)
+
+file1 <- list.files(path = "../../stanout/spl2", 
+                    pattern = "mogbetaco.+.csv$",
+                    full.names = T)
+
+file2 <- list.files(path = "../../stanout/spl2_shift", 
+                    pattern = "mogbetaco.+.csv$",
+                    full.names = T)
+
+ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
+                     mutate(data = .x)) %>% 
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
+         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
+                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
+  filter(param %in% c("beta", "delta", "prob"))
+
+ps_beta <- filter(ps, param == "beta") %>% 
+  select(beta = value, data) %>% 
+  mutate(idx = 1:n(), .by = data)
+
+ps_spl2_both <- ps %>% filter(param != "beta") %>% 
+  pivot_wider(names_from = param, values_from = value) %>% 
+  unnest(cols = c(delta, prob)) %>% 
+  mutate(idx = 1:n(), .by = where(is.character)) %>% 
+  left_join(ps_beta, by = join_by(idx, data)) %>% 
+  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
+         beta2 = beta + delta,
+         across(c(beta2, beta), ~exp(.)),
+#         delta = beta2 - beta,
+         across(location, str_replace, " ", "\n")) %>% 
+  pivot_longer(c(beta, prob, beta2)) %>%
+  summarise(across(value, list(mean = mean, 
+                               lower = lower, 
+                               upper = upper),
+                   .names = "{.fn}"),
+            .by = c(lang, location, name, data)) %>% 
+  mutate(ds = "SPL2",
+         group = str_c("Data: ", data, "; Language: ", lang)) %>% 
+  select(-lang)
+
+ps_spl2 <- ps_spl2_both %>% filter(data == "SPL2 (_^[shift])")
+
+# Combine all
+ps <- bind_rows(ps_cato, ps_c2l1, ps_gunn, ps_lift, ps_plantra, ps_spl2)
+```
+
+# Key-combination effect
+
+Interkey intervals at before sentence location either did (PLanTra, LIFT) scope over both the shift key and the character following the shift key or included only the first key following a the space immediately proceeding a sentence (CATO, C2L1, SPL2, GE2).
+In other words, the interkey interval before sentences are sum of two interkey intervals in the PLanTra and LIFT data, namely `_^[shift]^C`, where `^` represents a transition between keys, consists only of one interkey interval, namely `_^[shift]`, in the remaining datasets.
+Therefore, differences in hesitation patterns across datasets for transitions immediately proceeding a sentence could be explained without reference to edges of linguistic units.
+
+Therefore we tested to what extent the inclusion of one additional interkey interval at sentence boundaries affected the hesitation results.
+We used the SPL2 data for this comparison and calculated intervals including and excluding the keystroke following a sentence-initial shift press.
+Although we modelled all transition locations, we present only before-sentence transitions as there was, as one would expect, no difference for before and within-word interkey intervals.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans} (only for the data from participants writing in their first language).
 
@@
 cellmeans <- ps_spl2_both %>% 
+  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
   separate(group, into = c("group", "lang"), sep = "; ") %>%
-  select(-ds, -group) %>%
-  mutate(across(lang, ~str_remove(., "Language: ")),
+  filter(str_detect(lang, "L1")) %>% 
+  select(-ds, -group, -lang) %>%
+  mutate(#across(lang, ~str_remove(., "Language: ")),
          across(location, ~str_replace(., "\\n", " "))) %>% 
@@
   mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
-  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
-         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
+  filter(lang %in% c("EN", "overall")) %>% 
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
   pivot_wider(names_from = data, values_from = value) %>% 
@@
   left_join(cellmeans) %>% 
-  arrange(param, location, lang) %>% 
   filter(location %in% c("overall", "before sentence")) %>% 
-  select(param, 
-         Language = lang,
-         `_^[shift] + C` = ends_with(" + C)"),
+  mutate(across(param, ~recode_factor(., 
+                                `beta` = "Fluent interkey intervals",
+                                `delta` = "Hesitation duration",
+                                `prob` = "Hesitation probability",
+                                .ordered = T))) %>% 
+  select(`Model parameter` = param, 
+         `_^[shift]^C` = ends_with(" + C)"),
          `_^[shift]` = ends_with("FT])"),
@@
 
-apa_table(tmp[,-1],
-          align = c("l", "r", "r", "r", "r"), 
-          caption = "Mixture model estimates for key transitions immediately preceding a sentence.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95% PIs in brackets.",
+apa_table(tmp,
+          align = c("l", "l", "r", "r", "r", "r"), 
+          caption = "Mixture model estimates for interkey interval immediately preceding a sentence.
+Cellmeans are shown for interkey intervals that do and do not include the interkey interval following a sentence-initial shift key press.
+Hesitation duration is shown in mseecs along with the hesitation probability.
+The difference due to the additional keystoke is shown fo the hesitation duration (log scale) and hesitation probability (logit scale); 95% PIs are shown in brackets.",
           note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
           font_size = "footnotesize",
-          longtable = T,
-          stub_indents = list("Fluent transitions" = 1,
-                              "Hesitation duration" = 2:3,
-                              "Hesitation probability" = 4:5) )
+          longtable = T)
 
@@
 
-The fluent transition duration and hesitation duration were not affected by whether or not the sentence-initial transition include the character following shift for neither language.
-However, we found strong evidence for an increased hesitation probability -- in both -- languages when the before-sentence key transition included the character following the shift key.
-Notably, the hesitation probability increased ceiling.
-It is not entirely surprised that the constrained mixture model identifies essentially all before-sentence transition as pauses: the constrained model does not distinguish between transition locations for the duration of fluent transitions.
-Therefore essentially all before-sentence transitions are identified as hesitation as they always include two keystrokes while all other transition location (which is the majority of the data) do not.
+The duration of fluent and hesitant interkey intervals was not affected by whether or not the sentence-initial transition include the character following the shift key.
+However, we found strong evidence for an increased hesitation probability when the before-sentence interkey interval included the character following the shift key.
+Notably, the hesitation probability approached ceiling when the interkey interval following the shift key was added.
+In other words, the mixture model identified almost all before-sentence transitions as being hesitant as they -- in the majority of cases -- included two keystrokes while interkey intervals for all other transition locations -- and therefore the majority of the data -- did not.

### 2024-10-26 report/appendix/sim.Rmd
After review update

@@
 
-A possible concern with these results -- substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-To address this concern we repeated a comparison for a single distribution model and a two-distributions mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart.
-In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.
+A possible concern with our results -- the substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always outperform single-distribution models.
+We addressed this concern before by using cross-validation techniques for model comparison which prevent overfit by penalising models with more parameters.
+To further address this concern we evaluated models -- similar to the ones used in the main text -- for two sets of simulated data.
+Data were simulated using two random number generators, one that samples data from a single distribution and another that samples data from a weighted combination of two distributions.
+These simulated datasets allow us to test the predictive performance of our models in a context where we know the true underlying data generating process.
 
-In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equation \ref{eq:bimodcon}).
-This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
+The first dataset was simulated from a weighted mixture of two log-normal distributions similar to the process described above (equation \ref{eq:bimodcon}).
+This process and the corresponding Bayesian model that we used to parameter estimation is summarised in equation \ref{eq:simmog}.
 
@@
 (\#eq:simmog)
-\text{y} \sim\text{ } & \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma^2_2)\\
+\text{y} \sim\text{ } & \theta \times \log\mathcal{N}(\beta + \delta, \sigma^2_1) +\\
+& (1 - \theta) \times \log\mathcal{N}(\beta, \sigma^2_2)\\
 \text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
@@
 
-This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants).
-The model includes two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.
+The equivalent Bayesian model is largely identical to the model used in the main text but does not assume different parameters for transition locations and does not include random effects for participants.
+This is because we neither simulated data for different transition locations (or other factors) or repeated measures for participants.
+The model assumes two data generating processes with each assuming a log-normal distribution with a mixing proportion $\theta$.
+The distribution of smaller values has a mean $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of larger values is constrained to have a mean that is $\delta$ units larger than $\beta$ and has a standard deviation $\sigma^2_1$ larger than the distribution with the central tendency $\beta$.
 
-In R, data that follow the model in equation \ref{eq:simmog} can be simulate using a helper function called `molg` (mixture of log Gaussians).
+In R we can simulate data for equation \ref{eq:simmog} using the helper function below.
 
 ```{r eval = F, echo = T}
-# Function for a mixture of log Gaussians
+# Function for a mixture of two log-Gaussians
 molg <- function(n, theta, mu1, mu2, sigma1, sigma2) {
@@
 
-N <- 1000 # number of subjects
-beta <- 5 # mean of fluent key intervals (in log)
-delta <- 1 # slowdown for hesitant interkey intervals (in log)
-theta <- .35 # propotion of hesitations
-sigma <- c(.25, .5) # trial-by-trial error
+N <- 1000 # number of participants
+beta <- 5 # mean of fluent interkey intervals (in log)
+delta <- 1 # increment for hesitant interkey intervals (in log)
+theta <- .35 # proportion of hesitations
+sigma <- c(.25, .5) # error variance
 
-# Generate data from a mixture of two log-Gaussians
+# Simulate data from a mixture of two log-Gaussians
 y <- molg(n = N, 
@@
 
-The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
+A second dataset was simulated coming from a single log-normal distribution following equation \ref{eq:simuv}. An equivalent Bayesian model was implemented for parameter estimation.
 
@@
 (\#eq:simuv)
-\text{y} \sim\text{ }& \text{log}\mathcal{N}(\beta, \sigma^2)\\
+\text{y} \sim\text{ }& \log\mathcal{N}(\beta, \sigma^2)\\
 \text{constraint: } & \sigma^2>0
@@
 
-Data following the distirbution in equation \ref{eq:simuv} can be simulated using a few lines of R code:
+Data following the distribution in equation \ref{eq:simuv} can be simulated using the following R code:
 
 ```{r echo = T, eval=F}
-N <- 1000 # number of subjects
-beta <- 5 # Population mean
-sigma <- .25 # trial-by-trial error
+N <- 1000 # number of participants
+beta <- 5 # population mean
+sigma <- .25 # error variance
 
-# Generate data from a log normal distribution
+# Simulate data from a single log-normal distribution
 y <- rlnorm(n = N, mean = beta, sd = sigma)
@@
 
-Again, this model is a simplified version of the single distribution models used in the main analysis.
-The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.
+Again, the Bayesian model corresponding to this process is a simplified version of the single-distribution model used in the main text.
+The model assumes a log-Gaussian distribution with a mean $\beta$ and a standard deviation $\sigma^2$.
 
-The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
-The simulated data are visualised in Figure \ref{fig:simdata}.
-Parameter values were chosen so that the simulated data are roughly similarly distributed to keystroke transitions.
+Both simulated datasets are visualised in Figure \ref{fig:simdata}.
+The parameter values used for each of the two data simulations can be seen in Table \ref{tab:simparam}.
+Parameter values were chosen so that the simulated data are distributed roughly similar to interkey interval data.
 
-```{r simdata, fig.cap= "Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process.
-The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
+```{r simdata, fig.cap= "Data simulated with a two-distribution (yellow) and a single-distribution (grey) random data-generating process.
+The x-axis showing the outcome y was log-scaled for visibility.", fig.height=3.5, fig.width=5}
 files <- list.files("../../mixture-model-sim/data", full.names = T)
@@
 
-For each of these two data sets we simulated 1,000 observations.
-We run 2 models -- a two-distribution mixture model and a single-distribution model -- each for both data sets.
-Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
+For each of these two datasets we simulated 1,000 observations.
+We fitted 2 models -- the two-distribution mixture model and the single-distribution model described above -- for each dataset.
+Models were run with 3 chains, with each 6,000 iterations of which 3,000 iterations were discarded as warmup samples.
 Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}.
-The parameters are shown by type of data generating process along with the true parameter values.
-Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they single-distribution model.
+True and estimated parameter values are shown for each data generating process and Bayesian model.
+Both models successfully uncovered the model parameters for the datasets simulated with the corresponding underlying process, but less so when the model was applied to the data generated with the other incorrect underlying process.
 
@@
 
-names(table) <- c("Model", "Parameter with true value", "Two-distributions model", "Single-distributions model")
+names(table) <- c("Model", "Parameter with true value", "Two-distribution data", "Single-distribution data")
 
@@
           escape = FALSE,
-      col_spanners = list("Estimate with 95\\% PI" = 2:3),
-      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
+      col_spanners = list("Estimates with 95\\% PIs" = 2:3),
+      caption = "Uncovered parameter estimates with 95\\% probability intervals (PI) and true parameter values for each simulated dataset by model.",
       align = c("l", rep("r", 2)),
-      stub_indents = list(`Two-distributions data` = 1:5, 
-                          `Single-distribution data` = 6:7))
+      stub_indents = list(`Two-distribution model` = 1:5, 
+                          `Single-distribution model` = 6:7))
 
@@
 
-We used LOO-CV to compare the predictive performance of the two models for each data generating process.
-The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
-The results rule out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a single-distribution process.
-However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
+We used leave-one-out cross-validation to compare the predictive performance of the two models for each data-generating process.
+Model comparisons can be found in Table \ref{tab:loossim}.
+For the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
 In fact, the mixture model's predictive performance is 11.6 standard errors higher compared to the single-distribution model.
-In comparison, for the single-distribution data, the two-distributions mixture model was negligibly different from the single-distribution model with a difference of 0.77 standard errors.
-Thus, the single distribution is more parsimonious.
+However, the mixture model showed a slightly lower predictive performance (i.e.
+a difference of 0.77 standard errors) for the single-process data.
+Thus, the single distribution model is the more parsimonious choice for the single-distribution data.
+These results rule out the possibility that mixture models always lead to higher predictive performance.
 
@@
           escape = FALSE,
-      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
+      caption ="Model comparisons by dataset. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
      note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",

### 2024-10-26 report/appendix/transloc_constr.Rmd
After review update

@@
 files <- list.files(str_c("../../stanout/", 
-                          c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetacon.+.csv", full.names = T) # "spl2_shift",
+                          c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), 
+                    pattern = "mogbetacon.+.csv", full.names = T) # "spl2_shift",
 
@@
 tmp <- ps_loc_diffs %>% 
-  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
+  pivot_longer(starts_with("diff"), 
+               names_to = c("diffid", ".value"),
+               names_sep = "_") %>% 
   mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
@@
           #                              `\\textbf{SPL2 (L1; shift + C)}` = 13:14),
-          col_spanners = list(" " = c(1),
-                          #    "Fluent transitions" = c(2, 3), 
-                              "Hesitation slowdown" = c(2, 3), 
-                              "Hesitation probability" = c(4, 5)),
+          col_spanners = list("Hesitation duration" = 2:3, 
+                              "Hesitation probability" = 4:5),
           note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
-          caption = "Effect of transition location on interkey intervals. Differences are shown on log scale (for durations) and logit scale for hesitation probability. 95\\% PIs in brackets.")
+          caption = "Transition-location effect. Differences are shown for the hesitation duration (log scale) and the hesitation probability (logit scale); 95\\% PIs in brackets.")
 

### 2024-10-26 report/manuscript.Rmd
After review update

@@
 ---
-title             : "Typing in tandem: finite mixture models reveal fundamental parallelism in multi-sentence text composition"
-shorttitle        : "Typing in tandem"
+title             : "Typing in tandem: language planning in multi-sentence text production is fundamentally parallel"
+#shorttitle        : "Typing in tandem"
+shorttitle        : "Revision of XGE-2024-2060 as invited by the action editor, Michele Diaz, Ph.D."
 
@@
     email         : "jens.roeser@ntu.ac.uk"
+    role          : 
+      - Conceptualisation
+      - Data curation
+      - Formal analysis
+      - Funding acquisition
+      - Investigation
+      - Methodology
+      - Project administration
+      - Software
+      - Visualisation    
+      - Writing -- original draft
+      - Writing -- review \& editing
   - name          : "Rianne Conijn"
     affiliation   : "2"
+    role          :
+      - Conceptualisation
+      - Investigation
+      - Writing -- original draft
+      - Writing -- review \& editing
   - name          : "Evgeny Chukharev"
     affiliation   : "3"
+    role          :
+      - Funding acquisition
+      - Investigation
+      - Resources  
+      - Project administration
+      - Software
+      - Supervision
   - name          : "Gunn Helen Ofstad"
     affiliation   : "4"
+    role          :
+      - Investigation
+      - Resources
+
   - name          : "Mark Torrance"
     affiliation   : "1"
+    role          :
+      - Conceptualisation
+      - Data curation
+      - Formal analysis
+      - Funding acquisition
+      - Investigation
+      - Methodology
+      - Resources  
+      - Supervision
+      - Visualisation    
+      - Writing -- original draft
+      - Writing -- review \& editing
 
@@
 
-# We are grateful for all authors who made their available (to us), in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues.
+authornote: |
+  We are grateful for all authors who made their data available to us, in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues, and Vibeke Rønneberg, Per Henning Uppstad, and colleagues at the Norwegian Reading National Centre, University of Stavanger.
 This work was supported by the National Science Foundation under Grant No.
-2016868: "ProWrite: Biometric feedback for improving college students' writing processes." and UKRI ESRC under Grant No.
+2016868: "ProWrite: Biometric feedback for improving college students’ writing processes." and UKRI ESRC under Grant No.
 ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production difficulties shape text formation."
-# 
-# Data and analysis scripts written in R and Stan can be found on OSF (https://osf.io/z65dw/; Roeser et al., 2024).
+
+  Data and analysis scripts written in R and Stan can be found on OSF (https://osf.io/z65dw/; Roeser et al., 2024).
 For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see https://rpubs.com/jensroes/mixture-models-tutorial.
+
+  Parts of this work have been shared through conference presentations, namely at the SIG writing conference in Paris, France (2024) and in Umea, Sweden (2022), and at EARLI SIG Online Measurements in Southampton, UK (2022).
 
@@
 
-keywords: "Cascade; parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"
+keywords: "Parallel processing; writing; mixture models; language production; keystroke logging"
 
 bibliography      : ["references.bib"]
+nocite: | 
+  @brysbaert2021improving, @dickey1970weighted, @wagenmakers2010bayesian, @jeffreys1961theory, @lee2014bayesian, @dienes2014using
 
@@
 
-beta_label <- "Fluent interkey\nintervals (in msecs)"
-delta_label <- "Slowdown for hesitant\ninterkey intervals (in secs)"
+beta_label <- "Fluent interkey\nintervals"
+delta_label <- "Hesitant\ninterkey intervals"
 theta_label <- "Probability of hesitant\ninterkey intervals"
@@
 
+```{r child = 'appendix/parallel.Rmd'}
+```
+
+\newpage
+
 ```{r child = 'appendix/models.Rmd'}
@@
 ```{r child = 'appendix/fit2data.Rmd'}
-```
-
-\newpage
-
-```{r child = 'appendix/transloc.Rmd', eval=F}
-```
-
-\newpage
-
-```{r child = 'appendix/shift.Rmd', eval = F}
 ```

### 2024-10-26 report/manuscript.tex
After review update

@@
 \usepackage{xcolor}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx}
@@
 {}{\PatchFailed}
-\keywords{Cascading parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes}
+\keywords{Parallel processing; writing; mixture models; language production; keystroke logging}
 \usepackage{csquotes}
@@
 \DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
+\makeatletter
+\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-1em}%
+  {\normalfont\normalsize\bfseries\typesectitle}}
+
+\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-\z@\relax}%
+  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
+\makeatother
+
 \ifLuaTeX
@@
 \hypersetup{
-  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
-  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev3},
+  pdftitle={Typing in tandem: language planning in multi-sentence text production is fundamentally parallel},
+  pdfauthor={Jens Roeser1, Rianne Conijn2, Evgeny Chukharev3, Gunn Helen Ofstad4, \& Mark Torrance1},
   pdflang={en-EN},
-  pdfkeywords={Cascading parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes},
+  pdfkeywords={Parallel processing; writing; mixture models; language production; keystroke logging},
   hidelinks,
@@
 
-\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
-\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev\textsuperscript{3}}
+\title{Typing in tandem: language planning in multi-sentence text production is fundamentally parallel}
+\author{Jens Roeser\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, Evgeny Chukharev\textsuperscript{3}, Gunn Helen Ofstad\textsuperscript{4}, \& Mark Torrance\textsuperscript{1}}
 \date{}
 
-\shorttitle{Modelling writing hesitations}
+\shorttitle{Typing in tandem}
 
@@
 
+We are grateful for all authors who made their data available to us, in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues, and Vibeke Rønneberg, Per Henning Uppstad, and colleagues at the Norwegian Reading National Centre, University of Stavanger .
+This work was supported by the National Science Foundation under Grant No.~2016868: \enquote{ProWrite: Biometric feedback for improving college students' writing processes.} and UKRI ESRC under Grant No.~ES/W011832/1: \enquote{Can you use it in a sentence?: Establishing how word-production difficulties shape text formation.}
+
+Data and analysis scripts written in R and Stan can be found on OSF (\url{https://osf.io/z65dw/}; Roeser et al., 2024).
+For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see \url{https://rpubs.com/jensroes/mixture-models-tutorial}.
+
+Parts of this work have been shared through conference presentations, namely at the SIG writing conference in Paris, France (2024) and in Umea, Sweden (2022), and at EARLI SIG Online Measurements in Southampton, UK (2022).
+
+The authors made the following contributions.
+Jens Roeser: Conceptualisation, Data curation, Formal analysis, Funding acquisition, Investigation, Methodology, Project administration, Software, Visualisation, Writing -- original draft, Writing -- review \& editing; Rianne Conijn: Conceptualisation, Investigation, Writing -- original draft, Writing -- review \& editing; Evgeny Chukharev: Funding acquisition, Investigation, Resources, Project administration, Software, Supervision; Gunn Helen Ofstad: Investigation, Resources; Mark Torrance: Conceptualisation, Data curation, Formal analysis, Funding acquisition, Investigation, Methodology, Resources, Supervision, Visualisation, Writing -- original draft, Writing -- review \& editing.
+
 Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Human-Technology Interaction Group, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa\\\textsuperscript{4} Educational Sciences and Humanities, University of Stavanger, Norway}
 
 \abstract{%
-Classical serial models view the process of producing a text as a chain of discrete pauses and writing bursts.
-In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel with execution.
-These two theories make different assumptions about the underlying process that generates inter-keystroke intervals.
-Across six datasets comprising multi-sentence text composition we instantiated these assumptions as two sets of Bayesian mixed-effects models: (1) single-distribution models as assumed by the serial processing account, and (2) finite two-distribution mixture models as assumed by the parallel processing account.
-We analysed keystroke intervals at before-sentence, before word, and within word transitions.
-Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view across all datasets.
+Classical serial models view the process of producing a text as a chain of discrete pauses, during which the next span of text is planned, and bursts of activity during which this text is output onto the page or computer screen.
+In contrast, semi-parallel models of writing assume that planning is not complete at production onset and operates in parallel with execution.
+Across six datasets from participants composing multi-sentence texts we instantiated these assumptions in two sets of Bayesian mixed effects models: (1) single-distribution models as assumed by the serial processing account, and (2) a finite (two-distribution) mixture model as assumed by the parallel processing account.
+We analysed interkey intervals at before-sentence, before word, and within word transitions.
+Model comparisons demonstrated strong evidence in favour of the parallel view across all datasets.
 When pausing occurred, sentence initial pauses were longer than word initial pauses which is consistent with the idea that larger linguistic edges are associated with higher level planning.
-However, we found -- across populations -- that keystroke latencies at word and even at sentence boundaries were often too rapid to permit planning of what was written next.
-Our results cannot be explained by the serial but are in line with the cascading, semi-parallel view of writing.
+However, we found -- across populations -- that interkey intervals at word and even at sentence boundaries were often too rapid to permit planning of what was written next.
+Our results cannot be explained by the serial processing but are in line with the semi-parallel view of writing.
 }
@@
 
-Translating ideas into language involves a sequence of processes starting with communicative intent, proceeding through generation of a message, retrieving appropriate lexical items, deciding with which part of the message to start the utterance which, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), is determined by syntactic requirements of our language, then retrieving orthographic and / or phonological codes, and that are finally submitted to the motor programs that output speech, key or pen-strokes or signs (Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
+Translating ideas into language involves a sequence of processes starting with communicative intent, proceeding through generation of a message, retrieving appropriate lexical items, deciding with which part of the message to start the utterance (which, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), is determined by syntactic requirements of our language) and then retrieving orthographic and / or phonological codes, these are finally submitted to the motor programs that output speech, key or pen-strokes or signs (Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).
 
 The coordination of these processes in writing is arguably more demanding than in spoken language.
-This is because writing requires spelling retrieval which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen are later-developing and more effortful than those required to articulate language sounds in speech leading to a reduced output speed compared to speech (see for a recent review Kandel, 2023).
-Consequently when writing, more than in speech, we have to mentally buffer linguistic plans before we can express them in language.
-Written communication is also less time constrained: speaking underlies fluency requirements, because hesitating during speech has a communicative effect (e.g.
+This is because writing requires spelling retrieval, which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen.
+Pen control and typing skills are later-developing and more effortful than those required to articulate language sounds in speech and, possibly as a consequence, written output rate is typically slower (for a recent review see Kandel, 2023).
+Consequently mentally buffering of linguistic plans prior to output is likely to play a greater role in writing than in speech.
+
+Written communication is also less time constrained: speaking comes with fluency requirements because hesitating during speech has a communicative effect (e.g.
 Clark \& Fox Tree, 2002).
 There is therefore stronger pressure to plan ahead in speech than in writing (Roeser et al., 2019; Torrance \& Nottbusch, 2012).
-Indeed writers can pause when they wish -- even in the middle of a word -- and for how long they want (within reason) without compromising communication.
-Written text production -- for example writing argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
+Indeed writers can pause when they wish -- even in the middle of a word -- and for how long they want without compromising communication.
+Written production of multi-sentence texts -- composing an argumentative essay, for example -- therefore involves a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).
 
 The study reported in this paper compares statistical models that instantiate two different theories of how cognitive planning processes are coordinated in writing.
-Through the analysis of 6 different datasets from writers of different ages and competencies composing multi-sentence texts, we compare these statistical models and therefore the cognitive model of how keystrokes in writing are being generated they are mapping on.
+Through the analysis of 6 different datasets from writers of different ages and competencies composing multi-sentence texts, we compare these statistical models and therefore the cognitive models of text production onto which they map.
 Also, we evaluate competing predictions of these models by revisiting the often-reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006).
@@
 We then introduce two possible cognitive frameworks in which these can be interpreted.
-Finally we argue that analysis of pause-patterns at linguistic edges are important to permit differentiation between these two theories.
-
-We focus on data form spontaneous text production (see Gernsbacher \& Givón, 1995) for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production and have, therefore, real-world relevance (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
-Second, since the emergence of keystroke-logging software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; for a recent special issue see Conijn and Torrance (2023).\footnote{Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.} These data are therefore important for contemporary theories of writing and freely available for statistical modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
-Third, spontaneous text production -- unlike picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Meulemans et al., 2022; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021) -- content generation is not artificially constrained to individual sentence units (or less) by one or more images and linguistic form is not constrained to a sequence of words, respectively.
-Instead text composition provides data from both inter and intra-sentence transitions.\footnote{Also, pauses in text production do not reflect visual encoding of the stimulus (although possibly reflect looking back during writing which we will return to in the discussion).}
+Finally we argue that analysis of pause-patterns at linguistic edges is necessary to permit differentiation between these two theories.
+
+We focus on data form spontaneous text production (see Gernsbacher \& Givón, 1995) for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).
+Second, since the emergence of keystroke-logging software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature; for a recent special issue see Conijn and Torrance (2023).\footnote{Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.} These data are therefore important for contemporary theories of writing and freely available for statistical modelling and machine learning (Conijn et al., 2019; Conijn, 2020).
+Third, spontaneous text production is not constrained in the same way as it would be in experiments in which text is elicited in response to picture stimuli (Damian \& Stadthagen-Gonzalez, 2009; Meulemans et al., 2022; Nottbusch, 2010; Roeser et al., 2019) or is copied (Roeser, De Maeyer, et al., 2024; Van Waes et al., 2019, 2021).
+Instead the content-determination processes are free to run entirely internally, without the need to consult an image or a text that must be copied.\footnote{Also, pauses in text production do not reflect visual encoding of the stimulus (although possibly reflect looking back during writing which we will return to in the discussion).}
 
 Time intervals between adjacent keystrokes during typing vary as a function of various factors.
-Explaining the factors that influence the duration of inter-keystroke intervals (IKIs) requires a theory of how the various processes that transform intent into keypresses are structured.
-Processing -- planning what to say and how to say it -- may be serial.
+Explaining the factors that influence the duration of inter-keystroke intervals (interkey intervals) requires a theory of how the various processes that transform intent into keypresses are structured.
+Planning what to say and how to say it may be serial.
 This means that the processing sequence operates one unit at a time.
-Processing of the next language unit can only start when the previous unit was fully processed from communicative intention to completion of the written output.
-Therefore longer IKIs represent periods when the writer is planning what to write next.
-The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
-This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).
-
-This serial account is consistent with the frequent finding that average IKI duration immediately before larger linguistic units compared to smaller linguistic units.
+Under this theory, planning of the next language unit can only start when the previous unit has been fully processed, from communicative intention to completion of the written output.
+Therefore interkey intervals represent periods when the writer is planning what to write next: interkey interval will be longer at linguistic edges where planning scopes over larger text spans and their duration represents the time necessary to complete that planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002).
+
+This serial account is consistent with the frequent finding that the average interkey interval duration immediately before larger linguistic units is longer compared to smaller linguistic units.
 For example, in spontaneous text the mean pause duration before a writer starts a sentence has typically been reported to be longer than before a mid-sentence word and these are, in turn, longer than between mid-word key presses (Deane \& Zhang, 2015; Mohsen, 2021; Spelman Miller, 2006; Spelman Miller et al., 2008; Torrance, Rønneberg, et al., 2016; Wengelin, 2002; Xu \& Qi, 2017).
 This is understood as being due to time needed to plan message and syntax for the next sentence, which is added to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke of that word (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019).
-However this account is not consistent with what we know about writing.
-Consider these two observations:
+However this account is not consistent with two other sets for findings:
 
@@
 Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of a real-time process.
-For speech it is well known that the pre-planning duration depends -- among other factors -- on the need to avoid intra-sentence hesitation that are needed to plan upcoming words in parallel to production (e.g.
+For speech it is well known that the pre-planning duration depends -- among other factors -- on the need to avoid intra-sentential hesitations; speakers have a preference towards minimising language pre-planning depending on the extent to which upcoming words can be planned in parallel to production (Griffin, 2003; e.g.
 Levelt \& Meyer, 2000; Wheeldon \& Konopka, 2023).
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech {[}Nottbusch (2010); dam09; Roeser et al.
-(2019){]}.
+For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019).
 For example, in Roeser et al.
@@
 
-Second, even though the average sentence-initial IKI tends to be longer, they still tend to be very rapid.
-For example, Medimorec and Risko (2017) found that, in undergraduate students' writing on familiar topics 71\% of sentence-initial IKIs were less than 1 seconds in duration.
+Second, even though the average sentence-initial interkey interval tends to be longer, they still tend to be very rapid.
+For example, Medimorec and Risko (2017) found that, in undergraduate students' writing on familiar topics 71\% of sentence-initial interkey intervals were less than 1 seconds in duration.
 Also Rønneberg et al.
@@
 Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent.
-The fact that these very short sentence initial pauses exist, and in fact are quite common, is not consistent with the assumption that planning of the upcoming sentence is started and finished in the period between completion of the previous sentence and starting to write the next.
-
-If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations.
-The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that operate partly in parallel and largely without executive control.
+The fact that these very short sentence-initial pauses exist, and in fact are quite common, is not consistent with the assumption that planning of the upcoming sentence is started and finished in the period between completion of the previous sentence and starting to write the next.
+
+If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers composting multi-sentence texts typically occurs remarkably fluently, with very few hesitations.
+The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that operates partly in parallel and largely without executive control.
 Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991).
-Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
-Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes.
-In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing the output of the previous sentence.
-
-This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals.
-IKIs result from one of two underlying mental processes.
-When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then IKIs are determined just by the time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
-However if one or more upstream processes provide output more slowly, then IKIs become dependent directly on upstream processing times and not on finger movement.
-IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).
-The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not.
+Van Galen (1991), for example, argued for a cascade of modular processes each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016).
+Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes buffers allow fingers to catch up with message and language processing.
+Under this account when writers move from one sentence to another without hesitation this is due to then having completed required planning for the next sentence in parallel with output of the the sentence that they have just completed.
+
+This semi-parallel cascaded account of the composition process therefore gives a rather different understanding of inter-keystroke intervals.
+Interkey intervals result from one of two sets of underlying mental processes.
+When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then interkey intervals are determined just by the time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade).
+However if one or more upstream processes provide output more slowly, then interkey intervals are determined just by time taken to complete upstream processes and not on finger movement.
+These two distinct data-generating processes mean that interkey intervals are a mixture of two distributions, one associated with the very rapid processing necessary for actioning keystrokes, and another that forms as a result of delays resulting from some combination of processing at the semantic, syntactic, lexical, or orthographic level not being complete at the time of the immediately preceding keystroke (Roeser, De Maeyer, et al., 2024).
+The parallel account provides an explanation for why writers sometimes hesitate before sentence (or word) onset, but often do not.
 This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.
 
-In language comprehension, Vasishth and colleagues (Vasishth, Jäger, et al., 2017; Vasishth, Chopin, et al., 2017) demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental.
-Similar for keystroke data Roeser et al.
-(2021) demonstrated -- using a similar approach as used in the present paper\footnote{For similar approaches see Almond et al.
+Capturing this understanding of interkey intervals requires statistical models that are premised on more than one data-generating process.
+These models have been adopted in a number of psycholinguistic studies.
+For example, in language comprehension, Vasishth and colleagues (Vasishth, Jäger, et al., 2017; Vasishth, Chopin, et al., 2017) demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implementation.
+In the context of written production Roeser, De Maeyer, et al.
+(2024) demonstrated\footnote{For similar approaches see Almond et al.
 (2012), Baaijen et al.
@@
 (2022), Li (2021), and Van Waes et al.
-(2021).} -- how models that assume this mixture of two underlying data generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g.~\enquote{dkdkdkdk}).
-As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition.
+(2021).} how models that allow the possibility of two underlying data-generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g.~\enquote{dkdkdkdk}).
+As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition: copy-typing provides a typing context that is constrained to visual encoding of the target string and the retrieval of motor codes for letter combinations.
+For relatively experienced writers, visual encoding is likely to happen in parallel to writing (De Smet et al., 2018); hesitations occur when motor codes, for example for low frequency bigrams, were not retrieved in time (Roeser, De Maeyer, et al., 2024).
 However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes.
@@
 
-In the study that is presented in this paper we test using Bayesian hierarchical models of keystroke intervals -- similar to Roeser et al.
-(2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts.
+\hypertarget{present-study}{%
+\subsection{Present Study}\label{present-study}}
+
+In the study that is presented in this paper we use Bayesian hierarchical models of keystroke intervals -- similar to Roeser, De Maeyer, et al.
+(2024) -- to identify whether the semi-parallel view of writing generalises to contexts in which participants compose texts.
 On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).
@@
 The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location.
-In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word locations compared to within-word locations.
+In other words, we expect interkey intervals to be proportionally longer at before-sentence locations compared to before-word locations compared to within-word locations.
 This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation.
-The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not (always) pause to plan their text but plan in parallel to writing.
+The semi-parallel cascading view, in contrast, assumes that although interkey intervals at larger linguistic edges might be longer, it is in principle possible that writers do not (always) pause to plan their text but plan in parallel to writing.
 In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production.
 Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occur to some extent probabilistically across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability.
-We hypothesise that the statistical model of the parallel cascading view provides a better out-of-sample generalisation than the implementation of the serial model of writing in the context of unconstrained spontaneous text composition.
+We hypothesise that the statistical model of the parallel view provides a better out-of-sample generalisation than the implementation of the serial model of writing in the context of unconstrained spontaneous text composition.
+We tested these models across 6 datasets.
+
+For each dataset we evaluated four Bayesian statistical models.
+The first three models were premised on the (traditional) single-distribution understanding of interkey intervals.
+These models are consistent with the assumption that processing leading to a keystroke is fundamentally serial.
+In the first model we assumed the interkey intervals were drawn from a Gaussian (normal)distribution.
+In the second we assumed a log-Gaussian distribution.
+In both cases, consistent with previous studies that have modelled inter-keystroke intervals, we assumed equal variance across different linguistic edges (within words, at word boundaries, and at sentence boundaries).
+To provide a fairer comparison with the two-distribution model that followed we tested a third unequal variance model that allowed intervals associated with higher-order linguistic edges to have greater variability.
+
+The fourth model was premised on the assumption that the processing that results in a keystroke runs, in part, in parallel with preceding output.
+As we have discussed, this model assumes two data-generating processes.
+We therefore followed the finite (two-distribution) mixture-modelling approach described by Roeser, De Maeyer, et al.
+(2024).
+This model assumes that interkey intervals result from a mixture of: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels that cause delays in this information flow.
+Therefore interkey intervals are not directly and sufficiently determined by transition location -- as in the serial account -- but are also associated with different probabilities that a hesitation occurs.
 
@@
 
-In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts.
-Full modelling details can be found in Appendix \textbackslash ref\{statistical-models; we summarise the most important model properties in the following.
-The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data.
-In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.
-
-The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in Roeser et al.
-(2021).
-Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes.
-These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs.
-This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.
-
-We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts.
-We discussed advantages with this approach in the introduction section.
-These datasets include samples from populations with various writing experience and languages (e.g.~young / L2 writers, students) performing different writing tasks (e.g.~essays, syntheses).
+We analysed interkey intervals from six existing studies in which participants composed spontaneous, multi-sentence texts in response to writing prompts.
+The datasets include samples from populations with various writing experience and languages (e.g.~young / L2 writers, students) performing different writing tasks (e.g.~essays, syntheses).
 Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.
+
+This study was not preregistered.
 
@@
 
-TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2.
-For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say \enquote{unpublished}.
-I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk).
-Was the GUNNEXP2 sample Norwegian?
-Should we include information like time limit or text length in the table?
-
 Six datasets with keystroke data from free text production were used for analysis.
 An overview can be found in Table \ref{tab:datasets}; descriptions below.
-Dataset used are C2l1 (Rønneberg et al., 2022), CATO (Torrance, Rønneberg, et al., 2016), GUNNEXP2 (Torrance \& Ofstad, n.d.), SPL2 (Torrance et al., n.d.), LIFT published in Vandermeulen, Steendam, et al.
+Datasets used are C2l1 (Rønneberg et al., 2022), CATO (Torrance, Rønneberg, et al., 2016), GE2 (Ofstad Oxborough \& Torrance, 2011), SPL2 (Torrance et al., n.d.), LIFT published in Vandermeulen, Steendam, et al.
 (2020) and described in Vandermeulen, De Maeyer, et al.
-(2020), PLanTra data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
-In C2L1 Norwegian 6th graders composed argumentative essays.
-In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text.
-In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced.
-The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each.
-The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability.
-The GUNNEXP2 data \ldots.
-{[}MARK, please add a sentence here to describe the production context{]}.
-
-For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
-LIFT and PLanTra data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021) and SPL2 data were collected using CyWrite (Chukharev-Hudilainen et al., 2019).
+(2020), and PLanTra published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a).
 
@@
 
-\begin{tabular}{lp{3cm}p{2cm}p{2.5cm}rrp{2.5cm}p{2.25cm}p{2.5cm}}
+\begin{tabular}{p{1.5cm}p{2.75cm}p{1.75cm}p{2.25cm}p{1.75cm}p{1.75cm}p{.5cm}p{2.5cm}p{1.75cm}p{2cm}}
 \toprule
-Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Sample} & \multicolumn{1}{c}{Country} & \multicolumn{1}{c}{Language}\\
+Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{Words in text$^a$} & \multicolumn{1}{c}{Age} & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Sample} & \multicolumn{1}{c}{Country} & \multicolumn{1}{c}{Language}\\
 \midrule
-C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.8 & 6th graders & Norway & Norwegian\\
-CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.9 & Secondary school students & Norway & Norwegian\\
-GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite & ????? & 45 & ???? & Undergraduate students & ???? Norway & ???? Norwegian\\
-LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 17 & Secondary school students & The Netherlands & Dutch\\
-PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23 & Master students & Belgium & English (L2)\\
-SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.6 & Undergraduate students & USA & English (L1) / Spanish (L2)\\
+C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essay & 128.5 (52.9) & 11.8 (0.47) & 126 & 6th graders & Norway & Norwegian\\
+CATO & Torrance et al. (2016) & EyeWrite & Expository essay & 288 (119) & 16.9 (0.88) & 52 & Secondary school students (dyslexic, non dyslexic) & Norway & Norwegian\\
+GE2 & Torrance and Ofstad (n.d) & EyeWrite & Argumentative essay & 540 (141) & 19.1 (1.4)$^c$ & 45 & Undergraduate students & UK & English\\
+LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis of several printed sources$^b$ & 374 (56)$^d$ & 16.7 (1.5) & 658 & Secondary school students & The Netherlands & Dutch\\
+PLanTra & Rossetti and Van Waes (2022b) & InputLog & Paraphrase technical text$^b$ & 295 (79) & 23 (2.7) & 47 & Master students & Belgium & English (L2)\\
+SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essay & 668 (301) & 20.6 (1.62) & 39 & Undergraduate students & USA & English\\
 \bottomrule
+\addlinespace
 \end{tabular}
@@
 }
+
+\begin{tablenotes}[para]
+\normalsize{\textit{Note.} $^a$Mean number of words in the final text; standard deviations in parentheses.
+$^b$Reading activity was marked in the dataset and removed before analysis.
+$^c$Estimated from a similar sample from the same population.
+$^d$Estimated based on the number of characters in the final product following Brysbaert et al.
+(2021).}
+\end{tablenotes}
 
@@
 
+In C2L1 Norwegian 6th graders composed argumentative essays.
+In CATO upper Norwegian secondary students with and without dyslexia composed expository texts either normally or with masked letters to prevent writers reading their unfolding text.
+In SPL2 undergraduate students produced argumentative essays in their first language (L1; English) and in a second language L2 (Spanish) in which they were able to compose text but in which their mastery fell well behind their first language.
+Order of language (L1 / L2) and two writing prompts were counterbalanced.
+The LIFT data are from pre-university students producing argumentative and informative text syntheses on four topics each.
+The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability, before and after receiving either online instruction on how to apply plain language principles to sustainability content or an online instruction exclusively on the topic of sustainability.
+The GE2 data are from Undergraduate students producing two argumentative essays on general-knowledge topics in either masked text or normal writing conditions, with masking, topic and order counterbalanced.
+
+For C2L1, CATO, and GE2, keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).
+LIFT and PLanTra data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021) and SPL2 data were collected using CyWrite (Chukharev-Hudilainen et al., 2019).
+
 \hypertarget{data-extraction}{%
@@
 
-From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
-In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions.
-At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section.
-Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
-Also we removed transitions that were followed by an editing operation.
+From the keystroke data we extracted interkey intervals between adjacent keys at locations that have been identified repeatedly in previous research as locations where observed mean interkey intervals (or counts of interkey intervals over a predetermined threshold) have tended to vary substantially (e.g.
+Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}.
+In particular we analysed the interkey intervals that resulted in the insertion of a character that started a new sentence (before-sentence transition); interkey intervals that started a new word (other than those at the beginning of a sentence; before-word transitions); and transitions between characters within words all transitions.
+At before-sentence locations, interkey intervals were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GE2) but were timed to the character key press in the PLanTra and LIFT datasets (i.e.~the duration included the time to perform the capitalising shift keypress); we return to this difference in the Results section.
+Transitions that occurred at the very start of the text or at the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
+Importantly we also removed transitions that were followed by an editing operation.
+We therefore just modelled the times between keypresses in ongoing production.
 
@@
 
-\caption{\label{tab:keyloc}Transition location classification.}
+\caption{\label{tab:keyloc}Transition location classification. The final text in all cases was "The cat meowed. Then it slept."}
 
@@
 T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
-Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
+Before word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
 Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
@@
 We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data).
-We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
-We further removed keystroke intervals that were too short to represent a behavioural response (\(\le\) 50 msecs) or too long (\(\ge\) 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}.
+We removed participants that produced fewer than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).
+We further removed keystroke intervals that were too short to represent intentional typing (\(\le\) 50 msecs, see Gentner et al., 1980; Rumelhart \& Norman, 1982) or were of a length where such that they were unlikely to be associated with ongoing text production (\(\ge\) 30 secs); percentages can be found in Table \ref{tab:datareduction}.
 From the remaining data we randomly sampled a maximum of 100 observations per participant, condition, and transition location (when more than 100 were available).
-This was done to reduce the computation time of the Bayesian models require to complete sampling.
+This was done to reduce the computation time of the Bayesian models required to complete sampling.
 For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets.
@@
 \caption{\label{tab:datareduction}Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data included by transition location (`--` if all data were used).
+Shown are the percentage of extreme data that were removed and the percentage of randomly sampled data that were analysed by transition location (`--` if all data were included).
 Standard error in parentheses.}
@@
 \toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Randomly sampled data in \%} \\
+ \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Extreme values in \%} & \multicolumn{3}{c}{Data analysed in \%} \\
 \cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-6}
@@
 CATO & 0.65 (0.15) & 0.023 (0.016) & 15 (0.95) & 49 (2.2) & --\\
-GUNNEXP2 & 2.2 (0.17) & 0.0072 (0.0055) & 6.2 (0.43) & 23 (1.4) & --\\
+GE2 & 2.2 (0.17) & 0.0072 (0.0055) & 6.2 (0.43) & 23 (1.4) & --\\
 LIFT & 2.6 (0.16) & 0.0048 (0.0038) & 3.2 (0.22) & 13 (0.86) & 99 (0.15)\\
@@
 
-\hypertarget{bayesian-statitical-modelling}{%
-\subsection{Bayesian statitical modelling}\label{bayesian-statitical-modelling}}
-
-We reanalysed keystroke data -- transition durations measured as the time between first and second keypress -- from six datasets in a series of five Bayesian models.
+\hypertarget{statistical-modelling}{%
+\subsection{Statistical modelling}\label{statistical-modelling}}
+
+We analysed keystroke data in a series of four Bayesian mixed-effects models.
 An overview of all models can be found in Table \ref{tab:models}.
-More detailed model descriptions and motivation can be found in Appendix \ref{statistical-models}.
+Full modelling details can be found in Appendix \ref{statistical-models} but we summarised the most relevant properties in this section.
+
+The first three models -- models M1, M2, and M3 -- were single-distribution models consistent with the serial account.
+They were single distribution models, with fixed effects for each combination of transition location and dataset-specific manipulations.
+Models M1 (Gaussian) and M2 (log-Gaussian) are consistent with standard models used in the literature and therefore serve as baseline models.
+For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; see Appendix \ref{statistical-models} for rationale.
 
@@
 
-\caption{\label{tab:models}Overview of cognitive models of writing. All models included by-participant random effect.}
+\caption{\label{tab:models}Model overview. All models included by-participant random effect and study-specific manipulations.}
 
@@
 \textbf{Serial} & \\
-\ \ \ M1 & Single Gaussian model with effects of transition location (and experiment-specific manipulation).\\
+\ \ \ M1 & Single Gaussian model with effects of transition location.\\
 \ \ \ M2 & Single distribution log-Gaussian model equivialent to M1.\\
 \ \ \ M3 & Equivalent to M2 but with different variance components for each transition location (unequal variance).\\
-\textbf{Cascading} & \\
-\ \ \ M4 & Two-distributions mixture of a log-Gaussian for fluent transition intervals and another log-Gaussian that is wider than the previous distribution hesitant / disfluent key transitions; the distribution of hesitant keystroke intervals assumes -- similar to M3 -- different values for transition location but the distribution of fluent intervals has the same mean across all design cell.
-Mixing proportions capture the relative number of disfluent transitions for the distribution of hesitant transitions for each combination of transition location.\\
-\ \ \ M5 & Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location.\\
+\textbf{Parallel} & \\
+\ \ \ M4 & Two-distributions mixture of a log-Gaussian for fluent interkey intervals and another wider log-Gaussian for hesitant / disfluent interkey intervals; the distribution of hesitant interkey intervals assumes -- similar to M3 -- different values for transition locations.
+The mixing weight of these distributions capture the relative number of disfluent transitions for each transition location.\\
 \bottomrule
@@
 
-Models M1, M2, and M3 are single distribution models -- consistent with the serial account -- include fixed effects for each combination of transition location and dataset-specific manipulations.
-In particular models M1 and M2 are consistent with standard models used in the literature and therefore serve as baseline models.
-For model M3 we relaxed the equal-variance assumption for transition locations, thus allowing transition locations on larger linguistic boundaries to assume a larger standard deviation; this is because slower human behaviour is generally known to be associated with a larger variability (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973); see Appendix \ref{statistical-models} for rationale.
-
-Models M4 and M5 are extensions of the unequal-variance single distribution model M3.
-Importantly models M4 and M5 assume that keystroke data result from a combination of two data generating processes -- consistent with the cascading account -- rather than one.
-These model are called finite mixture models in the literature (Gelman et al., 2014; MacLahlan \& Peel, 2000; Roeser et al., 2021) and capture the assumption that processing at higher levels of activation leads to longer pauses which can in principle happen at transition before sentence, before words or even mid-word.
-In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that pauses at larger linguistic edges are more likely but not obligated by the cognitive system that generates keystrokes.
-This is achieved by modelling key transitions as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
+Model M4 is a two-distribution mixture model consistent with the parallel cascading account.
+This model assumes that keystroke data result from a combination of two data generating processes; these models are referred to as finite mixture models in the literature (Gelman et al., 2014; MacLahlan \& Peel, 2000).
+Both single-distribution models and the two-distribution mixture model capture that processing difficulty at higher levels of activation leads to longer pauses but only the two-distribution model captures that planning-related pauses may occur at any interkey interval, even mid-word, and that planning may not be reflected in sentence or word-initial pauses when planning happens in parallel to output.
+In other words, instead of assuming that there is one process that shifts the distribution of transition durations for larger linguistic edges, we allow for the possibility that key-transitions at larger linguistic edges are more likely to reflect processing delays but the cognitive system does not obligate planning to occur before sentences or words.
+This is achieved by modelling interkey intervals as coming from a weighted mixture of two distributions associated with two different states, illustrated in equation \ref{eq:bimodcon2}:
 
@@
 These fluent keystroke transitions are merely constrained by a person's ability to move their finger.
-The population estimate for fluent key transitions (i.e.~typing speed) is captured by the \(\beta\) parameter.
+Fluent interkey intervals (i.e.~typing speed) are captured by the \(\beta\) parameter.
+\(\beta\) is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon2} referring to the same unknown parameter.
 \item
-  Any interruptions at upstream levels of mental representation result in a delay in the information flow and therefore create a lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
-The population estimate of the hesitation slowdown is captured by \(\delta\) and its frequency is captured by the mixing weight (proportion) \(\theta\).
+  Any interruption at upstream levels of mental representation delays the information flow reflected in a resulting lag between keystroke intervals, for example when words or their spelling could not be retrieved in time.
+The slowdown for such hesitations is captured by \(\delta\) and its frequency by the mixing proportion \(\theta\).
+The slowdown \(\delta\) was allowed to vary by transition locations because hesitations at larger linguistic units are likely to be associated with higher level planning which causes longer delays in the output.
+The \(\delta\) parameter was constrained to be positive, so that it captures how much longer hesitant interkey intervals are in addition to \(\beta\).
 \end{enumerate}
@@
 \label{eq:bimodcon2}
-  \text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-& (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \times \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
@@
 
-The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant key transitions, and the second line represents the distribution of fluent -- uninterrupted -- key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parameterised to represent the probability that a key transition is associated with the distribution of hesitant transitions.
+The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant interkey intervals that include the \(\delta\) parameter presenting the slowdown for hesitant transitions; the second line represents the distribution of fluent -- uninterrupted -- interkey intervals that is characterised by \(\beta\) which is also present in the first file.
+Either of these two distributions is associated with the mixing proportion \(\theta\) that ranges between 0 and 1.
+\(\theta\) was parameterised here to represent the probability that an interkey interval is associated with the distribution of hesitant transitions.
 This probability is inversely related to the mixing weight of the distribution of fluent transitions by \(1-\theta\) .
 In other words, a larger weight for either distribution inevitably means a lower weight for the other.
-We call this parameter the probability of hesitant transitions.
-When a parameter was assumed to vary by transition location (levels: before sentence, before word, within word) this was indicated as subscript; similar for participants.
-The hesitation probability was allowed to vary by participants constrained by a hyper-parameter for each transition location.
-
-The constrained implementation M4 assume that the size of a fluent key-transition does not vary by transition location.
-In other words, the value that the \(\beta\) parameter is taking on is the same for transitions at before-sentence, before-word, and within-word locations.
-However, letter bigrams may be executed systematically faster than transitions between space and either a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations.
-This is for two reasons: first, and importantly, highly frequent bi- and even trigrams are stored, retrieved and executed as one motor code (Behmer Jr.
-\& Crump, 2016; Grudin \& Larochelle, 1982; Ostry, 1983; Terzuolo \& Viviani, 1980) because of the strong association between letters.
-This association is reducing the length of transition durations.
-It is less likely that the same association exists for space-letter and space-shift pairs.
-Second, before-sentence transition in two datasets (PLanTra, LIFT) are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transition; we return to this in the results section.
-We therefore implemented an unconstrained mixture model M5 that assumes that the distribution over durations of fluent transitions -- represented as \(\beta\) -- may vary across transition locations.
+We call this parameter the probability of hesitant interkey intervals.
+When a parameter is allowed to vary by transition location (levels: before sentence, before word, within word), this was indicated as subscript; similar for participants.
+For example, \(\theta_\text{location[i], participant[i]}\) means that the mixing proportion \(\theta\) is allowed to take on a different value for both the transition location and participant associated with the \(i^\text{th}\) keystroke interval \(i \in 1 \ldots N\) where \(N\) is the total number of keystroke intervals.
+Because the hesitation probability was allowed to vary by participants as well as location type, we included a hyper-parameter for the mixing proportion \(\theta\) for each transition location.
+
+As with model M3, M4 allowed the variance associated with the key-intervals indicated as \(\sigma_{e'}^2\) to vary by transition location.
+Further each of the two distributions in M4 had its own variance constrained such that the variance associated with the distribution of hesitant keystrokes \(\sigma_{e'}^2\) was larger than the distribution of fluent keystroke intervals \(\sigma_{e}^2\).
+This is important because slower human behaviour similar to hesitant keystroke transitions is generally known to be associated with a larger variability (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
+Model M4 and all other models included a random intercepts term \(u_\text{participants}\) for participants which is constrained to be distributed around 0 with an unknown standard deviation, i.e.~\(\mathcal{N}(0, \sigma_\text{p}^2)\), hence indicating participant specific deviations from the average typing speed \(\beta\).
 
 All models were implemented in the Bayesian framework (Gelman et al., 2014; McElreath, 2016).
-The Bayesian framework is ideal for the estimation of parameter values.
-This is because Bayesian parameter estimates are expresses as probability distributions that capture the associated uncertainty (Farrell \& Lewandowsky, 2018; Gelman et al., 2014; Lee \& Wagenmakers, 2014).
-To achieve this, Bayesian models require the inclusion of prior information, i.e.~existing knowledge about parameter values.
-For small datasets priors influence the inferred parameter values; for larger datasets weakly informative and vague priors are quickly overcome by the data (i.e.~automatic Ockham's razor, Jefferys \& Berger, 1992).
-In other words the choice of priors has less impact on the posterior, certainly for weakly informative priors.
 In the present analysis, we used weakly informative priors to aid model convergence by constraining the parameter space to plausible values (see e.g.
@@
 Also, as the sample size of most of our datasets is large, weakly informative priors have no or a negligible effect on the posterior.
-
-Stan code for mixture models was based on (Roeser et al., 2021; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (ADD URL HERE); for a tutorial see \href{rpubs.com/jensroes/1000459}{https://rpubs.com/jensroes/1000459}.\footnote{The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
+Stan code for mixture models was based on Roeser, De Maeyer, et al.
+(2024; see also Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017) and can be found on OSF (Roeser, Conijn, et al., 2024); for a tutorial see \href{rpubs.com/jensroes/mixture-models-tutorial}{https://rpubs.com/jensroes/mixture-models-tutorial}.\footnote{The R (R Core Team, 2020) package rstan (Stan Development Team, 2018) was used to interface with the probabilistic programming language Stan (Carpenter et al., 2016) which was used to implement all models.
 Models were run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning.
@@
 
+To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling (Vehtari et al., 2015, 2017).
+Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared by the difference between models \(\Delta\widehat{elpd}\).
+We also summarised this difference as normalised over its standard error \(\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid\), the \emph{z}-score of the difference between models (Sivula et al., 2020).
+Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.
+
 \hypertarget{results}{%
@@
 
-To compare the out-of-sample predictive performance of our models we used leave-one-out cross-validation based on Pareto smoothed importance-sampling (Vehtari et al., 2015, 2017).
-Predictive performance was estimated as the sum of the expected log predictive density (\(\widehat{elpd}\)) and compared by the difference between models \(\Delta\widehat{elpd}\).
-We also summarised this difference as normalised over its standard error \(\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid\), hence the \emph{z}-score of the difference between models (Sivula et al., 2020).
-Similar to other cross-validation techniques, the advantage of leave-one-out cross-validation is that more complex models -- models with more parameters -- are penalised to prevent overfit.
-
-Model differences are shown in Table \ref{tab:loos}.
-All datasets revealed the same pattern.
-Importantly, the constrained two-distributions mixture model rendered a higher predictive performance compared to the unequal variance single-distribution model; the difference \(\Delta\widehat{elpd}\) between models ranged within {[}18, 40{]} standard errors, indicating a substantially higher predictive performance for the two-distributions model.
-The unconstrained mixture model M5 rendered a higher predictive performance than the constrained model M4 which did not allow the distribution of fluent keystroke-intervals to vary by transition location; the difference in predictive performance was between {[}7, 15{]} standard errors higher for the unconstrained model.
-Among the single distribution models we found higher predictive performance for the unequal variance model compared to the log-Gaussian model; the difference \(\Delta\widehat{elpd}\) ranged within {[}43, 68{]} standard errors.
-The weakest model was the single distribution Gaussian which had -- compared the single distribution log-Gaussian -- a lower predictive performance in a range of {[}43, 69{]} standard errors.
-
-\begin{landscape}
+The differences in the predictive performance of models for all datasets are shown in Table \ref{tab:loos}.
+For all six dataset the two-distributions mixture model (M4) provided greater predictive performance than all three single-distribution models.
+Differences in predictive performance (\(\Delta\widehat{elpd}\)) between models M4 and the nearest competing single-distribution model (M3) ranged between {[}18, 40{]} standard errors, indicating a substantially higher predictive performance for the two-distributions model.
+Among the single distribution models we found higher predictive performance for the unequal variance model and the lowest predictive performance for the single distribution Gaussian.
+Comparisons showing the fit of model predictions to the data can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
+Data predicted by the two-distribution mixture model closely align with the observed data; data predicted by single distribution Gaussian models showed an inferior fit to the data.
+
+\begin{table}[tbp]
 
 \begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} $\Delta\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{p{1.5cm}p{5cm}rrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\begin{threeparttable}
+
 \caption{\label{tab:loos}Model comparisons.
 Models were compared incrementally from the simplest to the most complex model.
-The comparison is shown along with the distribution type of the more complex model.
-A negative difference in $\Delta\widehat{elpd}$ indicates higher predictive performance for the more complex model.
-Standard error is shown in parentheses.}\\
+Comparison are shown with the distribution type of the model with the higher predictive performance.
+A negative difference in $\Delta\widehat{elpd}$ indicates higher predictive performance for the more complex model; standard error shown in parentheses.}
+
+\small{
+
+\begin{tabular}{p{1.75cm}p{4cm}p{4.25cm}p{3.5cm}}
 \toprule
-Comparison & \multicolumn{1}{c}{Distribution} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
+Dataset & Mixture of two log-Gaussians (M3 -- M4) & Single log-Gaussian (unequal variance; M2 -- M3) & Single log-Gaussian (M1 -- M2)\\
 \midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loos} continued}}\\
-\toprule
-Comparison & \multicolumn{1}{c}{Distribution} & \multicolumn{1}{c}{C2L1} & \multicolumn{1}{c}{CATO} & \multicolumn{1}{c}{GUNNEXP2} & \multicolumn{1}{c}{LIFT} & \multicolumn{1}{c}{PLanTra} & \multicolumn{1}{c}{SPL2}\\
-\midrule
-\endhead
-M4 -- M5 & Mixture of two log-Gaussians (unconstrained; M5) & -546 (51) & -607 (43) & -554 (38) & -500 (37) & -130 (18) & -564 (41)\\
-M3 -- M4 & Mixture of two log-Gaussians (constrained; M4) & -1,637 (72) & -1,782 (75) & -2,427 (92) & -5,105 (127) & -2,384 (90) & -1,181 (64)\\
-M2 -- M3 & Single log-Gaussian (unequal variance; M3) & -38,301 (823) & -38,972 (913) & -41,155 (653) & -76,961 (1,656) & -35,204 (556) & -30,430 (450)\\
-M1 -- M2 & Single log-Gaussian (M2) & -37,517 (798) & -37,673 (881) & -38,843 (616) & -74,851 (1,624) & -33,823 (526) & -28,717 (418)\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\end{landscape}
-
-We also tested to what extent model predictions fit observed data.
-Comparisons can be found in Appendix \ref{fit-to-data} and echo the findings in Table \ref{tab:loos}.
-Data predicted by the two-distribution mixture models are similar to observed data; data predicted by single distribution Gaussian models showed a poor fit to the data.
-
-\hypertarget{transition-location-effect}{%
-\subsection{Transition location effect}\label{transition-location-effect}}
-
-As demonstrated in the model comparisons, the two-distributions mixture-models is a better statistical representation on the underlying process generating keystroke data than compared to any of the single-distribution models.
-Information about this process is captured by the posterior estimates of the models' parameter values.
-To reiterate, there are three conceptually important parameters: (1) the magnitude of the slowdown for transition hesitations (represented as \(\delta\) in equation \ref{eq:bimodcon2}), (2) the probability of hesitant transitions (represented as \(\theta\) in equation \ref{eq:bimodcon2}), and (3) the duration of fluent transitions (the typing speed, which was indicated as \(\beta\) in equation \ref{eq:bimodcon2}).
-
-Figure \ref{fig:mixmodel} illustrates the mixture model approach on the basis of the posterior of the constrained implementation applied to the CATO data: for each transition durations we obtain two posterior probability distributions, one associated with fluent transitions between keys and another for key transitions where upstream difficulty result in hesitant transitions.
-The hesitation probability is indicated as the relative height of either distribution.
-The transition locations are characterized by the hesitation parameters: as we illustrate the constrained mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location but their height differs as the probability of fluent transition is the inverse of the height of hesitant transitions; i.e.~\(1 - \theta\).
-More interesting are the distributions associated with hesitations; indicated in yellow: Within word transitions have a negligibly small hesitation probability barely visible in the visualisation.
-Hesitations before words and sentences are both roughly equally likely than fluent transitions.
-In other words, half of the time participants did not pause before a word or sentence; we explore this below.
-However when hesitations occurred, these are longer at before-sentence transitions compared to before word transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/mixmodel-1}
-
-}
-
-\caption{Mixture model plot for each transition location.
-Shown are the distributions for both mixture components (fluent and hesitant transitions) and their weighting.
-Posterior of CATO data.}\label{fig:mixmodel}
-\end{figure}
-
-We evaluate the posterior of the constrained mixture model; however we also summarise the posterior of the unconstrained mixture model below and the full analysis can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-We demonstrated above that the predictive performance of the unconstrained mixture model was substantially higher than the predictive performance of the constrained mixture model.
-However, the constrained mixture model is the theoretically more parsimonious model of the cascading view; in other words, differences between transition locations for the component of fluent transitions were to some extent speculative and unexpected.
-We therefore present the posterior parameter values for the constrained two-distributions mixture model in the following.
-
-The mixture-model posterior parameter estimates for all datasets is visualised in Figure \ref{fig:crossstudypost2}.
-Although models were fitted with all data set-specific conditions, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.
-We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.}, and excluded conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
-There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-gunnexp2}).}.
-For posteriors of all conditions within data sets see Appendix \ref{posterior-parameter-estimates}.
-The resulting posterior allows us to examine differences and similarities associated with transition locations for all dataset.
-In particular we can see how the linguistic location in the text is associated with hesitation duration and probability.
-The estimate for fluent transitions is shown by dataset, across transition location.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost2-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.}\label{fig:crossstudypost2}
-\end{figure}
-
-Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets.
-We summarise the results of the pairwise comparisons for transitions locations; full results and parameter estimates can be found in Appendix \ref{transition-location-effect-constrained-mixture-model}.
-We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs \textgreater{} 100 for all datasets) but only for half of the data sets also longer (GUNNEXP2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
-Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GUNNEXP2 and SPL2, both BFs \textgreater{} 100) but longer (all BFs \textgreater{} 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
-
-To reiterate, for most datasets we found that pauses are before sentences are longer but not necessarily more likely than before words (except for datasets SPL2 and GUNNEXP2).
-In fact, LIFT showed some indication that pauses were less likely before sentences than words.
-This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for key transitions at larger linguistic edges.
-It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.
-
-Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character while the remaining data sets measured before-sentence transitions as the difference between space key and shift.
-To control for the use of complex key combinations, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability (BFs \textgreater{} 100) but it did not affect the duration of fluent or hesitant transitions (BFs \textless{} 0.3); see Appendix \ref{key-combination-effect-constrained-mixture-model}.
-This conflicts the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did included the character following shift (PLanTra, LIFT).
-In other words, patterns in our results cannot be explained on the basis of how before-sentence key-transitions were operationalised.
-
-Finally, the duration of fluent transitions was not distinguished by location in the constrained distribution model.
-Figure \ref{fig:crossstudypost2} highlights that two datasets (C2L1, CATO) show substantially longer fluent transitions.
-This is presumably reflecting that these data were sampled from was the youngest and least experienced population of writers in our data pool.
-Hesitation duration and probability followed the same patterns as the majority of datasets.
-
-The predictive performance of the unconstrained model was higher compared to the constrained model (see Table \ref{tab:loos}).
-We, therefore, summarise briefly differences in the results for the transition-location effects; full results can be found in Appendix \ref{transition-location-effect-unconstrained-mixture-model}.
-Higher predictive performance for the unconstrained model suggests that variability across transition locations for fluent transition durations; in particular fluent before-word transitions were longer compared to mid-word transitions (all BFs \textgreater{} 100) but not before-sentence compared to before-word transitions (all BFs \textless{} 0.08, but SPL2 and GUNNEXP2: BFs \textgreater{} 100).
-Patterns for hesitation probability was generally similar in both the constrained and unconstrained model.
-However, for hesitation durations the unconstrained model showed evidence for longer durations at before-word locations compared to within-word location in all but one dataset (all BFs \textgreater{} 10, except for LIFT: BF = 0.7); differences in hesitation duration for before-sentence transitions compared to before-word transitions was similar to the constrained model.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The classical serial view (Flower \& Hayes, 1980) characterizes writing as a sequence of planning and execution cycles that results in a writing process that consists of bursts interrupted by pauses.
-The cascading model of writing, in contrast, captures that planning, at least in competent writers can occur in parallel to production (Olive, 2014).
-We used Bayesian statistical models, in particular hierarchical single-distribution and two-distribution mixture models, to directly compare the serial and the cascading model of writing, respectively.
-On the basis of six datasets with key data from free text production tasks, we presented compelling and consistent statistical evidence in favour of the cascading view.
-
-Patters observed in the parameter estimates are generally similar across data sets (with caveats).
-Under the cascading view of writing, key transitions that directly precede sentences or words are not necessarily associated with a pause because writers can plan in parallel to writing.
-On the basis of parameter estimates of the mixture model we tested to what extent larger linguistic edges are associated with the preparation of upcoming planning unit.
-We found that in the majority of datasets, pauses were as likely to occur before sentences and words, in both experienced and novice writers, while mid-word pauses are generally rare.
-Also, with few exceptions, pauses were typically longer before sentences than before words (which were longer than mid-word pauses).
-In other words, when writers do pause, the involvement of higher-level processing at sentence boundaries is typically leading to longer pauses.
-However, the pausing probability results can only be explained by the cascading view of writing, not by the serial view.
-This is because our results for the hesitation probability suggest that writers often do not pause before sentences; indeed, hesitation probabilities are largely identical before sentences and before words.
-This suggest that writers frequently plan the next word and even sentence in parallel to production which mitigates the need to pause to plan up-coming linguistic information.
-
-Interestingly even the youngest writers in our data pool (C2L1) showed evidence of parallel planning expressed in fluent transitions between words and even sentences.
-Key-transitions were substantially longer compared to all other datasets with a larger slowdown before words.
-Yet the overall pausing probability showed a patterns comparable to most other data sets.
-In particular, a larger pause probability was observed for transitions before sentences / words compared to mid-word transitions.
-This is interesting because, for example, Olive (2014) described a cascading model that is operating in a serial fashion for inexperienced or struggling writers explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
-However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers but pauses were more likely before words and sentences but similarly long indicating a tendency to prepare utterances in planning units possibly smaller than the clause.
-The main difference for these young writers was that their writing speed (fluent transitions) was substantially slower compared to more experienced writers in the sample.
-Also for L2 writers (PlanTra) we observe pausing patterns that do not resemble a serialised writing process as indicated by a larger number of fluent transitions across word and sentence boundaries.
-
-We observed two more differences between data sets -- other than the patterns observed for C2L1 summarised above -- that are worth highlighting.
-These can largely be explained on the basis of differences associated with the population (writing competence / experience) and text genre (see also Conijn et al., 2019) that these samples were obtained from: The first exception concern the GUNNEXP2 and SPL2 data sets: pauses were more likely to occur at before-sentence transitions compared to before-word transitions while in the remaining data sets hesitations were equally probably at before-word and before-sentence locations.
-\textbf{TODO: I don't have an explanation for this.} Second, pauses were not always longer before sentence compared to before words compared to mid-word paues; there were two exceptions: (1) no difference was found for before-sentence compared to before-word transitions in C2L1 possibly due to a a word-level planning strategy in this sample of young writers as inexperienced writers tend to use a more localised strategy for text production.
-(2) pause durations did not vary by transition location in the LIFT dataset although the posterior showed the same general patterns as observed in most other datasets.
-The absence of a difference could be explained as specific to the synthesis writing task because there is no need for content creation {[}\textbf{REFERENCE?}{]} but the results for the PlanTra dataset rule out this possibility.
-
-A possible concern with our result -- substantially better predictive performance for two-distributions mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-This concern is preempted by using cross-validation techniques for model comparison which is preventing model overfit.
-In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model.
-We simulated two data sets that were either based on a single distribution or two-distributions random data generating process.
-This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values.
-Modelling details can be found in Appendix \ref{simulation}.
-As might be expected, we found, that the two-distribution mixture was successful at uncovering the parameter values of the data generated with a two-distributions mixture models; we observed the same for the single-distribution model of the data generated with a single distribution process.
-Parameters were not successfully uncovered when we switched model type and dataset.
-Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the single distribution model, when applied to data based on a single distributions process but a substantially higher predictive performance for data simulated from a two-distributions mixture process.
-Therefore we can rule out the possibility that the overwhelming statistical support for the two-distributions mixture models can be explained in terms of overfitting.
-
-Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units.
-Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text.
-There is limited research that used eye tracking to investigate regressive gaze behaviour during writing (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
-There are at least four reasons why eye movements away from the inscription point might occure during writing: First, reading is necessary for revision (i.e.~writers must check if writing goals have been met).
-Second, writers might use text to cue or reinstating ideas after interruption of writing process.
-Third, writers might look for or correct spelling or grammatical errors.
-We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text.
-Fourth, writers might look at the source their are writing from.
-For two datasets (PlanTra, LIFT) which involved writing from sources, we removed transitions that were made when the sources (rather than the writing document) was in focus.
-
-Frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example Torrance, Johansson, et al.
-(2016) reported that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words (\enquote{hopping,} see also Chukharev-Hudilainen et al., 2019).
-This is except, lookbacks are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words.
-Finally even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.
-This can be seen in the result of two datasets in our pool (CATO, GUNNEXP2): these experiments included a manipulation where the produced text was masked (or not).
-In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}.
-This suggest that pausing patterns do not change if reading previous text is not possible.
-
-A practical advantage of the mixture-model approach (also Roeser et al., 2021) is that we are not required to stipulate interval thresholds as plausible candidate for lower pause bound to handle the complex distribution of latency from spontaneous text production.
-Such threshold are widely used by writing researchers, a strategy that was inherited from early research in speech production (for review see Rochester, 1973), with studies dating back to at least the mid 1990s (Foulin, 1998).
-There is one central problems with this approach: A threshold requires a definition of what passes as a pause (Van Waes et al., 2016; Wengelin, 2006), i.e.~a pause criterion threshold often set to 2 secs (Chanquoy et al., 1996; Kaufer et al., 1986; Sullivan \& Lindgren, 2002; Wengelin, 2002) or some other lower bound (Chukharev-Hudilainen, 2014; Connelly et al., 2012; Leijten \& Van Waes, 2013).
-Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause (see e.g.
-Wengelin, 2006) they would be arbitrary.
-For example, a sentence-initial pause of 2 secs has a very different interpretation from a 2 secs pause that occurs before or within a word.
-Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision.
-So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs.
-However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose (Chenu et al., 2014).
-Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values (see also Almond et al., 2012; Baaijen et al., 2012; Hall et al., 2022; Li, 2021).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In contrast to the serial pause-and-burst view, the cascading view emphasis that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution.
-Using the Bayesian framework we implemented models of both the serial model of writing and the cascading view.
-Model comparisons and the inspection of posterior parameter estimates supported with a cascading view of writing but provided strong evidence against the serial view.
-We found that key transitions across word and even sentences boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information.
-This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses).
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-
-\setlength{\parindent}{-0.5in}
-
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{references}{}
-
-\endgroup
-
-\newpage
-
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\newpage
-
-\hypertarget{statistical-models}{%
-\section{Statistical models}\label{statistical-models}}
-
-We are using the Bayesian framework (e.g.
-Farrell \& Lewandowsky, 2018; Gelman et al., 2014; Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing.
-In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models.
-The models presented in this section build on one another so that later model include assumptions of earlier models.
-
-\hypertarget{serial-model-of-writing}{%
-\subsection{Serial model of writing}\label{serial-model-of-writing}}
-
-\hypertarget{single-distribution-gaussian}{%
-\subsubsection{Single distribution Gaussian}\label{single-distribution-gaussian}}
-
-Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit.
-The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors.
-For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes.
-There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis.
-We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model.
-This can be expressed as \(\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)\).
-Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.
-
-We can extend this simple model of word-planning to other linguistic location.
-We introduce earlier that larger linguistic edges are associated with planning on higher levels.
-For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
-Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors.
-We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries.
-However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs.
-Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning.
-The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).
-
-We implemented the average keytransition associated with a given text locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location.
-Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs.
-Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\).
-The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\).
-This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).
-
-The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodgaus}
-\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant[i]} \sim \text{N}(0, \sigma_\text{p}^2)\\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-where \(i\) iterates over every key-transition \(i \in 1 \dots N\) with \(N\) being the total number of keystroke intervals, and thus \(\text{location[i]}\) is the transition location associated with the \(i\)\^{}\text{th} keystroke and \(\text{participant[i]}\) is the participant associated with the \(i\)\^{}\text{th} keystroke.
-
-Standard-deviation parameters were constrained to be positive because standard deviations can never be negative.
-
-\hypertarget{single-distribution-log-gaussian}{%
-\subsubsection{Single distribution log-Gaussian}\label{single-distribution-log-gaussian}}
-
-The previous model assumes that the data-generating process is a Gaussian distribution.
-The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.
-There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values.
-IKIs, as the distance between two subsequent key-down events, must be positive.
-The lower bound is delimited by a person's ability to move their fingers and keyboard polling.
-(2) the log-scale is known to be a better match for data from human behaviour and motor responses.
-In particular, a Gaussian distribution assumes that units are scaled linearly.
-For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
-This does not necessarily map onto the psychological interpretation for short and long keyintervals.
-For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2).
-In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals.
-Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).
-
-The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{log}\mathcal{N}()\) was replaced by \(\text{logN}()\).
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimodloggaus}
-\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
-\text{where: } &
-\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
-& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{single-distribution-unequal-variance-log-gaussian}{%
-\subsubsection{Single distribution unequal-variance log-Gaussian}\label{single-distribution-unequal-variance-log-gaussian}}
-
-The third model representing the serial view of writing is a single distribution unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section.
-The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text.
-The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance).
-This assumption is however not in line with what we know about data from human behaviour.
-Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973).
-For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes.
-Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies by transition location.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:unimoduv}
-\text{iki}_i \sim\text{ } & \text{log}\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
-\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
- & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
- \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
-\end{aligned}
-\end{equation}
-
-\hypertarget{parallel-cascading-model-of-writing}{%
-\subsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}
-
-The following two models are are extensions of the models introduced for the serial view.
-Crucially, the cascading view allows planning to happen in parallel to production.
-Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset.
-This is done by assuming that IKIs come from a weighted combination of two distributions.
-
-\hypertarget{two-distributions-log-gaussian-mixture-model-constrained}{%
-\subsubsection{Two-distributions log-Gaussian mixture model (constrained)}\label{two-distributions-log-gaussian-mixture-model-constrained}}
-
-This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses.
-Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory.
-This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Activation can flow into keystrokes without interruption.
-These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}.
-Note that \(\beta\) is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.
-\item
-  Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time.
-The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
-The magnitude of the slowdown is associated with transition location.
-This is because delays at larger linguistic units are likely to be associated with higher level planning.
-By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
-\end{enumerate}
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimodcon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 < \theta < 1
-\end{aligned}
-\end{equation}
-
-The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions.
-Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1.
-\(\theta\) is here parameterised to represent the probability that an IKI is associated with the distribution of long IKIs.
-This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\) .
-In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution.
-The weights of both distributions must sum to 1.
-We will call this parameter the probability of hesitant transitions.
-
-The probability of hesitant transitions is assumed to vary as a function of both transition location and participants.
-In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges.
-As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less.
-This is akin to what is known as a random by-participant slopes model.
-
-Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
-In addition we constrained the variances so that \(\sigma_{e'}\) associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions \(\sigma_e\) (see Vasishth, Chopin, et al., 2017; Vasishth, Jäger, et al., 2017).
-This was achieved by introducing a parameter \(\sigma_\text{diff}\).
-The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.
-
-\hypertarget{two-distributions-log-gaussian-mixture-model-unconstrained}{%
-\subsubsection{Two-distributions log-Gaussian mixture model (unconstrained)}\label{two-distributions-log-gaussian-mixture-model-unconstrained}}
-
-The size of a fluent key-transition does not necessarily vary by transition location.
-In other words, the parameter \(\beta\) is the same for before-sentence, before-word, and within-word transitions.
-This is what we called a constrained model.
-However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section).
-This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter.
-Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data.
-We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.
-
-In this model we assume that \(\beta\) varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:bimoduncon}
-\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
-    \text{where: }  & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
-\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
-        & \sigma_{e'}^2 > \sigma_{e}^2\\
-        & 0 < \theta < 1
-\end{aligned}
-\end{equation}
-
-\newpage
-
-\hypertarget{fit-to-data}{%
-\section{Fit to data}\label{fit-to-data}}
-
-Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data.
-From each model we obtained 100 simulations (predictions).
-These can be found in Figure \ref{fig:prediction}.
-The model predictions fitted the data substantially better for the mixture models.
-The worse fit was observed for the single distribution Gaussian model which predicted large amounts of negative keystroke intervals which are implausible.
-Using a log-normal distribution removed these negative keystroke intervals.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/fitplots}
-
-}
-
-\caption{Fit to data with models (distribution type) above the plots.
-Comparison of 50 simulations to observed data summarised by model.
-For illustration the x-axis was truncated at 2 secs.}\label{fig:prediction}
-\end{figure}
-
-\newpage
-
-\hypertarget{transition-location-effect-unconstrained-mixture-model}{%
-\section{Transition location effect (unconstrained mixture model)}\label{transition-location-effect-unconstrained-mixture-model}}
-
-We present the mixture-model posterior estimates for each of the three parameter values in the three facets of Figure \ref{fig:crossstudypost}.
-Similar to the constrained mixture model, we aggregated the posterior across conditions\footnote{We aggregated across pre-post test for the PLanTra dataset as well as genre and topic of the LIFT data set.} and removed conditions that might conflate comparisons\footnote{We removed the masked writing condition in the GUNNEXP2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.}.
-The resulting posterior allows us to examine differences between transition locations for each data set associated with each of the three mixture-model parameters.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost-1}
-
-}
-
-\caption{Mixture model parameter estimates across studies.
-Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GUNNEXP2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.
-Axes for transitions durations are log-scaled for visability.}\label{fig:crossstudypost}
-\end{figure}
-
-In the following we evaluate differences between transition locations for all three mixture model parameters.
-Figure \ref{fig:crossstudypost} shows largely the same patterns (with caveats) for keystroke interval estimates by transition location across data sets for all three mixture model parameters.
-We found that hesitations appear more frequently at before-word transitions than within words across data sets (all BFs \textgreater{} 100).
-Also hesitations are longer at before-sentence transitions compared to before-word transitions (all BFs \textgreater{} 100; except for C2L1: BF = 0.35 and LIFT = 2.14) and before-word transitions compared to within-word transitions (all BF \textgreater{} 10; except for LIFT: BF = 0.71).
-Further, we observed that fluent key-transitions are slower at before-word and before-sentence locations compared to within-word locations (all BFs \textgreater{} 100) but there is generally no evidence for a difference for fluent transitions for before-sentence transitions compared to before-word transitions (BFs \textless{} 0.08; except for SPL2: BF \textgreater{} 100 and GUNNEXP2: BF \textgreater{} 100).
-The full results of these pairwise comparisons can be found in Table \ref{tab:loceffect}.
-
-Hesitation duration tends to be longer at before-sentence locations compared to before-word locations (except for data sets C2L1 and LIFT) and longer for before-word locations compared to within-word locations (except for the LIFT data set).
-However, for most data set there is negligible evidence for the idea that writers pause more frequently at before-sentence locations compared to before-word locations (except for data sets SPL2 and GUNNEXP2).
-This is interesting because it is generally believed that pausing behaviour is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges following the pattern before-sentence \(>\) before-word \(>\) within-word.
-In fact, the data set LIFT showed less pausing before sentences compared to before words.
-In brief, while keystroke transitions and pauses tend to be longer and more frequent at before-word locations compared to within-word transitions, it is not clear in which contexts keystroke transitions are before-sentence locations are slower, and their hesitations are longer and more frequent.
-
-Outstanding is the overall substantially longer fluent transitions for the C2L1 data.
-This is presumably reflecting that the population that this sample is from was the youngest among our data sets presumably involving the least experienced writers in our data pool.
-Hesitation duration and frequencies were similar to the other data sets.
-There are some inconsistencies for fluent before-sentence transitions compared to before-word transitions.
-Some data sets show before-sentence slowdowns for fluent transition compared to words.
-These inconsistencies could, to some extent, be explained on the basis that data sets differ as to whether before-sentence transitions involve complex key combination involve the mean or sum of transitions between space, shift and / or the sentence-initial letters.
-In particular, some data include the character key following the shift key at before sentence location (PLanTra, LIFT) but others did not scope over the character following the shift key (CATO, C2L1, SPL2, GUNNEXP2).
-Notice though that for the data sets PLanTra and LIFT, there was no evidence for a consistent differences between before-sentence and before-word transitions (except for longer hesitations in the PLanTra dataset).
-
-To address this finding, and inconsistency in how before-sentence transitions were timed, we test to what extent this difference affected the modelling results for the SPL2 data set.
-The results are shown in Appendix \ref{key-combination-effect-unconstrained-mixture-model}.
-We found that including the character following the shift key substantially extends both the transition duration and the hesitation duration but not the hesitation frequency.
-However, this conflicts with the absence of differences in data sets that included the character following shift at before-sentence transitions (PLanTra, LIFT).
-In other words, it is unlikely that patterns in our results can be explained on the basis of how before-keystroke transitions were operationalised (complex key-combinations at before-sentence locations).
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffect}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffect} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fluent transitions} & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.13, 0.15] & 0.07 & 0.21 [-0.13, 0.54] & 0.35 & 0.54 [-0.31, 1.43] & 0.89\\
-\ \ \ before vs within word & 0.4 [0.38, 0.42] & > 100 & 0.49 [0.39, 0.58] & > 100 & 1.1 [0.76, 1.44] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.03 [-0.18, 0.11] & 0.08 & 1.19 [0.88, 1.49] & > 100 & 0.41 [-0.44, 1.27] & 0.67\\
-\ \ \ before vs within word & 0.35 [0.31, 0.38] & > 100 & 0.35 [0.14, 0.54] & 17.48 & 1.83 [1.19, 2.49] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.22 [0.12, 0.32] & > 100 & 0.93 [0.78, 1.07] & > 100 & 1.32 [0.86, 1.78] & > 100\\
-\ \ \ before vs within word & 0.23 [0.21, 0.25] & > 100 & 0.38 [0.21, 0.54] & > 100 & 1.97 [1.59, 2.37] & > 100\\
-\textbf{LIFT} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & -0.04 [-0.13, 0.05] & 0.07 & 0.35 [0.05, 0.72] & 2.14 & -0.49 [-1.03, 0] & 1.63\\
-\ \ \ before vs within word & 0.21 [0.13, 0.27] & > 100 & 0.26 [-0.16, 0.55] & 0.71 & 1.56 [1.02, 2.16] & > 100\\
-\textbf{PLanTra} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.01 [-0.09, 0.11] & 0.07 & 0.65 [0.46, 0.83] & > 100 & 0.13 [-0.25, 0.53] & 0.25\\
-\ \ \ before vs within word & 0.16 [0.1, 0.21] & > 100 & 0.37 [0.2, 0.54] & > 100 & 1.83 [1.48, 2.17] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.17, 0.31] & > 100 & 1.39 [1.25, 1.52] & > 100 & 0.69 [0.17, 1.19] & 8.01\\
-\ \ \ before vs within word & 0.31 [0.28, 0.34] & > 100 & 0.35 [0.14, 0.54] & 16.25 & 1.94 [1.4, 2.5] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\newpage
-
-\hypertarget{key-combination-effect-unconstrained-mixture-model}{%
-\section{Key-combination effect (unconstrained mixture model)}\label{key-combination-effect-unconstrained-mixture-model}}
-
-The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences sumed across two key intervals in the PLanTra and LIFT data, namely \texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval, namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets.
-Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
-Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.
-
-Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
-We compared the SPL2 data including and excluding the keystroke after shift.
-Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
-The results of this comparison can be found in Table \ref{tab:shiftcellmeans2}.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans2}Mixture model estimates for key transitions.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans2} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  & \\
-\ \ \ L1 & 159 [147, 172] & 0.48 [0.34, 0.63] & > 100\\
-\ \ \ L2 & 159 [147, 172] & 0.41 [0.19, 0.63] & 46.08\\
-Hesitation duration &  &  & \\
-\ \ \ L1 & 1,281 [1,082, 1,496] & -0.46 [-0.61, -0.3] & > 100\\
-\ \ \ L2 & 1,802 [1,557, 2,073] & -0.34 [-0.51, -0.17] & > 100\\
-Hesitation probability &  &  & \\
-\ \ \ L1 & .78 [.69, .86] & 0.49 [-0.04, 1.03] & 1.43\\
-\ \ \ L2 & .92 [.87, .96] & 0.48 [-0.16, 1.13] & 0.94\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift.
-Fluent key transitions were substantially longer when including the interval following the shift key.
-The slowdown for hesitations was affected too but the difference is numerically small.
-There was no conclusive evidence for an increased hesitation probability.
-
-\newpage
-
-\hypertarget{key-combination-effect-constrained-mixture-model}{%
-\section{Key-combination effect (constrained mixture model)}\label{key-combination-effect-constrained-mixture-model}}
-
-The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
-In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely \texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval, namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets.
-Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.
-
-Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions.
-We compared the SPL2 data including and excluding the keystroke after shift.
-Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations.
-The results of this comparison can be found in Table \ref{tab:shiftcellmeans}.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions immediately preceding a sentence.
-Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown and the probability of hesitant transitions.
-The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
-\toprule
-Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Fluent transitions &  &  &  & \\
-\ \ \ overall & 156 [145, 169] & 159 [147, 172] & -0.02 [-0.13, 0.09] & 0.06\\
-Hesitation duration &  &  &  & \\
-\ \ \ L1 & 1,167 [1,037, 1,305] & 1,281 [1,082, 1,496] & -0.06 [-0.21, 0.09] & 0.11\\
-\ \ \ L2 & 2,019 [1,772, 2,292] & 1,802 [1,557, 2,073] & 0.12 [-0.02, 0.27] & 0.29\\
-Hesitation probability &  &  &  & \\
-\ \ \ L1 & 1.00 [.99, 1.00] & .78 [.69, .86] & 4.91 [3.68, 6.35] & > 100\\
-\ \ \ L2 & 1.00 [.99, 1.00] & .92 [.87, .96] & 3.4 [2.12, 4.85] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-The fluent transition duration and hesitation duration were not affected by whether or not the sentence-initial transition include the character following shift for neither language.
-However, we found strong evidence for an increased hesitation probability -- in both -- languages when the before-sentence key transition included the character following the shift key.
-Notably, the hesitation probability increased ceiling.
-It is not entirely surprised that the constrained mixture model identifies essentially all before-sentence transition as pauses: the constrained model does not distinguish between transition locations for the duration of fluent transitions.
-Therefore essentially all before-sentence transitions are identified as hesitation as they always include two keystrokes while all other transition location (which is the majority of the data) do not.
-
-\newpage
-
-\hypertarget{l2-effect-spl2}{%
-\section{L2 effect (SPL2)}\label{l2-effect-spl2}}
-
-For the SPL2 data (only the \texttt{\_\^{}{[}shift{]}} sentence-transitions) we calculated the L2 effect (i.e.~the difference between writing in L2 and L1).
-The results can be found in Table \ref{tab:l2effect}.
-The results show longer hesitation duration and more pauses across all transition locations when writing in L2.
-Only within-word transitions did show negligible evidence for an increased hesitation duration in L2.
-
-\begin{table}[b]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:l2effect}Mixture model estimates for language effect.
-Cell means are shown the slowdown for hesitant transitions when writing in L1 and L2 and the probability of hesitant transitions.
-The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}
-
-\small{
-
-\begin{tabular}{lrrrr}
-\toprule
-Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 1,281 [1,082, 1,496] & 1,802 [1,557, 2,073] & 0.31 [0.17, 0.45] & > 100\\
-\ \ \ before word & 289 [254, 327] & 560 [507, 617] & 0.47 [0.39, 0.55] & > 100\\
-\ \ \ within word & 120 [53, 222] & 202 [156, 258] & 0.27 [-0.07, 0.55] & 0.68\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .78 [.69, .86] & .92 [.87, .96] & 1.23 [0.52, 1.97] & > 100\\
-\ \ \ before word & .51 [.41, .62] & .89 [.83, .93] & 2.02 [1.4, 2.66] & > 100\\
-\ \ \ within word & .04 [.02, .07] & .15 [.10, .22] & 1.52 [0.78, 2.29] & > 100\\
+C2L1 & -1,637 (72) & -785 (60) & -37,517 (798)\\
+CATO & -1,782 (75) & -1,299 (71) & -37,673 (881)\\
+GE2 & -2,427 (92) & -2,312 (89) & -38,843 (616)\\
+LIFT & -5,105 (127) & -2,111 (99) & -74,851 (1,624)\\
+PLanTra & -2,384 (90) & -1,381 (80) & -33,823 (526)\\
+SPL2 & -1,181 (64) & -1,713 (69) & -28,717 (418)\\
 \bottomrule
@@
 \begin{tablenotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\normalsize{\textit{Note.} $\Delta\widehat{elpd}$ = difference in predictive performance -- estimated as expected log pointwise predictive density}
 \end{tablenotes}
@@
 
-\newpage
-
-\hypertarget{masking-effect-cato-gunnexp2}{%
-\section{Masking effect (CATO, GUNNEXP2)}\label{masking-effect-cato-gunnexp2}}
-
-Studies associated with two datasets (CATO, GUNNEXP2) investigated to what extent masking the previously written text affects keystroke behaviour.
-The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}.
-There was some evidence for longer before-sentence hesitation durations when the text was masked in non-dyslexic writers.
-Evidence for all other comparisons was negligible.
-
-\begin{landscape}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:maskingeffect}Mixture model estimates for masking effect.
-Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of hesitant transitions.
-The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Hesitation duration &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & 2,231 [1,543, 3,111] & 1,680 [1,239, 2,205] & -0.25 [-0.64, 0.13] & 0.44\\
-\ \ \ before sentence & CATO & non dyslexic & 1,733 [1,241, 2,325] & 1,502 [1,134, 1,942] & -0.12 [-0.47, 0.22] & 0.23\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & 1,124 [954, 1,307] & 1,418 [1,218, 1,634] & 0.21 [0.05, 0.36] & 2.59\\
-\ \ \ before word & CATO & dyslexic & 444 [405, 485] & 407 [369, 447] & -0.06 [-0.13, 0.01] & 0.14\\
-\ \ \ before word & CATO & non dyslexic & 345 [312, 381] & 326 [295, 361] & -0.04 [-0.11, 0.04] & 0.06\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & 363 [325, 403] & 379 [341, 420] & 0.03 [-0.05, 0.11] & 0.05\\
-\ \ \ within word & CATO & dyslexic & 278 [166, 400] & 177 [116, 252] & -0.24 [-0.52, 0.06] & 0.57\\
-\ \ \ within word & CATO & non dyslexic & 409 [244, 593] & 392 [230, 591] & -0.03 [-0.42, 0.38] & 0.2\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & 182 [105, 279] & 209 [135, 306] & 0.08 [-0.24, 0.4] & 0.19\\
-Hesitattion probability &  &  &  &  &  & \\
-\ \ \ before sentence & CATO & dyslexic & .60 [.38, .79] & .80 [.60, .93] & 0.2 [-0.06, 0.46] & 0.45\\
-\ \ \ before sentence & CATO & non dyslexic & .55 [.34, .76] & .68 [.47, .84] & 0.13 [-0.16, 0.4] & 0.21\\
-\ \ \ before sentence & GUNNEXP2 & non dyslexic & .88 [.82, .93] & .90 [.85, .94] & 0.16 [-0.46, 0.79] & 0.36\\
-\ \ \ before word & CATO & dyslexic & .75 [.64, .85] & .72 [.59, .82] & -0.04 [-0.19, 0.12] & 0.09\\
-\ \ \ before word & CATO & non dyslexic & .64 [.51, .75] & .62 [.49, .74] & -0.02 [-0.19, 0.16] & 0.09\\
-\ \ \ before word & GUNNEXP2 & non dyslexic & .47 [.40, .54] & .45 [.38, .52] & -0.06 [-0.46, 0.35] & 0.21\\
-\ \ \ within word & CATO & dyslexic & .07 [.03, .12] & .11 [.06, .19] & 0.05 [-0.02, 0.13] & 0.08\\
-\ \ \ within word & CATO & non dyslexic & .03 [.02, .06] & .03 [.02, .06] & 0 [-0.03, 0.03] & 0.01\\
-\ \ \ within word & GUNNEXP2 & non dyslexic & .05 [.03, .08] & .04 [.03, .06] & -0.29 [-0.88, 0.28] & 0.48\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
+\hypertarget{transition-location-effect}{%
+\subsection{Transition location effect}\label{transition-location-effect}}
+
+Figure \ref{fig:mixmodel} illustrates the mixture model results for every dataset: for each interkey interval we obtained two posterior probability distributions, one associated with fluent transitions between keys and another for interkey intervals where upstream difficulty resulted in hesitant interkey intervals.
+The hesitation probability -- the probability of interkey intervals associated with hesitations -- is indicated as the height of the distribution shown in yellow (on the right of each panel).
+The transition locations are characterised by the hesitation parameters: as we illustrate the mixture model, the distributions of fluent transitions -- indicated in grey -- are constant across transition location (within each dataset) but their height differs as the probability of fluent interkey intervals is the inverse of the height of hesitant interkey intervals; i.e.~\(1 - \theta\).
+Within-word interkey intervals have a negligibly small hesitation probability barely visible in the visualisation.
+Hesitations before words and sentences are both roughly equally likely than fluent transitions.
+In other words, half of the time participants did not pause before a word or sentence; we explore this below.
+However, when hesitations occurred, these are longer at before-sentence interkey intervals compared to before word interkey intervals.
+The slowdown for hesitant interkey intervals -- represented earlier as \(\delta\) -- is the distance between the distribution of fluent interkey intervals (\(\beta\)) shown in grey and the distribution of hesitant interkey intervals shown in yellow.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/mixmodel-1}
 
@@
 
-\end{ThreePartTable}
-\end{center}
-\end{landscape}
-
-\newpage
-
-\hypertarget{posterior-parameter-estimates}{%
-\section{Posterior parameter estimates}\label{posterior-parameter-estimates}}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics{figures/psplots1constr.pdf}
-\caption{Posterior parameter distribution}
+\caption{Mixture model plot for each transition location.
+Shown are the distributions for both mixture components (fluent and hesitant interkey intervals) and their weighting.
+Interkey intervals are shown in seconds on log scaled intervals.}\label{fig:mixmodel}
 \end{figure}
-\newpage
-\begin{figure}[!htb]
-\ContinuedFloat
-\captionsetup{list=off,format=cont}
-\centering
-\includegraphics{figures/psplots2constr.pdf}
-\label{fig:fullps1}
-\caption{Posterior parameter distribution}
+
+Figure \ref{fig:mixmodel} highlighted how model parameters relate to each other in pairs of distributions but to compare the effect of transition location on model parameters we need to compare the posterior distribution of the three parameters of interest individually.
+Therefore Figure \ref{fig:crossstudypost2} visualises the same posterior as Figure \ref{fig:mixmodel} but expressed directly in terms of posterior probability distributions of the individual mixture-model parameters.
+These are shown for all datasets by transition location.
+Although models were fitted with all dataset-specific conditions, we aggregated the posterior across conditions\footnote{We aggregated across pre and post tests for the PLanTra dataset as well as genre and topic of the LIFT data set.
+We demonstrate in Appendix \ref{pre-post-test-plantra} and \ref{genre-effect-lift}, respectively, that there is negligible evidence for differences between these conditions.}, and excluded conditions that might confound comparisons\footnote{We removed the masked writing condition in the GE2 and CATO, the dyslexic group in the CATO data set, and L2 writing in the SPL2 data set.
+There was evidence for L2 writing effects (see Appendix \ref{l2-effect-spl2}); evidence for marking effects was weak (see Appendix \ref{masking-effect-cato-ge2}).}.
+For posteriors of all conditions within datasets see Appendix \ref{posterior-parameter-estimates}.
+The resulting posterior allows us to examine differences and similarities associated with transition locations for all datasets.
+In particular we can see how transition location in the text is associated with hesitant interkey intervals and hesitation probability.
+For completeness we also report the estimate for fluent interkey intervals.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/crossstudypost2-1}
+
+}
+
+\caption{Mixture model parameter estimates across studies.
+Distributions of parameter estimates are represented as posterior mean and 95\% probability interval (PI).
+Interkey-interval duration measures are shown in msecs and hesitation probability is shown in proportions.
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.}\label{fig:crossstudypost2}
 \end{figure}
 
-\newpage
-
-\hypertarget{pre-post-test-plantra}{%
-\section{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
-
-The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}. Evidence for all comparisons was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
-Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant transitions and the probability of hesitant transitions.
-The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 879 [723, 1,052] & 909 [740, 1,100] & -0.03 [-0.23, 0.17] & 0.1\\
-\ \ \ before word & 438 [395, 485] & 434 [389, 483] & 0.01 [-0.08, 0.1] & 0.05\\
-\ \ \ within word & 245 [180, 323] & 269 [201, 353] & -0.06 [-0.3, 0.18] & 0.14\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .58 [.50, .66] & .54 [.47, .62] & 0.15 [-0.26, 0.58] & 0.27\\
-\ \ \ before word & .56 [.50, .61] & .49 [.43, .55] & 0.27 [-0.04, 0.59] & 0.72\\
-\ \ \ within word & .11 [.08, .14] & .09 [.07, .12] & 0.13 [-0.27, 0.54] & 0.26\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\newpage
-
-\hypertarget{genre-effect-lift}{%
-\section{Genre effect (LIFT)}\label{genre-effect-lift}}
-
-From the LIFT data, we compared the difference between genres, i.e.~when writing an informative text as opposed to writing an argumentative text.
-The results are shown in Table \ref{tab:genreeffect}.
-Cellmeans and differences were average across writing topic.
-Evidence for a genre comparisons was negligible.
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:genreeffect}Mixture model estimates for genre effect.
-Cell means are shown for argumentative and informative texts in msecs for the hesitation slowdown and the probability of hesitant transitions.
-The effect for genre is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions.
-95\% PIs in brackets.}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
-\toprule
-Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
-\midrule
-\endhead
-Hesitation duration &  &  &  & \\
-\ \ \ before sentence & 373 [301, 454] & 411 [330, 504] & -0.07 [-0.33, 0.19] & 0.16\\
-\ \ \ before word & 287 [250, 326] & 284 [248, 323] & 0.01 [-0.15, 0.16] & 0.09\\
-\ \ \ within word & 199 [118, 306] & 246 [155, 361] & -0.12 [-0.6, 0.36] & 0.29\\
-Hesitation probability &  &  &  & \\
-\ \ \ before sentence & .35 [.28, .43] & .34 [.27, .42] & 0.05 [-0.57, 0.73] & 0.36\\
-\ \ \ before word & .46 [.39, .54] & .48 [.41, .55] & -0.05 [-0.51, 0.44] & 0.25\\
-\ \ \ within word & .07 [.05, .10] & .07 [.05, .10] & -0.02 [-0.81, 0.76] & 0.46\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\newpage
-
-\hypertarget{transition-location-effect-constrained-mixture-model}{%
-\section{Transition location effect (constrained mixture model)}\label{transition-location-effect-constrained-mixture-model}}
-
-\begin{center}
-\begin{ThreePartTable}
-
-\begin{TableNotes}[para]
-\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
-\end{TableNotes}
-
-\footnotesize{
-
-\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:loceffectcon}Effect of transition location on keystroke intervals.
-Differences are shown on log scale (for durations) and logit scale for hesitation probability.
-95\% PIs in brackets.}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:loceffectcon} continued}}\\
-\toprule
- \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hesitation slowdown} & \multicolumn{2}{c}{Hesitation probability} \\
-\cmidrule(r){1-1} \cmidrule(r){2-3} \cmidrule(r){4-5}
-Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
-\midrule
-\endhead
-\textbf{C2L1} &  &  &  & \\
-\ \ \ before sentence vs word & 0.24 [0.02, 0.48] & 1.11 & 0.44 [-0.56, 1.66] & 0.68\\
-\ \ \ before vs within word & -0.03 [-0.14, 0.09] & 0.07 & 3.95 [3.54, 4.38] & > 100\\
-\textbf{CATO (non-dyslexic, unmasked)} &  &  &  & \\
-\ \ \ before sentence vs word & 1.27 [0.98, 1.55] & > 100 & -0.36 [-1.41, 0.69] & 0.67\\
-\ \ \ before vs within word & -0.1 [-0.38, 0.21] & 0.2 & 4.05 [3.2, 4.91] & > 100\\
-\textbf{GUNNEXP2 (unmasked)} &  &  &  & \\
-\ \ \ before sentence vs word & 0.9 [0.76, 1.04] & > 100 & 2.15 [1.6, 2.75] & > 100\\
-\ \ \ before vs within word & 0.44 [0.17, 0.69] & 26.78 & 2.76 [2.24, 3.28] & > 100\\
-\textbf{LIFT} &  &  &  & \\
-\ \ \ before sentence vs word & 0.21 [-0.05, 0.48] & 0.5 & -0.53 [-1.04, 0.08] & 1.61\\
-\ \ \ before vs within word & 0.17 [-0.32, 0.53] & 0.41 & 2.48 [1.81, 3.22] & > 100\\
-\textbf{PLanTra} &  &  &  & \\
-\ \ \ before sentence vs word & 0.57 [0.4, 0.74] & > 100 & 0.16 [-0.25, 0.57] & 0.28\\
-\ \ \ before vs within word & 0.36 [0.16, 0.56] & 55.53 & 2.3 [1.9, 2.7] & > 100\\
-\textbf{SPL2 (L1)} &  &  &  & \\
-\ \ \ before sentence vs word & 1.17 [1.02, 1.3] & > 100 & 1.23 [0.59, 1.88] & > 100\\
-\ \ \ before vs within word & 0.48 [0.16, 0.76] & 9.07 & 3.3 [2.56, 4.08] & > 100\\
-\bottomrule
-\addlinespace
-\insertTableNotes
-\end{longtable}
-
-}
-
-\end{ThreePartTable}
-\end{center}
-
-\newpage
-
-\hypertarget{simulation}{%
-\section{Simulation}\label{simulation}}
-
-A possible concern with these results -- substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit.
-We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models.
-To address this concern we repeated a comparison for a single distribution model and a two-distributions mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart.
-In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.
-
-In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}).
-This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simmog}
-\text{y} \sim\text{ } & \theta \cdot \text{logN}(\beta + \delta, \sigma^2_1) +\\
-& (1 - \theta) \cdot \text{logN}(\beta, \sigma^2_2)\\
-\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
-        & \sigma_{1}^2 > \sigma_{2}^2
-\end{aligned}
-\end{equation}
-
-This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants).
-The model includes two log-normal distributions with a mixing proportion \(\theta\) of which the distribution of shorter values has a mean of \(\beta\) and a standard deviation \(\sigma^2_2\); the second distribution of longer values is constrained to have a mean that is larger by a factor of \(\delta\) and has a larger standard deviation \(\sigma^2_1\).
-
-The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.
-
-\begin{equation}
-\begin{aligned}
-\label{eq:simuv}
-\text{y} \sim\text{ }& \text{logN}(\beta, \sigma^2)\\
-\text{constraint: } & \sigma^2>0
-\end{aligned}
-\end{equation}
-
-Again, this model is a simplified version of the single distribution models used in the main analysis.
-The model assumes a log-Gaussian distribution with a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).
-
-The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}.
-The simulated data are visualised in Figure \ref{fig:simdata}.
-Parameter values were chosen so that the simulated data are roughly similarly distributed to keystroke transitions.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
-
-}
-
-\caption{Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process.
-The x-axis showing the outcome y was log-scaled for visability.}\label{fig:simdata}
-\end{figure}
-
-For each of these two data sets we simulated 1,000 observations.
-We run 2 models -- a two-distribution mixture model and a single-distribution model -- each for both data sets.
-Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup.
-Estimates with 95\% probability intervals are shown in Table \ref{tab:simparam}.
-The parameters are shown by type of data generating process along with the true parameter values.
-Parameter value estimates are shown by type of Bayesian model.
-The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process.
-Particularly the mixing proportion \(\theta\) and the slowdown parameter \(\delta\) were not uncovered at all by they single-distribution model.
+Figure \ref{fig:crossstudypost2} shows largely the same patterns (with caveats) for all datasets.
+We summarise the evidence for differences between transitions locations in Table \ref{tab:loceffectconbf}.
+We found that hesitations were more frequent -- higher hesitation probability -- before words than within words (BFs \textgreater{} 100 for all datasets) but only for half of the datasets hesitation durations were longer (GE2: BF = 27; PLanTra: BF = 56; SPL2: BF = 9.1; negligible evidence for C2L1: BF = 0.1; CATO: BF = 0.2; LIFT: BF = 0.4).
+Hesitations were equally likely before sentences and words (C2L1 and CATO: BFs = 0.7; PLanTra: BF = 0.3; LIFT: BF = 1.6; except for GE2 and SPL2; both BFs \textgreater{} 100) but longer (all BFs \textgreater{} 100, except for C2L1: BF = 1.11 and LIFT = 0.5).
+Full results and parameter estimates can be found in Appendix \ref{transition-location-effect-1}.
 
@@
 
-\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}
-
-\begin{tabular}{lrr}
+\caption{\label{tab:loceffectconbf}BFs for the effect of transition location on interkey intervals shown for the hesitation duration and hesitation probability.
+Comparisons for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.}
+
+\begin{tabular}{lp{3cm}p{3cm}p{3cm}p{3cm}}
 \toprule
- & \multicolumn{2}{c}{Estimate with 95\% PI} \\
-\cmidrule(r){2-3}
-Parameter with true value & \multicolumn{1}{c}{Two-distributions model} & \multicolumn{1}{c}{Single-distributions model}\\
+ & \multicolumn{2}{c}{Hesitation duration} & \multicolumn{2}{c}{Hesitation probability} \\
+\cmidrule(r){2-3} \cmidrule(r){4-5}
+Dataset & before sentence vs word & before vs within word & before sentence vs word & before vs within word\\
 \midrule
-Two-distributions data &  & \\
-\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
-\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
-\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
-\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
-\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
-Single-distribution data &  & \\
-\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
-\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-We used LOO-CV to compare the predictive performance of the two models for each data generating process.
-The model comparisons can be found for each data generating process in Table \ref{tab:loossim}.
-The results rule out the possibility that the mixture model does always lead to higher predictive performance.
-Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a single-distribution process.
-However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
-In fact, the ratio of \(\Delta\widehat{elpd}\) and its standard error -- a \emph{z}-score approximation -- reveals a value of 11.6 which is strong evidence in in favour of the mixture model compared to the single-distribution model (Sivula et al., 2020).
-In comparison, for the single-distribution data, the single-distribution model was negligibly better than the two-distribution mixture model which was evidence by a value of 0.77 for the ratio of \(\Delta\widehat{elpd}\) and its standard error.
-Thus, as the difference \(\Delta\widehat{elpd}\) between the model is negligible, the single distribution is preferred by the law of parsimony.
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:loossim}Model comparisons by data set.
-The top row shows the models with the highest predictive performance for each data generating process.
-Standard error is shown in parentheses.}
-
-\begin{tabular}{lrr}
-\toprule
-Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
-\midrule
-Data: Two-distribution mixture process &  & \\
-\ \ \ Two-distribution mixture model & -- & -6,068 (41)\\
-\ \ \ Single-distribution model & -191.3 (16.5) & -6,259 (38)\\
-Data: Single-distribution process &  & \\
-\ \ \ Single-distribution model & -- & -5,030 (24)\\
-\ \ \ Two-distribution mixture model & -0.5 (0.7) & -5,030 (24)\\
+C2L1 & 1.11 & 0.07 & 0.68 & > 100\\
+CATO & > 100 & 0.2 & 0.67 & > 100\\
+GE2 & > 100 & 26.78 & > 100 & > 100\\
+LIFT & 0.5 & 0.41 & 1.61 & > 100\\
+PLanTra & > 100 & 55.53 & 0.28 & > 100\\
+SPL2 & > 100 & 9.07 & > 100 & > 100\\
 \bottomrule
@@
 \begin{tablenotes}[para]
-\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
+\normalsize{\textit{Note.} BF = evidence in favour of the alternative hypothesis over the null hypothesis calculated using the Savage-Dickey method (Dickey et al., 1970; Wagenmakers et al., 2010).
+BFs greater than 3 indicate moderate evidence, greater than 10 strong evidence, and greater than 30 very strong evidence for a statistically meaningful effect (Jeffreys, 1961; Lee \& Wagenmakers, 2014).
+A BF smaller than 0.33 suggests evidence in favour of a null effect (Dienes, 2014) and a BF around 1 is inconclusive evidence.}
 \end{tablenotes}
@@
 
+To reiterate, for most datasets we found that pauses before sentences are longer but not always more likely than before words (except for SPL2 and GE2).
+In fact, LIFT showed some indication that pauses were less likely before sentences than words.
+This is interesting because it is generally believed that pausing behaviour is associated with linguistic boundaries such that more and longer pauses are predicted for interkey intervals at larger linguistic edges (see Introduction section).
+It is less clear in which contexts pre-sentence pausing is more likely than pausing before words.
+
+Differences between datasets could, to some extent, be explained by the fact that two datasets (PLanTra, LIFT) defined before-sentence transitions as the sum of transitions between space, shift and the sentence-initial character, while the remaining data sets measured before-sentence transitions as the interval between space key and shift.
+To test to what extent complex key combinations affects the parameter estimates, we addressed this possibility for the SPL2 dataset and showed that including the character following the shift key substantially increased the hesitation probability for sentence-initial transitions (BFs \textgreater{} 100) but it did not affect the duration hesitant interkey intervals (BFs \textless{} 0.3); see Appendix \ref{key-combination-effect}.
+This conflicts with the earlier finding that before-sentence hesitations were not more likely than before-word hesitations for datasets that did include the character following shift (PLanTra, LIFT).
+In other words, patterns in our results cannot be explained on the basis of how sentence-initial key transitions were operationalised.
+
+Finally, Figure \ref{fig:crossstudypost2} highlights that fluent interkey intervals in two datasets (C2L1, CATO) were substantially longer than for other datasets.
+This is presumably reflecting that these data were sampled from the youngest and least experienced population of writers in our data pool.
+Interestingly, however, hesitation duration and hesitation probability followed the same patterns as the majority of datasets.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The classical serial view (Flower \& Hayes, 1980) characterises composing spontaneous multi-sentence text as a sequence of planning and execution cycles.
+Text is therefore produced as a series of production bursts bounded by pauses.
+The parallel model of writing, in contrast, sees planning, at least in competent writers, as occurring to a large extent in parallel with output (Olive, 2014).
+The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is that pause duration doesn't necessarily correlate with planning necessary for next output.
+We implemented these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing.
+Fitting these models with six sets of keystroke data from free text production tasks -- with a total of 953 participants producing 5,223,280 interkey intervals -- we have found compelling and consistent statistical evidence in favour of the parallel view.
+In all cases, the predictive performance of the two-distribution mixture-models (consistent with the parallel view of writing) outperformed the single-distribution models.
+
+The two distribution model assumes two data generating processes, one associated with the immediate demands of motor planning required for the production of a keystroke, and one that captures all upstream processing from message to abstract letter representation.
+A direct implication of parallel, cascaded processing is that the distribution over interkey intervals is not additive -- it is not a combination of time to motor plan plus time for upstream processing.
+The interval between keypresses is \emph{either} determined by motor execution \emph{or} by processing that occurred upstream of motor execution (in cases where provision of output from the upstream processes lags behind the rate at which it can be output by the fingers).
+Under a serial account -- where all planning necessary for production of next-text occurs in the interkey interval immediately before output of -- interkey interval will be the sum of times for each processing step required for generation of that text.
+Both accounts predict longer mean interkey intervals at higher-order text edges (interkey intervals greater before sentences than before mid-sentence words, and before mid-sentence words than before mid-word characters).
+This has been consistently observed in a number of previous studies (Deane \& Zhang, 2015; Mohsen, 2021; Spelman Miller, 2006; Spelman Miller et al., 2008; Torrance, Rønneberg, et al., 2016; Wengelin, 2002; Xu \& Qi, 2017).
+However, the serial account does not permit the possibility of a distribution of short interkey intervals at higher-order edges: It strongly predicts that \emph{all} sentence boundaries will be associated with interkey intervals that are longer than those required just for the motor planning of the next character.
+Under the parallel-processing account it is very likely that some -- perhaps a large majority -- of interkey intervals before sentences will be longer.
+However there will be a distinct subset that is drawn from a distribution associated just with interkey intervals necessary for motor execution.
+
+This is what we found.
+At all linguistic edges we found a distribution of interkey intervals at durations to motor execution at durations substantially below a threshold that might plausibly be necessary for, for example, lexical / orthographic retrieval (Grudin \& Larochelle, 1982; Ostry, 1983; Terzuolo \& Viviani, 1980; Van Waes et al., 2021).
+For mid-word interkey intervals, as might be expected, the mixture model predicted that the vast majority of interkey intervals fall within this distribution.
+Across the six datasets that we sampled, the central tendency for the duration of mid-word interkey intervals was broadly similar to mean mid-word interkey intervals observed in previous studies.
+Importantly, however, these short interkey intervals were also present at higher-order text edges.
+Across all datasets we found, at minimum, a non-trivial minority of sentence-initial interkey intervals were sampled from the short distribution (i.e.~with durations that could only be associated with motor execution).
+In two cases -- for writers sampled in the PLanTra and Lift datasets -- the majority of interkey intervals fell within this distribution.
+Thus, although as might be expected there was a considerably increased tendency for longer interkey intervals at higher order boundaries: Planning associated with planning a new sentence is, probabilistically, much more likely to spill beyond output of the previous sentence and therefore affect duration of the sentence-initial interkey interval.
+However the presence of a non-trivial proportion of interkey intervals with a duration that is inconsistent with anything other than motor execution can only be explained with recourse to theories that permit preparation in parallel with preceding output.
+
+Our results showed that the hesitation probabilities are largely identical before sentences and before words, which suggests that writers frequently plan the next word and even sentence in parallel to production -- consistent with the parallel view of writing -- which mitigates the need to pause to plan up-coming linguistic information.
+
+This parallelism was evident even in the youngest writers in our data pool (C2L1; mean age: 11.8 years).
+As might be expected for children who compose predominantly by handwriting and whose typing skills will still be developing, motor execution rate (mean duration of fluent transitions) was substantially slower compared to the more experienced typists in other datasets.
+In all other respects, however, findings for this dataset were similar to that of other datasets and consistent with parallel processing.
+A two-distribution mixture model provided best fit to these data.
+At both before-word and before-sentence locations a non-trivial number of interkey intervals were within the short distribution - i.e.~at durations too short to permit lexical retrieval or syntactic planning.
+Probability of longer intervals was greater before-sentence and before-word transitions compared to mid-word transitions.
+This is interesting because, for example, Olive (2014) described a parallel model that operates in a serial fashion for inexperienced or struggling writers, explained on the basis of task demands that reduce the ability to plan in parallel to processing which leads to a serialisation of planning units.
+However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers: Pauses were both equally long and likely at before-word and before-sentence transitions but more likely compared to mid-word transitions.
+Also for L2 writers (PLanTra) we observe pausing patterns that do not resemble a serialised writing process as evidenced by a large number of fluent transitions at word and sentence boundaries.
+
+We observed two more differences among datasets that are worth highlighting.
+Writers sampled in the GE2 and SPL2 datasets were more likely to pause before sentences than before words (see also Medimorec \& Risko, 2017; Wengelin, 2002).
+In the remaining datasets from writers who were either younger (and not educationally-selected) or writing in second language, hesitation was equally probable pre-word and pre-sentence.
+This is most readily attributed simply to increased demands of lexical and / or orthographic processing in these samples.
+
+Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production.
+(2) pause durations did not vary across transition locations in the LIFT dataset.
+The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning novel contents and text structure: the structure of one the source texts might have been used during synthesis (\enquote{structural isomorphism,} Hidi \& Anderson, 1986).
+However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents.
+
+We are, therefore, interpreting the superior predictive performance of the two-distribution mixture model as strong evidence for parallelism in written composition.
+An alternative, more prosaic explanation might be that two-distribution models outperform single distributions simply because they estimate more parameters; two distributions will always provide better fit than one distribution models due just to overfitting.
+There are two related reasons not to prefer this explanation.
+First, the cross-validation methods that we used to compare model predictive performance favour more parsimonious models and are therefore robust to overfitting.
+Second, in the analysis described in Appendix \ref{simulation} we directly tested the hypothesis that modelling data randomly sampled from a single log-normal distribution as a two-distribution mixture will provide better fit than modelling as a single distribution.
+We found that the two-distribution mixture model was successful at uncovering the parameter values of the data generated with a two-distributions mixture process; we observed the same for the single-distribution model applied to the data generated with a single distribution process.
+Parameters were not successfully uncovered when we switched model type and dataset.
+Cross-validation rendered a substantially higher predictive performance for the mixture model compared to the single-distribution model for data sampled from a two-distributions mixture process, but not when applied to data based on a single-distribution process.
+Therefore we can rule out the possibility that the statistical support for the two-distributions mixture-models can be explained on the basis of their larger number of parameters (i.e.model overfitting).
+
+Throughout this paper we have been attributing longer (second-distribution) interkey intervals to spill-over from upstream pre-motor processing that could not be completed in its entirety in parallel with previous output.
+These longer inter-key intervals may also, on occasion, be associated with activity that is initiated after the preceding process and is entirely bounded by the duration of the pause.
+We are thinking here specifically of the tendency, particularly at higher order text boundaries, for writers to glance back into their existing text (Alamargot et al., 2010; Beers et al., 2010; Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Van Waes et al., 2010).
+Torrance, Johansson, et al.
+(2016) found that lookbacks appear with a frequency of 45\% before sentences, 12\% before words and 5\% within words of which 36\% were associated with sustained reading but mostly less patterned forward and backward saccades between words (\enquote{hopping,} see also Chukharev-Hudilainen et al., 2019).
+There are broadly two reasons why writers might look back into their existing text.\footnote{Eye movement between key presses in writing-from-sources tasks will also be associated with reading source texts.
+Note, however, that for two datasets (PLanTra, LIFT) that involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.} Lookback may be associated with monitoring just-produced text for errors.
+In the present study we excluded interkey intervals that preceded deletion or cursor movement, and also intervals at timescales that might be associated with strategic reviewing of large spans of text (e.g., reading text-already-written from top to bottom).
+However it may be that some longer intervals were associated with monitoring that failed to find errors that needed correcting.
+A second function for lookback may be to support planning next-text (Hayes, 1996; Johansson et al., 2010; Torrance, Johansson, et al., 2016).
+The preceding text, and particularly the preceding sentence, is a rich source of cues to support retrieval of both content and lexical items necessary to prepare what to write next.
+This will become particularly important where upstream processing lags behind the rate at which it can be output, and one or more of the buffers necessary to resolve this \enquote{temporal friction} (Van Galen, 1991) becomes overloaded.
+Comparison of conditions unmasked and masked text conditions in the CATO and GE2 datasets -- conditions in which writers could and could not read the text that they had already produced -- suggested that, if anything, proportion of longer interkey intervals at sentence boundaries was greater when text was masked (details in Appendix \ref{masking-effect-cato-ge2}).
+This finding is consistent both with lookback being only a minor contributor to interkey interval and, where it does occur, serving a rapid memory-refresh function.
+There is also some evidence, at least in the context of handwritten production, that eye movements back into already-written text themselves occasionally occur in parallel with output (Alamargot et al., 2007).
+
+The findings we report in this paper are, we argue, the first robust evidence for parallelism in written composition of multi-sentence texts.
+The conclusion that text production processes are fundamentally parallel will not come as a surprise to most researchers.
+Parallel processing has a long history in theories of speech production (Chang et al., 2006; Dell, 1986; Dell et al., 1999; Dell \& O'Seaghdha, 1992; Garrett, 1975; Levelt, 1989).
+The transitory nature of activation and storage during fluent spoken production prevent, for example, generating phonological code a long time before it is required to drive articulation (Christiansen \& Chater, 2016).
+Spoken production is \enquote{just-in-time}.
+Perhaps because, at least in principle, there is not the same push to fluency in written production -- in principle writers can pause at any point with no detriment to communication -- writing extended text has tended to be seen as fundamentally serial, with pauses to prepare next-text followed by bursts of output during which this is executed (Flower \& Hayes, 1980).
+Our findings suggest that this is not the case.
+The language production processes that underlie text production follow the same just-in-time principle as speech.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+The parallel view of writing, in contrast to the serial view, claims that writers do not necessarily pause to plan the upcoming language unit but plan in parallel with output.
+In this paper we have demonstrated, across six datasets, that statistical models premised on a parallel-processing understanding of planning in written composition outperform models premised on serial processing.
+We also found that interkey intervals across word and even sentence boundaries are frequently at timescales that are not consistent with planning of upcoming linguistic information.
+This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g.~young / L2 writers, students) and across different composition tasks.
+Our argument that written composition -- a very common form of language production -- is a largely parallel process is probably uncontroversial.
+However, the findings we presented in this paper constitute the first robust evidence in support of this claim.
+
 \newpage
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\begingroup
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Alamargot2007}{}}%
+Alamargot, D., Dansac, C., Chesnet, D., \& Fayol, M.
+(2007).
+{Parallel Processing Before and After Pauses: A Combined Analysis of Graphomotor and Eye Movements During Procedural Text Production}.
+In M.
+Torrance, L.
+Van Waes, \& D.
+Galbraith (Eds.), \emph{Writing and cognition} (pp.
+11--29).
+BRILL.
+\url{https://doi.org/10.1163/9781849508223_003}
+
 \leavevmode\vadjust pre{\hypertarget{ref-alamargot2010using}{}}%
@@
 Baaijen, V. M., Galbraith, D., \& De Glopper, K. (2012). Keystroke analysis: Reflections on procedures and measures. \emph{Written Communication}, \emph{29}(3), 246--277.
+
+\leavevmode\vadjust pre{\hypertarget{ref-baayen2008analyzing}{}}%
+Baayen, R. H. (2008). \emph{Analyzing linguistic data: A practical introduction to statistics using {R}}. Cambridge University Press.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-behmer2016crunching}{}}%
-Behmer Jr., L.
-P., \& Crump, M.
-J.
-C.
-(2016).
-Crunching big data with fingertips: How typists tune their performance toward the statistics of natural language.
-In M.
-N.
-Jones (Ed.), \emph{Big data in cognitive science} (Vol.
-1, pp.
-329--345).
-Psychology Press.
-\url{https://doi.org/10.4324/9781315413570-27}
-
 \leavevmode\vadjust pre{\hypertarget{ref-bock2014syntactically}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-brysbaert2021improving}{}}%
+Brysbaert, M., Sui, L., Duyck, W., \& Dirix, N.
+(2021).
+Improving reading rate prediction with word length information: Evidence from {D}utch.
+\emph{Quarterly Journal of Experimental Psychology}, \emph{74}(11), 2013--2018.
+
 \leavevmode\vadjust pre{\hypertarget{ref-carpenter2016stan}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chanquoy1996writing}{}}%
-Chanquoy, L., Foulin, J.-N., \& Fayol, M.
-(1996).
-Writing in adults: {A} real-time approach.
-In G.
-Rijlaarsdam, H.
-Van den Bergh, \& M.
-Couzijn (Eds.), \emph{Theories, models and methodology in writing research} (pp.
-36--44).
-Amsterdam University Press.
-
 \leavevmode\vadjust pre{\hypertarget{ref-chenoweth2001fluency}{}}%
 Chenoweth, N. A., \& Hayes, J. R. (2001). Fluency in writing: Generating text in {L1} and {L2}. \emph{Written Communication}, \emph{18}(1), 80--98.
-
-\leavevmode\vadjust pre{\hypertarget{ref-chenoweth2003inner}{}}%
-Chenoweth, N. A., \& Hayes, J. R. (2003). The inner voice in writing. \emph{Written Communication}, \emph{20}(1), 99--118.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chukharev2014pauses}{}}%
-Chukharev-Hudilainen, E. (2014). Pauses in spontaneous written communication: {A} keystroke logging study. \emph{Journal of Writing Research}, \emph{6}(1), 61--84.
-
 \leavevmode\vadjust pre{\hypertarget{ref-chukharev2019combined}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-conijn2021timecourse}{}}%
-Conijn, R., \& Torrance, M. (Eds.). (2023). Timecourse method. \emph{Reading and Writing}. \url{https://link.springer.com/collections/gedbaiibja}
-
-\leavevmode\vadjust pre{\hypertarget{ref-connelly2012predicting}{}}%
-Connelly, V., Dockrell, J.
-E., Walter, K., \& Critten, S.
-(2012).
-Predicting the quality of composition and written language bursts from oral language, spelling, and handwriting skills in children with and without specific language impairment.
-\emph{Written Communication}, \emph{29}(3), 278--302.
+Conijn, R., \& Torrance, M. (Eds.). (2023). \emph{Timecourse method}. \url{https://link.springer.com/collections/gedbaiibja}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-dell1986}{}}%
+Dell, G. S. (1986). A spreading-activation theory of retrieval in sentence production. \emph{Psychological Review}, \emph{93}(3), 283--321.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dell1999connectionist}{}}%
+Dell, G. S., Chang, F., \& Griffin, Z. M. (1999). Connectionist models of language production: Lexical access and grammatical encoding. \emph{Cognitive Science}, \emph{23}(4), 517--542.
+
 \leavevmode\vadjust pre{\hypertarget{ref-dell2007case}{}}%
@@
 \emph{Journal of Memory and Language}, \emph{56}(4), 490--520.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dell1992stages}{}}%
+Dell, G. S., \& O'Seaghdha, P. G. (1992). Stages of lexical access in language production. \emph{Cognition}, \emph{42}, 287--314.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dickey1970weighted}{}}%
+Dickey, J.
+M., Lientz, B.
+P., et al.
+(1970).
+The weighted likelihood ratio, sharp hypotheses about chances, the order of a markov chain.
+\emph{The Annals of Mathematical Statistics}, \emph{41}(1), 214--226.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dienes2014using}{}}%
+Dienes, Z. (2014). Using {B}ayes to get the most out of non-significant results. \emph{Frontiers in Psychology}, \emph{5}(781), 1--17.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-foulin1998extent}{}}%
-Foulin, J.-N.
-(1998).
-To what extent does pause location predict pause duration in adults' and children's writing?
-\emph{Cahiers de Psychologie Cognitive/Current Psychology of Cognition}, \emph{17}, 601--620.
+\leavevmode\vadjust pre{\hypertarget{ref-gar75}{}}%
+Garrett, M.
+F.
+(1975).
+Levels of processing in sentence production.
+In G.
+H.
+Bower (Ed.), \emph{The psychology of learning and motivation: Advances in research and theory} (Vol.
+9, pp.
+133--177).
+Academic Press.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-gentner1980finger}{}}%
+Gentner, D. R., Grudin, J., \& Conway, E. (1980). \emph{Finger movements in transcription typing} (No. 8001). Center for Human Information Processing, University of California.
+
 \leavevmode\vadjust pre{\hypertarget{ref-gernsbacher1995coherence}{}}%
@@
 Hall, S., Baaijen, V. M., \& Galbraith, D. (2022). Constructing theoretically informed measures of pause duration in experimentally manipulated writing. \emph{Reading and Writing}, 1--29.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hayes2013new}{}}%
+Hayes, J.
+R.
+(1996).
+A new framework for understanding cognition and affect in writing.
+In C.
+M.
+Levy \& S.
+E.
+Ransdell (Eds.), \emph{The science of writing: Theories, methods, individual differences, and applications} (pp.
+1--27).
+Lawrence Erlbaum.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-jefferys1992ockham}{}}%
-Jefferys, W. H., \& Berger, J. O. (1992). Ockham's razor and {B}ayesian analysis. \emph{American Scientist}, \emph{80}(1), 64--72.
+\leavevmode\vadjust pre{\hypertarget{ref-hidi1986producing}{}}%
+Hidi, S., \& Anderson, V. (1986). Producing written summaries: Task demands, cognitive operations, and implications for instruction. \emph{Review of Educational Research}, \emph{56}(4), 473--493.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jeffreys1961theory}{}}%
+Jeffreys, H. (1961). \emph{The theory of probability} (Vol. 3). Oxford University Press, Clarendon Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-johansson2010looking}{}}%
+Johansson, R., Wengelin, Å., Johansson, V., \& Holmqvist, K.
+(2010).
+Looking at the keyboard or the monitor: Relationship with text production processes.
+\emph{Reading and Writing}, \emph{23}, 835--851.
 
@@
 Leijten, M., \& Van Waes, L. (2013). Keystroke logging in writing research: Using {Inputlog} to analyze and visualize writing processes. \emph{Written Communication}, \emph{30}(3), 358--392.
+
+\leavevmode\vadjust pre{\hypertarget{ref-levelt1989speaking}{}}%
+Levelt, W. J. M. (1989). \emph{Speaking: {F}rom intention to articulation} (Vol. 1). MIT press.
 
@@
 Nottbusch, G. (2010). Grammatical planning, execution, and control in written sentence production. \emph{Reading and Writing}, \emph{23}(7), 777--801.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nottbusch2005syllabic}{}}%
+Nottbusch, G., Grimm, A., Weingarten, R., \& Will, U. (2005). Syllabic sructures in typing: Evidence from deaf writers. \emph{Reading and Writing}, \emph{18}, 497--526.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-torranceb}{}}%
+Ofstad Oxborough, G.
+H., \& Torrance, M.
+(2011).
+Multilevel analysis of latency in writing.
+\emph{21st Annual Meeting of the Society for Text and Discourse}.
+\url{http://textanddiscourse2011.conference.univ-poitiers.fr/PROG_DEFIN.pdf}
+
 \leavevmode\vadjust pre{\hypertarget{ref-olive2014toward}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rochester1973significance}{}}%
-Rochester, S. R. (1973). The significance of pauses in spontaneous speech. \emph{Journal of Psycholinguistic Research}, \emph{2}, 51--81.
+\leavevmode\vadjust pre{\hypertarget{ref-roeser2024osf}{}}%
+Roeser, J., Conijn, R., Chukharev, E., Ofstad, G.
+H., \& Torrance, M.
+(2024).
+\emph{Typing in tandem: Language planning in multi-sentence text production is fundamentally parallel}.
+\url{https://osf.io/z65dw/}.
+\url{https://doi.org/10.17605/OSF.IO/Z65DW}
 
 \leavevmode\vadjust pre{\hypertarget{ref-roeser2021modelling}{}}%
-Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L.
-(2021).
-Modelling typing disfluencies as finite mixture process.
-\emph{Reading and Writing}, 1--26.
-\url{https://doi.org/10.1007/s11145-021-10203-z}
+Roeser, J., De Maeyer, S., Leijten, M., \& Van Waes, L. (2024). Modelling typing disfluencies as finite mixture process. \emph{Reading and Writing}, \emph{37}(2), 359--384.
 
@@
 \emph{Cognition}, \emph{127}(2), 235--241.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rumelhart1982simulating}{}}%
+Rumelhart, D. E., \& Norman, D. A. (1982). Simulating a skilled typist: A study of skilled cognitive-motor performance. \emph{Cognitive Science}, \emph{6}(1), 1--36.
 
@@
 Stan Development Team. (2018). \emph{{RStan}: The {R} interface to {Stan}}. \url{https://mc-stan.org/}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sullivan2002self}{}}%
-Sullivan, K. P. H., \& Lindgren, E. (2002). Self-assessment in autonomous computer-aided second language writing. \emph{ELT Journal}, \emph{56}(3), 258--266.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-torranceb}{}}%
-Torrance, M., \& Ofstad, G. (n.d.). \emph{Cascading processes in multi-sentence text production}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-torrancea}{}}%
-Torrance, M., Roeser, J., \& Chukharev-Hudilainen, E. (n.d.). \emph{Lookback in {L1 and L2} writing: An eye movement study}.
+Torrance, M., Roeser, J., \& Chukharev, E. (n.d.). \emph{Lookback supports cascaded, just-in-time processing in second language written composition}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-van2016keystroke}{}}%
-Van Waes, L., Leijten, M., Lindgren, E., \& Wengelin, Å. (2016). \emph{Keystroke logging in writing research: Analyzing online writing processes}. 410--426.
-
 \leavevmode\vadjust pre{\hypertarget{ref-van2019multilingual}{}}%
@@
 \url{https://doi.org/10.1037/0033-295X.114.3.830}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers2010bayesian}{}}%
+Wagenmakers, E.-J., Lodewyckx, T., Kuriyal, H., \& Grasman, R.
+(2010).
+Bayesian hypothesis testing for psychologists: A tutorial on the savage--dickey method.
+\emph{Cognitive Psychology}, \emph{60}(3), 158--189.
 
@@
 
+\endgroup
+
+\newpage
+
+\hypertarget{appendix-appendix}{%
+\appendix}
+
+\newpage
+
+\hypertarget{interkey-interval-predictions-under-a-parallel-theory-of-text-production}{%
+\section{Interkey-interval predictions under a parallel theory of text production}\label{interkey-interval-predictions-under-a-parallel-theory-of-text-production}}
+
+Figure \ref{fig:model} shows how a parallel understanding of language production (Olive, 2014; following Van Galen, 1991) predicts interkey intervals.
+It demonstrates a fundamental property, in which upstream processes provide downstream processes with information as it becomes available (i.e.~on a just-in-time basis), and there is the possibility of buffering.
+For the purposes of illustration, all processing above the final buffer is lumped together as \enquote{message and language processing} and all processing below this buffer as \enquote{motor processing and execution}.
+We make no claims as to exactly how these processes are divided.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics{figures/cascade_figure.pdf}
+  \caption{Illustration of a parallel theory of text production.}
+  \label{fig:model}
+\end{figure}
+
+In both cases the specific processing that is responsible for determining interkey interval is shown in black.
+Figure \ref{fig:model}A shows the case where upstream processing runs faster than, and therefore ahead of, motor execution.
+Message and language processing provides necessary information to the execution processes in advance of (or at exactly the same time as) when it can be used.
+In this case interkey intervals -- both IKI1 and IKI2 -- are determined just by the time taken to prepare and execute the keystroke.
+It is entirely unaffected by the duration of upstream processing.
+Figure \ref{fig:model}B shows the case where upstream processes deliver information to motor processes at a rate that lags behind that at which it could be potentially motor-planned and executed (i.e.~where \enquote{just in time} processing fails).
+This results in a delay in output -- a \enquote{hesitation} in our terminology.
+IKI1 in Figure \ref{fig:model}B is therefore longer than IKI1 in Figure \ref{fig:model}A.
+Just-in-time processing is then resumed, and IKI2 in Figure \ref{fig:model}B is of the same duration as the interkey intervals in Figure \ref{fig:model}A.
+
+In the case where upstream processes run ahead of motor processing, the interkey interval is obviously determined just by the speed at which the writer can move their finger (can motor-plan and execute).
+The essential point that Figure \ref{fig:model} illustrates is that in the case where upstream processes lag behind motor output (Figure \ref{fig:model}B), the interkey interval is determined \emph{just} by time taken for upstream processing.
+Effects on the interkey interval are not additive.
+Interkey intervals are therefore determined by one of two entirely separate data-generating processes.
+This is the basis for our claim that a finite mixture of two distributions represents a statistical model of parallel processing.
+
+\newpage
+
+\hypertarget{statistical-models}{%
+\section{Statistical models}\label{statistical-models}}
+
+We implemented four statistical models using the Bayesian framework (e.g.
+Farrell \& Lewandowsky, 2018; Gelman et al., 2014; Lee \& Wagenmakers, 2014) to evaluate the serial or the parallel view.
+In other words, we used statistical models to map between keystroke data and the theoretically assumed cognitive process that underlies the generation of theses data.
+These models are presented in more detail in this section.
+
+\hypertarget{single-distribution-gaussian-model}{%
+\subsection{Single-distribution Gaussian model}\label{single-distribution-gaussian-model}}
+
+Under the serial view, all planning relevant for the to-be-produced utterance must be completed prior to typing onset.
+The duration of the resulting interkey interval depends on, among others, psycholinguistic factors; i.e.~interkey intervals that are located before easily retrievable high-frequency words tend to be short; similar for words with fewer graphemes, syllables, and morphemes (Nottbusch et al., 2005; Roeser et al., 2019; Torrance et al., 2018).
+We can capture the variability associated with word-features by assuming that before-word interkey intervals can be described as coming from a normal (Gaussian) distribution that can be fully characterised by an unknown central tendency \(\mu\) of interkey intervals associated with word-level planning and a dispersion \(\sigma_\text{e}^2\), that captures the residual variance in the data that is not captured by the model.
+This can be expressed as \(\text{iki}_\text{before word} \sim \mathcal{N}(\mu, \sigma_\text{e}^2)\) where \(\text{iki}_\text{before word}\) are all interkey intervals immediately proceeding a word.
+Of importance is the estimated value of the central tendency parameter \(\mu\) representing in this case the average time needed to mentally prepare a word.
+
+We can extend this simple model of word planning to other linguistic locations.
+Edges of larger linguistic units are typically understood as being associated with planning on higher levels.
+For example, at sentence boundaries, planning needs to happen for word-level properties -- which we captured before with the parameter \(\mu\) -- but also for higher level information such as what meaning it is that should be conveyed, which word to plan and output first, and potentially even dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019).
+We can capture factors that we assume to impact interkey intervals by decomposing \(\mu\) into the general form \(\mu = \alpha + \beta \times \text{x}\).
+For example, to capture the additional planning needed at sentence boundaries we can add the predictor \(\text{x}_\text{sentence[0,1]}\) so that \(\mu = \alpha + \beta \times \text{x}_\text{sentence[0,1]}\).
+This states that when the value of \(\text{x}_\text{sentence}\) takes on 0, the equation reduces to \(\mu = \alpha\) which is then the average interkey interval for word boundaries but, when \(\text{x}_\text{sentence}\) takes on the value 1, the average interkey interval for word boundaries \(\alpha\) is incremented by a changed in the outcome variable of \(\beta\) msecs.
+The value of the \(\beta\) parameter represents, therefore, the additional planning necessary for sentences.
+The application of such a statistical model to the data therefore provides us with an estimate of the parameter value that can be used for statistical inference (e.g.~to determine whether there is evidence for a statistically meaningful difference for interkey intervals immediately proceeding words vs sentences).
+
+We implemented such a model -- a standard Gaussian mixed-effects model -- as described in equation \ref{eq:unimodgaus}.
+This model assumes that interkey intervals \(\text{iki}_i\) with \(i \in 1 \ldots N\) where \(N\) is the total number of interkey intervals come from a Gaussian distribution with a mean \(\mu\),
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimodgaus}
+\text{iki}_i \sim\text{ } & \mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2)\\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+which is a linear function of the \(i^\text{th}\) transition location (we used there levels: before sentence, before word, within word) captured by \(\beta_\text{location}\), and the random-intercept term \(u_\text{participant}\) which is constrained to come from a normal distribution with a mean of 0 and the standard deviation \(\sigma_\text{p}^2\).
+For the random-intercept term, \(\text{participant[i]}\) is the participant associated with the \(i^\text{th}\) interkey interval.
+The posterior of \(u_\text{participant}\) is therefore the difference between the posterior interkey-interval estimate of each participant (i.e.~a positive value indicates that a participant is slower than average; a negative value indicates that a participant is faster than average) and allows to capture the fact that some writers are faster and other writers are slower.
+Importantly this model returns posterior probability distributions with interkey-interval estimates for each \(\beta\), one for each transition location.
+
+Standard-deviation parameters were constrained to be positive because standard deviations can, by definition, never be negative.
+
+\hypertarget{single-distribution-log-gaussian-model}{%
+\subsection{Single-distribution log-Gaussian model}\label{single-distribution-log-gaussian-model}}
+
+The previous model assumes a Gaussian probability function as the underlying data-generating process.
+The model presented in this section is identical to the previous model but instead of assuming a Gaussian probability function, we assume that the process that generates the follows a log-normal (log-Gaussian) distribution.
+
+There are, at least, two reason for why a log-Gaussian distribution is more appropriate than a standard Gaussian: (1) the log-Gaussian have a natural lower bound.
+This is a desirable property because the distance between two subsequent keystroke events is by definition positive.
+The lower bound of the distribution of interkey intervals is delimited by a writer's ability to move their fingers and keyboard polling.
+(2) the log-scale is known to be a better match for data from human motor behaviour and response time data than the normal distribution (Baayen, 2008; Wagenmakers \& Brown, 2007).
+This is because in a Gaussian distribution the distance between adjacent units is linear.
+In other words a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs.
+This does not map onto the psychological interpretation for short and long interkey intervals.
+For example effects that result from difficulty on the motor level (e.g.~executing familiar key combinations such as \emph{ng} compared to \emph{gn}) are typically smaller than differences that are due to higher levels of processing (e.g.~struggling to retrieve a word in an L1 or L2).
+In other words, an effect of 25 msecs is relatively large in the context of lower-level motor events but small in the context of higher-level cognitive activity such as planning what to say next.
+Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end (motor activity) is more psychologically meaningful than a 25 msecs difference on the upper end of the distribution interkey intervals (retrieving words, planning contents).
+
+This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\mathcal{N}()\) was replaced by \(\log\mathcal{N}()\) to represent a log-normal distribution.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimodloggaus}
+\text{iki}_i \sim\text{ } & \log\mathcal{N}(\mu_i, \sigma_\text{e}^2) \\
+\text{where: } &
+\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+& u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+\hypertarget{single-distribution-unequal-variance-log-gaussian-model}{%
+\subsection{Single-distribution unequal-variance log-Gaussian model}\label{single-distribution-unequal-variance-log-gaussian-model}}
+
+The third model is identical to the model presented in the previous section but relaxed the equal-variance assumption associated with transition location.
+In particular the previous models assumed that the variances associated with each transition location were identical.
+This assumption does however not fit with what we know about data from human motor behaviour.
+Longer latencies are known to be associated with a larger variance for human motor behaviour (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).
+For interkey intervals pauses at edges of larger linguistic units can therefore be assumed to be associated with a larger variance.
+In equation \ref{eq:unimoduv}, we introduced this assumption by allowing the standard deviations \(\sigma_{e_\text{location}}^2)\) to vary by transition location.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:unimoduv}
+\text{iki}_i \sim\text{ } & \log\mathcal{N}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
+\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
+ & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
+ \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
+\end{aligned}
+\end{equation}
+
+\hypertarget{two-distribution-log-gaussian-mixture-model}{%
+\subsection{Two-distribution log-Gaussian mixture model}\label{two-distribution-log-gaussian-mixture-model}}
+
+The following model is an extension of the previous single-distribution model.
+In contrast to the models introduced for the serial view of writing, the parallel view assumes that planning occurs while previously planned language units are being executed in writing.
+For the model presented in this section we removed the constraint that all planning must be completed before writing onset.
+In other words, instead of assuming that different linguistic edges shift the distribution over average interkey intervals towards larger values -- as in the previous models -- the parallel view assumes the frequency of observing a hesitant intekey interval and the size of this hesitation depends on text location of the key transition.
+This was achieved by introducing the assumption that interkey intervals come from a weighted combination (i.e.~mixture) of two distributions associated with two different states:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Activation flows into keystrokes without interruption.
+These fluent interkey intervals are merely constrained by a writer's ability to move their fingers and were captured by \(\beta\) in equation \ref{eq:bimodcon}.
+In other words, \(\beta\) represents the average typing speed for fluent transitions between keys.
+Note that the \(\beta\) parameter is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon}.
+\item
+  Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals when information was not available in time, for example when competition occurs during lexical retrieval or when its orthographic representation was not easily available.
+The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}.
+This \(\delta\) parameter was constrained to be positive and added to the distribution of fluent key transitions \(\beta\).
+It therefore represents the magnitude of the slowdown associated with hesitant transitions.
+The slowdown \(\delta\) was allowed to vary by transition locations because hesitations at larger linguistic units are more likely to be associated with higher level planning which may delay output.
+\end{enumerate}
+
+The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant interkey intervals; the second line represents fluent interkey intervals.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:bimodcon}
+\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \log\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+  & (1 - \theta_\text{location[i], participant[i]}) \times \log\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
+\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
+\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
+        & \sigma_{e'}^2 > \sigma_{e}^2\\
+        & 0 < \theta < 1
+\end{aligned}
+\end{equation}
+
+These two distributions are associated with the mixing weight \(\theta\) which must be larger than 0 and smaller than 1.
+\(\theta\) is parameterised to represent the weighting of the distribution in the first line, hence representing the \emph{hesitation probability}.
+This probability is inversely related to the mixing weight of the distribution of short interkey intervals by \(1-\theta\) as the weights of both distributions must sum to 1.
+In line with the literature discussed in the introduction, we assume that the hesitation probability is likely to vary across linguistic locations.
+As hesitation frequency is subject to individual differences and writing style (and skills), we also assumed that some participants are more and others are less likely to hesitate at certain transition locations (Van Waes et al., 2021).
+
+Lastly, we carried over the unequal variance assumption and let the standard deviations \(\sigma_{e'}^2\) and \(\sigma_{e}^2\) vary by transition location.
+We constrained the standard deviations so that \(\sigma_{e'}\) which is associated with the distribution of hesitant interkey intervals is larger than the standard deviation associated with fluent transitions \(\sigma_e\) for reasons discussed above.
+
+\newpage
+
+\hypertarget{fit-to-data}{%
+\section{Fit to data}\label{fit-to-data}}
+
+Model fit was assessed by testing to what extent data simulated from the model match the observed data.
+From each model we obtained 50 simulations (predictions).
+Simulated and observed data are visualised in Figure \ref{fig:prediction}.
+Data predicted by the mixture model were substantially more similar to the observed data compared to any of the single-distribution models.
+The worse fit was observed for the single-distribution Gaussian model which predicted a larger number of negative keystroke intervals.
+
+\begin{figure}
+
+{\centering \includegraphics{figures/fitplots}
+
+}
+
+\caption{Predicted data (50 simulations in each cell) in red compared to interkey interval data shown in black shown by model and datasets.
+Better fit i shown on the left and weaker fit is shown on the right.
+X-axes were truncated at 2 secs.}\label{fig:prediction}
+\end{figure}
+
+\newpage
+
+\hypertarget{key-combination-effect}{%
+\section{Key-combination effect}\label{key-combination-effect}}
+
+Interkey intervals at before sentence location either did (PLanTra, LIFT) scope over both the shift key and the character following the shift key or included only the first key following a the space immediately proceeding a sentence (CATO, C2L1, SPL2, GE2).
+In other words, the interkey interval before sentences are sum of two interkey intervals in the PLanTra and LIFT data, namely \texttt{\_\^{}{[}shift{]}\^{}C}, where \texttt{\^{}} represents a transition between keys, consists only of one interkey interval, namely \texttt{\_\^{}{[}shift{]}}, in the remaining datasets.
+Therefore, differences in hesitation patterns across datasets for transitions immediately proceeding a sentence could be explained without reference to edges of linguistic units.
+
+Therefore we tested to what extent the inclusion of one additional interkey interval at sentence boundaries affected the hesitation results.
+We used the SPL2 data for this comparison and calculated intervals including and excluding the keystroke following a sentence-initial shift press.
+Although we modelled all transition locations, we present only before-sentence transitions as there was, as one would expect, no difference for before and within-word interkey intervals.
+The results of this comparison can be found in Table \ref{tab:shiftcellmeans} (only for the data from participants writing in their first language).
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{llrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:shiftcellmeans}Mixture model estimates for interkey interval immediately preceding a sentence.
+Cellmeans are shown for interkey intervals that do and do not include the interkey interval following a sentence-initial shift key press.
+Hesitation duration is shown in mseecs along with the hesitation probability.
+The difference due to the additional keystoke is shown fo the hesitation duration (log scale) and hesitation probability (logit scale); 95\% PIs are shown in brackets.}\\
+\toprule
+Model parameter & \multicolumn{1}{c}{\_\textasciicircum{}[shift]\textasciicircum{}C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
+\toprule
+Model parameter & \multicolumn{1}{c}{\_\textasciicircum{}[shift]\textasciicircum{}C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Fluent interkey intervals & 156 [145, 169] & 159 [147, 172] & -0.02 [-0.13, 0.09] & 0.06\\
+Hesitation duration & 1,324 [1,186, 1,470] & 1,440 [1,236, 1,662] & -0.06 [-0.21, 0.09] & 0.11\\
+Hesitation probability & 1.00 [.99, 1.00] & .78 [.69, .86] & 4.91 [3.68, 6.35] & > 100\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+
+The duration of fluent and hesitant interkey intervals was not affected by whether or not the sentence-initial transition include the character following the shift key.
+However, we found strong evidence for an increased hesitation probability when the before-sentence interkey interval included the character following the shift key.
+Notably, the hesitation probability approached ceiling when the interkey interval following the shift key was added.
+In other words, the mixture model identified almost all before-sentence transitions as being hesitant as they -- in the majority of cases -- included two keystrokes while interkey intervals for all other transition locations -- and therefore the majority of the data -- did not.
+
+\newpage
+
+\hypertarget{l2-effect-spl2}{%
+\section{L2 effect (SPL2)}\label{l2-effect-spl2}}
+
+For the SPL2 data we calculated the L2 effect (i.e.~the effect of writing in L2 or L1).
+The results can be found in Table \ref{tab:l2effect}.
+The results show longer hesitation duration and more hesitations across all transition locations when writing in L2.
+Only within-word transitions showed negligible evidence for more or longer hesitations.
+
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:l2effect}Mixture model estimates for language effect.
+Cellmeans are shown for the hesitation duration and for the probability of hesitant interkey intervals for texts produced in the writer's L1 and L2.
+Language differences are shown for the hesitation duration (log scale) and the hesitation probability (logit scale); 95\% PIs in brackets.}
+
+\small{
+
+\begin{tabular}{lrrrr}
+\toprule
+Transition location & \multicolumn{1}{c}{L1} & \multicolumn{1}{c}{L2} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+Hesitation duration &  &  &  & \\
+\ \ \ before sentence & 1,440 [1,236, 1,662] & 1,961 [1,710, 2,240] & 0.31 [0.17, 0.45] & > 100\\
+\ \ \ before word & 448 [405, 495] & 719 [656, 788] & 0.47 [0.39, 0.55] & > 100\\
+\ \ \ within word & 279 [210, 384] & 362 [310, 422] & 0.27 [-0.07, 0.55] & 0.68\\
+Hesitation probability &  &  &  & \\
+\ \ \ before sentence & .78 [.69, .86] & .92 [.87, .96] & 1.23 [0.52, 1.97] & > 100\\
+\ \ \ before word & .51 [.41, .62] & .89 [.83, .93] & 2.02 [1.4, 2.66] & > 100\\
+\ \ \ within word & .04 [.02, .07] & .15 [.10, .22] & 1.52 [0.78, 2.29] & > 100\\
+\bottomrule
+\addlinespace
+\end{tabular}
+
+}
+
+\begin{tablenotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{tablenotes}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+\newpage
+
+\hypertarget{masking-effect-cato-ge2}{%
+\section{Masking effect (CATO, GE2)}\label{masking-effect-cato-ge2}}
+
+Studies associated with the CATO and GE2 datasets investigated to what extent hiding (masking) previously written text from the reader affects keystroke behaviour.
+Mixture model results for the masking effect are shown in Table \ref{tab:maskingeffect}.
+There is some evidence that when the text was masked dylexic writers hesitated longer before starting to type sentences.
+Evidence for all other comparisons was negligible.
+
+\begin{landscape}
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lllrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:maskingeffect}Mixture-model estimates for masking effect.
+Cellmeans are shown for the masked and unmasked writing task for hesitation duration (in msecs) and hesitation probability.
+The masking effect is shown on the log scale for durations and on the logit scale for probabilities; 95\% PIs in brackets.}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:maskingeffect} continued}}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Dataset} & \multicolumn{1}{c}{Group} & \multicolumn{1}{c}{Unmasked} & \multicolumn{1}{c}{Masked} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Hesitation duration &  &  &  &  &  & \\
+\ \ \ before sentence & CATO & dyslexic & 2,423 [1,731, 3,302] & 1,871 [1,428, 2,398] & -0.25 [-0.64, 0.13] & 0.44\\
+\ \ \ before sentence & CATO & non dyslexic & 1,924 [1,429, 2,518] & 1,693 [1,323, 2,134] & -0.12 [-0.47, 0.22] & 0.23\\
+\ \ \ before sentence & GE2 & non dyslexic & 1,282 [1,108, 1,469] & 1,575 [1,369, 1,796] & 0.21 [0.05, 0.36] & 2.59\\
+\ \ \ before word & CATO & dyslexic & 635 [590, 683] & 598 [553, 644] & -0.06 [-0.13, 0.01] & 0.14\\
+\ \ \ before word & CATO & non dyslexic & 536 [497, 579] & 518 [480, 558] & -0.04 [-0.11, 0.04] & 0.06\\
+\ \ \ before word & GE2 & non dyslexic & 520 [476, 567] & 536 [492, 584] & 0.03 [-0.05, 0.11] & 0.05\\
+\ \ \ within word & CATO & dyslexic & 469 [356, 593] & 368 [305, 445] & -0.24 [-0.52, 0.06] & 0.57\\
+\ \ \ within word & CATO & non dyslexic & 600 [434, 787] & 583 [420, 784] & -0.03 [-0.42, 0.38] & 0.2\\
+\ \ \ within word & GE2 & non dyslexic & 339 [261, 438] & 367 [290, 466] & 0.08 [-0.24, 0.4] & 0.19\\
+Hesitation probability &  &  &  &  &  & \\
+\ \ \ before sentence & CATO & dyslexic & .60 [.38, .79] & .80 [.60, .93] & 0.2 [-0.06, 0.46] & 0.45\\
+\ \ \ before sentence & CATO & non dyslexic & .55 [.34, .76] & .68 [.47, .84] & 0.13 [-0.16, 0.4] & 0.21\\
+\ \ \ before sentence & GE2 & non dyslexic & .88 [.82, .93] & .90 [.85, .94] & 0.16 [-0.46, 0.79] & 0.36\\
+\ \ \ before word & CATO & dyslexic & .75 [.64, .85] & .72 [.59, .82] & -0.04 [-0.19, 0.12] & 0.09\\
+\ \ \ before word & CATO & non dyslexic & .64 [.51, .75] & .62 [.49, .74] & -0.02 [-0.19, 0.16] & 0.09\\
+\ \ \ before word & GE2 & non dyslexic & .47 [.40, .54] & .45 [.38, .52] & -0.06 [-0.46, 0.35] & 0.21\\
+\ \ \ within word & CATO & dyslexic & .07 [.03, .12] & .11 [.06, .19] & 0.05 [-0.02, 0.13] & 0.08\\
+\ \ \ within word & CATO & non dyslexic & .03 [.02, .06] & .03 [.02, .06] & 0 [-0.03, 0.03] & 0.01\\
+\ \ \ within word & GE2 & non dyslexic & .05 [.03, .08] & .04 [.03, .06] & -0.29 [-0.88, 0.28] & 0.48\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+\end{landscape}
+
+\newpage
+
+\hypertarget{posterior-parameter-estimates}{%
+\section{Posterior parameter estimates}\label{posterior-parameter-estimates}}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics{figures/psplots1constr.pdf}
+\caption{Distribution of posterior parameter estimates with 95\% probability intervals (PI).
+Fluent interkey intervals and hesitation durations are shown in msecs; probability of hesitations is the proportion of hesitant interkey intervals.}
+\end{figure}
+\newpage
+\begin{figure}[!htb]
+\ContinuedFloat
+\captionsetup{list=off,format=cont}
+\centering
+\includegraphics{figures/psplots2constr.pdf}
+\label{fig:fullps1}
+\caption{Distribution of posterior parameter estimates with 95\% probability intervals (PI).
+Fluent interkey intervals and hesitation durations are shown in msecs; probability of hesitations is the proportion of hesitant interkey intervals.}
+
+\end{figure}
+
+\newpage
+
+\hypertarget{pre-post-test-plantra}{%
+\section{Pre-post test (PLanTra)}\label{pre-post-test-plantra}}
+
+The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}. Evidence for all comparisons was negligible.
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:retesteffect}Mixture model estimates for post-test effect.
+Cellmeans are shown for pre and post-test for the duration of hesitant interkey intervals in msecs and its probability.
+Differences are shown on the log scale for hesitation duration and on the logit scale for the hesitation probability; 95\% PIs in brackets.}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:retesteffect} continued}}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Pre-test} & \multicolumn{1}{c}{Post-test} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Hesitation duration &  &  &  & \\
+\ \ \ before sentence & 1,034 [875, 1,210] & 1,064 [893, 1,257] & -0.03 [-0.23, 0.17] & 0.1\\
+\ \ \ before word & 593 [544, 645] & 589 [540, 643] & 0.01 [-0.08, 0.1] & 0.05\\
+\ \ \ within word & 399 [333, 480] & 424 [353, 509] & -0.06 [-0.3, 0.18] & 0.14\\
+Hesitation probability &  &  &  & \\
+\ \ \ before sentence & .58 [.50, .66] & .54 [.47, .62] & 0.15 [-0.26, 0.58] & 0.27\\
+\ \ \ before word & .56 [.50, .61] & .49 [.43, .55] & 0.27 [-0.04, 0.59] & 0.72\\
+\ \ \ within word & .11 [.08, .14] & .09 [.07, .12] & 0.13 [-0.27, 0.54] & 0.26\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+
+\newpage
+
+\hypertarget{genre-effect-lift}{%
+\section{Genre effect (LIFT)}\label{genre-effect-lift}}
+
+From the LIFT data, we assessed differences between genres, i.e.~writing an informative as opposed to argumentative text.
+Results are shown in Table \ref{tab:genreeffect}.
+Cellmeans and differences were average across writing topic.
+Evidence for all comparisons was negligible.
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:genreeffect}Mixture-model estimates for genre effect.
+Cellmeans for hesitation duration (in msecs) and the probability of hesitant transitions are shown for argumentative and informative texts.
+The genre effect is shown on the log scale for hesitationn durations and on the logit scale for the probability of hesitant transitions; 95\% PIs in brackets.}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:genreeffect} continued}}\\
+\toprule
+Transition location & \multicolumn{1}{c}{Argumentative} & \multicolumn{1}{c}{Informative} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
+\midrule
+\endhead
+Hesitation duration &  &  &  & \\
+\ \ \ before sentence & 524 [451, 606] & 562 [481, 656] & -0.07 [-0.33, 0.19] & 0.16\\
+\ \ \ before word & 438 [400, 479] & 435 [398, 476] & 0.01 [-0.15, 0.16] & 0.09\\
+\ \ \ within word & 351 [269, 458] & 397 [306, 513] & -0.12 [-0.6, 0.36] & 0.29\\
+Hesitation probability &  &  &  & \\
+\ \ \ before sentence & .35 [.28, .43] & .34 [.27, .42] & 0.05 [-0.57, 0.73] & 0.36\\
+\ \ \ before word & .46 [.39, .54] & .48 [.41, .55] & -0.05 [-0.51, 0.44] & 0.25\\
+\ \ \ within word & .07 [.05, .10] & .07 [.05, .10] & -0.02 [-0.81, 0.76] & 0.46\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+
+\newpage
+
+\hypertarget{transition-location-effect-1}{%
+\section{Transition location effect}\label{transition-location-effect-1}}
+
+\begin{center}
+\begin{ThreePartTable}
+
+\begin{TableNotes}[para]
+\normalsize{\textit{Note.} PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.}
+\end{TableNotes}
+
+\footnotesize{
+
+\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:loceffectcon}Transition-location effect. Differences are shown for the hesitation duration (log scale) and the hesitation probability (logit scale); 95\% PIs in brackets.}\\
+\toprule
+ & \multicolumn{2}{c}{Hesitation duration} & \multicolumn{2}{c}{Hesitation probability} \\
+\cmidrule(r){2-3} \cmidrule(r){4-5}
+Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:loceffectcon} continued}}\\
+\toprule
+ & \multicolumn{2}{c}{Hesitation duration} & \multicolumn{2}{c}{Hesitation probability} \\
+\cmidrule(r){2-3} \cmidrule(r){4-5}
+Comparison & Est. [95\% PIs] & BF & Est. [95\% PIs] & BF\\
+\midrule
+\endhead
+\textbf{C2L1} &  &  &  & \\
+\ \ \ before sentence vs word & 0.24 [0.02, 0.48] & 1.11 & 0.44 [-0.56, 1.66] & 0.68\\
+\ \ \ before vs within word & -0.03 [-0.14, 0.09] & 0.07 & 3.95 [3.54, 4.38] & > 100\\
+\textbf{CATO (non-dyslexic, unmasked)} &  &  &  & \\
+\ \ \ before sentence vs word & 1.27 [0.98, 1.55] & > 100 & -0.36 [-1.41, 0.69] & 0.67\\
+\ \ \ before vs within word & -0.1 [-0.38, 0.21] & 0.2 & 4.05 [3.2, 4.91] & > 100\\
+\textbf{GE2 (unmasked)} &  &  &  & \\
+\ \ \ before sentence vs word & 0.9 [0.76, 1.04] & > 100 & 2.15 [1.6, 2.75] & > 100\\
+\ \ \ before vs within word & 0.44 [0.17, 0.69] & 26.78 & 2.76 [2.24, 3.28] & > 100\\
+\textbf{LIFT} &  &  &  & \\
+\ \ \ before sentence vs word & 0.21 [-0.05, 0.48] & 0.5 & -0.53 [-1.04, 0.08] & 1.61\\
+\ \ \ before vs within word & 0.17 [-0.32, 0.53] & 0.41 & 2.48 [1.81, 3.22] & > 100\\
+\textbf{PLanTra} &  &  &  & \\
+\ \ \ before sentence vs word & 0.57 [0.4, 0.74] & > 100 & 0.16 [-0.25, 0.57] & 0.28\\
+\ \ \ before vs within word & 0.36 [0.16, 0.56] & 55.53 & 2.3 [1.9, 2.7] & > 100\\
+\textbf{SPL2 (L1)} &  &  &  & \\
+\ \ \ before sentence vs word & 1.17 [1.02, 1.3] & > 100 & 1.23 [0.59, 1.88] & > 100\\
+\ \ \ before vs within word & 0.48 [0.16, 0.76] & 9.07 & 3.3 [2.56, 4.08] & > 100\\
+\bottomrule
+\addlinespace
+\insertTableNotes
+\end{longtable}
+
+}
+
+\end{ThreePartTable}
+\end{center}
+
+\newpage
+
+\hypertarget{simulation}{%
+\section{Simulation}\label{simulation}}
+
+A possible concern with our results -- the substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always outperform single-distribution models.
+We addressed this concern before by using cross-validation techniques for model comparison which prevent overfit by penalising models with more parameters.
+To further address this concern we evaluated models -- similar to the ones used in the main text -- for two sets of simulated data.
+Data were simulated using two random number generators, one that samples data from a single distribution and another that samples data from a weighted combination of two distributions.
+These simulated datasets allow us to test the predictive performance of our models in a context where we know the true underlying data generating process.
+
+The first dataset was simulated from a weighted mixture of two log-normal distributions similar to the process described above (equation \ref{eq:bimodcon}).
+This process and the corresponding Bayesian model that we used to parameter estimation is summarised in equation \ref{eq:simmog}.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:simmog}
+\text{y} \sim\text{ } & \theta \times \log\mathcal{N}(\beta + \delta, \sigma^2_1) +\\
+& (1 - \theta) \times \log\mathcal{N}(\beta, \sigma^2_2)\\
+\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
+        & \sigma_{1}^2 > \sigma_{2}^2
+\end{aligned}
+\end{equation}
+
+The equivalent Bayesian model is largely identical to the model used in the main text but does not assume different parameters for transition locations and does not include random effects for participants.
+This is because we neither simulated data for different transition locations (or other factors) or repeated measures for participants.
+The model assumes two data generating processes with each assuming a log-normal distribution with a mixing proportion \(\theta\).
+The distribution of smaller values has a mean \(\beta\) and a standard deviation \(\sigma^2_2\); the second distribution of larger values is constrained to have a mean that is \(\delta\) units larger than \(\beta\) and has a standard deviation \(\sigma^2_1\) larger than the distribution with the central tendency \(\beta\).
+
+In R we can simulate data for equation \ref{eq:simmog} using the helper function below.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# Function for a mixture of two log{-}Gaussians}
+\NormalTok{molg }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(n, theta, mu1, mu2, sigma1, sigma2) \{}
+\NormalTok{  y0 }\OtherTok{\textless{}{-}} \FunctionTok{rlnorm}\NormalTok{(n, }\AttributeTok{mean =}\NormalTok{ mu1, }\AttributeTok{sd =}\NormalTok{ sigma1)}
+\NormalTok{  y1 }\OtherTok{\textless{}{-}} \FunctionTok{rlnorm}\NormalTok{(n, }\AttributeTok{mean =}\NormalTok{ mu2, }\AttributeTok{sd =}\NormalTok{ sigma2)}
+\NormalTok{  mix }\OtherTok{\textless{}{-}} \FunctionTok{rbinom}\NormalTok{(n, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ theta)}
+\NormalTok{  y }\OtherTok{\textless{}{-}}\NormalTok{ y0 }\SpecialCharTok{*}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ mix) }\SpecialCharTok{+}\NormalTok{ y1 }\SpecialCharTok{*}\NormalTok{ mix }
+\NormalTok{\}}
+
+\NormalTok{N }\OtherTok{\textless{}{-}} \DecValTok{1000} \CommentTok{\# number of participants}
+\NormalTok{beta }\OtherTok{\textless{}{-}} \DecValTok{5} \CommentTok{\# mean of fluent interkey intervals (in log)}
+\NormalTok{delta }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# increment for hesitant interkey intervals (in log)}
+\NormalTok{theta }\OtherTok{\textless{}{-}}\NormalTok{ .}\DecValTok{35} \CommentTok{\# proportion of hesitations}
+\NormalTok{sigma }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(.}\DecValTok{25}\NormalTok{, .}\DecValTok{5}\NormalTok{) }\CommentTok{\# error variance}
+
+\CommentTok{\# Simulate data from a mixture of two log{-}Gaussians}
+\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{molg}\NormalTok{(}\AttributeTok{n =}\NormalTok{ N, }
+          \AttributeTok{theta =}\NormalTok{ theta,}
+          \AttributeTok{mu1 =}\NormalTok{ beta,}
+          \AttributeTok{mu2 =}\NormalTok{ beta }\SpecialCharTok{+}\NormalTok{ delta,}
+          \AttributeTok{sigma1 =}\NormalTok{ sigma[}\DecValTok{1}\NormalTok{],}
+          \AttributeTok{sigma2 =}\NormalTok{ sigma[}\DecValTok{2}\NormalTok{])}
+\end{Highlighting}
+\end{Shaded}
+
+A second dataset was simulated coming from a single log-normal distribution following equation \ref{eq:simuv}. An equivalent Bayesian model was implemented for parameter estimation.
+
+\begin{equation}
+\begin{aligned}
+\label{eq:simuv}
+\text{y} \sim\text{ }& \log\mathcal{N}(\beta, \sigma^2)\\
+\text{constraint: } & \sigma^2>0
+\end{aligned}
+\end{equation}
+
+Data following the distribution in equation \ref{eq:simuv} can be simulated using the following R code:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{N }\OtherTok{\textless{}{-}} \DecValTok{1000} \CommentTok{\# number of participants}
+\NormalTok{beta }\OtherTok{\textless{}{-}} \DecValTok{5} \CommentTok{\# population mean}
+\NormalTok{sigma }\OtherTok{\textless{}{-}}\NormalTok{ .}\DecValTok{25} \CommentTok{\# error variance}
+
+\CommentTok{\# Simulate data from a single log{-}normal distribution}
+\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{rlnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ N, }\AttributeTok{mean =}\NormalTok{ beta, }\AttributeTok{sd =}\NormalTok{ sigma)}
+\end{Highlighting}
+\end{Shaded}
+
+Again, the Bayesian model corresponding to this process is a simplified version of the single-distribution model used in the main text.
+The model assumes a log-Gaussian distribution with a mean \(\beta\) and a standard deviation \(\sigma^2\).
+
+Both simulated datasets are visualised in Figure \ref{fig:simdata}.
+The parameter values used for each of the two data simulations can be seen in Table \ref{tab:simparam}.
+Parameter values were chosen so that the simulated data are distributed roughly similar to interkey interval data.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}
+
+}
+
+\caption{Data simulated with a two-distribution (yellow) and a single-distribution (grey) random data-generating process.
+The x-axis showing the outcome y was log-scaled for visibility.}\label{fig:simdata}
+\end{figure}
+
+For each of these two datasets we simulated 1,000 observations.
+We fitted 2 models -- the two-distribution mixture model and the single-distribution model described above -- for each dataset.
+Models were run with 3 chains, with each 6,000 iterations of which 3,000 iterations were discarded as warmup samples.
+Estimates with 95\% probability intervals are shown in Table \ref{tab:simparam}.
+True and estimated parameter values are shown for each data generating process and Bayesian model.
+Both models successfully uncovered the model parameters for the datasets simulated with the corresponding underlying process, but less so when the model was applied to the data generated with the other incorrect underlying process.
+
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability intervals (PI) and true parameter values for each simulated dataset by model.}
+
+\begin{tabular}{lrr}
+\toprule
+ & \multicolumn{2}{c}{Estimates with 95\% PIs} \\
+\cmidrule(r){2-3}
+Parameter with true value & \multicolumn{1}{c}{Two-distribution data} & \multicolumn{1}{c}{Single-distribution data}\\
+\midrule
+Two-distribution model &  & \\
+\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
+\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
+\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
+\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
+\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
+Single-distribution model &  & \\
+\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
+\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
+\bottomrule
+\end{tabular}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+We used leave-one-out cross-validation to compare the predictive performance of the two models for each data-generating process.
+Model comparisons can be found in Table \ref{tab:loossim}.
+For the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance.
+In fact, the mixture model's predictive performance is 11.6 standard errors higher compared to the single-distribution model.
+However, the mixture model showed a slightly lower predictive performance (i.e.~a difference of 0.77 standard errors) for the single-process data.
+Thus, the single distribution model is the more parsimonious choice for the single-distribution data.
+These results rule out the possibility that mixture models always lead to higher predictive performance.
+
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:loossim}Model comparisons by dataset.
+The top row shows the models with the highest predictive performance for each data generating process.
+Standard error is shown in parentheses.}
+
+\begin{tabular}{lrr}
+\toprule
+Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
+\midrule
+Data: Two-distribution mixture process &  & \\
+\ \ \ Two-distribution mixture model & -- & -6,068 (41)\\
+\ \ \ Single-distribution model & -191.3 (16.5) & -6,259 (38)\\
+Data: Single-distribution process &  & \\
+\ \ \ Single-distribution model & -- & -5,030 (24)\\
+\ \ \ Two-distribution mixture model & -0.5 (0.7) & -5,030 (24)\\
+\bottomrule
+\addlinespace
+\end{tabular}
+
+\begin{tablenotes}[para]
+\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
+\end{tablenotes}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+\newpage
+
 \end{document}

### 2024-10-26 report/sections/introduction.Rmd
After review update

@@
 @levelt2000word; @griffin2003reversed; @wheeldon_konopka_2023].
-For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech [@not10; dam09; @roeser2018advance].
+For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech [@not10; @dam09; @roeser2018advance].
 For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e.

### 2024-10-26 report/sections/methods.Rmd
After review update

@@
                            "EyeWrite"),
-                `Writing task` = c("Argumentative", 
-                                 "Expository",
-                                 "Argumentative",
-                                 "Text simplification",
-                                 "Synthesis", 
-                                 "Argumentative"),
-                 N = as.character(c(126, 26*2, 39, 47, 658, 45)),
+                `Writing task` = c("Argumentative essay", 
+                                 "Expository essay",
+                                 "Argumentative essay",
+                                 "Paraphrase technical text$^b$",
+                                 "Synthesis of several printed sources$^b$", 
+                                 "Argumentative essay"),
+                 `Words in text$^a$` = c("128.5 (52.9)", "288 (119)", "668 (301)", "295 (79)", "374 (56)$^d$", "540 (141)"),
 #                n_texts = c(1, 2, 2, 2, NA),
@@
 #                n_words = c(),
-                `Mean age` = c(11.8, 16.9, 20.6, 23, 17.0, 19),
+                `Age` = c("11.8 (0.47)", "16.9 (0.88)", "20.6 (1.62)", "23 (2.7)", "16.7 (1.5)", "19.1 (1.4)$^c$"),
+                 N = as.character(c(126, 26*2, 39, 47, 658, 45)),
                  Sample = c("6th graders", "Secondary school students (dyslexic, non dyslexic)", "Undergraduate students", "Master students", "Secondary school students", "Undergraduate students"),
@@
   arrange(Dataset) %>% 
-  apa_table(align = c("l", "p{3cm}", "p{2cm}", "p{2.5cm}", "r", "r", "p{2.5cm}", "p{2.25cm}", "p{2.5cm}"), 
+  apa_table(align = c("p{1.5cm}", "p{2.75cm}", "p{1.75cm}", "p{2.25cm}", "p{1.75cm}", "p{1.75cm}", "p{.5cm}", "p{2.5cm}", "p{1.75cm}", "p{2cm}"), 
             escape = FALSE, 
+            longtable = FALSE,
+            row.names = T,
             font_size = "footnotesize",
-            caption = "Datasets in brief.")
+            caption = 'Datasets in brief.', 
+            note = "$^a$Mean number of words in the final text; standard deviations in parentheses.
+$^b$Reading activity was marked in the dataset and removed before analysis.
+$^c$Estimated from a similar sample from the same population.
+$^d$Estimated based on the number of characters in the final product following Brysbaert et al.
+(2021).")
 
-#                Manipulation = c("--", 
-#                               "Weak decoders / control; texted masked / unmasked",
-#                               "Text produced in L1 / L2",
-#                               "Pre / post test with training on plain language principles", 
-#                              "Various topics and genres",
-#                               "masked / unmasked")
 ```
@@
 At before-sentence locations, interkey intervals were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GE2) but were timed to the character key press in the  PLanTra and LIFT datasets (i.e.
-the duration included the time to perform the capitalisting shift keypress); we return to this difference in the Results section.
+the duration included the time to perform the capitalising shift keypress); we return to this difference in the Results section.
 Transitions that occurred at the very start of the text or at the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.
@@
 
-table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
+table <- tibble(`Transition type` = c("Within word", "Before word", "Before sentence"),
        Description = c("Transitions between any letter",
@@
 
-apa_table(table, caption = "Transition location classification.",
+apa_table(table, caption = 'Transition location classification. The final text in all cases was "The cat meowed. Then it slept."',
             align = c("p{3cm}", "p{5cm}", "p{6cm}"), 
@@
           caption = "Data reduction.
-Mean percentage of extreme data removed and the mean percentage of randomly sampled data analysed by transition location (`--` if all data were used).
+Shown are the percentage of extreme data that were removed and the percentage of randomly sampled data that were analysed by transition location (`--` if all data were included).
 Standard error in parentheses.",
@@
                        "Equivalent to M2 but with different variance components for each transition location (unequal variance).",
-                       "Two-distributions mixture of a log-Gaussian for fluent interkey intervals and another wider log-Gaussian for hesitant / disfluent interkey intervals; the distribution of hesitant interkey intervals assumes -- similar to M3 -- different values for transition locations.
-The mixing weight of these distributions capture the relative number of disfluent transitions for each transition location.",
+                       "Two-distributions mixture of a log-Gaussian for fluent interkey intervals and another wider log-Gaussian for hesitant interkey intervals; the distribution of hesitant interkey intervals assumes -- similar to M3 -- different values for transition locations.
+The mixing weight (i.e.
+proportion of observations estimated to be associated with each distribution) captures the relative number of hesitant transitions for each transition location.",
                        "Equivialent to M4 but distribution of fluent transitions is allowed to vary by transition location.")) %>% 

### 2024-10-26 report/sections/results.Rmd
After review update

@@
 A negative difference in $\\Delta\\widehat{elpd}$ indicates higher predictive performance for the more complex model; standard error shown in parentheses.",
-            align =c("p{1.75cm}", "p{4cm}", "p{4.25cm}", "p{3.5cm}"), 
+            align =c("p{2cm}", "p{4cm}", "p{4.25cm}", "p{3.5cm}"), 
             escape = FALSE,
@@
 # get files
-files <- list.files("../../stanout", pattern = "mog_constr_posterior_for_densityplot.csv", recursive = T, full.names = T)
+files <- list.files("../../stanout", 
+                    pattern = "mog_constr_posterior_for_densityplot.csv", 
+                    recursive = T, full.names = T)
 
@@
                           left = "Density",
-                          bottom = "Interkey interval in secs")
+                          bottom = "Interkey intervals in secs")
 #ggsave("mixdists.png", plots_fin, width = 12, height = 6)
@@
          !(str_detect(data, "CATO") & task == "masked" & !is.na(task)),
-         !(str_detect(data, "GE2") & xn == "masked" & !is.na(xn)))
+         !(str_detect(data, "GE2") & xn == "masked" & !is.na(xn))) %>% 
+  mutate(across(data, ~factor(., levels = unique(data)[6:1], ordered = T)),
+         across(location, factor))
 
@@
 
-```{r crossstudypost2, fig.cap="Mixture model parameter estimates across studies.
+```{r crossstudypost2old, fig.cap="Mixture model parameter estimates across studies.
 Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
-Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group."}
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.", eval = F}
 
@@
 grid.arrange(patchworkGrob(plot), 
-             left = "Posterior estimate with 95% PIs",
+             left = "Posterior estimates with 95% PIs",
              bottom = "Transition location")
 
+```
+
+```{r crossstudypost2, fig.cap="Mixture model parameter estimates across studies.
+Distributions of parameter estimates are represented as posterior mean and 95\\% probability interval (PI).
+Interkey-interval duration measures are shown in msecs and hesitation probability is shown in proportions.
+Estimates for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.", fig.height=5.5}
+
+posd <- position_dodge(-.5)
+dotsize <- 2.5
+grouplabel <- "Transition\nlocations:"
+shapes <- c(8, 9, 1)
+width <- 0
+linewidth <- .35
+nrow <- 1
+beta <- filter(ps, name == "beta", 
+               location == "before\nword") %>%
+  mutate(name = beta_label,
+         location = "overall") %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = data)) +
+  geom_errorbar(width = width, alpha = .75) +
+  geom_point(size = dotsize) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_colour_colorblind() +
+  scale_y_continuous(labels = scales::comma) +
+  theme(axis.title = element_blank(),
+        axis.ticks.y = element_blank()) +
+  coord_flip() +
+  labs(colour = grouplabel,
+       shape = grouplabel) +
+  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
+         shape = guide_legend(nrow=nrow,byrow=TRUE))
+
+delta <- filter(ps, name == "beta2") %>% 
+  mutate(name = delta_label) %>% 
+  #mutate(across(c(mean, lower, upper), ~./1000)) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = data,
+             shape = location)) +
+  geom_errorbar(width = width, alpha = .75, position = posd) +
+  geom_point(aes(colour = location), 
+            size = dotsize, 
+            position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_colour_colorblind() +
+  scale_y_continuous(labels = scales::comma) +
+  labs(colour = grouplabel,
+       shape = grouplabel) +
+  theme(axis.title = element_blank(),
+        axis.ticks.y = element_blank(),
+        axis.text.y = element_blank()) +
+  coord_flip() +
+  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
+         shape = guide_legend(nrow=nrow,byrow=TRUE))
+
+theta <- filter(ps, name == "prob") %>% 
+  mutate(name = theta_label) %>% 
+  ggplot(aes(y = mean, 
+             ymin = lower, 
+             ymax = upper,
+             x = data,
+             shape = location)) +
+  geom_errorbar(width = width, alpha = .75, position = posd) +
+  geom_point(aes(colour = location), 
+            size = dotsize, 
+            position = posd) +
+  facet_grid(~name) +
+  scale_shape_manual(values = shapes) +
+  scale_colour_colorblind() +
+  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
+  labs(colour = grouplabel,
+       shape = grouplabel) +
+  theme(axis.title = element_blank(),
+        axis.ticks.y = element_blank(),
+        axis.text.y = element_blank()) + 
+  coord_flip() +
+  guides(colour=guide_legend(nrow=nrow,byrow=TRUE),
+         shape = guide_legend(nrow=nrow,byrow=TRUE))
+
+plot <- beta + delta + theta + plot_layout(guides = "collect",
+                                           widths = c(.4, 1, 1))
+
+plot <- grid.arrange(patchworkGrob(plot), 
+             bottom = "Posterior estimates with 95% PIs")
+
+ggsave("../posteriorplot.pdf", plot = plot, width = 8, height = 6)
 ```
@@
           col.names = c("Dataset", rep(c("before sentence vs word", "before vs within word"), 2)),
-          col_spanners = list(" " = c(1),
-                              "Hesitation slowdown" = c(2, 3), 
-                              "Hesitation probability" = c(4, 5)),
-          note = "BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
-          caption = "BFs for the effect of transition location on interkey intervals shown for the hesitation slowdown and hesitation probability.
+          col_spanners = list("Hesitation duration" = 2:3, 
+                              "Hesitation probability" = 4:5),
+          note = "BF = evidence in favour of the alternative hypothesis over the null hypothesis calculated using the Savage-Dickey method (Dickey et al., 1970; Wagenmakers et al., 2010).
+BFs greater than 3 indicate moderate evidence, greater than 10 strong evidence, and greater than 30 very strong evidence for a statistically meaningful effect (Jeffreys, 1961; Lee \\& Wagenmakers, 2014).
+A BF smaller than 0.33 suggests evidence in favour of a null effect (Dienes, 2014) and a BF around 1 is inconclusive evidence.",
+          caption = "BFs for the effect of transition location on interkey intervals shown for the hesitation duration and hesitation probability.
 Comparisons for the CATO dataset were calculated for the non-dyslexic group, unmasked condition; also the GE2 estimtes represent the unmasked condition; SPL2 estimates are for the L1 group.")

### 2024-10-26 slides/sections/discussion.Rmd
After review update

@@
 
-- Writers do not necessarily pause but plan utterances in parallel to writing.
+- First robust evidence that written composition is largely a parallel process.
+- Writers do not necessarily pause at larger linguistic locations but plan utterances in parallel to writing.
 - Evidenced by 
   - stronger predictive performance of mixture models [@roeser2021modelling].
-  - pauses are generally equally likely before before sentences compare to before words (but often longer).
-- Largely consistent for populations with different levels of writing experience, languages (e.g. young / L2 writers, students), composition tasks. 
-- First robust evidence that written composition is largely a parallel process.
+  - pauses are not generally more likely before sentences compare to words (but often longer).
+- Pauses are consistently more likely at word-initial location compared to word-medial but (1) not always longer and (2) for sentence-initial interkey intervals pausing behaviour varies as a function of writing experience, languages (e.g.
+young / L2 writers, students), composition tasks.

### 2024-10-26 slides/sections/evidence.Rmd
After review update

@@
 
-- Interkey intervals are longer at larger linguistic edges following relatively fluent production bursts [@chenoweth2001fluency; @hayes2012evidence; @kaufer1986composing].
-- Mean pause duration before a writer starts a sentence is typically longer than before a mid-sentence word and these are, in turn, longer than between mid-word key presses [e.g.
-@conijn2019understanding].
-- Planning message and syntax for the next sentence adds to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke [@baaijen2012keystroke; @roeser2018advance].
+- Pause duration before a writer starts a sentence is typically longer than before a mid-sentence word and these are longer than between mid-word key presses [e.g. @conijn2019understanding]. 
+- Long interkey intervals are typically followed by production bursts [@hayes2012evidence; @kaufer1986composing].
+- Planning message and syntax for a sentence adds to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke [@baaijen2012keystroke; @roeser2018advance].
 
@@
 
+- Variations in interkey intervals are not sufficiently explained by text location.
 - Writers often do not hesitate before sentence (or word) onset.
-- In some cases sentences are not planned, planning was postponed until after sentence onset, or planning was completed in parallel with previous output [@not10; dam09; @roeser2018advance].
+- In some cases sentences are not planned, planning was postponed until after sentence onset, or completed in parallel with previous output [@not10; @roeser2018advance].
 - Sentence initial durations are often shorter than what one would expect if sentence-initial pauses reflect planning.
-  - @medimorec2017pauses: 71% of sentence-initial interkey intervals shorter than 1 sec. 
-  - @ronneberg2022process: 50% sentence-initial intervals were around 430 msecs but 1.2 secs for the remainder. 
+- Fluent output is maintained by a cascade of processes that run in parallel [@olive2014toward; @van1991handwriting].
+  <!-- - @medimorec2017pauses: 71% shorter than 1 sec -->
+  <!-- - @ronneberg2022process: 50% around 430 msecs and around 1.2 secs  -->
   <!-- - For comparison, these short pauses are less than mean written naming response latency for single objects [@torrance2018timed] and longer pauses were similar to response latencies for short sentences when describing arrays of images [@roeser2018advance].
 -->
-- Fluent output is maintained by a cascade of processes that run in parallel [@olive2014toward;@van1991handwriting].
-
-## Two distinct data-generating processes leading to a mixture of two distributions of interkey intervals [@roeser2021modelling].
-
-- Minimum interkey intervals for executing motor movements; 100-150 msecs [@conijn2019understanding;@waes2019].
-- If upstream processes provide output more slowly, then interkey intervals are determined by time taken to complete upstream processes and not on finger movement. 
-- Neither process (or distribution of keystrokes) can be explained on the basis of the keystroke location in the text.

### 2024-10-26 slides/sections/implementation.Rmd
After review update

@@
-## Implentation
+## Implementation and model comparison
 
-- Bayesian inference [e.g. @lambert2018student]
-- Random effects for participants [see @roeser2021modelling] with code based on @vasishth2017
-- Stan [@carpenter2016stan] using `rstan` [@rstan] to interface with $R$
-- Leave-one-out cross-validation estimated as the sum of the expected log predictive density $\widehat{elpd}$ and compared by the difference between models $\Delta\widehat{elpd}$ [@vehtari2015pareto; @vehtari2017practical]
-- Difference between models was summarised as $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, the *z*-score of the difference between models [@sivula2020uncertainty].
+<!-- - Bayesian inference [e.g. @lambert2018student] models implemented in Stan [@carpenter2016stan] using `rstan` [@rstan] to interface with $R$ -->
+- Code based on @vasishth2017 [see @roeser2021modelling]
+- Random effects for participants
+- Leave-one-out cross-validation: sum of the *expected log predictive density* $\widehat{elpd}$ 
+- *Difference between models* $\Delta\widehat{elpd}$ [@vehtari2015pareto; @vehtari2017practical]
+- Difference between models was summarised as
+
+$\mid\frac{\Delta\widehat{elpd}}{\text{SE}_\text{diff}}\mid$,
+
+i.e. the *z*-score of the change in predictive performance [@sivula2020uncertainty].

### 2024-10-26 slides/sections/locationeffects.Rmd
After review update

@@
 
+```{r loceffectconbf}
+
+files <- list.files(str_c("../../stanout/", 
+                          c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetacon.+.csv", full.names = T) # "spl2_shift",
+
+ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
+                                 mutate(data = .x)) %>% 
+  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
+         across(data, ~recode(., 
+                              "cato" = "CATO",
+                              "spl2" = "SPL2", 
+                              "gunnexp2" = "GE2",
+                              #                "spl2_shift" = "*SPL2* (L1; shift + C)", 
+                              "plantra" = "PLanTra",
+                              "lift" = "LIFT",
+                              "c2l1" = "C2L1"))) %>% 
+  filter(!(str_detect(data, "SPL2") & lang == "ES"),
+         !(str_detect(data, "CATO") & group == "dyslexic"),
+         !(str_detect(data, "CATO") & task == "masked"),
+         !(str_detect(data, "GE2") & xn == "masked"),
+         param %in% c("delta", "theta")) %>%
+  select(param, location, data, value) %>% 
+  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
+  pivot_wider(names_from = location, values_from = value) %>%
+  unnest(-c(data, param)) %>% 
+  mutate(diff.1 = `before sentence` - `before word`,
+         diff.2 = `before word` - `within word`) %>% 
+  summarise(across(starts_with("diff"), 
+                   list(mean = mean, 
+                        lower = lower, 
+                        upper = upper, 
+                        BF = ~log(BF(.)))),
+            .by = c(data, param))
+
+tmp <- ps_loc_diffs %>% 
+  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
+  #mutate(across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
+  mutate(across(BF, ~round(.,2))) %>% 
+  select(-mean, -lower, -upper) %>% 
+  pivot_wider(names_from = c(diffid, param), values_from = BF) %>% 
+  select(data, contains("delta"), contains("theta")) %>% 
+  arrange(data)
+
+tmp %>% 
+  mutate(across(-data, ~cell_spec(., color = ifelse(. > 1 | . < 0, "white", "black"), 
+                                  background = ifelse(. > 1, "blue",
+                                                  ifelse(. < 0, "red", "white"))))) %>% 
+  kable(col.names = c("Dataset", rep(c("before sentence vs word", "before vs within word"), 2)),
+        caption = "Values indicate log BFs in support of the alternative hypothesis over the null hypothesis.", 
+        align = c("l", rep("r", 4)),
+        escape = F,
+        format = "html") %>% 
+  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
+                full_width = T, 
+                position = "right",
+                font_size = 14) %>% 
+  add_header_above(c(" " = 1,
+                     "Hesitation slowdown" = 2,
+                     "Hesitation probability" = 2)) %>% 
+  row_spec(0:6, extra_css = "line-height: 2.25;")  
+```
+
+## Location estimates by model parameter
+
 ```{r crossstudypost2, fig.cap="", fig.height=4.5, fig.width=10}
@@
 ```
-
-## Location estimates by model parameter
-
-```{r loceffectconbf, results='asis'}
-
-files <- list.files(str_c("../../stanout/", 
-                          c("lift", "spl2", "gunnexp2", "plantra", "cato", "c2l1")), pattern = "mogbetacon.+.csv", full.names = T) # "spl2_shift",
-
-ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
-                                 mutate(data = .x)) %>% 
-  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
-         across(data, ~recode(., 
-                              "cato" = "CATO",
-                              "spl2" = "SPL2", 
-                              "gunnexp2" = "GE2",
-                              #                "spl2_shift" = "*SPL2* (L1; shift + C)", 
-                              "plantra" = "PLanTra",
-                              "lift" = "LIFT",
-                              "c2l1" = "C2L1"))) %>% 
-  filter(!(str_detect(data, "SPL2") & lang == "ES"),
-         !(str_detect(data, "CATO") & group == "dyslexic"),
-         !(str_detect(data, "CATO") & task == "masked"),
-         !(str_detect(data, "GE2") & xn == "masked"),
-         param %in% c("delta", "theta")) %>%
-  select(param, location, data, value) %>% 
-  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
-  pivot_wider(names_from = location, values_from = value) %>%
-  unnest(-c(data, param)) %>% 
-  mutate(diff.1 = `before sentence` - `before word`,
-         diff.2 = `before word` - `within word`) %>% 
-  summarise(across(starts_with("diff"), 
-                   list(mean = mean, 
-                        lower = lower, 
-                        upper = upper, 
-                        BF = BF)),
-            .by = c(data, param))
-
-tmp <- ps_loc_diffs %>% 
-  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
-  mutate(across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
-  select(-mean, -lower, -upper) %>% 
-  pivot_wider(names_from = c(diffid, param), values_from = BF) %>% 
-  select(data, contains("delta"), contains("theta")) %>% 
-  arrange(data)
-
-tmp %>% 
-  kable(col.names = c("Dataset", rep(c("before sentence vs word", "before vs within word"), 2)),
-        caption = "Values indicate BFs in support of the alternative hypothesis over the null hypothesis.", align = c("l", rep("r", 4))) %>% 
-  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
-                full_width = T, 
-                position = "right",
-                font_size = 14) %>% 
-  add_header_above(c(" " = 1,
-                     "Hesitation slowdown" = 2,
-                     "Hesitation probability" = 2))
-
-```

### 2024-10-26 slides/sections/modelcomparisons.Rmd
After review update

@@
   kable(align = c("l", rep("r", 3)),
-        caption = "Vales are the absolute ratio of the difference in predictive performance measured as $\\widehat{elpd}$ and its standard error $\\mid\\frac{\\Delta\\widehat{elpd}}{\\text{SE}}\\mid$ which corresponds to the $z$-score of the difference between models [@sivula2020uncertainty].") %>%
+        caption = "Values are the absolute ratio of the difference in predictive performance measured as $\\widehat{elpd}$ [@vehtari2015pareto; @vehtari2017practical] and its standard error $\\mid\\frac{\\Delta\\widehat{elpd}}{\\text{SE}}\\mid$ which corresponds to the $z$-score of the change in predictive performance [@sivula2020uncertainty].") %>%
   kable_styling(bootstrap_options = c("striped", "hover", "responsive"),

### 2024-10-26 slides/sections/modeloverview.Rmd
After review update

@@
                        "Single distribution log-Gaussian with different variance components per text location",
-                       "Two-distributions mixture of log-Gaussians.")) %>% 
+                       "Two-distributions mixture of log-Gaussians")) %>% 
   select(Models, Description) %>% 

### 2024-10-26 slides/sections/mog_app.Rmd
After review update

@@
+# Mixture model parameters
+
 ---
@@
 inputPanel(
-  sliderInput("beta_0", "Fluent typing", min = 100, max = 500, value = 150, ticks = FALSE),
-  sliderInput("delta", "Hesitation slowdown", min = 50, max = 300, value = 100, ticks = FALSE),
-  sliderInput("theta", "Hesitation probability", min = .001, max = .999, value = .3, ticks = FALSE),  
-  switchInput("params", 
-              size = "mini", 
-              label = "Show means",
-              onLabel = "on", 
-              offLabel = "off", 
-              labelWidth = "50%",
-              width = "50%")
+  sliderInput("beta_0", "Fluent typing", min = 50, max = 500, value = 300, step = 10, ticks = FALSE),
+  sliderInput("delta", "Hesitation slowdown", min = 50, max = 1500, value = 100, step = 50, ticks = FALSE),
+  sliderInput("theta", "Hesitation probability", min = .001, max = .999, value = .3, step = .05, ticks = FALSE)  
+  # switchInput("params", 
+  #             size = "mini", 
+  #             label = "Show means",
+  #             onLabel = "on", 
+  #             offLabel = "off", 
+  #             labelWidth = "50%",
+  #             width = "50%", 
+  #             value = FALSE)
 )
@@
   sigmae_p <- sigma + sigma_diff
-  data <- data.frame(x = c(0, 1500))
+  data <- data.frame(x = c(0, 2500))
 
   # Formula shown in plot
-  formula <- list(~'iki'[i]~"~"~theta~paste("\u00B7")~  "logN("*beta~+~delta*","~sigma[2]^2*")",
-                  ~+~"(1" - theta*")"~paste("\u00B7")~"logN("*beta*","~sigma[1]^2*")")
+  # formula <- list(~'iki'[i]~"~"~theta~paste("\u00B7")~  "logN("*beta~+~delta*","~sigma[2]^2*")",
+  #                 ~+~"(1" - theta*")"~paste("\u00B7")~"logN("*beta*","~sigma[1]^2*")")
+  # 
+  # p2 <- ggplot() +
+  #   # annotate("text", 
+  #   #          x = 1.5, 
+  #   #          y = seq(1, .25, 
+  #   #                  length = length(formula)),
+  #   #          label = formula, 
+  #   #          parse = TRUE, 
+  #   #          size = 7) +
+  #   geom_rect(aes(xmin = -Inf, 
+  #                 xmax = Inf, 
+  #                 ymin = -Inf, 
+  #                 ymax = Inf), 
+  #             fill = "blue", 
+  #             alpha = 0.05, 
+  #             color = "black") +
+  #   theme_void() +
+  #   coord_cartesian(xlim = c(0, 3), ylim = c(-2.25, 2.4))
 
-  p2 <- ggplot() +
-    annotate("text", 
-             x = 1.5, 
-             y = seq(1, .25, 
-                     length = length(formula)),
-             label = formula, 
-             parse = TRUE, 
-             size = 6) +
-    geom_rect(aes(xmin = -Inf, 
-                  xmax = Inf, 
-                  ymin = -Inf, 
-                  ymax = Inf), 
-              fill = "blue", 
-              alpha = 0.05, 
-              color = "black") +
-    theme_void() +
-    coord_cartesian(xlim = c(0, 3), ylim = c(-2.25, 2.4))
+  # if(input$params){
+  #   beta_value <- round(exp(beta_0),0)
+  #   theta_value <- dezero(theta, 2)
+  #   delta_value <- round(delta, 2)
+  #   beta_label <- substitute(beta==beta0, list(
+  #     beta0 = beta_value))
+  #   theta_label <- substitute(theta==theta_value, list(
+  #     theta_value = theta_value))
+  #   delta_label <- substitute(delta==delta_value, list(
+  #     delta_value = delta_value))
+  #   
+  #   
+  #   p2 <- p2 + 
+  #     annotate("text",
+  #              x = 2.25,
+  #              y = -1,
+  #              label = deparse(beta_label), 
+  #              parse = TRUE, 
+  #              size = 6) +
+  #     annotate("text",
+  #              x = 2.25,
+  #              y = -1.5,
+  #              label = deparse(theta_label), 
+  #              parse = TRUE, 
+  #              size = 6) +
+  #     annotate("text",
+  #              x = 2.25,
+  #              y = -2,
+  #              label = deparse(delta_label), 
+  #              parse = TRUE, 
+  #              size = 6)
+  #   
+  # }
 
-  if(input$params){
-    beta_value <- round(exp(beta_0),0)
-    theta_value <- dezero(theta, 2)
-    delta_value <- round(delta, 2)
-    beta_label <- substitute(beta==beta0, list(
-      beta0 = beta_value))
-    theta_label <- substitute(theta==theta_value, list(
-      theta_value = theta_value))
-    delta_label <- substitute(delta==delta_value, list(
-      delta_value = delta_value))
-
-    p2 <- p2 + 
-      annotate("text",
-               x = 2.25,
-               y = -1,
-               label = deparse(beta_label), 
-               parse = TRUE, 
-               size = 6) +
-      annotate("text",
-               x = 2.25,
-               y = -1.5,
-               label = deparse(theta_label), 
-               parse = TRUE, 
-               size = 6) +
-      annotate("text",
-               x = 2.25,
-               y = -2,
-               label = deparse(delta_label), 
-               parse = TRUE, 
-               size = 6)
-
-  }
-
-  p1 <- ggplot(data, aes(x)) +
+  ggplot(data, aes(x)) +
     stat_function(geom = "line", fun = plot_mix_comps,
@@
     scale_x_continuous(labels = scales::comma, breaks = seq(0, 2500, 200)) +
-    coord_cartesian(ylim = c(0, .007), 
-                    xlim = c(0, 800)) +
+    coord_cartesian(ylim = c(0, .005), 
+                    xlim = c(-10, 2100)) +
     labs(title = "",
@@
 
-  p1 + p2 + 
-    plot_layout(widths = c(3, 1.25)) + 
-    plot_layout(guides = 'collect')
+#  p1 + p2 + 
+#    plot_layout(widths = c(3, 1.25)) + 
+#    plot_layout(guides = 'collect')
 
@@
 :::
+
+## Parallel view: mixture of two processes {.smaller}
+
+<div style="float: left; width: 40%;">
+
+```{r fig.width=5}
+# function for mixture models
+plot_mix_comps <- function(x, mu, sigma, lam) lam * dnorm(x, mu, sigma)
+
+# Parameters
+data <- data.frame(x = c(50, 1500))
+theta_conj <- .3
+theta_simp <- .1
+alpha <- 250
+delta <- c(250, 500)
+alpha2_w <- alpha + delta[1]
+alpha2_s <- alpha + delta[2]
+sigma <- 40
+sigma_diff <- c(30, 60)
+sigmae <- sigma
+sigmae_p_w <- sigma + sigma_diff[1]
+sigmae_p_s <- sigma + sigma_diff[2]
+
+ggplot(data, aes(x)) +
+  stat_function(geom = "line", fun = plot_mix_comps,
+                aes(colour = word_lbl),
+                linewidth = linewidth,
+                args = list(alpha, sigmae, lam = 1 - theta_simp)) +
+  stat_function(geom = "line", fun = plot_mix_comps,
+                aes(colour = word_lbl),
+                linewidth = linewidth,
+                args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
+  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
+            args = list(alpha, sigmae, lam = 1 - theta_simp)) +
+  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
+            args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
+  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
+  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
+  scale_x_continuous(labels = scales::comma, 
+                     limits = c(50, 1000), 
+                     breaks = seq(0, 1500, 200)) +
+  labs(x = "Interkey intervals [in ms]", 
+       colour = "Text location",
+       fill = "Text location") +
+  theme(axis.text.y = element_blank(),
+        axis.title.y = element_blank(),
+        axis.ticks.y = element_blank())
+
+```
+</div>
+
+<div style="float: right; width: 50%;">
+\
+\
+\
+\
+
+$$
+\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma_{e'}^2) + \\
+   (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
+$$
+
+</div>
+
+## Parallel view: mixture of two processes {.smaller}
+
+<div style="float: left; width: 40%;">
+
+```{r fig.width=5}
+ggplot(data, aes(x)) +
+  stat_function(geom = "line", fun = plot_mix_comps,
+                aes(colour = word_lbl),
+                linewidth = linewidth,
+                args = list(alpha, sigmae, lam = 1 - theta_simp)) +
+  stat_function(geom = "line", fun = plot_mix_comps,
+                aes(colour = sent_lbl),
+                linewidth = linewidth,
+                args = list(alpha, sigmae, lam = 1 - theta_conj)) +
+  stat_function(geom = "line", fun = plot_mix_comps,
+                aes(colour = word_lbl),
+                linewidth = linewidth,
+                args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
+  stat_function(geom = "line", fun = plot_mix_comps,
+                aes(colour = sent_lbl),
+                linewidth = linewidth,
+                args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
+  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
+            args = list(alpha, sigmae, lam = 1 - theta_simp)) +
+  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
+            args = list(alpha, sigmae, lam = 1 - theta_conj)) +
+  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
+            args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
+  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
+            args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
+  geom_curve(aes(x = alpha2_w, 
+                 xend = alpha2_s, y = 0.0008, yend = 0.0015),
+             curvature = -0.2, 
+             size = .15,
+             arrow = arrow(length = unit(0.3, "cm"))) +
+  annotate("text", 
+           x = (alpha2_w + alpha2_s)/2,
+           y=.0022, 
+           label=paste0("theta[diff]", "~and~", "delta[diff]"), size = 4, parse=TRUE) +
+  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
+  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
+  scale_x_continuous(labels = scales::comma, 
+                     limits = c(50, 1000), 
+                     breaks = seq(0, 1500, 200)) +
+  labs(x = "Interkey intervals [in ms]", 
+       colour = "Text location",
+       fill = "Text location") +
+  theme(axis.text.y = element_blank(),
+        axis.title.y = element_blank(),
+        axis.ticks.y = element_blank())
+
+```
+</div>
+
+<div style="float: right; width: 50%;">
+\
+\
+\
+\
+
+$$
+\text{iki}_{i} \sim\text{ } \theta_\text{location[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
+   (1 - \theta_\text{location[i]}) \times \text{log}\mathcal{N}(\beta, \sigma_{e_\text{location[i]}}^2)
+$$

### 2024-10-26 slides/sections/sampledata.Rmd
After review update

@@
 
-```{r eval = F}
-read_csv('tutorial/data/process_data_by_event.csv') %>% 
+```{r eval = F, message=F}
+read_csv('../../tutorial/data/process_data_by_event.csv') %>% 
   filter(location != "post-paragraph", 
@@
          idx = sample(n())) %>% 
-  filter(idx <= 50) %>% 
+  filter(idx <= 20) %>% 
   select(participant, iki = event_duration, location, pretext, text) %>% 
@@
   mutate(recent_text = substr(pretext, 
-                              start = nchar(pretext) - 40,
+                              start = nchar(pretext) - 15,
                               stop = nchar(pretext))) %>% 
@@
   relocate(participant, location, recent_text, new_letter, iki) %>% 
-  write_csv("slides/data/sampledata.csv")
+  write_csv("../data/sampledata.csv")
 ```
@@
 ```{r}
-data <- read_csv("slides/data/sampledata.csv")
-slice(data, sample(nrow(data), size = 10))
+data <- read_csv("../data/sampledata.csv")
+slice(data, sample(nrow(data), size = 20)) %>% 
+  kable() %>% 
+  kable_styling(full_width = T, font_size = 16)
 ```
 
-## How are these data distributed?
+## How are interkey intervals distributed?
 
-```{r fig.width=10, fig.height=6}
+```{r fig.width=10, fig.height=4.5}
 ggplot(data, aes(x = iki, colour = location, fill = location)) +
@@
 ```
+
+## Usual practice
+
+```{r fig.width=10, fig.height=4.5}
+ggplot(data, aes(x = iki, colour = location, fill = location)) +
+  geom_density(alpha = .25) +
+  scale_colour_colorblind() +
+  scale_fill_colorblind() +
+  scale_x_log10() +
+  geom_vline(xintercept = c(500, 1000, 2000), colour = "red", linetype = "dashed") +
+  annotate(geom = "text",
+           x = c(500, 1000, 2000), 
+           y = c(1.5, 1.25, 1),
+           label = str_c(c(500, 1000, 2000), " msecs")) +
+  labs(y = "density")
+```
+
+## Usual practice
+
+```{r echo = T}
+data_with_pause <- data %>% 
+  mutate(is_pause.500 = iki > 500,
+         is_pause.1k = iki > 1000,
+         is_pause.2k = iki > 2000)
+```
+
+```{r echo = T}
+data_with_pause %>% 
+  summarise(across(starts_with("is_pause"), list(prop = mean)), .by = location) %>% 
+  rename_with(.fn = ~str_remove(., "^.+\\.")) %>% 
+  arrange(location)
+```
+
+## Usual practice
+
+```{r echo = T}
+data_with_pause %>% 
+  pivot_longer(starts_with("is_pause"), names_to = c(".value", "threshold"), names_sep = "\\.") %>% 
+  filter(is_pause) %>% 
+  summarise(avg = mean(iki), sd = sd(iki), n = n(), .by = c(threshold, location)) %>% 
+  arrange(threshold)
+```

### 2024-10-26 slides/sections/simulation.Rmd
After review update

@@
 
-- Simulate data from a single log-normal distribution
-- Simulate data from a mixture of to log-normal distribution
+- Simulate data from (i) a single log-normal distribution and (ii) a mixture of two log-normal distributions.
+- Analysed both in (i) a single process model and (ii) a mixture model.
 - Data simulated with mixture process: 

### 2024-10-26 slides/sections/statisticalmodels.Rmd
After review update

@@
-## Serial view: single-process model {.smaller}
-
-<div style="float: left; width: 40%;">
-
-```{r fig.width=5}
-```
-</div>
-
-<div style="float: right; width: 50%;">
-\
-\
-\
-\
-
-$$
-\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
-$$
-
-</div>
-
 ## Serial view: single-process model  {.smaller}
@@
 sent_mean <- 950
-sd <- c(40, 60)
+sd <- c(40, 40)
 word_lbl <- "word initial"
@@
             args = list(mean = sent_mean, sd = sd[2])) +
-  geom_curve(aes(x = word_mean, xend = sent_mean, y = 0.01, yend = 0.00675),
+  geom_curve(aes(x = word_mean, xend = sent_mean, y = 0.01, yend = 0.01),
              curvature = -0.2, 
@@
              arrow = arrow(length = unit(0.3, "cm"))) +
-  annotate("text", x=(sent_mean+word_mean)/2, y=.0105, label=paste0("beta[diff]"), size = 4, parse=TRUE) +
+  annotate("text", x=(sent_mean+word_mean)/2, y=.0115, label=paste0("beta[diff]"), size = 4, parse=TRUE) +
   scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
@@
 $$
-\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta_\text{textlocation[i]}, \sigma_{e_\text{textlocation[i]}}^2)
+\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta_\text{textlocation[i]}, \sigma_{e}^2)
 $$
@@
 
+## Parallel view can be implemented as two distributions mixture-process [@roeser2021modelling]
+
+- Executing motor movements [100-150 msecs, @conijn2019understanding;@waes2019].
+- If upstream processes provide output more slowly, then interkey intervals are determined by time taken to complete upstream processes and not motor movement.
+
 ## Parallel view: mixture of two processes {.smaller}
@@
 <div style="float: left; width: 40%;">
+
+```{r fig.width=5}
+ggplot(data = data, aes(x)) +
+  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
+                linewidth = linewidth,
+                aes(colour = "word initial"), show.legend = F) +
+  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
+            args = list(mean = word_mean, sd = sd[1])) +
+  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
+  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
+  scale_y_continuous(limits = c(0, .012)) +
+  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
+  labs(x = "Interkey intervals [in ms]", 
+       colour = "Text location",
+       fill = "Text location") +
+  theme(axis.text.y = element_blank(),
+        axis.title.y = element_blank(),
+        axis.ticks.y = element_blank())
+
+```
 
@@
 $$
-\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma_{e'}^2) + \\
-   (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
+\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2) \\
+\theta=1
 $$
@@
 
-</div>
-
-## Parallel view: mixture of two processes {.smaller}
-
-<div style="float: left; width: 40%;">
-
-```{r fig.width=5}
-ggplot(data, aes(x)) +
-  stat_function(geom = "line", fun = plot_mix_comps,
-                aes(colour = word_lbl),
-                linewidth = linewidth,
-                args = list(alpha, sigmae, lam = 1 - theta_simp)) +
-  stat_function(geom = "line", fun = plot_mix_comps,
-                aes(colour = sent_lbl),
-                linewidth = linewidth,
-                args = list(alpha, sigmae, lam = 1 - theta_conj)) +
-  stat_function(geom = "line", fun = plot_mix_comps,
-                aes(colour = word_lbl),
-                linewidth = linewidth,
-                args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
-  stat_function(geom = "line", fun = plot_mix_comps,
-                aes(colour = sent_lbl),
-                linewidth = linewidth,
-                args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
-  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
-            args = list(alpha, sigmae, lam = 1 - theta_simp)) +
-  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
-            args = list(alpha, sigmae, lam = 1 - theta_conj)) +
-  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
-            args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
-  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
-            args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
-  geom_curve(aes(x = alpha2_w, 
-                 xend = alpha2_s, y = 0.0008, yend = 0.0015),
-             curvature = -0.2, 
-             size = .15,
-             arrow = arrow(length = unit(0.3, "cm"))) +
-  annotate("text", 
-           x = (alpha2_w + alpha2_s)/2,
-           y=.0022, 
-           label=paste0("theta[diff]", "~and~", "delta[diff]"), size = 4, parse=TRUE) +
-  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
-  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
-  scale_x_continuous(labels = scales::comma, 
-                     limits = c(50, 1000), 
-                     breaks = seq(0, 1500, 200)) +
-  labs(x = "Interkey intervals [in ms]", 
-       colour = "Text location",
-       fill = "Text location") +
-  theme(axis.text.y = element_blank(),
-        axis.title.y = element_blank(),
-        axis.ticks.y = element_blank())
-
-```
-</div>
-
-<div style="float: right; width: 50%;">
-\
-\
-\
-\
-
-$$
-\text{iki}_{i} \sim\text{ } \theta_\text{location[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
-   (1 - \theta_\text{location[i]}) \times \text{log}\mathcal{N}(\beta, \sigma_{e_\text{location[i]}}^2)
-$$
+- Fluent typing speed: $\beta$
+- Hesitation slowdown (pause duration): $\delta$
+- Hesitation probability (pause frequency): $\theta$
 

### 2024-10-26 slides/sections/textlocations.Rmd
After review update

@@
 ```{r keyloc, results='asis'}
-tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
+tibble(Location = c("Within word", "Below word", "Before sentence"),
                 Description = c("Transitions between any letter",
@@
                 Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d.
-T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.",
-                            "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
-                            "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept.")) %>% 
+T$^{\\wedge}$h$^{\\wedge}$e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.",
+                            "The $^{\\wedge}$cat $^{\\wedge}$meowed. Then $^{\\wedge}$it $^{\\wedge}$slept.", 
+                            "The cat meowed. $^{\\wedge}$Then it slept.")) %>% 
+  select(-Description) %>% 
   kable() %>% 
@@
   add_footnote("Note: Key intervals that terminated in a revision were removed.") %>% 
-  row_spec(0:3, extra_css = "line-height: 3.25;")  # Adjust the number for more spacing
+  row_spec(0:3, extra_css = "line-height: 3.25;")
 

### 2024-10-26 tutorial/tutorial.Rmd
After review update

@@
 data {
-  int<lower=1> N;                    // number of observations
-  int<lower=1> nS;                   // number of ppts
-  int<lower=1, upper=nS> ppt[N];     // ppts identifier
-  int<lower=1> K;                    // number of conditions
-  int<lower=1> condition[N];         // condition identifiers
-  vector[N] y;                       // outcome
+	int<lower=1> N;               // Number of observations
+  int<lower=1> nS;              // Number of subjects
+  int<lower=1, upper=nS> ppt[N];// Participant identifiers
+  int<lower=1> K;               // Number of conditions
+  int<lower=1> condition[N];    // Condition identifiers
+  vector[N] y;                  // Outcome: interkey intervals
 }
@@
   pivot_wider(names_from = location, values_from = value) %>%
-  select(-id) %>% # drop id column
   transmute(parameter = str_c(parameter, "[diff]"),
@@
 and then apply the `BF` function from above first to the difference between before-sentence and before-word transitions.
-For interpretation it is important to remember that $\theta$ is a probability (which we transformed from logit to probability above); hence the difference between two probabilities can only range between -1 and 1 (in the most extreme case).
+For interpretation it is important to remember that $\theta$ is a probability (which we transformed from logit to probability above mostly for ease of explanation; in practice it is a better idea to work with the logit scaled values); hence the difference between two probabilities can only range between -1 and 1 (in the most extreme case).
 In other words, a prior SD of 1 is saying anything goes (which is not helpful).
@@
 For a model such as `y ~ 1 + condition`, `hypothesis` can calculate the evidence for the hypothesis that the effect for `condition` is, for example, equal to 0.
-`hypothesis` can then be applied to the model with a statement that describes the hypothesis that should be evaluated.
-
-```{r eval = F}
-hypothesis(brm_model, "conditionb = 0")
-```
-
+`hypothesis` can then be applied to the model with a statement that describes the hypothesis that should be evaluated, for example `hypothesis(brm_model, "conditionb = 0")` for a null hypothesis.
 To calculate the evidence ratio the argument `sample_prior = TRUE` (or `sample_prior = 'yes'`) needs to be specified in `brm`.]
